### 2016-05-23 removals/manuscript/removal_ms.tex
Complete first draft

%---------------------------------------------
% This document is for pdflatex
%---------------------------------------------
\documentclass[11pt]{article}

\usepackage{amsmath,amsfonts,amssymb,graphicx,natbib,setspace,authblk}
\usepackage{float}
\usepackage[running]{lineno}
\usepackage[vmargin=1in,hmargin=1in]{geometry}

\usepackage{enumitem}
\setlist{topsep=.125em,itemsep=-0.15em,leftmargin=0.75cm}

\usepackage{gensymb}

\usepackage[compact]{titlesec}

\usepackage{bm,mathrsfs}

\usepackage{ifpdf}
\ifpdf
\DeclareGraphicsExtensions{.pdf,.png,.jpg}
\usepackage{epstopdf}
\else
\DeclareGraphicsExtensions{.eps}
\fi

\renewcommand{\floatpagefraction}{0.98}
\renewcommand{\topfraction}{0.99}
\renewcommand{\textfraction}{0.05}

\clubpenalty = 10000
\widowpenalty = 10000

\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand{\ba}{\begin{equation} \begin{aligned}}
\newcommand{\ea}{\end{aligned} \end{equation}}

\def\X{\mathbf{X}}

\floatstyle{boxed}
\newfloat{Box}{tbph}{box}

\title{Weak interspecific interactions in a sagebrush steppe: Evidence from observational data, population models, and a removal experiment}

\author[a]{Peter B. Adler\thanks{Corresponding author. Department of Wildland Resources and the Ecology Center, Utah State University, Logan, Utah Email: peter.adler@usu.edu}}
\author[a]{Andrew Kleinhesselink}
\author[b]{Stephen P. Ellner}
\author[b]{Giles Hooker}
\author[a]{Brittany Teller}
\affil[a]{Department of Wildland Resources and the Ecology Center, Utah State University, Logan, Utah 84322}
\affil[b]{Department of Ecology and Evolutionary Biology, Cornell University, Ithaca, New York}

\renewcommand\Authands{ and }

\date{Last compile: \today}

\sloppy

\renewcommand{\baselinestretch}{1.25}

\begin{document}

\maketitle

\textbf{\large{Keywords:}} Coexistence, competition, integral projection model.

\bigskip \textbf{Running title:} Weak interspecific interactions

\smallskip \textbf{Article type:} Letter

\smallskip \textbf{Authorship statement:} PBA and ARK designed the experiment and supervised data collection, PBA, SPE, GH and BT analyzed data, PBA wrote the first draft of the manuscript and all authors contributed to editing.
\smallskip

\setlength{\parindent}{8ex}

\newpage

\begin{doublespacing}

\linenumbers

\section*{Summary}

\begin{enumerate}
\item. A previous study of four dominant species from a sagebrush steppe community used historical data sets and multispecies population models to demonstrate weak interspecific interactions, implying very little competitive release. 
\item. We conducted a removal experiment to test the model's ability to predict competitive release. We established new quadrats in the same location where the historical data were collected. We assigned half the new quadrats to a  perennial grass removal treatment and half to a treatment in which we removed the dominant shrub, \textit{Artemisia tripartita}.
\item. We modeled survival, growth and recruitment as functions of local neighborhood species composition along with an indicator variable for removal treatment. For survival and recruitment, the removal treatment effects were never significantly positive, indicating that competitive release was not underestimated by the interspecific effects included in the ``baseline'' model. 
\item. For growth, the removal treatment effect was significant and positive for two species, \textit{Poa secunda} and \textit{Pseudoroegneria spicata}, indicating that the baseline model underestimated competitive release. However, including information about the location of these grass individuals with respect to removed \textit{Artemisia} failed to improve the growth regressions, raising questions about the mechanisms driving the positive response to removal.
\item. For three of the four species, individual based models and integral projections models  based on the vital rate regressions showed that removal treatment effects had little impact on population growth rates or simulated equilibrium cover following a species removal. For \textit{P. spicata}, the population models that included effects of \textit{Artemisia} removal projected much stronger competitive release than the baseline models. 
\item. \emph{Synthesis}: Although models based on observational data did not perfectly predict species responses to removal, overall the results increase our confidence that competitive release is in fact small for at least three of our four study species. The implication is that resource competition is not the primary factor determining the dynamics of this community.

\end{enumerate}

\section*{Introduction}

Stable coexistence requires that intraspecific interactions must limit populations more than interspecific interactions (Chesson 2000). In other words, when species are limited by different factors, intraspecific limitations will be stronger than interspecific limitations. For example, species might differ in their use of multiple limiting resources (Tilman), respond differently to spatial and temporal variation in the environment (Chesson 2000), or host different natural enemies (Janzen, Connell). Whenever we are studying populations of species that appear to be coexisting stably, such as populations that have co-occurred in close proximity for extended periods, we should expect intraspecific limitations to be stronger than interspecific limitations. However, we still might expect interspecific interactions to be strong and negative.

In fact, interspecific competition is arguably the primary obsession of plant community ecologists. We motivate research on plant species coexistence by noting that all plant species compete for water, light, and a few mineral resources (e.g. Silvertown 2004). Resource competition plays a central role in both Tilman's mechanistic resource competition models (Tilman 1982) and Grime's Competitor-Stress tolerator-Ruderal strategies (Grime). Neutral theory (Hubbell 2001) goes even further, assuming that interspecific competition is exactly as strong as interspecific competition. Recent work has emphasized the role of interspecific facilitation in some plant communities (Calloway? Maestre?), but it is unclear whether this research has shaken the default assumption that plant species compete intensely.

Competitive release provides an intuitive, population-level measure of the strength of interspecific competition. Following removal of one or more species from a community, how much will the abundance of remaining species increase? Large increases in the abundance of the remaining species represent a large competitive release, and imply strong interspecific interaction. Plant ecologists have conducted dozens of removal experiments and have observed almost every possible result. In some cases, competitive release is complete: remaining species fully compensate for removed biomass (Leps 1999, Jutila and Grace 2002). In other cases, the degree of competitive release depended on the functional group or dominance of the removed and remaining species (Sala et al. 1989, Belsky 1992, Symstad and Tilman 2001, Smith and Knapp 2003). Finally, even negative effects of removals have been observed (Keddy 1989, Gilbert et al. 2009), consistent with the recent facilitation literature. Results from removal experiments do come with caveats. Experiments typically do not run long enough to document long-term outcomes, responses may only be measured at one life stage whereas population responses depend on the whole life cycle, and removal treatments require disturbance which may cause a variety of impacts unrelated to resource competition (Aarssen and Epp 1990).

An alternative approach to studying competitive release is to fit multispecies population models to long-term, observational data, and then simulate population responses to removals. Applying this approach to arid and semiarid plant communities, we found evidence of strong intraspecific limitation and very weak interspecific interactions, leading to very stable coexistence and implying very little competitive release \citep{Adler:2010fu} ( Chu and Adler 2015). Since we only studied the most common, co-occurring species, we expected to find evidence of stable coexistence, but we were surprised that interspecific interactions and species' responses to the loss of a dominant competitor were so weak. Compared to traditional field experiments, our approach has the advantage of projecting full life-cycle responses over long time periods, but its reliance on observational data sets means that inference is weak. We would have more confidence in our results if they were confirmed by an experiment.

We conducted a removal experiment in a sagebrush steppe to test whether our models, which are based on observational data, underestimate the strength of interspecific competition and competitive release. We studied the responses of three perennial grass species to removal of sagebrush and the response of sagebrush to removal of all grasses. We addressed two research questions: First, are the survival, growth or recruitment of our target species significantly different in control vs. removal treatments after accounting for differences in local species composition? Second, how do short and long-term population-level responses to removals compare to projections from multispecies population models based on observational data?

\section*{Methods}

\subsection*{Study site and data set description}

The U.S. Sheep Experiment Station (USSES) is located 9.6 km north of Dubois, Idaho (44.2°N, 112.1°W), 1500 m above sea level. During the period of data collection (1926 – 1957), mean annual precipitation was 270 mm and mean temperatures ranged from -8\degree C (January) to 21\degree C (July). The vegetation is dominated by the shrub, \textit{Artemisia tripartita}, and the C3  perennial bunchgrasses \textit{Pseudoroegneria spicata}, \textit{Hesperostipa comata}, and \textit{Poa secunda}. These four species, the focus of our models, accounted for over 70\% of basal cover (grasses) and 60\% of canopy cover (shrubs and forbs).

Scientists at the USSES established 26 1-m$^2$ quadrats between 1926 and 1932. Eighteen quadrats were distributed among four ungrazed exclosures, and eight were distributed in two paddocks grazed at medium intensity spring through fall. All quadrats were located on similar topography and soils. In most years until 1957, all individual plants in each quadrat were mapped using a pantograph (Blaisdell 1958). The historical data set is public and available online \citep*{zachmann-etal-2010}. In 2007, we relocated 14 of the original quadrats, all of which are located inside permanent livestock exclosures, and resumed annual mapped censusing using the traditional pantograph method.

We extracted data on survival, growth, and recruitment from the mapped quadrats based on plants' spatial locations. Our approach tracks genets, which may be composed of multiple polygons, as they fragment and/or coalesce. Each mapped polygon is classified as a surviving genet or a new recruit based on its spatial location relative to genets present in previous years (Lauenroth and Adler 2008). We modeled  vital rates using data from 22 year-to-year transitions between 1926 and 1957 and eight year-to-year transitions from 2007 to 2015. For the first two transitions in the 1920's only four quadrats were observed, while at least 14 quadrats were observed for all subsequent transitions.

\subsection*{Removal experiment}

In spring 2011, in a large exclosure containing six of the permanent, historical quadrats, we selected locations for an additional 16 quadrats for the removal treatments. We used the following criteria to select the quadrat locations. Each plot had to have xx? cumulative cover/density? of the three target perennial grass species as well as yy? cover/density of \textit{Artemisia}. We mapped the quadrats in June, 2011 and then implemented the removal treatments in September (?), 2011. We randomly assigned 8 of the sixteen new quadrats to the sagebrush removal treatment and the remaining 8 to the grass removal treatment. We cut sagebrush stems at ground level and applied herbicide to the cut stems. In the grass removal plots, we herbicided all perennial grasses. We removed plants zz cm beyond the quadrat boundaries to minimize edge effects. The sagebrush removal was virtually 100\% effective; grass removals required additional herbicide applications in April and May of each year,  with fewer grasses remaining each additional year.

% add climate data for 2011-2015

Although the primary focus of our analysis is comparison of observed removal responses with response predicted by our population models, we also conducted a more traditional statistical analysis of differences in cover trends on control vs. removal plots. The response variable for this analysis was year-to-year change in cover at the quadrat scale, calculated as $log(Cover_{t+1}/Cover_t)$. We tested for the effect of removal treatment on this metric of cover change with a linear mixed effects model implemented in the \textit{lme4} package (Bates et al. 2015) of \textit{R}, including year and quadrat as random effects. We considered the removal treatment effect significant when the 95\% confidence interval on the estimated coefficient did not overlap zero.

\subsection*{Statistical models of vital rates}

Our regression models for survival, growth and recruitment follow previous work (Adler et al. 2010, Chu and Adler 2015) with a few exceptions, such as a new intercept term for the removal treatment.  We model the survival probability of an individual genet as a function of genet size, the neighborhood-scale crowding experienced by the genet from both conspecific and heterospecific genets, temporal variation among years, and permanent spatial variation among groups of quadrats (‘group’ here means a set of nearby quadrats located within one pasture or grazing exclosure), along with the removal treatment effect. More formally, the survival probability, $S$, of genet $i$ in species $j$, group $g$, and removal treatment $h$, from time $t$ to $t+1$  as
\begin{equation}
\mbox{logit}(S_{ijgh,t}) = \gamma_{j,t}^S + \varphi_{jg}^S+  \chi_{jh}^S  + \beta_{j,t}^S u_{ij,t} +  \boldsymbol{\omega}_{j,t}^S \boldsymbol{W}_{ij,t}
\label{eqn:survReg}
\end{equation}
where $\gamma$ is a time-dependent intercept, $\varphi$ is the coefficient for the effect of spatial group,$\chi$ is the removal treatment effect, $\beta$ is the coefficient that represents the effect of log genet size, $u$, on survival, and $\boldsymbol{\omega}$ is a vector of per capita interaction coefficients which determine the impact of crowding, $\boldsymbol{W}$ , by each species on the focal species. The vector $\boldsymbol{W}$ includes crowding from the four dominant species,  \textit{A. tripartita}, \textit{P. spicata}, \textit{H. comata}, and \textit{Poa secunda}, as well as two aggregated covariates representing crowding from (1)  all other perennial grasses and shrubs, which were mapped as polygons, and (2) all forb species, which were mapped as points.

Our growth model has a similar structure. The change in genet size from time $t$ to $t+1$ , conditional on survival, is given by:
\begin{equation}
\mbox{logit}(u_{ijgh,t+1}) = \gamma_{j,t}^G + \varphi_{jg}^G+  \chi_{jh}^G  + \beta_{j,t}^G u_{ij,t} +  \boldsymbol{\omega}_{j,t}^G \boldsymbol{W}_{ij,t} + \varepsilon_{ij,t}^G .
\label{eqn:growReg}
\end{equation}
To capture non-constant error variance in growth, we modeled the variance $\varepsilon$  about the growth curve (\ref{eqn:growReg})  as a nonlinear function of predicted genet size:
\begin{equation}
Var(\varepsilon_{ij,t}^G) = a \exp ^{(bu_{ij,t+1})} .
\label{eqn:growVar}
\end{equation}

We model the crowding experienced by a focal genet as a function of the distance to and size of neighbor genets. In previous work, we assumed that the decay of crowding with neighbor distance followed a Gaussian function (Chu and Adler 2015), but here we use a data-driven approach (Teller et al. 2016). We model the crowding experienced by genet $i$ of species $j$ from neighbors of species $m$ as the sum of neighbor areas across a set of concentric annuli, $k$, centered at the plant,
\begin{equation}
w_{ijm,k} = F_{jm}(d_{k})A_{i,k}      % can I put subscripts on a function?
\label{eqn:wik}
\end{equation}
where $F_{jm}$ is the competition kernel, $d_{k}$ is the average of the inner and outer radii of annulus $k$,
and $A_{im,k}$ is the total area of genets of species $m$
in annulus $k$ around genet $i$. The total crowding on genet $i$ exerted by species $m$ is
\begin{equation}
W_{ijm}  =\sum_k {w_{ijm,k}} .
\end{equation} 
Note that $W_{ijj}$ gives intraspecific crowding. The $W$'s are then introduced as covariates in the survival (\ref{eqn:survReg}) and growth (\ref{eqn:growReg}) regressions.

We assume that the competition kernel, $F(d)$, is non-negative and decreasing, which means that distant plants have less effect than close plants. Otherwise, we let the data dictate the shape of $F$ by fitting splines.  The demographic models such as \eqref{eqn:survReg} have $\gamma$, $\varphi$, $\chi$ , $\beta$, $\boldsymbol{\omega}$, and a vector of spline coefficients $\vec{b}$ as parameters to be fitted. We implemented this in the statistical computing environment, \texttt{R},  by making $\vec{b}$ the argument of an objective function that computes $W$ using the input value of $\vec{b}$, calls the model-fitting function
\texttt{lmer} and/or \texttt{glmer} to fit the other model parameters, and returns the negative log likelihood. We fit models for both survival and growth, and used the sum of the two log likelihoods. This assumes that one measure of crowding affects survival and growth. We estimated the $F$ functions using survival and growth models that depended only on intraspecific crowding. Based on previous work showing that competition kernels based on the effects of neighbors outperformed models based on the response of target plants (Adler et al. 2010), we set all $F_{jm}$ equal to $F_{jj}$. We used data from all historical and contemporary control treatment plots to estimate the competition kernels, which are shown in Fig. \ref{fig:CompKernels}.

Once we had estimated the parameters of the competitions kernels, we calculated the observed values of $W$ and fit the full survival and growth regressions, which include interspecific interaction coefficients, $\boldsymbol{\omega}$. All genets in a quadrat were included in calculating $W$, but plants located within 5 cm of quadrat edges were not used in fitting. We fit the models using the \textit{R-INLA} package for \texttt{R} (Rue et al. 20xx).

We model recruitment at the quadrat level rather than at the individual genet level because the mapped data do not allow us to determine which recruits were produced by which potential parent genets. We assume that the number of individuals, $y$, of species $j$ recruiting at time $t+1$ in the location $q$ follows a negative binomial distribution:
\begin{equation}
y_{jq,t+1}= NegBin(\lambda_{jq,t+1},\theta) 	   
\label{eqn:recrDataModel}
\end{equation}
where $\lambda$ is the mean intensity and $\theta$ is the size parameter. In turn,  $\lambda$ depends on the composition of the quadrat in the previous year:
\begin{equation}
\lambda_{jq,t+1} = C'_{jq,t} \exp{ (\gamma_{j,t}^R +  \varphi_{jg}^R + \chi_{jm}^R +\boldsymbol{ \omega}^R \boldsymbol{\sqrt{C'}_{q,t}} ) }
\label{eqn:recrProcessModel}
\end{equation}
where $R$ refers to Recruitment, $C'_{jq,t}$ is the ‘effective cover’ (cm$^2$) of species $j$ in quadrat $q$ at time $t$, $\gamma$ is a time-dependent intercept, $\varphi$ is a coefficient for the effect of group location, $\boldsymbol{\omega}$ is a vector of coefficients that determine the strength of intra- and interspecific density-dependence, and $\boldsymbol{C'}$ is the vector of ‘effective’ cover of each species in the community. Eqn (\ref{eqn:recrProcessModel}) is essentially the Ricker model for discrete time population growth (Ricker 1954), except that we use the square root of cover in the density-dependent term. We chose not to directly use a Ricker equation as Adler et al. (2010) did, as recruitment models with the square-root transformation had smaller Deviance Information Criterion (DIC) values than the ones without the transformation (Spiegelhalter et al. 2002). Following previous work (Adler et al. 2010), we treated year and group variables as random factors, allowing intercepts to vary among years and spatial locations.

Due to the possibility that plants outside the mapped quadrat could contribute recruits to the focal quadrat or interact with plants in the focal quadrat, we estimated effective cover as a mixture of the observed cover, $C$, in the focal quadrat, $q$, and the mean cover, $\bar{C}$, across the group, $g$, in which the quadrat is located:
$C'_{jq,t}=p_j C_{jq,t}+(1-p_j) \bar{C}_{jg,t}$								
where $p$ is a mixing fraction between 0 and 1.

We ran Markov Chain Monte Carlo (MCMC) simulations implemented in WinBUGS 1.4 (Lunn et al. 2000) to estimate the recruitment model parameters. Each model was run for 20,000 iterations and 2 chains with different initial values for parameters. We discarded the initial 10,000 samples. Convergence was observed graphically for all parameters and diagnosed using the Brooks-Gelman statistic (Brooks and Gelman 1998).

For all three vital rate models, we considered the effect of removal treatments to be significant when the confidence interval on the estimate of $\chi$ did not overlap zero. In an effort to further understand the removal treatment effects, we explored additional variations on the growth model with a linear random effects model. For hypothesis testing with these models, we focused on the 95\% confidence interval of the coefficient of interest. We also compared the species composition of the control and removal plots at both the quadrat and neighborhood ($\boldsymbol{W}$) scales.

\subsection*{Individual-based models}

The vital rate regressions provide a qualitative test of removal treatment effects: Does removal have an effect on a vital rate that goes above and beyond the effects of species interactions that are explicit in the model? However, the population-level response of a species requires integrating across all three vital rates. We used an individual-based model (IBM) to compare observed and predicted changes in population size (cover) from one year to the next.

To simulate changes in cover in each quadrat from year $t$ to year $t+1$, we initialized the IBM with the actual configuration of the four focal species (genet sizes and locations) observed in year $t$. We then projected the model forward one year, applying the random year effects corresponding to year $t$ and the spatial random effects corresponding to the quadrat of interest. We first applied the recruitment regression to determine the expected number of new recruits at time $t+1$, which we then multiplied by the mean size of recruits observed for the focal species to get expected cover of new recruits. We then used the survival regression to predict the survival probability of each genet, and the growth regression to predict the change in size of each genet. Expected cover at $t+1$ is the sum of the survival probabilities multiplied by the expected sizes plus the area of new recruits. 
% need to REDO with observed W's rather than W's that assume round genets

For control plots, we used a version of the model with all removal treatment effects set to zero. For removal treatment plots, we generated predictions from two versions of the model, with and without treatment effects set to zero. Because we were interested in comparing the predictions of these two models to observations, and were not interested in the effects of demographic stochasticity, we used a deterministic version of the model (e.g. recruitment is the $\lambda$ of (\ref{eqn:recrProcessModel}), rather than a random draw from a Poisson distribution with a mean of $\lambda$). We calculated the mean across quadrats of observed log cover change, $log(Cover_{t+1}/Cover_t)$ and compared it to the mean across quadrats of the predicted cover change.

\subsection*{Integral projection models}

To quantify the long-term, population-level consequences of the removal effects, we built and then simulated an integral projection model (IPM). In our IPM, the population of species $j$ is represented by a density function $n(u_{j,t})$ which gives the density of genets of size $u$ at time $t$, with genet size on a natural-log scale, i.e. $n(u_{j,t})du$ is the number of genets whose area (on arithmetic scale) is between $exp(u_j)$ and $exp(u_j+du)$. The density function for size $v$ at time $t+1$ is given by
\begin{equation}
n(v_{j,t+1})=\int_{L_j}^{U_j} k_j (v_j,u_j,\boldsymbol{\bar{W}}_j  (u_j))n(u_{j,t})   
\label{eqn:IPM}
\end{equation}
where the kernel $k_j (v_j,u_j,\boldsymbol{\bar{W}}_j  (u_j))$ describes all possible transitions from size $u$ to $v$ and $\boldsymbol{\bar{W}}_j$  is a vector whose elements are the average crowding experienced from each of the other species by an individual of size $u_j$ of species $j$ (we describe below how ($\boldsymbol{\bar{W}}_j$ ) is calculated). The integral is evaluated over all possible sizes from a lower size limit $L$ to an upper size limit $U$ that extends past the range of observed sizes.

The kernel is constructed from the fitted survival ($S$), growth ($G$), and recruitment ($R$) models:
\begin{equation}
 k_j (v_j,u_j,\boldsymbol{\bar{W}}_j ) = S_j(u_j,\boldsymbol{\bar{W}}_j  (u_j)) G_j(v_j,u_j,\boldsymbol{\bar{W}}_j  (u_j)) + R_j (v_j,u_j,\boldsymbol{\bar{W}}_j) .
\label{eqn:IPM} 
\end{equation}
$S$ is given by (Eqn. \ref{eqn:survReg}) and $G$ is given by (Eqn. \ref{eqn:growReg}). The recruitment function, $R$, defined in (\ref{eqn:recrDataModel} and \ref{eqn:recrProcessModel}), gives the number of new recruits produced per quadrat. To incorporate this recruitment function into the IPM, we assumed that fecundity increases linearly with size, $R_j (v_j,u_j,\boldsymbol{\bar{W}}_j)=exp(u_j ) R_j (v_j,\boldsymbol{\bar{W}}_j)$ (Adler et al. 2010).

In the vital rate regressions and in the IBM, we calculated a neighborhood crowding index, $W_ijm$ , unique to each individual $i$ based on the spatial locations and sizes of neighboring plants. This measure cannot be extended to the IPM, because the IPM is spatially implicit. Therefore, we used an approximation that captures the essential features of spatially-explicit neighborhood interactions (Adler et al. 2010). We found that in both the observed data and IBM simulations, heterospecific individuals were approximately randomly distributed, but conspecific individuals displayed non-random, size-dependent spatial patterns: while small genets were randomly distributed, large genets were segregated from each other. The overdispersion of large conspecific genets was introduced into the IPM with a 'no-overlap' rule (Adler et al. 2010, 2012). In \ref{suppMethods} we explain our implementation of the no-overlap rule with our nonparametric competition kernels (STEVE).

We simulated the IPM to estimate equilibrium abundances under three different scenarios. The first scenario featured all four species and our "baseline" model for control plots (no removal effects). We initialized all species with arbitrarily low cover and ran the model for 500 time steps (years) to allow the community to reach an equilibrium. To incorporate temporal environmental variation, at each time step we randomly selected one of the 30 sets of random year effects for the survival, growth and recruitment models ("kernel selection" in the terminology of Metcalf et al. 20xx))). We calculated the equilibrium cover for each species by simulating an additional 2000 time steps and averaging each species' cover over this period. For the second scenario, we set the cover of \textit{A. tripartita} to zero and then repeated the procedure. This simulation estimates the competitive release of grasses following \textit{A. tripartita} removal according to the baseline model (no removal effects). For the third scenario, we again set  \textit{A. triparitia} to zero, but this time we projected the model using the vital rate regressions that include removal effects. By comparing the output of this simulation with the previous scenario we can quantify the long-term consequences of differences between our baseline model and a model that includes additional removal treatment effects.

\section*{Results}

\subsection*{Cover trends}

Trends in cover by species and removal treatment were variable. For example, grass removal had little effect on the cover of \textit{A. tripartita} (Fig. \ref{fig:CoverTrends}) , but did appear to cause an increase in the cover of \textit{P. spicata}. We tested the effects of removal on cover by analyzing year-to-year changes in (log) cover, a measure of population growth rate (Fig. \ref{fig:CoverChange}). Grass removal had no significant effect on \textit{A. tripartita} (95\% confidence interval on removal effect: -0.33, 0.27) or \textit{Poa secunda} (95\% CI: 0.02, 0.87) population growth but did increase \textit{H. comata} (95\% CI: -0.40, 0.28) and \textit{P. spicata} (95\% CI: 0.04, 0.50) population growth (Table \ref{table:coefficients}). However, these treatment comparisons do not address our research questions about whether responses to removal were adequately predicted by models fit to observational data (i.e. control plots).

\begin{figure}[tbp]
\centering
\includegraphics[width=1\textwidth]{treatment_trends_cover}
\caption{Cover trends by treatment for the four dominant species.  Note the difference in y-axis scale between the canopy cover of \textit{A. tripartita} and the basal cover of the dominant grasses. ``Shrub removal'' refers to removal of \textit{A. tripartita}, and ``Grass removal'' refers to removal of all perennial grasses. The cover of species in treatments from which they are removed (e.g. \textit{A. tripartita} cover in shrub removal plots) is included to show values at the start of the experiment in 2011 and the efficacy of the removal treatments.}
\label{fig:CoverTrends}
\end{figure}

\begin{figure}[tbp]
\centering
\includegraphics[width=1\textwidth]{cover_change_chrono}
\caption{Observed and predicted year-to-year changes in cover for the four dominant species. Mean observed changes (averaged across quadrats) are shown in solid symbols, mean predicted changes are shown in open symbols. Values for control and removal plots are blue and red, respectively. Predictions from the ``Baseline'' model do account for effects of local species composition on vital rates, but do not include additional effects of removals. Predictions from the ``Removal'' model do include additional treatment effects on vital rates. }
\label{fig:CoverChange}
\end{figure}

\subsection*{Vital rates}

Our vital rate models take into account the effects of local species composition and should therefore be able to predict the effect of removals on survival, growth and recruitment without any additional parameters. If the removal treatment effects that we added to these models are statistically significant, it would indicate an effect of removals that goes above and beyond the response predicted by a model based on data from the control plots. For the survival models, the 95\% confidence intervals on the removals effects overlapped zero for all four species (Fig. \ref{fig:VitalRateTest}; Tables \ref{ARTRsurvival} to \ref{PSSPsurvival}), meaning that the survival probabilities of plants in removal plots can be predicted based on data from control plots. However, for the growth models, the 95\% confidence intervals on the removal effects were positive for \textit{Poa secunda}  and \textit{P. spicata}, meaning that individuals of these species grew more rapidly in removal plots than controls even after accounting for neighborhood interactions (Fig. \ref{fig:VitalRateTest}; Tables \ref{ARTRgrowth} to \ref{PSSPgrowth}). In contrast, \textit{A. tripartita} and  \textit{Poa secunda} had lower recruitment in removal than controls plots (upper limit of 95\% CI $<0$; Fig. \ref{fig:VitalRateTest}; Table \ref{table:recruitment}). The very low value (-26.8) and large confidence intervals (-69.6, -3.2) of the removal effect for  \textit{A. tripartita} recruitment reflect the fact that zero recruitment occurred in the four years of observations on the eight grass removal plots, making parameter estimation difficult.

 \begin{figure}[tbp]
 \centering
 \includegraphics[width=1\textwidth]{treatment_tests}
 \caption{Effects of removal treatments on the survival, growth, and recruitment of the four dominant species. Circles show the estimated coefficient of the removal effect, bars show 95\% confidence intervals. The removal effect shows response of \textit{A. tripartita} ("ARTR") to perennial grass removal, and response of the perennial grasses to \textit{A. tripartita} removal, after accounting for local neighborhood interactions. }
 \label{fig:VitalRateTest}
 \end{figure}

The positive removal effects on \textit{Poa secunda}  and \textit{P. spicata} growth are particularly interesting, because they suggest that models based on observational data may underestimate competitive release. We conducted additional analyses to further explore removal effects on the growth of these two species. First, we classified all individuals of these species as located either inside or outside the limits of \textit{A. tripartita} canopies present in 2011 (before removal). If removal is releasing grasses from \textit{A. tripartita} competition, we would expect individuals located under the shrub canopies to experience a greater effect of removal. However, the location of individuals with respect to \textit{A. tripartita} canopies had no effect on growth of either species (Tables \ref{table:POSEgrowth-inARTR} and \ref{table:PSSPgrowth-inARTR}), meaning that individuals located outside \textit{A. tripartita} canopies responded to removals as much as individuals underneath (removed) canopies.

Second, we evaluated the possibility that removal effects might be transient. We removed aboveground plant material, but roots of the removed plants remain in the ground and decompose, potentially creating a flush of mineral nutrients (refs?). We tested for a transient removal effect by adding a year-by-treatment interaction. For both  \textit{Poa secunda}  and \textit{P. spicata} , we found no clear sign of a decrease in the removal effect over time, and in fact the removal effects for \textit{P. spicata} were strongest in the last two years of observations (Tables \ref{table:POSEgrowth-byYr} and \ref{table:PSSPgrowth-byYr}).

Third, responses to removals might be poorly predicted by models based on observational data if the removals create local crowding outside the range of historical variation or if the removals break covariances in crowding among species. However, the distribution of crowding observed within control plots easily spans the crowding values observed in removal plots (Fig. \ref{fig:Wscatters}), and we failed to find evidence for strong interspecific covariances in crowding (\ref{suppMethods}).

Finally, we evaluated pre-existing differences in community composition in removal and control plots. In selecting removal plots, our criteria required candidate plots to include a certain amount of our target removal species. Because we adopted the historical plots as our controls, the same criteria were not applied (a control plot might not contain any \textit{A. tripartita}). To quantify resulting differences in pre-treatment (2011) composition, we compared neighborhood crowding at the individual plant level (the $\boldsymbol{W}$ of the growth regression) across treatments. \textit{A. tripartita} crowding and crowding by all subdominant shrub and grass species appeared greater than in the controls, whereas crowding by \textit{H. comata} was lower in removal plots than controls, though it was absent from most \textit{P. spicata} neighborhood in both treatments (Fig. \ref{fig:W-by-treatment}).

\subsection*{Population-level responses}

To quantify the short-term population-level consequences of removals, we used an IBM to project the cover of each species in each quadrat one year ahead using versions of the model with and without removal treatment effects. For \textit{A. tripartita}, both observed and predicted population growth rates were close to zero, and differences in model predictions were small (Fig. \ref{fig:CoverChange}). \textit{H. comata} experienced greater variation in growth rates, but the inclusion of removal treatment effects had little effect on population growth rates. For \textit{Poa secunda} and especially \textit{P. spicata}, the model with removal effects consistently predicted higher population growth than the baseline model  (Fig. \ref{fig:CoverChange}). For \textit{P. spicata}, including the removal effects increased the annual multiplicative rate of increase by roughly 35\%.

To quantify the long-term population-level consequences of removals, we used an IPM to project stochastic equilium cover. We concentrated on grass species' responses to \textit{A. tripartita} removal, since the growth regressions indicated our baseline model might underestimate competitive release in this situation. Using the baseline model, with no removal treatment coeffiicients, the IPM predicts little effect of \textit{A. tripartita} removal on the cover of any grass species (compare gray and brown boxes in Fig. \ref{fig:IPMresults}). Using a model with removal treatment coefficients, the IPM predicts little effect of shrub removal on \textit{Poa secunda} and \textit{H. comata}, but a 2.75 fold increase in the cover of \textit{P. spicata} (Fig. \ref{fig:IPMresults}).

 \begin{figure}[tbp]
 \centering
 \includegraphics[width=1\textwidth]{boxplots}
 \caption{Equilibrium cover of the four dominant species simulated by the IPM. Boxplots show interannual variation reflecting random year effects. Gray boxes show cover simulated by a model with \textit{A. tripartita} present and no removal treatment coefficients, brown boxes show results from the same model but with \textit{A. tripartita} cover set to zero (a species removal), and green boxes show results from a model with  \textit{A. tripartita} set to zero and removal treatment coefficients included.  }
 \label{fig:IPMresults}
 \end{figure}

\section*{Discussion}

Our population models based on historical, observational data reveal weak interspecific interactions (Adler et al. 2010, Chu and Adler 2015) and, as a result, predict very little competitive release among the four dominant species in a sagebrush steppe. Despite the fact that previous removal studies have reported variable results, ranging from negative effects of removal on remaining species to complete competitive release, we were suspicious of our model's predictions and conducted a removal experiment to test them. The results provide substantial, but not complete, support for our model's predictions.

For three of the four species, we find little or no evidence that models based on observational data underestimate the intensity of interspecific competition. Removal treatment effects were never significant for \textit{H. comata}; our baseline model, which accounts for local neighborhood composition, adequately explained vital rate responses to \textit{A. tripartita} removal. The same was true for the survival and growth models for \textit{A. tripartita}, but at the recruitment stage the grass removal treatment effect was significant and negative, implying that our baseline model might overestimate--not underestimate--competitive effects of the perennial grasses on \textit{Artemisia} recruitment. \textit{Poa secunda} showed mixed responses to \textit{Artemisia} removal: removal effects were not significant at the survival stage, were positive at the growth stage, and negative at the recruitment stage (where the estimated effect of \textit{Artemisia} on \textit{Poa} is positive ). At the population level, models with and without removal effects projected similar patterns of short (Fig. \ref{fig:CoverChange}) and long-term (Fig. \ref{fig:IPMresults}) dynamics for \textit{H. comata} and \textit{A. tripartita}. For \textit{Poa secunda}, the model with removal effects predicted higher population growth over the short but not the long-term, perhaps because the short-term projections mostly reflect the positive effect of \textit{Artemisia} removal on growth, while the long-term equilibrium are also sensitive to the negative effects of removal on recruitment. Although the baseline model did not always successfully predict removal effects on vital rates, it does appear to successfully capture the population-level consequences of removals on these three species and did not severly underestimate the magnitude of competitive release.

The story is different for the fourth species, \textit{P. spicata}. Both individual growth and short- and long-term population growth were higher in the sagebrush removal treatments than in the control plots, suggesting that our analyses of observational data may underestimate the competitive release that \textit{P. spicata} experiences following \textit{Artemisia} removal. However, the mechanism by which \textit{Artemisia} removal promotes \textit{P. spicata} remains unclear. If  \textit{Artemisia} removal is releasing \textit{P. spicata} from competition, we should expect \textit{P. spicata} individuals that were located beneath \textit{Artemisia} shrubs that we removed to respond more than individuals located away from any shrubs. Yet we found no evidence for such a difference. An alternative explanation involves pre-treatment differences in composition between control plots and  the \textit{Artemisia} removal  plots, which had higher cover of \textit{Artemisia} (see 2011 means in Fig. \ref{fig:CoverTrends}) and higher cover of subdominant grasses and shrubs (Fig. \ref{fig:W-by-treatment}). Perhaps these pre-treatment differences reflect subtle differences in edaphic and environmental factors which promote \textit{P. spicata} growth. While \textit{P. spicata} appears to show stronger competitive release than our baseline model predicted, the evidence is not conclusive.

The small competitive release of many species in our system implies that classic resource competition is not the primary factor determining community structure. Interactions with above- or belowground enemies can also regulate populations and, given enough species-specificity in the effects of these enemies (Clark paper, Janzen, Connell), decouple interspecific dynamics. Species-specific responses to spatial and temporal environmental variation can also weaken interspecific interactions (Chesson et al. 2000), though in a previous study we failed to find evidence of a temporal storage effect in this community (Adler et al. 2009). More likely, plants may alter the environment, and even resource availability, in ways that facilitate their neighbors. For example, shade cast by \textit{Artemisia} canopies may reduce evapotranspiration (e.g. refs...). \textit{Artemisia} removal might offest this facilitative effect by reducing resource uptake, resulting in little net change in water balance following removal.  Unfortunately, we did not monitor soil moisture at the spatial and temporal resolution necessary to test this hypothesis.

Evidence for weak interspecific interactions and limited competitive release carries interesting implications. If most interspecific interactions are weak, then single-species models may be adequate for forecasting population responses to environmental change (Adler et al. 2012, Kleinhesselink and Adler 2015). Similarly, differences among species in relative abundance must reflect variation in the strength of intraspecific density dependence rather than variation in the strength of interspecific interactions (Yenni et al. 2015). In such a system, effective management and conservation requires understanding the sources and species-specificity of intraspecific density dependence. In fact, an experiment to test our model's skill at predicting release from intraspecific competition may be even more important than the test of interspecific competitive release described here.

Our results have given us more confidence that competitive release is quite small for some species in some communities. But for many species in other communities, interspecific competition may still be an important driver. In closed canopy-forests where competition for light is intense, we might have found different results than in our system, in which competition for a number of belowground resources is more important than competition for light. However, a number of recent studies from forests have found that heterospecific competitive effects are dramatically weaker than conspecific effects (Uriarte refs, Comita et al. 2010, Johnson et al. 2012). We also expect stronger competitive release in successional systems, or communities in the early stages of community assembly where the process of competitive exclusion is still underway (examples?). Finally, we might have found stronger competitive release if we could have studied rare species in our system, as long as those species are rare because they are competitively suppressed. If they are rare because they occupy unique, but narrow, niches, interspecific interactions would still be weak.  Determining where interspecific interactions and competitive release are strong, and where they are not, remains a fascinating research problem.

\section*{Acknowledgements}

Funding: NSF DEB-1353078 and DEB-1054040, Utah Agricultural Experiment Station (get journal paper number). The USDA-ARS Sheep Experiment Station generously provided access to historical data and the field experiment site.

\newpage
\bibliographystyle{Ecology}
\bibliography{IPMrefs}

\end{doublespacing}

%%%%%%%%%%%%%%%%% For word count

%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% APPENDICES !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

\clearpage 
\newpage

\setcounter{page}{1}
\setcounter{equation}{0}
\setcounter{figure}{0}
\setcounter{section}{0}
\setcounter{table}{0}
\renewcommand{\theequation}{SI.\arabic{equation}}
\renewcommand{\thetable}{SI-\arabic{table}}
\renewcommand{\thefigure}{SI-\arabic{figure}}
\renewcommand{\thesection}{Section SI.\arabic{section}}

\centerline{\Large \textbf{Supporting Information }}
\centerline{Adler et al., ``Weak interspecific interactions''}

\vspace{0.4in}

\section{Supplementary Methods} \label{suppMethods}

We explored interspecific covariance in local crowding experienced by individual plants, by regressing the $W$ values exerted by one neighbor species, the response variable, against the $W$ values of all other species, the independent variables. Because some $W=0$, we conducted two separate regressions. First, using all $W$'s, we performed a generalized linear regression with a logit link function to evaluate whether the probability that the focal species' $W=0$ is influenced by the value of other species' $W$'s. Second, for the set of records in which the focal species $W>0$, we performed a linear regression. We repeated these regressions for individuals of each focal speices. Due to large samples size, $W$ values were often statistically significant predictors. However, they explained very little variance. The maximum reduction in deviance for the generalized linear regressions and $R^2$ for the linear regressions were both less than 8\%. The \texttt{R} code for this analysis is included as ..\texttt{\textbackslash Wdistrib\textbackslash exploreSurvivalWs.r})

Spatially-implicit approximations of local crowding for the IPM. Steve? Paste in the MeanFieldSplineIPM.tex?

\section{Additional Tables}

% supplementary tables

\begin{table}[h]
\caption{Cover change models}
\centering
%\begin{center}
\begin{tabular}{l c c c c }
\hline
 & Model 1 & Model 2 & Model 3 & Model 4 \\
\hline
(Intercept)           & $-0.03$          & $-0.04$          & $-0.02$          & $-0.09$          \\
                      & $[-0.33;\ 0.27]$ & $[-0.32;\ 0.24]$ & $[-0.31;\ 0.27]$ & $[-0.33;\ 0.15]$ \\
TreatmentNo\_grass    & $-0.09$          &                  &                  &                  \\
                      & $[-0.45;\ 0.27]$ &                  &                  &                  \\
TreatmentNo\_shrub    &                  & $0.44^{*}$       & $-0.06$          & $0.27^{*}$       \\
                      &                  & $[0.02;\ 0.87]$  & $[-0.40;\ 0.28]$ & $[0.04;\ 0.50]$  \\
\hline
AIC                   & 23.20            & 59.58            & 162.06           & 117.50           \\
BIC                   & 33.15            & 66.24            & 173.51           & 129.22           \\
Log Likelihood        & -6.60            & -24.79           & -76.03           & -53.75           \\
Num. obs.             & 54               & 28               & 73               & 77               \\
Num. groups: quad     & 12               & 5                & 14               & 14               \\
Num. groups: year     & 8                & 8                & 8                & 8                \\
Var: quad (Intercept) & 0.08             & 0.00             & 0.00             & 0.00             \\
Var: year (Intercept) & 0.01             & 0.00             & 0.10             & 0.09             \\
Var: Residual         & 0.04             & 0.32             & 0.40             & 0.19             \\
\hline
\multicolumn{5}{l}{\scriptsize{$^*$ 0 outside the confidence interval}}
\end{tabular}
\label{table:coefficients}
%\end{center}
\end{table}

% latex table generated in R 3.2.2 by xtable 1.8-2 package
% Thu May 19 08:31:09 2016
\begin{table}[ht]
\centering
\caption{Summary of fixed effects for the ARTR survival model} 
\label{ARTRsurvival}
\begin{tabular}{rrrrrrrr}
  \hline
 & mean & sd & 0.025quant & 0.5quant & 0.975quant & mode & kld \\ 
  \hline
(Intercept) & -0.0705 & 0.2256 & -0.5109 & -0.0723 & 0.3799 & -0.0757 & 0.0000 \\ 
  logarea & 0.7137 & 0.0605 & 0.5980 & 0.7124 & 0.8370 & 0.7098 & 0.0000 \\ 
  TreatmentNo\_grass & -0.2581 & 0.8998 & -1.9434 & -0.2884 & 1.5959 & -0.3511 & 0.0000 \\ 
  W.ARTR & -1.4362 & 0.2154 & -1.8709 & -1.4322 & -1.0242 & -1.4240 & 0.0000 \\ 
  W.HECO & -0.0818 & 0.0430 & -0.1630 & -0.0830 & 0.0062 & -0.0855 & 0.0000 \\ 
  W.POSE & -0.1159 & 0.0859 & -0.2851 & -0.1158 & 0.0523 & -0.1155 & 0.0000 \\ 
  W.PSSP & 0.0292 & 0.0276 & -0.0225 & 0.0283 & 0.0863 & 0.0263 & 0.0000 \\ 
  W.allcov & -0.0081 & 0.0093 & -0.0265 & -0.0081 & 0.0099 & -0.0080 & 0.0000 \\ 
  W.allpts & 0.1296 & 0.1579 & -0.1787 & 0.1290 & 0.4409 & 0.1278 & 0.0000 \\ 
   \hline
\end{tabular}
\end{table}

% latex table generated in R 3.2.2 by xtable 1.8-2 package
% Thu May 19 08:31:59 2016
\begin{table}[ht]
\centering
\caption{Summary of fixed effects for the HECO survival model} 
\label{HECOsurvival}
\begin{tabular}{rrrrrrrr}
  \hline
 & mean & sd & 0.025quant & 0.5quant & 0.975quant & mode & kld \\ 
  \hline
(Intercept) & 1.4652 & 0.2285 & 1.0126 & 1.4652 & 1.9170 & 1.4650 & 0.0000 \\ 
  logarea & 1.3190 & 0.1036 & 1.1266 & 1.3149 & 1.5346 & 1.3068 & 0.0000 \\ 
  TreatmentNo\_shrub & 0.7313 & 0.4426 & -0.1292 & 0.7284 & 1.6074 & 0.7226 & 0.0000 \\ 
  W.ARTR & -0.0049 & 0.0037 & -0.0119 & -0.0050 & 0.0027 & -0.0052 & 0.0000 \\ 
  W.HECO & -0.6707 & 0.0725 & -0.8206 & -0.6680 & -0.5354 & -0.6627 & 0.0000 \\ 
  W.POSE & 0.0965 & 0.0690 & -0.0272 & 0.0921 & 0.2440 & 0.0830 & 0.0000 \\ 
  W.PSSP & -0.0015 & 0.0197 & -0.0386 & -0.0020 & 0.0390 & -0.0032 & 0.0000 \\ 
  W.allcov & -0.0032 & 0.0058 & -0.0140 & -0.0034 & 0.0087 & -0.0038 & 0.0000 \\ 
  W.allpts & -0.2630 & 0.1184 & -0.4889 & -0.2654 & -0.0236 & -0.2701 & 0.0000 \\ 
  logarea:W.ARTR & -0.0005 & 0.0028 & -0.0057 & -0.0006 & 0.0052 & -0.0008 & 0.0000 \\ 
  logarea:W.HECO & 0.2751 & 0.0600 & 0.1522 & 0.2770 & 0.3879 & 0.2809 & 0.0000 \\ 
  logarea:W.POSE & 0.0067 & 0.0513 & -0.0833 & 0.0026 & 0.1182 & -0.0060 & 0.0000 \\ 
  logarea:W.PSSP & -0.0273 & 0.0147 & -0.0543 & -0.0280 & 0.0036 & -0.0294 & 0.0000 \\ 
  logarea:W.allcov & -0.0043 & 0.0041 & -0.0118 & -0.0045 & 0.0044 & -0.0050 & 0.0000 \\ 
  logarea:W.allpts & -0.1474 & 0.0903 & -0.3187 & -0.1496 & 0.0364 & -0.1541 & 0.0000 \\ 
   \hline
\end{tabular}
\end{table}

% latex table generated in R 3.2.2 by xtable 1.8-2 package
% Thu May 19 08:33:14 2016
\begin{table}[ht]
\centering
\caption{Summary of fixed effects for the POSE survival model} 
\label{POSEsurvival}
\begin{tabular}{rrrrrrrr}
  \hline
 & mean & sd & 0.025quant & 0.5quant & 0.975quant & mode & kld \\ 
  \hline
(Intercept) & 1.3190 & 0.2047 & 0.9110 & 1.3199 & 1.7215 & 1.3213 & 0.0000 \\ 
  logarea & 1.0711 & 0.0694 & 0.9387 & 1.0693 & 1.2137 & 1.0657 & 0.0000 \\ 
  TreatmentNo\_shrub & -0.0442 & 0.1832 & -0.4030 & -0.0445 & 0.3158 & -0.0451 & 0.0000 \\ 
  W.ARTR & 0.0001 & 0.0019 & -0.0036 & 0.0001 & 0.0038 & 0.0001 & 0.0000 \\ 
  W.HECO & -0.0075 & 0.0142 & -0.0353 & -0.0076 & 0.0204 & -0.0077 & 0.0000 \\ 
  W.POSE & -1.2650 & 0.0825 & -1.4296 & -1.2641 & -1.1053 & -1.2624 & 0.0000 \\ 
  W.PSSP & 0.0244 & 0.0123 & 0.0005 & 0.0243 & 0.0486 & 0.0242 & 0.0000 \\ 
  W.allcov & -0.0031 & 0.0027 & -0.0083 & -0.0031 & 0.0022 & -0.0031 & 0.0000 \\ 
  W.allpts & 0.0145 & 0.0527 & -0.0889 & 0.0144 & 0.1181 & 0.0143 & 0.0000 \\ 
   \hline
\end{tabular}
\end{table}

% latex table generated in R 3.2.2 by xtable 1.8-2 package
% Thu May 19 08:34:39 2016
\begin{table}[ht]
\centering
\caption{Summary of fixed effects for the PSSP survival model} 
\label{PSSPsurvival}
\begin{tabular}{rrrrrrrr}
  \hline
 & mean & sd & 0.025quant & 0.5quant & 0.975quant & mode & kld \\ 
  \hline
(Intercept) & 1.2208 & 0.1772 & 0.8756 & 1.2191 & 1.5757 & 1.2158 & 0.0000 \\ 
  logarea & 1.5691 & 0.1099 & 1.3570 & 1.5674 & 1.7908 & 1.5641 & 0.0000 \\ 
  TreatmentNo\_shrub & -0.2570 & 0.2348 & -0.7203 & -0.2563 & 0.2016 & -0.2548 & 0.0000 \\ 
  W.ARTR & 0.0094 & 0.0026 & 0.0044 & 0.0094 & 0.0147 & 0.0093 & 0.0000 \\ 
  W.HECO & -0.0005 & 0.0192 & -0.0370 & -0.0009 & 0.0386 & -0.0018 & 0.0000 \\ 
  W.POSE & 0.0183 & 0.0467 & -0.0675 & 0.0162 & 0.1162 & 0.0116 & 0.0000 \\ 
  W.PSSP & -0.4486 & 0.0312 & -0.5112 & -0.4481 & -0.3885 & -0.4472 & 0.0000 \\ 
  W.allcov & 0.0074 & 0.0035 & 0.0007 & 0.0073 & 0.0146 & 0.0072 & 0.0000 \\ 
  W.allpts & 0.0792 & 0.0797 & -0.0726 & 0.0776 & 0.2405 & 0.0742 & 0.0000 \\ 
  logarea:W.ARTR & -0.0014 & 0.0021 & -0.0053 & -0.0014 & 0.0029 & -0.0015 & 0.0000 \\ 
  logarea:W.HECO & -0.0174 & 0.0158 & -0.0466 & -0.0182 & 0.0156 & -0.0196 & 0.0000 \\ 
  logarea:W.POSE & 0.0147 & 0.0386 & -0.0540 & 0.0120 & 0.0978 & 0.0065 & 0.0000 \\ 
  logarea:W.PSSP & 0.0127 & 0.0239 & -0.0334 & 0.0124 & 0.0605 & 0.0119 & 0.0000 \\ 
  logarea:W.allcov & -0.0027 & 0.0027 & -0.0078 & -0.0028 & 0.0029 & -0.0029 & 0.0000 \\ 
  logarea:W.allpts & -0.0255 & 0.0647 & -0.1483 & -0.0270 & 0.1063 & -0.0302 & 0.0000 \\ 
   \hline
\end{tabular}
\end{table}

% latex table generated in R 3.2.2 by xtable 1.8-2 package
% Thu May 19 09:51:32 2016
\begin{table}[ht]
\centering
\caption{Summary of fixed effects for the ARTR growth model} 
\label{ARTRgrowth}
\begin{tabular}{rrrrrrrr}
  \hline
 & mean & sd & 0.025quant & 0.5quant & 0.975quant & mode & kld \\ 
  \hline
(Intercept) & 0.7252 & 0.2473 & 0.2395 & 0.7252 & 1.2107 & 0.7251 & 0.0000 \\ 
  logarea.t0 & 0.8661 & 0.0407 & 0.7860 & 0.8661 & 0.9460 & 0.8661 & 0.0000 \\ 
  TreatmentNo\_grass & 0.1014 & 0.1714 & -0.2352 & 0.1014 & 0.4377 & 0.1014 & 0.0000 \\ 
  W.ARTR & -0.0743 & 0.0989 & -0.2685 & -0.0743 & 0.1198 & -0.0743 & 0.0000 \\ 
  W.HECO & 0.0123 & 0.0151 & -0.0174 & 0.0123 & 0.0420 & 0.0123 & 0.0000 \\ 
  W.POSE & -0.0417 & 0.0325 & -0.1055 & -0.0417 & 0.0221 & -0.0417 & 0.0000 \\ 
  W.PSSP & 0.0067 & 0.0086 & -0.0101 & 0.0067 & 0.0235 & 0.0067 & 0.0000 \\ 
  W.allcov & -0.0063 & 0.0032 & -0.0127 & -0.0063 & 0.0000 & -0.0063 & 0.0000 \\ 
  W.allpts & 0.0043 & 0.0503 & -0.0944 & 0.0043 & 0.1030 & 0.0043 & 0.0000 \\ 
   \hline
\end{tabular}
\end{table}

% latex table generated in R 3.2.2 by xtable 1.8-2 package
% Thu May 19 09:51:55 2016
\begin{table}[ht]
\centering
\caption{Summary of fixed effects for the HECO growth model} 
\label{HECOgrowth}
\begin{tabular}{rrrrrrrr}
  \hline
 & mean & sd & 0.025quant & 0.5quant & 0.975quant & mode & kld \\ 
  \hline
(Intercept) & 0.3524 & 0.0858 & 0.1812 & 0.3528 & 0.5208 & 0.3535 & 0.0000 \\ 
  logarea.t0 & 0.8341 & 0.0212 & 0.7922 & 0.8341 & 0.8761 & 0.8340 & 0.0000 \\ 
  TreatmentNo\_shrub & 0.0228 & 0.1523 & -0.2768 & 0.0230 & 0.3210 & 0.0235 & 0.0000 \\ 
  W.ARTR & -0.0030 & 0.0010 & -0.0050 & -0.0030 & -0.0009 & -0.0030 & 0.0000 \\ 
  W.HECO & -0.0472 & 0.0191 & -0.0847 & -0.0472 & -0.0097 & -0.0472 & 0.0000 \\ 
  W.POSE & 0.0206 & 0.0161 & -0.0111 & 0.0206 & 0.0523 & 0.0206 & 0.0000 \\ 
  W.PSSP & -0.0162 & 0.0065 & -0.0289 & -0.0162 & -0.0035 & -0.0162 & 0.0000 \\ 
  W.allcov & -0.0047 & 0.0019 & -0.0085 & -0.0047 & -0.0009 & -0.0047 & 0.0000 \\ 
  W.allpts & -0.0544 & 0.0374 & -0.1279 & -0.0544 & 0.0190 & -0.0545 & 0.0000 \\ 
   \hline
\end{tabular}
\end{table}

% latex table generated in R 3.2.2 by xtable 1.8-2 package
% Thu May 19 09:52:23 2016
\begin{table}[ht]
\centering
\caption{Summary of fixed effects for the POSE growth model} 
\label{POSEgrowth}
\begin{tabular}{rrrrrrrr}
  \hline
 & mean & sd & 0.025quant & 0.5quant & 0.975quant & mode & kld \\ 
  \hline
(Intercept) & 0.4954 & 0.0697 & 0.3566 & 0.4957 & 0.6328 & 0.4963 & 0.0000 \\ 
  logarea.t0 & 0.6788 & 0.0225 & 0.6335 & 0.6790 & 0.7227 & 0.6794 & 0.0000 \\ 
  TreatmentNo\_shrub & 0.1789 & 0.0679 & 0.0456 & 0.1789 & 0.3120 & 0.1789 & 0.0000 \\ 
  W.ARTR & -0.0003 & 0.0008 & -0.0019 & -0.0003 & 0.0014 & -0.0003 & 0.0000 \\ 
  W.HECO & 0.0034 & 0.0062 & -0.0088 & 0.0034 & 0.0156 & 0.0034 & 0.0000 \\ 
  W.POSE & -0.2681 & 0.0382 & -0.3432 & -0.2681 & -0.1931 & -0.2681 & 0.0000 \\ 
  W.PSSP & -0.0077 & 0.0056 & -0.0186 & -0.0077 & 0.0032 & -0.0077 & 0.0000 \\ 
  W.allcov & -0.0002 & 0.0012 & -0.0026 & -0.0002 & 0.0021 & -0.0002 & 0.0000 \\ 
  W.allpts & -0.0289 & 0.0229 & -0.0739 & -0.0289 & 0.0161 & -0.0289 & 0.0000 \\ 
   \hline
\end{tabular}
\end{table}

% latex table generated in R 3.2.2 by xtable 1.8-2 package
% Thu May 19 09:53:11 2016
\begin{table}[ht]
\centering
\caption{Summary of fixed effects for the PSSP growth model} 
\label{PSSPgrowth}
\begin{tabular}{rrrrrrrr}
  \hline
 & mean & sd & 0.025quant & 0.5quant & 0.975quant & mode & kld \\ 
  \hline
(Intercept) & 0.3862 & 0.0784 & 0.2292 & 0.3867 & 0.5403 & 0.3878 & 0.0000 \\ 
  logarea.t0 & 0.8299 & 0.0164 & 0.7974 & 0.8299 & 0.8621 & 0.8300 & 0.0000 \\ 
  TreatmentNo\_shrub & 0.2767 & 0.0757 & 0.1281 & 0.2767 & 0.4252 & 0.2768 & 0.0000 \\ 
  W.ARTR & -0.0024 & 0.0008 & -0.0039 & -0.0024 & -0.0008 & -0.0024 & 0.0000 \\ 
  W.HECO & -0.0110 & 0.0070 & -0.0247 & -0.0110 & 0.0027 & -0.0110 & 0.0000 \\ 
  W.POSE & -0.0111 & 0.0129 & -0.0365 & -0.0111 & 0.0142 & -0.0111 & 0.0000 \\ 
  W.PSSP & -0.1498 & 0.0131 & -0.1754 & -0.1498 & -0.1242 & -0.1498 & 0.0000 \\ 
  W.allcov & -0.0035 & 0.0011 & -0.0055 & -0.0035 & -0.0014 & -0.0035 & 0.0000 \\ 
  W.allpts & -0.0498 & 0.0228 & -0.0946 & -0.0498 & -0.0051 & -0.0498 & 0.0000 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}
\caption{Poa secunda growth with Artemisia canopy effect}
\begin{center}
\begin{tabular}{l c }
\hline
 & Model 1 \\
\hline
(Intercept)                      & $0.50^{*}$        \\
                                 & $[0.36;\ 0.63]$   \\
logarea.t0                       & $0.68^{*}$        \\
                                 & $[0.63;\ 0.73]$   \\
TreatmentNo\_shrub               & $0.21^{*}$        \\
                                 & $[0.06;\ 0.37]$   \\
W.ARTR                           & $-0.00$           \\
                                 & $[-0.00;\ 0.00]$  \\
W.HECO                           & $0.00$            \\
                                 & $[-0.01;\ 0.02]$  \\
W.POSE                           & $-0.27^{*}$       \\
                                 & $[-0.34;\ -0.19]$ \\
W.PSSP                           & $-0.01$           \\
                                 & $[-0.02;\ 0.00]$  \\
W.allcov                         & $-0.00$           \\
                                 & $[-0.00;\ 0.00]$  \\
W.allpts                         & $-0.03$           \\
                                 & $[-0.07;\ 0.02]$  \\
inARTR                           & $-0.09$           \\
                                 & $[-0.33;\ 0.14]$  \\
\hline
AIC                              & 15070.92          \\
BIC                              & 15162.94          \\
Log Likelihood                   & -7521.46          \\
Num. obs.                        & 5286              \\
Num. groups: year                & 30                \\
Var: year (Intercept)            & 0.11              \\
Var: year logarea.t0             & 0.01              \\
Cov: year (Intercept) logarea.t0 & -0.02             \\
Var: Residual                    & 0.97              \\
\hline
\multicolumn{2}{l}{\scriptsize{$^*$ 0 outside the confidence interval}}
\end{tabular}
\label{table:POSEgrowth-inARTR}
\end{center}
\end{table}

\begin{table}
\caption{\textit{P. spicata} growth with \textit{Artemisia} canopy effect}
\begin{center}
\begin{tabular}{l c }
\hline
 & Model 1 \\
\hline
(Intercept)                      & $0.41^{*}$        \\
                                 & $[0.27;\ 0.54]$   \\
logarea.t0                       & $0.83^{*}$        \\
                                 & $[0.79;\ 0.87]$   \\
TreatmentNo\_shrub               & $0.35^{*}$        \\
                                 & $[0.18;\ 0.52]$   \\
W.ARTR                           & $-0.00^{*}$       \\
                                 & $[-0.00;\ -0.00]$ \\
W.HECO                           & $-0.01$           \\
                                 & $[-0.03;\ 0.00]$  \\
W.POSE                           & $-0.01$           \\
                                 & $[-0.04;\ 0.01]$  \\
W.PSSP                           & $-0.14^{*}$       \\
                                 & $[-0.16;\ -0.11]$ \\
W.allcov                         & $-0.00^{*}$       \\
                                 & $[-0.01;\ -0.00]$ \\
W.allpts                         & $-0.06^{*}$       \\
                                 & $[-0.10;\ -0.01]$ \\
inARTR                           & $-0.21$           \\
                                 & $[-0.45;\ 0.03]$  \\
\hline
AIC                              & 14834.74          \\
BIC                              & 14928.30          \\
Log Likelihood                   & -7403.37          \\
Num. obs.                        & 5904              \\
Num. groups: year                & 30                \\
Var: year (Intercept)            & 0.12              \\
Var: year logarea.t0             & 0.01              \\
Cov: year (Intercept) logarea.t0 & -0.03             \\
Var: Residual                    & 0.70              \\
\hline
\multicolumn{2}{l}{\scriptsize{$^*$ 0 outside the confidence interval}}
\end{tabular}
\label{table:PSSPgrowth-inARTR}
\end{center}
\end{table}

\begin{table}
\caption{Poa secunda growth with year-by-treatment interaction}
\begin{center}
\begin{tabular}{l c }
\hline
 & Model 1 \\
\hline
(Intercept)                      & $0.50^{*}$        \\
                                 & $[0.36;\ 0.63]$   \\
trtYears1                        & $0.05$            \\
                                 & $[-0.18;\ 0.28]$  \\
trtYears2                        & $0.25^{*}$        \\
                                 & $[0.01;\ 0.50]$   \\
trtYears3                        & $0.19$            \\
                                 & $[-0.09;\ 0.47]$  \\
trtYears4                        & $0.29$            \\
                                 & $[-0.02;\ 0.60]$  \\
logarea.t0                       & $0.68^{*}$        \\
                                 & $[0.63;\ 0.73]$   \\
W.ARTR                           & $-0.00$           \\
                                 & $[-0.00;\ 0.00]$  \\
W.HECO                           & $0.00$            \\
                                 & $[-0.01;\ 0.02]$  \\
W.POSE                           & $-0.27^{*}$       \\
                                 & $[-0.34;\ -0.19]$ \\
W.PSSP                           & $-0.01$           \\
                                 & $[-0.02;\ 0.00]$  \\
W.allcov                         & $-0.00$           \\
                                 & $[-0.00;\ 0.00]$  \\
W.allpts                         & $-0.03$           \\
                                 & $[-0.07;\ 0.02]$  \\
\hline
AIC                              & 15076.15          \\
BIC                              & 15181.32          \\
Log Likelihood                   & -7522.08          \\
Num. obs.                        & 5286              \\
Num. groups: year                & 30                \\
Var: year (Intercept)            & 0.11              \\
Var: year logarea.t0             & 0.01              \\
Cov: year (Intercept) logarea.t0 & -0.02             \\
Var: Residual                    & 0.97              \\
\hline
\multicolumn{2}{l}{\scriptsize{$^*$ 0 outside the confidence interval}}
\end{tabular}
\label{table:POSEgrowth-byYr}
\end{center}
\end{table}

\begin{table}
\caption{P. spicata growth with year-by-treatment interaction}
\begin{center}
\begin{tabular}{l c }
\hline
 & Model 1 \\
\hline
(Intercept)                      & $0.41^{*}$        \\
                                 & $[0.27;\ 0.54]$   \\
trtYears1                        & $0.25$            \\
                                 & $[-0.02;\ 0.52]$  \\
trtYears2                        & $-0.06$           \\
                                 & $[-0.33;\ 0.21]$  \\
trtYears3                        & $0.44^{*}$        \\
                                 & $[0.15;\ 0.73]$   \\
trtYears4                        & $0.53^{*}$        \\
                                 & $[0.23;\ 0.84]$   \\
logarea.t0                       & $0.83^{*}$        \\
                                 & $[0.79;\ 0.87]$   \\
W.ARTR                           & $-0.00^{*}$       \\
                                 & $[-0.00;\ -0.00]$ \\
W.HECO                           & $-0.01$           \\
                                 & $[-0.03;\ 0.00]$  \\
W.POSE                           & $-0.01$           \\
                                 & $[-0.04;\ 0.01]$  \\
W.PSSP                           & $-0.14^{*}$       \\
                                 & $[-0.16;\ -0.11]$ \\
W.allcov                         & $-0.00^{*}$       \\
                                 & $[-0.01;\ -0.00]$ \\
W.allpts                         & $-0.06^{*}$       \\
                                 & $[-0.10;\ -0.01]$ \\
\hline
AIC                              & 14834.13          \\
BIC                              & 14941.06          \\
Log Likelihood                   & -7401.06          \\
Num. obs.                        & 5904              \\
Num. groups: year                & 30                \\
Var: year (Intercept)            & 0.12              \\
Var: year logarea.t0             & 0.01              \\
Cov: year (Intercept) logarea.t0 & -0.02             \\
Var: Residual                    & 0.69              \\
\hline
\multicolumn{2}{l}{\scriptsize{$^*$ 0 outside the confidence interval}}
\end{tabular}
\label{table:PSSPgrowth-byYr}
\end{center}
\end{table}

% latex table generated in R 3.2.2 by xtable 1.8-2 package
% Mon May 23 12:08:13 2016
\begin{table}[ht]
\centering
\caption{Summary of fixed effects for the recruitment model} 
\label{table:recruitment}
\begin{tabular}{rrrrrrr}
  \hline
 & mean & sd & X2.5. & X97.5. & Rhat & n.eff \\ 
  \hline
intcpt.mu[1] & 0.4020 & 0.7346 & -1.1469 & 1.7240 & 1.1164 &    18 \\ 
  intcpt.mu[2] & 3.3186 & 0.5271 & 2.3009 & 4.3691 & 1.0041 &   500 \\ 
  intcpt.mu[3] & 3.2278 & 0.3828 & 2.4909 & 3.9860 & 1.0021 &  2000 \\ 
  intcpt.mu[4] & 2.8859 & 0.4080 & 2.0990 & 3.7150 & 1.0710 &    28 \\ 
  intcpt.trt[2,2] & -0.4870 & 0.5332 & -1.5241 & 0.5814 & 1.0005 &  2000 \\ 
  intcpt.trt[2,3] & -1.3272 & 0.3844 & -2.1322 & -0.5825 & 1.0018 &  2000 \\ 
  intcpt.trt[2,4] & -0.2148 & 0.3154 & -0.8458 & 0.3810 & 1.0015 &  1500 \\ 
  intcpt.trt[3,1] & -26.7881 & 18.2985 & -69.6230 & -3.1958 & 1.0010 &  2000 \\ 
  dd[1,1] & -0.5972 & 0.1128 & -0.7979 & -0.3581 & 1.1004 &    23 \\ 
  dd[1,2] & 0.0790 & 0.0646 & -0.0480 & 0.2020 & 1.0165 &   110 \\ 
  dd[1,3] & 0.0187 & 0.0408 & -0.0625 & 0.0989 & 1.0033 &   530 \\ 
  dd[1,4] & 0.1143 & 0.0473 & 0.0245 & 0.2090 & 1.0319 &    86 \\ 
  dd[2,1] & -0.4855 & 0.1619 & -0.7977 & -0.1613 & 1.0015 &  2000 \\ 
  dd[2,2] & -1.7763 & 0.1300 & -2.0320 & -1.5270 & 1.0158 &   120 \\ 
  dd[2,3] & 0.0335 & 0.0899 & -0.1377 & 0.2058 & 1.0031 &   570 \\ 
  dd[2,4] & -0.3687 & 0.0932 & -0.5474 & -0.1816 & 1.0024 &   810 \\ 
  dd[3,1] & -0.7223 & 0.3736 & -1.4671 & -0.0399 & 1.0139 &   140 \\ 
  dd[3,2] & -0.0166 & 0.2475 & -0.5859 & 0.4237 & 1.0092 &   320 \\ 
  dd[3,3] & -1.9665 & 0.2122 & -2.4881 & -1.6289 & 1.0226 &  1000 \\ 
  dd[3,4] & -0.1495 & 0.1629 & -0.4771 & 0.1599 & 1.0006 &  2000 \\ 
  dd[4,1] & -0.1196 & 0.3023 & -0.7493 & 0.4611 & 1.0325 &    53 \\ 
  dd[4,2] & -0.3597 & 0.2374 & -0.8784 & 0.0571 & 1.0034 &   800 \\ 
  dd[4,3] & 0.1429 & 0.1557 & -0.1669 & 0.4593 & 1.0019 &  1300 \\ 
  dd[4,4] & -1.7602 & 0.1818 & -2.0890 & -1.3710 & 1.0532 &    41 \\ 
  theta[1] & 0.6247 & 0.0732 & 0.4926 & 0.7761 & 1.0030 &   600 \\ 
  theta[2] & 1.1588 & 0.1500 & 0.8948 & 1.4830 & 1.0008 &  2000 \\ 
  theta[3] & 1.1570 & 0.1085 & 0.9687 & 1.3840 & 1.0015 &  2000 \\ 
  theta[4] & 1.0910 & 0.1053 & 0.8967 & 1.3130 & 1.0017 &  2000 \\ 
  u[1] & 0.8209 & 0.1188 & 0.4884 & 0.9571 & 1.0663 &    60 \\ 
  u[2] & 0.9995 & 0.0005 & 0.9981 & 1.0000 & 1.0028 &  2000 \\ 
  u[3] & 0.6802 & 0.1945 & 0.1677 & 0.9186 & 1.0775 &   260 \\ 
  u[4] & 0.6791 & 0.1808 & 0.2467 & 0.9301 & 1.0028 &   650 \\ 
   \hline
\end{tabular}
\end{table}

\clearpage

\section{Additional Figures}

 \begin{figure}[h]
 \centering
 \includegraphics[width=0.7\textwidth]{CompKernels}
 \caption{Competition kernels for the four dominant species.  }
 \label{fig:CompKernels}
 \end{figure}

  \begin{figure}[tbp]
  \centering
  \includegraphics[width=1\textwidth]{PSSP_W_scatters}
  \caption{Bivariate scatter plots comparing crowding exerted by each neighbor species on \textit{P. spicata}. Blue symbols show values from controls plots, red symbols show values from \textit{Artemisia} removal plots.``W.allcov'' refers to the aggregated crowding by all shrubs and perennial grasses beyond the focal dominant species, and ``W.allpts'' refers to the aggregated crowding by forb species.  }
  \label{fig:Wscatters}
  \end{figure}

  \begin{figure}[tbp]
 \centering
 \includegraphics[width=1\textwidth]{PSSP_W_byTrt}
 \caption{Distribution of crowding exerted by each neighbor species on \textit{P. spicata} individuals in controls plots (blue) and removal quadrats (red) in 2011, before \textit{Artemisia}  removals were conducted. Inset bar graphs show the probability that a particular neighbor species was present ($W>0$) in a focal plant's local neighborhood in control (blue) and removal (red) quadrats. ``W.allcov'' refers to the aggregated crowding by all shrubs and perennial grasses beyond the focal dominant species, and ``W.allpts'' refers to the aggregated crowding by forb species.  }
 \label{fig:W-by-treatment}
 \end{figure}

\end{document}

### 2016-05-23 removals/Wdistrib/exploreSurvivalWs.R

@@ -208,7 +208,7 @@ for(doSpp in 1:4){
 
   nonzero.rsq[doSpp,] <-Rsq
   zero.dev[doSpp,] <- ave.dev
-
+  
 } # next doSpp
 
 print(nonzero.rsq)

### 2016-05-23 removals/manuscript/removal_ms.tex
tiny changes

@@
 \author[b]{Stephen P. Ellner}
-\author[b]{Giles Hooker}
-\author[a]{Brittany Teller}
+\author[c]{Giles Hooker}
+\author[a]{Brittany Teller, and others? Robin? Bret?}
 \affil[a]{Department of Wildland Resources and the Ecology Center, Utah State University, Logan, Utah 84322}
 \affil[b]{Department of Ecology and Evolutionary Biology, Cornell University, Ithaca, New York}
+\affil[c]{Department of Biological Statistics and Computational Biology, Cornell University, Ithaca, New York}
 
@@
 \begin{enumerate}
-\item.
-A previous study of four dominant species from a sagebrush steppe community used historical data sets and multispecies population models to demonstrate weak interspecific interactions, implying very little competitive release.
-\item.
-We conducted a removal experiment to test the model's ability to predict competitive release.
+\item A previous study of four dominant species from a sagebrush steppe community used historical data sets and multispecies population models to demonstrate weak interspecific interactions, implying very little competitive release.
+\item We conducted a removal experiment to test the model's ability to predict competitive release.
 We established new quadrats in the same location where the historical data were collected.
 We assigned half the new quadrats to a  perennial grass removal treatment and half to a treatment in which we removed the dominant shrub, \textit{Artemisia tripartita}.
-\item.
-We modeled survival, growth and recruitment as functions of local neighborhood species composition along with an indicator variable for removal treatment.
+\item We modeled survival, growth and recruitment as functions of local neighborhood species composition along with an indicator variable for removal treatment.
 For survival and recruitment, the removal treatment effects were never significantly positive, indicating that competitive release was not underestimated by the interspecific effects included in the ``baseline'' model.
-\item.
-For growth, the removal treatment effect was significant and positive for two species, \textit{Poa secunda} and \textit{Pseudoroegneria spicata}, indicating that the baseline model underestimated competitive release.
+\item For growth, the removal treatment effect was significant and positive for two species, \textit{Poa secunda} and \textit{Pseudoroegneria spicata}, indicating that the baseline model underestimated competitive release.
 However, including information about the location of these grass individuals with respect to removed \textit{Artemisia} failed to improve the growth regressions, raising questions about the mechanisms driving the positive response to removal.
-\item.
-For three of the four species, individual based models and integral projections models  based on the vital rate regressions showed that removal treatment effects had little impact on population growth rates or simulated equilibrium cover following a species removal.
+\item For three of the four species, individual based models and integral projections models  based on the vital rate regressions showed that removal treatment effects had little impact on population growth rates or simulated equilibrium cover following a species removal.
 For \textit{P.
 spicata}, the population models that included effects of \textit{Artemisia} removal projected much stronger competitive release than the baseline models.
-\item.
-\emph{Synthesis}: Although models based on observational data did not perfectly predict species responses to removal, overall the results increase our confidence that competitive release is in fact small for at least three of our four study species.
+\item \emph{Synthesis}: Although models based on observational data did not perfectly predict species responses to removal, overall the results increase our confidence that competitive release is in fact small for at least three of our four study species.
 The implication is that resource competition is not the primary factor determining the dynamics of this community.

### 2016-05-25 removals/growth/import2ibm_1step.r
Still working on new ibm scripts

@@ -1,0 +1,46 @@
+# Import and format growth parameters
+# then define growth function
+
+# growth parameters
+Gpars=list(intcpt=rep(NA,Nspp),intcpt.yr=matrix(0,Nyrs,Nspp),intcpt.gr=matrix(0,6,Nspp),
+  intcpt.trt=rep(0,Nspp),slope=rep(NA,Nspp),slope.yr=matrix(0,Nyrs,Nspp),
+  nb=matrix(0,Nspp,Nspp+2),sigma2.a=rep(NA,Nspp),sigma2.b=rep(NA,Nspp))
+
+for(i in 1:Nspp){
+
+  infile=paste0("growth/",sppList[i],"_growth.csv")
+
+  Gdata=read.csv(infile)
+  # main intercept
+  Gpars$intcpt[i]=Gdata$X.Intercept.[1]
+  # Treatment effect
+  tmp=grep("Treatment",names(Gdata))
+  if(length(tmp)>0) Gpars$intcpt.trt[i]=Gdata[1,tmp]
+  # random year effects on intercept
+  Gpars$intcpt.yr[,i]=Gdata$Intercept.yr
+  # group effects
+  tmp=which(names(Gdata)=="Group")
+  if(length(tmp)>0) Gpars$intcpt.gr[,i]=Gdata$Group[!is.na(Gdata$Group)] 
+  # size slope
+  Gpars$slope[i]=Gdata$logarea.t0[1]
+  # random effects on slope
+  tmp=which(names(Gdata)=="logarea.yr")
+  if(length(tmp)>0) Gpars$slope.yr[,i]=Gdata[,tmp]
+  # get competition coefficients
+  tmp=grep("W.",(names(Gdata)))
+  Gpars$nb[i,]=as.numeric(Gdata[1,tmp])
+  # variance parameters
+  Gpars$sigma2.a[i]=Gdata$sigma.a[1]
+  Gpars$sigma2.b[i]=Gdata$sigma.b[1]
+} # next i
+rm(Gdata)
+
+# growth
+grow=function(Gpars,doSpp,doGroup,doYear,sizes,crowding,Treatment){
+  trt.Effect=Gpars$intcpt.trt[doSpp]
+  if(Treatment=="Control") trt.Effect=0
[…]

### 2016-05-25 removals/recruitment/import2ibm_1step.r

@@ -1,0 +1,55 @@
+
+# import parameters
+# recruitment parameters
+Rpars=list(intcpt.mu=rep(0,Nspp),intcpt.yr=matrix(0,Nyrs,Nspp),intcpt.trt=matrix(0,3,Nspp),intcpt.tau=rep(100,Nspp),
+  intcpt.gr=matrix(NA,6,Nspp),g.tau=rep(NA,Nspp),
+  dd=matrix(NA,Nspp,Nspp),theta=rep(NA,Nspp),sizeMean=rep(NA,Nspp),sizeVar=rep(NA,Nspp),
+  recSizes=list(1))
+
+#   if(trtEffects==T){
+   infile=paste0("recruitment/recruit_params_m1.csv")
+#   }else{
+#     infile=paste0("recruitment/recruit_params_m0.csv")
+#   } 
+
+ Rdata=read.csv(infile)
+ 
+ # subset out non-essential parameters
+ tmp=c(grep("lambda",row.names(Rdata)),grep("deviance",row.names(Rdata)),
+  grep("DIC",row.names(Rdata)))   
+ Rdata=Rdata[-tmp,]
+ 
+ # format parameters
+ tmp=paste("Rpars$",row.names(Rdata),"<-",Rdata[,1],sep="")
+ eval(parse(n=dim(Rdata)[1],text=tmp))
+ for(i in 1:Nspp){
+   infile=paste(root,"/driversdata/data/idaho/speciesData/",sppList[i],"/recSize.csv",sep="")
+   recSize=read.csv(infile)
+   Rpars$sizeMean[i]=mean(log(recSize$area))
+   Rpars$sizeVar[i]=var(log(recSize$area))
+   #Rpars$recSizes[[i]]=recSize$area
+ }
+Rpars$dd=t(Rpars$dd) # c[i,j] = effect of j on i
+
+# reformat treatment effects
+Rpars$intcpt.trt <- c(Rpars$intcpt.trt[3,1], # ARTR in no_grass
+                      Rpars$intcpt.trt[2,2:4])   # grasses in no_shrub
+
+rm(Rdata)
+
+# recruitment
+recruit=function(Rpars,totArea,doGroup,doYear,Treatment){
+    # dd is a matrix of dim Nspp X Nspp
+    # sizes and spp are vectors of the same length (=N plants)
+    trt.Effect = Rpars$intcpt.trt
+    if(Treatment=="Control") trt.Effect[] = 0
[…]

### 2016-05-25 removals/removal_analysis_wrapper.r

@@ -87,7 +87,7 @@ setwd("..")
 library(boot)
 library(R2WinBUGS)
 setwd("recruitment")
-source("call_recruit_m1.r")
+#source("call_recruit_m1.r")
 
 # add treatment test data for ARTR 
 pars.summary <- read.csv("recruit_params_m1.csv")

### 2016-05-25 removals/survival/import2ibm_1step.r

@@ -14,10 +14,8 @@ for(i in 1:Nspp){
   # main intercept
   Spars$intcpt[i]=Sdata$X.Intercept.[1]
   # Treatment effect
-  if(trtEffects==T){
-    tmp=grep("Treatment",names(Sdata))
-    if(length(tmp)>0) Spars$intcpt.trt[i]=Sdata[1,tmp]
-  }
+  tmp=grep("Treatment",names(Sdata))
+  if(length(tmp)>0) Spars$intcpt.trt[i]=Sdata[1,tmp]
   # Group effects
   tmp=which(names(Sdata)=="Group")
   if(length(tmp)>0) Spars$intcpt.gr[,i]=Sdata$Group[!is.na(Sdata$Group)] # get spatial average
@@ -35,10 +33,11 @@ Spars$yrList=Sdata$year
 rm(Sdata)
 
 # survival function
-survive=function(Spars,doSpp,doGroup,doYear,sizes,crowding){
-   logsizes=log(sizes)
-   mu=Spars$intcpt[doSpp]+Spars$intcpt.gr[doGroup,doSpp]+Spars$intcpt.yr[doYear,doSpp]+Spars$intcpt.trt[doSpp]+
-    Spars$slope[doSpp]*logsizes+Spars$nb[doSpp,]*crowding
+survive=function(Spars,doSpp,doGroup,doYear,sizes,crowding,Treatment){
+  trt.Effect=Spars$intcpt.trt[doSpp]
+  if(Treatment=="Control") trt.Effect=0
+   mu=Spars$intcpt[doSpp]+Spars$intcpt.gr[doGroup,doSpp]+Spars$intcpt.yr[doYear,doSpp]+trt.Effect+
+    Spars$slope[doSpp]*sizes+sum(Spars$nb[doSpp,]*crowding)
    out=inv.logit(mu)
    #out=rbinom(length(sizes),1,out)
    out

### 2016-06-24 removals/manuscript/removal_ms.tex
revising ms.

@@
 \widowpenalty = 10000
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
+%%% Just for commenting
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\usepackage[usenames]{color}
+\newcommand{\new}{\textcolor{red}}
+\newcommand{\comment}{\textcolor{blue}}
 
@@
 \item \emph{Synthesis}: Although models based on observational data did not perfectly predict species responses to removal, overall the results increase our confidence that competitive release is in fact small for at least three of our four study species.
-The implication is that resource competition is not the primary factor determining the dynamics of this community.
+The implication is that \new{interspecific} resource competition is not the primary factor determining the dynamics of this community.
 
@@
 Stable coexistence requires that intraspecific interactions must limit populations more than interspecific interactions (Chesson 2000). 
-In other words, when species are limited by different factors, intraspecific limitations will be stronger than interspecific limitations.
-For example, species might differ in their use of multiple limiting resources (Tilman), respond differently to spatial and temporal variation in the environment (Chesson 2000), or host different natural enemies (Janzen, Connell).
-Whenever we are studying populations of species that appear to be coexisting stably, such as populations that have co-occurred in close proximity for extended periods, we should expect intraspecific limitations to be stronger than interspecific limitations.
+\comment{(SPE: snip here)}
+%In other words, when species are limited by different factors, intraspecific limitations will be stronger than interspecific limitations. 
+%For example, species might differ in their use of multiple limiting resources (Tilman), respond differently to spatial and temporal variation in the
+%environment (Chesson 2000), or host different natural enemies (Janzen, Connell). 
+Whenever we study populations of species that appear to be coexisting stably, such as populations that have co-occurred in close proximity for extended periods, we should expect intraspecific limitations to be stronger than interspecific limitations.
 However, we still might expect interspecific interactions to be strong and negative.
@@
 
-We conducted a removal experiment in a sagebrush steppe to test whether our models, which are based on observational data, underestimate the strength of interspecific competition and competitive release.
-We studied the responses of three perennial grass species to removal of sagebrush and the response of sagebrush to removal of all grasses.
+\new{Here we present a removal experiment, conducted in a sagebrush steppe} 
+% We conducted a removal experiment in a sagebrush steppe 
+to test whether our models, which are based on observational data, underestimate the strength of interspecific competition and competitive release.
+We studied the responses of three perennial grass species to removal of sagebrush, and the response of sagebrush to removal of all grasses.
 We addressed two research questions: First, are the survival, growth or recruitment of our target species significantly different in control vs.
@@
 The U.S.
-Sheep Experiment Station (USSES) is located 9.6 km north of Dubois, Idaho (44.2°N, 112.1°W), 1500 m above sea level.
+Sheep Experiment Station (USSES) is located 9.6 km north of Dubois, Idaho (44.2\degree N, 112.1\degree W), 1500 m above sea level.
 During the period of data collection (1926 – 1957), mean annual precipitation was 270 mm and mean temperatures ranged from -8\degree C (January) to 21\degree C (July).
@@
 The historical data set is public and available online \citep*{zachmann-etal-2010}.
-In 2007, we relocated 14 of the original quadrats, all of which are located inside permanent livestock exclosures, and resumed annual mapped censusing using the traditional pantograph method.
+In 2007, we located 14 of the original quadrats, all of which are inside permanent livestock exclosures, and resumed annual mapped censusing using the traditional pantograph method.
 
@@
 Each mapped polygon is classified as a surviving genet or a new recruit based on its spatial location relative to genets present in previous years (Lauenroth and Adler 2008).
-We modeled  vital rates using data from 22 year-to-year transitions between 1926 and 1957 and eight year-to-year transitions from 2007 to 2015.
-For the first two transitions in the 1920's only four quadrats were observed, while at least 14 quadrats were observed for all subsequent transitions.
+We modeled vital rates using data from 22 year-to-year transitions between 1926 and 1957, and eight year-to-year transitions from 2007 to 2015.
+Only four quadrates were observed for the first two transitions in the 1920's, while at least 14 quadrats were observed for all subsequent transitions.
 
 \subsection*{Removal experiment}
-
-In spring 2011, in a large exclosure containing six of the permanent, historical quadrats, we selected locations for an additional 16 quadrats for the removal treatments.
+In spring 2011, in a large exclosure containing six of the historical permanent quadrats, we selected locations for an additional 16 quadrats for the removal treatments.
 We used the following criteria to select the quadrat locations.
@@
 We mapped the quadrats in June, 2011 and then implemented the removal treatments in September (?), 2011.
-We randomly assigned 8 of the sixteen new quadrats to the sagebrush removal treatment and the remaining 8 to the grass removal treatment.
+We randomly assigned 8 of the sixteen new quadrats to the sagebrush removal treatment, and the remaining 8 to the grass removal treatment.
 We cut sagebrush stems at ground level and applied herbicide to the cut stems.
@@
 
-Although the primary focus of our analysis is comparison of observed removal responses with response predicted by our population models, we also conducted a more traditional statistical analysis of differences in cover trends on control vs.
+Although the primary focus of our analysis is comparison of observed removal responses with responses predicted by our population models, we also conducted a more traditional statistical analysis of differences in cover trends on control vs.
 removal plots.
 The response variable for this analysis was year-to-year change in cover at the quadrat scale, calculated as $log(Cover_{t+1}/Cover_t)$.
-We tested for the effect of removal treatment on this metric of cover change with a linear mixed effects model implemented in the \textit{lme4} package (Bates et al.
+We tested for the effect of removal treatment on this measure of cover change with a linear mixed effects model implemented in the \textit{lme4} package (Bates et al.
 2015) of \textit{R}, including year and quadrat as random effects.
-We considered the removal treatment effect significant when the 95\% confidence interval on the estimated coefficient did not overlap zero.
+We considered the removal treatment effect to be significant when the 95\% confidence interval on the estimated coefficient did not overlap zero.
 
@@
 2010, Chu and Adler 2015) with a few exceptions, such as a new intercept term for the removal treatment.
-We model the survival probability of an individual genet as a function of genet size, the neighborhood-scale crowding experienced by the genet from both conspecific and heterospecific genets, temporal variation among years, and permanent spatial variation among groups of quadrats (‘group’ here means a set of nearby quadrats located within one pasture or grazing exclosure), along with the removal treatment effect.
-More formally, the survival probability, $S$, of genet $i$ in species $j$, group $g$, and removal treatment $h$, from time $t$ to $t+1$  as
+We model the survival probability of an individual genet as a function of genet size, the neighborhood-scale crowding experienced by the genet from both conspecific and heterospecific genets, temporal variation among years, and permanent spatial variation among groups of quadrats (`group'; here means a set of nearby quadrats located within one pasture or grazing exclosure), along with the removal treatment effect.
+Formally, we modeled the survival probability, $S$, of genet $i$ in species $j$, group $g$, and removal treatment $h$, from time $t$ to $t+1$  as
 \begin{equation}
-\mbox{logit}(S_{ijgh,t}) = \gamma_{j,t}^S + \varphi_{jg}^S+  \chi_{jh}^S  + \beta_{j,t}^S u_{ij,t} +  \boldsymbol{\omega}_{j,t}^S \boldsymbol{W}_{ij,t}
+\mbox{logit}(S_{ijgh,t}) = \gamma_{j,t}^S + \varphi_{jg}^S+  \chi_{jh}^S  + \beta_{j,t}^S u_{ij,t} +  
+\left \langle \boldsymbol{\omega}_{j,t}^S, \boldsymbol{W}_{ij,t} \right \rangle 
 \label{eqn:survReg}
 \end{equation}
-where $\gamma$ is a time-dependent intercept, $\varphi$ is the coefficient for the effect of spatial group,$\chi$ is the removal treatment effect, $\beta$ is the coefficient that represents the effect of log genet size, $u$, on survival, and $\boldsymbol{\omega}$ is a vector of per capita interaction coefficients which determine the impact of crowding, $\boldsymbol{W}$ , by each species on the focal species.
+where $\gamma$ is a time-dependent intercept, $\varphi$ is the coefficient for the effect of spatial group,$\chi$ is the removal treatment effect, $\beta$ is the coefficient that represents the effect of log genet size, $u$, on survival.
+$\boldsymbol{\omega}$ is a vector of interaction coefficients which determine the impact of crowding, $\boldsymbol{W}$, by each species on the focal species.
 The vector $\boldsymbol{W}$ includes crowding from the four dominant species,  \textit{A.
@@
 spicata}, \textit{H.
-comata}, and \textit{Poa secunda}, as well as two aggregated covariates representing crowding from (1)  all other perennial grasses and shrubs, which were mapped as polygons, and (2) all forb species, which were mapped as points.
+comata}, and \textit{Poa secunda}, as well as two covariates representing total crowding from (1) all other perennial grasses and shrubs, which were mapped as polygons, and (2) all forb species, which were mapped as points.
+\new{$\left \langle \boldsymbol{x, y} \right \rangle$ denotes the inner product of vectors $\boldsymbol{x}$ and $\boldsymbol{y}$, 
+calculated as \texttt{sum(x*y)} in R.}
 
@@
 \begin{equation}
-\mbox{logit}(u_{ijgh,t+1}) = \gamma_{j,t}^G + \varphi_{jg}^G+  \chi_{jh}^G  + \beta_{j,t}^G u_{ij,t} +  \boldsymbol{\omega}_{j,t}^G \boldsymbol{W}_{ij,t} + \varepsilon_{ij,t}^G .
+\mbox{logit}(u_{ijgh,t+1}) = \gamma_{j,t}^G + \varphi_{jg}^G+  \chi_{jh}^G  + \beta_{j,t}^G u_{ij,t} + 
+\left \langle  \boldsymbol{\omega}_{j,t}^G, \boldsymbol{W}_{ij,t} \right \rangle + \varepsilon_{ij,t}^G .
 \label{eqn:growReg}
@@
 \begin{equation}
-w_{ijm,k} = F_{jm}(d_{k})A_{i,k}      % can I put subscripts on a function?
+w_{ijm,k} = F_{jm}(d_{k})A_{i,k}     
 \label{eqn:wik}
 \end{equation}
-where $F_{jm}$ is the competition kernel, $d_{k}$ is the average of the inner and outer radii of annulus $k$,
-and $A_{im,k}$ is the total area of genets of species $m$
-in annulus $k$ around genet $i$. The total crowding on genet $i$ exerted by species $m$ is
+where $F_{jm}$ is the competition kernel (described below) for effects of species $j$ on species $m$, 
+$d_{k}$ is the average of the inner and outer radii of annulus $k$, 
+and $A_{im,k}$ is the total area of genets of species $m$ in annulus $k$ around genet $i$. The total crowding on 
+genet $i$ exerted by species $m$ is
 \begin{equation}
@@
 \end{equation} 
-Note that $W_{ijj}$ gives intraspecific crowding. The $W$'s are then introduced as covariates in the survival (\ref{eqn:survReg}) and growth (\ref{eqn:growReg}) regressions.
-
-We assume that the competition kernel, $F(d)$, is non-negative and decreasing, which means that distant plants have less effect than close plants.
-Otherwise, we let the data dictate the shape of $F$ by fitting splines.
-The demographic models such as \eqref{eqn:survReg} have $\gamma$, $\varphi$, $\chi$ , $\beta$, $\boldsymbol{\omega}$, and a vector of spline coefficients $\vec{b}$ as parameters to be fitted.
-We implemented this in the statistical computing environment, \texttt{R},  by making $\vec{b}$ the argument of an objective function that computes $W$ using the input value of $\vec{b}$, calls the model-fitting function
-\texttt{lmer} and/or \texttt{glmer} to fit the other model parameters, and returns the negative log likelihood.
-We fit models for both survival and growth, and used the sum of the two log likelihoods.
-This assumes that one measure of crowding affects survival and growth.
-We estimated the $F$ functions using survival and growth models that depended only on intraspecific crowding.
-Based on previous work showing that competition kernels based on the effects of neighbors outperformed models based on the response of target plants (Adler et al.
-2010), we set all $F_{jm}$ equal to $F_{jj}$.
-We used data from all historical and contemporary control treatment plots to estimate the competition kernels, which are shown in Fig.
+Note that $W_{ijj}$ gives intraspecific crowding. The $W$'s are then the components of the $\boldsymbol{W}$ vectors 
+introduced as covariates in the survival (\ref{eqn:survReg}) and growth (\ref{eqn:growReg}) regressions.
+
+\new{We assume that competition kernels $F_{jm}(d)$ are non-negative and decreasing, which means that distant plants have less effect 
+than close plants. Otherwise, we let the data dictate the shape of the kernel by fitting a spline model 
+using the methods of Teller et al. (2016). The shape of $F_{jm}$ is determined by a set of spline basis coefficients $\vec{b}_{jm}$
+and a smoothing parameter $\eta$ that controls the complexity of the fitted kernel. 
+Demographic models such as \eqref{eqn:survReg} then have $\gamma$, $\varphi$, $\chi$ , 
+$\beta$, $\boldsymbol{\omega}$, $\vec{b}$ and $\eta$ as parameters to be fitted. We implemented this in the statistical computing environment, \texttt{R}, 
+by making the spline coefficients and $\eta$ the arguments of an objective function that computes $\boldsymbol{W}$ using the input coefficients, 
+calls the model-fitting function \texttt{lmer} and/or \texttt{glmer} to fit the other parameters in the survival and growth regressions, and returns an approximate AIC value and model degrees of freedom ($df$) for survival and growth combined.
+We used the $\vec{b}$ values at the smoothest
+local minimum of AIC as a function of $df$, as in Teller et al. (2016). This approach assumes that one measure of crowding affects 
+survival and growth. In addition, 
+for fitting the kernels we assumed that survival and growth depended only on intraspecific crowding, and thus only fitted the
+within-species competition kernels $F_{jj}$. Based on previous work (Adler et al. 2010), we set all $F_{jm}$ equal to $F_{jj}$, meaning that 
+the within-species competition kernel for species $j$ is also used to determine the effect of all other species on species $j$.
+We used data from all historical plots and contemporary control-treatment plots to estimate the competition kernels, which are shown in Fig.
 \ref{fig:CompKernels}.
-
-Once we had estimated the parameters of the competitions kernels, we calculated the observed values of $W$ and fit the full survival and growth regressions, which include interspecific interaction coefficients, $\boldsymbol{\omega}$.
+}
+\comment{SPE: see Robin's comment about whether you mean $F_{mm}$ or $F_{jj}$; I'm not sure which one you did.}
+
+Once we had estimated the competitions kernels, we used them to calculate the values of $\boldsymbol{W}$ for each individual, 
+and fit the full survival and growth regressions, which include the interspecific interaction coefficients, $\boldsymbol{\omega}$. 
 All genets in a quadrat were included in calculating $W$, but plants located within 5 cm of quadrat edges were not used in fitting. 
-We fit the models using the \textit{R-INLA} package for \texttt{R} (Rue et al.
-20xx).
+We fit the models using the \textit{R-INLA} package for \texttt{R} (Rue et al. 20xx).
 
@@
 \begin{equation}
-\lambda_{jq,t+1} = C'_{jq,t} \exp{ (\gamma_{j,t}^R +  \varphi_{jg}^R + \chi_{jm}^R +\boldsymbol{ \omega}^R \boldsymbol{\sqrt{C'}_{q,t}} ) }
+\lambda_{jq,t+1} = C'_{jq,t} \exp{ \left(\gamma_{j,t}^R +  \varphi_{jg}^R + \chi_{jm}^R + 
+\left \langle \boldsymbol{\omega}^R , \boldsymbol{\sqrt{C'}_{q,t}} \right \rangle \right) }
 \label{eqn:recrProcessModel}
 \end{equation}
-where $R$ refers to Recruitment, $C'_{jq,t}$ is the ‘effective cover’ (cm$^2$) of species $j$ in quadrat $q$ at time $t$, $\gamma$ is a time-dependent intercept, $\varphi$ is a coefficient for the effect of group location, $\boldsymbol{\omega}$ is a vector of coefficients that determine the strength of intra- and interspecific density-dependence, and $\boldsymbol{C'}$ is the vector of ‘effective’ cover of each species in the community.
+where the superscript $R$ refers to Recruitment, $C'_{jq,t}$ is the `effective cover' (cm$^2$) of species $j$ in quadrat $q$ at time $t$, $\gamma$ is a time-dependent intercept, $\varphi$ is a coefficient for the effect of group location,
+$\chi$ is the removal treatment effect, $\boldsymbol{\omega}$ is a vector of coefficients that determine the strength of intra- and interspecific density-dependence, and $\boldsymbol{C'}$ is the vector of ‘effective’ cover of each species in the community.
 Eqn (\ref{eqn:recrProcessModel}) is essentially the Ricker model for discrete time population growth (Ricker 1954), except that we use the square root of cover in the density-dependent term.
-We chose not to directly use a Ricker equation as Adler et al.
-(2010) did, as recruitment models with the square-root transformation had smaller Deviance Information Criterion (DIC) values than the ones without the transformation (Spiegelhalter et al.
-2002).
+We previously used a Ricker equation (Adler et al.
+2010), but using the square-root transformation gave a lower Deviance Information Criterion (DIC) value (Spiegelhalter et al.
+2002) for the recruitment model.
 Following previous work (Adler et al.
@@
 $C'_{jq,t}=p_j C_{jq,t}+(1-p_j) \bar{C}_{jg,t}$								
-where $p$ is a mixing fraction between 0 and 1.
-
-We ran Markov Chain Monte Carlo (MCMC) simulations implemented in WinBUGS 1.4 (Lunn et al.
+where $p$ is a mixing fraction between 0 and 1 that was estimated as part of fitting the model.
+
+We ran Markov Chain Monte Carlo (MCMC) simulations in WinBUGS 1.4 (Lunn et al.
 2000) to estimate the recruitment model parameters.
-Each model was run for 20,000 iterations and 2 chains with different initial values for parameters.
+Each model was run for 20,000 iterations and two independent chains with different initial values for parameters.
 We discarded the initial 10,000 samples.
-Convergence was observed graphically for all parameters and diagnosed using the Brooks-Gelman statistic (Brooks and Gelman 1998).
-
-For all three vital rate models, we considered the effect of removal treatments to be significant when the confidence interval on the estimate of $\chi$ did not overlap zero.
+Convergence was observed graphically for all parameters, and confirmed by the Brooks-Gelman statistic (Brooks and Gelman 1998).
+
+For all three vital rate models, we considered the effect of removal treatments to be significant when the 95\% confidence interval (or posterior
+interval) on the estimate of $\chi$ did not overlap zero.
 In an effort to further understand the removal treatment effects, we explored additional variations on the growth model with a linear random effects model.
@@
 The vital rate regressions provide a qualitative test of removal treatment effects: Does removal have an effect on a vital rate that goes above and beyond the effects of species interactions that are explicit in the model?
-However, the population-level response of a species requires integrating across all three vital rates.
-We used an individual-based model (IBM) to compare observed and predicted changes in population size (cover) from one year to the next.
+However, the population-level response by each species integrates the effects of the removal treatment on all three vital rates.
+We used an individual-based model (IBM) to compare observed and predicted changes
+in population size (cover) from one year to the next.
 
@@
 We then projected the model forward one year, applying the random year effects corresponding to year $t$ and the spatial random effects corresponding to the quadrat of interest.
+Because we were interested in comparing model predictions to observations, and were not interested in the effects of demographic stochasticity, we used a deterministic version of the model (e.g., recruitment is the $\lambda$ of (\ref{eqn:recrProcessModel}), rather than a random draw from a Poisson distribution with a mean of $\lambda$).
 We first applied the recruitment regression to determine the expected number of new recruits at time $t+1$, which we then multiplied by the mean size of recruits observed for the focal species to get expected cover of new recruits.
 We then used the survival regression to predict the survival probability of each genet, and the growth regression to predict the change in size of each genet.
-Expected cover at $t+1$ is the sum of the survival probabilities multiplied by the expected sizes plus the area of new recruits.
+Expected cover at $t+1$ is expected area of new recruits, plus the sum of the survival probabilities multiplied by the expected sizes.
 % need to REDO with observed W's rather than W's that assume round genets
@@
 For removal treatment plots, we generated predictions from two versions of the model, with and without treatment effects set to zero.
-Because we were interested in comparing the predictions of these two models to observations, and were not interested in the effects of demographic stochasticity, we used a deterministic version of the model (e.g.
-recruitment is the $\lambda$ of (\ref{eqn:recrProcessModel}), rather than a random draw from a Poisson distribution with a mean of $\lambda$).
 We calculated the mean across quadrats of observed log cover change, $log(Cover_{t+1}/Cover_t)$ and compared it to the mean across quadrats of the predicted cover change.
@@
 \end{equation}
-where the kernel $k_j (v_j,u_j,\boldsymbol{\bar{W}}_j  (u_j))$ describes all possible transitions from size $u$ to $v$ and $\boldsymbol{\bar{W}}_j$  is a vector whose elements are the average crowding experienced from each of the other species by an individual of size $u_j$ of species $j$ (we describe below how ($\boldsymbol{\bar{W}}_j$ ) is calculated).
+where the kernel $k_j (v_j,u_j,\boldsymbol{\bar{W}}_j  (u_j))$ describes all possible transitions from size $u$ to $v$ and $\boldsymbol{\bar{W}}_j$  is a vector whose elements are the average crowding experienced from each of the other species by an individual of size $u_j$ of species $j$ (we describe below how $\boldsymbol{\bar{W}}_j$ is calculated).
 The integral is evaluated over all possible sizes from a lower size limit $L$ to an upper size limit $U$ that extends past the range of observed sizes.
@@
 \begin{equation}
- k_j (v_j,u_j,\boldsymbol{\bar{W}}_j ) = S_j(u_j,\boldsymbol{\bar{W}}_j  (u_j)) G_j(v_j,u_j,\boldsymbol{\bar{W}}_j  (u_j)) + R_j (v_j,u_j,\boldsymbol{\bar{W}}_j) .
+k_j (v_j,u_j,\boldsymbol{\bar{W}}_j ) = S_j(u_j,\boldsymbol{\bar{W}}_j  (u_j)) G_j(v_j,u_j,\boldsymbol{\bar{W}}_j  (u_j)) 
++ R_j (v_j,u_j,\boldsymbol{\bar{W}}_j) .
 \label{eqn:IPM} 
 \end{equation}
-$S$ is given by (Eqn.
-\ref{eqn:survReg}) and $G$ is given by (Eqn.
-\ref{eqn:growReg}).
-The recruitment function, $R$, defined in (\ref{eqn:recrDataModel} and \ref{eqn:recrProcessModel}), gives the number of new recruits produced per quadrat.
-To incorporate this recruitment function into the IPM, we assumed that fecundity increases linearly with size, $R_j (v_j,u_j,\boldsymbol{\bar{W}}_j)=exp(u_j ) R_j (v_j,\boldsymbol{\bar{W}}_j)$ (Adler et al.
+$S$ is given by (Eqn. \ref{eqn:survReg}) and $G$ is given by (Eqn. \ref{eqn:growReg}). The recruitment rate, $\lambda$, 
+defined in (\ref{eqn:recrDataModel} and \ref{eqn:recrProcessModel}), gives the expected 
+number of new recruits produced per quadrat, per unit area of parents. To incorporate this recruitment function into the IPM, 
+we assumed that individual fecundity increases linearly with individual size (area, equal to $e^u$). The recruitment kernel is thus
+$R_j(v_j,u_j,\boldsymbol{\bar{W}}_j)=C_0(v_j)e^{u_j}\lambda_j$ (Adler et al. 2010) where $C_0$ is the
+size distribution of new recruits, and $\lambda_j$ is calculated from (\ref{eqn:recrDataModel} and \ref{eqn:recrProcessModel}) with
+population densities converted into total cover per m$^2$, $C$.
+
+In the vital rate regressions and in the IBM, the neighborhood crowding index, $W_ijm$ , unique to each individual $i$ based on the spatial locations and sizes of neighboring plants.
+This measure cannot be extended to the IPM, because the IPM is spatially implicit.
+Therefore, we used a spatially
+implicit mean-field approximation that captures the essential features of spatially-explicit neighborhood interactions (Adler et al.
 2010).
-
-In the vital rate regressions and in the IBM, we calculated a neighborhood crowding index, $W_ijm$ , unique to each individual $i$ based on the spatial locations and sizes of neighboring plants.
-This measure cannot be extended to the IPM, because the IPM is spatially implicit.
-Therefore, we used an approximation that captures the essential features of spatially-explicit neighborhood interactions (Adler et al.
-2010).
-We found that in both the observed data and IBM simulations, heterospecific individuals were approximately randomly distributed, but conspecific individuals displayed non-random, size-dependent spatial patterns: while small genets were randomly distributed, large genets were segregated from each other.
+We found that in both the observed data and IBM simulations, heterospecific individuals were approximately randomly distributed with respect to each other,
+but conspecific individuals displayed non-random, size-dependent spatial patterns: while small genets were randomly distributed, large genets 
+were segregated from each other.
 The overdispersion of large conspecific genets was introduced into the IPM with a 'no-overlap' rule (Adler et al.
 2010, 2012).
-In \ref{suppMethods} we explain our implementation of the no-overlap rule with our nonparametric competition kernels (STEVE).
+In \ref{suppMethods} we explain our implementation of the no-overlap rule with our nonparametric competition kernels.
 

### 2016-06-24 removals/manuscript/removal_ms.tex
edits on the ms

@@
 \end{equation}
-where $F_{jm}$ is the competition kernel (described below) for effects of species $j$ on species $m$, 
+where $F_{jm}$ is the competition kernel (described below) for effects of species $m$ on species $j$, 
 $d_{k}$ is the average of the inner and outer radii of annulus $k$, 
@@
 W_{ijm}  =\sum_k {w_{ijm,k}} .
+\label{eqn:wijm}
 \end{equation} 
@@
 
-\new{We assume that competition kernels $F_{jm}(d)$ are non-negative and decreasing, which means that distant plants have less effect 
+\new{We assume that competition kernels $F_{jm}(d)$ are non-negative and decreasing, so that distant plants have less effect 
 than close plants. Otherwise, we let the data dictate the shape of the kernel by fitting a spline model 
@@
 $\beta$, $\boldsymbol{\omega}$, $\vec{b}$ and $\eta$ as parameters to be fitted. We implemented this in the statistical computing environment, \texttt{R}, 
-by making the spline coefficients and $\eta$ the arguments of an objective function that computes $\boldsymbol{W}$ using the input coefficients, 
-calls the model-fitting function \texttt{lmer} and/or \texttt{glmer} to fit the other parameters in the survival and growth regressions, and returns an approximate AIC value and model degrees of freedom ($df$) for survival and growth combined.
-We used the $\vec{b}$ values at the smoothest
+by making the spline coefficients and $\eta$ the arguments of an objective function that computes $\boldsymbol{W}$ using the input spline coefficients, 
+calls the model-fitting functions \texttt{lmer} and/or \texttt{glmer} to fit the other parameters in the survival and growth regressions, 
+and returns an approximate AIC value and model degrees of freedom ($df$) for survival and growth combined. We used the $\vec{b}$ values at the smoothest 
 local minimum of AIC as a function of $df$, as in Teller et al. (2016). This approach assumes that one measure of crowding affects 
-survival and growth. In addition, 
-for fitting the kernels we assumed that survival and growth depended only on intraspecific crowding, and thus only fitted the
+survival and growth. In addition, for fitting the kernels we assumed that survival and growth depended only on intraspecific crowding, and thus only fitted the
 within-species competition kernels $F_{jj}$. Based on previous work (Adler et al. 2010), we set all $F_{jm}$ equal to $F_{jj}$, meaning that 
@@
 \begin{equation}
-\lambda_{jq,t+1} = C'_{jq,t} \exp{ \left(\gamma_{j,t}^R +  \varphi_{jg}^R + \chi_{jm}^R + 
+\lambda_{jq,t+1} = C'_{jq,t} \exp{\left(\gamma_{j,t}^R +  \varphi_{jg}^R + \chi_{jm}^R + 
 \left \langle \boldsymbol{\omega}^R , \boldsymbol{\sqrt{C'}_{q,t}} \right \rangle \right) }
@@
 where the superscript $R$ refers to Recruitment, $C'_{jq,t}$ is the `effective cover' (cm$^2$) of species $j$ in quadrat $q$ at time $t$, $\gamma$ is a time-dependent intercept, $\varphi$ is a coefficient for the effect of group location,
-$\chi$ is the removal treatment effect, $\boldsymbol{\omega}$ is a vector of coefficients that determine the strength of intra- and interspecific density-dependence, and $\boldsymbol{C'}$ is the vector of ‘effective’ cover of each species in the community.
+$\chi$ is the removal treatment effect, $\boldsymbol{\omega}$ is a vector of coefficients that determine the strength of intra- and interspecific density-dependence, and $\boldsymbol{C'}$ is the vector of ``effective'' cover of each species in the community.
 Eqn (\ref{eqn:recrProcessModel}) is essentially the Ricker model for discrete time population growth (Ricker 1954), except that we use the square root of cover in the density-dependent term.
 We previously used a Ricker equation (Adler et al.
-2010), but using the square-root transformation gave a lower Deviance Information Criterion (DIC) value (Spiegelhalter et al.
+2010), but using the square-root transformation instead gave a lower Deviance Information Criterion (DIC) value (Spiegelhalter et al.
 2002) for the recruitment model.
@@
 
-Due to the possibility that plants outside the mapped quadrat could contribute recruits to the focal quadrat or interact with plants in the focal quadrat, we estimated effective cover as a mixture of the observed cover, $C$, in the focal quadrat, $q$, and the mean cover, $\bar{C}$, across the group, $g$, in which the quadrat is located:
-$C'_{jq,t}=p_j C_{jq,t}+(1-p_j) \bar{C}_{jg,t}$								
-where $p$ is a mixing fraction between 0 and 1 that was estimated as part of fitting the model.
+Due to the possibility that plants outside the mapped quadrat could contribute recruits to the focal quadrat or interact with plants in the focal quadrat, we estimated effective cover as a mixture of the observed cover, $C$, in the focal quadrat, $q$, and the mean cover, $\bar{C}$, across the group, $g$, in
+which the quadrat is located: $C'_{jq,t}=p_j C_{jq,t}+(1-p_j) \bar{C}_{jg,t}$, where $p$ is a mixing fraction between 0 and 1 that was estimated as part of fitting the model.
 
@@
 
+\comment{SPE: how were group effects incorporated into the IPM, and the IBM?}
+
 To quantify the long-term, population-level consequences of the removal effects, we built and then simulated an integral projection model (IPM).
 In our IPM, the population of species $j$ is represented by a density function $n(u_{j,t})$ which gives the density of genets of size $u$ at time $t$, with genet size on a natural-log scale, i.e.
-$n(u_{j,t})du$ is the number of genets whose area (on arithmetic scale) is between $exp(u_j)$ and $exp(u_j+du)$.
-The density function for size $v$ at time $t+1$ is given by
+$n(u_{j,t})du$ is the number of genets whose log area is between $u$ and $u+du$, hence their area
+(on arithmetic scale) is between $exp(u_j)$ and $exp(u_j+du)$. The size distribution function at time $t+1$ is given by
 \begin{equation}
-n(v_{j,t+1})=\int_{L_j}^{U_j} k_j (v_j,u_j,\boldsymbol{\bar{W}}_j  (u_j))n(u_{j,t})   
+n(v_{j,t+1})=\int_{L_j}^{U_j} k_j (v_j,u_j,{\vec{n}})n(u_{j,t})   
 \label{eqn:IPM}
 \end{equation}
-where the kernel $k_j (v_j,u_j,\boldsymbol{\bar{W}}_j  (u_j))$ describes all possible transitions from size $u$ to $v$ and $\boldsymbol{\bar{W}}_j$  is a vector whose elements are the average crowding experienced from each of the other species by an individual of size $u_j$ of species $j$ (we describe below how $\boldsymbol{\bar{W}}_j$ is calculated).
-The integral is evaluated over all possible sizes from a lower size limit $L$ to an upper size limit $U$ that extends past the range of observed sizes.
-
-The kernel is constructed from the fitted survival ($S$), growth ($G$), and recruitment ($R$) models:
+where the kernel $k_j (v_j,u_j,\vec{n}_j)$ describes all possible transitions from size $u$ to $v$ and $\vec{n}$ 
+is the set of size-distribution functions for all species in the community. The integral is evaluated over all possible sizes from a 
+lower size limit $L$ to an upper size limit $U$ that extends past the range of observed sizes.
+
+The kernel is constructed from the fitted survival ($S$), growth ($G$), and recruitment ($R$) models, 
 \begin{equation}
-k_j (v_j,u_j,\boldsymbol{\bar{W}}_j ) = S_j(u_j,\boldsymbol{\bar{W}}_j  (u_j)) G_j(v_j,u_j,\boldsymbol{\bar{W}}_j  (u_j)) 
-+ R_j (v_j,u_j,\boldsymbol{\bar{W}}_j) .
+k_j(v_j,u_j,\vec{n})=S_j(u_j,\vec{n})G_j(v_j,u_j,\vec{n})+R_j(v_j,u_j,\vec{n}) du_j
 \label{eqn:IPM} 
 \end{equation}
-$S$ is given by (Eqn. \ref{eqn:survReg}) and $G$ is given by (Eqn. \ref{eqn:growReg}). The recruitment rate, $\lambda$, 
-defined in (\ref{eqn:recrDataModel} and \ref{eqn:recrProcessModel}), gives the expected 
-number of new recruits produced per quadrat, per unit area of parents. To incorporate this recruitment function into the IPM, 
-we assumed that individual fecundity increases linearly with individual size (area, equal to $e^u$). The recruitment kernel is thus
-$R_j(v_j,u_j,\boldsymbol{\bar{W}}_j)=C_0(v_j)e^{u_j}\lambda_j$ (Adler et al. 2010) where $C_0$ is the
-size distribution of new recruits, and $\lambda_j$ is calculated from (\ref{eqn:recrDataModel} and \ref{eqn:recrProcessModel}) with
-population densities converted into total cover per m$^2$, $C$.
-
-In the vital rate regressions and in the IBM, the neighborhood crowding index, $W_ijm$ , unique to each individual $i$ based on the spatial locations and sizes of neighboring plants.
-This measure cannot be extended to the IPM, because the IPM is spatially implicit.
-Therefore, we used a spatially
-implicit mean-field approximation that captures the essential features of spatially-explicit neighborhood interactions (Adler et al.
-2010).
+$S$ is given by (Eqn. \ref{eqn:survReg}) and $G$ is given by (Eqn. \ref{eqn:growReg}), using an expected neighborhood crowding
+index calculated from the size distributions of the competing species. In the vital rate regressions and in the IBM, each 
+individual has a unique neighborhood crowding index based on the spatial locations and sizes of neighboring plants. This 
+measure cannot be extended to the IPM, which tracks individual size but not location. Instead we used a 
+mean-field approximation that captures the essential features of spatially-explicit neighborhood interactions (Adler et al. 2010). 
 We found that in both the observed data and IBM simulations, heterospecific individuals were approximately randomly distributed with respect to each other, 
-but conspecific individuals displayed non-random, size-dependent spatial patterns: while small genets were randomly distributed, large genets 
-were segregated from each other.
-The overdispersion of large conspecific genets was introduced into the IPM with a 'no-overlap' rule (Adler et al.
-2010, 2012).
-In \ref{suppMethods} we explain our implementation of the no-overlap rule with our nonparametric competition kernels.
+but conspecific individuals displayed a non-random, size-dependent patterns: small genets were randomly distributed, while large genets 
+were segregated from each other. The overdispersion of large conspecific genets was introduced into the IPM 
+with a 'no-overlap' rule (Adler et al. 2010, 2012). In \ref{suppMethods} we explain how the spline competition kernels with 
+the no-overlap rule kernels were used to compute neigbhorhood crowding index as a function of individual size for the IPM.
+
+For recruitment in the IPM, the factor $ \Phi_j  = \exp{\left(\gamma_{j,t}^R +  \varphi_{jg}^R + \chi_{jm}^R + 
+\left \langle \boldsymbol{\omega}^R , \boldsymbol{\sqrt{C'_{q,t}}} \right \rangle \right) }$
+in eqn. (\ref{eqn:recrDataModel}) gives the expected 
+area of new recruits per quadrat, per unit area of parents. To incorporate this recruitment function into the IPM, 
+we assumed that individual fecundity increases linearly with individual area (equal to $e^u$ for a size-$u$ individual). The recruitment kernel is thus
+$R_j(v_j,u_j,\vec{n})=c_{0,j}(v_j)e^{u_j}\Phi_j$ (Adler et al. 2010) where $c_{0,j}$ is the
+size distribution of new recruits. $\Phi_j$ is calculated from eqn. (\ref{eqn:recrDataModel}) with total cover values
+computed from the population densities as $C'_{j,t} = \int e^u n_j(u,t) du$.
 
 We simulated the IPM to estimate equilibrium abundances under three different scenarios.
-The first scenario featured all four species and our "baseline" model for control plots (no removal effects).
+The first scenario featured all four species and our ``baseline'' model for control plots (no removal effects).
 We initialized all species with arbitrarily low cover and ran the model for 500 time steps (years) to allow the community to reach an equilibrium.
-To incorporate temporal environmental variation, at each time step we randomly selected one of the 30 sets of random year effects for the survival, growth and recruitment models ("kernel selection" in the terminology of Metcalf et al.
-20xx))).
+To incorporate temporal environmental variation, at each time step we randomly selected one of the 30 sets of random year effects for the survival, growth and recruitment models (``kernel selection'' in the terminology of Metcalf et al.
+2016))).
 We calculated the equilibrium cover for each species by simulating an additional 2000 time steps and averaging each species' cover over this period.
@@
 
-Trends in cover by species and removal treatment were variable.
-For example, grass removal had little effect on the cover of \textit{A.
-tripartita} (Fig.
-\ref{fig:CoverTrends}) , but did appear to cause an increase in the cover of \textit{P.
+Trends in cover by species and removal treatment were variable. For example, grass removal had little effect on the cover of \textit{A. tripartita} 
+(Fig.
+\ref{fig:CoverTrends}), but \new{shrub removal} did appear to cause an increase in the cover of \textit{P.
 spicata}.
@@
 \ref{fig:CoverChange}).
-Grass removal had no significant effect on \textit{A.
-tripartita} (95\% confidence interval on removal effect: -0.33, 0.27) or \textit{Poa secunda} (95\% CI: 0.02, 0.87) population growth but did increase \textit{H.
-comata} (95\% CI: -0.40, 0.28) and \textit{P.
-spicata} (95\% CI: 0.04, 0.50) population growth (Table \ref{table:coefficients}).
-However, these treatment comparisons do not address our research questions about whether responses to removal were adequately predicted by models fit to observational data (i.e.
-control plots).
+\new{Removal treatment} had no significant effect on
+\textit{A. tripartita} (95\% confidence interval on removal treatment effect: -0.33, 0.27) or 
+\textit{Poa secunda} (95\% CI: -0.40, 0.28) population growth, but did increase 
+\textit{H. comata} (95\% CI: 0.02, 0.87) and
+ \textit{P. spicata} (95\% CI: 0.04, 0.50) population growth 
+(Table \ref{table:coefficients}). However, these treatment comparisons do not address our questions about whether responses to removal were 
+adequately predicted by models fit to observational data on non-removal plots. 
+\comment{SPE: I changed the CI numbers in this paragraph to match what I think the Table means.}
 
@@
 
-To quantify the long-term population-level consequences of removals, we used an IPM to project stochastic equilium cover.
+To quantify the long-term population-level consequences of removals, we used an IPM to project stochastic equilibrium cover.
 We concentrated on grass species' responses to \textit{A.
@@
 
-The small competitive release of many species in our system implies that classic resource competition is not the primary factor determining community structure.
+The small competitive release of many species in our system implies that classic resource competition 
+is not the primary factor determining community structure. 
+\comment{SPE: I think this depends on what you mean by ``classic resource competition''.
+If you mean Tilman-style competition for one limiting resource distributed homogeneously, then I agree.
+But a MacArthur-Levins model with resource spectrum and distinct utilization curves could also be called classic resource competition,
+and that model will have weak to nonexistent competitive release if niche overlap is low.} 
 Interactions with above- or belowground enemies can also regulate populations and, given enough species-specificity in the effects of these enemies (Clark paper, Janzen, Connell), decouple interspecific dynamics.
@@
 But for many species in other communities, interspecific competition may still be an important driver.
-In closed canopy-forests where competition for light is intense, we might have found different results than in our system, in which competition for a number of belowground resources is more important than competition for light.
+In closed-canopy forests where competition for light is intense, we might have found different results than in our system, in which competition for a number of belowground resources is more important than competition for light.
 However, a number of recent studies from forests have found that heterospecific competitive effects are dramatically weaker than conspecific effects (Uriarte refs, Comita et al.
@@
 Determining where interspecific interactions and competitive release are strong, and where they are not, remains a fascinating research problem.
+\comment{SPE: Equally interesting, I think, is the question of why interspecific
+interactions are so weak (when they are), as opposed to what you would expect from ``all plant species compete for water, light, and a few mineral resources''.}
 
@@
 
-Funding: NSF DEB-1353078 and DEB-1054040, Utah Agricultural Experiment Station (get journal paper number).
+Funding: NSF DEB-1353078, DEB-1054040, DEB 1353039 (SPE), Utah Agricultural Experiment Station (get journal paper number).
 The USDA-ARS Sheep Experiment Station generously provided access to historical data and the field experiment site.
@@
 \end{doublespacing}
-
-%%%%%%%%%%%%%%%%% For word count
 
@@
 
+\subsection{Interspecific covariance in local crowding} 
 We explored interspecific covariance in local crowding experienced by individual plants, by regressing the $W$ values exerted by one neighbor species, the response variable, against the $W$ values of all other species, the independent variables.
 Because some $W=0$, we conducted two separate regressions.
-First, using all $W$'s, we performed a generalized linear regression with a logit link function to evaluate whether the probability that the focal species' $W=0$ is influenced by the value of other species' $W$'s.
-Second, for the set of records in which the focal species $W>0$, we performed a linear regression.
-We repeated these regressions for individuals of each focal speices.
+First, using all $W$'s, we fitted a generalized linear model with a logit link function to evaluate whether the probability that the focal species' probability of having $W=0$ is influenced by the value of other species' $W$'s.
+Second, for the set of records in which the focal species has $W>0$, we performed a linear regression.
+We repeated these regressions for individuals of each focal species.
 Due to large samples size, $W$ values were often statistically significant predictors.
@@
 The maximum reduction in deviance for the generalized linear regressions and $R^2$ for the linear regressions were both less than 8\%.
-The \texttt{R} code for this analysis is included as ..\texttt{\textbackslash Wdistrib\textbackslash exploreSurvivalWs.r})
-
-Spatially-implicit approximations of local crowding for the IPM. Steve? Paste in the MeanFieldSplineIPM.tex?
-
+The \texttt{R} code for this analysis is included as ..\texttt{\textbackslash Wdistrib\textbackslash exploreSurvivalWs.r}) \comment{It's not clear what the independent and dependent variables are, in the analyses described
+here. It might help to use the notation from the Methods section?}
+
+\subsection{Mean field approximation of local crowding for the IPM} 
+\citet{adler-etal-2010} developed a mean field approximation for local crowding when the
+competition kernels are all Gaussian functions, $F_{jm}(d) = e^{-\alpha_{jm} d^2}$. The approximation is explained in 
+the online SI to \citet{adler-etal-2010} and in section 5.3 of \citet{ipmbook-2016}. 
+Here we explain how that approximation was modified for the IPMs in this paper, which
+used fitted nonparametric competition kernels.
+
+For $j \ne m$ (between-species competition), overlap between individuals is allowed. The mean field approximation is 
+that from the perspective of any focal plant in species $j$, individuals of species $m$ are distributed at random in space, 
+independent of each other and of their size.
+
+Consider the region between the circles of radius $x$ and $x+dx$ centered on a focal genet of species $j$. The area of this annulus
+is $2 \pi x \; dx$  to leading order for $dx \approx 0$. A species $m$ genet 
+in the annulus puts competitive pressure $F_{jm}(x)$ times its area on
+the focal genet. The expected total competitive pressure from all such genets 
+is therefore is $F_{jm}(x) 2 \pi x \; dx$ times the expected fractional cover of species $m$ in the annulus 
+(fractional cover is the total area of species $m$ genets, as a fraction of the total area). The excepted fractional cover $C_m$ of species $m$
+in the annulus equals its fractional cover in the habitat as a whole, because of the assumption of random distribution
+spatial distributions. We therefore have $C_m  = \int e^u n_m(u,t) du/A$ where $A$ is the total area of the habitat. 
+The total expected competitive pressure on a species-$j$ genet due to species $m$ is then 
+\begin{equation}
+W_{jm} = \int_0^\infty{C_m F_{jm}(x) 2 \pi x \; dx}  = C_m \left [2 \pi \int_0^{\infty} x F(x) \, dx \right ].
+\label{eqn:wbarm}
+\end{equation} 
+The quantity in square brackets is a constant (that is, it only depends on what the kernel function
+is) so it can be computed once and for all for each kernel used in the IPM. The integral is finite because
+all fitted kernels fall to zero at a finite distance from the focal plant.
+
+Our kernel fitting method only uses competition kernel values at the ``mid-ring'' distances
+halfway between the inner and outer radii of a series of annuli around each focal
+plant, scaled so that the value at the innermost mid-ring distance equals 1. 
+In the IPM we defined the kernel at other distances by linear interpolation between values at 
+mid-ring distances, except that for the innermost ring a kernel value of 1 was specified at the
+outer radius of the ring and at distance $x=0$.
+
+Now consider within-species competition. We assume that conspecifics cannot overlap. Genet shapes are irregular, but we 
+nonetheless implement the no-overlap rule by assuming that a genet of log area $u_i$ is a 
+circle of radius $r_i$ where $\pi r_i^2 = e^{u_i}$. The no-overlap rule is then that the centroids of two conspecific individuals 
+must be separated by at least the sum of their radii.
+
+For any one focal genet, the no-overlap restriction on its neighbors' locations affects 
+only a negligibly small part of the habitat. The expected cover of individuals in the places
+where they can occur (relative to one focal plant) is thus assumed to equal their expected locations
+in the habitat as a whole.
+
+Let $C_m(u)$ be the total cover of species $m$ genets of radius $r$ or smaller, 
+\begin{equation}
+C_m(r) = \int_L^{\log(\pi r^2)}{\! \! \! e^z n_m(z,t) \, dz} .
+\label{eqn:cm}
+\end{equation}
+A focal genet of radius $r$ cannot have any conspecific neighbors centered 
+at distances less than $r$. It can have a neighbor centered at distance $x>r$ if that neighbor's
+radius is no more than $x-r$. Adding up the expected cover of all such possible neighbors
+for a focal genet of radius $r$,    
+\begin{equation}
+W_{mm}(r) = 2 \pi \int_r^{\infty}F_{mm}(x) x C_m(x-r) \, dx
+\label{eqn:wbarmr} 
+\end{equation}
+This integral is again finite and computable because the kernels $F$ fall to 0 at finite $x$.
+
+\clearpage 
+\newpage  
 \section{Additional Tables}
@@
 %\begin{center}
+\comment{SPE: are the four columns actually the four species?}
 \begin{tabular}{l c c c c }
@@
  & Model 1 & Model 2 & Model 3 & Model 4 \\
+Species & ARTR & HECO & POSE & PSSP \\
 \hline
@@
                       & $[-0.33;\ 0.27]$ & $[-0.32;\ 0.24]$ & $[-0.31;\ 0.27]$ & $[-0.33;\ 0.15]$ \\
-TreatmentNo\_grass    & $-0.09$          &                  &                  &                  \\
+Treatment: No grasses    & $-0.09$          &                  &                  &                  \\
                       & $[-0.45;\ 0.27]$ &                  &                  &                  \\
-TreatmentNo\_shrub    &                  & $0.44^{*}$       & $-0.06$          & $0.27^{*}$       \\
+Treatment: No shrub    &                  & $0.44^{*}$       & $-0.06$          & $0.27^{*}$       \\
                       &                  & $[0.02;\ 0.87]$  & $[-0.40;\ 0.28]$ & $[0.04;\ 0.50]$  \\

### 2016-07-01 removals/growth/write_params.r
Recording max CI for growth and survival

@@ -27,6 +27,13 @@ formatGrowthPars <-function(model,outfile){
   tmp[1,]=model$summary.fixed$mean
   params=cbind(params,tmp)
   
+  # record the edge of the 95% CI furthest from zero
+  j <- grep("Treatment",row.names(model$summary.fixed))
+  tmp <- model$summary.fixed[j,c("0.025quant","0.975quant")]
+  k <- which(abs(tmp)==max(abs(tmp)))
+  params$TreatMaxCI <- NA
+  params$TreatMaxCI[1] <- as.numeric(tmp[k])
+  
   #variance 
   params$sigma.a=NA; params$sigma.a[1]=coef(outVar)[1] 
   params$sigma.b=NA; params$sigma.b[1]=coef(outVar)[2]

### 2016-07-01 removals/removal_analysis_wrapper.r

@@ -18,7 +18,7 @@ statsOutput <- paste0(getwd(),"/stats_tables.tex")
 source("treatment_trends_removals.r")
 
 # clean up
-tmp=ls() ; tmp=tmp[tmp!="root"]
+tmp=ls() ; tmp=tmp[tmp!="root" & tmp!="statsOutput"]
 rm(list=tmp)
 
 ###
@@ -141,7 +141,7 @@ write.csv(trtTests,"treatment_test_results.csv",row.names=F)
 source("treatment_test_figure.r")
 
 # clean up
-tmp=ls() ; tmp=tmp[tmp!="root"]
+tmp=ls() ; tmp=tmp[tmp!="root" & tmp!="statsOutput"]
 rm(list=tmp)
 
 ###
@@ -153,7 +153,7 @@ library("TeachingDemos") # for inset plots
 source("Wdistrib/exploreSurvivalWs.R")
 
 # clean up
-tmp=ls() ; tmp=tmp[tmp!="root"]
+tmp=ls() ; tmp=tmp[tmp!="root" & tmp!="statsOutput"]
 rm(list=tmp)
 
 ###
@@ -169,8 +169,9 @@ source("ibm/ibm_removal_1step.r")
 source("ibm/summarize_sims1step.r")
 
 # clean up
-tmp=ls() ; tmp=tmp[tmp!="root"]
+tmp=ls() ; tmp=tmp[tmp!="root" & tmp!="statsOutput"]
 rm(list=tmp)
+
 
 ###
 ### 5. get equilibrium cover from an IPM ###############################

### 2016-07-01 removals/survival/write_params.r

@@ -1,5 +1,5 @@
 
-# function to format and output parameters of growth models
+# function to format and output parameters of survival models
 
 formatSurvPars <-function(model,outfile){
 
@@ -20,6 +20,13 @@ formatSurvPars <-function(model,outfile){
   colnames(tmp)=model$names.fixed
   tmp[1,]=model$summary.fixed$mean
   params=cbind(params,tmp)
+  
+  # record the edge of the 95% CI furthest from zero
+  j <- grep("Treatment",row.names(model$summary.fixed))
+  tmp <- model$summary.fixed[j,c("0.025quant","0.975quant")]
+  k <- which(abs(tmp)==max(abs(tmp)))
+  params$TreatMaxCI <- NA
+  params$TreatMaxCI[1] <- as.numeric(tmp[k])
 
   write.table(params,outfile,row.names=F,sep=",")
     

### 2016-08-15 removals/survival/fetchSurvData.r
Just fiddling

@@ -15,7 +15,7 @@ fetchSdat <- function(doSpp,speciesList,datadir,distWts){
   ringD$year<-ringD$year
   
   # merge D with ringD (D contains fewer rows)
-  D1<-merge(D1,ringD,by.x=c("quad","year","trackID"),by.y=c("quad","year","genetID"))
+  D1<-merge(D1,ringD,all.x=T,by.x=c("quad","year","trackID"),by.y=c("quad","year","genetID"))
   D1=D1[order(D1$X),]
   rm(ringD)
   row.names(D1) <- NULL  

### 2016-08-17 removals/Wdistrib/exploreSurvivalWs.R
Added quantile regression no overlap analysis

@@ -56,7 +56,7 @@ for(iSpp in 1:4){
   #   Dall$W.HECO[ii] <- 0 ; Dall$W.POSE[ii] <- 0 ; Dall$W.PSSP[ii] <- 0
 
   allD[[iSpp]] <- Dall[,c("year","Treatment","Group","W.ARTR", "W.HECO","W.POSE","W.PSSP","W.allcov","W.allpts")]
-  rm(Dall,D1,D2,tmp)
+  rm(Dall, D1,D2,tmp)
 
 }
 
@@ -75,8 +75,83 @@ for(i in 1:4){
 }
 dev.off()
 
+
 #####################
-#  2. Model W's (with help from Giles Hooker)
+#  2. Calculate no overlap zone (with code from Steve Ellner)
+####################
+
+# import all rings data
+setwd("..")
+source("survival/fetchSurvData-rings.r")
+setwd("Wdistrib")
+
+ringD <-list()
+for(iSpp in 1:4){
+  doSpp <- sppList[iSpp]
+  # get old data
+  D1 <- fetchSdatRings(doSpp=doSpp,speciesList=sppList,datadir=dataDir1,distWts=dists)
+  D1$year <- D1$year+1900
+  # import modern data
+  D2 <- fetchSdatRings(doSpp=doSpp,speciesList=sppList,datadir=dataDir2,distWts=dists)
+  D2 <- D2[,-c(13,15)]
+  # combine old and modern
+  Dall <- rbind(D1,D2)
+  # merge in treatment data
+  tmp <- read.csv(paste(dataDir2,"/quad_info.csv",sep=""))
+  tmp <- tmp[,c("quad","Treatment","Group")]
+  Dall <- merge(Dall,tmp, all.x=T)
+  Dall <-subset(Dall,Treatment=="Control") # onlu use control treatment data
+  ringD[[iSpp]] <- Dall
+}
+rm(Dall)
+
+midRings=c(seq(1,19,2),seq(22.5,47.5,5),seq(55,145,10))
+
+# loop through species again and analyze no overlap radius
+fits<-list(0)
+png("NoOverlap.png",height=11,width=8.5,res=400,units="in")
+par(mfrow=c(4,4),mgp=c(2,1,0),mar=c(4,4,2,1),cex.axis=1.2,cex.lab=1.2)
+for(iSpp in 1:4){
+  
+  doSpp <- sppList[iSpp]
+  data <- ringD[[iSpp]]
[…]

### 2016-08-17 removals/removal_analysis_wrapper.r

@@ -149,6 +149,7 @@ rm(list=tmp)
 ###
 
 library("TeachingDemos") # for inset plots
+library("quantreg")
 
 source("Wdistrib/exploreSurvivalWs.R")
 

### 2016-08-17 removals/survival/fetchSurvData-rings.r

@@ -1,0 +1,37 @@
+
+fetchSdatRings <- function(doSpp,speciesList,datadir,distWts){
+
+  infile=paste(datadir,"/speciesdata/",doSpp,"/survD.csv",sep="")
+  survD=read.csv(file=infile)
+  D1=survD[survD$allEdge==0,];
+  D1$year <- D1$year
+  D1$logarea=log(D1$area)
+  D1$quad=as.character(D1$quad)
+  
+  # import neighbor data
+  ringD <- read.csv(paste(datadir,"/speciesdata/",doSpp,"/",doSpp,"_nbhood_rings.csv",sep=""))
+  tmpD <- read.csv(paste(datadir,"/speciesdata/",doSpp,"/",doSpp,"_nbhood_rings_allothers.csv",sep=""))
+  ringD<-merge(ringD,tmpD)
+  ringD$year<-ringD$year
+  
+  # merge D with ringD (D contains fewer rows)
+  D1<-merge(D1,ringD,all.x=T,by.x=c("quad","year","trackID"),by.y=c("quad","year","genetID"))
+  D1=D1[order(D1$X),]
+  rm(ringD)
+  row.names(D1) <- NULL  
+  
+  # calculate W's (MAKE SURE NOT TO REORDER D!)
+  W <- matrix(NA,NROW(D1),length(speciesList))
+  colnames(W) <- paste("W.",speciesList,sep="")
+
+  # do big 4
+  for(iSpp in 1:4){
+    neighborCols=which(substr(names(D1),1,4)==speciesList[iSpp]) # pull out annulus data for the focal species 
+    dist_wts <- distWts[,paste0(speciesList[iSpp])]
+    C <- data.matrix(D1[,neighborCols]) #matrix of conspecific areas in the annuli 
+    W[,iSpp] <- C%*%dist_wts 
+  }
+  
+  return(D1)
+
+}

### 2016-08-17 removals/manuscript/removal_ms.tex
Edits on ms

@@
 
-\new{Here we present a removal experiment, conducted in a sagebrush steppe} 
-% We conducted a removal experiment in a sagebrush steppe 
-to test whether our models, which are based on observational data, underestimate the strength of interspecific competition and competitive release.
+To test whether our models, which are based on observational data, underestimate the strength of interspecific competition and competitive release, we conducted a removal experiment in a sagebrush steppe.
 We studied the responses of three perennial grass species to removal of sagebrush, and the response of sagebrush to removal of all grasses.
@@
 removal treatments after accounting for differences in local species composition?
-Second, how do short and long-term population-level responses to removals compare to projections from multispecies population models based on observational data?
+Because our demographic models include the effects of neighborhood crowding, they should be able to predict how experimental removal of neighbors alters demographic performance.
+If the full effect of experimental removals is stronger  than predicted, it means that our models underestimate competitive release.
+Our second question integrates across the full life cycle: how do short and long-term population-level responses to removals compare to projections from multispecies population models based on observational data?
+If models based on observational data effectively predict competitive release, the population dynamics they project should be similar to projections from models that include information from the experimental removal treatments.
 
@@
 We used the following criteria to select the quadrat locations.
-Each plot had to have xx?
-cumulative cover/density?
-of the three target perennial grass species as well as yy?
-cover/density of \textit{Artemisia}.
-We mapped the quadrats in June, 2011 and then implemented the removal treatments in September (?), 2011.
+Each plot had to have 5\% canopy cover of \textit{Artemisia tripartita} and at least 3\% combined basal cover of the three focal perennial grasses.
+We avoided plots falling on hill slopes, areas with greater than 20\% bare rock, or with over 10\% cover of the woody shrubs \textit{Purshia tridentata} or \textit{Amelanchier utahensis}.
+We mapped the quadrats in June, 2011 and then implemented the removal treatments in September, 2011.
 We randomly assigned 8 of the sixteen new quadrats to the sagebrush removal treatment, and the remaining 8 to the grass removal treatment.
-We cut sagebrush stems at ground level and applied herbicide to the cut stems.
-In the grass removal plots, we herbicided all perennial grasses.
-We removed plants zz cm beyond the quadrat boundaries to minimize edge effects.
-The sagebrush removal was virtually 100\% effective; grass removals required additional herbicide applications in April and May of each year, with fewer grasses remaining each additional year.
+We cut sagebrush stems at ground level and applied a 5\% solution of Roundup Pro (Glyphosate, Monsanto) to the cut stems.
+In the grass removal plots, we painted the same herbicide solution on all perennial grasses.
+We removed plants 50 cm beyond the quadrat boundaries to minimize edge effects.
+The sagebrush removal was virtually 100\% effective; grass removals required additional herbicide applications in April and May of each year, with fewer grasses remaining each year.
 
@@
 Our regression models for survival, growth and recruitment follow previous work (Adler et al.
-2010, Chu and Adler 2015) with a few exceptions, such as a new intercept term for the removal treatment.
-We model the survival probability of an individual genet as a function of genet size, the neighborhood-scale crowding experienced by the genet from both conspecific and heterospecific genets, temporal variation among years, and permanent spatial variation among groups of quadrats (`group'; here means a set of nearby quadrats located within one pasture or grazing exclosure), along with the removal treatment effect.
+2010, Chu and Adler 2015).
+We model the survival probability of an individual genet as a function of genet size, the neighborhood-scale crowding experienced by the genet from both conspecific and heterospecific genets, temporal variation among years, and permanent spatial variation among groups of quadrats (`group'; here means a set of nearby quadrats located within one pasture or grazing exclosure).
+Where the current analysis departs from previous work is the inclusion of a categorical removal treatment effect.
+If the conspecific and heterospecific crowding effects explain demographic rates in removal treatment plots as well as they do in control plots, then there will be no need for the removal treatment effect, and the estimated coefficient will be small and not significantly different from zero.
+However, if demographic rates in removal plots differ from control plots even after accounting for crowding effects, then the removal treatment effect will be significant.
+For example, we might find that, holding crowding constant, grasses grow more quickly in the removal plots than in the control plots.
+This result would indicate that our "baseline model," a model including all estimated parameters except for the removal treatment effect, underestimates competitive release.
+We refer to the model that includes the estimated removal treatment effect as the "removal model."
+
 Formally, we modeled the survival probability, $S$, of genet $i$ in species $j$, group $g$, and removal treatment $h$, from time $t$ to $t+1$  as
@@
 \end{equation}
-where $\gamma$ is a time-dependent intercept, $\varphi$ is the coefficient for the effect of spatial group,$\chi$ is the removal treatment effect, $\beta$ is the coefficient that represents the effect of log genet size, $u$, on survival.
+where $\gamma$ is a time-dependent intercept, $\varphi$ is the coefficient for the effect of spatial group, $\chi$ is the removal treatment effect, $\beta$ is the coefficient that represents the effect of log genet size, $u$, on survival.
 $\boldsymbol{\omega}$ is a vector of interaction coefficients which determine the impact of crowding, $\boldsymbol{W}$, by each species on the focal species.
@@
 
-\new{We assume that competition kernels $F_{jm}(d)$ are non-negative and decreasing, so that distant plants have less effect 
+We assume that competition kernels $F_{jm}(d)$ are non-negative and decreasing, so that distant plants have less effect 
 than close plants. Otherwise, we let the data dictate the shape of the kernel by fitting a spline model 
@@
 survival and growth. In addition, for fitting the kernels we assumed that survival and growth depended only on intraspecific crowding, and thus only fitted the
-within-species competition kernels $F_{jj}$. Based on previous work (Adler et al. 2010), we set all $F_{jm}$ equal to $F_{jj}$, meaning that 
+within-species competition kernels $F_{jj}$. Based on previous work (Adler et al. 2010), we set all $F_{mj}$ equal to $F_{jj}$, meaning that 
 the within-species competition kernel for species $j$ is also used to determine the effect of all other species on species $j$.
@@
 \ref{fig:CompKernels}.
-}
-\comment{SPE: see Robin's comment about whether you mean $F_{mm}$ or $F_{jj}$; I'm not sure which one you did.}
 
@@
 triparitia} to zero, but this time we projected the model using the vital rate regressions that include removal effects.
-By comparing the output of this simulation with the previous scenario we can quantify the long-term consequences of differences between our baseline model and a model that includes additional removal treatment effects.
+If our baseline model successfully predicts competitive release observed in the removal experiment, then the results of the second and third scenarios should be similar.
+However, to the extent that the baseline model underestimates (or overestimates) the actual competitive release, the results of the second and third scenarios will differ.
 
@@
 (Fig.
-\ref{fig:CoverTrends}), but \new{shrub removal} did appear to cause an increase in the cover of \textit{P.
+\ref{fig:CoverTrends}), but shrub removal did appear to cause an increase in the cover of \textit{P.
 spicata}.
@@
 \ref{fig:CoverChange}).
-\new{Removal treatment} had no significant effect on
-\textit{A. tripartita} (95\% confidence interval on removal treatment effect: -0.33, 0.27) or 
+The removal treatments had no significant effect on
+\textit{A. tripartita} (95\% confidence interval on removal treatment effect: -0.45, 0.27) or 
 \textit{Poa secunda} (95\% CI: -0.40, 0.28) population growth, but did increase 
@@
 adequately predicted by models fit to observational data on non-removal plots.
-\comment{SPE: I changed the CI numbers in this paragraph to match what I think the Table means.}
 
@@
 \ref{fig:W-by-treatment}).
+These pre-existing differences in composition raise the possibility that plots selected for the removal treatments might differ from the control plots in some unobserved covariate that could influence plant performance.
 
@@
 %\begin{center}
-\comment{SPE: are the four columns actually the four species?}
 \begin{tabular}{l c c c c }
 \hline
- & Model 1 & Model 2 & Model 3 & Model 4 \\
 Species & ARTR & HECO & POSE & PSSP \\

### 2016-09-01 doc/precip_experiment_ms.Rmd
adding the first skeleton outline of the precip prediction analysis

---
bibliography: ~/Documents/precip_experiment/doc/precip_experiment.bib
csl: components/ecology.csl
fontsize: 12pt
geometry: margin=1in
header-includes:
- \usepackage{lineno}
- \usepackage{booktabs}
- \linenumbers
- \usepackage{setspace}
- \doublespacing
- \usepackage[documeZnt]{ragged2e}
- \usepackage{amsmath}
- \usepackage{lscape}
- \usepackage{bm}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
layout: 12pt
linkcolor: black
output:
  pdf_document:
    fig_caption: yes
urlcolor: black
---

```{r caching, include=FALSE}
library('png')
library('grid')
library("methods")
library("knitr")
#library("kfigr")
basename <- "manuscript"
opts_chunk$set(fig.path = paste("components/figure/", basename, "-", sep=""),
               cache.path = paste("components/cache/", basename, "/", sep=""))
opts_chunk$set(cache = FALSE)
opts_chunk$set(tidy=FALSE, warning=FALSE, message=FALSE, 
               comment = NA, verbose = TRUE, echo=FALSE)

# PDF-based figures
opts_chunk$set(dev='pdf')
```

```{R echo = FALSE}
# ds_stats <- readRDS('../data/temp_data/dataset_short_summary.RDS')
# cover_stats <- readRDS('../data/temp_data/cover_summary.RDS')
# eq_df <- readRDS('../data/temp_data/eq.RDS' ) 
# model <- readRDS('../analysis/model_out/lmer_clim_model.rds')
# extremes <- readRDS('../data/temp_data/extremes.RDS')
# species_summary <- readRDS('../data/temp_data/species_summary.RDS')
# cover_vs_production <- readRDS('../data/temp_data/cover_vs_production.RDS')
# production_n <- as.numeric(cover_vs_production[2])
# cover_n <- as.numeric(cover_vs_production[1])
```

\renewcommand*{\thefootnote}{\fnsymbol{footnote}}

\begin{centering}
\textbf{\large{Predicting the effects of precipitation change on plant communities}}

\textsc{\small{Andrew R. Kleinhesselink\footnote{Corresponding author: arklein@aggiemail.usu.edu}\textsuperscript{1}, Peter B. Adler\textsuperscript{1} and Andrew Tredennick\textsuperscript{1}}}

\textit{\small{\textsuperscript{1}Department of Wildland Resources and the Ecology Center, 5230 Old Main Hill, Utah State University, Logan, Utah 84322 USA}} \\

\end{centering}

\renewcommand*{\thefootnote}{\arabic{footnote}}
\setcounter{footnote}{0}

Abstract
--------

Global climate change has put the issue of generating and validating predictions at the forefront of ecology. Critical issues for ecological predictions are whether models derived from historical data can predict ecological responses to climate conditions that fall outside of the observed range of climate variation, and whether biotic interactions with other species need to be well understood in order to make accurate predictions. In this experiment we tested our ability to predict how a natural grassland plant community would respond to a multi-year climate manipulation experiment. We based our predictions on statistical models that relate annual climate and local competition to the growth, survival and recruitment of four plant species. Models for each demographic rate were fit using 20+ years of observational data collected during the middle part of the 20th century. For each demographic rate we tested three models with increasing levels of ecological complexity: a model without interactions between neighboring species; one with interactions from all species being treated equivalently; and one with species specific interactions. We then used each of these models to predict observations made from 2007 to 2016 in ambient and experimentally manipulated climate conditions. Our experiment shows that ....

*Key words: ecological prediction, climate change, sagebrush steppe, rain-out shelter, drought*

Introduction
-------------

Making successful predictions about the natural world has great practical utility but also is a sign of comprehensive and successful scientific understanding. In ecology the goal of prediction seems ever more urgent.  Every ecological system on earth is experiencing and is projecting to continue experiencing unprecedented rates of climate change driven by anthropogenic greenhouse gas emmissions. As a society our discussions of how to mitigate the causes of and adapt to the effects of climate change are explicitly dependent on ecological predictions.  However, ecological predictions of the consequences of climate change are far from simple and there have been few concrete tests of how well our predictions actually work in the field.

Outline of problem:

Proposed experiment:

In this experiment we tested our ability to predict how a natural grassland plant community would respond to an experimentally imposed multi-year climate manipulation. We based our predictions on statistical models that relate annual climate and local competition to the growth, survival and recruitment of four species. Models for each demographic rate were fit using 20+ years of observational data collected during the middle part of the 20th century. For each demographic rate we tested models with three levels of ecological complexity, one without any interactions between neighboring species, one including intra-specific interactions, and one including both intra-specific and inter-specific interactions. We then generated predictions from each of these models for the response of each species to the conditions in the modern experiment.

Materials and Methods
---------------------
#### Multi-year plant community data  
Long-term data on individual abundance and size of plants in a sagebrush steppe community have been collecting since the 1930s at the US Sheep Experiment Station in Dubois Idaho.  The plots were recensused beginning in 2007 using methods identical to the original studies. These data allow for the fitting of statistical models that estimate the effects of climate and competitive effects on individual plants in the plots.  In theory these statistical models can be used to project the effects of both a change in the competitive environment at this site--or a change in the climate.

### Rain-out shelter and irrigation experiment 
In the summer of 2011, we established eight pairs of new plots within the largest exclosure at the USSES. Areas for plot pairs were selected to be similar in physical environment and vegetation composition to the original plots located at this site.  Plot locations were determined by running eight transects across the grazing exclosure and establishing new plot pairs at a random location on each transect. We rejected areas that fell on hill slopes, areas with greater than 20% bare rock, or areas with over 10% cover of the woody shrubs Purshia tridentata or Amelanchiar utahensis. Plots within each pair were established 6 m away from one another and the corners of the plots were marked with steel bars. To get baseline pre-treatment data, we then censused the plant communities within the new plots with the pantograph technique as in the original plots.

In the fall of 2011, we randomly assigned one plot in each plot pair to recieve an experimental drought treatment and the other plot within each pair to recieve an experimental irrigation treatment.  We imposed an experimental drought on the drought plots with rain-out shelters consisting of transparent plastic gutters suspended over the plot.  The plastic gutters intercept rainfall falling on the plot and direct that water to collection tank. We used battery powered pumps to pump collected water out of the tanks and onto the paired irrigation treatments.  Pumps were switched on and off automatically with float switches located in the tanks.  The float switches were activated when tanks contained approximately 10 l of water. Shelters covered a 5 m^2 area around the 1 m square drought plot and we pumped the collected water through sprinklers onto a equal sized area surrounding the focal irrigation plot.  The shelter gutters were spaced to cover exactly 50% of the focal drought plot, thereby imposing a 50% reduction in rainfall from ambient conditions.  Likewise irrigation plots received 150% the ambient rainfall of control plots.

Pumps were disconnected when temperatures were regularly below freezing, from approximately October to May of each year.

### Soil water model

### Estimating climate effects on vital rates

### Generating predictions

Results
-------

Discussion
-----------

### Lessons for making ecological predictions

Conclusion
----------

Acknowledgments
---------------
This work was funded by a National Science Foundation award DEB-1054040 to PB. AK was supported by an NSF Graduate Research Fellowship and the Utah State University Graduate Studies Office. We wish to thank Kevin, Joe, Marina, Britta, Alice Willson, Carlee Coleman, Bret Taylor and the US Sheep Experiment Station.

\pagebreak{}

Tables
------

\begin{table}[ht]
\centering
\caption{Summary of datasets used in the analysis}
\label{tab:T1}
\begin{tabular}{llllll}
  \toprule
  \input{ ../data/temp_data/summary_table.txt}
\end{tabular}
\end{table}

\pagebreak{}

\begin{table}[ht]
\centering
\caption{Coefficients from a linear mixed effects model fit to the annual sagebrush data (n = `r ds_stats$n_obs`). Production indicates the difference between data estimating sagebrush cover (\protect\%) and production (g per m\protect\textsuperscript{2}). Estimates for a and b correspond to the variables in the Gompertz population model described in \ref{eq1}. Climate variables are defined in the main text.} 
\label{tab:T2}
\begin{tabular}{rrrr}
  \toprule
  \input{ ../data/temp_data/model_summary.txt}
\end{tabular}
\end{table}

\pagebreak{}

Figures
-------

\begin{figure}[!ht]
  \centering
      \includegraphics[width=6in]{../figs/hypothesis.png}
  \caption{Hypothetical effects of interannual climate variation on species' populations across its range. The bottom axis corresponds to average climate at each site. Figures in the top row show change in cover on the y-axis and annual temperature anomalies at each site on the x-axis. Years with above average temperatures are expected to increase population size at the coldest site (A), while they decrease population size at the warmest site (C). The middle panel plots the sensitivities of each site against the average climate gradient.  We interpret a negative slope between sensitivities to annual climate variation and average climate as support for the hypothesis that temperature controls the species' distribution.}
  \label{fig:hypothesis}
\end{figure}

\pagebreak{}

\begin{figure}[!ht]
  \centering
      \includegraphics[width=6in]{../figs/map_locations2.png}
  \caption{Map of sites with multi-year sagebrush cover data in the western USA. Point size corresponds to number of observations at each site. Gray areas show the distribution of sagebrush based on the USGS SAGEMAP dataset (http://sagemap.wr.usgs.gov/). Inset shows an example of multi-year sagebrush cover data from three monitoring plots at Camp Williams, Utah.}
  \label{fig:map}
\end{figure}

\pagebreak{}

\begin{figure}[!ht]
  \centering
      \includegraphics[width=6in]{../figs/effects_plot.png}
  \caption{Sensitivity of sagebrush abundance to annual climate covariates plotted against average site climate. Sensitivity is defined as the log change in sagebrush abundance produced by a 10 cm increase in precipitation (A,C) or a 1\textdegree C increase in temperature (B,D). Effects below zero indicate where above average temperatures or precipitation would decrease population size, while effects above zero indicate where above average annual temperature or precipitation would increase population size. Gray areas show 95\protect\% confidence intervals.}
  \label{fig:effects}
\end{figure}

\pagebreak{}

\begin{figure}[!ht]
  \centering
      \includegraphics[width=6in]{../figs/predict_both.png}
  \caption{Predicted effects of a short-term increase in precipitation (A) or temperature (B) on sagebrush abundance at each of the monitoring sites.  Panel A shows the predicted proportional change after four years with 10\protect\% above average precipitation plotted against site mean annual precipitation. Panel B shows the predicted proportional change after four years of 2\protect\textdegree C above average temperatures plotted against site mean annual temperature. Gray bars show bootstrapped 95\protect\% confidence intervals. Confidence intervals do not take into account uncertainty in random effects.}
  \label{fig:predictions}
\end{figure}

\pagebreak{}

References
----------

### 2016-09-05 removals/Wdistrib/exploreSurvivalWs.R
Updated scripts to work with addition of 2016 data

@@ -140,15 +140,16 @@ for(iSpp in 1:4){
     }    
     data$firstRing=firstRing
     
-    plot(data$radius, firstRing,main=paste0(doSpp,":",nbSpp),xlab="Genet radius (cm)",ylab="Mid-radius of 1st occupied ring (cm)");
+    plot(data$radius, firstRing,main=paste0(doSpp,":",nbSpp),ylim=c(0,20),xlab="Genet radius (cm)",ylab="Mid-radius of 1st occupied ring (cm)");
     abline(0,1,col="blue"); 
     fit = rq(firstRing~radius-1,data=data,tau=0.05); 
     abline(fit); fits[[jSpp+4*(iSpp-1)]]=fit; 
     
-    # try nonparametric fit
-    firstRing[is.na(firstRing)] <- 150
-    fit2 = rqss(firstRing~qss(radius-1,lambda=1),data=data,tau=0.05); 
-    fits.rqss[[jSpp+4*(iSpp-1)]] <- fit2
+#     # try nonparametric fit
+#     data$firstRing[is.na(data$firstRing)] <- 150
+#     fit2 = rqss(firstRing~qss(radius-1,lambda=0.01),data=data,tau=0.05); 
+#     fits.rqss[[jSpp+4*(iSpp-1)]] <- fit2
+#     #plot with predict
     
     fac = round(coef(fit)[1], digits=2); 
     legend("topright",legend=as.character(fac),bty="n",cex=1.2,inset=0.1); 

### 2016-09-05 removals/ibm/ibm_removal_1step.r

@@ -8,7 +8,7 @@ outfile="ibm/simulations1step/ObsPred_1step.csv"
 Nspp=length(sppList)
 
 curDir <- getwd()
-Nyrs <- 30
+Nyrs <- 31
 # set up survival parameters and function
 source("survival/import2ibm_1step.r")
 # set up growth parameters and function
@@ -82,9 +82,9 @@ cov.obs <- reshape(cov.obs,idvar=c("quad","year"),direction="wide",timevar="spec
 cov.obs[is.na(cov.obs)] <- 0
 cov.obs <- cov.obs[,c(1,2,6,3,4,5)] # reorder columns
 
-# get 2015 quadrat cover totals (these are not in the survival data file)
+# get 2016 quadrat cover totals (these are not in the survival data file)
 tmp <- read.csv("QuadYearCover.csv")
-tmp <- subset(tmp, year==2015)
+tmp <- subset(tmp, year==2016)
 # set removed spp to zero
 tmp$cover[tmp$species=="Artemisia tripartita" & tmp$Treatment=="No_shrub"] <- 0
 tmp$cover[tmp$species!="Artemisia tripartita" & tmp$Treatment=="No_grass"] <- 0
@@ -136,7 +136,7 @@ cov.pred[is.na(cov.pred)] <- 0
 out.recruit <- out.recruit.trt <- data.frame(quad=cov.obs$quad,year=as.numeric(cov.obs$year),matrix(NA,dim(cov.obs)[1],4))
 names(out.recruit)[3:6] <- sppList
 names(out.recruit.trt)[3:6] <- sppList
-out.recruit <- subset(out.recruit,year<2015); out.recruit.trt <- subset(out.recruit.trt,year<2015)
+out.recruit <- subset(out.recruit,year<2016); out.recruit.trt <- subset(out.recruit.trt,year<2016)
 for(k in 1:dim(out.recruit)[1]){
     
     totArea <- cov.obs[which(cov.obs$year==out.recruit$year[k] & cov.obs$quad==out.recruit$quad[k]),3:6]

### 2016-09-12 precip/doc/precip_experiment_ms.Rmd
moving all my files to a new directory within the main project directory. This will facilitate organized merging with Peters directory

--- 2016-09-01 doc/precip_experiment_ms.Rmd
@@
 
-\textsc{\small{Andrew R. Kleinhesselink\footnote{Corresponding author: arklein@aggiemail.usu.edu}\textsuperscript{1}, Peter B. Adler\textsuperscript{1} and Andrew Tredennick\textsuperscript{1}}}
+\textsc{\small{Andrew R.
+Kleinhesselink\footnote{Corresponding author: arklein@aggiemail.usu.edu}\textsuperscript{1}, Peter B.
+Adler\textsuperscript{1}, Andrew Tredennick\textsuperscript{1}, Caitlin Andrews{2}\textsuperscript{2} and John Bradford\textsuperscript{2} }
 
@@
 #### Multi-year plant community data  
-Long-term data on individual abundance and size of plants in a sagebrush steppe community have been collecting since the 1930s at the US Sheep Experiment Station in Dubois Idaho.
-The plots were recensused beginning in 2007 using methods identical to the original studies.
-These data allow for the fitting of statistical models that estimate the effects of climate and competitive effects on individual plants in the plots.
-In theory these statistical models can be used to project the effects of both a change in the competitive environment at this site--or a change in the climate.
+Long-term data on individual abundance and size of plants come from the US Sheep Experiment Station (USSES) in Dubois, Idaho, USA (44\degree 12' N, 112\degree 1' W, 1500 m A.S.L.).
+This recieves 307 mm of precipitation annually and has a mean annual temperature of 6.2\degree C.
+The vegetation is dominated by three perennial grasses, Pseudoroegneria spicata (Syn.
+Pascopyrum spicata), Poa secunda (Syn.
+Poa sandbergii), and Hesperostipa comata (Syn.
+Stipa comata), and one small woody shrub, Artemisia tripartita [Asteraceae].
+
+Our historical data come from permanent plot monitoring at the USSES conducted annually from 1923 to 1957.
+The permanent plots are 1 x 1 m and the corners were marked with steel bars so that they could be precisely relocated year after year.
+The size and position of each individual plant within the plots was recorded on paper maps using a mechanical tracing device called a pantograph.
+This method accurately reproduces the exact spatial configuration of plants within a plot in a given year to less than 1 cm accuracy.
+
+Beginning in 2007, 18 of the original permanent plots were relocated and re-measured annually using the same methods as the original investigators.
+The permanent plots measured in this contemporary period were located within three areas that had been ungrazed since the 1920's.
 
@@
 
-### Estimating climate effects on vital rates
+### Estimating effects of climate and competition on vital rates
+
+We fit our predictive models using the multi-year records of individual plant growth, survival and recruitment in the (20) original permanent plots from 19XX to 19XX.
+Each estimated the effects of temperature, precipitation and soil moisture on
 
@@
 -------
+
+### Changes in climate and community composition
 

### 2016-09-12 R/climate/export_climate_station_data_for_SOILWAT.R

@@ -1,56 +1,0 @@
-# format for Caitlin Andrews and John Bradford:
-
-rm(list = ls () )
-library( dplyr ) 
-library( tidyr )
-
-station_dat <- read.csv('data/USSES_climate.csv')
-station_dat$date <- as.POSIXct( strptime( station_dat$DATE, format = '%Y%m%d', tz = 'MST')    )
-
-station_dat <- station_dat %>% select( date, STATION, STATION_NAME, LATITUDE, LONGITUDE, ELEVATION, PRCP, SNWD, SNOW, TMAX, TMIN )  
-
-station_dat <- station_dat %>% mutate( LONGITUDE = LONGITUDE[which.max(date)], LATITUDE = LATITUDE[ which.max(date )], ELEVATION = ELEVATION[ which.max(date )] )
-
-head(station_dat)
-
-station_dat$year <- strftime(station_dat$date, '%Y')
-station_dat$DOY <- as.numeric( strftime( station_dat$date, '%j'))
-station_dat$PPT <- station_dat$PRCP/10
-
-station_dat %>% group_by(year)  %>% filter( year > 1980 , year < 2011, PPT >= 0 ) %>% summarise(AP = sum(PPT, na.rm = T)) %>% summarise( MAP = mean(AP))
-
-df <- expand.grid( DOY = 1:365 , year = c(min(station_dat$year):max(station_dat$year)))
-
-station_dat <- merge( df,station_dat, by = c('year', 'DOY'), all.x = T, all.y = T)
-
-year_list <- split( station_dat[ , c('DOY', 'TMAX', 'TMIN', 'PPT') ], station_dat$year)
-
-write_with_header <- function(x, file, header, f = write.table, ...){
-
-  datafile <- file(file, open = 'wt')
-
-  on.exit(close(datafile))
-
-  if(!missing(header)) writeLines(header,con=datafile)
-
-  f(x, datafile,...)
-}
-
-make_header <- function( prefix, df, station, year) { 
-  
-  paste0( '#', prefix, station, ' year = ', year, '\n#', 'DOY', ' ', 'Tmax(C)', ' ', 'Tmin(C)', ' ', 'PPT(cm)')  
-  
-}
-
-
-for ( i in 1:length( year_list) ) { 
-  
-  temp_df <- year_list[[i]]
-  temp_year <- names(year_list)[[i]]
-  
-  temp_fname <- file.path( 'data/temp_data/weather_files', paste0( 'weath.', temp_year) )
[…]

### 2016-09-12 R/climate/plot_seasonal_and_daily_temperatures.R

@@ -1,36 +1,0 @@
-################################################################################################## 
-#
-# Plot daily temperatures for each plot each month.  Good way to check that times are correct. 
-#
-##################################################################################################
-
-rm(list = ls())
-library( dplyr)
-library(ggplot2)
-
-temps <- readRDS('data/temp_data/decagon_data_with_station_data.RDS')
-
-df  <- temps %>% 
-  filter( bad_values == "0", depth == 'air temperature', !is.na(v) , stat == 'raw') 
-
-test <- df %>% ungroup() %>% filter( f == f[1] )
-
-daily_t <- ggplot(test, aes( x  = hour , y = v , group = simple_date ) ) +  
-  geom_line(stat = 'smooth', method = 'loess', alpha = 0.1) + 
-  geom_vline(aes(xintercept = 16)) + 
-  ylim ( -50, 60 ) + 
-  facet_wrap( ~ month )
-
-plots <- df %>% ungroup() %>% 
-  group_by (plot , year ) %>% 
-  do ( gg = daily_t %+% . + ggtitle ( paste( 'plot = ', unique( .$plot) , 'year = ', unique( .$year) ))) 
-
-
-# print -----------------------------------------------------------------------------------------------
-
-pdf ( 'figures/check_daily_temperatures.pdf', width = 7 , height = 10)
-print( plots$gg ) 
-dev.off()
-
-
-

### 2016-09-12 R/climate/prepare_climate_covariates.R

@@ -1,96 +1,0 @@
-#####################################################################################
-#
-# Make climate variables from quarterly climate 
-#
-#####################################################################################
-
-rm(list = ls() )
-
-library(dplyr)
-library(tidyr)
-library(ggplot2)
-library(zoo)
-
-# -- specify climate variables -------------------------------------------------------# 
-
-clim_vars <- c('T.sp.1', 'T.sp.2', 'P.w.sp.1', 'P.w.sp.2', 'T.su.1', 'T.su.2', 'P.a.0', 'P.a.1')
-clim_vars <- sort(clim_vars)
-
-# ------- load files ------------------------------------------------------------------ 
-
-quarterly_clim <- readRDS('data/temp_data/quarterly_climate.RDS')
-
-# ------ calculate quarterly lags -----------------------------------------------------# 
-#
-#   Variable names follow these conventions: 
-#   
-#     First letter gives variable type: 
-#       "P" is cumulative precipitation
-#       "T" is average mean monthly temperature
-#
-#     Letters after the first period give the season aggregation window:
-#
-#       w  = winter (Q1) 
-#       sp = spring (Q2)
-#       su = summer (Q3)
-#       f  = fall   (Q4)
-#       a  = annual (Q1-4)
-#
-#     e.g. "P.sp" is the cumulative precipitation of the spring quarter and "P.w.sp" is
-#     the cumulative precipitation of the winter and spring. 
-#   
-#     Number after the second period indicates the year of the transition, For example, 
-#     "P.sp.1" gives the cumulative precipitation of the spring preceding the first year of the 
-#     transition, whereas "T.f.w.2" gives the average temperature of the fall and winter 
-#     preceding the second year of the transition. "0" refers to year before first year, 
-#     i.e. "lag effect" (sensu Adler). 
-#
-# -------------------------------------------------------------------------------------# 
-q_precip <- 
-  quarterly_clim %>% 
-  filter( var == 'TPCP_ttl') %>%
-  group_by(Treatment) %>% 
-  arrange(Treatment, year, quarter) %>%
-  mutate(P.sp.2 = val, 
[…]

### 2016-09-12 R/get_all_demographic_data.R

@@ -1,16 +1,0 @@
-####################################################################################
-#
-# run scripts to fetch all demographic data 
-# 
-####################################################################################
-
-get_growth_files <- dir('R', pattern = 'get.*growth\\.R', recursive = TRUE, full.names = TRUE)
-
-get_survival_files <- dir('R', pattern = 'get.*survival\\.R', recursive = TRUE, full.names = TRUE)
-
-get_recruitment_files <- dir('R', pattern = 'get.*recruitment\\.R', recursive = TRUE, full.names = TRUE)
-
-lapply( c(get_growth_files, get_survival_files, get_recruitment_files), source) 
-
-#TODO write functions to export to HPC server 
-

### 2016-09-12 R/growth/fetchGrowthData.R

@@ -1,50 +1,0 @@
-
-fetchGdat <- function(doSpp,speciesList,datadir,distWts){
-
-  growDfile=paste(datadir,"/speciesData/",doSpp,"/growDnoNA.csv",sep="")
-  growD=read.csv(file=growDfile)
-  D1=growD[growD$allEdge==0,];
-  D1$year <- D1$year
-  D1$logarea.t0=log(D1$area.t0)
-  D1$logarea.t1=log(D1$area.t1)
-  D1$quad=as.character(D1$quad)
-  
-  # import neighbor data
-  ringD <- read.csv(paste(datadir,"/speciesData/",doSpp,"/",doSpp,"_nbhood_rings.csv",sep=""))
-  tmpD <- read.csv(paste(datadir,"/speciesData/",doSpp,"/",doSpp,"_nbhood_rings_allothers.csv",sep=""))
-  ringD<-merge(ringD,tmpD)
-  ringD$year<-ringD$year
-  
-  # merge D with ringD (D contains fewer rows)
-  D1<-merge(D1,ringD,by.x=c("quad","year","trackID"),by.y=c("quad","year","genetID"))
-  D1=D1[order(D1$X),]
-  rm(ringD,growD)
-  row.names(D1) <- NULL  
-  
-  # calculate W's (MAKE SURE NOT TO REORDER D!)
-  W <- matrix(NA,NROW(D1),length(speciesList))
-  colnames(W) <- paste("W.",speciesList,sep="")
-
-  # do big 4
-  for(iSpp in 1:4){
-    neighborCols=which(substr(names(D1),1,4)==speciesList[iSpp]) # pull out annulus data for the focal species 
-    dist_wts <- distWts[,paste0(speciesList[iSpp])]
-    C <- data.matrix(D1[,neighborCols]) #matrix of conspecific areas in the annuli 
-    W[,iSpp] <- C%*%dist_wts 
-  }
-  
-  # do allcov and allpts
-  for(iSpp in 5:6){
-    neighborCols=which(substr(names(D1),1,6)==speciesList[iSpp]) # pull out annulus data for the focal species 
-    dist_wts <- distWts[,paste0(speciesList[iSpp])]
-    C <- data.matrix(D1[,neighborCols]) #matrix of conspecific areas in the annuli 
-    W[,iSpp] <- C%*%dist_wts 
-  }
-  
-  #format
[…]

### 2016-09-12 R/growth/get_HECO_growth.R

@@ -1,77 +1,0 @@
-# PBA March 2016
-# modified by ARK July 2016
-# call from removal_analysis_wrapper.r
-
-#########################################
-#  1. Import data and calculate W's
-#########################################
-
-root <- '~'
-doSpp <- "HECO"
-sppList <- c("ARTR","HECO","POSE","PSSP","allcov","allpts")
-dataDir1 <- paste(root,"/driversdata/data/idaho",sep="")
-dataDir2 <- paste(root,"/driversdata/data/idaho_modern",sep="")
-nonCompLength.s=5 #Number of columns in SppData that are not measures of competitors 
-
-# set up distance weights------------------------------------------------
-dists <- read.csv(paste(dataDir2,"/speciesData/IdahoModDistanceWeights_noExptl.csv",sep=""));
-dists$allcov <- rowMeans(dists[,1:4])  # for "other" polygons use average of big 4
-dists$allpts <- dists$POSE  # set forb dist wts = smallest grass (POSE)
-
-# import old data--------------------------------------------------------
-
-source("R/growth/fetchGrowthData.R")
-
-D1 <- fetchGdat(doSpp=doSpp,speciesList=sppList,datadir=dataDir1,distWts=dists)
-D1$Treatment <- "Control"
-D1$Period <- "Historical"
-
-# import modern data--------------------------------------------------------
-
-D2 <- fetchGdat(doSpp=doSpp,speciesList=sppList,datadir=dataDir2,distWts=dists)
-D2$Period <- "Modern"
-
-# merge in treatment data
-tmp <- read.csv(paste(dataDir2,"/quad_info.csv",sep=""))
-tmp <- tmp[,c("quad","Treatment")]
-D2 <- merge(D2,tmp, all.x=T)
-
-# account for removal in baseline years
-if(doSpp!="ARTR"){
-  ii <- which(D2$year>=2011 & D2$Treatment=="No_shrub")
-  D2$W.ARTR[ii] <- 0
-}else{
-  ii <- which(D2$year>=2011 & D2$Treatment=="No_grass")
-   D2$W.HECO[ii] <- 0 ; D2$W.POSE[ii] <- 0 ; D2$W.PSSP[ii] <- 0
-}
-
-# combine old and modern
-allD <- rbind(D1,D2)
-rm(D1,D2,tmp)
-
-# clean up dataset ----------------------------------------------
-allD$year[allD$year<2000] <- allD$year[allD$year<2000] + 1900
-
-if(doSpp=="ARTR"){
[…]

### 2016-09-12 R/observed_soil_moisture/correct_decagon_readings.R

@@ -1,134 +1,0 @@
-rm(list = ls()) 
-
-library( ggplot2 ) 
-library(tidyr)
-library(dplyr)
-library(zoo)
-
-df <- readRDS(file = 'data/temp_data/decagon_data_corrected_dates.RDS')
-
-df <- df %>% ungroup() 
-
-# Filter out bad readings ------------------------------------------------------------------------
-
-get_run_lengths <- function( x ) { 
-  unlist ( lapply( rle(x)$lengths, function(x) rep(x, x) ) ) 
-}
-
-# My complicated steps for data classification:  ---------------------------------------------------------------
-#   calculate rolling mean at 5 values 
-#   calculate standard deviations from rolling mean 
-#   calculate rolling mean of standard deviations 
-#   classify values as high variability or low variabilty based on rolling mean of standard deviations 
-#   classify sections as bad windows when 40/100 values are high variability 
-#   calculate run lengths of good windows
-#   only keep good windows of run lengths > 50 values in a row 
-
-low_VWC_cutoff <- 0.005
-high_VWC_cutoff <- 0.01
-C_cutoff <- 400
-
-corrected <- 
-  df %>% 
-  group_by(plot, position, measure ) %>% 
-  arrange( new_date) %>% 
-  mutate( Tdiff = as.numeric( new_date - lag(new_date), tz = 'MST' ) ) %>% 
-  mutate( frame_length = n() ) %>% 
-  filter( frame_length > 100 ) %>% 
-  mutate( rllm = rollapply( value, 100, mean, na.rm = TRUE, fill = NA, align = 'center')) %>% 
-  mutate( dff = (value - rllm)^2) %>% 
-  mutate( rllsd = rollapply( dff, 50, mean, na.rm = TRUE, fill = NA, align = 'center')) %>%
-  mutate( highv = ifelse( measure == 'VWC' & !is.na(rllsd) & rllm < 0.10 & rllsd > low_VWC_cutoff, 1, 0)) %>% 
-  mutate( highv = ifelse( measure == 'VWC' & !is.na(rllsd) & rllm >= 0.10 & rllsd > high_VWC_cutoff, 1, highv)) %>% 
-  mutate( highv = ifelse( measure == 'C' & !is.na(rllsd) & rllsd > C_cutoff, 1, highv )) %>% 
-  mutate( out_range = ifelse( measure == 'C' & !is.na(value) & (value < -30 | value > 65 ) , 1, 0 )) %>% 
[…]

### 2016-09-12 R/observed_soil_moisture/format_plot_15_2014_2.R

@@ -1,17 +1,0 @@
-# special formatting script for plot 15 
-
-df <- read.table(file = 'data/soil_moist_data/2014_2/15_15Sep14-1802.txt', sep = '\t', header = F)
-
-cn <- df[ 1, ]
-
-df <- df [ -1 , ]
-
-df$date <- strptime( df$V1, '%m/%d/%y %I:%M %p', tz = 'MST' ) 
-
-df <- df[ order(df$date), ]
-
-df <- df [ , -7 ]
-
-df <- rbind( cn, df )
-
-write.table( df, 'data/soil_moist_data/2014_2/15_reorderd.txt', sep = '\t', row.names = F, col.names = F, quote = F)

### 2016-09-12 R/observed_soil_moisture/plot_raw_decagon_data.R

@@ -1,29 +1,0 @@
-rm(list = ls()) 
-
-library( ggplot2 ) 
-library(tidyr)
-library(dplyr)
-
-df <- readRDS(file = 'data/temp_data/decagon_data.RDS')
-
-plot_ts <- function( x ) { 
-
-  ggplot ( data = x, aes( x = date, y = value , group = port, color = Treatment )) + 
-    geom_line()   + 
-    facet_wrap( ~ depth, ncol = 1 ) + 
-    ggtitle( paste( 'Plot Group:', unique(x$PrecipGroup)) )
-  
-}
-
-p <- df %>% group_by(PrecipGroup, measure) %>% do( p = plot_ts(x = . )  )
-
-
-pdf('figures/raw_decagon_time_series.pdf', width = 10, height = 7)
-
-for ( i in p$p ) { 
-  
-  print( i )   
-  
-}
-dev.off() 
-

### 2016-09-12 analysis/growth/get_initial_values.R

@@ -1,97 +1,0 @@
-##################################################################################################################
-#
-# Get init vals for stan models
-#
-##################################################################################################################
-
-#---model descriptions --------------------------------------------------------------------------------------------
-#  
-#   m1: null model:                       
-#   m2: climate model:                    w/ climate 
-#   m3: single species model:                         w/ intra-specific competition 
-#   m4: single species climate model:     w/ climate, w/ intra-specific competition
-#   m5: full model:                       w/ climate, w/ intra-specific competition, w/ interspecific competition 
-#
-# ----------------------------------------------------------------------------------------------------------------- 
-
-rm(list = ls())
-
-library(lme4)
-
-make_df <- function( x ) { 
-  N <- x$N
-  lens <- lapply( x, length )
-  nrs <- lapply(  x, nrow ) 
-  nrs [ sapply( nrs, is.null )  ]  <- 0
-  
-  data.frame( x [ which(lens == N | nrs == N) ]  )
-} 
-
-set_init_vals_list <-  function( model, C_names, W_names ) {  
-  
-  init_vals <- as.list( fixef(model)[1:2] )
-  
-  init_vals <- c(init_vals, as.numeric(data.frame( VarCorr(model) )$sdcor[ c(1,2,4)]))
-  
-  names( init_vals )[1:5] <- c('a_mu', 'b1_mu', 'sig_a', 'sig_b1', 'sig_G')
-  
-  b2 <- fixef(model)[C_names]
-  w <- fixef(model)[W_names]
-
-  if( length(b2[!is.na(b2)]) > 0 )  { 
-    init_vals <- c(init_vals, list( b2 = as.numeric(b2[!is.na(b2)])))
-  } 
-  if( length(w[!is.na(w)]) > 0 )  { 
-    init_vals <- c(init_vals, list( w = as.numeric(w[!is.na(w)])))  
-  }
-  
-  # growth variance is size specific in these models 
-  # Choose initial values similar to those from Tredennick's paper 
-  init_vals$tau <- 1           
-  init_vals$tauSize <- 0
-  
-  return( init_vals )
-
-}
-
-
-get_init_vals_growth_models <- function( spp, datalist, ... ) {
-
-  df <- make_df(x = datalist) 
-  
-  C_names <- names(df)[ grep('^C', names(df))] # climate effects 
[…]

### 2016-09-12 analysis/growth/growth_fxns.R

@@ -1,134 +1,0 @@
-##
-##  R Functions for Parallelized Stan Fits for survival logistic model (for IPM)
-##
-##
-##  Author:       Andrew Tredennick
-##  Email:        atredenn@gmail.com
-##  Date created: 12-6-2015
-##
-
-####
-####  Regularization Without Cross Validation (collects betas)
-####
-reg.fcn <- function(i){
-  library(rstan)
-  library(ggmcmc)
-  library(plyr)
-  clim_covs <- growD[,clim_vars]
-  groups <- as.numeric(growD$Group)
-  G <- length(unique(growD$Group))
-  gid <- as.numeric(growD$Group)
-  nyrs <- length(unique(growD$year))
-  W <- cbind(growD$W, growD$W*log(growD$area.t0))
-  yid <- as.numeric(as.factor(growD$year))
-  sd_now <- sd_vec[i]
-  datalist <- list(N=nrow(growD), Yrs=nyrs, yid=yid,
-                   Covs=ncol(clim_covs), Y=log(growD$area.t1), 
-                   X=log(growD$area.t0), C=clim_covs, W=W, G=G, gid=gid,
-                   tau_beta=sd_now, gid_out=gid_out)
-  
-  pars <- c("b2")
-  inits <- list()
-  inits[[1]] <- list(a_mu=0, a=rep(0,nyrs), b1_mu=0.01,
-                     b1=rep(0.01,nyrs), gint=rep(0,G), 
-                     w=c(0,0), sig_b1=0.5, sig_a=0.5, 
-                     tau=0.5, tauSize=0.5, sig_G=0.5, 
-                     b2=rep(0,ncol(clim_covs)))
-  inits[[2]] <- list(a_mu=0.5, a=rep(0.5,nyrs), b1_mu=0.1,
-                     b1=rep(0.1,nyrs), gint=rep(0.4,G), 
-                     w=c(0.1,0.1), sig_b1=0.2, sig_a=0.2, 
-                     tau=0.2, tauSize=0.2, sig_G=0.2, 
-                     b2=rep(0.5,ncol(clim_covs)))
-  inits[[3]] <- list(a_mu=1, a=rep(1,nyrs), b1_mu=0.2,
-                     b1=rep(0.2,nyrs), gint=rep(0.2,G), 
[…]

### 2016-09-12 analysis/run_stan_models.R

@@ -1,147 +1,0 @@
-##########################################################################
-#
-##  R script for growth as run on Utah State University's
-##  High Performance Computing system.
-## 
-##  Adapted from Andrew Tredennick's script 
-##
-##  Script takes command line argument 'species' and 'vital_rate'.  Runs 
-##  m separate models for each species and vital rate. Models that include
-##  climate effects are run for each level of n prior standard deviation 
-##  for regularization of the climate effects parameters. Script will launch 
-##  n*c models:  where 'c' is the the number of models with climate effects 
-##  and 'n' the number of prior standard deviations. 
-##
-##  Saved output:
-##    Save full model output from each model  
-##
-##  Original Author:      Andrew Tredennick
-##  Modified By:          Andrew Kleinhesselink 
-##  Email:                arklein@aggiemail.usu.edu
-##  Date created:         12-6-2015
-##  Date modified:        09-08-2016
-# 
-##########################################################################
-
-rm(list = ls())
-
-library(rstan)
-library(matrixStats)
-library(loo )
-
-tweak_inits <- function(inits){ 
-  
-  inits + runif(length(inits), -0.001, 0.001)
-  
-}
-
-# -- read in growth data lists and initial values ---------------------------------------------------------# 
-
-# ---Set SD Prior from Command Line Arguments -------------------------------------------------------------#
-
-args <- commandArgs(trailingOnly = TRUE)
-
-do_spp <- args[1]
-do_vr  <- args[2]
-do_sd  <- args[3]
-nchains <- args[4]
-
-# Set up regularization grid ------------------------------------------------------------------------------#
-
-n.beta <- 24
-sd_vec <- seq(0.1,1.5, length.out = n.beta)
-
-# read in data and models ---------------------------------------------------------------------------------#
-do_vr <- 'growth'
-do_spp <- 'ARTR'
-data_path <- 'data/temp_data'
-model_path <- 'analysis'
-output_path <- 'analysis/stan_fits'
-
-data_file <- dir(data_path, pattern = paste0(do_vr, '_data_lists_for_stan.RDS'), full.names = TRUE )
-init_file <- dir(data_path, pattern = paste0(do_vr, '_init_vals.RDS'), full.names = TRUE)
-
[…]

### 2016-09-12 analysis/treatment_trends_precip.r

@@ -1,222 +1,0 @@
-
-# call from removal_analysis_wrapper.r
-
-sppList=c("Artemisia tripartita","Hesperostipa comata","Poa secunda","Pseudoroegneria spicata")
-dataDir <- file.path(root,"/driversdata/data/idaho_modern/")
-
-# import data and calculate treatment trends ######################################
-
-
-covD<-read.csv(paste(dataDir,"allrecords_cover.csv",sep=""))
-trts<-read.csv(paste(dataDir,"quad_info.csv",sep=""))
-
-# use this to make sure we don't miss zeros
-allquadyrs<-unique(covD[,c("quad","year")],MARGIN=2)
-tmp<-expand.grid(species=sppList,year=sort(unique(covD$year)))
-allquadyrs<-merge(allquadyrs,tmp,all=T)
-allquadyrs<-merge(allquadyrs,trts[,c("Group","quad","Treatment")])
-
-# only do precip treatments
-keep <- which(is.element(allquadyrs$Treatment,c("Control","Drought","Irrigation")))
-allquadyrs <- allquadyrs[keep,]
-
-# aggregate to quadrat level
-keep<-which(is.element(covD$species,sppList))
-sppD<-covD[keep,]
-sppD<-merge(sppD,allquadyrs,all.y=T)
-sppD$area[is.na(sppD$area)]<-0
-sppD$area<-sppD$area*100
-sppD.q<-aggregate(sppD$area,by=list(species=sppD$species,Group=sppD$Group,Treatment=sppD$Treatment,
-                  quad=sppD$quad,year=sppD$year),FUN=sum)
-names(sppD.q)[NCOL(sppD.q)]<-"cover"
-write.csv(sppD.q,"data/temp_data/QuadYearCover.csv",row.names = FALSE)  # save these 
-
-# focus on all control plots or just those in the big exclosure?
-sppD.q <- subset(sppD.q,Group=="E1")
-
-#calculate treatment means by year
-spp.mean <- aggregate(sppD.q$cover,by=list(species=sppD.q$species,Treatment=sppD.q$Treatment,
-                  year=sppD.q$year),FUN=mean)
-names(spp.mean)[NCOL(spp.mean)] <- "cover"
[…]

### 2016-09-12 precip/R/climate/plot_seasonal_and_daily_temperatures.R

@@ -1,0 +1,36 @@
+################################################################################################## 
+#
+# Plot daily temperatures for each plot each month.  Good way to check that times are correct. 
+#
+##################################################################################################
+
+rm(list = ls())
+library( dplyr)
+library(ggplot2)
+
+temps <- readRDS('data/temp_data/decagon_data_with_station_data.RDS')
+
+df  <- temps %>% 
+  filter( bad_values == "0", depth == 'air temperature', !is.na(v) , stat == 'raw') 
+
+test <- df %>% ungroup() %>% filter( f == f[1] )
+
+daily_t <- ggplot(test, aes( x  = hour , y = v , group = simple_date ) ) +  
+  geom_line(stat = 'smooth', method = 'loess', alpha = 0.1) + 
+  geom_vline(aes(xintercept = 16)) + 
+  ylim ( -50, 60 ) + 
+  facet_wrap( ~ month )
+
+plots <- df %>% ungroup() %>% 
+  group_by (plot , year ) %>% 
+  do ( gg = daily_t %+% . + ggtitle ( paste( 'plot = ', unique( .$plot) , 'year = ', unique( .$year) ))) 
+
+
+# print -----------------------------------------------------------------------------------------------
+
+pdf ( 'figures/check_daily_temperatures.pdf', width = 7 , height = 10)
+print( plots$gg ) 
+dev.off()
+
+
+

### 2016-09-12 precip/R/growth/get_ARTR_growth.R

@@ -1,0 +1,83 @@
+# PBA March 2016
+# modified by ARK July 2016
+# call from precip_analysis_wrapper.r
+# simple test for differences in growth rate between precip treatments in the contemporary period 
+
+#rm(list = ls())
+
+#########################################
+#  1. Import data, merge treatment effects and save out 
+#########################################
+
+doSpp <- "ARTR"
+sppList <- c("ARTR","HECO","POSE","PSSP","allcov","allpts")
+dataDir1 <- paste("~/driversdata/data/idaho",sep="")
+dataDir2 <- paste("~/driversdata/data/idaho_modern",sep="")
+nonCompLength.s=5 #Number of columns in SppData that are not measures of competitors 
+
+# set up distance weights------------------------------------------------
+#dists <- read.csv('~/driversdata/data/idaho/speciesData/IdahoDistanceWeights.csv')
+dists <- read.csv('~/driversdata/data/idaho_modern/speciesData/IdahoModDistanceWeights_noExptl.csv')
+dists$allcov <- rowMeans(dists[,1:4])  # for "other" polygons use average of big 4
+dists$allpts <- dists$POSE  # set forb dist wts = smallest grass (POSE)
+
+# import old data--------------------------------------------------------
+
+source("R/growth/fetchGrowthData.R")
+
+D1 <- fetchGdat(doSpp=doSpp,speciesList=sppList,datadir=dataDir1,distWts=dists)
+D1$Treatment <- "Control"
+D1$Period <- "Historical"
+
+# import modern data--------------------------------------------------------
+D2 <- fetchGdat(doSpp=doSpp,speciesList=sppList,datadir=dataDir2,distWts=dists)
+D2$Period <- "Modern"
+
+# merge in treatment data
+tmp <- read.csv(paste(dataDir2,"/quad_info.csv",sep=""))
+tmp <- tmp[,c("quad","Treatment")]
+D2 <- merge(D2,tmp, all.x=T)
+
+# account for removal in baseline years
+if(doSpp!="ARTR"){
+  ii <- which(D2$year>=2011 & D2$Treatment=="No_shrub")
+  D2$W.ARTR[ii] <- 0
+}else{
+  ii <- which(D2$year>=2011 & D2$Treatment=="No_grass")
+   D2$W.HECO[ii] <- 0 ; D2$W.POSE[ii] <- 0 ; D2$W.PSSP[ii] <- 0
+}
+
+# combine old and modern
+allD <- rbind(D1,D2)
+rm(D1,D2,tmp)
[…]

### 2016-09-12 precip/R/growth/get_PSSP_growth.R

@@ -1,0 +1,83 @@
+# PBA March 2016
+# modified by ARK July 2016
+# call from removal_analysis_wrapper.r
+
+#########################################
+#  1. Import data and calculate W's
+#########################################
+
+root <- '~'
+doSpp <- "PSSP"
+sppList <- c("ARTR","HECO","POSE","PSSP","allcov","allpts")
+dataDir1 <- paste(root,"/driversdata/data/idaho",sep="")
+dataDir2 <- paste(root,"/driversdata/data/idaho_modern",sep="")
+nonCompLength.s=5 #Number of columns in SppData that are not measures of competitors 
+
+# set up distance weights------------------------------------------------
+
+dists <- read.csv(paste(dataDir2,"/speciesData/IdahoModDistanceWeights_noExptl.csv",sep=""));
+dists$allcov <- rowMeans(dists[,1:4])  # for "other" polygons use average of big 4
+dists$allpts <- dists$POSE  # set forb dist wts = smallest grass (POSE)
+
+# import old data--------------------------------------------------------
+source("./R/growth/fetchGrowthData.R")
+
+D1 <- fetchGdat(doSpp=doSpp,speciesList=sppList,datadir=dataDir1,distWts=dists)
+D1$Treatment <- "Control"
+D1$Period <- 'Historical'
+
+# import modern data--------------------------------------------------------
+D2 <- fetchGdat(doSpp=doSpp,speciesList=sppList,datadir=dataDir2,distWts=dists)
+D2$Period <- "Modern"
+
+# merge in treatment data
+tmp <- read.csv(paste(dataDir2,"/quad_info.csv",sep=""))
+tmp <- tmp[,c("quad","Treatment")]
+D2 <- merge(D2,tmp, all.x=T)
+
+# account for removal in baseline years
+if(doSpp!="ARTR"){
+  ii <- which(D2$year>=2011 & D2$Treatment=="No_shrub")
+  D2$W.ARTR[ii] <- 0
+}else{
+  ii <- which(D2$year>=2011 & D2$Treatment=="No_grass")
+   D2$W.HECO[ii] <- 0 ; D2$W.POSE[ii] <- 0 ; D2$W.PSSP[ii] <- 0
+}
+
+# combine old and modern
+allD <- rbind(D1,D2)
+rm(D1,D2,tmp)
+
+# merge data on removals at individual level
+tmp <- read.csv(paste(dataDir2,"/speciesData/",doSpp,"/",doSpp,"_within_ARTRremovals.csv",sep=""))
[…]

### 2016-09-12 precip/R/observed_soil_moisture/spring_spot_measurements.R

@@ -1,0 +1,116 @@
+##### Spring 2015 soil moisture
+##### 
+
+rm(list = ls())
+library(ggplot2 )
+library(dplyr)
+library(tidyr)
+library(lme4)
+
+q_info <- read.csv('data/quad_info.csv')
+
+calibration <- read.csv('data/soil_moist_data/spot_measurements/2015-05-07_soil_probe_calibration.csv')
+
+p1 <- read.csv('data/soil_moist_data/spot_measurements/2012-06-06_spot_measurements.csv', skip = 3)
+
+p1$date <- '2012-06-06'
+p1$Plot <- gsub( p1$Plot, pattern = '-', replacement = '_')
+p1$rep <- c(1:2)
+p1 <- p1 %>% rename( plot = Plot )
+
+p2 <- read.csv('data/soil_moist_data/spot_measurements/2015-04-29_spot_measurements.csv', skip = 2)
+p3 <- read.csv('data/soil_moist_data/spot_measurements/2015-05-07_spot_measurements.csv') 
+p4 <- read.csv('data/soil_moist_data/spot_measurements/2016-05-10_spot_measurements.csv')
+p5 <- read.csv('data/soil_moist_data/spot_measurements/2015-06-09_spot_measurements.csv')
+
+p2$date <- '2015-04-29'
+
+df <- rbind( p2, p3, p4, p5)
+
+df <- df %>% gather( key = rep, PCT, E1:W3 )
+
+df <- rbind( p1, df )
+
+q_info$plot <- gsub( q_info$QuadName, pattern = 'X', replacement = '')
+
+df <- merge (df, q_info , by = 'plot')
+
+df$date <- as.POSIXct(df$date, tz = 'MST')
+df <- df %>% rename(VWC = PCT)
+
+soil_density <- read.csv('data/soil_moist_data/gravimetric_samples/exclosure_soil_samples.csv')
+
+soil_density <- soil_density %>% 
+  mutate( layer = ifelse( depth > 15, 'deep', 'shallow') ) %>% 
+  mutate( wet_soil = wet_weight - bag_weight, soil = dry_weight - bag_weight ) %>% 
+  mutate( water_weight = wet_soil - soil, soil_density = soil/soil_volume_cm3, GWC = water_weight/soil) %>% 
+  group_by(layer) %>% 
+  summarise( avg_soil_density = mean(soil_density))
+
+soil_d_value <- soil_density$avg_soil_density[ soil_density$layer == 'shallow']
+  
[…]

### 2016-09-12 precip/R/test_pwd.R

@@ -1,0 +1,4 @@
+# test pwd 
+
+print( getwd())
+print(system('pwd'))

### 2016-09-12 precip/R/treatment_test_figure.r

@@ -1,0 +1,37 @@
+
+# call from removal_analysis_wrapper.r
+
+if(exists("trtTests")==F)  trtTests <- read.csv("treatment_test_results.csv")
+
+# plotting function
+plotEffects <- function(doStage){
+  ii<-which(trtTests$stage==doStage)
+  myLims <- c(min(trtTests$CI.02.5[ii],na.rm=T),max(trtTests$CI.97.5[ii],na.rm=T))
+  if(doStage=="recruitment") myLims <- c(-5,0.6)
+  plot(c(1:4),trtTests$effect[ii],ylim=myLims,xlim=c(0.5,4.5),xlab="",ylab="",xaxt="n",pch=16)
+  axis(1,at=c(1:4),labels=c("ARTR","HECO","POSE","PSSP"),las=2)
+  abline(h=0,col="gray")
+  points(x=c(1:4),y=trtTests$effect[ii],pch=16,cex=2) # add some larger points
+  arrows(x0=c(1:4),y0=trtTests$CI.02.5[ii],x1=c(1:4),y1=trtTests$CI.97.5[ii],
+         code=3,angle=90,length=0.05) 
+}
+
+png("../figures/treatment_tests.png",height=3.25,width=7,res=400,units="in")
+
+  par(mfrow=c(1,3),tcl=-0.2,mgp=c(2,0.5,0),mar=c(4,2,3,1),oma=c(2,2,0,0))
+  
+  #plotEffects(doStage="survival")
+  #mtext("(A) Survival",side=3,line=1,adj=0,cex=0.9)
+  
+  plotEffects(doStage="growth")
+  mtext("(B) Growth",side=3,line=1,adj=0,cex=0.9)
+  
+  #plotEffects(doStage="recruitment")
+  #mtext("(C) Recruitment",side=3,line=1,adj=0,cex=0.9)
+  
+  mtext("Species",side=1,line=0,outer=T)
+  mtext("Removal effect",side=2,line=0,outer=T)
+
+dev.off()
+
+

### 2016-09-12 precip/analysis/treatment_trends_precip.r

@@ -1,0 +1,222 @@
+
+# call from removal_analysis_wrapper.r
+
+sppList=c("Artemisia tripartita","Hesperostipa comata","Poa secunda","Pseudoroegneria spicata")
+dataDir <- file.path(root,"/driversdata/data/idaho_modern/")
+
+# import data and calculate treatment trends ######################################
+
+
+covD<-read.csv(paste(dataDir,"allrecords_cover.csv",sep=""))
+trts<-read.csv(paste(dataDir,"quad_info.csv",sep=""))
+
+# use this to make sure we don't miss zeros
+allquadyrs<-unique(covD[,c("quad","year")],MARGIN=2)
+tmp<-expand.grid(species=sppList,year=sort(unique(covD$year)))
+allquadyrs<-merge(allquadyrs,tmp,all=T)
+allquadyrs<-merge(allquadyrs,trts[,c("Group","quad","Treatment")])
+
+# only do precip treatments
+keep <- which(is.element(allquadyrs$Treatment,c("Control","Drought","Irrigation")))
+allquadyrs <- allquadyrs[keep,]
+
+# aggregate to quadrat level
+keep<-which(is.element(covD$species,sppList))
+sppD<-covD[keep,]
+sppD<-merge(sppD,allquadyrs,all.y=T)
+sppD$area[is.na(sppD$area)]<-0
+sppD$area<-sppD$area*100
+sppD.q<-aggregate(sppD$area,by=list(species=sppD$species,Group=sppD$Group,Treatment=sppD$Treatment,
+                  quad=sppD$quad,year=sppD$year),FUN=sum)
+names(sppD.q)[NCOL(sppD.q)]<-"cover"
+write.csv(sppD.q,"data/temp_data/QuadYearCover.csv",row.names = FALSE)  # save these 
+
+# focus on all control plots or just those in the big exclosure?
+sppD.q <- subset(sppD.q,Group=="E1")
+
+#calculate treatment means by year
+spp.mean <- aggregate(sppD.q$cover,by=list(species=sppD.q$species,Treatment=sppD.q$Treatment,
+                  year=sppD.q$year),FUN=mean)
+names(spp.mean)[NCOL(spp.mean)] <- "cover"
[…]

### 2016-09-13 removals/manuscript/removal_ms.tex
Minor changes to removal ms

@@
 \author[a]{Andrew Kleinhesselink}
-\author[b]{Stephen P. Ellner}
-\author[c]{Giles Hooker}
-\author[a]{Brittany Teller, and others? Robin? Bret?}
+\author[b]{Giles Hooker}
+\author[c]{Bret Taylor}
+\author[a]{Brittany Teller}
+\author[d]{Stephen P. Ellner}
 \affil[a]{Department of Wildland Resources and the Ecology Center, Utah State University, Logan, Utah 84322}
-\affil[b]{Department of Ecology and Evolutionary Biology, Cornell University, Ithaca, New York}
-\affil[c]{Department of Biological Statistics and Computational Biology, Cornell University, Ithaca, New York}
+\affil[b]{Department of Biological Statistics and Computational Biology, Cornell University, Ithaca, New York}
+\affil[c]{USDA-ARS....}
+\affil[d]{Department of Ecology and Evolutionary Biology, Cornell University, Ithaca, New York}
 
@@
 Stable coexistence requires that intraspecific interactions must limit populations more than interspecific interactions (Chesson 2000).
-\comment{(SPE: snip here)}
-%In other words, when species are limited by different factors, intraspecific limitations will be stronger than interspecific limitations. 
-%For example, species might differ in their use of multiple limiting resources (Tilman), respond differently to spatial and temporal variation in the
-%environment (Chesson 2000), or host different natural enemies (Janzen, Connell). 
 Whenever we study populations of species that appear to be coexisting stably, such as populations that have co-occurred in close proximity for extended periods, we should expect intraspecific limitations to be stronger than interspecific limitations.
 However, we still might expect interspecific interactions to be strong and negative.
-
 In fact, interspecific competition is arguably the primary obsession of plant community ecologists.
@@
 Competitive release provides an intuitive, population-level measure of the strength of interspecific competition.
-Following removal of one or more species from a community, how much will the abundance of remaining species increase?
+Following removal of one or more species from a community, how much will the abundance of the remaining species increase?
 Large increases in the abundance of the remaining species represent a large competitive release, and imply strong interspecific interaction.
@@
 Since we only studied the most common, co-occurring species, we expected to find evidence of stable coexistence, but we were surprised that interspecific interactions and species' responses to the loss of a dominant competitor were so weak.
-Compared to traditional field experiments, our approach has the advantage of projecting full life-cycle responses over long time periods, but its reliance on observational data sets means that inference is weak.
+Compared to traditional field experiments, our approach has the advantage of projecting full life-cycle responses over long time periods, but its reliance on observational data sets limits the strength of our inference.
 We would have more confidence in our results if they were confirmed by an experiment.
@@
 comata}, and \textit{Poa secunda}, as well as two covariates representing total crowding from (1) all other perennial grasses and shrubs, which were mapped as polygons, and (2) all forb species, which were mapped as points.
-\new{$\left \langle \boldsymbol{x, y} \right \rangle$ denotes the inner product of vectors $\boldsymbol{x}$ and $\boldsymbol{y}$, 
-calculated as \texttt{sum(x*y)} in R.}
+$\left \langle \boldsymbol{x, y} \right \rangle$ denotes the inner product of vectors $\boldsymbol{x}$ and $\boldsymbol{y}$, 
+calculated as \texttt{sum(x*y)} in R.
 
@@
 Expected cover at $t+1$ is expected area of new recruits, plus the sum of the survival probabilities multiplied by the expected sizes.
-% need to REDO with observed W's rather than W's that assume round genets
 
@@
 \subsection*{Integral projection models}
-
-\comment{SPE: how were group effects incorporated into the IPM, and the IBM?}
 
@@
 We simulated the IPM to estimate equilibrium abundances under three different scenarios.
+For all IPM simulations, we set the spatial random effects to zero, meaning that we are simulating dynamics on a hypothetical, average site.
 The first scenario featured all four species and our ``baseline'' model for control plots (no removal effects).
@@
 \ref{fig:VitalRateTest}; Tables \ref{ARTRgrowth} to \ref{PSSPgrowth}).
-In contrast, \textit{A.
-tripartita} and  \textit{Poa secunda} had lower recruitment in removal than controls plots (upper limit of 95\% CI $<0$; Fig.
+In contrast, \textit{Poa secunda} had lower recruitment in removal than controls plots (upper limit of 95\% CI $<0$; Fig.
 \ref{fig:VitalRateTest}; Table \ref{table:recruitment}).
-The very low value (-26.8) and large confidence intervals (-69.6, -3.2) of the removal effect for  \textit{A.
-tripartita} recruitment reflect the fact that zero recruitment occurred in the four years of observations on the eight grass removal plots, making parameter estimation difficult.
 
@@
 
-Funding: NSF DEB-1353078, DEB-1054040, DEB 1353039 (SPE), Utah Agricultural Experiment Station (get journal paper number).
+Funding was provided by NSF grants DEB-1353078,and  DEB-1054040 to PBA and DEB 1353039 to SPE, and by the Utah Agricultural Experiment Station (get journal paper number).
 The USDA-ARS Sheep Experiment Station generously provided access to historical data and the field experiment site.
+We thank Robin Snyder for suggestions that improved the manuscript.
 
@@
 The maximum reduction in deviance for the generalized linear regressions and $R^2$ for the linear regressions were both less than 8\%.
-The \texttt{R} code for this analysis is included as ..\texttt{\textbackslash Wdistrib\textbackslash exploreSurvivalWs.r}) \comment{It's not clear what the independent and dependent variables are, in the analyses described
-here. It might help to use the notation from the Methods section?}
+The \texttt{R} code for this analysis is included as ..\texttt{\textbackslash Wdistrib\textbackslash exploreSurvivalWs.r}) \comment{It's not clear what the independent and dependent variables are, in the analyses described here.
+It might help to use the notation from the Methods section?}
 

### 2016-09-20 removals/manuscript/removal_ms.tex
Revising manuscript

@@
 We assigned half the new quadrats to a  perennial grass removal treatment and half to a treatment in which we removed the dominant shrub, \textit{Artemisia tripartita}.
-\item We modeled survival, growth and recruitment as functions of local neighborhood species composition along with an indicator variable for removal treatment.
-For survival and recruitment, the removal treatment effects were never significantly positive, indicating that competitive release was not underestimated by the interspecific effects included in the ``baseline'' model.
+We modeled survival, growth and recruitment as functions of local neighborhood species composition along with an indicator variable for removal treatment.
+If our ``baseline" model, which accounts for local plant-plant interactions, accurately predicts responses to removal, then the fitted removal indicator effect should be non-signficant.
+\item  For survival and recruitment, the removal treatment effects were never significantly positive, indicating that competitive release was not underestimated by the interspecific effects included in the baseline model.
+For  \textit{Poa secunda} recruitment, the removal treatment was actually negative, indicating that our baseline model overestimated competitive release.
 \item For growth, the removal treatment effect was significant and positive for two species, \textit{Poa secunda} and \textit{Pseudoroegneria spicata}, indicating that the baseline model underestimated competitive release.
@@
 For \textit{P.
-spicata}, the population models that included effects of \textit{Artemisia} removal projected much stronger competitive release than the baseline models.
+spicata}, the population models that included effects of \textit{Artemisia} removal projected greater competitive release than the baseline models.
 \item \emph{Synthesis}: Although models based on observational data did not perfectly predict species responses to removal, overall the results increase our confidence that competitive release is in fact small for at least three of our four study species.
@@
 An alternative approach to studying competitive release is to fit multispecies population models to long-term, observational data, and then simulate population responses to removals.
-Applying this approach to arid and semiarid plant communities, we found evidence of strong intraspecific limitation and very weak interspecific interactions, leading to very stable coexistence and implying very little competitive release \citep{Adler:2010fu} ( Chu and Adler 2015).
+Applying this approach to arid and semiarid plant communities, we found evidence of strong intraspecific limitation and very weak interspecific interactions, leading to very stable coexistence and implying little competitive release \citep{Adler:2010fu} ( Chu and Adler 2015).
 Since we only studied the most common, co-occurring species, we expected to find evidence of stable coexistence, but we were surprised that interspecific interactions and species' responses to the loss of a dominant competitor were so weak.
@@
 The sagebrush removal was virtually 100\% effective; grass removals required additional herbicide applications in April and May of each year, with fewer grasses remaining each year.
-
-% add climate data for 2011-2015
 
@@
 
+One reason we might observe small effects of the removal treatment is the limited power of our experiment.
+Perhaps large uncertainty in the estimates of these effects makes it difficult to reject the null hypothesis represented by our baseline model.
+To address this possibility, we also simulated a version of the model in which we replaced the mean estimate of each removal treatment effect with the parameter value at either the 5\% or 95\% confidence interval, whichever was farthest from zero.
+This version of the model shows the population-level changes we would expect given maximum removal effects.
+
 \subsection*{Integral projection models}
@@
 
+As with the IBM, we repeated these simulations using maximum removal treatment effects represented by parameters values at the 5\% or 95\% confidence limit farthest from zero.
+
 \section*{Results}
 
+During the period of our experiment, annual water-year (October-September) precipitation fell within the 5\% - 95\% quantiles of the long-term mean, although more years fell below the mean than above it (\ref{fig:climate}a).
+However, annual mean temperatures were consistently warm relative to the long-term mean, and two years were warmer than the 95\% quantile (\ref{fig:climate}b).
+
 \subsection*{Cover trends}
@@
 Trends in cover by species and removal treatment were variable. For example, grass removal had little effect on the cover of \textit{A. tripartita} 
-(Fig.
-\ref{fig:CoverTrends}), but shrub removal did appear to cause an increase in the cover of \textit{P.
-spicata}.
+(Fig. \ref{fig:CoverTrends}), but shrub removal did appear to cause an increase in the cover of \textit{P. spicata}. 
 We tested the effects of removal on cover by analyzing year-to-year changes in (log) cover, a measure of population growth rate (Fig.
@@
 The removal treatments had no significant effect on
-\textit{A. tripartita} (95\% confidence interval on removal treatment effect: -0.45, 0.27) or 
-\textit{Poa secunda} (95\% CI: -0.40, 0.28) population growth, but did increase 
-\textit{H. comata} (95\% CI: 0.02, 0.87) and
- \textit{P. spicata} (95\% CI: 0.04, 0.50) population growth 
+\textit{A. tripartita} (95\% confidence interval on removal treatment effect: -0.75, 0.56) or 
+\textit{Poa secunda} (95\% CI: -0.41, 0.25) population growth, but did increase 
+\textit{H. comata} (95\% CI: 0.01, 0.65) and
+ \textit{P. spicata} (95\% CI: 0.01, 0.40) population growth 
 (Table \ref{table:coefficients}). However, these treatment comparisons do not address our questions about whether responses to removal were 
@@
 Predictions from the ``Baseline'' model do account for effects of local species composition on vital rates, but do not include additional effects of removals.
-Predictions from the ``Removal'' model do include additional treatment effects on vital rates.
+Predictions from the ``Removal'' model include additional treatment effects on vital rates.
 }
@@
 
+%$  $ FIX COVERCHANGE LEGEND
+
 \subsection*{Vital rates}
@@
 If the removal treatment effects that we added to these models are statistically significant, it would indicate an effect of removals that goes above and beyond the response predicted by a model based on data from the control plots.
-For the survival models, the 95\% confidence intervals on the removals effects overlapped zero for all four species (Fig.
+For the survival models, the 95\% confidence intervals on the removal effects overlapped zero for all four species (Fig.
 \ref{fig:VitalRateTest}; Tables \ref{ARTRsurvival} to \ref{PSSPsurvival}), meaning that the survival probabilities of plants in removal plots can be predicted based on data from control plots.
@@
 Second, we evaluated the possibility that removal effects might be transient.
-We removed aboveground plant material, but roots of the removed plants remain in the ground and decompose, potentially creating a flush of mineral nutrients (refs?).
+We removed aboveground plant material, but roots of the removed plants remain in the ground and decompose, potentially creating a flush of mineral nutrients.
 We tested for a transient removal effect by adding a year-by-treatment interaction.
 For both  \textit{Poa secunda}  and \textit{P.
-spicata} , we found no clear sign of a decrease in the removal effect over time, and in fact the removal effects for \textit{P.
-spicata} were strongest in the last two years of observations (Tables \ref{table:POSEgrowth-byYr} and \ref{table:PSSPgrowth-byYr}).
+spicata} , we found no sign of a decrease in the removal effect over time (Tables \ref{table:POSEgrowth-byYr} and \ref{table:PSSPgrowth-byYr}).
 
@@
 To quantify the short-term population-level consequences of removals, we used an IBM to project the cover of each species in each quadrat one year ahead using versions of the model with and without removal treatment effects.
+Our models tended to under-predict population growth (Fig.
+\ref{fig:CoverChange}) and population size (Fig.
+\ref{fig:IBM1step}) for all species, but the difference between predictions from the baseline and removal models varied among species.
 For \textit{A.
-tripartita}, both observed and predicted population growth rates were close to zero, and differences in model predictions were small (Fig.
-\ref{fig:CoverChange}).
+tripartita}, both observed and predicted population growth rates were close to zero, and differences in model predictions were small.
 \textit{H.
@@
 spicata}, the model with removal effects consistently predicted higher population growth than the baseline model (Fig.
-\ref{fig:CoverChange}).
-For \textit{P.
-spicata}, including the removal effects increased the annual multiplicative rate of increase by roughly 35\%.
+\ref{fig:CoverChange}), and reduced the underestimates of cover (Fig.
+\ref{fig:IBM1step}).
+The version of the model with maximum removal treatment effects (removal coefficients set to the 95\% confidence limit furthest from zero), were qualitatively similar, but further amplified the contrast between predictions of the baseline and removal models for and \textit{P.
+spicata} REF MAX CI FIG
+
+ \begin{figure}[tbp]
+ \centering
+ \includegraphics[width=1\textwidth]{cover_projections_1step}
+ \caption{Observed and predicted cover for each of the four modeled species.
+Solid lines show mean observed cover averaged across control (blue) and removal treatment (red) quadrats.
+Dashed lines show one-step-ahead predictions from the baseline IBM (no removal treatment coefficients), averaged across quadrats.
+Dotted lines show one-step-ahead predictions from an IBM that includes removal treatment coefficients.
+}
+ \label{fig:IBM1step}
+ \end{figure}
+
+% add max CI figure in SI
 
@@
 We concentrated on grass species' responses to \textit{A.
-tripartita} removal, since the growth regressions indicated our baseline model might underestimate competitive release in this situation.
+tripartita} removal, since the growth regressions indicated our baseline model might underestimate competitive release in this situation (our baseline vital rate models worked well in predicting \textit{A.
+tripartita} responses to grass removal).
 Using the baseline model, with no removal treatment coeffiicients, the IPM predicts little effect of \textit{A.
@@
 Using a model with removal treatment coefficients, the IPM predicts little effect of shrub removal on \textit{Poa secunda} and \textit{H.
-comata}, but a 2.75 fold increase in the cover of \textit{P.
+comata}, but a 2-fold increase in the cover of \textit{P.
 spicata} (Fig.
 \ref{fig:IPMresults}).
+Running a version of the model with maximum removal treatment effects (removal coefficients set to the 95\% confidence limit furthest from zero), resulted in the same qualitative outcome, but a larger increase in \textit{P.
+spicata} % (Fig.
+\ref{fig:IPMresults-maxCI})
+
+% add max CI figure in SI
 
@@
 \hline
-(Intercept)           & $-0.03$          & $-0.04$          & $-0.02$          & $-0.09$          \\
-                      & $[-0.33;\ 0.27]$ & $[-0.32;\ 0.24]$ & $[-0.31;\ 0.27]$ & $[-0.33;\ 0.15]$ \\
-Treatment: No grasses    & $-0.09$          &                  &                  &                  \\
-                      & $[-0.45;\ 0.27]$ &                  &                  &                  \\
-Treatment: No shrub    &                  & $0.44^{*}$       & $-0.06$          & $0.27^{*}$       \\
-                      &                  & $[0.02;\ 0.87]$  & $[-0.40;\ 0.28]$ & $[0.04;\ 0.50]$  \\
-\hline
-AIC                   & 23.20            & 59.58            & 162.06           & 117.50           \\
-BIC                   & 33.15            & 66.24            & 173.51           & 129.22           \\
-Log Likelihood        & -6.60            & -24.79           & -76.03           & -53.75           \\
-Num. obs.             & 54               & 28               & 73               & 77               \\
-Num. groups: quad     & 12               & 5                & 14               & 14               \\
-Num. groups: year     & 8                & 8                & 8                & 8                \\
-Var: quad (Intercept) & 0.08             & 0.00             & 0.00             & 0.00             \\
-Var: year (Intercept) & 0.01             & 0.00             & 0.10             & 0.09             \\
-Var: Residual         & 0.04             & 0.32             & 0.40             & 0.19             \\
+(Intercept)           & $0.00$           & $-0.05$          & $0.01$           & $-0.10$          \\
+                      & $[-0.41;\ 0.42]$ & $[-0.21;\ 0.10]$ & $[-0.26;\ 0.28]$ & $[-0.28;\ 0.08]$ \\
+TreatmentNo\_grass    & $-0.09$          &                  &                  &                  \\
+                      & $[-0.75;\ 0.56]$ &                  &                  &                  \\
+TreatmentNo\_shrub    &                  & $0.33^{*}$       & $-0.08$          & $0.21^{*}$       \\
+                      &                  & $[0.01;\ 0.65]$  & $[-0.41;\ 0.25]$ & $[0.01;\ 0.40]$  \\
+\hline
+AIC                   & 284.97           & 160.41           & 337.77           & 253.15           \\
+BIC                   & 298.99           & 172.62           & 352.65           & 268.46           \\
+Log Likelihood        & -137.49          & -75.20           & -163.88          & -121.57          \\
+Num. obs.             & 122              & 85               & 145              & 158              \\
+Num. groups: quad     & 19               & 11               & 21               & 22               \\
+Num. groups: year     & 9                & 9                & 9                & 9                \\
+Var: quad (Intercept) & 0.43             & 0.00             & 0.02             & 0.00             \\
+Var: year (Intercept) & 0.00             & 0.02             & 0.12             & 0.06             \\
+Var: Residual         & 0.41             & 0.32             & 0.48             & 0.24             \\
 \hline
@@
 % latex table generated in R 3.2.2 by xtable 1.8-2 package
-% Thu May 19 08:31:09 2016
+% Mon Sep 19 08:32:17 2016
 \begin{table}[ht]
 \centering
-\caption{Summary of fixed effects for the ARTR survival model} 
+\caption{Summary of fixed effects for the \textit{A. tripartita} survival model} 
 \label{ARTRsurvival}
@@
   \hline
-(Intercept) & -0.0705 & 0.2256 & -0.5109 & -0.0723 & 0.3799 & -0.0757 & 0.0000 \\ 
-  logarea & 0.7137 & 0.0605 & 0.5980 & 0.7124 & 0.8370 & 0.7098 & 0.0000 \\ 
-  TreatmentNo\_grass & -0.2581 & 0.8998 & -1.9434 & -0.2884 & 1.5959 & -0.3511 & 0.0000 \\ 
-  W.ARTR & -1.4362 & 0.2154 & -1.8709 & -1.4322 & -1.0242 & -1.4240 & 0.0000 \\ 
-  W.HECO & -0.0818 & 0.0430 & -0.1630 & -0.0830 & 0.0062 & -0.0855 & 0.0000 \\ 
-  W.POSE & -0.1159 & 0.0859 & -0.2851 & -0.1158 & 0.0523 & -0.1155 & 0.0000 \\ 
-  W.PSSP & 0.0292 & 0.0276 & -0.0225 & 0.0283 & 0.0863 & 0.0263 & 0.0000 \\ 
-  W.allcov & -0.0081 & 0.0093 & -0.0265 & -0.0081 & 0.0099 & -0.0080 & 0.0000 \\ 
-  W.allpts & 0.1296 & 0.1579 & -0.1787 & 0.1290 & 0.4409 & 0.1278 & 0.0000 \\ 
+(Intercept) & -0.0207 & 0.2261 & -0.4554 & -0.0246 & 0.4367 & -0.0321 & 0.0000 \\ 
+  logarea & 0.7208 & 0.0575 & 0.6109 & 0.7195 & 0.8378 & 0.7171 & 0.0000 \\ 
+  TreatmentNo\_grass & -1.3218 & 0.7343 & -2.7293 & -1.3340 & 0.1563 & -1.3591 & 0.0000 \\ 
+  W.ARTR & -1.5214 & 0.2075 & -1.9391 & -1.5179 & -1.1237 & -1.5107 & 0.0000 \\ 
+  W.HECO & -0.0694 & 0.0423 & -0.1485 & -0.0708 & 0.0178 & -0.0738 & 0.0000 \\ 
+  W.POSE & -0.0049 & 0.0761 & -0.1469 & -0.0077 & 0.1525 & -0.0134 & 0.0000 \\ 
+  W.PSSP & 0.0281 & 0.0275 & -0.0235 & 0.0271 & 0.0849 & 0.0252 & 0.0000 \\ 
+  W.allcov & -0.0055 & 0.0083 & -0.0219 & -0.0054 & 0.0109 & -0.0054 & 0.0000 \\ 
+  W.allpts & 0.1694 & 0.1579 & -0.1386 & 0.1686 & 0.4812 & 0.1672 & 0.0000 \\ 
    \hline
@@
 % latex table generated in R 3.2.2 by xtable 1.8-2 package
-% Thu May 19 08:31:59 2016
+% Mon Sep 19 08:33:00 2016
 \begin{table}[ht]
 \centering
-\caption{Summary of fixed effects for the HECO survival model} 
+\caption{Summary of fixed effects for the \textit{H. comata} survival model} 
 \label{HECOsurvival}
@@
   \hline
-(Intercept) & 1.4652 & 0.2285 & 1.0126 & 1.4652 & 1.9170 & 1.4650 & 0.0000 \\ 
-  logarea & 1.3190 & 0.1036 & 1.1266 & 1.3149 & 1.5346 & 1.3068 & 0.0000 \\ 
-  TreatmentNo\_shrub & 0.7313 & 0.4426 & -0.1292 & 0.7284 & 1.6074 & 0.7226 & 0.0000 \\ 
-  W.ARTR & -0.0049 & 0.0037 & -0.0119 & -0.0050 & 0.0027 & -0.0052 & 0.0000 \\ 
-  W.HECO & -0.6707 & 0.0725 & -0.8206 & -0.6680 & -0.5354 & -0.6627 & 0.0000 \\ 
-  W.POSE & 0.0965 & 0.0690 & -0.0272 & 0.0921 & 0.2440 & 0.0830 & 0.0000 \\ 
-  W.PSSP & -0.0015 & 0.0197 & -0.0386 & -0.0020 & 0.0390 & -0.0032 & 0.0000 \\ 
-  W.allcov & -0.0032 & 0.0058 & -0.0140 & -0.0034 & 0.0087 & -0.0038 & 0.0000 \\ 
-  W.allpts & -0.2630 & 0.1184 & -0.4889 & -0.2654 & -0.0236 & -0.2701 & 0.0000 \\ 
-  logarea:W.ARTR & -0.0005 & 0.0028 & -0.0057 & -0.0006 & 0.0052 & -0.0008 & 0.0000 \\ 
-  logarea:W.HECO & 0.2751 & 0.0600 & 0.1522 & 0.2770 & 0.3879 & 0.2809 & 0.0000 \\ 
-  logarea:W.POSE & 0.0067 & 0.0513 & -0.0833 & 0.0026 & 0.1182 & -0.0060 & 0.0000 \\ 
-  logarea:W.PSSP & -0.0273 & 0.0147 & -0.0543 & -0.0280 & 0.0036 & -0.0294 & 0.0000 \\ 
-  logarea:W.allcov & -0.0043 & 0.0041 & -0.0118 & -0.0045 & 0.0044 & -0.0050 & 0.0000 \\ 
-  logarea:W.allpts & -0.1474 & 0.0903 & -0.3187 & -0.1496 & 0.0364 & -0.1541 & 0.0000 \\ 
+(Intercept) & 1.4189 & 0.2115 & 0.9991 & 1.4190 & 1.8371 & 1.4192 & 0.0000 \\ 
+  logarea & 1.2885 & 0.0813 & 1.1399 & 1.2843 & 1.4601 & 1.2758 & 0.0000 \\ 
+  TreatmentNo\_shrub & 0.3958 & 0.4175 & -0.4193 & 0.3942 & 1.2190 & 0.3909 & 0.0000 \\ 
+  W.ARTR & -0.0064 & 0.0028 & -0.0120 & -0.0064 & -0.0009 & -0.0064 & 0.0000 \\ 
+  W.HECO & -0.6647 & 0.0673 & -0.7996 & -0.6638 & -0.5351 & -0.6619 & 0.0000 \\ 
+  W.POSE & 0.0732 & 0.0534 & -0.0304 & 0.0728 & 0.1790 & 0.0720 & 0.0000 \\ 
+  W.PSSP & 0.0190 & 0.0184 & -0.0167 & 0.0188 & 0.0554 & 0.0185 & 0.0000 \\ 
+  W.allcov & 0.0021 & 0.0047 & -0.0072 & 0.0021 & 0.0113 & 0.0021 & 0.0000 \\ 
+  W.allpts & -0.1185 & 0.0919 & -0.2998 & -0.1183 & 0.0610 & -0.1177 & 0.0000 \\ 
    \hline
@@
 % latex table generated in R 3.2.2 by xtable 1.8-2 package
-% Thu May 19 08:33:14 2016
+% Mon Sep 19 08:34:30 2016
 \begin{table}[ht]
 \centering
-\caption{Summary of fixed effects for the POSE survival model} 
+\caption{Summary of fixed effects for the \textit{Poa secunda} survival model} 
 \label{POSEsurvival}
@@
   \hline
-(Intercept) & 1.3190 & 0.2047 & 0.9110 & 1.3199 & 1.7215 & 1.3213 & 0.0000 \\ 
-  logarea & 1.0711 & 0.0694 & 0.9387 & 1.0693 & 1.2137 & 1.0657 & 0.0000 \\ 
-  TreatmentNo\_shrub & -0.0442 & 0.1832 & -0.4030 & -0.0445 & 0.3158 & -0.0451 & 0.0000 \\ 
+(Intercept) & 1.3162 & 0.1965 & 0.9242 & 1.3171 & 1.7024 & 1.3186 & 0.0000 \\ 
+  logarea & 1.0585 & 0.0640 & 0.9365 & 1.0569 & 1.1902 & 1.0537 & 0.0000 \\ 
+  TreatmentNo\_shrub & -0.2572 & 0.1723 & -0.5949 & -0.2573 & 0.0811 & -0.2576 & 0.0000 \\ 
   W.ARTR & 0.0001 & 0.0019 & -0.0036 & 0.0001 & 0.0038 & 0.0001 & 0.0000 \\ 
-  W.HECO & -0.0075 & 0.0142 & -0.0353 & -0.0076 & 0.0204 & -0.0077 & 0.0000 \\ 
-  W.POSE & -1.2650 & 0.0825 & -1.4296 & -1.2641 & -1.1053 & -1.2624 & 0.0000 \\ 
-  W.PSSP & 0.0244 & 0.0123 & 0.0005 & 0.0243 & 0.0486 & 0.0242 & 0.0000 \\ 
-  W.allcov & -0.0031 & 0.0027 & -0.0083 & -0.0031 & 0.0022 & -0.0031 & 0.0000 \\ 
-  W.allpts & 0.0145 & 0.0527 & -0.0889 & 0.0144 & 0.1181 & 0.0143 & 0.0000 \\ 
+  W.HECO & -0.0153 & 0.0138 & -0.0424 & -0.0154 & 0.0119 & -0.0154 & 0.0000 \\ 
+  W.POSE & -1.2632 & 0.0816 & -1.4260 & -1.2623 & -1.1053 & -1.2605 & 0.0000 \\ 
+  W.PSSP & 0.0249 & 0.0122 & 0.0011 & 0.0249 & 0.0490 & 0.0247 & 0.0000 \\ 
+  W.allcov & -0.0024 & 0.0026 & -0.0075 & -0.0024 & 0.0029 & -0.0024 & 0.0000 \\ 
+  W.allpts & 0.0018 & 0.0524 & -0.1009 & 0.0018 & 0.1048 & 0.0017 & 0.0000 \\ 
    \hline
@@
 % latex table generated in R 3.2.2 by xtable 1.8-2 package
-% Thu May 19 08:34:39 2016
+% Mon Sep 19 08:35:48 2016
 \begin{table}[ht]
 \centering
-\caption{Summary of fixed effects for the PSSP survival model} 
+\caption{Summary of fixed effects for the \textit{P. spicata} survival model} 
 \label{PSSPsurvival}
@@
   \hline
-(Intercept) & 1.2208 & 0.1772 & 0.8756 & 1.2191 & 1.5757 & 1.2158 & 0.0000 \\ 
-  logarea & 1.5691 & 0.1099 & 1.3570 & 1.5674 & 1.7908 & 1.5641 & 0.0000 \\ 
-  TreatmentNo\_shrub & -0.2570 & 0.2348 & -0.7203 & -0.2563 & 0.2016 & -0.2548 & 0.0000 \\ 
-  W.ARTR & 0.0094 & 0.0026 & 0.0044 & 0.0094 & 0.0147 & 0.0093 & 0.0000 \\ 
-  W.HECO & -0.0005 & 0.0192 & -0.0370 & -0.0009 & 0.0386 & -0.0018 & 0.0000 \\ 
-  W.POSE & 0.0183 & 0.0467 & -0.0675 & 0.0162 & 0.1162 & 0.0116 & 0.0000 \\ 
-  W.PSSP & -0.4486 & 0.0312 & -0.5112 & -0.4481 & -0.3885 & -0.4472 & 0.0000 \\ 
-  W.allcov & 0.0074 & 0.0035 & 0.0007 & 0.0073 & 0.0146 & 0.0072 & 0.0000 \\ 
-  W.allpts & 0.0792 & 0.0797 & -0.0726 & 0.0776 & 0.2405 & 0.0742 & 0.0000 \\ 
-  logarea:W.ARTR & -0.0014 & 0.0021 & -0.0053 & -0.0014 & 0.0029 & -0.0015 & 0.0000 \\ 
-  logarea:W.HECO & -0.0174 & 0.0158 & -0.0466 & -0.0182 & 0.0156 & -0.0196 & 0.0000 \\ 
-  logarea:W.POSE & 0.0147 & 0.0386 & -0.0540 & 0.0120 & 0.0978 & 0.0065 & 0.0000 \\ 
-  logarea:W.PSSP & 0.0127 & 0.0239 & -0.0334 & 0.0124 & 0.0605 & 0.0119 & 0.0000 \\ 
-  logarea:W.allcov & -0.0027 & 0.0027 & -0.0078 & -0.0028 & 0.0029 & -0.0029 & 0.0000 \\ 
-  logarea:W.allpts & -0.0255 & 0.0647 & -0.1483 & -0.0270 & 0.1063 & -0.0302 & 0.0000 \\ 
+(Intercept) & 1.2121 & 0.1699 & 0.8815 & 1.2103 & 1.5533 & 1.2066 & 0.0000 \\ 
+  logarea & 1.5492 & 0.0976 & 1.3612 & 1.5474 & 1.7473 & 1.5437 & 0.0000 \\ 
+  TreatmentNo\_shrub & -0.1885 & 0.2208 & -0.6237 & -0.1879 & 0.2431 & -0.1867 & 0.0000 \\ 
+  W.ARTR & 0.0101 & 0.0021 & 0.0061 & 0.0101 & 0.0142 & 0.0101 & 0.0000 \\ 
+  W.HECO & 0.0046 & 0.0185 & -0.0319 & 0.0046 & 0.0406 & 0.0047 & 0.0000 \\ 
+  W.POSE & 0.0245 & 0.0384 & -0.0502 & 0.0242 & 0.1007 & 0.0236 & 0.0000 \\ 
+  W.PSSP & -0.4430 & 0.0291 & -0.5010 & -0.4427 & -0.3867 & -0.4420 & 0.0000 \\ 
+  W.allcov & 0.0100 & 0.0028 & 0.0046 & 0.0100 & 0.0155 & 0.0100 & 0.0000 \\ 
+  W.allpts & 0.0943 & 0.0605 & -0.0246 & 0.0944 & 0.2129 & 0.0944 & 0.0000 \\ 
    \hline
@@
 % latex table generated in R 3.2.2 by xtable 1.8-2 package
-% Thu May 19 09:51:32 2016
+% Mon Sep 19 08:36:00 2016
 \begin{table}[ht]
 \centering
-\caption{Summary of fixed effects for the ARTR growth model} 
+\caption{Summary of fixed effects for the \textit{A. tripartita} growth model} 
 \label{ARTRgrowth}
@@
   \hline
-(Intercept) & 0.7252 & 0.2473 & 0.2395 & 0.7252 & 1.2107 & 0.7251 & 0.0000 \\ 
-  logarea.t0 & 0.8661 & 0.0407 & 0.7860 & 0.8661 & 0.9460 & 0.8661 & 0.0000 \\ 
-  TreatmentNo\_grass & 0.1014 & 0.1714 & -0.2352 & 0.1014 & 0.4377 & 0.1014 & 0.0000 \\ 
-  W.ARTR & -0.0743 & 0.0989 & -0.2685 & -0.0743 & 0.1198 & -0.0743 & 0.0000 \\ 
-  W.HECO & 0.0123 & 0.0151 & -0.0174 & 0.0123 & 0.0420 & 0.0123 & 0.0000 \\ 
-  W.POSE & -0.0417 & 0.0325 & -0.1055 & -0.0417 & 0.0221 & -0.0417 & 0.0000 \\ 
-  W.PSSP & 0.0067 & 0.0086 & -0.0101 & 0.0067 & 0.0235 & 0.0067 & 0.0000 \\ 
-  W.allcov & -0.0063 & 0.0032 & -0.0127 & -0.0063 & 0.0000 & -0.0063 & 0.0000 \\ 
-  W.allpts & 0.0043 & 0.0503 & -0.0944 & 0.0043 & 0.1030 & 0.0043 & 0.0000 \\ 
+(Intercept) & 0.7378 & 0.2437 & 0.2591 & 0.7378 & 1.2161 & 0.7377 & 0.0000 \\ 
+  logarea.t0 & 0.8663 & 0.0393 & 0.7892 & 0.8663 & 0.9434 & 0.8663 & 0.0000 \\ 
+  TreatmentNo\_grass & 0.1037 & 0.1436 & -0.1783 & 0.1037 & 0.3854 & 0.1037 & 0.0000 \\ 
+  W.ARTR & -0.0364 & 0.0892 & -0.2116 & -0.0364 & 0.1387 & -0.0364 & 0.0000 \\ 
+  W.HECO & 0.0033 & 0.0136 & -0.0233 & 0.0033 & 0.0299 & 0.0033 & 0.0000 \\ 
+  W.POSE & -0.0573 & 0.0268 & -0.1100 & -0.0573 & -0.0047 & -0.0573 & 0.0000 \\ 
+  W.PSSP & 0.0062 & 0.0084 & -0.0102 & 0.0062 & 0.0226 & 0.0062 & 0.0000 \\ 
+  W.allcov & -0.0055 & 0.0030 & -0.0114 & -0.0055 & 0.0004 & -0.0055 & 0.0000 \\ 
+  W.allpts & 0.0029 & 0.0477 & -0.0908 & 0.0029 & 0.0965 & 0.0029 & 0.0000 \\ 
    \hline
@@
 % latex table generated in R 3.2.2 by xtable 1.8-2 package
-% Thu May 19 09:51:55 2016
+% Mon Sep 19 08:36:24 2016
 \begin{table}[ht]
 \centering
-\caption{Summary of fixed effects for the HECO growth model} 
+\caption{Summary of fixed effects for the \textit{H. comata} growth model} 
 \label{HECOgrowth}
@@
   \hline
-(Intercept) & 0.3524 & 0.0858 & 0.1812 & 0.3528 & 0.5208 & 0.3535 & 0.0000 \\ 
-  logarea.t0 & 0.8341 & 0.0212 & 0.7922 & 0.8341 & 0.8761 & 0.8340 & 0.0000 \\ 
-  TreatmentNo\_shrub & 0.0228 & 0.1523 & -0.2768 & 0.0230 & 0.3210 & 0.0235 & 0.0000 \\ 
-  W.ARTR & -0.0030 & 0.0010 & -0.0050 & -0.0030 & -0.0009 & -0.0030 & 0.0000 \\ 
-  W.HECO & -0.0472 & 0.0191 & -0.0847 & -0.0472 & -0.0097 & -0.0472 & 0.0000 \\ 
-  W.POSE & 0.0206 & 0.0161 & -0.0111 & 0.0206 & 0.0523 & 0.0206 & 0.0000 \\ 
-  W.PSSP & -0.0162 & 0.0065 & -0.0289 & -0.0162 & -0.0035 & -0.0162 & 0.0000 \\ 
-  W.allcov & -0.0047 & 0.0019 & -0.0085 & -0.0047 & -0.0009 & -0.0047 & 0.0000 \\ 
-  W.allpts & -0.0544 & 0.0374 & -0.1279 & -0.0544 & 0.0190 & -0.0545 & 0.0000 \\ 
+(Intercept) & 0.3782 & 0.0797 & 0.2212 & 0.3783 & 0.5343 & 0.3784 & 0.0000 \\ 
+  logarea.t0 & 0.8272 & 0.0205 & 0.7869 & 0.8272 & 0.8674 & 0.8272 & 0.0000 \\ 
+  TreatmentNo\_shrub & -0.0162 & 0.1336 & -0.2785 & -0.0162 & 0.2460 & -0.0162 & 0.0000 \\ 
+  W.ARTR & -0.0031 & 0.0010 & -0.0051 & -0.0031 & -0.0010 & -0.0031 & 0.0000 \\ 
+  W.HECO & -0.0479 & 0.0195 & -0.0862 & -0.0479 & -0.0098 & -0.0479 & 0.0000 \\ 
+  W.POSE & 0.0242 & 0.0162 & -0.0077 & 0.0242 & 0.0560 & 0.0242 & 0.0000 \\ 
+  W.PSSP & -0.0151 & 0.0065 & -0.0279 & -0.0151 & -0.0023 & -0.0151 & 0.0000 \\ 
+  W.allcov & -0.0038 & 0.0019 & -0.0076 & -0.0038 & 0.0000 & -0.0038 & 0.0000 \\ 
+  W.allpts & -0.0513 & 0.0380 & -0.1258 & -0.0513 & 0.0232 & -0.0513 & 0.0000 \\ 
    \hline
@@
 % latex table generated in R 3.2.2 by xtable 1.8-2 package
-% Thu May 19 09:52:23 2016
+% Mon Sep 19 08:36:58 2016
 \begin{table}[ht]
 \centering
-\caption{Summary of fixed effects for the POSE growth model} 
+\caption{Summary of fixed effects for the \textit{Poa secunda} growth model} 
 \label{POSEgrowth}
@@
   \hline
-(Intercept) & 0.4954 & 0.0697 & 0.3566 & 0.4957 & 0.6328 & 0.4963 & 0.0000 \\ 
-  logarea.t0 & 0.6788 & 0.0225 & 0.6335 & 0.6790 & 0.7227 & 0.6794 & 0.0000 \\ 
-  TreatmentNo\_shrub & 0.1789 & 0.0679 & 0.0456 & 0.1789 & 0.3120 & 0.1789 & 0.0000 \\ 
-  W.ARTR & -0.0003 & 0.0008 & -0.0019 & -0.0003 & 0.0014 & -0.0003 & 0.0000 \\ 
-  W.HECO & 0.0034 & 0.0062 & -0.0088 & 0.0034 & 0.0156 & 0.0034 & 0.0000 \\ 
-  W.POSE & -0.2681 & 0.0382 & -0.3432 & -0.2681 & -0.1931 & -0.2681 & 0.0000 \\ 
-  W.PSSP & -0.0077 & 0.0056 & -0.0186 & -0.0077 & 0.0032 & -0.0077 & 0.0000 \\ 
-  W.allcov & -0.0002 & 0.0012 & -0.0026 & -0.0002 & 0.0021 & -0.0002 & 0.0000 \\ 
-  W.allpts & -0.0289 & 0.0229 & -0.0739 & -0.0289 & 0.0161 & -0.0289 & 0.0000 \\ 
+(Intercept) & 0.4923 & 0.0681 & 0.3581 & 0.4923 & 0.6259 & 0.4924 & 0.0000 \\ 
+  logarea.t0 & 0.6750 & 0.0232 & 0.6292 & 0.6750 & 0.7203 & 0.6751 & 0.0000 \\ 
+  TreatmentNo\_shrub & 0.2161 & 0.0644 & 0.0897 & 0.2161 & 0.3425 & 0.2161 & 0.0000 \\ 
+  W.ARTR & -0.0002 & 0.0008 & -0.0019 & -0.0002 & 0.0014 & -0.0002 & 0.0000 \\ 
+  W.HECO & 0.0039 & 0.0063 & -0.0084 & 0.0039 & 0.0161 & 0.0039 & 0.0000 \\ 
+  W.POSE & -0.2641 & 0.0387 & -0.3400 & -0.2641 & -0.1883 & -0.2641 & 0.0000 \\ 
+  W.PSSP & -0.0073 & 0.0056 & -0.0183 & -0.0073 & 0.0038 & -0.0073 & 0.0000 \\ 
+  W.allcov & -0.0003 & 0.0012 & -0.0026 & -0.0003 & 0.0020 & -0.0003 & 0.0000 \\ 
+  W.allpts & -0.0254 & 0.0231 & -0.0709 & -0.0254 & 0.0200 & -0.0254 & 0.0000 \\ 
    \hline
@@
 % latex table generated in R 3.2.2 by xtable 1.8-2 package
-% Thu May 19 09:53:11 2016
+% Mon Sep 19 08:37:48 2016
 \begin{table}[ht]
 \centering
-\caption{Summary of fixed effects for the PSSP growth model} 
+\caption{Summary of fixed effects for the \textit{P. spicata} growth model} 
 \label{PSSPgrowth}
@@
   \hline
-(Intercept) & 0.3862 & 0.0784 & 0.2292 & 0.3867 & 0.5403 & 0.3878 & 0.0000 \\ 
-  logarea.t0 & 0.8299 & 0.0164 & 0.7974 & 0.8299 & 0.8621 & 0.8300 & 0.0000 \\ 
-  TreatmentNo\_shrub & 0.2767 & 0.0757 & 0.1281 & 0.2767 & 0.4252 & 0.2768 & 0.0000 \\ 
-  W.ARTR & -0.0024 & 0.0008 & -0.0039 & -0.0024 & -0.0008 & -0.0024 & 0.0000 \\ 
-  W.HECO & -0.0110 & 0.0070 & -0.0247 & -0.0110 & 0.0027 & -0.0110 & 0.0000 \\ 
-  W.POSE & -0.0111 & 0.0129 & -0.0365 & -0.0111 & 0.0142 & -0.0111 & 0.0000 \\ 
-  W.PSSP & -0.1498 & 0.0131 & -0.1754 & -0.1498 & -0.1242 & -0.1498 & 0.0000 \\ 
-  W.allcov & -0.0035 & 0.0011 & -0.0055 & -0.0035 & -0.0014 & -0.0035 & 0.0000 \\ 
-  W.allpts & -0.0498 & 0.0228 & -0.0946 & -0.0498 & -0.0051 & -0.0498 & 0.0000 \\ 
+(Intercept) & 0.3987 & 0.0660 & 0.2690 & 0.3987 & 0.5281 & 0.3988 & 0.0000 \\ 
+  logarea.t0 & 0.8249 & 0.0149 & 0.7956 & 0.8249 & 0.8543 & 0.8249 & 0.0000 \\ 
+  TreatmentNo\_shrub & 0.1948 & 0.0714 & 0.0546 & 0.1948 & 0.3349 & 0.1948 & 0.0000 \\ 
+  W.ARTR & -0.0025 & 0.0008 & -0.0041 & -0.0025 & -0.0010 & -0.0025 & 0.0000 \\ 
+  W.HECO & -0.0118 & 0.0072 & -0.0260 & -0.0118 & 0.0023 & -0.0118 & 0.0000 \\ 
+  W.POSE & -0.0067 & 0.0130 & -0.0323 & -0.0067 & 0.0189 & -0.0067 & 0.0000 \\ 
+  W.PSSP & -0.1480 & 0.0133 & -0.1740 & -0.1480 & -0.1220 & -0.1480 & 0.0000 \\ 
+  W.allcov & -0.0036 & 0.0011 & -0.0057 & -0.0036 & -0.0015 & -0.0036 & 0.0000 \\ 
+  W.allpts & -0.0522 & 0.0236 & -0.0984 & -0.0522 & -0.0060 & -0.0522 & 0.0000 \\ 
    \hline
@@
 \begin{table}
-\caption{Poa secunda growth with Artemisia canopy effect}
+\caption{\textit{Poa secunda} growth with \textit{Artemisia} canopy effect}
 \begin{center}
@@
 \hline
-(Intercept)                      & $0.50^{*}$        \\
-                                 & $[0.36;\ 0.63]$   \\
+(Intercept)                      & $0.49^{*}$        \\
+                                 & $[0.36;\ 0.62]$   \\
 logarea.t0                       & $0.68^{*}$        \\
-                                 & $[0.63;\ 0.73]$   \\
-TreatmentNo\_shrub               & $0.21^{*}$        \\
-                                 & $[0.06;\ 0.37]$   \\
+                                 & $[0.63;\ 0.72]$   \\
+TreatmentNo\_shrub               & $0.24^{*}$        \\
+                                 & $[0.09;\ 0.38]$   \\
 W.ARTR                           & $-0.00$           \\
@@
                                  & $[-0.01;\ 0.02]$  \\
-W.POSE                           & $-0.27^{*}$       \\
+W.POSE                           & $-0.26^{*}$       \\
+                                 & $[-0.34;\ -0.19]$ \\
+W.PSSP                           & $-0.01$           \\
+                                 & $[-0.02;\ 0.00]$  \\
+W.allcov                         & $-0.00$           \\
+                                 & $[-0.00;\ 0.00]$  \\
+W.allpts                         & $-0.02$           \\
+                                 & $[-0.07;\ 0.02]$  \\
+inARTR                           & $-0.06$           \\
+                                 & $[-0.28;\ 0.16]$  \\
+\hline
+AIC                              & 15641.80          \\
+BIC                              & 15734.29          \\
+Log Likelihood                   & -7806.90          \\
+Num. obs.                        & 5469              \\
+Num. groups: year                & 31                \\
+Var: year (Intercept)            & 0.11              \\
+Var: year logarea.t0             & 0.01              \\
+Cov: year (Intercept) logarea.t0 & -0.02             \\
+Var: Residual                    & 0.98              \\
+\hline
+\multicolumn{2}{l}{\scriptsize{$^*$ 0 outside the confidence interval}}
+\end{tabular}
+\label{table:POSEgrowth-inARTR}
+\end{center}
+\end{table}
+
+\begin{table}
+\caption{\textit{P. spicata} growth with \textit{Artemisia} canopy effect}
+\begin{center}
+\begin{tabular}{l c }
+\hline
+ & Model 1 \\
+\hline
+(Intercept)                      & $0.42^{*}$        \\
+                                 & $[0.29;\ 0.56]$   \\
+logarea.t0                       & $0.82^{*}$        \\
+                                 & $[0.79;\ 0.86]$   \\
+TreatmentNo\_shrub               & $0.24^{*}$        \\
+                                 & $[0.09;\ 0.40]$   \\
+W.ARTR                           & $-0.00^{*}$       \\
+                                 & $[-0.00;\ -0.00]$ \\
+W.HECO                           & $-0.01^{*}$       \\
+                                 & $[-0.03;\ -0.00]$ \\
+W.POSE                           & $-0.01$           \\
+                                 & $[-0.03;\ 0.02]$  \\
+W.PSSP                           & $-0.14^{*}$       \\
+                                 & $[-0.16;\ -0.11]$ \\
+W.allcov                         & $-0.00^{*}$       \\
+                                 & $[-0.01;\ -0.00]$ \\
+W.allpts                         & $-0.06^{*}$       \\
+                                 & $[-0.10;\ -0.01]$ \\
+inARTR                           & $-0.14$           \\
+                                 & $[-0.35;\ 0.08]$  \\
+\hline
+AIC                              & 15282.37          \\
+BIC                              & 15376.25          \\
+Log Likelihood                   & -7627.18          \\
+Num. obs.                        & 6039              \\
+Num. groups: year                & 31                \\
+Var: year (Intercept)            & 0.12              \\
+Var: year logarea.t0             & 0.01              \\
+Cov: year (Intercept) logarea.t0 & -0.03             \\
+Var: Residual                    & 0.71              \\
+\hline
+\multicolumn{2}{l}{\scriptsize{$^*$ 0 outside the confidence interval}}
+\end{tabular}
+\label{table:PSSPgrowth-inARTR}
+\end{center}
+\end{table}
+
+\begin{table}
+\caption{\textit{Poa secunda} growth with year-by-treatment interaction}
+\begin{center}
+\begin{tabular}{l c }
+\hline
+ & Model 1 \\
+\hline
+(Intercept)                      & $0.49^{*}$        \\
+                                 & $[0.37;\ 0.62]$   \\
+trtYears1                        & $0.06$            \\
+                                 & $[-0.17;\ 0.29]$  \\
+trtYears2                        & $0.24$            \\
+                                 & $[-0.01;\ 0.49]$  \\
+trtYears3                        & $0.23$            \\
+                                 & $[-0.05;\ 0.51]$  \\
+trtYears4                        & $0.25$            \\
+                                 & $[-0.07;\ 0.56]$  \\
+trtYears5                        & $0.45^{*}$        \\
+                                 & $[0.11;\ 0.79]$   \\
+logarea.t0                       & $0.68^{*}$        \\
+                                 & $[0.63;\ 0.72]$   \\
+W.ARTR                           & $-0.00$           \\
+                                 & $[-0.00;\ 0.00]$  \\
+W.HECO                           & $0.00$            \\
+                                 & $[-0.01;\ 0.02]$  \\
+W.POSE                           & $-0.26^{*}$       \\
                                  & $[-0.34;\ -0.19]$ \\
@@
                                  & $[-0.07;\ 0.02]$  \\
-inARTR                           & $-0.09$           \\
-                                 & $[-0.33;\ 0.14]$  \\
-\hline
-AIC                              & 15070.92          \\
-BIC                              & 15162.94          \\
-Log Likelihood                   & -7521.46          \\
-Num. obs.                        & 5286              \\
-Num. groups: year                & 30                \\
+\hline
+AIC                              & 15648.62          \\
+BIC                              & 15760.94          \\
+Log Likelihood                   & -7807.31          \\
+Num. obs.                        & 5469              \\
+Num. groups: year                & 31                \\
 Var: year (Intercept)            & 0.11              \\
@@
 Cov: year (Intercept) logarea.t0 & -0.02             \\
-Var: Residual                    & 0.97              \\
+Var: Residual                    & 0.98              \\
 \hline
@@
 \end{tabular}
-\label{table:POSEgrowth-inARTR}
+\label{table:POSEgrowth-byYr}
 \end{center}
@@
 \begin{table}
-\caption{\textit{P. spicata} growth with \textit{Artemisia} canopy effect}
+\caption{\textit{P. spicata} growth with year-by-treatment interaction}
 \begin{center}
@@
 \hline
-(Intercept)                      & $0.41^{*}$        \\
-                                 & $[0.27;\ 0.54]$   \\
-logarea.t0                       & $0.83^{*}$        \\
-                                 & $[0.79;\ 0.87]$   \\
-TreatmentNo\_shrub               & $0.35^{*}$        \\
-                                 & $[0.18;\ 0.52]$   \\
+(Intercept)                      & $0.42^{*}$        \\
+                                 & $[0.28;\ 0.55]$   \\
+trtYears1                        & $0.26$            \\
+                                 & $[-0.01;\ 0.54]$  \\
+trtYears2                        & $-0.05$           \\
+                                 & $[-0.32;\ 0.22]$  \\
+trtYears3                        & $0.44^{*}$        \\
+                                 & $[0.14;\ 0.73]$   \\
+trtYears4                        & $0.42^{*}$        \\
+                                 & $[0.12;\ 0.72]$   \\
+trtYears5                        & $-0.11$           \\
+                                 & $[-0.42;\ 0.20]$  \\
+logarea.t0                       & $0.82^{*}$        \\
+                                 & $[0.79;\ 0.86]$   \\
 W.ARTR                           & $-0.00^{*}$       \\
                                  & $[-0.00;\ -0.00]$ \\
-W.HECO                           & $-0.01$           \\
-                                 & $[-0.03;\ 0.00]$  \\
+W.HECO                           & $-0.01^{*}$       \\
+                                 & $[-0.03;\ -0.00]$ \\
 W.POSE                           & $-0.01$           \\
-                                 & $[-0.04;\ 0.01]$  \\
+                                 & $[-0.03;\ 0.02]$  \\
 W.PSSP                           & $-0.14^{*}$       \\
@@
                                  & $[-0.10;\ -0.01]$ \\
-inARTR                           & $-0.21$           \\
-                                 & $[-0.45;\ 0.03]$  \\
-\hline
-AIC                              & 14834.74          \\
-BIC                              & 14928.30          \\
-Log Likelihood                   & -7403.37          \\
-Num. obs.                        & 5904              \\
-Num. groups: year                & 30                \\
+\hline
+AIC                              & 15281.67          \\
+BIC                              & 15395.67          \\
+Log Likelihood                   & -7623.83          \\
+Num. obs.                        & 6039              \\
+Num. groups: year                & 31                \\
 Var: year (Intercept)            & 0.12              \\
@@
 Cov: year (Intercept) logarea.t0 & -0.03             \\
-Var: Residual                    & 0.70              \\
-\hline
-\multicolumn{2}{l}{\scriptsize{$^*$ 0 outside the confidence interval}}
-\end{tabular}
-\label{table:PSSPgrowth-inARTR}
-\end{center}
-\end{table}
-
-\begin{table}
-\caption{Poa secunda growth with year-by-treatment interaction}
-\begin{center}
-\begin{tabular}{l c }
-\hline
- & Model 1 \\
-\hline
-(Intercept)                      & $0.50^{*}$        \\
-                                 & $[0.36;\ 0.63]$   \\
-trtYears1                        & $0.05$            \\
-                                 & $[-0.18;\ 0.28]$  \\
-trtYears2                        & $0.25^{*}$        \\
-                                 & $[0.01;\ 0.50]$   \\
-trtYears3                        & $0.19$            \\
-                                 & $[-0.09;\ 0.47]$  \\
-trtYears4                        & $0.29$            \\
-                                 & $[-0.02;\ 0.60]$  \\
-logarea.t0                       & $0.68^{*}$        \\
-                                 & $[0.63;\ 0.73]$   \\
-W.ARTR                           & $-0.00$           \\
-                                 & $[-0.00;\ 0.00]$  \\
-W.HECO                           & $0.00$            \\
-                                 & $[-0.01;\ 0.02]$  \\
-W.POSE                           & $-0.27^{*}$       \\
-                                 & $[-0.34;\ -0.19]$ \\
-W.PSSP                           & $-0.01$           \\
-                                 & $[-0.02;\ 0.00]$  \\
-W.allcov                         & $-0.00$           \\
-                                 & $[-0.00;\ 0.00]$  \\
-W.allpts                         & $-0.03$           \\
-                                 & $[-0.07;\ 0.02]$  \\
-\hline
-AIC                              & 15076.15          \\
-BIC                              & 15181.32          \\
-Log Likelihood                   & -7522.08          \\
-Num. obs.                        & 5286              \\
-Num. groups: year                & 30                \\
-Var: year (Intercept)            & 0.11              \\
-Var: year logarea.t0             & 0.01              \\
-Cov: year (Intercept) logarea.t0 & -0.02             \\
-Var: Residual                    & 0.97              \\
-\hline
-\multicolumn{2}{l}{\scriptsize{$^*$ 0 outside the confidence interval}}
-\end{tabular}
-\label{table:POSEgrowth-byYr}
-\end{center}
-\end{table}
-
-\begin{table}
-\caption{P. spicata growth with year-by-treatment interaction}
-\begin{center}
-\begin{tabular}{l c }
-\hline
- & Model 1 \\
-\hline
-(Intercept)                      & $0.41^{*}$        \\
-                                 & $[0.27;\ 0.54]$   \\
-trtYears1                        & $0.25$            \\
-                                 & $[-0.02;\ 0.52]$  \\
-trtYears2                        & $-0.06$           \\
-                                 & $[-0.33;\ 0.21]$  \\
-trtYears3                        & $0.44^{*}$        \\
-                                 & $[0.15;\ 0.73]$   \\
-trtYears4                        & $0.53^{*}$        \\
-                                 & $[0.23;\ 0.84]$   \\
-logarea.t0                       & $0.83^{*}$        \\
-                                 & $[0.79;\ 0.87]$   \\
-W.ARTR                           & $-0.00^{*}$       \\
-                                 & $[-0.00;\ -0.00]$ \\
-W.HECO                           & $-0.01$           \\
-                                 & $[-0.03;\ 0.00]$  \\
-W.POSE                           & $-0.01$           \\
-                                 & $[-0.04;\ 0.01]$  \\
-W.PSSP                           & $-0.14^{*}$       \\
-                                 & $[-0.16;\ -0.11]$ \\
-W.allcov                         & $-0.00^{*}$       \\
-                                 & $[-0.01;\ -0.00]$ \\
-W.allpts                         & $-0.06^{*}$       \\
-                                 & $[-0.10;\ -0.01]$ \\
-\hline
-AIC                              & 14834.13          \\
-BIC                              & 14941.06          \\
-Log Likelihood                   & -7401.06          \\
-Num. obs.                        & 5904              \\
-Num. groups: year                & 30                \\
-Var: year (Intercept)            & 0.12              \\
-Var: year logarea.t0             & 0.01              \\
-Cov: year (Intercept) logarea.t0 & -0.02             \\
-Var: Residual                    & 0.69              \\
+Var: Residual                    & 0.71              \\
 \hline
@@
 % latex table generated in R 3.2.2 by xtable 1.8-2 package
-% Mon May 23 12:08:13 2016
+% Mon Sep 19 08:37:48 2016
 \begin{table}[ht]
 \centering
-\caption{Summary of fixed effects for the recruitment model} 
+\caption{Summary of fixed effects for the recruitment model (symbols correspond to Eqns. \ref{eqn:recrDataModel} and \ref{eqn:recrProcessModel})} 
 \label{table:recruitment}
@@
   \hline
-intcpt.mu[1] & 0.4020 & 0.7346 & -1.1469 & 1.7240 & 1.1164 &    18 \\ 
-  intcpt.mu[2] & 3.3186 & 0.5271 & 2.3009 & 4.3691 & 1.0041 &   500 \\ 
-  intcpt.mu[3] & 3.2278 & 0.3828 & 2.4909 & 3.9860 & 1.0021 &  2000 \\ 
-  intcpt.mu[4] & 2.8859 & 0.4080 & 2.0990 & 3.7150 & 1.0710 &    28 \\ 
-  intcpt.trt[2,2] & -0.4870 & 0.5332 & -1.5241 & 0.5814 & 1.0005 &  2000 \\ 
-  intcpt.trt[2,3] & -1.3272 & 0.3844 & -2.1322 & -0.5825 & 1.0018 &  2000 \\ 
-  intcpt.trt[2,4] & -0.2148 & 0.3154 & -0.8458 & 0.3810 & 1.0015 &  1500 \\ 
-  intcpt.trt[3,1] & -26.7881 & 18.2985 & -69.6230 & -3.1958 & 1.0010 &  2000 \\ 
-  dd[1,1] & -0.5972 & 0.1128 & -0.7979 & -0.3581 & 1.1004 &    23 \\ 
-  dd[1,2] & 0.0790 & 0.0646 & -0.0480 & 0.2020 & 1.0165 &   110 \\ 
-  dd[1,3] & 0.0187 & 0.0408 & -0.0625 & 0.0989 & 1.0033 &   530 \\ 
-  dd[1,4] & 0.1143 & 0.0473 & 0.0245 & 0.2090 & 1.0319 &    86 \\ 
-  dd[2,1] & -0.4855 & 0.1619 & -0.7977 & -0.1613 & 1.0015 &  2000 \\ 
-  dd[2,2] & -1.7763 & 0.1300 & -2.0320 & -1.5270 & 1.0158 &   120 \\ 
-  dd[2,3] & 0.0335 & 0.0899 & -0.1377 & 0.2058 & 1.0031 &   570 \\ 
-  dd[2,4] & -0.3687 & 0.0932 & -0.5474 & -0.1816 & 1.0024 &   810 \\ 
-  dd[3,1] & -0.7223 & 0.3736 & -1.4671 & -0.0399 & 1.0139 &   140 \\ 
-  dd[3,2] & -0.0166 & 0.2475 & -0.5859 & 0.4237 & 1.0092 &   320 \\ 
-  dd[3,3] & -1.9665 & 0.2122 & -2.4881 & -1.6289 & 1.0226 &  1000 \\ 
-  dd[3,4] & -0.1495 & 0.1629 & -0.4771 & 0.1599 & 1.0006 &  2000 \\ 
-  dd[4,1] & -0.1196 & 0.3023 & -0.7493 & 0.4611 & 1.0325 &    53 \\ 
-  dd[4,2] & -0.3597 & 0.2374 & -0.8784 & 0.0571 & 1.0034 &   800 \\ 
-  dd[4,3] & 0.1429 & 0.1557 & -0.1669 & 0.4593 & 1.0019 &  1300 \\ 
-  dd[4,4] & -1.7602 & 0.1818 & -2.0890 & -1.3710 & 1.0532 &    41 \\ 
-  theta[1] & 0.6247 & 0.0732 & 0.4926 & 0.7761 & 1.0030 &   600 \\ 
-  theta[2] & 1.1588 & 0.1500 & 0.8948 & 1.4830 & 1.0008 &  2000 \\ 
-  theta[3] & 1.1570 & 0.1085 & 0.9687 & 1.3840 & 1.0015 &  2000 \\ 
-  theta[4] & 1.0910 & 0.1053 & 0.8967 & 1.3130 & 1.0017 &  2000 \\ 
-  u[1] & 0.8209 & 0.1188 & 0.4884 & 0.9571 & 1.0663 &    60 \\ 
-  u[2] & 0.9995 & 0.0005 & 0.9981 & 1.0000 & 1.0028 &  2000 \\ 
-  u[3] & 0.6802 & 0.1945 & 0.1677 & 0.9186 & 1.0775 &   260 \\ 
-  u[4] & 0.6791 & 0.1808 & 0.2467 & 0.9301 & 1.0028 &   650 \\ 
+ $\gamma$[1] & 0.3341 & 0.7071 & -1.1262 & 1.6391 & 1.0018 &  1100 \\ 
+  $\gamma$[2] & 3.4488 & 0.4468 & 2.5010 & 4.3401 & 1.0904 &    25 \\ 
+  $\gamma$[3] & 3.2440 & 0.3650 & 2.4840 & 3.9190 & 1.0536 &    34 \\ 
+ $\gamma$[4] & 2.9054 & 0.3689 & 2.1820 & 3.6160 & 1.0149 &   110 \\ 
+  $\chi$[2,2] & -0.0953 & 0.4093 & -0.8854 & 0.7570 & 1.0010 &  2000 \\ 
+ $\chi$[2,3] & -1.2787 & 0.3325 & -1.9370 & -0.6392 & 1.0041 &   420 \\ 
+   $\chi$[2,4] & 0.0951 & 0.2617 & -0.4121 & 0.6063 & 1.0023 &   810 \\ 
+  $\chi$[3,1] & -1.4366 & 0.8544 & -3.1491 & 0.1132 & 1.0006 &  2000 \\ 
+  $\omega$[1,1] & -0.5881 & 0.1018 & -0.7639 & -0.3720 & 1.0197 &    82 \\ 
+  $\omega$[1,2] & 0.0635 & 0.0651 & -0.0528 & 0.1997 & 1.0329 &    59 \\ 
+  $\omega$[1,3] & 0.0220 & 0.0422 & -0.0610 & 0.1040 & 1.0036 &   920 \\ 
+  $\omega$[1,4] & 0.1164 & 0.0456 & 0.0347 & 0.2123 & 1.0197 &    96 \\ 
+  $\omega$[2,1] & -0.4800 & 0.1522 & -0.7746 & -0.1906 & 1.0029 &  2000 \\ 
+  $\omega$[2,2] & -1.7368 & 0.1297 & -1.9841 & -1.4819 & 1.0006 &  2000 \\ 
+  $\omega$[2,3] & 0.0388 & 0.0865 & -0.1287 & 0.2138 & 1.0152 &   100 \\ 
+  $\omega$[2,4] & -0.3682 & 0.0926 & -0.5434 & -0.1881 & 1.0005 &  2000 \\ 
+  $\omega$[3,1] & -0.6589 & 0.3342 & -1.3131 & -0.0155 & 1.0096 &   280 \\ 
+  $\omega$[3,2] & -0.1021 & 0.1981 & -0.4828 & 0.2936 & 1.0089 &   210 \\ 
+  $\omega$[3,3] & -1.8943 & 0.1611 & -2.2060 & -1.5850 & 1.0084 &  2000 \\ 
+  $\omega$[3,4] & -0.1953 & 0.1508 & -0.4968 & 0.0995 & 1.0074 &  2000 \\ 
+  $\omega$[4,1] & -0.1867 & 0.3000 & -0.7608 & 0.4486 & 1.0099 &   190 \\ 
+  $\omega$[4,2] & -0.3746 & 0.1896 & -0.7224 & -0.0115 & 1.0453 &    55 \\ 
+  $\omega$[4,3] & 0.1131 & 0.1437 & -0.1583 & 0.4170 & 1.0321 &    55 \\ 
+  $\omega$[4,4] & -1.7379 & 0.1549 & -2.0400 & -1.4409 & 1.0116 &  1400 \\ 
+   $\theta$[1] & 0.6198 & 0.0740 & 0.4835 & 0.7766 & 1.0008 &  2000 \\ 
+   $\theta$[2] & 1.1197 & 0.1428 & 0.8666 & 1.4450 & 1.0008 &  2000 \\ 
+   $\theta$[3] & 1.1718 & 0.1068 & 0.9685 & 1.3890 & 1.0008 &  2000 \\ 
+  $\theta$[4] & 1.0961 & 0.1060 & 0.9009 & 1.3280 & 1.0022 &   890 \\ 
+  $p$[1] & 0.8137 & 0.1228 & 0.4813 & 0.9525 & 1.0399 &    60 \\ 
+  $p$[2] & 0.9996 & 0.0005 & 0.9984 & 1.0000 & 1.0007 &  2000 \\ 
+  $p$[3] & 0.7906 & 0.1288 & 0.4273 & 0.9382 & 1.0176 &  2000 \\ 
+  $p$[4] & 0.7502 & 0.1480 & 0.3920 & 0.9453 & 1.0150 &  2000 \\ 
    \hline
@@
  \label{fig:CompKernels}
+ \end{figure}
+
+  \begin{figure}[h]
+  \centering
+  \includegraphics[width=0.7\textwidth]{climate}
+  \caption{Annual precipitation (a) and mean temperature (b) during the period of the experiment, shown against the long-term (1927-2016) means (solid blue and red lines) and 5\% and 95\% quantiles (dashed lines).
+}
+  \label{fig:climate}
   \end{figure}

### 2016-09-20 removals/climate_fig.r

@@ -1,0 +1,39 @@
+
+dat <- read.csv("../../driversdata/data/idaho_modern/climateData/monthly_climate_from_daily.csv")
+
+dat$climYr <- ifelse(dat$month>9,dat$year+1,dat$year)
+
+dat <- subset(dat,climYr>1926) # remove incomplete year
+
+annualPpt <- aggregate(dat[,c("TPPT")],by=list(dat$climYr),FUN="sum")
+names(annualPpt) <- c("climYr","precip")
+annualPpt[,2] <- annualPpt[,2]*25.4 # convert in to mm
+meanPpt <- mean(annualPpt[,2],na.rm=T)
+ppt.90 <- quantile(annualPpt[,2],c(0.05,0.95))
+modernPpt <- subset(annualPpt,climYr>2010)
+
+annualTemp <- aggregate(dat[,c("TMEAN")],by=list(dat$climYr),FUN="mean")
+names(annualTemp) <- c("climYr","temperature")
+annualTemp[,2] <- (annualTemp[,2]-32)*5/9 # convert F to C
+meanTemp <- mean(annualTemp[,2],na.rm=T)
+temp.90 <- quantile(annualTemp[,2],c(0.05,0.95),na.rm=T)
+modernTemp <- subset(annualTemp,climYr>2010)
+
+# make figure
+png("climate.png",height=3.5,width=8,res=400,units="in")
+
+par(mfrow=c(1,2),tcl=-0.2,mgp=c(2,0.5,0),mar=c(3,4,2,1))
+
+plot(modernPpt,type="o",pch=16,ylim=c(150,500),xlab="Year",ylab="Annual precipitation (mm)")
+abline(h=meanPpt,col="dodgerblue3",lwd=1)
+abline(h=ppt.90[1],col="dodgerblue3",lwd=1,lty="dashed")
+abline(h=ppt.90[2],col="dodgerblue3",lwd=1,lty="dashed")
+mtext("(a)",side=3,line=0.5,adj=0)
+
+plot(modernTemp,type="o",pch=16,ylim=c(4,9.5),xlab="Year",ylab="Mean temperature (C)")
+abline(h=meanTemp,col="red3",lwd=1)
+abline(h=temp.90[1],col="red3",lwd=1,lty="dashed")
+abline(h=temp.90[2],col="red3",lwd=1,lty="dashed")
+mtext("(b)",side=3,line=0.5,adj=0)
+
+dev.off()

### 2016-09-20 removals/removal_analysis_wrapper.r

@@ -17,6 +17,9 @@ library(lme4)
 statsOutput <- paste0(getwd(),"/stats_tables.tex")
 source("treatment_trends_removals.r")
 
+# make climate figure
+source("climate_fig.r")
+
 # clean up
 tmp=ls() ; tmp=tmp[tmp!="root" & tmp!="statsOutput"]
 rm(list=tmp)

### 2016-09-20 removals/manuscript/removal_ms.tex
Manuscript edits

@@
 The version of the model with maximum removal treatment effects (removal coefficients set to the 95\% confidence limit furthest from zero), were qualitatively similar, but further amplified the contrast between predictions of the baseline and removal models for and \textit{P.
-spicata} REF MAX CI FIG
+spicata} (Fig.
+\ref{fig:IBM1step-maxCI}).
 
@@
  \end{figure}
-
-% add max CI figure in SI
 
@@
 Running a version of the model with maximum removal treatment effects (removal coefficients set to the 95\% confidence limit furthest from zero), resulted in the same qualitative outcome, but a larger increase in \textit{P.
-spicata} % (Fig.
-\ref{fig:IPMresults-maxCI})
-
-% add max CI figure in SI
+spicata} (Fig.
+\ref{fig:IPMresults-maxCI}).
 
@@
 For three of the four species, we find little or no evidence that models based on observational data underestimate the intensity of interspecific competition.
-Removal treatment effects were never significant for \textit{H.
-comata}; our baseline model, which accounts for local neighborhood composition, adequately explained vital rate responses to \textit{A.
-tripartita} removal.
-The same was true for the survival and growth models for \textit{A.
-tripartita}, but at the recruitment stage the grass removal treatment effect was significant and negative, implying that our baseline model might overestimate--not underestimate--competitive effects of the perennial grasses on \textit{Artemisia} recruitment.
-\textit{Poa secunda} showed mixed responses to \textit{Artemisia} removal: removal effects were not significant at the survival stage, were positive at the growth stage, and negative at the recruitment stage (where the estimated effect of \textit{Artemisia} on \textit{Poa} is positive ).
+Removal treatment effects were never significant for \textit{A.
+tripartita} or \textit{H.
+comata}; our baseline model, which accounts for local neighborhood composition, adequately explained vital rate responses to  removal.
+\textit{Poa secunda} showed mixed responses to \textit{Artemisia} removal: removal effects were not significant at the survival stage, were positive at the growth stage, and negative at the recruitment stage (where the estimated effect of \textit{Artemisia} on \textit{Poa} is positive).
 At the population level, models with and without removal effects projected similar patterns of short (Fig.
-\ref{fig:CoverChange}) and long-term (Fig.
+\ref{fig:IBM1step}) and long-term (Fig.
 \ref{fig:IPMresults}) dynamics for \textit{H.
@@
 For \textit{Poa secunda}, the model with removal effects predicted higher population growth over the short but not the long-term, perhaps because the short-term projections mostly reflect the positive effect of \textit{Artemisia} removal on growth, while the long-term equilibrium are also sensitive to the negative effects of removal on recruitment.
-Although the baseline model did not always successfully predict removal effects on vital rates, it does appear to successfully capture the population-level consequences of removals on these three species and did not severly underestimate the magnitude of competitive release.
+Although the baseline model did not always successfully predict removal effects on vital rates, it does appear to successfully capture the population-level consequences of removals on these three species and did not severely underestimate the magnitude of competitive release.
 
@@
 \ref{fig:W-by-treatment}).
-Perhaps these pre-treatment differences reflect subtle differences in edaphic and environmental factors which promote \textit{P.
+Perhaps these pre-treatment differences reflect subtle differences in edaphic and environmental factors which could promote \textit{P.
 spicata} growth.
@@
 
-The small competitive release of many species in our system implies that classic resource competition 
-is not the primary factor determining community structure. 
-\comment{SPE: I think this depends on what you mean by ``classic resource competition''.
-If you mean Tilman-style competition for one limiting resource distributed homogeneously, then I agree.
-But a MacArthur-Levins model with resource spectrum and distinct utilization curves could also be called classic resource competition,
-and that model will have weak to nonexistent competitive release if niche overlap is low.} 
-Interactions with above- or belowground enemies can also regulate populations and, given enough species-specificity in the effects of these enemies (Clark paper, Janzen, Connell), decouple interspecific dynamics.
-Species-specific responses to spatial and temporal environmental variation can also weaken interspecific interactions (Chesson et al.
+The small competitive release of many species in our system implies that each of the four dominant species we studied are limited by quite different factors.
+This low niche overlap could reflect strong resource partitioning (Tilman 1982), perhaps combined with variation in the times and locations at which species are taking up those resources.
+Species-specific responses to spatial and temporal variation in non-resource environmental variables can also weaken interspecific interactions (Chesson et al.
 2000), though in a previous study we failed to find evidence of a temporal storage effect in this community (Adler et al.
 2009).
+Interactions with above- or belowground enemies can regulate populations and, given enough species-specificity in the effects of these enemies (Clark paper, Janzen, Connell), decouple interspecific dynamics.
 More likely, plants may alter the environment, and even resource availability, in ways that facilitate their neighbors.
@@
 2012, Kleinhesselink and Adler 2015).
-Similarly, differences among species in relative abundance must reflect variation in the strength of intraspecific density dependence rather than variation in the strength of interspecific interactions (Yenni et al.
+Similarly, differences among species in relative abundance should reflect variation in the strength of intraspecific density dependence more than variation in the strength of interspecific interactions (Yenni et al.
 2015).
 In such a system, effective management and conservation requires understanding the sources and species-specificity of intraspecific density dependence.
-In fact, an experiment to test our model's skill at predicting release from intraspecific competition may be even more important than the test of interspecific competitive release described here.
+In fact, an experiment to test our model's skill at predicting release from intraspecific competition could be more important than the test of interspecific competitive release described here.
 
@@
 If they are rare because they occupy unique, but narrow, niches, interspecific interactions would still be weak.
-Determining where interspecific interactions and competitive release are strong, and where they are not, remains a fascinating research problem.
-\comment{SPE: Equally interesting, I think, is the question of why interspecific
-interactions are so weak (when they are), as opposed to what you would expect from ``all plant species compete for water, light, and a few mineral resources''.}
+Determining why interspecific interactions and competitive release are strong at some times and places and why they are weak in other situations remains a fascinating research problem.
 
@@
 
+ \begin{figure}[tbp]
+ \centering
+ \includegraphics[width=1\textwidth]{cover_projections_1step_maxCI}
+ \caption{Observed and predicted cover for each of the four modeled species assuming maximum effects of removal.
+Solid lines show mean observed cover averaged across control (blue) and removal treatment (red) quadrats.
+Dashed lines show one-step-ahead predictions from the baseline IBM (no removal treatment coefficients), averaged across quadrats.
+Dotted lines show one-step-ahead predictions from an IBM that includes removal treatment coefficients representing the outermost values of the 95\% confidence intervals.
+}
+ \label{fig:IBM1step-maxCI}
+ \end{figure}
+
+  \begin{figure}[tbp]
+  \centering
+  \includegraphics[width=1\textwidth]{boxplots-maxCI}
+  \caption{Equilibrium cover of the four dominant species simulated by the IPM with maximum removal treatment effects.
+Boxplots show interannual variation reflecting random year effects.
+Gray boxes show cover simulated by a model with \textit{A.
+tripartita} present and no removal treatment coefficients, brown boxes show results from the same model but with \textit{A.
+tripartita} cover set to zero (a species removal), green boxes show results from a model with  \textit{A.
+tripartita} set to zero and (mean) removal treatment coefficients included, and blue boxes show results with removal treatments coefficients assigned the outermost value of the 95\% confidence intervals.
+}
+  \label{fig:IPMresults-maxCI}
+  \end{figure}
+
 \end{document}

### 2016-09-21 removals/manuscript/removal_ms.tex
More ms revisions, added in references

@@
 \item \emph{Synthesis}: Although models based on observational data did not perfectly predict species responses to removal, overall the results increase our confidence that competitive release is in fact small for at least three of our four study species.
-The implication is that \new{interspecific} resource competition is not the primary factor determining the dynamics of this community.
+The implication is that interspecific resource competition is not the primary factor determining the dynamics of this community.
 
@@
 
-Stable coexistence requires that intraspecific interactions must limit populations more than interspecific interactions (Chesson 2000).
+Stable coexistence requires that intraspecific interactions must limit populations more than interspecific interactions \citep{chesson_mechanisms_2000}.
 Whenever we study populations of species that appear to be coexisting stably, such as populations that have co-occurred in close proximity for extended periods, we should expect intraspecific limitations to be stronger than interspecific limitations.
@@
 We motivate research on plant species coexistence by noting that all plant species compete for water, light, and a few mineral resources (e.g.
-Silvertown 2004).
-Resource competition plays a central role in both Tilman's mechanistic resource competition models (Tilman 1982) and Grime's Competitor-Stress tolerator-Ruderal strategies (Grime).
-Neutral theory (Hubbell 2001) goes even further, assuming that interspecific competition is exactly as strong as interspecific competition.
-Recent work has emphasized the role of interspecific facilitation in some plant communities (Calloway?
-Maestre?), but it is unclear whether this research has shaken the default assumption that plant species compete intensely.
+\citealt{silvertown_plant_2004}).
+Resource competition plays a central role in both Tilman's mechanistic resource competition models \citep{tilman_resource_1982} and Grime's Competitor-Stress tolerator-Ruderal strategies \citep{grime_plant_1979}.
+Neutral theory \citep{hubbell_unified_2001} goes even further, assuming that interspecific competition is exactly as strong as interspecific competition.
+Recent work has emphasized the role of interspecific facilitation in some plant communities \citep{he_global_2013,brooker_facilitation_2008}, but it is unclear whether this research has shaken the default assumption that plant species compete intensely.
 
@@
 Plant ecologists have conducted dozens of removal experiments and have observed almost every possible result.
-In some cases, competitive release is complete: remaining species fully compensate for removed biomass (Leps 1999, Jutila and Grace 2002).
-In other cases, the degree of competitive release depended on the functional group or dominance of the removed and remaining species (Sala et al.
-1989, Belsky 1992, Symstad and Tilman 2001, Smith and Knapp 2003).
-Finally, even negative effects of removals have been observed (Keddy 1989, Gilbert et al.
-2009), consistent with the recent facilitation literature.
+In some cases, competitive release is complete: remaining species fully compensate for removed biomass \citep{leps_nutrient_1999,jutila_effects_2002}.
+In other cases, the degree of competitive release depended on the functional group or dominance of the removed and remaining species \citep{smith_dominant_2003,sala_resource_1989,belsky_effects_1992}.
+Finally, even negative effects of removals have been observed \citep{keddy_effects_1989,gilbert_dominant_2009}, consistent with the recent facilitation literature.
 Results from removal experiments do come with caveats.
-Experiments typically do not run long enough to document long-term outcomes, responses may only be measured at one life stage whereas population responses depend on the whole life cycle, and removal treatments require disturbance which may cause a variety of impacts unrelated to resource competition (Aarssen and Epp 1990).
+Experiments typically do not run long enough to document long-term outcomes, responses may only be measured at one life stage whereas population responses depend on the whole life cycle, and removal treatments require disturbance which may cause a variety of impacts unrelated to resource competition \citep{aarssen_neighbour_1990}.
 
 An alternative approach to studying competitive release is to fit multispecies population models to long-term, observational data, and then simulate population responses to removals.
-Applying this approach to arid and semiarid plant communities, we found evidence of strong intraspecific limitation and very weak interspecific interactions, leading to very stable coexistence and implying little competitive release \citep{Adler:2010fu} ( Chu and Adler 2015).
+Applying this approach to arid and semiarid plant communities, we found evidence of strong intraspecific limitation and very weak interspecific interactions, leading to very stable coexistence and implying little competitive release \citep{adler_coexistence_2010,chu_large_2015}.
 Since we only studied the most common, co-occurring species, we expected to find evidence of stable coexistence, but we were surprised that interspecific interactions and species' responses to the loss of a dominant competitor were so weak.
@@
 In most years until 1957, all individual plants in each quadrat were mapped using a pantograph (Blaisdell 1958).
-The historical data set is public and available online \citep*{zachmann-etal-2010}.
+The historical data set is public and available online \citep{zachmann_mapped_2010}.
 In 2007, we located 14 of the original quadrats, all of which are inside permanent livestock exclosures, and resumed annual mapped censusing using the traditional pantograph method.
@@
 Our approach tracks genets, which may be composed of multiple polygons, as they fragment and/or coalesce.
-Each mapped polygon is classified as a surviving genet or a new recruit based on its spatial location relative to genets present in previous years (Lauenroth and Adler 2008).
-We modeled vital rates using data from 22 year-to-year transitions between 1926 and 1957, and eight year-to-year transitions from 2007 to 2015.
+Each mapped polygon is classified as a surviving genet or a new recruit based on its spatial location relative to genets present in previous years \citep{lauenroth_demography_2008}.
+We modeled vital rates using data from 22 year-to-year transitions between 1926 and 1957, and nine year-to-year transitions from 2007 to 2016.
 Only four quadrates were observed for the first two transitions in the 1920's, while at least 14 quadrats were observed for all subsequent transitions.
@@
 The response variable for this analysis was year-to-year change in cover at the quadrat scale, calculated as $log(Cover_{t+1}/Cover_t)$.
-We tested for the effect of removal treatment on this measure of cover change with a linear mixed effects model implemented in the \textit{lme4} package (Bates et al.
-2015) of \textit{R}, including year and quadrat as random effects.
+We tested for the effect of removal treatment on this measure of cover change with a linear mixed effects model implemented in the \textit{lme4} package \citep{Bates2015} of \textit{R}, including year and quadrat as random effects.
 We considered the removal treatment effect to be significant when the 95\% confidence interval on the estimated coefficient did not overlap zero.
@@
 
-Our regression models for survival, growth and recruitment follow previous work (Adler et al.
-2010, Chu and Adler 2015).
+Our regression models for survival, growth and recruitment follow previous work \citep{adler_coexistence_2010,chu_large_2015}.
 We model the survival probability of an individual genet as a function of genet size, the neighborhood-scale crowding experienced by the genet from both conspecific and heterospecific genets, temporal variation among years, and permanent spatial variation among groups of quadrats (`group'; here means a set of nearby quadrats located within one pasture or grazing exclosure).
@@
 We model the crowding experienced by a focal genet as a function of the distance to and size of neighbor genets.
-In previous work, we assumed that the decay of crowding with neighbor distance followed a Gaussian function (Chu and Adler 2015), but here we use a data-driven approach (Teller et al.
-2016).
+In previous work, we assumed that the decay of crowding with neighbor distance followed a Gaussian function \citep{chu_large_2015}, but here we use a data-driven approach \citep{teller_linking_2016}.
 We model the crowding experienced by genet $i$ of species $j$ from neighbors of species $m$ as the sum of neighbor areas across a set of concentric annuli, $k$, centered at the plant,
@@
 and returns an approximate AIC value and model degrees of freedom ($df$) for survival and growth combined. We used the $\vec{b}$ values at the smoothest 
-local minimum of AIC as a function of $df$, as in Teller et al. (2016). This approach assumes that one measure of crowding affects 
+local minimum of AIC as a function of $df$, as in \cite{teller_linking_2016}. This approach assumes that one measure of crowding affects 
 survival and growth. In addition, for fitting the kernels we assumed that survival and growth depended only on intraspecific crowding, and thus only fitted the
-within-species competition kernels $F_{jj}$. Based on previous work (Adler et al. 2010), we set all $F_{mj}$ equal to $F_{jj}$, meaning that 
+within-species competition kernels $F_{jj}$. Based on previous work \citep{adler_coexistence_2010}, we set all $F_{mj}$ equal to $F_{jj}$, meaning that 
 the within-species competition kernel for species $j$ is also used to determine the effect of all other species on species $j$.
@@
 All genets in a quadrat were included in calculating $W$, but plants located within 5 cm of quadrat edges were not used in fitting. 
-We fit the models using the \textit{R-INLA} package for \texttt{R} (Rue et al. 20xx).
+We fit the models using the \textit{R-INLA} package for \texttt{R} \citep{rue_approximate_2009}.
 
@@
 $\chi$ is the removal treatment effect, $\boldsymbol{\omega}$ is a vector of coefficients that determine the strength of intra- and interspecific density-dependence, and $\boldsymbol{C'}$ is the vector of ``effective'' cover of each species in the community.
-Eqn (\ref{eqn:recrProcessModel}) is essentially the Ricker model for discrete time population growth (Ricker 1954), except that we use the square root of cover in the density-dependent term.
-We previously used a Ricker equation (Adler et al.
-2010), but using the square-root transformation instead gave a lower Deviance Information Criterion (DIC) value (Spiegelhalter et al.
-2002) for the recruitment model.
-Following previous work (Adler et al.
-2010), we treated year and group variables as random factors, allowing intercepts to vary among years and spatial locations.
+Eqn (\ref{eqn:recrProcessModel}) is essentially the Ricker model for discrete time population growth \citep{ricker_stock_1954}, except that we use the square root of cover in the density-dependent term.
+We previously used a Ricker equation \citep{adler_coexistence_2010}, but using the square-root transformation instead gave a lower Deviance Information Criterion (DIC) value \citep{spiegelhalter_bayesian_2002} for the recruitment model.
+Following previous work \citep{adler_coexistence_2010}, we treated year and group variables as random factors, allowing intercepts to vary among years and spatial locations.
 
@@
 
-We ran Markov Chain Monte Carlo (MCMC) simulations in WinBUGS 1.4 (Lunn et al.
-2000) to estimate the recruitment model parameters.
+We ran Markov Chain Monte Carlo (MCMC) simulations in WinBUGS 1.4 \citep{lunn_winbugs_2000} to estimate the recruitment model parameters.
 Each model was run for 20,000 iterations and two independent chains with different initial values for parameters.
 We discarded the initial 10,000 samples.
-Convergence was observed graphically for all parameters, and confirmed by the Brooks-Gelman statistic (Brooks and Gelman 1998).
+Convergence was observed graphically for all parameters, and confirmed by the Brooks-Gelman statistic \citep{brooks_general_1998}.
 
@@
 measure cannot be extended to the IPM, which tracks individual size but not location. Instead we used a 
-mean-field approximation that captures the essential features of spatially-explicit neighborhood interactions (Adler et al. 2010). 
+mean-field approximation that captures the essential features of spatially-explicit neighborhood interactions \citep{adler_coexistence_2010}. 
 We found that in both the observed data and IBM simulations, heterospecific individuals were approximately randomly distributed with respect to each other, 
@@
 were segregated from each other. The overdispersion of large conspecific genets was introduced into the IPM 
-with a 'no-overlap' rule (Adler et al. 2010, 2012). In \ref{suppMethods} we explain how the spline competition kernels with 
-the no-overlap rule kernels were used to compute neigbhorhood crowding index as a function of individual size for the IPM.
+with a 'no-overlap' rule \citep{adler_coexistence_2010}.
+In \ref{suppMethods} we explain how the spline competition kernels with the no-overlap rule kernels were used to compute neigbhorhood crowding index as a function of individual size for the IPM.
 
@@
 in eqn. (\ref{eqn:recrDataModel}) gives the expected 
-area of new recruits per quadrat, per unit area of parents. To incorporate this recruitment function into the IPM, 
-we assumed that individual fecundity increases linearly with individual area (equal to $e^u$ for a size-$u$ individual). The recruitment kernel is thus
-$R_j(v_j,u_j,\vec{n})=c_{0,j}(v_j)e^{u_j}\Phi_j$ (Adler et al. 2010) where $c_{0,j}$ is the
-size distribution of new recruits. $\Phi_j$ is calculated from eqn. (\ref{eqn:recrDataModel}) with total cover values
-computed from the population densities as $C'_{j,t} = \int e^u n_j(u,t) du$.
+area of new recruits per quadrat, per unit area of parents.
+To incorporate this recruitment function into the IPM, we assumed that individual fecundity increases linearly with individual area (equal to $e^u$ for a size-$u$ individual).
+The recruitment kernel is thus
+$R_j(v_j,u_j,\vec{n})=c_{0,j}(v_j)e^{u_j}\Phi_j$ \citep{adler_coexistence_2010} where $c_{0,j}$ is the
+size distribution of new recruits. $\Phi_j$ is calculated from eqn. (\ref{eqn:recrDataModel}) with total cover values computed from the population densities as $C'_{j,t} = \int e^u n_j(u,t) du$.
 
@@
 To incorporate temporal environmental variation, at each time step we randomly selected one of the 30 sets of random year effects for the survival, growth and recruitment models (``kernel selection'' in the terminology of Metcalf et al.
-2016))).
+2016).
 We calculated the equilibrium cover for each species by simulating an additional 2000 time steps and averaging each species' cover over this period.
@@
 \end{figure}
-
-%$  $ FIX COVERCHANGE LEGEND
 
@@
 
-Our population models based on historical, observational data reveal weak interspecific interactions (Adler et al.
-2010, Chu and Adler 2015) and, as a result, predict very little competitive release among the four dominant species in a sagebrush steppe.
+Our population models based on historical, observational data reveal weak interspecific interactions \citep{adler_coexistence_2010,chu_large_2015} and, as a result, predict very little competitive release among the four dominant species in a sagebrush steppe.
 Despite the fact that previous removal studies have reported variable results, ranging from negative effects of removal on remaining species to complete competitive release, we were suspicious of our model's predictions and conducted a removal experiment to test them.
@@
 This low niche overlap could reflect strong resource partitioning (Tilman 1982), perhaps combined with variation in the times and locations at which species are taking up those resources.
-Species-specific responses to spatial and temporal variation in non-resource environmental variables can also weaken interspecific interactions (Chesson et al.
-2000), though in a previous study we failed to find evidence of a temporal storage effect in this community (Adler et al.
-2009).
-Interactions with above- or belowground enemies can regulate populations and, given enough species-specificity in the effects of these enemies (Clark paper, Janzen, Connell), decouple interspecific dynamics.
+Species-specific responses to spatial and temporal variation in non-resource environmental variables can also weaken interspecific interactions \citep{chesson_mechanisms_2000}, though in a previous study we failed to find evidence of a temporal storage effect in this community \citep{adler_weak_2009}.
+Interactions with above- or belowground enemies can regulate populations and, given enough species-specificity in the effects of these enemies, decouple interspecific dynamics \citep{hersh_evaluating_2011,janzen_herbivores_1970,connell_role_1971}.
 More likely, plants may alter the environment, and even resource availability, in ways that facilitate their neighbors.
 For example, shade cast by \textit{Artemisia} canopies may reduce evapotranspiration (e.g.
-refs...).
+\citealt{Barbier2008}).
 \textit{Artemisia} removal might offest this facilitative effect by reducing resource uptake, resulting in little net change in water balance following removal.
@@
 Evidence for weak interspecific interactions and limited competitive release carries interesting implications.
-If most interspecific interactions are weak, then single-species models may be adequate for forecasting population responses to environmental change (Adler et al.
-2012, Kleinhesselink and Adler 2015).
-Similarly, differences among species in relative abundance should reflect variation in the strength of intraspecific density dependence more than variation in the strength of interspecific interactions (Yenni et al.
-2015).
+If most interspecific interactions are weak, then single-species models may be adequate for forecasting population responses to environmental change \citep{adler_forecasting_2012,kleinhesselink_indirect_2015}.
+Similarly, differences among species in relative abundance should reflect variation in the strength of intraspecific density dependence more than variation in the strength of interspecific interactions \citep{yenni_strong_2012}.
 In such a system, effective management and conservation requires understanding the sources and species-specificity of intraspecific density dependence.
@@
 In closed-canopy forests where competition for light is intense, we might have found different results than in our system, in which competition for a number of belowground resources is more important than competition for light.
-However, a number of recent studies from forests have found that heterospecific competitive effects are dramatically weaker than conspecific effects (Uriarte refs, Comita et al.
-2010, Johnson et al.
-2012).
-We also expect stronger competitive release in successional systems, or communities in the early stages of community assembly where the process of competitive exclusion is still underway (examples?).
+However, a number of recent studies from forests have found that heterospecific competitive effects are dramatically weaker than conspecific effects \citep{comita_asymmetric_2010,kunstler_plant_2015,johnson_conspecific_2012}.
+We also expect stronger competitive release in successional systems, or communities in the early stages of community assembly where the process of competitive exclusion is still underway \citep{kokkoris_patterns_1999}.
 Finally, we might have found stronger competitive release if we could have studied rare species in our system, as long as those species are rare because they are competitively suppressed.
@@
 Because some $W=0$, we conducted two separate regressions.
-First, using all $W$'s, we fitted a generalized linear model with a logit link function to evaluate whether the probability that the focal species' probability of having $W=0$ is influenced by the value of other species' $W$'s.
-Second, for the set of records in which the focal species has $W>0$, we performed a linear regression.
-We repeated these regressions for individuals of each focal species.
-Due to large samples size, $W$ values were often statistically significant predictors.
+First, using all $W$'s, we fitted a generalized linear model with a logit link function to evaluate whether the probability that the focal species' $W=0$ is influenced by the value of other species' $W$'s.
+In this model, the dependent variable is a Bernoulli variate coding for the zero or non-zero value of the focal species' crowding, and the independent variables are the $W$'s for all other species.
+Second, for the set of records in which the focal species has $W>0$, we performed a linear regression, where the focal species' $W$ is the dependent variable, and the other species' $W$'s are the independent variables.
+We repeated these regressions for each focal species.
+Due to large samples size, interspecific $W$ values were often statistically significant predictors of intraspecific.
 However, they explained very little variance.
 The maximum reduction in deviance for the generalized linear regressions and $R^2$ for the linear regressions were both less than 8\%.
-The \texttt{R} code for this analysis is included as ..\texttt{\textbackslash Wdistrib\textbackslash exploreSurvivalWs.r}) \comment{It's not clear what the independent and dependent variables are, in the analyses described here.
-It might help to use the notation from the Methods section?}
+The \texttt{R} code for this analysis is included as ..\texttt{\textbackslash Wdistrib\textbackslash exploreSurvivalWs.r}.
 
 \subsection{Mean field approximation of local crowding for the IPM} 
-\citet{adler-etal-2010} developed a mean field approximation for local crowding when the
+\citet{adler_coexistence_2010} developed a mean field approximation for local crowding when the
 competition kernels are all Gaussian functions, $F_{jm}(d) = e^{-\alpha_{jm} d^2}$. The approximation is explained in 
-the online SI to \citet{adler-etal-2010} and in section 5.3 of \citet{ipmbook-2016}. 
+the online SI to \citet{adler_coexistence_2010} and in section 5.3 of \citet{Ellner2016}. 
 Here we explain how that approximation was modified for the IPMs in this paper, which
@@
   \centering
-  \includegraphics[width=0.7\textwidth]{climate}
+  \includegraphics[width=1\textwidth]{climate}
   \caption{Annual precipitation (a) and mean temperature (b) during the period of the experiment, shown against the long-term (1927-2016) means (solid blue and red lines) and 5\% and 95\% quantiles (dashed lines).

### 2016-09-29 removals/manuscript/removal_ms.tex

@@
 We initialized all species with arbitrarily low cover and ran the model for 500 time steps (years) to allow the community to reach an equilibrium.
-To incorporate temporal environmental variation, at each time step we randomly selected one of the 30 sets of random year effects for the survival, growth and recruitment models (``kernel selection'' in the terminology of Metcalf et al.
-2016).
+To incorporate temporal environmental variation, at each time step we randomly selected one of the 30 sets of random year effects for the survival, growth and recruitment models (``kernel selection'' in the terminology of \citealt{metcalf_statistical_2015}).
 We calculated the equilibrium cover for each species by simulating an additional 2000 time steps and averaging each species' cover over this period.

### 2016-09-30 removals/manuscript/removal_ms.tex
Minor removal ms edits

@@
 However, to the extent that the baseline model underestimates (or overestimates) the actual competitive release, the results of the second and third scenarios will differ.
-
 As with the IBM, we repeated these simulations using maximum removal treatment effects represented by parameters values at the 5\% or 95\% confidence limit farthest from zero.
+We did not simulate long-term \textit{A.
+tripartita} response to grass removal because the vital rate regressions for \textit{A.
+tripartita} showed that the baseline model effectively predicted removal effects.
 
@@
 
+ The equilibrium cover of \text{Artemisia} simulated by the IPM (6.0\%) is consistent with observed cover during the historical data collection period (6.6\% from 1931-1956), but is low compared to contemporary observed cover (15-25\%,\ref{fig:CoverTrends}).
+Could the small competitive release following simulated \text{Artemisia} removal simply reflect the low \text{Artemisia} cover?
+To explore this possibility, we increased \text{Artemisia} cover in the baseline simulation by increasing average \text{Artemisia} fecundity.
+The increased fecundity led to a doubling of \text{Artemisia} average cover, but had essentially zero effect on the cover of the grasses (changes of less than 0.1\% cover).
+
 \section*{Discussion}
@@
 Despite the fact that previous removal studies have reported variable results, ranging from negative effects of removal on remaining species to complete competitive release, we were suspicious of our model's predictions and conducted a removal experiment to test them.
-The results provide substantial, but not complete, support for our model's predictions.
+The results provide substantial support for our model's predictions.
 
@@
 \textit{Artemisia} removal might offest this facilitative effect by reducing resource uptake, resulting in little net change in water balance following removal.
-Unfortunately, we did not monitor soil moisture at the spatial and temporal resolution necessary to test this hypothesis.
+Unfortunately, we did not monitor soil moisture at the fine spatial and temporal resolution necessary to test this hypothesis.
 
@@
 
-Our results have given us more confidence that competitive release is quite small for some species in some communities.
+Our results have given us more confidence that competitive release is quite small for the dominant species in our study system.
 But for many species in other communities, interspecific competition may still be an important driver.
 In closed-canopy forests where competition for light is intense, we might have found different results than in our system, in which competition for a number of belowground resources is more important than competition for light.
-However, a number of recent studies from forests have found that heterospecific competitive effects are dramatically weaker than conspecific effects \citep{comita_asymmetric_2010,kunstler_plant_2015,johnson_conspecific_2012}.
+However, a number of recent studies from forests have also found that heterospecific competitive effects are dramatically weaker than conspecific effects \citep{comita_asymmetric_2010,kunstler_plant_2015,johnson_conspecific_2012}.
 We also expect stronger competitive release in successional systems, or communities in the early stages of community assembly where the process of competitive exclusion is still underway \citep{kokkoris_patterns_1999}.
-Finally, we might have found stronger competitive release if we could have studied rare species in our system, as long as those species are rare because they are competitively suppressed.
+Finally, we might have found stronger competitive release if we could have studied rare species in our system, as long as those species are rare because of niche overlap with stronger competitors.
 If they are rare because they occupy unique, but narrow, niches, interspecific interactions would still be weak.
-Determining why interspecific interactions and competitive release are strong at some times and places and why they are weak in other situations remains a fascinating research problem.
+Determining why interspecific interactions and competitive release are strong in some ecological contexts but weak in others remains a fascinating research problem.
 
@@
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-% APPENDICES !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
+% APPENDICES !
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

### 2016-09-30 removals/manuscript/removal_ms.tex
Minor manuscript edits

@@
 
-\smallskip \textbf{Authorship statement:} PBA and ARK designed the experiment and supervised data collection, PBA, SPE, GH and BT analyzed data, PBA wrote the first draft of the manuscript and all authors contributed to editing.
+\smallskip \textbf{Authorship statement:} PBA and ARK designed the experiment and supervised data collection with the help of B.
+Taylor, PBA, SPE, GH and B.
+Teller analyzed data, PBA wrote the first draft of the manuscript and all authors contributed to editing.
 \smallskip
@@
 Resource competition plays a central role in both Tilman's mechanistic resource competition models \citep{tilman_resource_1982} and Grime's Competitor-Stress tolerator-Ruderal strategies \citep{grime_plant_1979}.
-Neutral theory \citep{hubbell_unified_2001} goes even further, assuming that interspecific competition is exactly as strong as interspecific competition.
+Neutral theory \citep{hubbell_unified_2001} goes even further, assuming that interspecific competition is exactly as strong as intraspecific competition.
 Recent work has emphasized the role of interspecific facilitation in some plant communities \citep{he_global_2013,brooker_facilitation_2008}, but it is unclear whether this research has shaken the default assumption that plant species compete intensely.
@@
 We modeled vital rates using data from 22 year-to-year transitions between 1926 and 1957, and nine year-to-year transitions from 2007 to 2016.
-Only four quadrates were observed for the first two transitions in the 1920's, while at least 14 quadrats were observed for all subsequent transitions.
+Only four quadrats were observed for the first two transitions in the 1920's, while at least 14 quadrats were observed for all subsequent transitions.
+
+All data and R code necessary to reproduce our analysis will be deposited in the Dryad Digital Repository once the manuscript is accepted.
+The current version of the computer code is available at https://github.com/pbadler/ExperimentTests/tree/master/removals and the data are available at https://bitbucket.org/ellner/driversdata.
 
@@
 \end{equation}
-where $\gamma$ is a time-dependent intercept, $\varphi$ is the coefficient for the effect of spatial group, $\chi$ is the removal treatment effect, $\beta$ is the coefficient that represents the effect of log genet size, $u$, on survival.
+where $\gamma$ is a time-dependent intercept, $\varphi$ is the coefficient for the effect of spatial location group, $\chi$ is the removal treatment effect, $\beta$ is the coefficient that represents the effect of log genet size, $u$, on survival.
 $\boldsymbol{\omega}$ is a vector of interaction coefficients which determine the impact of crowding, $\boldsymbol{W}$, by each species on the focal species.
@@
 \begin{equation}
-\mbox{logit}(u_{ijgh,t+1}) = \gamma_{j,t}^G + \varphi_{jg}^G+  \chi_{jh}^G  + \beta_{j,t}^G u_{ij,t} + 
+u_{ijgh,t+1} = \gamma_{j,t}^G + \varphi_{jg}^G+  \chi_{jh}^G  + \beta_{j,t}^G u_{ij,t} + 
 \left \langle  \boldsymbol{\omega}_{j,t}^G, \boldsymbol{W}_{ij,t} \right \rangle + \varepsilon_{ij,t}^G .
@@
 All genets in a quadrat were included in calculating $W$, but plants located within 5 cm of quadrat edges were not used in fitting. 
-We fit the models using the \textit{R-INLA} package for \texttt{R} \citep{rue_approximate_2009}.
+We fit the models using the \textit{R-INLA} package for R \citep{rue_approximate_2009}.
+To explore variations on the growth regressions, such as the addition of a year-by-treatment interaction or information about the locations of individual grasses relative to removed \textit{Artemisia} shrubs, we fit models using the \textit{lmer} function in the \textit{lme4} package for R
 
@@
 \end{equation}
-where the superscript $R$ refers to Recruitment, $C'_{jq,t}$ is the `effective cover' (cm$^2$) of species $j$ in quadrat $q$ at time $t$, $\gamma$ is a time-dependent intercept, $\varphi$ is a coefficient for the effect of group location,
+where the superscript $R$ refers to Recruitment, $C'_{jq,t}$ is the `effective cover' (cm$^2$) of species $j$ in quadrat $q$ at time $t$, $\gamma$ is a time-dependent intercept, $\varphi$ is a coefficient for the effect of spatial location,
 $\chi$ is the removal treatment effect, $\boldsymbol{\omega}$ is a vector of coefficients that determine the strength of intra- and interspecific density-dependence, and $\boldsymbol{C'}$ is the vector of ``effective'' cover of each species in the community.
@@
 We previously used a Ricker equation \citep{adler_coexistence_2010}, but using the square-root transformation instead gave a lower Deviance Information Criterion (DIC) value \citep{spiegelhalter_bayesian_2002} for the recruitment model.
-Following previous work \citep{adler_coexistence_2010}, we treated year and group variables as random factors, allowing intercepts to vary among years and spatial locations.
-
-Due to the possibility that plants outside the mapped quadrat could contribute recruits to the focal quadrat or interact with plants in the focal quadrat, we estimated effective cover as a mixture of the observed cover, $C$, in the focal quadrat, $q$, and the mean cover, $\bar{C}$, across the group, $g$, in
-which the quadrat is located: $C'_{jq,t}=p_j C_{jq,t}+(1-p_j) \bar{C}_{jg,t}$, where $p$ is a mixing fraction between 0 and 1 that was estimated as part of fitting the model.
+Following previous work \citep{adler_coexistence_2010}, we treated year and spatial group variables as random factors, allowing intercepts to vary among years and spatial locations.
+
+Due to the possibility that plants outside the mapped quadrat could contribute recruits to the focal quadrat or interact with plants in the focal quadrat, we estimated effective cover as a mixture of the observed cover, $C$, in the focal quadrat, $q$, and the mean cover, $\bar{C}$, across the spatial location, $g$, in which the quadrat is located: $C'_{jq,t}=p_j C_{jq,t}+(1-p_j) \bar{C}_{jg,t}$, where $p$ is a mixing fraction between 0 and 1 that was estimated as part of fitting the model.
 

### 2016-10-01 removals/manuscript/removal_ms.tex
adding my comments and edits to the removal manuscript.  This is in branch andy-edits which then could then be merged with the master branch if you wanted to use my edits directly with git

@@
 
-\title{Weak interspecific interactions in a sagebrush steppe: Evidence from observational data, population models, and a removal experiment}
+\title{When historical observations, models, and experiments agree: Interspecific competition is weak in a sagebrush steppe}
 
@@
 \author[b]{Giles Hooker}
-\author[c]{Bret Taylor}
+\author[c]{J. Bret Taylor}
 \author[a]{Brittany Teller}
 \author[d]{Stephen P. Ellner}
-\affil[a]{Department of Wildland Resources and the Ecology Center, Utah State University, Logan, Utah 84322}
+\affil[a]{Department of Wildland Resources and the Ecology Center, Utah State University, Logan, Utah}
 \affil[b]{Department of Biological Statistics and Computational Biology, Cornell University, Ithaca, New York}
-\affil[c]{USDA-ARS....}
+\affil[c]{USDA, Agricultural Research Service, U. S. Sheep Experiment Station, 19 Office Loop, Dubois, ID, USA}
 \affil[d]{Department of Ecology and Evolutionary Biology, Cornell University, Ithaca, New York}
@@
 
-\smallskip \textbf{Authorship statement:} PBA and ARK designed the experiment and supervised data collection with the help of B.
-Taylor, PBA, SPE, GH and B.
-Teller analyzed data, PBA wrote the first draft of the manuscript and all authors contributed to editing.
+\smallskip \textbf{Authorship statement:} PBA and ARK designed the experiment and supervised data collection with the help of JBT, PBA, SPE, GH and BT analyzed data, PBA wrote the first draft of the manuscript and all authors contributed to editing.
 \smallskip
@@
 \begin{enumerate}
-\item A previous study of four dominant species from a sagebrush steppe community used historical data sets and multispecies population models to demonstrate weak interspecific interactions, implying very little competitive release.
-\item We conducted a removal experiment to test the model's ability to predict competitive release.
+\item A previous study of four dominant species from a sagebrush steppe community used historical data sets and multispecies population models to demonstrate weak interspecific interactions. 
+\item The implication of this result is that species abundances at this site are determined mainly by the species themselves and there should be little competitive release when other species are removed.
+\item We conducted a removal experiment to test this prediction.
 We established new quadrats in the same location where the historical data were collected.
@@
 However, including information about the location of these grass individuals with respect to removed \textit{Artemisia} failed to improve the growth regressions, raising questions about the mechanisms driving the positive response to removal.
-\item For three of the four species, individual based models and integral projections models based on the vital rate regressions showed that removal treatment effects had little impact on population growth rates or simulated equilibrium cover following a species removal.
+\item For three species, individual based models and integral projections models based on the vital rate regressions showed that removal treatment effects had little impact on population growth rates or simulated equilibrium cover following a species removal.
 For \textit{P.
 spicata}, the population models that included effects of \textit{Artemisia} removal projected greater competitive release than the baseline models.
-\item \emph{Synthesis}: Although models based on observational data did not perfectly predict species responses to removal, overall the results increase our confidence that competitive release is in fact small for at least three of our four study species.
-The implication is that interspecific resource competition is not the primary factor determining the dynamics of this community.
+\item \emph{Synthesis}: Although models based on observational data did not perfectly predict species responses to removal, the experiment increases our confidence that competitive release is small for at least three of our four study species.
+The surprising implication is that species abundances at this site are a function of the species themselves and do not depend on the abundances of interspecific competitors.
 
@@
 However, we still might expect interspecific interactions to be strong and negative.
-In fact, interspecific competition is arguably the primary obsession of plant community ecologists.
-We motivate research on plant species coexistence by noting that all plant species compete for water, light, and a few mineral resources (e.g.
+In fact, ecologists often motivate research on plant species coexistence by noting that all plant species compete for water, light, and a few mineral resources (e.g.
 \citealt{silvertown_plant_2004}).
@@
 Neutral theory \citep{hubbell_unified_2001} goes even further, assuming that interspecific competition is exactly as strong as intraspecific competition.
-Recent work has emphasized the role of interspecific facilitation in some plant communities \citep{he_global_2013,brooker_facilitation_2008}, but it is unclear whether this research has shaken the default assumption that plant species compete intensely.
+While a growing list of studies also emphasize the role of interspecific facilitation in some plant communities \citep{he_global_2013,brooker_facilitation_2008}, it is unclear whether this has overturned the default assumption that plant species compete intensely with other species.
 
@@
 \subsection*{Removal experiment}
-In spring 2011, in a large exclosure containing six of the historical permanent quadrats, we selected locations for an additional 16 quadrats for the removal treatments.
-We used the following criteria to select the quadrat locations.
-Each plot had to have 5\% canopy cover of \textit{Artemisia tripartita} and at least 3\% combined basal cover of the three focal perennial grasses.
-We avoided plots falling on hill slopes, areas with greater than 20\% bare rock, or with over 10\% cover of the woody shrubs \textit{Purshia tridentata} or \textit{Amelanchier utahensis}.
-We mapped the quadrats in June, 2011 and then implemented the removal treatments in September, 2011.
+In spring 2011, in a large exclosure containing six of the historical permanent quadrats, we established an additional 16 quadrats for the removal treatments.
+New plots were selected so as to be similar to historical plots, but also needed to have enough of the focal species present so that removal experiments could be conducted.
+We evaluated potential locations for the new plots near the original plots and then used only those locations with at least 5\% canopy cover of \textit{Artemisia tripartita} and at least 3\% combined basal cover of the three focal perennial grasses.
+We rejected locations falling on hill slopes, areas with greater than 20\% bare rock, or with over 10\% cover of the woody shrubs \textit{Purshia tridentata} or \textit{Amelanchier utahensis}.
+We mapped the new plots in June, 2011 and then implemented the removal treatments in September, 2011.
 We randomly assigned 8 of the sixteen new quadrats to the sagebrush removal treatment, and the remaining 8 to the grass removal treatment.
@@
 \end{equation}
+
+% I could be wrong but I think the left-hand side of the equation needs to be 
+% indexed to {t + 1} not t.
+
 where $\gamma$ is a time-dependent intercept, $\varphi$ is the coefficient for the effect of spatial location group, $\chi$ is the removal treatment effect, $\beta$ is the coefficient that represents the effect of log genet size, $u$, on survival.
@@
 Because we were interested in comparing model predictions to observations, and were not interested in the effects of demographic stochasticity, we used a deterministic version of the model (e.g., recruitment is the $\lambda$ of (\ref{eqn:recrProcessModel}), rather than a random draw from a Poisson distribution with a mean of $\lambda$).
+
+%% Wait, how are recruit positions within the plot distributed?
+
 We first applied the recruitment regression to determine the expected number of new recruits at time $t+1$, which we then multiplied by the mean size of recruits observed for the focal species to get expected cover of new recruits.
@@
 Our vital rate models take into account the effects of local species composition and should therefore be able to predict the effect of removals on survival, growth and recruitment without any additional parameters.
-If the removal treatment effects that we added to these models are statistically significant, it would indicate an effect of removals that goes above and beyond the response predicted by a model based on data from the control plots.
+If the removal treatment effects that we added to these models are statistically significant, it would indicate removal effects are stronger or weaker from those predicted by a model based on observational data.
+
+% There’s a bit of a leap here from statistical significance to predictions
+% I think this is intuitively correct and probably right most of the time, but 
+% I wonder whether a main effect of removal treatment necessarily implies 
+% that predictions don’t match observations. (I guess it probably does).  
+% However, in the case that treatment effects are not significant, I think there
+% could be cases where predictions still technically differ from observations.
+For % instance, do predictions technically differ from observations if they predict the % mean, but underestimate the variance?
+Something to consider.
+
 For the survival models, the 95\% confidence intervals on the removal effects overlapped zero for all four species (Fig.
@@
 \ref{fig:VitalRateTest}; Tables \ref{ARTRgrowth} to \ref{PSSPgrowth}).
-In contrast, \textit{Poa secunda} had lower recruitment in removal than controls plots (upper limit of 95\% CI $<0$; Fig.
-\ref{fig:VitalRateTest}; Table \ref{table:recruitment}).
+
+% What about other species’ recruitment results?
+
+In contrast, \textit{Poa secunda} had lower recruitment in removal than controls plots (upper limit of 95\% CI $<0$; Fig. \ref{fig:VitalRateTest}; Table \ref{table:recruitment}).
 
@@
 
+%
+% Peter the dashed and dotted lines in this figure don’t show up as distinct to me 
+% (baseline vs. removal predicted).  I can’t tell which is which
+%
+
 The positive removal effects on \textit{Poa secunda}  and \textit{P.
-spicata} growth are particularly interesting, because they suggest that models based on observational data may underestimate competitive release.
+spicata} growth suggest that models based on observational data underestimate competitive release.
 We conducted additional analyses to further explore removal effects on the growth of these two species.
@@
 
-Third, responses to removals might be poorly predicted by models based on observational data if the removals create local crowding outside the range of historical variation or if the removals break covariances in crowding among species.
-However, the distribution of crowding observed within control plots easily spans the crowding values observed in removal plots (Fig.
-\ref{fig:Wscatters}), and we failed to find evidence for strong interspecific covariances in crowding (\ref{suppMethods}).
+Third, responses to removals might be poorly predicted by models based on observational data if the removal plots lead to lower plant densities (crowding) than the range of the historical variation or if the removals break covariances in plant densities among species.
+However, the crowding values observed in removal plots (Fig.
+\ref{fig:Wscatters}) fell within the range of crowding observed in control plots, and we failed to find evidence for strong interspecific covariances in crowding (\ref{suppMethods}).
 
@@
 comata} was lower in removal plots than controls, though it was absent from most \textit{P.
-spicata} neighborhood in both treatments (Fig.
+spicata} neighborhoods in both treatments (Fig.
 \ref{fig:W-by-treatment}).
@@
 Our population models based on historical, observational data reveal weak interspecific interactions \citep{adler_coexistence_2010,chu_large_2015} and, as a result, predict very little competitive release among the four dominant species in a sagebrush steppe.
-Despite the fact that previous removal studies have reported variable results, ranging from negative effects of removal on remaining species to complete competitive release, we were suspicious of our model's predictions and conducted a removal experiment to test them.
+Given the obsession plant community ecologists have with competition between species, we were suspicious of our model's predictions and conducted a removal experiment to test them.
 The results provide substantial support for our model's predictions.
@@
 For three of the four species, we find little or no evidence that models based on observational data underestimate the intensity of interspecific competition.
-Removal treatment effects were never significant for \textit{A.
+After accounting for local neighborhood crowding, the estimated main effect of removal treatment was never significant for \textit{A.
 tripartita} or \textit{H.
-comata}; our baseline model, which accounts for local neighborhood composition, adequately explained vital rate responses to  removal.
+Comata}.
 \textit{Poa secunda} showed mixed responses to \textit{Artemisia} removal: removal effects were not significant at the survival stage, were positive at the growth stage, and negative at the recruitment stage (where the estimated effect of \textit{Artemisia} on \textit{Poa} is positive).
@@
 
+% another hypotheses about PSSP
+% I predict you will cut this! :)
+
+Another explanation for the under-prediction of \textit{P.
+spicata} competitive release may come from consideration of the scale at which plots are monitored in this experiment.
+In our models the competition experienced by a focal plant is weighted by distance of competitors from the focal plant.
+However because our plots are 1 m across we have no information about the effects of competition at scales greater than on 1 m.
+Given that \textit{Artemisia} roots can spread laterally beyond their canopy’s extent (Kleinhesselink personal observation), it is possible that even \textit{P.
+spicata} plants in plots without \textit{Artemisia} canopies, a situation in theory identical to our removals, are experiencing some competition below ground from \textit{Artemisia} outside the plots.
+When we conducted removals we removed plants within 50 cm of the removal plots.
+These “outside the bounds” removals could give an added boost to \textit{P.
+spicata} performance that would not be captured by our historical data.
+Additionally, there is growing evidence that plant roots are territorial, they recognize self and non-self roots and respond differently to the roots of different species \citep{Krinnetz and Caldwell}.
+Moreover, roots of many species appear to grow towards interspecific competition and avoid intraspecific neighbors—resulting in patterns of intraspecific root aggregation \citep{bertelheimer_aggregative_2006}.
+If \textit{Artemisia} roots behave in this way, it could mean that our removal plots contain much lower densities of \textit{Artemisia} roots belowground than control plots without \textit{Artemisia} canopies aboveground.
+If this is the case, we could find that as \textit{Artemisia} roots from plants outside our plots re-invade the territory within the plots, the growth of \textit{P.
+spicata} will gradually be reduced and come to match that in control plots without \textit{Artemisia}.
+
 The small competitive release of many species in our system implies that each of the four dominant species we studied are limited by quite different factors.
@@
 Similarly, differences among species in relative abundance should reflect variation in the strength of intraspecific density dependence more than variation in the strength of interspecific interactions \citep{yenni_strong_2012}.
+
 In such a system, effective management and conservation requires understanding the sources and species-specificity of intraspecific density dependence.
 In fact, an experiment to test our model's skill at predicting release from intraspecific competition could be more important than the test of interspecific competitive release described here.
+However, effectively manipulating interspecific competition while simultaneously allowing species abundance to change is logically challenging.
 
@@
 However, a number of recent studies from forests have also found that heterospecific competitive effects are dramatically weaker than conspecific effects \citep{comita_asymmetric_2010,kunstler_plant_2015,johnson_conspecific_2012}.
-We also expect stronger competitive release in successional systems, or communities in the early stages of community assembly where the process of competitive exclusion is still underway \citep{kokkoris_patterns_1999}.
+
+% added this nutnet tie-in that I think is cool; it could go up after the Yenni 
+% reference above.
+
+In grasslands, our results may also help explain a recent study showing that plot level abundance of widespread species often show little variation across continents, even when the identity of competitors between these locations is very different \citep{firn_abundance_2011}.
+If intraspecific competition is very strong, as in our study site, then the effect of novel competitors and new interspecific interactions may be expected to have little impact on the relative abundances of species in these systems.
+
+We do expect stronger competitive release in successional systems, or communities in the early stages of community assembly where the process of competitive exclusion is still underway \citep{kokkoris_patterns_1999}.
+
 Finally, we might have found stronger competitive release if we could have studied rare species in our system, as long as those species are rare because of niche overlap with stronger competitors.

### 2016-10-02 removals/manuscript/removal_ms.tex
adding more revisions to removal manuscript on andy edits branch.  Adding more specific references for Pseudoroegeneria competition, spatial segregation and hydraulic lift

--- 2016-10-01 removals/manuscript/removal_ms.tex
@@
 
-% I could be wrong but I think the left-hand side of the equation needs to be 
-% indexed to {t + 1} not t.
+%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+% I could be wrong but I think the left-hand side of the equation above
+% needs to be  indexed to {t + 1} not t.  
+%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 
@@
 
-%% Wait, how are recruit positions within the plot distributed?
+% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+% Wait, how are recruit positions within the plot distributed? 
+% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 
@@
 
+% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 % There’s a bit of a leap here from statistical significance to predictions
@@
 % could be cases where predictions still technically differ from observations. 
-For % instance, do predictions technically differ from observations if they predict the % mean, but underestimate the variance?
-Something to consider.
+% For % instance, do predictions technically differ from observations if they predict the 
+% mean, but underestimate the variance? Something to consider.
+% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 
@@
 
+% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 % What about other species’ recruitment results? 
+% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 
@@
 
-%
+% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 % Peter the dashed and dotted lines in this figure don’t show up as distinct to me 
 % (baseline vs. removal predicted).  I can’t tell which is which
-%
+% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 
@@
 
-% another hypotheses about PSSP
-% I predict you will cut this! :)
+% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+% Here's my   hypotheses about PSSP. I predict you will cut this! :)   
+% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 
@@
 spicata} competitive release may come from consideration of the scale at which plots are monitored in this experiment.
-In our models the competition experienced by a focal plant is weighted by distance of competitors from the focal plant.
-However because our plots are 1 m across we have no information about the effects of competition at scales greater than on 1 m.
+In our models, the competition experienced by a focal plant is weighted by distance of competitors from the focal plant.
+However we have no information about the effects of competition at scales greater than on 1 m.
 Given that \textit{Artemisia} roots can spread laterally beyond their canopy’s extent (Kleinhesselink personal observation), it is possible that even \textit{P.
-spicata} plants in plots without \textit{Artemisia} canopies, a situation in theory identical to our removals, are experiencing some competition below ground from \textit{Artemisia} outside the plots.
+spicata} plants in plots without \textit{Artemisia} canopies, a situation in theory identical to our removals, are experiencing some competition with \textit{Artemisia} whose roots extend into the plots from outside.
 When we conducted removals we removed plants within 50 cm of the removal plots.
-These “outside the bounds” removals could give an added boost to \textit{P.
-spicata} performance that would not be captured by our historical data.
-Additionally, there is growing evidence that plant roots are territorial, they recognize self and non-self roots and respond differently to the roots of different species \citep{Krinnetz and Caldwell}.
-Moreover, roots of many species appear to grow towards interspecific competition and avoid intraspecific neighbors—resulting in patterns of intraspecific root aggregation \citep{bertelheimer_aggregative_2006}.
-If \textit{Artemisia} roots behave in this way, it could mean that our removal plots contain much lower densities of \textit{Artemisia} roots belowground than control plots without \textit{Artemisia} canopies aboveground.
-If this is the case, we could find that as \textit{Artemisia} roots from plants outside our plots re-invade the territory within the plots, the growth of \textit{P.
-spicata} will gradually be reduced and come to match that in control plots without \textit{Artemisia}.
+These marginal removals could give an added boost to \textit{P.
+spicata} performance that would not be captured by the crowding covariate in our model.
+Additionally, there is evidence from many species, including desert shrubs that plant roots from individuals are often spatially segregated \citep{schenk_spatial_1999, brisson_effect_1994}.
+If this is the case for \textit{Artemisia}, then the roots within a plot may belong to only one or two \textit{Artemisia} plants.
+If we remove those individuals, it seems possible that over time roots from other individual \textit{Artemisia} plants outside our plots will invade the removal plots over time.
+A gradually lessening of the removal effect \textit{P.
+spicata} in the removal plots over time would be support this hypothesis.
 
@@
 For example, shade cast by \textit{Artemisia} canopies may reduce evapotranspiration (e.g.
-\citealt{Barbier2008}).
+\citealt{Barbier2008}),
+
+%adding these refs on hydraulic lift:
+
+and \textit{Artemisia} may also increase soil moisture through hydraulic lift \citep{richards_hydraulic_1987}, and this may also increase nitrogen availability \citep{cardon_sagebrush_2013}.
 \textit{Artemisia} removal might offest this facilitative effect by reducing resource uptake, resulting in little net change in water balance following removal.
 Unfortunately, we did not monitor soil moisture at the fine spatial and temporal resolution necessary to test this hypothesis.
+
+% --------------------------------------------------------
+% section on caldwell's resource competition studies: 
+% --------------------------------------------------------
+
+It is interesting to note, that the one grass species which showed a stronger than expected growth response to \textit{Artemisia} removal, \textit{P.
+spicata}, has been shown in a number of studies to compete strongly for soil resources with big sagebrush \textit{Artemisia tridentata}, a close relative of the shrub in our study \citep{caldwell_competition_1985,caldwell_rapid_1987,caldwell_exploitation_1991}.
+\textit{Artemisia tridentata} in these studies was shown to be more effective at competing with \textit{P.
+spicata} for phosphorous than with another perennial grass species \citep{caldwell_competition_1985}.
+Thus it seems plausible that reduced competition for soil nutrients may be responsible for the strong positive effect of \textit{Artemisia} removal on \textit{P.
+spicata} in our study.
 
@@
 
-% added this nutnet tie-in that I think is cool; it could go up after the Yenni 
-% reference above.
+% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+% I added this nutnet tie-in that I think is cool; it could go up 
+% after the Yenni reference above instead.   
+% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~`
 

### 2016-11-02 removals/manuscript/removal_ms.tex
Minor updates to figures and figure legends

--- 2016-10-01 removals/manuscript/removal_ms.tex
@@
 \textit{A.
-tripartita} cover in shrub removal plots) is included to show values at the start of the experiment in 2011 and the efficacy of the removal treatments.}
+tripartita} cover in shrub removal plots) is included to show values at the start of the experiment in 2011 and the efficacy of the removal treatments.
+Source file: \texttt{treatment\_trends\_removals.r}.}
 \label{fig:CoverTrends}
@@
 Predictions from the ``Removal'' model include additional treatment effects on vital rates.
+Source file: \texttt{ibm\\summarize\_sims1step.r}.
 }
@@
 Dotted lines show one-step-ahead predictions from an IBM that includes removal treatment coefficients.
-}
+Source file: \texttt{ibm\\summarize\_sims1step.r} }
  \label{fig:IBM1step}
@@
 Dotted lines show one-step-ahead predictions from an IBM that includes removal treatment coefficients representing the outermost values of the 95\% confidence intervals.
-}
+Source file: \texttt{ibm\\summarize\_sims1step.r} }
  \label{fig:IBM1step-maxCI}

### 2016-11-11 precip/manuscript/precip_ms.tex
adding draft of precip manuscript based on Peters removal latex script

--- 2016-10-01 removals/manuscript/removal_ms.tex
@@
 
-\author[a]{Peter B. Adler\thanks{Corresponding author. Department of Wildland Resources and the Ecology Center, Utah State University, Logan, Utah Email: peter.adler@usu.edu}}
-\author[a]{Andrew Kleinhesselink}
-\author[b]{Giles Hooker}
-\author[c]{J. Bret Taylor}
-\author[a]{Brittany Teller}
-\author[d]{Stephen P. Ellner}
+\author[a]{Andrew R. Kleinhesselink\thanks{Corresponding author. Department of Wildland Resources and the Ecology Center, Utah State University, Logan, Utah Email: arklein@aggiemail.usu.edu}}
+\author[a]{Peter B. Adler}
+\author[a]{Andrew Tredennick}
 \affil[a]{Department of Wildland Resources and the Ecology Center, Utah State University, Logan, Utah}
-\affil[b]{Department of Biological Statistics and Computational Biology, Cornell University, Ithaca, New York}
-\affil[c]{USDA, Agricultural Research Service, U. S. Sheep Experiment Station, 19 Office Loop, Dubois, ID, USA}
-\affil[d]{Department of Ecology and Evolutionary Biology, Cornell University, Ithaca, New York}
 
@@
 
-\smallskip \textbf{Authorship statement:} PBA and ARK designed the experiment and supervised data collection with the help of JBT, PBA, SPE, GH and BT analyzed data, PBA wrote the first draft of the manuscript and all authors contributed to editing.
+\smallskip \textbf{Authorship statement:} ARK and PBA designed the experiment and supervised data collection. AT helped with the design of the data analysis.
 \smallskip
@@
 \begin{enumerate}
-\item A previous study of four dominant species from a sagebrush steppe community used historical data sets and multispecies population models to demonstrate weak interspecific interactions, implying very little competitive release.
-\item We conducted a removal experiment to test the model's ability to predict competitive release.
-We established new quadrats in the same location where the historical data were collected.
-We assigned half the new quadrats to a  perennial grass removal treatment and half to a treatment in which we removed the dominant shrub, \textit{Artemisia tripartita}.
-We modeled survival, growth and recruitment as functions of local neighborhood species composition along with an indicator variable for removal treatment.
-If our ``baseline" model, which accounts for local plant-plant interactions, accurately predicts responses to removal, then the fitted removal indicator effect should be non-signficant.
-\item  For survival and recruitment, the removal treatment effects were never significantly positive, indicating that competitive release was not underestimated by the interspecific effects included in the baseline model.
+\item Climate is thought to be an important driver of population ecology however it can be difficult to determine exactly which aspects of climate affect populations and at what vital rates.
+Tests of whether statistical models fit to climate and population data can accurately predict population responses to natural and experimental climate manipulations have rarely been conducted.
+\item Using long-term observational data collected from the 1920's to the 1950's from a sagebrush steppe plant community, previous research has reported a wide variety of climate effects on four dominant plant species.
+We tested whether statistical models fit to the historical data could predict the population response of plants in these original plots to natural and experimentally imposed climate variation some 60 years after the historical data collection ended.
+New plots in the same location where the historical data were collected were assigned to either drought or irrigation treatments imposed using rainfall shelters and automated sprinklers.
+The original plots were also monitored and served as a control.
+We modeled survival, growth and recruitment as functions of spring temperature and spring and summer soil moisture.
+We then used statistical models fitted to the historical data to predict the year and treatment effects observed in contemporary data.
+\item  For survival and recruitment, the treatment effects were never significantly positive, indicating that competitive release was not underestimated by the interspecific effects included in the baseline model.
 For  \textit{Poa secunda} recruitment, the removal treatment was actually negative, indicating that our baseline model overestimated competitive release.
@@
 
-Stable coexistence requires that intraspecific interactions must limit populations more than interspecific interactions \citep{chesson_mechanisms_2000}.
-Whenever we study populations of species that appear to be coexisting stably, such as populations that have co-occurred in close proximity for extended periods, we should expect intraspecific limitations to be stronger than interspecific limitations.
-However, we still might expect interspecific interactions to be strong and negative.
-In fact, interspecific competition is arguably the primary obsession of plant community ecologists.
-We motivate research on plant species coexistence by noting that all plant species compete for water, light, and a few mineral resources (e.g.
-\citealt{silvertown_plant_2004}).
-Resource competition plays a central role in both Tilman's mechanistic resource competition models \citep{tilman_resource_1982} and Grime's Competitor-Stress tolerator-Ruderal strategies \citep{grime_plant_1979}.
-Neutral theory \citep{hubbell_unified_2001} goes even further, assuming that interspecific competition is exactly as strong as intraspecific competition.
-Recent work has emphasized the role of interspecific facilitation in some plant communities \citep{he_global_2013,brooker_facilitation_2008}, but it is unclear whether this research has shaken the default assumption that plant species compete intensely.
-
-Competitive release provides an intuitive, population-level measure of the strength of interspecific competition.
-Following removal of one or more species from a community, how much will the abundance of the remaining species increase?
-Large increases in the abundance of the remaining species represent a large competitive release, and imply strong interspecific interaction.
-Plant ecologists have conducted dozens of removal experiments and have observed almost every possible result.
-In some cases, competitive release is complete: remaining species fully compensate for removed biomass \citep{leps_nutrient_1999,jutila_effects_2002}.
-In other cases, the degree of competitive release depended on the functional group or dominance of the removed and remaining species \citep{smith_dominant_2003,sala_resource_1989,belsky_effects_1992}.
-Finally, even negative effects of removals have been observed \citep{keddy_effects_1989,gilbert_dominant_2009}, consistent with the recent facilitation literature.
-Results from removal experiments do come with caveats.
-Experiments typically do not run long enough to document long-term outcomes, responses may only be measured at one life stage whereas population responses depend on the whole life cycle, and removal treatments require disturbance which may cause a variety of impacts unrelated to resource competition \citep{aarssen_neighbour_1990}.
-
-An alternative approach to studying competitive release is to fit multispecies population models to long-term, observational data, and then simulate population responses to removals.
-Applying this approach to arid and semiarid plant communities, we found evidence of strong intraspecific limitation and very weak interspecific interactions, leading to very stable coexistence and implying little competitive release \citep{adler_coexistence_2010,chu_large_2015}.
-Since we only studied the most common, co-occurring species, we expected to find evidence of stable coexistence, but we were surprised that interspecific interactions and species' responses to the loss of a dominant competitor were so weak.
-Compared to traditional field experiments, our approach has the advantage of projecting full life-cycle responses over long time periods, but its reliance on observational data sets limits the strength of our inference.
-We would have more confidence in our results if they were confirmed by an experiment.
-
-To test whether our models, which are based on observational data, underestimate the strength of interspecific competition and competitive release, we conducted a removal experiment in a sagebrush steppe.
-We studied the responses of three perennial grass species to removal of sagebrush, and the response of sagebrush to removal of all grasses.
-We addressed two research questions: First, are the survival, growth or recruitment of our target species significantly different in control vs.
-removal treatments after accounting for differences in local species composition?
-Because our demographic models include the effects of neighborhood crowding, they should be able to predict how experimental removal of neighbors alters demographic performance.
-If the full effect of experimental removals is stronger than predicted, it means that our models underestimate competitive release.
-Our second question integrates across the full life cycle: how do short and long-term population-level responses to removals compare to projections from multispecies population models based on observational data?
-If models based on observational data effectively predict competitive release, the population dynamics they project should be similar to projections from models that include information from the experimental removal treatments.
+Climate is widely considered one of the most powerful external forces driving changes in species abundance across space and time \citep{teller_linking_2016}.
+The effects of climate on populations and ecosystems are most apparent at the largest scales in space and time: climate determines the distribution of ecosystems, treelines and the the range limits of many species \citep{}.
+The paleoecological record shows that long-term climate change lead to changes in species abundance and changes in species' ranges.
+Nevertheless, for many species it is difficult to determine how annual climate variation affects their populations and demographic rates \citep{ehrlen_advancing_2016}.
+Observational data on species performance across many different years can provide some information on how climate might affect population abundance \citep{dalgleish_climate_2010,ehrlen_advancing_2016}.
+But many years of data are needed to reliably detect climate effects, especially when climate effects are weak and often correlated amongst themselves \citep{teller_linking_2016,gerber_optimal_2015}.
+Understanding the effects of climate on populations is increasingly important, however, if we are to understand and predict the effects of climate change on populations.
+If models fit to historical data can be used to accurately predict the effects of contemporary climate variation on populations, as well as the effects of experimental climate manipulations, it would be strong confirmation that the climate effects they describe are real.
+
+To test whether our models, which are based on observational data, accurately predict the effects of natural climate variation, we conducted a precipitation manipulation experiment in a sagebrush steppe.
+We studied the responses of three perennial grass species and a shrub species to natural climate variation over 10 years and experimental drought and irrigation treatments over four years.
+We addressed two research questions: First, how much do the  survival, growth or recruitment of our target species differ between the precipitation manipulation treatments?
+If our experiment does affect species vital rates we interpret that as strong evidence that variation in precipitation should have an effect on populations.
+Our second question is whether statistical models parameterized from historical observational data can accurately predict the effects of modern climate variation and precipitation manipulation on species performance?
+If models based on observational effectively capture the effects of climate on species performance, they should also predict the effects of natural climate variation in the contemporary dataset as well as the effects of precipitation treatments in the experiment.
 
@@
 
-All data and R code necessary to reproduce our analysis will be deposited in the Dryad Digital Repository once the manuscript is accepted.
-The current version of the computer code is available at https://github.com/pbadler/ExperimentTests/tree/master/removals and the data are available at https://bitbucket.org/ellner/driversdata.
-
-\subsection*{Removal experiment}
-In spring 2011, in a large exclosure containing six of the historical permanent quadrats, we selected locations for an additional 16 quadrats for the removal treatments.
-We used the following criteria to select the quadrat locations.
-Each plot had to have 5\% canopy cover of \textit{Artemisia tripartita} and at least 3\% combined basal cover of the three focal perennial grasses.
+\subsection*{Precipitation experiment}
+In spring 2011, in a large exclosure containing six of the historical permanent quadrats, we selected locations for an additional 16 quadrats for the precipitation experiment.
 We avoided plots falling on hill slopes, areas with greater than 20\% bare rock, or with over 10\% cover of the woody shrubs \textit{Purshia tridentata} or \textit{Amelanchier utahensis}.
-We mapped the quadrats in June, 2011 and then implemented the removal treatments in September, 2011.
-We randomly assigned 8 of the sixteen new quadrats to the sagebrush removal treatment, and the remaining 8 to the grass removal treatment.
-We cut sagebrush stems at ground level and applied a 5\% solution of Roundup Pro (Glyphosate, Monsanto) to the cut stems.
-In the grass removal plots, we painted the same herbicide solution on all perennial grasses.
-We removed plants 50 cm beyond the quadrat boundaries to minimize edge effects.
-The sagebrush removal was virtually 100\% effective; grass removals required additional herbicide applications in April and May of each year, with fewer grasses remaining each year.
-
-Although the primary focus of our analysis is comparison of observed removal responses with responses predicted by our population models, we also conducted a more traditional statistical analysis of differences in cover trends on control vs.
-removal plots.
+New plots were established in pairs, and one plot per pair was randomly assigned to either the precipitation reduction or the precipitation addition treatment.
+We mapped the quadrats in June, 2011 and then built the rainfall shelters and set-up the irrigation systems in the fall of 2011.
+We used a rain-out shelter and automatic irrigation design described in \citep{gherardi, yahdjian_rainout_2002}.
+Each rain-out shelter covered an area of 2.5 by 2 m and consisted of transparent acrylic shingles held up over the plot to channel 50\% of incoming rainfall off of the plot and into 75 l reservoirs.
+The collected water was automatically pumped out of the reservoirs and sprayed onto paired irrigation treatment plots.
+Pumping was initiated automatically with float switches that were triggered when water levels in the reservoirs were approximately 20 l, or equivalently irrigation was triggered once for every 6 mm of rainfall collected.
+We disconnected the irrigation pumps in late fall each year and re-connected them in April.
+The drought shelters remained in place throughout the year.
+
+We monitored soil moisture and air temperature in four of the precipitation experiment plot pairs using Decagon Devices (Pulman Washington) 5TM and EC-5 soil moisture sensors and 5TE temperature sensors.
+We installed two soil moisture sensors in each monitored plot at 5 cm and two at 25 cm deep in the soil.
+Air temperature was measured underneath the roofing of the shelter at 30 cm above ground.
+Data were logged automatically every four hours.
+
+We conducted a simple statistical analysis to determine the effect of the rain-out shelters and irrigation on the observed soil moisture.
+
+We conducted a simple statistical analysis of differences in cover trends on treatment plots.
 The response variable for this analysis was year-to-year change in cover at the quadrat scale, calculated as $log(Cover_{t+1}/Cover_t)$.
-We tested for the effect of removal treatment on this measure of cover change with a linear mixed effects model implemented in the \textit{lme4} package \citep{Bates2015} of \textit{R}, including year and quadrat as random effects.
-We considered the removal treatment effect to be significant when the 95\% confidence interval on the estimated coefficient did not overlap zero.
+We tested for the effect of precipitation treatment on this measure of cover change with a linear mixed effects model in the \textit{lme4} package \citep{Bates2015} of \textit{R}, including year and quadrat as random effects.
+We considered the precipitation treatment effects to be significant when the 95\% confidence interval on the estimated coefficient did not overlap zero.
+
+\subsection*{Soil moisture modeling}
+
+We expected that our precipitation manipulation experiment would have effects on plants through the mechanism of altering available soil moisture during the growing season.
+Because we do not have direct measurements of soil moisture for each year of observed plant cover, we used the SOILWAT soil moisture model to estimate daily soil moisture at the USSES from 1925 to the present \citep{Parton 1978}.
+We used an enhanced version of soilwat that has recently been developed for use in semi-arid shrubland ecosystems \citep{Bradford}.
+SOILWAT uses daily weather data, ecosystem specific vegetation properties and site specific soil properties to estimate water balance processes.
+SOILWAT specifically estimates rainfall interception by vegetation, evaporation of intercepted water, snow melt and snow redistribution, infiltration into the soil, percolation through the soil, bare-soil evaporation, transpiration from each soil layer, and drainage.
+We parameterized SOILWAT with the generic sagebrush steppe vegetation parameters and site specific soil texture and bulk density data.
+We used daily weather data collected at the USSES from 1925 until the present as weather forcing data for the SOILWAT predictions.
+The USSES climate station (station id: GHCND:USC00102707) is located within 2 km of the research plots.
+We downloaded daily and monthly tmin, tmax, and precipitation data from the National Climate Data Centers online database.
+
+Because the
+
+We averaged daily soil moisture predictions for 5, 25 and 30 cm soil layers, and daily mean temperature, and then aggregated these quarterly.
 
@@
 We model the survival probability of an individual genet as a function of genet size, the neighborhood-scale crowding experienced by the genet from both conspecific and heterospecific genets, temporal variation among years, and permanent spatial variation among groups of quadrats (`group'; here means a set of nearby quadrats located within one pasture or grazing exclosure).
-Where the current analysis departs from previous work is the inclusion of a categorical removal treatment effect.
-If the conspecific and heterospecific crowding effects explain demographic rates in removal treatment plots as well as they do in control plots, then there will be no need for the removal treatment effect, and the estimated coefficient will be small and not significantly different from zero.
-However, if demographic rates in removal plots differ from control plots even after accounting for crowding effects, then the removal treatment effect will be significant.
-For example, we might find that, holding crowding constant, grasses grow more quickly in the removal plots than in the control plots.
-This result would indicate that our "baseline model," a model including all estimated parameters except for the removal treatment effect, underestimates competitive release.
-We refer to the model that includes the estimated removal treatment effect as the "removal model."
+In this analysis we only include crowding from the four main focal species.
+We incorporated eight climate effects in each of our vital rate regressions: the
 
@@
 
-We ran Markov Chain Monte Carlo (MCMC) simulations in WinBUGS 1.4 \citep{lunn_winbugs_2000} to estimate the recruitment model parameters.
-Each model was run for 20,000 iterations and two independent chains with different initial values for parameters.
-We discarded the initial 10,000 samples.
+We ran Hamiltonian-Markov Chain Monte Carlo (HMCMC) simulations in STAN 6.4 \citep{stan} to estimate the recruitment model parameters.
+Each model was run for 2,000 iterations and four independent chains with different initial values for parameters.
+We discarded the initial 1,000 samples.
 Convergence was observed graphically for all parameters, and confirmed by the Brooks-Gelman statistic \citep{brooks_general_1998}.
 
-For all three vital rate models, we considered the effect of removal treatments to be significant when the 95\% confidence interval (or posterior
-interval) on the estimate of $\chi$ did not overlap zero.
-In an effort to further understand the removal treatment effects, we explored additional variations on the growth model with a linear random effects model.
-For hypothesis testing with these models, we focused on the 95\% confidence interval of the coefficient of interest.
-We also compared the species composition of the control and removal plots at both the quadrat and neighborhood ($\boldsymbol{W}$) scales.
+For all three vital rate models, we considered the effect of climate covariates to be significant when 95\% posterior
+intervals did not overlap zero.
+
+All data and R code necessary to reproduce our analysis will be deposited in the Dryad Digital Repository once the manuscript is accepted.
+The current version of the computer code is available at https://github.com/pbadler/ExperimentTests/tree/master/removals and the data are available at https://bitbucket.org/ellner/driversdata.
 
@@
 \bibliographystyle{Ecology}
-\bibliography{IPMrefs}
+\bibliography{precip_experiment}
 

### 2016-12-13 precip/manuscript/precip_ms.tex
more commits fun.
scaling the competition and size parameters for better fitting

@@
 
-\title{Weak interspecific interactions in a sagebrush steppe: Evidence from observational data, population models, and a removal experiment}
+\title{Can historical data predict ecological responses to novel climate conditions?}
 
@@
 
-\textbf{\large{Keywords:}} Coexistence, competition, integral projection model.
-
-\bigskip \textbf{Running title:} Weak interspecific interactions
+\textbf{\large{Keywords:}} Climate change, demographic models, rain-out shelter.
+
+\bigskip \textbf{Running title:} Predicting climate response
 
@@
 
-\smallskip \textbf{Authorship statement:} ARK and PBA designed the experiment and supervised data collection. AT helped with the design of the data analysis.
+\smallskip \textbf{Authorship statement:} ARK and PBA designed the experiment and supervised data collection. AT helped with the data analysis.
 \smallskip
@@
 \begin{enumerate}
-\item Climate is thought to be an important driver of population ecology however it can be difficult to determine exactly which aspects of climate affect populations and at what vital rates.
-Tests of whether statistical models fit to climate and population data can accurately predict population responses to natural and experimental climate manipulations have rarely been conducted.
+\item Climate is thought to be an important driver of population ecology however it can be difficult to determine exactly which aspects of climate affect populations and at which stages in an organisms life-cycle.
+Moreover, there have been few tests of whether statistical models fit to historical climate and population data can accurately predict the future response of populations to natural and experimentally imposed climate conditions.
 \item Using long-term observational data collected from the 1920's to the 1950's from a sagebrush steppe plant community, previous research has reported a wide variety of climate effects on four dominant plant species.
-We tested whether statistical models fit to the historical data could predict the population response of plants in these original plots to natural and experimentally imposed climate variation some 60 years after the historical data collection ended.
-New plots in the same location where the historical data were collected were assigned to either drought or irrigation treatments imposed using rainfall shelters and automated sprinklers.
+We tested whether statistical models fit to these historical observational data could predict the population response of these plant populations to natural and experimentally imposed climate variation some 60 years after the historical data collection ended.
+\\item We established sixteen new plots in the same location as the historical data and assigned each to either drought or irrigation treatments imposed using rainfall shelters and automatic sprinklers.
 The original plots were also monitored and served as a control.
-We modeled survival, growth and recruitment as functions of spring temperature and spring and summer soil moisture.
-We then used statistical models fitted to the historical data to predict the year and treatment effects observed in contemporary data.
-\item  For survival and recruitment, the treatment effects were never significantly positive, indicating that competitive release was not underestimated by the interspecific effects included in the baseline model.
-For  \textit{Poa secunda} recruitment, the removal treatment was actually negative, indicating that our baseline model overestimated competitive release.
+We fit statistical models for survival, growth and recruitment of each species as functions of seasonal temperature and soil moisture in the historical data.
+We then used these models to predict the treatment effects observed in contemporary data.
+\item  For   \textit{Poa secunda} recruitment, the removal treatment was actually negative, indicating that our baseline model overestimated competitive release.
 \item For growth, the removal treatment effect was significant and positive for two species, \textit{Poa secunda} and \textit{Pseudoroegneria spicata}, indicating that the baseline model underestimated competitive release.
@@
 Air temperature was measured underneath the roofing of the shelter at 30 cm above ground.
+We also installed sensors in a nearby area receiving ambient rainfall to serve as a measure of un-manipulated control conditions.
 Data were logged automatically every four hours.
@@
 We conducted a simple statistical analysis to determine the effect of the rain-out shelters and irrigation on the observed soil moisture.
+We averaged soil moisture readings by day plot group  and then fit a simple linear model to the soil moisture data to determine the effects
 

### 2016-12-13 precip/analysis/figure_scripts/plot_one_step_ahead_cover.R

@@ -32,8 +32,6 @@ years$Period[ years$year <= 1960 ] <- 'Historical'
 species_list <- c('ARTR', 'HECO', 'POSE', 'PSSP')
 
 ylims <- list( c(0,35), c(0,7.5), c(0,7.5), c(0,7.5))
-i = 1
-
 
 for( i in 1:4) {  
   spp   <- species_list[i]  
@@ -64,10 +62,15 @@ for( i in 1:4) {
     gather( simulation, size , matches('^X[0-9]')) %>% 
     group_by( year, Treatment, simulation, quad ) %>% 
     summarize( total_size = sum( size )) 
+  
+  rdf <- 
+    rdf %>% 
+    gather( simulation, rec_area, matches('^X[0-9]')) 
+  
+  saveRDS( merge(rdf[ , c('year', 'Treatment', 'simulation', 'rec_area', 'quad')], cover1, all.x = T), paste0( 'output/ibm/simulations/', spp, '_one_step_ahead_climate_model_cover_per_quad.RDS'))
 
   cover2 <- 
     rdf %>% 
-    gather( simulation, rec_area, matches('^X[0-9]')) %>% 
     group_by( year, Treatment, simulation) %>% 
     summarise( rec_area = mean(rec_area))
   
@@ -104,7 +107,7 @@ for( i in 1:4) {
     summarise( true_cov = 100*(mean(totCover )/10000)) 
   
   # ------------------------------------------------------------------------------------------------------ #
-  df <- read.csv('data/temp_data/ARTR_growth_and_survival_cleaned_dataframe.csv')
+  df <- read.csv(paste0('data/temp_data/', spp, '_growth_and_survival_cleaned_dataframe.csv'))
   
   df$true_cov <- 100*(exp(df$logarea.t0)/10000)
   
@@ -123,58 +126,40 @@ for( i in 1:4) {
   
   plot_cover <-
     merge( obs_cover, pred_cover, all.x = TRUE) %>%
-    mutate( predicted = ifelse(Treatment != 'Control' & year == 2011, observed, predicted )) %>%
-    mutate( predicted = ifelse(year == 2007, observed, predicted )) %>%
-    mutate( predicted = ifelse(year == 1929, observed, predicted )) %>%
-    mutate( predicted = ifelse(year == 1945, observed, predicted )) %>%
-    mutate( predicted = ifelse(year == 1949, observed, predicted )) %>%
[…]

### 2016-12-13 precip/analysis/figure_scripts/plot_predicted_vs_observed_cover_changes.R

@@ -1,0 +1,85 @@
+# 
+library(dplyr)
+library(tidyr)
+library(stringr)
+
+rm(list = ls() ) 
+spp = 'ARTR'
+
+# get last year of cover which is not in the survival dataframe ------------------------------------- #   
+oldCover <- read.csv(paste0( '~/driversdata/data/idaho/speciesData/', spp, '/quadratCover.csv'))
+newCover <- read.csv(paste0( '~/driversdata/data/idaho_modern/speciesData/', spp, '/quadratCover.csv'))
+oldCover$Period <- "Historical"
+newCover$Period <- "Modern"
+last_cover  <- rbind(oldCover, newCover)
+quad <- read.csv('~/driversdata/data/idaho_modern/quad_info.csv')
+
+last_cover <-
+  last_cover %>%
+  left_join(quad) %>%
+  mutate( quad = as.numeric(str_extract(pattern = '[0-9]+$', quad))) %>% 
+  filter( Treatment %in% c('Control', 'Drought', 'Irrigation')) %>%
+  mutate( year = ifelse(year < 100, year + 1900, year)) %>% 
+  mutate( observed = totCover)
+
+# ------------------------------------------------------------------------------------------------------ #
+df <- read.csv(paste0('data/temp_data/', spp, '_growth_and_survival_cleaned_dataframe.csv'))
+
+df$true_cov <- exp(df$logarea.t0)
+
+trueCov <- 
+  df %>% 
+  group_by( Period, year, Treatment, quad ) %>% 
+  summarise( observed = sum( true_cov ) )
+
+obs_cover <- merge( last_cover, trueCov, by = c('Period', 'year', 'Treatment', 'quad'), all.x = T)
+
+obs_cover <- obs_cover %>% mutate( observed = ifelse( is.na(observed.y), observed.x, observed.y)) # fill in data from end years in dataframe
+
+
+# get predicted cover -------------------------------------------------------------------------------------# 
+pred <- readRDS('output/ibm/simulations/ARTR_one_step_ahead_climate_model_cover_per_quad.RDS')
+pred$total_size[ is.na(pred$total_size  ) ]  <- 0 
+
+pred <- 
+  pred %>% 
+  mutate( predicted = rec_area + total_size ) %>%  
+  arrange( simulation, Treatment, quad, year) %>% 
[…]

### 2016-12-13 precip/analysis/figure_scripts/plot_treatment_stan_fits.R

@@ -30,8 +30,7 @@ plot_posterior_year_effects <- function(df){
 setwd('~/Documents/ExperimentTests/precip/')
 
 mfiles <- dir('output/stan_fits', '.*_treatment_fit.RDS', full.names = TRUE)
-i = 1
-
+i = 7
 for( i in 1:length(mfiles)){ 
   
   bname <- basename(mfiles[i])
@@ -39,7 +38,7 @@ for( i in 1:length(mfiles)){
 
   spp <- mpars[1]
   vr <- mpars[2]
-    
+   
   temp_fit <- readRDS(mfiles[i])
   
   df <- readRDS(paste0( 'data/temp_data/', vr, '_data_lists_for_stan.RDS'))[[spp]]

### 2016-12-13 precip/analysis/fit_stan_climate_models.R

@@ -14,11 +14,7 @@ for( i in 1:length(dl_files)){
   
   for( j in 1:length(dl)){ 
     
-    if(is.null( ncol(dl[[j]]$C)) ){ 
-      myfit <- stan(str_replace(m_files[i], '.stan', '_special.stan'), data  = dl[[j]], thin = 4, cores = 4, iter = 2000, seed = 1)
-    }else{
-      myfit <- stan(m_files[i], data  = dl[[j]], thin = 4, cores = 4, iter = 2000, seed = 1)
-    }
+    myfit <- stan(m_files[i], data  = dl[[j]], thin = 4, cores = 4, iter = 2000, seed = 1)
     
     ss <-  get_sampler_params(myfit) 
     

### 2016-12-13 precip/analysis/fit_stan_treatment_models.R

@@ -2,7 +2,7 @@ rm(list = ls())
 library(rstan)
 
 df <- expand.grid(species = c('ARTR', 'HECO', 'POSE', 'PSSP'), vital_rate = c('growth', 'recruitment', 'survival'))
-
+i = 1
 for(i in 1:nrow(df)){ 
   
   spp <- df$species[i]
@@ -18,7 +18,8 @@ for(i in 1:nrow(df)){
   
   if ( dv > 0 ) { 
     # try again if divergence 
-    myfit <- stan( fit = myfit, data = dat, cores = 4, iter = 4000, thin = 8, seed = 1)
+    inits <- apply(myfit, 2, relist, skeleton = rstan:::create_skeleton(myfit@model_pars, myfit@par_dims)) # find initial values from end of chain 
+    myfit <- stan( fit = myfit, init = inits, data = dat, cores = 4, iter = 4000, thin = 8, seed = 1)
   }
   
   ss <-  get_sampler_params(myfit) 
@@ -27,7 +28,8 @@ for(i in 1:nrow(df)){
   
   if ( dv > 0 ){ 
     # try again if divergent 
-    myfit <- stan( fit = myfit, data = dat, cores = 4, iter = 2000, thin = 4, 
+    inits <- apply(myfit, 2, relist, skeleton = rstan:::create_skeleton(myfit@model_pars, myfit@par_dims))
+    myfit <- stan( fit = myfit, init = inits, data = dat, cores = 4, iter = 2000, thin = 4, 
                   control = list(adapt_delta = 0.85, stepsize = 0.8, max_treedepth = 20), seed = 1 )
   } 
   
@@ -37,8 +39,9 @@ for(i in 1:nrow(df)){
   
   if ( dv > 0 ){ 
     # try again if divergent 
-    myfit <- stan( fit = myfit, data = dat, cores = 4, iter = 4000, thin = 8, 
-                   control = list(adapt_delta = 0.99, stepsize = 0.2, max_treedepth = 20), seed = 1 )
+    inits <- apply(myfit, 2, relist, skeleton = rstan:::create_skeleton(myfit@model_pars, myfit@par_dims)) 
+    myfit <- stan( fit = myfit, init = inits, data = dat, cores = 4, iter = 4000, thin = 8, 
+                   control = list(adapt_delta = 0.99, stepsize = 0.1, max_treedepth = 20), seed = 1 )
   } 
   
   saveRDS(myfit, paste0('output/stan_fits/', spp, '_', vr, '_treatment_fit.RDS'))

### 2016-12-13 precip/analysis/growth/import2ipm_noOverlap.r

@@ -1,0 +1,53 @@
+# Import and format growth parameters
+# then define growth function
+
+# growth parameters
+Gpars=list(intcpt=rep(NA,Nspp),intcpt.yr=matrix(0,Nyrs,Nspp),intcpt.gr=matrix(0,6,Nspp),
+  slope=rep(NA,Nspp),slope.yr=matrix(0,Nyrs,Nspp),
+  nb=matrix(0,Nspp,Nspp),sigma2.a=rep(NA,Nspp),sigma2.b=rep(NA,Nspp))
+for(i in 1:Nspp){
+
+  infile=paste0("growth/",sppList[i],"_growth.csv")
+
+  Gdata=read.csv(infile)
+  # main intercept
+  Gpars$intcpt[i]=Gdata$X.Intercept.[1]
+  # random year effects on intercept
+  Gpars$intcpt.yr[,i]=Gdata$Intercept.yr
+  # group effects
+  tmp=which(names(Gdata)=="Group")
+  if(length(tmp)>0) Gpars$intcpt.gr[,i]=Gdata$Group[!is.na(Gdata$Group)] 
+  # size slope
+  Gpars$slope[i]=Gdata$logarea.t0[1]
+  # random effects on slope
+  tmp=which(names(Gdata)=="logarea.yr")
+  if(length(tmp)>0) Gpars$slope.yr[,i]=Gdata[,tmp]
+  # get competition coefficients
+  tmp=paste("W",sppList,sep=".")
+  tmp=which(is.element(names(Gdata),tmp))
+  if(length(tmp)>0) Gpars$nb[i,]=as.numeric(Gdata[1,tmp])
+  # Treatment effects
+  if(trtEffects==T){
+    tmp=grep("Treatment",names(Gdata))
+    if(max.CI==T) {
+      Gpars$intcpt[i]=Gpars$intcpt[i] + Gdata$TreatMaxCI[1] # add max treatment intercept
+    }else{
+      Gpars$intcpt[i]=Gpars$intcpt[i] + Gdata[1,tmp[1]] # add treatment intercept
+    }
+  }
+  # variance parameters
+  Gpars$sigma2.a[i]=Gdata$sigma.a[1]
+  Gpars$sigma2.b[i]=Gdata$sigma.b[1]
+} # next i
+rm(Gdata)
+
+# growth function
[…]

### 2016-12-13 precip/analysis/ibm/summarize_validate_sims1step.r

@@ -1,0 +1,215 @@
+
+# call from validate wrapper
+setwd("ibm/")
+
+sppList=c("ARTR","HECO","POSE","PSSP")
+Nspp=length(sppList)
+sppNames=c("A. tripartita","H. comata","Poa secunda","P. spicata")
+myCol=c("black","forestgreen","blue","red")
+
+# function to load observed and predicted cover
+getSims<-function(qList,trtEffects){
+   out=NULL
+   for(i in qList){
+     if(trtEffects==F){
+       infile=paste("simulations1step/",i,"_validation_cov_removals_noTrt.csv",sep="")
+     }else{
+       infile=paste("simulations1step/",i,"_validation_cov_removals_Trt.csv",sep="")
+     }
+     tmpD=read.csv(infile)
+     out <- rbind(out,tmpD)
+   }
+   return(out)
+}
+
+# function to load data and calculate log(cover_t+1 / cover_t)
+getPopGrowth<-function(qList,trtEffects){
+   out=NULL
+   for(i in qList){
+     if(trtEffects==F){
+       infile=paste("simulations1step/",i,"_validation_cov_removals_noTrt.csv",sep="")
+     }else{
+       infile=paste("simulations1step/",i,"_validation_cov_removals_Trt.csv",sep="")
+     }
+     tmpD=read.csv(infile)
+     # make sure that the denominator (t0) is always the observed cover, not predicted cover
+     change=data.frame(year=tmpD$year[1:(NROW(tmpD)-1)],
+            log(tmpD[2:NROW(tmpD),2:NCOL(tmpD)]/cbind(tmpD[1:(NROW(tmpD)-1),2:5],tmpD[1:(NROW(tmpD)-1),2:5])))
+     change[change==Inf]<-NA; change[change==-Inf]<-NA
+     out=rbind(out,change)
+   }
+   return(out)
+}
+
+# control plots
+qList <- paste0("Q",c(1:6,19:26))  # all contemporary control plots#
+#qList <- c("Q1","Q2","Q3","Q4","Q5","Q6" ) # just Group 1 
+quadD <- getSims(qList,trtEffects=F)
[…]

### 2016-12-13 precip/analysis/ipm/IPM-getEquilibrium.r

@@ -1,0 +1,100 @@
+# Multispecies, spatially implicit IPM
+# This version makes it possible to assume "no overlap"
+# for intraspecific competition only or intra- and interspecific competition
+
+#============================================================================================#
+# Calculate the equilibrium areas.
+#============================================================================================# 
+
+## initial population density vector
+nt=v
+for(i in 1:Nspp) nt[[i]][]=0  # make sure all species = 0
+for(i in init.species) nt[[i]][]=0.1  # give init.species density > 0
+new.nt=nt
+
+# set up matrix to record cover
+covSave = matrix(NA,tlimit,Nspp)
+covSave[1,]=sumCover(v,nt,h,A)
+
+# set up list to store size distributions
+sizeSave=list(NULL)
+for(i in 1:Nspp){
+  sizeSave[[i]]=matrix(NA,length(v[[i]]),(tlimit))
+  sizeSave[[i]][,1]=nt[[i]]/sum(nt[[i]])
+}
+
+# initial densities 
+Nsave=matrix(NA,tlimit,Nspp)
+Nsave[1,]=sumN(nt,h)
+
+yrSave=rep(NA,tlimit)
+for (i in 2:(tlimit)){
+  
+  #draw from observed year effects
+  doYear=sample(c(1:Nyrs),1)
+  yrSave[i]=doYear
+  
+  #get recruits per area
+  cover=covSave[i-1,] ; N=Nsave[i-1,]
+  rpa=get.rpa(Rpars,cover,doYear)
+  
+  #calculate size-specific crowding
+  for(ii in 1:Nspp){ 
+    
+    # first do all overlap W's
+    Xbar=cover*A/N       # multiply by A to get cover back in cm^2
+    varX=varN(v,nt,h,Xbar,N) 
+    muW = Xbar*N*W.constant/A # divide by A to get units in m^2 per m^2, not cm^2
+    muW[is.na(muW)]=0
+    Wmat[[ii]]=matrix(muW,nrow=length(v[[ii]]),ncol=Nspp,byrow=T)
+    
+    #now do conspecific no overlap W
+    Ctot[ii]=h[ii]*sum(expv[[ii]]*nt[[ii]]) 
[…]

### 2016-12-13 precip/analysis/ipm/get_W_functions.r

@@ -1,0 +1,38 @@
+
+# call from removal_analysis_wrapper.r
+
+distWts <- read.csv("~/driversdata/data/idaho_modern/speciesData/IdahoModDistanceWeights_noExptl.csv")
+#distWts <- read.csv(paste0(root,"/driversdata/data/idaho/speciesdata/IdahoDistanceWeights.csv"))
+
+# make distWts constant over first annulus (0 to 2 cm)
+distWts <- rbind(distWts,c(1,1,1,1,0))
+distWts <- distWts[order(distWts$midRings),]
+
+Wfuns <- list(4) # store approximate spline functions
+zc <- numeric(4) # store distance at which spline function goes to zero
+W.constant <- rep(0,4) # store constant for heterospecific mean field approximation
+for(i in 1:4){
+  
+  Wfuns[[i]] <- approxfun(x=distWts$midRings, y=distWts[,i], rule=2) # W as a continuous function
+
+  tmp <- which(distWts[,i]==0)
+  if(length(tmp)>0){
+    zc[i] <- distWts$midRings[min(tmp)]
+  }else{
+    zc[i] <- 150  # max distance
+  }
+  
+  # if W is continuous
+  W.constant[i] <- 2*pi*integrate(f=function(z) z*Wfuns[[i]](z), lower=0, upper=zc[i])$value
+  
+}
+names(Wfuns) <- names(distWts[,1:4])
+
+# take a look at the functions
+# x<-seq(0,25,0.1)
+# y<-Wfuns[[1]](x)
+# plot(x,y,type="l")
+# for(i in 2:4){
+#   lines(x,Wfuns[[i]](x))
+# }
+

### 2016-12-13 precip/analysis/modify_climate_datalists.R

@@ -1,7 +1,7 @@
 rm(list = ls())
 
 var_table <- read.csv('output/selected_climate_covariates.csv')
-i  = 1
+
 for( i in 1:nrow(var_table)){ 
   
   spp    <- var_table$species[i]

### 2016-12-13 precip/analysis/precip_analysis_wrapper.R

@@ -1,0 +1,215 @@
+
+# call all scripts for precip experiment analysis
+
+rm(list=ls(all=TRUE))
+graphics.off();
+
+###
+### 1. get treatment trends #################################
+###
+
+library(texreg) # to save output
+library(lme4)
+
+statsOutput <- paste0(getwd(),"/stats_tables.tex")
+#source("treatment_trends_removals.r")
+
+# make climate figure
+#source("climate_fig.r")
+
+# clean up
+tmp=ls() ; tmp=tmp[tmp!="root" & tmp!="statsOutput"]
+rm(list=tmp)
+
+###
+### 2. fit vital rate regressions ###########################
+###
+
+library(texreg) # to save output
+library(xtable)
+library(lme4)
+library(INLA)
+
+# table to store Treatment effects
+trtTests <- data.frame("species"="c","stage"="c","effect"=1,"CI.02.5"=1,"CI.97.5"=1,stringsAsFactors = F)
+
+# read in distance weights
+dists <- read.csv("~/driversdata/data/idaho_modern/speciesData/IdahoModDistanceWeights_noExptl.csv")
+
+###
+### fit survival models (takes ~ 10 minutes)
+###
+
+###
+### fit growth models (takes < 5 mins)
+###
+
+
+###
+### fit recruitment models (this can take a few hours)
+###
+
+
+###
+### write results to file
+###
+
+
+# make treatment effect figure
+
+# clean up
+tmp=ls() ; tmp=tmp[tmp!="root" & tmp!="statsOutput"]
+rm(list=tmp)
+
+###
+### 3. explore neighborhood composition ###################################
+###
+
+
+# clean up
+tmp=ls() ; tmp=tmp[tmp!="root" & tmp!="statsOutput"]
+rm(list=tmp)
+
+###
+### 4. get IBM predictions for quadrat cover ###############################
+###
+
+sppList <-  c("ARTR","HECO","POSE","PSSP")
+
+# read in distance weights
+dists <- read.csv("~/driversdata/data/idaho_modern/speciesData/IdahoModDistanceWeights_noExptl.csv")
+
+#max.CI <- F  # TRUE means use maximum removal effect
+
+# do contemporary control plots
+quadList <- paste0("Q",c(1:6,19:26))  
[…]

### 2016-12-13 precip/analysis/recruitment/import2ipm.r

@@ -1,0 +1,81 @@
+# import parameters
+# recruitment parameters
+Rpars=list(intcpt.mu=rep(0,Nspp),intcpt.yr=matrix(0,Nyrs,Nspp),intcpt.trt=matrix(0,3,Nspp),intcpt.tau=rep(100,Nspp),
+  intcpt.gr=matrix(NA,6,Nspp),g.tau=rep(NA,Nspp),
+  dd=matrix(NA,Nspp,Nspp),theta=rep(NA,Nspp),sizeMean=rep(NA,Nspp),sizeVar=rep(NA,Nspp),
+  recSizes=list(1))
+
+#   if(trtEffects==T){
+  infile=paste0("recruitment/recruit_params_m1.csv")
+#   }else{
+#     infile=paste0("recruitment/recruit_params_m0.csv")
+#   }   
+ Rdata=read.csv(infile)
+ 
+ # subset out non-essential parameters
+ tmp=c(grep("lambda",row.names(Rdata)),grep("deviance",row.names(Rdata)),
+  grep("DIC",row.names(Rdata)))   
+ Rdata=Rdata[-tmp,]
+ 
+ # format parameters
+ tmp=paste("Rpars$",row.names(Rdata),"<-",Rdata[,1],sep="")
+ eval(parse(n=dim(Rdata)[1],text=tmp))
+ for(i in 1:Nspp){
+   infile=paste(root,"/driversdata/data/idaho/speciesData/",sppList[i],"/recSize.csv",sep="")
+   recSize=read.csv(infile)
+   Rpars$sizeMean[i]=mean(log(recSize$area))
+   Rpars$sizeVar[i]=var(log(recSize$area))
+   #Rpars$recSizes[[i]]=recSize$area
+ }
+Rpars$dd=t(Rpars$dd) # c[i,j] = effect of j on i
+
+# get edge of confidence interval for treatment effects
+if(max.CI==T){
+  keep <- grep("intcpt.trt",row.names(Rdata))
+  tmp <- Rdata[keep,c("X2.5.","X97.5.")]
+  tmp$max <- ifelse(abs(tmp$X2.5.)>abs(tmp$X97.5.),tmp$X2.5.,tmp$X97.5.)
+  Rpars$intcpt.trt[2,2] <- tmp$max[which(row.names(tmp)=="intcpt.trt[2,2]")]
+  Rpars$intcpt.trt[2,3] <- tmp$max[which(row.names(tmp)=="intcpt.trt[2,3]")]
[…]

### 2016-12-13 precip/analysis/survival/import2ipm_noOverlap.r

@@ -1,0 +1,52 @@
+# import & format survival parameters
+# then define survival function
+
+# survival parameters
+Spars=list(intcpt=rep(NA,Nspp),intcpt.yr=matrix(0,Nyrs,Nspp),intcpt.gr=matrix(0,6,Nspp),
+  intcpt.trt=rep(0,Nspp),slope=rep(NA,Nspp),slope.yr=matrix(0,Nyrs,Nspp),
+  nb=matrix(0,Nspp,Nspp))
+
+for(i in 1:Nspp){
+  
+  infile=paste0("analysis/survival/",sppList[i],"_surv.csv")
+  
+  Sdata=read.csv(infile)
+  # main intercept
+  Spars$intcpt[i]=Sdata$X.Intercept.[1]
+  # Treatment effect
+  if(trtEffects==T){
+    tmp=grep("Treatment",names(Sdata))
+    if(max.CI==T) {
+      Spars$intcpt[i]=Spars$intcpt[i] + Sdata$TreatMaxCI[1] # add max treatment intercept
+    }else{
+      Spars$intcpt[i]=Spars$intcpt[i] + Sdata[1,tmp[1]] # add treatment intercept
+    }
+  }
+  # Group effects
+  tmp=which(names(Sdata)=="Group")
+  if(length(tmp)>0) Spars$intcpt.gr[,i]=Sdata$Group[!is.na(Sdata$Group)] # get spatial average
+  # random year effects
+  Spars$intcpt.yr[,i]=Sdata$Intercept.yr
+  Spars$slope[i]=Sdata$logarea[1]
+  # random effects on slope
+  tmp=which(names(Sdata)=="logarea.yr")
+  if(length(tmp)>0) Spars$slope.yr[,i]=Sdata[,tmp]
+  # get competition coefficients
+  tmp=paste("W",sppList,sep=".")
+  tmp=which(is.element(names(Sdata),tmp))
+  if(length(tmp)>0) Spars$nb[i,]=as.numeric(Sdata[1,tmp])
+} # next i
+Spars$yrList=Sdata$year
+rm(Sdata)
+
+
+## survival function: probability an individual of size u survives  (u is on log scale)
+S=function(u,W,Spars,doYear,doSpp){
+  mu=Spars$intcpt[doSpp]+Spars$intcpt.yr[doYear,doSpp]+Spars$intcpt.trt[doSpp]+
[…]

### 2016-12-14 precip/manuscript/precip_ms.tex
adding edits to the abstract

@@
 
-\title{Can historical data predict ecological responses to novel climate conditions?}
+\title{Can historical data predict population responses to climate change experiments?}
 
@@
 \begin{enumerate}
-\item Climate is thought to be an important driver of population ecology however it can be difficult to determine exactly which aspects of climate affect populations and at which stages in an organisms life-cycle.
-Moreover, there have been few tests of whether statistical models fit to historical climate and population data can accurately predict the future response of populations to natural and experimentally imposed climate conditions.
-\item Using long-term observational data collected from the 1920's to the 1950's from a sagebrush steppe plant community, previous research has reported a wide variety of climate effects on four dominant plant species.
-We tested whether statistical models fit to these historical observational data could predict the population response of these plant populations to natural and experimentally imposed climate variation some 60 years after the historical data collection ended.
+\item Climate is thought to be an important driver of population ecology however it can be difficult to determine which aspects of climate affect populations and at which stages in an organisms life-cycle.
+Moreover, there have been few tests of whether statistical models fit to observational data on climate and populations can accurately predict the future responses of populations to natural or experimentally imposed climate conditions.
+\item Using observational data collected from the 1920's to the 1950's from a sagebrush steppe plant community, previous research has reported a wide variety of climate effects on four dominant plant species.
+We tested whether statistical models based on these historical data could predict how these species respond to natural and experimentally imposed climate variation from 2006 to 2016, some 60 years after the historical data collection ended.
 \\item We established sixteen new plots in the same location as the historical data and assigned each to either drought or irrigation treatments imposed using rainfall shelters and automatic sprinklers.
-The original plots were also monitored and served as a control.
-We fit statistical models for survival, growth and recruitment of each species as functions of seasonal temperature and soil moisture in the historical data.
-We then used these models to predict the treatment effects observed in contemporary data.
-\item  For   \textit{Poa secunda} recruitment, the removal treatment was actually negative, indicating that our baseline model overestimated competitive release.
-\item For growth, the removal treatment effect was significant and positive for two species, \textit{Poa secunda} and \textit{Pseudoroegneria spicata}, indicating that the baseline model underestimated competitive release.
-However, including information about the location of these grass individuals with respect to removed \textit{Artemisia} failed to improve the growth regressions, raising questions about the mechanisms driving the positive response to removal.
-\item For three of the four species, individual based models and integral projections models based on the vital rate regressions showed that removal treatment effects had little impact on population growth rates or simulated equilibrium cover following a species removal.
-For \textit{P.
-spicata}, the population models that included effects of \textit{Artemisia} removal projected greater competitive release than the baseline models.
-\item \emph{Synthesis}: Although models based on observational data did not perfectly predict species responses to removal, overall the results increase our confidence that competitive release is in fact small for at least three of our four study species.
-The implication is that interspecific resource competition is not the primary factor determining the dynamics of this community.
+The original plots were also monitored and served as ambient climate controls.
+Using the historical data, we fit statistical models for survival, growth and recruitment of each species in response to seasonal temperature and soil moisture.
+We used these models to predict variation in species vital rates in the experiment and used a simple individual based population model to predict one year ahead population growth rates in each of the experimental plots.
+\item Over the course of the experiment, average plot cover of \textit{Hesperostipa comata} and \textit{Pseudoroegneria spicata} declined, while cover of \textit{Poa secunda} showed a trend towards increase in the irrigated plots.
+At the level of individual vital rates, experimental drought led to decreases in \textit{} growth, survival, while irrigation led to increases in \textit{}.
+\item Vital rate models fit to the historical data correctly predicted significant increases in X and X and significant decreaes in X and X.
+However, the historical models predicted significant changes in X X and X that were not observed in the experiment.
+Moreover, the experiment led to a significant change in X that was not predicted by the historical data.
+\item Our individual based population model more accurately predicted changes in the population growth rates of \textit{} in the experimental plots than a model without climate effects.
+However for the other X species predictions made from the climate model were less accurate than models without climate.
+\item \emph{Synthesis}: Out of all four species examined, climate covariates only appeared to increase our predictive accuracy for one species: \textit{Hesperostipa comata}.
+In general the relatively weak responses to the experimental treatments agrees with the low explanatory power of climate covariates in explaining the vital rates and population dynamics of these species in this system.
+The highly variable environment and water limited conditions may select for species with a conservative non-responsive strategy to variation in climate both natural and experimentally imposed.
 

### 2016-12-16 precip/manuscript/precip_ms.tex
adding edits to the introduction of the manuscript

@@
 \item Using observational data collected from the 1920's to the 1950's from a sagebrush steppe plant community, previous research has reported a wide variety of climate effects on four dominant plant species.
-We tested whether statistical models based on these historical data could predict how these species respond to natural and experimentally imposed climate variation from 2006 to 2016, some 60 years after the historical data collection ended.
+We tested whether statistical models based on these observational data could predict how these species respond to a experimentally imposed drought and irrigation from 2006 to 2016.
 \\item We established sixteen new plots in the same location as the historical data and assigned each to either drought or irrigation treatments imposed using rainfall shelters and automatic sprinklers.
 The original plots were also monitored and served as ambient climate controls.
-Using the historical data, we fit statistical models for survival, growth and recruitment of each species in response to seasonal temperature and soil moisture.
-We used these models to predict variation in species vital rates in the experiment and used a simple individual based population model to predict one year ahead population growth rates in each of the experimental plots.
-\item Over the course of the experiment, average plot cover of \textit{Hesperostipa comata} and \textit{Pseudoroegneria spicata} declined, while cover of \textit{Poa secunda} showed a trend towards increase in the irrigated plots.
-At the level of individual vital rates, experimental drought led to decreases in \textit{} growth, survival, while irrigation led to increases in \textit{}.
-\item Vital rate models fit to the historical data correctly predicted significant increases in X and X and significant decreaes in X and X.
+In order to describe the demographic effects of the climate manipulations, we used all data, observational and experimental, to fit a set of statistical models describing each species survival, growth and recruitment in response to local competitive interactions and the effects of the climate manipulation experiment.
+Next we fit another set of models to only the observational data that used temperature and soil moisture as covariates to describe annual variation in the vital rates.
+Finally, we used the second set of models fitted only to the observational data, to predict the effects of the experimental treatments on each species vital rates and used the vital rates to parameterize an individual based population model and predict one step ahead cover in each of the experimental plots.
+\item The experimental drought and irrigation treatments successfully lowered and increased soil moisture respectively.
+Over the course of the experiment, average plot cover of \textit{Hesperostipa comata} and \textit{Pseudoroegneria spicata} declined, while cover of \textit{Poa secunda} showed a trend towards increase in the irrigated plots.
+At the level of individual vital rates, experimental drought reduced \textit{Hesperostipa comata} and \textit{Poa secunda} growth and survival.
+Drought increased and irrigation decreased \textit{Artemisia tripartita} growth.
+Recruitment rates for all three grasses were lower in both the drought and irrigation plots than in the control plots.
+\item Vital rate models fit to the observational data correctly predicted significant increases in X and X and significant decreaes in X and X.
 However, the historical models predicted significant changes in X X and X that were not observed in the experiment.
@@
 Climate is widely considered one of the most powerful external forces driving changes in species abundance across space and time \citep{teller_linking_2016}.
-The effects of climate on populations and ecosystems are most apparent at the largest scales in space and time: climate determines the distribution of ecosystems, treelines and the the range limits of many species \citep{}.
-The paleoecological record shows that long-term climate change lead to changes in species abundance and changes in species' ranges.
-Nevertheless, for many species it is difficult to determine how annual climate variation affects their populations and demographic rates \citep{ehrlen_advancing_2016}.
-Observational data on species performance across many different years can provide some information on how climate might affect population abundance \citep{dalgleish_climate_2010,ehrlen_advancing_2016}.
-But many years of data are needed to reliably detect climate effects, especially when climate effects are weak and often correlated amongst themselves \citep{teller_linking_2016,gerber_optimal_2015}.
+The effects of climate on populations and ecosystems are most apparent at the largest scales in space and time: climate determines the distribution of ecosystems, treelines and the the range limits of many species \citep{}, while The paleoecological record shows that long-term climate change lead to changes in species abundance and changes in species' ranges \citep{davis}.
+Nevertheless, for many species it is difficult to determine how annual climate variation affects populations and demographic rates \citep{ehrlen_advancing_2016}.
+Observational data on species performance across years with varying climate can provide some information on how climate might affect population abundance \citep{dalgleish_climate_2010,ehrlen_advancing_2016}.
+However, many years of data are needed to reliably detect climate effects, especially when annual variation in demographic rates is high \citep{teller_linking_2016,gerber_optimal_2015}.
 Understanding the effects of climate on populations is increasingly important, however, if we are to understand and predict the effects of climate change on populations.
-If models fit to historical data can be used to accurately predict the effects of contemporary climate variation on populations, as well as the effects of experimental climate manipulations, it would be strong confirmation that the climate effects they describe are real.
+If models fit to historical data can be used to accurately predict the effects of experimental climate manipulations, it would be strong confirmation that the climate effects they describe are real and will hold in the future \citep{}.
+
+Many plant species occur across a wide range of climates \cite{} and individual plants in many environments must tolerate large fluctuations in seasonal temperature and soil moisture.
+Unlike most animals, plants must endure these conditions in place and often become dormant in less optimal seasons \textit{}.
+At first pass, these observation would seem to indicate that annual variation in climate should have little effect on plant population performance.
+
+Nevertheless, there is abundant evidence that plant performance shows high year to year variation, both at the level of individual growth, survival and reproduction and in terms of total population abundancend often this variation c.
+In many regions interannual variation in precipitation can be directly linked to variation in net primary productivity \citep{knapp and smith, Hsu and Adler}.
+The growth rates of trees are also often tightly linked to annual precipitation, so much so that annual growth rings in their stems can serve as accurate records of historical climate thousands of years before the present \citep{yang}.
+Likewise, many smaller shrub and sub-shrub species show strong variation in growth in response to climate that is recorded in their tissues \citep{Franklin, Srur}.
+Among annual plants in desert ecosystems, germination and reproductive output is tightly linked to interannual precipitation variation \citep{venables}.
+
+Linking annual climate variation to demographic performance in plants is a high priority for building population models for plants that can predict their future response to climate change \citep{crone_ability_2013,ehrlen_advancing_2016}.
+However, despite the clear signs that climate drives net primary productivity at the ecosystem level as and variation in individual growth rate in trees and other woody species, as well as the germination and fecundity of annual plants, there are few studies that have linked the effects of interannual climate variation to population models for plants, and fewer still that have shown that these population models can be used to accurately predict future plant responses to short term climate variation.
+Adler et al.
+\citep{adler byrn and leiker} showed that population models based on historically observed correlations between plant population growth rates and precipitation did have some predictive power in describing species response to a short-term climate manipulation in a North American grassland.
+Three species showed responses to experimentally imposed drought and irrigation that were well predicted by population models fitted to historical observations.
+However, another three species, showed responses to the experimental conditions that were not well predicted by historical observations.
+The authors suggested that limited replication in the historical data for two of these species and changing competitive conditions in the community may have led to the poor predictions.
+
+The effects of annual climate variation on species in a sagebrush steppe plant community at the US Sheep Experiment Station near Dubois, Idaho have been described in at least X different studies since
+
+Section on climate models in this system.
 

### 2016-12-16 precip/manuscript/precip_ms.tex
adding more manuscript edits to intro and methods

@@
 \begin{enumerate}
-\item Climate is thought to be an important driver of population ecology however it can be difficult to determine which aspects of climate affect populations and at which stages in an organisms life-cycle.
-Moreover, there have been few tests of whether statistical models fit to observational data on climate and populations can accurately predict the future responses of populations to natural or experimentally imposed climate conditions.
-\item Using observational data collected from the 1920's to the 1950's from a sagebrush steppe plant community, previous research has reported a wide variety of climate effects on four dominant plant species.
-We tested whether statistical models based on these observational data could predict how these species respond to a experimentally imposed drought and irrigation from 2006 to 2016.
-\\item We established sixteen new plots in the same location as the historical data and assigned each to either drought or irrigation treatments imposed using rainfall shelters and automatic sprinklers.
+\item Climate is an important driver of population ecology however there have been few tests of whether observational data that links population performance with climate variation can be used to predict the responses of populations to experimentally imposed climate conditions.
+\item Using longterm historical observational data from a sagebrush steppe plant community, previous research has reported a wide variety of climate effects on four dominant plant species.
+We tested whether statistical models based on these observational data could be used to predict how these species respond to a five year drought and irrigation experiment conducted from 2011 to 2016.
+\\item We established sixteen new plots at the same field cite as the original historical data and assigned each to either drought or irrigation treatments imposed using rainfall shelters and automatic sprinklers.
 The original plots were also monitored and served as ambient climate controls.
 In order to describe the demographic effects of the climate manipulations, we used all data, observational and experimental, to fit a set of statistical models describing each species survival, growth and recruitment in response to local competitive interactions and the effects of the climate manipulation experiment.
-Next we fit another set of models to only the observational data that used temperature and soil moisture as covariates to describe annual variation in the vital rates.
-Finally, we used the second set of models fitted only to the observational data, to predict the effects of the experimental treatments on each species vital rates and used the vital rates to parameterize an individual based population model and predict one step ahead cover in each of the experimental plots.
+Next we fit another set of models to only the observational data that used seasonal soil moisture covariates to describe annual variation in the vital rates.
+Finally, we used the observation-based models to predict the effects of the experimental treatments on each species vital rates.
+We also used an individual based population model to predict one year ahead changes in population size in the experimental plots.
 \item The experimental drought and irrigation treatments successfully lowered and increased soil moisture respectively.
 Over the course of the experiment, average plot cover of \textit{Hesperostipa comata} and \textit{Pseudoroegneria spicata} declined, while cover of \textit{Poa secunda} showed a trend towards increase in the irrigated plots.
-At the level of individual vital rates, experimental drought reduced \textit{Hesperostipa comata} and \textit{Poa secunda} growth and survival.
-Drought increased and irrigation decreased \textit{Artemisia tripartita} growth.
-Recruitment rates for all three grasses were lower in both the drought and irrigation plots than in the control plots.
+At the level of individual vital rates, experimental drought reduced growth of \textit{Hesperostipa comata} and growth and survival of \textit{Poa secunda}.
+Drought increased, while irrigation decreased, growth of the shrub \textit{Artemisia tripartita}.
 \item Vital rate models fit to the observational data correctly predicted significant increases in X and X and significant decreaes in X and X.
@@
 Climate is widely considered one of the most powerful external forces driving changes in species abundance across space and time \citep{teller_linking_2016}.
-The effects of climate on populations and ecosystems are most apparent at the largest scales in space and time: climate determines the distribution of ecosystems, treelines and the the range limits of many species \citep{}, while The paleoecological record shows that long-term climate change lead to changes in species abundance and changes in species' ranges \citep{davis}.
-Nevertheless, for many species it is difficult to determine how annual climate variation affects populations and demographic rates \citep{ehrlen_advancing_2016}.
+The effects of climate on populations and ecosystems are most apparent at the largest scales in space and time: climate determines the distribution of ecosystems, treelines and the the range limits of many species \citep{}, while the recent historical and the paleoecological record shows that long-term climate change leads to changes in species range limits \citep{parmesan, davis}.
+Understanding and predicting the effects of climate on populations is an increasingly important goal if we are to understand and predict the effects of climate change on earth's ecosystems.
+
+Unfortunately, for many species it is difficult to determine how annual climate variation affects populations and demographic rates \citep{teller_linking_2016, ehrlen_advancing_2016}.
 Observational data on species performance across years with varying climate can provide some information on how climate might affect population abundance \citep{dalgleish_climate_2010,ehrlen_advancing_2016}.
 However, many years of data are needed to reliably detect climate effects, especially when annual variation in demographic rates is high \citep{teller_linking_2016,gerber_optimal_2015}.
-Understanding the effects of climate on populations is increasingly important, however, if we are to understand and predict the effects of climate change on populations.
-If models fit to historical data can be used to accurately predict the effects of experimental climate manipulations, it would be strong confirmation that the climate effects they describe are real and will hold in the future \citep{}.
+Generating predictions from observational historical data for the future novel conditions created by climate change is fraught with risk.
+Climate change will not only change mean annual temperature and precipitation, but also affect annual variation in these measures, and possibly the covariance between them \citep{}.
+Models based on the historical response of populations to annual climate variation, will therefore be extrapolating beyond the range of observed climate variation when they are used to generate predictions in the future \citep{}.
+If models fit to historical data can be used to accurately predict the effects of experimental climate manipulations, especially manipulations that generate extreme variation in climate, it would be strong confirmation that the climate effects they describe are real and will hold in the future \cite{adler_can_the_past_2014}.
 
@@
 Linking annual climate variation to demographic performance in plants is a high priority for building population models for plants that can predict their future response to climate change \citep{crone_ability_2013,ehrlen_advancing_2016}.
-However, despite the clear signs that climate drives net primary productivity at the ecosystem level as and variation in individual growth rate in trees and other woody species, as well as the germination and fecundity of annual plants, there are few studies that have linked the effects of interannual climate variation to population models for plants, and fewer still that have shown that these population models can be used to accurately predict future plant responses to short term climate variation.
+However, despite the clear signs that climate drives net primary productivity at the ecosystem level and the variation in individual growth rate in trees and other woody species, as well as the germination and fecundity of annual plants, there are few studies that have connected the effects of interannual climate variation to population models for plants, and fewer still that have tested whether these population models can be used to accurately predict the future responses of plants to short term climate variation.
 Adler et al.
@@
 
-The effects of annual climate variation on species in a sagebrush steppe plant community at the US Sheep Experiment Station near Dubois, Idaho have been described in at least X different studies since
-
-Section on climate models in this system.
-
-To test whether our models, which are based on observational data, accurately predict the effects of natural climate variation, we conducted a precipitation manipulation experiment in a sagebrush steppe.
-We studied the responses of three perennial grass species and a shrub species to natural climate variation over 10 years and experimental drought and irrigation treatments over four years.
-We addressed two research questions: First, how much do the  survival, growth or recruitment of our target species differ between the precipitation manipulation treatments?
+The demography and competitive interactions between three dominant grass species and a dominant shrub species at in a sagebrush steppe plant community at the US Sheep Experiment Station near Dubois, Idaho have been described in at least X different studies since 2005.
+X of these studies report significant effects of historical climate variation on the vital rates and overall population growth of these species.
+Both precipitation and temperature have been shown to have strong species-specific effects on this system.
+Although past studies used different statistical models and methods for variable selection they indicate X.
+This well-studied system offers the ideal opportunity to test whether the associations between annual climate variation and plant demography we observe in long-term observational data can be used to predict the future responses of plant populations to climate change.
+
+In this study, we report how the four dominant plant species at the USSES responded to a five year drought and irrigation experiment and use the results to address two research questions: first, how much do the growth, recruitment and survival of our target species differ between the precipitation manipulation treatments?
 If our experiment does affect species vital rates we interpret that as strong evidence that variation in precipitation should have an effect on populations.
-Our second question is whether statistical models parameterized from historical observational data can accurately predict the effects of modern climate variation and precipitation manipulation on species performance?
-If models based on observational effectively capture the effects of climate on species performance, they should also predict the effects of natural climate variation in the contemporary dataset as well as the effects of precipitation treatments in the experiment.
+Second, we test whether statistical models parameterized from observational data only can accurately predict each species response to the experimental precipitation manipulation?
+If models based on observational effectively capture the effects of climate on species performance, they should also predict the effects of precipitation treatments in the experiment.
 
@@
 During the period of data collection (1926 – 1957), mean annual precipitation was 270 mm and mean temperatures ranged from -8\degree C (January) to 21\degree C (July).
-The vegetation is dominated by the shrub, \textit{Artemisia tripartita}, and the C3  perennial bunchgrasses \textit{Pseudoroegneria spicata}, \textit{Hesperostipa comata}, and \textit{Poa secunda}.
-These four species, the focus of our models, accounted for over 70\% of basal cover (grasses) and 60\% of canopy cover (shrubs and forbs).
+The vegetation is dominated by a shrub, \textit{Artemisia tripartita}, and three perennial C3 grasses: \textit{Pseudoroegneria spicata}, \textit{Hesperostipa comata}, and \textit{Poa secunda}.
+These dominant species account for over 70\% of basal cover and 60\% of canopy cover at this site.
 
@@
 In 2007, we located 14 of the original quadrats, all of which are inside permanent livestock exclosures, and resumed annual mapped censusing using the traditional pantograph method.
+Daily temperature and precipitation has been monitored throughout this period at a climate station located at the USSES headquarters (station id: GHCND:USC00102707) which located within 2 km of the research plots.
+We downloaded daily and monthly tmin, tmax, and precipitation data from the National Climate Data Centers online database.
 
 We extracted data on survival, growth, and recruitment from the mapped quadrats based on plants' spatial locations.
-Our approach tracks genets, which may be composed of multiple polygons, as they fragment and/or coalesce.
+Our approach tracks genets representing individual plants.
+For the shrub, each genet is associated with the basal position of a stem.
+For the bunchgrasses, each genet represents a spatially distinct polygon in the mapped quadrat.
+These genets may fragment and/or coalesce over time.
 Each mapped polygon is classified as a surviving genet or a new recruit based on its spatial location relative to genets present in previous years \citep{lauenroth_demography_2008}.
-We modeled vital rates using data from 22 year-to-year transitions between 1926 and 1957, and nine year-to-year transitions from 2007 to 2016.
-Only four quadrats were observed for the first two transitions in the 1920's, while at least 14 quadrats were observed for all subsequent transitions.
+We modeled vital rates using data from 21 year-to-year transitions between 1929 and 1957, and four year-to-year transitions from 2007 to 2011.
 
 \subsection*{Precipitation experiment}
-In spring 2011, in a large exclosure containing six of the historical permanent quadrats, we selected locations for an additional 16 quadrats for the precipitation experiment.
+In spring 2011, we selected locations for an additional 16 quadrats for the precipitation experiment.
+We located these in a large exclosure containing six of the historical permanent quadrats.
 We avoided plots falling on hill slopes, areas with greater than 20\% bare rock, or with over 10\% cover of the woody shrubs \textit{Purshia tridentata} or \textit{Amelanchier utahensis}.
@@
 Each rain-out shelter covered an area of 2.5 by 2 m and consisted of transparent acrylic shingles held up over the plot to channel 50\% of incoming rainfall off of the plot and into 75 l reservoirs.
-The collected water was automatically pumped out of the reservoirs and sprayed onto paired irrigation treatment plots.
+The collected water was pumped out of reservoirs and sprayed onto paired irrigation treatment plots.
 Pumping was initiated automatically with float switches that were triggered when water levels in the reservoirs were approximately 20 l, or equivalently irrigation was triggered once for every 6 mm of rainfall collected.
@@
 Air temperature was measured underneath the roofing of the shelter at 30 cm above ground.
-We also installed sensors in a nearby area receiving ambient rainfall to serve as a measure of un-manipulated control conditions.
+For each pair of manipulated plots, we also installed sensors in a nearby area to measure ambient rainfall and temperature.
 Data were logged automatically every four hours.
-
-We conducted a simple statistical analysis to determine the effect of the rain-out shelters and irrigation on the observed soil moisture.
-We averaged soil moisture readings by day plot group  and then fit a simple linear model to the soil moisture data to determine the effects
+We augmented automatic monitoring of the climate in these plots with by taking direct measurements of soil moisture with a handheld EC-5 soil moisture sensor around each plot at X different times in the spring and summer between 2012 and 2016.
+
+Since we only collected automatic soil moisture and temperature data in half of the 16 experimental plots, we conducted a simple statistical analysis to determine the average effect of the rain-out shelters and irrigation treatments on the observed observed plots and then applied these average effects to all the plots in the experiment.
+We expected that differences in the soil moisture between drought irrigation plots might become more apparent after and during periods of rainfall we first classified the daily climate during the experimental period into dry and rainy days.
+Rainy days were defined as any days when average temperatures were above 3 degrees C and precipitation was over x mm.
+The days after a day classified as rainy were also classified as rainy.
+We then averaged datalogger readings by day and plot within each plot group (one drought, one irrigation and one ambient plot each).
+For each plot group, we then subtracted the drought and irrigation readings from the ambient measurements.
+We then modeled the average deviations in soil moisture and temperature in the experimental treatments as a function of season, raininess and their interaction.
+We used the parameters from this model to describe the average effects of the experimental treatments on soil moisture in all 16 experimental plots in the study
 
 We conducted a simple statistical analysis of differences in cover trends on treatment plots.
-The response variable for this analysis was year-to-year change in cover at the quadrat scale, calculated as $log(Cover_{t+1}/Cover_t)$.
-We tested for the effect of precipitation treatment on this measure of cover change with a linear mixed effects model in the \textit{lme4} package \citep{Bates2015} of \textit{R}, including year and quadrat as random effects.
-We considered the precipitation treatment effects to be significant when the 95\% confidence interval on the estimated coefficient did not overlap zero.
+The response variable for this analysis was change in cover at the quadrat scale measured at the start of the experiment in spring 2011 prior to manipulation, to the last year of the experiment in 2016.
+We calculated this change in cover as $log(Cover_{t+1}/Cover_t)$ per plot.
+We tested for the effect of precipitation treatment on this change in cover with a linear model in \textit{R}.
+We considered the precipitation treatment effects to be significant when??.
 
@@
 We expected that our precipitation manipulation experiment would have effects on plants through the mechanism of altering available soil moisture during the growing season.
-Because we do not have direct measurements of soil moisture for each year of observed plant cover, we used the SOILWAT soil moisture model to estimate daily soil moisture at the USSES from 1925 to the present \citep{Parton 1978}.
+Because we do not have direct measurements of soil moisture for each year of observed plant cover in the historical record, we used the SOILWAT soil moisture model to estimate daily soil moisture at the USSES from 1925 to the present \citep{Parton 1978}.
 We used an enhanced version of soilwat that has recently been developed for use in semi-arid shrubland ecosystems \citep{Bradford}.
@@
 We used daily weather data collected at the USSES from 1925 until the present as weather forcing data for the SOILWAT predictions.
-The USSES climate station (station id: GHCND:USC00102707) is located within 2 km of the research plots.
-We downloaded daily and monthly tmin, tmax, and precipitation data from the National Climate Data Centers online database.
-
-Because the
-
-We averaged daily soil moisture predictions for 5, 25 and 30 cm soil layers, and daily mean temperature, and then aggregated these quarterly.
+
+We averaged daily soil moisture predictions from SOILWAT from the 5, 25 and 30 and 40 cm soil layers, and daily mean temperature, and then aggregated these seasonally.
 

### 2016-12-16 precip/manuscript/precip_ms.tex
adding more manuscript edits

@@
 Although past studies used different statistical models and methods for variable selection they indicate X.
-This well-studied system offers the ideal opportunity to test whether the associations between annual climate variation and plant demography we observe in long-term observational data can be used to predict the future responses of plant populations to climate change.
+This well-studied system offers the ideal opportunity to test whether statistical associations between annual climate variation and plant demography in long-term observational data can be used to predict the future responses of plant populations to climate change.
 
@@
 
-Since we only collected automatic soil moisture and temperature data in half of the 16 experimental plots, we conducted a simple statistical analysis to determine the average effect of the rain-out shelters and irrigation treatments on the observed observed plots and then applied these average effects to all the plots in the experiment.
-We expected that differences in the soil moisture between drought irrigation plots might become more apparent after and during periods of rainfall we first classified the daily climate during the experimental period into dry and rainy days.
-Rainy days were defined as any days when average temperatures were above 3 degrees C and precipitation was over x mm.
-The days after a day classified as rainy were also classified as rainy.
-We then averaged datalogger readings by day and plot within each plot group (one drought, one irrigation and one ambient plot each).
-For each plot group, we then subtracted the drought and irrigation readings from the ambient measurements.
-We then modeled the average deviations in soil moisture and temperature in the experimental treatments as a function of season, raininess and their interaction.
-We used the parameters from this model to describe the average effects of the experimental treatments on soil moisture in all 16 experimental plots in the study
-
-We conducted a simple statistical analysis of differences in cover trends on treatment plots.
-The response variable for this analysis was change in cover at the quadrat scale measured at the start of the experiment in spring 2011 prior to manipulation, to the last year of the experiment in 2016.
-We calculated this change in cover as $log(Cover_{t+1}/Cover_t)$ per plot.
-We tested for the effect of precipitation treatment on this change in cover with a linear model in \textit{R}.
+We conducted a simple statistical to determine the net effect of the experimental treatments on cover in the experiment.
+First we calculated the log change in cover for each of the four focal species in each quadrat from from the start of the experiment in spring prior to manipulation, to the last year of the experiment.
+Log change in cover was defined as , $log(Cover_{2016}/Cover_{2011})$ where $Cover_{2016}$ is the cover of each species in 2011 and $Cover_{2015}$ is cover in 2011.
+We tested for the effect of precipitation treatment on this measure with a linear model in \textit{R}.
 We considered the precipitation treatment effects to be significant when??.
@@
 
-We expected that our precipitation manipulation experiment would have effects on plants through the mechanism of altering available soil moisture during the growing season.
-Because we do not have direct measurements of soil moisture for each year of observed plant cover in the historical record, we used the SOILWAT soil moisture model to estimate daily soil moisture at the USSES from 1925 to the present \citep{Parton 1978}.
+We expected that our precipitation manipulation experiment would affect plants by altering available soil moisture during the growing season.
+Because we do not have direct soil moisture measures for each year of observed plant cover in the historical record, we used the SOILWAT soil moisture model to estimate daily soil moisture at the USSES from 1925 to the present \citep{Parton 1978}.
 We used an enhanced version of soilwat that has recently been developed for use in semi-arid shrubland ecosystems \citep{Bradford}.
@@
 
-We averaged daily soil moisture predictions from SOILWAT from the 5, 25 and 30 and 40 cm soil layers, and daily mean temperature, and then aggregated these seasonally.
+We averaged daily soil moisture predictions from SOILWAT from upper 40 cm of soil and then aggregated these seasonally to serve as the covariates in the vital rate regressions for each species.
+In order to predict species' performance in the experimental plots we took the raw SOILWAT predictions for the ambient conditions for each day during the experimental period and then adjusted these up or down using coefficients from a linear model fit to the changes in soil moisture directly observed in the drought and irrigation plots.
+
+Because we did not monitor soil moisture directly in all X control, drought and irrigation plots, we used a model to describe the average treatment effects on soil moisture during the course of the experiment.
+To do this we first averaged observed soil moisture data from the dataloggers installed in four of the plot groups by date and plot and then within each plot group we found the average deviation from ambient soil moisture by subtracting the average soil moisture in the drought and irrigation plots from the average soil moisture in the nearby control areas.
+This left us with a list of average daily deviations from ambient soil moisture in the drought and irrigation plots.
+
+We expected that the effects the drought and irrigation experiment would have on soil moisture would depend on the season and whether it had rained recently.
+So we classified the daily climate during the experimental period into four seasons and classified each day as dry or rainy.
+Rainy days were defined as any days when average temperatures were above 3 degrees C and precipitation was over x mm.
+The day immediately following rainfall was also classified as rainy.
+
+We modeled the average daily deviations in soil moisture and temperature in the treatments as a function of season, whether it had rained recently and the interaction between these two factors.
+After all this, we could then use this model to adjust the ambient soil moisture values predicted from the SOILWAT model above.
+These adjusted soil moisture values reflected the average season and rainfall dependent effects of the experimental treatments on soil moisture and could be used as covariates for predicting the effects of our manipulation on each species demographic rates.
 
@@
 In this analysis we only include crowding from the four main focal species.
-We incorporated eight climate effects in each of our vital rate regressions: the
-
-Formally, we modeled the survival probability, $S$, of genet $i$ in species $j$, group $g$, and removal treatment $h$, from time $t$ to $t+1$  as
+
+Formally, we modeled the survival probability, $S$, of genet $i$ in species $j$, group $g$, and removal treatme nt $h$, from time $t$ to $t+1$  as
 \begin{equation}

### 2016-12-17 precip/manuscript/precip_ms.tex
adding manuscript changes to the methods mainly

@@
 
-Our regression models for survival, growth and recruitment follow previous work \citep{adler_coexistence_2010,chu_large_2015}.
+For each of the vital rates and each species we fit three separate models.
+First we fit a treatment model fit to all observations in the historical data as well as the contemporary experiment.
+This model included parameters estimating the effects of the drought and irrigation treatments on each species vital rate.
+This model was used to describe the basic results of the experiment.
+Next we fit two models to the historical observational data only (including the first four years of the modern data 2007 to 2010).
+We used these two models to generate predictions for the effects of the experiment.
+In the first prediction model, year to year variation in vital rates was treated as random effect.
+In the second prediction model, the "climate" model, we included paramaters that described the effects of year to year variation in seasonal soil moisture.
+These three models allow for two meaningful comparisons.
+First we can compare the predictions made by the two prediction models directly to the raw data in the experimental plots.
+In addition, we can compare the direction and magnitude of the soil moisture parameters estimated from the observational data by the climate model to the coefficients describing the treatment effects in the treatment model.
+
+All three versions of the models of the models above follow the same basic structure and are developed from previous work \citep{adler_coexistence_2010,chu_large_2015}.
 We model the survival probability of an individual genet as a function of genet size, the neighborhood-scale crowding experienced by the genet from both conspecific and heterospecific genets, temporal variation among years, and permanent spatial variation among groups of quadrats (`group'; here means a set of nearby quadrats located within one pasture or grazing exclosure).
@@
 
-Formally, we modeled the survival probability, $S$, of genet $i$ in species $j$, group $g$, and removal treatme nt $h$, from time $t$ to $t+1$  as
-\begin{equation}
-\mbox{logit}(S_{ijgh,t}) = \gamma_{j,t}^S + \varphi_{jg}^S+  \chi_{jh}^S  + \beta_{j,t}^S u_{ij,t} +  
+Formally, we modeled the survival probability, $S$, of genet $i$ in species $j$, group $g$, and removal treatmnt $h$, from time $t$ to $t+1$  as
+\begin{equation}
+\mbox{logit}(S_{ijgh,t}) = \varphi_{jg}^S + \gamma_{j,t}^S  + \beta_{j,t}^S u_{ij,t} +  
 \left \langle \boldsymbol{\omega}_{j,t}^S, \boldsymbol{W}_{ij,t} \right \rangle 
@@
 \end{equation}
-where $\gamma$ is a time-dependent intercept, $\varphi$ is the coefficient for the effect of spatial location group, $\chi$ is the removal treatment effect, $\beta$ is the coefficient that represents the effect of log genet size, $u$, on survival.
+where $\varphi$ is the spatial group dependent intercept, $\gamma$ is a year-effect, $\beta$ is year-dependent coefficient that represents the effect of log genet size, $u$, on survival in year $t$.
 $\boldsymbol{\omega}$ is a vector of interaction coefficients which determine the impact of crowding, $\boldsymbol{W}$, by each species on the focal species.
@@
 spicata}, \textit{H.
-comata}, and \textit{Poa secunda}, as well as two covariates representing total crowding from (1) all other perennial grasses and shrubs, which were mapped as polygons, and (2) all forb species, which were mapped as points.
+comata}, and \textit{Poa secunda}.
 $\left \langle \boldsymbol{x, y} \right \rangle$ denotes the inner product of vectors $\boldsymbol{x}$ and $\boldsymbol{y}$, 
@@
 
+To describe the treatment effects in the experiment a new term is added to the above model,
+\begin{equation}
+ \boldsymbol{T}\chi_{j}^S
+\label{eqn:survT}
+\end{equation}
+where $\chi$ is a vector of treatment effect coefficients describing the effects of each experimental treatment $h$ on the survival rate, and $\boldsymbol{T}$ is a design matrix indicating the treatment level of each observation in the data.
+The design matrix also includes terms for the interaction between plant size $u$ and the treatment effects.
+These interaction terms allow the effect of each treatment to vary with plant size.
+
+In the climate model, the $\chi$ and $\xi$ terms are replaced with, 
+\begin{equation}
+\boldsymbol{C}\xi_{j}^S 
+\label{eqn:survC}
+\end{equation}
+where $\xi$ gives a vector of coefficients describing the effects of a set of soil moisture covariates $\boldsymbol{C}$ in treatment $h$ and year $t$ on the survival rate of species $j$.
+$\boldsymbol{C}$ is a vector of seasonal average soil moisture at a set of species-specific seasons and time lags and can include interaction effects between plant size, $u$ and the soil moisture covariates.
+These interaction effects allow the effect of soil moisture to vary with plant size.
+
 Our growth model has a similar structure. The change in genet size from time $t$ to $t+1$ , conditional on survival, is given by:
 \begin{equation}
-u_{ijgh,t+1} = \gamma_{j,t}^G + \varphi_{jg}^G+  \chi_{jh}^G  + \beta_{j,t}^G u_{ij,t} + 
+u_{ijgh,t+1} = \varphi_{jg}^G + \gamma_{j,t}^G + \chi_{jh}^G  + \beta_{j,t}^G u_{ij,t} + 
 \left \langle  \boldsymbol{\omega}_{j,t}^G, \boldsymbol{W}_{ij,t} \right \rangle + \varepsilon_{ij,t}^G .
@@
 
+As in the survival regression above, specific specific parameters describing the treatment effects on growth are added in the treatment model, 
+\begin{equation}
+\boldsymbol{T} \chi_{j}^G 
+\label{eqn:growT}
+\end{equation}
+
+where $\chi$ is a treatment effect describing the effect of experimental treatment $h$ on growth, including treatment by size interactions.
+
+Similarly, in the climate model soil moisture influences the growth equation through these terms,  
+\begin{equation}
+\boldsymbol{C}\xi_{j}^G 
+\label{eqn:growC}
+\end{equation}
+where $\xi$ is a vector of coefficients describing the effects of soil moisture covariates in the matrix $\boldsymbol{C}$ for treatment $h$ and year $t$ on growth of species $j$.
+Again this can include interactions between soil moisture and plant size $u$.
+
 We model the crowding experienced by a focal genet as a function of the distance to and size of neighbor genets.
-In previous work, we assumed that the decay of crowding with neighbor distance followed a Gaussian function \citep{chu_large_2015}, but here we use a data-driven approach \citep{teller_linking_2016}.
+In previous work, we assumed that the decay of crowding with neighbor distance followed a Gaussian function \citep{chu_large_2015}, but here we use a data-driven approach \citep{teller_linking_2016, adler_in_prep}.
 We model the crowding experienced by genet $i$ of species $j$ from neighbors of species $m$ as the sum of neighbor areas across a set of concentric annuli, $k$, centered at the plant,
@@
 the within-species competition kernel for species $j$ is also used to determine the effect of all other species on species $j$.
-We used data from all historical plots and contemporary control-treatment plots to estimate the competition kernels, which are shown in Fig.
-\ref{fig:CompKernels}.
+We used data from all historical plots and contemporary control-treatment plots to estimate the competition kernels \citep{adler_in_prep}.
 
@@
 All genets in a quadrat were included in calculating $W$, but plants located within 5 cm of quadrat edges were not used in fitting.
-We fit the models using the \textit{R-INLA} package for R \citep{rue_approximate_2009}.
-To explore variations on the growth regressions, such as the addition of a year-by-treatment interaction or information about the locations of individual grasses relative to removed \textit{Artemisia} shrubs, we fit models using the \textit{lmer} function in the \textit{lme4} package for R
-
-We model recruitment at the quadrat level rather than at the individual genet level because the mapped data do not allow us to determine which recruits were produced by which potential parent genets.
+
+We model recruitment at the quadrat level rather than at the individual genet level because the mapped data do not allow us to determine which recruits were produced by which potential parent plants.
 We assume that the number of individuals, $y$, of species $j$ recruiting at time $t+1$ in the location $q$ follows a negative binomial distribution:
@@
 \begin{equation}
-\lambda_{jq,t+1} = C'_{jq,t} \exp{\left(\gamma_{j,t}^R +  \varphi_{jg}^R + \chi_{jm}^R + 
+\lambda_{jq,t+1} = C'_{jq,t} \exp{\left(\varphi_{jg}^R + \gamma_{j,t}^R + 
 \left \langle \boldsymbol{\omega}^R , \boldsymbol{\sqrt{C'}_{q,t}} \right \rangle \right) }
@@
 \end{equation}
-where the superscript $R$ refers to Recruitment, $C'_{jq,t}$ is the `effective cover' (cm$^2$) of species $j$ in quadrat $q$ at time $t$, $\gamma$ is a time-dependent intercept, $\varphi$ is a coefficient for the effect of spatial location,
-$\chi$ is the removal treatment effect, $\boldsymbol{\omega}$ is a vector of coefficients that determine the strength of intra- and interspecific density-dependence, and $\boldsymbol{C'}$ is the vector of ``effective'' cover of each species in the community.
-Eqn (\ref{eqn:recrProcessModel}) is essentially the Ricker model for discrete time population growth \citep{ricker_stock_1954}, except that we use the square root of cover in the density-dependent term.
-We previously used a Ricker equation \citep{adler_coexistence_2010}, but using the square-root transformation instead gave a lower Deviance Information Criterion (DIC) value \citep{spiegelhalter_bayesian_2002} for the recruitment model.
-Following previous work \citep{adler_coexistence_2010}, we treated year and spatial group variables as random factors, allowing intercepts to vary among years and spatial locations.
-
-Due to the possibility that plants outside the mapped quadrat could contribute recruits to the focal quadrat or interact with plants in the focal quadrat, we estimated effective cover as a mixture of the observed cover, $C$, in the focal quadrat, $q$, and the mean cover, $\bar{C}$, across the spatial location, $g$, in which the quadrat is located: $C'_{jq,t}=p_j C_{jq,t}+(1-p_j) \bar{C}_{jg,t}$, where $p$ is a mixing fraction between 0 and 1 that was estimated as part of fitting the model.
-
-We ran Hamiltonian-Markov Chain Monte Carlo (HMCMC) simulations in STAN 6.4 \citep{stan} to estimate the recruitment model parameters.
+where the superscript $R$ refers to Recruitment, $C'_{jq,t}$ is the `effective cover' (cm$^2$) of species $j$ in quadrat $q$ at time $t$, $\varphi$ is a group dependent intercept, $\gamma$ is a random year effect, $\boldsymbol{\omega}$ is a vector of coefficients that determine the strength of intra- and interspecific density-dependence, and $\boldsymbol{C'}$ is the vector of ``effective'' cover of each species in the community.
+Following previous work \citep{adler_coexistence_2010}, we treated year as a random factor allowing intercepts to vary among years.
+
+Because plants outside the mapped quadrat could contribute recruits to the focal quadrat or interact with plants in the focal quadrat, we estimated effective cover as a mixture of the observed cover, $C$, in the focal quadrat, $q$, and the mean cover, $\bar{C}$, across the spatial location, $g$, in which the quadrat is located: $C'_{jq,t}=p_j C_{jq,t}+(1-p_j) \bar{C}_{jg,t}$, where $p$ is a mixing fraction between 0 and 1 that was estimated as part of fitting the model.
+
+In the treatment model, a new term is added to the exponential term in the equation above, 
+\begin{equation}
+ \boldsymbol{T}\boldsymbol{chi}_{j}^R
+\end{equation}
+where $chi$ describes the effect of the treatment levels on recruitment.
+
+Likewise in the climate model this term is added,
+\begin{equation}
+\boldsymbol{C}\xi_{j}^R 
+\label{eqn:recC}
+\end{equation}
+where the $\boldsymbol{\xi}$ gives a set of coefficients for the soil year, and treatment specific soil moisture covariates in $\boldsymbol{C}$.
+
+We fit all vital rate models using Hamiltonian-Markov Chain Monte Carlo (HMCMC) simulations in STAN 6.4 \citep{stan} and  \citep{rstan}.
 Each model was run for 2,000 iterations and four independent chains with different initial values for parameters.
 We discarded the initial 1,000 samples.
-Convergence was observed graphically for all parameters, and confirmed by the Brooks-Gelman statistic \citep{brooks_general_1998}.
-
-For all three vital rate models, we considered the effect of climate covariates to be significant when 95\% posterior
-intervals did not overlap zero.
-
-All data and R code necessary to reproduce our analysis will be deposited in the Dryad Digital Repository once the manuscript is accepted.
-The current version of the computer code is available at https://github.com/pbadler/ExperimentTests/tree/master/removals and the data are available at https://bitbucket.org/ellner/driversdata.
+Convergence was observed graphically for all parameters, and confirmed by assessing the split $\widehat{R}$ statistic which at convergence is equal to one \citep{stan manual}.
+
+To assess the effects of the experimental treatments on species survival and growth we fit treatment models with and without the size by treatment interactions in the treatment effect term for (\ref{eqn:survT}) and growth models (\ref{eqn:growT}).
+We judged whether including interaction terms improved model fit by comparing the Watanabe-Aikake Information Criteria (WAIC) scores for each version of the model and retained the version with the lower WAIC score \citep{vetari_practical_2015}.
+WAIC scores approximate cross-validation predictive accuracy for a model and like traditional AIC allow for model comparison.
+Lower WAIC scores indicate a more parsimonious model.
+When a treatment model for survival or growth for a species included a size by treatment effect, we also included a size by soil moisture effect in the climate model for that species and vital rate.
+This allowed us to more directly compare the effects in the experimental data to the effects predicted from the climate model fit to the observational data.
+
+\subsection*{Selecting soil moisture covariates}
+
+After generating a time series of predicted daily soil moisture from the SOILWAT model, we averaged daily soil moisture across spring, summer and fall seasons in each year.
+We considered each of the three seasonal soil moisture variables at three different time periods relative to the demographic transition from year $t$ to year $t+1$.
+Soil moisture in the year between $t$ and $t+1$ is indicated with a "1" subscript.
+Soil moisture in the year before $t$ is indicated with a "0" subscript.
+And soil moisture preceding this year is indicated with a "lag" subscript.
+For example, for the year 2010, $spring_1$ would indicate soil moisture in the spring of 2010, $spring_0$ would indicate soil moisture during the spring of 2009 and $spring_{lag}$ would indicate soil moisture during the spring of 2008.
+
+In order to select among the nine potential soil moisture covariates (three seasons and three lags each) for each species and vital rate, we first fit a model with random year effects but without climate covariates to the observational data up to 2010.
+We then extracted the mean of the year effects estimates for each fitted year in the data.
+These random effects represented unexplained deviations in the average vital rate in a given year.
+We then found the correlations between each of the soil moisture variable and the random year effects.
+For each vital rate and species, we selected the three covariates with highest correlations with these year effects.
+This screening technique has been used in previous demographic studies at this site \citep{dalgleish_climate_2010} and is often used in dendrochronology to screen for potential climate influence on tree-ring growth \citep{wang trees}; however, it is subject to the criticism that it is a form of data dredging \citep{dredging}.
+Nevertheless, we felt that this approach was justified in this study because ultimately we did not make inference from these fitted parameters until after we tested their ability to predict the data in the experimental plots.
+
+Once we selected soil moisture covariates for each of the species and vital rates we fit the climate models including the climate terms for survival \ref{eqn:survC}, growth \ref{eqn:growC} and recruitment \ref{eqn:recC}.
+We fit these models only to the observational data from the historical period and (1928 to 1957) and the first four years of the modern period (2007 to 2010).
+Thus in generating predictions for the treatment effects observed in from 2011 to 2016 we are predicting truly out of sample data.
 
@@
 
-The vital rate regressions provide a qualitative test of removal treatment effects: Does removal have an effect on a vital rate that goes above and beyond the effects of species interactions that are explicit in the model?
-However, the population-level response by each species integrates the effects of the removal treatment on all three vital rates.
-We used an individual-based model (IBM) to compare observed and predicted changes
-in population size (cover) from one year to the next.
+The vital rate regressions allow us to evaluate whether the experimental treatments and the soil moisture variables considered had an effect on species performance.
+But the population response of each species ultimately depends on the integrated effects of treatment or soil moisture on all three vital rates.
+To evaluate whether the climate models could predict the overall population response of these species in the drought and irrigation experiment we used an individual-based model (IBM) to compare observed and predicted changes in population size (cover) from one year to the next in the experimental plots.
 
@@
 
-One reason we might observe small effects of the removal treatment is the limited power of our experiment.
-Perhaps large uncertainty in the estimates of these effects makes it difficult to reject the null hypothesis represented by our baseline model.
-To address this possibility, we also simulated a version of the model in which we replaced the mean estimate of each removal treatment effect with the parameter value at either the 5\% or 95\% confidence interval, whichever was farthest from zero.
-This version of the model shows the population-level changes we would expect given maximum removal effects.
-
-\subsection*{Integral projection models}
-
-To quantify the long-term, population-level consequences of the removal effects, we built and then simulated an integral projection model (IPM).
-In our IPM, the population of species $j$ is represented by a density function $n(u_{j,t})$ which gives the density of genets of size $u$ at time $t$, with genet size on a natural-log scale, i.e.
-$n(u_{j,t})du$ is the number of genets whose log area is between $u$ and $u+du$, hence their area
-(on arithmetic scale) is between $exp(u_j)$ and $exp(u_j+du)$. The size distribution function at time $t+1$ is given by
-\begin{equation}
-n(v_{j,t+1})=\int_{L_j}^{U_j} k_j (v_j,u_j,{\vec{n}})n(u_{j,t})   
-\label{eqn:IPM}
-\end{equation}
-where the kernel $k_j (v_j,u_j,\vec{n}_j)$ describes all possible transitions from size $u$ to $v$ and $\vec{n}$ 
-is the set of size-distribution functions for all species in the community. The integral is evaluated over all possible sizes from a 
-lower size limit $L$ to an upper size limit $U$ that extends past the range of observed sizes.
-
-The kernel is constructed from the fitted survival ($S$), growth ($G$), and recruitment ($R$) models, 
-\begin{equation}
-k_j(v_j,u_j,\vec{n})=S_j(u_j,\vec{n})G_j(v_j,u_j,\vec{n})+R_j(v_j,u_j,\vec{n}) du_j
-\label{eqn:IPM} 
-\end{equation}
-$S$ is given by (Eqn. \ref{eqn:survReg}) and $G$ is given by (Eqn. \ref{eqn:growReg}), using an expected neighborhood crowding
-index calculated from the size distributions of the competing species. In the vital rate regressions and in the IBM, each 
-individual has a unique neighborhood crowding index based on the spatial locations and sizes of neighboring plants. This 
-measure cannot be extended to the IPM, which tracks individual size but not location. Instead we used a 
-mean-field approximation that captures the essential features of spatially-explicit neighborhood interactions \citep{adler_coexistence_2010}. 
-We found that in both the observed data and IBM simulations, heterospecific individuals were approximately randomly distributed with respect to each other, 
-but conspecific individuals displayed a non-random, size-dependent patterns: small genets were randomly distributed, while large genets 
-were segregated from each other. The overdispersion of large conspecific genets was introduced into the IPM 
-with a 'no-overlap' rule \citep{adler_coexistence_2010}.
-In \ref{suppMethods} we explain how the spline competition kernels with the no-overlap rule kernels were used to compute neigbhorhood crowding index as a function of individual size for the IPM.
-
-For recruitment in the IPM, the factor $ \Phi_j  = \exp{\left(\gamma_{j,t}^R +  \varphi_{jg}^R + \chi_{jm}^R + 
-\left \langle \boldsymbol{\omega}^R , \boldsymbol{\sqrt{C'_{q,t}}} \right \rangle \right) }$
-in eqn. (\ref{eqn:recrDataModel}) gives the expected 
-area of new recruits per quadrat, per unit area of parents.
-To incorporate this recruitment function into the IPM, we assumed that individual fecundity increases linearly with individual area (equal to $e^u$ for a size-$u$ individual).
-The recruitment kernel is thus
-$R_j(v_j,u_j,\vec{n})=c_{0,j}(v_j)e^{u_j}\Phi_j$ \citep{adler_coexistence_2010} where $c_{0,j}$ is the
-size distribution of new recruits. $\Phi_j$ is calculated from eqn. (\ref{eqn:recrDataModel}) with total cover values computed from the population densities as $C'_{j,t} = \int e^u n_j(u,t) du$.
-
-We simulated the IPM to estimate equilibrium abundances under three different scenarios.
-For all IPM simulations, we set the spatial random effects to zero, meaning that we are simulating dynamics on a hypothetical, average site.
-The first scenario featured all four species and our ``baseline'' model for control plots (no removal effects).
-We initialized all species with arbitrarily low cover and ran the model for 500 time steps (years) to allow the community to reach an equilibrium.
-To incorporate temporal environmental variation, at each time step we randomly selected one of the 30 sets of random year effects for the survival, growth and recruitment models (``kernel selection'' in the terminology of \citealt{metcalf_statistical_2015}).
-We calculated the equilibrium cover for each species by simulating an additional 2000 time steps and averaging each species' cover over this period.
-For the second scenario, we set the cover of \textit{A.
-tripartita} to zero and then repeated the procedure.
-This simulation estimates the competitive release of grasses following \textit{A.
-tripartita} removal according to the baseline model (no removal effects).
-For the third scenario, we again set  \textit{A.
-triparitia} to zero, but this time we projected the model using the vital rate regressions that include removal effects.
-If our baseline model successfully predicts competitive release observed in the removal experiment, then the results of the second and third scenarios should be similar.
-However, to the extent that the baseline model underestimates (or overestimates) the actual competitive release, the results of the second and third scenarios will differ.
-As with the IBM, we repeated these simulations using maximum removal treatment effects represented by parameters values at the 5\% or 95\% confidence limit farthest from zero.
-We did not simulate long-term \textit{A.
-tripartita} response to grass removal because the vital rate regressions for \textit{A.
-tripartita} showed that the baseline model effectively predicted removal effects.
+\subsection*{Assessing soil moisture effects and predictive accuracy}
+
+After fitting the random year effects (null) and climate models to the observational data, we generated predictions from these models for each of the vital rates in the experimental plots.
+We then assessed performance of the climate and null models by calculating the mean square error between the predictions and observations for each of the vital rates.
+In addition we evaluated whether the climate model predicted treatment effects of similar direction and magnitude to those observed in the experiment.
+We did this by extracting the soil moisture coefficients contained in $\xi$ for each of the vital rates and then multiplying those by the appropriate soil moisture covariates for each year and treatment level in the experiment.
+The average of this product across all the years of the experiment then gives the predicted treatment coefficient for treatment level $h$.
+These predicted treatment effects can then be compared directly to the treatment effects fitted in the treatment model in the term $\chi$.
+
+For treatment and vital rate models we considered the effect of climate covariates or treatment effects to be significant when the 95\% Bayesian credible intervals on the posterior estimates did not overlap zero.
+
+We evaluated cover predictions from the climate and year effects model similarly.  We first fit the
+
+All data and R code necessary to reproduce our analysis will be deposited in the Dryad Digital Repository once the manuscript is accepted.
+The current version of the computer code is available at https://github.com/pbadler/ExperimentTests/tree/master/precip and the data are available at https://bitbucket.org/ellner/driversdata.
 
@@
 However, annual mean temperatures were consistently warm relative to the long-term mean, and two years were warmer than the 95\% quantile (\ref{fig:climate}b).
+
+The experimental drought shelter successfully reduced the soil moisture in the study plots in the spring and summer, while irrigation successfully increased soil moisture \ref{fig:spotm}.
+Our continuous data
 
@@
 
-Our population models based on historical, observational data reveal weak interspecific interactions \citep{adler_coexistence_2010,chu_large_2015} and, as a result, predict very little competitive release among the four dominant species in a sagebrush steppe.
+Our experiment showed that including climate covariates in our population models did little to improve population level predictions across the course of the experiment.
+In general the climate models did not perform as well as null models in predicting response to the experimental drought and irrigation treatments at the vital rate or the population level.
+However, there were some glimmers of hope.
+First at the population level we generally found that including climate covariates did not improve predictions in the control and irrigation plots, but they did appear to improve predictions in the drought plots ().
+This suggests that models fit to observational data do contain useful information about the effects of climate on individual performance.
+The fact that these predictions showed up in the more extreme conditions produced by the drought plots could be a sign that in ambient conditions the effect of climate is weak enough that it is washed out at the population level by finer scale sources of variation.
+This pattern was observed for species x, x and x.
+
+ the climate signal  improve population growth predictions in the drought treatment but not experiment showed some relatively strong reson based on historical, observational data reveal weak interspecific interactions \citep{adler_coexistence_2010,chu_large_2015} and, as a result, predict very little competitive release among the four dominant species in a sagebrush steppe.
 Despite the fact that previous removal studies have reported variable results, ranging from negative effects of removal on remaining species to complete competitive release, we were suspicious of our model's predictions and conducted a removal experiment to test them.
@@
 
-Funding was provided by NSF grants DEB-1353078,and  DEB-1054040 to PBA and DEB 1353039 to SPE, and by the Utah Agricultural Experiment Station (get journal paper number).
+Funding was provided by an NSF GRFP to AK, NSF grants DEB-1353078, and DEB-1054040 to PBA and by the Utah Agricultural Experiment Station (get journal paper number).
 The USDA-ARS Sheep Experiment Station generously provided access to historical data and the field experiment site.
-We thank Robin Snyder for suggestions that improved the manuscript.
+Joe and Kevin, Lau, John Bradford, Caitlin (soilwat),
 

### 2016-12-18 precip/manuscript/precip_ms.tex
a few revisions to methods

@@
 where $\xi$ gives a vector of coefficients describing the effects of a set of soil moisture covariates $\boldsymbol{C}$ in treatment $h$ and year $t$ on the survival rate of species $j$.
-$\boldsymbol{C}$ is a vector of seasonal average soil moisture at a set of species-specific seasons and time lags and can include interaction effects between plant size, $u$ and the soil moisture covariates.
-These interaction effects allow the effect of soil moisture to vary with plant size.
+$\boldsymbol{C}$ is a vector of seasonal average soil moisture and can include interaction effects between plant size, $u$, and the soil moisture covariates that allow the effects of soil moisture to vary with plant size.
 
@@
 \end{equation} 
-Note that $W_{ijj}$ gives intraspecific crowding. The $W$'s are then the components of the $\boldsymbol{W}$ vectors 
-introduced as covariates in the survival (\ref{eqn:survReg}) and growth (\ref{eqn:growReg}) regressions.
-
-We assume that competition kernels $F_{jm}(d)$ are non-negative and decreasing, so that distant plants have less effect 
-than close plants. Otherwise, we let the data dictate the shape of the kernel by fitting a spline model 
+Note that $W_{ijj}$ gives intraspecific crowding.
+The $W$'s are then the components of the $\boldsymbol{W}$ vectors introduced as covariates in the survival (\ref{eqn:survReg}) and growth (\ref{eqn:growReg}) regressions.
+
+We assume that competition kernels $F_{jm}(d)$ are non-negative and decreasing, so that distant plants have less effect than close plants.
+Otherwise, we let the data dictate the shape of the kernel by fitting a spline model
 using the methods of Teller et al. (2016). The shape of $F_{jm}$ is determined by a set of spline basis coefficients $\vec{b}_{jm}$
@@
 
-After fitting the random year effects (null) and climate models to the observational data, we generated predictions from these models for each of the vital rates in the experimental plots.
+After fitting the random year effects (null model) and climate models to the observational data, we generated predictions from these models for each of the vital rates in the experimental plots.
 We then assessed performance of the climate and null models by calculating the mean square error between the predictions and observations for each of the vital rates.
@@
 We did this by extracting the soil moisture coefficients contained in $\xi$ for each of the vital rates and then multiplying those by the appropriate soil moisture covariates for each year and treatment level in the experiment.
-The average of this product across all the years of the experiment then gives the predicted treatment coefficient for treatment level $h$.
-These predicted treatment effects can then be compared directly to the treatment effects fitted in the treatment model in the term $\chi$.
+We then averaged these products across the years in the experiment to get the effects of the experiment predicted by the soil moisture model.
+These predicted treatment effects can then be compared directly to the treatment parameters in the term $\chi$.
 
 For treatment and vital rate models we considered the effect of climate covariates or treatment effects to be significant when the 95\% Bayesian credible intervals on the posterior estimates did not overlap zero.
-
-We evaluated cover predictions from the climate and year effects model similarly.  We first fit the
 
@@
 The experimental drought shelter successfully reduced the soil moisture in the study plots in the spring and summer, while irrigation successfully increased soil moisture \ref{fig:spotm}.
-Our continuous data
+Our seasonal model of soil moisture in the experiment indicated that
 

### 2016-12-18 precip/manuscript/precip_ms.tex
adding results text, need to add figures and tables next

--- 2016-12-17 precip/manuscript/precip_ms.tex
@@
 
+Although the main focus of the current analysis the effects of soil moisture, we also modeled the effects of inter- and intra-specific competition in our vital rate models.
 We model the crowding experienced by a focal genet as a function of the distance to and size of neighbor genets.
@@
 We fit all vital rate models using Hamiltonian-Markov Chain Monte Carlo (HMCMC) simulations in STAN 6.4 \citep{stan} and  \citep{rstan}.
+The prior distributions and model code used are described more completely in appendix A.
 Each model was run for 2,000 iterations and four independent chains with different initial values for parameters.
@@
 
-\subsection*{Individual-based models}
-
-The vital rate regressions allow us to evaluate whether the experimental treatments and the soil moisture variables considered had an effect on species performance.
-But the population response of each species ultimately depends on the integrated effects of treatment or soil moisture on all three vital rates.
-To evaluate whether the climate models could predict the overall population response of these species in the drought and irrigation experiment we used an individual-based model (IBM) to compare observed and predicted changes in population size (cover) from one year to the next in the experimental plots.
-
-To simulate changes in cover in each quadrat from year $t$ to year $t+1$, we initialized the IBM with the actual configuration of the four focal species (genet sizes and locations) observed in year $t$.
-We then projected the model forward one year, applying the random year effects corresponding to year $t$ and the spatial random effects corresponding to the quadrat of interest.
-Because we were interested in comparing model predictions to observations, and were not interested in the effects of demographic stochasticity, we used a deterministic version of the model (e.g., recruitment is the $\lambda$ of (\ref{eqn:recrProcessModel}), rather than a random draw from a Poisson distribution with a mean of $\lambda$).
-We first applied the recruitment regression to determine the expected number of new recruits at time $t+1$, which we then multiplied by the mean size of recruits observed for the focal species to get expected cover of new recruits.
-We then used the survival regression to predict the survival probability of each genet, and the growth regression to predict the change in size of each genet.
-Expected cover at $t+1$ is expected area of new recruits, plus the sum of the survival probabilities multiplied by the expected sizes.
-
-For control plots, we used a version of the model with all removal treatment effects set to zero.
-For removal treatment plots, we generated predictions from two versions of the model, with and without treatment effects set to zero.
-We calculated the mean across quadrats of observed log cover change, $log(Cover_{t+1}/Cover_t)$ and compared it to the mean across quadrats of the predicted cover change.
-
-\subsection*{Assessing soil moisture effects and predictive accuracy}
-
-After fitting the random year effects (null) and climate models to the observational data, we generated predictions from these models for each of the vital rates in the experimental plots.
-We then assessed performance of the climate and null models by calculating the mean square error between the predictions and observations for each of the vital rates.
-In addition we evaluated whether the climate model predicted treatment effects of similar direction and magnitude to those observed in the experiment.
+\subsection*{Predicting cover from individual-based models}
+
+The vital rate regressions allow us to evaluate whether soil moisture and the experimental treatments had an effect on species performance.
+But the population response ultimately depends on the integrated effects of treatment or soil moisture on all three vital rates.
+To evaluate whether the climate models could predict the overall population response of these species in the drought and irrigation experiment we used an individual-based model (IBM) to compare observed and predicted changes in population size, defined as plot-level percent cover, from one year to the next in the experimental plots.
+
+To simulate changes in cover in each quadrat from year $t$ to year $t+1$, we initialized the IBM with the observed genet sizes and locations of the four focal species observed in year $t$ in each quadrat.
+For every individual genet in a quadrat, we projected its size and survival probability in the next year using the growth and survival models and the appropriate crowding and soil moisture or treatment covariates for that year and quadrat.
+Likewise we projected the number of new recruits in the quadrat in the next year using the recruitment model.
+We calculated the expected cover in year $t+1$ as the sum area of new recruits, plus the sum of the expected area of existing plants multiplied by each plant's expected survival probability.
+
+We accounted for the uncertainty in our random year effects in our projections, by drawing random year effects for each predicted year from a normal distribution with a mean of zero and a standard deviation drawn from the posterior estimate of the standard deviation of the random year effects \ref{eqn:}.
+We generated predictions using the full posterior distributions of each model parameter which allowed us to carry forward all of the uncertainty of the fitted vital rate models into our cover predictions.
+Because we were interested in comparing model predictions to observations, and were not interested in the effects of demographic stochasticity, we used a deterministic version of the models (e.g., recruitment is the $\lambda$ of (\ref{eqn:recrProcessModel}), rather than a random draw from a negative binomial distribution with a mean of $\lambda$).
+
+We generated predictions for the full time series of observations, including the five experimental years, from the year effects model, the climate model and the treatment model.
+After generating predictions for each year, we found the mean cover across all quadrats in each treatment level and then calculated the predicted log cover change as $log(Cover_{t+1}/Cover_{t})$.
+
+\subsection*{Quantifying predictive accuracy}
+
+After fitting the year effects and climate models to the observational data, we generated predictions from these models for each of the vital rates in the experimental plots.
+We then assessed performance of the climate and null models by calculating the mean square error (MSE) between the predicted and observed responses in the experimental data as,
+
+\begin{equation}
+MSE = \frac{1}{n} \sum_{i=1}^{n} (y_i - E(y_i|\theta))^2, 
+\label{eqn:MSE}
+\end{equation}
+where $y_i$ is the outcome of observation $i$ and $E(y_i|\theta)$ gives the expected outcome given the parameters in the model $\theta$.
+The MSE is easy to interpret, but is not always appropriate for models fit with non-normal error structures \citep{gelman_understanding_2014}.
+A more general statistic for assessing model predictions is the log pointwise predictive density (lppd) \citep{gelman_understanding_2014}.
+The lppd for a given model is defined as,
+
+\begin{equation}
+lppd = \sum_{i=1}^{n} log \int p(y_i| \theta)p_{post}(\theta) d\theta, 
+\label{eqn:lppd}
+\end{equation}
+where the integral on the right side gives the probability of observing the outcome $y$ at each data point $i$ given the full posterior distribution of the parameters in the model $p_{post}(\theta)$.
+In practice we computed the lppd from the posterior simulations generated by STAN as,
+
+\begin{equation}
+\widehat{lppd} = \sum_{i=1}^{n} log \left(\frac{1}{S} \sum_{s=1}^{S} p(y_i | \theta^S) \right),
+\label{eqn:clppd}
+\end{equation}
+where the summation of $p(y_i|\theta^S)$ gives the total probability of observing the the actual response $y_i$ given the simulated posterior distribution $\theta^S$ across the full set of model simulations $S$.
+The log of this sum is then averaged across the set of all observations $i$.
+Higher lppd scores indicate that the model better predicts the observations.
+
+In addition, we evaluated whether the climate model predicted treatment effects of similar direction and magnitude to those observed in the experiment.
 We did this by extracting the soil moisture coefficients contained in $\xi$ for each of the vital rates and then multiplying those by the appropriate soil moisture covariates for each year and treatment level in the experiment.
-The average of this product across all the years of the experiment then gives the predicted treatment coefficient for treatment level $h$.
-These predicted treatment effects can then be compared directly to the treatment effects fitted in the treatment model in the term $\chi$.
-
-For treatment and vital rate models we considered the effect of climate covariates or treatment effects to be significant when the 95\% Bayesian credible intervals on the posterior estimates did not overlap zero.
-
-We evaluated cover predictions from the climate and year effects model similarly.  We first fit the
+We then averaged these across all five years in the experiment to find the average treatment effect predicted by the climate model.
+We compared these to the posteriors of the treatment parameters, $chi$, from the treatment model.
+As a measure of agreement between our predictions and observed response we calculated the correlation between the predicted and observed parameter values.
+
+We considered the effect of climate covariates or treatment effects to be significant when the 95\% Bayesian credible intervals on the posterior estimates did not overlap zero.
 
@@
 
-During the period of our experiment, annual water-year (October-September) precipitation fell within the 5\% - 95\% quantiles of the long-term mean, although more years fell below the mean than above it (\ref{fig:climate}a).
-However, annual mean temperatures were consistently warm relative to the long-term mean, and two years were warmer than the 95\% quantile (\ref{fig:climate}b).
-
-The experimental drought shelter successfully reduced the soil moisture in the study plots in the spring and summer, while irrigation successfully increased soil moisture \ref{fig:spotm}.
-Our continuous data
-
-\subsection*{Cover trends}
-
-Trends in cover by species and removal treatment were variable. For example, grass removal had little effect on the cover of \textit{A. tripartita} 
+\subsection*{Effects on soil moisture}
+
+We successfully changed the average soil moisture measured in the plots we covered with drought shelters and irrigated, in the directions we expected (fig); our spot measurements conducted during the springs of X, X and X showed that drought plots were X percent drier, while irrigation plots were X percent wetter than ambient plots.
+This difference was highly significant () but there was a lot of spatial variation in the soil moisture measurements within treatments (\ref{fig:}).
+The automatically recorded soil moisture data from only half of the experimental plots also showed that the drought treatment decreased and the irrigation treatment increased soil moisture, but the size of these effects were dependent on season and whether it had rained recently ().
+The best model for predicting the average treatment effects on soil moisture was : .
+We did not find evidence that the drought shelters and the irrigation treatments significantly changed air temperature within the plots ().
+
+During the course of our experiment annual precipitation recorded at the local climate station varied from X to X percent of period of record averages (1925 to 2016).
+The year of X was exceptionally dry while the year of X was.
+The precipitation during the summer of X was the wettest on record.
+Mean annual average air temperature during this period were consistently well above average for the period of record, with the years of X X and X being among the warmest ever recorded at this site.
+
+The SOILWAT soil moisture model indicated that soil moisture fluctated from between X and X on average each year, with the months of X and X being the wettest while soil moisture during the months of X and X was driest (fig).
+Predicted seasonal soil moisture for each year was positively correlated with precipitation, and negatively correlated with average temperatures.
+The average soil moisture predicted from soilwat for the entire study period from X to X is shown in figure X.
+Soil moisture predictions for the ambient conditions from 2011 to 2016 fall on the low side of the period of record averages.
+This was driven by higher than average temperatures and severely dry winters and spring conditions in the years of X.
+
+When we adjust the seasonally averaged soil moisture predictions by the treatment effects, we find that the soil moisture conditions predicted in the drought plots during the course of the experiment are lower than those predicted for any of the previous years in the historical record.
+Likewise the predicted soil moisture in the irrigation plots was far above the average for the historical record in the years of X and X.
+We interpret this as evidence that our drought and irrigation plots created truly novel soil moisture conditions in the years X and X that were unlikely to have ever been experienced by the study species during the entire period of the observational data.
+We visualize this by conducting a principle component analysis (PCA) on all the seasonal soil moisture, and plotting each year on the first two principle components.
+The PCA shows that the ambient, drought and irrigation plots experienced warmer than average seasonal conditions during the course of the experiment (further left on PCA1).
+The drought plots were considerably drier, and the irrigated plots considerably wetter (figure PCA).
+
+\subsection*{Effects on cover and vital rates}
+
+The cover of \textit{H.
+comata} and \textit{P.
+spicata} fell significantly by X and X percent respectively from the pre-treatment cover in 2011 to 2016.
+The cover of \textit{P.
+secunda} showed a trend towards decrease in the drought plots and an increase in the irrigated plots but these changes were not significant.
+In contrast the cover of \textit{A.
+tripartita} changed in the opposite direction from the grasses; it increased in the drought plots and decreased in the irrigated plots but these changes were not significant (fig).
+
+Our treatment models fit to the experimental and observational data indicated a variety of treatment effects on the vital rates of each species (table x).
+Among the survival models, WAIC scores indicated that we should retain size x treatment effects for \textit{P.
+secunda}.
+Among the growth models, we retained size by treatment effects for \textit{A.
+tripartita} and \textit{P.
+secunda}.
+DISCUSS SIGNIFICANT treatment effects: For \textit{A.
+tripartita} we observed a significant decrease in growth in the x treatment, while we observed no change in the X treatment.
+\textit{H.
+comata} growth, survival and recruitment, while growth.
+\textit{P.
+secunda} growth and survival recruitment.
+\textit{P.
+spicata} growth survival and recruitment.
+Treatment by size effects on \textit{A.
+tripartita} and \textit{P.
+secunda} showed that ()
+
+We observed a number of strong correlations between the year effects in the vital rate models and the seasonal soil moisture variables and table x reports the strongest three correlations for each species and vital rate (table).
+In general the soil moisture during the spring and summer period appeared to be more strongly correlated with the year effects for each species than the fall soil moisture, but see \textit{P.
+spicata} vr.
+All three time lags show up in the choosen variables, but 1 out of X variables choosen involved soil moisture in the year immediately before the first year of the transition, (subscript "0"), while variables involving soil moisture in the year before that (subscript "lag") and after that (subscript "1") were choosen less often.
+
+We observed vital rate and species-specific effects of soil moisture on each of these species after fitting the vital rate models with the soil moisture covariates discussed above (fig).
+DISSCUSS significant soil moisture effects: for \textit{A.
+tripartita} growth was significantly affected by ...
+For \textit{H.
+comata}.
+For \textit{P.
+secunda}.
+And for \textit{P.
+spicata}.
+Soil moisture by size effects on \textit{A.
+tripartita} and \textit{P.
+secunda} showed that ()
+
+\subsection*{Evaluating the predictions}
+
+For each vital rate and species, predictions from the year effects and climate models were nearly indistinguishable (fig).
+This indicated that the additional soil moisture covariates in the climate models had very little effect on the predictions.
+Not surprisingly \citep{chu_direct_2016,treddenick_} our models show that plant size and intra-specific crowding are far more influential in determining growth, survival and recruitment than climate effects.
+In all cases, we found MSE and lppd scores for the climate models were slightly worse than the MSE and the lppd of the year effects model (table).
+This indicated that adding the soil moisture parameters did not improve model predictions over all (table).
+When we break the data up into treatment groups we do see that the climate model outperformed the year effects model for predicting X species VR in treatment T.
+
+While our climate models failed to improve predictions, we did find that they predicted treatment responses that were largely in the same direction as the observed treatment responses (fig).
+The overall correlation between the predicted treatment parameters and the observed treatment parameters was X with stronger correlation for the climate effects on the size by treatment effects fit for \textit{P.
+secunda} growth and survival and \textit{A.
+tripartita} growth (fig).
+A table of MSE between predicted and expected treatment parameters shows that averaged over all species and vital rates, we did the best job of predicting the response to the drought treatment.
+Averaged across all treatements and vital rates, we did the best job predicting the response of \textit{species} to the treatments.
+Finally if we average across all species and treatments, we did the best job of predicting VR parameters in response to the treatment.
+
+As for the vital rate predictions, our one-step ahead cover predictions from the year effects models and the climate models were fairly similar (fig).
+Models with and without soil moisture parameters tended to underpredict the cover of \textit{A.
+tripartita} and \textit{P.secunda} during the experiment (fig).
+When we compare the MSE of the predicted population growth rates for each year and quadrat from the model with and without the soil moisture parameters, we find that adding soil moisture to the vital rate models always made the one step ahead cover predictions slightly less accurate (table).
+The pattern when we consider the correlation between the predicted and observed population growth rates was slightly different (table?).
+When we broke the population growth rate data up by treatment, we found that adding soil moisture to the models mostly did not improve predictions of cover change in the control and irrigation plots.
+However, including the soil moisture parameters led to better predictions of changes in population growth rates for three of the species (S , S, and S) in the drought plots (fig).
+
+As expected from previous research most of our demographic models estimated strong negative intra-specific crowding effects and weaker negative inter-specific crowding effects on the focal species (appendix) \citep{adler_coexistence_2010,chu_direct_2016,chu_large_2015}.
+
 (Fig. \ref{fig:CoverTrends}), but shrub removal did appear to cause an increase in the cover of \textit{P. spicata}. 
@@
 Our experiment showed that including climate covariates in our population models did little to improve population level predictions across the course of the experiment.
-In general the climate models did not perform as well as null models in predicting response to the experimental drought and irrigation treatments at the vital rate or the population level.
+Averaged across the control, drought and irrigation treatments, the climate models did not perform as well as null models in predicting response the experiment at the vital rate or at the level of population growth rates.
 However, there were some glimmers of hope.

### 2016-12-18 precip/manuscript/precip_ms.tex

@@
 \subsection*{Quantifying predictive accuracy} 
-
 After fitting the year effects and climate models to the observational data, we generated predictions from these models for each of the vital rates in the experimental plots.
@@
 We did not find evidence that the drought shelters and the irrigation treatments significantly changed air temperature within the plots ().
-
-During the course of our experiment annual precipitation recorded at the local climate station varied from X to X percent of period of record averages (1925 to 2016).
-The year of X was exceptionally dry while the year of X was.
-The precipitation during the summer of X was the wettest on record.
-Mean annual average air temperature during this period were consistently well above average for the period of record, with the years of X X and X being among the warmest ever recorded at this site.
 
@@
 secunda} growth and survival and \textit{A.
-tripartita} growth (fig).
+tripartita} growth (fig) than for the treatment effects on the intercept.
 A table of MSE between predicted and expected treatment parameters shows that averaged over all species and vital rates, we did the best job of predicting the response to the drought treatment.
@@
 The pattern when we consider the correlation between the predicted and observed population growth rates was slightly different (table?).
-When we broke the population growth rate data up by treatment, we found that adding soil moisture to the models mostly did not improve predictions of cover change in the control and irrigation plots.
-However, including the soil moisture parameters led to better predictions of changes in population growth rates for three of the species (S , S, and S) in the drought plots (fig).
+When we considered the population growth rate data separately by treatment, we found that adding soil moisture to the models mostly did not improve predictions of cover change in the control and irrigation plots.
+However, we did find that adding soil moisture parameters produced better predictions of population growth rates for three of the species (S , S, and S) in the drought plots (fig).
 

### 2016-12-19 precip/manuscript/precip_ms.tex
adding more manuscript revisions.  Struggled to get the soil moisture treatment effects script to work.  Need more work on that

@@
 \fi
+
+\graphicspath{{../figures/}}
 
@@
 \end{equation}
-
 where $\chi$ is a treatment effect describing the effect of experimental treatment $h$ on growth, including treatment by size interactions.
@@
 We fit all vital rate models using Hamiltonian-Markov Chain Monte Carlo (HMCMC) simulations in STAN 6.4 \citep{stan} and  \citep{rstan}.
-The prior distributions and model code used are described more completely in appendix A.
+The priors and model code are described more completely in appendix A.
 Each model was run for 2,000 iterations and four independent chains with different initial values for parameters.
@@
 But the population response ultimately depends on the integrated effects of treatment or soil moisture on all three vital rates.
-To evaluate whether the climate models could predict the overall population response of these species in the drought and irrigation experiment we used an individual-based model (IBM) to compare observed and predicted changes in population size, defined as plot-level percent cover, from one year to the next in the experimental plots.
+To evaluate whether the climate models could predict the responses of these species in the drought and irrigation experiment at the overall population level we used an individual-based model (IBM) to compare observed and predicted changes in population size from one year to the next.
 
@@
 Likewise we projected the number of new recruits in the quadrat in the next year using the recruitment model.
-We calculated the expected cover in year $t+1$ as the sum area of new recruits, plus the sum of the expected area of existing plants multiplied by each plant's expected survival probability.
-
-We accounted for the uncertainty in our random year effects in our projections, by drawing random year effects for each predicted year from a normal distribution with a mean of zero and a standard deviation drawn from the posterior estimate of the standard deviation of the random year effects \ref{eqn:}.
+We calculated the expected cover in year $t+1$ as the total area of new recruits, plus the sum of the predicted area of each existing plant at time $t+1$ multiplied by each plant's expected survival probability from time $t$ to $t+1$.
+
+We accounted for the uncertainty in our random year effects when generating predictions, by drawing random year effects for each predicted year from a normal distribution with a mean of zero and a standard deviation drawn from the posterior estimate of the standard deviation of the random year effects \ref{eqn:}.
 We generated predictions using the full posterior distributions of each model parameter which allowed us to carry forward all of the uncertainty of the fitted vital rate models into our cover predictions.
@@
 
-We successfully changed the average soil moisture measured in the plots we covered with drought shelters and irrigated, in the directions we expected (fig); our spot measurements conducted during the springs of X, X and X showed that drought plots were X percent drier, while irrigation plots were X percent wetter than ambient plots.
-This difference was highly significant () but there was a lot of spatial variation in the soil moisture measurements within treatments (\ref{fig:}).
+Our treatments successfully changed the soil moisture in the experimental plots in the directions expected (fig.
+\ref{fig:VWCspot}).
+Our spot measurements of soil moisture from the spring showed that drought plots were roughly 50\% drier, while irrigated plots were roughly 40\% wetter than ambient plots.
+This difference was highly significant (table \ref{table:spotVWC}).
+
 The automatically recorded soil moisture data from only half of the experimental plots also showed that the drought treatment decreased and the irrigation treatment increased soil moisture, but the size of these effects were dependent on season and whether it had rained recently ().
@@
 
-\begin{figure}[tbp]
-\centering
-\includegraphics[width=1\textwidth]{treatment_trends_cover}
-\caption{Cover trends by treatment for the four dominant species.
-Note the difference in y-axis scale between the canopy cover of \textit{A.
-tripartita} and the basal cover of the dominant grasses.
-``Shrub removal'' refers to removal of \textit{A.
-tripartita}, and ``Grass removal'' refers to removal of all perennial grasses.
-The cover of species in treatments from which they are removed (e.g.
-\textit{A.
-tripartita} cover in shrub removal plots) is included to show values at the start of the experiment in 2011 and the efficacy of the removal treatments.}
-\label{fig:CoverTrends}
-\end{figure}
-
-\begin{figure}[tbp]
-\centering
-\includegraphics[width=1\textwidth]{cover_change_chrono}
-\caption{Observed and predicted year-to-year changes in cover for the four dominant species.
-Mean observed changes (averaged across quadrats) are shown in solid symbols, mean predicted changes are shown in open symbols.
-Values for control and removal plots are blue and red, respectively.
-Predictions from the ``Baseline'' model do account for effects of local species composition on vital rates, but do not include additional effects of removals.
-Predictions from the ``Removal'' model include additional treatment effects on vital rates.
-}
-\label{fig:CoverChange}
-\end{figure}
-
 \subsection*{Vital rates}
@@
 \ref{fig:VitalRateTest}; Table \ref{table:recruitment}).
-
- \begin{figure}[tbp]
- \centering
- \includegraphics[width=1\textwidth]{treatment_tests}
- \caption{Effects of removal treatments on the survival, growth, and recruitment of the four dominant species.
-Circles show the estimated coefficient of the removal effect, bars show 95\% confidence intervals.
-The removal effect shows response of \textit{A.
-tripartita} ("ARTR") to perennial grass removal, and response of the perennial grasses to \textit{A.
-tripartita} removal, after accounting for local neighborhood interactions.
-}
- \label{fig:VitalRateTest}
- \end{figure}
 
@@
 
- \begin{figure}[tbp]
- \centering
- \includegraphics[width=1\textwidth]{cover_projections_1step}
- \caption{Observed and predicted cover for each of the four modeled species.
-Solid lines show mean observed cover averaged across control (blue) and removal treatment (red) quadrats.
-Dashed lines show one-step-ahead predictions from the baseline IBM (no removal treatment coefficients), averaged across quadrats.
-Dotted lines show one-step-ahead predictions from an IBM that includes removal treatment coefficients.
-}
- \label{fig:IBM1step}
- \end{figure}
-
 To quantify the long-term population-level consequences of removals, we used an IPM to project stochastic equilibrium cover.
@@
 \ref{fig:IPMresults-maxCI}).
-
- \begin{figure}[tbp]
- \centering
- \includegraphics[width=1\textwidth]{boxplots}
- \caption{Equilibrium cover of the four dominant species simulated by the IPM.
-Boxplots show interannual variation reflecting random year effects.
-Gray boxes show cover simulated by a model with \textit{A.
-tripartita} present and no removal treatment coefficients, brown boxes show results from the same model but with \textit{A.
-tripartita} cover set to zero (a species removal), and green boxes show results from a model with  \textit{A.
-tripartita} set to zero and removal treatment coefficients included.
-}
- \label{fig:IPMresults}
- \end{figure}
 
@@
 \end{doublespacing}
+
+%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+% TABLES !
+%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
+\begin{table}[h]
+	\caption{Spring soil moisture}
+	\centering
+	%\begin{center}
+	\begin{tabular}{l c c }
+		\hline
+		& Model 1 \\
+		\hline
+		(Intercept)                  & $8.81^{*}$    	& $[5.78;\ 11.83]$  \\
+		TreatmentDrought             & $-3.97^{*}$      & $[-4.84;\ -3.09]$ \\
+		TreatmentIrrigation          & $3.26^{*}$       & $[2.39;\ 4.14]$   \\
+		\hline
+		AIC                          & 3191.87           \\
+		BIC                          & 3222.92           \\
+		Log Likelihood               & -1588.93          \\
+		Num. obs.                    & 624               \\
+		Num. groups: plot            & 24                \\
+		Num. groups: PrecipGroup     & 8                 \\
+		Num. groups: date            & 5                 \\
+		Var: plot (Intercept)        & 0.45              \\
+		Var: PrecipGroup (Intercept) & 0.23              \\
+		Var: date (Intercept)        & 11.24             \\
+		Var: Residual                & 8.90              \\
+		\hline
+		\multicolumn{2}{l}{\scriptsize{$^*$ 0 outside the confidence interval}}
+	\end{tabular}
+	\label{table:spotVWC}
+	%\end{center}
+\end{table}
+
+%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+% FIGURES !
+%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
+\begin{figure}[tbp]
+	\centering
+	\includegraphics[width=1\textwidth]{VWC_spot_measurements}
+	\caption{Soil moisture in the upper 5 cm of drought and irrigated plots compared to ambient controls.
+Soil moisture was measured at six locations around each plot at five different dates during the spring.
+Control plots were nearby areas of experiencing ambient soil moisture.
+Box plots show the mean soil moisture and the upper lower interquartile range.
+Dots show individual soil moisture measurements.
+Readings of volumetric soil moisture less than zero were occasionally obtained in very dry soil.}
+	\label{fig:VWCspot}
+\end{figure}
+
+\begin{figure}[tbp]
+	\centering
+	\includegraphics[width=1\textwidth]{avg_daily_soil_moisture}
+	\caption{Average soil moisture in the control, drought, and irrigation treatments during each year of the experiment.
+Soil moisture was monitored in four drought plots and four irrigated plots and four ambient control plots.
+Averages represent average soil moisture recorded every two hours over the course of the day at 5 cm and 25 cm depths.}
+	\label{fig:CoverTrends}
+\end{figure}
+
+\begin{figure}[tbp]
+	\centering
+	\includegraphics[width=1\textwidth]{cover_change_chrono}
+	\caption{Observed and predicted year-to-year changes in cover for the four dominant species.
+Mean observed changes (averaged across quadrats) are shown in solid symbols, mean predicted changes are shown in open symbols.
+Values for control and removal plots are blue and red, respectively.
+Predictions from the ``Baseline'' model do account for effects of local species composition on vital rates, but do not include additional effects of removals.
+Predictions from the ``Removal'' model include additional treatment effects on vital rates.
+}
+	\label{fig:CoverChange}
+\end{figure}
 
@@
 
- \begin{figure}[h]
- \centering
- \includegraphics[width=0.7\textwidth]{CompKernels}
- \caption{Competition kernels for the four dominant species.  }
- \label{fig:CompKernels}
- \end{figure}
-
-  \begin{figure}[h]
-  \centering
-  \includegraphics[width=1\textwidth]{climate}
-  \caption{Annual precipitation (a) and mean temperature (b) during the period of the experiment, shown against the long-term (1927-2016) means (solid blue and red lines) and 5\% and 95\% quantiles (dashed lines).
-}
-  \label{fig:climate}
-  \end{figure}
-
-  \begin{figure}[tbp]
-  \centering
-  \includegraphics[width=1\textwidth]{PSSP_W_scatters}
-  \caption{Bivariate scatter plots comparing crowding exerted by each neighbor species on \textit{P.
-spicata}.
-Blue symbols show values from controls plots, red symbols show values from \textit{Artemisia} removal plots.``W.allcov'' refers to the aggregated crowding by all shrubs and perennial grasses beyond the focal dominant species, and ``W.allpts'' refers to the aggregated crowding by forb species.
-}
-  \label{fig:Wscatters}
-  \end{figure}
-
-  \begin{figure}[tbp]
- \centering
- \includegraphics[width=1\textwidth]{PSSP_W_byTrt}
- \caption{Distribution of crowding exerted by each neighbor species on \textit{P.
-spicata} individuals in controls plots (blue) and removal quadrats (red) in 2011, before \textit{Artemisia}  removals were conducted.
-Inset bar graphs show the probability that a particular neighbor species was present ($W>0$) in a focal plant's local neighborhood in control (blue) and removal (red) quadrats.
-``W.allcov'' refers to the aggregated crowding by all shrubs and perennial grasses beyond the focal dominant species, and ``W.allpts'' refers to the aggregated crowding by forb species.
-}
- \label{fig:W-by-treatment}
- \end{figure}
-
- \begin{figure}[tbp]
- \centering
- \includegraphics[width=1\textwidth]{cover_projections_1step_maxCI}
- \caption{Observed and predicted cover for each of the four modeled species assuming maximum effects of removal.
-Solid lines show mean observed cover averaged across control (blue) and removal treatment (red) quadrats.
-Dashed lines show one-step-ahead predictions from the baseline IBM (no removal treatment coefficients), averaged across quadrats.
-Dotted lines show one-step-ahead predictions from an IBM that includes removal treatment coefficients representing the outermost values of the 95\% confidence intervals.
-}
- \label{fig:IBM1step-maxCI}
- \end{figure}
-
-  \begin{figure}[tbp]
-  \centering
-  \includegraphics[width=1\textwidth]{boxplots-maxCI}
-  \caption{Equilibrium cover of the four dominant species simulated by the IPM with maximum removal treatment effects.
-Boxplots show interannual variation reflecting random year effects.
-Gray boxes show cover simulated by a model with \textit{A.
-tripartita} present and no removal treatment coefficients, brown boxes show results from the same model but with \textit{A.
-tripartita} cover set to zero (a species removal), green boxes show results from a model with  \textit{A.
-tripartita} set to zero and (mean) removal treatment coefficients included, and blue boxes show results with removal treatments coefficients assigned the outermost value of the 95\% confidence intervals.
-}
-  \label{fig:IPMresults-maxCI}
-  \end{figure}
-
 \end{document}

### 2016-12-19 precip/R/climate/aggregate_spot_VWC.R

@@ -32,7 +32,7 @@ spotVWC <-
   group_by( season, date, PrecipGroup,rainfall, Treatment ) %>% 
   summarise( avg_VWC = mean(VWC, na.rm = TRUE)) %>% 
   group_by(PrecipGroup) %>% 
-  mutate( avg_VWC = scale(avg_VWC)) %>%  # scale within Precip Group and Depth 
+  mutate( avg_VWC = scale(avg_VWC, mean(avg_VWC[Treatment == 'Control'], na.rm = T), sd(avg_VWC[Treatment == 'Control'], na.rm = T))) %>%  # scale within Precip Group and Depth 
   spread( Treatment, avg_VWC) %>%
   mutate( Drought = Drought - Control, Irrigation = Irrigation - Control ) %>% 
   arrange( PrecipGroup, date) 

### 2016-12-19 precip/analysis/analysis_pipeline.R

@@ -56,7 +56,7 @@ source('analysis/figure_scripts/plot_treatment_parameter_estimates.R')
 
 # 6.  check for divergence 
 
- source('analysis/check_for_divergent_transitions.R')
+source('analysis/check_for_divergent_transitions.R')
 
 # 7. plot observed cover and predicted cover 
 

### 2016-12-20 precip/manuscript/precip_ms.tex
adding revisions to the soil moisture statistics and the manuscript

@@
 Data were logged automatically every four hours.
-We augmented automatic monitoring of the climate in these plots with by taking direct measurements of soil moisture with a handheld EC-5 soil moisture sensor around each plot at X different times in the spring and summer between 2012 and 2016.
+We augmented automatic monitoring of the climate in these plots with by taking direct measurements of soil moisture with a handheld EC-5 soil moisture sensor at six points around each plot on 6/6/2012, 4/29/2015, 5/7/2015, 6/9/2015 and 5/10/2016.
+We analyzed these spot measurements for significant treatment effects on soil moisture using a linear mixed effects model with the \textit{lmer} package in \textit{R} \citep{bates}, with plot, plot group, and date as random effects in the model.
 
@@
 We tested for the effect of precipitation treatment on this measure with a linear model in \textit{R}.
-We considered the precipitation treatment effects to be significant when??.
 
@@
 
-We averaged daily soil moisture predictions from SOILWAT from upper 40 cm of soil and then aggregated these seasonally to serve as the covariates in the vital rate regressions for each species.
-In order to predict species' performance in the experimental plots we took the raw SOILWAT predictions for the ambient conditions for each day during the experimental period and then adjusted these up or down using coefficients from a linear model fit to the changes in soil moisture directly observed in the drought and irrigation plots.
-
-Because we did not monitor soil moisture directly in all X control, drought and irrigation plots, we used a model to describe the average treatment effects on soil moisture during the course of the experiment.
-To do this we first averaged observed soil moisture data from the dataloggers installed in four of the plot groups by date and plot and then within each plot group we found the average deviation from ambient soil moisture by subtracting the average soil moisture in the drought and irrigation plots from the average soil moisture in the nearby control areas.
-This left us with a list of average daily deviations from ambient soil moisture in the drought and irrigation plots.
-
-We expected that the effects the drought and irrigation experiment would have on soil moisture would depend on the season and whether it had rained recently.
-So we classified the daily climate during the experimental period into four seasons and classified each day as dry or rainy.
-Rainy days were defined as any days when average temperatures were above 3 degrees C and precipitation was over x mm.
+We averaged daily soil moisture predictions from SOILWAT from upper 40 cm of soil and then averaged these seasonally to serve as the covariates in the vital rate regressions for each species.
+Because we did not monitor soil moisture directly in all control, drought and irrigation plots, we used a model to describe the average treatment effects on soil moisture during the course of the experiment.
+To do this we first averaged observed soil moisture data by date and plot and then standardized these by the mean and standard deviation of the control soil moisture conditions observed within each plot group.
+We then found the difference between the soil moisture in the treated plots and the ambient conditions.
+We then modeled these treatment effects as a function of season and whether a day was rainy or dry.
+We expected that our drought and irrigation treatments might be more effective during rainy weather than during dry weather.
+Rainy days were defined as any days when any precipitation was recorded and average temperatures were above 3 degrees C.
 The day immediately following rainfall was also classified as rainy.
-
-We modeled the average daily deviations in soil moisture and temperature in the treatments as a function of season, whether it had rained recently and the interaction between these two factors.
-After all this, we could then use this model to adjust the ambient soil moisture values predicted from the SOILWAT model above.
+We fit this model using the \textit{lmer} package in \textit{R} \citep{bates} with random effects for plot group and date.
+We then used this model to predict the treatment effects on soil moisture for the entire study period from the ambient soil moisture values predicted from the SOILWAT model described above.
 These adjusted soil moisture values reflected the average season and rainfall dependent effects of the experimental treatments on soil moisture and could be used as covariates for predicting the effects of our manipulation on each species demographic rates.
@@
 Our treatments successfully changed the soil moisture in the experimental plots in the directions expected (fig.
-\ref{fig:VWCspot}).
-Our spot measurements of soil moisture from the spring showed that drought plots were roughly 50\% drier, while irrigated plots were roughly 40\% wetter than ambient plots.
-This difference was highly significant (table \ref{table:spotVWC}).
-
-The automatically recorded soil moisture data from only half of the experimental plots also showed that the drought treatment decreased and the irrigation treatment increased soil moisture, but the size of these effects were dependent on season and whether it had rained recently ().
-The best model for predicting the average treatment effects on soil moisture was : .
-We did not find evidence that the drought shelters and the irrigation treatments significantly changed air temperature within the plots ().
-
-The SOILWAT soil moisture model indicated that soil moisture fluctated from between X and X on average each year, with the months of X and X being the wettest while soil moisture during the months of X and X was driest (fig).
-Predicted seasonal soil moisture for each year was positively correlated with precipitation, and negatively correlated with average temperatures.
-The average soil moisture predicted from soilwat for the entire study period from X to X is shown in figure X.
-Soil moisture predictions for the ambient conditions from 2011 to 2016 fall on the low side of the period of record averages.
-This was driven by higher than average temperatures and severely dry winters and spring conditions in the years of X.
-
-When we adjust the seasonally averaged soil moisture predictions by the treatment effects, we find that the soil moisture conditions predicted in the drought plots during the course of the experiment are lower than those predicted for any of the previous years in the historical record.
-Likewise the predicted soil moisture in the irrigation plots was far above the average for the historical record in the years of X and X.
-We interpret this as evidence that our drought and irrigation plots created truly novel soil moisture conditions in the years X and X that were unlikely to have ever been experienced by the study species during the entire period of the observational data.
-We visualize this by conducting a principle component analysis (PCA) on all the seasonal soil moisture, and plotting each year on the first two principle components.
-The PCA shows that the ambient, drought and irrigation plots experienced warmer than average seasonal conditions during the course of the experiment (further left on PCA1).
-The drought plots were considerably drier, and the irrigated plots considerably wetter (figure PCA).
+\ref{fig:spotVWC}).
+Spring spot measurements of soil moisture from all the plots showed that on average the drought plots were roughly 50\% drier, while irrigated plots were roughly 40\% wetter than ambient conditions (table \ref{table:spotVWC}).
+
+The continuously recorded soil moisture data also recorded treatment effects on soil moisture, but these effects were weaker on average than the more complete spot measurements and depended on season and recent rainfall (table \ref{table:soil_moisture_model}; fig \ref{fig:dailyVWC}).
+We saw weaker effects during the spring than during the fall and summer: the drought plots were about 20-30\% drier in the fall and summer but only 7 to 14\% drier during the spring.
+The irrigated plots were 30\% wetter during the fall and summer but only 20 to 25\% wetter during the spring.
+Treatment differences were slightly larger during rainy periods, especially in the spring.
+We did not find evidence that the drought shelters and the irrigation treatments consistently affected air temperature within the plots.
+
+The SOILWAT soil moisture model predicted average monthly soil volumetric water content of between 10 ml/ml and 15 ml/ml each month, with the month of April being the wettest and the month of July, August and September being the driest.
+Annual variation in seasonal soil moisture for each year was positively correlated with seasonal precipitation and negatively correlated with seasonal temperature.
+During the course of the experiment, SOILWAT reproduced much of the daily variation observed in soil moisture recorded by our automatic data loggers, but the average soil moisture predicted by SOILWAT was about 5 ml/ml wetter than the soil moisture content observed.
+
+After adjusting the SOILWAT seasonal soil moisture predictions by the treatment effects, we found that the soil moisture predicted in the drought plots during the course of the experiment was well below the historical seasonal averages: the summer of 2012 and 2013, the fall of 2013, and the spring and winter of 2014 fell below the 5th percentile limit for drought in the historical period \ref{fig:seasonalVWC}.
+Soil moisture in our irrigation plots was generally above the historical average soil moisture but conditions never exceeded the 90th percentile for soil moisture in the historical period.
 
@@
 comata} and \textit{P.
-spicata} fell significantly by X and X percent respectively from the pre-treatment cover in 2011 to 2016.
+spicata} fell significantly in the drought plots from the pre-treatment cover 2011 to 2016 (tables \ref{table:changeHECO}, \ref{table:changePSSP}, fig \ref{fig:coverChange}).
 The cover of \textit{P.
-secunda} showed a trend towards decrease in the drought plots and an increase in the irrigated plots but these changes were not significant.
-In contrast the cover of \textit{A.
-tripartita} changed in the opposite direction from the grasses; it increased in the drought plots and decreased in the irrigated plots but these changes were not significant (fig).
-
-Our treatment models fit to the experimental and observational data indicated a variety of treatment effects on the vital rates of each species (table x).
-Among the survival models, WAIC scores indicated that we should retain size x treatment effects for \textit{P.
+secunda} showed a slight decrease in the drought plots and an increase in the irrigated plots but these changes were not significant (table \ref{table:changePOSE}).
+In contrast to the grasses, the cover of \textit{A.
+tripartita} increased slightly in all three treatments (fig \ref{fig:coverChange}).
+
+Our treatment models fit to the experimental and observational data indicated a variety of treatment effects on the vital rates of each species.
+After comparing WAIC scores for the growth models with and without the size by treatment effects, we retained size by treatment interaction effects in the growth models for \textit{A.
+tripartita} and \textit{P.
+secunda}, and the survival model for \textit{P.
 secunda}.
-Among the growth models, we retained size by treatment effects for \textit{A.
-tripartita} and \textit{P.
-secunda}.
-DISCUSS SIGNIFICANT treatment effects: For \textit{A.
-tripartita} we observed a significant decrease in growth in the x treatment, while we observed no change in the X treatment.
+For \textit{A.
+tripartita} the model showed a significant size by treatment effect in the drought plots: the effect of drought was positive for plants of average size and smaller \ref{fig:growthTreat}, but plants larger than the mean by more than 1.5 standard deviations grew slightly less in the drought treatment than in the controls.
+\textit{A.
+tripartita} showed the opposite response in the irrigation plots, (although the irrigation parameters were not technically significant at the 95\% confidence level), irrigation reduced growth the most for small plants while irrigation had positive effects on plants more than 1.5 standard deviations greater than the mean size.
 \textit{H.
-comata} growth, survival and recruitment, while growth.
+comata} growth showed a strong (but not significant) decrease in the drought plots and no response to the irrigation plots.
+Like \textit{A.
+tripartita}, \textit{P.secunda} growth also showed a size by treatment effect in the drought plots, with the negative effects on growth becoming greater for larger plants.
 \textit{P.
-secunda} growth and survival recruitment.
+secunda} showed the opposite response in the irrigation plots with larger plants showing the largest increase in growth in response to irrigation (although these effects were technically not significant).
 \textit{P.
-spicata} growth survival and recruitment.
-Treatment by size effects on \textit{A.
-tripartita} and \textit{P.
-secunda} showed that ()
+spicata} growth was relatively unaffected by the drought and irrigation treatments.
+
+The effects of drought were negative and signficant on the survival of all three grass species (fig \ref{fig:)
 
@@
 As expected from previous research most of our demographic models estimated strong negative intra-specific crowding effects and weaker negative inter-specific crowding effects on the focal species (appendix) \citep{adler_coexistence_2010,chu_direct_2016,chu_large_2015}.
-
-(Fig. \ref{fig:CoverTrends}), but shrub removal did appear to cause an increase in the cover of \textit{P. spicata}. 
-We tested the effects of removal on cover by analyzing year-to-year changes in (log) cover, a measure of population growth rate (Fig.
-\ref{fig:CoverChange}).
-The removal treatments had no significant effect on
-\textit{A. tripartita} (95\% confidence interval on removal treatment effect: -0.75, 0.56) or 
-\textit{Poa secunda} (95\% CI: -0.41, 0.25) population growth, but did increase 
-\textit{H. comata} (95\% CI: 0.01, 0.65) and
- \textit{P. spicata} (95\% CI: 0.01, 0.40) population growth 
-(Table \ref{table:coefficients}). However, these treatment comparisons do not address our questions about whether responses to removal were 
-adequately predicted by models fit to observational data on non-removal plots.
-
-\subsection*{Vital rates}
-
-Our vital rate models take into account the effects of local species composition and should therefore be able to predict the effect of removals on survival, growth and recruitment without any additional parameters.
-If the removal treatment effects that we added to these models are statistically significant, it would indicate an effect of removals that goes above and beyond the response predicted by a model based on data from the control plots.
-For the survival models, the 95\% confidence intervals on the removal effects overlapped zero for all four species (Fig.
-\ref{fig:VitalRateTest}; Tables \ref{ARTRsurvival} to \ref{PSSPsurvival}), meaning that the survival probabilities of plants in removal plots can be predicted based on data from control plots.
-However, for the growth models, the 95\% confidence intervals on the removal effects were positive for \textit{Poa secunda}  and \textit{P.
-spicata}, meaning that individuals of these species grew more rapidly in removal plots than controls even after accounting for neighborhood interactions (Fig.
-\ref{fig:VitalRateTest}; Tables \ref{ARTRgrowth} to \ref{PSSPgrowth}).
-In contrast, \textit{Poa secunda} had lower recruitment in removal than controls plots (upper limit of 95\% CI $<0$; Fig.
-\ref{fig:VitalRateTest}; Table \ref{table:recruitment}).
-
-The positive removal effects on \textit{Poa secunda}  and \textit{P.
-spicata} growth are particularly interesting, because they suggest that models based on observational data may underestimate competitive release.
-We conducted additional analyses to further explore removal effects on the growth of these two species.
-First, we classified all individuals of these species as located either inside or outside the limits of \textit{A.
-tripartita} canopies present in 2011 (before removal).
-If removal is releasing grasses from \textit{A.
-tripartita} competition, we would expect individuals located under the shrub canopies to experience a greater effect of removal.
-However, the location of individuals with respect to \textit{A.
-tripartita} canopies had no effect on growth of either species (Tables \ref{table:POSEgrowth-inARTR} and \ref{table:PSSPgrowth-inARTR}), meaning that individuals located outside \textit{A.
-tripartita} canopies responded to removals as much as individuals underneath (removed) canopies.
-
-Second, we evaluated the possibility that removal effects might be transient.
-We removed aboveground plant material, but roots of the removed plants remain in the ground and decompose, potentially creating a flush of mineral nutrients.
-We tested for a transient removal effect by adding a year-by-treatment interaction.
-For both  \textit{Poa secunda}  and \textit{P.
-spicata} , we found no sign of a decrease in the removal effect over time (Tables \ref{table:POSEgrowth-byYr} and \ref{table:PSSPgrowth-byYr}).
-
-Third, responses to removals might be poorly predicted by models based on observational data if the removals create local crowding outside the range of historical variation or if the removals break covariances in crowding among species.
-However, the distribution of crowding observed within control plots easily spans the crowding values observed in removal plots (Fig.
-\ref{fig:Wscatters}), and we failed to find evidence for strong interspecific covariances in crowding (\ref{suppMethods}).
-
-Finally, we evaluated pre-existing differences in community composition in removal and control plots.
-In selecting removal plots, our criteria required candidate plots to include a certain amount of our target removal species.
-Because we adopted the historical plots as our controls, the same criteria were not applied (a control plot might not contain any \textit{A.
-tripartita}).
-To quantify resulting differences in pre-treatment (2011) composition, we compared neighborhood crowding at the individual plant level (the $\boldsymbol{W}$ of the growth regression) across treatments.
-\textit{A.
-tripartita} crowding and crowding by all subdominant shrub and grass species appeared greater than in the controls, whereas crowding by \textit{H.
-comata} was lower in removal plots than controls, though it was absent from most \textit{P.
-spicata} neighborhood in both treatments (Fig.
-\ref{fig:W-by-treatment}).
-These pre-existing differences in composition raise the possibility that plots selected for the removal treatments might differ from the control plots in some unobserved covariate that could influence plant performance.
-
-\subsection*{Population-level responses}
-
-To quantify the short-term population-level consequences of removals, we used an IBM to project the cover of each species in each quadrat one year ahead using versions of the model with and without removal treatment effects.
-Our models tended to under-predict population growth (Fig.
-\ref{fig:CoverChange}) and population size (Fig.
-\ref{fig:IBM1step}) for all species, but the difference between predictions from the baseline and removal models varied among species.
-For \textit{A.
-tripartita}, both observed and predicted population growth rates were close to zero, and differences in model predictions were small.
-\textit{H.
-comata} experienced greater variation in growth rates, but the inclusion of removal treatment effects had little effect on population growth rates.
-For \textit{Poa secunda} and especially \textit{P.
-spicata}, the model with removal effects consistently predicted higher population growth than the baseline model (Fig.
-\ref{fig:CoverChange}), and reduced the underestimates of cover (Fig.
-\ref{fig:IBM1step}).
-The version of the model with maximum removal treatment effects (removal coefficients set to the 95\% confidence limit furthest from zero), were qualitatively similar, but further amplified the contrast between predictions of the baseline and removal models for and \textit{P.
-spicata} (Fig.
-\ref{fig:IBM1step-maxCI}).
-
-To quantify the long-term population-level consequences of removals, we used an IPM to project stochastic equilibrium cover.
-We concentrated on grass species' responses to \textit{A.
-tripartita} removal, since the growth regressions indicated our baseline model might underestimate competitive release in this situation (our baseline vital rate models worked well in predicting \textit{A.
-tripartita} responses to grass removal).
-Using the baseline model, with no removal treatment coeffiicients, the IPM predicts little effect of \textit{A.
-tripartita} removal on the cover of any grass species (compare gray and brown boxes in Fig.
-\ref{fig:IPMresults}).
-Using a model with removal treatment coefficients, the IPM predicts little effect of shrub removal on \textit{Poa secunda} and \textit{H.
-comata}, but a 2-fold increase in the cover of \textit{P.
-spicata} (Fig.
-\ref{fig:IPMresults}).
-Running a version of the model with maximum removal treatment effects (removal coefficients set to the 95\% confidence limit furthest from zero), resulted in the same qualitative outcome, but a larger increase in \textit{P.
-spicata} (Fig.
-\ref{fig:IPMresults-maxCI}).
-
-The equilibrium cover of \text{Artemisia} simulated by the IPM (6.0\%) is consistent with observed cover during the historical data collection period (6.6\% from 1931-1956), but is low compared to contemporary observed cover (15-25\%,\ref{fig:CoverTrends}).
-Could the small competitive release following simulated \text{Artemisia} removal simply reflect the low \text{Artemisia} cover?
-To explore this possibility, we increased \text{Artemisia} cover in the baseline simulation by increasing average \text{Artemisia} fecundity.
-The increased fecundity led to a doubling of \text{Artemisia} average cover, but had essentially zero effect on the cover of the grasses (changes of less than 0.1\% cover).
 
@@
 
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-% TABLES !
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+\section{Tables}
 
 \begin{table}[h]
-	\caption{Spring soil moisture}
+	\caption{Treatment effects on spring soil moisture}
 	\centering
@@
 
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-% FIGURES !
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+\begin{table}
+	\caption{Model of treatment effects on soil moisture}
+	\begin{center}
+		\begin{tabular}{l c c }
+			\hline
+			& Model 1 \\
+			\hline
+			(Intercept)                       & $-0.57^{*}$       & $[-0.89;\ -0.26]$ \\
+			TreatmentIrrigation               & $1.23^{*}$        & $[1.18;\ 1.29]$   \\
+			rainfallrainy                     & $-0.05$           & $[-0.12;\ 0.01]$  \\
+			seasonspring                      & $0.27^{*}$        & $[0.23;\ 0.32]$   \\
+			seasonsummer                      & $0.15^{*}$        & $[0.10;\ 0.19]$   \\
+			seasonwinter                      & $0.25^{*}$        & $[0.21;\ 0.29]$   \\
+			TreatmentIrrigation:rainfallrainy & $0.18^{*}$        & $[0.13;\ 0.24]$   \\
+			TreatmentIrrigation:seasonspring  & $-0.23^{*}$       & $[-0.29;\ -0.16]$ \\
+			TreatmentIrrigation:seasonsummer  & $-0.26^{*}$       & $[-0.33;\ -0.20]$ \\
+			TreatmentIrrigation:seasonwinter  & $-0.33^{*}$       & $[-0.40;\ -0.27]$ \\
+			rainfallrainy:seasonspring        & $-0.23^{*}$       & $[-0.31;\ -0.16]$ \\
+			rainfallrainy:seasonsummer        & $-0.07$           & $[-0.14;\ 0.01]$  \\
+			rainfallrainy:seasonwinter        & $-0.07$           & $[-0.20;\ 0.07]$  \\
+			\hline
+			AIC                               & 14581.58          \\
+			BIC                               & 14695.49          \\
+			Log Likelihood                    & -7274.79          \\
+			Num. obs.                         & 9133              \\
+			Num. groups: date                 & 1596              \\
+			Num. groups: PrecipGroup          & 8                 \\
+			Var: date (Intercept)      		  & 0.00              \\
+			Var: PrecipGroup (Intercept)      & 0.19              \\
+			Var: Residual                     & 2.50              \\
+			\hline
+			\multicolumn{2}{l}{\scriptsize{$^*$ 0 outside the confidence interval}}
+		\end{tabular}
+		\label{table:soil_moisture_model}
+	\end{center}
+\end{table}
+
+% latex table generated in R 3.3.1 by xtable 1.8-2 package
+% Mon Dec 19 18:10:15 2016
+\begin{table}[ht]
+	\centering
+	\begin{tabular}{rrrrr}
+		\hline
+		& Estimate & Std. Error & t value & Pr($>$$|$t$|$) \\ 
+		\hline
+		(Intercept) & -0.2835 & 0.4627 & -0.61 & 0.5487 \\ 
+		TreatmentDrought & 0.9378 & 0.7743 & 1.21 & 0.2434 \\ 
+		TreatmentIrrigation & 0.8882 & 0.7743 & 1.15 & 0.2682 \\ 
+		\hline
+	\end{tabular}
+	\caption{Treatment effects on log cover change for 	extit{A. tripartita} from 2011 to 2016. Intercept gives control effects.} 
+	\label{table:changeARTR}
+\end{table}
+
+% latex table generated in R 3.3.1 by xtable 1.8-2 package
+% Mon Dec 19 18:10:16 2016
+\begin{table}[ht]
+	\centering
+	\begin{tabular}{rrrrr}
+		\hline
+		& Estimate & Std. Error & t value & Pr($>$$|$t$|$) \\ 
+		\hline
+		(Intercept) & 0.3982 & 0.2587 & 1.54 & 0.1548 \\ 
+		TreatmentDrought & -2.9951 & 0.5784 & -5.18 & 0.0004 \\ 
+		TreatmentIrrigation & -0.1219 & 0.4953 & -0.25 & 0.8105 \\ 
+		\hline
+	\end{tabular}
+	\caption{Treatment effects on log cover change for 	extit{H. comata} from 2011 to 2016. Intercept gives control effects.} 
+	\label{table:changeHECO}
+\end{table}
+
+% latex table generated in R 3.3.1 by xtable 1.8-2 package
+% Mon Dec 19 18:10:16 2016
+\begin{table}[ht]
+	\centering
+	\begin{tabular}{rrrrr}
+		\hline
+		& Estimate & Std. Error & t value & Pr($>$$|$t$|$) \\ 
+		\hline
+		(Intercept) & -0.7247 & 0.4613 & -1.57 & 0.1298 \\ 
+		TreatmentDrought & 0.0273 & 0.8208 & 0.03 & 0.9737 \\ 
+		TreatmentIrrigation & 1.1459 & 0.7797 & 1.47 & 0.1552 \\ 
+		\hline
+	\end{tabular}
+	\caption{Treatment effects on log cover change for 	extit{P. secunda} from 2011 to 2016. Intercept gives control effects.} 
+	\label{table:changePOSE}
+\end{table}
+
+% latex table generated in R 3.3.1 by xtable 1.8-2 package
+% Mon Dec 19 18:10:16 2016
+\begin{table}[ht]
+	\centering
+	\begin{tabular}{rrrrr}
+		\hline
+		& Estimate & Std. Error & t value & Pr($>$$|$t$|$) \\ 
+		\hline
+		(Intercept) & 0.0188 & 0.2124 & 0.09 & 0.9303 \\ 
+		TreatmentDrought & -0.8851 & 0.3780 & -2.34 & 0.0287 \\ 
+		TreatmentIrrigation & 0.1453 & 0.3780 & 0.38 & 0.7044 \\ 
+		\hline
+	\end{tabular}
+	\caption{Treatment effects on log cover change for 	extit{P. spicata} from 2011 to 2016. Intercept gives control effects.} 
+	\label{table:changePSSP}
+\end{table}
+
+\section{Figures}
 
@@
 Control plots were nearby areas of experiencing ambient soil moisture.
-Box plots show the mean soil moisture and the upper lower interquartile range.
+Box plots show the median soil moisture and the interquartile range.
 Dots show individual soil moisture measurements.
 Readings of volumetric soil moisture less than zero were occasionally obtained in very dry soil.}
-	\label{fig:VWCspot}
+	\label{fig:spotVWC}
 \end{figure}
@@
 	\caption{Average soil moisture in the control, drought, and irrigation treatments during each year of the experiment.
-Soil moisture was monitored in four drought plots and four irrigated plots and four ambient control plots.
-Averages represent average soil moisture recorded every two hours over the course of the day at 5 cm and 25 cm depths.}
-	\label{fig:CoverTrends}
+Soil moisture was monitored in four drought plots, four irrigated plots and four ambient control plots.
+Two sensors were installed at 5 cm depth at each plot and two at 25 cm and data was logged every 2 hours.}
+	\label{fig:dailyVWC}
 \end{figure}
@@
 	\centering
-	\includegraphics[width=1\textwidth]{cover_change_chrono}
-	\caption{Observed and predicted year-to-year changes in cover for the four dominant species.
-Mean observed changes (averaged across quadrats) are shown in solid symbols, mean predicted changes are shown in open symbols.
-Values for control and removal plots are blue and red, respectively.
-Predictions from the ``Baseline'' model do account for effects of local species composition on vital rates, but do not include additional effects of removals.
-Predictions from the ``Removal'' model include additional treatment effects on vital rates.
+	\includegraphics[width=1\textwidth]{modern_soil_moisture_comparison}
+	\caption{Average seasonal soil moisture in the control, drought, and irrigation treatments during each year of the experiment.
+The dashed gray lines give the 5th percentile and 95th percentile limits for seasonal soil moisture in the historical record (1929 to 2010).
 }
-	\label{fig:CoverChange}
+	\label{fig:seasonalVWC}
+\end{figure}
+
+\begin{figure}[tbp]
+	\centering
+	\includegraphics[width=1\textwidth]{treatment_trends_cover}
+	\caption{Average cover of each of the focal species in the control, drought, and irrigation treatments during the experiment.
+Percent cover of \textit{A.
+tripartita} refers to canopy cover, while the cover of the three grasses refers to basal cover.
+}
+	\label{fig:coverTrends}
+\end{figure}
+
+\begin{figure}[tbp]
+	\centering
+	\includegraphics[width=1\textwidth]{start_to_finish_cover_change}
+	\caption{Log change in cover in each of the experimental plots from the pre-treatment monitoring in 2011 to the last year of the experiment in 2016.
+Box plots show the median cover change and the interquartile range.}
+	\label{fig:coverChange}
+\end{figure}
+
+\begin{figure}[tbp]
+	\centering
+	\includegraphics[width=1\textwidth]{treatment_effect_growth}
+	\caption{Parameter estimates for the effects of treatment on growth of all four species.
+We assessed a parameter as significant when the 95\% Bayesian credible intervals did not overlap zero.
+Size by treatment interactions were only fit for ARTR, and POSE.
+Plant size was centered on mean size and scaled by its standard deviation.
+ARTR = \textit{A.
+tripartita}, HECO = \textit{H.
+comata}, POSE = \textit{P.
+secunda}, PSSP = \textit{P.
+spicata}.
+}
+	\label{fig:growthTreat}
+\end{figure}
+
+\begin{figure}[tbp]
+	\centering
+	\includegraphics[width=1\textwidth]{treatment_effect_survival}
+	\caption{Parameter estimates for the effects of treatment on survival of all four species.
+We assessed a parameter as significant when the 95\% Bayesian credible intervals did not overlap zero.
+Size by treatment interactions were only fit for POSE.
+Plant size was centered on mean size and scaled by its standard deviation.
+ARTR = \textit{A.
+tripartita}, HECO = \textit{H.
+comata}, POSE = \textit{P.
+secunda}, PSSP = \textit{P.
+spicata}.
+}
+	\label{fig:survivalTreat}
+\end{figure}
+
+\begin{figure}[tbp]
+	\centering
+	\includegraphics[width=1\textwidth]{treatment_effect_recruitment}
+	\caption{Parameter estimates for the effects of treatment on recruitment of all four species.
+We assessed a parameter as significant when the 95\% Bayesian credible intervals did not overlap zero.
+ARTR = \textit{A.
+tripartita}, HECO = \textit{H.
+comata}, POSE = \textit{P.
+secunda}, PSSP = \textit{P.
+spicata}.
+}
+	\label{fig:recruitmentTreat}
 \end{figure}

### 2016-12-20 precip/analysis/cover_trends/treatment_trends_precip.r

@@ -168,7 +168,7 @@ png("figures/treatment_trends_cover.png",height=2.75,width=8,units="in",res=400)
 
   # mean cover
   for(doSpp in sppList){
-    tmp.mean<-subset(spp.mean,species==doSpp)
+    tmp.mean<-subset(spp.mean,species==doSpp & year > 2010)
     if(doSpp==sppList[1]){
       my.y <- c(0,max(tmp.mean[,3:5], na.rm = T))
     }else{
@@ -239,7 +239,6 @@ subset( sppD.q, species == 'Hesperostipa comata' & year == 2016) %>%
   arrange( year ) %>% 
   filter( cover > 0 )
 
-png("figures/start_to_finish_cover_change.png", height=3,width=8.5,units="in",res=400)
 
 temp <- subset( sppD.q , year == 2016 )
 temp$Treatment <- factor(temp$Treatment )
@@ -252,13 +251,43 @@ lc.HECO <- lm(dat =  subset( temp, species == 'HECO'), lc ~ Treatment )
 lc.POSE <- lm(dat =  subset( temp, species == 'POSE'), lc ~ Treatment )
 lc.PSSP <- lm(dat =  subset( temp, species == 'PSSP'), lc ~ Treatment )
 
+
 summary(lc.ARTR)
 summary(lc.HECO)
 summary(lc.POSE)
 summary(lc.PSSP)
 
-temp
+library(xtable)
+xt1 <- xtable(lc.ARTR, 
+              caption = 'Treatment effects on log cover change for \textit{A. tripartita} from 2011 to 2016. Intercept gives control effects.', 
+              label = 'table:changeARTR')
+xt2 <- xtable(lc.HECO, caption = 'Treatment effects on log cover change for \textit{H. comata} from 2011 to 2016. Intercept gives control effects.', 
+              label = 'table:changeHECO')
+xt3 <- xtable(lc.POSE, caption = 'Treatment effects on log cover change for \textit{P. secunda} from 2011 to 2016. Intercept gives control effects.', 
+              label = 'table:changePOSE')
+xt4 <- xtable(lc.PSSP, caption = 'Treatment effects on log cover change for \textit{P. spicata} from 2011 to 2016. Intercept gives control effects.', 
+              label = 'table:changePSSP')
 
+print(xt1, 'output/results_tables/ARTR_cover_change.txt', type = 'latex')
[…]

### 2016-12-20 precip/analysis/figure_scripts/plot_cover_over_time.R

@@ -30,6 +30,7 @@ modern_data$Period <- 'Modern'
 
 all_data <- rbind(historical_data, modern_data)
 
+
 all_data <-
   all_data %>% 
   filter(! Treatment %in% c('No_grass', 'No_shrub')) %>% 

### 2016-12-20 precip/manuscript/precip_ms.tex
adding manuscript edits to results climate model section

@@
 
-The continuously recorded soil moisture data also recorded treatment effects on soil moisture, but these effects were weaker on average than the more complete spot measurements and depended on season and recent rainfall (table \ref{table:soil_moisture_model}; fig \ref{fig:dailyVWC}).
-We saw weaker effects during the spring than during the fall and summer: the drought plots were about 20-30\% drier in the fall and summer but only 7 to 14\% drier during the spring.
-The irrigated plots were 30\% wetter during the fall and summer but only 20 to 25\% wetter during the spring.
+The continuously recorded soil moisture data also showed treatment effects on soil moisture, but these effects were weaker on average than the spot measurements and depended on season and recent rainfall (table \ref{table:soil_moisture_model}; fig \ref{fig:dailyVWC}).
+We saw weaker effects during the spring than during the fall and summer: the drought plots were about 20-30\% drier in the fall and summer but only 7 to 14\% drier during the spring, while the irrigated plots were 30\% wetter during the fall and summer but only 20-25\% wetter during the spring.
 Treatment differences were slightly larger during rainy periods, especially in the spring.
-We did not find evidence that the drought shelters and the irrigation treatments consistently affected air temperature within the plots.
-
-The SOILWAT soil moisture model predicted average monthly soil volumetric water content of between 10 ml/ml and 15 ml/ml each month, with the month of April being the wettest and the month of July, August and September being the driest.
+We did not find evidence that the drought shelters and the irrigation treatments consistently affected air temperature at 30 cm above the plots.
+
+The SOILWAT soil moisture model predicted average monthly soil volumetric water content of between 10 ml/ml and 15 ml/ml each month, with the month of April being the wettest and the month of July, August and September being the driest on average.
 Annual variation in seasonal soil moisture for each year was positively correlated with seasonal precipitation and negatively correlated with seasonal temperature.
-During the course of the experiment, SOILWAT reproduced much of the daily variation observed in soil moisture recorded by our automatic data loggers, but the average soil moisture predicted by SOILWAT was about 5 ml/ml wetter than the soil moisture content observed.
+During the course of the experiment, SOILWAT reproduced much of the daily variation observed in soil moisture recorded by our automatic data loggers, but the average soil moisture predicted by SOILWAT was about 5 ml/ml wetter than the soil moisture content recorded by the data loggers.
 
@@
 comata} and \textit{P.
-spicata} fell significantly in the drought plots from the pre-treatment cover 2011 to 2016 (tables \ref{table:changeHECO}, \ref{table:changePSSP}, fig \ref{fig:coverChange}).
+spicata} fell significantly in the drought plots from 2011 to 2016 (tables \ref{table:changeHECO}, \ref{table:changePSSP}, fig \ref{fig:coverChange}).
 The cover of \textit{P.
@@
 For \textit{A.
-tripartita} the model showed a significant size by treatment effect in the drought plots: the effect of drought was positive for plants of average size and smaller \ref{fig:growthTreat}, but plants larger than the mean by more than 1.5 standard deviations grew slightly less in the drought treatment than in the controls.
+tripartita} we found significant size by treatment effects of drought: drought had positive effects on plants of average size and smaller \ref{fig:growthTreat}, but plants larger than the mean by more than 1.5 standard deviations grew slightly less in the drought treatment than in the controls.
 \textit{A.
-tripartita} showed the opposite response in the irrigation plots, (although the irrigation parameters were not technically significant at the 95\% confidence level), irrigation reduced growth the most for small plants while irrigation had positive effects on plants more than 1.5 standard deviations greater than the mean size.
-\textit{H.
-comata} growth showed a strong (but not significant) decrease in the drought plots and no response to the irrigation plots.
+tripartita} showed the opposite response in the irrigation plots, (although the irrigation parameters were not technically significant at the 95\% confidence level): irrigation reduced growth the most for small plants while irrigation had positive effects on plants more than 1.5 standard deviations greater than the mean size.
+Drought led to a strong (but not signficant) decrease in \textit{H.
+comata} growth, while irrigation had no effect on growth.
 Like \textit{A.
-tripartita}, \textit{P.secunda} growth also showed a size by treatment effect in the drought plots, with the negative effects on growth becoming greater for larger plants.
+tripartita}, we saw size by treatment effects on \textit{P.secunda} growth, with the negative effects of drought becoming greater for larger plants.
 \textit{P.
@@
 
-The effects of drought were negative and signficant on the survival of all three grass species (fig \ref{fig:)
-
-We observed a number of strong correlations between the year effects in the vital rate models and the seasonal soil moisture variables and table x reports the strongest three correlations for each species and vital rate (table).
-In general the soil moisture during the spring and summer period appeared to be more strongly correlated with the year effects for each species than the fall soil moisture, but see \textit{P.
-spicata} vr.
-All three time lags show up in the choosen variables, but 1 out of X variables choosen involved soil moisture in the year immediately before the first year of the transition, (subscript "0"), while variables involving soil moisture in the year before that (subscript "lag") and after that (subscript "1") were choosen less often.
-
-We observed vital rate and species-specific effects of soil moisture on each of these species after fitting the vital rate models with the soil moisture covariates discussed above (fig).
-DISSCUSS significant soil moisture effects: for \textit{A.
-tripartita} growth was significantly affected by ...
-For \textit{H.
-comata}.
-For \textit{P.
-secunda}.
-And for \textit{P.
-spicata}.
-Soil moisture by size effects on \textit{A.
+Drought decreased the survival of all three grass species (fig \ref{fig:survivalTreat}).
+And \textit{P.
+secunda} showed a negative size by drought interaction effect, indicating that the survival of larger plants was more negatively affected by drought than that of the smaller plants, with the smallest plants (plants one standard deviation smaller than the mean) actually seeing a slightly positive effect of drought.
+\textit{A.
+tripartita} survival was relatively unaffected by the drought and irrigation treatments.
+
+Recruitment in our irrigation plots was significantly less than in control plots for two grass species \textit{P.
+secunda} and \textit{P.
+spicata} (fig \ref{fig:recruitmentTreat}).
+However, recruitment was also lower in the drought plots than in the the control plots (although not significantly so), indicating that the decrease in the irrigated plots may have not been entirely due to the irrigation itself.
+The recruitment data for \textit{A.
+tripartita} were relatively limited, with only 32 new recruits in total observed in all 30 plots over the course of the five year experiment.
+
+\subsection*{Soil moisture effects}
+
+We choose three seasonal soil moisture variables for each species' climate model based on their correlation with the year effects in the random year effects model fit to the observational data (table \ref{table:strongCor}).
+We included size by soil moisture variables for \textit{A.
 tripartita} and  \textit{P.
-secunda} showed that ()
+secunda} based on the treatment response we observed in the experiment.
+All three time lags and all three seasons show up in the choosen variables.
+After fitting the vital rate models with the selected soil moisture variable we observed a trend towards positive soil moisture effects on growth of all three grasses \ref{fig:climateGrowth}.
+For \textit{H.comata} the soil moisture of the most recent summer had a significantly positive effect while the soil moisture of the previous summer and the fall before that were also positive but not significant.
+For \textit{A.
+tripartita} the summer and fall soil moisture of the previous year had strong negative effects on growth.
+There were also strong positive size by climate interaction effects for these variables: soil moisture had a stronger negative effect on small plants and a positive effect only on the largest plants.
+
+Soil moisture had significant effects on the survival of all four species \ref{fig:climateSurvival}.
+As for growth the grasses showed mainly positive effects while \textit{A.
+tripartita} showed a significant negative effect of soil moisture in the summer of the previous year and a strong negative effect of spring soil moisture of the previous year.
+\textit{H.
+comata} showed a significant positive effect of spring lag soil moisture and a strong positive effect of spring soil moisture in the previous year.
+\textit{P.
+secunda} showed a significant positive effect of the previous spring's soil moisture and there was a significant interaction between this effect and plant size: as plant size increased this effect became more strongly positive.
+Finally for \textit{P.
+spicata} there was a significant positive effect of lag spring soil moisture on survival.
+
+There were only two significant effects of soil moisture on recruitment: lag fall soil moisture had a positive effect on \textit{P.
+secunda}, and lag summer soil moisture had a negative effect on \textit{P.
+spicata} recruitment.
+Soil moisture of the previous year's summer had also had a strong negative effect on \textit{P.
+spicata} recruitment.
 
@@
 
-\section{Tables}
+\clearpage
+\newpage
+
+\section*{Tables}
 
@@
 
-\section{Figures}
+% latex table generated in R 3.3.1 by xtable 1.8-2 package
+% Mon Dec 19 20:47:29 2016
+\begin{table}[ht]
+	\centering
+	\begin{tabular}{lllrrrr}
+		\hline
+		vital\_rate & species & climate variable & Int. cor. & p val. & Size cor. & Size p. val. \\ 
+		\hline
+		growth & ARTR & VWC.su.0 & -0.49 & 0.02 & 0.26 & 0.26 \\ 
+		growth & ARTR & VWC.f.0 & -0.28 & 0.23 & 0.40 & 0.08 \\ 
+		growth & ARTR & VWC.sp.1 & 0.17 & 0.45 & -0.33 & 0.14 \\ 
+		growth & HECO & VWC.su.1 & 0.69 & 0.00 &  &  \\ 
+		growth & HECO & VWC.su.0 & 0.50 & 0.02 &  &  \\ 
+		growth & HECO & VWC.f.lag & 0.37 & 0.10 &  &  \\ 
+		growth & POSE & VWC.f.lag & 0.31 & 0.17 & -0.11 & 0.64 \\ 
+		growth & POSE & VWC.su.lag & 0.29 & 0.20 & -0.20 & 0.38 \\ 
+		growth & POSE & VWC.sp.1 & 0.26 & 0.25 & -0.20 & 0.38 \\ 
+		growth & PSSP & VWC.f.lag & 0.34 & 0.13 &  &  \\ 
+		growth & PSSP & VWC.su.lag & 0.25 & 0.27 &  &  \\ 
+		growth & PSSP & VWC.f.0 & -0.22 & 0.34 &  &  \\ 
+		recruitment & ARTR & VWC.su.lag & -0.32 & 0.16 &  &  \\ 
+		recruitment & ARTR & VWC.su.0 & -0.26 & 0.25 &  &  \\ 
+		recruitment & ARTR & VWC.sp.1 & 0.22 & 0.34 &  &  \\ 
+		recruitment & HECO & VWC.su.lag & -0.31 & 0.18 &  &  \\ 
+		recruitment & HECO & VWC.su.0 & -0.30 & 0.18 &  &  \\ 
+		recruitment & HECO & VWC.f.lag & 0.19 & 0.40 &  &  \\ 
+		recruitment & POSE & VWC.sp.1 & 0.49 & 0.02 &  &  \\ 
+		recruitment & POSE & VWC.f.lag & 0.34 & 0.13 &  &  \\ 
+		recruitment & POSE & VWC.f.1 & 0.32 & 0.16 &  &  \\ 
+		recruitment & PSSP & VWC.su.lag & -0.52 & 0.02 &  &  \\ 
+		recruitment & PSSP & VWC.su.0 & -0.48 & 0.03 &  &  \\ 
+		recruitment & PSSP & VWC.sp.0 & 0.30 & 0.19 &  &  \\ 
+		survival & ARTR & VWC.su.0 & -0.60 & 0.00 &  &  \\ 
+		survival & ARTR & VWC.sp.0 & -0.41 & 0.06 &  &  \\ 
+		survival & ARTR & VWC.su.1 & -0.40 & 0.07 &  &  \\ 
+		survival & HECO & VWC.sp.0 & 0.44 & 0.04 &  &  \\ 
+		survival & HECO & VWC.sp.lag & 0.43 & 0.05 &  &  \\ 
+		survival & HECO & VWC.f.1 & 0.33 & 0.14 &  &  \\ 
+		survival & POSE & VWC.sp.0 & 0.44 & 0.04 & 0.22 & 0.34 \\ 
+		survival & POSE & VWC.sp.1 & 0.27 & 0.23 & -0.46 & 0.04 \\ 
+		survival & POSE & VWC.f.lag & -0.00 & 0.99 & 0.30 & 0.19 \\ 
+		survival & PSSP & VWC.sp.0 & 0.36 & 0.11 &  &  \\ 
+		survival & PSSP & VWC.sp.lag & 0.34 & 0.13 &  &  \\ 
+		survival & PSSP & VWC.su.1 & 0.26 & 0.26 &  &  \\ 
+		\hline
+	\end{tabular}
+	\caption{Correlations between intercept of year effects and soil moisture variables.
+P-values of correlations are given next to correlations.
+For ARTR growth and POSE growth and survival, the correlations between the year effects on size and the soil moisture variables are also given.}
+	\label{table:strongCor}
+\end{table}
+
+\clearpage
+\newpage
+
+\section*{Figures}
 
@@
 	\centering
-	\includegraphics[width=1\textwidth]{treatment_trends_cover}
-	\caption{Average cover of each of the focal species in the control, drought, and irrigation treatments during the experiment.
-Percent cover of \textit{A.
-tripartita} refers to canopy cover, while the cover of the three grasses refers to basal cover.
-}
-	\label{fig:coverTrends}
-\end{figure}
-
-\begin{figure}[tbp]
-	\centering
 	\includegraphics[width=1\textwidth]{start_to_finish_cover_change}
 	\caption{Log change in cover in each of the experimental plots from the pre-treatment monitoring in 2011 to the last year of the experiment in 2016.
-Box plots show the median cover change and the interquartile range.}
+Box plots show the median cover change and the interquartile range.
+ARTR = \textit{A.
+tripartita}, HECO = \textit{H.
+comata}, POSE = \textit{P.
+secunda}, PSSP = \textit{P.
+spicata}.}
 	\label{fig:coverChange}
@@
 \end{figure}
+
+\begin{figure}[tbp]
+	\centering
+	\includegraphics[width=1\textwidth]{treatment_trends_cover}
+	\caption{Average cover of each of the focal species in the control, drought, and irrigation treatments during the experiment.
+Percent cover of \textit{A.
+tripartita} refers to canopy cover, while the cover of the three grasses refers to basal cover.
+}
+	\label{fig:coverTrends}
+\end{figure}
+
+\begin{figure}[tbp]
+	\centering
+	\includegraphics[width=1\textwidth]{climate_effect_growth}
+	\caption{Parameter estimates for the selected seasonal soil moisture covariates on the growth of all four species.
+Parameters are ordered chronologically from most recent to the current growing season on the right to most distant on the left.
+Dark blue parameters show size x climate interaction effects.
+We assessed a parameter as significant when the 95\% Bayesian credible intervals did not overlap zero.
+ARTR = \textit{A.
+tripartita}, HECO = \textit{H.
+comata}, POSE = \textit{P.
+secunda}, PSSP = \textit{P.
+spicata}.
+}
+	\label{fig:climateGrowth}
+\end{figure}
+
+\begin{figure}[tbp]
+	\centering
+	\includegraphics[width=1\textwidth]{climate_effect_survival}
+	\caption{Parameter estimates for the selected seasonal soil moisture covariates on the survival of all four species.
+Parameters are ordered chronologically from most recent to the current growing season on the right to most distant on the left.
+Dark blue parameters show size x climate interaction effects.
+We assessed a parameter as significant when the 95\% Bayesian credible intervals did not overlap zero.
+ARTR = \textit{A.
+tripartita}, HECO = \textit{H.
+comata}, POSE = \textit{P.
+secunda}, PSSP = \textit{P.
+spicata}.
+}
+	\label{fig:climateSurvival}
+\end{figure}
+
+\begin{figure}[tbp]
+	\centering
+	\includegraphics[width=1\textwidth]{climate_effect_recruitment}
+	\caption{Parameter estimates for the selected seasonal soil moisture covariates on the survival of all four species.
+Parameters are ordered chronologically from most recent to the current growing season on the right to most distant on the left.
+Dark blue parameters show size x climate interaction effects.
+We assessed a parameter as significant when the 95\% Bayesian credible intervals did not overlap zero.
+ARTR = \textit{A.
+tripartita}, HECO = \textit{H.
+comata}, POSE = \textit{P.
+secunda}, PSSP = \textit{P.
+spicata}.
+}
+	\label{fig:climateRecruitment}
+\end{figure}
+
+\clearpage 
+\newpage
 

### 2016-12-20 precip/analysis/figure_scripts/plot_treatment_parameter_estimates.R

@@ -38,12 +38,20 @@ effect_plot <-
   geom_hline( aes( yintercept = 0 ), linetype = 2 , alpha = 0.5) + 
   facet_grid( type  ~  species ) + 
   ylab ( 'mean effect (+/- 95% Bayesian Credible Interval)') + 
+  xlab( '') + 
+  labs(color = '') + 
   scale_color_manual(values = my_colors[3:4]) + 
   my_theme
 
 gg <- treat %>% group_by(vital_rate) %>% do(gg =  effect_plot %+% .  + ggtitle( paste0('Effects on ', .$vital_rate)))
 
-pdf( 'figures/treatment_effect_estimates.pdf', height = 5, width = 8 ) 
-print( gg$gg ) 
+png( 'figures/treatment_effect_growth.png', height = 5, width = 7, res = 300, units = 'in' ) 
+print( gg$gg[[1]] ) 
+dev.off() 
+png( 'figures/treatment_effect_recruitment.png', height = 4, width = 7, res = 300, units = 'in' ) 
+print( gg$gg[[2]] ) 
+dev.off() 
+png( 'figures/treatment_effect_survival.png', height = 5, width = 7, res = 300, units = 'in' ) 
+print( gg$gg[[3]] ) 
 dev.off() 
 

### 2016-12-21 precip/manuscript/precip_ms.tex
adding a bunch of manuscript revisions and revisions to the plotting scripts

@@
 \usepackage{gensymb}
+
+\usepackage{longtable}
+\setlength{\LTcapwidth}{\linewidth}
 
@@
 We tested whether statistical models based on these observational data could be used to predict how these species respond to a five year drought and irrigation experiment conducted from 2011 to 2016.
-\\item We established sixteen new plots at the same field cite as the original historical data and assigned each to either drought or irrigation treatments imposed using rainfall shelters and automatic sprinklers.
+\item We established sixteen new plots at the same field cite as the original historical data and assigned each to either drought or irrigation treatments imposed using rainfall shelters and automatic sprinklers.
 The original plots were also monitored and served as ambient climate controls.
@@
 \item The experimental drought and irrigation treatments successfully lowered and increased soil moisture respectively.
-Over the course of the experiment, average plot cover of \textit{Hesperostipa comata} and \textit{Pseudoroegneria spicata} declined, while cover of \textit{Poa secunda} showed a trend towards increase in the irrigated plots.
-At the level of individual vital rates, experimental drought reduced growth of \textit{Hesperostipa comata} and growth and survival of \textit{Poa secunda}.
-Drought increased, while irrigation decreased, growth of the shrub \textit{Artemisia tripartita}.
-\item Vital rate models fit to the observational data correctly predicted significant increases in X and X and significant decreaes in X and X.
-However, the historical models predicted significant changes in X X and X that were not observed in the experiment.
-Moreover, the experiment led to a significant change in X that was not predicted by the historical data.
-\item Our individual based population model more accurately predicted changes in the population growth rates of \textit{} in the experimental plots than a model without climate effects.
-However for the other X species predictions made from the climate model were less accurate than models without climate.
-\item \emph{Synthesis}: Out of all four species examined, climate covariates only appeared to increase our predictive accuracy for one species: \textit{Hesperostipa comata}.
-In general the relatively weak responses to the experimental treatments agrees with the low explanatory power of climate covariates in explaining the vital rates and population dynamics of these species in this system.
-The highly variable environment and water limited conditions may select for species with a conservative non-responsive strategy to variation in climate both natural and experimentally imposed.
+Over the course of the experiment, average plot cover of \textit{Hesperostipa comata} and \textit{Pseudoroegneria spicata} declined significantly, while cover of \textit{Poa secunda} showed a trend towards increase in the irrigated plots.
+At the level of individual vital rates, experimental drought reduced growth and survival of \textit{Hesperostipa comata} and \textit{Poa secunda} and the survival of \textit{P.
+spicata}, while drought increased the growth of small \textit{A.
+tripartita}.
+\item Soil moisture parameters improved model predictions for six out of the twelve models we examined.
+Moreover, we observed strong positive correlations between the predicted treatment effects and the treatment effects we observed.
+\item At the population-level, including the effects of soil moisture improved the cover predictions made by individual based models for \textit{P. secunda} and \textit{P. spicata}.   
+\item \emph{Synthesis}: Observational climate data did hold valuable information for predicting species' responses to this climate change experiment.
+We were encouraged that treatment responses were often in the right direction to predicted responses even when the effect were not signficant.
+We generally were better able to predict species responses to the drought treatment than to the control or the irrigation treatment.
+This suggests that soil moisture is an important factor in predicting the population dynamics of these species but only when water is truly limiting.
 
@@
 \begin{equation}
-\boldsymbol{T} \chi_{j}^G 
+\boldsymbol{T}\chi_{j}^G 
 \label{eqn:growT}
@@
 \begin{equation}
- \boldsymbol{T}\boldsymbol{chi}_{j}^R
-\end{equation}
-where $chi$ describes the effect of the treatment levels on recruitment.
+ \boldsymbol{T}\boldsymbol{\chi}_{j}^R
+\end{equation}
+where $\chi$ describes the effect of the treatment levels on recruitment.
 
@@
 For \textit{A.
-tripartita} we found significant size by treatment effects of drought: drought had positive effects on plants of average size and smaller \ref{fig:growthTreat}, but plants larger than the mean by more than 1.5 standard deviations grew slightly less in the drought treatment than in the controls.
+tripartita} we found significant size by treatment effects of drought: drought had positive effects on plants of average size and smaller \ref{fig:growthTreat}, but plants larger than the mean size by more than 1.5 standard deviations grew slightly less in the drought treatment than in the controls.
 \textit{A.
-tripartita} showed the opposite response in the irrigation plots, (although the irrigation parameters were not technically significant at the 95\% confidence level): irrigation reduced growth the most for small plants while irrigation had positive effects on plants more than 1.5 standard deviations greater than the mean size.
+tripartita} showed the opposite response in the irrigated plots, (although the irrigation parameters were not technically significant at the 95\% confidence level): irrigation reduced growth for small plants while irrigation increased growth of plants more than 1.5 standard deviations greater than the mean size.
 Drought led to a strong (but not signficant) decrease in \textit{H.
@@
 
-\subsection*{Soil moisture effects}
+As expected from previous research most of our demographic models estimated strong negative intra-specific crowding effects and weaker negative inter-specific crowding effects on the focal species (appendix) \citep{adler_coexistence_2010,chu_direct_2016,chu_large_2015}.
+
+\subsection*{Effects of soil moisture on vital rates}
 
@@
 
+The intra- and interspecific crowding effects estimated in the climate model were similar to those estimated in the treatment model (appendix) \citep{adler_coexistence_2010,chu_direct_2016,chu_large_2015}.
+
 \subsection*{Evaluating the predictions}
 
-For each vital rate and species, predictions from the year effects and climate models were nearly indistinguishable (fig).
-This indicated that the additional soil moisture covariates in the climate models had very little effect on the predictions.
-Not surprisingly \citep{chu_direct_2016,treddenick_} our models show that plant size and intra-specific crowding are far more influential in determining growth, survival and recruitment than climate effects.
-In all cases, we found MSE and lppd scores for the climate models were slightly worse than the MSE and the lppd of the year effects model (table).
-This indicated that adding the soil moisture parameters did not improve model predictions over all (table).
-When we break the data up into treatment groups we do see that the climate model outperformed the year effects model for predicting X species VR in treatment T.
-
-While our climate models failed to improve predictions, we did find that they predicted treatment responses that were largely in the same direction as the observed treatment responses (fig).
-The overall correlation between the predicted treatment parameters and the observed treatment parameters was X with stronger correlation for the climate effects on the size by treatment effects fit for \textit{P.
-secunda} growth and survival and \textit{A.
-tripartita} growth (fig) than for the treatment effects on the intercept.
-A table of MSE between predicted and expected treatment parameters shows that averaged over all species and vital rates, we did the best job of predicting the response to the drought treatment.
-Averaged across all treatements and vital rates, we did the best job predicting the response of \textit{species} to the treatments.
-Finally if we average across all species and treatments, we did the best job of predicting VR parameters in response to the treatment.
-
-As for the vital rate predictions, our one-step ahead cover predictions from the year effects models and the climate models were fairly similar (fig).
-Models with and without soil moisture parameters tended to underpredict the cover of \textit{A.
-tripartita} and \textit{P.secunda} during the experiment (fig).
-When we compare the MSE of the predicted population growth rates for each year and quadrat from the model with and without the soil moisture parameters, we find that adding soil moisture to the vital rate models always made the one step ahead cover predictions slightly less accurate (table).
-The pattern when we consider the correlation between the predicted and observed population growth rates was slightly different (table?).
-When we considered the population growth rate data separately by treatment, we found that adding soil moisture to the models mostly did not improve predictions of cover change in the control and irrigation plots.
-However, we did find that adding soil moisture parameters produced better predictions of population growth rates for three of the species (S , S, and S) in the drought plots (fig).
-
-As expected from previous research most of our demographic models estimated strong negative intra-specific crowding effects and weaker negative inter-specific crowding effects on the focal species (appendix) \citep{adler_coexistence_2010,chu_direct_2016,chu_large_2015}.
+For most models adding climate covariates did not improve our ability to predict species responses in the experiment \ref{table:overallPreds}.
+However, the climate models did improve overall prediction MSE for growth of \textit{A.
+tripartita} and growth and survival of \textit{P.
+secunda} (table \ref{table:overallPreds}).
+In terms of lppd, the climate model outperformed the year effects model for \textit{A.
+tripartita} growth, \textit{H.
+comata} recruitment, \textit{P.
+secunda} growth and survival and \textit{P.
+spicata} recruitment.
+.
+
+When we look at the predictions for each treatment separately we see that climate covariates improved model predictions more often in the drought treatments than in the control or irrigation treatments \ref{table:treatmentPreds}.
+For all four species, the climate model outperformed the year effects model for predicting the response of growth to drought in terms of lppd \ref{table:treatmentPreds}.
+The climate model outperformed the year effects model for predicting irrigation effects on growth for all species except \textit{H.
+comata}.
+
+Overall our climate models often predicted the correct direction of the drought and irrigation treatments \ref{fig:parPredictions}.
+In four cases we both observed and predicted treatment effects significantly different from zero based on the 95\% Bayesian credible interval around the parameters: the drought response of \textit{H.
+comata} survival (fig \ref{fig:parPredHECOSurvival}), the drought response of \textit{P.
+secunda} growth (fig \ref{fig:parPredPOSEGrowth}), the irrigation response of \textit{P spicata} recruitment (fig \ref{fig:parPredPSSPRecruitment}) and the irrigation response of \textit{P.
+secunda} recruitment (fig \ref{fig:parPredPOSERecruitment}).
+In only one of these cases, for \textit{P.
+secunda} recruitment, was the predicted effect in the opposite direction from the observed treatment effect \ref{fig:parPredictions}.
+The overall correlation between the predicted and observed treatment effects for all treatments, species and vital rates was r = 0.54, whereas the correlation for the drought treatment effects (r = 0.77) was better than for the irrigation effects (r = 0.46).
+Also for the three models in which we included size by treatment or size by climate interactions the correlation of these size dependent effects were much stronger than the intercept parameter estimates \ref{fig:parPredictions}.
+
+Using the vital rate models for each species we generated one year ahead cover predictions for each quadrat in each year of the experiment.
+Average cover predicted by the climate model tended to be lower than the observed cover each year for \textit{A.
+tripartita} and \textit{P.secunda} (fig \ref{fig:coverPred}).
+Comparing the overall population growth rates predicted to those observed in the experiment, we see that the MSE of the climate model was lower than the MSE of the year effects model for \textit{P.
+secunda} and \textit{P.
+spicata} (table \ref{table:corPGR}).
+The predictions produced by the climate model for these species were also slightly more correlated with the observations than the predictions produced by the year effects model (table \ref{table:corPGR}).
+Considering each treatment and species separately we see that the predicted population growth rates for \textit{A.
+tripartita}, \textit{P.
+secunda} and \textit{P.
+spicata} were all consistently lower than the observed population growth rates (figs \ref{fig:pgrARTR}, \ref{fig:pgrPOSE}, \ref{fig:pgrPSSP}).
+The climate model showed lower MSE for \textit{A.
+tripartita}, \textit{P.
+secunda} and \textit{P.
+spicata} in the irrigation treatment, \textit{P.
+spicata} in the control treatment and \textit{H.
+comata} in the drought treatment (figs \ref{fig:pgrHECO}).
+However, the correlations between the predicted and observed log changes in cover did not always show the same pattern as MSE: the climate model made more strongly correlated predictions with the observations than the year effects model only for \textit{P.
+spicata} and \textit{P.
+secunda} in the control treatment and \textit{P.
+secunda} and \textit{H.
+comata} in the drought treatment.
 
@@
 
-Our experiment showed that including climate covariates in our population models did little to improve population level predictions across the course of the experiment.
-Averaged across the control, drought and irrigation treatments, the climate models did not perform as well as null models in predicting response the experiment at the vital rate or at the level of population growth rates.
-However, there were some glimmers of hope.
-First at the population level we generally found that including climate covariates did not improve predictions in the control and irrigation plots, but they did appear to improve predictions in the drought plots ().
-This suggests that models fit to observational data do contain useful information about the effects of climate on individual performance.
-The fact that these predictions showed up in the more extreme conditions produced by the drought plots could be a sign that in ambient conditions the effect of climate is weak enough that it is washed out at the population level by finer scale sources of variation.
-This pattern was observed for species x, x and x.
-
- the climate signal  improve population growth predictions in the drought treatment but not experiment showed some relatively strong reson based on historical, observational data reveal weak interspecific interactions \citep{adler_coexistence_2010,chu_large_2015} and, as a result, predict very little competitive release among the four dominant species in a sagebrush steppe.
-Despite the fact that previous removal studies have reported variable results, ranging from negative effects of removal on remaining species to complete competitive release, we were suspicious of our model's predictions and conducted a removal experiment to test them.
-The results provide substantial support for our model's predictions.
-
-For three of the four species, we find little or no evidence that models based on observational data underestimate the intensity of interspecific competition.
-Removal treatment effects were never significant for \textit{A.
-tripartita} or \textit{H.
-comata}; our baseline model, which accounts for local neighborhood composition, adequately explained vital rate responses to  removal.
-\textit{Poa secunda} showed mixed responses to \textit{Artemisia} removal: removal effects were not significant at the survival stage, were positive at the growth stage, and negative at the recruitment stage (where the estimated effect of \textit{Artemisia} on \textit{Poa} is positive).
-At the population level, models with and without removal effects projected similar patterns of short (Fig.
-\ref{fig:IBM1step}) and long-term (Fig.
-\ref{fig:IPMresults}) dynamics for \textit{H.
-comata} and \textit{A.
+Our experiment showed that observational data on the response of plant populations to interannual climate variation can indeed help us predict the direction of species response to experimental climate manipulations \ref{fig:parPredictions}.
+This was true even though adding climate parameters to the demographic models only improved vital rate predictions for half of the models \ref{table:overallPreds}.
+This should give us some hope that even when climate effects in demographic models to observational data are weak, they may contain useful qualitative information on the direction of climate effects in the future.
+
+* discuss biggest successes--astreces in fig \ref{fig:parPredictions}
+* discuss biggest mistake, P. secunda recruitment went in wrong direction.
+
+Scaling up to the population level, the climate models only produced better one step a head predictions of the overall response of the species to the experiment for two species: \textit{P.
+spicata} and \textit{P.
+secunda} (table \ref{table:corPGR}).
+However adding climate effects produced better population-level predictions in the drought treatments for \textit{P.
+secunda} and \textit{H.
+comata}.
+Both the survival and the growth of these species were positively affected by increased soil moisture (fig \ref{fig:climateGrowth}, \ref{fig:climateSurvival}), and so it makes intuitive sense that we would observe declines in these species cover in the drought treatment (fig \ref{fig:coverChange}).
+
+*Among our species, we had the most success predicting the response of the three grass species and less success predicting the response of the shrub species \textit{A.
 tripartita}.
-For \textit{Poa secunda}, the model with removal effects predicted higher population growth over the short but not the long-term, perhaps because the short-term projections mostly reflect the positive effect of \textit{Artemisia} removal on growth, while the long-term equilibrium are also sensitive to the negative effects of removal on recruitment.
-Although the baseline model did not always successfully predict removal effects on vital rates, it does appear to successfully capture the population-level consequences of removals on these three species and did not severely underestimate the magnitude of competitive release.
-
-The story is different for the fourth species, \textit{P.
-spicata}.
-Both individual growth and short- and long-term population growth were higher in the sagebrush removal treatments than in the control plots, suggesting that our analyses of observational data may underestimate the competitive release that \textit{P.
-spicata} experiences following \textit{Artemisia} removal.
-However, the mechanism by which \textit{Artemisia} removal promotes \textit{P.
-spicata} remains unclear.
-If  \textit{Artemisia} removal is releasing \textit{P.
-spicata} from competition, we should expect \textit{P.
-spicata} individuals that were located beneath \textit{Artemisia} shrubs that we removed to respond more than individuals located away from any shrubs.
-Yet we found no evidence for such a difference.
-An alternative explanation involves pre-treatment differences in composition between control plots and  the \textit{Artemisia} removal  plots, which had higher cover of \textit{Artemisia} (see 2011 means in Fig.
-\ref{fig:CoverTrends}) and higher cover of subdominant grasses and shrubs (Fig.
-\ref{fig:W-by-treatment}).
-Perhaps these pre-treatment differences reflect subtle differences in edaphic and environmental factors which could promote \textit{P.
-spicata} growth.
-While \textit{P.
-spicata} appears to show stronger competitive release than our baseline model predicted, the evidence is not conclusive.
-
-The small competitive release of many species in our system implies that each of the four dominant species we studied are limited by quite different factors.
-This low niche overlap could reflect strong resource partitioning (Tilman 1982), perhaps combined with variation in the times and locations at which species are taking up those resources.
-Species-specific responses to spatial and temporal variation in non-resource environmental variables can also weaken interspecific interactions \citep{chesson_mechanisms_2000}, though in a previous study we failed to find evidence of a temporal storage effect in this community \citep{adler_weak_2009}.
-Interactions with above- or belowground enemies can regulate populations and, given enough species-specificity in the effects of these enemies, decouple interspecific dynamics \citep{hersh_evaluating_2011,janzen_herbivores_1970,connell_role_1971}.
-More likely, plants may alter the environment, and even resource availability, in ways that facilitate their neighbors.
-For example, shade cast by \textit{Artemisia} canopies may reduce evapotranspiration (e.g.
-\citealt{Barbier2008}).
-\textit{Artemisia} removal might offest this facilitative effect by reducing resource uptake, resulting in little net change in water balance following removal.
-Unfortunately, we did not monitor soil moisture at the fine spatial and temporal resolution necessary to test this hypothesis.
-
-Evidence for weak interspecific interactions and limited competitive release carries interesting implications.
-If most interspecific interactions are weak, then single-species models may be adequate for forecasting population responses to environmental change \citep{adler_forecasting_2012,kleinhesselink_indirect_2015}.
-Similarly, differences among species in relative abundance should reflect variation in the strength of intraspecific density dependence more than variation in the strength of interspecific interactions \citep{yenni_strong_2012}.
-In such a system, effective management and conservation requires understanding the sources and species-specificity of intraspecific density dependence.
-In fact, an experiment to test our model's skill at predicting release from intraspecific competition could be more important than the test of interspecific competitive release described here.
-
-Our results have given us more confidence that competitive release is quite small for the dominant species in our study system.
-But for many species in other communities, interspecific competition may still be an important driver.
-In closed-canopy forests where competition for light is intense, we might have found different results than in our system, in which competition for a number of belowground resources is more important than competition for light.
-However, a number of recent studies from forests have also found that heterospecific competitive effects are dramatically weaker than conspecific effects \citep{comita_asymmetric_2010,kunstler_plant_2015,johnson_conspecific_2012}.
-We also expect stronger competitive release in successional systems, or communities in the early stages of community assembly where the process of competitive exclusion is still underway \citep{kokkoris_patterns_1999}.
-Finally, we might have found stronger competitive release if we could have studied rare species in our system, as long as those species are rare because of niche overlap with stronger competitors.
-If they are rare because they occupy unique, but narrow, niches, interspecific interactions would still be weak.
-Determining why interspecific interactions and competitive release are strong in some ecological contexts but weak in others remains a fascinating research problem.
+Why might ARTR be different?
+
+*We also had more success predicting species response to the drought treatment than to the irrigation treatment.
+We also tended to see species respond more strongly to the drought treatment than they did to the irrigation treatment.
+This tells us that in some cases, demographic models based on observational data may be of more use for prediction in extreme conditions than when conditions are close to average.
+Our drought treatment likely created extremely low soil moisture compared to the historical average \ref{fig:seasonalVWC}.
+We hypothesize that this made water the most limiting resource for plants in these plots.
+In contrast, in the control and irrigated plots soil moisture may not have always been limiting during the course of the experiment.
+It may make sense then that our climate models did not make more accurate predictions in these conditions than a random year effects model.
+
+*Mixed success at predicting species responses to high and low moisture availability has interesting implications.
+On the one hand it is re-assuring that observational data is sometimes useful.
+On the other hand, for predictions to be truly useful we would also like more information to help us sort out why our predictions for some species were good and for others no better than a null model.
+In other words, we have little ability to predict when our predictions are likely to be accurate.
+Among plants detailed physiological ecology may give us a guide to which factors and climate conditions are likely to affect which species.
+In our system, for instance, it may make sense that the grasses showed a stronger response to the drought treatment than the a woody shrub.
+Although these grasses are adapted to the arid conditions that characterize the sagebrush steppe, they thrive during the brief window in spring and early summer when the soil moisture and temperatures are warmer.
+\textit{A.
+tripartita} on the other hand grows throughout the summer and generally has deeper roots than the grasses.
+These traits may help it tolerate the water stress induced by drought.
+
+*Will our predictions be useful for the longer term?
+Because these species compete, one could argue that our predictions for any one species in this community will only be as good as the predictions we make for their competitors.
+For instance, while we observed little effect of drought on \textit{A.
+tripartita} in our experiment, it is possible that it will eventually respond positively to the drought treatment as cover of the grass species it comptetes with declines (fig \ref{fig:coverChange}).
+We know that grass species compete strongly with \textit{A.
+tripartita}.
+However, we also know from previous work in this community that each species is more limited by intra-specific competition than by inter-specific crowding.
+This fact ensures that the direct effects of climate change will generally be greater than the competition mediated indirect effects of climate change \cite{chu_direct_2016, kleinhesselink_indirect_2015}.
+
+*Our results give us more confidence that historical observational data can in theory be used to detect and predict the demographic effects of climate change.
+This should encourage more researchers to try and use observational data to predict the future in both experimental and natural settings.
+Nevertheless, our success at predicting the short-term response of two species to a small-scale climate manipulation is not likely to be very reassuring to applied ecologists and resource managers wishing to make accurate medium to long-term quantitative predictions about the effects of climate change on the species they manage.
+Clearly more work is needed to distinguish which predictions are worth having confidence in and which predictions we should have less confidence in.
+Towards that goal, perhaps the best way to build confidence in ecological predictions is conduct more tests like this one.
 
@@
 	\end{tabular}
-	\caption{Treatment effects on log cover change for 	extit{A. tripartita} from 2011 to 2016. Intercept gives control effects.} 
+	\caption{Treatment effects on log cover change for \textit{A. tripartita} from 2011 to 2016. Intercept gives control effects.} 
 	\label{table:changeARTR}
@@
 	\end{tabular}
-	\caption{Treatment effects on log cover change for 	extit{H. comata} from 2011 to 2016. Intercept gives control effects.} 
+	\caption{Treatment effects on log cover change for \textit{H. comata} from 2011 to 2016. Intercept gives control effects.} 
 	\label{table:changeHECO}
@@
 	\end{tabular}
-	\caption{Treatment effects on log cover change for 	extit{P. secunda} from 2011 to 2016. Intercept gives control effects.} 
+	\caption{Treatment effects on log cover change for \textit{P. secunda} from 2011 to 2016. Intercept gives control effects.} 
 	\label{table:changePOSE}
@@
 	\end{tabular}
-	\caption{Treatment effects on log cover change for 	extit{P. spicata} from 2011 to 2016. Intercept gives control effects.} 
+	\caption{Treatment effects on log cover change for \textit{P. spicata} from 2011 to 2016. Intercept gives control effects.} 
 	\label{table:changePSSP}
@@
 % latex table generated in R 3.3.1 by xtable 1.8-2 package
-% Mon Dec 19 20:47:29 2016
+% Tue Dec 20 12:39:32 2016
 \begin{table}[ht]
@@
 		\hline
-		growth & ARTR & VWC.su.0 & -0.49 & 0.02 & 0.26 & 0.26 \\ 
-		growth & ARTR & VWC.f.0 & -0.28 & 0.23 & 0.40 & 0.08 \\ 
-		growth & ARTR & VWC.sp.1 & 0.17 & 0.45 & -0.33 & 0.14 \\ 
-		growth & HECO & VWC.su.1 & 0.69 & 0.00 &  &  \\ 
-		growth & HECO & VWC.su.0 & 0.50 & 0.02 &  &  \\ 
-		growth & HECO & VWC.f.lag & 0.37 & 0.10 &  &  \\ 
-		growth & POSE & VWC.f.lag & 0.31 & 0.17 & -0.11 & 0.64 \\ 
-		growth & POSE & VWC.su.lag & 0.29 & 0.20 & -0.20 & 0.38 \\ 
-		growth & POSE & VWC.sp.1 & 0.26 & 0.25 & -0.20 & 0.38 \\ 
-		growth & PSSP & VWC.f.lag & 0.34 & 0.13 &  &  \\ 
-		growth & PSSP & VWC.su.lag & 0.25 & 0.27 &  &  \\ 
-		growth & PSSP & VWC.f.0 & -0.22 & 0.34 &  &  \\ 
-		recruitment & ARTR & VWC.su.lag & -0.32 & 0.16 &  &  \\ 
-		recruitment & ARTR & VWC.su.0 & -0.26 & 0.25 &  &  \\ 
-		recruitment & ARTR & VWC.sp.1 & 0.22 & 0.34 &  &  \\ 
-		recruitment & HECO & VWC.su.lag & -0.31 & 0.18 &  &  \\ 
-		recruitment & HECO & VWC.su.0 & -0.30 & 0.18 &  &  \\ 
-		recruitment & HECO & VWC.f.lag & 0.19 & 0.40 &  &  \\ 
-		recruitment & POSE & VWC.sp.1 & 0.49 & 0.02 &  &  \\ 
-		recruitment & POSE & VWC.f.lag & 0.34 & 0.13 &  &  \\ 
-		recruitment & POSE & VWC.f.1 & 0.32 & 0.16 &  &  \\ 
-		recruitment & PSSP & VWC.su.lag & -0.52 & 0.02 &  &  \\ 
-		recruitment & PSSP & VWC.su.0 & -0.48 & 0.03 &  &  \\ 
-		recruitment & PSSP & VWC.sp.0 & 0.30 & 0.19 &  &  \\ 
-		survival & ARTR & VWC.su.0 & -0.60 & 0.00 &  &  \\ 
-		survival & ARTR & VWC.sp.0 & -0.41 & 0.06 &  &  \\ 
-		survival & ARTR & VWC.su.1 & -0.40 & 0.07 &  &  \\ 
-		survival & HECO & VWC.sp.0 & 0.44 & 0.04 &  &  \\ 
-		survival & HECO & VWC.sp.lag & 0.43 & 0.05 &  &  \\ 
-		survival & HECO & VWC.f.1 & 0.33 & 0.14 &  &  \\ 
-		survival & POSE & VWC.sp.0 & 0.44 & 0.04 & 0.22 & 0.34 \\ 
-		survival & POSE & VWC.sp.1 & 0.27 & 0.23 & -0.46 & 0.04 \\ 
-		survival & POSE & VWC.f.lag & -0.00 & 0.99 & 0.30 & 0.19 \\ 
-		survival & PSSP & VWC.sp.0 & 0.36 & 0.11 &  &  \\ 
-		survival & PSSP & VWC.sp.lag & 0.34 & 0.13 &  &  \\ 
-		survival & PSSP & VWC.su.1 & 0.26 & 0.26 &  &  \\ 
+		growth & ARTR & su.0 & -0.49 & 0.02 & 0.26 & 0.26 \\ 
+		growth & ARTR & f.0 & -0.28 & 0.23 & 0.40 & 0.08 \\ 
+		growth & ARTR & sp.1 & 0.17 & 0.45 & -0.33 & 0.14 \\ 
+		growth & HECO & su.1 & 0.69 & 0.00 &  &  \\ 
+		growth & HECO & su.0 & 0.50 & 0.02 &  &  \\ 
+		growth & HECO & f.lag & 0.37 & 0.10 &  &  \\ 
+		growth & POSE & f.lag & 0.31 & 0.17 & -0.11 & 0.64 \\ 
+		growth & POSE & su.lag & 0.29 & 0.20 & -0.20 & 0.38 \\ 
+		growth & POSE & sp.1 & 0.26 & 0.25 & -0.20 & 0.38 \\ 
+		growth & PSSP & f.lag & 0.34 & 0.13 &  &  \\ 
+		growth & PSSP & su.lag & 0.25 & 0.27 &  &  \\ 
+		growth & PSSP & f.0 & -0.22 & 0.34 &  &  \\ 
+		recruitment & ARTR & su.lag & -0.32 & 0.16 &  &  \\ 
+		recruitment & ARTR & su.0 & -0.26 & 0.25 &  &  \\ 
+		recruitment & ARTR & sp.1 & 0.22 & 0.34 &  &  \\ 
+		recruitment & HECO & su.lag & -0.31 & 0.18 &  &  \\ 
+		recruitment & HECO & su.0 & -0.30 & 0.18 &  &  \\ 
+		recruitment & HECO & f.lag & 0.19 & 0.40 &  &  \\ 
+		recruitment & POSE & sp.1 & 0.49 & 0.02 &  &  \\ 
+		recruitment & POSE & f.lag & 0.34 & 0.13 &  &  \\ 
+		recruitment & POSE & f.1 & 0.32 & 0.16 &  &  \\ 
+		recruitment & PSSP & su.lag & -0.52 & 0.02 &  &  \\ 
+		recruitment & PSSP & su.0 & -0.48 & 0.03 &  &  \\ 
+		recruitment & PSSP & sp.0 & 0.30 & 0.19 &  &  \\ 
+		survival & ARTR & su.0 & -0.60 & 0.00 &  &  \\ 
+		survival & ARTR & sp.0 & -0.41 & 0.06 &  &  \\ 
+		survival & ARTR & su.1 & -0.40 & 0.07 &  &  \\ 
+		survival & HECO & sp.0 & 0.44 & 0.04 &  &  \\ 
+		survival & HECO & sp.lag & 0.43 & 0.05 &  &  \\ 
+		survival & HECO & f.1 & 0.33 & 0.14 &  &  \\ 
+		survival & POSE & sp.0 & 0.44 & 0.04 & 0.22 & 0.34 \\ 
+		survival & POSE & sp.1 & 0.27 & 0.23 & -0.46 & 0.04 \\ 
+		survival & POSE & f.lag & -0.00 & 0.99 & 0.30 & 0.19 \\ 
+		survival & PSSP & sp.0 & 0.36 & 0.11 &  &  \\ 
+		survival & PSSP & sp.lag & 0.34 & 0.13 &  &  \\ 
+		survival & PSSP & su.1 & 0.26 & 0.26 &  &  \\ 
 		\hline
 	\end{tabular}
-	\caption{Correlations between intercept of year effects and soil moisture variables.
-P-values of correlations are given next to correlations.
-For ARTR growth and POSE growth and survival, the correlations between the year effects on size and the soil moisture variables are also given.}
+	\caption{Selected climate variables for each vital rate model for each species.
+Correlations and p-values between the choosen variables and the intercept of year effects model are shown.
+For ARTR growth and POSE growth and survival, the correlations between the year effects on size and the soil moisture variables are also given.
+"f" = fall, "su" = summer, "sp" = spring.
+ARTR = \textit{A.
+tripartita}, HECO = \textit{H.
+comata}, POSE = \textit{P.
+secunda}, PSSP = \textit{P.
+spicata}.}
 	\label{table:strongCor}
@@
 
+% latex table generated in R 3.3.1 by xtable 1.8-2 package
+% Tue Dec 20 12:38:30 2016
+\begin{table}[ht]
+	\centering
+	\begin{tabular}{lllrrrl}
+		\hline
+		species & vital\_rate & score & climate model & year effects model & diff & improved \\ 
+		\hline
+		ARTR & growth & lppd & -186.08 & -189.17 & 3.09 & *** \\ 
+		ARTR & growth & MSE & 0.55 & 0.55 & -0.00 & *** \\ 
+		ARTR & recruitment & lppd & -79.05 & -77.81 & -1.24 &  \\ 
+		ARTR & recruitment & MSE & 169.86 & 9.47 & 160.39 &  \\ 
+		ARTR & survival & lppd & -37.55 & -36.42 & -1.13 &  \\ 
+		ARTR & survival & MSE & 0.06 & 0.06 & 0.00 &  \\ 
+		HECO & growth & lppd & -475.54 & -454.36 & -21.18 &  \\ 
+		HECO & growth & MSE & 1.26 & 1.18 & 0.09 &  \\ 
+		HECO & recruitment & lppd & -149.43 & -151.60 & 2.17 & *** \\ 
+		HECO & recruitment & MSE & 288.16 & 235.57 & 52.59 &  \\ 
+		HECO & survival & lppd & -158.99 & -147.06 & -11.93 &  \\ 
+		HECO & survival & MSE & 0.12 & 0.11 & 0.01 &  \\ 
+		POSE & growth & lppd & -1823.71 & -1831.56 & 7.85 & *** \\ 
+		POSE & growth & MSE & 1.72 & 1.73 & -0.01 & *** \\ 
+		POSE & recruitment & lppd & -260.05 & -257.46 & -2.59 &  \\ 
+		POSE & recruitment & MSE & 45.40 & 37.32 & 8.08 &  \\ 
+		POSE & survival & lppd & -698.06 & -718.60 & 20.54 & *** \\ 
+		POSE & survival & MSE & 0.14 & 0.14 & -0.00 & *** \\ 
+		PSSP & growth & lppd & -1232.93 & -1237.92 & 4.99 & *** \\ 
+		PSSP & growth & MSE & 1.51 & 1.51 & 0.00 &  \\ 
+		PSSP & recruitment & lppd & -271.34 & -273.94 & 2.60 & *** \\ 
+		PSSP & recruitment & MSE & 79.09 & 42.68 & 36.41 &  \\ 
+		PSSP & survival & lppd & -332.47 & -307.26 & -25.21 &  \\ 
+		PSSP & survival & MSE & 0.11 & 0.10 & 0.01 &  \\ 
+		\hline
+	\end{tabular}
+	\caption{Comparison of model predictions from climate model and year effects model for each species and vital rate.
+Two prediction scores are reported, MSE and lppd.
+Lower MSE indicates improved predictions whereas higher lppd indicates improved predictions.
+Instances where the climate model outperformed the random year effects model are marked with "***" in the last column.
+ARTR = \textit{A.
+tripartita}, HECO = \textit{H.
+comata}, POSE = \textit{P.
+secunda}, PSSP = \textit{P.
+spicata}.}
+	\label{table:overallPreds}
+\end{table}
+
+% latex table generated in R 3.3.1 by xtable 1.8-2 package
+% Tue Dec 20 19:24:15 2016
+\begin{table}[ht]
+	\centering
+	\begin{tabular}{rllrrrl}
+		\hline
+		& species & stat & year effects model & climate model & diff & improved \\ 
+		\hline
+		1 & ARTR & cor & 0.48 & 0.19 & -0.29 &  \\ 
+		2 & ARTR & MSE & 0.30 & 0.30 & 0.00 &  \\ 
+		3 & HECO & cor & 0.29 & 0.22 & -0.07 &  \\ 
+		4 & HECO & MSE & 0.49 & 0.57 & 0.07 &  \\ 
+		5 & POSE & cor & 0.45 & 0.53 & 0.07 & *** \\ 
+		6 & POSE & MSE & 0.42 & 0.41 & -0.01 & *** \\ 
+		7 & PSSP & cor & 0.36 & 0.38 & 0.03 & *** \\ 
+		8 & PSSP & MSE & 0.39 & 0.39 & -0.01 & *** \\ 
+		\hline
+	\end{tabular}
+	\caption{MSE of predicted log cover changes and correlations between log cover changes predicted and observed.
+Predictions for the cover changes in the experimental plots were generated either from the year effects or the climate models.
+Instances where the climate model made better predictions than the year effects model are indicated with the "***".
+ARTR = \textit{A.
+tripartita}, HECO = \textit{H.
+comata}, POSE = \textit{P.
+secunda}, PSSP = \textit{P.
+spicata}.}
+	\label{table:corPGR}
+\end{table}
+
 \clearpage
@@
 
-\begin{figure}[tbp]
+\begin{figure}[!htbp]
 	\centering
@@
 
-\begin{figure}[tbp]
-	\centering
-	\includegraphics[width=1\textwidth]{treatment_trends_cover}
-	\caption{Average cover of each of the focal species in the control, drought, and irrigation treatments during the experiment.
-Percent cover of \textit{A.
-tripartita} refers to canopy cover, while the cover of the three grasses refers to basal cover.
-}
-	\label{fig:coverTrends}
-\end{figure}
-
-\begin{figure}[tbp]
+\begin{figure}[!htbp]
 	\centering
@@
 
+\begin{figure}[!htbp]
+	\centering
+	\includegraphics[width=1\textwidth]{parameter_predictions}
+	\caption{The treatment effects predicted by the climate model compared to the treatment effects observed for the intercept parameters (left side) and size by climate/treatment effects (right side).
+The observed treatment effects come from the treatment model fitted to all the data including the five years of the experiment.
+The predicted parameters come from the climate model fitted to all years of observational data but do not include the five years of the experiment (2011 to 2016).
+Treatment parameters that were both observed and predicted to be significantly different from zero are shown with the "*" symbol.
+The correlation between predicted and observed parameters is given on each panel.
+\textit{P.
+secunda} recruitment was predicted to be positively affected by the irrigation treatment but was in fact negatively affected.
+The other significant effects were in the correct direction.
+We assessed a parameter as significant when the 95\% Bayesian credible intervals did not overlap zero.}
+	\label{fig:parPredictions}
+\end{figure}
+
+\begin{figure}[!htbp]
+	\centering
+	\includegraphics[width=1\textwidth]{predicted_and_observed_cover}
+	\caption{Observed average cover per quadrat in the experimental and control plots and one year ahead cover predictions from the climate model.
+Cover predictions for each year are generated from the IBM based on the observed distribution of plants in each quadrat in the current year.
+Quadrat cover was not predicted for the first year of the experiment in 2011.
+Note the different cover scales for ARTR and the three grass species.
+ARTR = \textit{A.
+tripartita}, HECO = \textit{H.
+comata}, POSE = \textit{P.
+secunda}, PSSP = \textit{P.
+spicata}.}
+	\label{fig:coverPred}
+\end{figure}
+
+\begin{figure}[!htbp]
+	\centering
+	\includegraphics[width=1\textwidth]{ARTR_predicted_pgr_comparison}
+	\caption{Observed and predicted one year ahead log change in \textit{A.
+tripartita} cover in the experiment.
+Changes in cover predicted by the climate model are shown on the left and those predicted by the year effects model are shown on the right.
+Correlations coefficients between predictions and observations and MSE are shown for each treatment and model.}
+	\label{fig:pgrART}
+\end{figure}
+
+\begin{figure}[!htbp]
+	\centering
+	\includegraphics[width=1\textwidth]{HECO_predicted_pgr_comparison}
+	\caption{Observed and predicted one year ahead log change in \textit{H.
+comata} cover in the experiment.
+Changes in cover predicted by the climate model are shown on the left and those predicted by the year effects model are shown on the right.
+Correlations coefficients between predictions and observations and MSE are shown for each treatment and model.
+}
+	\label{fig:pgrHECO}
+\end{figure}
+
+\begin{figure}[!htbp]
+	\centering
+	\includegraphics[width=1\textwidth]{POSE_predicted_pgr_comparison}
+	\caption{Observed and predicted one year ahead log change in \textit{P.
+secunda} cover in the experiment.
+Changes in cover predicted by the climate model are shown on the left and those predicted by the year effects model are shown on the right.
+Correlations coefficients between predictions and observations and MSE are shown for each treatment and model.}
+	\label{fig:pgrPOSE}
+\end{figure}
+
+\begin{figure}[!htbp]
+	\centering
+	\includegraphics[width=1\textwidth]{PSSP_predicted_pgr_comparison}
+	\caption{Observed and predicted one year ahead log change in \textit{P.
+spicata} cover in the experiment.
+Changes in cover predicted by the climate model are shown on the left and those predicted by the year effects model are shown on the right.
+Correlations coefficients between predictions and observations and MSE are shown for each treatment and model.}
+	\label{fig:pgrPSSP}
+\end{figure}
+
 \clearpage 
@@
 \section{Additional Tables} 
-
-% supplementary tables
-
-\begin{table}[h]
-\caption{Cover change models}
-\centering
-%\begin{center}
-\begin{tabular}{l c c c c }
-\hline
-Species & ARTR & HECO & POSE & PSSP \\
-\hline
-(Intercept)           & $0.00$           & $-0.05$          & $0.01$           & $-0.10$          \\
-                      & $[-0.41;\ 0.42]$ & $[-0.21;\ 0.10]$ & $[-0.26;\ 0.28]$ & $[-0.28;\ 0.08]$ \\
-TreatmentNo\_grass    & $-0.09$          &                  &                  &                  \\
-                      & $[-0.75;\ 0.56]$ &                  &                  &                  \\
-TreatmentNo\_shrub    &                  & $0.33^{*}$       & $-0.08$          & $0.21^{*}$       \\
-                      &                  & $[0.01;\ 0.65]$  & $[-0.41;\ 0.25]$ & $[0.01;\ 0.40]$  \\
-\hline
-AIC                   & 284.97           & 160.41           & 337.77           & 253.15           \\
-BIC                   & 298.99           & 172.62           & 352.65           & 268.46           \\
-Log Likelihood        & -137.49          & -75.20           & -163.88          & -121.57          \\
-Num. obs.             & 122              & 85               & 145              & 158              \\
-Num. groups: quad     & 19               & 11               & 21               & 22               \\
-Num. groups: year     & 9                & 9                & 9                & 9                \\
-Var: quad (Intercept) & 0.43             & 0.00             & 0.02             & 0.00             \\
-Var: year (Intercept) & 0.00             & 0.02             & 0.12             & 0.06             \\
-Var: Residual         & 0.41             & 0.32             & 0.48             & 0.24             \\
-\hline
-\multicolumn{5}{l}{\scriptsize{$^*$ 0 outside the confidence interval}}
-\end{tabular}
-\label{table:coefficients}
-%\end{center}
-\end{table}
-
-% latex table generated in R 3.2.2 by xtable 1.8-2 package
-% Mon Sep 19 08:32:17 2016
-\begin{table}[ht]
-\centering
-\caption{Summary of fixed effects for the \textit{A. tripartita} survival model} 
-\label{ARTRsurvival}
-\begin{tabular}{rrrrrrrr}
-  \hline
- & mean & sd & 0.025quant & 0.5quant & 0.975quant & mode & kld \\ 
-  \hline
-(Intercept) & -0.0207 & 0.2261 & -0.4554 & -0.0246 & 0.4367 & -0.0321 & 0.0000 \\ 
-  logarea & 0.7208 & 0.0575 & 0.6109 & 0.7195 & 0.8378 & 0.7171 & 0.0000 \\ 
-  TreatmentNo\_grass & -1.3218 & 0.7343 & -2.7293 & -1.3340 & 0.1563 & -1.3591 & 0.0000 \\ 
-  W.ARTR & -1.5214 & 0.2075 & -1.9391 & -1.5179 & -1.1237 & -1.5107 & 0.0000 \\ 
-  W.HECO & -0.0694 & 0.0423 & -0.1485 & -0.0708 & 0.0178 & -0.0738 & 0.0000 \\ 
-  W.POSE & -0.0049 & 0.0761 & -0.1469 & -0.0077 & 0.1525 & -0.0134 & 0.0000 \\ 
-  W.PSSP & 0.0281 & 0.0275 & -0.0235 & 0.0271 & 0.0849 & 0.0252 & 0.0000 \\ 
-  W.allcov & -0.0055 & 0.0083 & -0.0219 & -0.0054 & 0.0109 & -0.0054 & 0.0000 \\ 
-  W.allpts & 0.1694 & 0.1579 & -0.1386 & 0.1686 & 0.4812 & 0.1672 & 0.0000 \\ 
-   \hline
-\end{tabular}
-\end{table}
-
-% latex table generated in R 3.2.2 by xtable 1.8-2 package
-% Mon Sep 19 08:33:00 2016
-\begin{table}[ht]
-\centering
-\caption{Summary of fixed effects for the \textit{H. comata} survival model} 
-\label{HECOsurvival}
-\begin{tabular}{rrrrrrrr}
-  \hline
- & mean & sd & 0.025quant & 0.5quant & 0.975quant & mode & kld \\ 
-  \hline
-(Intercept) & 1.4189 & 0.2115 & 0.9991 & 1.4190 & 1.8371 & 1.4192 & 0.0000 \\ 
-  logarea & 1.2885 & 0.0813 & 1.1399 & 1.2843 & 1.4601 & 1.2758 & 0.0000 \\ 
-  TreatmentNo\_shrub & 0.3958 & 0.4175 & -0.4193 & 0.3942 & 1.2190 & 0.3909 & 0.0000 \\ 
-  W.ARTR & -0.0064 & 0.0028 & -0.0120 & -0.0064 & -0.0009 & -0.0064 & 0.0000 \\ 
-  W.HECO & -0.6647 & 0.0673 & -0.7996 & -0.6638 & -0.5351 & -0.6619 & 0.0000 \\ 
-  W.POSE & 0.0732 & 0.0534 & -0.0304 & 0.0728 & 0.1790 & 0.0720 & 0.0000 \\ 
-  W.PSSP & 0.0190 & 0.0184 & -0.0167 & 0.0188 & 0.0554 & 0.0185 & 0.0000 \\ 
-  W.allcov & 0.0021 & 0.0047 & -0.0072 & 0.0021 & 0.0113 & 0.0021 & 0.0000 \\ 
-  W.allpts & -0.1185 & 0.0919 & -0.2998 & -0.1183 & 0.0610 & -0.1177 & 0.0000 \\ 
-   \hline
-\end{tabular}
-\end{table}
-
-% latex table generated in R 3.2.2 by xtable 1.8-2 package
-% Mon Sep 19 08:34:30 2016
-\begin{table}[ht]
-\centering
-\caption{Summary of fixed effects for the \textit{Poa secunda} survival model} 
-\label{POSEsurvival}
-\begin{tabular}{rrrrrrrr}
-  \hline
- & mean & sd & 0.025quant & 0.5quant & 0.975quant & mode & kld \\ 
-  \hline
-(Intercept) & 1.3162 & 0.1965 & 0.9242 & 1.3171 & 1.7024 & 1.3186 & 0.0000 \\ 
-  logarea & 1.0585 & 0.0640 & 0.9365 & 1.0569 & 1.1902 & 1.0537 & 0.0000 \\ 
-  TreatmentNo\_shrub & -0.2572 & 0.1723 & -0.5949 & -0.2573 & 0.0811 & -0.2576 & 0.0000 \\ 
-  W.ARTR & 0.0001 & 0.0019 & -0.0036 & 0.0001 & 0.0038 & 0.0001 & 0.0000 \\ 
-  W.HECO & -0.0153 & 0.0138 & -0.0424 & -0.0154 & 0.0119 & -0.0154 & 0.0000 \\ 
-  W.POSE & -1.2632 & 0.0816 & -1.4260 & -1.2623 & -1.1053 & -1.2605 & 0.0000 \\ 
-  W.PSSP & 0.0249 & 0.0122 & 0.0011 & 0.0249 & 0.0490 & 0.0247 & 0.0000 \\ 
-  W.allcov & -0.0024 & 0.0026 & -0.0075 & -0.0024 & 0.0029 & -0.0024 & 0.0000 \\ 
-  W.allpts & 0.0018 & 0.0524 & -0.1009 & 0.0018 & 0.1048 & 0.0017 & 0.0000 \\ 
-   \hline
-\end{tabular}
-\end{table}
-
-% latex table generated in R 3.2.2 by xtable 1.8-2 package
-% Mon Sep 19 08:35:48 2016
-\begin{table}[ht]
-\centering
-\caption{Summary of fixed effects for the \textit{P. spicata} survival model} 
-\label{PSSPsurvival}
-\begin{tabular}{rrrrrrrr}
-  \hline
- & mean & sd & 0.025quant & 0.5quant & 0.975quant & mode & kld \\ 
-  \hline
-(Intercept) & 1.2121 & 0.1699 & 0.8815 & 1.2103 & 1.5533 & 1.2066 & 0.0000 \\ 
-  logarea & 1.5492 & 0.0976 & 1.3612 & 1.5474 & 1.7473 & 1.5437 & 0.0000 \\ 
-  TreatmentNo\_shrub & -0.1885 & 0.2208 & -0.6237 & -0.1879 & 0.2431 & -0.1867 & 0.0000 \\ 
-  W.ARTR & 0.0101 & 0.0021 & 0.0061 & 0.0101 & 0.0142 & 0.0101 & 0.0000 \\ 
-  W.HECO & 0.0046 & 0.0185 & -0.0319 & 0.0046 & 0.0406 & 0.0047 & 0.0000 \\ 
-  W.POSE & 0.0245 & 0.0384 & -0.0502 & 0.0242 & 0.1007 & 0.0236 & 0.0000 \\ 
-  W.PSSP & -0.4430 & 0.0291 & -0.5010 & -0.4427 & -0.3867 & -0.4420 & 0.0000 \\ 
-  W.allcov & 0.0100 & 0.0028 & 0.0046 & 0.0100 & 0.0155 & 0.0100 & 0.0000 \\ 
-  W.allpts & 0.0943 & 0.0605 & -0.0246 & 0.0944 & 0.2129 & 0.0944 & 0.0000 \\ 
-   \hline
-\end{tabular}
-\end{table}
-
-% latex table generated in R 3.2.2 by xtable 1.8-2 package
-% Mon Sep 19 08:36:00 2016
-\begin{table}[ht]
-\centering
-\caption{Summary of fixed effects for the \textit{A. tripartita} growth model} 
-\label{ARTRgrowth}
-\begin{tabular}{rrrrrrrr}
-  \hline
- & mean & sd & 0.025quant & 0.5quant & 0.975quant & mode & kld \\ 
-  \hline
-(Intercept) & 0.7378 & 0.2437 & 0.2591 & 0.7378 & 1.2161 & 0.7377 & 0.0000 \\ 
-  logarea.t0 & 0.8663 & 0.0393 & 0.7892 & 0.8663 & 0.9434 & 0.8663 & 0.0000 \\ 
-  TreatmentNo\_grass & 0.1037 & 0.1436 & -0.1783 & 0.1037 & 0.3854 & 0.1037 & 0.0000 \\ 
-  W.ARTR & -0.0364 & 0.0892 & -0.2116 & -0.0364 & 0.1387 & -0.0364 & 0.0000 \\ 
-  W.HECO & 0.0033 & 0.0136 & -0.0233 & 0.0033 & 0.0299 & 0.0033 & 0.0000 \\ 
-  W.POSE & -0.0573 & 0.0268 & -0.1100 & -0.0573 & -0.0047 & -0.0573 & 0.0000 \\ 
-  W.PSSP & 0.0062 & 0.0084 & -0.0102 & 0.0062 & 0.0226 & 0.0062 & 0.0000 \\ 
-  W.allcov & -0.0055 & 0.0030 & -0.0114 & -0.0055 & 0.0004 & -0.0055 & 0.0000 \\ 
-  W.allpts & 0.0029 & 0.0477 & -0.0908 & 0.0029 & 0.0965 & 0.0029 & 0.0000 \\ 
-   \hline
-\end{tabular}
-\end{table}
-
-% latex table generated in R 3.2.2 by xtable 1.8-2 package
-% Mon Sep 19 08:36:24 2016
-\begin{table}[ht]
-\centering
-\caption{Summary of fixed effects for the \textit{H. comata} growth model} 
-\label{HECOgrowth}
-\begin{tabular}{rrrrrrrr}
-  \hline
- & mean & sd & 0.025quant & 0.5quant & 0.975quant & mode & kld \\ 
-  \hline
-(Intercept) & 0.3782 & 0.0797 & 0.2212 & 0.3783 & 0.5343 & 0.3784 & 0.0000 \\ 
-  logarea.t0 & 0.8272 & 0.0205 & 0.7869 & 0.8272 & 0.8674 & 0.8272 & 0.0000 \\ 
-  TreatmentNo\_shrub & -0.0162 & 0.1336 & -0.2785 & -0.0162 & 0.2460 & -0.0162 & 0.0000 \\ 
-  W.ARTR & -0.0031 & 0.0010 & -0.0051 & -0.0031 & -0.0010 & -0.0031 & 0.0000 \\ 
-  W.HECO & -0.0479 & 0.0195 & -0.0862 & -0.0479 & -0.0098 & -0.0479 & 0.0000 \\ 
-  W.POSE & 0.0242 & 0.0162 & -0.0077 & 0.0242 & 0.0560 & 0.0242 & 0.0000 \\ 
-  W.PSSP & -0.0151 & 0.0065 & -0.0279 & -0.0151 & -0.0023 & -0.0151 & 0.0000 \\ 
-  W.allcov & -0.0038 & 0.0019 & -0.0076 & -0.0038 & 0.0000 & -0.0038 & 0.0000 \\ 
-  W.allpts & -0.0513 & 0.0380 & -0.1258 & -0.0513 & 0.0232 & -0.0513 & 0.0000 \\ 
-   \hline
-\end{tabular}
-\end{table}
-
-% latex table generated in R 3.2.2 by xtable 1.8-2 package
-% Mon Sep 19 08:36:58 2016
-\begin{table}[ht]
-\centering
-\caption{Summary of fixed effects for the \textit{Poa secunda} growth model} 
-\label{POSEgrowth}
-\begin{tabular}{rrrrrrrr}
-  \hline
- & mean & sd & 0.025quant & 0.5quant & 0.975quant & mode & kld \\ 
-  \hline
-(Intercept) & 0.4923 & 0.0681 & 0.3581 & 0.4923 & 0.6259 & 0.4924 & 0.0000 \\ 
-  logarea.t0 & 0.6750 & 0.0232 & 0.6292 & 0.6750 & 0.7203 & 0.6751 & 0.0000 \\ 
-  TreatmentNo\_shrub & 0.2161 & 0.0644 & 0.0897 & 0.2161 & 0.3425 & 0.2161 & 0.0000 \\ 
-  W.ARTR & -0.0002 & 0.0008 & -0.0019 & -0.0002 & 0.0014 & -0.0002 & 0.0000 \\ 
-  W.HECO & 0.0039 & 0.0063 & -0.0084 & 0.0039 & 0.0161 & 0.0039 & 0.0000 \\ 
-  W.POSE & -0.2641 & 0.0387 & -0.3400 & -0.2641 & -0.1883 & -0.2641 & 0.0000 \\ 
-  W.PSSP & -0.0073 & 0.0056 & -0.0183 & -0.0073 & 0.0038 & -0.0073 & 0.0000 \\ 
-  W.allcov & -0.0003 & 0.0012 & -0.0026 & -0.0003 & 0.0020 & -0.0003 & 0.0000 \\ 
-  W.allpts & -0.0254 & 0.0231 & -0.0709 & -0.0254 & 0.0200 & -0.0254 & 0.0000 \\ 
-   \hline
-\end{tabular}
-\end{table}
-
-% latex table generated in R 3.2.2 by xtable 1.8-2 package
-% Mon Sep 19 08:37:48 2016
-\begin{table}[ht]
-\centering
-\caption{Summary of fixed effects for the \textit{P. spicata} growth model} 
-\label{PSSPgrowth}
-\begin{tabular}{rrrrrrrr}
-  \hline
- & mean & sd & 0.025quant & 0.5quant & 0.975quant & mode & kld \\ 
-  \hline
-(Intercept) & 0.3987 & 0.0660 & 0.2690 & 0.3987 & 0.5281 & 0.3988 & 0.0000 \\ 
-  logarea.t0 & 0.8249 & 0.0149 & 0.7956 & 0.8249 & 0.8543 & 0.8249 & 0.0000 \\ 
-  TreatmentNo\_shrub & 0.1948 & 0.0714 & 0.0546 & 0.1948 & 0.3349 & 0.1948 & 0.0000 \\ 
-  W.ARTR & -0.0025 & 0.0008 & -0.0041 & -0.0025 & -0.0010 & -0.0025 & 0.0000 \\ 
-  W.HECO & -0.0118 & 0.0072 & -0.0260 & -0.0118 & 0.0023 & -0.0118 & 0.0000 \\ 
-  W.POSE & -0.0067 & 0.0130 & -0.0323 & -0.0067 & 0.0189 & -0.0067 & 0.0000 \\ 
-  W.PSSP & -0.1480 & 0.0133 & -0.1740 & -0.1480 & -0.1220 & -0.1480 & 0.0000 \\ 
-  W.allcov & -0.0036 & 0.0011 & -0.0057 & -0.0036 & -0.0015 & -0.0036 & 0.0000 \\ 
-  W.allpts & -0.0522 & 0.0236 & -0.0984 & -0.0522 & -0.0060 & -0.0522 & 0.0000 \\ 
-   \hline
-\end{tabular}
-\end{table}
-
-\begin{table}
-\caption{\textit{Poa secunda} growth with \textit{Artemisia} canopy effect}
-\begin{center}
-\begin{tabular}{l c }
-\hline
- & Model 1 \\
-\hline
-(Intercept)                      & $0.49^{*}$        \\
-                                 & $[0.36;\ 0.62]$   \\
-logarea.t0                       & $0.68^{*}$        \\
-                                 & $[0.63;\ 0.72]$   \\
-TreatmentNo\_shrub               & $0.24^{*}$        \\
-                                 & $[0.09;\ 0.38]$   \\
-W.ARTR                           & $-0.00$           \\
-                                 & $[-0.00;\ 0.00]$  \\
-W.HECO                           & $0.00$            \\
-                                 & $[-0.01;\ 0.02]$  \\
-W.POSE                           & $-0.26^{*}$       \\
-                                 & $[-0.34;\ -0.19]$ \\
-W.PSSP                           & $-0.01$           \\
-                                 & $[-0.02;\ 0.00]$  \\
-W.allcov                         & $-0.00$           \\
-                                 & $[-0.00;\ 0.00]$  \\
-W.allpts                         & $-0.02$           \\
-                                 & $[-0.07;\ 0.02]$  \\
-inARTR                           & $-0.06$           \\
-                                 & $[-0.28;\ 0.16]$  \\
-\hline
-AIC                              & 15641.80          \\
-BIC                              & 15734.29          \\
-Log Likelihood                   & -7806.90          \\
-Num. obs.                        & 5469              \\
-Num. groups: year                & 31                \\
-Var: year (Intercept)            & 0.11              \\
-Var: year logarea.t0             & 0.01              \\
-Cov: year (Intercept) logarea.t0 & -0.02             \\
-Var: Residual                    & 0.98              \\
-\hline
-\multicolumn{2}{l}{\scriptsize{$^*$ 0 outside the confidence interval}}
-\end{tabular}
-\label{table:POSEgrowth-inARTR}
-\end{center}
-\end{table}
-
-\begin{table}
-\caption{\textit{P. spicata} growth with \textit{Artemisia} canopy effect}
-\begin{center}
-\begin{tabular}{l c }
-\hline
- & Model 1 \\
-\hline
-(Intercept)                      & $0.42^{*}$        \\
-                                 & $[0.29;\ 0.56]$   \\
-logarea.t0                       & $0.82^{*}$        \\
-                                 & $[0.79;\ 0.86]$   \\
-TreatmentNo\_shrub               & $0.24^{*}$        \\
-                                 & $[0.09;\ 0.40]$   \\
-W.ARTR                           & $-0.00^{*}$       \\
-                                 & $[-0.00;\ -0.00]$ \\
-W.HECO                           & $-0.01^{*}$       \\
-                                 & $[-0.03;\ -0.00]$ \\
-W.POSE                           & $-0.01$           \\
-                                 & $[-0.03;\ 0.02]$  \\
-W.PSSP                           & $-0.14^{*}$       \\
-                                 & $[-0.16;\ -0.11]$ \\
-W.allcov                         & $-0.00^{*}$       \\
-                                 & $[-0.01;\ -0.00]$ \\
-W.allpts                         & $-0.06^{*}$       \\
-                                 & $[-0.10;\ -0.01]$ \\
-inARTR                           & $-0.14$           \\
-                                 & $[-0.35;\ 0.08]$  \\
-\hline
-AIC                              & 15282.37          \\
-BIC                              & 15376.25          \\
-Log Likelihood                   & -7627.18          \\
-Num. obs.                        & 6039              \\
-Num. groups: year                & 31                \\
-Var: year (Intercept)            & 0.12              \\
-Var: year logarea.t0             & 0.01              \\
-Cov: year (Intercept) logarea.t0 & -0.03             \\
-Var: Residual                    & 0.71              \\
-\hline
-\multicolumn{2}{l}{\scriptsize{$^*$ 0 outside the confidence interval}}
-\end{tabular}
-\label{table:PSSPgrowth-inARTR}
-\end{center}
-\end{table}
-
-\begin{table}
-\caption{\textit{Poa secunda} growth with year-by-treatment interaction}
-\begin{center}
-\begin{tabular}{l c }
-\hline
- & Model 1 \\
-\hline
-(Intercept)                      & $0.49^{*}$        \\
-                                 & $[0.37;\ 0.62]$   \\
-trtYears1                        & $0.06$            \\
-                                 & $[-0.17;\ 0.29]$  \\
-trtYears2                        & $0.24$            \\
-                                 & $[-0.01;\ 0.49]$  \\
-trtYears3                        & $0.23$            \\
-                                 & $[-0.05;\ 0.51]$  \\
-trtYears4                        & $0.25$            \\
-                                 & $[-0.07;\ 0.56]$  \\
-trtYears5                        & $0.45^{*}$        \\
-                                 & $[0.11;\ 0.79]$   \\
-logarea.t0                       & $0.68^{*}$        \\
-                                 & $[0.63;\ 0.72]$   \\
-W.ARTR                           & $-0.00$           \\
-                                 & $[-0.00;\ 0.00]$  \\
-W.HECO                           & $0.00$            \\
-                                 & $[-0.01;\ 0.02]$  \\
-W.POSE                           & $-0.26^{*}$       \\
-                                 & $[-0.34;\ -0.19]$ \\
-W.PSSP                           & $-0.01$           \\
-                                 & $[-0.02;\ 0.00]$  \\
-W.allcov                         & $-0.00$           \\
-                                 & $[-0.00;\ 0.00]$  \\
-W.allpts                         & $-0.03$           \\
-                                 & $[-0.07;\ 0.02]$  \\
-\hline
-AIC                              & 15648.62          \\
-BIC                              & 15760.94          \\
-Log Likelihood                   & -7807.31          \\
-Num. obs.                        & 5469              \\
-Num. groups: year                & 31                \\
-Var: year (Intercept)            & 0.11              \\
-Var: year logarea.t0             & 0.01              \\
-Cov: year (Intercept) logarea.t0 & -0.02             \\
-Var: Residual                    & 0.98              \\
-\hline
-\multicolumn{2}{l}{\scriptsize{$^*$ 0 outside the confidence interval}}
-\end{tabular}
-\label{table:POSEgrowth-byYr}
-\end{center}
-\end{table}
-
-\begin{table}
-\caption{\textit{P. spicata} growth with year-by-treatment interaction}
-\begin{center}
-\begin{tabular}{l c }
-\hline
- & Model 1 \\
-\hline
-(Intercept)                      & $0.42^{*}$        \\
-                                 & $[0.28;\ 0.55]$   \\
-trtYears1                        & $0.26$            \\
-                                 & $[-0.01;\ 0.54]$  \\
-trtYears2                        & $-0.05$           \\
-                                 & $[-0.32;\ 0.22]$  \\
-trtYears3                        & $0.44^{*}$        \\
-                                 & $[0.14;\ 0.73]$   \\
-trtYears4                        & $0.42^{*}$        \\
-                                 & $[0.12;\ 0.72]$   \\
-trtYears5                        & $-0.11$           \\
-                                 & $[-0.42;\ 0.20]$  \\
-logarea.t0                       & $0.82^{*}$        \\
-                                 & $[0.79;\ 0.86]$   \\
-W.ARTR                           & $-0.00^{*}$       \\
-                                 & $[-0.00;\ -0.00]$ \\
-W.HECO                           & $-0.01^{*}$       \\
-                                 & $[-0.03;\ -0.00]$ \\
-W.POSE                           & $-0.01$           \\
-                                 & $[-0.03;\ 0.02]$  \\
-W.PSSP                           & $-0.14^{*}$       \\
-                                 & $[-0.16;\ -0.11]$ \\
-W.allcov                         & $-0.00^{*}$       \\
-                                 & $[-0.01;\ -0.00]$ \\
-W.allpts                         & $-0.06^{*}$       \\
-                                 & $[-0.10;\ -0.01]$ \\
-\hline
-AIC                              & 15281.67          \\
-BIC                              & 15395.67          \\
-Log Likelihood                   & -7623.83          \\
-Num. obs.                        & 6039              \\
-Num. groups: year                & 31                \\
-Var: year (Intercept)            & 0.12              \\
-Var: year logarea.t0             & 0.01              \\
-Cov: year (Intercept) logarea.t0 & -0.03             \\
-Var: Residual                    & 0.71              \\
-\hline
-\multicolumn{2}{l}{\scriptsize{$^*$ 0 outside the confidence interval}}
-\end{tabular}
-\label{table:PSSPgrowth-byYr}
-\end{center}
-\end{table}
-
-% latex table generated in R 3.2.2 by xtable 1.8-2 package
-% Mon Sep 19 08:37:48 2016
-\begin{table}[ht]
-\centering
-\caption{Summary of fixed effects for the recruitment model (symbols correspond to Eqns. \ref{eqn:recrDataModel} and \ref{eqn:recrProcessModel})} 
-\label{table:recruitment}
-\begin{tabular}{rrrrrrr}
-  \hline
- & mean & sd & X2.5. & X97.5. & Rhat & n.eff \\ 
-  \hline
- $\gamma$[1] & 0.3341 & 0.7071 & -1.1262 & 1.6391 & 1.0018 &  1100 \\ 
-  $\gamma$[2] & 3.4488 & 0.4468 & 2.5010 & 4.3401 & 1.0904 &    25 \\ 
-  $\gamma$[3] & 3.2440 & 0.3650 & 2.4840 & 3.9190 & 1.0536 &    34 \\ 
- $\gamma$[4] & 2.9054 & 0.3689 & 2.1820 & 3.6160 & 1.0149 &   110 \\ 
-  $\chi$[2,2] & -0.0953 & 0.4093 & -0.8854 & 0.7570 & 1.0010 &  2000 \\ 
- $\chi$[2,3] & -1.2787 & 0.3325 & -1.9370 & -0.6392 & 1.0041 &   420 \\ 
-   $\chi$[2,4] & 0.0951 & 0.2617 & -0.4121 & 0.6063 & 1.0023 &   810 \\ 
-  $\chi$[3,1] & -1.4366 & 0.8544 & -3.1491 & 0.1132 & 1.0006 &  2000 \\ 
-  $\omega$[1,1] & -0.5881 & 0.1018 & -0.7639 & -0.3720 & 1.0197 &    82 \\ 
-  $\omega$[1,2] & 0.0635 & 0.0651 & -0.0528 & 0.1997 & 1.0329 &    59 \\ 
-  $\omega$[1,3] & 0.0220 & 0.0422 & -0.0610 & 0.1040 & 1.0036 &   920 \\ 
-  $\omega$[1,4] & 0.1164 & 0.0456 & 0.0347 & 0.2123 & 1.0197 &    96 \\ 
-  $\omega$[2,1] & -0.4800 & 0.1522 & -0.7746 & -0.1906 & 1.0029 &  2000 \\ 
-  $\omega$[2,2] & -1.7368 & 0.1297 & -1.9841 & -1.4819 & 1.0006 &  2000 \\ 
-  $\omega$[2,3] & 0.0388 & 0.0865 & -0.1287 & 0.2138 & 1.0152 &   100 \\ 
-  $\omega$[2,4] & -0.3682 & 0.0926 & -0.5434 & -0.1881 & 1.0005 &  2000 \\ 
-  $\omega$[3,1] & -0.6589 & 0.3342 & -1.3131 & -0.0155 & 1.0096 &   280 \\ 
-  $\omega$[3,2] & -0.1021 & 0.1981 & -0.4828 & 0.2936 & 1.0089 &   210 \\ 
-  $\omega$[3,3] & -1.8943 & 0.1611 & -2.2060 & -1.5850 & 1.0084 &  2000 \\ 
-  $\omega$[3,4] & -0.1953 & 0.1508 & -0.4968 & 0.0995 & 1.0074 &  2000 \\ 
-  $\omega$[4,1] & -0.1867 & 0.3000 & -0.7608 & 0.4486 & 1.0099 &   190 \\ 
-  $\omega$[4,2] & -0.3746 & 0.1896 & -0.7224 & -0.0115 & 1.0453 &    55 \\ 
-  $\omega$[4,3] & 0.1131 & 0.1437 & -0.1583 & 0.4170 & 1.0321 &    55 \\ 
-  $\omega$[4,4] & -1.7379 & 0.1549 & -2.0400 & -1.4409 & 1.0116 &  1400 \\ 
-   $\theta$[1] & 0.6198 & 0.0740 & 0.4835 & 0.7766 & 1.0008 &  2000 \\ 
-   $\theta$[2] & 1.1197 & 0.1428 & 0.8666 & 1.4450 & 1.0008 &  2000 \\ 
-   $\theta$[3] & 1.1718 & 0.1068 & 0.9685 & 1.3890 & 1.0008 &  2000 \\ 
-  $\theta$[4] & 1.0961 & 0.1060 & 0.9009 & 1.3280 & 1.0022 &   890 \\ 
-  $p$[1] & 0.8137 & 0.1228 & 0.4813 & 0.9525 & 1.0399 &    60 \\ 
-  $p$[2] & 0.9996 & 0.0005 & 0.9984 & 1.0000 & 1.0007 &  2000 \\ 
-  $p$[3] & 0.7906 & 0.1288 & 0.4273 & 0.9382 & 1.0176 &  2000 \\ 
-  $p$[4] & 0.7502 & 0.1480 & 0.3920 & 0.9453 & 1.0150 &  2000 \\ 
-   \hline
-\end{tabular}
-\end{table}
+% latex table generated in R 3.3.1 by xtable 1.8-2 package
+% Tue Dec 20 13:32:53 2016
+\begin{longtable}{rllllrrrl}
+	\hline
+	& species & vital\_rate & Treatment & score & climate model & year effects model & diff & improved \\ 
+	\hline
+	\endhead
+	\hline
+	\multicolumn{9}{l}{\footnotesize Continued on next page}
+	\endfoot
+	\endlastfoot
+	1 & ARTR & growth & Control & lppd & -107.53 & -108.40 & 0.87 & *** \\ 
+	2 & ARTR & growth & Control & MSE & 0.57 & 0.56 & 0.01 &  \\ 
+	3 & ARTR & growth & Drought & lppd & -39.71 & -40.01 & 0.30 & *** \\ 
+	4 & ARTR & growth & Drought & MSE & 0.53 & 0.49 & 0.03 &  \\ 
+	5 & ARTR & growth & Irrigation & lppd & -38.85 & -40.75 & 1.91 & *** \\ 
+	6 & ARTR & growth & Irrigation & MSE & 0.50 & 0.57 & -0.07 & *** \\ 
+	7 & ARTR & recruitment & Control & lppd & -32.61 & -31.05 & -1.56 &  \\ 
+	8 & ARTR & recruitment & Control & MSE & 61.88 & 10.59 & 51.29 &  \\ 
+	9 & ARTR & recruitment & Drought & lppd & -28.93 & -26.52 & -2.41 &  \\ 
+	10 & ARTR & recruitment & Drought & MSE & 523.95 & 11.62 & 512.33 &  \\ 
+	11 & ARTR & recruitment & Irrigation & lppd & -17.51 & -20.24 & 2.73 & *** \\ 
+	12 & ARTR & recruitment & Irrigation & MSE & 4.72 & 5.34 & -0.62 & *** \\ 
+	13 & ARTR & survival & Control & lppd & -24.19 & -23.06 & -1.13 &  \\ 
+	14 & ARTR & survival & Control & MSE & 0.07 & 0.07 & 0.00 &  \\ 
+	15 & ARTR & survival & Drought & lppd & -5.48 & -5.34 & -0.13 &  \\ 
+	16 & ARTR & survival & Drought & MSE & 0.04 & 0.04 & -0.00 & *** \\ 
+	17 & ARTR & survival & Irrigation & lppd & -7.88 & -8.01 & 0.13 & *** \\ 
+	18 & ARTR & survival & Irrigation & MSE & 0.06 & 0.06 & -0.00 & *** \\ 
+	19 & HECO & growth & Control & lppd & -377.77 & -369.46 & -8.31 &  \\ 
+	20 & HECO & growth & Control & MSE & 1.11 & 1.09 & 0.02 &  \\ 
+	21 & HECO & growth & Drought & lppd & -8.74 & -10.97 & 2.23 & *** \\ 
+	22 & HECO & growth & Drought & MSE & 3.12 & 4.35 & -1.23 & *** \\ 
+	23 & HECO & growth & Irrigation & lppd & -89.03 & -73.93 & -15.10 &  \\ 
+	24 & HECO & growth & Irrigation & MSE & 2.02 & 1.47 & 0.55 &  \\ 
+	25 & HECO & recruitment & Control & lppd & -93.35 & -93.47 & 0.12 & *** \\ 
+	26 & HECO & recruitment & Control & MSE & 613.41 & 499.11 & 114.29 &  \\ 
+	27 & HECO & recruitment & Drought & lppd & -25.39 & -27.56 & 2.17 & *** \\ 
+	28 & HECO & recruitment & Drought & MSE & 1.94 & 2.33 & -0.38 & *** \\ 
+	29 & HECO & recruitment & Irrigation & lppd & -30.69 & -30.56 & -0.12 &  \\ 
+	30 & HECO & recruitment & Irrigation & MSE & 5.21 & 7.62 & -2.41 & *** \\ 
+	31 & HECO & survival & Control & lppd & -124.59 & -112.70 & -11.90 &  \\ 
+	32 & HECO & survival & Control & MSE & 0.12 & 0.11 & 0.01 &  \\ 
+	33 & HECO & survival & Drought & lppd & -17.89 & -20.82 & 2.93 & *** \\ 
+	34 & HECO & survival & Drought & MSE & 0.22 & 0.26 & -0.05 & *** \\ 
+	35 & HECO & survival & Irrigation & lppd & -16.51 & -13.54 & -2.97 &  \\ 
+	36 & HECO & survival & Irrigation & MSE & 0.09 & 0.07 & 0.02 &  \\ 
+	37 & POSE & growth & Control & lppd & -1117.20 & -1117.29 & 0.08 & *** \\ 
+	38 & POSE & growth & Control & MSE & 1.50 & 1.51 & -0.00 & *** \\ 
+	39 & POSE & growth & Drought & lppd & -254.17 & -257.32 & 3.16 & *** \\ 
+	40 & POSE & growth & Drought & MSE & 2.66 & 2.69 & -0.03 & *** \\ 
+	41 & POSE & growth & Irrigation & lppd & -452.34 & -456.95 & 4.61 & *** \\ 
+	42 & POSE & growth & Irrigation & MSE & 1.87 & 1.90 & -0.03 & *** \\ 
+	43 & POSE & recruitment & Control & lppd & -127.64 & -128.73 & 1.09 & *** \\ 
+	44 & POSE & recruitment & Control & MSE & 35.20 & 44.82 & -9.61 & *** \\ 
+	45 & POSE & recruitment & Drought & lppd & -60.34 & -63.15 & 2.80 & *** \\ 
+	46 & POSE & recruitment & Drought & MSE & 23.60 & 33.88 & -10.28 & *** \\ 
+	47 & POSE & recruitment & Irrigation & lppd & -72.06 & -65.58 & -6.48 &  \\ 
+	48 & POSE & recruitment & Irrigation & MSE & 85.04 & 27.63 & 57.41 &  \\ 
+	49 & POSE & survival & Control & lppd & -366.49 & -380.49 & 14.00 & *** \\ 
+	50 & POSE & survival & Control & MSE & 0.12 & 0.12 & -0.00 & *** \\ 
+	51 & POSE & survival & Drought & lppd & -175.29 & -187.88 & 12.59 & *** \\ 
+	52 & POSE & survival & Drought & MSE & 0.21 & 0.22 & -0.01 & *** \\ 
+	53 & POSE & survival & Irrigation & lppd & -156.28 & -150.23 & -6.05 &  \\ 
+	54 & POSE & survival & Irrigation & MSE & 0.13 & 0.13 & 0.00 &  \\ 
+	55 & PSSP & growth & Control & lppd & -627.07 & -625.63 & -1.44 &  \\ 
+	56 & PSSP & growth & Control & MSE & 1.38 & 1.36 & 0.02 &  \\ 
+	57 & PSSP & growth & Drought & lppd & -292.90 & -297.61 & 4.71 & *** \\ 
+	58 & PSSP & growth & Drought & MSE & 1.82 & 1.85 & -0.03 & *** \\ 
+	59 & PSSP & growth & Irrigation & lppd & -312.96 & -314.67 & 1.71 & *** \\ 
+	60 & PSSP & growth & Irrigation & MSE & 1.51 & 1.51 & -0.00 & *** \\ 
+	61 & PSSP & recruitment & Control & lppd & -123.81 & -125.23 & 1.42 & *** \\ 
+	62 & PSSP & recruitment & Control & MSE & 30.59 & 34.79 & -4.20 & *** \\ 
+	63 & PSSP & recruitment & Drought & lppd & -87.34 & -86.58 & -0.77 &  \\ 
+	64 & PSSP & recruitment & Drought & MSE & 222.16 & 54.58 & 167.58 &  \\ 
+	65 & PSSP & recruitment & Irrigation & lppd & -60.18 & -62.12 & 1.94 & *** \\ 
+	66 & PSSP & recruitment & Irrigation & MSE & 20.90 & 44.57 & -23.67 & *** \\ 
+	67 & PSSP & survival & Control & lppd & -148.69 & -140.15 & -8.54 &  \\ 
+	68 & PSSP & survival & Control & MSE & 0.10 & 0.09 & 0.01 &  \\ 
+	69 & PSSP & survival & Drought & lppd & -83.26 & -78.09 & -5.17 &  \\ 
+	70 & PSSP & survival & Drought & MSE & 0.11 & 0.11 & 0.01 &  \\ 
+	71 & PSSP & survival & Irrigation & lppd & -100.52 & -89.02 & -11.50 &  \\ 
+	72 & PSSP & survival & Irrigation & MSE & 0.13 & 0.11 & 0.01 &  \\ 
+	\hline
+	\caption{Comparison of model predictions from climate model and year effects model for each species and vital rate and treatment.
+Two prediction scores are reported, MSE and lppd.
+Lower MSE indicates improved predictions whereas higher lppd indicates improved predictions.
+Instances where the climate model outperformed the random year effects model are marked with "***" in the last column.
+ARTR = \textit{A.
+tripartita}, HECO = \textit{H.
+comata}, POSE = \textit{P.
+secunda}, PSSP = \textit{P.
+spicata}.}
+	\label{table:treatmentPreds}
+\end{longtable}
 
 \clearpage
-
+\newpage
+\clearpage
+\newpage
 \section{Additional Figures}
 
+\begin{figure}[!htbp]
+	\centering
+	\includegraphics[width=1\textwidth]{pred_v_obs_treatment_ARTR_growth}
+	\caption{Comparison of treatment effects predicted and observed for \textit{A.
+tripartita} growth.
+Upper figure shows each parameter estimate separately while the lower figure shows the effect of treatment as a function of plant size.
+The observed treatment effects come from the treatment model fitted to all the data including the five years of the experiment.
+The predicted parameters come from the climate model fitted to all years of observational data but do not include the five years of the experiment (2011 to 2016).
+We assessed a parameter as significant when the 95\% Bayesian credible intervals did not overlap zero.}
+	\label{fig:parPredARTRGrowth}
+\end{figure}
+
+\begin{figure}[!htbp]
+	\centering
+	\includegraphics[width=1\textwidth]{pred_v_obs_treatment_HECO_growth}
+	\caption{Comparison of treatment effects predicted and observed for \textit{H.
+comata} growth.
+The observed treatment effects come from the treatment model fitted to all the data including the five years of the experiment.
+The predicted parameters come from the climate model fitted to all years of observational data but do not include the five years of the experiment (2011 to 2016).
+We assessed a parameter as significant when the 95\% Bayesian credible intervals did not overlap zero.}
+	\label{fig:parPredHECOGrowth}
+\end{figure}
+
+\begin{figure}[!htbp]
+	\centering
+	\includegraphics[width=1\textwidth]{pred_v_obs_treatment_POSE_growth}
+	\caption{Comparison of treatment effects predicted and observed for \textit{P.
+secunda} growth.
+Upper figure shows each parameter estimate separately while the lower figure shows the effect of treatment as a function of plant size.
+The observed treatment effects come from the treatment model fitted to all the data including the five years of the experiment.
+The predicted parameters come from the climate model fitted to all years of observational data but do not include the five years of the experiment (2011 to 2016).
+We assessed a parameter as significant when the 95\% Bayesian credible intervals did not overlap zero.}
+	\label{fig:parPredPOSEGrowth}
+\end{figure}
+
+\begin{figure}[!htbp]
+	\centering
+	\includegraphics[width=1\textwidth]{pred_v_obs_treatment_PSSP_growth}
+	\caption{Comparison of treatment effects predicted and observed for \textit{P.
+spicata} growth.
+The observed treatment effects come from the treatment model fitted to all the data including the five years of the experiment.
+The predicted parameters come from the climate model fitted to all years of observational data but do not include the five years of the experiment (2011 to 2016).
+We assessed a parameter as significant when the 95\% Bayesian credible intervals did not overlap zero.}
+	\label{fig:parPredPSSPGrowth}
+\end{figure}
+
+\begin{figure}[!htbp]
+	\centering
+	\includegraphics[width=1\textwidth]{pred_v_obs_treatment_ARTR_survival}
+	\caption{Comparison of treatment effects predicted and observed for \textit{A.
+survival} survival.
+The observed treatment effects come from the treatment model fitted to all the data including the five years of the experiment.
+The predicted parameters come from the climate model fitted to all years of observational data but do not include the five years of the experiment (2011 to 2016).
+We assessed a parameter as significant when the 95\% Bayesian credible intervals did not overlap zero.}
+	\label{fig:parPredARTRSurvival}
+\end{figure}
+
+\begin{figure}[!htbp]
+	\centering
+	\includegraphics[width=1\textwidth]{pred_v_obs_treatment_HECO_survival}
+	\caption{Comparison of treatment effects predicted and observed for \textit{H.
+comata} survival.
+The observed treatment effects come from the treatment model fitted to all the data including the five years of the experiment.
+The predicted parameters come from the climate model fitted to all years of observational data but do not include the five years of the experiment (2011 to 2016).
+We assessed a parameter as significant when the 95\% Bayesian credible intervals did not overlap zero.}
+	\label{fig:parPredHECOSurvival}
+\end{figure}
+
+\begin{figure}[!htbp]
+	\centering
+	\includegraphics[width=1\textwidth]{pred_v_obs_treatment_POSE_survival}
+	\caption{Comparison of treatment effects predicted and observed for \textit{P.
+secunda} survival.
+Upper figure shows each parameter estimate separately while the lower figure shows the effect of treatment as a function of plant size.
+The observed treatment effects come from the treatment model fitted to all the data including the five years of the experiment.
+The predicted parameters come from the climate model fitted to all years of observational data but do not include the five years of the experiment (2011 to 2016).
+We assessed a parameter as significant when the 95\% Bayesian credible intervals did not overlap zero.}
+	\label{fig:parPredPOSESurvival}
+\end{figure}
+
+\begin{figure}[!htbp]
+	\centering
+	\includegraphics[width=1\textwidth]{pred_v_obs_treatment_PSSP_survival}
+	\caption{Comparison of treatment effects predicted and observed for \textit{P.
+spicata} survival.
+The observed treatment effects come from the treatment model fitted to all the data including the five years of the experiment.
+The predicted parameters come from the climate model fitted to all years of observational data but do not include the five years of the experiment (2011 to 2016).
+We assessed a parameter as significant when the 95\% Bayesian credible intervals did not overlap zero.}
+	\label{fig:parPredPSSPSurvival}
+\end{figure}
+
 \end{document}

### 2016-12-21 precip/R/climate/plot_air_temp_diffs.R

@@ -1,0 +1,46 @@
+
+rm(list = ls()) 
+
+library( ggplot2 ) 
+library(tidyr)
+library(dplyr)
+library(lme4)
+library(zoo)
+
+df <- readRDS('~/driversdata/data/idaho_modern/soil_moisture_data/data/processed_data/decagon_data_with_station_data.RDS')
+load('analysis/figure_scripts/my_plotting_theme.Rdata')
+# 
+df <- df %>% mutate( v = ifelse(measure == 'VWC', v*100, v)) # convert to percent 
+df$year <- as.numeric( strftime( as.Date( df$simple_date), '%Y' ) )
+# summarize treatment differences:  -----------------------------------------------------------------------------------
+
+plot_T_vals <- df %>% 
+  filter( !is.na(v), bad_values == 0, measure == 'C', depth_label != '25 cm deep') %>%
+  group_by(Treatment_label, season_label, tod, depth_label) %>% 
+  summarise( avg = mean(v), stddev = sd(v), max = max(v), min = min(v), n = n(), ci = 1.96*(stddev/sqrt(n)), uci = avg + ci, lci = avg - ci ) 
+
+
+air_T_diffs <- 
+  df %>% 
+  filter( measure == 'C', depth == 'air temperature', stat == 'raw', bad_values == 0) %>% 
+  select( PrecipGroup, simple_date, season, tod, Treatment, v ) %>% 
+  group_by( PrecipGroup, Treatment, season, simple_date, tod ) %>%
+  summarise( avg_T = mean(v, na.rm = TRUE), n = n()) %>% 
+  filter( !is.na(avg_T), n == 6)  
+
+
+ggplot(subset( air_T_diffs, PrecipGroup == 3), aes( x = tod, y = avg_T, fill = Treatment )) + 
+  geom_boxplot( ) + 
+  facet_grid(  ~ season )+ 
+  scale_fill_manual(values = my_colors[2:4])
+
+
+ggplot(subset( air_T_diffs, PrecipGroup != 3), aes( x = tod, y = avg_T, fill = Treatment )) + 
+  geom_boxplot( ) + 
+  facet_grid(  ~ season )+ 
+  scale_fill_manual(values = my_colors[2:4])
+
+
[…]

### 2016-12-21 precip/analysis/figure_scripts/plot_one_step_ahead_cover.R

@@ -68,7 +68,7 @@ generate_cover_predictions <- function( spp, model ) {
   predicted_cover$area <- predicted_cover$rec_area + predicted_cover$total_size
   
   predicted_cover$model <- model
-  
+  predicted_cover$species <- spp 
   saveRDS( predicted_cover, paste0( 'output/ibm/simulations/', spp, '_one_step_ahead_', model, '_model_quadrat_cover.RDS'))
   
   return(predicted_cover)
@@ -89,8 +89,7 @@ ylims <- list( c(0,25), c(0,4), c(0,4), c(0,4))
 names(ylims) <- species_list
 iter <- expand.grid( species = species_list, model = model_list )
 
-for( i in 1:nrow(iter) ) { 
-  
+for( i in 7:nrow(iter) ) { 
   spp <- as.character( iter$species[i])
   model <- as.character( iter$model[i] )
   
@@ -143,6 +142,8 @@ for( i in 1:nrow(iter) ) {
   qarea <- merge( qarea, qarea2[, c('year', 'quad', 'Treatment', 'area')], by = c('quad', 'year', 'Treatment' ), all.x = T, all.y = T)
   qarea$area <- ifelse( is.na( qarea$area.x ), qarea$area.y, qarea$area.x)  # fill in missing area for second year 
   
+  qarea$species <- spp 
+  
   write.csv(qarea, paste0( 'output/ibm/simulations/', spp, '_observed_cover_per_quadrat.csv'), row.names = F)
   
   # merge observed and predicted cover ---------------------------------------------------
@@ -173,7 +174,7 @@ for( i in 1:nrow(iter) ) {
   
   plot_df <- 
     merge( observed_cover, predicted_cover, all.x = T) %>% 
-    mutate( predicted = ifelse(is.na(predicted), observed, predicted))
+    mutate( predicted = ifelse(is.na(predicted), observed, predicted)) ### fill in predicted cover with observed for first year in series 
   
   plot_df$Period <- ifelse( plot_df$year < 2011 , 'Historical', 'Modern')
   

### 2016-12-21 precip/analysis/figure_scripts/plot_predicted_and_observed.R

@@ -1,81 +1,0 @@
-rm(list =ls())
-library(stringr)
-library(dplyr)
-library(tidyr)
-library(ggplot2)
-library(boot)
-library(rstan)
-# functions ---------------------------------------------------------------------------- 
-
-#
-load('analysis/figure_scripts/my_plotting_theme.Rdata')
-
-setwd('~/Documents/ExperimentTests/precip/')
-
-years <- expand.grid(year = 1925:2017, Treatment = c(1:3), stat = c('true_cov', 'pred_cover'))
-years$Period[ years$year > 2010 ] <- 'Modern'
-years$Period[ years$year <= 2010 ] <- 'Historical'
-
-# output lists 
-survives <- list(NA)
-size     <- list(NA)
-recruits <- list(NA)
-
-species_list <- c('ARTR', 'HECO', 'POSE', 'PSSP')
-
-i = 1
-
-for( i in 1:4) {  
-  spp   <- species_list[i]  
-  
-  sdl   <- readRDS('data/temp_data/modified_survival_data_lists_for_stan.RDS')[[i]]
-  rdl   <- readRDS('data/temp_data/modified_recruitment_data_lists_for_stan.RDS')[[i]]
-  gdl   <- readRDS('data/temp_data/modified_growth_data_lists_for_stan.RDS')[[i]]
-  
-  m <- dir('output/stan_fits', paste0( spp, '.*_climate_fit.RDS'), full.names = TRUE)
-  m1 <- dir('output/stan_fits', paste0( spp, '.*year_effects_fit.RDS'), full.names = TRUE)
-  
-  sdf <- data.frame(sdl[ c('treathold', 'Yhold', 'gidhold', 'yearhold')] )
-  rdf <- data.frame(rdl[ c('treathold', 'Yhold', 'gidhold', 'yearhold')] )
-  gdf <- data.frame(gdl[ c('treathold', 'Yhold', 'gidhold', 'yearhold')] )
-  
-  sdf$climate <- summary(readRDS(m[3]), c('muhat'))$summary[, 1]
-  rdf$climate <- summary(readRDS(m[2]), c('lambda_pred'))$summary[, 1]
-  gdf$climate <- summary(readRDS(m[1]), c('muhat'))$summary[, 1]
-  
-  sdf$year_effects <- summary(readRDS(m1[3]), c('muhat'))$summary[, 1]
[…]

### 2016-12-21 precip/analysis/figure_scripts/plot_predicted_vs_observed_cover_changes.R

@@ -16,7 +16,7 @@ grd2 <- grd
 grd2$model <- 'year_effects'
 grd <- rbind(grd,grd2)
 i = 1
-out <- labels <- list(NA)
+out <- labels <- stats <-  list(NA)
 
 for(i in 1:nrow(grd)){ 
   
@@ -64,19 +64,22 @@ for(i in 1:nrow(grd)){
   # get mean squared error ---- 
   df <- merge(obs_pgr, pred_pgr)
   
-  df$Period <- ifelse(df$year > 2010, 'Modern', 'Historical')
-  T2 <- as.character( df$Treatment)
-  T2 <- ifelse(df$Period == 'Historical' , 'Historical', T2)
-  df$Treatment <- T2
-  df$Treatment <- factor( df$Treatment, levels = c('Historical', 'Control', 'Drought', 'Irrigation'), ordered = T)
-  df$Treatment <- factor( df$Treatment, labels = c('Training Data', 'Control', 'Drought', 'Irrigation'))
+  df$Period <- ifelse(df$year > 2011, 'Modern', 'Historical')
   
   df <- 
     df %>% 
     filter( obs*100/10000 > cover_threshholds, 
             pred*100/10000 > cover_threshholds )  # only use plots with greater than cover threshold 
   
-  allcombos <- expand.grid(Treatment = c('Training Data', 'Control', 'Drought', 'Irrigation'), year = c(1925:2016))
+  allcombos <- expand.grid(Treatment = c('Control', 'Drought', 'Irrigation'), year = c(1925:2016))
+  
+  df <- subset( df , Period == 'Modern' ) # confine it to the experimental years only 
+  
+  overall_MSE <- 
+    df %>% ungroup %>% summarise( MSE = mean((observed-predicted)^2))
+  
+  overall_cor <- 
+    df %>% ungroup %>% summarise( cor = cor(predicted, observed))
   
   MSE <- 
     df %>% 
@@ -87,10 +90,9 @@ for(i in 1:nrow(grd)){
     df %>% 
     group_by( Treatment ) %>% 
     summarise( cor = cor(predicted, observed))
-  
+
   xlim <- max( df$predicted )
   ylim <- min(  df$observed)
-  
   
   label_df <- merge(cor, MSE)
   
@@ -103,7 +105,6 @@ for(i in 1:nrow(grd)){
   
   df <- merge(allcombos, df, all.x = T)
   
-  
   pts <- 
[…]

### 2016-12-21 precip/analysis/figure_scripts/plot_significant_effect_comparison.R

@@ -41,6 +41,12 @@ all_effects <-
   mutate( type = ifelse(type == 'TRUE_FALSE' | type == 'FALSE_TRUE', 'mixed', type )) %>% 
   mutate( sig_label = ifelse(type == 'TRUE_TRUE', 'significant', 'not significant'))
 
+overall_cor <- all_effects %>% ungroup %>% summarise( cor = cor(mu.x, mu.y))
+overall_cor
+
+overall_cor <- all_effects  %>% group_by(Treatment) %>% summarise( cor = cor(mu.x, mu.y ))
+overall_cor
+
 cors <- all_effects %>% group_by(par) %>% summarise( cor = cor(mu.x, mu.y))
 cors$x.pos <- c(1.5,1.5) 
 cors$y.pos <- c(1.5,1.5)
@@ -49,6 +55,7 @@ cors$label <- paste0( 'r=', round( cors$cor, 2))
 all_effects <- unique(all_effects)
 
 nrow( all_effects )
+
 
 p1 <- 
   ggplot(all_effects, aes( x = mu.x, y = mu.y, color = Treatment, shape = sig_label,  group = 1 )  ) + 
@@ -70,7 +77,8 @@ p1 <-
 p1 <- p1 + geom_text( data = cors, aes( x  = x.pos, y = y.pos , label = label , color = NULL, shape = NULL), show.legend = F) 
 p1
 
-p1 + geom_text( data = subset( all_effects, type == 'TRUE_TRUE' & species == 'POSE' & vital_rate == 'recruitment' & Treatment == 'Irrigation'), 
+
+p1 <- p1 + geom_text( data = subset( all_effects, type == 'TRUE_TRUE' & species == 'POSE' & vital_rate == 'recruitment' & Treatment == 'Irrigation'), 
                 aes( label = paste(species, vital_rate, sep = '\n' )) , adj = -0.1, nudge_y = -0.2, show.legend = F) 
 
 p <- 
@@ -86,12 +94,12 @@ p <-
 
 g <- all_effects %>% group_by(type ) %>% do( p = p %+% . + ggtitle(.$type) )
 
-p1
+g$p
 
 
 
-pdf( 'figures/significant_parameter_comparison.pdf' , height = 6, width = 8)
+png( 'figures/parameter_predictions.png' , height = 6, width = 8 , res = 300 , units = 'in')
 
-print( p1 )
+print( p1 + labs( color = '' )) 
 
 dev.off()

### 2016-12-21 precip/analysis/get_MSE_scores.R

@@ -1,0 +1,85 @@
+rm(list =ls())
+library(stringr)
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+library(boot)
+library(rstan)
+# functions ---------------------------------------------------------------------------- 
+
+#
+load('analysis/figure_scripts/my_plotting_theme.Rdata')
+
+setwd('~/Documents/ExperimentTests/precip/')
+
+years <- expand.grid(year = 1925:2017, Treatment = c(1:3), stat = c('true_cov', 'pred_cover'))
+years$Period[ years$year > 2010 ] <- 'Modern'
+years$Period[ years$year <= 2010 ] <- 'Historical'
+
+# output lists 
+survives <- list(NA)
+size     <- list(NA)
+recruits <- list(NA)
+
+species_list <- c('ARTR', 'HECO', 'POSE', 'PSSP')
+
+i = 1
+out <- out2 <- list()
+
+for( i in 1:4) {  
+  spp   <- species_list[i]  
+  
+  sdl   <- readRDS('data/temp_data/modified_survival_data_lists_for_stan.RDS')[[i]]
+  rdl   <- readRDS('data/temp_data/modified_recruitment_data_lists_for_stan.RDS')[[i]]
+  gdl   <- readRDS('data/temp_data/modified_growth_data_lists_for_stan.RDS')[[i]]
+  
+  m <- dir('output/stan_fits', paste0( spp, '.*_climate_fit.RDS'), full.names = TRUE)
+  m1 <- dir('output/stan_fits', paste0( spp, '.*year_effects_fit.RDS'), full.names = TRUE)
+  
+  sdf <- data.frame(sdl[ c('treathold', 'Yhold', 'gidhold', 'yearhold')] )
+  rdf <- data.frame(rdl[ c('treathold', 'Yhold', 'gidhold', 'yearhold')] )
+  gdf <- data.frame(gdl[ c('treathold', 'Yhold', 'gidhold', 'yearhold')] )
+  
+  sdf$climate <- summary(readRDS(m[3]), c('muhat'))$summary[, 1]
+  rdf$climate <- summary(readRDS(m[2]), c('lambda_pred'))$summary[, 1]
+  gdf$climate <- summary(readRDS(m[1]), c('muhat'))$summary[, 1]
+  
+  sdf$year_effects <- summary(readRDS(m1[3]), c('muhat'))$summary[, 1]
[…]

### 2016-12-21 precip/analysis/make_prediction_score_tables.R

@@ -1,0 +1,88 @@
+# format prediction tables 
+alllppd <- read.csv('output/overall_lppd.csv')
+allMSE <- read.csv('output/all_MSE_scores.csv')
+
+allMSE$score <- 'MSE'
+alllppd$score <- 'lppd'
+
+alllppd$model <- factor(alllppd$model, labels = c('climate model', 'year effects model'))
+allMSE$model <- factor(allMSE$model, labels = c('climate model', 'year effects model'))
+
+all_scores <- rbind( allMSE %>% spread(model, MSE), alllppd %>% spread(model, lppd) )
+all_scores$diff <- all_scores$`climate model` - all_scores$`year effects model`
+
+overall_score_table <- 
+  all_scores %>% 
+  mutate( improved = ifelse( diff < 0 & score == 'MSE', '***', '')) %>% 
+  mutate( improved = ifelse( diff > 0 & score == 'lppd', '***', improved)) %>% 
+  arrange( species, vital_rate, score )
+
+
+oaxt <- xtable(overall_score_table, caption = 'Comparison of model predictions from climate model and year effects model for each species and vital rate.  Two prediction scores are reported, MSE and lppd. Lower MSE indicates improved predictions whereas higher lppd indicates improved predictions.  Instances where the climate model outperformed the random year effects model are marked with "***" in the last column. ARTR = \\textit{A. tripartita}, HECO = \\textit{H. comata}, POSE = \\textit{P. secunda}, PSSP = \\textit{P. spicata}.' , 
+label = 'table:overallPreds')
+
+print(oaxt, 'output/results_tables/overall_prediction_score.text', type = 'latex', include.rownames = F)
+
+# ---------- predictions by treatment level ------------------------------------------------------- # 
+
+tlppd <- read.csv('output/treatment_lppd.csv')
+tMSE <- read.csv('output/treatment_MSE_scores.csv')
+
+tlppd <- tlppd %>% dplyr::select(-climate_model_performance) %>% gather( model, lppd, climate:year)
+
+tMSE$score <- 'MSE'
+tlppd$score <- 'lppd'
+
+
+tlppd$model <- factor(tlppd$model, labels = c('climate model', 'year effects model'))
+tMSE$model <- factor(tMSE$model, labels = c('climate model', 'year effects model'))
+
+
+treatment_scores <- rbind( tMSE %>% spread(model, MSE), tlppd %>% spread(model, lppd) )
[…]

### 2016-12-22 precip/manuscript/precip_ms.tex
edits to the manuscript abstract intro and methods

@@
 \author[a]{Andrew Tredennick}
+\author[b]{Caitlin Andrews}
+\author[b]{John Bradford}
 \affil[a]{Department of Wildland Resources and the Ecology Center, Utah State University, Logan, Utah}
+\affil[b]{US Geological Survey, Southwest Biological Science Center, Flagstaff, Arizona}
 
@@
 
-\smallskip \textbf{Authorship statement:} ARK and PBA designed the experiment and supervised data collection. AT helped with the data analysis.
+\smallskip \textbf{Authorship statement:} ARK and PBA designed the experiment and supervised data collection. AT helped with the data analysis. JB and CA helped with the soil moisture model. 
 \smallskip
@@
 \begin{enumerate}
-\item Climate is an important driver of population ecology however there have been few tests of whether observational data that links population performance with climate variation can be used to predict the responses of populations to experimentally imposed climate conditions.
+\item Climate is an important driver of population ecology however there have been few tests of whether observational data correlating population performance with climate variation can be used to predict future responses of populations to experimentally imposed climate.
 \item Using longterm historical observational data from a sagebrush steppe plant community, previous research has reported a wide variety of climate effects on four dominant plant species.
-We tested whether statistical models based on these observational data could be used to predict how these species respond to a five year drought and irrigation experiment conducted from 2011 to 2016.
-\item We established sixteen new plots at the same field cite as the original historical data and assigned each to either drought or irrigation treatments imposed using rainfall shelters and automatic sprinklers.
-The original plots were also monitored and served as ambient climate controls.
-In order to describe the demographic effects of the climate manipulations, we used all data, observational and experimental, to fit a set of statistical models describing each species survival, growth and recruitment in response to local competitive interactions and the effects of the climate manipulation experiment.
-Next we fit another set of models to only the observational data that used seasonal soil moisture covariates to describe annual variation in the vital rates.
-Finally, we used the observation-based models to predict the effects of the experimental treatments on each species vital rates.
-We also used an individual based population model to predict one year ahead changes in population size in the experimental plots.
-\item The experimental drought and irrigation treatments successfully lowered and increased soil moisture respectively.
-Over the course of the experiment, average plot cover of \textit{Hesperostipa comata} and \textit{Pseudoroegneria spicata} declined significantly, while cover of \textit{Poa secunda} showed a trend towards increase in the irrigated plots.
-At the level of individual vital rates, experimental drought reduced growth and survival of \textit{Hesperostipa comata} and \textit{Poa secunda} and the survival of \textit{P.
-spicata}, while drought increased the growth of small \textit{A.
+We tested whether these climate effects observed in observational data were useful for predicting each species response to a drought and irrigation experiment.
+\item We used rainfall shelters and automatic sprinklers to manipulate the soil moisture in sixteen new plots established at the same field site as the original historical plots.
+We also re-monitored the original plots as controls.
+After running the experiment for five years, we analyzed how the growth, survival and recruitment of each species responded to the experimental drought and irrigation treatments.
+\item In order to test whether we could predict the observed responses to the experiment we fit two types of models to the observational data collected prior to the experiment: first we fit a no climate model that included the effects of plant size and competition on plant performance in each plot in each year; next we fit a climate model that in addition to size and competition also included the effects of three seasonal soil moisture variables.
+Next we used these models to predict the survival and growth of each individual plant, and the total recruitment in each plot during the course of the experiment.
+We also used an individual based population model parameterized with the vital rate predictions to predict one step ahead changes in population size in the experimental plots.
+We then compared the predictions made by each of these models to the observed experimental responses and compared the performance of the no climate model to the climate model.
+\item Over the course of the experiment, average cover of the perennial grasses, \textit{Hesperostipa comata} and \textit{Pseudoroegneria spicata} declined significantly in the drought treatment.
+At the level of individual vital rates, experimental drought reduced the survival of \textit{Hesperostipa comata} and \textit{P.
+spicata}  and the growth of the grass \textit{Poa secunda}.
+In contrast, drought increased the growth of the shrub \textit{A.
 tripartita}.
-\item Soil moisture parameters improved model predictions for six out of the twelve models we examined.
-Moreover, we observed strong positive correlations between the predicted treatment effects and the treatment effects we observed.
-\item At the population-level, including the effects of soil moisture improved the cover predictions made by individual based models for \textit{P. secunda} and \textit{P. spicata}.   
-\item \emph{Synthesis}: Observational climate data did hold valuable information for predicting species' responses to this climate change experiment.
-We were encouraged that treatment responses were often in the right direction to predicted responses even when the effect were not signficant.
-We generally were better able to predict species responses to the drought treatment than to the control or the irrigation treatment.
-This suggests that soil moisture is an important factor in predicting the population dynamics of these species but only when water is truly limiting.
+\item The climate better predicted the response to the experiment than the no climate model in six out of the twelve cases we examined.
+Across all species and vital rates, there was a strong positive correlation between the observed effects of the experiment and the effects predicted by the climate model fit to the observational data.
+At the population-level, the climate model made better predictions of the changes in species cover than the no climate model for two species: \textit{P.
+secunda} and \textit{P.
+spicata}.
+Moreover, in the drought treatments the climate model made better predictions for the changes in cover of all four species.
+\item \emph{Synthesis}: Observational climate data held valuable information for predicting species' responses to a modern climate change experiment in this ecosystem.
+Treatment responses often matched the direction of predicted responses even when the effects were not significant.
+We were better able to predict species' responses to the drought treatment than to the control and irrigation treatments.
+Suggesting that soil moisture is an important factor in predicting the population dynamics of these species but only when water is truly limiting.
 
@@
 
-Climate is widely considered one of the most powerful external forces driving changes in species abundance across space and time \citep{teller_linking_2016}.
-The effects of climate on populations and ecosystems are most apparent at the largest scales in space and time: climate determines the distribution of ecosystems, treelines and the the range limits of many species \citep{}, while the recent historical and the paleoecological record shows that long-term climate change leads to changes in species range limits \citep{parmesan, davis}.
-Understanding and predicting the effects of climate on populations is an increasingly important goal if we are to understand and predict the effects of climate change on earth's ecosystems.
-
-Unfortunately, for many species it is difficult to determine how annual climate variation affects populations and demographic rates \citep{teller_linking_2016, ehrlen_advancing_2016}.
-Observational data on species performance across years with varying climate can provide some information on how climate might affect population abundance \citep{dalgleish_climate_2010,ehrlen_advancing_2016}.
-However, many years of data are needed to reliably detect climate effects, especially when annual variation in demographic rates is high \citep{teller_linking_2016,gerber_optimal_2015}.
-Generating predictions from observational historical data for the future novel conditions created by climate change is fraught with risk.
-Climate change will not only change mean annual temperature and precipitation, but also affect annual variation in these measures, and possibly the covariance between them \citep{}.
-Models based on the historical response of populations to annual climate variation, will therefore be extrapolating beyond the range of observed climate variation when they are used to generate predictions in the future \citep{}.
-If models fit to historical data can be used to accurately predict the effects of experimental climate manipulations, especially manipulations that generate extreme variation in climate, it would be strong confirmation that the climate effects they describe are real and will hold in the future \cite{adler_can_the_past_2014}.
-
-Many plant species occur across a wide range of climates \cite{} and individual plants in many environments must tolerate large fluctuations in seasonal temperature and soil moisture.
-Unlike most animals, plants must endure these conditions in place and often become dormant in less optimal seasons \textit{}.
-At first pass, these observation would seem to indicate that annual variation in climate should have little effect on plant population performance.
-
-Nevertheless, there is abundant evidence that plant performance shows high year to year variation, both at the level of individual growth, survival and reproduction and in terms of total population abundancend often this variation c.
-In many regions interannual variation in precipitation can be directly linked to variation in net primary productivity \citep{knapp and smith, Hsu and Adler}.
-The growth rates of trees are also often tightly linked to annual precipitation, so much so that annual growth rings in their stems can serve as accurate records of historical climate thousands of years before the present \citep{yang}.
-Likewise, many smaller shrub and sub-shrub species show strong variation in growth in response to climate that is recorded in their tissues \citep{Franklin, Srur}.
-Among annual plants in desert ecosystems, germination and reproductive output is tightly linked to interannual precipitation variation \citep{venables}.
-
-Linking annual climate variation to demographic performance in plants is a high priority for building population models for plants that can predict their future response to climate change \citep{crone_ability_2013,ehrlen_advancing_2016}.
-However, despite the clear signs that climate drives net primary productivity at the ecosystem level and the variation in individual growth rate in trees and other woody species, as well as the germination and fecundity of annual plants, there are few studies that have connected the effects of interannual climate variation to population models for plants, and fewer still that have tested whether these population models can be used to accurately predict the future responses of plants to short term climate variation.
-Adler et al.
-\citep{adler byrn and leiker} showed that population models based on historically observed correlations between plant population growth rates and precipitation did have some predictive power in describing species response to a short-term climate manipulation in a North American grassland.
+Climate is one of the most powerful drivers of changes in species abundance across space and time \citep{post_synchronization_2002,davis_range_2001,walther_ecological_2002}.
+The effects of climate on populations and ecosystems are most apparent at the largest scales: climate determines the distribution of ecosystems \citep{whittaker_communities_1975}, treelines \citep{korner_alpine_2012} and the the range limits of many species \citep{parmesan_globally_2003,davis_range_2001}.
+Understanding and predicting the effects of climate on populations is an increasingly important goal if we are to anticipate the effects of climate change on earth's ecosystems \citep{tredennick_we_2016,petchey_ecological_2015,ehrlen_advancing_2016,teller_linking_2016}.
+
+Ecologists often resort to one of two methods for predicting the effects of future climate change on populations and communities: they may use experiments to manipulate aspects of climate directly and observe the response of populations \citep{elmendorf_experiment_2015,knapp_pushing_2016,compagnoni_warming_2014}; or they may use longterm observational data on species performance and abundance collected over many years and relate this to ambient annual variation in climate  \citep{koons_climate_2012,lunn_demography_2016,dalgleish_climate_2010,jenouvrier_demographic_2009}.
+The strength of the experimental approach is in the stronger inference that comes from manipulating some aspects of climate while controlling for other factors; for instance knowing that loss of snow cover, and not necessarily changes in soil temperature or moisture are the factors causing a species performance to change with warming \citep{compagnoni_warming_2014}.
+It also allows for the creation of conditions that may be more extreme than those observed historically \cite{knapp_pushing_2016} but are possible in the future.
+However, it is often expensive to control climate at even the smallest scales, and larger scale climate manipulation is often impossible.
+Moreover, experimental manipulation can come with artifacts that may make them less then ideal models for understanding and predicting the effects of future variation in climate \citep{wolkovich_warming_2012}.
+
+Using observational data to predict the effects of climate variation on populations has many advantages over using experiments.
+The primary one being the ever increasing availability of long-term ecological data and ever increasingly detailed climate data.
+Observational studies may also be the only way to study the effects of climate on large and or migratory species, for which it would be difficult to manipulate climate \citep{koons_climate_2012,jenouvrier_demographic_2009,aubry_climate_2013}.
+However, there are many potential drawbacks to using essentially correlative relationships between species performance and climate to predict future species' responses.
+First, many years of data are needed to reliably detect climate effects, especially when annual variation in demographic rates is high \citep{teller_linking_2016,gerber_optimal_2015}.
+\citep{teller_linking_2016} estimate that even cutting edge statistical approaches for fitting relationships between climate and species performance require at least 20-25 years of independent climate observations before they perform well.
+Moreover, even strong correlations between species performance and the climate covariates we choose to include in our models may not be due to direct causation and may not hold up in future years of observation \citep{hilborn_correlation_2016}.
+
+And yet even overcoming this challenge, predicting the effects of climate change would face yet another challenge: in many parts of the world, climate change will lead to mean annual temperatures that will be above the range of historical variation.
+If species performance responds non-linearly to temperature, we may be unable to predict how species respond to temperatures outside the range of observed variation \citep{doak_demographic_2010}.
+In addition, climate change will not just change the average climate, but is also likely to increase the variance in precipitation and lead to extreme events, which will have their own consequences independent of changes in climate averages \citep{drake_population_2005,jentsch_new_2007}.
+Any models based on observations drawn from the historical range of variation will therefore be extrapolating beyond both the range of observed averages and variance when used to predict the future \citep{williams_novel_2007}.
+Demonstrating that models fit to historical data can accurately predict the effects of experimental climate manipulations, especially manipulations that generate extreme conditions, would be strong confirmation that the climate effects they describe are not spurious and will hold even in the novel conditions of the future \cite{adler_can_2013}.
+
+There is abundant evidence that plant performance shows high year to year variation, both at the level of individual growth, survival and reproduction and total population abundance.
+Precipitation is often especially important in driving interannual variation in net primary productivity among terrestrial plants \citep{knapp_variation_2001,hsu_anticipating_2014}.
+And the growth rates of trees are also often so tightly linked to annual precipitation that growth rings in their stems can serve as accurate records of historical climate variation thousands of years before the present \citep{yang_3500-year_2014}.
+Among smaller plants precipitation clearly affects the growth rates of small shrub and sub-shrub species \citep{srur_annual_2009,franklin_growth_2013}, and the germination and reproductive output of desert annuals is often tightly linked to annual precipitation \citep{venable_bet_2007}.
+
+Despite the clear signs that climate drives net primary productivity at the ecosystem level and the variation in individual growth rate in many plants, there have been relatively few studies that clearly link observed climate variation to species performance in population models \citep{ehrlen_advancing_2016}.
+Fewer still have tested whether the population-level effects of climate variation shown in observational data can actually be used to predict the responses of plant populations to future climate variation.
+\citep{adler_can_2013} showed that population models based on observed correlations between plant population growth rates and precipitation did have some predictive power in describing species response to a short-term climate manipulation in a North American grassland.
 Three species showed responses to experimentally imposed drought and irrigation that were well predicted by population models fitted to historical observations.
@@
 
-The demography and competitive interactions between three dominant grass species and a dominant shrub species at in a sagebrush steppe plant community at the US Sheep Experiment Station near Dubois, Idaho have been described in at least X different studies since 2005.
-X of these studies report significant effects of historical climate variation on the vital rates and overall population growth of these species.
-Both precipitation and temperature have been shown to have strong species-specific effects on this system.
-Although past studies used different statistical models and methods for variable selection they indicate X.
-This well-studied system offers the ideal opportunity to test whether statistical associations between annual climate variation and plant demography in long-term observational data can be used to predict the future responses of plant populations to climate change.
-
-In this study, we report how the four dominant plant species at the USSES responded to a five year drought and irrigation experiment and use the results to address two research questions: first, how much do the growth, recruitment and survival of our target species differ between the precipitation manipulation treatments?
-If our experiment does affect species vital rates we interpret that as strong evidence that variation in precipitation should have an effect on populations.
-Second, we test whether statistical models parameterized from observational data only can accurately predict each species response to the experimental precipitation manipulation?
-If models based on observational effectively capture the effects of climate on species performance, they should also predict the effects of precipitation treatments in the experiment.
+The demography of and competitive interactions between three dominant perennial bunchgrasses and a shrub in a sagebrush steppe plant community at the US Sheep Experiment Station near Dubois, Idaho have been described in at least seven different studies since 2010 \citep{adler_coexistence_2010,adler_forecasting_2012,adler_weak_2009,adler_weak_2016,chu_direct_2016,chu_large_2015,dalgleish_climate_2010}.
+And several of these studies report significant effects of climate variation on the vital rates and overall population growth rates of these species \citep{dalgleish_climate_2010,adler_forecasting_2012,adler_weak_2009,chu_direct_2016}.
+This well-studied system offers the ideal opportunity to test whether statistical associations between annual climate and plant demography in longterm observational data can be used to predict the responses of plant populations to experimental climate change experiments.
+
+In this study, we report how the four dominant plant species at the USSES respond to a five year drought and irrigation experiment and use the results to address two research questions: first, how much do the growth, recruitment and survival of our target species differ between the precipitation manipulation treatments?
+If our experiment does affect species vital rates we interpret that as strong evidence that changes in precipitation should have an effect on populations in the future.
+Second, can we predict each species' response to the experimental conditions based on how they respond to natural climate variation in the observational data?
+If models based on observational data can predict the response of species to this experiment this should give us more confidence in longterm population monitoring data to predict species response to climate change.
 
@@
 The U.S.
-Sheep Experiment Station (USSES) is located 9.6 km north of Dubois, Idaho (44.2\degree N, 112.1\degree W), 1500 m above sea level.
-During the period of data collection (1926 – 1957), mean annual precipitation was 270 mm and mean temperatures ranged from -8\degree C (January) to 21\degree C (July).
+Sheep Experiment Station (USSES) is located at Dubois, Idaho (44.2\degree N, 112.1\degree W), 1500 m above sea level.
+During the period of data collection (1926 – 2016), mean annual precipitation was 270 mm and mean temperatures ranged from -8\degree C (January) to 21\degree C (July).
 The vegetation is dominated by a shrub, \textit{Artemisia tripartita}, and three perennial C3 grasses: \textit{Pseudoroegneria spicata}, \textit{Hesperostipa comata}, and \textit{Poa secunda}.
@@
 All quadrats were located on similar topography and soils.
-In most years until 1957, all individual plants in each quadrat were mapped using a pantograph (Blaisdell 1958).
+In most years until 1957, all individual plants in each quadrat were mapped using a pantograph \citep{blaisdell_}.
 The historical data set is public and available online \citep{zachmann_mapped_2010}.
@@
 We mapped the quadrats in June, 2011 and then built the rainfall shelters and set-up the irrigation systems in the fall of 2011.
-We used a rain-out shelter and automatic irrigation design described in \citep{gherardi, yahdjian_rainout_2002}.
-Each rain-out shelter covered an area of 2.5 by 2 m and consisted of transparent acrylic shingles held up over the plot to channel 50\% of incoming rainfall off of the plot and into 75 l reservoirs.
+We used a rain-out shelter and automatic irrigation design described in \citep{gherardi_automated_2013}.
+Each rain-out shelter covered an area of 2.5 by 2 m and consisted of transparent acrylic shingles held up 1.5 to 1 m over the plot to channel 50\% of incoming rainfall off of the plot and into 75 l reservoirs.
 The collected water was pumped out of reservoirs and sprayed onto paired irrigation treatment plots.
@@
 Data were logged automatically every four hours.
-We augmented automatic monitoring of the climate in these plots with by taking direct measurements of soil moisture with a handheld EC-5 soil moisture sensor at six points around each plot on 6/6/2012, 4/29/2015, 5/7/2015, 6/9/2015 and 5/10/2016.
+We augmented automatic monitoring of the climate in these plots with direct measurements of soil moisture with a handheld EC-5 soil moisture sensor at six points around all 16 plots on 6/6/2012, 4/29/2015, 5/7/2015, 6/9/2015 and 5/10/2016.
 We analyzed these spot measurements for significant treatment effects on soil moisture using a linear mixed effects model with the \textit{lmer} package in \textit{R} \citep{bates}, with plot, plot group, and date as random effects in the model.
@@
 We expected that our precipitation manipulation experiment would affect plants by altering available soil moisture during the growing season.
-Because we do not have direct soil moisture measures for each year of observed plant cover in the historical record, we used the SOILWAT soil moisture model to estimate daily soil moisture at the USSES from 1925 to the present \citep{Parton 1978}.
+Because we do not have direct soil moisture measures for each year of observed plant cover in the historical record, we used the SOILWAT soil moisture model to estimate daily soil moisture at the USSES from 1925 to the present \citep{Parton_1978}.
 We used an enhanced version of soilwat that has recently been developed for use in semi-arid shrubland ecosystems \citep{Bradford}.
@@
 
+\subsection*{Overview of the analysis}
+
+Our analysis consists of two separate datasets and three different categories of vital rate models.
+We refer to the first dataset as the observational data.
+It consists of all the historical data collected from 1925 to 1957 as well as the modern data collected from the same plots from 2007 to 2010.
+These data record the response of plants in each plot to the ambient climate variation.
+We refer to the second dataset as the experimental data.
+It consists of the data collected from 2011 to 2016 from the 16 new experimentally manipulated plots, as well as from 14 of the original historical plots which serve as ambient climate controls.
+
+In order describe the effects of the experimental treatments on each vital rate, we fit "treatment" models.
+The treatment models included parameters for the effects of the drought and irrigation treatments on each vital rate.
+We fit these models to both datasets together, all the experimental and all the observational data.
+We did this because we wanted to focus our predictions on the effects of the experimental treatments on the vital rates, rather than any differences between the datasets in the effects of crowding and plant size on the vital rates.
+
+Next in order to test how well we could predict the responses in the experimental plots, we fit two classes of models to the observational dataset only.
+Thus when we used these models to predict the response of each species to the five years of experimental data we are generating predictions for out of sample data.
+We refer to the first type of model as the "no climate" models.
+These models include parameters for the effects of competition on each vital rate and the size dependence of survival and growth but they do not include climate or treatment effects.
+Lastly, we fit the "climate" models which include the effects of annual variation in soil moisture on each vital rate.
+The no climate model provides us a baseline by which to measure the accuracy of the predictions from the climate model.
+Because much of the variation in growth, survival and recruitment in this system can be explained by plant size and competition, we expect that these two models will make similar predictions for individual plant performance in the experiment.
+However, if the climate model makes more accurate predictions than the no climate model, this indicates that the climate parameters it includes contain useful information for prediction.
+
 \subsection*{Statistical models of vital rates}
 
-For each of the vital rates and each species we fit three separate models.
-First we fit a treatment model fit to all observations in the historical data as well as the contemporary experiment.
-This model included parameters estimating the effects of the drought and irrigation treatments on each species vital rate.
-This model was used to describe the basic results of the experiment.
-Next we fit two models to the historical observational data only (including the first four years of the modern data 2007 to 2010).
-We used these two models to generate predictions for the effects of the experiment.
-In the first prediction model, year to year variation in vital rates was treated as random effect.
-In the second prediction model, the "climate" model, we included paramaters that described the effects of year to year variation in seasonal soil moisture.
-These three models allow for two meaningful comparisons.
-First we can compare the predictions made by the two prediction models directly to the raw data in the experimental plots.
-In addition, we can compare the direction and magnitude of the soil moisture parameters estimated from the observational data by the climate model to the coefficients describing the treatment effects in the treatment model.
-
-All three versions of the models of the models above follow the same basic structure and are developed from previous work \citep{adler_coexistence_2010,chu_large_2015}.
+All three categories of models described above follow the same basic structure and differ only in how they treat climate and treatment effects and  \citep{adler_coexistence_2010,chu_large_2015}.
 We model the survival probability of an individual genet as a function of genet size, the neighborhood-scale crowding experienced by the genet from both conspecific and heterospecific genets, temporal variation among years, and permanent spatial variation among groups of quadrats (`group'; here means a set of nearby quadrats located within one pasture or grazing exclosure).
@@
 
-Formally, we modeled the survival probability, $S$, of genet $i$ in species $j$, group $g$, and removal treatmnt $h$, from time $t$ to $t+1$  as
+Formally, we modeled the survival probability, $S$, of genet $i$ in species $j$, group $g$, and from time $t$ to $t+1$  as
 \begin{equation}
-\mbox{logit}(S_{ijgh,t}) = \varphi_{jg}^S + \gamma_{j,t}^S  + \beta_{j,t}^S u_{ij,t} +  
+\mbox{logit}(S_{ijg,t}) = \varphi_{jg}^S + \gamma_{j,t}^S  + \beta_{j,t}^S u_{ij,t} +  
 \left \langle \boldsymbol{\omega}_{j,t}^S, \boldsymbol{W}_{ij,t} \right \rangle 
@@
 $\left \langle \boldsymbol{x, y} \right \rangle$ denotes the inner product of vectors $\boldsymbol{x}$ and $\boldsymbol{y}$, 
-calculated as \texttt{sum(x*y)} in R.
-
-To describe the treatment effects in the experiment a new term is added to the above model,
-\begin{equation}
- \boldsymbol{T}\chi_{j}^S
-\label{eqn:survT}
-\end{equation}
-where $\chi$ is a vector of treatment effect coefficients describing the effects of each experimental treatment $h$ on the survival rate, and $\boldsymbol{T}$ is a design matrix indicating the treatment level of each observation in the data.
-The design matrix also includes terms for the interaction between plant size $u$ and the treatment effects.
-These interaction terms allow the effect of each treatment to vary with plant size.
-
-In the climate model, the $\chi$ and $\xi$ terms are replaced with, 
-\begin{equation}
-\boldsymbol{C}\xi_{j}^S 
-\label{eqn:survC}
-\end{equation}
-where $\xi$ gives a vector of coefficients describing the effects of a set of soil moisture covariates $\boldsymbol{C}$ in treatment $h$ and year $t$ on the survival rate of species $j$.
-$\boldsymbol{C}$ is a vector of seasonal average soil moisture and can include interaction effects between plant size, $u$, and the soil moisture covariates that allow the effects of soil moisture to vary with plant size.
+calculated as \texttt{sum(x*y)} in R. This model is the no climate model for survival.
+
+In the treatment model, a new term is added to the above model, $\boldsymbol{T}\chi_{j}^S$ where $\chi$ is a vector of treatment effect coefficients for each experimental treatment level $h$ on the survival rate, and $\boldsymbol{T}$ is a design matrix indicating the treatment level of each observation in the data.
+The design matrix also includes terms for the interaction between plant size $u$ and the treatment effects which allow the effect of each treatment to vary with plant size.
+
+In the climate model, the above term is replaced with $\boldsymbol{C}\xi_{j}^S$, where $\xi$ gives a vector of coefficients describing the effects of a set of soil moisture covariates $\boldsymbol{C}$ in treatment $h$ and year $t$ on the survival rate of species $j$.
+$\boldsymbol{C}$ can include interaction effects between plant size, $u$, and the soil moisture covariates allowing the effects of soil moisture to vary with plant size.
 
@@
 \begin{equation}
-u_{ijgh,t+1} = \varphi_{jg}^G + \gamma_{j,t}^G + \chi_{jh}^G  + \beta_{j,t}^G u_{ij,t} + 
+u_{ijg,t+1} = \varphi_{jg}^G + \gamma_{j,t}^G + \chi_{jh}^G  + \beta_{j,t}^G u_{ij,t} + 
 \left \langle  \boldsymbol{\omega}_{j,t}^G, \boldsymbol{W}_{ij,t} \right \rangle + \varepsilon_{ij,t}^G .
@@
 \end{equation}
-To capture non-constant error variance in growth, we modeled the variance $\varepsilon$  about the growth curve (\ref{eqn:growReg})  as a nonlinear function of predicted genet size:
-\begin{equation}
-Var(\varepsilon_{ij,t}^G) = a \exp ^{(bu_{ij,t+1})} .
-\label{eqn:growVar}
-\end{equation}
-
-As in the survival regression above, specific specific parameters describing the treatment effects on growth are added in the treatment model, 
-\begin{equation}
-\boldsymbol{T}\chi_{j}^G 
-\label{eqn:growT}
-\end{equation}
-where $\chi$ is a treatment effect describing the effect of experimental treatment $h$ on growth, including treatment by size interactions.
-
-Similarly, in the climate model soil moisture influences the growth equation through these terms,  
-\begin{equation}
-\boldsymbol{C}\xi_{j}^G 
-\label{eqn:growC}
-\end{equation}
-where $\xi$ is a vector of coefficients describing the effects of soil moisture covariates in the matrix $\boldsymbol{C}$ for treatment $h$ and year $t$ on growth of species $j$.
+
+As in the survival regression above, parameters describing the treatment effects on growth are added in the treatment model, $\boldsymbol{T}\chi_{j}^G$, where $\chi$ is a treatment effect describing the effect of experimental treatment $h$ on growth, including treatment by size interactions.
+
+Similarly, in the climate model, the above term is replaced with $\boldsymbol{C}\xi_{j}^G$, where $\xi$ is a vector of coefficients describing the effects of soil moisture covariates in the matrix $\boldsymbol{C}$ for treatment $h$ and year $t$ on growth of species $j$.
 Again this can include interactions between soil moisture and plant size $u$.
@@
 We model the crowding experienced by a focal genet as a function of the distance to and size of neighbor genets.
-In previous work, we assumed that the decay of crowding with neighbor distance followed a Gaussian function \citep{chu_large_2015}, but here we use a data-driven approach \citep{teller_linking_2016, adler_in_prep}.
-We model the crowding experienced by genet $i$ of species $j$ from neighbors of species $m$ as the sum of neighbor areas across a set of concentric annuli, $k$, centered at the plant,
+These effects are well described in previous work \citep{teller_linking_2016,adler_weak_2016}.
+Briefly, we model the crowding experienced by genet $i$ of species $j$ from neighbors of species $m$ as the sum of neighbor areas across a set of concentric annuli, $k$, centered at the plant,
 \begin{equation}
@@
 Otherwise, we let the data dictate the shape of the kernel by fitting a spline model
-using the methods of Teller et al. (2016). The shape of $F_{jm}$ is determined by a set of spline basis coefficients $\vec{b}_{jm}$
-and a smoothing parameter $\eta$ that controls the complexity of the fitted kernel. 
-Demographic models such as \eqref{eqn:survReg} then have $\gamma$, $\varphi$, $\chi$ , 
-$\beta$, $\boldsymbol{\omega}$, $\vec{b}$ and $\eta$ as parameters to be fitted. We implemented this in the statistical computing environment, \texttt{R}, 
-by making the spline coefficients and $\eta$ the arguments of an objective function that computes $\boldsymbol{W}$ using the input spline coefficients, 
-calls the model-fitting functions \texttt{lmer} and/or \texttt{glmer} to fit the other parameters in the survival and growth regressions, 
-and returns an approximate AIC value and model degrees of freedom ($df$) for survival and growth combined. We used the $\vec{b}$ values at the smoothest 
-local minimum of AIC as a function of $df$, as in \cite{teller_linking_2016}. This approach assumes that one measure of crowding affects 
-survival and growth. In addition, for fitting the kernels we assumed that survival and growth depended only on intraspecific crowding, and thus only fitted the
-within-species competition kernels $F_{jj}$. Based on previous work \citep{adler_coexistence_2010}, we set all $F_{mj}$ equal to $F_{jj}$, meaning that 
-the within-species competition kernel for species $j$ is also used to determine the effect of all other species on species $j$.
-We used data from all historical plots and contemporary control-treatment plots to estimate the competition kernels \citep{adler_in_prep}.
-
-Once we had estimated the competitions kernels, we used them to calculate the values of $\boldsymbol{W}$ for each individual, 
-and fit the full survival and growth regressions, which include the interspecific interaction coefficients, $\boldsymbol{\omega}$. 
+using the methods of Teller et al.
+(2016).
+We used data from all historical plots and contemporary control-treatment plots to estimate the competition kernels and these are described in more detail in \citep{adler_weak_2016}.
+
+Once we had estimated the competitions kernels, we used them to calculate the values of $\boldsymbol{W}$ for each individual, and fit the full survival and growth regressions, which include the interspecific interaction coefficients, $\boldsymbol{\omega}$.
 All genets in a quadrat were included in calculating $W$, but plants located within 5 cm of quadrat edges were not used in fitting.
@@
 
-In the treatment model, a new term is added to the exponential term in the equation above, 
-\begin{equation}
- \boldsymbol{T}\boldsymbol{\chi}_{j}^R
-\end{equation}
-where $\chi$ describes the effect of the treatment levels on recruitment.
-
-Likewise in the climate model this term is added,
-\begin{equation}
-\boldsymbol{C}\xi_{j}^R 
-\label{eqn:recC}
-\end{equation}
-where the $\boldsymbol{\xi}$ gives a set of coefficients for the soil year, and treatment specific soil moisture covariates in $\boldsymbol{C}$.
-
-We fit all vital rate models using Hamiltonian-Markov Chain Monte Carlo (HMCMC) simulations in STAN 6.4 \citep{stan} and  \citep{rstan}.
+In the treatment model for recruitment, a new term is added to the exponential term in the equation above, $\boldsymbol{T}\boldsymbol{\chi}_{j}^R$ where $\chi$ describes the effect of each treatment level on recruitment.
+
+Likewise in the climate model this term is replaced by $\boldsymbol{C}\xi_{j}^R$ where the $\boldsymbol{\xi}$ gives a set of coefficients for the year, and treatment specific soil moisture covariates in $\boldsymbol{C}$.
+
+We fit all vital rate models using Hamiltonian-Markov Chain Monte Carlo (HMCMC) simulations in the programs STAN 10.1 and rStan (\citep{stan} and  \citep{rstan}).
 The priors and model code are described more completely in appendix A.
@@
 We discarded the initial 1,000 samples.
-Convergence was observed graphically for all parameters, and confirmed by assessing the split $\widehat{R}$ statistic which at convergence is equal to one \citep{stan manual}.
-
-To assess the effects of the experimental treatments on species survival and growth we fit treatment models with and without the size by treatment interactions in the treatment effect term for (\ref{eqn:survT}) and growth models (\ref{eqn:growT}).
-We judged whether including interaction terms improved model fit by comparing the Watanabe-Aikake Information Criteria (WAIC) scores for each version of the model and retained the version with the lower WAIC score \citep{vetari_practical_2015}.
-WAIC scores approximate cross-validation predictive accuracy for a model and like traditional AIC allow for model comparison.
+Convergence was observed graphically for all parameters, and confirmed by assessing the split $\widehat{R}$ statistic which at convergence is equal to one \citep{stan_manual}.
+
+We fit the treatment models for species survival and growth with and without the size by treatment interactions in the treatment effect term $\chi$.
+We then judged whether including the interaction terms improved model fit by comparing the Watanabe-Aikake Information Criteria (WAIC) scores of each version of the model and retained the version with the lower WAIC score \citep{vetari_practical_2015}.
+WAIC are similar to AIC scores and allow for comparison of Bayesian models.
 Lower WAIC scores indicate a more parsimonious model.
-When a treatment model for survival or growth for a species included a size by treatment effect, we also included a size by soil moisture effect in the climate model for that species and vital rate.
-This allowed us to more directly compare the effects in the experimental data to the effects predicted from the climate model fit to the observational data.
+When a treatment model for survival or growth of a species included a size by treatment effect in $\chi$, we also included a size by soil moisture effect in the $\xi$ term in the climate model for that species and vital rate.
+This allowed us to more directly compare the predictions from the climate model to the effects in the treatment model.
 
@@
 
-After generating a time series of predicted daily soil moisture from the SOILWAT model, we averaged daily soil moisture across spring, summer and fall seasons in each year.
+After generating a time series of predicted daily soil moisture from SOILWAT, we averaged daily soil moisture across spring, summer and fall seasons in each year.
 We considered each of the three seasonal soil moisture variables at three different time periods relative to the demographic transition from year $t$ to year $t+1$.
@@
 And soil moisture preceding this year is indicated with a "lag" subscript.
-For example, for the year 2010, $spring_1$ would indicate soil moisture in the spring of 2010, $spring_0$ would indicate soil moisture during the spring of 2009 and $spring_{lag}$ would indicate soil moisture during the spring of 2008.
-
-In order to select among the nine potential soil moisture covariates (three seasons and three lags each) for each species and vital rate, we first fit a model with random year effects but without climate covariates to the observational data up to 2010.
-We then extracted the mean of the year effects estimates for each fitted year in the data.
-These random effects represented unexplained deviations in the average vital rate in a given year.
-We then found the correlations between each of the soil moisture variable and the random year effects.
-For each vital rate and species, we selected the three covariates with highest correlations with these year effects.
-This screening technique has been used in previous demographic studies at this site \citep{dalgleish_climate_2010} and is often used in dendrochronology to screen for potential climate influence on tree-ring growth \citep{wang trees}; however, it is subject to the criticism that it is a form of data dredging \citep{dredging}.
-Nevertheless, we felt that this approach was justified in this study because ultimately we did not make inference from these fitted parameters until after we tested their ability to predict the data in the experimental plots.
-
-Once we selected soil moisture covariates for each of the species and vital rates we fit the climate models including the climate terms for survival \ref{eqn:survC}, growth \ref{eqn:growC} and recruitment \ref{eqn:recC}.
-We fit these models only to the observational data from the historical period and (1928 to 1957) and the first four years of the modern period (2007 to 2010).
-Thus in generating predictions for the treatment effects observed in from 2011 to 2016 we are predicting truly out of sample data.
+For example, for the year 2010, $spring_1$ indicates soil moisture in the spring of 2010, $spring_0$ indicates soil moisture during spring of 2009 and $spring_{lag}$ indicate soil moisture during spring 2008.
+
+We wanted to avoid fitting nine soil moisture covariates (three seasons and three lags each) for each species and vital rate, so we used only three soil moisture covariates per species and vital rate.
+We selected these three by calculating the correlations of each soil moisture variable with the random year effects from the no climate model fit and then selecting the three soil moisture variables with the strongest correlations with these year effects.
+This screening technique has been used in previous demographic studies at this site \citep{dalgleish_climate_2010} and is often used in dendrochronology to screen for potential climate influence on tree-ring growth \citep{wang_temporal_2003}.
+We felt this approach was justified because we did not make inference on these fitted parameters until after we validated their ability to predict the out of sample data in the experimental plots.
 
@@
 
-We accounted for the uncertainty in our random year effects when generating predictions, by drawing random year effects for each predicted year from a normal distribution with a mean of zero and a standard deviation drawn from the posterior estimate of the standard deviation of the random year effects \ref{eqn:}.
-We generated predictions using the full posterior distributions of each model parameter which allowed us to carry forward all of the uncertainty of the fitted vital rate models into our cover predictions.
+We generated predictions using 1000 samples from the posterior distributions of each model parameter which allowed us to carry forward all of the uncertainty of the fitted vital rate models into our cover predictions.
 Because we were interested in comparing model predictions to observations, and were not interested in the effects of demographic stochasticity, we used a deterministic version of the models (e.g., recruitment is the $\lambda$ of (\ref{eqn:recrProcessModel}), rather than a random draw from a negative binomial distribution with a mean of $\lambda$).
 
-We generated predictions for the full time series of observations, including the five experimental years, from the year effects model, the climate model and the treatment model.
-After generating predictions for each year, we found the mean cover across all quadrats in each treatment level and then calculated the predicted log cover change as $log(Cover_{t+1}/Cover_{t})$.
+After generating predictions for each year from the climate and no climate models, we found the predicted quadrat-level changes in cover as $log(Cover_{t+1}/Cover_{t})$.
 
 \subsection*{Quantifying predictive accuracy} 
-After fitting the year effects and climate models to the observational data, we generated predictions from these models for each of the vital rates in the experimental plots.
-We then assessed performance of the climate and null models by calculating the mean square error (MSE) between the predicted and observed responses in the experimental data as,
+We assessed the predictive performance of the climate and no climate by calculating the mean square error (MSE) between the predicted and observed responses in the experimental data as,
 
@@
 We compared these to the posteriors of the treatment parameters, $chi$, from the treatment model.
-As a measure of agreement between our predictions and observed response we calculated the correlation between the predicted and observed parameter values.
+As a measure of agreement between our predictions and observed response we calculated the correlation between the predicted and observed treatment effects.
 
@@
 
-We choose three seasonal soil moisture variables for each species' climate model based on their correlation with the year effects in the random year effects model fit to the observational data (table \ref{table:strongCor}).
+We choose three seasonal soil moisture variables for each species' climate model based on their correlation with the random year effects in the no climate model fit to the observational data (table \ref{table:strongCor}).
 We included size by soil moisture variables for \textit{A.
@@
 secunda} (table \ref{table:overallPreds}).
-In terms of lppd, the climate model outperformed the year effects model for \textit{A.
+In terms of lppd, the climate model outperformed the no climate model for \textit{A.
 tripartita} growth, \textit{H.
@@
 When we look at the predictions for each treatment separately we see that climate covariates improved model predictions more often in the drought treatments than in the control or irrigation treatments \ref{table:treatmentPreds}.
-For all four species, the climate model outperformed the year effects model for predicting the response of growth to drought in terms of lppd \ref{table:treatmentPreds}.
-The climate model outperformed the year effects model for predicting irrigation effects on growth for all species except \textit{H.
+For all four species, the climate model outperformed the no climate model for predicting the response of growth to drought in terms of lppd \ref{table:treatmentPreds}.
+The climate model outperformed the no climate model for predicting irrigation effects on growth for all species except \textit{H.
 comata}.
@@
 
-Using the vital rate models for each species we generated one year ahead cover predictions for each quadrat in each year of the experiment.
+Using the vital rate models for each species we generated one step ahead cover predictions for each quadrat in each year of the experiment.
 Average cover predicted by the climate model tended to be lower than the observed cover each year for \textit{A.
 tripartita} and \textit{P.secunda} (fig \ref{fig:coverPred}).
-Comparing the overall population growth rates predicted to those observed in the experiment, we see that the MSE of the climate model was lower than the MSE of the year effects model for \textit{P.
+Comparing the overall population growth rates predicted to those observed in the experiment, we see that the MSE of the climate model was lower than the MSE of the no climate model for \textit{P.
 secunda} and \textit{P.
 spicata} (table \ref{table:corPGR}).
-The predictions produced by the climate model for these species were also slightly more correlated with the observations than the predictions produced by the year effects model (table \ref{table:corPGR}).
+The predictions produced by the climate model for these species were also slightly more correlated with the observations than the predictions produced by the no climate model (table \ref{table:corPGR}).
 Considering each treatment and species separately we see that the predicted population growth rates for \textit{A.
@@
 comata} in the drought treatment (figs \ref{fig:pgrHECO}).
-However, the correlations between the predicted and observed log changes in cover did not always show the same pattern as MSE: the climate model made more strongly correlated predictions with the observations than the year effects model only for \textit{P.
+However, the correlations between the predicted and observed log changes in cover did not always show the same pattern as MSE: the climate model made more strongly correlated predictions with the observations than the no climate model only for \textit{P.
 spicata} and \textit{P.
@@
 In contrast, in the control and irrigated plots soil moisture may not have always been limiting during the course of the experiment.
-It may make sense then that our climate models did not make more accurate predictions in these conditions than a random year effects model.
+It may make sense then that our climate models did not make more accurate predictions in these conditions than the no climate model.
 
@@
 On the one hand it is re-assuring that observational data is sometimes useful.
-On the other hand, for predictions to be truly useful we would also like more information to help us sort out why our predictions for some species were good and for others no better than a null model.
+On the other hand, for predictions to be truly useful we would also like more information to help us sort out why our predictions for some species were good and for others no better than a no climate model.
 In other words, we have little ability to predict when our predictions are likely to be accurate.
@@
 \newpage
-\bibliographystyle{Ecology}
+\bibliographystyle{ecology}
 \bibliography{precip_experiment}
@@
 	\caption{Selected climate variables for each vital rate model for each species.
-Correlations and p-values between the choosen variables and the intercept of year effects model are shown.
+Correlations and p-values between the choosen variables and the intercept of the no climate model are shown.
 For ARTR growth and POSE growth and survival, the correlations between the year effects on size and the soil moisture variables are also given.
@@
 		\hline
-		species & vital\_rate & score & climate model & year effects model & diff & improved \\ 
+		species & vital\_rate & score & climate model & no climate model & diff & improved \\ 
 		\hline
@@
 	\end{tabular}
-	\caption{Comparison of model predictions from climate model and year effects model for each species and vital rate.
+	\caption{Comparison of model predictions from climate model and no climate model for each species and vital rate.
 Two prediction scores are reported, MSE and lppd.
 Lower MSE indicates improved predictions whereas higher lppd indicates improved predictions.
-Instances where the climate model outperformed the random year effects model are marked with "***" in the last column.
+Instances where the climate model outperformed the random no climate model are marked with "***" in the last column.
 ARTR = \textit{A.
@@
 		\hline
-		& species & stat & year effects model & climate model & diff & improved \\ 
+		& species & stat & no climate model & climate model & diff & improved \\ 
 		\hline
@@
 	\caption{MSE of predicted log cover changes and correlations between log cover changes predicted and observed.
-Predictions for the cover changes in the experimental plots were generated either from the year effects or the climate models.
-Instances where the climate model made better predictions than the year effects model are indicated with the "***".
+Predictions for the cover changes in the experimental plots were generated either from the no climate or the climate models.
+Instances where the climate model made better predictions than the no climate model are indicated with the "***".
 ARTR = \textit{A.
@@
 	\includegraphics[width=1\textwidth]{predicted_and_observed_cover}
-	\caption{Observed average cover per quadrat in the experimental and control plots and one year ahead cover predictions from the climate model.
+	\caption{Observed average cover per quadrat in the experimental and control plots and one step ahead cover predictions from the climate model.
 Cover predictions for each year are generated from the IBM based on the observed distribution of plants in each quadrat in the current year.
@@
 	\includegraphics[width=1\textwidth]{ARTR_predicted_pgr_comparison}
-	\caption{Observed and predicted one year ahead log change in \textit{A.
+	\caption{Observed and predicted one step ahead log change in \textit{A.
 tripartita} cover in the experiment.
-Changes in cover predicted by the climate model are shown on the left and those predicted by the year effects model are shown on the right.
+Changes in cover predicted by the climate model are shown on the left and those predicted by the no climate model are shown on the right.
 Correlations coefficients between predictions and observations and MSE are shown for each treatment and model.}
@@
 	\includegraphics[width=1\textwidth]{HECO_predicted_pgr_comparison}
-	\caption{Observed and predicted one year ahead log change in \textit{H.
+	\caption{Observed and predicted one step ahead log change in \textit{H.
 comata} cover in the experiment.
-Changes in cover predicted by the climate model are shown on the left and those predicted by the year effects model are shown on the right.
+Changes in cover predicted by the climate model are shown on the left and those predicted by the no climate model are shown on the right.
 Correlations coefficients between predictions and observations and MSE are shown for each treatment and model.
@@
 	\includegraphics[width=1\textwidth]{POSE_predicted_pgr_comparison}
-	\caption{Observed and predicted one year ahead log change in \textit{P.
+	\caption{Observed and predicted one step ahead log change in \textit{P.
 secunda} cover in the experiment.
-Changes in cover predicted by the climate model are shown on the left and those predicted by the year effects model are shown on the right.
+Changes in cover predicted by the climate model are shown on the left and those predicted by the no climate model are shown on the right.
 Correlations coefficients between predictions and observations and MSE are shown for each treatment and model.}
@@
 	\includegraphics[width=1\textwidth]{PSSP_predicted_pgr_comparison}
-	\caption{Observed and predicted one year ahead log change in \textit{P.
+	\caption{Observed and predicted one step ahead log change in \textit{P.
 spicata} cover in the experiment.
-Changes in cover predicted by the climate model are shown on the left and those predicted by the year effects model are shown on the right.
+Changes in cover predicted by the climate model are shown on the left and those predicted by the no climate model are shown on the right.
 Correlations coefficients between predictions and observations and MSE are shown for each treatment and model.}
@@
 	\hline
-	& species & vital\_rate & Treatment & score & climate model & year effects model & diff & improved \\ 
+	& species & vital\_rate & Treatment & score & climate model & no climate model & diff & improved \\ 
 	\hline
@@
 	\hline
-	\caption{Comparison of model predictions from climate model and year effects model for each species and vital rate and treatment.
+	\caption{Comparison of model predictions from climate model and no climate model for each species and vital rate and treatment.
 Two prediction scores are reported, MSE and lppd.
 Lower MSE indicates improved predictions whereas higher lppd indicates improved predictions.
-Instances where the climate model outperformed the random year effects model are marked with "***" in the last column.
+Instances where the climate model outperformed the no climate model are marked with "***" in the last column.
 ARTR = \textit{A.

### 2016-12-22 precip/manuscript/precip_ms.tex
edits to the results

@@
 
-The continuously recorded soil moisture data also showed treatment effects on soil moisture, but these effects were weaker on average than the spot measurements and depended on season and recent rainfall (table \ref{table:soil_moisture_model}; fig \ref{fig:dailyVWC}).
-We saw weaker effects during the spring than during the fall and summer: the drought plots were about 20-30\% drier in the fall and summer but only 7 to 14\% drier during the spring, while the irrigated plots were 30\% wetter during the fall and summer but only 20-25\% wetter during the spring.
-Treatment differences were slightly larger during rainy periods, especially in the spring.
+The continuously recorded soil moisture data also showed treatment effects, but these were weaker on average than the spot measurements and depended on season and recent rainfall (table \ref{table:soil_moisture_model}; fig \ref{fig:dailyVWC}).
+We saw weaker effects during the spring than during the fall and summer: the drought plots were about 20-30\% drier than ambient in the fall and summer but only 7 to 14\% drier during the spring, while the irrigated plots were 30\% wetter during the fall and summer but only 20-25\% wetter during the spring.
+Treatment differences were slightly larger during rainy periods, especially in the spring (rainfall effect in table \ref{table:soil_moisture_model}).
 We did not find evidence that the drought shelters and the irrigation treatments consistently affected air temperature at 30 cm above the plots.
@@
 Annual variation in seasonal soil moisture for each year was positively correlated with seasonal precipitation and negatively correlated with seasonal temperature.
-During the course of the experiment, SOILWAT reproduced much of the daily variation observed in soil moisture recorded by our automatic data loggers, but the average soil moisture predicted by SOILWAT was about 5 ml/ml wetter than the soil moisture content recorded by the data loggers.
-
-After adjusting the SOILWAT seasonal soil moisture predictions by the treatment effects, we found that the soil moisture predicted in the drought plots during the course of the experiment was well below the historical seasonal averages: the summer of 2012 and 2013, the fall of 2013, and the spring and winter of 2014 fell below the 5th percentile limit for drought in the historical period \ref{fig:seasonalVWC}.
-Soil moisture in our irrigation plots was generally above the historical average soil moisture but conditions never exceeded the 90th percentile for soil moisture in the historical period.
+During the course of the experiment, SOILWAT reproduced much of the daily variation observed in soil moisture recorded by our automatic data loggers, but the average soil moisture predicted by SOILWAT was about 5 ml/ml higher than the soil moisture content observed in the field.
+
+After adjusting the SOILWAT seasonal soil moisture predictions by the treatment effects, we found that the soil moisture predicted in the drought plots during the course of the experiment was well below the historical seasonal averages: the summer of 2012 and 2013, the fall of 2013, and the spring and winter of 2014 fell below the 5th percentile limit for drought in the historical period (fig \ref{fig:seasonalVWC}).
+Soil moisture in our irrigation plots was generally above the historical average soil moisture but conditions never exceeded the 90th percentile for soil moisture in the historical period (fig \ref{fig:seasonalVWC}).
 
@@
 comata} and \textit{P.
-spicata} fell significantly in the drought plots from 2011 to 2016 (tables \ref{table:changeHECO}, \ref{table:changePSSP}, fig \ref{fig:coverChange}).
+spicata} fell significantly in the drought plots from 2011 to 2016 (tables \ref{table:changeHECO}, \ref{table:changePSSP}; fig \ref{fig:coverChange}).
 The cover of \textit{P.
@@
 Our treatment models fit to the experimental and observational data indicated a variety of treatment effects on the vital rates of each species.
-After comparing WAIC scores for the growth models with and without the size by treatment effects, we retained size by treatment interaction effects in the growth models for \textit{A.
+Based on the WAIC scores with and without the size by treatment effects, we retained size by treatment effects in the growth models for \textit{A.
 tripartita} and \textit{P.
@@
 For \textit{A.
-tripartita} we found significant size by treatment effects of drought: drought had positive effects on plants of average size and smaller \ref{fig:growthTreat}, but plants larger than the mean size by more than 1.5 standard deviations grew slightly less in the drought treatment than in the controls.
+tripartita} we found significant size by treatment effects of drought: drought had positive effects on plants of average size and smaller (fig \ref{fig:growthTreat}), but plants larger than the mean size by more than 1.5 standard deviations grew slightly less in the drought treatment than in the controls.
 \textit{A.
-tripartita} showed the opposite response in the irrigated plots, (although the irrigation parameters were not technically significant at the 95\% confidence level): irrigation reduced growth for small plants while irrigation increased growth of plants more than 1.5 standard deviations greater than the mean size.
+tripartita} showed the opposite response in the irrigated plots, (although the irrigation parameters were not significant at the 95\% confidence level): irrigation reduced growth for small plants while irrigation increased growth of plants more than 1.5 standard deviations larger than the mean size.
 Drought led to a strong (but not signficant) decrease in \textit{H.
@@
 \textit{P.
-secunda} showed the opposite response in the irrigation plots with larger plants showing the largest increase in growth in response to irrigation (although these effects were technically not significant).
+secunda} showed the opposite response in the irrigation plots with larger plants showing the largest increase in growth in response to irrigation (although not significant).
 \textit{P.
@@
 
-Drought decreased the survival of all three grass species (fig \ref{fig:survivalTreat}).
+Survival of all three grass species (fig \ref{fig:survivalTreat}) decreased in the drought plots.
 And \textit{P.
-secunda} showed a negative size by drought interaction effect, indicating that the survival of larger plants was more negatively affected by drought than that of the smaller plants, with the smallest plants (plants one standard deviation smaller than the mean) actually seeing a slightly positive effect of drought.
+secunda} showed a negative size by drought interaction effect: the survival of larger plants was more negatively affected by drought than that of the smaller plants.
 \textit{A.
@@
 The recruitment data for \textit{A.
-tripartita} were relatively limited, with only 32 new recruits in total observed in all 30 plots over the course of the five year experiment.
-
-As expected from previous research most of our demographic models estimated strong negative intra-specific crowding effects and weaker negative inter-specific crowding effects on the focal species (appendix) \citep{adler_coexistence_2010,chu_direct_2016,chu_large_2015}.
+tripartita} were relatively limited, with only 32 new recruits in total observed in all 30 plots over the course of the five year experiment and we observed no treatment effects.
+
+Consistent with previous research most of our demographic models estimated strong negative intra-specific crowding effects and weaker negative inter-specific crowding effects on the focal species (appendix) \citep{adler_coexistence_2010,chu_direct_2016,chu_large_2015,adler_weak_2016}.
 
@@
 
-We choose three seasonal soil moisture variables for each species' climate model based on their correlation with the random year effects in the no climate model fit to the observational data (table \ref{table:strongCor}).
+We choose three seasonal soil moisture variables for each species' based on their correlation with the random year effects in the no climate model (table \ref{table:strongCor}).
 We included size by soil moisture variables for \textit{A.
@@
 secunda} based on the treatment response we observed in the experiment.
-All three time lags and all three seasons show up in the choosen variables.
-After fitting the vital rate models with the selected soil moisture variable we observed a trend towards positive soil moisture effects on growth of all three grasses \ref{fig:climateGrowth}.
-For \textit{H.comata} the soil moisture of the most recent summer had a significantly positive effect while the soil moisture of the previous summer and the fall before that were also positive but not significant.
+All three time lags and all three seasons show up in the selected variables.
+After fitting the vital rate models with the selected soil moisture variable we observed a trend towards positive soil moisture effects on growth of all three grasses (fig \ref{fig:climateGrowth}).
+For \textit{H.comata} the soil moisture of the most recent summer ($summer_1$) had a significantly positive effect while the soil moisture during $summer_0$ and $fall_lag$ were also positive but not significant.
 For \textit{A.
-tripartita} the summer and fall soil moisture of the previous year had strong negative effects on growth.
-There were also strong positive size by climate interaction effects for these variables: soil moisture had a stronger negative effect on small plants and a positive effect only on the largest plants.
-
-Soil moisture had significant effects on the survival of all four species \ref{fig:climateSurvival}.
+tripartita} $fall_0$ and $summer_0$ had strong negative effects on growth.
+There were also strong positive size by climate interaction effects for these variables: soil moisture had a stronger negative effect on small plants and a positive effect only on the largest plants (fig \ref{fig:parPredARTRGrowth}).
+
+Soil moisture had significant effects on the survival of all four species (fig  \ref{fig:climateSurvival}).
 As for growth the grasses showed mainly positive effects while \textit{A.
-tripartita} showed a significant negative effect of soil moisture in the summer of the previous year and a strong negative effect of spring soil moisture of the previous year.
+tripartita} showed a significant negative effect of $summer_0$ and a strong negative effect of $spring_0$.
 \textit{H.
-comata} showed a significant positive effect of spring lag soil moisture and a strong positive effect of spring soil moisture in the previous year.
+comata} showed a significant positive effect of $spring_lag$ soil moisture and a strong positive effect of $spring_0$ and $fall_1$.
 \textit{P.
-secunda} showed a significant positive effect of the previous spring's soil moisture and there was a significant interaction between this effect and plant size: as plant size increased this effect became more strongly positive.
+secunda} showed a significant positive effect of the previous $spring_0$ and there was an interaction between this effect and plant size: as plant size increased this effect became more positive.
 Finally for \textit{P.
-spicata} there was a significant positive effect of lag spring soil moisture on survival.
-
-There were only two significant effects of soil moisture on recruitment: lag fall soil moisture had a positive effect on \textit{P.
-secunda}, and lag summer soil moisture had a negative effect on \textit{P.
+spicata} there was a significant positive effect of $spring_lag$ soil moisture on survival.
+
+There were only two significant effects of soil moisture on recruitment: $fall_lag$ soil moisture had a positive effect on \textit{P.
+secunda}, and $summer_lag$ soil moisture had a negative effect on \textit{P.
+spicata} recruitment (fig \ref{fig:climateRecruitment}).
+Soil moisture of $summer_0$ also had a strong negative effect on \textit{P.
 spicata} recruitment.
-Soil moisture of the previous year's summer had also had a strong negative effect on \textit{P.
-spicata} recruitment.
-
-The intra- and interspecific crowding effects estimated in the climate model were similar to those estimated in the treatment model (appendix) \citep{adler_coexistence_2010,chu_direct_2016,chu_large_2015}.
+
+The intra- and interspecific crowding effects estimated in the climate model were similar to those estimated in the treatment model (appendix).
 
@@
 
-For most models adding climate covariates did not improve our ability to predict species responses in the experiment \ref{table:overallPreds}.
+For most models adding climate covariates did not improve our ability to predict species responses in the experiment (table \ref{table:overallPreds}).
 However, the climate models did improve overall prediction MSE for growth of \textit{A.
@@
 secunda} (table \ref{table:overallPreds}).
-In terms of lppd, the climate model outperformed the no climate model for \textit{A.
+In terms of lppd, the climate model outperformed the no climate model in six out of twelve models: for \textit{A.
 tripartita} growth, \textit{H.
@@
 secunda} growth and survival and \textit{P.
-spicata} recruitment.
-.
-
-When we look at the predictions for each treatment separately we see that climate covariates improved model predictions more often in the drought treatments than in the control or irrigation treatments \ref{table:treatmentPreds}.
-For all four species, the climate model outperformed the no climate model for predicting the response of growth to drought in terms of lppd \ref{table:treatmentPreds}.
-The climate model outperformed the no climate model for predicting irrigation effects on growth for all species except \textit{H.
+spicata} recruitment (table \ref{table:overallPreds}).
+
+When we look at the predictions for each treatment separately we see that climate covariates improved model predictions more often in the drought treatments than in the control or irrigation treatments (table \ref{table:treatmentPreds}).
+For all four species, the climate model outperformed the no climate model for predicting the response of growth to drought in terms of lppd (table \ref{table:treatmentPreds}).
+The climate model also outperformed the no climate model for predicting irrigation effects on growth for all species except \textit{H.
 comata}.
 
-Overall our climate models often predicted the correct direction of the drought and irrigation treatments \ref{fig:parPredictions}.
-In four cases we both observed and predicted treatment effects significantly different from zero based on the 95\% Bayesian credible interval around the parameters: the drought response of \textit{H.
+Overall our climate models often predicted the correct direction of the drought and irrigation treatments (fig \ref{fig:parPredictions}).
+In four cases we both observed and predicted treatment effects significantly different from zero based on the 95\% Bayesian credible interval around the parameter mean: the drought response of \textit{H.
 comata} survival (fig \ref{fig:parPredHECOSurvival}), the drought response of \textit{P.
@@
 In only one of these cases, for \textit{P.
-secunda} recruitment, was the predicted effect in the opposite direction from the observed treatment effect \ref{fig:parPredictions}.
-The overall correlation between the predicted and observed treatment effects for all treatments, species and vital rates was r = 0.54, whereas the correlation for the drought treatment effects (r = 0.77) was better than for the irrigation effects (r = 0.46).
-Also for the three models in which we included size by treatment or size by climate interactions the correlation of these size dependent effects were much stronger than the intercept parameter estimates \ref{fig:parPredictions}.
+secunda} recruitment, was the predicted effect in the opposite direction from the observed treatment effect (fig \ref{fig:parPredictions}).
+The overall correlation between the predicted and observed treatment effects for all treatments, species and vital rates was r = 0.54, whereas the correlation for the drought treatment effects, r = 0.77, was better than for the irrigation effects, r = 0.46.
+Also the correlation between the size by climate and size by treatment effects for \textit{A.
+tripartita} growth and \textit{P.
+secunda} growth and survival was much stronger than the correlation between the intercept parameter estimates (fig \ref{fig:parPredictions}).
 
@@
 The predictions produced by the climate model for these species were also slightly more correlated with the observations than the predictions produced by the no climate model (table \ref{table:corPGR}).
-Considering each treatment and species separately we see that the predicted population growth rates for \textit{A.
+Considering each treatment and species separately, the predicted population growth rates for \textit{A.
 tripartita}, \textit{P.
@@
 spicata} in the control treatment and \textit{H.
-comata} in the drought treatment (figs \ref{fig:pgrHECO}).
+comata} in the drought treatment (fig \ref{fig:pgrHECO}).
 However, the correlations between the predicted and observed log changes in cover did not always show the same pattern as MSE: the climate model made more strongly correlated predictions with the observations than the no climate model only for \textit{P.
@@
 
-Our experiment showed that observational data on the response of plant populations to interannual climate variation can indeed help us predict the direction of species response to experimental climate manipulations \ref{fig:parPredictions}.
-This was true even though adding climate parameters to the demographic models only improved vital rate predictions for half of the models \ref{table:overallPreds}.
-This should give us some hope that even when climate effects in demographic models to observational data are weak, they may contain useful qualitative information on the direction of climate effects in the future.
+Our experiment showed that observational data on the response of plant populations to interannual climate variation can indeed help us predict the direction of species responses to experimental climate manipulations (fig \ref{fig:parPredictions}).
+This was true even though adding climate parameters to the demographic models only improved vital rate predictions for half of the models (fig \ref{table:overallPreds}).
+This should give us some hope that even when climate effects in demographic models fit to observational data are weak, they may contain useful qualitative information on the direction of climate effects in the future.
 
@@
 
-Funding was provided by an NSF GRFP to AK, NSF grants DEB-1353078, and DEB-1054040 to PBA and by the Utah Agricultural Experiment Station (get journal paper number).
+Funding was provided by NSF grants DEB-1353078, and DEB-1054040 to PBA, an NSF GRFP award to AK, and by the Utah Agricultural Experiment Station (get journal paper number).
 The USDA-ARS Sheep Experiment Station generously provided access to historical data and the field experiment site.
-Joe and Kevin, Lau, John Bradford, Caitlin (soilwat),
+Laureano Gherardi provided us with valuable advice in constructing the rain-out and irrigation experiment.
 

### 2016-12-22 precip/manuscript/precip_ms.tex
almost done with the discussion

@@
 
-\begin{enumerate}
-\item Climate is an important driver of population ecology however there have been few tests of whether observational data correlating population performance with climate variation can be used to predict future responses of populations to experimentally imposed climate.
-\item Using longterm historical observational data from a sagebrush steppe plant community, previous research has reported a wide variety of climate effects on four dominant plant species.
-We tested whether these climate effects observed in observational data were useful for predicting each species response to a drought and irrigation experiment.
-\item We used rainfall shelters and automatic sprinklers to manipulate the soil moisture in sixteen new plots established at the same field site as the original historical plots.
-We also re-monitored the original plots as controls.
+Climate is an important driver of population ecology however there have been few tests of whether observational data correlating population performance with climate variation can be used to predict how populations respond to climate change experiments.
+We tested whether longterm observational data relating the effects of climate variation to the survival, growth and recruitment of four common plants in a sagebrush steppe could be produce accurate predictions for how each species would respond to a drought and irrigation experiment.
+We used rainout shelters and automatic sprinklers to manipulate the soil moisture in sixteen new plots established at the same field site as the longterm observational data.
 After running the experiment for five years, we analyzed how the growth, survival and recruitment of each species responded to the experimental drought and irrigation treatments.
-\item In order to test whether we could predict the observed responses to the experiment we fit two types of models to the observational data collected prior to the experiment: first we fit a no climate model that included the effects of plant size and competition on plant performance in each plot in each year; next we fit a climate model that in addition to size and competition also included the effects of three seasonal soil moisture variables.
-Next we used these models to predict the survival and growth of each individual plant, and the total recruitment in each plot during the course of the experiment.
-We also used an individual based population model parameterized with the vital rate predictions to predict one step ahead changes in population size in the experimental plots.
-We then compared the predictions made by each of these models to the observed experimental responses and compared the performance of the no climate model to the climate model.
-\item Over the course of the experiment, average cover of the perennial grasses, \textit{Hesperostipa comata} and \textit{Pseudoroegneria spicata} declined significantly in the drought treatment.
+We made predictions using two types of models fit to the observational data collected prior to the experiment: a no climate model that only included the effects of plant size and competition on plant performance in each plot in each year; and a climate model that also included the effects of three seasonal soil moisture variables.
+We then used these models to predict the performance of each species during the course of the experiment.
+We also generated one step a head predictions of population size in each experimental plot using an individual based population model.
+We compared predictions made by the the no climate and climate models to the observed experimental responses.
+Over the course of the experiment, average cover of the perennial grasses, \textit{Hesperostipa comata} and \textit{Pseudoroegneria spicata} declined significantly in the drought treatment.
 At the level of individual vital rates, experimental drought reduced the survival of \textit{Hesperostipa comata} and \textit{P.
@@
 tripartita}.
-\item The climate better predicted the response to the experiment than the no climate model in six out of the twelve cases we examined.
+The climate model better predicted the response to the experiment than the no climate model in six out of twelve cases.
 Across all species and vital rates, there was a strong positive correlation between the observed effects of the experiment and the effects predicted by the climate model fit to the observational data.
-At the population-level, the climate model made better predictions of the changes in species cover than the no climate model for two species: \textit{P.
+At the population-level, the climate model made better predictions of the changes in species cover than the no climate model for \textit{P.
 secunda} and \textit{P.
 spicata}.
-Moreover, in the drought treatments the climate model made better predictions for the changes in cover of all four species.
-\item \emph{Synthesis}: Observational climate data held valuable information for predicting species' responses to a modern climate change experiment in this ecosystem.
+Observational climate data held valuable information for predicting species' responses to a climate change experiment in this ecosystem.
 Treatment responses often matched the direction of predicted responses even when the effects were not significant.
 We were better able to predict species' responses to the drought treatment than to the control and irrigation treatments.
-Suggesting that soil moisture is an important factor in predicting the population dynamics of these species but only when water is truly limiting.
+This suggests that soil moisture is an important factor in predicting the population dynamics of these species but only when water is truly limiting.
 
@@
 If species performance responds non-linearly to temperature, we may be unable to predict how species respond to temperatures outside the range of observed variation \citep{doak_demographic_2010}.
-In addition, climate change will not just change the average climate, but is also likely to increase the variance in precipitation and lead to extreme events, which will have their own consequences independent of changes in climate averages \citep{drake_population_2005,jentsch_new_2007}.
+In addition, climate change will not just change the average climate, but is also likely to increase the variance in precipitation and lead to extreme events, which will have their own consequences independent of changes in climate means \citep{gherardi_enhanced_2015,jentsch_new_2007}.
 Any models based on observations drawn from the historical range of variation will therefore be extrapolating beyond both the range of observed averages and variance when used to predict the future \citep{williams_novel_2007}.
@@
 There is abundant evidence that plant performance shows high year to year variation, both at the level of individual growth, survival and reproduction and total population abundance.
-Precipitation is often especially important in driving interannual variation in net primary productivity among terrestrial plants \citep{knapp_variation_2001,hsu_anticipating_2014}.
-And the growth rates of trees are also often so tightly linked to annual precipitation that growth rings in their stems can serve as accurate records of historical climate variation thousands of years before the present \citep{yang_3500-year_2014}.
-Among smaller plants precipitation clearly affects the growth rates of small shrub and sub-shrub species \citep{srur_annual_2009,franklin_growth_2013}, and the germination and reproductive output of desert annuals is often tightly linked to annual precipitation \citep{venable_bet_2007}.
-
-Despite the clear signs that climate drives net primary productivity at the ecosystem level and the variation in individual growth rate in many plants, there have been relatively few studies that clearly link observed climate variation to species performance in population models \citep{ehrlen_advancing_2016}.
+Precipitation is often especially important in driving interannual variation in net primary productivity among terrestrial plants \citep{knapp_variation_2001,hsu_anticipating_2014}, the annual growth rings in trees \citep{yang_3500-year_2014} and smaller plants \citep{srur_annual_2009,franklin_growth_2013}, as well as the germination and reproductive output of many annuals \citep{venable_bet_2007}.
+
+Despite the clear signs that precipitation drives net primary productivity at the ecosystem level and annual variation in individual growth rate in many plants, there have been relatively few studies that clearly link observed climate variation to species performance in population models \citep{ehrlen_advancing_2016}.
 Fewer still have tested whether the population-level effects of climate variation shown in observational data can actually be used to predict the responses of plant populations to future climate variation.
@@
 This was true even though adding climate parameters to the demographic models only improved vital rate predictions for half of the models (fig \ref{table:overallPreds}).
-This should give us some hope that even when climate effects in demographic models fit to observational data are weak, they may contain useful qualitative information on the direction of climate effects in the future.
-
-* discuss biggest successes--astreces in fig \ref{fig:parPredictions}
-* discuss biggest mistake, P. secunda recruitment went in wrong direction.
-
-Scaling up to the population level, the climate models only produced better one step a head predictions of the overall response of the species to the experiment for two species: \textit{P.
+This should give us some hope that even when climate effects in demographic models fit to observational data are weak or not significant, they may contain useful qualitative information on the direction of climate effects in the future.
+
+*Although a number of previous studies of the plants in this system have looked the effects of climate on demography and survival, this is the first study to demonstrate effects of climate experimentally.
+The significant decline in \textit{H.
+comata} cover we report in the experiment (fig \ref{fig:coverChange}) is consistent with the population-level sensitivities to precipitation estimated by previous studies (\citep{chu_direct_2016},\citep{dalgleish_climate_2010}, \citep{adler_forecasting_2012}).
+Each of these studies report \textit{H.
+comata} having the strongest positive population response to increased precipitation among all the species.
+The response of this species' vital rates to drought in the experiment also matches this result (fig \ref{fig:growthTreat}, \ref{fig:survivalTreat}).
+On the other hand if we had only conducted an irrigation experiment our results may not have shown this consistency with previous work as \textit{H.
+comata} showed no positive response to irrigation.
+Previous work has also generally reported postive population responses to precipitation for the other grasses, \textit{P.
+secunda} and \textit{P.
+spicata} (\citep{adler_forecasting_2012}, \citep{chu_direct_2016}, \citep{dalgleish_climate_2010}).
+These results are consistent with significant decrease we saw in \textit{P.
+spicata} cover and survival and \textit{P.
+secunda} growth and survival in the drought plots in our experiment (figs \ref{fig:coverChange}, \ref{fig:growthTreat}, \ref{fig:survivalTreat}).
+As for \textit{H.
+comata} the magnitudes of the negative drought effects were greater than the magnitudes of the positive irrigation effects on these species.
+
+In contrast to the effects on the grasses the effects on \textit{A.
+tripartita} are more complicated.
+The effects of drought and irrigation on \textit{A.
+tripartita} growth were highly dependent on plant size (figs \ref{fig:growthTreat}): drought had a positive effect except possibly for the largest plants (fig \ref{fig:parPredARTRGrowth}).
+This result matches the negative effects of precipitation on this species estimated by \citep{adler_forecasting_2012} and \citep{chu_direct_2016}.
+These studies also report relative strong indirect effects of precipitation on \textit{A.
+tripartita} mediated by its competition with grasses.
+It is possible that some of the positive effect on \textit{A.
+tripartita} growth we see in the drought plots is the result of reduced grass competition these plots (fig \ref{fig:coverPred}; \citep{chu_direct_2016}).
+Our growth model should take into account any changes in grass abundance that could be driving a positive response from \textit{A.
+tripartita}.
+So this leaves us with the question of why this species would show a positive direct response to drought?
+Although there is some evidence that saturated soils can be detrimental to sagebrush growth \citep{} there was no evidence in our data that ambient moisture would have been so wet as to be limiting to sagebrush during this experiment (fig \ref{fig:dailyVWC}).
+Alternatively our drought treatments could be reducing snow cover in the winter in spring and allowing for earlier growth of sagebrush.
+In a high elevation sagebrush meadow earlier snowmelt was correlated with increased sagebrush performance \citep{harte_}.
+
+One trend that appears to stand out in these results is the relatively asymmetrical effects of drought and irrigation.
+Although our irrigation plots raised soil moisture, we saw little positive effect of added moisture on the same species that were reduced by drought (figs \ref{fig:growthTreat}, \ref{fig:survivalTreat}).
+We expected that in this water limited system, increasing soil moisture would lead to a stronger increase in plant performance.
+But perhaps we should not have been surprised by this result.
+It is possible that adding water to this system does little to increase grass growth and survival for a number of reasons.
+First, these species are presumably able to escape drought, by growing early in the year, or avoid drought stress through high water use efficiency \citep{bazzaz, franks}.
+Indeed our soil moisture data generally show a pulse of soil moisture in the spring when many grasses are actively growing (fig \ref{fig:dailyVWC}).
+During this window of active growth soil moisture in sagebrush steppe may be quite high and perhaps other resources become limiting \citep{caldwell} \citep{ino}.
+Moreover, our treatments had relatively smaller effects on soil moisture during this period \ref{table:soil_moisture_model}.
+If the species in our system are well adapted to the ambient water limitation, it makes some sense that they would show relatively small responses to increased moisture.
+
+A possibly related result was recently observed by \citep{gherardi_} in a warm desert grassland: the effect of increased soil moisture on grass productivity in that system showed a concave down pattern of response to increased soil moisture.
+Increasing soil moisture had weaker effects on grass productivity than decreasing soil moisture in equal proportion.
+This pattern could explain why we repeatedly saw large decreases in either the cover or performance of the three grasses in drought plots but did not see any statistically significant increase in performance in the irrigated plots.
+
+Reflecting on the possible mechanisms by which soil moisture could have affected these species sheds some light on our sucesses and failures of prediction.
+The drought effects we observed on the three grasses were often stronger than the effects we predicted, while the irrigation responses observed were often weaker than predicted (figs \ref{fig:parPredHECOGrowth}, \ref{fig:parPredHECOSurvival}, \ref{fig:parPredPOSEGrowth}, \ref{fig:parPredPOSESurvival}, \ref{fig:parPredPSSPGrowth}, \ref{fig:parPredPSSPSurvival}).
+If we had fit our growth and survival models with a non-linear function for soil moisture, perhaps informed by more mechanistic understanding of water limitation in this system, they may have made more accurate predictions of the drought and irrigation effects \citep{ehrlen_advancing_2016}.
+
+Out of all the climate effects we predicted and observed, we only observed four cases where vital rate predictions and observations were both significantly different from zero (fig  \ref{fig:parPredictions}).
+In three of these cases, we successfully predicted the direction of the treatment effect.
+However, in one cases we predicted a treatment effect that was significantly
+
+Using the IBM, we also demonstrated that the climate model predictions of the vital rates could be used to generate population-level predictions that were also better than the no climate models in some cases: the climate model improved one step ahead predictions of cover changes of \textit{P.
 spicata} and \textit{P.
 secunda} (table \ref{table:corPGR}).
-However adding climate effects produced better population-level predictions in the drought treatments for \textit{P.
-secunda} and \textit{H.
-comata}.
-Both the survival and the growth of these species were positively affected by increased soil moisture (fig \ref{fig:climateGrowth}, \ref{fig:climateSurvival}), and so it makes intuitive sense that we would observe declines in these species cover in the drought treatment (fig \ref{fig:coverChange}).
-
-*Among our species, we had the most success predicting the response of the three grass species and less success predicting the response of the shrub species \textit{A.
-tripartita}.
-Why might ARTR be different?
-
-*We also had more success predicting species response to the drought treatment than to the irrigation treatment.
-We also tended to see species respond more strongly to the drought treatment than they did to the irrigation treatment.
-This tells us that in some cases, demographic models based on observational data may be of more use for prediction in extreme conditions than when conditions are close to average.
-Our drought treatment likely created extremely low soil moisture compared to the historical average \ref{fig:seasonalVWC}.
-We hypothesize that this made water the most limiting resource for plants in these plots.
-In contrast, in the control and irrigated plots soil moisture may not have always been limiting during the course of the experiment.
-It may make sense then that our climate models did not make more accurate predictions in these conditions than the no climate model.
+In the drought treatment our one step ahead cover predictions for \textit{H.
+comata} and \textit{P.
+secunda} were also better than the no climate model.
+In this analysis, we either used the climate model, or the no climate model to generate predictions for the IBM.
+In theory we could have refined our predictions by  climate model to predict all three vital rates in the IBM.
+Improvements of our cover predictions may have been possible if we only included climate models in the IBM when those models improved vital rate predictions--for instance, the \textit{A tripartita} IBM would only include the climate model for growth but not for survival and recruitment.
 
@@
 
-*Will our predictions be useful for the longer term?
-Because these species compete, one could argue that our predictions for any one species in this community will only be as good as the predictions we make for their competitors.
-For instance, while we observed little effect of drought on \textit{A.
-tripartita} in our experiment, it is possible that it will eventually respond positively to the drought treatment as cover of the grass species it comptetes with declines (fig \ref{fig:coverChange}).
-We know that grass species compete strongly with \textit{A.
-tripartita}.
-However, we also know from previous work in this community that each species is more limited by intra-specific competition than by inter-specific crowding.
-This fact ensures that the direct effects of climate change will generally be greater than the competition mediated indirect effects of climate change \cite{chu_direct_2016, kleinhesselink_indirect_2015}.
-
-*Our results give us more confidence that historical observational data can in theory be used to detect and predict the demographic effects of climate change.
-This should encourage more researchers to try and use observational data to predict the future in both experimental and natural settings.
-Nevertheless, our success at predicting the short-term response of two species to a small-scale climate manipulation is not likely to be very reassuring to applied ecologists and resource managers wishing to make accurate medium to long-term quantitative predictions about the effects of climate change on the species they manage.
+*Will our predictions be useful for the predicting the effects of climate change in the future?
+Because these species compete, one could argue that our predictions for any one species in this community will only be as good as the predictions we make for their competitors \citep{tylianakis}.
+For instance, while we observed weak treatment effects on \textit{A.
+tripartita} in our experiment (fig \ref{fig:coverChange}) could drought ultimately lead to a decline in the cover of its competitors which will then allow it to increase in abundance in the future?
+While this is possible for some species, research at our site and several other grasslands suggest that species are more limited by intra-specific competition than inter-specific competition \cite{chu_direct_2016,adler_weak_2016}.
+In theory this fact should mean that direct effects rather than indirect effects dominate species responses to climate change (\citep{levine_competitors_2010,kleinhesselink_indirect_2015,adler_forecasting_2012}).
+
+*Our results give us more confidence that historical observational data can be used to detect and predict the demographic effects of climate change.
+This should encourage more researchers to try and use observational data to predict the future in both experimental and natural settings \citep{houlahan_priority_2016}.
+Making quantitative predictions, however humble they are, and testing them is arguably a key to scientific progress in all fields.
+Nevertheless, our success at predicting the short-term response of two species to a small-scale climate manipulation is not likely to be very reassuring to applied ecologists and resource managers wishing to make accurate predictions about the effects of climate change in large complex systems.
 Clearly more work is needed to distinguish which predictions are worth having confidence in and which predictions we should have less confidence in.
-Towards that goal, perhaps the best way to build confidence in ecological predictions is conduct more tests like this one.
+Towards that goal, perhaps the best way forward is to conduct more tests like this one.
 
@@
 \centerline{\Large \textbf{Supporting Information }}
-\centerline{Adler et al., ``Weak interspecific interactions''}
+\centerline{Kleinhesselink et al., ``Predicting climate response''}
 

### 2016-12-22 precip/manuscript/pgr_predictions.tex
commiting revisions to the discussion and formatting revisions on the tables

--- 2016-12-22 precip/manuscript/precip_ms.tex
@@
-%---------------------------------------------
-% This document is for pdflatex
-%---------------------------------------------
-\documentclass[11pt]{article}
-
-\usepackage{amsmath,amsfonts,amssymb,graphicx,natbib,setspace,authblk}
-\usepackage{float}
-\usepackage[running]{lineno}
-\usepackage[vmargin=1in,hmargin=1in]{geometry}
-
-\usepackage{enumitem}
-\setlist{topsep=.125em,itemsep=-0.15em,leftmargin=0.75cm}
-
-\usepackage{gensymb}
-
-\usepackage{longtable}
-\setlength{\LTcapwidth}{\linewidth}
-
-\usepackage[compact]{titlesec}
-
-\usepackage{bm,mathrsfs}
-
-\usepackage{ifpdf}
-\ifpdf
-\DeclareGraphicsExtensions{.pdf,.png,.jpg}
-\usepackage{epstopdf}
-\else
-\DeclareGraphicsExtensions{.eps}
-\fi
-
-\graphicspath{{../figures/}}
-
-\renewcommand{\floatpagefraction}{0.98}
-\renewcommand{\topfraction}{0.99}
-\renewcommand{\textfraction}{0.05}
-
-\clubpenalty = 10000
-\widowpenalty = 10000
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
-%%% Just for commenting
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\usepackage[usenames]{color}
-\newcommand{\new}{\textcolor{red}}
-\newcommand{\comment}{\textcolor{blue}}
-
-\newcommand{\be}{\begin{equation}}
-\newcommand{\ee}{\end{equation}}
-\newcommand{\ba}{\begin{equation} \begin{aligned}}
-\newcommand{\ea}{\end{aligned} \end{equation}}
-
-\def\X{\mathbf{X}}
-
-\floatstyle{boxed}
-\newfloat{Box}{tbph}{box}
-
-\title{Can historical data predict population responses to climate change experiments?}
-
-\author[a]{Andrew R. Kleinhesselink\thanks{Corresponding author. Department of Wildland Resources and the Ecology Center, Utah State University, Logan, Utah Email: arklein@aggiemail.usu.edu}}
-\author[a]{Peter B. Adler}
-\author[a]{Andrew Tredennick}
-\author[b]{Caitlin Andrews}
-\author[b]{John Bradford}
-\affil[a]{Department of Wildland Resources and the Ecology Center, Utah State University, Logan, Utah}
-\affil[b]{US Geological Survey, Southwest Biological Science Center, Flagstaff, Arizona}
-
-\renewcommand\Authands{ and }
-
-\date{Last compile: \today}
-
-\sloppy
-
-\renewcommand{\baselinestretch}{1.25}
-
-\begin{document}
-
-\maketitle
-
-\textbf{\large{Keywords:}} Climate change, demographic models, rain-out shelter.
-
-\bigskip \textbf{Running title:} Predicting climate response
-
-\smallskip \textbf{Article type:} Letter
-
-\smallskip \textbf{Authorship statement:} ARK and PBA designed the experiment and supervised data collection. AT helped with the data analysis. JB and CA helped with the soil moisture model. 
-\smallskip
-
-\setlength{\parindent}{8ex}
-
-\newpage
-
-\begin{doublespacing}
-
-\linenumbers
-
-\section*{Summary}
-
-\begin{enumerate}
-\item Climate is an important driver of population ecology however there have been few tests of whether observational data correlating population performance with climate variation can be used to predict future responses of populations to experimentally imposed climate.
-\item Using longterm historical observational data from a sagebrush steppe plant community, previous research has reported a wide variety of climate effects on four dominant plant species.
-We tested whether these climate effects observed in observational data were useful for predicting each species response to a drought and irrigation experiment.
-\item We used rainfall shelters and automatic sprinklers to manipulate the soil moisture in sixteen new plots established at the same field site as the original historical plots.
-We also re-monitored the original plots as controls.
-After running the experiment for five years, we analyzed how the growth, survival and recruitment of each species responded to the experimental drought and irrigation treatments.
-\item In order to test whether we could predict the observed responses to the experiment we fit two types of models to the observational data collected prior to the experiment: first we fit a no climate model that included the effects of plant size and competition on plant performance in each plot in each year; next we fit a climate model that in addition to size and competition also included the effects of three seasonal soil moisture variables.
-Next we used these models to predict the survival and growth of each individual plant, and the total recruitment in each plot during the course of the experiment.
-We also used an individual based population model parameterized with the vital rate predictions to predict one step ahead changes in population size in the experimental plots.
-We then compared the predictions made by each of these models to the observed experimental responses and compared the performance of the no climate model to the climate model.
-\item Over the course of the experiment, average cover of the perennial grasses, \textit{Hesperostipa comata} and \textit{Pseudoroegneria spicata} declined significantly in the drought treatment.
-At the level of individual vital rates, experimental drought reduced the survival of \textit{Hesperostipa comata} and \textit{P.
-spicata}  and the growth of the grass \textit{Poa secunda}.
-In contrast, drought increased the growth of the shrub \textit{A.
-tripartita}.
-\item The climate better predicted the response to the experiment than the no climate model in six out of the twelve cases we examined.
-Across all species and vital rates, there was a strong positive correlation between the observed effects of the experiment and the effects predicted by the climate model fit to the observational data.
-At the population-level, the climate model made better predictions of the changes in species cover than the no climate model for two species: \textit{P.
-secunda} and \textit{P.
-spicata}.
-Moreover, in the drought treatments the climate model made better predictions for the changes in cover of all four species.
-\item \emph{Synthesis}: Observational climate data held valuable information for predicting species' responses to a modern climate change experiment in this ecosystem.
-Treatment responses often matched the direction of predicted responses even when the effects were not significant.
-We were better able to predict species' responses to the drought treatment than to the control and irrigation treatments.
-Suggesting that soil moisture is an important factor in predicting the population dynamics of these species but only when water is truly limiting.
-
-\end{enumerate}
-
-\section*{Introduction}
-
-Climate is one of the most powerful drivers of changes in species abundance across space and time \citep{post_synchronization_2002,davis_range_2001,walther_ecological_2002}.
-The effects of climate on populations and ecosystems are most apparent at the largest scales: climate determines the distribution of ecosystems \citep{whittaker_communities_1975}, treelines \citep{korner_alpine_2012} and the the range limits of many species \citep{parmesan_globally_2003,davis_range_2001}.
-Understanding and predicting the effects of climate on populations is an increasingly important goal if we are to anticipate the effects of climate change on earth's ecosystems \citep{tredennick_we_2016,petchey_ecological_2015,ehrlen_advancing_2016,teller_linking_2016}.
-
-Ecologists often resort to one of two methods for predicting the effects of future climate change on populations and communities: they may use experiments to manipulate aspects of climate directly and observe the response of populations \citep{elmendorf_experiment_2015,knapp_pushing_2016,compagnoni_warming_2014}; or they may use longterm observational data on species performance and abundance collected over many years and relate this to ambient annual variation in climate  \citep{koons_climate_2012,lunn_demography_2016,dalgleish_climate_2010,jenouvrier_demographic_2009}.
-The strength of the experimental approach is in the stronger inference that comes from manipulating some aspects of climate while controlling for other factors; for instance knowing that loss of snow cover, and not necessarily changes in soil temperature or moisture are the factors causing a species performance to change with warming \citep{compagnoni_warming_2014}.
-It also allows for the creation of conditions that may be more extreme than those observed historically \cite{knapp_pushing_2016} but are possible in the future.
-However, it is often expensive to control climate at even the smallest scales, and larger scale climate manipulation is often impossible.
-Moreover, experimental manipulation can come with artifacts that may make them less then ideal models for understanding and predicting the effects of future variation in climate \citep{wolkovich_warming_2012}.
-
-Using observational data to predict the effects of climate variation on populations has many advantages over using experiments.
-The primary one being the ever increasing availability of long-term ecological data and ever increasingly detailed climate data.
-Observational studies may also be the only way to study the effects of climate on large and or migratory species, for which it would be difficult to manipulate climate \citep{koons_climate_2012,jenouvrier_demographic_2009,aubry_climate_2013}.
-However, there are many potential drawbacks to using essentially correlative relationships between species performance and climate to predict future species' responses.
-First, many years of data are needed to reliably detect climate effects, especially when annual variation in demographic rates is high \citep{teller_linking_2016,gerber_optimal_2015}.
-\citep{teller_linking_2016} estimate that even cutting edge statistical approaches for fitting relationships between climate and species performance require at least 20-25 years of independent climate observations before they perform well.
-Moreover, even strong correlations between species performance and the climate covariates we choose to include in our models may not be due to direct causation and may not hold up in future years of observation \citep{hilborn_correlation_2016}.
-
-And yet even overcoming this challenge, predicting the effects of climate change would face yet another challenge: in many parts of the world, climate change will lead to mean annual temperatures that will be above the range of historical variation.
-If species performance responds non-linearly to temperature, we may be unable to predict how species respond to temperatures outside the range of observed variation \citep{doak_demographic_2010}.
-In addition, climate change will not just change the average climate, but is also likely to increase the variance in precipitation and lead to extreme events, which will have their own consequences independent of changes in climate averages \citep{drake_population_2005,jentsch_new_2007}.
-Any models based on observations drawn from the historical range of variation will therefore be extrapolating beyond both the range of observed averages and variance when used to predict the future \citep{williams_novel_2007}.
-Demonstrating that models fit to historical data can accurately predict the effects of experimental climate manipulations, especially manipulations that generate extreme conditions, would be strong confirmation that the climate effects they describe are not spurious and will hold even in the novel conditions of the future \cite{adler_can_2013}.
-
-There is abundant evidence that plant performance shows high year to year variation, both at the level of individual growth, survival and reproduction and total population abundance.
-Precipitation is often especially important in driving interannual variation in net primary productivity among terrestrial plants \citep{knapp_variation_2001,hsu_anticipating_2014}.
-And the growth rates of trees are also often so tightly linked to annual precipitation that growth rings in their stems can serve as accurate records of historical climate variation thousands of years before the present \citep{yang_3500-year_2014}.
-Among smaller plants precipitation clearly affects the growth rates of small shrub and sub-shrub species \citep{srur_annual_2009,franklin_growth_2013}, and the germination and reproductive output of desert annuals is often tightly linked to annual precipitation \citep{venable_bet_2007}.
-
-Despite the clear signs that climate drives net primary productivity at the ecosystem level and the variation in individual growth rate in many plants, there have been relatively few studies that clearly link observed climate variation to species performance in population models \citep{ehrlen_advancing_2016}.
-Fewer still have tested whether the population-level effects of climate variation shown in observational data can actually be used to predict the responses of plant populations to future climate variation.
-\citep{adler_can_2013} showed that population models based on observed correlations between plant population growth rates and precipitation did have some predictive power in describing species response to a short-term climate manipulation in a North American grassland.
-Three species showed responses to experimentally imposed drought and irrigation that were well predicted by population models fitted to historical observations.
-However, another three species, showed responses to the experimental conditions that were not well predicted by historical observations.
-The authors suggested that limited replication in the historical data for two of these species and changing competitive conditions in the community may have led to the poor predictions.
-
-The demography of and competitive interactions between three dominant perennial bunchgrasses and a shrub in a sagebrush steppe plant community at the US Sheep Experiment Station near Dubois, Idaho have been described in at least seven different studies since 2010 \citep{adler_coexistence_2010,adler_forecasting_2012,adler_weak_2009,adler_weak_2016,chu_direct_2016,chu_large_2015,dalgleish_climate_2010}.
-And several of these studies report significant effects of climate variation on the vital rates and overall population growth rates of these species \citep{dalgleish_climate_2010,adler_forecasting_2012,adler_weak_2009,chu_direct_2016}.
-This well-studied system offers the ideal opportunity to test whether statistical associations between annual climate and plant demography in longterm observational data can be used to predict the responses of plant populations to experimental climate change experiments.
-
-In this study, we report how the four dominant plant species at the USSES respond to a five year drought and irrigation experiment and use the results to address two research questions: first, how much do the growth, recruitment and survival of our target species differ between the precipitation manipulation treatments?
-If our experiment does affect species vital rates we interpret that as strong evidence that changes in precipitation should have an effect on populations in the future.
-Second, can we predict each species' response to the experimental conditions based on how they respond to natural climate variation in the observational data?
-If models based on observational data can predict the response of species to this experiment this should give us more confidence in longterm population monitoring data to predict species response to climate change.
-
-\section*{Methods}
-
-\subsection*{Study site and data set description}
-
-The U.S.
-Sheep Experiment Station (USSES) is located at Dubois, Idaho (44.2\degree N, 112.1\degree W), 1500 m above sea level.
-During the period of data collection (1926 – 2016), mean annual precipitation was 270 mm and mean temperatures ranged from -8\degree C (January) to 21\degree C (July).
-The vegetation is dominated by a shrub, \textit{Artemisia tripartita}, and three perennial C3 grasses: \textit{Pseudoroegneria spicata}, \textit{Hesperostipa comata}, and \textit{Poa secunda}.
-These dominant species account for over 70\% of basal cover and 60\% of canopy cover at this site.
-
-Scientists at the USSES established 26 1-m$^2$ quadrats between 1926 and 1932.
-Eighteen quadrats were distributed among four ungrazed exclosures, and eight were distributed in two paddocks grazed at medium intensity spring through fall.
-All quadrats were located on similar topography and soils.
-In most years until 1957, all individual plants in each quadrat were mapped using a pantograph \citep{blaisdell_}.
-The historical data set is public and available online \citep{zachmann_mapped_2010}.
-In 2007, we located 14 of the original quadrats, all of which are inside permanent livestock exclosures, and resumed annual mapped censusing using the traditional pantograph method.
-Daily temperature and precipitation has been monitored throughout this period at a climate station located at the USSES headquarters (station id: GHCND:USC00102707) which located within 2 km of the research plots.
-We downloaded daily and monthly tmin, tmax, and precipitation data from the National Climate Data Centers online database.
-
-We extracted data on survival, growth, and recruitment from the mapped quadrats based on plants' spatial locations.
-Our approach tracks genets representing individual plants.
-For the shrub, each genet is associated with the basal position of a stem.
-For the bunchgrasses, each genet represents a spatially distinct polygon in the mapped quadrat.
-These genets may fragment and/or coalesce over time.
-Each mapped polygon is classified as a surviving genet or a new recruit based on its spatial location relative to genets present in previous years \citep{lauenroth_demography_2008}.
-We modeled vital rates using data from 21 year-to-year transitions between 1929 and 1957, and four year-to-year transitions from 2007 to 2011.
-
-\subsection*{Precipitation experiment}
-In spring 2011, we selected locations for an additional 16 quadrats for the precipitation experiment.
-We located these in a large exclosure containing six of the historical permanent quadrats.
-We avoided plots falling on hill slopes, areas with greater than 20\% bare rock, or with over 10\% cover of the woody shrubs \textit{Purshia tridentata} or \textit{Amelanchier utahensis}.
-New plots were established in pairs, and one plot per pair was randomly assigned to either the precipitation reduction or the precipitation addition treatment.
-We mapped the quadrats in June, 2011 and then built the rainfall shelters and set-up the irrigation systems in the fall of 2011.
-We used a rain-out shelter and automatic irrigation design described in \citep{gherardi_automated_2013}.
-Each rain-out shelter covered an area of 2.5 by 2 m and consisted of transparent acrylic shingles held up 1.5 to 1 m over the plot to channel 50\% of incoming rainfall off of the plot and into 75 l reservoirs.
-The collected water was pumped out of reservoirs and sprayed onto paired irrigation treatment plots.
-Pumping was initiated automatically with float switches that were triggered when water levels in the reservoirs were approximately 20 l, or equivalently irrigation was triggered once for every 6 mm of rainfall collected.
-We disconnected the irrigation pumps in late fall each year and re-connected them in April.
-The drought shelters remained in place throughout the year.
-
-We monitored soil moisture and air temperature in four of the precipitation experiment plot pairs using Decagon Devices (Pulman Washington) 5TM and EC-5 soil moisture sensors and 5TE temperature sensors.
-We installed two soil moisture sensors in each monitored plot at 5 cm and two at 25 cm deep in the soil.
-Air temperature was measured underneath the roofing of the shelter at 30 cm above ground.
-For each pair of manipulated plots, we also installed sensors in a nearby area to measure ambient rainfall and temperature.
-Data were logged automatically every four hours.
-We augmented automatic monitoring of the climate in these plots with direct measurements of soil moisture with a handheld EC-5 soil moisture sensor at six points around all 16 plots on 6/6/2012, 4/29/2015, 5/7/2015, 6/9/2015 and 5/10/2016.
-We analyzed these spot measurements for significant treatment effects on soil moisture using a linear mixed effects model with the \textit{lmer} package in \textit{R} \citep{bates}, with plot, plot group, and date as random effects in the model.
-
-We conducted a simple statistical to determine the net effect of the experimental treatments on cover in the experiment.
-First we calculated the log change in cover for each of the four focal species in each quadrat from from the start of the experiment in spring prior to manipulation, to the last year of the experiment.
-Log change in cover was defined as , $log(Cover_{2016}/Cover_{2011})$ where $Cover_{2016}$ is the cover of each species in 2011 and $Cover_{2015}$ is cover in 2011.
-We tested for the effect of precipitation treatment on this measure with a linear model in \textit{R}.
-
-\subsection*{Soil moisture modeling}
-
-We expected that our precipitation manipulation experiment would affect plants by altering available soil moisture during the growing season.
-Because we do not have direct soil moisture measures for each year of observed plant cover in the historical record, we used the SOILWAT soil moisture model to estimate daily soil moisture at the USSES from 1925 to the present \citep{Parton_1978}.
-We used an enhanced version of soilwat that has recently been developed for use in semi-arid shrubland ecosystems \citep{Bradford}.
-SOILWAT uses daily weather data, ecosystem specific vegetation properties and site specific soil properties to estimate water balance processes.
-SOILWAT specifically estimates rainfall interception by vegetation, evaporation of intercepted water, snow melt and snow redistribution, infiltration into the soil, percolation through the soil, bare-soil evaporation, transpiration from each soil layer, and drainage.
-We parameterized SOILWAT with the generic sagebrush steppe vegetation parameters and site specific soil texture and bulk density data.
-We used daily weather data collected at the USSES from 1925 until the present as weather forcing data for the SOILWAT predictions.
-
-We averaged daily soil moisture predictions from SOILWAT from upper 40 cm of soil and then averaged these seasonally to serve as the covariates in the vital rate regressions for each species.
-Because we did not monitor soil moisture directly in all control, drought and irrigation plots, we used a model to describe the average treatment effects on soil moisture during the course of the experiment.
-To do this we first averaged observed soil moisture data by date and plot and then standardized these by the mean and standard deviation of the control soil moisture conditions observed within each plot group.
-We then found the difference between the soil moisture in the treated plots and the ambient conditions.
-We then modeled these treatment effects as a function of season and whether a day was rainy or dry.
-We expected that our drought and irrigation treatments might be more effective during rainy weather than during dry weather.
-Rainy days were defined as any days when any precipitation was recorded and average temperatures were above 3 degrees C.
-The day immediately following rainfall was also classified as rainy.
-We fit this model using the \textit{lmer} package in \textit{R} \citep{bates} with random effects for plot group and date.
-We then used this model to predict the treatment effects on soil moisture for the entire study period from the ambient soil moisture values predicted from the SOILWAT model described above.
-These adjusted soil moisture values reflected the average season and rainfall dependent effects of the experimental treatments on soil moisture and could be used as covariates for predicting the effects of our manipulation on each species demographic rates.
-
-\subsection*{Overview of the analysis}
-
-Our analysis consists of two separate datasets and three different categories of vital rate models.
-We refer to the first dataset as the observational data.
-It consists of all the historical data collected from 1925 to 1957 as well as the modern data collected from the same plots from 2007 to 2010.
-These data record the response of plants in each plot to the ambient climate variation.
-We refer to the second dataset as the experimental data.
-It consists of the data collected from 2011 to 2016 from the 16 new experimentally manipulated plots, as well as from 14 of the original historical plots which serve as ambient climate controls.
-
-In order describe the effects of the experimental treatments on each vital rate, we fit "treatment" models.
-The treatment models included parameters for the effects of the drought and irrigation treatments on each vital rate.
-We fit these models to both datasets together, all the experimental and all the observational data.
-We did this because we wanted to focus our predictions on the effects of the experimental treatments on the vital rates, rather than any differences between the datasets in the effects of crowding and plant size on the vital rates.
-
-Next in order to test how well we could predict the responses in the experimental plots, we fit two classes of models to the observational dataset only.
-Thus when we used these models to predict the response of each species to the five years of experimental data we are generating predictions for out of sample data.
-We refer to the first type of model as the "no climate" models.
-These models include parameters for the effects of competition on each vital rate and the size dependence of survival and growth but they do not include climate or treatment effects.
-Lastly, we fit the "climate" models which include the effects of annual variation in soil moisture on each vital rate.
-The no climate model provides us a baseline by which to measure the accuracy of the predictions from the climate model.
-Because much of the variation in growth, survival and recruitment in this system can be explained by plant size and competition, we expect that these two models will make similar predictions for individual plant performance in the experiment.
-However, if the climate model makes more accurate predictions than the no climate model, this indicates that the climate parameters it includes contain useful information for prediction.
-
-\subsection*{Statistical models of vital rates}
-
-All three categories of models described above follow the same basic structure and differ only in how they treat climate and treatment effects and  \citep{adler_coexistence_2010,chu_large_2015}.
-We model the survival probability of an individual genet as a function of genet size, the neighborhood-scale crowding experienced by the genet from both conspecific and heterospecific genets, temporal variation among years, and permanent spatial variation among groups of quadrats (`group'; here means a set of nearby quadrats located within one pasture or grazing exclosure).
-In this analysis we only include crowding from the four main focal species.
-
-Formally, we modeled the survival probability, $S$, of genet $i$ in species $j$, group $g$, and from time $t$ to $t+1$  as
-\begin{equation}
-\mbox{logit}(S_{ijg,t}) = \varphi_{jg}^S + \gamma_{j,t}^S  + \beta_{j,t}^S u_{ij,t} +  
-\left \langle \boldsymbol{\omega}_{j,t}^S, \boldsymbol{W}_{ij,t} \right \rangle 
-\label{eqn:survReg}
-\end{equation}
-where $\varphi$ is the spatial group dependent intercept, $\gamma$ is a year-effect, $\beta$ is year-dependent coefficient that represents the effect of log genet size, $u$, on survival in year $t$.
-$\boldsymbol{\omega}$ is a vector of interaction coefficients which determine the impact of crowding, $\boldsymbol{W}$, by each species on the focal species.
-The vector $\boldsymbol{W}$ includes crowding from the four dominant species, \textit{A.
-tripartita}, \textit{P.
-spicata}, \textit{H.
-comata}, and \textit{Poa secunda}.
-$\left \langle \boldsymbol{x, y} \right \rangle$ denotes the inner product of vectors $\boldsymbol{x}$ and $\boldsymbol{y}$, 
-calculated as \texttt{sum(x*y)} in R. This model is the no climate model for survival.
-
-In the treatment model, a new term is added to the above model, $\boldsymbol{T}\chi_{j}^S$ where $\chi$ is a vector of treatment effect coefficients for each experimental treatment level $h$ on the survival rate, and $\boldsymbol{T}$ is a design matrix indicating the treatment level of each observation in the data.
-The design matrix also includes terms for the interaction between plant size $u$ and the treatment effects which allow the effect of each treatment to vary with plant size.
-
-In the climate model, the above term is replaced with $\boldsymbol{C}\xi_{j}^S$, where $\xi$ gives a vector of coefficients describing the effects of a set of soil moisture covariates $\boldsymbol{C}$ in treatment $h$ and year $t$ on the survival rate of species $j$.
-$\boldsymbol{C}$ can include interaction effects between plant size, $u$, and the soil moisture covariates allowing the effects of soil moisture to vary with plant size.
-
-Our growth model has a similar structure. The change in genet size from time $t$ to $t+1$ , conditional on survival, is given by:
-\begin{equation}
-u_{ijg,t+1} = \varphi_{jg}^G + \gamma_{j,t}^G + \chi_{jh}^G  + \beta_{j,t}^G u_{ij,t} + 
-\left \langle  \boldsymbol{\omega}_{j,t}^G, \boldsymbol{W}_{ij,t} \right \rangle + \varepsilon_{ij,t}^G .
-\label{eqn:growReg}
-\end{equation}
-
-As in the survival regression above, parameters describing the treatment effects on growth are added in the treatment model, $\boldsymbol{T}\chi_{j}^G$, where $\chi$ is a treatment effect describing the effect of experimental treatment $h$ on growth, including treatment by size interactions.
-
-Similarly, in the climate model, the above term is replaced with $\boldsymbol{C}\xi_{j}^G$, where $\xi$ is a vector of coefficients describing the effects of soil moisture covariates in the matrix $\boldsymbol{C}$ for treatment $h$ and year $t$ on growth of species $j$.
-Again this can include interactions between soil moisture and plant size $u$.
-
-Although the main focus of the current analysis the effects of soil moisture, we also modeled the effects of inter- and intra-specific competition in our vital rate models.
-We model the crowding experienced by a focal genet as a function of the distance to and size of neighbor genets.
-These effects are well described in previous work \citep{teller_linking_2016,adler_weak_2016}.
-Briefly, we model the crowding experienced by genet $i$ of species $j$ from neighbors of species $m$ as the sum of neighbor areas across a set of concentric annuli, $k$, centered at the plant,
-\begin{equation}
-w_{ijm,k} = F_{jm}(d_{k})A_{i,k}     
-\label{eqn:wik}
-\end{equation}
-where $F_{jm}$ is the competition kernel (described below) for effects of species $m$ on species $j$, 
-$d_{k}$ is the average of the inner and outer radii of annulus $k$, 
-and $A_{im,k}$ is the total area of genets of species $m$ in annulus $k$ around genet $i$. The total crowding on 
-genet $i$ exerted by species $m$ is
-\begin{equation}
-W_{ijm}  =\sum_k {w_{ijm,k}} .
-\label{eqn:wijm}
-\end{equation} 
-Note that $W_{ijj}$ gives intraspecific crowding.
-The $W$'s are then the components of the $\boldsymbol{W}$ vectors introduced as covariates in the survival (\ref{eqn:survReg}) and growth (\ref{eqn:growReg}) regressions.
-
-We assume that competition kernels $F_{jm}(d)$ are non-negative and decreasing, so that distant plants have less effect than close plants.
-Otherwise, we let the data dictate the shape of the kernel by fitting a spline model
-using the methods of Teller et al.
-(2016).
-We used data from all historical plots and contemporary control-treatment plots to estimate the competition kernels and these are described in more detail in \citep{adler_weak_2016}.
-
-Once we had estimated the competitions kernels, we used them to calculate the values of $\boldsymbol{W}$ for each individual, and fit the full survival and growth regressions, which include the interspecific interaction coefficients, $\boldsymbol{\omega}$.
-All genets in a quadrat were included in calculating $W$, but plants located within 5 cm of quadrat edges were not used in fitting.
-
-We model recruitment at the quadrat level rather than at the individual genet level because the mapped data do not allow us to determine which recruits were produced by which potential parent plants.
-We assume that the number of individuals, $y$, of species $j$ recruiting at time $t+1$ in the location $q$ follows a negative binomial distribution:
-\begin{equation}
-y_{jq,t+1}= NegBin(\lambda_{jq,t+1},\theta) 	   
-\label{eqn:recrDataModel}
-\end{equation}
-where $\lambda$ is the mean intensity and $\theta$ is the size parameter. In turn, $\lambda$ depends on the composition of the quadrat in the previous year:
-\begin{equation}
-\lambda_{jq,t+1} = C'_{jq,t} \exp{\left(\varphi_{jg}^R + \gamma_{j,t}^R + 
-\left \langle \boldsymbol{\omega}^R , \boldsymbol{\sqrt{C'}_{q,t}} \right \rangle \right) }
-\label{eqn:recrProcessModel}
-\end{equation}
-where the superscript $R$ refers to Recruitment, $C'_{jq,t}$ is the `effective cover' (cm$^2$) of species $j$ in quadrat $q$ at time $t$, $\varphi$ is a group dependent intercept, $\gamma$ is a random year effect, $\boldsymbol{\omega}$ is a vector of coefficients that determine the strength of intra- and interspecific density-dependence, and $\boldsymbol{C'}$ is the vector of ``effective'' cover of each species in the community.
-Following previous work \citep{adler_coexistence_2010}, we treated year as a random factor allowing intercepts to vary among years.
-
-Because plants outside the mapped quadrat could contribute recruits to the focal quadrat or interact with plants in the focal quadrat, we estimated effective cover as a mixture of the observed cover, $C$, in the focal quadrat, $q$, and the mean cover, $\bar{C}$, across the spatial location, $g$, in which the quadrat is located: $C'_{jq,t}=p_j C_{jq,t}+(1-p_j) \bar{C}_{jg,t}$, where $p$ is a mixing fraction between 0 and 1 that was estimated as part of fitting the model.
-
-In the treatment model for recruitment, a new term is added to the exponential term in the equation above, $\boldsymbol{T}\boldsymbol{\chi}_{j}^R$ where $\chi$ describes the effect of each treatment level on recruitment.
-
-Likewise in the climate model this term is replaced by $\boldsymbol{C}\xi_{j}^R$ where the $\boldsymbol{\xi}$ gives a set of coefficients for the year, and treatment specific soil moisture covariates in $\boldsymbol{C}$.
-
-We fit all vital rate models using Hamiltonian-Markov Chain Monte Carlo (HMCMC) simulations in the programs STAN 10.1 and rStan (\citep{stan} and  \citep{rstan}).
-The priors and model code are described more completely in appendix A.
-Each model was run for 2,000 iterations and four independent chains with different initial values for parameters.
-We discarded the initial 1,000 samples.
-Convergence was observed graphically for all parameters, and confirmed by assessing the split $\widehat{R}$ statistic which at convergence is equal to one \citep{stan_manual}.
-
-We fit the treatment models for species survival and growth with and without the size by treatment interactions in the treatment effect term $\chi$.
-We then judged whether including the interaction terms improved model fit by comparing the Watanabe-Aikake Information Criteria (WAIC) scores of each version of the model and retained the version with the lower WAIC score \citep{vetari_practical_2015}.
-WAIC are similar to AIC scores and allow for comparison of Bayesian models.
-Lower WAIC scores indicate a more parsimonious model.
-When a treatment model for survival or growth of a species included a size by treatment effect in $\chi$, we also included a size by soil moisture effect in the $\xi$ term in the climate model for that species and vital rate.
-This allowed us to more directly compare the predictions from the climate model to the effects in the treatment model.
-
-\subsection*{Selecting soil moisture covariates}
-
-After generating a time series of predicted daily soil moisture from SOILWAT, we averaged daily soil moisture across spring, summer and fall seasons in each year.
-We considered each of the three seasonal soil moisture variables at three different time periods relative to the demographic transition from year $t$ to year $t+1$.
-Soil moisture in the year between $t$ and $t+1$ is indicated with a "1" subscript.
-Soil moisture in the year before $t$ is indicated with a "0" subscript.
-And soil moisture preceding this year is indicated with a "lag" subscript.
-For example, for the year 2010, $spring_1$ indicates soil moisture in the spring of 2010, $spring_0$ indicates soil moisture during spring of 2009 and $spring_{lag}$ indicate soil moisture during spring 2008.
-
-We wanted to avoid fitting nine soil moisture covariates (three seasons and three lags each) for each species and vital rate, so we used only three soil moisture covariates per species and vital rate.
-We selected these three by calculating the correlations of each soil moisture variable with the random year effects from the no climate model fit and then selecting the three soil moisture variables with the strongest correlations with these year effects.
-This screening technique has been used in previous demographic studies at this site \citep{dalgleish_climate_2010} and is often used in dendrochronology to screen for potential climate influence on tree-ring growth \citep{wang_temporal_2003}.
-We felt this approach was justified because we did not make inference on these fitted parameters until after we validated their ability to predict the out of sample data in the experimental plots.
-
-\subsection*{Predicting cover from individual-based models}
-
-The vital rate regressions allow us to evaluate whether soil moisture and the experimental treatments had an effect on species performance.
-But the population response ultimately depends on the integrated effects of treatment or soil moisture on all three vital rates.
-To evaluate whether the climate models could predict the responses of these species in the drought and irrigation experiment at the overall population level we used an individual-based model (IBM) to compare observed and predicted changes in population size from one year to the next.
-
-To simulate changes in cover in each quadrat from year $t$ to year $t+1$, we initialized the IBM with the observed genet sizes and locations of the four focal species observed in year $t$ in each quadrat.
-For every individual genet in a quadrat, we projected its size and survival probability in the next year using the growth and survival models and the appropriate crowding and soil moisture or treatment covariates for that year and quadrat.
-Likewise we projected the number of new recruits in the quadrat in the next year using the recruitment model.
-We calculated the expected cover in year $t+1$ as the total area of new recruits, plus the sum of the predicted area of each existing plant at time $t+1$ multiplied by each plant's expected survival probability from time $t$ to $t+1$.
-
-We generated predictions using 1000 samples from the posterior distributions of each model parameter which allowed us to carry forward all of the uncertainty of the fitted vital rate models into our cover predictions.
-Because we were interested in comparing model predictions to observations, and were not interested in the effects of demographic stochasticity, we used a deterministic version of the models (e.g., recruitment is the $\lambda$ of (\ref{eqn:recrProcessModel}), rather than a random draw from a negative binomial distribution with a mean of $\lambda$).
-
-After generating predictions for each year from the climate and no climate models, we found the predicted quadrat-level changes in cover as $log(Cover_{t+1}/Cover_{t})$.
-
-\subsection*{Quantifying predictive accuracy} 
-We assessed the predictive performance of the climate and no climate by calculating the mean square error (MSE) between the predicted and observed responses in the experimental data as,
-
-\begin{equation}
-MSE = \frac{1}{n} \sum_{i=1}^{n} (y_i - E(y_i|\theta))^2, 
-\label{eqn:MSE}
-\end{equation}
-where $y_i$ is the outcome of observation $i$ and $E(y_i|\theta)$ gives the expected outcome given the parameters in the model $\theta$.
-The MSE is easy to interpret, but is not always appropriate for models fit with non-normal error structures \citep{gelman_understanding_2014}.
-A more general statistic for assessing model predictions is the log pointwise predictive density (lppd) \citep{gelman_understanding_2014}.
-The lppd for a given model is defined as,
-
-\begin{equation}
-lppd = \sum_{i=1}^{n} log \int p(y_i| \theta)p_{post}(\theta) d\theta, 
-\label{eqn:lppd}
-\end{equation}
-where the integral on the right side gives the probability of observing the outcome $y$ at each data point $i$ given the full posterior distribution of the parameters in the model $p_{post}(\theta)$.
-In practice we computed the lppd from the posterior simulations generated by STAN as,
-
-\begin{equation}
-\widehat{lppd} = \sum_{i=1}^{n} log \left(\frac{1}{S} \sum_{s=1}^{S} p(y_i | \theta^S) \right),
-\label{eqn:clppd}
-\end{equation}
-where the summation of $p(y_i|\theta^S)$ gives the total probability of observing the the actual response $y_i$ given the simulated posterior distribution $\theta^S$ across the full set of model simulations $S$.
-The log of this sum is then averaged across the set of all observations $i$.
-Higher lppd scores indicate that the model better predicts the observations.
-
-In addition, we evaluated whether the climate model predicted treatment effects of similar direction and magnitude to those observed in the experiment.
-We did this by extracting the soil moisture coefficients contained in $\xi$ for each of the vital rates and then multiplying those by the appropriate soil moisture covariates for each year and treatment level in the experiment.
-We then averaged these across all five years in the experiment to find the average treatment effect predicted by the climate model.
-We compared these to the posteriors of the treatment parameters, $chi$, from the treatment model.
-As a measure of agreement between our predictions and observed response we calculated the correlation between the predicted and observed treatment effects.
-
-We considered the effect of climate covariates or treatment effects to be significant when the 95\% Bayesian credible intervals on the posterior estimates did not overlap zero.
-
-All data and R code necessary to reproduce our analysis will be deposited in the Dryad Digital Repository once the manuscript is accepted.
-The current version of the computer code is available at https://github.com/pbadler/ExperimentTests/tree/master/precip and the data are available at https://bitbucket.org/ellner/driversdata.
-
-\section*{Results}
-
-\subsection*{Effects on soil moisture}
-
-Our treatments successfully changed the soil moisture in the experimental plots in the directions expected (fig.
-\ref{fig:spotVWC}).
-Spring spot measurements of soil moisture from all the plots showed that on average the drought plots were roughly 50\% drier, while irrigated plots were roughly 40\% wetter than ambient conditions (table \ref{table:spotVWC}).
-
-The continuously recorded soil moisture data also showed treatment effects, but these were weaker on average than the spot measurements and depended on season and recent rainfall (table \ref{table:soil_moisture_model}; fig \ref{fig:dailyVWC}).
-We saw weaker effects during the spring than during the fall and summer: the drought plots were about 20-30\% drier than ambient in the fall and summer but only 7 to 14\% drier during the spring, while the irrigated plots were 30\% wetter during the fall and summer but only 20-25\% wetter during the spring.
-Treatment differences were slightly larger during rainy periods, especially in the spring (rainfall effect in table \ref{table:soil_moisture_model}).
-We did not find evidence that the drought shelters and the irrigation treatments consistently affected air temperature at 30 cm above the plots.
-
-The SOILWAT soil moisture model predicted average monthly soil volumetric water content of between 10 ml/ml and 15 ml/ml each month, with the month of April being the wettest and the month of July, August and September being the driest on average.
-Annual variation in seasonal soil moisture for each year was positively correlated with seasonal precipitation and negatively correlated with seasonal temperature.
-During the course of the experiment, SOILWAT reproduced much of the daily variation observed in soil moisture recorded by our automatic data loggers, but the average soil moisture predicted by SOILWAT was about 5 ml/ml higher than the soil moisture content observed in the field.
-
-After adjusting the SOILWAT seasonal soil moisture predictions by the treatment effects, we found that the soil moisture predicted in the drought plots during the course of the experiment was well below the historical seasonal averages: the summer of 2012 and 2013, the fall of 2013, and the spring and winter of 2014 fell below the 5th percentile limit for drought in the historical period (fig \ref{fig:seasonalVWC}).
-Soil moisture in our irrigation plots was generally above the historical average soil moisture but conditions never exceeded the 90th percentile for soil moisture in the historical period (fig \ref{fig:seasonalVWC}).
-
-\subsection*{Effects on cover and vital rates}
-
-The cover of \textit{H.
-comata} and \textit{P.
-spicata} fell significantly in the drought plots from 2011 to 2016 (tables \ref{table:changeHECO}, \ref{table:changePSSP}; fig \ref{fig:coverChange}).
-The cover of \textit{P.
-secunda} showed a slight decrease in the drought plots and an increase in the irrigated plots but these changes were not significant (table \ref{table:changePOSE}).
-In contrast to the grasses, the cover of \textit{A.
-tripartita} increased slightly in all three treatments (fig \ref{fig:coverChange}).
-
-Our treatment models fit to the experimental and observational data indicated a variety of treatment effects on the vital rates of each species.
-Based on the WAIC scores with and without the size by treatment effects, we retained size by treatment effects in the growth models for \textit{A.
-tripartita} and \textit{P.
-secunda}, and the survival model for \textit{P.
-secunda}.
-For \textit{A.
-tripartita} we found significant size by treatment effects of drought: drought had positive effects on plants of average size and smaller (fig \ref{fig:growthTreat}), but plants larger than the mean size by more than 1.5 standard deviations grew slightly less in the drought treatment than in the controls.
-\textit{A.
-tripartita} showed the opposite response in the irrigated plots, (although the irrigation parameters were not significant at the 95\% confidence level): irrigation reduced growth for small plants while irrigation increased growth of plants more than 1.5 standard deviations larger than the mean size.
-Drought led to a strong (but not signficant) decrease in \textit{H.
-comata} growth, while irrigation had no effect on growth.
-Like \textit{A.
-tripartita}, we saw size by treatment effects on \textit{P.secunda} growth, with the negative effects of drought becoming greater for larger plants.
-\textit{P.
-secunda} showed the opposite response in the irrigation plots with larger plants showing the largest increase in growth in response to irrigation (although not significant).
-\textit{P.
-spicata} growth was relatively unaffected by the drought and irrigation treatments.
-
-Survival of all three grass species (fig \ref{fig:survivalTreat}) decreased in the drought plots.
-And \textit{P.
-secunda} showed a negative size by drought interaction effect: the survival of larger plants was more negatively affected by drought than that of the smaller plants.
-\textit{A.
-tripartita} survival was relatively unaffected by the drought and irrigation treatments.
-
-Recruitment in our irrigation plots was significantly less than in control plots for two grass species \textit{P.
-secunda} and \textit{P.
-spicata} (fig \ref{fig:recruitmentTreat}).
-However, recruitment was also lower in the drought plots than in the the control plots (although not significantly so), indicating that the decrease in the irrigated plots may have not been entirely due to the irrigation itself.
-The recruitment data for \textit{A.
-tripartita} were relatively limited, with only 32 new recruits in total observed in all 30 plots over the course of the five year experiment and we observed no treatment effects.
-
-Consistent with previous research most of our demographic models estimated strong negative intra-specific crowding effects and weaker negative inter-specific crowding effects on the focal species (appendix) \citep{adler_coexistence_2010,chu_direct_2016,chu_large_2015,adler_weak_2016}.
-
-\subsection*{Effects of soil moisture on vital rates}
-
-We choose three seasonal soil moisture variables for each species' based on their correlation with the random year effects in the no climate model (table \ref{table:strongCor}).
-We included size by soil moisture variables for \textit{A.
-tripartita} and  \textit{P.
-secunda} based on the treatment response we observed in the experiment.
-All three time lags and all three seasons show up in the selected variables.
-After fitting the vital rate models with the selected soil moisture variable we observed a trend towards positive soil moisture effects on growth of all three grasses (fig \ref{fig:climateGrowth}).
-For \textit{H.comata} the soil moisture of the most recent summer ($summer_1$) had a significantly positive effect while the soil moisture during $summer_0$ and $fall_lag$ were also positive but not significant.
-For \textit{A.
-tripartita} $fall_0$ and $summer_0$ had strong negative effects on growth.
-There were also strong positive size by climate interaction effects for these variables: soil moisture had a stronger negative effect on small plants and a positive effect only on the largest plants (fig \ref{fig:parPredARTRGrowth}).
-
-Soil moisture had significant effects on the survival of all four species (fig  \ref{fig:climateSurvival}).
-As for growth the grasses showed mainly positive effects while \textit{A.
-tripartita} showed a significant negative effect of $summer_0$ and a strong negative effect of $spring_0$.
-\textit{H.
-comata} showed a significant positive effect of $spring_lag$ soil moisture and a strong positive effect of $spring_0$ and $fall_1$.
-\textit{P.
-secunda} showed a significant positive effect of the previous $spring_0$ and there was an interaction between this effect and plant size: as plant size increased this effect became more positive.
-Finally for \textit{P.
-spicata} there was a significant positive effect of $spring_lag$ soil moisture on survival.
-
-There were only two significant effects of soil moisture on recruitment: $fall_lag$ soil moisture had a positive effect on \textit{P.
-secunda}, and $summer_lag$ soil moisture had a negative effect on \textit{P.
-spicata} recruitment (fig \ref{fig:climateRecruitment}).
-Soil moisture of $summer_0$ also had a strong negative effect on \textit{P.
-spicata} recruitment.
-
-The intra- and interspecific crowding effects estimated in the climate model were similar to those estimated in the treatment model (appendix).
-
-\subsection*{Evaluating the predictions}
-
-For most models adding climate covariates did not improve our ability to predict species responses in the experiment (table \ref{table:overallPreds}).
-However, the climate models did improve overall prediction MSE for growth of \textit{A.
-tripartita} and growth and survival of \textit{P.
-secunda} (table \ref{table:overallPreds}).
-In terms of lppd, the climate model outperformed the no climate model in six out of twelve models: for \textit{A.
-tripartita} growth, \textit{H.
-comata} recruitment, \textit{P.
-secunda} growth and survival and \textit{P.
-spicata} recruitment (table \ref{table:overallPreds}).
-
-When we look at the predictions for each treatment separately we see that climate covariates improved model predictions more often in the drought treatments than in the control or irrigation treatments (table \ref{table:treatmentPreds}).
-For all four species, the climate model outperformed the no climate model for predicting the response of growth to drought in terms of lppd (table \ref{table:treatmentPreds}).
-The climate model also outperformed the no climate model for predicting irrigation effects on growth for all species except \textit{H.
-comata}.
-
-Overall our climate models often predicted the correct direction of the drought and irrigation treatments (fig \ref{fig:parPredictions}).
-In four cases we both observed and predicted treatment effects significantly different from zero based on the 95\% Bayesian credible interval around the parameter mean: the drought response of \textit{H.
-comata} survival (fig \ref{fig:parPredHECOSurvival}), the drought response of \textit{P.
-secunda} growth (fig \ref{fig:parPredPOSEGrowth}), the irrigation response of \textit{P spicata} recruitment (fig \ref{fig:parPredPSSPRecruitment}) and the irrigation response of \textit{P.
-secunda} recruitment (fig \ref{fig:parPredPOSERecruitment}).
-In only one of these cases, for \textit{P.
-secunda} recruitment, was the predicted effect in the opposite direction from the observed treatment effect (fig \ref{fig:parPredictions}).
-The overall correlation between the predicted and observed treatment effects for all treatments, species and vital rates was r = 0.54, whereas the correlation for the drought treatment effects, r = 0.77, was better than for the irrigation effects, r = 0.46.
-Also the correlation between the size by climate and size by treatment effects for \textit{A.
-tripartita} growth and \textit{P.
-secunda} growth and survival was much stronger than the correlation between the intercept parameter estimates (fig \ref{fig:parPredictions}).
-
-Using the vital rate models for each species we generated one step ahead cover predictions for each quadrat in each year of the experiment.
-Average cover predicted by the climate model tended to be lower than the observed cover each year for \textit{A.
-tripartita} and \textit{P.secunda} (fig \ref{fig:coverPred}).
-Comparing the overall population growth rates predicted to those observed in the experiment, we see that the MSE of the climate model was lower than the MSE of the no climate model for \textit{P.
-secunda} and \textit{P.
-spicata} (table \ref{table:corPGR}).
-The predictions produced by the climate model for these species were also slightly more correlated with the observations than the predictions produced by the no climate model (table \ref{table:corPGR}).
-Considering each treatment and species separately, the predicted population growth rates for \textit{A.
-tripartita}, \textit{P.
-secunda} and \textit{P.
-spicata} were all consistently lower than the observed population growth rates (figs \ref{fig:pgrARTR}, \ref{fig:pgrPOSE}, \ref{fig:pgrPSSP}).
-The climate model showed lower MSE for \textit{A.
-tripartita}, \textit{P.
-secunda} and \textit{P.
-spicata} in the irrigation treatment, \textit{P.
-spicata} in the control treatment and \textit{H.
-comata} in the drought treatment (fig \ref{fig:pgrHECO}).
-However, the correlations between the predicted and observed log changes in cover did not always show the same pattern as MSE: the climate model made more strongly correlated predictions with the observations than the no climate model only for \textit{P.
-spicata} and \textit{P.
-secunda} in the control treatment and \textit{P.
-secunda} and \textit{H.
-comata} in the drought treatment.
-
-\section*{Discussion}
-
-Our experiment showed that observational data on the response of plant populations to interannual climate variation can indeed help us predict the direction of species responses to experimental climate manipulations (fig \ref{fig:parPredictions}).
-This was true even though adding climate parameters to the demographic models only improved vital rate predictions for half of the models (fig \ref{table:overallPreds}).
-This should give us some hope that even when climate effects in demographic models fit to observational data are weak, they may contain useful qualitative information on the direction of climate effects in the future.
-
-* discuss biggest successes--astreces in fig \ref{fig:parPredictions}
-* discuss biggest mistake, P. secunda recruitment went in wrong direction.
-
-Scaling up to the population level, the climate models only produced better one step a head predictions of the overall response of the species to the experiment for two species: \textit{P.
-spicata} and \textit{P.
-secunda} (table \ref{table:corPGR}).
-However adding climate effects produced better population-level predictions in the drought treatments for \textit{P.
-secunda} and \textit{H.
-comata}.
-Both the survival and the growth of these species were positively affected by increased soil moisture (fig \ref{fig:climateGrowth}, \ref{fig:climateSurvival}), and so it makes intuitive sense that we would observe declines in these species cover in the drought treatment (fig \ref{fig:coverChange}).
-
-*Among our species, we had the most success predicting the response of the three grass species and less success predicting the response of the shrub species \textit{A.
-tripartita}.
-Why might ARTR be different?
-
-*We also had more success predicting species response to the drought treatment than to the irrigation treatment.
-We also tended to see species respond more strongly to the drought treatment than they did to the irrigation treatment.
-This tells us that in some cases, demographic models based on observational data may be of more use for prediction in extreme conditions than when conditions are close to average.
-Our drought treatment likely created extremely low soil moisture compared to the historical average \ref{fig:seasonalVWC}.
-We hypothesize that this made water the most limiting resource for plants in these plots.
-In contrast, in the control and irrigated plots soil moisture may not have always been limiting during the course of the experiment.
-It may make sense then that our climate models did not make more accurate predictions in these conditions than the no climate model.
-
-*Mixed success at predicting species responses to high and low moisture availability has interesting implications.
-On the one hand it is re-assuring that observational data is sometimes useful.
-On the other hand, for predictions to be truly useful we would also like more information to help us sort out why our predictions for some species were good and for others no better than a no climate model.
-In other words, we have little ability to predict when our predictions are likely to be accurate.
-Among plants detailed physiological ecology may give us a guide to which factors and climate conditions are likely to affect which species.
-In our system, for instance, it may make sense that the grasses showed a stronger response to the drought treatment than the a woody shrub.
-Although these grasses are adapted to the arid conditions that characterize the sagebrush steppe, they thrive during the brief window in spring and early summer when the soil moisture and temperatures are warmer.
-\textit{A.
-tripartita} on the other hand grows throughout the summer and generally has deeper roots than the grasses.
-These traits may help it tolerate the water stress induced by drought.
-
-*Will our predictions be useful for the longer term?
-Because these species compete, one could argue that our predictions for any one species in this community will only be as good as the predictions we make for their competitors.
-For instance, while we observed little effect of drought on \textit{A.
-tripartita} in our experiment, it is possible that it will eventually respond positively to the drought treatment as cover of the grass species it comptetes with declines (fig \ref{fig:coverChange}).
-We know that grass species compete strongly with \textit{A.
-tripartita}.
-However, we also know from previous work in this community that each species is more limited by intra-specific competition than by inter-specific crowding.
-This fact ensures that the direct effects of climate change will generally be greater than the competition mediated indirect effects of climate change \cite{chu_direct_2016, kleinhesselink_indirect_2015}.
-
-*Our results give us more confidence that historical observational data can in theory be used to detect and predict the demographic effects of climate change.
-This should encourage more researchers to try and use observational data to predict the future in both experimental and natural settings.
-Nevertheless, our success at predicting the short-term response of two species to a small-scale climate manipulation is not likely to be very reassuring to applied ecologists and resource managers wishing to make accurate medium to long-term quantitative predictions about the effects of climate change on the species they manage.
-Clearly more work is needed to distinguish which predictions are worth having confidence in and which predictions we should have less confidence in.
-Towards that goal, perhaps the best way to build confidence in ecological predictions is conduct more tests like this one.
-
-\section*{Acknowledgements}
-
-Funding was provided by NSF grants DEB-1353078, and DEB-1054040 to PBA, an NSF GRFP award to AK, and by the Utah Agricultural Experiment Station (get journal paper number).
-The USDA-ARS Sheep Experiment Station generously provided access to historical data and the field experiment site.
-Laureano Gherardi provided us with valuable advice in constructing the rain-out and irrigation experiment.
-
-\newpage
-\bibliographystyle{ecology}
-\bibliography{precip_experiment}
-
-\end{doublespacing}
-
-\clearpage
-\newpage
-
-\section*{Tables}
-
-\begin{table}[h]
-	\caption{Treatment effects on spring soil moisture}
+% latex table generated in R 3.3.1 by xtable 1.8-2 package
+% Thu Dec 22 04:07:36 2016
+\begin{table}[!h]
 \centering
-	%\begin{center}
-	\begin{tabular}{l c c }
-		\hline
-		& Model 1 \\
-		\hline
-		(Intercept)                  & $8.81^{*}$    	& $[5.78;\ 11.83]$  \\
-		TreatmentDrought             & $-3.97^{*}$      & $[-4.84;\ -3.09]$ \\
-		TreatmentIrrigation          & $3.26^{*}$       & $[2.39;\ 4.14]$   \\
-		\hline
-		AIC                          & 3191.87           \\
-		BIC                          & 3222.92           \\
-		Log Likelihood               & -1588.93          \\
-		Num. obs.                    & 624               \\
-		Num. groups: plot            & 24                \\
-		Num. groups: PrecipGroup     & 8                 \\
-		Num. groups: date            & 5                 \\
-		Var: plot (Intercept)        & 0.45              \\
-		Var: PrecipGroup (Intercept) & 0.23              \\
-		Var: date (Intercept)        & 11.24             \\
-		Var: Residual                & 8.90              \\
-		\hline
-		\multicolumn{2}{l}{\scriptsize{$^*$ 0 outside the confidence interval}}
-	\end{tabular}
-	\label{table:spotVWC}
-	%\end{center}
-\end{table}
-
-\begin{table}
-	\caption{Model of treatment effects on soil moisture}
-	\begin{center}
-		\begin{tabular}{l c c }
-			\hline
-			& Model 1 \\
-			\hline
-			(Intercept)                       & $-0.57^{*}$       & $[-0.89;\ -0.26]$ \\
-			TreatmentIrrigation               & $1.23^{*}$        & $[1.18;\ 1.29]$   \\
-			rainfallrainy                     & $-0.05$           & $[-0.12;\ 0.01]$  \\
-			seasonspring                      & $0.27^{*}$        & $[0.23;\ 0.32]$   \\
-			seasonsummer                      & $0.15^{*}$        & $[0.10;\ 0.19]$   \\
-			seasonwinter                      & $0.25^{*}$        & $[0.21;\ 0.29]$   \\
-			TreatmentIrrigation:rainfallrainy & $0.18^{*}$        & $[0.13;\ 0.24]$   \\
-			TreatmentIrrigation:seasonspring  & $-0.23^{*}$       & $[-0.29;\ -0.16]$ \\
-			TreatmentIrrigation:seasonsummer  & $-0.26^{*}$       & $[-0.33;\ -0.20]$ \\
-			TreatmentIrrigation:seasonwinter  & $-0.33^{*}$       & $[-0.40;\ -0.27]$ \\
-			rainfallrainy:seasonspring        & $-0.23^{*}$       & $[-0.31;\ -0.16]$ \\
-			rainfallrainy:seasonsummer        & $-0.07$           & $[-0.14;\ 0.01]$  \\
-			rainfallrainy:seasonwinter        & $-0.07$           & $[-0.20;\ 0.07]$  \\
-			\hline
-			AIC                               & 14581.58          \\
-			BIC                               & 14695.49          \\
-			Log Likelihood                    & -7274.79          \\
-			Num. obs.                         & 9133              \\
-			Num. groups: date                 & 1596              \\
-			Num. groups: PrecipGroup          & 8                 \\
-			Var: date (Intercept)      		  & 0.00              \\
-			Var: PrecipGroup (Intercept)      & 0.19              \\
-			Var: Residual                     & 2.50              \\
-			\hline
-			\multicolumn{2}{l}{\scriptsize{$^*$ 0 outside the confidence interval}}
-		\end{tabular}
-		\label{table:soil_moisture_model}
-	\end{center}
-\end{table}
-
-% latex table generated in R 3.3.1 by xtable 1.8-2 package
-% Mon Dec 19 18:10:15 2016
-\begin{table}[ht]
-	\centering
-	\begin{tabular}{rrrrr}
-		\hline
-		& Estimate & Std. Error & t value & Pr($>$$|$t$|$) \\ 
-		\hline
-		(Intercept) & -0.2835 & 0.4627 & -0.61 & 0.5487 \\ 
-		TreatmentDrought & 0.9378 & 0.7743 & 1.21 & 0.2434 \\ 
-		TreatmentIrrigation & 0.8882 & 0.7743 & 1.15 & 0.2682 \\ 
-		\hline
-	\end{tabular}
-	\caption{Treatment effects on log cover change for \textit{A. tripartita} from 2011 to 2016. Intercept gives control effects.} 
-	\label{table:changeARTR}
-\end{table}
-
-% latex table generated in R 3.3.1 by xtable 1.8-2 package
-% Mon Dec 19 18:10:16 2016
-\begin{table}[ht]
-	\centering
-	\begin{tabular}{rrrrr}
-		\hline
-		& Estimate & Std. Error & t value & Pr($>$$|$t$|$) \\ 
-		\hline
-		(Intercept) & 0.3982 & 0.2587 & 1.54 & 0.1548 \\ 
-		TreatmentDrought & -2.9951 & 0.5784 & -5.18 & 0.0004 \\ 
-		TreatmentIrrigation & -0.1219 & 0.4953 & -0.25 & 0.8105 \\ 
-		\hline
-	\end{tabular}
-	\caption{Treatment effects on log cover change for \textit{H. comata} from 2011 to 2016. Intercept gives control effects.} 
-	\label{table:changeHECO}
-\end{table}
-
-% latex table generated in R 3.3.1 by xtable 1.8-2 package
-% Mon Dec 19 18:10:16 2016
-\begin{table}[ht]
-	\centering
-	\begin{tabular}{rrrrr}
-		\hline
-		& Estimate & Std. Error & t value & Pr($>$$|$t$|$) \\ 
-		\hline
-		(Intercept) & -0.7247 & 0.4613 & -1.57 & 0.1298 \\ 
-		TreatmentDrought & 0.0273 & 0.8208 & 0.03 & 0.9737 \\ 
-		TreatmentIrrigation & 1.1459 & 0.7797 & 1.47 & 0.1552 \\ 
-		\hline
-	\end{tabular}
-	\caption{Treatment effects on log cover change for \textit{P. secunda} from 2011 to 2016. Intercept gives control effects.} 
-	\label{table:changePOSE}
-\end{table}
-
-% latex table generated in R 3.3.1 by xtable 1.8-2 package
-% Mon Dec 19 18:10:16 2016
-\begin{table}[ht]
-	\centering
-	\begin{tabular}{rrrrr}
-		\hline
-		& Estimate & Std. Error & t value & Pr($>$$|$t$|$) \\ 
-		\hline
-		(Intercept) & 0.0188 & 0.2124 & 0.09 & 0.9303 \\ 
-		TreatmentDrought & -0.8851 & 0.3780 & -2.34 & 0.0287 \\ 
-		TreatmentIrrigation & 0.1453 & 0.3780 & 0.38 & 0.7044 \\ 
-		\hline
-	\end{tabular}
-	\caption{Treatment effects on log cover change for \textit{P. spicata} from 2011 to 2016. Intercept gives control effects.} 
-	\label{table:changePSSP}
-\end{table}
-
-% latex table generated in R 3.3.1 by xtable 1.8-2 package
-% Tue Dec 20 12:39:32 2016
-\begin{table}[ht]
-	\centering
-	\begin{tabular}{lllrrrr}
-		\hline
-		vital\_rate & species & climate variable & Int. cor. & p val. & Size cor. & Size p. val. \\ 
-		\hline
-		growth & ARTR & su.0 & -0.49 & 0.02 & 0.26 & 0.26 \\ 
-		growth & ARTR & f.0 & -0.28 & 0.23 & 0.40 & 0.08 \\ 
-		growth & ARTR & sp.1 & 0.17 & 0.45 & -0.33 & 0.14 \\ 
-		growth & HECO & su.1 & 0.69 & 0.00 &  &  \\ 
-		growth & HECO & su.0 & 0.50 & 0.02 &  &  \\ 
-		growth & HECO & f.lag & 0.37 & 0.10 &  &  \\ 
-		growth & POSE & f.lag & 0.31 & 0.17 & -0.11 & 0.64 \\ 
-		growth & POSE & su.lag & 0.29 & 0.20 & -0.20 & 0.38 \\ 
-		growth & POSE & sp.1 & 0.26 & 0.25 & -0.20 & 0.38 \\ 
-		growth & PSSP & f.lag & 0.34 & 0.13 &  &  \\ 
-		growth & PSSP & su.lag & 0.25 & 0.27 &  &  \\ 
-		growth & PSSP & f.0 & -0.22 & 0.34 &  &  \\ 
-		recruitment & ARTR & su.lag & -0.32 & 0.16 &  &  \\ 
-		recruitment & ARTR & su.0 & -0.26 & 0.25 &  &  \\ 
-		recruitment & ARTR & sp.1 & 0.22 & 0.34 &  &  \\ 
-		recruitment & HECO & su.lag & -0.31 & 0.18 &  &  \\ 
-		recruitment & HECO & su.0 & -0.30 & 0.18 &  &  \\ 
-		recruitment & HECO & f.lag & 0.19 & 0.40 &  &  \\ 
-		recruitment & POSE & sp.1 & 0.49 & 0.02 &  &  \\ 
-		recruitment & POSE & f.lag & 0.34 & 0.13 &  &  \\ 
-		recruitment & POSE & f.1 & 0.32 & 0.16 &  &  \\ 
-		recruitment & PSSP & su.lag & -0.52 & 0.02 &  &  \\ 
-		recruitment & PSSP & su.0 & -0.48 & 0.03 &  &  \\ 
-		recruitment & PSSP & sp.0 & 0.30 & 0.19 &  &  \\ 
-		survival & ARTR & su.0 & -0.60 & 0.00 &  &  \\ 
-		survival & ARTR & sp.0 & -0.41 & 0.06 &  &  \\ 
-		survival & ARTR & su.1 & -0.40 & 0.07 &  &  \\ 
-		survival & HECO & sp.0 & 0.44 & 0.04 &  &  \\ 
-		survival & HECO & sp.lag & 0.43 & 0.05 &  &  \\ 
-		survival & HECO & f.1 & 0.33 & 0.14 &  &  \\ 
-		survival & POSE & sp.0 & 0.44 & 0.04 & 0.22 & 0.34 \\ 
-		survival & POSE & sp.1 & 0.27 & 0.23 & -0.46 & 0.04 \\ 
-		survival & POSE & f.lag & -0.00 & 0.99 & 0.30 & 0.19 \\ 
-		survival & PSSP & sp.0 & 0.36 & 0.11 &  &  \\ 
-		survival & PSSP & sp.lag & 0.34 & 0.13 &  &  \\ 
-		survival & PSSP & su.1 & 0.26 & 0.26 &  &  \\ 
-		\hline
-	\end{tabular}
-	\caption{Selected climate variables for each vital rate model for each species.
-Correlations and p-values between the choosen variables and the intercept of the no climate model are shown.
-For ARTR growth and POSE growth and survival, the correlations between the year effects on size and the soil moisture variables are also given.
-"f" = fall, "su" = summer, "sp" = spring.
+\caption{MSE of predicted log cover changes and correlations between log cover changes predicted and observed.
+Predictions for the cover changes in the experimental plots were generated either from the year effects or the climate models.
+Instances where the climate model made better predictions than the year effects model are indicated with the "***".
 ARTR = \textit{A.
@@
 spicata}.}
-	\label{table:strongCor}
-\end{table}
-
-% latex table generated in R 3.3.1 by xtable 1.8-2 package
-% Tue Dec 20 12:38:30 2016
-\begin{table}[ht]
-	\centering
-	\begin{tabular}{lllrrrl}
-		\hline
-		species & vital\_rate & score & climate model & no climate model & diff & improved \\ 
-		\hline
-		ARTR & growth & lppd & -186.08 & -189.17 & 3.09 & *** \\ 
-		ARTR & growth & MSE & 0.55 & 0.55 & -0.00 & *** \\ 
-		ARTR & recruitment & lppd & -79.05 & -77.81 & -1.24 &  \\ 
-		ARTR & recruitment & MSE & 169.86 & 9.47 & 160.39 &  \\ 
-		ARTR & survival & lppd & -37.55 & -36.42 & -1.13 &  \\ 
-		ARTR & survival & MSE & 0.06 & 0.06 & 0.00 &  \\ 
-		HECO & growth & lppd & -475.54 & -454.36 & -21.18 &  \\ 
-		HECO & growth & MSE & 1.26 & 1.18 & 0.09 &  \\ 
-		HECO & recruitment & lppd & -149.43 & -151.60 & 2.17 & *** \\ 
-		HECO & recruitment & MSE & 288.16 & 235.57 & 52.59 &  \\ 
-		HECO & survival & lppd & -158.99 & -147.06 & -11.93 &  \\ 
-		HECO & survival & MSE & 0.12 & 0.11 & 0.01 &  \\ 
-		POSE & growth & lppd & -1823.71 & -1831.56 & 7.85 & *** \\ 
-		POSE & growth & MSE & 1.72 & 1.73 & -0.01 & *** \\ 
-		POSE & recruitment & lppd & -260.05 & -257.46 & -2.59 &  \\ 
-		POSE & recruitment & MSE & 45.40 & 37.32 & 8.08 &  \\ 
-		POSE & survival & lppd & -698.06 & -718.60 & 20.54 & *** \\ 
-		POSE & survival & MSE & 0.14 & 0.14 & -0.00 & *** \\ 
-		PSSP & growth & lppd & -1232.93 & -1237.92 & 4.99 & *** \\ 
-		PSSP & growth & MSE & 1.51 & 1.51 & 0.00 &  \\ 
-		PSSP & recruitment & lppd & -271.34 & -273.94 & 2.60 & *** \\ 
-		PSSP & recruitment & MSE & 79.09 & 42.68 & 36.41 &  \\ 
-		PSSP & survival & lppd & -332.47 & -307.26 & -25.21 &  \\ 
-		PSSP & survival & MSE & 0.11 & 0.10 & 0.01 &  \\ 
-		\hline
-	\end{tabular}
-	\caption{Comparison of model predictions from climate model and no climate model for each species and vital rate.
-Two prediction scores are reported, MSE and lppd.
-Lower MSE indicates improved predictions whereas higher lppd indicates improved predictions.
-Instances where the climate model outperformed the random no climate model are marked with "***" in the last column.
-ARTR = \textit{A.
-tripartita}, HECO = \textit{H.
-comata}, POSE = \textit{P.
-secunda}, PSSP = \textit{P.
-spicata}.}
-	\label{table:overallPreds}
-\end{table}
-
-% latex table generated in R 3.3.1 by xtable 1.8-2 package
-% Tue Dec 20 19:24:15 2016
-\begin{table}[ht]
-	\centering
+\label{table:corPGR}
 \begin{tabular}{rllrrrl}
   \hline
-		& species & stat & no climate model & climate model & diff & improved \\ 
+ & species & stat & year effects model & climate model & diff & improved \\ 
   \hline
@@
 \end{tabular}
-	\caption{MSE of predicted log cover changes and correlations between log cover changes predicted and observed.
-Predictions for the cover changes in the experimental plots were generated either from the no climate or the climate models.
-Instances where the climate model made better predictions than the no climate model are indicated with the "***".
-ARTR = \textit{A.
-tripartita}, HECO = \textit{H.
-comata}, POSE = \textit{P.
-secunda}, PSSP = \textit{P.
-spicata}.}
-	\label{table:corPGR}
 \end{table}
-
-\clearpage
-\newpage
-
-\section*{Figures}
-
-\begin{figure}[!htbp]
-	\centering
-	\includegraphics[width=1\textwidth]{VWC_spot_measurements}
-	\caption{Soil moisture in the upper 5 cm of drought and irrigated plots compared to ambient controls.
-Soil moisture was measured at six locations around each plot at five different dates during the spring.
-Control plots were nearby areas of experiencing ambient soil moisture.
-Box plots show the median soil moisture and the interquartile range.
-Dots show individual soil moisture measurements.
-Readings of volumetric soil moisture less than zero were occasionally obtained in very dry soil.}
-	\label{fig:spotVWC}
-\end{figure}
-
-\begin{figure}[!htbp]
-	\centering
-	\includegraphics[width=1\textwidth]{avg_daily_soil_moisture}
-	\caption{Average soil moisture in the control, drought, and irrigation treatments during each year of the experiment.
-Soil moisture was monitored in four drought plots, four irrigated plots and four ambient control plots.
-Two sensors were installed at 5 cm depth at each plot and two at 25 cm and data was logged every 2 hours.}
-	\label{fig:dailyVWC}
-\end{figure}
-
-\begin{figure}[!htbp]
-	\centering
-	\includegraphics[width=1\textwidth]{modern_soil_moisture_comparison}
-	\caption{Average seasonal soil moisture in the control, drought, and irrigation treatments during each year of the experiment.
-The dashed gray lines give the 5th percentile and 95th percentile limits for seasonal soil moisture in the historical record (1929 to 2010).
-}
-	\label{fig:seasonalVWC}
-\end{figure}
-
-\begin{figure}[!htbp]
-	\centering
-	\includegraphics[width=1\textwidth]{start_to_finish_cover_change}
-	\caption{Log change in cover in each of the experimental plots from the pre-treatment monitoring in 2011 to the last year of the experiment in 2016.
-Box plots show the median cover change and the interquartile range.
-ARTR = \textit{A.
-tripartita}, HECO = \textit{H.
-comata}, POSE = \textit{P.
-secunda}, PSSP = \textit{P.
-spicata}.}
-	\label{fig:coverChange}
-\end{figure}
-
-\begin{figure}[!htbp]
-	\centering
-	\includegraphics[width=1\textwidth]{treatment_effect_growth}
-	\caption{Parameter estimates for the effects of treatment on growth of all four species.
-We assessed a parameter as significant when the 95\% Bayesian credible intervals did not overlap zero.
-Size by treatment interactions were only fit for ARTR, and POSE.
-Plant size was centered on mean size and scaled by its standard deviation.
-ARTR = \textit{A.
-tripartita}, HECO = \textit{H.
-comata}, POSE = \textit{P.
-secunda}, PSSP = \textit{P.
-spicata}.
-}
-	\label{fig:growthTreat}
-\end{figure}
-
-\begin{figure}[!htbp]
-	\centering
-	\includegraphics[width=1\textwidth]{treatment_effect_survival}
-	\caption{Parameter estimates for the effects of treatment on survival of all four species.
-We assessed a parameter as significant when the 95\% Bayesian credible intervals did not overlap zero.
-Size by treatment interactions were only fit for POSE.
-Plant size was centered on mean size and scaled by its standard deviation.
-ARTR = \textit{A.
-tripartita}, HECO = \textit{H.
-comata}, POSE = \textit{P.
-secunda}, PSSP = \textit{P.
-spicata}.
-}
-	\label{fig:survivalTreat}
-\end{figure}
-
-\begin{figure}[!htbp]
-	\centering
-	\includegraphics[width=1\textwidth]{treatment_effect_recruitment}
-	\caption{Parameter estimates for the effects of treatment on recruitment of all four species.
-We assessed a parameter as significant when the 95\% Bayesian credible intervals did not overlap zero.
-ARTR = \textit{A.
-tripartita}, HECO = \textit{H.
-comata}, POSE = \textit{P.
-secunda}, PSSP = \textit{P.
-spicata}.
-}
-	\label{fig:recruitmentTreat}
-\end{figure}
-
-\begin{figure}[!htbp]
-	\centering
-	\includegraphics[width=1\textwidth]{climate_effect_growth}
-	\caption{Parameter estimates for the selected seasonal soil moisture covariates on the growth of all four species.
-Parameters are ordered chronologically from most recent to the current growing season on the right to most distant on the left.
-Dark blue parameters show size x climate interaction effects.
-We assessed a parameter as significant when the 95\% Bayesian credible intervals did not overlap zero.
-ARTR = \textit{A.
-tripartita}, HECO = \textit{H.
-comata}, POSE = \textit{P.
-secunda}, PSSP = \textit{P.
-spicata}.
-}
-	\label{fig:climateGrowth}
-\end{figure}
-
-\begin{figure}[!htbp]
-	\centering
-	\includegraphics[width=1\textwidth]{climate_effect_survival}
-	\caption{Parameter estimates for the selected seasonal soil moisture covariates on the survival of all four species.
-Parameters are ordered chronologically from most recent to the current growing season on the right to most distant on the left.
-Dark blue parameters show size x climate interaction effects.
-We assessed a parameter as significant when the 95\% Bayesian credible intervals did not overlap zero.
-ARTR = \textit{A.
-tripartita}, HECO = \textit{H.
-comata}, POSE = \textit{P.
-secunda}, PSSP = \textit{P.
-spicata}.
-}
-	\label{fig:climateSurvival}
-\end{figure}
-
-\begin{figure}[!htbp]
-	\centering
-	\includegraphics[width=1\textwidth]{climate_effect_recruitment}
-	\caption{Parameter estimates for the selected seasonal soil moisture covariates on the survival of all four species.
-Parameters are ordered chronologically from most recent to the current growing season on the right to most distant on the left.
-Dark blue parameters show size x climate interaction effects.
-We assessed a parameter as significant when the 95\% Bayesian credible intervals did not overlap zero.
-ARTR = \textit{A.
-tripartita}, HECO = \textit{H.
-comata}, POSE = \textit{P.
-secunda}, PSSP = \textit{P.
-spicata}.
-}
-	\label{fig:climateRecruitment}
-\end{figure}
-
-\begin{figure}[!htbp]
-	\centering
-	\includegraphics[width=1\textwidth]{parameter_predictions}
-	\caption{The treatment effects predicted by the climate model compared to the treatment effects observed for the intercept parameters (left side) and size by climate/treatment effects (right side).
-The observed treatment effects come from the treatment model fitted to all the data including the five years of the experiment.
-The predicted parameters come from the climate model fitted to all years of observational data but do not include the five years of the experiment (2011 to 2016).
-Treatment parameters that were both observed and predicted to be significantly different from zero are shown with the "*" symbol.
-The correlation between predicted and observed parameters is given on each panel.
-\textit{P.
-secunda} recruitment was predicted to be positively affected by the irrigation treatment but was in fact negatively affected.
-The other significant effects were in the correct direction.
-We assessed a parameter as significant when the 95\% Bayesian credible intervals did not overlap zero.}
-	\label{fig:parPredictions}
-\end{figure}
-
-\begin{figure}[!htbp]
-	\centering
-	\includegraphics[width=1\textwidth]{predicted_and_observed_cover}
-	\caption{Observed average cover per quadrat in the experimental and control plots and one step ahead cover predictions from the climate model.
-Cover predictions for each year are generated from the IBM based on the observed distribution of plants in each quadrat in the current year.
-Quadrat cover was not predicted for the first year of the experiment in 2011.
-Note the different cover scales for ARTR and the three grass species.
-ARTR = \textit{A.
-tripartita}, HECO = \textit{H.
-comata}, POSE = \textit{P.
-secunda}, PSSP = \textit{P.
-spicata}.}
-	\label{fig:coverPred}
-\end{figure}
-
-\begin{figure}[!htbp]
-	\centering
-	\includegraphics[width=1\textwidth]{ARTR_predicted_pgr_comparison}
-	\caption{Observed and predicted one step ahead log change in \textit{A.
-tripartita} cover in the experiment.
-Changes in cover predicted by the climate model are shown on the left and those predicted by the no climate model are shown on the right.
-Correlations coefficients between predictions and observations and MSE are shown for each treatment and model.}
-	\label{fig:pgrART}
-\end{figure}
-
-\begin{figure}[!htbp]
-	\centering
-	\includegraphics[width=1\textwidth]{HECO_predicted_pgr_comparison}
-	\caption{Observed and predicted one step ahead log change in \textit{H.
-comata} cover in the experiment.
-Changes in cover predicted by the climate model are shown on the left and those predicted by the no climate model are shown on the right.
-Correlations coefficients between predictions and observations and MSE are shown for each treatment and model.
-}
-	\label{fig:pgrHECO}
-\end{figure}
-
-\begin{figure}[!htbp]
-	\centering
-	\includegraphics[width=1\textwidth]{POSE_predicted_pgr_comparison}
-	\caption{Observed and predicted one step ahead log change in \textit{P.
-secunda} cover in the experiment.
-Changes in cover predicted by the climate model are shown on the left and those predicted by the no climate model are shown on the right.
-Correlations coefficients between predictions and observations and MSE are shown for each treatment and model.}
-	\label{fig:pgrPOSE}
-\end{figure}
-
-\begin{figure}[!htbp]
-	\centering
-	\includegraphics[width=1\textwidth]{PSSP_predicted_pgr_comparison}
-	\caption{Observed and predicted one step ahead log change in \textit{P.
-spicata} cover in the experiment.
-Changes in cover predicted by the climate model are shown on the left and those predicted by the no climate model are shown on the right.
-Correlations coefficients between predictions and observations and MSE are shown for each treatment and model.}
-	\label{fig:pgrPSSP}
-\end{figure}
-
-\clearpage 
-\newpage
-
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-% APPENDICES !
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-
-\clearpage 
-\newpage
-
-\setcounter{page}{1}
-\setcounter{equation}{0}
-\setcounter{figure}{0}
-\setcounter{section}{0}
-\setcounter{table}{0}
-\renewcommand{\theequation}{SI.\arabic{equation}}
-\renewcommand{\thetable}{SI-\arabic{table}}
-\renewcommand{\thefigure}{SI-\arabic{figure}}
-\renewcommand{\thesection}{Section SI.\arabic{section}}
-
-\centerline{\Large \textbf{Supporting Information }}
-\centerline{Adler et al., ``Weak interspecific interactions''}
-
-\vspace{0.4in}
-
-\section{Supplementary Methods} \label{suppMethods}
-
-\subsection{Interspecific covariance in local crowding} 
-We explored interspecific covariance in local crowding experienced by individual plants, by regressing the $W$ values exerted by one neighbor species, the response variable, against the $W$ values of all other species, the independent variables.
-Because some $W=0$, we conducted two separate regressions.
-First, using all $W$'s, we fitted a generalized linear model with a logit link function to evaluate whether the probability that the focal species' $W=0$ is influenced by the value of other species' $W$'s.
-In this model, the dependent variable is a Bernoulli variate coding for the zero or non-zero value of the focal species' crowding, and the independent variables are the $W$'s for all other species.
-Second, for the set of records in which the focal species has $W>0$, we performed a linear regression, where the focal species' $W$ is the dependent variable, and the other species' $W$'s are the independent variables.
-We repeated these regressions for each focal species.
-Due to large samples size, interspecific $W$ values were often statistically significant predictors of intraspecific.
-However, they explained very little variance.
-The maximum reduction in deviance for the generalized linear regressions and $R^2$ for the linear regressions were both less than 8\%.
-The \texttt{R} code for this analysis is included as ..\texttt{\textbackslash Wdistrib\textbackslash exploreSurvivalWs.r}.
-
-\subsection{Mean field approximation of local crowding for the IPM} 
-\citet{adler_coexistence_2010} developed a mean field approximation for local crowding when the
-competition kernels are all Gaussian functions, $F_{jm}(d) = e^{-\alpha_{jm} d^2}$. The approximation is explained in 
-the online SI to \citet{adler_coexistence_2010} and in section 5.3 of \citet{Ellner2016}. 
-Here we explain how that approximation was modified for the IPMs in this paper, which
-used fitted nonparametric competition kernels.
-
-For $j \ne m$ (between-species competition), overlap between individuals is allowed. The mean field approximation is 
-that from the perspective of any focal plant in species $j$, individuals of species $m$ are distributed at random in space, 
-independent of each other and of their size.
-
-Consider the region between the circles of radius $x$ and $x+dx$ centered on a focal genet of species $j$. The area of this annulus
-is $2 \pi x \; dx$  to leading order for $dx \approx 0$. A species $m$ genet 
-in the annulus puts competitive pressure $F_{jm}(x)$ times its area on
-the focal genet. The expected total competitive pressure from all such genets 
-is therefore is $F_{jm}(x) 2 \pi x \; dx$ times the expected fractional cover of species $m$ in the annulus 
-(fractional cover is the total area of species $m$ genets, as a fraction of the total area). The excepted fractional cover $C_m$ of species $m$
-in the annulus equals its fractional cover in the habitat as a whole, because of the assumption of random distribution
-spatial distributions. We therefore have $C_m  = \int e^u n_m(u,t) du/A$ where $A$ is the total area of the habitat. 
-The total expected competitive pressure on a species-$j$ genet due to species $m$ is then 
-\begin{equation}
-W_{jm} = \int_0^\infty{C_m F_{jm}(x) 2 \pi x \; dx}  = C_m \left [2 \pi \int_0^{\infty} x F(x) \, dx \right ].
-\label{eqn:wbarm}
-\end{equation} 
-The quantity in square brackets is a constant (that is, it only depends on what the kernel function
-is) so it can be computed once and for all for each kernel used in the IPM. The integral is finite because
-all fitted kernels fall to zero at a finite distance from the focal plant.
-
-Our kernel fitting method only uses competition kernel values at the ``mid-ring'' distances
-halfway between the inner and outer radii of a series of annuli around each focal
-plant, scaled so that the value at the innermost mid-ring distance equals 1. 
-In the IPM we defined the kernel at other distances by linear interpolation between values at 
-mid-ring distances, except that for the innermost ring a kernel value of 1 was specified at the
-outer radius of the ring and at distance $x=0$.
-
-Now consider within-species competition. We assume that conspecifics cannot overlap. Genet shapes are irregular, but we 
-nonetheless implement the no-overlap rule by assuming that a genet of log area $u_i$ is a 
-circle of radius $r_i$ where $\pi r_i^2 = e^{u_i}$. The no-overlap rule is then that the centroids of two conspecific individuals 
-must be separated by at least the sum of their radii.
-
-For any one focal genet, the no-overlap restriction on its neighbors' locations affects 
-only a negligibly small part of the habitat. The expected cover of individuals in the places
-where they can occur (relative to one focal plant) is thus assumed to equal their expected locations
-in the habitat as a whole.
-
-Let $C_m(u)$ be the total cover of species $m$ genets of radius $r$ or smaller, 
-\begin{equation}
-C_m(r) = \int_L^{\log(\pi r^2)}{\! \! \! e^z n_m(z,t) \, dz} .
-\label{eqn:cm}
-\end{equation}
-A focal genet of radius $r$ cannot have any conspecific neighbors centered 
-at distances less than $r$. It can have a neighbor centered at distance $x>r$ if that neighbor's
-radius is no more than $x-r$. Adding up the expected cover of all such possible neighbors
-for a focal genet of radius $r$,    
-\begin{equation}
-W_{mm}(r) = 2 \pi \int_r^{\infty}F_{mm}(x) x C_m(x-r) \, dx
-\label{eqn:wbarmr} 
-\end{equation}
-This integral is again finite and computable because the kernels $F$ fall to 0 at finite $x$.
-
-\clearpage 
-\newpage  
-\section{Additional Tables} 
-% latex table generated in R 3.3.1 by xtable 1.8-2 package
-% Tue Dec 20 13:32:53 2016
-\begin{longtable}{rllllrrrl}
-	\hline
-	& species & vital\_rate & Treatment & score & climate model & no climate model & diff & improved \\ 
-	\hline
-	\endhead
-	\hline
-	\multicolumn{9}{l}{\footnotesize Continued on next page}
-	\endfoot
-	\endlastfoot
-	1 & ARTR & growth & Control & lppd & -107.53 & -108.40 & 0.87 & *** \\ 
-	2 & ARTR & growth & Control & MSE & 0.57 & 0.56 & 0.01 &  \\ 
-	3 & ARTR & growth & Drought & lppd & -39.71 & -40.01 & 0.30 & *** \\ 
-	4 & ARTR & growth & Drought & MSE & 0.53 & 0.49 & 0.03 &  \\ 
-	5 & ARTR & growth & Irrigation & lppd & -38.85 & -40.75 & 1.91 & *** \\ 
-	6 & ARTR & growth & Irrigation & MSE & 0.50 & 0.57 & -0.07 & *** \\ 
-	7 & ARTR & recruitment & Control & lppd & -32.61 & -31.05 & -1.56 &  \\ 
-	8 & ARTR & recruitment & Control & MSE & 61.88 & 10.59 & 51.29 &  \\ 
-	9 & ARTR & recruitment & Drought & lppd & -28.93 & -26.52 & -2.41 &  \\ 
-	10 & ARTR & recruitment & Drought & MSE & 523.95 & 11.62 & 512.33 &  \\ 
-	11 & ARTR & recruitment & Irrigation & lppd & -17.51 & -20.24 & 2.73 & *** \\ 
-	12 & ARTR & recruitment & Irrigation & MSE & 4.72 & 5.34 & -0.62 & *** \\ 
-	13 & ARTR & survival & Control & lppd & -24.19 & -23.06 & -1.13 &  \\ 
-	14 & ARTR & survival & Control & MSE & 0.07 & 0.07 & 0.00 &  \\ 
-	15 & ARTR & survival & Drought & lppd & -5.48 & -5.34 & -0.13 &  \\ 
-	16 & ARTR & survival & Drought & MSE & 0.04 & 0.04 & -0.00 & *** \\ 
-	17 & ARTR & survival & Irrigation & lppd & -7.88 & -8.01 & 0.13 & *** \\ 
-	18 & ARTR & survival & Irrigation & MSE & 0.06 & 0.06 & -0.00 & *** \\ 
-	19 & HECO & growth & Control & lppd & -377.77 & -369.46 & -8.31 &  \\ 
-	20 & HECO & growth & Control & MSE & 1.11 & 1.09 & 0.02 &  \\ 
-	21 & HECO & growth & Drought & lppd & -8.74 & -10.97 & 2.23 & *** \\ 
-	22 & HECO & growth & Drought & MSE & 3.12 & 4.35 & -1.23 & *** \\ 
-	23 & HECO & growth & Irrigation & lppd & -89.03 & -73.93 & -15.10 &  \\ 
-	24 & HECO & growth & Irrigation & MSE & 2.02 & 1.47 & 0.55 &  \\ 
-	25 & HECO & recruitment & Control & lppd & -93.35 & -93.47 & 0.12 & *** \\ 
-	26 & HECO & recruitment & Control & MSE & 613.41 & 499.11 & 114.29 &  \\ 
-	27 & HECO & recruitment & Drought & lppd & -25.39 & -27.56 & 2.17 & *** \\ 
-	28 & HECO & recruitment & Drought & MSE & 1.94 & 2.33 & -0.38 & *** \\ 
-	29 & HECO & recruitment & Irrigation & lppd & -30.69 & -30.56 & -0.12 &  \\ 
-	30 & HECO & recruitment & Irrigation & MSE & 5.21 & 7.62 & -2.41 & *** \\ 
-	31 & HECO & survival & Control & lppd & -124.59 & -112.70 & -11.90 &  \\ 
-	32 & HECO & survival & Control & MSE & 0.12 & 0.11 & 0.01 &  \\ 
-	33 & HECO & survival & Drought & lppd & -17.89 & -20.82 & 2.93 & *** \\ 
-	34 & HECO & survival & Drought & MSE & 0.22 & 0.26 & -0.05 & *** \\ 
-	35 & HECO & survival & Irrigation & lppd & -16.51 & -13.54 & -2.97 &  \\ 
-	36 & HECO & survival & Irrigation & MSE & 0.09 & 0.07 & 0.02 &  \\ 
-	37 & POSE & growth & Control & lppd & -1117.20 & -1117.29 & 0.08 & *** \\ 
-	38 & POSE & growth & Control & MSE & 1.50 & 1.51 & -0.00 & *** \\ 
-	39 & POSE & growth & Drought & lppd & -254.17 & -257.32 & 3.16 & *** \\ 
-	40 & POSE & growth & Drought & MSE & 2.66 & 2.69 & -0.03 & *** \\ 
-	41 & POSE & growth & Irrigation & lppd & -452.34 & -456.95 & 4.61 & *** \\ 
-	42 & POSE & growth & Irrigation & MSE & 1.87 & 1.90 & -0.03 & *** \\ 
-	43 & POSE & recruitment & Control & lppd & -127.64 & -128.73 & 1.09 & *** \\ 
-	44 & POSE & recruitment & Control & MSE & 35.20 & 44.82 & -9.61 & *** \\ 
-	45 & POSE & recruitment & Drought & lppd & -60.34 & -63.15 & 2.80 & *** \\ 
-	46 & POSE & recruitment & Drought & MSE & 23.60 & 33.88 & -10.28 & *** \\ 
-	47 & POSE & recruitment & Irrigation & lppd & -72.06 & -65.58 & -6.48 &  \\ 
-	48 & POSE & recruitment & Irrigation & MSE & 85.04 & 27.63 & 57.41 &  \\ 
-	49 & POSE & survival & Control & lppd & -366.49 & -380.49 & 14.00 & *** \\ 
-	50 & POSE & survival & Control & MSE & 0.12 & 0.12 & -0.00 & *** \\ 
-	51 & POSE & survival & Drought & lppd & -175.29 & -187.88 & 12.59 & *** \\ 
-	52 & POSE & survival & Drought & MSE & 0.21 & 0.22 & -0.01 & *** \\ 
-	53 & POSE & survival & Irrigation & lppd & -156.28 & -150.23 & -6.05 &  \\ 
-	54 & POSE & survival & Irrigation & MSE & 0.13 & 0.13 & 0.00 &  \\ 
-	55 & PSSP & growth & Control & lppd & -627.07 & -625.63 & -1.44 &  \\ 
-	56 & PSSP & growth & Control & MSE & 1.38 & 1.36 & 0.02 &  \\ 
-	57 & PSSP & growth & Drought & lppd & -292.90 & -297.61 & 4.71 & *** \\ 
-	58 & PSSP & growth & Drought & MSE & 1.82 & 1.85 & -0.03 & *** \\ 
-	59 & PSSP & growth & Irrigation & lppd & -312.96 & -314.67 & 1.71 & *** \\ 
-	60 & PSSP & growth & Irrigation & MSE & 1.51 & 1.51 & -0.00 & *** \\ 
-	61 & PSSP & recruitment & Control & lppd & -123.81 & -125.23 & 1.42 & *** \\ 
-	62 & PSSP & recruitment & Control & MSE & 30.59 & 34.79 & -4.20 & *** \\ 
-	63 & PSSP & recruitment & Drought & lppd & -87.34 & -86.58 & -0.77 &  \\ 
-	64 & PSSP & recruitment & Drought & MSE & 222.16 & 54.58 & 167.58 &  \\ 
-	65 & PSSP & recruitment & Irrigation & lppd & -60.18 & -62.12 & 1.94 & *** \\ 
-	66 & PSSP & recruitment & Irrigation & MSE & 20.90 & 44.57 & -23.67 & *** \\ 
-	67 & PSSP & survival & Control & lppd & -148.69 & -140.15 & -8.54 &  \\ 
-	68 & PSSP & survival & Control & MSE & 0.10 & 0.09 & 0.01 &  \\ 
-	69 & PSSP & survival & Drought & lppd & -83.26 & -78.09 & -5.17 &  \\ 
-	70 & PSSP & survival & Drought & MSE & 0.11 & 0.11 & 0.01 &  \\ 
-	71 & PSSP & survival & Irrigation & lppd & -100.52 & -89.02 & -11.50 &  \\ 
-	72 & PSSP & survival & Irrigation & MSE & 0.13 & 0.11 & 0.01 &  \\ 
-	\hline
-	\caption{Comparison of model predictions from climate model and no climate model for each species and vital rate and treatment.
-Two prediction scores are reported, MSE and lppd.
-Lower MSE indicates improved predictions whereas higher lppd indicates improved predictions.
-Instances where the climate model outperformed the no climate model are marked with "***" in the last column.
-ARTR = \textit{A.
-tripartita}, HECO = \textit{H.
-comata}, POSE = \textit{P.
-secunda}, PSSP = \textit{P.
-spicata}.}
-	\label{table:treatmentPreds}
-\end{longtable}
-
-\clearpage
-\newpage
-\clearpage
-\newpage
-\section{Additional Figures}
-
-\begin{figure}[!htbp]
-	\centering
-	\includegraphics[width=1\textwidth]{pred_v_obs_treatment_ARTR_growth}
-	\caption{Comparison of treatment effects predicted and observed for \textit{A.
-tripartita} growth.
-Upper figure shows each parameter estimate separately while the lower figure shows the effect of treatment as a function of plant size.
-The observed treatment effects come from the treatment model fitted to all the data including the five years of the experiment.
-The predicted parameters come from the climate model fitted to all years of observational data but do not include the five years of the experiment (2011 to 2016).
-We assessed a parameter as significant when the 95\% Bayesian credible intervals did not overlap zero.}
-	\label{fig:parPredARTRGrowth}
-\end{figure}
-
-\begin{figure}[!htbp]
-	\centering
-	\includegraphics[width=1\textwidth]{pred_v_obs_treatment_HECO_growth}
-	\caption{Comparison of treatment effects predicted and observed for \textit{H.
-comata} growth.
-The observed treatment effects come from the treatment model fitted to all the data including the five years of the experiment.
-The predicted parameters come from the climate model fitted to all years of observational data but do not include the five years of the experiment (2011 to 2016).
-We assessed a parameter as significant when the 95\% Bayesian credible intervals did not overlap zero.}
-	\label{fig:parPredHECOGrowth}
-\end{figure}
-
-\begin{figure}[!htbp]
-	\centering
-	\includegraphics[width=1\textwidth]{pred_v_obs_treatment_POSE_growth}
-	\caption{Comparison of treatment effects predicted and observed for \textit{P.
-secunda} growth.
-Upper figure shows each parameter estimate separately while the lower figure shows the effect of treatment as a function of plant size.
-The observed treatment effects come from the treatment model fitted to all the data including the five years of the experiment.
-The predicted parameters come from the climate model fitted to all years of observational data but do not include the five years of the experiment (2011 to 2016).
-We assessed a parameter as significant when the 95\% Bayesian credible intervals did not overlap zero.}
-	\label{fig:parPredPOSEGrowth}
-\end{figure}
-
-\begin{figure}[!htbp]
-	\centering
-	\includegraphics[width=1\textwidth]{pred_v_obs_treatment_PSSP_growth}
-	\caption{Comparison of treatment effects predicted and observed for \textit{P.
-spicata} growth.
-The observed treatment effects come from the treatment model fitted to all the data including the five years of the experiment.
-The predicted parameters come from the climate model fitted to all years of observational data but do not include the five years of the experiment (2011 to 2016).
-We assessed a parameter as significant when the 95\% Bayesian credible intervals did not overlap zero.}
-	\label{fig:parPredPSSPGrowth}
-\end{figure}
-
-\begin{figure}[!htbp]
-	\centering
-	\includegraphics[width=1\textwidth]{pred_v_obs_treatment_ARTR_survival}
-	\caption{Comparison of treatment effects predicted and observed for \textit{A.
-survival} survival.
-The observed treatment effects come from the treatment model fitted to all the data including the five years of the experiment.
-The predicted parameters come from the climate model fitted to all years of observational data but do not include the five years of the experiment (2011 to 2016).
-We assessed a parameter as significant when the 95\% Bayesian credible intervals did not overlap zero.}
-	\label{fig:parPredARTRSurvival}
-\end{figure}
-
-\begin{figure}[!htbp]
-	\centering
-	\includegraphics[width=1\textwidth]{pred_v_obs_treatment_HECO_survival}
-	\caption{Comparison of treatment effects predicted and observed for \textit{H.
-comata} survival.
-The observed treatment effects come from the treatment model fitted to all the data including the five years of the experiment.
-The predicted parameters come from the climate model fitted to all years of observational data but do not include the five years of the experiment (2011 to 2016).
-We assessed a parameter as significant when the 95\% Bayesian credible intervals did not overlap zero.}
-	\label{fig:parPredHECOSurvival}
-\end{figure}
-
-\begin{figure}[!htbp]
-	\centering
-	\includegraphics[width=1\textwidth]{pred_v_obs_treatment_POSE_survival}
-	\caption{Comparison of treatment effects predicted and observed for \textit{P.
-secunda} survival.
-Upper figure shows each parameter estimate separately while the lower figure shows the effect of treatment as a function of plant size.
-The observed treatment effects come from the treatment model fitted to all the data including the five years of the experiment.
-The predicted parameters come from the climate model fitted to all years of observational data but do not include the five years of the experiment (2011 to 2016).
-We assessed a parameter as significant when the 95\% Bayesian credible intervals did not overlap zero.}
-	\label{fig:parPredPOSESurvival}
-\end{figure}
-
-\begin{figure}[!htbp]
-	\centering
-	\includegraphics[width=1\textwidth]{pred_v_obs_treatment_PSSP_survival}
-	\caption{Comparison of treatment effects predicted and observed for \textit{P.
-spicata} survival.
-The observed treatment effects come from the treatment model fitted to all the data including the five years of the experiment.
-The predicted parameters come from the climate model fitted to all years of observational data but do not include the five years of the experiment (2011 to 2016).
-We assessed a parameter as significant when the 95\% Bayesian credible intervals did not overlap zero.}
-	\label{fig:parPredPSSPSurvival}
-\end{figure}
-
-\end{document}

### 2016-12-22 precip/manuscript/pgr_predictions_by_treatment.tex
commiting revisions to the discussion and formatting revisions on the tables

% latex table generated in R 3.3.1 by xtable 1.8-2 package
% Thu Dec 22 04:07:33 2016
\begin{table}[H]
\centering
\caption{MSE of predicted log cover changes and correlations between log cover changes predicted and observed. Predictions for the cover changes in the experimental plots were generated either from the year effects or the climate models. Instances where the climate model made better predictions than the year effects model are indicated with the "***". ARTR = \textit{A. tripartita}, HECO = \textit{H. comata}, POSE = \textit{P. secunda}, PSSP = \textit{P. spicata}.} 
\label{table:corPGR}
\begin{tabular}{rlllrrrl}
  \hline
 & species & Treatment & stat & no climate model & climate model & diff & improved \\ 
  \hline
1 & ARTR & Control & cor & 0.36 & -0.02 & -0.39 &  \\ 
  2 & ARTR & Control & MSE & 0.25 & 0.26 & 0.00 &  \\ 
  3 & ARTR & Drought & cor & 0.60 & 0.43 & -0.17 &  \\ 
  4 & ARTR & Drought & MSE & 0.35 & 0.38 & 0.03 &  \\ 
  5 & ARTR & Irrigation & cor & 0.61 & 0.09 & -0.52 &  \\ 
  6 & ARTR & Irrigation & MSE & 0.32 & 0.29 & -0.04 & *** \\ 
  7 & HECO & Control & cor & 0.46 & 0.27 & -0.19 &  \\ 
  8 & HECO & Control & MSE & 0.29 & 0.32 & 0.04 &  \\ 
  9 & HECO & Drought & cor & -0.24 & 0.72 & 0.95 & *** \\ 
  10 & HECO & Drought & MSE & 2.07 & 1.50 & -0.58 & *** \\ 
  11 & HECO & Irrigation & cor & 0.53 & -0.18 & -0.71 &  \\ 
  12 & HECO & Irrigation & MSE & 0.52 & 0.86 & 0.34 &  \\ 
  13 & POSE & Control & cor & 0.43 & 0.48 & 0.05 & *** \\ 
  14 & POSE & Control & MSE & 0.38 & 0.40 & 0.01 &  \\ 
  15 & POSE & Drought & cor & 0.56 & 0.72 & 0.16 & *** \\ 
  16 & POSE & Drought & MSE & 0.45 & 0.46 & 0.00 &  \\ 
  17 & POSE & Irrigation & cor & 0.55 & 0.41 & -0.14 &  \\ 
  18 & POSE & Irrigation & MSE & 0.47 & 0.42 & -0.05 & *** \\ 
  19 & PSSP & Control & cor & 0.28 & 0.38 & 0.10 & *** \\ 
  20 & PSSP & Control & MSE & 0.34 & 0.33 & -0.01 & *** \\ 
  21 & PSSP & Drought & cor & 0.48 & 0.45 & -0.03 &  \\ 
  22 & PSSP & Drought & MSE & 0.51 & 0.51 & 0.00 &  \\ 
  23 & PSSP & Irrigation & cor & 0.43 & 0.30 & -0.12 &  \\ 
  24 & PSSP & Irrigation & MSE & 0.38 & 0.37 & -0.01 & *** \\ 
   \hline
\end{tabular}
\end{table}

### 2016-12-22 precip/manuscript/precip_ms.tex
commiting revisions to the discussion and formatting revisions on the tables

@@
 
-\section*{Summary}
+\section*{Abstract}
 
@@
 
-\end{enumerate}
-
 \section*{Introduction}
@@
 We augmented automatic monitoring of the climate in these plots with direct measurements of soil moisture with a handheld EC-5 soil moisture sensor at six points around all 16 plots on 6/6/2012, 4/29/2015, 5/7/2015, 6/9/2015 and 5/10/2016.
-We analyzed these spot measurements for significant treatment effects on soil moisture using a linear mixed effects model with the \textit{lmer} package in \textit{R} \citep{bates}, with plot, plot group, and date as random effects in the model.
+We analyzed these spot measurements for significant treatment effects on soil moisture using a linear mixed effects model with the \textit{lmer} package in \textit{R}, with plot, plot group, and date as random effects in the model (\citep{bates_fitting_2015}).
 
@@
 
-We fit all vital rate models using Hamiltonian-Markov Chain Monte Carlo (HMCMC) simulations in the programs STAN 10.1 and rStan (\citep{stan} and  \citep{rstan}).
+We fit all vital rate models using Hamiltonian-Markov Chain Monte Carlo (HMCMC) simulations in the programs STAN 10.1 and rStan \citep{}.
 The priors and model code are described more completely in appendix A.
@@
 We discarded the initial 1,000 samples.
-Convergence was observed graphically for all parameters, and confirmed by assessing the split $\widehat{R}$ statistic which at convergence is equal to one \citep{stan_manual}.
+Convergence was observed graphically for all parameters, and confirmed by assessing the split $\widehat{R}$ statistic which at convergence is equal to one \citep{}.
 
@@
 
-*Although a number of previous studies of the plants in this system have looked the effects of climate on demography and survival, this is the first study to demonstrate effects of climate experimentally.
+Although a number of previous studies of the plants in this system have looked the effects of climate on demography and survival, this is the first study to demonstrate effects of climate experimentally.
 The significant decline in \textit{H.
@@
 Each of these studies report \textit{H.
-comata} having the strongest positive population response to increased precipitation among all the species.
-The response of this species' vital rates to drought in the experiment also matches this result (fig \ref{fig:growthTreat}, \ref{fig:survivalTreat}).
+comata} having the strongest positive population response to increased precipitation among all the species; this is matched by our finding that this species has the strongest negative response to experimental drought (fig \ref{fig:growthTreat}, \ref{fig:survivalTreat}).
 On the other hand if we had only conducted an irrigation experiment our results may not have shown this consistency with previous work as \textit{H.
@@
 secunda} growth and survival in the drought plots in our experiment (figs \ref{fig:coverChange}, \ref{fig:growthTreat}, \ref{fig:survivalTreat}).
-As for \textit{H.
-comata} the magnitudes of the negative drought effects were greater than the magnitudes of the positive irrigation effects on these species.
-
-In contrast to the effects on the grasses the effects on \textit{A.
+The magnitudes of the negative drought effects were greater than the magnitudes of the positive irrigation effects on these grasses.
+
+In contrast to the effects on the grasses, the effects on \textit{A.
 tripartita} are more complicated.
@@
 tripartita} growth we see in the drought plots is the result of reduced grass competition these plots (fig \ref{fig:coverPred}; \citep{chu_direct_2016}).
-Our growth model should take into account any changes in grass abundance that could be driving a positive response from \textit{A.
+However, our growth model includes interspecific crowding and so should take into account any changes in grass abundance that could be driving a positive response from \textit{A.
 tripartita}.
 So this leaves us with the question of why this species would show a positive direct response to drought?
-Although there is some evidence that saturated soils can be detrimental to sagebrush growth \citep{} there was no evidence in our data that ambient moisture would have been so wet as to be limiting to sagebrush during this experiment (fig \ref{fig:dailyVWC}).
-Alternatively our drought treatments could be reducing snow cover in the winter in spring and allowing for earlier growth of sagebrush.
-In a high elevation sagebrush meadow earlier snowmelt was correlated with increased sagebrush performance \citep{harte_}.
+Although there is some evidence that saturated soils can be detrimental to big sagebrush (\textit{A.
+tridentata}), a closely related species \citep{sturges_response_1989}, there was no evidence in our data that soils were saturated with moisture in the ambient or irrigation treatments (fig \ref{fig:dailyVWC}).
+In a similar study to ours, \citep{germino_desert_2014} showed that big sagebrush increased in response to deep soil moisture but actually decreased in response to shallow soil moisture.
+Perhaps \textit{A.
+tripartita} responds similarly and benefits directly from the drier soils.
+Another possible explanation is that our drought treatments reduced snow cover in the winter and early spring, an effect that has been shown to benefit big sagebrush in other ecosystems \citep{perfors_enhanced_2003}.
 
@@
 It is possible that adding water to this system does little to increase grass growth and survival for a number of reasons.
-First, these species are presumably able to escape drought, by growing early in the year, or avoid drought stress through high water use efficiency \citep{bazzaz, franks}.
+First, these species are presumably able to escape drought, by growing early in the year, or avoid drought stress later in the year through high water use efficiency \citep{bazzaz_physiological_1979,franks_plasticity_2011}.
+If the species in our system are well adapted to the ambient water limitation, it makes some sense that they would show relatively small responses to increased moisture.
 Indeed our soil moisture data generally show a pulse of soil moisture in the spring when many grasses are actively growing (fig \ref{fig:dailyVWC}).
-During this window of active growth soil moisture in sagebrush steppe may be quite high and perhaps other resources become limiting \citep{caldwell} \citep{ino}.
-Moreover, our treatments had relatively smaller effects on soil moisture during this period \ref{table:soil_moisture_model}.
-If the species in our system are well adapted to the ambient water limitation, it makes some sense that they would show relatively small responses to increased moisture.
-
-A possibly related result was recently observed by \citep{gherardi_} in a warm desert grassland: the effect of increased soil moisture on grass productivity in that system showed a concave down pattern of response to increased soil moisture.
-Increasing soil moisture had weaker effects on grass productivity than decreasing soil moisture in equal proportion.
-This pattern could explain why we repeatedly saw large decreases in either the cover or performance of the three grasses in drought plots but did not see any statistically significant increase in performance in the irrigated plots.
-
-Reflecting on the possible mechanisms by which soil moisture could have affected these species sheds some light on our sucesses and failures of prediction.
+During this window of active growth soil moisture in sagebrush steppe may be quite high and perhaps other resources become limiting \citep{caldwell_rapid_1987}.
+Moreover, our treatments had relatively smaller effects on soil moisture during the spring (table ).
+
+A similar result was recently reported from another precipitation experiment in a warm desert grassland: the effect of increased soil moisture on grass productivity in that system showed a concave down pattern of response to increased soil moisture \citep{gherardi_enhanced_2015}.
+Increasing precipitation had a weaker effect on annual grass productivity than decreasing precipitation by an equal amount.
+A pattern like this is qualitatively similar to our finding significant decreases in either the cover or performance of the three grasses in drought plots, but not seeing any statistically significant increase in performance in the irrigated plots.
+
+Reflecting on the possible mechanisms by which soil moisture could have affected these species may also shed some light on our sucesses and failures at prediction.
 The drought effects we observed on the three grasses were often stronger than the effects we predicted, while the irrigation responses observed were often weaker than predicted (figs \ref{fig:parPredHECOGrowth}, \ref{fig:parPredHECOSurvival}, \ref{fig:parPredPOSEGrowth}, \ref{fig:parPredPOSESurvival}, \ref{fig:parPredPSSPGrowth}, \ref{fig:parPredPSSPSurvival}).
-If we had fit our growth and survival models with a non-linear function for soil moisture, perhaps informed by more mechanistic understanding of water limitation in this system, they may have made more accurate predictions of the drought and irrigation effects \citep{ehrlen_advancing_2016}.
+If we had fit our growth and survival models with a non-linear function for soil moisture, perhaps informed by more mechanistic understanding of water limitation in this system, we may have made more accurate predictions of the drought and irrigation effects \citep{ehrlen_advancing_2016}.
 
 Out of all the climate effects we predicted and observed, we only observed four cases where vital rate predictions and observations were both significantly different from zero (fig  \ref{fig:parPredictions}).
-In three of these cases, we successfully predicted the direction of the treatment effect.
-However, in one cases we predicted a treatment effect that was significantly
-
-Using the IBM, we also demonstrated that the climate model predictions of the vital rates could be used to generate population-level predictions that were also better than the no climate models in some cases: the climate model improved one step ahead predictions of cover changes of \textit{P.
+In three of these cases, we successfully predicted the direction of the treatment effects.
+However, for \textit{P.
+secunda} recruitment we predicted a response to irrigation that was significantly positive and yet observed a significantly negative response (figs \ref{fig:parPredictions}).
+This is arguably our greatest error in prediction.
+But there may be some mitigating factors that can excuse our error.
+First we note that we observed decreases in recruitment in both the drought and irrigation plots for \textit{P.
+secunda} and also for \textit{P.
+spicata} (fig \ref{fig:recruitmentTreat}).
+So while we observed a significant decrease in \textit{P.
+secunda} recruitment in the irrigated plots this may have been due to underlying differences in the experimental plots from the historical control plots.
+Because recruitment is modeled at the quadrat level we also had fewer observations and predictions to work with for these recruitment models than we had for the other vital rates.
+
+Using the IBM, we demonstrated that the climate model predictions of the vital rates could be used to generate population-level predictions that were also better than the no climate models in some cases: the climate model improved one step ahead predictions of cover changes of \textit{P.
 spicata} and \textit{P.
@@
 
-*Mixed success at predicting species responses to high and low moisture availability has interesting implications.
-On the one hand it is re-assuring that observational data is sometimes useful.
-On the other hand, for predictions to be truly useful we would also like more information to help us sort out why our predictions for some species were good and for others no better than a no climate model.
-In other words, we have little ability to predict when our predictions are likely to be accurate.
-Among plants detailed physiological ecology may give us a guide to which factors and climate conditions are likely to affect which species.
+Mixed success at predicting species responses to high and low moisture availability has interesting implications.
+On the one hand it is reassuring that observational data captures important climate effects.
+On the other hand, will knowing that some species' responses are unpredictable, but not knowing which ones, discourage ecologists from making any predictions at all?
+Among plants, perhaps improved ecophysiological models or trait data can be used as a screen to generate hypothesis about which species are more likely to be affected by climate and in which ways.
 In our system, for instance, it may make sense that the grasses showed a stronger response to the drought treatment than the a woody shrub.
@@
 \textit{A.
-tripartita} on the other hand grows throughout the summer and generally has deeper roots than the grasses.
-These traits may help it tolerate the water stress induced by drought.
-
-*Will our predictions be useful for the predicting the effects of climate change in the future?
-Because these species compete, one could argue that our predictions for any one species in this community will only be as good as the predictions we make for their competitors \citep{tylianakis}.
+tripartita} on the other hand grows throughout the summer and has deeper roots than the grasses, these traits may buffer it from some of the annual variability in soil moisture experienced by the grasses.
+
+Will our predictions be useful for the predicting the effects of climate change in the future?
+Because these species compete, one could argue that our predictions for any one species in this community will only be as good as the predictions we make for their competitors, pathogens and predators \citep{tylianakis_global_2008}.
 For instance, while we observed weak treatment effects on \textit{A.
-tripartita} in our experiment (fig \ref{fig:coverChange}) could drought ultimately lead to a decline in the cover of its competitors which will then allow it to increase in abundance in the future?
+tripartita} in our experiment (fig \ref{fig:coverChange}) could drought ultimately lead to a decline in the cover of its competitors which will cause it to increase in abundance in the future?
 While this is possible for some species, research at our site and several other grasslands suggest that species are more limited by intra-specific competition than inter-specific competition \cite{chu_direct_2016,adler_weak_2016}.
-In theory this fact should mean that direct effects rather than indirect effects dominate species responses to climate change (\citep{levine_competitors_2010,kleinhesselink_indirect_2015,adler_forecasting_2012}).
-
-*Our results give us more confidence that historical observational data can be used to detect and predict the demographic effects of climate change.
+In theory, this should mean that we can predict species short term responses to climate change with single species models (\citep{levine_competitors_2010,kleinhesselink_indirect_2015,adler_forecasting_2012}).
+
+Our results give us more confidence that historical observational data can be used to detect and predict the demographic effects of climate change among sagebrush steppe plants.
 This should encourage more researchers to try and use observational data to predict the future in both experimental and natural settings \citep{houlahan_priority_2016}.
-Making quantitative predictions, however humble they are, and testing them is arguably a key to scientific progress in all fields.
 Nevertheless, our success at predicting the short-term response of two species to a small-scale climate manipulation is not likely to be very reassuring to applied ecologists and resource managers wishing to make accurate predictions about the effects of climate change in large complex systems.
-Clearly more work is needed to distinguish which predictions are worth having confidence in and which predictions we should have less confidence in.
+Clearly more work is needed to learn how best to meet the challenge of prediction in ecological systems.
 Towards that goal, perhaps the best way forward is to conduct more tests like this one.
@@
 
-\begin{table}[h]
-	\caption{Treatment effects on spring soil moisture}
-	\centering
-	%\begin{center}
-	\begin{tabular}{l c c }
-		\hline
-		& Model 1 \\
-		\hline
-		(Intercept)                  & $8.81^{*}$    	& $[5.78;\ 11.83]$  \\
-		TreatmentDrought             & $-3.97^{*}$      & $[-4.84;\ -3.09]$ \\
-		TreatmentIrrigation          & $3.26^{*}$       & $[2.39;\ 4.14]$   \\
-		\hline
-		AIC                          & 3191.87           \\
-		BIC                          & 3222.92           \\
-		Log Likelihood               & -1588.93          \\
-		Num. obs.                    & 624               \\
-		Num. groups: plot            & 24                \\
-		Num. groups: PrecipGroup     & 8                 \\
-		Num. groups: date            & 5                 \\
-		Var: plot (Intercept)        & 0.45              \\
-		Var: PrecipGroup (Intercept) & 0.23              \\
-		Var: date (Intercept)        & 11.24             \\
-		Var: Residual                & 8.90              \\
-		\hline
-		\multicolumn{2}{l}{\scriptsize{$^*$ 0 outside the confidence interval}}
-	\end{tabular}
-	\label{table:spotVWC}
-	%\end{center}
-\end{table}
-
-\begin{table}
-	\caption{Model of treatment effects on soil moisture}
-	\begin{center}
-		\begin{tabular}{l c c }
-			\hline
-			& Model 1 \\
-			\hline
-			(Intercept)                       & $-0.57^{*}$       & $[-0.89;\ -0.26]$ \\
-			TreatmentIrrigation               & $1.23^{*}$        & $[1.18;\ 1.29]$   \\
-			rainfallrainy                     & $-0.05$           & $[-0.12;\ 0.01]$  \\
-			seasonspring                      & $0.27^{*}$        & $[0.23;\ 0.32]$   \\
-			seasonsummer                      & $0.15^{*}$        & $[0.10;\ 0.19]$   \\
-			seasonwinter                      & $0.25^{*}$        & $[0.21;\ 0.29]$   \\
-			TreatmentIrrigation:rainfallrainy & $0.18^{*}$        & $[0.13;\ 0.24]$   \\
-			TreatmentIrrigation:seasonspring  & $-0.23^{*}$       & $[-0.29;\ -0.16]$ \\
-			TreatmentIrrigation:seasonsummer  & $-0.26^{*}$       & $[-0.33;\ -0.20]$ \\
-			TreatmentIrrigation:seasonwinter  & $-0.33^{*}$       & $[-0.40;\ -0.27]$ \\
-			rainfallrainy:seasonspring        & $-0.23^{*}$       & $[-0.31;\ -0.16]$ \\
-			rainfallrainy:seasonsummer        & $-0.07$           & $[-0.14;\ 0.01]$  \\
-			rainfallrainy:seasonwinter        & $-0.07$           & $[-0.20;\ 0.07]$  \\
-			\hline
-			AIC                               & 14581.58          \\
-			BIC                               & 14695.49          \\
-			Log Likelihood                    & -7274.79          \\
-			Num. obs.                         & 9133              \\
-			Num. groups: date                 & 1596              \\
-			Num. groups: PrecipGroup          & 8                 \\
-			Var: date (Intercept)      		  & 0.00              \\
-			Var: PrecipGroup (Intercept)      & 0.19              \\
-			Var: Residual                     & 2.50              \\
-			\hline
-			\multicolumn{2}{l}{\scriptsize{$^*$ 0 outside the confidence interval}}
-		\end{tabular}
-		\label{table:soil_moisture_model}
-	\end{center}
-\end{table}
-
-% latex table generated in R 3.3.1 by xtable 1.8-2 package
-% Mon Dec 19 18:10:15 2016
-\begin{table}[ht]
-	\centering
-	\begin{tabular}{rrrrr}
-		\hline
-		& Estimate & Std. Error & t value & Pr($>$$|$t$|$) \\ 
-		\hline
-		(Intercept) & -0.2835 & 0.4627 & -0.61 & 0.5487 \\ 
-		TreatmentDrought & 0.9378 & 0.7743 & 1.21 & 0.2434 \\ 
-		TreatmentIrrigation & 0.8882 & 0.7743 & 1.15 & 0.2682 \\ 
-		\hline
-	\end{tabular}
-	\caption{Treatment effects on log cover change for \textit{A. tripartita} from 2011 to 2016. Intercept gives control effects.} 
-	\label{table:changeARTR}
-\end{table}
-
-% latex table generated in R 3.3.1 by xtable 1.8-2 package
-% Mon Dec 19 18:10:16 2016
-\begin{table}[ht]
-	\centering
-	\begin{tabular}{rrrrr}
-		\hline
-		& Estimate & Std. Error & t value & Pr($>$$|$t$|$) \\ 
-		\hline
-		(Intercept) & 0.3982 & 0.2587 & 1.54 & 0.1548 \\ 
-		TreatmentDrought & -2.9951 & 0.5784 & -5.18 & 0.0004 \\ 
-		TreatmentIrrigation & -0.1219 & 0.4953 & -0.25 & 0.8105 \\ 
-		\hline
-	\end{tabular}
-	\caption{Treatment effects on log cover change for \textit{H. comata} from 2011 to 2016. Intercept gives control effects.} 
-	\label{table:changeHECO}
-\end{table}
-
-% latex table generated in R 3.3.1 by xtable 1.8-2 package
-% Mon Dec 19 18:10:16 2016
-\begin{table}[ht]
-	\centering
-	\begin{tabular}{rrrrr}
-		\hline
-		& Estimate & Std. Error & t value & Pr($>$$|$t$|$) \\ 
-		\hline
-		(Intercept) & -0.7247 & 0.4613 & -1.57 & 0.1298 \\ 
-		TreatmentDrought & 0.0273 & 0.8208 & 0.03 & 0.9737 \\ 
-		TreatmentIrrigation & 1.1459 & 0.7797 & 1.47 & 0.1552 \\ 
-		\hline
-	\end{tabular}
-	\caption{Treatment effects on log cover change for \textit{P. secunda} from 2011 to 2016. Intercept gives control effects.} 
-	\label{table:changePOSE}
-\end{table}
-
-% latex table generated in R 3.3.1 by xtable 1.8-2 package
-% Mon Dec 19 18:10:16 2016
-\begin{table}[ht]
-	\centering
-	\begin{tabular}{rrrrr}
-		\hline
-		& Estimate & Std. Error & t value & Pr($>$$|$t$|$) \\ 
-		\hline
-		(Intercept) & 0.0188 & 0.2124 & 0.09 & 0.9303 \\ 
-		TreatmentDrought & -0.8851 & 0.3780 & -2.34 & 0.0287 \\ 
-		TreatmentIrrigation & 0.1453 & 0.3780 & 0.38 & 0.7044 \\ 
-		\hline
-	\end{tabular}
-	\caption{Treatment effects on log cover change for \textit{P. spicata} from 2011 to 2016. Intercept gives control effects.} 
-	\label{table:changePSSP}
-\end{table}
-
-% latex table generated in R 3.3.1 by xtable 1.8-2 package
-% Tue Dec 20 12:39:32 2016
-\begin{table}[ht]
-	\centering
-	\begin{tabular}{lllrrrr}
-		\hline
-		vital\_rate & species & climate variable & Int. cor. & p val. & Size cor. & Size p. val. \\ 
-		\hline
-		growth & ARTR & su.0 & -0.49 & 0.02 & 0.26 & 0.26 \\ 
-		growth & ARTR & f.0 & -0.28 & 0.23 & 0.40 & 0.08 \\ 
-		growth & ARTR & sp.1 & 0.17 & 0.45 & -0.33 & 0.14 \\ 
-		growth & HECO & su.1 & 0.69 & 0.00 &  &  \\ 
-		growth & HECO & su.0 & 0.50 & 0.02 &  &  \\ 
-		growth & HECO & f.lag & 0.37 & 0.10 &  &  \\ 
-		growth & POSE & f.lag & 0.31 & 0.17 & -0.11 & 0.64 \\ 
-		growth & POSE & su.lag & 0.29 & 0.20 & -0.20 & 0.38 \\ 
-		growth & POSE & sp.1 & 0.26 & 0.25 & -0.20 & 0.38 \\ 
-		growth & PSSP & f.lag & 0.34 & 0.13 &  &  \\ 
-		growth & PSSP & su.lag & 0.25 & 0.27 &  &  \\ 
-		growth & PSSP & f.0 & -0.22 & 0.34 &  &  \\ 
-		recruitment & ARTR & su.lag & -0.32 & 0.16 &  &  \\ 
-		recruitment & ARTR & su.0 & -0.26 & 0.25 &  &  \\ 
-		recruitment & ARTR & sp.1 & 0.22 & 0.34 &  &  \\ 
-		recruitment & HECO & su.lag & -0.31 & 0.18 &  &  \\ 
-		recruitment & HECO & su.0 & -0.30 & 0.18 &  &  \\ 
-		recruitment & HECO & f.lag & 0.19 & 0.40 &  &  \\ 
-		recruitment & POSE & sp.1 & 0.49 & 0.02 &  &  \\ 
-		recruitment & POSE & f.lag & 0.34 & 0.13 &  &  \\ 
-		recruitment & POSE & f.1 & 0.32 & 0.16 &  &  \\ 
-		recruitment & PSSP & su.lag & -0.52 & 0.02 &  &  \\ 
-		recruitment & PSSP & su.0 & -0.48 & 0.03 &  &  \\ 
-		recruitment & PSSP & sp.0 & 0.30 & 0.19 &  &  \\ 
-		survival & ARTR & su.0 & -0.60 & 0.00 &  &  \\ 
-		survival & ARTR & sp.0 & -0.41 & 0.06 &  &  \\ 
-		survival & ARTR & su.1 & -0.40 & 0.07 &  &  \\ 
-		survival & HECO & sp.0 & 0.44 & 0.04 &  &  \\ 
-		survival & HECO & sp.lag & 0.43 & 0.05 &  &  \\ 
-		survival & HECO & f.1 & 0.33 & 0.14 &  &  \\ 
-		survival & POSE & sp.0 & 0.44 & 0.04 & 0.22 & 0.34 \\ 
-		survival & POSE & sp.1 & 0.27 & 0.23 & -0.46 & 0.04 \\ 
-		survival & POSE & f.lag & -0.00 & 0.99 & 0.30 & 0.19 \\ 
-		survival & PSSP & sp.0 & 0.36 & 0.11 &  &  \\ 
-		survival & PSSP & sp.lag & 0.34 & 0.13 &  &  \\ 
-		survival & PSSP & su.1 & 0.26 & 0.26 &  &  \\ 
-		\hline
-	\end{tabular}
-	\caption{Selected climate variables for each vital rate model for each species.
-Correlations and p-values between the choosen variables and the intercept of the no climate model are shown.
-For ARTR growth and POSE growth and survival, the correlations between the year effects on size and the soil moisture variables are also given.
-"f" = fall, "su" = summer, "sp" = spring.
-ARTR = \textit{A.
-tripartita}, HECO = \textit{H.
-comata}, POSE = \textit{P.
-secunda}, PSSP = \textit{P.
-spicata}.}
-	\label{table:strongCor}
-\end{table}
-
-% latex table generated in R 3.3.1 by xtable 1.8-2 package
-% Tue Dec 20 12:38:30 2016
-\begin{table}[ht]
-	\centering
-	\begin{tabular}{lllrrrl}
-		\hline
-		species & vital\_rate & score & climate model & no climate model & diff & improved \\ 
-		\hline
-		ARTR & growth & lppd & -186.08 & -189.17 & 3.09 & *** \\ 
-		ARTR & growth & MSE & 0.55 & 0.55 & -0.00 & *** \\ 
-		ARTR & recruitment & lppd & -79.05 & -77.81 & -1.24 &  \\ 
-		ARTR & recruitment & MSE & 169.86 & 9.47 & 160.39 &  \\ 
-		ARTR & survival & lppd & -37.55 & -36.42 & -1.13 &  \\ 
-		ARTR & survival & MSE & 0.06 & 0.06 & 0.00 &  \\ 
-		HECO & growth & lppd & -475.54 & -454.36 & -21.18 &  \\ 
-		HECO & growth & MSE & 1.26 & 1.18 & 0.09 &  \\ 
-		HECO & recruitment & lppd & -149.43 & -151.60 & 2.17 & *** \\ 
-		HECO & recruitment & MSE & 288.16 & 235.57 & 52.59 &  \\ 
-		HECO & survival & lppd & -158.99 & -147.06 & -11.93 &  \\ 
-		HECO & survival & MSE & 0.12 & 0.11 & 0.01 &  \\ 
-		POSE & growth & lppd & -1823.71 & -1831.56 & 7.85 & *** \\ 
-		POSE & growth & MSE & 1.72 & 1.73 & -0.01 & *** \\ 
-		POSE & recruitment & lppd & -260.05 & -257.46 & -2.59 &  \\ 
-		POSE & recruitment & MSE & 45.40 & 37.32 & 8.08 &  \\ 
-		POSE & survival & lppd & -698.06 & -718.60 & 20.54 & *** \\ 
-		POSE & survival & MSE & 0.14 & 0.14 & -0.00 & *** \\ 
-		PSSP & growth & lppd & -1232.93 & -1237.92 & 4.99 & *** \\ 
-		PSSP & growth & MSE & 1.51 & 1.51 & 0.00 &  \\ 
-		PSSP & recruitment & lppd & -271.34 & -273.94 & 2.60 & *** \\ 
-		PSSP & recruitment & MSE & 79.09 & 42.68 & 36.41 &  \\ 
-		PSSP & survival & lppd & -332.47 & -307.26 & -25.21 &  \\ 
-		PSSP & survival & MSE & 0.11 & 0.10 & 0.01 &  \\ 
-		\hline
-	\end{tabular}
-	\caption{Comparison of model predictions from climate model and no climate model for each species and vital rate.
-Two prediction scores are reported, MSE and lppd.
-Lower MSE indicates improved predictions whereas higher lppd indicates improved predictions.
-Instances where the climate model outperformed the random no climate model are marked with "***" in the last column.
-ARTR = \textit{A.
-tripartita}, HECO = \textit{H.
-comata}, POSE = \textit{P.
-secunda}, PSSP = \textit{P.
-spicata}.}
-	\label{table:overallPreds}
-\end{table}
-
-% latex table generated in R 3.3.1 by xtable 1.8-2 package
-% Tue Dec 20 19:24:15 2016
-\begin{table}[ht]
-	\centering
-	\begin{tabular}{rllrrrl}
-		\hline
-		& species & stat & no climate model & climate model & diff & improved \\ 
-		\hline
-		1 & ARTR & cor & 0.48 & 0.19 & -0.29 &  \\ 
-		2 & ARTR & MSE & 0.30 & 0.30 & 0.00 &  \\ 
-		3 & HECO & cor & 0.29 & 0.22 & -0.07 &  \\ 
-		4 & HECO & MSE & 0.49 & 0.57 & 0.07 &  \\ 
-		5 & POSE & cor & 0.45 & 0.53 & 0.07 & *** \\ 
-		6 & POSE & MSE & 0.42 & 0.41 & -0.01 & *** \\ 
-		7 & PSSP & cor & 0.36 & 0.38 & 0.03 & *** \\ 
-		8 & PSSP & MSE & 0.39 & 0.39 & -0.01 & *** \\ 
-		\hline
-	\end{tabular}
-	\caption{MSE of predicted log cover changes and correlations between log cover changes predicted and observed.
-Predictions for the cover changes in the experimental plots were generated either from the no climate or the climate models.
-Instances where the climate model made better predictions than the no climate model are indicated with the "***".
-ARTR = \textit{A.
-tripartita}, HECO = \textit{H.
-comata}, POSE = \textit{P.
-secunda}, PSSP = \textit{P.
-spicata}.}
-	\label{table:corPGR}
-\end{table}
+\input{spot_measurements}
+\input{soil_moisture_model}
+\input{ARTR_cover_change}
+\input{HECO_cover_change}
+\input{POSE_cover_change}
+\input{PSSP_cover_change}
+\input{strong_correlations}
+\input{overall_prediction_score}
+\input{pgr_predictions}
 
@@
 Changes in cover predicted by the climate model are shown on the left and those predicted by the no climate model are shown on the right.
-Correlations coefficients between predictions and observations and MSE are shown for each treatment and model.}
+Correlations coefficients between predictions and observations and MSE are shown for each treatment and model.
+Gray line shows 1:1 line.}
 	\label{fig:pgrART}
@@
 Correlations coefficients between predictions and observations and MSE are shown for each treatment and model.
-}
+Gray line shows 1:1 line.}
 	\label{fig:pgrHECO}
@@
 Changes in cover predicted by the climate model are shown on the left and those predicted by the no climate model are shown on the right.
-Correlations coefficients between predictions and observations and MSE are shown for each treatment and model.}
+Correlations coefficients between predictions and observations and MSE are shown for each treatment and model.Gray line shows 1:1 line.}
 	\label{fig:pgrPOSE}
@@
 Changes in cover predicted by the climate model are shown on the left and those predicted by the no climate model are shown on the right.
-Correlations coefficients between predictions and observations and MSE are shown for each treatment and model.}
+Correlations coefficients between predictions and observations and MSE are shown for each treatment and model.Gray line shows 1:1 line.}
 	\label{fig:pgrPSSP}
@@
 \newpage
-\section{Additional Tables} 
-% latex table generated in R 3.3.1 by xtable 1.8-2 package
-% Tue Dec 20 13:32:53 2016
-\begin{longtable}{rllllrrrl}
-	\hline
-	& species & vital\_rate & Treatment & score & climate model & no climate model & diff & improved \\ 
-	\hline
-	\endhead
-	\hline
-	\multicolumn{9}{l}{\footnotesize Continued on next page}
-	\endfoot
-	\endlastfoot
-	1 & ARTR & growth & Control & lppd & -107.53 & -108.40 & 0.87 & *** \\ 
-	2 & ARTR & growth & Control & MSE & 0.57 & 0.56 & 0.01 &  \\ 
-	3 & ARTR & growth & Drought & lppd & -39.71 & -40.01 & 0.30 & *** \\ 
-	4 & ARTR & growth & Drought & MSE & 0.53 & 0.49 & 0.03 &  \\ 
-	5 & ARTR & growth & Irrigation & lppd & -38.85 & -40.75 & 1.91 & *** \\ 
-	6 & ARTR & growth & Irrigation & MSE & 0.50 & 0.57 & -0.07 & *** \\ 
-	7 & ARTR & recruitment & Control & lppd & -32.61 & -31.05 & -1.56 &  \\ 
-	8 & ARTR & recruitment & Control & MSE & 61.88 & 10.59 & 51.29 &  \\ 
-	9 & ARTR & recruitment & Drought & lppd & -28.93 & -26.52 & -2.41 &  \\ 
-	10 & ARTR & recruitment & Drought & MSE & 523.95 & 11.62 & 512.33 &  \\ 
-	11 & ARTR & recruitment & Irrigation & lppd & -17.51 & -20.24 & 2.73 & *** \\ 
-	12 & ARTR & recruitment & Irrigation & MSE & 4.72 & 5.34 & -0.62 & *** \\ 
-	13 & ARTR & survival & Control & lppd & -24.19 & -23.06 & -1.13 &  \\ 
-	14 & ARTR & survival & Control & MSE & 0.07 & 0.07 & 0.00 &  \\ 
-	15 & ARTR & survival & Drought & lppd & -5.48 & -5.34 & -0.13 &  \\ 
-	16 & ARTR & survival & Drought & MSE & 0.04 & 0.04 & -0.00 & *** \\ 
-	17 & ARTR & survival & Irrigation & lppd & -7.88 & -8.01 & 0.13 & *** \\ 
-	18 & ARTR & survival & Irrigation & MSE & 0.06 & 0.06 & -0.00 & *** \\ 
-	19 & HECO & growth & Control & lppd & -377.77 & -369.46 & -8.31 &  \\ 
-	20 & HECO & growth & Control & MSE & 1.11 & 1.09 & 0.02 &  \\ 
-	21 & HECO & growth & Drought & lppd & -8.74 & -10.97 & 2.23 & *** \\ 
-	22 & HECO & growth & Drought & MSE & 3.12 & 4.35 & -1.23 & *** \\ 
-	23 & HECO & growth & Irrigation & lppd & -89.03 & -73.93 & -15.10 &  \\ 
-	24 & HECO & growth & Irrigation & MSE & 2.02 & 1.47 & 0.55 &  \\ 
-	25 & HECO & recruitment & Control & lppd & -93.35 & -93.47 & 0.12 & *** \\ 
-	26 & HECO & recruitment & Control & MSE & 613.41 & 499.11 & 114.29 &  \\ 
-	27 & HECO & recruitment & Drought & lppd & -25.39 & -27.56 & 2.17 & *** \\ 
-	28 & HECO & recruitment & Drought & MSE & 1.94 & 2.33 & -0.38 & *** \\ 
-	29 & HECO & recruitment & Irrigation & lppd & -30.69 & -30.56 & -0.12 &  \\ 
-	30 & HECO & recruitment & Irrigation & MSE & 5.21 & 7.62 & -2.41 & *** \\ 
-	31 & HECO & survival & Control & lppd & -124.59 & -112.70 & -11.90 &  \\ 
-	32 & HECO & survival & Control & MSE & 0.12 & 0.11 & 0.01 &  \\ 
-	33 & HECO & survival & Drought & lppd & -17.89 & -20.82 & 2.93 & *** \\ 
-	34 & HECO & survival & Drought & MSE & 0.22 & 0.26 & -0.05 & *** \\ 
-	35 & HECO & survival & Irrigation & lppd & -16.51 & -13.54 & -2.97 &  \\ 
-	36 & HECO & survival & Irrigation & MSE & 0.09 & 0.07 & 0.02 &  \\ 
-	37 & POSE & growth & Control & lppd & -1117.20 & -1117.29 & 0.08 & *** \\ 
-	38 & POSE & growth & Control & MSE & 1.50 & 1.51 & -0.00 & *** \\ 
-	39 & POSE & growth & Drought & lppd & -254.17 & -257.32 & 3.16 & *** \\ 
-	40 & POSE & growth & Drought & MSE & 2.66 & 2.69 & -0.03 & *** \\ 
-	41 & POSE & growth & Irrigation & lppd & -452.34 & -456.95 & 4.61 & *** \\ 
-	42 & POSE & growth & Irrigation & MSE & 1.87 & 1.90 & -0.03 & *** \\ 
-	43 & POSE & recruitment & Control & lppd & -127.64 & -128.73 & 1.09 & *** \\ 
-	44 & POSE & recruitment & Control & MSE & 35.20 & 44.82 & -9.61 & *** \\ 
-	45 & POSE & recruitment & Drought & lppd & -60.34 & -63.15 & 2.80 & *** \\ 
-	46 & POSE & recruitment & Drought & MSE & 23.60 & 33.88 & -10.28 & *** \\ 
-	47 & POSE & recruitment & Irrigation & lppd & -72.06 & -65.58 & -6.48 &  \\ 
-	48 & POSE & recruitment & Irrigation & MSE & 85.04 & 27.63 & 57.41 &  \\ 
-	49 & POSE & survival & Control & lppd & -366.49 & -380.49 & 14.00 & *** \\ 
-	50 & POSE & survival & Control & MSE & 0.12 & 0.12 & -0.00 & *** \\ 
-	51 & POSE & survival & Drought & lppd & -175.29 & -187.88 & 12.59 & *** \\ 
-	52 & POSE & survival & Drought & MSE & 0.21 & 0.22 & -0.01 & *** \\ 
-	53 & POSE & survival & Irrigation & lppd & -156.28 & -150.23 & -6.05 &  \\ 
-	54 & POSE & survival & Irrigation & MSE & 0.13 & 0.13 & 0.00 &  \\ 
-	55 & PSSP & growth & Control & lppd & -627.07 & -625.63 & -1.44 &  \\ 
-	56 & PSSP & growth & Control & MSE & 1.38 & 1.36 & 0.02 &  \\ 
-	57 & PSSP & growth & Drought & lppd & -292.90 & -297.61 & 4.71 & *** \\ 
-	58 & PSSP & growth & Drought & MSE & 1.82 & 1.85 & -0.03 & *** \\ 
-	59 & PSSP & growth & Irrigation & lppd & -312.96 & -314.67 & 1.71 & *** \\ 
-	60 & PSSP & growth & Irrigation & MSE & 1.51 & 1.51 & -0.00 & *** \\ 
-	61 & PSSP & recruitment & Control & lppd & -123.81 & -125.23 & 1.42 & *** \\ 
-	62 & PSSP & recruitment & Control & MSE & 30.59 & 34.79 & -4.20 & *** \\ 
-	63 & PSSP & recruitment & Drought & lppd & -87.34 & -86.58 & -0.77 &  \\ 
-	64 & PSSP & recruitment & Drought & MSE & 222.16 & 54.58 & 167.58 &  \\ 
-	65 & PSSP & recruitment & Irrigation & lppd & -60.18 & -62.12 & 1.94 & *** \\ 
-	66 & PSSP & recruitment & Irrigation & MSE & 20.90 & 44.57 & -23.67 & *** \\ 
-	67 & PSSP & survival & Control & lppd & -148.69 & -140.15 & -8.54 &  \\ 
-	68 & PSSP & survival & Control & MSE & 0.10 & 0.09 & 0.01 &  \\ 
-	69 & PSSP & survival & Drought & lppd & -83.26 & -78.09 & -5.17 &  \\ 
-	70 & PSSP & survival & Drought & MSE & 0.11 & 0.11 & 0.01 &  \\ 
-	71 & PSSP & survival & Irrigation & lppd & -100.52 & -89.02 & -11.50 &  \\ 
-	72 & PSSP & survival & Irrigation & MSE & 0.13 & 0.11 & 0.01 &  \\ 
-	\hline
-	\caption{Comparison of model predictions from climate model and no climate model for each species and vital rate and treatment.
-Two prediction scores are reported, MSE and lppd.
-Lower MSE indicates improved predictions whereas higher lppd indicates improved predictions.
-Instances where the climate model outperformed the no climate model are marked with "***" in the last column.
-ARTR = \textit{A.
-tripartita}, HECO = \textit{H.
-comata}, POSE = \textit{P.
-secunda}, PSSP = \textit{P.
-spicata}.}
-	\label{table:treatmentPreds}
-\end{longtable}
+
+\input{treatment_prediction_score}
 

### 2016-12-22 precip/analysis/figure_scripts/plot_predicted_vs_observed_cover_changes.R

@@ -143,7 +143,7 @@ for(i in 1:nrow(grd)){
   
 }
  
- 
+tail(out)
 
 out <- do.call(rbind, out)
 label_df <- do.call(rbind, labels )
@@ -161,6 +161,7 @@ for( i in 1:length(species_list)) {
     geom_point() +
     geom_smooth(method = 'lm', se = F, alpha = 0.2, linetype = 2, color = 1, size = 1) +
     geom_text( data = subset(label_df, species == spp), aes( x = pos.x, y = pos.y , color = NULL , label = unique(label)), hjust = 1, vjust = -0.5 , show.legend = F) +
+    geom_abline(aes(intercept = 0, slope = 1), color = 'gray') + 
     facet_grid(  Treatment ~ model  )+
     xlab( 'Annual population growth rate predicted') +
     ylab( 'Annual population growth rate observed') +
@@ -189,4 +190,35 @@ fit_table <- fit_table %>% dplyr::select(species, stat, `year effects model` , `
 
 corxt <- xtable(fit_table, caption = 'MSE of predicted log cover changes and correlations between log cover changes predicted and observed. Predictions for the cover changes in the experimental plots were generated either from the year effects or the climate models. Instances where the climate model made better predictions than the year effects model are indicated with the "***". ARTR = \\textit{A. tripartita}, HECO = \\textit{H. comata}, POSE = \\textit{P. secunda}, PSSP = \\textit{P. spicata}.',
        label = 'table:corPGR')
-print(corxt, 'output/results_tables/pgr_predictions.text', type = 'latex')
+
+print(corxt, 'manuscript/pgr_predictions.tex', type = 'latex', caption.placement ="top", table.placement = 'H')
+
+# ----------- all treatments and species ---------------
+out$model <- str_replace(out$model, '_', ' ')
+
+out$model <- factor(out$model, labels = c('climate model', 'no climate model'))
+
+out <- out[complete.cases(out),]
+
+
+levels(out$model)
+
+scores_by_treatment <- 
+  out %>% 
+  group_by( Treatment, species, model)  %>% 
+  summarise( cor = cor( predicted , observed ), MSE = mean((predicted-observed)^2)) %>% 
+  gather( stat, val, cor:MSE) %>% 
+  spread(model, val) %>% 
+  ungroup() %>% 
+  mutate(diff = `climate model` - `no climate model`) %>% 
+  mutate( improved = ifelse( stat == 'cor' & diff > 0 , '***', '')) %>% 
[…]

### 2016-12-22 precip/analysis/make_prediction_score_tables.R

@@ -21,7 +21,7 @@ overall_score_table <-
 oaxt <- xtable(overall_score_table, caption = 'Comparison of model predictions from climate model and year effects model for each species and vital rate.  Two prediction scores are reported, MSE and lppd. Lower MSE indicates improved predictions whereas higher lppd indicates improved predictions.  Instances where the climate model outperformed the random year effects model are marked with "***" in the last column. ARTR = \\textit{A. tripartita}, HECO = \\textit{H. comata}, POSE = \\textit{P. secunda}, PSSP = \\textit{P. spicata}.' , 
 label = 'table:overallPreds')
 
-print(oaxt, 'output/results_tables/overall_prediction_score.text', type = 'latex', include.rownames = F)
+print(oaxt, 'manuscript/overall_prediction_score.tex', type = 'latex', include.rownames = F, caption.placement ="top")
 
 # ---------- predictions by treatment level ------------------------------------------------------- # 
 
@@ -61,8 +61,8 @@ command <- paste0("\\hline\n\\endhead\n",
                   "\\endlastfoot\n")
 add.to.row$command <- command
 
-print(txt, 'output/results_tables/treatment_prediction_score.text', type = 'latex', show.rownames = F, hline.after=c(-1), add.to.row = add.to.row,
-      tabular.environment = "longtable", floating = F) 
+print(txt, 'manuscript/treatment_prediction_score.tex', type = 'latex', show.rownames = F, hline.after=c(-1), add.to.row = add.to.row,
+      tabular.environment = "longtable", floating = F, caption.placement ="top") 
 
 
 species_improvement <- 

### 2016-12-22 precip/analysis/treatment_trends_precip.R

@@ -268,10 +268,10 @@ xt3 <- xtable(lc.POSE, caption = 'Treatment effects on log cover change for \tex
 xt4 <- xtable(lc.PSSP, caption = 'Treatment effects on log cover change for \textit{P. spicata} from 2011 to 2016. Intercept gives control effects.', 
               label = 'table:changePSSP')
 
-print(xt1, 'output/results_tables/ARTR_cover_change.txt', type = 'latex')
-print(xt2, 'output/results_tables/HECO_cover_change.txt', type = 'latex')
-print(xt3, 'output/results_tables/POSE_cover_change.txt', type = 'latex')
-print(xt4, 'output/results_tables/PSSP_cover_change.txt', type = 'latex')
+print(xt1, 'manuscript/ARTR_cover_change.tex', type = 'latex', caption.placement ="top")
+print(xt2, 'manuscript/HECO_cover_change.tex', type = 'latex', caption.placement ="top")
+print(xt3, 'manuscript/POSE_cover_change.tex', type = 'latex', caption.placement ="top")
+print(xt4, 'manuscript/PSSP_cover_change.tex', type = 'latex', caption.placement ="top")
 
 temp$percent_change <- 100*(temp$coverDiff/temp$cover.2011 )
 

### 2016-12-22 precip/manuscript/precip_ms.tex
fixing some labeling of the figures and changing the supporting information to an appendix in the ms

@@
 We tested whether longterm observational data relating the effects of climate variation to the survival, growth and recruitment of four common plants in a sagebrush steppe could be produce accurate predictions for how each species would respond to a drought and irrigation experiment.
+
 We used rainout shelters and automatic sprinklers to manipulate the soil moisture in sixteen new plots established at the same field site as the longterm observational data.
@@
 We compared predictions made by the the no climate and climate models to the observed experimental responses.
+
 Over the course of the experiment, average cover of the perennial grasses, \textit{Hesperostipa comata} and \textit{Pseudoroegneria spicata} declined significantly in the drought treatment.
@@
 spicata}.
+
 Observational climate data held valuable information for predicting species' responses to a climate change experiment in this ecosystem.
@@
 All quadrats were located on similar topography and soils.
-In most years until 1957, all individual plants in each quadrat were mapped using a pantograph \citep{blaisdell_}.
+In most years until 1957, all individual plants in each quadrat were mapped using a pantograph \citep{blaisdell_seasonal_1958}.
 The historical data set is public and available online \citep{zachmann_mapped_2010}.
@@
 We expected that our precipitation manipulation experiment would affect plants by altering available soil moisture during the growing season.
-Because we do not have direct soil moisture measures for each year of observed plant cover in the historical record, we used the SOILWAT soil moisture model to estimate daily soil moisture at the USSES from 1925 to the present \citep{Parton_1978}.
-We used an enhanced version of soilwat that has recently been developed for use in semi-arid shrubland ecosystems \citep{Bradford}.
+Because we do not have direct soil moisture measures for each year of observed plant cover in the historical record, we used the SOILWAT soil moisture model to estimate daily soil moisture at the USSES from 1925 to the present \citep{sala_long-term_1992}.
+We used an enhanced version of SOILWAT that has recently been developed for use in semi-arid shrubland ecosystems \citep{bradford_ecohydrology_2014}.
 SOILWAT uses daily weather data, ecosystem specific vegetation properties and site specific soil properties to estimate water balance processes.
@@
 The day immediately following rainfall was also classified as rainy.
-We fit this model using the \textit{lmer} package in \textit{R} \citep{bates} with random effects for plot group and date.
+We fit this model using the \textit{lmer} package in \textit{R} with random effects for plot group and date \citep{bates_fitting_2015}.
 We then used this model to predict the treatment effects on soil moisture for the entire study period from the ambient soil moisture values predicted from the SOILWAT model described above.
@@
 After fitting the vital rate models with the selected soil moisture variable we observed a trend towards positive soil moisture effects on growth of all three grasses (fig \ref{fig:climateGrowth}).
-For \textit{H.comata} the soil moisture of the most recent summer ($summer_1$) had a significantly positive effect while the soil moisture during $summer_0$ and $fall_lag$ were also positive but not significant.
+For \textit{H.comata} the soil moisture of the most recent summer ($summer_1$) had a significantly positive effect while the soil moisture during $summer_0$ and $fall_{lag}$ were also positive but not significant.
 For \textit{A.
@@
 Finally for \textit{P.
-spicata} there was a significant positive effect of $spring_lag$ soil moisture on survival.
-
-There were only two significant effects of soil moisture on recruitment: $fall_lag$ soil moisture had a positive effect on \textit{P.
-secunda}, and $summer_lag$ soil moisture had a negative effect on \textit{P.
+spicata} there was a significant positive effect of $spring_{lag}$ soil moisture on survival.
+
+There were only two significant effects of soil moisture on recruitment: $fall_{lag}$ soil moisture had a positive effect on \textit{P.
+secunda}, and $summer_{lag}$ soil moisture had a negative effect on \textit{P.
 spicata} recruitment (fig \ref{fig:climateRecruitment}).
@@
 Parameters are ordered chronologically from most recent to the current growing season on the right to most distant on the left.
-Dark blue parameters show size x climate interaction effects.
+Red parameters show size x climate interaction effects.
 We assessed a parameter as significant when the 95\% Bayesian credible intervals did not overlap zero.
@@
 	\includegraphics[width=1\textwidth]{climate_effect_recruitment}
-	\caption{Parameter estimates for the selected seasonal soil moisture covariates on the survival of all four species.
+	\caption{Parameter estimates for the selected seasonal soil moisture covariates on the recruitment of all four species.
 Parameters are ordered chronologically from most recent to the current growing season on the right to most distant on the left.
-Dark blue parameters show size x climate interaction effects.
 We assessed a parameter as significant when the 95\% Bayesian credible intervals did not overlap zero.
@@
 	\caption{The treatment effects predicted by the climate model compared to the treatment effects observed for the intercept parameters (left side) and size by climate/treatment effects (right side).
+Parameters from all species and vital rates are shown together.
 The observed treatment effects come from the treatment model fitted to all the data including the five years of the experiment.
@@
 Gray line shows 1:1 line.}
-	\label{fig:pgrART}
+	\label{fig:pgrARTR}
 \end{figure}
@@
 \setcounter{table}{0}
-\renewcommand{\theequation}{SI.\arabic{equation}}
-\renewcommand{\thetable}{SI-\arabic{table}}
-\renewcommand{\thefigure}{SI-\arabic{figure}}
-\renewcommand{\thesection}{Section SI.\arabic{section}}
-
-\centerline{\Large \textbf{Supporting Information }}
+\renewcommand{\theequation}{A.\arabic{equation}}
+\renewcommand{\thetable}{A-\arabic{table}}
+\renewcommand{\thefigure}{A-\arabic{figure}}
+\renewcommand{\thesection}{Section A.\arabic{section}}
+
+\centerline{\Large \textbf{Appendix}}
 \centerline{Kleinhesselink et al., ``Predicting climate response''}
@@
 
-\section{Supplementary Methods} \label{suppMethods}
-
-\subsection{Interspecific covariance in local crowding} 
-We explored interspecific covariance in local crowding experienced by individual plants, by regressing the $W$ values exerted by one neighbor species, the response variable, against the $W$ values of all other species, the independent variables.
-Because some $W=0$, we conducted two separate regressions.
-First, using all $W$'s, we fitted a generalized linear model with a logit link function to evaluate whether the probability that the focal species' $W=0$ is influenced by the value of other species' $W$'s.
-In this model, the dependent variable is a Bernoulli variate coding for the zero or non-zero value of the focal species' crowding, and the independent variables are the $W$'s for all other species.
-Second, for the set of records in which the focal species has $W>0$, we performed a linear regression, where the focal species' $W$ is the dependent variable, and the other species' $W$'s are the independent variables.
-We repeated these regressions for each focal species.
-Due to large samples size, interspecific $W$ values were often statistically significant predictors of intraspecific.
-However, they explained very little variance.
-The maximum reduction in deviance for the generalized linear regressions and $R^2$ for the linear regressions were both less than 8\%.
-The \texttt{R} code for this analysis is included as ..\texttt{\textbackslash Wdistrib\textbackslash exploreSurvivalWs.r}.
-
-\subsection{Mean field approximation of local crowding for the IPM} 
-\citet{adler_coexistence_2010} developed a mean field approximation for local crowding when the
-competition kernels are all Gaussian functions, $F_{jm}(d) = e^{-\alpha_{jm} d^2}$. The approximation is explained in 
-the online SI to \citet{adler_coexistence_2010} and in section 5.3 of \citet{Ellner2016}. 
-Here we explain how that approximation was modified for the IPMs in this paper, which
-used fitted nonparametric competition kernels.
-
-For $j \ne m$ (between-species competition), overlap between individuals is allowed. The mean field approximation is 
-that from the perspective of any focal plant in species $j$, individuals of species $m$ are distributed at random in space, 
-independent of each other and of their size.
-
-Consider the region between the circles of radius $x$ and $x+dx$ centered on a focal genet of species $j$. The area of this annulus
-is $2 \pi x \; dx$  to leading order for $dx \approx 0$. A species $m$ genet 
-in the annulus puts competitive pressure $F_{jm}(x)$ times its area on
-the focal genet. The expected total competitive pressure from all such genets 
-is therefore is $F_{jm}(x) 2 \pi x \; dx$ times the expected fractional cover of species $m$ in the annulus 
-(fractional cover is the total area of species $m$ genets, as a fraction of the total area). The excepted fractional cover $C_m$ of species $m$
-in the annulus equals its fractional cover in the habitat as a whole, because of the assumption of random distribution
-spatial distributions. We therefore have $C_m  = \int e^u n_m(u,t) du/A$ where $A$ is the total area of the habitat. 
-The total expected competitive pressure on a species-$j$ genet due to species $m$ is then 
-\begin{equation}
-W_{jm} = \int_0^\infty{C_m F_{jm}(x) 2 \pi x \; dx}  = C_m \left [2 \pi \int_0^{\infty} x F(x) \, dx \right ].
-\label{eqn:wbarm}
-\end{equation} 
-The quantity in square brackets is a constant (that is, it only depends on what the kernel function
-is) so it can be computed once and for all for each kernel used in the IPM. The integral is finite because
-all fitted kernels fall to zero at a finite distance from the focal plant.
-
-Our kernel fitting method only uses competition kernel values at the ``mid-ring'' distances
-halfway between the inner and outer radii of a series of annuli around each focal
-plant, scaled so that the value at the innermost mid-ring distance equals 1. 
-In the IPM we defined the kernel at other distances by linear interpolation between values at 
-mid-ring distances, except that for the innermost ring a kernel value of 1 was specified at the
-outer radius of the ring and at distance $x=0$.
-
-Now consider within-species competition. We assume that conspecifics cannot overlap. Genet shapes are irregular, but we 
-nonetheless implement the no-overlap rule by assuming that a genet of log area $u_i$ is a 
-circle of radius $r_i$ where $\pi r_i^2 = e^{u_i}$. The no-overlap rule is then that the centroids of two conspecific individuals 
-must be separated by at least the sum of their radii.
-
-For any one focal genet, the no-overlap restriction on its neighbors' locations affects 
-only a negligibly small part of the habitat. The expected cover of individuals in the places
-where they can occur (relative to one focal plant) is thus assumed to equal their expected locations
-in the habitat as a whole.
-
-Let $C_m(u)$ be the total cover of species $m$ genets of radius $r$ or smaller, 
-\begin{equation}
-C_m(r) = \int_L^{\log(\pi r^2)}{\! \! \! e^z n_m(z,t) \, dz} .
-\label{eqn:cm}
-\end{equation}
-A focal genet of radius $r$ cannot have any conspecific neighbors centered 
-at distances less than $r$. It can have a neighbor centered at distance $x>r$ if that neighbor's
-radius is no more than $x-r$. Adding up the expected cover of all such possible neighbors
-for a focal genet of radius $r$,    
-\begin{equation}
-W_{mm}(r) = 2 \pi \int_r^{\infty}F_{mm}(x) x C_m(x-r) \, dx
-\label{eqn:wbarmr} 
-\end{equation}
-This integral is again finite and computable because the kernels $F$ fall to 0 at finite $x$.
+\section{Additional Tables} \label{appendix}
+
+\input{treatment_prediction_score}
 
@@
 
-\input{treatment_prediction_score}
-
-\clearpage
-\newpage
-\clearpage
-\newpage
 \section{Additional Figures}
@@
 	\caption{Comparison of treatment effects predicted and observed for \textit{A.
-survival} survival.
+tripartita} survival.
 The observed treatment effects come from the treatment model fitted to all the data including the five years of the experiment.
@@
 
+\begin{figure}[!htbp]
+	\centering
+	\includegraphics[width=1\textwidth]{pred_v_ob_treatment_ARTR_recruitment}
+	\caption{Comparison of treatment effects predicted and observed for \textit{A.
+tripartita} recruitment.
+The observed treatment effects come from the treatment model fitted to all the data including the five years of the experiment.
+The predicted parameters come from the climate model fitted to all years of observational data but do not include the five years of the experiment (2011 to 2016).
+We assessed a parameter as significant when the 95\% Bayesian credible intervals did not overlap zero.}
+	\label{fig:parPredARTRRecruitment}
+\end{figure}
+
+\begin{figure}[!htbp]
+	\centering
+	\includegraphics[width=1\textwidth]{pred_v_ob_treatment_HECO_recruitment}
+	\caption{Comparison of treatment effects predicted and observed for \textit{H.
+comata} recruitment.
+The observed treatment effects come from the treatment model fitted to all the data including the five years of the experiment.
+The predicted parameters come from the climate model fitted to all years of observational data but do not include the five years of the experiment (2011 to 2016).
+We assessed a parameter as significant when the 95\% Bayesian credible intervals did not overlap zero.}
+	\label{fig:parPredHECORecruitment}
+\end{figure}
+
+\begin{figure}[!htbp]
+	\centering
+	\includegraphics[width=1\textwidth]{pred_v_ob_treatment_POSE_recruitment}
+	\caption{Comparison of treatment effects predicted and observed for \textit{P.
+secunda} recruitment.
+The observed treatment effects come from the treatment model fitted to all the data including the five years of the experiment.
+The predicted parameters come from the climate model fitted to all years of observational data but do not include the five years of the experiment (2011 to 2016).
+We assessed a parameter as significant when the 95\% Bayesian credible intervals did not overlap zero.}
+	\label{fig:parPredPOSERecruitment}
+\end{figure}
+
+\begin{figure}[!htbp]
+	\centering
+	\includegraphics[width=1\textwidth]{pred_v_ob_treatment_PSSP_recruitment}
+	\caption{Comparison of treatment effects predicted and observed for \textit{P.
+spicata} recruitment.
+The observed treatment effects come from the treatment model fitted to all the data including the five years of the experiment.
+The predicted parameters come from the climate model fitted to all years of observational data but do not include the five years of the experiment (2011 to 2016).
+We assessed a parameter as significant when the 95\% Bayesian credible intervals did not overlap zero.}
+	\label{fig:parPredPSSPRecruitment}
+\end{figure}
+
 \end{document}

### 2016-12-23 precip/manuscript/precip_ms.tex
adding discussion revision

@@
 
-\smallskip \textbf{Authorship statement:} ARK and PBA designed the experiment and supervised data collection. AT helped with the data analysis. JB and CA helped with the soil moisture model. 
+\smallskip \textbf{Authorship statement:} ARK and PBA designed the experiment and supervised data collection.
+AT helped with the data analysis.
+JB and CA helped with the soil moisture model.
+ARK wrote the first draft of the manuscript which all authors edited.
 \smallskip
@@
 
-Climate is an important driver of population ecology however there have been few tests of whether observational data correlating population performance with climate variation can be used to predict how populations respond to climate change experiments.
-We tested whether longterm observational data relating the effects of climate variation to the survival, growth and recruitment of four common plants in a sagebrush steppe could be produce accurate predictions for how each species would respond to a drought and irrigation experiment.
-
-We used rainout shelters and automatic sprinklers to manipulate the soil moisture in sixteen new plots established at the same field site as the longterm observational data.
-After running the experiment for five years, we analyzed how the growth, survival and recruitment of each species responded to the experimental drought and irrigation treatments.
-We made predictions using two types of models fit to the observational data collected prior to the experiment: a no climate model that only included the effects of plant size and competition on plant performance in each plot in each year; and a climate model that also included the effects of three seasonal soil moisture variables.
-We then used these models to predict the performance of each species during the course of the experiment.
-We also generated one step a head predictions of population size in each experimental plot using an individual based population model.
-We compared predictions made by the the no climate and climate models to the observed experimental responses.
+Climate is an important driver of population dynamics and annual variation in demographic rates often correlate with variation in weather.
+However, the predictive potential of such correlations is largely unknown.
+We tested how well population models fit using long-term observational could predict the responses of four dominant plant species in a sagebrush steppe to a drought and irrigation experiment.
+
+We used rainout shelters and automatic sprinklers to manipulate the soil moisture in sixteen plots established at the same field site where long-term observational data was collected.
+We report how the growth, survival and recruitment of each species responded to the experimental drought and irrigation treatments after five years.
+We then attempted to predict these treatment responses using two models fit to the observational data collected prior to the experiment: a no climate model that only included the effects of plant size and competition on plant performance in each plot in each year, and a climate model that also included the effects of three seasonal soil moisture variables.
+We compared predictions made by the no climate and climate models to the actual experimental responses.
+We also generated one-step-ahead predictions of population size in each experimental plot using individual based population parameterized with the vital rate estimates from the climate and no climate models.
 
@@
 tripartita}.
-The climate model better predicted the response to the experiment than the no climate model in six out of twelve cases.
-Across all species and vital rates, there was a strong positive correlation between the observed effects of the experiment and the effects predicted by the climate model fit to the observational data.
-At the population-level, the climate model made better predictions of the changes in species cover than the no climate model for \textit{P.
+The climate model made better predictions of the experimental responses than the no climate model in six out of twelve cases.
+Across all species and vital rates, there was a strong positive correlation between the observed responses to the treatments and the responses predicted by the climate model.
+At the population-level, the climate model predicted changes in species cover more accurately than the no climate model for \textit{P.
 secunda} and \textit{P.
@@
 Treatment responses often matched the direction of predicted responses even when the effects were not significant.
-We were better able to predict species' responses to the drought treatment than to the control and irrigation treatments.
-This suggests that soil moisture is an important factor in predicting the population dynamics of these species but only when water is truly limiting.
+We were better able to predict species' responses to the drought treatment than to the control and irrigation treatments, suggesting that soil moisture is an important factor for predicting the population dynamics of these species but only when water is truly limiting.
 
@@
 
-Ecologists often resort to one of two methods for predicting the effects of future climate change on populations and communities: they may use experiments to manipulate aspects of climate directly and observe the response of populations \citep{elmendorf_experiment_2015,knapp_pushing_2016,compagnoni_warming_2014}; or they may use longterm observational data on species performance and abundance collected over many years and relate this to ambient annual variation in climate  \citep{koons_climate_2012,lunn_demography_2016,dalgleish_climate_2010,jenouvrier_demographic_2009}.
-The strength of the experimental approach is in the stronger inference that comes from manipulating some aspects of climate while controlling for other factors; for instance knowing that loss of snow cover, and not necessarily changes in soil temperature or moisture are the factors causing a species performance to change with warming \citep{compagnoni_warming_2014}.
+Ecologists often resort to one of two methods for predicting the effects of future climate change on populations and communities: they may use experiments to manipulate aspects of climate directly and observe the response of populations \citep{elmendorf_experiment_2015,knapp_pushing_2016,compagnoni_warming_2014}; or they may use long term observational data on species performance and abundance collected over many years and relate this to ambient annual variation in climate  \citep{koons_climate_2012,lunn_demography_2016,dalgleish_climate_2010,jenouvrier_demographic_2009}.
+The strength of the experimental approach is in the stronger inference that comes from manipulating some aspects of climate while controlling for other factors; for instance, knowing that loss of snow cover, and not necessarily changes in soil temperature or moisture are factors causing a species performance to change with warming \citep{compagnoni_warming_2014}.
 It also allows for the creation of conditions that may be more extreme than those observed historically \cite{knapp_pushing_2016} but are possible in the future.
@@
 
-Using observational data to predict the effects of climate variation on populations has many advantages over using experiments.
-The primary one being the ever increasing availability of long-term ecological data and ever increasingly detailed climate data.
+The use of observational data to predict the effects of climate variation on populations has its own advantages and disadvantages.
+One important advantage is cost: analyses of already existing long-term ecological data and ever increasingly detailed climate data are cheaper than experiments.
 Observational studies may also be the only way to study the effects of climate on large and or migratory species, for which it would be difficult to manipulate climate \citep{koons_climate_2012,jenouvrier_demographic_2009,aubry_climate_2013}.
-However, there are many potential drawbacks to using essentially correlative relationships between species performance and climate to predict future species' responses.
+The principal disadvantage is the reliance on essentially correlative relationships between species performance and climate to predict future species' responses.
 First, many years of data are needed to reliably detect climate effects, especially when annual variation in demographic rates is high \citep{teller_linking_2016,gerber_optimal_2015}.
 \citep{teller_linking_2016} estimate that even cutting edge statistical approaches for fitting relationships between climate and species performance require at least 20-25 years of independent climate observations before they perform well.
-Moreover, even strong correlations between species performance and the climate covariates we choose to include in our models may not be due to direct causation and may not hold up in future years of observation \citep{hilborn_correlation_2016}.
-
-And yet even overcoming this challenge, predicting the effects of climate change would face yet another challenge: in many parts of the world, climate change will lead to mean annual temperatures that will be above the range of historical variation.
-If species performance responds non-linearly to temperature, we may be unable to predict how species respond to temperatures outside the range of observed variation \citep{doak_demographic_2010}.
-In addition, climate change will not just change the average climate, but is also likely to increase the variance in precipitation and lead to extreme events, which will have their own consequences independent of changes in climate means \citep{gherardi_enhanced_2015,jentsch_new_2007}.
-Any models based on observations drawn from the historical range of variation will therefore be extrapolating beyond both the range of observed averages and variance when used to predict the future \citep{williams_novel_2007}.
-Demonstrating that models fit to historical data can accurately predict the effects of experimental climate manipulations, especially manipulations that generate extreme conditions, would be strong confirmation that the climate effects they describe are not spurious and will hold even in the novel conditions of the future \cite{adler_can_2013}.
-
-There is abundant evidence that plant performance shows high year to year variation, both at the level of individual growth, survival and reproduction and total population abundance.
-Precipitation is often especially important in driving interannual variation in net primary productivity among terrestrial plants \citep{knapp_variation_2001,hsu_anticipating_2014}, the annual growth rings in trees \citep{yang_3500-year_2014} and smaller plants \citep{srur_annual_2009,franklin_growth_2013}, as well as the germination and reproductive output of many annuals \citep{venable_bet_2007}.
-
-Despite the clear signs that precipitation drives net primary productivity at the ecosystem level and annual variation in individual growth rate in many plants, there have been relatively few studies that clearly link observed climate variation to species performance in population models \citep{ehrlen_advancing_2016}.
-Fewer still have tested whether the population-level effects of climate variation shown in observational data can actually be used to predict the responses of plant populations to future climate variation.
-\citep{adler_can_2013} showed that population models based on observed correlations between plant population growth rates and precipitation did have some predictive power in describing species response to a short-term climate manipulation in a North American grassland.
+Moreover, even strong correlations between species performance and the climate covariates we choose to include in our models may not reflect direct causation, leading to failures when predicting future, out of sample performance \citep{hilborn_correlation_2016}.
+
+The extrapolation of climate-demography correlations presents another potential problem.
+In many systems, future precipitation and temperature will fall outside the range of historical variation.
+If species performance responds non-linearly to these drivers, fitting linear models for species responses to climate may produce larger errors when future conditions are outside the range of observed variation \citep{doak_demographic_2010}.
+In addition, climate change will not only alter average weather, but is also likely to increase the variance in precipitation and the frequency of extreme events, which will have their own consequences independent of changes in means \citep{gherardi_enhanced_2015,jentsch_new_2007}.
+Any models based on observations drawn from the historical range of variation will therefore be extrapolating beyond both the range of observed averages and variances when used to predict the future \citep{williams_novel_2007}.
+
+Here, we combine the strengths of experimental and observational approaches by testing the ability of models fit to historical data to predict the effects of experimental climate manipulations that generate extreme conditions.
+A demonstration that the observational approach can skillfully predict experimental responses would provide strong confirmation that observed climate-demography correlations are not spurious and will hold even in novel conditions in the future \cite{adler_can_2013}.
+\citep{adler_can_2013} showed that population models based on observed correlations between plant population growth rates and precipitation did have some predictive power in describing species response to a short-term climate manipulation.
 Three species showed responses to experimentally imposed drought and irrigation that were well predicted by population models fitted to historical observations.
-However, another three species, showed responses to the experimental conditions that were not well predicted by historical observations.
-The authors suggested that limited replication in the historical data for two of these species and changing competitive conditions in the community may have led to the poor predictions.
-
-The demography of and competitive interactions between three dominant perennial bunchgrasses and a shrub in a sagebrush steppe plant community at the US Sheep Experiment Station near Dubois, Idaho have been described in at least seven different studies since 2010 \citep{adler_coexistence_2010,adler_forecasting_2012,adler_weak_2009,adler_weak_2016,chu_direct_2016,chu_large_2015,dalgleish_climate_2010}.
-And several of these studies report significant effects of climate variation on the vital rates and overall population growth rates of these species \citep{dalgleish_climate_2010,adler_forecasting_2012,adler_weak_2009,chu_direct_2016}.
-This well-studied system offers the ideal opportunity to test whether statistical associations between annual climate and plant demography in longterm observational data can be used to predict the responses of plant populations to experimental climate change experiments.
+However, the responses of another three species, were not well predicted by historical observations.
+
+Among plant populations, interannual variation in precipitation and or soil moisture often drives variation in net primary productivity \citep{knapp_variation_2001,hsu_anticipating_2014}, the annual growth rates of the woody tissue in trees and shrubs (\citep{yang_3500-year_2014},\citep{srur_annual_2009,franklin_growth_2013}), and the germination and reproductive output of annuals \citep{venable_bet_2007}.
+Despite clear signs that precipitation shold be important for plant populations, there have been relatively few studies that clearly link observed variation in precipitation to species performance in population models \citep{ehrlen_advancing_2016}.
+
+The sagebrush steppe plant community at the US Sheep Experiment Station near Dubois, Idaho offers an ideal opportunity to test whether the climate effects in plant populations models derived from observational data can also be used to predict species responses to controlled precipitation experiments.
+The demography of three perennial bunchgrasses and a shrub species at the USSES have been described in detail in seven different studies since 2010, several of which report significant effects of seasonal precipitation on the vital rates and overall population growth rates of these species (\citep{adler_coexistence_2010,adler_forecasting_2012,adler_weak_2009,adler_weak_2016,chu_direct_2016,chu_large_2015,dalgleish_climate_2010}).
 
 In this study, we report how the four dominant plant species at the USSES respond to a five year drought and irrigation experiment and use the results to address two research questions: first, how much do the growth, recruitment and survival of our target species differ between the precipitation manipulation treatments?
-If our experiment does affect species vital rates we interpret that as strong evidence that changes in precipitation should have an effect on populations in the future.
+Significant experimental effects on species vital rates imply that future changes in precipitation will impact populations.
 Second, can we predict each species' response to the experimental conditions based on how they respond to natural climate variation in the observational data?
-If models based on observational data can predict the response of species to this experiment this should give us more confidence in longterm population monitoring data to predict species response to climate change.
+If models based on observational data can predict the response of species to this experiment we will gain confidence in using long-term population monitoring data to predict species responses to future climate change.
 
@@
 We refer to the first dataset as the observational data.
-It consists of all the historical data collected from 1925 to 1957 as well as the modern data collected from the same plots from 2007 to 2010.
+It consists of all the historical data collected from 1925 to 1957 as well as the contemporary data collected from the same plots from 2007 to 2010.
 These data record the response of plants in each plot to the ambient climate variation.
@@
 In order describe the effects of the experimental treatments on each vital rate, we fit "treatment" models.
-The treatment models included parameters for the effects of the drought and irrigation treatments on each vital rate.
-We fit these models to both datasets together, all the experimental and all the observational data.
-We did this because we wanted to focus our predictions on the effects of the experimental treatments on the vital rates, rather than any differences between the datasets in the effects of crowding and plant size on the vital rates.
-
-Next in order to test how well we could predict the responses in the experimental plots, we fit two classes of models to the observational dataset only.
-Thus when we used these models to predict the response of each species to the five years of experimental data we are generating predictions for out of sample data.
-We refer to the first type of model as the "no climate" models.
-These models include parameters for the effects of competition on each vital rate and the size dependence of survival and growth but they do not include climate or treatment effects.
-Lastly, we fit the "climate" models which include the effects of annual variation in soil moisture on each vital rate.
-The no climate model provides us a baseline by which to measure the accuracy of the predictions from the climate model.
+The treatment models included parameters representing the effects of the drought and irrigation treatments on each vital rate.
+We fit these models to using all the experimental and all the observational data.
+We combined the datasets because we wanted to focus our predictions on the effects of the experimental treatments on the vital rates, rather than any differences between the historical and contemporary periods in the effects of crowding and plant size on the vital rates.
+
+Next, in order to test how well we could predict the responses in the experimental plots, we fit two classes of models to the observational dataset only.
+The "no climate" models include parameters for the effects of competition on each vital rate and the size dependence of survival and growth but they do not include climate or treatment effects.
+The "climate" models include the effects of annual variation in soil moisture on each vital rate.
+The no climate model provides us a baseline by which to measure the accuracy of predictions from the climate model.
 Because much of the variation in growth, survival and recruitment in this system can be explained by plant size and competition, we expect that these two models will make similar predictions for individual plant performance in the experiment.
 However, if the climate model makes more accurate predictions than the no climate model, this indicates that the climate parameters it includes contain useful information for prediction.
+Note that because these models are fit using only the observation dataset, when we use  these models to predict experimental responses we are generating true out-of-sample predictions.
 
@@
 Our experiment showed that observational data on the response of plant populations to interannual climate variation can indeed help us predict the direction of species responses to experimental climate manipulations (fig \ref{fig:parPredictions}).
-This was true even though adding climate parameters to the demographic models only improved vital rate predictions for half of the models (fig \ref{table:overallPreds}).
+The historical climate-demography correlations helped predict the direction of experimental responses even though adding climate parameters to the demographic models only improved vital rate predictions for half of the models (fig \ref{table:overallPreds}).
 This should give us some hope that even when climate effects in demographic models fit to observational data are weak or not significant, they may contain useful qualitative information on the direction of climate effects in the future.
 
-Although a number of previous studies of the plants in this system have looked the effects of climate on demography and survival, this is the first study to demonstrate effects of climate experimentally.
-The significant decline in \textit{H.
-comata} cover we report in the experiment (fig \ref{fig:coverChange}) is consistent with the population-level sensitivities to precipitation estimated by previous studies (\citep{chu_direct_2016},\citep{dalgleish_climate_2010}, \citep{adler_forecasting_2012}).
-Each of these studies report \textit{H.
-comata} having the strongest positive population response to increased precipitation among all the species; this is matched by our finding that this species has the strongest negative response to experimental drought (fig \ref{fig:growthTreat}, \ref{fig:survivalTreat}).
-On the other hand if we had only conducted an irrigation experiment our results may not have shown this consistency with previous work as \textit{H.
+\subsection*{Comparison of experimental and natural climate effects}
+While previous studies in this system used the observational data to describe the effects of climate on demography and survival, this is the first study to demonstrate effects of climate experimentally.
+While previous studies of this system fit a variety of climate effects and used different modeling approaches, we see many points of similarity between these studies in the responses of the four dominant species to precipitation (\citep{chu_direct_2016},\citep{dalgleish_climate_2010}, \citep{adler_forecasting_2012}).
+First, in all three studies the strongest positive effects of precipitation among the four species are reported for \textit{H.
+comata}; this matches the negative effects of our drought experiment on this species (fig \ref{fig:coverChange}).
+This effect is driven by a negative growth and survival response to drought (fig \ref{fig:growthTreat}, \ref{fig:survivalTreat}).
+On the other hand, if we had only conducted an irrigation experiment our results may not have shown this consistency with previous work as \textit{H.
 comata} showed no positive response to irrigation.
-Previous work has also generally reported postive population responses to precipitation for the other grasses, \textit{P.
+Previous studies also reported positive effects of precipitation on the other grasses, \textit{P.
 secunda} and \textit{P.
-spicata} (\citep{adler_forecasting_2012}, \citep{chu_direct_2016}, \citep{dalgleish_climate_2010}).
-These results are consistent with significant decrease we saw in \textit{P.
-spicata} cover and survival and \textit{P.
-secunda} growth and survival in the drought plots in our experiment (figs \ref{fig:coverChange}, \ref{fig:growthTreat}, \ref{fig:survivalTreat}).
-The magnitudes of the negative drought effects were greater than the magnitudes of the positive irrigation effects on these grasses.
-
-In contrast to the effects on the grasses, the effects on \textit{A.
+spicata}.
+Again our results are consistent with this result: drought led to declines in cover of \textit{P.
+spicata}, \ref{fig:coverChange}, and in the growth and survival of \textit{P.
+secunda} (figs \ref{fig:growthTreat}, \ref{fig:survivalTreat}).
+As for the \textit{H.
+comata} the magnitudes of drought effects were greater than the irrigation effects on these grasses.
+
+The effects of precipitation on \textit{A.
 tripartita} are more complicated.
-The effects of drought and irrigation on \textit{A.
-tripartita} growth were highly dependent on plant size (figs \ref{fig:growthTreat}): drought had a positive effect except possibly for the largest plants (fig \ref{fig:parPredARTRGrowth}).
-This result matches the negative effects of precipitation on this species estimated by \citep{adler_forecasting_2012} and \citep{chu_direct_2016}.
+Previous research reported negative direct effects of precipitation on this species (\citep{adler_forecasting_2012}, \citep{chu_direct_2016}).
+This effect has always seemed odd because it is hard to imagine why precipitation would have a direct negative effect on a species in this dry ecosystem.
+But again the largely positive (but size dependent) effects of drought treatments on \textit{A.
+tripartita} growth should give us more confidence in the negative effects of precipitation shown in the historical data.
 These studies also report relative strong indirect effects of precipitation on \textit{A.
@@
 It is possible that some of the positive effect on \textit{A.
-tripartita} growth we see in the drought plots is the result of reduced grass competition these plots (fig \ref{fig:coverPred}; \citep{chu_direct_2016}).
+tripartita} growth in our drought plots is the result of reduced grass cover (fig \ref{fig:coverPred}; \citep{chu_direct_2016}).
 However, our growth model includes interspecific crowding and so should take into account any changes in grass abundance that could be driving a positive response from \textit{A.
 tripartita}.
-So this leaves us with the question of why this species would show a positive direct response to drought?
-Although there is some evidence that saturated soils can be detrimental to big sagebrush (\textit{A.
-tridentata}), a closely related species \citep{sturges_response_1989}, there was no evidence in our data that soils were saturated with moisture in the ambient or irrigation treatments (fig \ref{fig:dailyVWC}).
-In a similar study to ours, \citep{germino_desert_2014} showed that big sagebrush increased in response to deep soil moisture but actually decreased in response to shallow soil moisture.
-Perhaps \textit{A.
-tripartita} responds similarly and benefits directly from the drier soils.
-Another possible explanation is that our drought treatments reduced snow cover in the winter and early spring, an effect that has been shown to benefit big sagebrush in other ecosystems \citep{perfors_enhanced_2003}.
-
-One trend that appears to stand out in these results is the relatively asymmetrical effects of drought and irrigation.
-Although our irrigation plots raised soil moisture, we saw little positive effect of added moisture on the same species that were reduced by drought (figs \ref{fig:growthTreat}, \ref{fig:survivalTreat}).
-We expected that in this water limited system, increasing soil moisture would lead to a stronger increase in plant performance.
-But perhaps we should not have been surprised by this result.
-It is possible that adding water to this system does little to increase grass growth and survival for a number of reasons.
-First, these species are presumably able to escape drought, by growing early in the year, or avoid drought stress later in the year through high water use efficiency \citep{bazzaz_physiological_1979,franks_plasticity_2011}.
-If the species in our system are well adapted to the ambient water limitation, it makes some sense that they would show relatively small responses to increased moisture.
+This leaves us with the question of why this species would show a positive direct response to drought.
+Although there is some evidence that saturated soils in the spring are detrimental for big sagebrush (\textit{A.
+tridentata}), a closely related species \citep{sturges_response_1989,germino_desert_2014}, soil saturation would conservatively seem to require soils to be above 30 or 40\% volumetric water content for several weeks, something that we did not observe (fig \ref{fig:dailyVWC}).
+Another possible explanation is that our drought treatments reduced snow cover in the winter and early spring, an effect that has been shown to benefit related sagebrush species in other ecosystems \citep{perfors_enhanced_2003}.
+
+Overall we were somewhat surprised by the weak effects that reducing water availability by 50\% and increasing water availability by 150\% had in this arid system.
+Cross-biome studies of the relationship between precipitation and ANPP generally show that arid systems are highly sensitive to water limitation \citep{huxman_convergence_2004}.
+We have two explanations for the seemingly weak effects of precipitation we observed on demography.
+First, we measure the size of the pernnial bunchgrasses in this system by their basal cover, which may not have a strong relationship with their annual production.
+It is likely that we would find larger effects of precipitation on these grasses if we had a more complete measurement of aboveground biomass.
+Moreover, much of the growth of these species may be going into roots.
+
+Another explanation for the weak effects of precipitation are that perennial species in this cold desert ecosystems are well adapted to tolerate drought, either through escaping drought by growing early in the year, or by avoiding drought stress later in the year through high water use efficiency \citep{bazzaz_physiological_1979,franks_plasticity_2011}.
 Indeed our soil moisture data generally show a pulse of soil moisture in the spring when many grasses are actively growing (fig \ref{fig:dailyVWC}).
-During this window of active growth soil moisture in sagebrush steppe may be quite high and perhaps other resources become limiting \citep{caldwell_rapid_1987}.
-Moreover, our treatments had relatively smaller effects on soil moisture during the spring (table ).
-
-A similar result was recently reported from another precipitation experiment in a warm desert grassland: the effect of increased soil moisture on grass productivity in that system showed a concave down pattern of response to increased soil moisture \citep{gherardi_enhanced_2015}.
-Increasing precipitation had a weaker effect on annual grass productivity than decreasing precipitation by an equal amount.
-A pattern like this is qualitatively similar to our finding significant decreases in either the cover or performance of the three grasses in drought plots, but not seeing any statistically significant increase in performance in the irrigated plots.
-
-Reflecting on the possible mechanisms by which soil moisture could have affected these species may also shed some light on our sucesses and failures at prediction.
-The drought effects we observed on the three grasses were often stronger than the effects we predicted, while the irrigation responses observed were often weaker than predicted (figs \ref{fig:parPredHECOGrowth}, \ref{fig:parPredHECOSurvival}, \ref{fig:parPredPOSEGrowth}, \ref{fig:parPredPOSESurvival}, \ref{fig:parPredPSSPGrowth}, \ref{fig:parPredPSSPSurvival}).
-If we had fit our growth and survival models with a non-linear function for soil moisture, perhaps informed by more mechanistic understanding of water limitation in this system, we may have made more accurate predictions of the drought and irrigation effects \citep{ehrlen_advancing_2016}.
-
-Out of all the climate effects we predicted and observed, we only observed four cases where vital rate predictions and observations were both significantly different from zero (fig  \ref{fig:parPredictions}).
-In three of these cases, we successfully predicted the direction of the treatment effects.
-However, for \textit{P.
-secunda} recruitment we predicted a response to irrigation that was significantly positive and yet observed a significantly negative response (figs \ref{fig:parPredictions}).
-This is arguably our greatest error in prediction.
-But there may be some mitigating factors that can excuse our error.
-First we note that we observed decreases in recruitment in both the drought and irrigation plots for \textit{P.
-secunda} and also for \textit{P.
-spicata} (fig \ref{fig:recruitmentTreat}).
-So while we observed a significant decrease in \textit{P.
-secunda} recruitment in the irrigated plots this may have been due to underlying differences in the experimental plots from the historical control plots.
-Because recruitment is modeled at the quadrat level we also had fewer observations and predictions to work with for these recruitment models than we had for the other vital rates.
-
-Using the IBM, we demonstrated that the climate model predictions of the vital rates could be used to generate population-level predictions that were also better than the no climate models in some cases: the climate model improved one step ahead predictions of cover changes of \textit{P.
+Likewise, \textit{A.
+tripartita} is more deeply rooted than the grasses and able to continue its growth throughout the growing season by drawing from deeper soil water \citep{germino_desert_2014}.
+The adaptations of native perennial plants in cold deserts could make them less sensitive to water availability than species in a more mesic ecosystem.
+
+\subsection*{Can the past predict the future?}
+Our second research question was whether we could use long term observational data on species response to precipitation to predict the response of each species to the experiment.
+Using the IBM, we to generate predicted changes in population size for each year, we found that climate model predictions were indeed better than the no climate models for two species: \textit{P.
 spicata} and \textit{P.
@@
 secunda} were also better than the no climate model.
-In this analysis, we either used the climate model, or the no climate model to generate predictions for the IBM.
-In theory we could have refined our predictions by  climate model to predict all three vital rates in the IBM.
-Improvements of our cover predictions may have been possible if we only included climate models in the IBM when those models improved vital rate predictions--for instance, the \textit{A tripartita} IBM would only include the climate model for growth but not for survival and recruitment.
-
-Mixed success at predicting species responses to high and low moisture availability has interesting implications.
-On the one hand it is reassuring that observational data captures important climate effects.
-On the other hand, will knowing that some species' responses are unpredictable, but not knowing which ones, discourage ecologists from making any predictions at all?
-Among plants, perhaps improved ecophysiological models or trait data can be used as a screen to generate hypothesis about which species are more likely to be affected by climate and in which ways.
-In our system, for instance, it may make sense that the grasses showed a stronger response to the drought treatment than the a woody shrub.
-Although these grasses are adapted to the arid conditions that characterize the sagebrush steppe, they thrive during the brief window in spring and early summer when the soil moisture and temperatures are warmer.
-\textit{A.
-tripartita} on the other hand grows throughout the summer and has deeper roots than the grasses, these traits may buffer it from some of the annual variability in soil moisture experienced by the grasses.
-
-Will our predictions be useful for the predicting the effects of climate change in the future?
-Because these species compete, one could argue that our predictions for any one species in this community will only be as good as the predictions we make for their competitors, pathogens and predators \citep{tylianakis_global_2008}.
-For instance, while we observed weak treatment effects on \textit{A.
-tripartita} in our experiment (fig \ref{fig:coverChange}) could drought ultimately lead to a decline in the cover of its competitors which will cause it to increase in abundance in the future?
-While this is possible for some species, research at our site and several other grasslands suggest that species are more limited by intra-specific competition than inter-specific competition \cite{chu_direct_2016,adler_weak_2016}.
-In theory, this should mean that we can predict species short term responses to climate change with single species models (\citep{levine_competitors_2010,kleinhesselink_indirect_2015,adler_forecasting_2012}).
-
-Our results give us more confidence that historical observational data can be used to detect and predict the demographic effects of climate change among sagebrush steppe plants.
-This should encourage more researchers to try and use observational data to predict the future in both experimental and natural settings \citep{houlahan_priority_2016}.
-Nevertheless, our success at predicting the short-term response of two species to a small-scale climate manipulation is not likely to be very reassuring to applied ecologists and resource managers wishing to make accurate predictions about the effects of climate change in large complex systems.
-Clearly more work is needed to learn how best to meet the challenge of prediction in ecological systems.
+Moreover, we also found that climate models produced better predictions of species vital rates for half of the species/vital rate combinations we tested (table \ref{table:overallPreds}).
+The rate of overall improvement in predictive ability produced by the climate models over the no climate models was similar to the results reported by \citep{adler_can_2013} who also reported improved population-level predictions for half of the species predicted.
+Likewise, in a within sample cross-validation analysis, \citep{tredennick_we_2016} found that including climate covariates improved population level predictions for two out of four species in a mixed grass prairie in Montana.
+
+We also compared the treatment parameters from the treatment model fit to the experimental data to the treatment parameters predicted by the climate model fit only to the observational data \ref{fig:parPredictions}.
+Among all the climate effects we predicted and observed, there were only four cases where vital rate predictions and observations were both significantly different from zero (fig \ref{fig:parPredictions}) and three of these cases we successfully predicted the direction of the treatment effects.
+However, for \textit{P.
+secunda} recruitment we predicted a positive response of irrigation, but observed a negative response (figs \ref{fig:parPredictions}).
+From a statistical standpoint this is our arguably our greatest error in prediction.
+However, recruitment decreased in both the drought and irrigation plots for \textit{P.
+secunda} and also for \textit{P.
+spicata} (fig \ref{fig:recruitmentTreat}).
+So its likely the decrease in \textit{P.
+secunda} recruitment in the irrigated plots was due to underlying differences in the set of experimental plots from the historical control plots rather than the precipitation treatments.
+
+The drought effects we observed on the three grasses were often stronger than the effects we predicted, while the irrigation effects observed were often weaker than predicted (figs \ref{fig:parPredHECOGrowth}, \ref{fig:parPredHECOSurvival}, \ref{fig:parPredPOSEGrowth}, \ref{fig:parPredPOSESurvival}, \ref{fig:parPredPSSPGrowth}, \ref{fig:parPredPSSPSurvival}).
+In this water limited system, we expected that experimental irrigation would lead to increases in plant performance, but we saw few cases where irrigation benefited any of the plants.
+A pattern qualitatively similar to this shows up in both natural and experimental data comparing precipitation to ANPP: decreases in grassland ANPP induced by drought are often of greater magnitude than increases in ANPP induced by experimental irrigation or by above average precipitation (\citep{hsu_anticipating_2014,gherardi_enhanced_2015}).
+If we had fit our growth and survival models with a non-linear function for soil moisture, perhaps informed by more mechanistic understanding of water limitation on the physiology of these plants, we may have made more accurate predictions of the drought and irrigation effects \citep{ehrlen_advancing_2016}.
+
+\subsection*{Conclusion}
+
+Our results give us more confidence that observational data can be used to detect and predict the effects of annual soil moisture variation on sagebrush steppe plants.
+This should encourage more researchers to try and use observational data to predict population response to climate in both experimental and natural settings \citep{houlahan_priority_2016,ehrlen_advancing_2016}.
+Nevertheless, our success at predicting the short-term response of two out of four species to a simple precipitation manipulation is not likely to impress applied ecologists and policymakers who often need accurate predictions for the effects of climate change in large complex systems.
+Clearly more work is needed to learn how to accurately predict the ecological responses of species to climate change.
 Towards that goal, perhaps the best way forward is to conduct more tests like this one.

### 2017-01-11 removals/manuscript/removal_ms.tex
Working on final?? edits

@@
 tripartita} cover in shrub removal plots) is included to show values at the start of the experiment in 2011 and the efficacy of the removal treatments.
-Source file: \texttt{treatment\_trends\_removals.r}.}
+Source file: \texttt{treatment\textunderscore trends\textunderscore removals.r}.}
 \label{fig:CoverTrends}
@@
 \caption{Observed and predicted year-to-year changes in cover for the four dominant species.
-Mean observed changes (averaged across quadrats) are shown in solid symbols, mean predicted changes are shown in open symbols.
-Values for control and removal plots are blue and red, respectively.
-Predictions from the ``Baseline'' model do account for effects of local species composition on vital rates, but do not include additional effects of removals.
-Predictions from the ``Removal'' model include additional treatment effects on vital rates.
-Source file: \texttt{ibm\\summarize\_sims1step.r}.
+Mean observed changes (averaged across quadrats) are shown in black for control (solid symbols) and removal (open symbols) treatment.
+Predictions from the ``Baseline'' model (blue) account for effects of local species composition on vital rates, but do not include additional effects of removals.
+Predictions from the ``Removal'' model (red) include additional treatment effects on vital rates.
+Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r}.
 }
@@
 The removal effect shows response of \textit{A.
-tripartita} ("ARTR") to perennial grass removal, and response of the perennial grasses to \textit{A.
+tripartita} (``ARTR") to perennial grass removal, and response of the perennial grasses to \textit{A.
 tripartita} removal, after accounting for local neighborhood interactions.
-}
+ Source file: \texttt{treatment\textunderscore test\textunderscore figure.r}.}
  \label{fig:VitalRateTest}
@@
 For both  \textit{Poa secunda}  and \textit{P.
-spicata} , we found no sign of a decrease in the removal effect over time (Tables \ref{table:POSEgrowth-byYr} and \ref{table:PSSPgrowth-byYr}).
+spicata}, we found no sign of a decrease in the removal effect over time (Tables \ref{table:POSEgrowth-trtYears} and \ref{table:PSSPgrowth-trtYears}).
 
@@
  \caption{Observed and predicted cover for each of the four modeled species.
-Solid lines show mean observed cover averaged across control (blue) and removal treatment (red) quadrats.
-Dashed lines show one-step-ahead predictions from the baseline IBM (no removal treatment coefficients), averaged across quadrats.
-Dotted lines show one-step-ahead predictions from an IBM that includes removal treatment coefficients.
-Source file: \texttt{ibm\\summarize\_sims1step.r} }
+Solid black lines show mean observed cover averaged across control (closed symbols) and removal treatment (open symbols) quadrats.
+Blue lines and symbols show one-step-ahead predictions from the baseline IBM (no removal treatment coefficients), averaged across quadrats.
+Red lines and symbols show one-step-ahead predictions from an IBM that includes removal treatment coefficients.
+Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r} }
  \label{fig:IBM1step}
@@
 tripartita} set to zero and removal treatment coefficients included.
-}
+Source file: \texttt{ipm\textbackslash IPM-figures.r}}
  \label{fig:IPMresults}
@@
 \begin{table}
-\caption{\textit{Poa secunda} growth with \textit{Artemisia} canopy effect}
-\begin{center}
-\begin{tabular}{l c }
-\hline
- & Model 1 \\
-\hline
-(Intercept)                      & $0.49^{*}$        \\
-                                 & $[0.36;\ 0.62]$   \\
-logarea.t0                       & $0.68^{*}$        \\
-                                 & $[0.63;\ 0.72]$   \\
-TreatmentNo\_shrub               & $0.24^{*}$        \\
-                                 & $[0.09;\ 0.38]$   \\
-W.ARTR                           & $-0.00$           \\
-                                 & $[-0.00;\ 0.00]$  \\
-W.HECO                           & $0.00$            \\
-                                 & $[-0.01;\ 0.02]$  \\
-W.POSE                           & $-0.26^{*}$       \\
-                                 & $[-0.34;\ -0.19]$ \\
-W.PSSP                           & $-0.01$           \\
-                                 & $[-0.02;\ 0.00]$  \\
-W.allcov                         & $-0.00$           \\
-                                 & $[-0.00;\ 0.00]$  \\
-W.allpts                         & $-0.02$           \\
-                                 & $[-0.07;\ 0.02]$  \\
-inARTR                           & $-0.06$           \\
-                                 & $[-0.28;\ 0.16]$  \\
-\hline
-AIC                              & 15641.80          \\
-BIC                              & 15734.29          \\
-Log Likelihood                   & -7806.90          \\
-Num. obs.                        & 5469              \\
-Num. groups: year                & 31                \\
-Var: year (Intercept)            & 0.11              \\
-Var: year logarea.t0             & 0.01              \\
-Cov: year (Intercept) logarea.t0 & -0.02             \\
-Var: Residual                    & 0.98              \\
-\hline
-\multicolumn{2}{l}{\scriptsize{$^*$ 0 outside the confidence interval}}
-\end{tabular}
+\centering
+\caption{Summary of fixed effects for the \textit{Po. secunda} growth model with individual-level \textit{A. tripartita removal} data} 
 \label{table:POSEgrowth-inARTR}
-\end{center}
+\begin{tabular}{rrrrrrrr}
+  \hline
+ & mean & sd & 0.025quant & 0.5quant & 0.975quant & mode & kld \\ 
+  \hline
+(Intercept) & 0.4932 & 0.0622 & 0.3700 & 0.4934 & 0.6154 & 0.4937 & 0.0000 \\ 
+  logarea.t0 & 0.6746 & 0.0220 & 0.6306 & 0.6747 & 0.7177 & 0.6750 & 0.0000 \\ 
+  TreatmentNo\_shrub & 0.2370 & 0.0739 & 0.0919 & 0.2370 & 0.3820 & 0.2370 & 0.0000 \\ 
+  W.ARTR & -0.0002 & 0.0008 & -0.0018 & -0.0002 & 0.0014 & -0.0002 & 0.0000 \\ 
+  W.HECO & 0.0038 & 0.0062 & -0.0083 & 0.0038 & 0.0160 & 0.0038 & 0.0000 \\ 
+  W.POSE & -0.2644 & 0.0383 & -0.3395 & -0.2644 & -0.1893 & -0.2644 & 0.0000 \\ 
+  W.PSSP & -0.0075 & 0.0056 & -0.0184 & -0.0075 & 0.0034 & -0.0075 & 0.0000 \\ 
+  W.allcov & -0.0003 & 0.0012 & -0.0026 & -0.0003 & 0.0020 & -0.0003 & 0.0000 \\ 
+  W.allpts & -0.0258 & 0.0229 & -0.0707 & -0.0258 & 0.0191 & -0.0258 & 0.0000 \\ 
+  inARTR & -0.0592 & 0.1133 & -0.2817 & -0.0592 & 0.1632 & -0.0592 & 0.0000 \\ 
+   \hline
+\end{tabular}
 \end{table}
@@
 \begin{table}
-\caption{\textit{P. spicata} growth with \textit{Artemisia} canopy effect}
-\begin{center}
-\begin{tabular}{l c }
-\hline
- & Model 1 \\
-\hline
-(Intercept)                      & $0.42^{*}$        \\
-                                 & $[0.29;\ 0.56]$   \\
-logarea.t0                       & $0.82^{*}$        \\
-                                 & $[0.79;\ 0.86]$   \\
-TreatmentNo\_shrub               & $0.24^{*}$        \\
-                                 & $[0.09;\ 0.40]$   \\
-W.ARTR                           & $-0.00^{*}$       \\
-                                 & $[-0.00;\ -0.00]$ \\
-W.HECO                           & $-0.01^{*}$       \\
-                                 & $[-0.03;\ -0.00]$ \\
-W.POSE                           & $-0.01$           \\
-                                 & $[-0.03;\ 0.02]$  \\
-W.PSSP                           & $-0.14^{*}$       \\
-                                 & $[-0.16;\ -0.11]$ \\
-W.allcov                         & $-0.00^{*}$       \\
-                                 & $[-0.01;\ -0.00]$ \\
-W.allpts                         & $-0.06^{*}$       \\
-                                 & $[-0.10;\ -0.01]$ \\
-inARTR                           & $-0.14$           \\
-                                 & $[-0.35;\ 0.08]$  \\
-\hline
-AIC                              & 15282.37          \\
-BIC                              & 15376.25          \\
-Log Likelihood                   & -7627.18          \\
-Num. obs.                        & 6039              \\
-Num. groups: year                & 31                \\
-Var: year (Intercept)            & 0.12              \\
-Var: year logarea.t0             & 0.01              \\
-Cov: year (Intercept) logarea.t0 & -0.03             \\
-Var: Residual                    & 0.71              \\
-\hline
-\multicolumn{2}{l}{\scriptsize{$^*$ 0 outside the confidence interval}}
-\end{tabular}
+\centering
+\caption{Summary of fixed effects for the \textit{Ps. spicata} growth model with individual-level \textit{A. tripartita} removal data} 
 \label{table:PSSPgrowth-inARTR}
-\end{center}
+\begin{tabular}{rrrrrrrr}
+  \hline
+ & mean & sd & 0.025quant & 0.5quant & 0.975quant & mode & kld \\ 
+  \hline
+(Intercept) & 0.4129 & 0.0633 & 0.2885 & 0.4128 & 0.5381 & 0.4124 & 0.0000 \\ 
+  logarea.t0 & 0.8261 & 0.0173 & 0.7919 & 0.8262 & 0.8602 & 0.8262 & 0.0000 \\ 
+  TreatmentNo\_shrub & 0.2444 & 0.0798 & 0.0878 & 0.2444 & 0.4009 & 0.2444 & 0.0000 \\ 
+  W.ARTR & -0.0029 & 0.0008 & -0.0044 & -0.0029 & -0.0015 & -0.0029 & 0.0000 \\ 
+  W.HECO & -0.0138 & 0.0068 & -0.0272 & -0.0138 & -0.0005 & -0.0138 & 0.0000 \\ 
+  W.POSE & -0.0071 & 0.0127 & -0.0319 & -0.0071 & 0.0178 & -0.0071 & 0.0000 \\ 
+  W.PSSP & -0.1372 & 0.0126 & -0.1619 & -0.1372 & -0.1125 & -0.1372 & 0.0000 \\ 
+  W.allcov & -0.0039 & 0.0010 & -0.0059 & -0.0039 & -0.0018 & -0.0039 & 0.0000 \\ 
+  W.allpts & -0.0576 & 0.0224 & -0.1017 & -0.0576 & -0.0136 & -0.0576 & 0.0000 \\ 
+  inARTR & -0.1301 & 0.1102 & -0.3465 & -0.1301 & 0.0862 & -0.1301 & 0.0000 \\ 
+   \hline
+\end{tabular}
 \end{table}
@@
 \begin{table}
-\caption{\textit{Poa secunda} growth with year-by-treatment interaction}
-\begin{center}
-\begin{tabular}{l c }
-\hline
- & Model 1 \\
-\hline
-(Intercept)                      & $0.49^{*}$        \\
-                                 & $[0.37;\ 0.62]$   \\
-trtYears1                        & $0.06$            \\
-                                 & $[-0.17;\ 0.29]$  \\
-trtYears2                        & $0.24$            \\
-                                 & $[-0.01;\ 0.49]$  \\
-trtYears3                        & $0.23$            \\
-                                 & $[-0.05;\ 0.51]$  \\
-trtYears4                        & $0.25$            \\
-                                 & $[-0.07;\ 0.56]$  \\
-trtYears5                        & $0.45^{*}$        \\
-                                 & $[0.11;\ 0.79]$   \\
-logarea.t0                       & $0.68^{*}$        \\
-                                 & $[0.63;\ 0.72]$   \\
-W.ARTR                           & $-0.00$           \\
-                                 & $[-0.00;\ 0.00]$  \\
-W.HECO                           & $0.00$            \\
-                                 & $[-0.01;\ 0.02]$  \\
-W.POSE                           & $-0.26^{*}$       \\
-                                 & $[-0.34;\ -0.19]$ \\
-W.PSSP                           & $-0.01$           \\
-                                 & $[-0.02;\ 0.00]$  \\
-W.allcov                         & $-0.00$           \\
-                                 & $[-0.00;\ 0.00]$  \\
-W.allpts                         & $-0.03$           \\
-                                 & $[-0.07;\ 0.02]$  \\
-\hline
-AIC                              & 15648.62          \\
-BIC                              & 15760.94          \\
-Log Likelihood                   & -7807.31          \\
-Num. obs.                        & 5469              \\
-Num. groups: year                & 31                \\
-Var: year (Intercept)            & 0.11              \\
-Var: year logarea.t0             & 0.01              \\
-Cov: year (Intercept) logarea.t0 & -0.02             \\
-Var: Residual                    & 0.98              \\
-\hline
-\multicolumn{2}{l}{\scriptsize{$^*$ 0 outside the confidence interval}}
-\end{tabular}
-\label{table:POSEgrowth-byYr}
-\end{center}
+\centering
+\caption{Summary of fixed effects for the \textit{Po. secunda} model with treatment*year effects} 
+\label{table:POSEgrowth-trtYears}
+\begin{tabular}{rrrrrrrr}
+  \hline
+ & mean & sd & 0.025quant & 0.5quant & 0.975quant & mode & kld \\ 
+  \hline
+(Intercept) & 0.4963 & 0.0616 & 0.3752 & 0.4963 & 0.6172 & 0.4964 & 0.0000 \\ 
+  trtYears1 & 0.0798 & 0.1272 & -0.1699 & 0.0798 & 0.3293 & 0.0798 & 0.0000 \\ 
+  trtYears2 & 0.2425 & 0.1350 & -0.0225 & 0.2425 & 0.5073 & 0.2425 & 0.0000 \\ 
+  trtYears3 & 0.2512 & 0.1554 & -0.0539 & 0.2512 & 0.5561 & 0.2512 & 0.0000 \\ 
+  trtYears4 & 0.2988 & 0.1693 & -0.0337 & 0.2988 & 0.6310 & 0.2988 & 0.0000 \\ 
+  trtYears5 & 0.4787 & 0.1845 & 0.1164 & 0.4787 & 0.8407 & 0.4787 & 0.0000 \\ 
+  logarea.t0 & 0.6731 & 0.0234 & 0.6270 & 0.6732 & 0.7191 & 0.6732 & 0.0000 \\ 
+  W.ARTR & -0.0003 & 0.0009 & -0.0019 & -0.0003 & 0.0014 & -0.0003 & 0.0000 \\ 
+  W.HECO & 0.0035 & 0.0064 & -0.0090 & 0.0035 & 0.0160 & 0.0035 & 0.0000 \\ 
+  W.POSE & -0.2624 & 0.0394 & -0.3398 & -0.2624 & -0.1851 & -0.2624 & 0.0000 \\ 
+  W.PSSP & -0.0080 & 0.0057 & -0.0192 & -0.0080 & 0.0032 & -0.0080 & 0.0000 \\ 
+  W.allcov & -0.0004 & 0.0012 & -0.0027 & -0.0004 & 0.0020 & -0.0004 & 0.0000 \\ 
+  W.allpts & -0.0267 & 0.0236 & -0.0730 & -0.0267 & 0.0196 & -0.0267 & 0.0000 \\ 
+  inARTR & -0.0563 & 0.1169 & -0.2859 & -0.0563 & 0.1730 & -0.0563 & 0.0000 \\ 
+   \hline
+\end{tabular}
 \end{table}
@@
 \begin{table}
-\caption{\textit{P. spicata} growth with year-by-treatment interaction}
-\begin{center}
-\begin{tabular}{l c }
-\hline
- & Model 1 \\
-\hline
-(Intercept)                      & $0.42^{*}$        \\
-                                 & $[0.28;\ 0.55]$   \\
-trtYears1                        & $0.26$            \\
-                                 & $[-0.01;\ 0.54]$  \\
-trtYears2                        & $-0.05$           \\
-                                 & $[-0.32;\ 0.22]$  \\
-trtYears3                        & $0.44^{*}$        \\
-                                 & $[0.14;\ 0.73]$   \\
-trtYears4                        & $0.42^{*}$        \\
-                                 & $[0.12;\ 0.72]$   \\
-trtYears5                        & $-0.11$           \\
-                                 & $[-0.42;\ 0.20]$  \\
-logarea.t0                       & $0.82^{*}$        \\
-                                 & $[0.79;\ 0.86]$   \\
-W.ARTR                           & $-0.00^{*}$       \\
-                                 & $[-0.00;\ -0.00]$ \\
-W.HECO                           & $-0.01^{*}$       \\
-                                 & $[-0.03;\ -0.00]$ \\
-W.POSE                           & $-0.01$           \\
-                                 & $[-0.03;\ 0.02]$  \\
-W.PSSP                           & $-0.14^{*}$       \\
-                                 & $[-0.16;\ -0.11]$ \\
-W.allcov                         & $-0.00^{*}$       \\
-                                 & $[-0.01;\ -0.00]$ \\
-W.allpts                         & $-0.06^{*}$       \\
-                                 & $[-0.10;\ -0.01]$ \\
-\hline
-AIC                              & 15281.67          \\
-BIC                              & 15395.67          \\
-Log Likelihood                   & -7623.83          \\
-Num. obs.                        & 6039              \\
-Num. groups: year                & 31                \\
-Var: year (Intercept)            & 0.12              \\
-Var: year logarea.t0             & 0.01              \\
-Cov: year (Intercept) logarea.t0 & -0.03             \\
-Var: Residual                    & 0.71              \\
-\hline
-\multicolumn{2}{l}{\scriptsize{$^*$ 0 outside the confidence interval}}
-\end{tabular}
-\label{table:PSSPgrowth-byYr}
-\end{center}
+\centering
+\caption{Summary of fixed effects for the \textit{Ps. spicata} model with treatment*year effects} 
+\label{table:PSSPgrowth-trtYears}
+\begin{tabular}{rrrrrrrr}
+  \hline
+ & mean & sd & 0.025quant & 0.5quant & 0.975quant & mode & kld \\ 
+  \hline
+(Intercept) & 0.4122 & 0.0626 & 0.2892 & 0.4120 & 0.5364 & 0.4115 & 0.0000 \\ 
+  trtYears1 & 0.3256 & 0.1440 & 0.0428 & 0.3256 & 0.6081 & 0.3257 & 0.0000 \\ 
+  trtYears2 & -0.0353 & 0.1467 & -0.3235 & -0.0353 & 0.2525 & -0.0353 & 0.0000 \\ 
+  trtYears3 & 0.4829 & 0.1586 & 0.1714 & 0.4829 & 0.7940 & 0.4829 & 0.0000 \\ 
+  trtYears4 & 0.5059 & 0.1602 & 0.1915 & 0.5059 & 0.8202 & 0.5059 & 0.0000 \\ 
+  trtYears5 & -0.0588 & 0.1682 & -0.3890 & -0.0588 & 0.2712 & -0.0588 & 0.0000 \\ 
+  logarea.t0 & 0.8266 & 0.0173 & 0.7922 & 0.8267 & 0.8607 & 0.8268 & 0.0000 \\ 
+  W.ARTR & -0.0030 & 0.0008 & -0.0044 & -0.0030 & -0.0015 & -0.0030 & 0.0000 \\ 
+  W.HECO & -0.0139 & 0.0068 & -0.0272 & -0.0139 & -0.0005 & -0.0139 & 0.0000 \\ 
+  W.POSE & -0.0064 & 0.0127 & -0.0313 & -0.0064 & 0.0185 & -0.0064 & 0.0000 \\ 
+  W.PSSP & -0.1362 & 0.0126 & -0.1609 & -0.1362 & -0.1115 & -0.1362 & 0.0000 \\ 
+  W.allcov & -0.0040 & 0.0010 & -0.0061 & -0.0040 & -0.0020 & -0.0040 & 0.0000 \\ 
+  W.allpts & -0.0579 & 0.0224 & -0.1020 & -0.0579 & -0.0139 & -0.0579 & 0.0000 \\ 
+  inARTR & -0.1235 & 0.1104 & -0.3403 & -0.1235 & 0.0931 & -0.1235 & 0.0000 \\ 
+   \hline
+\end{tabular}
 \end{table}
@@
  \includegraphics[width=0.7\textwidth]{CompKernels}
- \caption{Competition kernels for the four dominant species.  }
+ \caption{Competition kernels for the four dominant species. Source file: \texttt{Wdistrib\textbackslash exploreSurvivalWs.r} }
  \label{fig:CompKernels}
@@
   \caption{Annual precipitation (a) and mean temperature (b) during the period of the experiment, shown against the long-term (1927-2016) means (solid blue and red lines) and 5\% and 95\% quantiles (dashed lines).
-}
+Source file:  \texttt{climate\textunderscore fig.r}}
   \label{fig:climate}
@@
 Blue symbols show values from controls plots, red symbols show values from \textit{Artemisia} removal plots.``W.allcov'' refers to the aggregated crowding by all shrubs and perennial grasses beyond the focal dominant species, and ``W.allpts'' refers to the aggregated crowding by forb species.
-}
+\texttt{Wdistrib\textbackslash exploreSurvivalWs.r} }
   \label{fig:Wscatters}
@@
 ``W.allcov'' refers to the aggregated crowding by all shrubs and perennial grasses beyond the focal dominant species, and ``W.allpts'' refers to the aggregated crowding by forb species.
-}
+\texttt{Wdistrib\textbackslash exploreSurvivalWs.r} }
  \label{fig:W-by-treatment}
@@
  \caption{Observed and predicted cover for each of the four modeled species assuming maximum effects of removal. 
-Solid lines show mean observed cover averaged across control (blue) and removal treatment (red) quadrats.
-Dashed lines show one-step-ahead predictions from the baseline IBM (no removal treatment coefficients), averaged across quadrats.
-Dotted lines show one-step-ahead predictions from an IBM that includes removal treatment coefficients representing the outermost values of the 95\% confidence intervals.
-Source file: \texttt{ibm\\summarize\_sims1step.r} }
+  Solid black lines show mean observed cover averaged across control (closed symbols) and removal treatment (open symbols) quadrats.
+Blue lines and symbols show one-step-ahead predictions from the baseline IBM (no removal treatment coefficients), averaged across quadrats.
+Red lines and symbols show one-step-ahead predictions from an IBM that includes removal treatment coefficients.
+Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r} }
  \label{fig:IBM1step-maxCI}
@@
 tripartita} set to zero and (mean) removal treatment coefficients included, and blue boxes show results with removal treatments coefficients assigned the outermost value of the 95\% confidence intervals.
-}
+Source file: \texttt{ipm\textbackslash IPM-figures.r}}
   \label{fig:IPMresults-maxCI}

### 2017-01-12 removals/manuscript/removal_ms.tex
Added new 1:1 figure

@@
 \item A previous study of four dominant species from a sagebrush steppe community used historical data sets and multispecies population models to demonstrate weak interspecific interactions, implying very little competitive release.
-\item We conducted a removal experiment to test the model's ability to predict competitive release.
+\item We conducted a removal experiment to test the model's prediction of little competitive release.
 We established new quadrats in the same location where the historical data were collected.
@@
 
-An alternative approach to studying competitive release is to fit multispecies population models to long-term, observational data, and then simulate population responses to removals.
+An alternative approach to studying competitive release is to fit multispecies population models that estimate species interactions using long-term, observational data, and then simulate population responses to removals.
 Applying this approach to arid and semiarid plant communities, we found evidence of strong intraspecific limitation and very weak interspecific interactions, leading to very stable coexistence and implying little competitive release \citep{adler_coexistence_2010,chu_large_2015}.
-Since we only studied the most common, co-occurring species, we expected to find evidence of stable coexistence, but we were surprised that interspecific interactions and species' responses to the loss of a dominant competitor were so weak.
+Since we only studied the most common, co-occurring species, we expected to find evidence of stable coexistence, but we were surprised that interspecific interactions and population responses to the loss of dominant species were so weak.
 Compared to traditional field experiments, our approach has the advantage of projecting full life-cycle responses over long time periods, but its reliance on observational data sets limits the strength of our inference.
@@
 Because our demographic models include the effects of neighborhood crowding, they should be able to predict how experimental removal of neighbors alters demographic performance.
-If the full effect of experimental removals is stronger than predicted, it means that our models underestimate competitive release.
+If the full effect of experimental removals is stronger than predicted, it means that our statistical models underestimate competitive release.
 Our second question integrates across the full life cycle: how do short and long-term population-level responses to removals compare to projections from multispecies population models based on observational data?
@@
 
-\begin{figure}[tbp]
-\centering
-\includegraphics[width=1\textwidth]{cover_change_chrono}
-\caption{Observed and predicted year-to-year changes in cover for the four dominant species.
-Mean observed changes (averaged across quadrats) are shown in black for control (solid symbols) and removal (open symbols) treatment.
-Predictions from the ``Baseline'' model (blue) account for effects of local species composition on vital rates, but do not include additional effects of removals.
-Predictions from the ``Removal'' model (red) include additional treatment effects on vital rates.
-Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r}.
-}
-\label{fig:CoverChange}
-\end{figure}
-
 \subsection*{Vital rates}
@@
 \ref{fig:IBM1step-maxCI}).
+Across species, our baseline model tended to underpredict annual growth rates in control plots, but the correlation between observations and predictions was strong (Fig.
+\ref{fig:ObsPred1to1}A).
+In removal plots, we see greater underprediction of growth rates in certain years, especially for \textit{H.
+comata} and \textit{P.
+spicata}, but including removal treatment effects only slightly reduces this bias (Fig.
+\ref{fig:ObsPred1to1}B).
 
@@
  \label{fig:IBM1step}
+ \end{figure}
+
+  \begin{figure}[tbp]
+  \centering
+  \includegraphics[width=1\textwidth]{cover_change_1to1}
+  \caption{Comparison of observed and predicted annual per capita growth rates across species in (A) control and (B) removal treatments.
+The black line is a 1:1 line.
+Solid symbols correspond to predictions form the baseline IBM (no removal effects) and open symbols in (B) show predictions from an IBM that includes removal treatment effects.
+Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r} }
+  \label{fig:ObsPred1to1}
   \end{figure}
@@
   \centering
+  \includegraphics[width=1\textwidth]{cover_change_chrono}
+  \caption{Observed and predicted year-to-year changes in cover for the four dominant species.
+Mean observed changes (averaged across quadrats) are shown in black for control (solid symbols) and removal (open symbols) treatment.
+Predictions from the ``Baseline'' model (blue) account for effects of local species composition on vital rates, but do not include additional effects of removals.
+Predictions from the ``Removal'' model (red) include additional treatment effects on vital rates.
+Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r}.
+}
+  \label{fig:CoverChange}
+  \end{figure}
+
+  \begin{figure}[tbp]
+  \centering
   \includegraphics[width=1\textwidth]{PSSP_W_scatters}

### 2017-01-12 removals/manuscript/removal_ms.tex
More minor edits to figures

@@
 For \textit{Poa secunda} and especially \textit{P.
-spicata}, the model with removal effects consistently predicted higher population growth than the baseline model (Fig.
-\ref{fig:CoverChange}), and reduced the underestimates of cover (Fig.
+spicata}, the model with removal effects consistently predicted higher population growth than the baseline model  and reduced the underestimates of cover (Fig.
 \ref{fig:IBM1step}).
-The version of the model with maximum removal treatment effects (removal coefficients set to the 95\% confidence limit furthest from zero), were qualitatively similar, but further amplified the contrast between predictions of the baseline and removal models for and \textit{P.
-spicata} (Fig.
-\ref{fig:IBM1step-maxCI}).
-Across species, our baseline model tended to underpredict annual growth rates in control plots, but the correlation between observations and predictions was strong (Fig.
+Viewing all annual growth rates across species and years shows the same pattern of underprediction for some species, but also reveals a strong correlation between observations and predictions in control plots (Fig.
 \ref{fig:ObsPred1to1}A).
-In removal plots, we see greater underprediction of growth rates in certain years, especially for \textit{H.
+In removal plots, underprediction of growth rates in certain years was larger, especially for \textit{H.
 comata} and \textit{P.
-spicata}, but including removal treatment effects only slightly reduces this bias (Fig.
+spicata}; including removal treatment effects only slightly reduced this bias (Fig.
 \ref{fig:ObsPred1to1}B).
+For the version of the model with maximum removal treatment effects (removal coefficients set to the 95\% confidence limit furthest from zero), the patterns were qualitatively similar, with a somewhat greater contrast between predictions of the baseline and removal models for \textit{P.
+secunda} and \textit{P.
+spicata} (Figs.
+\ref{fig:IBM1step-maxCI},\ref{fig:ObsPred1to1-maxCI}).
 
@@
    \centering
+   \includegraphics[width=1\textwidth]{cover_change_1to1_maxCI}
+   \caption{Comparison of observed and predicted annual per capita growth rates across species in (A) control and (B) removal treatments assuming maximum effects of removals.
+The black line is a 1:1 line.
+Solid symbols correspond to predictions form the baseline IBM (no removal effects) and open symbols in (B) show predictions from an IBM that includes removal treatment effects.
+Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r} }
+   \label{fig:ObsPred1to1-maxCI}
+   \end{figure}
+
+  \begin{figure}[tbp]
+  \centering
   \includegraphics[width=1\textwidth]{boxplots-maxCI}

### 2017-01-12 removals/manuscript/removal_ms.tex
More small edits

@@
 
-\title{Weak interspecific interactions in a sagebrush steppe: Evidence from observational data, population models, and a removal experiment}
+\title{Weak interspecific interactions in a sagebrush steppe: evidence from observations, models, and experiments}
 
@@
 \begin{enumerate}
-\item A previous study of four dominant species from a sagebrush steppe community used historical data sets and multispecies population models to demonstrate weak interspecific interactions, implying very little competitive release.
-\item We conducted a removal experiment to test the model's prediction of little competitive release.
+\item A previous study of four dominant species from a sagebrush steppe community used historical data sets and multispecies population models to demonstrate weak interspecific interactions.
+The models predict little competitive release following species removals.
+\item We conducted a removal experiment to test this prediction.
 We established new quadrats in the same location where the historical data were collected.
@@
 However, including information about the location of these grass individuals with respect to removed \textit{Artemisia} failed to improve the growth regressions, raising questions about the mechanisms driving the positive response to removal.
-\item For three of the four species, individual based models and integral projections models based on the vital rate regressions showed that removal treatment effects had little impact on population growth rates or simulated equilibrium cover following a species removal.
+\item For three species, individual based models and integral projections models based on the vital rate regressions showed that removal treatment effects had little impact on population growth rates or simulated equilibrium cover following a species removal.
 For \textit{P.
 spicata}, the population models that included effects of \textit{Artemisia} removal projected greater competitive release than the baseline models.
-\item \emph{Synthesis}: Although models based on observational data did not perfectly predict species responses to removal, overall the results increase our confidence that competitive release is in fact small for at least three of our four study species.
+\item \emph{Synthesis}: Although models based on observational data did not perfectly predict species responses to removal, the experimental results increase our confidence that competitive release is in fact small for at least three of our four study species.
 The implication is that interspecific resource competition is not the primary factor determining the dynamics of this community.
@@
 However, we still might expect interspecific interactions to be strong and negative.
-In fact, interspecific competition is arguably the primary obsession of plant community ecologists.
-We motivate research on plant species coexistence by noting that all plant species compete for water, light, and a few mineral resources (e.g.
+In fact, plant community ecologists often motivate research on coexistence by noting that all plant species compete for water, light, and a few mineral resources (e.g.
 \citealt{silvertown_plant_2004}).
@@
 Neutral theory \citep{hubbell_unified_2001} goes even further, assuming that interspecific competition is exactly as strong as intraspecific competition.
-Recent work has emphasized the role of interspecific facilitation in some plant communities \citep{he_global_2013,brooker_facilitation_2008}, but it is unclear whether this research has shaken the default assumption that plant species compete intensely.
+A growing list of studies has emphasized the role of interspecific facilitation in some plant communities \citep{he_global_2013,brooker_facilitation_2008}, but it is unclear whether this research has overturned the default assumption that plants compete intensely with other species.
 
@@
 Sheep Experiment Station (USSES) is located 9.6 km north of Dubois, Idaho (44.2\degree N, 112.1\degree W), 1500 m above sea level.
-During the period of data collection (1926 – 1957), mean annual precipitation was 270 mm and mean temperatures ranged from -8\degree C (January) to 21\degree C (July).
+During the period of data collection (1926 \textendash 1957), mean annual precipitation was 270 mm and mean temperatures ranged from -8\degree C (January) to 21\degree C (July).
 The vegetation is dominated by the shrub, \textit{Artemisia tripartita}, and the C3  perennial bunchgrasses \textit{Pseudoroegneria spicata}, \textit{Hesperostipa comata}, and \textit{Poa secunda}.
@@
 \subsection*{Removal experiment}
-In spring 2011, in a large exclosure containing six of the historical permanent quadrats, we selected locations for an additional 16 quadrats for the removal treatments.
-We used the following criteria to select the quadrat locations.
-Each plot had to have 5\% canopy cover of \textit{Artemisia tripartita} and at least 3\% combined basal cover of the three focal perennial grasses.
-We avoided plots falling on hill slopes, areas with greater than 20\% bare rock, or with over 10\% cover of the woody shrubs \textit{Purshia tridentata} or \textit{Amelanchier utahensis}.
-We mapped the quadrats in June, 2011 and then implemented the removal treatments in September, 2011.
+In spring 2011, in a large exclosure containing six of the historical permanent quadrats, we established an additional 16 quadrats for the removal treatments.
+New quadrats were selected to be similar to historical plots, but also needed to have enough of the focal species present so that removal experiments could be conducted.
+We evaluated potential locations for the new quadrats near the original quadrats and then used only those locations with at least 5\% canopy cover of \textit{Artemisia tripartita} and at least 3\% combined basal cover of the three focal perennial grasses.
+We rejected locations falling on hill slopes, areas with greater than 20\% bare rock, or with over 10\% cover of the woody shrubs \textit{Purshia tridentata} or \textit{Amelanchier utahensis}.
+We mapped the new quadrats in June, 2011 and then implemented the removal treatments in September, 2011.
 We randomly assigned 8 of the sixteen new quadrats to the sagebrush removal treatment, and the remaining 8 to the grass removal treatment.
@@
 \begin{equation}
-\mbox{logit}(S_{ijgh,t}) = \gamma_{j,t}^S + \varphi_{jg}^S+  \chi_{jh}^S  + \beta_{j,t}^S u_{ij,t} +  
+\mbox{logit}(S_{ijgh,t+1}) = \gamma_{j,t}^S + \varphi_{jg}^S+  \chi_{jh}^S  + \beta_{j,t}^S u_{ij,t} +  
 \left \langle \boldsymbol{\omega}_{j,t}^S, \boldsymbol{W}_{ij,t} \right \rangle 
@@
 Expected cover at $t+1$ is expected area of new recruits, plus the sum of the survival probabilities multiplied by the expected sizes.
+Because we do not carry these simulations forward beyond one time step, we do not need to assign locations to the new recruits.
 
@@
 
-Our vital rate models take into account the effects of local species composition and should therefore be able to predict the effect of removals on survival, growth and recruitment without any additional parameters.
+Our vital rate models take into account the effects of local species composition and should therefore be able to explain the effect of removals on survival, growth and recruitment without any additional parameters.
 If the removal treatment effects that we added to these models are statistically significant, it would indicate an effect of removals that goes above and beyond the response predicted by a model based on data from the control plots.
@@
 \ref{fig:VitalRateTest}; Table \ref{table:recruitment}).
+For the other species, we could not detect significant removal treatment effects on recruitment.
 
@@
 The positive removal effects on \textit{Poa secunda}  and \textit{P.
-spicata} growth are particularly interesting, because they suggest that models based on observational data may underestimate competitive release.
+spicata} growth suggest that models based on observational data may underestimate competitive release.
 We conducted additional analyses to further explore removal effects on the growth of these two species.
@@
 
-Third, responses to removals might be poorly predicted by models based on observational data if the removals create local crowding outside the range of historical variation or if the removals break covariances in crowding among species.
+Third, responses to removals might be poorly predicted by models based on observational data if the removals reduce plant densities (crowding) below the range of historical variation, or if the removals break covariances in plant densities among species.
 However, the distribution of crowding observed within control plots easily spans the crowding values observed in removal plots (Fig.
@@
 For three of the four species, we find little or no evidence that models based on observational data underestimate the intensity of interspecific competition.
-Removal treatment effects were never significant for \textit{A.
+After accounting for local neighborhood crowding, the removal treatment effects were never significant for \textit{A.
 tripartita} or \textit{H.
@@
 \citealt{Barbier2008}).
-\textit{Artemisia} removal might offest this facilitative effect by reducing resource uptake, resulting in little net change in water balance following removal.
+\textit{Artemisia} removal might offset this facilitative effect by reducing resource uptake, resulting in little net change in water balance following removal.
 Unfortunately, we did not monitor soil moisture at the fine spatial and temporal resolution necessary to test this hypothesis.
@@
 Similarly, differences among species in relative abundance should reflect variation in the strength of intraspecific density dependence more than variation in the strength of interspecific interactions \citep{yenni_strong_2012}.
-In such a system, effective management and conservation requires understanding the sources and species-specificity of intraspecific density dependence.
+A recent global, grassland study showing little variation in the abundance of invasive species in native and invaded ranges on different continents  \citep{firn_abundance_2011} is consistent with the idea that intraspecific interactions are the primary determinant of species abundances.
+In systems where this is true, effective management and conservation requires understanding the sources and species-specificity of intraspecific density dependence.
 In fact, an experiment to test our model's skill at predicting release from intraspecific competition could be more important than the test of interspecific competitive release described here.
@@
 
-\newpage
+%\newpage
 \bibliographystyle{Ecology}

### 2017-01-13 removals/manuscript/removal_ms.tex
Moved data into ExperimentTest repository, changed paths in scripts accordingly

@@
 
-Funding was provided by NSF grants DEB-1353078,and  DEB-1054040 to PBA and DEB 1353039 to SPE, and by the Utah Agricultural Experiment Station (get journal paper number).
+Funding was provided by NSF grants DEB-1353078,and  DEB-1054040 to PBA and DEB-1353039 to SPE, and by the Utah Agricultural Experiment Station (get journal paper number).
 The USDA-ARS Sheep Experiment Station generously provided access to historical data and the field experiment site.

### 2017-01-13 removals/survival/PSSPsurvival.R

@@ -8,8 +8,8 @@
 
 doSpp <- "PSSP"
 sppList <- c("ARTR","HECO","POSE","PSSP","allcov","allpts")
-dataDir1 <- paste(root,"/driversdata/data/idaho",sep="")
-dataDir2 <- paste(root,"/driversdata/data/idaho_modern",sep="")
+dataDir1 <- paste(root,"/ExperimentTests/data/idaho",sep="")
+dataDir2 <- paste(root,"/ExperimentTests/data/idaho_modern",sep="")
 nonCompLength.s=5 #Number of columns in SppData that are not measures of competitors 
 
 # set up distance weights------------------------------------------------

### 2017-01-13 removals/manuscript/removal_ms.tex
Minor edits

@@
 We modeled survival, growth and recruitment as functions of local neighborhood species composition along with an indicator variable for removal treatment.
-If our ``baseline" model, which accounts for local plant-plant interactions, accurately predicts responses to removal, then the fitted removal indicator effect should be non-signficant.
+If our ``baseline" model, which accounts for local plant-plant interactions, explains responses to removal, then the fitted removal indicator effect should be non-signficant.
 \item  For survival and recruitment, the removal treatment effects were never significantly positive, indicating that competitive release was not underestimated by the interspecific effects included in the baseline model.
@@
 \item For growth, the removal treatment effect was significant and positive for two species, \textit{Poa secunda} and \textit{Pseudoroegneria spicata}, indicating that the baseline model underestimated competitive release.
-However, including information about the location of these grass individuals with respect to removed \textit{Artemisia} failed to improve the growth regressions, raising questions about the mechanisms driving the positive response to removal.
+However, including information about the location of these grass individuals with respect to removed \textit{A.
+tripartita} failed to improve the growth regressions, raising questions about the mechanisms driving the positive response to removal.
 \item For three species, individual based models and integral projections models based on the vital rate regressions showed that removal treatment effects had little impact on population growth rates or simulated equilibrium cover following a species removal.
 For \textit{P.
-spicata}, the population models that included effects of \textit{Artemisia} removal projected greater competitive release than the baseline models.
+spicata}, the population models that included effects of \textit{A.
+tripartita} removal projected greater competitive release than the baseline models.
 \item \emph{Synthesis}: Although models based on observational data did not perfectly predict species responses to removal, the experimental results increase our confidence that competitive release is in fact small for at least three of our four study species.
@@
 Our second question integrates across the full life cycle: how do short and long-term population-level responses to removals compare to projections from multispecies population models based on observational data?
-If models based on observational data effectively predict competitive release, the population dynamics they project should be similar to projections from models that include information from the experimental removal treatments.
+If models based on observational data adequately predict competitive release, the population dynamics they project should be similar to projections from models that include information from the experimental removal treatments.
 
@@
 During the period of data collection (1926 \textendash 1957), mean annual precipitation was 270 mm and mean temperatures ranged from -8\degree C (January) to 21\degree C (July).
-The vegetation is dominated by the shrub, \textit{Artemisia tripartita}, and the C3  perennial bunchgrasses \textit{Pseudoroegneria spicata}, \textit{Hesperostipa comata}, and \textit{Poa secunda}.
+The vegetation is dominated by the shrub, \textit{Artemisia tripartita}, and the C3  perennial bunchgrasses \textit{Hesperostipa comata}, \textit{Pseudoroegneria spicata},  and \textit{Poa secunda}.
 These four species, the focus of our models, accounted for over 70\% of basal cover (grasses) and 60\% of canopy cover (shrubs and forbs).
@@
 All data and R code necessary to reproduce our analysis will be deposited in the Dryad Digital Repository once the manuscript is accepted.
-The current version of the computer code is available at https://github.com/pbadler/ExperimentTests/tree/master/removals and the data are available at https://bitbucket.org/ellner/driversdata.
+The current version of the computer code and all necessary data are available at \texttt{https://github.com/pbadler/ExperimentTests.git}.
 
@@
 The response variable for this analysis was year-to-year change in cover at the quadrat scale, calculated as $log(Cover_{t+1}/Cover_t)$.
-We tested for the effect of removal treatment on this measure of cover change with a linear mixed effects model implemented in the \textit{lme4} package \citep{Bates2015} of \textit{R}, including year and quadrat as random effects.
+We tested for the effect of removal treatment on this measure of cover change with a linear mixed effects model implemented in the \texttt{lme4} package \citep{Bates2015} of R, including year and quadrat as random effects.
 We considered the removal treatment effect to be significant when the 95\% confidence interval on the estimated coefficient did not overlap zero.
@@
 For example, we might find that, holding crowding constant, grasses grow more quickly in the removal plots than in the control plots.
-This result would indicate that our "baseline model," a model including all estimated parameters except for the removal treatment effect, underestimates competitive release.
-We refer to the model that includes the estimated removal treatment effect as the "removal model."
+This result would indicate that our ``baseline model," a model including all estimated parameters except for the removal treatment effect, underestimates competitive release.
+We refer to the model that includes the estimated removal treatment effect as the ``removal model."
 
@@
 The vector $\boldsymbol{W}$ includes crowding from the four dominant species,  \textit{A.
-tripartita}, \textit{P.
-spicata}, \textit{H.
-comata}, and \textit{Poa secunda}, as well as two covariates representing total crowding from (1) all other perennial grasses and shrubs, which were mapped as polygons, and (2) all forb species, which were mapped as points.
+tripartita}, \textit{H.
+comata}, \textit{P.
+spicata}, and \textit{P.
+secunda}, as well as two covariates representing total crowding from (1) all other perennial grasses and shrubs, which were mapped as polygons, and (2) all forb species, which were mapped as points.
 $\left \langle \boldsymbol{x, y} \right \rangle$ denotes the inner product of vectors $\boldsymbol{x}$ and $\boldsymbol{y}$, 
@@
 All genets in a quadrat were included in calculating $W$, but plants located within 5 cm of quadrat edges were not used in fitting. 
-We fit the models using the \textit{R-INLA} package for R \citep{rue_approximate_2009}.
-To explore variations on the growth regressions, such as the addition of a year-by-treatment interaction or information about the locations of individual grasses relative to removed \textit{Artemisia} shrubs, we fit models using the \textit{lmer} function in the \textit{lme4} package for R
+We fit the models using the \texttt{R-INLA} package for R \citep{rue_approximate_2009}.
+We used the same approach to explore variations on the growth regressions, such as the addition of a year-by-treatment interaction or information about the locations of individual grasses relative to removed \textit{A.
+tripartita}.
+We also compared the species composition of the control and removal plots at both the quadrat and neighborhood ($\boldsymbol{W}$) scales.
 
@@
 Convergence was observed graphically for all parameters, and confirmed by the Brooks-Gelman statistic \citep{brooks_general_1998}.
-
-For all three vital rate models, we considered the effect of removal treatments to be significant when the 95\% confidence interval (or posterior
-interval) on the estimate of $\chi$ did not overlap zero.
-In an effort to further understand the removal treatment effects, we explored additional variations on the growth model with a linear random effects model.
-For hypothesis testing with these models, we focused on the 95\% confidence interval of the coefficient of interest.
-We also compared the species composition of the control and removal plots at both the quadrat and neighborhood ($\boldsymbol{W}$) scales.
+For all three vital rate models, we considered the effect of removal treatments, or other coefficients of interest, to be significant when the 95\% confidence interval (or posterior interval) on the estimate of $\chi$ did not overlap zero.
 
@@
 were segregated from each other. The overdispersion of large conspecific genets was introduced into the IPM 
-with a 'no-overlap' rule \citep{adler_coexistence_2010}.
+with a `no-overlap' rule \citep{adler_coexistence_2010}.
 In \ref{suppMethods} we explain how the spline competition kernels with the no-overlap rule kernels were used to compute neigbhorhood crowding index as a function of individual size for the IPM.
@@
 If our baseline model successfully predicts competitive release observed in the removal experiment, then the results of the second and third scenarios should be similar.
-However, to the extent that the baseline model underestimates (or overestimates) the actual competitive release, the results of the second and third scenarios will differ.
+To the extent that the baseline model underestimates (or overestimates) the actual competitive release, the results of the second and third scenarios will differ.
 As with the IBM, we repeated these simulations using maximum removal treatment effects represented by parameters values at the 5\% or 95\% confidence limit farthest from zero.
@@
 \textit{A. tripartita} (95\% confidence interval on removal treatment effect: -0.75, 0.56) or 
-\textit{Poa secunda} (95\% CI: -0.41, 0.25) population growth, but did increase 
+\textit{P. secunda} (95\% CI: -0.41, 0.25) population growth, but did increase 
 \textit{H. comata} (95\% CI: 0.01, 0.65) and
@@
 \ref{fig:VitalRateTest}; Tables \ref{ARTRsurvival} to \ref{PSSPsurvival}), meaning that the survival probabilities of plants in removal plots can be predicted based on data from control plots.
-However, for the growth models, the 95\% confidence intervals on the removal effects were positive for \textit{Poa secunda}  and \textit{P.
+For the growth models, the 95\% confidence intervals on the removal effects were positive for \textit{P.
+secunda}  and \textit{P.
 spicata}, meaning that individuals of these species grew more rapidly in removal plots than controls even after accounting for neighborhood interactions (Fig.
@@
 
-The positive removal effects on \textit{Poa secunda}  and \textit{P.
+The positive removal effects on \textit{P.
+secunda}  and \textit{P.
 spicata} growth suggest that models based on observational data may underestimate competitive release.
@@
 We tested for a transient removal effect by adding a year-by-treatment interaction.
-For both  \textit{Poa secunda}  and \textit{P.
+For both  \textit{P.
+secunda}  and \textit{P.
 spicata}, we found no sign of a decrease in the removal effect over time (Tables \ref{table:POSEgrowth-trtYears} and \ref{table:PSSPgrowth-trtYears}).
@@
 \textit{A.
-tripartita} crowding and crowding by all subdominant shrub and grass species appeared greater than in the controls, whereas crowding by \textit{H.
-comata} was lower in removal plots than controls, though it was absent from most \textit{P.
+tripartita} crowding and crowding by all subdominant shrub and grass species appeared greater in removal plots than in the controls, whereas crowding by \textit{H.
+comata} was lower in removals than controls, though it was absent from most \textit{P.
 spicata} neighborhoods in both treatments (Fig.
@@
 For \textit{A.
-tripartita}, both observed and predicted population growth rates were close to zero, and differences in model predictions were small.
+tripartita}, both observed and predicted population growth rates were close to zero, and differences in predictions from the two models were small.
 \textit{H.
@@
 Red lines and symbols show one-step-ahead predictions from an IBM that includes removal treatment coefficients.
-Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r} }
+Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r}.
+}
  \label{fig:IBM1step}
@@
 Solid symbols correspond to predictions form the baseline IBM (no removal effects) and open symbols in (B) show predictions from an IBM that includes removal treatment effects.
-Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r} }
+Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r}.
+}
   \label{fig:ObsPred1to1}
@@
 We concentrated on grass species' responses to \textit{A.
-tripartita} removal, since the growth regressions indicated our baseline model might underestimate competitive release in this situation (our baseline vital rate models worked well in predicting \textit{A.
+tripartita} removal, since the growth regressions indicated our baseline model might underestimate competitive release in this situation (our baseline vital rate models worked well for predicting \textit{A.
 tripartita} responses to grass removal).
@@
 tripartita} present and no removal treatment coefficients, brown boxes show results from the same model but with \textit{A.
-tripartita} cover set to zero (a species removal), and green boxes show results from a model with  \textit{A.
+tripartita} (``ARTR") cover set to zero (a species removal), and green boxes show results from a model with  \textit{A.
 tripartita} set to zero and removal treatment coefficients included.
@@
 
- The equilibrium cover of \text{Artemisia} simulated by the IPM (6.0\%) is consistent with observed cover during the historical data collection period (6.6\% from 1931-1956), but is low compared to contemporary observed cover (15-25\%,\ref{fig:CoverTrends}).
-Could the small competitive release following simulated \text{Artemisia} removal simply reflect the low \text{Artemisia} cover?
-To explore this possibility, we increased \text{Artemisia} cover in the baseline simulation by increasing average \text{Artemisia} fecundity.
-The increased fecundity led to a doubling of \text{Artemisia} average cover, but had essentially zero effect on the cover of the grasses (changes of less than 0.1\% cover).
+ The equilibrium cover of \textit{A.
+tripartita} simulated by the IPM (6.0\%) is consistent with observed cover during the historical data collection period (6.6\% from 1931-1956), but is low compared to contemporary observed cover (15-25\%, Fig.
+\ref{fig:CoverTrends}).
+Could the small competitive release following simulated \textit{A.
+tripartita} removal simply reflect the low \textit{A.
+tripartita} cover?
+To explore this possibility, we increased \text{Artemisia} cover in the baseline simulation by increasing average \textit{A.
+tripartita} fecundity.
+The increased fecundity led to a doubling of \textit{A.
+tripartita} average cover, but had essentially zero effect on the cover of the grasses (changes of less than 0.1\% cover).
 
@@
 comata}; our baseline model, which accounts for local neighborhood composition, adequately explained vital rate responses to  removal.
-\textit{Poa secunda} showed mixed responses to \textit{Artemisia} removal: removal effects were not significant at the survival stage, were positive at the growth stage, and negative at the recruitment stage (where the estimated effect of \textit{Artemisia} on \textit{Poa} is positive).
+\textit{P.
+secunda} showed mixed responses to \textit{A.
+tripartita} removal: removal effects were not significant at the survival stage, were positive at the growth stage, and negative at the recruitment stage (where the estimated effect of \textit{A.
+tripartita} on \textit{P.
+secunda} is positive).
 At the population level, models with and without removal effects projected similar patterns of short (Fig.
@@
 tripartita}.
-For \textit{Poa secunda}, the model with removal effects predicted higher population growth over the short but not the long-term, perhaps because the short-term projections mostly reflect the positive effect of \textit{Artemisia} removal on growth, while the long-term equilibrium are also sensitive to the negative effects of removal on recruitment.
+For \textit{P.
+secunda}, the model with removal effects predicted higher population growth over the short but not the long-term, perhaps because the short-term projections mostly reflect the positive effect of \textit{A.
+tripartita} removal on growth, while the long-term equilibrium are also sensitive to the negative effects of removal on recruitment.
 Although the baseline model did not always successfully predict removal effects on vital rates, it does appear to successfully capture the population-level consequences of removals on these three species and did not severely underestimate the magnitude of competitive release.
@@
 Both individual growth and short- and long-term population growth were higher in the sagebrush removal treatments than in the control plots, suggesting that our analyses of observational data may underestimate the competitive release that \textit{P.
-spicata} experiences following \textit{Artemisia} removal.
-However, the mechanism by which \textit{Artemisia} removal promotes \textit{P.
+spicata} experiences following \textit{A.
+tripartita} removal.
+However, the mechanism by which \textit{A.
+tripartita} removal promotes \textit{P.
 spicata} remains unclear.
-If  \textit{Artemisia} removal is releasing \textit{P.
+If  \textit{A.
+tripartita} removal is releasing \textit{P.
 spicata} from competition, we should expect \textit{P.
-spicata} individuals that were located beneath \textit{Artemisia} shrubs that we removed to respond more than individuals located away from any shrubs.
+spicata} individuals that were located beneath \textit{A.
+tripartita} shrubs that we removed to respond more than individuals located away from any shrubs.
 Yet we found no evidence for such a difference.
-An alternative explanation involves pre-treatment differences in composition between control plots and  the \textit{Artemisia} removal  plots, which had higher cover of \textit{Artemisia} (see 2011 means in Fig.
+An alternative explanation involves pre-treatment differences in composition between control plots and  the \textit{A.
+tripartita} removal  plots, which had higher cover of \textit{A.
+tripartita} (see 2011 means in Fig.
 \ref{fig:CoverTrends}) and higher cover of subdominant grasses and shrubs (Fig.
 \ref{fig:W-by-treatment}).
-Perhaps these pre-treatment differences reflect subtle differences in edaphic and environmental factors which could promote \textit{P.
+Perhaps these pre-treatment differences reflect subtle differences in edaphic conditions which could promote \textit{P.
 spicata} growth.
-While \textit{P.
+In summary, while \textit{P.
 spicata} appears to show stronger competitive release than our baseline model predicted, the evidence is not conclusive.
@@
 More likely, plants may alter the environment, and even resource availability, in ways that facilitate their neighbors.
-For example, shade cast by \textit{Artemisia} canopies may reduce evapotranspiration (e.g.
+For example, shade cast by \textit{{A.
+tripartita} canopies may reduce evapotranspiration (e.g.
 \citealt{Barbier2008}).
-\textit{Artemisia} removal might offset this facilitative effect by reducing resource uptake, resulting in little net change in water balance following removal.
+\textit{{A.
+tripartita} removal might offset this facilitative effect by reducing resource uptake, resulting in little net change in water balance following removal.
 Unfortunately, we did not monitor soil moisture at the fine spatial and temporal resolution necessary to test this hypothesis.
 
-Evidence for weak interspecific interactions and limited competitive release carries interesting implications.
+Evidence for weak interspecific interactions and limited competitive release has interesting implications.
 If most interspecific interactions are weak, then single-species models may be adequate for forecasting population responses to environmental change \citep{adler_forecasting_2012,kleinhesselink_indirect_2015}.
 Similarly, differences among species in relative abundance should reflect variation in the strength of intraspecific density dependence more than variation in the strength of interspecific interactions \citep{yenni_strong_2012}.
-A recent global, grassland study showing little variation in the abundance of invasive species in native and invaded ranges on different continents  \citep{firn_abundance_2011} is consistent with the idea that intraspecific interactions are the primary determinant of species abundances.
+A recent global study showing little variation in the abundance of invasive grassland species in native and invaded ranges on different continents  \citep{firn_abundance_2011} is consistent with the idea that intraspecific interactions are the primary determinant of species abundances.
 In systems where this is true, effective management and conservation requires understanding the sources and species-specificity of intraspecific density dependence.
@@
 In closed-canopy forests where competition for light is intense, we might have found different results than in our system, in which competition for a number of belowground resources is more important than competition for light.
-However, a number of recent studies from forests have also found that heterospecific competitive effects are dramatically weaker than conspecific effects \citep{comita_asymmetric_2010,kunstler_plant_2015,johnson_conspecific_2012}.
+However, a number of recent studies from forests have also found that heterospecific competitive effects are dramatically weaker than conspecific effects \citep{comita_asymmetric_2010,Kunstler2016,johnson_conspecific_2012}.
 We also expect stronger competitive release in successional systems, or communities in the early stages of community assembly where the process of competitive exclusion is still underway \citep{kokkoris_patterns_1999}.
@@
 
-Funding was provided by NSF grants DEB-1353078,and  DEB-1054040 to PBA and DEB-1353039 to SPE, and by the Utah Agricultural Experiment Station (get journal paper number).
+Funding was provided by NSF grants DEB-1353078,and  DEB-1054040 to PBA and DEB-1353039 to SPE, and by the Utah Agricultural Experiment Station.
 The USDA-ARS Sheep Experiment Station generously provided access to historical data and the field experiment site.
@@
 \begin{table}[h]
-\caption{Cover change models}
+\caption{Statistical models of year-to-year changes in log(cover) for the four focal species.}
 \centering
@@
 \hline
-Species & ARTR & HECO & POSE & PSSP \\
+Species & \textit{A. tripartita} & \textit{H. comata} & \textit{P. secunda} & \textit{P. spicata} \\
 \hline
@@
 \centering
-\caption{Summary of fixed effects for the \textit{A. tripartita} survival model} 
+\caption{Summary of fixed effects for the \textit{A. tripartita} survival model. ``logarea" is the effect of plant size, ``Treatment*" is the removal effect, and the ``W.*" coefficients 
+are effects of neighborhood crowding (the $\omega$s in eqn.
+\ref{eqn:survReg}).
+``quant" refers to quantile and ``kld" reports the Kullback-Leibler divergence between the Gaussian and the (simplified)Laplace approximation to the marginal posterior densities.
+}
 \label{ARTRsurvival}
@@
 \centering
-\caption{Summary of fixed effects for the \textit{H. comata} survival model} 
+\caption{Summary of fixed effects for the \textit{H. comata} survival model (see Table \ref{ARTRsurvival} for an explanation of coefficient names
+and column headers).} 
 \label{HECOsurvival}
@@
 \centering
-\caption{Summary of fixed effects for the \textit{Poa secunda} survival model} 
+\caption{Summary of fixed effects for the \textit{Poa secunda} survival model (see Table \ref{ARTRsurvival} for an explanation of coefficient names
+and column headers).} 
 \label{POSEsurvival}
@@
 \centering
-\caption{Summary of fixed effects for the \textit{P. spicata} survival model} 
+\caption{Summary of fixed effects for the \textit{P. spicata} survival model (see Table \ref{ARTRsurvival} for an explanation of coefficient names
+and column headers).} 
 \label{PSSPsurvival}
@@
 \centering
-\caption{Summary of fixed effects for the \textit{A. tripartita} growth model} 
+\caption{Summary of fixed effects for the \textit{A. tripartita} growth model. ``logarea.t0" is the effect of plant size, ``Treatment*" is the removal effect, and the ``W.*" coefficients 
+are effects of neighborhood crowding (the $\omega$s in eqn.
+\ref{eqn:growReg}.
+``quant" refers to quantile and ``kld" reports the Kullback-Leibler divergence between the Gaussian and the (simplified) Laplace approximation to the marginal posterior densities.}
 \label{ARTRgrowth}
@@
 \centering
-\caption{Summary of fixed effects for the \textit{H. comata} growth model} 
+\caption{Summary of fixed effects for the \textit{H. comata} growth model (see Table \ref{ARTRgrowth} for an explanation of coefficient names
+and column headers).} 
 \label{HECOgrowth}
@@
 \centering
-\caption{Summary of fixed effects for the \textit{Poa secunda} growth model} 
+\caption{Summary of fixed effects for the \textit{Poa secunda} growth model (see Table \ref{ARTRgrowth} for an explanation of coefficient names
+and column headers).} 
 \label{POSEgrowth}
@@
 \centering
-\caption{Summary of fixed effects for the \textit{P. spicata} growth model} 
+\caption{Summary of fixed effects for the \textit{P. spicata} growth model (see Table \ref{ARTRgrowth} for an explanation of coefficient names
+and column headers).} 
 \label{PSSPgrowth}
@@
 \centering
-\caption{Summary of fixed effects for the \textit{Po. secunda} growth model with individual-level \textit{A. tripartita removal} data} 
+\caption{Summary of fixed effects for the \textit{P.
+secunda} growth model with individual-level \textit{A.
+tripartita removal} data (the ``inARTR" coefficient).
+See Table \ref{ARTRgrowth} for an explanation of other coefficient names and column headers).
+}
 \label{table:POSEgrowth-inARTR}
@@
 \centering
-\caption{Summary of fixed effects for the \textit{Ps. spicata} growth model with individual-level \textit{A. tripartita} removal data} 
+\caption{Summary of fixed effects for the \textit{P.
+spicata} growth model with individual-level \textit{A.
+tripartita} removal data (the ``inARTR" coefficient).
+See Table \ref{ARTRgrowth} for an explanation of other coefficient names and column headers).}
 \label{table:PSSPgrowth-inARTR}
@@
 \centering
-\caption{Summary of fixed effects for the \textit{Po. secunda} model with treatment*year effects} 
+\caption{Summary of fixed effects for the \textit{P.
+secunda} model with treatment*year effects (the ``trtYears*" coefficients).
+See Table \ref{ARTRgrowth} for an explanation of other coefficient names and column headers).}
 \label{table:POSEgrowth-trtYears}
@@
 \centering
-\caption{Summary of fixed effects for the \textit{Ps. spicata} model with treatment*year effects} 
+\caption{Summary of fixed effects for the \textit{P.
+spicata} model with treatment*year effects (the ``trtYears*" coefficients).
+See Table \ref{ARTRgrowth} for an explanation of other coefficient names and column headers).}
 \label{table:PSSPgrowth-trtYears}
@@
  \includegraphics[width=0.7\textwidth]{CompKernels}
- \caption{Competition kernels for the four dominant species. Source file: \texttt{Wdistrib\textbackslash exploreSurvivalWs.r} }
+ \caption{Competition kernels for the four dominant species. Source file: \texttt{Wdistrib\textbackslash exploreSurvivalWs.r}. }
  \label{fig:CompKernels}
@@
   \caption{Annual precipitation (a) and mean temperature (b) during the period of the experiment, shown against the long-term (1927-2016) means (solid blue and red lines) and 5\% and 95\% quantiles (dashed lines).
-Source file:  \texttt{climate\textunderscore fig.r}}
+Source file:  \texttt{climate\textunderscore fig.r}.}
   \label{fig:climate}
@@
 Blue symbols show values from controls plots, red symbols show values from \textit{Artemisia} removal plots.``W.allcov'' refers to the aggregated crowding by all shrubs and perennial grasses beyond the focal dominant species, and ``W.allpts'' refers to the aggregated crowding by forb species.
-\texttt{Wdistrib\textbackslash exploreSurvivalWs.r} }
+\texttt{Wdistrib\textbackslash exploreSurvivalWs.r}.
+}
   \label{fig:Wscatters}
@@
  \caption{Distribution of crowding exerted by each neighbor species on \textit{P.
-spicata} individuals in controls plots (blue) and removal quadrats (red) in 2011, before \textit{Artemisia}  removals were conducted.
+spicata} individuals in controls plots (blue) and removal quadrats (red) in 2011, before \textit{A.
+tripartita}  removals were conducted.
 Inset bar graphs show the probability that a particular neighbor species was present ($W>0$) in a focal plant's local neighborhood in control (blue) and removal (red) quadrats.
 ``W.allcov'' refers to the aggregated crowding by all shrubs and perennial grasses beyond the focal dominant species, and ``W.allpts'' refers to the aggregated crowding by forb species.
-\texttt{Wdistrib\textbackslash exploreSurvivalWs.r} }
+\texttt{Wdistrib\textbackslash exploreSurvivalWs.r}.
+}
  \label{fig:W-by-treatment}
@@
 Red lines and symbols show one-step-ahead predictions from an IBM that includes removal treatment coefficients.
-Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r} }
+Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r}.
+}
  \label{fig:IBM1step-maxCI}
@@
 Solid symbols correspond to predictions form the baseline IBM (no removal effects) and open symbols in (B) show predictions from an IBM that includes removal treatment effects.
-Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r} }
+Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r}.
+}
    \label{fig:ObsPred1to1-maxCI}
@@
 tripartita} set to zero and (mean) removal treatment coefficients included, and blue boxes show results with removal treatments coefficients assigned the outermost value of the 95\% confidence intervals.
-Source file: \texttt{ipm\textbackslash IPM-figures.r}}
+Source file: \texttt{ipm\textbackslash IPM-figures.r}.}
   \label{fig:IPMresults-maxCI}

### 2017-01-13 removals/manuscript/removal_ms.tex
Bibtek problems
Final edits to ms.

@@
 More likely, plants may alter the environment, and even resource availability, in ways that facilitate their neighbors.
-For example, shade cast by \textit{{A.
+For example, shade cast by \textit{A.
 tripartita} canopies may reduce evapotranspiration (e.g.
 \citealt{Barbier2008}).
-\textit{{A.
+\textit{A.
 tripartita} removal might offset this facilitative effect by reducing resource uptake, resulting in little net change in water balance following removal.
@@
 \bibliographystyle{Ecology}
-\bibliography{IPMrefs}
+\bibliography{RemovalRefs}
 
@@
 \ref{eqn:survReg}).
-``quant" refers to quantile and ``kld" reports the Kullback-Leibler divergence between the Gaussian and the (simplified)Laplace approximation to the marginal posterior densities.
+``quant" refers to quantile and ``kld" reports the Kullback-Leibler divergence between the Gaussian and the (simplified) Laplace approximation to the marginal posterior densities.
 }
@@
 \centering
-\caption{Summary of fixed effects for the \textit{Poa secunda} survival model (see Table \ref{ARTRsurvival} for an explanation of coefficient names
+\caption{Summary of fixed effects for the \textit{P. secunda} survival model (see Table \ref{ARTRsurvival} for an explanation of coefficient names
 and column headers).} 
@@
 \centering
-\caption{Summary of fixed effects for the \textit{Poa secunda} growth model (see Table \ref{ARTRgrowth} for an explanation of coefficient names
+\caption{Summary of fixed effects for the \textit{P. secunda} growth model (see Table \ref{ARTRgrowth} for an explanation of coefficient names
 and column headers).} 

### 2017-01-13 removals/manuscript/cover.tex
Cover letter and formatting

\documentclass[11pt]{letter}
%\renewcommand{\baselinestretch}{2}
%\setlength{\headheight}{2in}
\setlength{\topmargin}{0in}
\setlength{\headsep}{0in}
\setlength{\oddsidemargin}{0in}
\setlength{\evensidemargin}{0in}
\setlength{\textheight}{8.5in}
\setlength{\textwidth}{6.5in}   
%\setlength{\textheight}{6.5in}

\usepackage{graphicx}

\begin{document}

Peter Adler\\
 Department of Wildland Resources and the Ecology Center\\
 5230 Old Main\\
 Utah State University, Logan, UT 84322\\
 Tel. 801-910-3816 \quad Fax:  435-797-3796\\
 email: peter.adler@usu.edu

 \medskip

Dear Editors,

We are hereby submitting our manuscript, ``Weak interspecific interactions in a sagebrush steppe: evidence from observations, models, and experiments''
for publication as a Standard Paper in the Journal of Ecology.

Interspecific competition is arguably the primary obsession of plant community ecologists, but in a series of paper my collaborators and I have published based on long-term demographic data and multispecies population models for semiarid plant communities (Adler et al. 2010, Adler et al. 2012, Chu and Adler 2015), we found surprisingly weak interspecific interactions. These models predict little competitive release: removal of a dominant species would have little impact on the abundances of remaining species.

We conducted a five year removal experiment in a sagebrush steppe plant community to test this prediction. We found that models based on long-term observational data adequately 
predicted vital rate and population-level responses to removals for three of our four study species. For the fourth species, the model based on observational data underestimated population growth in the removal treatment, but results of additional analyses question whether this mismatch actually reflects competitive release.

Overall, the results of our experiment increase our confidence that the structure and dynamics of this plant community are determined more by variation in intraspecific interactions than by strong interspecific competition for resources. We expect this conclusion will provide food for thought for a broad audience of plant ecologists raised on a steady diet of studies emphasizing the importance of interspecific competition.

For the convenience of reviewers, we have placed the Figures and Tables in the main text, but the final 
version (if the paper is accepted) will be structured as specified in the Author Guidelines. All data and R code needed
to reproduce the analysis are currently 
available on Github (see Methods), and will be deposited in a permanent archive upon acceptance of the manuscript.

None of this material has been published or is under consideration elsewhere. I am the corresponding author and my contact
information is listed above. All authors have seen and approved the manuscript.

We look forward to hearing from you.

Sincerely,

Peter Adler and coauthors

\bibliographystyle{Ecology}
\bibliography{RemovalRefs}

\newpage

\textbf{References}

Adler, P.B., Dalgleish, H.J. \& Ellner, S.P. (2012) Forecasting plant community impacts of climate variability and change: when do competitive interactions matter? Journal of Ecology, 100, 478-487.

Adler, P.B., Ellner, S.P. \& Levine, J.M. (2010) Coexistence of perennial plants: an embarrassment of niches. Ecology Letters, 13, 1019-1029.

Chu, C. \& Adler, P.B. (2015) Large niche differences emerge at the recruitment stage to stabilize grassland coexistence. Ecological Monographs, 85, 373-392.

\end{document}

### 2017-01-13 removals/manuscript/removal_ms.tex
Cover letter and formatting

@@
 
-\date{Last compile: \today}
+% \date{Last compile: \today}
 
@@
 
-\textbf{\large{Keywords:}} Coexistence, competition, integral projection model.
+\textbf{\large{Keywords:}} Coexistence, competition, integral projection model, removal experiment, sagebrush steppe.
 
@@
 
-\smallskip \textbf{Article type:} Letter
-
-\smallskip \textbf{Authorship statement:} PBA and ARK designed the experiment and supervised data collection with the help of JBT, PBA, SPE, GH and BT analyzed data, PBA wrote the first draft of the manuscript and all authors contributed to editing.
-\smallskip
-
-\setlength{\parindent}{8ex}
+\smallskip \textbf{Article type:} Standard Paper
 
@@
 Determining why interspecific interactions and competitive release are strong in some ecological contexts but weak in others remains a fascinating research problem.
+
+\section*{Author contributions}
+
+PBA and ARK designed the experiment and supervised data collection with the help of JBT, PBA, SPE, GH and BT analyzed data, PBA wrote the first draft of the manuscript and all authors contributed to editing.
 

### 2017-01-30 removals/manuscript/cover.tex
Final edits to ms and cover letter

@@
 
-Interspecific competition is arguably the primary obsession of plant community ecologists, but in a series of paper my collaborators and I have published based on long-term demographic data and multispecies population models for semiarid plant communities (Adler et al.
+Interspecific competition is arguably the primary obsession of plant community ecologists.
+However, in a series of papers based on long-term demographic data and multispecies population models for semiarid plant communities (Adler et al.
 2010, Adler et al.
-2012, Chu and Adler 2015), we found surprisingly weak interspecific interactions.
-These models predict little competitive release: removal of a dominant species would have little impact on the abundances of remaining species.
+2012, Chu and Adler 2015), my collaborators and I found surprisingly weak interspecific interactions.
+These results imply little competitive release; removal of a dominant species should have little impact on the abundances of remaining species.
 
-We conducted a five year removal experiment in a sagebrush steppe plant community to test this prediction. We found that models based on long-term observational data adequately 
+To test this prediction, we conducted a five year removal experiment in a sagebrush steppe plant community. We found that models based on long-term observational data adequately 
 predicted vital rate and population-level responses to removals for three of our four study species.
-For the fourth species, the model based on observational data underestimated population growth in the removal treatment, but results of additional analyses question whether this mismatch actually reflects competitive release.
+For the fourth species, the model based on observational data underestimated population growth in the removal treatment, but additional analyses suggest that this mismatch may not actually reflect competitive release.
 
 Overall, the results of our experiment increase our confidence that the structure and dynamics of this plant community are determined more by variation in intraspecific interactions than by strong interspecific competition for resources.
-We expect this conclusion will provide food for thought for a broad audience of plant ecologists raised on a steady diet of studies emphasizing the importance of interspecific competition.
+We expect this conclusion may surprise a broad audience of plant ecologists accustomed to studies emphasizing the importance of interspecific competition.
 

### 2017-01-30 removals/manuscript/removal_ms.tex
Final edits to ms and cover letter

@@
 \item \emph{Synthesis}: Although models based on observational data did not perfectly predict species responses to removal, the experimental results increase our confidence that competitive release is in fact small for at least three of our four study species.
-The implication is that interspecific resource competition is not the primary factor determining the dynamics of this community.
+The implication is that interspecific resource competition is not a primary factor determining the dynamics of this community.
+The success of the models also supports the use of multispecies models, based on long-term observational studies, to characterize competitive interactions in a community.
 
@@
 Our second question integrates across the full life cycle: how do short and long-term population-level responses to removals compare to projections from multispecies population models based on observational data?
-If models based on observational data adequately predict competitive release, the population dynamics they project should be similar to projections from models that include information from the experimental removal treatments.
+If models based on observational data adequately predict competitive release, the population dynamics they project should be similar to projections from models that include information from the experimental removal treatments, a result which would validate the use of models and long-term data sets to study interspecific interactions.
 
@@
 \mbox{logit}(S_{ijgh,t+1}) = \gamma_{j,t}^S + \varphi_{jg}^S+  \chi_{jh}^S  + \beta_{j,t}^S u_{ij,t} +  
-\left \langle \boldsymbol{\omega}_{j,t}^S, \boldsymbol{W}_{ij,t} \right \rangle 
+% \left \langle \boldsymbol{\omega}_{j,t}^S, \boldsymbol{W}_{ij,t} \right \rangle 
+\sum \limits_{j} \omega_{j}^S {W}_{ij,t}
 \label{eqn:survReg}
@@
 where $\gamma$ is a time-dependent intercept, $\varphi$ is the coefficient for the effect of spatial location group, $\chi$ is the removal treatment effect, $\beta$ is the coefficient that represents the effect of log genet size, $u$, on survival.
-$\boldsymbol{\omega}$ is a vector of interaction coefficients which determine the impact of crowding, $\boldsymbol{W}$, by each species on the focal species.
-The vector $\boldsymbol{W}$ includes crowding from the four dominant species,  \textit{A.
+%$\boldsymbol{\omega}$ is a vector of interaction coefficients which determine the impact of crowding, $\boldsymbol{W}$, by each species on the focal species. 
+$\omega_j$ is an interaction coefficient which determines the impact of crowding, $W_j$, by species $j$ on the focal species. 
+Our model includes crowding from the four dominant species,  \textit{A.
 tripartita}, \textit{H.
@@
 secunda}, as well as two covariates representing total crowding from (1) all other perennial grasses and shrubs, which were mapped as polygons, and (2) all forb species, which were mapped as points.
-$\left \langle \boldsymbol{x, y} \right \rangle$ denotes the inner product of vectors $\boldsymbol{x}$ and $\boldsymbol{y}$, 
-calculated as \texttt{sum(x*y)} in R.
+%$\left \langle \boldsymbol{x, y} \right \rangle$ denotes the inner product of vectors $\boldsymbol{x}$ and $\boldsymbol{y}$, 
+%calculated as \texttt{sum(x*y)} in R.
 
@@
 u_{ijgh,t+1} = \gamma_{j,t}^G + \varphi_{jg}^G+  \chi_{jh}^G  + \beta_{j,t}^G u_{ij,t} + 
-\left \langle  \boldsymbol{\omega}_{j,t}^G, \boldsymbol{W}_{ij,t} \right \rangle + \varepsilon_{ij,t}^G .
+%\left \langle  \boldsymbol{\omega}_{j,t}^G, \boldsymbol{W}_{ij,t} \right \rangle + \varepsilon_{ij,t}^G .
+\sum \limits_{j} \omega_{j}^G {W}_{ij,t} + \varepsilon_{ij,t}^G .
 \label{eqn:growReg}
@@
 \end{equation} 
-Note that $W_{ijj}$ gives intraspecific crowding. The $W$'s are then the components of the $\boldsymbol{W}$ vectors 
+Note that $W_{ijj}$ gives intraspecific crowding. The $W$'s are then % the components of the $\boldsymbol{W}$ vectors 
 introduced as covariates in the survival (\ref{eqn:survReg}) and growth (\ref{eqn:growReg}) regressions.
@@
 $\beta$, $\boldsymbol{\omega}$, $\vec{b}$ and $\eta$ as parameters to be fitted. We implemented this in the statistical computing environment, \texttt{R}, 
-by making the spline coefficients and $\eta$ the arguments of an objective function that computes $\boldsymbol{W}$ using the input spline coefficients, 
+by making the spline coefficients and $\eta$ the arguments of an objective function that computes ${W}$ using the input spline coefficients, 
 calls the model-fitting functions \texttt{lmer} and/or \texttt{glmer} to fit the other parameters in the survival and growth regressions, 
@@
 survival and growth. In addition, for fitting the kernels we assumed that survival and growth depended only on intraspecific crowding, and thus only fitted the
-within-species competition kernels $F_{jj}$. Based on previous work \citep{adler_coexistence_2010}, we set all $F_{mj}$ equal to $F_{jj}$, meaning that 
-the within-species competition kernel for species $j$ is also used to determine the effect of all other species on species $j$.
+within-species competition kernels $F_{jj}$.
+Based on previous work with parametric kernels, \citep{adler_coexistence_2010}, we know that interspecific effects are rarely significantly different from zero, and there is not enough signal to fit interspecific kernels nonparametrically.
+Therefore, we set all $F_{mj}$ equal to $F_{jj}$, meaning that
+the within-species competition kernel for species $j$ is also used to determine the effect of species $j$ on all other species.
 We used data from all historical plots and contemporary control-treatment plots to estimate the competition kernels, which are shown in Fig.
@@
 
-Once we had estimated the competitions kernels, we used them to calculate the values of $\boldsymbol{W}$ for each individual, 
-and fit the full survival and growth regressions, which include the interspecific interaction coefficients, $\boldsymbol{\omega}$. 
-All genets in a quadrat were included in calculating $W$, but plants located within 5 cm of quadrat edges were not used in fitting. 
+Once we had estimated the competitions kernels, we used them to calculate the values of all ${W_j}$ for each individual, 
+and fitted the full survival and growth regressions, which include the interspecific interaction coefficients, $\omega_j$. 
+All genets in a quadrat were included in calculating the $W$'s, but plants located within 5 cm of quadrat edges were not used in fitting. 
 We fit the models using the \texttt{R-INLA} package for R \citep{rue_approximate_2009}.
-We used the same approach to explore variations on the growth regressions, such as the addition of a year-by-treatment interaction or information about the locations of individual grasses relative to removed \textit{A.
+We used the same approach wen exploring variations on the growth regressions, such as the addition of a year-by-treatment interaction or information about the locations of individual grasses relative to removed \textit{A.
 tripartita}.
-We also compared the species composition of the control and removal plots at both the quadrat and neighborhood ($\boldsymbol{W}$) scales.
+We also compared the species composition of the control and removal plots at both the quadrat and neighborhood ($W$) scales.
 
@@
 \lambda_{jq,t+1} = C'_{jq,t} \exp{\left(\gamma_{j,t}^R +  \varphi_{jg}^R + \chi_{jm}^R + 
-\left \langle \boldsymbol{\omega}^R , \boldsymbol{\sqrt{C'}_{q,t}} \right \rangle \right) }
+%\left \langle \boldsymbol{\omega}^R , \boldsymbol{\sqrt{C'}_{q,t}} \right \rangle \right) }
+\sum \limits_{j} \omega_{j}^R \sqrt{C'}_{ij,t} \right)}
 \label{eqn:recrProcessModel}
@@
 where the superscript $R$ refers to Recruitment, $C'_{jq,t}$ is the `effective cover' (cm$^2$) of species $j$ in quadrat $q$ at time $t$, $\gamma$ is a time-dependent intercept, $\varphi$ is a coefficient for the effect of spatial location,
-$\chi$ is the removal treatment effect, $\boldsymbol{\omega}$ is a vector of coefficients that determine the strength of intra- and interspecific density-dependence, and $\boldsymbol{C'}$ is the vector of ``effective'' cover of each species in the community.
+$\chi$ is the removal treatment effect, $\omega_j$ is the coefficient that determines the density-dependent effect of species $j$ on the focal species, and $C'_j$ is the ``effective'' cover of species $j$ in the quadrat.
 Eqn (\ref{eqn:recrProcessModel}) is essentially the Ricker model for discrete time population growth \citep{ricker_stock_1954}, except that we use the square root of cover in the density-dependent term.
@@
 The vital rate regressions provide a qualitative test of removal treatment effects: Does removal have an effect on a vital rate that goes above and beyond the effects of species interactions that are explicit in the model?
-However, the population-level response by each species integrates the effects of the removal treatment on all three vital rates.
-We used an individual-based model (IBM) to compare observed and predicted changes
-in population size (cover) from one year to the next.
+However, the population-level effects of removal integrate the effects of the removal treatment on all three vital rates.
+We used an individual-based model (IBM) to predict changes in population size (cover) from one year to the next in removal and control quadrats, and compared these predictions with observed changes.
 
@@
 We then projected the model forward one year, applying the random year effects corresponding to year $t$ and the spatial random effects corresponding to the quadrat of interest.
-Because we were interested in comparing model predictions to observations, and were not interested in the effects of demographic stochasticity, we used a deterministic version of the model (e.g., recruitment is the $\lambda$ of (\ref{eqn:recrProcessModel}), rather than a random draw from a Poisson distribution with a mean of $\lambda$).
+Because the main goal of this comparison is to look for systematic deviations between model predictions and observations, we used a deterministic version of the model (e.g., recruitment is the $\lambda$ of (\ref{eqn:recrProcessModel}), rather than a random draw from a Poisson distribution with a mean of $\lambda$).
 We first applied the recruitment regression to determine the expected number of new recruits at time $t+1$, which we then multiplied by the mean size of recruits observed for the focal species to get expected cover of new recruits.
@@
 \begin{equation}
-n(v_{j,t+1})=\int_{L_j}^{U_j} k_j (v_j,u_j,{\vec{n}})n(u_{j,t})   
+n(v_{j,t+1})=\int_{L_j}^{U_j} k_j (v_j,u_j,{\vec{n}})n(u_{j,t})du_j   
 \label{eqn:IPM}
@@
 \begin{equation}
-k_j(v_j,u_j,\vec{n})=S_j(u_j,\vec{n})G_j(v_j,u_j,\vec{n})+R_j(v_j,u_j,\vec{n}) du_j
+k_j(v_j,u_j,\vec{n})=S_j(u_j,\vec{n})G_j(v_j,u_j,\vec{n})+R_j(v_j,u_j,\vec{n}) 
 \label{eqn:IPM} 
@@
 with a `no-overlap' rule \citep{adler_coexistence_2010}.
-In \ref{suppMethods} we explain how the spline competition kernels with the no-overlap rule kernels were used to compute neigbhorhood crowding index as a function of individual size for the IPM.
+In \ref{suppMethods} we explain how the spline competition kernels with the no-overlap rule were used to compute neighborhood crowding index as a function of individual size for the IPM.
 
@@
 $R_j(v_j,u_j,\vec{n})=c_{0,j}(v_j)e^{u_j}\Phi_j$ \citep{adler_coexistence_2010} where $c_{0,j}$ is the
-size distribution of new recruits. $\Phi_j$ is calculated from eqn. (\ref{eqn:recrDataModel}) with total cover values computed from the population densities as $C'_{j,t} = \int e^u n_j(u,t) du$.
+size distribution of new recruits. $\Phi_j$ is calculated from eqn. (\ref{eqn:recrDataModel}) with total cover values computed from the population densities as $C'_{j,t} = \int e^u n(u_{j,t}) du$.
 
@@
 For the third scenario, we again set  \textit{A.
-triparitia} to zero, but this time we projected the model using the vital rate regressions that include removal effects.
+triparitia} cover to zero, but this time we projected the model using the vital rate regressions that include removal effects.
 If our baseline model successfully predicts competitive release observed in the removal experiment, then the results of the second and third scenarios should be similar.
@@
 (Table \ref{table:coefficients}). However, these treatment comparisons do not address our questions about whether responses to removal were 
-adequately predicted by models fit to observational data on non-removal plots.
+adequately predicted by models fit to historical, observational data.
 
@@
 Our vital rate models take into account the effects of local species composition and should therefore be able to explain the effect of removals on survival, growth and recruitment without any additional parameters.
-If the removal treatment effects that we added to these models are statistically significant, it would indicate an effect of removals that goes above and beyond the response predicted by a model based on data from the control plots.
+If the removal treatment effects that we added to these models are statistically significant, it would indicate an effect of removals either larger or smaller than the effect predicted by a model based on data from the historical control plots.
 For the survival models, the 95\% confidence intervals on the removal effects overlapped zero for all four species (Fig.
@@
 Finally, we evaluated pre-existing differences in community composition in removal and control plots.
-In selecting removal plots, our criteria required candidate plots to include a certain amount of our target removal species.
+In selecting removal plots, our criteria required candidate plots to include at least a certain amount of our target removal species.
 Because we adopted the historical plots as our controls, the same criteria were not applied (a control plot might not contain any \textit{A.
@@
 tripartita} set to zero and removal treatment coefficients included.
-Source file: \texttt{ipm\textbackslash IPM-figures.r}}
+Source file: \texttt{ipm\textbackslash IPM-figures.r}.}
  \label{fig:IPMresults}
@@
 Evidence for weak interspecific interactions and limited competitive release has interesting implications.
-If most interspecific interactions are weak, then single-species models may be adequate for forecasting population responses to environmental change \citep{adler_forecasting_2012,kleinhesselink_indirect_2015}.
+If most interspecific interactions are weak, then single-species models may be adequate for forecasting population responses to environmental change \citep{adler_forecasting_2012,kleinhesselink_indirect_2015,Chu2016}.
 Similarly, differences among species in relative abundance should reflect variation in the strength of intraspecific density dependence more than variation in the strength of interspecific interactions \citep{yenni_strong_2012}.
@@
 In systems where this is true, effective management and conservation requires understanding the sources and species-specificity of intraspecific density dependence.
-In fact, an experiment to test our model's skill at predicting release from intraspecific competition could be more important than the test of interspecific competitive release described here.
+A useful first step in this direction would be to test whether our model is also successful at predicting the effects of release from intraspecific competition
 
@@
 We also expect stronger competitive release in successional systems, or communities in the early stages of community assembly where the process of competitive exclusion is still underway \citep{kokkoris_patterns_1999}.
-Finally, we might have found stronger competitive release if we could have studied rare species in our system, as long as those species are rare because of niche overlap with stronger competitors.
+Finally, we might have found stronger competitive release if we could have studied rare species in our system, but only if those species are rare because of niche overlap with stronger competitors.
 If they are rare because they occupy unique, but narrow, niches, interspecific interactions would still be weak.
@@
 
-PBA and ARK designed the experiment and supervised data collection with the help of JBT, PBA, SPE, GH and BT analyzed data, PBA wrote the first draft of the manuscript and all authors contributed to editing.
+PBA and ARK designed the experiment and supervised data collection with the help of JBT; PBA, SPE, GH and BT analyzed data; PBA wrote the first draft of the manuscript and all authors contributed to editing.
 
@@
 
-Funding was provided by NSF grants DEB-1353078,and  DEB-1054040 to PBA and DEB-1353039 to SPE, and by the Utah Agricultural Experiment Station.
+Funding was provided by NSF grants DEB-1353078,and  DEB-1054040 to PBA and DEB-1353039 to SPE and GH, and by the Utah Agricultural Experiment Station.
 The USDA-ARS Sheep Experiment Station generously provided access to historical data and the field experiment site.

### 2017-01-31 removals/manuscript/removal_ms.tex
Final version for submission--fixed stats notation

@@
 For example, we might find that, holding crowding constant, grasses grow more quickly in the removal plots than in the control plots.
-This result would indicate that our ``baseline model," a model including all estimated parameters except for the removal treatment effect, underestimates competitive release.
-We refer to the model that includes the estimated removal treatment effect as the ``removal model."
-
-Formally, we modeled the survival probability, $S$, of genet $i$ in species $j$, group $g$, and removal treatment $h$, from time $t$ to $t+1$  as
+This result would indicate that our ``baseline model'', a model including all estimated parameters except for the removal treatment effect, underestimates competitive release.
+We refer to the model that includes the estimated removal treatment effect as the ``removal model".
+
+Formally, we modeled the survival probability, $S$, of genet $i$ in species $j$ (which is in quadrat group $g$ and removal treatment $h$), 
+from time $t$ to $t+1$  as
 \begin{equation}
-\mbox{logit}(S_{ijgh,t+1}) = \gamma_{j,t}^S + \varphi_{jg}^S+  \chi_{jh}^S  + \beta_{j,t}^S u_{ij,t} +  
-% \left \langle \boldsymbol{\omega}_{j,t}^S, \boldsymbol{W}_{ij,t} \right \rangle 
-\sum \limits_{j} \omega_{j}^S {W}_{ij,t}
+\mbox{logit}(S_{ij,t+1}) = \gamma_{j,t}^S + \varphi_{jg}^S+  \chi_{jh}^S  + \beta_{j,t}^S u_{ij,t} +  
+\sum \limits_{m} \omega_{jm}^S {W}_{ijm,t}
 \label{eqn:survReg}
 \end{equation}
-where $\gamma$ is a time-dependent intercept, $\varphi$ is the coefficient for the effect of spatial location group, $\chi$ is the removal treatment effect, $\beta$ is the coefficient that represents the effect of log genet size, $u$, on survival.
-%$\boldsymbol{\omega}$ is a vector of interaction coefficients which determine the impact of crowding, $\boldsymbol{W}$, by each species on the focal species. 
-$\omega_j$ is an interaction coefficient which determines the impact of crowding, $W_j$, by species $j$ on the focal species. 
+where $\gamma$ is a time-dependent intercept (year effect), $\varphi$ is the coefficient for the 
+effect of spatial location group, $\chi$ is the removal treatment effect,and  $\beta$ is the coefficient that 
+represents the effect of log genet size, $u$, on survival. 
+$W_{ijm,t}$ is the crowding of individual $i$ in species $j$ by neighbors of species $m$ (explained below), and 
+$\omega_{jm}$ is an interaction coefficient which determines the impact of crowding by species $m$ on the focal species $j$.
+
+Our growth model has a similar structure. The change in genet size from time $t$ to $t+1$ , conditional on survival, is given by:
+\begin{equation}
+u_{ij,t+1} = \gamma_{j,t}^G + \varphi_{jg}^G+  \chi_{jh}^G  + \beta_{j,t}^G u_{ij,t} + 
+\sum \limits_{m} \omega_{jm}^G {W}_{ijm,t} + \varepsilon_{ij,t}^G .
+\label{eqn:growReg}
+\end{equation}
+To capture the non-constant error variance in growth, we modeled the variance $\varepsilon$ about the growth 
+curve (\ref{eqn:growReg}) as a nonlinear function of predicted genet size:
+\begin{equation}
+Var(\varepsilon_{ij,t}^G) = a \exp(b\hat{u}_{ij,t+1}).
+\label{eqn:growVar}
+\end{equation}
+where $\hat{u}_{ij,t+1}$ is the right-hand side of (\ref{eqn:growReg}) without $\varepsilon$.
+
 Our model includes crowding from the four dominant species, \textit{A.
@@
 secunda}, as well as two covariates representing total crowding from (1) all other perennial grasses and shrubs, which were mapped as polygons, and (2) all forb species, which were mapped as points.
-%$\left \langle \boldsymbol{x, y} \right \rangle$ denotes the inner product of vectors $\boldsymbol{x}$ and $\boldsymbol{y}$, 
-%calculated as \texttt{sum(x*y)} in R.
-
-Our growth model has a similar structure. The change in genet size from time $t$ to $t+1$ , conditional on survival, is given by:
+
+We modeled the crowding experienced by a focal genet in each year $t$ as a function of the distance to and size of neighbor genets. In previous work, 
+we assumed that the decay of crowding with neighbor distance followed a Gaussian function \citep{chu_large_2015}, but here we use a data-driven 
+approach \citep{teller_linking_2016}. We model the crowding experienced by genet $i$ of species $j$ from neighbors of species $m$ as the sum of neighbor 
+areas across a set of concentric annuli, $k$, centered at the plant,
 \begin{equation}
-u_{ijgh,t+1} = \gamma_{j,t}^G + \varphi_{jg}^G+  \chi_{jh}^G  + \beta_{j,t}^G u_{ij,t} + 
-%\left \langle  \boldsymbol{\omega}_{j,t}^G, \boldsymbol{W}_{ij,t} \right \rangle + \varepsilon_{ij,t}^G .
-\sum \limits_{j} \omega_{j}^G {W}_{ij,t} + \varepsilon_{ij,t}^G .
-\label{eqn:growReg}
-\end{equation}
-To capture non-constant error variance in growth, we modeled the variance $\varepsilon$  about the growth curve (\ref{eqn:growReg})  as a nonlinear function of predicted genet size:
-\begin{equation}
-Var(\varepsilon_{ij,t}^G) = a \exp ^{(bu_{ij,t+1})} .
-\label{eqn:growVar}
-\end{equation}
-
-We model the crowding experienced by a focal genet as a function of the distance to and size of neighbor genets.
-In previous work, we assumed that the decay of crowding with neighbor distance followed a Gaussian function \citep{chu_large_2015}, but here we use a data-driven approach \citep{teller_linking_2016}.
-We model the crowding experienced by genet $i$ of species $j$ from neighbors of species $m$ as the sum of neighbor areas across a set of concentric annuli, $k$, centered at the plant,
-\begin{equation}
-w_{ijm,k} = F_{jm}(d_{k})A_{i,k}     
+w_{ijmk,t} = F_{jm}(d_{k})A_{ijmk,t}     
 \label{eqn:wik}
@@
 $d_{k}$ is the average of the inner and outer radii of annulus $k$, 
-and $A_{im,k}$ is the total area of genets of species $m$ in annulus $k$ around genet $i$. The total crowding on 
-genet $i$ exerted by species $m$ is
+and $A_{ijmk,t}$ is the total area of genets of species $m$ in annulus $k$ around the focal genet in year $t$. The total crowding on 
+the focal genet exerted by species $m$ is
 \begin{equation}
-W_{ijm}  =\sum_k {w_{ijm,k}} .
+W_{ijm,t}  =\sum_k {w_{ijmk,t}} .
 \label{eqn:wijm}
 \end{equation} 
-Note that $W_{ijj}$ gives intraspecific crowding. The $W$'s are then % the components of the $\boldsymbol{W}$ vectors 
-introduced as covariates in the survival (\ref{eqn:survReg}) and growth (\ref{eqn:growReg}) regressions.
+Note that $W_{ijj,t}$ gives intraspecific crowding. The $W$'s are then 
+included as covariates in the survival (\ref{eqn:survReg}) and growth (\ref{eqn:growReg}) regressions.
 
@@
 than close plants. Otherwise, we let the data dictate the shape of the kernel by fitting a spline model 
-using the methods of Teller et al. (2016). The shape of $F_{jm}$ is determined by a set of spline basis coefficients $\vec{b}_{jm}$
+using the methods of \citep{teller_linking_2016}. The shape of $F_{jm}$ is determined by a set of spline basis coefficients $\vec{b}_{jm}$
 and a smoothing parameter $\eta$ that controls the complexity of the fitted kernel. 
@@
 $\beta$, $\boldsymbol{\omega}$, $\vec{b}$ and $\eta$ as parameters to be fitted. We implemented this in the statistical computing environment, \texttt{R}, 
-by making the spline coefficients and $\eta$ the arguments of an objective function that computes ${W}$ using the input spline coefficients, 
+by making the spline coefficients and $\eta$ the arguments of an objective function that calculates  each $w_{ijmk,t}$ value using the input spline coefficients, 
 calls the model-fitting functions \texttt{lmer} and/or \texttt{glmer} to fit the other parameters in the survival and growth regressions, 
 and returns an approximate AIC value and model degrees of freedom ($df$) for survival and growth combined. We used the $\vec{b}$ values at the smoothest 
-local minimum of AIC as a function of $df$, as in \cite{teller_linking_2016}. This approach assumes that one measure of crowding affects 
+(smallest $df$) local minimum of AIC as a function of $df$, as in \cite{teller_linking_2016}. This approach assumes that one measure of crowding affects 
 survival and growth. In addition, for fitting the kernels we assumed that survival and growth depended only on intraspecific crowding, and thus only fitted the
 within-species competition kernels $F_{jj}$.
-Based on previous work with parametric kernels, \citep{adler_coexistence_2010}, we know that interspecific effects are rarely significantly different from zero, and there is not enough signal to fit interspecific kernels nonparametrically.
+Based on previous work with parametric kernels, \citep{adler_coexistence_2010}, we know that interspecific effects are rarely significantly different from zero, so there is not enough signal to fit interspecific kernels nonparametrically.
 Therefore, we set all $F_{mj}$ equal to $F_{jj}$, meaning that
-the within-species competition kernel for species $j$ is also used to determine the effect of species $j$ on all other species.
-We used data from all historical plots and contemporary control-treatment plots to estimate the competition kernels, which are shown in Fig.
-\ref{fig:CompKernels}.
-
-Once we had estimated the competitions kernels, we used them to calculate the values of all ${W_j}$ for each individual, 
-and fitted the full survival and growth regressions, which include the interspecific interaction coefficients, $\omega_j$. 
-All genets in a quadrat were included in calculating the $W$'s, but plants located within 5 cm of quadrat edges were not used in fitting. 
-We fit the models using the \texttt{R-INLA} package for R \citep{rue_approximate_2009}.
-We used the same approach wen exploring variations on the growth regressions, such as the addition of a year-by-treatment interaction or information about the locations of individual grasses relative to removed \textit{A.
-tripartita}.
-We also compared the species composition of the control and removal plots at both the quadrat and neighborhood ($W$) scales.
-
-We model recruitment at the quadrat level rather than at the individual genet level because the mapped data do not allow us to determine which recruits were produced by which potential parent genets.
-We assume that the number of individuals, $y$, of species $j$ recruiting at time $t+1$ in the location $q$ follows a negative binomial distribution:
+the within-species competition kernel for species $j$ is also used to determine the distance-dependence of species $j$ neighbors on all other species. 
+We used data from all historical plots and contemporary control-treatment plots to estimate the competition kernels, which are shown in Fig. \ref{fig:CompKernels}.
+
+Once we had estimated the competitions kernels, we used them to calculate the $w_{ijmk,t}$ values for each individual in each year, 
+and fitted the full survival and growth regressions, which include the interspecific interaction coefficients, $\omega_{jm}$. 
+All genets in a quadrat were included in calculating the $w$'s, but plants located within 5 cm of quadrat edges were not used in fitting the regressions. 
+We fit the models using the \texttt{R-INLA} package for R \citep{rue_approximate_2009}. We used the same approach when exploring variations 
+on the growth regressions, such as the addition of a year-by-treatment interaction or information about the locations of 
+individual grasses relative to removed \textit{A. tripartita}. We also compared the species composition of the control and removal plots at both 
+the quadrat and neighborhood ($W$) scales.
+
+We model recruitment at the quadrat level rather than at the individual genet level because the mapped data do not allow 
+us to determine which recruits were produced by which potential parent genets. We assume that the number of individuals, $y$, of species $j$ recruiting at 
+time $t+1$ in the location $q$ follows a negative binomial distribution:
 \begin{equation}
@@
 \end{equation}
-where $\lambda$ is the mean intensity and $\theta$ is the size parameter. In turn, $\lambda$ depends on the composition of the quadrat in the previous year:
+where $\lambda$ is the mean and $\theta$ is the size parameter (\texttt{dnegbin(p,$\theta$)} in the BUGS language, with $p=\theta/(\theta+\lambda)$). 
+In turn, $\lambda$ depends on the composition of the quadrat in the previous year:
 \begin{equation}
-\lambda_{jq,t+1} = C'_{jq,t} \exp{\left(\gamma_{j,t}^R +  \varphi_{jg}^R + \chi_{jm}^R + 
-%\left \langle \boldsymbol{\omega}^R , \boldsymbol{\sqrt{C'}_{q,t}} \right \rangle \right) }
-\sum \limits_{j} \omega_{j}^R \sqrt{C'}_{ij,t} \right)}
+\lambda_{jq,t+1} = C'_{jq,t} \exp{\left(\gamma_{j,t}^R +  \varphi_{jg}^R + \chi_{jh}^R + 
+\sum \limits_{m} \omega_{jm}^R \sqrt{C'_{mq,t}} \right)}
 \label{eqn:recrProcessModel}
 \end{equation}
-where the superscript $R$ refers to Recruitment, $C'_{jq,t}$ is the `effective cover' (cm$^2$) of species $j$ in quadrat $q$ at time $t$, $\gamma$ is a time-dependent intercept, $\varphi$ is a coefficient for the effect of spatial location,
-$\chi$ is the removal treatment effect, $\omega_j$ is the coefficient that determines the density-dependent effect of species $j$ on the focal species, and $C'_j$ is the ``effective'' cover of species $j$ in the quadrat.
+where the superscript $R$ refers to Recruitment, $C'_{jq,t}$ is the `effective cover' (cm$^2$) of species $j$ in quadrat $q$ at time $t$, $\gamma$ 
+is a time-dependent intercept, $\varphi$ is a coefficient for the effect of spatial location group,
+$\chi$ is the removal treatment effect, and $\omega_{jm}$ is the coefficient that determines the effect of species $m$ cover on recruitment by species $j$. 
 Eqn (\ref{eqn:recrProcessModel}) is essentially the Ricker model for discrete time population growth \citep{ricker_stock_1954}, except that we use the square root of cover in the density-dependent term.
@@
 
-Due to the possibility that plants outside the mapped quadrat could contribute recruits to the focal quadrat or interact with plants in the focal quadrat, we estimated effective cover as a mixture of the observed cover, $C$, in the focal quadrat, $q$, and the mean cover, $\bar{C}$, across the spatial location, $g$, in which the quadrat is located: $C'_{jq,t}=p_j C_{jq,t}+(1-p_j) \bar{C}_{jg,t}$, where $p$ is a mixing fraction between 0 and 1 that was estimated as part of fitting the model.
+Due to the possibility that plants outside the mapped quadrat could contribute recruits to the focal quadrat or interact with plants in the focal quadrat, we estimated effective cover as a mixture of the observed cover, $C$, in the focal quadrat, $q$, and the mean cover, $\bar{C}$, across the spatial location group, $g$, in which the
+quadrat is located: $C'_{jq,t}=p_j C_{jq,t}+(1-p_j) \bar{C}_{jg,t}$, where $p$ is a mixing fraction between 0 and 1 that was estimated as part of fitting the model.
 

### 2017-01-31 removals/manuscript/removal_ms.tex
Fixed typos. This is the submitted version.

@@
 
-For recruitment in the IPM, the factor $ \Phi_j  = \exp{\left(\gamma_{j,t}^R +  \varphi_{jg}^R + \chi_{jm}^R + 
-\left \langle \boldsymbol{\omega}^R , \boldsymbol{\sqrt{C'_{q,t}}} \right \rangle \right) }$
+For recruitment in the IPM, the factor $ \Phi_j  =\exp{\left(\gamma_{j,t}^R +  \varphi_{jg}^R + \chi_{jh}^R + 
+\sum \limits_{m} \omega_{jm}^R \sqrt{C'_{mq,t}} \right)}$
 in eqn. (\ref{eqn:recrDataModel}) gives the expected 
@@
 We did not simulate long-term \textit{A.
-tripartita} response to grass removal because the vital rate regressions for \textit{A.
-tripartita} showed that the baseline model effectively predicted removal effects.
+tripartita} response to grass removal because the removal treatment effects in the vital rate regressions were never significantly different from zero, indicating that the baseline model effectively predicted competitive release.
 
@@
 tripartita}).
-To quantify resulting differences in pre-treatment (2011) composition, we compared neighborhood crowding at the individual plant level (the $\boldsymbol{W}$ of the growth regression) across treatments.
+To quantify resulting differences in pre-treatment (2011) composition, we compared neighborhood crowding at the individual plant level (the $W_{jm}$'s of the growth regressions) across treatments.
 \textit{A.
@@
 The black line is a 1:1 line.
-Solid symbols correspond to predictions form the baseline IBM (no removal effects) and open symbols in (B) show predictions from an IBM that includes removal treatment effects.
+Solid symbols correspond to predictions from the baseline IBM (no removal effects) and open symbols in (B) show predictions from an IBM that includes removal treatment effects.
 Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r}.
@@
 tripartita} cover?
-To explore this possibility, we increased \text{Artemisia} cover in the baseline simulation by increasing average \textit{A.
+To explore this possibility, we increased \textit{A.
+tripartita} cover in the baseline simulation by increasing average \textit{A.
 tripartita} fecundity.
@@
 \centering
-\caption{Summary of fixed effects for the \textit{H. comata} survival model (see Table \ref{ARTRsurvival} for an explanation of coefficient names
-and column headers).} 
+\caption{Summary of fixed effects for the \textit{H. comata} survival model. See Table \ref{ARTRsurvival} for an explanation of coefficient names
+and column headers.} 
 \label{HECOsurvival}
@@
 \centering
-\caption{Summary of fixed effects for the \textit{P. secunda} survival model (see Table \ref{ARTRsurvival} for an explanation of coefficient names
-and column headers).} 
+\caption{Summary of fixed effects for the \textit{P. secunda} survival model. See Table \ref{ARTRsurvival} for an explanation of coefficient names
+and column headers.} 
 \label{POSEsurvival}
@@
 \centering
-\caption{Summary of fixed effects for the \textit{P. spicata} survival model (see Table \ref{ARTRsurvival} for an explanation of coefficient names
-and column headers).} 
+\caption{Summary of fixed effects for the \textit{P. spicata} survival model. See Table \ref{ARTRsurvival} for an explanation of coefficient names
+and column headers.} 
 \label{PSSPsurvival}
@@
 \centering
-\caption{Summary of fixed effects for the \textit{H. comata} growth model (see Table \ref{ARTRgrowth} for an explanation of coefficient names
-and column headers).} 
+\caption{Summary of fixed effects for the \textit{H. comata} growth model. See Table \ref{ARTRgrowth} for an explanation of coefficient names
+and column headers.} 
 \label{HECOgrowth}
@@
 \centering
-\caption{Summary of fixed effects for the \textit{P. secunda} growth model (see Table \ref{ARTRgrowth} for an explanation of coefficient names
-and column headers).} 
+\caption{Summary of fixed effects for the \textit{P. secunda} growth model. See Table \ref{ARTRgrowth} for an explanation of coefficient names
+and column headers.} 
 \label{POSEgrowth}
@@
 \centering
-\caption{Summary of fixed effects for the \textit{P. spicata} growth model (see Table \ref{ARTRgrowth} for an explanation of coefficient names
-and column headers).} 
+\caption{Summary of fixed effects for the \textit{P. spicata} growth model. See Table \ref{ARTRgrowth} for an explanation of coefficient names
+and column headers.} 
 \label{PSSPgrowth}
@@
 tripartita removal} data (the ``inARTR" coefficient).
-See Table \ref{ARTRgrowth} for an explanation of other coefficient names and column headers).
+See Table \ref{ARTRgrowth} for an explanation of other coefficient names and column headers.
 }
@@
 tripartita} removal data (the ``inARTR" coefficient).
-See Table \ref{ARTRgrowth} for an explanation of other coefficient names and column headers).}
+See Table \ref{ARTRgrowth} for an explanation of other coefficient names and column headers.}
 \label{table:PSSPgrowth-inARTR}
@@
 secunda} model with treatment*year effects (the ``trtYears*" coefficients).
-See Table \ref{ARTRgrowth} for an explanation of other coefficient names and column headers).}
+See Table \ref{ARTRgrowth} for an explanation of other coefficient names and column header.}
 \label{table:POSEgrowth-trtYears}
@@
 spicata} model with treatment*year effects (the ``trtYears*" coefficients).
-See Table \ref{ARTRgrowth} for an explanation of other coefficient names and column headers).}
+See Table \ref{ARTRgrowth} for an explanation of other coefficient names and column headers.}
 \label{table:PSSPgrowth-trtYears}
@@
 \centering
-\caption{Summary of fixed effects for the recruitment model (symbols correspond to Eqns. \ref{eqn:recrDataModel} and \ref{eqn:recrProcessModel})} 
+\caption{Summary of fixed effects for the recruitment model (symbols correspond to Eqns. \ref{eqn:recrDataModel} and \ref{eqn:recrProcessModel}).} 
 \label{table:recruitment}

### 2017-04-20 removals/manuscript/removal_ms.tex
More revisions (formatting appendices)

@@
 
-\section*{Summary}
-
-\begin{enumerate}
-\item A previous study of four dominant species from a sagebrush steppe community used historical data sets and multispecies population models to demonstrate weak interspecific interactions.
+\section*{Abstract}
+
+A previous study of four dominant species from a sagebrush steppe community used historical data sets and multispecies population models to demonstrate weak interspecific interactions.
 The models predict little competitive release following species removals.
-\item We conducted a removal experiment to test this prediction.
+We conducted a removal experiment to test our model's ability to predict the outcome of interspecific interactions.
 We established new quadrats in the same location where the historical data were collected.
@@
 We modeled survival, growth and recruitment as functions of local neighborhood species composition along with an indicator variable for removal treatment.
-If our ``baseline" model, which accounts for local plant-plant interactions, explains responses to removal, then the fitted removal indicator effect should be non-signficant.
-\item  For survival and recruitment, the removal treatment effects were never significantly positive, indicating that competitive release was not underestimated by the interspecific effects included in the baseline model.
-For  \textit{Poa secunda} recruitment, the removal treatment was actually negative, indicating that our baseline model overestimated competitive release.
-\item For growth, the removal treatment effect was significant and positive for two species, \textit{Poa secunda} and \textit{Pseudoroegneria spicata}, indicating that the baseline model underestimated competitive release.
+If our ``baseline" model, which accounts for local plant-plant interactions, explains responses to removal, then the fitted removal indicator effect should be non-significant.
+For survival and recruitment, the removal treatment effects were never significantly positive, indicating that competitive effects were not underestimated by the baseline model.
+For  \textit{Poa secunda} recruitment, the removal treatment was actually negative, indicating that our baseline model overestimated competitive effects.
+For growth, the removal treatment effect was significant and positive for two species, \textit{Poa secunda} and \textit{Pseudoroegneria spicata}, indicating that the baseline model underestimated competitive effects.
 However, including information about the location of these grass individuals with respect to removed \textit{A.
 tripartita} failed to improve the growth regressions, raising questions about the mechanisms driving the positive response to removal.
-\item For three species, individual based models and integral projections models based on the vital rate regressions showed that removal treatment effects had little impact on population growth rates or simulated equilibrium cover following a species removal.
+For three species, individual based models and integral projections models based on the vital rate regressions showed that removal treatment effects had little impact on population growth rates or simulated equilibrium cover following a species removal.
 For \textit{P.
@@
 tripartita} removal projected greater competitive release than the baseline models.
-\item \emph{Synthesis}: Although models based on observational data did not perfectly predict species responses to removal, the experimental results increase our confidence that competitive release is in fact small for at least three of our four study species.
-The implication is that interspecific resource competition is not a primary factor determining the dynamics of this community.
+The ability of models based on observational data to predict population responses to experimental removals increased our confidence that competitive release is in fact small for at least three of our four study species.
+The implication is that interspecific resource competition is not a primary factor determining the structure and dynamics of this community.
 The success of the models also supports the use of multispecies models, based on long-term observational studies, to characterize competitive interactions in a community.
-
-\end{enumerate}
 
@@
 Neutral theory \citep{hubbell_unified_2001} goes even further, assuming that interspecific competition is exactly as strong as intraspecific competition.
-A growing list of studies has emphasized the role of interspecific facilitation in some plant communities \citep{he_global_2013,brooker_facilitation_2008}, but it is unclear whether this research has overturned the default assumption that plants compete intensely with other species.
+Although a growing list of studies has emphasized the role of interspecific facilitation in some plant communities \citep{he_global_2013,brooker_facilitation_2008}, it is unclear whether this research has overturned the default assumption that plants compete intensely with other species.
 
@@
 An alternative approach to studying competitive release is to fit multispecies population models that estimate species interactions using long-term, observational data, and then simulate population responses to removals.
-Applying this approach to arid and semiarid plant communities, we found evidence of strong intraspecific limitation and very weak interspecific interactions, leading to very stable coexistence and implying little competitive release \citep{adler_coexistence_2010,chu_large_2015}.
-Since we only studied the most common, co-occurring species, we expected to find evidence of stable coexistence, but we were surprised that interspecific interactions and population responses to the loss of dominant species were so weak.
+Applying this approach to arid and semiarid plant communities, we found evidence of strong intraspecific limitation and weak interspecific interactions, leading to very stable coexistence and implying little competitive release \citep{adler_coexistence_2010,chu_large_2015}.
+In fact, at our sagebrush steppe study site  \citep{adler_coexistence_2010}, our model predicts removal of sagebrush would actually decrease the equilibrium cover of the three dominant grass species by 0.8 to 5.5\%, reflecting the facilitative effect of sagebrush on grass recruitment and very weak competitive effects on growth and survival.
+Since we only studied the most common, co-occurring species, we expected to find evidence of stable coexistence, but the predicted negative effect of sagebrush removal on grass cover is surprising given the long history of sagebrush eradication to increase grass forage production in sagebrush rangelands \citep{Cook1963, Mueggler1958, Robertson1947}.
 Compared to traditional field experiments, our approach has the advantage of projecting full life-cycle responses over long time periods, but its reliance on observational data sets limits the strength of our inference.
-We would have more confidence in our results if they were confirmed by an experiment.
-
-To test whether our models, which are based on observational data, underestimate the strength of interspecific competition and competitive release, we conducted a removal experiment in a sagebrush steppe.
+We would have more confidence in our model's projections of competitive release if we could experimentally confirm the ability of our models to describe interspecific interactions.
+
+To test whether our sagebrush community model, which is based on observational data, underestimates the strength of interspecific competition and competitive release, we conducted a removal experiment in a sagebrush steppe.
 We studied the responses of three perennial grass species to removal of sagebrush, and the response of sagebrush to removal of all grasses.
+The primary purpose of the experiment was not to directly measure competitive release, which is difficult to do over a short time period in a semiarid ecosystem with low and variable plant cover, but rather to validate the parameters in our model that represent key interspecific interactions and ultimately determine the magnitude of competitive release.
 We addressed two research questions: First, are the survival, growth or recruitment of our target species significantly different in control vs.
@@
 Our second question integrates across the full life cycle: how do short and long-term population-level responses to removals compare to projections from multispecies population models based on observational data?
-If models based on observational data adequately predict competitive release, the population dynamics they project should be similar to projections from models that include information from the experimental removal treatments, a result which would validate the use of models and long-term data sets to study interspecific interactions.
+If models based on observational data adequately describe facilitative and competitive interactions, the population dynamics they project should be similar to projections from models that include information from the experimental removal treatments, a result which would validate the use of models and long-term data sets to study interspecific interactions.
 
@@
 We considered the removal treatment effect to be significant when the 95\% confidence interval on the estimated coefficient did not overlap zero.
+Because variance in residuals of plots with low cover was high, we explored fitting by weighted least squares, to account for the apparent heteroscedasiticity.
+The results of these tests were qualitatively similar, so opted to only report the simpler analysis.
 
@@
 Our regression models for survival, growth and recruitment follow previous work \citep{adler_coexistence_2010,chu_large_2015}.
-We model the survival probability of an individual genet as a function of genet size, the neighborhood-scale crowding experienced by the genet from both conspecific and heterospecific genets, temporal variation among years, and permanent spatial variation among groups of quadrats (`group'; here means a set of nearby quadrats located within one pasture or grazing exclosure).
-Where the current analysis departs from previous work is the inclusion of a categorical removal treatment effect.
+We model the survival probability and the growth of individual genets as a function of genet size, the neighborhood-scale crowding experienced by the genet from both conspecific and heterospecific genets, temporal variation among years, and permanent spatial variation among groups of quadrats (`group'; here means a set of nearby quadrats located within one pasture or grazing exclosure).
+The current analysis departs from previous work with the inclusion of a categorical removal treatment effect.
 If the conspecific and heterospecific crowding effects explain demographic rates in removal treatment plots as well as they do in control plots, then there will be no need for the removal treatment effect, and the estimated coefficient will be small and not significantly different from zero.
@@
 
-Formally, we modeled the survival probability, $S$, of genet $i$ in species $j$ (which is in quadrat group $g$ and removal treatment $h$), 
+More formally, we modeled the survival probability, $S$, of genet $i$ in species $j$ (which is in quadrat group $g$ and removal treatment $h$), 
 from time $t$ to $t+1$  as
@@
 secunda}, as well as two covariates representing total crowding from (1) all other perennial grasses and shrubs, which were mapped as polygons, and (2) all forb species, which were mapped as points.
-
-We modeled the crowding experienced by a focal genet in each year $t$ as a function of the distance to and size of neighbor genets. In previous work, 
-we assumed that the decay of crowding with neighbor distance followed a Gaussian function \citep{chu_large_2015}, but here we use a data-driven 
-approach \citep{teller_linking_2016}. We model the crowding experienced by genet $i$ of species $j$ from neighbors of species $m$ as the sum of neighbor 
-areas across a set of concentric annuli, $k$, centered at the plant,
-\begin{equation}
-w_{ijmk,t} = F_{jm}(d_{k})A_{ijmk,t}     
-\label{eqn:wik}
-\end{equation}
-where $F_{jm}$ is the competition kernel (described below) for effects of species $m$ on species $j$, 
-$d_{k}$ is the average of the inner and outer radii of annulus $k$, 
-and $A_{ijmk,t}$ is the total area of genets of species $m$ in annulus $k$ around the focal genet in year $t$. The total crowding on 
-the focal genet exerted by species $m$ is
-\begin{equation}
-W_{ijm,t}  =\sum_k {w_{ijmk,t}} .
-\label{eqn:wijm}
-\end{equation} 
-Note that $W_{ijj,t}$ gives intraspecific crowding. The $W$'s are then 
-included as covariates in the survival (\ref{eqn:survReg}) and growth (\ref{eqn:growReg}) regressions.
-
-We assume that competition kernels $F_{jm}(d)$ are non-negative and decreasing, so that distant plants have less effect 
-than close plants. Otherwise, we let the data dictate the shape of the kernel by fitting a spline model 
-using the methods of \citep{teller_linking_2016}. The shape of $F_{jm}$ is determined by a set of spline basis coefficients $\vec{b}_{jm}$
-and a smoothing parameter $\eta$ that controls the complexity of the fitted kernel. 
-Demographic models such as \eqref{eqn:survReg} then have $\gamma$, $\varphi$, $\chi$ , 
-$\beta$, $\boldsymbol{\omega}$, $\vec{b}$ and $\eta$ as parameters to be fitted. We implemented this in the statistical computing environment, \texttt{R}, 
-by making the spline coefficients and $\eta$ the arguments of an objective function that calculates  each $w_{ijmk,t}$ value using the input spline coefficients, 
-calls the model-fitting functions \texttt{lmer} and/or \texttt{glmer} to fit the other parameters in the survival and growth regressions, 
-and returns an approximate AIC value and model degrees of freedom ($df$) for survival and growth combined. We used the $\vec{b}$ values at the smoothest 
-(smallest $df$) local minimum of AIC as a function of $df$, as in \cite{teller_linking_2016}. This approach assumes that one measure of crowding affects 
-survival and growth. In addition, for fitting the kernels we assumed that survival and growth depended only on intraspecific crowding, and thus only fitted the
-within-species competition kernels $F_{jj}$.
-Based on previous work with parametric kernels, \citep{adler_coexistence_2010}, we know that interspecific effects are rarely significantly different from zero, so there is not enough signal to fit interspecific kernels nonparametrically.
-Therefore, we set all $F_{mj}$ equal to $F_{jj}$, meaning that
-the within-species competition kernel for species $j$ is also used to determine the distance-dependence of species $j$ neighbors on all other species. 
-We used data from all historical plots and contemporary control-treatment plots to estimate the competition kernels, which are shown in Fig. \ref{fig:CompKernels}.
-
-Once we had estimated the competitions kernels, we used them to calculate the $w_{ijmk,t}$ values for each individual in each year, 
-and fitted the full survival and growth regressions, which include the interspecific interaction coefficients, $\omega_{jm}$. 
-All genets in a quadrat were included in calculating the $w$'s, but plants located within 5 cm of quadrat edges were not used in fitting the regressions. 
-We fit the models using the \texttt{R-INLA} package for R \citep{rue_approximate_2009}. We used the same approach when exploring variations 
-on the growth regressions, such as the addition of a year-by-treatment interaction or information about the locations of 
-individual grasses relative to removed \textit{A. tripartita}. We also compared the species composition of the control and removal plots at both 
-the quadrat and neighborhood ($W$) scales.
+Appendix A provides a detailed explanation of how we modeled crowding based on the size of and distance to neighboring plants around each focal genet.
 
@@
 Viewing all annual growth rates across species and years shows the same pattern of underprediction for some species, but also reveals a strong correlation between observations and predictions in control plots (Fig.
-\ref{fig:ObsPred1to1}A).
+CUT.
 In removal plots, underprediction of growth rates in certain years was larger, especially for \textit{H.
 comata} and \textit{P.
-spicata}; including removal treatment effects only slightly reduced this bias (Fig.
-\ref{fig:ObsPred1to1}B).
+spicata}; including removal treatment effects only slightly reduced this bias (CUT).
 For the version of the model with maximum removal treatment effects (removal coefficients set to the 95\% confidence limit furthest from zero), the patterns were qualitatively similar, with a somewhat greater contrast between predictions of the baseline and removal models for \textit{P.
@@
 spicata} (Figs.
-\ref{fig:IBM1step-maxCI},\ref{fig:ObsPred1to1-maxCI}).
+\ref{fig:IBM1step-maxCI},CUT).
 
@@
  \centering
- \includegraphics[width=1\textwidth]{cover_projections_1step}
+ %\includegraphics[width=1\textwidth]{cover_projections_1step}
  \caption{Observed and predicted cover for each of the four modeled species.
@@
  \label{fig:IBM1step}
- \end{figure}
-
-  \begin{figure}[tbp]
-  \centering
-  \includegraphics[width=1\textwidth]{cover_change_1to1}
-  \caption{Comparison of observed and predicted annual per capita growth rates across species in (A) control and (B) removal treatments.
-The black line is a 1:1 line.
-Solid symbols correspond to predictions from the baseline IBM (no removal effects) and open symbols in (B) show predictions from an IBM that includes removal treatment effects.
-Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r}.
-}
-  \label{fig:ObsPred1to1}
  \end{figure}
@@
 \setcounter{table}{0}
-\renewcommand{\theequation}{SI.\arabic{equation}}
-\renewcommand{\thetable}{SI-\arabic{table}}
-\renewcommand{\thefigure}{SI-\arabic{figure}}
-\renewcommand{\thesection}{Section SI.\arabic{section}}
-
-\centerline{\Large \textbf{Supporting Information }}
+\renewcommand{\theequation}{A-\arabic{equation}}
+\renewcommand{\thetable}{B-\arabic{table}}
+\renewcommand{\thefigure}{C-\arabic{figure}}
+\renewcommand{\thesection}{A.\arabic{section}}
+
+\centerline{\Large \textbf{Appendices}}
 \centerline{Adler et al., ``Weak interspecific interactions''}
@@
 
-\section{Supplementary Methods} \label{suppMethods}
-
-\subsection{Interspecific covariance in local crowding} 
+\section*{Appendix A. Detailed Methods} \label{suppMethods}
+
+\subsection*{Modeling local crowding}
+
+We modeled the crowding experienced by a focal genet in each year $t$ as a function of the distance to and size of neighbor genets. In previous work, 
+we assumed that the decay of crowding with neighbor distance followed a Gaussian function \citep{chu_large_2015}, but here we use a data-driven 
+approach \citep{teller_linking_2016}. We model the crowding experienced by genet $i$ of species $j$ from neighbors of species $m$ as the sum of neighbor 
+areas across a set of concentric annuli, $k$, centered at the plant,
+\begin{equation}
+w_{ijmk,t} = F_{jm}(d_{k})A_{ijmk,t}     
+\label{eqn:wik}
+\end{equation}
+where $F_{jm}$ is the competition kernel (described below) for effects of species $m$ on species $j$, 
+$d_{k}$ is the average of the inner and outer radii of annulus $k$, 
+and $A_{ijmk,t}$ is the total area of genets of species $m$ in annulus $k$ around the focal genet in year $t$. The total crowding on 
+the focal genet exerted by species $m$ is
+\begin{equation}
+W_{ijm,t}  =\sum_k {w_{ijmk,t}} .
+\label{eqn:wijm}
+\end{equation} 
+Note that $W_{ijj,t}$ gives intraspecific crowding. The $W$'s are then 
+included as covariates in the survival (\ref{eqn:survReg}) and growth (\ref{eqn:growReg}) regressions.
+
+We assume that competition kernels $F_{jm}(d)$ are non-negative and decreasing, so that distant plants have less effect 
+than close plants. Otherwise, we let the data dictate the shape of the kernel by fitting a spline model 
+using the methods of \citep{teller_linking_2016}. The shape of $F_{jm}$ is determined by a set of spline basis coefficients $\vec{b}_{jm}$
+and a smoothing parameter $\eta$ that controls the complexity of the fitted kernel. 
+Demographic models such as \eqref{eqn:survReg} then have $\gamma$, $\varphi$, $\chi$ , 
+$\beta$, $\boldsymbol{\omega}$, $\vec{b}$ and $\eta$ as parameters to be fitted. We implemented this in the statistical computing environment, \texttt{R}, 
+by making the spline coefficients and $\eta$ the arguments of an objective function that calculates  each $w_{ijmk,t}$ value using the input spline coefficients, 
+calls the model-fitting functions \texttt{lmer} and/or \texttt{glmer} to fit the other parameters in the survival and growth regressions, 
+and returns an approximate AIC value and model degrees of freedom ($df$) for survival and growth combined. We used the $\vec{b}$ values at the smoothest 
+(smallest $df$) local minimum of AIC as a function of $df$, as in \cite{teller_linking_2016}. This approach assumes that one measure of crowding affects 
+survival and growth. In addition, for fitting the kernels we assumed that survival and growth depended only on intraspecific crowding, and thus only fitted the
+within-species competition kernels $F_{jj}$.
+Based on previous work with parametric kernels, \citep{adler_coexistence_2010}, we know that interspecific effects are rarely significantly different from zero, so there is not enough signal to fit interspecific kernels nonparametrically.
+Therefore, we set all $F_{mj}$ equal to $F_{jj}$, meaning that
+the within-species competition kernel for species $j$ is also used to determine the distance-dependence of species $j$ neighbors on all other species. 
+We used data from all historical plots and contemporary control-treatment plots to estimate the competition kernels, which are shown in Fig. \ref{fig:CompKernels}.
+
+Once we had estimated the competitions kernels, we used them to calculate the $w_{ijmk,t}$ values for each individual in each year, 
+and fitted the full survival and growth regressions, which include the interspecific interaction coefficients, $\omega_{jm}$. 
+All genets in a quadrat were included in calculating the $w$'s, but plants located within 5 cm of quadrat edges were not used in fitting the regressions. 
+We fit the models using the \texttt{R-INLA} package for R \citep{rue_approximate_2009}. We used the same approach when exploring variations 
+on the growth regressions, such as the addition of a year-by-treatment interaction or information about the locations of 
+individual grasses relative to removed \textit{A. tripartita}. We also compared the species composition of the control and removal plots at both 
+the quadrat and neighborhood ($W$) scales.
+
+\subsection*{Interspecific covariance in local crowding} 
 We explored interspecific covariance in local crowding experienced by individual plants, by regressing the $W$ values exerted by one neighbor species, the response variable, against the $W$ values of all other species, the independent variables.
@@
 
-\subsection{Mean field approximation of local crowding for the IPM} 
+\subsection*{Mean field approximation of local crowding for the IPM} 
 \citet{adler_coexistence_2010} developed a mean field approximation for local crowding when the
@@
 \newpage
-\section{Additional Tables}
+
+\renewcommand{\thesection}{B.\arabic{section}}
+
+\section*{Appendix B. Supplementary Tables}
 
@@
 
-\section{Additional Figures}
+\renewcommand{\thesection}{C.\arabic{section}}
+
+\section*{Appendix C. Supplementary Figures}
 
@@
   \centering
-   \includegraphics[width=1\textwidth]{cover_change_1to1_maxCI}
-   \caption{Comparison of observed and predicted annual per capita growth rates across species in (A) control and (B) removal treatments assuming maximum effects of removals.
-The black line is a 1:1 line.
-Solid symbols correspond to predictions form the baseline IBM (no removal effects) and open symbols in (B) show predictions from an IBM that includes removal treatment effects.
-Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r}.
-}
-   \label{fig:ObsPred1to1-maxCI}
-   \end{figure}
-
-  \begin{figure}[tbp]
-  \centering
   \includegraphics[width=1\textwidth]{boxplots-maxCI}

### 2017-04-21 removals/manuscript/removal_ms.tex
Intermediate save of in-progress revisions.

@@
 An alternative approach to studying competitive release is to fit multispecies population models that estimate species interactions using long-term, observational data, and then simulate population responses to removals.
-Applying this approach to arid and semiarid plant communities, we found evidence of strong intraspecific limitation and weak interspecific interactions, leading to very stable coexistence and implying little competitive release \citep{adler_coexistence_2010,chu_large_2015}.
-In fact, at our sagebrush steppe study site  \citep{adler_coexistence_2010}, our model predicts removal of sagebrush would actually decrease the equilibrium cover of the three dominant grass species by 0.8 to 5.5\%, reflecting the facilitative effect of sagebrush on grass recruitment and very weak competitive effects on growth and survival.
-Since we only studied the most common, co-occurring species, we expected to find evidence of stable coexistence, but the predicted negative effect of sagebrush removal on grass cover is surprising given the long history of sagebrush eradication to increase grass forage production in sagebrush rangelands \citep{Cook1963, Mueggler1958, Robertson1947}.
+Applying this approach to arid and semiarid plant communities, we found evidence of strong intraspecific limitation and weak interspecific interactions, leading to very stable coexistence and implying little competitive release \citep{adler_coexistence_2010,chu_large_2015}.\new{ In fact, at our sagebrush steppe study site  \citep{adler_coexistence_2010}, our model predicts removal of sagebrush would actually decrease the equilibrium cover of the three dominant grass species by 0.8 to 5.5\%, reflecting the facilitative effect of sagebrush on grass recruitment and very weak competitive effects on growth and survival.
+Since we only studied the most common, co-occurring species, we expected to find evidence of stable coexistence, but the predicted negative effect of sagebrush removal on grass cover is surprising given the long history of sagebrush eradication to increase grass forage production in
+ sagebrush rangelands \citep{Cook1963, Mueggler1958, Robertson1947}.} % end new
 Compared to traditional field experiments, our approach has the advantage of projecting full life-cycle responses over long time periods, but its reliance on observational data sets limits the strength of our inference.
@@
 
-To test whether our sagebrush community model, which is based on observational data, underestimates the strength of interspecific competition and competitive release, we conducted a removal experiment in a sagebrush steppe.
+To test whether our sagebrush community model, which is based on observational data, \new{underestimates the strength of interspecific competition and competitive release}, we conducted a removal experiment in a sagebrush steppe.
 We studied the responses of three perennial grass species to removal of sagebrush, and the response of sagebrush to removal of all grasses.
-The primary purpose of the experiment was not to directly measure competitive release, which is difficult to do over a short time period in a semiarid ecosystem with low and variable plant cover, but rather to validate the parameters in our model that represent key interspecific interactions and ultimately determine the magnitude of competitive release.
-We addressed two research questions: First, are the survival, growth or recruitment of our target species significantly different in control vs.
+\new{The primary purpose of the experiment was not to directly measure competitive release, which is difficult to do over a short time period in a semiarid ecosystem with low and variable plant cover, but rather to validate the parameters in our model that represent key interspecific interactions and ultimately determine the magnitude of competitive release.} We addressed two research questions: First, are the survival, growth or recruitment of our target species significantly different in control vs.
 removal treatments after accounting for differences in local species composition?
@@
 
-\subsection*{Statistical models of vital rates}
-
+\subsection*{Overview of statistical models of vital rates}
+
+\new{
 Our regression models for survival, growth and recruitment follow previous work \citep{adler_coexistence_2010,chu_large_2015}.
-We model the survival probability and the growth of individual genets as a function of genet size, the neighborhood-scale crowding experienced by the genet from both conspecific and heterospecific genets, temporal variation among years, and permanent spatial variation among groups of quadrats (`group'; here means a set of nearby quadrats located within one pasture or grazing exclosure).
 The current analysis departs from previous work with the inclusion of a categorical removal treatment effect.
-If the conspecific and heterospecific crowding effects explain demographic rates in removal treatment plots as well as they do in control plots, then there will be no need for the removal treatment effect, and the estimated coefficient will be small and not significantly different from zero.
-However, if demographic rates in removal plots differ from control plots even after accounting for crowding effects, then the removal treatment effect will be significant.
-For example, we might find that, holding crowding constant, grasses grow more quickly in the removal plots than in the control plots.
-This result would indicate that our ``baseline model'', a model including all estimated parameters except for the removal treatment effect, underestimates competitive release.
-We refer to the model that includes the estimated removal treatment effect as the ``removal model".
-
-More formally, we modeled the survival probability, $S$, of genet $i$ in species $j$ (which is in quadrat group $g$ and removal treatment $h$), 
-from time $t$ to $t+1$  as
+Although the three vital rate models differ in functional form (see Appendix A for complete model descriptions), all of them have a linear predictor, $\mu$, containing the following terms:}
+
 \begin{equation}
-\mbox{logit}(S_{ij,t+1}) = \gamma_{j,t}^S + \varphi_{jg}^S+  \chi_{jh}^S  + \beta_{j,t}^S u_{ij,t} +  
-\sum \limits_{m} \omega_{jm}^S {W}_{ijm,t}
+\mu_ij = \gamma_j+ \varphi_{jg}+  \chi_j  + \sum \limits_{m} \omega_{jm} {W}_{ijm}
 \label{eqn:survReg}
 \end{equation}
-where $\gamma$ is a time-dependent intercept (year effect), $\varphi$ is the coefficient for the 
-effect of spatial location group, $\chi$ is the removal treatment effect,and  $\beta$ is the coefficient that 
-represents the effect of log genet size, $u$, on survival. 
-$W_{ijm,t}$ is the crowding of individual $i$ in species $j$ by neighbors of species $m$ (explained below), and 
-$\omega_{jm}$ is an interaction coefficient which determines the impact of crowding by species $m$ on the focal species $j$.
-
-Our growth model has a similar structure. The change in genet size from time $t$ to $t+1$ , conditional on survival, is given by:
-\begin{equation}
-u_{ij,t+1} = \gamma_{j,t}^G + \varphi_{jg}^G+  \chi_{jh}^G  + \beta_{j,t}^G u_{ij,t} + 
-\sum \limits_{m} \omega_{jm}^G {W}_{ijm,t} + \varepsilon_{ij,t}^G .
-\label{eqn:growReg}
-\end{equation}
-To capture the non-constant error variance in growth, we modeled the variance $\varepsilon$ about the growth 
-curve (\ref{eqn:growReg}) as a nonlinear function of predicted genet size:
-\begin{equation}
-Var(\varepsilon_{ij,t}^G) = a \exp(b\hat{u}_{ij,t+1}).
-\label{eqn:growVar}
-\end{equation}
-where $\hat{u}_{ij,t+1}$ is the right-hand side of (\ref{eqn:growReg}) without $\varepsilon$.
-
-Our model includes crowding from the four dominant species, \textit{A.
-tripartita}, \textit{H.
-comata}, \textit{P.
-spicata}, and \textit{P.
-secunda}, as well as two covariates representing total crowding from (1) all other perennial grasses and shrubs, which were mapped as polygons, and (2) all forb species, which were mapped as points.
-Appendix A provides a detailed explanation of how we modeled crowding based on the size of and distance to neighboring plants around each focal genet.
-
-We model recruitment at the quadrat level rather than at the individual genet level because the mapped data do not allow 
-us to determine which recruits were produced by which potential parent genets. We assume that the number of individuals, $y$, of species $j$ recruiting at 
-time $t+1$ in the location $q$ follows a negative binomial distribution:
-\begin{equation}
-y_{jq,t+1}= NegBin(\lambda_{jq,t+1},\theta) 	   
-\label{eqn:recrDataModel}
-\end{equation}
-where $\lambda$ is the mean and $\theta$ is the size parameter (\texttt{dnegbin(p,$\theta$)} in the BUGS language, with $p=\theta/(\theta+\lambda)$). 
-In turn, $\lambda$ depends on the composition of the quadrat in the previous year:
-\begin{equation}
-\lambda_{jq,t+1} = C'_{jq,t} \exp{\left(\gamma_{j,t}^R +  \varphi_{jg}^R + \chi_{jh}^R + 
-\sum \limits_{m} \omega_{jm}^R \sqrt{C'_{mq,t}} \right)}
-\label{eqn:recrProcessModel}
-\end{equation}
-where the superscript $R$ refers to Recruitment, $C'_{jq,t}$ is the `effective cover' (cm$^2$) of species $j$ in quadrat $q$ at time $t$, $\gamma$ 
-is a time-dependent intercept, $\varphi$ is a coefficient for the effect of spatial location group,
-$\chi$ is the removal treatment effect, and $\omega_{jm}$ is the coefficient that determines the effect of species $m$ cover on recruitment by species $j$. 
-Eqn (\ref{eqn:recrProcessModel}) is essentially the Ricker model for discrete time population growth \citep{ricker_stock_1954}, except that we use the square root of cover in the density-dependent term.
-We previously used a Ricker equation \citep{adler_coexistence_2010}, but using the square-root transformation instead gave a lower Deviance Information Criterion (DIC) value \citep{spiegelhalter_bayesian_2002} for the recruitment model.
-Following previous work \citep{adler_coexistence_2010}, we treated year and spatial group variables as random factors, allowing intercepts to vary among years and spatial locations.
-
-Due to the possibility that plants outside the mapped quadrat could contribute recruits to the focal quadrat or interact with plants in the focal quadrat, we estimated effective cover as a mixture of the observed cover, $C$, in the focal quadrat, $q$, and the mean cover, $\bar{C}$, across the spatial location group, $g$, in which the
-quadrat is located: $C'_{jq,t}=p_j C_{jq,t}+(1-p_j) \bar{C}_{jg,t}$, where $p$ is a mixing fraction between 0 and 1 that was estimated as part of fitting the model.
-
-We ran Markov Chain Monte Carlo (MCMC) simulations in WinBUGS 1.4 \citep{lunn_winbugs_2000} to estimate the recruitment model parameters.
-Each model was run for 20,000 iterations and two independent chains with different initial values for parameters.
-We discarded the initial 10,000 samples.
-Convergence was observed graphically for all parameters, and confirmed by the Brooks-Gelman statistic \citep{brooks_general_1998}.
-For all three vital rate models, we considered the effect of removal treatments, or other coefficients of interest, to be significant when the 95\% confidence interval (or posterior interval) on the estimate of $\chi$ did not overlap zero.
+
+\new{where $i$ refers to an individual genet (for survival and growth) or quadrat (for recruitment) of species $j$, $\gamma$ is a time-dependent intercept (year effect), $\varphi$ is the coefficient for the effect of spatial location (`$g$' represents a group of nearby quadrats located within one pasture or grazing exclosure), $\chi$ is the removal treatment effect,
+$W_{ijm,t}$ is the crowding on species $j$ exerted by neighbors of species $m$ at the individual (survival and growth) or
+quadrat (recruitment) scale, and $\omega_{jm}$ is an interaction coefficient which determines the impact of crowding by species $m$ on the 
+focal species $j$. }
+
+\new{If the conspecific and heterospecific crowding effects, the $\omega_{jm}$'s, explain demographic rates in removal treatment plots as well as they do in control plots, then there will be no need for the removal treatment effect, and the estimated coefficient will be small and not significantly different from zero.
+However, if demographic rates in removal plots differ from control plots even after accounting for crowding effects, then the removal treatment effect will be significant.
+For example, we might find that, holding crowding constant, grasses grow more quickly in the sagebrush removal plots than in the control plots.
+This result would indicate that predictions from a ``baseline model'', a model including all estimated parameters except for the removal treatment effect, underestimates competitive release.
+We refer to the model that includes the estimated removal treatment effect as the ``removal model".
+Note that these are not estimated as separate models; we estimate all parameters together, but when we generate predictions from the baseline model, we simply set the removal treatment effects to zero.} % end new
 
@@
 Because the main goal of this comparison is to look for systematic deviations between model predictions and observations, we used a deterministic version of the model (e.g., recruitment is the $\lambda$ of (\ref{eqn:recrProcessModel}), rather than a random draw from a Poisson distribution with a mean of $\lambda$).
+\comment{STEVE DO WE NEED TO EXPLAIN OUR JENSEN'S INEQUALITY CORRECTION HERE?}
 We first applied the recruitment regression to determine the expected number of new recruits at time $t+1$, which we then multiplied by the mean size of recruits observed for the focal species to get expected cover of new recruits.
@@
 Because we do not carry these simulations forward beyond one time step, we do not need to assign locations to the new recruits.
-
-For control plots, we used a version of the model with all removal treatment effects set to zero.
-For removal treatment plots, we generated predictions from two versions of the model, with and without treatment effects set to zero.
-We calculated the mean across quadrats of observed log cover change, $log(Cover_{t+1}/Cover_t)$ and compared it to the mean across quadrats of the predicted cover change.
+For control plots, we used the baseline version of the model with all removal treatment effects set to zero.
+For removal treatment plots, we generated predictions from both the baseline and removal versions of the model, with and without treatment effects set to zero, respectively.
 
@@
 Perhaps large uncertainty in the estimates of these effects makes it difficult to reject the null hypothesis represented by our baseline model.
-To address this possibility, we also simulated a version of the model in which we replaced the mean estimate of each removal treatment effect with the parameter value at either the 5\% or 95\% confidence interval, whichever was farthest from zero.
+To address this possibility, we also simulated a version of the model in which we replaced the mean estimate of each removal treatment effect with the parameter value at either the 5\% or 95\% confidence limit, whichever was farthest from zero.
 This version of the model shows the population-level changes we would expect given maximum removal effects.
@@
 \subsection*{Integral projection models}
+
+\comment{Should we move much of this to Appendix A?}
 
@@
 mean-field approximation that captures the essential features of spatially-explicit neighborhood interactions \citep{adler_coexistence_2010}. 
-We found that in both the observed data and IBM simulations, heterospecific individuals were approximately randomly distributed with respect to each other, 
-but conspecific individuals displayed a non-random, size-dependent patterns: small genets were randomly distributed, while large genets 
+We found that in both the observed data and IBM simulations, heterospecific individuals were approximately randomly distributed with respect to each other, but conspecific individuals displayed a non-random, size-dependent patterns: small genets were randomly distributed, while large genets
 were segregated from each other. The overdispersion of large conspecific genets was introduced into the IPM 
@@
 For all IPM simulations, we set the spatial random effects to zero, meaning that we are simulating dynamics on a hypothetical, average site.
-The first scenario featured all four species and our ``baseline'' model for control plots (no removal effects).
+The first scenario featured all four species and our baseline model for control plots (no removal effects).
 We initialized all species with arbitrarily low cover and ran the model for 500 time steps (years) to allow the community to reach an equilibrium.
@@
 tripartita} to zero and then repeated the procedure.
-This simulation estimates the competitive release of grasses following \textit{A.
+This simulation projects the competitive release of grasses following \textit{A.
 tripartita} removal according to the baseline model (no removal effects).
@@
 As with the IBM, we repeated these simulations using maximum removal treatment effects represented by parameters values at the 5\% or 95\% confidence limit farthest from zero.
-We did not simulate long-term \textit{A.
+We do not show simulations of long-term \textit{A.
 tripartita} response to grass removal because the removal treatment effects in the vital rate regressions were never significantly different from zero, indicating that the baseline model effectively predicted competitive release.
+\comment{Just do it?}
 
@@
 (Fig. \ref{fig:CoverTrends}), but shrub removal did appear to cause an increase in the cover of \textit{P. spicata}. 
-We tested the effects of removal on cover by analyzing year-to-year changes in (log) cover, a measure of population growth rate (Fig.
-\ref{fig:CoverChange}).
-The removal treatments had no significant effect on
-\textit{A. tripartita} (95\% confidence interval on removal treatment effect: -0.75, 0.56) or 
-\textit{P. secunda} (95\% CI: -0.41, 0.25) population growth, but did increase 
-\textit{H. comata} (95\% CI: 0.01, 0.65) and
- \textit{P. spicata} (95\% CI: 0.01, 0.40) population growth 
-(Table \ref{table:coefficients}). However, these treatment comparisons do not address our questions about whether responses to removal were 
+However, when we tested the effects of removal on cover by analyzing year-to-year changes in (log) cover, a measure of population growth rate (Fig.
+\ref{fig:CoverChange}), none of the removal treatment effects were significantly different from zero (Table \ref{table:coefficients}).
+For \textit{P.
+spicata}, the 95\% confidence interval only just contained zero (95\% CI: 0.01, 0.40), providing marginal evidence for a positive effect of removal on population growth.
+Fitting with weighted least squares to address heteroscedasticity in residuals indicated even less evidence for a positive response of  \textit{P.
+spicata} to removal.
+However, these treatment comparisons do not directly address our questions about whether responses to removal were
 adequately predicted by models fit to historical, observational data.
@@
 
+\subsection*{Statistical models of vital rates}
+
+We modeled the survival probability, $S$, of genet $i$ in species $j$ (which is in quadrat group $g$ and removal treatment $h$), 
+from time $t$ to $t+1$  as
+\begin{equation}
+\mbox{logit}(S_{ij,t+1}) = \gamma_{j,t}^S + \varphi_{jg}^S+  \chi_{jh}^S  + \beta_{j,t}^S u_{ij,t} +  
+\sum \limits_{m} \omega_{jm}^S {W}_{ijm,t}
+\label{eqn:survReg}
+\end{equation}
+where $\gamma$ is a time-dependent intercept (year effect), $\varphi$ is the coefficient for the 
+effect of spatial location group, $\chi$ is the removal treatment effect,and  $\beta$ is the coefficient that 
+represents the effect of log genet size, $u$, on survival. 
+$W_{ijm,t}$ is the crowding of individual $i$ in species $j$ by neighbors of species $m$ (explained below), and 
+$\omega_{jm}$ is an interaction coefficient which determines the impact of crowding by species $m$ on the focal species $j$.
+
+Our growth model has a similar structure. The change in genet size from time $t$ to $t+1$ , conditional on survival, is given by:
+\begin{equation}
+u_{ij,t+1} = \gamma_{j,t}^G + \varphi_{jg}^G+  \chi_{jh}^G  + \beta_{j,t}^G u_{ij,t} + 
+\sum \limits_{m} \omega_{jm}^G {W}_{ijm,t} + \varepsilon_{ij,t}^G .
+\label{eqn:growReg}
+\end{equation}
+To capture the non-constant error variance in growth, we modeled the variance $\varepsilon$ about the growth 
+curve (\ref{eqn:growReg}) as a nonlinear function of predicted genet size:
+\begin{equation}
+Var(\varepsilon_{ij,t}^G) = a \exp(b\hat{u}_{ij,t+1}).
+\label{eqn:growVar}
+\end{equation}
+where $\hat{u}_{ij,t+1}$ is the right-hand side of (\ref{eqn:growReg}) without $\varepsilon$.
+
+Our model includes crowding from the four dominant species, \textit{A.
+tripartita}, \textit{H.
+comata}, \textit{P.
+spicata}, and \textit{P.
+secunda}, as well as two covariates representing total crowding from (1) all other perennial grasses and shrubs, which were mapped as polygons, and (2) all forb species, which were mapped as points.
+The next section (\textit{Modeling local crowding}) provides a detailed explanation of how we modeled crowding based on the size of and distance to neighboring plants around each focal genet.
+
+We model recruitment at the quadrat level rather than at the individual genet level because the mapped data do not allow 
+us to determine which recruits were produced by which potential parent genets. We assume that the number of individuals, $y$, of species $j$ recruiting at 
+time $t+1$ in the location $q$ follows a negative binomial distribution:
+\begin{equation}
+y_{jq,t+1}= NegBin(\lambda_{jq,t+1},\theta) 	   
+\label{eqn:recrDataModel}
+\end{equation}
+where $\lambda$ is the mean and $\theta$ is the size parameter (\texttt{dnegbin(p,$\theta$)} in the BUGS language, with $p=\theta/(\theta+\lambda)$). 
+In turn, $\lambda$ depends on the composition of the quadrat in the previous year:
+\begin{equation}
+\lambda_{jq,t+1} = C'_{jq,t} \exp{\left(\gamma_{j,t}^R +  \varphi_{jg}^R + \chi_{jh}^R + 
+\sum \limits_{m} \omega_{jm}^R \sqrt{C'_{mq,t}} \right)}
+\label{eqn:recrProcessModel}
+\end{equation}
+where the superscript $R$ refers to Recruitment, $C'_{jq,t}$ is the `effective cover' (cm$^2$) of species $j$ in quadrat $q$ at time $t$, $\gamma$ 
+is a time-dependent intercept, $\varphi$ is a coefficient for the effect of spatial location group,
+$\chi$ is the removal treatment effect, and $\omega_{jm}$ is the coefficient that determines the effect of species $m$ cover on recruitment by species $j$. 
+Eqn (\ref{eqn:recrProcessModel}) is essentially the Ricker model for discrete time population growth \citep{ricker_stock_1954}, except that we use the square root of cover in the density-dependent term.
+We previously used a Ricker equation \citep{adler_coexistence_2010}, but using the square-root transformation instead gave a lower Deviance Information Criterion (DIC) value \citep{spiegelhalter_bayesian_2002} for the recruitment model.
+Following previous work \citep{adler_coexistence_2010}, we treated year and spatial group variables as random factors, allowing intercepts to vary among years and spatial locations.
+
+Due to the possibility that plants outside the mapped quadrat could contribute recruits to the focal quadrat or interact with plants in the focal quadrat, we estimated effective cover as a mixture of the observed cover, $C$, in the focal quadrat, $q$, and the mean cover, $\bar{C}$, across the spatial location group, $g$, in which the
+quadrat is located: $C'_{jq,t}=p_j C_{jq,t}+(1-p_j) \bar{C}_{jg,t}$, where $p$ is a mixing fraction between 0 and 1 that was estimated as part of fitting the model.
+
+We ran Markov Chain Monte Carlo (MCMC) simulations in WinBUGS 1.4 \citep{lunn_winbugs_2000} to estimate the recruitment model parameters.
+Each model was run for 20,000 iterations and two independent chains with different initial values for parameters.
+We discarded the initial 10,000 samples.
+Convergence was observed graphically for all parameters, and confirmed by the Brooks-Gelman statistic \citep{brooks_general_1998}.
+For all three vital rate models, we considered the effect of removal treatments, or other coefficients of interest, to be significant when the 95\% confidence interval (or posterior interval) on the estimate of $\chi$ did not overlap zero.
+
 \subsection*{Modeling local crowding}
@@
 \begin{table}[h]
-\caption{Statistical models of year-to-year changes in log(cover) for the four focal species.}
+\caption{Statistical models of year-to-year changes in log(cover) for the four focal species. Values in square brackets show 95\%
+ confidence intervals.}
 \centering
@@
 \hline
-(Intercept)           & $0.00$           & $-0.05$          & $0.01$           & $-0.10$          \\
-                      & $[-0.41;\ 0.42]$ & $[-0.21;\ 0.10]$ & $[-0.26;\ 0.28]$ & $[-0.28;\ 0.08]$ \\
-TreatmentNo\_grass    & $-0.09$          &                  &                  &                  \\
-                      & $[-0.75;\ 0.56]$ &                  &                  &                  \\
-TreatmentNo\_shrub    &                  & $0.33^{*}$       & $-0.08$          & $0.21^{*}$       \\
-                      &                  & $[0.01;\ 0.65]$  & $[-0.41;\ 0.25]$ & $[0.01;\ 0.40]$  \\
-\hline
-AIC                   & 284.97           & 160.41           & 337.77           & 253.15           \\
-BIC                   & 298.99           & 172.62           & 352.65           & 268.46           \\
-Log Likelihood        & -137.49          & -75.20           & -163.88          & -121.57          \\
-Num. obs.             & 122              & 85               & 145              & 158              \\
+(Intercept)           & $-0.03$          & $0.08$           & $-0.14$          & $0.02$           \\
+                      & $[-0.48;\ 0.41]$ & $[-0.14;\ 0.30]$ & $[-0.56;\ 0.27]$ & $[-0.24;\ 0.29]$ \\
+TreatmentNo\_grass    & $-0.05$          &                  &                  &                  \\
+                      & $[-0.74;\ 0.64]$ &                  &                  &                  \\
+TreatmentNo\_shrub    &                  & $0.24$           & $-0.05$          & $0.18$           \\
+                      &                  & $[-0.12;\ 0.60]$ & $[-0.35;\ 0.25]$ & $[-0.01;\ 0.38]$ \\
+\hline
+AIC                   & 227.74           & 114.65           & 205.41           & 168.17           \\
+BIC                   & 240.01           & 124.69           & 218.18           & 181.40           \\
+Log Likelihood        & -108.87          & -52.33           & -97.70           & -79.09           \\
+Num. obs.             & 86               & 55               & 95               & 104              \\
 Num. groups: quad     & 19               & 11               & 21               & 22               \\
-Num. groups: year     & 9                & 9                & 9                & 9                \\
-Var: quad (Intercept) & 0.43             & 0.00             & 0.02             & 0.00             \\
-Var: year (Intercept) & 0.00             & 0.02             & 0.12             & 0.06             \\
-Var: Residual         & 0.41             & 0.32             & 0.48             & 0.24             \\
+Num. groups: year     & 5                & 5                & 5                & 5                \\
+Var: quad (Intercept) & 0.44             & 0.00             & 0.02             & 0.00             \\
+Var: year (Intercept) & 0.00             & 0.02             & 0.19             & 0.07             \\
+Var: Residual         & 0.52             & 0.36             & 0.38             & 0.24             \\
 \hline

### 2017-04-21 removals/treatment_trends_removals.r

@@ -109,7 +109,7 @@ dPSSP <- subset(logChange,species=="Pseudoroegneria spicata" & !is.na(pcgr) & Tr
 dPSSP$year <- as.factor(dPSSP$year)
 mPSSP <- lmer(pcgr ~ Treatment + (1|quad) + (1|year),data=dPSSP) # fit unscaled 
 
-texreg(list(mARTR_wt,mHECO_wt,mPOSE,mPSSP_wt), ci.force=TRUE,caption="Cover change models",
+texreg(list(mARTR,mHECO,mPOSE,mPSSP), ci.force=TRUE,caption="Cover change models",
       caption.above=TRUE,file=statsOutput)
 
 # figures ########################################################################

### 2017-04-21 removals/manuscript/removal_ms.tex
More revisions, all the way through.

@@
 
-A previous study of four dominant species from a sagebrush steppe community used historical data sets and multispecies population models to demonstrate weak interspecific interactions.
-The models predict little competitive release following species removals.
-We conducted a removal experiment to test our model's ability to predict the outcome of interspecific interactions.
-We established new quadrats in the same location where the historical data were collected.
-We assigned half the new quadrats to a  perennial grass removal treatment and half to a treatment in which we removed the dominant shrub, \textit{Artemisia tripartita}.
+A previous study of four dominant species from a sagebrush steppe community used historical data sets and multispecies population models to demonstrate weak interspecific interactions and predict little competitive release.
+We conducted a removal experiment to test our model's ability to accurately describe interspecific interactions.
+Our removal treatments were 1) removal of all perennial grass and 2) removal of the dominant shrub, \textit{Artemisia tripartita}.
 We modeled survival, growth and recruitment as functions of local neighborhood species composition along with an indicator variable for removal treatment.
 If our ``baseline" model, which accounts for local plant-plant interactions, explains responses to removal, then the fitted removal indicator effect should be non-significant.
-For survival and recruitment, the removal treatment effects were never significantly positive, indicating that competitive effects were not underestimated by the baseline model.
-For  \textit{Poa secunda} recruitment, the removal treatment was actually negative, indicating that our baseline model overestimated competitive effects.
+For survival and recruitment, the removal treatment effects were never significantly positive, indicating that the basline model did not underestimate competitive effects.
+For  \textit{Poa secunda} recruitment, the removal treatment was negative, indicating that our baseline model overestimated competition.
 For growth, the removal treatment effect was significant and positive for two species, \textit{Poa secunda} and \textit{Pseudoroegneria spicata}, indicating that the baseline model underestimated competitive effects.
-However, including information about the location of these grass individuals with respect to removed \textit{A.
+However, including information about the location of grass individuals with respect to removed \textit{A.
 tripartita} failed to improve the growth regressions, raising questions about the mechanisms driving the positive response to removal.
-For three species, individual based models and integral projections models based on the vital rate regressions showed that removal treatment effects had little impact on population growth rates or simulated equilibrium cover following a species removal.
+For three species, population models based on the vital rate regressions showed little impact of removals on population growth rates or simulated equilibrium cover.
 For \textit{P.
@@
 tripartita} removal projected greater competitive release than the baseline models.
-The ability of models based on observational data to predict population responses to experimental removals increased our confidence that competitive release is in fact small for at least three of our four study species.
-The implication is that interspecific resource competition is not a primary factor determining the structure and dynamics of this community.
-The success of the models also supports the use of multispecies models, based on long-term observational studies, to characterize competitive interactions in a community.
+The ability of models based on observational data to predict population responses to experimental removals increased our confidence that interspecific interactions are in fact weak for at least three of our four study species.
+The implication is that other factors besides interspecific resource competition play important roles in determining the structure and dynamics of this community.
+The success of the models also supports the use of multispecies models, based on long-term observational studies, to characterize competitive interactions.
 
@@
 (Fig. \ref{fig:CoverTrends}), but shrub removal did appear to cause an increase in the cover of \textit{P. spicata}. 
-However, when we tested the effects of removal on cover by analyzing year-to-year changes in (log) cover, a measure of population growth rate (Fig.
+However, \new{when we tested the effects of removal on cover by analyzing year-to-year changes in (log) cover, a measure of population growth rate (Fig.
 \ref{fig:CoverChange}), none of the removal treatment effects were significantly different from zero (Table \ref{table:coefficients}).
 For \textit{P.
-spicata}, the 95\% confidence interval only just contained zero (95\% CI: 0.01, 0.40), providing marginal evidence for a positive effect of removal on population growth.
+spicata}, the 95\% confidence interval only just contained zero (95\% CI: -0.01, 0.38), providing marginal evidence for a positive effect of removal on population growth.
 Fitting with weighted least squares to address heteroscedasticity in residuals indicated even less evidence for a positive response of  \textit{P.
-spicata} to removal.
-However, these treatment comparisons do not directly address our questions about whether responses to removal were
+spicata} to removal.} However, these treatment comparisons do not directly address our questions about whether responses to removal were
 adequately predicted by models fit to historical, observational data.
@@
 
-To quantify the short-term population-level consequences of removals, we used an IBM to project the cover of each species in each quadrat one year ahead using versions of the model with and without removal treatment effects.
-Our models tended to under-predict population growth (Fig.
-\ref{fig:CoverChange}) and population size (Fig.
-\ref{fig:IBM1step}) for all species, but the difference between predictions from the baseline and removal models varied among species.
+To quantify the short-term population-level consequences of removals, \new{we used an IBM to project the cover of each species in each quadrat one year ahead using versions of the model with and without removal treatment effects.
+Plotting observed vs.
+predicted cover for each quadrat in each year shows points clustered around the 1:1 line, indicating reasonable accuracy and little bias for both control and removal quadrats (Fig.
+\ref{fig:obsVpred}).
+For \textit{H.
+comata}, differences in predicted cover of removal plots between the baseline and removal models were very small.
+For the other three species, the removal model predicted higher cover in removal plots than the baseline model predicted, but only for \textit{P.
+spicata} were predictions from the removal model closer to the 1:1 line.
+Using the version of the model with maximum removal treatment effects (removal coefficients set to the 95\% confidence limit furthest from zero) increased differences between baseline and removal model predictions} (Fig.
+\ref{fig:obsVpred_maxCI}).
+\comment{I think this figure may be useless?
+cut?}
+
+ \begin{figure}[tbp]
+ \centering
+ \includegraphics[width=1\textwidth]{obsVpred_quad_year}
+ \caption{The relationship between observed and predicted cover for each quadrat, year, and species.
+The 1:1 line is dashed, the solid black line shows the trend for control plots with predictions from the baseline IBM model (black symbols), the blue line shows the trend for removal plots with predictions from the baseline model (blue symbols), and the red line shows the trend for removal plots with predictions from the removal model (red symbols).
+Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r}.
+}
+ \label{fig:obsVpred}
+ \end{figure}
+
+\new{Time series of predicted and observed cover, averaged over quadrats, show similar patterns.
+Predicted population size closely tracked observed population size (Fig.
+\ref{fig:IBM1step}) in most years for all species, but the difference between predictions from the baseline and removal models varied among species.} For \textit{H.
+comata}, differences in predictions from the two models were virtually indistinguishable.
 For \textit{A.
-tripartita}, both observed and predicted population growth rates were close to zero, and differences in predictions from the two models were small.
-\textit{H.
-comata} experienced greater variation in growth rates, but the inclusion of removal treatment effects had little effect on population growth rates.
-For \textit{Poa secunda} and especially \textit{P.
-spicata}, the model with removal effects consistently predicted higher population growth than the baseline model  and reduced the underestimates of cover (Fig.
+tripartita}, \textit{Poa secunda} and especially \textit{P.
+spicata}, the model with removal effects consistently predicted higher population growth than the baseline model (Fig.
 \ref{fig:IBM1step}).
-Viewing all annual growth rates across species and years shows the same pattern of underprediction for some species, but also reveals a strong correlation between observations and predictions in control plots (Fig.
-CUT.
-In removal plots, underprediction of growth rates in certain years was larger, especially for \textit{H.
-comata} and \textit{P.
-spicata}; including removal treatment effects only slightly reduced this bias (CUT).
-For the version of the model with maximum removal treatment effects (removal coefficients set to the 95\% confidence limit furthest from zero), the patterns were qualitatively similar, with a somewhat greater contrast between predictions of the baseline and removal models for \textit{P.
+Using the version of the model with maximum removal treatment effects, the patterns were qualitatively similar, with a somewhat greater contrast between predictions of the baseline and removal models for \textit{P.
 secunda} and \textit{P.
-spicata} (Figs.
-\ref{fig:IBM1step-maxCI},CUT).
+spicata} (Fig.
+\ref{fig:IBM1step-maxCI}).
 
@@
  \centering
- %\includegraphics[width=1\textwidth]{cover_projections_1step}
+ \includegraphics[width=1\textwidth]{cover_projections_1step}
  \caption{Observed and predicted cover for each of the four modeled species.
@@
 \ref{fig:IPMresults}).
+In fact, simulated cover of \textit{H.
+comata} and \textit{PSSP} declines slightly following \textit{A.
+tripartita} removal.
 Using a model with removal treatment coefficients, the IPM predicts little effect of shrub removal on \textit{Poa secunda} and \textit{H.
@@
 Our population models based on historical, observational data reveal weak interspecific interactions \citep{adler_coexistence_2010,chu_large_2015} and, as a result, predict very little competitive release among the four dominant species in a sagebrush steppe.
-Despite the fact that previous removal studies have reported variable results, ranging from negative effects of removal on remaining species to complete competitive release, we were suspicious of our model's predictions and conducted a removal experiment to test them.
-The results provide substantial support for our model's predictions.
+Despite the fact that previous removal studies have reported variable results, ranging from negative effects of removal on remaining species to complete competitive release, we were suspicious of our model's \new{ability to accurately describe interspecific interactions, so we conducted a removal experiment to test it.} The results provide substantial support for our model's predictions.
 
@@
 tripartita} removal on growth, while the long-term equilibrium are also sensitive to the negative effects of removal on recruitment.
-Although the baseline model did not always successfully predict removal effects on vital rates, it does appear to successfully capture the population-level consequences of removals on these three species and did not severely underestimate the magnitude of competitive release.
+Although the baseline model did not always successfully predict removal effects on vital rates, it does appear to successfully capture the population-level consequences of removals on these three species.
 
@@
 spicata}.
-Both individual growth and short- and long-term population growth were higher in the sagebrush removal treatments than in the control plots, suggesting that our analyses of observational data may underestimate the competitive release that \textit{P.
-spicata} experiences following \textit{A.
+Both individual growth and projected short- and long-term population growth were higher in the sagebrush removal treatments than in the control plots, suggesting that our analyses of observational data may underestimate the competitive effect of \textit{A.
+tripartita} on \textit{P.
+spicata}, and the resulting positive effect of \textit{A.
 tripartita} removal.
@@
 
-The small competitive release of many species in our system implies that each of the four dominant species we studied are limited by quite different factors.
+\new{How can we reconcile the insensitivity of grasses to sagebrush removal in our study with evidence from rangeland ecology research showing that sagebrush eradication increases grass production?
+The most likely explanation involves ecological context: our study focused on \textit{Artemisia tripartita} in an ungrazed plant communities, whereas most research on sagebrush and forage production focused on \textit{Artemisia tridentata} in heavily grazed communities, and often involved grass seeding following sagebrush removal (e.g.
+\citealt{Cook1963}).
+\textit{A.
+tripartita} is smaller in stature than \textit{A.
+tridentata} and may interact differently with perennial grasses.
+More importantly, cover of sagebrush can increase dramatically on overgrazed sites, especially under a spring grazing regime.
+Competitive release of grasses following sagebrush removal often increases in a nonlinear fashion, with little increase in production at sites with low sagebrush cover, and accelerating responses as cover passes some threshold \citep{McDaniel2005}.
+The cover of sagebrush in our ungrazed quadrats may have fallen below that threshold.
+}
+
+The weak interspecific interactions among many of the dominant species in our system implies that each of species we studied is limited by quite different factors.
 This low niche overlap could reflect strong resource partitioning (Tilman 1982), perhaps combined with variation in the times and locations at which species are taking up those resources.
@@
 
-Our results have given us more confidence that competitive release is quite small for the dominant species in our study system.
-But for many species in other communities, interspecific competition may still be an important driver.
+For many species in other communities, interspecific competition may still be an important driver.
 In closed-canopy forests where competition for light is intense, we might have found different results than in our system, in which competition for a number of belowground resources is more important than competition for light.
 However, a number of recent studies from forests have also found that heterospecific competitive effects are dramatically weaker than conspecific effects \citep{comita_asymmetric_2010,Kunstler2016,johnson_conspecific_2012}.
-We also expect stronger competitive release in successional systems, or communities in the early stages of community assembly where the process of competitive exclusion is still underway \citep{kokkoris_patterns_1999}.
-Finally, we might have found stronger competitive release if we could have studied rare species in our system, but only if those species are rare because of niche overlap with stronger competitors.
+We also expect stronger competition in successional systems, or communities in the early stages of community assembly where the process of competitive exclusion is still underway \citep{kokkoris_patterns_1999}.
+Finally, we might have found stronger evidence of competition if we could have studied rare species in our system, but only if those species are rare because of niche overlap with stronger competitors.
 If they are rare because they occupy unique, but narrow, niches, interspecific interactions would still be weak.
@@
   \centering
-  \includegraphics[width=1\textwidth]{cover_change_chrono}
-  \caption{Observed and predicted year-to-year changes in cover for the four dominant species.
-Mean observed changes (averaged across quadrats) are shown in black for control (solid symbols) and removal (open symbols) treatment.
-Predictions from the ``Baseline'' model (blue) account for effects of local species composition on vital rates, but do not include additional effects of removals.
-Predictions from the ``Removal'' model (red) include additional treatment effects on vital rates.
-Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r}.
-}
-  \label{fig:CoverChange}
-  \end{figure}
-
-  \begin{figure}[tbp]
-  \centering
   \includegraphics[width=1\textwidth]{PSSP_W_scatters}
@@
 \centering
+\includegraphics[width=1\textwidth]{obsVpred_quad_year_maxCI}
+\caption{The relationship between observed and predicted cover for each quadrat, year, and species assuming maximum effects of removal.
+The 1:1 line is dashed, the solid black line shows the trend for control plots with predictions from the baseline IBM model (black symbols), the blue line shows the trend for removal plots with predictions from the baseline model (blue symbols), and the red line shows the trend for removal plots with predictions from the removal model (red symbols).
+Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r}.
+}
+\label{fig:obsVpred_maxCI}
+\end{figure}
+
+ \begin{figure}[tbp]
+ \centering
  \includegraphics[width=1\textwidth]{cover_projections_1step_maxCI}

### 2017-04-24 removals/manuscript/removal_ms.tex
More manuscript revisions

@@
 
-A previous study of four dominant species from a sagebrush steppe community used historical data sets and multispecies population models to demonstrate weak interspecific interactions and predict little competitive release.
+\new{A previous study of four dominant species from a sagebrush steppe community used historical data sets and multispecies population models to demonstrate weak interspecific interactions and predict little competitive release.
 We conducted a removal experiment to test our model's ability to accurately describe interspecific interactions.
-Our removal treatments were 1) removal of all perennial grass and 2) removal of the dominant shrub, \textit{Artemisia tripartita}.
-We modeled survival, growth and recruitment as functions of local neighborhood species composition along with an indicator variable for removal treatment.
+Our removal treatments were 1) removal of all perennial grass and 2) removal of the dominant shrub, \textit{Artemisia tripartita}.} We modeled survival, growth and recruitment as functions of local neighborhood species composition along with an indicator variable for removal treatment.
 If our ``baseline" model, which accounts for local plant-plant interactions, explains responses to removal, then the fitted removal indicator effect should be non-significant.
@@
 An alternative approach to studying competitive release is to fit multispecies population models that estimate species interactions using long-term, observational data, and then simulate population responses to removals.
-Applying this approach to arid and semiarid plant communities, we found evidence of strong intraspecific limitation and weak interspecific interactions, leading to very stable coexistence and implying little competitive release \citep{adler_coexistence_2010,chu_large_2015}.\new{ In fact, at our sagebrush steppe study site  \citep{adler_coexistence_2010}, our model predicts removal of sagebrush would actually decrease the equilibrium cover of the three dominant grass species by 0.8 to 5.5\%, reflecting the facilitative effect of sagebrush on grass recruitment and very weak competitive effects on growth and survival.
+Applying this approach to arid and semiarid plant communities, we found evidence of strong intraspecific limitation and weak interspecific interactions, leading to very stable coexistence and implying little competitive release \citep{adler_coexistence_2010,chu_large_2015}.
+\new{ In fact, at our sagebrush steppe study site  \citep{adler_coexistence_2010}, our model predicts removal of sagebrush would actually reduce the equilibrium cover of the three dominant grass species very slightly, reflecting the facilitative effect of sagebrush on grass recruitment and very weak competitive effects on growth and survival.
 Since we only studied the most common, co-occurring species, we expected to find evidence of stable coexistence, but the predicted negative effect of sagebrush removal on grass cover is surprising given the long history of sagebrush eradication to increase grass forage production in
@@
 All data and R code necessary to reproduce our analysis will be deposited in the Dryad Digital Repository once the manuscript is accepted.
-The current version of the computer code and all necessary data are available at \texttt{https://github.com/pbadler/ExperimentTests.git}.
+The current version of the computer code and all necessary data are available at \texttt{https://github.com/pbadler/ExperimentTests/tree/master/removals}.
 
 \subsection*{Removal experiment}
+
 In spring 2011, in a large exclosure containing six of the historical permanent quadrats, we established an additional 16 quadrats for the removal treatments.
@@
 We considered the removal treatment effect to be significant when the 95\% confidence interval on the estimated coefficient did not overlap zero.
-Because variance in residuals of plots with low cover was high, we explored fitting by weighted least squares, to account for the apparent heteroscedasiticity.
+Because variance in residuals of plots with low cover was high for two species, we explored fitting by weighted least squares, to account for the apparent heteroscedasiticity.
 The results of these tests were qualitatively similar, so opted to only report the simpler analysis.
@@
 Although the three vital rate models differ in functional form (see Appendix A for complete model descriptions), all of them have a linear predictor, $\mu$, containing the following terms:}
-
 \begin{equation}
-\mu_ij = \gamma_j+ \varphi_{jg}+  \chi_j  + \sum \limits_{m} \omega_{jm} {W}_{ijm}
+\mu_{ij} = \gamma_j+ \varphi_{jg}+  \chi_j  + \sum \limits_{m} \omega_{jm} {W}_{ijm}
 \label{eqn:survReg}
 \end{equation}
-
 \new{where $i$ refers to an individual genet (for survival and growth) or quadrat (for recruitment) of species $j$, $\gamma$ is a time-dependent intercept (year effect), $\varphi$ is the coefficient for the effect of spatial location (`$g$' represents a group of nearby quadrats located within one pasture or grazing exclosure), $\chi$ is the removal treatment effect,
@@
 $R_j(v_j,u_j,\vec{n})=c_{0,j}(v_j)e^{u_j}\Phi_j$ \citep{adler_coexistence_2010} where $c_{0,j}$ is the
-size distribution of new recruits. $\Phi_j$ is calculated from eqn. (\ref{eqn:recrDataModel}) with total cover values computed from the population densities as $C'_{j,t} = \int e^u n(u_{j,t}) du$.
+size distribution of new recruits. $\Phi_j$ is calculated from eqn. \ref{eqn:recrDataModel} with total cover values computed from the population densities as $C'_{j,t} = \int e^u n(u_{j,t}) du$.
 
@@
 (Fig. \ref{fig:CoverTrends}), but shrub removal did appear to cause an increase in the cover of \textit{P. spicata}. 
-However, \new{when we tested the effects of removal on cover by analyzing year-to-year changes in (log) cover, a measure of population growth rate (Fig.
-\ref{fig:CoverChange}), none of the removal treatment effects were significantly different from zero (Table \ref{table:coefficients}).
+However, \new{when we tested the effects of removal on cover by analyzing year-to-year changes in (log) cover, a measure of population growth rate, none of the removal treatment effects were significantly different from zero (Table \ref{table:coefficients}).
 For \textit{P.
@@
 However, the distribution of crowding observed within control plots easily spans the crowding values observed in removal plots (Fig.
-\ref{fig:Wscatters}), and we failed to find evidence for strong interspecific covariances in crowding (\ref{suppMethods}).
+\ref{fig:Wscatters}), and we failed to find evidence for strong interspecific covariances in crowding (Appendix A).
 
@@
 tripartita} present and no removal treatment coefficients, brown boxes show results from the same model but with \textit{A.
-tripartita} (``ARTR") cover set to zero (a species removal), and green boxes show results from a model with  \textit{A.
+tripartita} cover set to zero (a species removal), and green boxes show results from a model with \textit{A.
 tripartita} set to zero and removal treatment coefficients included.
@@
 
-\section*{Appendix A. Detailed Methods} \label{suppMethods}
+\section*{Appendix A. Detailed Methods} \label{AppendixA}
 
@@
 % supplementary tables
-
 \begin{table}[h]

### 2017-05-10 removals/manuscript/removal_ms.tex
trivia
edits on the manuscript

@@
 
-\new{A previous study of four dominant species from a sagebrush steppe community used historical data sets and multispecies population models to demonstrate weak interspecific interactions and predict little competitive release.
+%\new{A previous study of four dominant species from a sagebrush steppe community used historical data sets and multispecies population models to demonstrate weak interspecific interactions and predict little competitive release.
 We conducted a removal experiment to test our model's ability to accurately describe interspecific interactions.
@@
 The success of the models also supports the use of multispecies models, based on long-term observational studies, to characterize competitive interactions.
+}
+
+\new{A previous study of four dominant species in a sagebrush steppe community used historical data sets and multispecies population models to demonstrate weak interspecific interactions, predicted that little competitive release would result from removal of allospecific competitors.
+We conducted a removal experiment to test whether our model, with weak interspecific competition, accurately describes interspecific interactions.
+Our removal treatments were 1) removal of all perennial grasses and 2) removal of the dominant shrub, \textit{Artemisia tripartita}.
+We modeled survival, growth and recruitment as functions of the locations, sizes, and species identities of neighboring plants, along with an indicator variable for removal treatment.
+If our ``baseline" model, which accounts for local plant-plant interactions, accurately explains the observed responses to removals, then the fitted removal indicator effect should be non-significant.
+For survival and recruitment, the removal treatment effects were never significantly positive, indicating that the baseline model did not underestimate intraspecific competitive effects.
+For \textit{Poa secunda} recruitment, the removal treatment was negative, indicating that our baseline model overestimated intraspecific competition.
+For growth, the removal treatment effect was significant and positive for two species, \textit{Poa secunda} and \textit{Pseudoroegneria spicata}, indicating that the baseline model underestimated intraspecific competitive effects.
+However, including information about the location of grass individuals with respect to removed \textit{A.
+tripartita} failed to improve the growth regressions, raising questions about the mechanisms driving the positive response to removal.
+For three species, population models based on the vital rate regressions showed little impact of removals on population growth rates or simulated equilibrium cover.
+For \textit{P.
+spicata}, the population models that included effects of \textit{A.
+tripartita} removal projected greater competitive release than the baseline models.
+The ability of models based on observational data to predict population responses to experimental removals increased our confidence that interspecific interactions are in fact weak for at least three of our four study species.
+The implication is that factors besides interspecific resource competition play important roles in determining the structure and dynamics of this community.
+The success of the models also supports the use of multispecies models based on long-term observational data for  characterizing competitive interactions.} \comment{346 words, max=350 at Ecology}
 
@@
 
-\subsection*{Overview of statistical models of vital rates}
-
-\new{
-Our regression models for survival, growth and recruitment follow previous work \citep{adler_coexistence_2010,chu_large_2015}.
+\subsection*{Overview of statistical models for vital rates}
+
+\new{Our regression models for survival, growth and recruitment follow previous work \citep{adler_coexistence_2010,chu_large_2015}.
 The current analysis departs from previous work with the inclusion of a categorical removal treatment effect.
@@
 We refer to the model that includes the estimated removal treatment effect as the ``removal model".
-Note that these are not estimated as separate models; we estimate all parameters together, but when we generate predictions from the baseline model, we simply set the removal treatment effects to zero.} % end new
+Note that these are not estimated as separate models; we estimate all parameters together, but when we generate predictions from the baseline model, we simply set the removal treatment effects to zero.}
 
 \subsection*{Individual-based models}
-
 The vital rate regressions provide a qualitative test of removal treatment effects: Does removal have an effect on a vital rate that goes above and beyond the effects of species interactions that are explicit in the model?
@@
 
-To simulate changes in cover in each quadrat from year $t$ to year $t+1$, we initialized the IBM with the actual configuration of the four focal species (genet sizes and locations) observed in year $t$.
+To simulate changes in cover in each quadrat from year $t$ to year $t+1$, we initialized the IBM with the actual configuration of the four focal species observed in year $t$ (genet sizes and locations).
 We then projected the model forward one year, applying the random year effects corresponding to year $t$ and the spatial random effects corresponding to the quadrat of interest.
-Because the main goal of this comparison is to look for systematic deviations between model predictions and observations, we used a deterministic version of the model (e.g., recruitment is the $\lambda$ of (\ref{eqn:recrProcessModel}), rather than a random draw from a Poisson distribution with a mean of $\lambda$).
-\comment{STEVE DO WE NEED TO EXPLAIN OUR JENSEN'S INEQUALITY CORRECTION HERE?}
-We first applied the recruitment regression to determine the expected number of new recruits at time $t+1$, which we then multiplied by the mean size of recruits observed for the focal species to get expected cover of new recruits.
-We then used the survival regression to predict the survival probability of each genet, and the growth regression to predict the change in size of each genet.
-Expected cover at $t+1$ is expected area of new recruits, plus the sum of the survival probabilities multiplied by the expected sizes.
-Because we do not carry these simulations forward beyond one time step, we do not need to assign locations to the new recruits.
-For control plots, we used the baseline version of the model with all removal treatment effects set to zero.
-For removal treatment plots, we generated predictions from both the baseline and removal versions of the model, with and without treatment effects set to zero, respectively.
+Because the main goal of this
+comparison is to look for systematic deviations between model predictions and observations, we used a deterministic version of the model (e.g., 
+recruitment is the $\lambda$ of (\ref{eqn:recrProcessModel}), rather than a Poisson random variable with mean $\lambda$). \new{Specifically, we used the recruitment regression to calculate 
+the expected number of new recruits at time $t+1$, which we then multiplied by the observed mean size of recruits for the focal species to get expected cover of new recruits. 
+We used the survival regression to predict the survival probability of each genet, and the growth regression to predict the expected size (on arithmetic scale) of each genet at time $t+1$ (note that the
+expected size on arithmetic scale is not exp of the expected log size, because of Jensen's inequality). 
+Expected cover at $t+1$ is expected area of new recruits, plus the product of survival probability and expected new size summed over all genets present at time $t$. }
+Because we do not project forward beyond one time step, we do not need to assign locations to the new recruits. For control plots, we used the baseline version of the model with all 
+removal treatment effects set to zero. For removal treatment plots, we generated predictions from both the baseline and removal versions of the model, 
+with and without treatment effects set to zero, respectively.
 
@@
 Perhaps large uncertainty in the estimates of these effects makes it difficult to reject the null hypothesis represented by our baseline model.
-To address this possibility, we also simulated a version of the model in which we replaced the mean estimate of each removal treatment effect with the parameter value at either the 5\% or 95\% confidence limit, whichever was farthest from zero.
+To address this possibility, we also simulated a version of the model in which we replaced the point estimate of each removal treatment effect with either the 5\% or 95\% confidence limit for the treatment effect,whichever was farthest from zero.
 This version of the model shows the population-level changes we would expect given maximum removal effects.
@@
 
-\comment{Should we move much of this to Appendix A?}
-
-To quantify the long-term, population-level consequences of the removal effects, we built and then simulated an integral projection model (IPM).
-In our IPM, the population of species $j$ is represented by a density function $n(u_{j,t})$ which gives the density of genets of size $u$ at time $t$, with genet size on a natural-log scale, i.e.
-$n(u_{j,t})du$ is the number of genets whose log area is between $u$ and $u+du$, hence their area
-(on arithmetic scale) is between $exp(u_j)$ and $exp(u_j+du)$. The size distribution function at time $t+1$ is given by
+\comment{PBA: should we move much of this to Appendix A? SPE: done}
+
+To quantify the long-term population-level consequences of the removal effects, we built and simulated an integral projection model (IPM). The population 
+of species $j$ is represented by a density function $n_j(u,t)$ which gives the density of genets of size $u$ at time $t$ (``size'' is natural log of genet area, so 
+that $n_j(u,t)h$ is the number of genets whose area is between between $\exp(u)$ and $\exp(u+h)$). The size distribution function at time $t+1$ is given by
 \begin{equation}
-n(v_{j,t+1})=\int_{L_j}^{U_j} k_j (v_j,u_j,{\vec{n}})n(u_{j,t})du_j   
+n_j(v,t+1)=\int\limits_{L_j}^{U_j} k_j (v,u,{\vec{n}})n_j(u,t)du   
 \label{eqn:IPM}
 \end{equation}
-where the kernel $k_j (v_j,u_j,\vec{n}_j)$ describes all possible transitions from size $u$ to $v$ and $\vec{n}$ 
-is the set of size-distribution functions for all species in the community. The integral is evaluated over all possible sizes from a 
-lower size limit $L$ to an upper size limit $U$ that extends past the range of observed sizes.
+where the kernel $k_j (v_j,u_j,\vec{n}_j)$ describes all possible transitions from size $u$ to $v$, and $\vec{n}$ 
+is the set of size-distribution functions for all species in the community. The integral is evaluated over a range of sizes [$L_j,U_j$] 
+that extends past the range of observed sizes.
 
@@
 \begin{equation}
-k_j(v_j,u_j,\vec{n})=S_j(u_j,\vec{n})G_j(v_j,u_j,\vec{n})+R_j(v_j,u_j,\vec{n}) 
+k_j(v,u,\vec{n})=S_j(u,\vec{n})G_j(v,u,\vec{n})+R_j(v,u,\vec{n}) 
 \label{eqn:IPM} 
 \end{equation}
-$S$ is given by (Eqn. \ref{eqn:survReg}) and $G$ is given by (Eqn. \ref{eqn:growReg}), using an expected neighborhood crowding
-index calculated from the size distributions of the competing species. In the vital rate regressions and in the IBM, each 
-individual has a unique neighborhood crowding index based on the spatial locations and sizes of neighboring plants. This 
-measure cannot be extended to the IPM, which tracks individual size but not location. Instead we used a 
-mean-field approximation that captures the essential features of spatially-explicit neighborhood interactions \citep{adler_coexistence_2010}. 
-We found that in both the observed data and IBM simulations, heterospecific individuals were approximately randomly distributed with respect to each other, but conspecific individuals displayed a non-random, size-dependent patterns: small genets were randomly distributed, while large genets
-were segregated from each other. The overdispersion of large conspecific genets was introduced into the IPM 
-with a `no-overlap' rule \citep{adler_coexistence_2010}.
-In \ref{suppMethods} we explain how the spline competition kernels with the no-overlap rule were used to compute neighborhood crowding index as a function of individual size for the IPM.
-
-For recruitment in the IPM, the factor $ \Phi_j  =\exp{\left(\gamma_{j,t}^R +  \varphi_{jg}^R + \chi_{jh}^R + 
-\sum \limits_{m} \omega_{jm}^R \sqrt{C'_{mq,t}} \right)}$
-in eqn. (\ref{eqn:recrDataModel}) gives the expected 
-area of new recruits per quadrat, per unit area of parents.
-To incorporate this recruitment function into the IPM, we assumed that individual fecundity increases linearly with individual area (equal to $e^u$ for a size-$u$ individual).
-The recruitment kernel is thus
-$R_j(v_j,u_j,\vec{n})=c_{0,j}(v_j)e^{u_j}\Phi_j$ \citep{adler_coexistence_2010} where $c_{0,j}$ is the
-size distribution of new recruits. $\Phi_j$ is calculated from eqn. \ref{eqn:recrDataModel} with total cover values computed from the population densities as $C'_{j,t} = \int e^u n(u_{j,t}) du$.
+which are very similar to those used in our previous studies of this system \citep{adler_coexistence_2010,adler_forecasting_2012}; see 
+Appendix \ref{AppendixA} for a complete description. As in these previous studies, the IPM uses a mean-field approximation for
+neighborhood comeptition that captures the essential features of spatially-explicit local interactions; 
+in Appendix \ref{sec:kernelMethods} we explain how the spline competition kernels were used 
+to compute neighborhood competition as a function of individual size for the IPM.
+
+%$S$ is given by (Eqn. \ref{eqn:survReg}) and $G$ is given by (Eqn. \ref{eqn:growReg}), using an expected neighborhood crowding
+%index calculated from the size distributions of the competing species. In the vital rate regressions and in the IBM, each 
+%individual has a unique neighborhood crowding index based on the spatial locations and sizes of neighboring plants. This 
+%measure cannot be extended to the IPM, which tracks individual size but not location. Instead we used a 
+%mean-field approximation that captures the essential features of spatially-explicit neighborhood interactions \citep{adler_coexistence_2010}. 
+%We found that in both the observed data and IBM simulations, heterospecific individuals were approximately randomly distributed with respect to each other, 
+%but conspecific individuals displayed a non-random, size-dependent patterns: small genets were randomly distributed, while large genets 
+%were segregated from each other. The overdispersion of large conspecific genets was introduced into the IPM 
+%with a `no-overlap' rule \citep{adler_coexistence_2010}. In Appendix \ref{sec:kernelMethods} we explain how the spline competition kernels with the no-overlap rule were used 
+%to compute neighborhood crowding index as a function of individual size for the IPM.  
+%
+%For recruitment in the IPM, the factor $ \Phi_j  =\exp{\left(\gamma_{j,t}^R +  \varphi_{jg}^R + \chi_{jh}^R + 
+%\sum \limits_{m} \omega_{jm}^R \sqrt{C'_{mq,t}} \right)}$
+%in eqn. (\ref{eqn:recrDataModel}) gives the expected 
+%area of new recruits per quadrat, per unit area of parents. To incorporate this recruitment function into the IPM, we assumed that individual fecundity increases 
+%linearly with individual area (equal to $e^u$ for a size-$u$ individual). The recruitment kernel is thus
+%$R_j(v_j,u_j,\vec{n})=c_{0,j}(v_j)e^{u_j}\Phi_j$ \citep{adler_coexistence_2010} where $c_{0,j}$ is the
+%size distribution of new recruits. $\Phi_j$ is calculated from eqn. \ref{eqn:recrDataModel} with total cover values computed from the population densities as $C'_{j,t} = \int e^u n(u_{j,t}) du$.
 
@@
 
-PBA and ARK designed the experiment and supervised data collection with the help of JBT; PBA, SPE, GH and BT analyzed data; PBA wrote the first draft of the manuscript and all authors contributed to editing.
+PBA and ARK designed the experiment and supervised data collection with the help of JBT; PBA, SPE, GH and BT analyzed data and constructed models; 
+PBA wrote the first draft of the manuscript and all authors contributed to editing.
 
@@
 \setcounter{table}{0}
-\renewcommand{\theequation}{A-\arabic{equation}}
-\renewcommand{\thetable}{B-\arabic{table}}
-\renewcommand{\thefigure}{C-\arabic{figure}}
-\renewcommand{\thesection}{A.\arabic{section}}
 
@@
 
-\section*{Appendix A. Detailed Methods} \label{AppendixA}
-
-\subsection*{Statistical models of vital rates}
+\renewcommand{\theequation}{A-\arabic{equation}}
+\renewcommand{\thetable}{A-\arabic{table}}
+\renewcommand{\thefigure}{A-\arabic{figure}}
+\renewcommand{\thesection}{\Alph{section}}
+
+\section{Detailed Methods} 
+\label{AppendixA}
+
+\subsection{Statistical models of vital rates}
 
@@
 
-\subsection*{Modeling local crowding}
-
+To incorporate the fitted recruitment function into the IPM, we assumed that individual fecundity increases 
+linearly with individual area (equal to $e^u$ for a size-$u$ individual). The recruitment kernel is thus
+$R_j(v,u,\vec{n})=c_{0,j}(v)e^{u}\Phi_j$ \citep{adler_coexistence_2010} where $c_{0,j}$ is the
+size distribution of new recruits. $\Phi_j$ is calculated from eqn. \ref{eqn:recrDataModel} with total cover values computed from the population densities as $C'_{j,t} = \int e^u n_j(u,t) du$.
+
+\subsection{Modeling local crowding}
 We modeled the crowding experienced by a focal genet in each year $t$ as a function of the distance to and size of neighbor genets. In previous work, 
@@
 
-\subsection*{Interspecific covariance in local crowding} 
+\subsection{Interspecific covariance in local crowding} 
 We explored interspecific covariance in local crowding experienced by individual plants, by regressing the $W$ values exerted by one neighbor species, the response variable, against the $W$ values of all other species, the independent variables.
@@
 
-\subsection*{Mean field approximation of local crowding for the IPM} 
+\subsection{Mean field approximation of local crowding for the IPM} 
+\label{sec:kernelMethods} 
 \citet{adler_coexistence_2010} developed a mean field approximation for local crowding when the
@@
 used fitted nonparametric competition kernels.
+
+The mean field approximation is based on the observed spatial distribution patterns of the species \citep{adler_coexistence_2010}. 
+In both the observed data and IBM simulations, heterospecific individuals were approximately randomly distributed with respect to each other, 
+but conspecific individuals displayed a non-random, size-dependent patterns: small genets were randomly distributed, while large genets 
+were segregated from each other. The overdispersion of large conspecific genets is incorporated into the IPM 
+with a `no-overlap' rule explained below, as in \citep{adler_coexistence_2010}.
 
@@
 
-\renewcommand{\thesection}{B.\arabic{section}}
-
-\section*{Appendix B. Supplementary Tables}
+\renewcommand{\theequation}{B-\arabic{equation}}
+\renewcommand{\thetable}{B-\arabic{table}}
+\renewcommand{\thefigure}{B-\arabic{figure}}
+
+\section{Supplementary Tables}
 
@@
 
-\renewcommand{\thesection}{C.\arabic{section}}
-
-\section*{Appendix C. Supplementary Figures}
+\renewcommand{\theequation}{C-\arabic{equation}}
+\renewcommand{\thetable}{C-\arabic{table}}
+\renewcommand{\thefigure}{C-\arabic{figure}}
+
+\section{Supplementary Figures}
 
@@
 Red lines and symbols show one-step-ahead predictions from an IBM that includes removal treatment coefficients.
-Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r}.
-}
+  Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r}. }
  \label{fig:IBM1step-maxCI}

### 2017-05-11 removals/manuscript/removal_ms.tex
edits and comments on removals ms

@@
 We used the survival regression to predict the survival probability of each genet, and the growth regression to predict the expected size (on arithmetic scale) of each genet at time $t+1$ (note that the
-expected size on arithmetic scale is not exp of the expected log size, because of Jensen's inequality). 
+expected size on arithmetic scale is not exp(expected log size), because of Jensen's inequality). 
 Expected cover at $t+1$ is expected area of new recruits, plus the product of survival probability and expected new size summed over all genets present at time $t$. }
@@
 
-\comment{PBA: should we move much of this to Appendix A? SPE: done}
+\comment{PBA: should we move much of this to Appendix A? SPE: DONE}
 
@@
 tripartita} response to grass removal because the removal treatment effects in the vital rate regressions were never significantly different from zero, indicating that the baseline model effectively predicted competitive release.
-\comment{Just do it?}
+\comment{PBA: Just do it?
+SPE: Just do it, and then explain why there is no difference between ``grass removal'' and ``grass removal with removal treatment effect''.
+}
 
@@
 During the period of our experiment, annual water-year (October-September) precipitation fell within the 5\% - 95\% quantiles of the long-term mean, although more years fell below the mean than above it (\ref{fig:climate}a).
-However, annual mean temperatures were consistently warm relative to the long-term mean, and two years were warmer than the 95\% quantile (\ref{fig:climate}b).
+However, annual mean temperatures were consistently warm relative to the long-term mean, and two years were warmer than the 95\% quantile (Figure \ref{fig:climate}b).
 
@@
 (Fig. \ref{fig:CoverTrends}), but shrub removal did appear to cause an increase in the cover of \textit{P. spicata}. 
-However, \new{when we tested the effects of removal on cover by analyzing year-to-year changes in (log) cover, a measure of population growth rate, none of the removal treatment effects were significantly different from zero (Table \ref{table:coefficients}).
-For \textit{P.
-spicata}, the 95\% confidence interval only just contained zero (95\% CI: -0.01, 0.38), providing marginal evidence for a positive effect of removal on population growth.
-Fitting with weighted least squares to address heteroscedasticity in residuals indicated even less evidence for a positive response of  \textit{P.
-spicata} to removal.} However, these treatment comparisons do not directly address our questions about whether responses to removal were
-adequately predicted by models fit to historical, observational data.
+However, \new{when we tested the effects of removal on cover by analyzing year-to-year changes in (log) cover, a measure of population 
+growth rate, none of the removal treatment effects were significantly different from zero (Table \ref{table:coefficients}). For \textit{P. spicata}, 
+the 95\% confidence interval only just contained zero (95\% CI: -0.01, 0.38), providing marginal evidence for a positive effect of removal on population growth. 
+Fitting with weighted least squares to address heteroscedasticity in residuals indicated even less evidence for a positive 
+response of  \textit{P. spicata} to removal.} 
+\comment{SPE: I think you need to explain how did ``analyzing year-to-year changes in (log) cover'' -- what was the response, what were the
+covariates? I think reviewers will also want to know what was done. And without that information, 
+it's not clear what the residuals would be, or what weighted least squares would entail. } 
+However, these treatment comparisons do not directly address our questions about whether 
+responses to removal were adequately predicted by models fit to historical, observational data.
 
@@
 \includegraphics[width=1\textwidth]{treatment_trends_cover}
-\caption{Cover trends by treatment for the four dominant species.
-Note the difference in y-axis scale between the canopy cover of \textit{A.
-tripartita} and the basal cover of the dominant grasses.
-``Shrub removal'' refers to removal of \textit{A.
-tripartita}, and ``Grass removal'' refers to removal of all perennial grasses.
-The cover of species in treatments from which they are removed (e.g.
-\textit{A.
-tripartita} cover in shrub removal plots) is included to show values at the start of the experiment in 2011 and the efficacy of the removal treatments.
-Source file: \texttt{treatment\textunderscore trends\textunderscore removals.r}.}
+\caption{Cover trends by treatment for the four dominant species.  Note the difference in y-axis scale between the canopy cover of \textit{A. tripartita} and 
+the basal cover of the dominant grasses. ``Shrub removal'' refers to removal of \textit{A. tripartita}, and ``Grass removal'' refers to removal of all perennial grasses. 
+The cover of species in treatments from which they are removed (e.g. \textit{A. tripartita} cover in shrub removal plots) is included to show values at the start 
+of the experiment in 2011 and the efficacy of the removal treatments. Source file: \texttt{treatment\textunderscore trends\textunderscore removals.r}.}
 \label{fig:CoverTrends}
@@
 Our vital rate models take into account the effects of local species composition and should therefore be able to explain the effect of removals on survival, growth and recruitment without any additional parameters.
-If the removal treatment effects that we added to these models are statistically significant, it would indicate an effect of removals either larger or smaller than the effect predicted by a model based on data from the historical control plots.
-For the survival models, the 95\% confidence intervals on the removal effects overlapped zero for all four species (Fig.
-\ref{fig:VitalRateTest}; Tables \ref{ARTRsurvival} to \ref{PSSPsurvival}), meaning that the survival probabilities of plants in removal plots can be predicted based on data from control plots.
-For the growth models, the 95\% confidence intervals on the removal effects were positive for \textit{P.
-secunda}  and \textit{P.
-spicata}, meaning that individuals of these species grew more rapidly in removal plots than controls even after accounting for neighborhood interactions (Fig.
-\ref{fig:VitalRateTest}; Tables \ref{ARTRgrowth} to \ref{PSSPgrowth}).
-In contrast, \textit{Poa secunda} had lower recruitment in removal than controls plots (upper limit of 95\% CI $<0$; Fig.
-\ref{fig:VitalRateTest}; Table \ref{table:recruitment}).
-For the other species, we could not detect significant removal treatment effects on recruitment.
+If the removal treatment effects that we added to these models are statistically significant, it would indicate an effect of removals either larger or smaller than the effect predicted by a model based on data
+from the historical control plots. 
+\comment{SPE: is the baseline model just based on the historical control plots data, or also on the modern data from control plots?} 
+For the survival models, the 95\% confidence intervals on the removal effects overlapped zero for all four species (Fig. \ref{fig:VitalRateTest}; 
+Tables \ref{ARTRsurvival} to \ref{PSSPsurvival}), meaning that the survival probabilities of plants in removal plots can be predicted based on data 
+from control plots. For the growth models, the 95\% confidence intervals on the removal effects were positive for \textit{P. secunda}  and \textit{P. spicata},
+ meaning that individuals of these species grew more rapidly in removal plots than controls even after accounting for neighborhood interactions 
+ (Fig. \ref{fig:VitalRateTest}; Tables \ref{ARTRgrowth} to \ref{PSSPgrowth}). In contrast, \textit{Poa secunda} had lower recruitment in removal 
+ than controls plots (upper limit of 95\% CI $<0$; Fig. \ref{fig:VitalRateTest}; Table \ref{table:recruitment}). For the other species, 
+ we could not detect significant removal treatment effects on recruitment.
 
@@
 However, the distribution of crowding observed within control plots easily spans the crowding values observed in removal plots (Fig.
-\ref{fig:Wscatters}), and we failed to find evidence for strong interspecific covariances in crowding (Appendix A).
+\ref{fig:Wscatters}), and we failed to find evidence for strong interspecific covariances in crowding (Appendix \ref{AppendixA}).
 
@@
 Plotting observed vs.
-predicted cover for each quadrat in each year shows points clustered around the 1:1 line, indicating reasonable accuracy and little bias for both control and removal quadrats (Fig.
-\ref{fig:obsVpred}).
+predicted cover for each quadrat in each year (Fig.
+\ref{fig:obsVpred}) shows points clustered around the 1:1 line, indicating reasonable accuracy and little bias for both control and removal quadrats.
 For \textit{H.
@@
 \ref{fig:obsVpred_maxCI}).
-\comment{I think this figure may be useless?
-cut?}
+\comment{PBA: I think this figure may be useless?
+cut?
+SPE: yes, cut.}
 
@@
 
-To quantify the long-term population-level consequences of removals, we used an IPM to project stochastic equilibrium cover.
+To quantify the long-term population-level consequences of permanent competitor removals, we used an IPM to project the long-term average cover of each species.
 We concentrated on grass species' responses to \textit{A.
@@
 \ref{fig:IPMresults}).
-In fact, simulated cover of \textit{H.
-comata} and \textit{PSSP} declines slightly following \textit{A.
+In fact, projected average cover of \textit{H.
+comata} and \textit{P.
+spicata} declines slightly following \textit{A.
 tripartita} removal.
@@
 
-Our population models based on historical, observational data reveal weak interspecific interactions \citep{adler_coexistence_2010,chu_large_2015} and, as a result, predict very little competitive release among the four dominant species in a sagebrush steppe.
-Despite the fact that previous removal studies have reported variable results, ranging from negative effects of removal on remaining species to complete competitive release, we were suspicious of our model's \new{ability to accurately describe interspecific interactions, so we conducted a removal experiment to test it.} The results provide substantial support for our model's predictions.
-
-For three of the four species, we find little or no evidence that models based on observational data underestimate the intensity of interspecific competition.
-After accounting for local neighborhood crowding, the removal treatment effects were never significant for \textit{A.
-tripartita} or \textit{H.
-comata}; our baseline model, which accounts for local neighborhood composition, adequately explained vital rate responses to  removal.
-\textit{P.
-secunda} showed mixed responses to \textit{A.
-tripartita} removal: removal effects were not significant at the survival stage, were positive at the growth stage, and negative at the recruitment stage (where the estimated effect of \textit{A.
-tripartita} on \textit{P.
-secunda} is positive).
-At the population level, models with and without removal effects projected similar patterns of short (Fig.
-\ref{fig:IBM1step}) and long-term (Fig.
-\ref{fig:IPMresults}) dynamics for \textit{H.
-comata} and \textit{A.
-tripartita}.
-For \textit{P.
-secunda}, the model with removal effects predicted higher population growth over the short but not the long-term, perhaps because the short-term projections mostly reflect the positive effect of \textit{A.
-tripartita} removal on growth, while the long-term equilibrium are also sensitive to the negative effects of removal on recruitment.
-Although the baseline model did not always successfully predict removal effects on vital rates, it does appear to successfully capture the population-level consequences of removals on these three species.
-
-The story is different for the fourth species, \textit{P.
-spicata}.
-Both individual growth and projected short- and long-term population growth were higher in the sagebrush removal treatments than in the control plots, suggesting that our analyses of observational data may underestimate the competitive effect of \textit{A.
-tripartita} on \textit{P.
-spicata}, and the resulting positive effect of \textit{A.
-tripartita} removal.
-However, the mechanism by which \textit{A.
-tripartita} removal promotes \textit{P.
-spicata} remains unclear.
-If  \textit{A.
-tripartita} removal is releasing \textit{P.
-spicata} from competition, we should expect \textit{P.
-spicata} individuals that were located beneath \textit{A.
-tripartita} shrubs that we removed to respond more than individuals located away from any shrubs.
-Yet we found no evidence for such a difference.
-An alternative explanation involves pre-treatment differences in composition between control plots and  the \textit{A.
-tripartita} removal  plots, which had higher cover of \textit{A.
-tripartita} (see 2011 means in Fig.
-\ref{fig:CoverTrends}) and higher cover of subdominant grasses and shrubs (Fig.
-\ref{fig:W-by-treatment}).
-Perhaps these pre-treatment differences reflect subtle differences in edaphic conditions which could promote \textit{P.
-spicata} growth.
-In summary, while \textit{P.
-spicata} appears to show stronger competitive release than our baseline model predicted, the evidence is not conclusive.
-
-\new{How can we reconcile the insensitivity of grasses to sagebrush removal in our study with evidence from rangeland ecology research showing that sagebrush eradication increases grass production?
-The most likely explanation involves ecological context: our study focused on \textit{Artemisia tripartita} in an ungrazed plant communities, whereas most research on sagebrush and forage production focused on \textit{Artemisia tridentata} in heavily grazed communities, and often involved grass seeding following sagebrush removal (e.g.
-\citealt{Cook1963}).
-\textit{A.
-tripartita} is smaller in stature than \textit{A.
-tridentata} and may interact differently with perennial grasses.
-More importantly, cover of sagebrush can increase dramatically on overgrazed sites, especially under a spring grazing regime.
-Competitive release of grasses following sagebrush removal often increases in a nonlinear fashion, with little increase in production at sites with low sagebrush cover, and accelerating responses as cover passes some threshold \citep{McDaniel2005}.
-The cover of sagebrush in our ungrazed quadrats may have fallen below that threshold.
-}
-
-The weak interspecific interactions among many of the dominant species in our system implies that each of species we studied is limited by quite different factors.
-This low niche overlap could reflect strong resource partitioning (Tilman 1982), perhaps combined with variation in the times and locations at which species are taking up those resources.
-Species-specific responses to spatial and temporal variation in non-resource environmental variables can also weaken interspecific interactions \citep{chesson_mechanisms_2000}, though in a previous study we failed to find evidence of a temporal storage effect in this community \citep{adler_weak_2009}.
-Interactions with above- or belowground enemies can regulate populations and, given enough species-specificity in the effects of these enemies, decouple interspecific dynamics \citep{hersh_evaluating_2011,janzen_herbivores_1970,connell_role_1971}.
-More likely, plants may alter the environment, and even resource availability, in ways that facilitate their neighbors.
-For example, shade cast by \textit{A.
-tripartita} canopies may reduce evapotranspiration (e.g.
-\citealt{Barbier2008}).
-\textit{A.
-tripartita} removal might offset this facilitative effect by reducing resource uptake, resulting in little net change in water balance following removal.
+%Our population models based on historical, observational data reveal weak interspecific interactions \citep{adler_coexistence_2010,chu_large_2015} and, as a result, 
+%predict very little competitive release among the four dominant species in a sagebrush steppe. Despite the fact that previous removal studies have reported variable 
+%results, ranging from negative effects of removal on remaining species to complete competitive release, we were suspicious of our model's \new{ability to accurately 
+%describe interspecific interactions, so we conducted a removal experiment to test it.} The results provide substantial support for our model's predictions.
+
+Our population models based on historical, observational data reveal weak interspecific interactions \citep{adler_coexistence_2010,chu_large_2015}. As a result, the 
+models predict very little competitive release among the four dominant species in a sagebrush steppe. Because the models were based solely on observational
+data, we were suspicious of our model's ability to accurately describe interspecific interactions, so we conducted a removal experiment to test it. 
+The results provide substantial support for our model's predictions, and for the weakness of interspecific interactions in this community. \comment{SPE has
+revised this paragraph. The previous version is in the tex file, commented out.}
+
+For three of the four species, we find little or no evidence that models based on observational data underestimate the intensity of interspecific competition. After accounting 
+for local neighborhood crowding, the removal treatment effects were never significant for \textit{A. tripartita} or \textit{H. comata}; 
+our baseline model, which accounts for local neighborhood composition, adequately explained vital rate responses to removal.  
+\textit{P. secunda} showed mixed responses to \textit{A. tripartita} removal: removal effects were not significant at the survival stage, 
+were positive at the growth stage, and negative at the recruitment stage (where the estimated effect of \textit{A. tripartita} 
+on \textit{P. secunda} is positive). At the population level, models with and without removal effects projected similar patterns of 
+short (Fig. \ref{fig:IBM1step}) and long-term (Fig. \ref{fig:IPMresults}) dynamics for \textit{H. comata} and \textit{A. tripartita}. 
+For \textit{P. secunda}, the model with removal effects predicted higher population growth over the short but not the long-term, perhaps 
+because the short-term projections mostly reflect the positive effect of \textit{A. tripartita} removal on growth, while the long-term 
+equilibrium are also sensitive to the negative effects of removal on recruitment. Although the baseline model did 
+not always successfully predict removal effects on vital rates, it does appear to successfully capture the population-level 
+consequences of removals on these three species.
+
+The story is different for the fourth species, \textit{P. spicata}. Both individual growth and projected short- and long-term population 
+growth were higher in the sagebrush removal treatments than in the control plots, suggesting that our analyses of observational data may 
+underestimate the competitive effect of \textit{A. tripartita} on \textit{P. spicata}, and the resulting positive effect of \textit{A. tripartita} 
+removal. However, the mechanism by which \textit{A. tripartita} removal promotes \textit{P. spicata} remains unclear. If  \textit{A. tripartita} 
+removal is releasing \textit{P. spicata} from competition, we should expect \textit{P. spicata} individuals that were located beneath \textit{A. tripartita} 
+shrubs that we removed to respond more than individuals located away from any shrubs. Yet we found no evidence for such a difference. 
+An alternative explanation involves pre-treatment differences in composition between control plots and  the \textit{A. tripartita} 
+removal plots, which had higher cover of \textit{A. tripartita} (see 2011 means in Fig. \ref{fig:CoverTrends}) and higher cover of subdominant 
+grasses and shrubs (Fig. \ref{fig:W-by-treatment}). Perhaps these pre-treatment differences reflect subtle differences in 
+edaphic conditions which could promote \textit{P. spicata} growth. In summary, while \textit{P. spicata} appears to 
+show stronger competitive release than our baseline model predicted, the evidence is not conclusive.
+
+\new{How can we reconcile the insensitivity of grasses to sagebrush removal in our study with evidence from rangeland ecology 
+research showing that sagebrush eradication increases grass production? The most likely explanation involves ecological context: our 
+study focused on \textit{Artemisia tripartita} in an ungrazed plant communities, whereas most research on sagebrush 
+and forage production has focused on \textit{Artemisia tridentata} in heavily grazed communities, and often has involved grass seeding 
+following sagebrush removal (e.g. \citealt{Cook1963}). \textit{A. tripartita} is smaller in stature than \textit{A. tridentata} 
+and may interact differently with perennial grasses. More importantly, cover of sagebrush can increase dramatically on overgrazed sites, 
+especially under a spring grazing regime. Competitive release of grasses following sagebrush removal often increases in 
+a nonlinear fashion, with little increase in production at sites with low initial sagebrush cover, 
+and accelerating responses as initial sagebrush cover passes some threshold \citep{McDaniel2005}. The cover of 
+sagebrush in our ungrazed quadrats may have fallen below that threshold. }
+
+The weak interspecific interactions among many of the dominant species in our system implies that these species  
+are limited by quite different factors. This low niche overlap could reflect strong resource partitioning \citep{tilman_resource_1982}, 
+perhaps combined with variation in the times and locations at which species are taking up those resources. 
+Species-specific responses to spatial and temporal variation in non-resource environmental variables can also weaken interspecific 
+interactions \citep{chesson_mechanisms_2000}, though in a previous study we failed to find evidence of a temporal storage effect 
+in this community \citep{adler_weak_2009}. Interactions with above- or belowground enemies can regulate populations and, given 
+enough species-specificity in the effects of these enemies, decouple the dynamics of different species dynamics \citep{hersh_evaluating_2011,
+janzen_herbivores_1970,connell_role_1971}.  More likely, plants may alter the environment, and even resource availability, 
+in ways that facilitate their neighbors. For example, shade cast by \textit{A. tripartita} canopies may reduce 
+evapotranspiration (e.g. \citealt{Barbier2008}). \textit{A. tripartita} removal might offset this facilitative effect by reducing 
+resource uptake, resulting in little net change in water balance following removal. 
 Unfortunately, we did not monitor soil moisture at the fine spatial and temporal resolution necessary to test this hypothesis.
 
-Evidence for weak interspecific interactions and limited competitive release has interesting implications.
-If most interspecific interactions are weak, then single-species models may be adequate for forecasting population responses to environmental change \citep{adler_forecasting_2012,kleinhesselink_indirect_2015,Chu2016}.
-Similarly, differences among species in relative abundance should reflect variation in the strength of intraspecific density dependence more than variation in the strength of interspecific interactions \citep{yenni_strong_2012}.
-A recent global study showing little variation in the abundance of invasive grassland species in native and invaded ranges on different continents  \citep{firn_abundance_2011} is consistent with the idea that intraspecific interactions are the primary determinant of species abundances.
-In systems where this is true, effective management and conservation requires understanding the sources and species-specificity of intraspecific density dependence.
-A useful first step in this direction would be to test whether our model is also successful at predicting the effects of release from intraspecific competition
-
-For many species in other communities, interspecific competition may still be an important driver.
-In closed-canopy forests where competition for light is intense, we might have found different results than in our system, in which competition for a number of belowground resources is more important than competition for light.
-However, a number of recent studies from forests have also found that heterospecific competitive effects are dramatically weaker than conspecific effects \citep{comita_asymmetric_2010,Kunstler2016,johnson_conspecific_2012}.
-We also expect stronger competition in successional systems, or communities in the early stages of community assembly where the process of competitive exclusion is still underway \citep{kokkoris_patterns_1999}.
-Finally, we might have found stronger evidence of competition if we could have studied rare species in our system, but only if those species are rare because of niche overlap with stronger competitors.
-If they are rare because they occupy unique, but narrow, niches, interspecific interactions would still be weak.
-Determining why interspecific interactions and competitive release are strong in some ecological contexts but weak in others remains a fascinating research problem.
+Evidence for weak interspecific interactions and limited competitive release has interesting implications. If most 
+interspecific interactions are weak, then single-species models may be adequate for forecasting population responses 
+to environmental change \citep{adler_forecasting_2012,kleinhesselink_indirect_2015,Chu2016}. Similarly, differences among 
+species in relative abundance should reflect variation in the strength of intraspecific density dependence more than 
+variation in the strength of interspecific interactions \citep{yenni_strong_2012}. A recent global study showing little 
+variation in the abundance of invasive grassland species in native and invaded ranges on different 
+continents  \citep{firn_abundance_2011} is consistent with the idea that intraspecific interactions are the primary determinant 
+of species abundances. In systems where this is true, effective management and conservation requires understanding the sources 
+and species-specificity of intraspecific density dependence. A useful first step in this direction would be to test 
+whether our model is also successful at predicting the effects of release from intraspecific competition
+
+For many species in other communities, interspecific competition may still be an important driver. In closed-canopy 
+forests where competition for light is intense, we might have found different results than in our system, in which 
+competition for a number of belowground resources is more important than competition for light. However, a 
+number of recent studies from forests have also found that heterospecific competitive effects are dramatically 
+weaker than conspecific effects \citep{comita_asymmetric_2010,Kunstler2016,johnson_conspecific_2012}.  We also 
+expect stronger competition in successional systems, or communities in the early stages of community assembly 
+where the process of competitive exclusion is still underway \citep{kokkoris_patterns_1999}. Finally, we might have
+found stronger evidence of competition if we could have studied rare species in our system, but only if those species 
+are rare because of niche overlap with stronger competitors. If they are rare because they occupy unique, but narrow, niches, 
+interspecific interactions would still be weak. Determining why interspecific interactions and competitive release are 
+strong in some ecological contexts but weak in others remains a fascinating research problem.
 
 \section*{Author contributions}
-
-PBA and ARK designed the experiment and supervised data collection with the help of JBT; PBA, SPE, GH and BT analyzed data and constructed models; 
-PBA wrote the first draft of the manuscript and all authors contributed to editing.
+PBA and ARK designed the experiment and supervised data collection with the help of JBT; PBA, SPE, GH and BT analyzed data \new{and 
+developed models}; PBA wrote the first draft of the manuscript and all authors contributed to editing.
 
 \section*{Acknowledgements}
-
 Funding was provided by NSF grants DEB-1353078,and  DEB-1054040 to PBA and DEB-1353039 to SPE and GH, and by the Utah Agricultural Experiment Station. 
@@
 $R_j(v,u,\vec{n})=c_{0,j}(v)e^{u}\Phi_j$ \citep{adler_coexistence_2010} where $c_{0,j}$ is the
-size distribution of new recruits. $\Phi_j$ is calculated from eqn. \ref{eqn:recrDataModel} with total cover values computed from the population densities as $C'_{j,t} = \int e^u n_j(u,t) du$.
+size distribution of new recruits. $\Phi_j$ is calculated from eqn. (\ref{eqn:recrDataModel}) with total cover values computed from the population densities as $C'_{j,t} = \int e^u n_j(u,t) du$.
 
@@
 than close plants. Otherwise, we let the data dictate the shape of the kernel by fitting a spline model 
-using the methods of \citep{teller_linking_2016}. The shape of $F_{jm}$ is determined by a set of spline basis coefficients $\vec{b}_{jm}$
+using the methods of \citet{teller_linking_2016}. The shape of $F_{jm}$ is determined by a set of spline basis coefficients $\vec{b}_{jm}$
 and a smoothing parameter $\eta$ that controls the complexity of the fitted kernel. 
@@
 \subsection{Interspecific covariance in local crowding} 
-We explored interspecific covariance in local crowding experienced by individual plants, by regressing the $W$ values exerted by one neighbor species, the response variable, against the $W$ values of all other species, the independent variables.
-Because some $W=0$, we conducted two separate regressions.
-First, using all $W$'s, we fitted a generalized linear model with a logit link function to evaluate whether the probability that the focal species' $W=0$ is influenced by the value of other species' $W$'s.
-In this model, the dependent variable is a Bernoulli variate coding for the zero or non-zero value of the focal species' crowding, and the independent variables are the $W$'s for all other species.
-Second, for the set of records in which the focal species has $W>0$, we performed a linear regression, where the focal species' $W$ is the dependent variable, and the other species' $W$'s are the independent variables.
-We repeated these regressions for each focal species.
-Due to large samples size, interspecific $W$ values were often statistically significant predictors of intraspecific.
-However, they explained very little variance.
-The maximum reduction in deviance for the generalized linear regressions and $R^2$ for the linear regressions were both less than 8\%.
-The \texttt{R} code for this analysis is included as ..\texttt{\textbackslash Wdistrib\textbackslash exploreSurvivalWs.r}.
+We explored interspecific covariance in local crowding experienced by individual plants, by regressing the $W$ values exerted by one neighbor species,
+ the response variable, against the $W$ values of all other species, the independent variables. Because some $W=0$, we conducted two 
+ separate regressions. First, using all $W$'s, we fitted a generalized linear model with a logit link function to evaluate whether t
+ he probability that the focal species' $W$ equals 0 is influenced by the value of other species' $W$'s. In this model, the dependent 
+ variable is a Bernoulli variate coding for zero versus non-zero value of the focal species' crowding, and the independent 
+ variables are the $W$'s for all other species. Second, for the set of records in which the focal species has $W>0$, we 
+ performed a linear regression, where the focal species' $W$ is the dependent variable, and the other species' $W$'s are 
+ the independent variables. We repeated these regressions for each focal species. Due to large samples size, interspecific $W$ 
+ values were often statistically significant predictors of intraspecific. However, they explained very little variance. The 
+ maximum reduction in deviance for the generalized linear regressions and $R^2$ for the linear regressions were both less 
+ than 8\%. The \texttt{R} code for this analysis is included as ..\texttt{\textbackslash Wdistrib\textbackslash exploreSurvivalWs.r}.
 
@@
 \hline
-AIC                   & 227.74           & 114.65           & 205.41           & 168.17           \\
-BIC                   & 240.01           & 124.69           & 218.18           & 181.40           \\
+%AIC                   & 227.74           & 114.65           & 205.41           & 168.17           \\
+%BIC                   & 240.01           & 124.69           & 218.18           & 181.40           \\
 Log Likelihood        & -108.87          & -52.33           & -97.70           & -79.09           \\
@@
   \includegraphics[width=1\textwidth]{boxplots-maxCI}
-  \caption{Equilibrium cover of the four dominant species simulated by the IPM with maximum removal treatment effects.
-Boxplots show interannual variation reflecting random year effects.
-Gray boxes show cover simulated by a model with \textit{A.
-tripartita} present and no removal treatment coefficients, brown boxes show results from the same model but with \textit{A.
-tripartita} cover set to zero (a species removal), green boxes show results from a model with  \textit{A.
-tripartita} set to zero and (mean) removal treatment coefficients included, and blue boxes show results with removal treatments coefficients assigned the outermost value of the 95\% confidence intervals.
-Source file: \texttt{ipm\textbackslash IPM-figures.r}.}
+  \caption{Equilibrium cover of the four dominant species simulated by the IPM with maximum removal treatment effects. Boxplots show interannual variation in cover, 
+  resulting from random year effects in the demographic models. Gray boxes show cover simulated by a model with \textit{A. tripartita} present and no removal treatment 
+  coefficients, brown boxes show results from the same model but with \textit{A. tripartita} cover set to zero (a species removal), green boxes show results from a 
+  model with  \textit{A. tripartita} set to zero and (mean) removal treatment coefficients included, and blue boxes show results with removal treatments 
+  coefficients assigned the outermost value of the 95\% confidence intervals.  Source file: \texttt{ipm\textbackslash IPM-figures.r}.}
   \label{fig:IPMresults-maxCI}

### 2017-05-12 removals/manuscript/removal_ms.tex
SPE edits

@@
 
-During the period of our experiment, annual water-year (October-September) precipitation fell within the 5\% - 95\% quantiles of the long-term mean, although more years fell below the mean than above it (\ref{fig:climate}a).
-However, annual mean temperatures were consistently warm relative to the long-term mean, and two years were warmer than the 95\% quantile (Figure \ref{fig:climate}b).
+During the period of our experiment, annual water-year (October-September) precipitation fell within the 5\% - 95\% quantiles of the 
+long-term mean, although more years fell below the mean than above it (\ref{fig:climate}a). However, annual mean temperatures 
+were consistently warm relative to the long-term mean, and two years were warmer than the 95\% quantile (Figure \ref{fig:climate}b).
 
@@
 However, these treatment comparisons do not directly address our questions about whether 
-responses to removal were adequately predicted by models fit to historical, observational data.
+responses to removal were adequately predicted by models fit to historical, observational data. \comment{SPE: is it ``historical, observational data'' or was the ``baseline'' model
+fitted to observational data both historical and modern?}
 
@@
 
-Our vital rate models take into account the effects of local species composition and should therefore be able to explain the effect of removals on survival, growth and recruitment without any additional parameters.
-If the removal treatment effects that we added to these models are statistically significant, it would indicate an effect of removals either larger or smaller than the effect predicted by a model based on data
-from the historical control plots. 
-\comment{SPE: is the baseline model just based on the historical control plots data, or also on the modern data from control plots?} 
+Our vital rate models take into account the effects of local species composition and should therefore be able 
+to explain the effect of removals on survival, growth and recruitment without any additional parameters. 
+\new{SPE: As in our previous studies using only historical data, the neighborhood competition coefficients in the 
+fitted models imply that interspecific competition is much weaker than intraspecific competition, in many
+cases by 1-2 orders of magnitude (Table \ref{table:alphas}).}  
+If the removal treatment effects that we added to these models are statistically significant, it would indicate an effect of 
+removals either larger or smaller than the effect predicted by a model based on data from the historical control plots. 
+\comment{SPE: is the baseline model just based on the historical data, or also on the modern data from control plots? 
+For me, ``historical control plots'' is ambiguous on this point.} 
 For the survival models, the 95\% confidence intervals on the removal effects overlapped zero for all four species (Fig. \ref{fig:VitalRateTest}; 
@@
 
-For three of the four species, we find little or no evidence that models based on observational data underestimate the intensity of interspecific competition. After accounting 
+For three of the four species, we find little or no evidence that models based on observational data underestimate 
+the intensity of interspecific competition. After accounting 
 for local neighborhood crowding, the removal treatment effects were never significant for \textit{A. tripartita} or \textit{H. comata}; 
@@
 \caption{Summary of fixed effects for the \textit{P.
-secunda} model with treatment*year effects (the ``trtYears*" coefficients).
+secunda} growth model with treatment*year effects (the ``trtYears*" coefficients).
 See Table \ref{ARTRgrowth} for an explanation of other coefficient names and column header.}
@@
 \caption{Summary of fixed effects for the \textit{P.
-spicata} model with treatment*year effects (the ``trtYears*" coefficients).
+spicata} growth model with treatment*year effects (the ``trtYears*" coefficients).
 See Table \ref{ARTRgrowth} for an explanation of other coefficient names and column headers.}
@@
 
+\newpage 
+\begin{table}[ht]
+\centering
+\caption{Summary of competition coefficients in the demographic models. These are the ``W.abcd'' fixed effects coefficients from
+the Tables in this section summarizing the survival, growth and recruitment models for each species.  
+In each $4 \times 4$ array of coefficients, the entry in row $i$, column $j$ is the 
+coefficient for the effect of species $j$ cover on species $i$. Intraspecific competition is weak relative to intraspecific competition if each
+species affects itself more than others, i.e. if the largest (in magnitude) element in each column is the element on the diagonal of the array.
+The one exception to this pattern is the effect of ARTR on recruitment by POSE, but the nonconforming
+coefficient ($\omega$[3,1] = -0.6589 in Table \ref{table:recruitment}) has low imprecision (sd = 0.3342, Table \ref{table:recruitment}.} 
+\label{table:alphas}
+\begin{tabular}{lrrrr}
+       & ARTR &  HECO &  POSE &  PSSP \\
+\hline 
+\underline{Survival} & & & & \\        
+ARTR   & -1.5214  & -0.0694  & -0.0049 &  0.0281\\
+HECO   & -0.0064  &  -0.6647 &  0.0732 &  0.0190\\
+POSE  &   0.0001&  -0.0153 &  -1.2632&   0.0249\\
+PSSP  &  0.0101 &  0.0046 & 0.0245 &  -0.4430\\
+\underline{Growth} & & & & \\ 
+ARTR   & -0.0364 &  0.0033& -0.0573&  0.0062\\
+HECO   &  -0.0031 & -0.0479& 0.0242 & -0.0151\\
+POSE  &  -0.0002& 0.0039&  -0.2641&   -0.0073\\
+PSSP  & -0.0025 & -0.0118 & -0.0067 & -0.1480\\
+\underline{Recruitment} & & & & \\ 
+ARTR  & -0.5881 & 0.0635 & 0.0220 & 0.1164 \\
+HECO   &  -0.4800 & -1.7368& 0.0388 &  -0.3682\\
+POSE  &  -0.6589& -0.1021& -1.8943 &   -0.1953\\
+PSSP  & -0.1867 &-0.3746  & 0.1131& -1.7379\\
+\hline
+\end{tabular}
+\end{table}
+
 \clearpage

### 2017-05-12 removals/manuscript/cover.tex
Incorporated Ellner and Hooker edits to removal manuscript

@@
 
-We are hereby submitting our manuscript, ``Weak interspecific interactions in a sagebrush steppe: evidence from observations, models, and experiments''
-for publication as a Standard Paper in the Journal of Ecology.
+Please consider our manuscript, ``Weak interspecific interactions in a sagebrush steppe: evidence from observations, models, and experiments''
+for publication as an Article in \textit{Ecology}.
 
@@
 We expect this conclusion may surprise a broad audience of plant ecologists accustomed to studies emphasizing the importance of interspecific competition.
+Our results also support the use of observational data to build multispecies population models and explore community dynamics.
 
@@
 
-\bibliographystyle{Ecology}
-\bibliography{RemovalRefs}
-
-\newpage
-
 \textbf{References}
 
-Adler, P.B., Dalgleish, H.J. \& Ellner, S.P. (2012) Forecasting plant community impacts of climate variability and change: when do competitive interactions matter? Journal of Ecology, 100, 478-487.
+Adler, P.B., Dalgleish, H.J. and Ellner, S.P. 2012. Forecasting plant community impacts of climate variability and change: when do competitive interactions matter? Journal of Ecology 100: 478-487.
 
-Adler, P.B., Ellner, S.P. \& Levine, J.M. (2010) Coexistence of perennial plants: an embarrassment of niches. Ecology Letters, 13, 1019-1029.
+Adler, P.B., Ellner, S.P. and Levine, J.M. 2010. Coexistence of perennial plants: an embarrassment of niches. Ecology Letters 13: 1019-1029.
 
-Chu, C. \& Adler, P.B. (2015) Large niche differences emerge at the recruitment stage to stabilize grassland coexistence. Ecological Monographs, 85, 373-392.
+Chu, C. and Adler, P.B. 2015. Large niche differences emerge at the recruitment stage to stabilize grassland coexistence. Ecological Monographs 85: 373-392.
 

### 2017-05-12 removals/manuscript/removal_ms.tex
Incorporated Ellner and Hooker edits to removal manuscript

@@
 
-%\new{A previous study of four dominant species from a sagebrush steppe community used historical data sets and multispecies population models to demonstrate weak interspecific interactions and predict little competitive release.
-We conducted a removal experiment to test our model's ability to accurately describe interspecific interactions.
-Our removal treatments were 1) removal of all perennial grass and 2) removal of the dominant shrub, \textit{Artemisia tripartita}.} We modeled survival, growth and recruitment as functions of local neighborhood species composition along with an indicator variable for removal treatment.
-If our ``baseline" model, which accounts for local plant-plant interactions, explains responses to removal, then the fitted removal indicator effect should be non-significant.
-For survival and recruitment, the removal treatment effects were never significantly positive, indicating that the basline model did not underestimate competitive effects.
-For  \textit{Poa secunda} recruitment, the removal treatment was negative, indicating that our baseline model overestimated competition.
-For growth, the removal treatment effect was significant and positive for two species, \textit{Poa secunda} and \textit{Pseudoroegneria spicata}, indicating that the baseline model underestimated competitive effects.
-However, including information about the location of grass individuals with respect to removed \textit{A.
-tripartita} failed to improve the growth regressions, raising questions about the mechanisms driving the positive response to removal.
-For three species, population models based on the vital rate regressions showed little impact of removals on population growth rates or simulated equilibrium cover.
-For \textit{P.
-spicata}, the population models that included effects of \textit{A.
-tripartita} removal projected greater competitive release than the baseline models.
-The ability of models based on observational data to predict population responses to experimental removals increased our confidence that interspecific interactions are in fact weak for at least three of our four study species.
-The implication is that other factors besides interspecific resource competition play important roles in determining the structure and dynamics of this community.
-The success of the models also supports the use of multispecies models, based on long-term observational studies, to characterize competitive interactions.
-}
-
-\new{A previous study of four dominant species in a sagebrush steppe community used historical data sets and multispecies population models to demonstrate weak interspecific interactions, predicted that little competitive release would result from removal of allospecific competitors.
+A previous study of four dominant species in a sagebrush steppe community used historical data sets and multispecies population models to demonstrate weak interspecific interactions and predicted that little competitive release would result from removal of heterospecific competitors.
 We conducted a removal experiment to test whether our model, with weak interspecific competition, accurately describes interspecific interactions.
@@
 tripartita} failed to improve the growth regressions, raising questions about the mechanisms driving the positive response to removal.
-For three species, population models based on the vital rate regressions showed little impact of removals on population growth rates or simulated equilibrium cover.
+For three species, population models based on the vital rate regressions showed little impact of removals on population growth rates or equilibrium cover.
 For \textit{P.
@@
 The implication is that factors besides interspecific resource competition play important roles in determining the structure and dynamics of this community.
-The success of the models also supports the use of multispecies models based on long-term observational data for  characterizing competitive interactions.} \comment{346 words, max=350 at Ecology}
+The success of the models also supports the use of multispecies models based on long-term observational data for  characterizing competitive interactions.
 
@@
 Applying this approach to arid and semiarid plant communities, we found evidence of strong intraspecific limitation and weak interspecific interactions, leading to very stable coexistence and implying little competitive release \citep{adler_coexistence_2010,chu_large_2015}.
-\new{ In fact, at our sagebrush steppe study site  \citep{adler_coexistence_2010}, our model predicts removal of sagebrush would actually reduce the equilibrium cover of the three dominant grass species very slightly, reflecting the facilitative effect of sagebrush on grass recruitment and very weak competitive effects on growth and survival.
-Since we only studied the most common, co-occurring species, we expected to find evidence of stable coexistence, but the predicted negative effect of sagebrush removal on grass cover is surprising given the long history of sagebrush eradication to increase grass forage production in
- sagebrush rangelands \citep{Cook1963, Mueggler1958, Robertson1947}.} % end new
+In fact, at our sagebrush steppe study site  \citep{adler_coexistence_2010}, our model predicts removal of sagebrush would actually reduce the equilibrium cover of the three dominant grass species very slightly, reflecting the facilitative effect of sagebrush on grass recruitment and very weak competitive effects on growth and survival.
+Since we only studied the most common, co-occurring species, we expected to find evidence of stable coexistence, but the predicted negative effect of sagebrush removal on grass cover is surprising given the long history of sagebrush eradication to increase grass forage production in sagebrush rangelands \citep{Cook1963, Mueggler1958, Robertson1947}.
 Compared to traditional field experiments, our approach has the advantage of projecting full life-cycle responses over long time periods, but its reliance on observational data sets limits the strength of our inference.
@@
 
-To test whether our sagebrush community model, which is based on observational data, \new{underestimates the strength of interspecific competition and competitive release}, we conducted a removal experiment in a sagebrush steppe.
+To test whether our sagebrush community model, which is based on observational data, underestimates the strength of interspecific competition and competitive release, we conducted a removal experiment in a sagebrush steppe.
 We studied the responses of three perennial grass species to removal of sagebrush, and the response of sagebrush to removal of all grasses.
-\new{The primary purpose of the experiment was not to directly measure competitive release, which is difficult to do over a short time period in a semiarid ecosystem with low and variable plant cover, but rather to validate the parameters in our model that represent key interspecific interactions and ultimately determine the magnitude of competitive release.} We addressed two research questions: First, are the survival, growth or recruitment of our target species significantly different in control vs.
+The primary purpose of the experiment was not to directly measure competitive release, which is difficult to do over a short time period in a semiarid ecosystem with low and variable plant cover, but rather to validate the parameters in our model that represent key interspecific interactions and ultimately determine the magnitude of competitive release.
+We addressed two research questions: First, are the survival, growth or recruitment of our target species significantly different in control vs.
 removal treatments after accounting for differences in local species composition?
@@
 
-\new{Our regression models for survival, growth and recruitment follow previous work \citep{adler_coexistence_2010,chu_large_2015}.
+Our regression models for survival, growth and recruitment follow previous work \citep{adler_coexistence_2010,chu_large_2015}.
 The current analysis departs from previous work with the inclusion of a categorical removal treatment effect.
-Although the three vital rate models differ in functional form (see Appendix A for complete model descriptions), all of them have a linear predictor, $\mu$, containing the following terms:}
+Although the three vital rate models differ in functional form (see Appendix A for complete model descriptions), all of them have a linear predictor, $\mu$, containing the following terms:
 \begin{equation}
@@
 \end{equation}
-\new{where $i$ refers to an individual genet (for survival and growth) or quadrat (for recruitment) of species $j$, $\gamma$ is a time-dependent intercept (year effect), $\varphi$ is the coefficient for the effect of spatial location (`$g$' represents a group of nearby quadrats located within one pasture or grazing exclosure), $\chi$ is the removal treatment effect,
+where $i$ refers to an individual genet (for survival and growth) or quadrat (for recruitment) of species $j$, $\gamma$ is a time-dependent intercept (year effect), $\varphi$ is the coefficient for the effect of spatial location (`$g$' represents a group of nearby quadrats located within one pasture or grazing exclosure), $\chi$ is the removal treatment effect,
 $W_{ijm,t}$ is the crowding on species $j$ exerted by neighbors of species $m$ at the individual (survival and growth) or
 quadrat (recruitment) scale, and $\omega_{jm}$ is an interaction coefficient which determines the impact of crowding by species $m$ on the 
-focal species $j$. }
-
-\new{If the conspecific and heterospecific crowding effects, the $\omega_{jm}$'s, explain demographic rates in removal treatment plots as well as they do in control plots, then there will be no need for the removal treatment effect, and the estimated coefficient will be small and not significantly different from zero.
+focal species $j$.
+
+If the conspecific and heterospecific crowding effects, the $\omega_{jm}$'s, explain demographic rates in removal treatment plots as well as they do in control plots, then there will be no need for the removal treatment effect, and the estimated coefficient will be small and not significantly different from zero.
 However, if demographic rates in removal plots differ from control plots even after accounting for crowding effects, then the removal treatment effect will be significant.
@@
 We refer to the model that includes the estimated removal treatment effect as the ``removal model".
-Note that these are not estimated as separate models; we estimate all parameters together, but when we generate predictions from the baseline model, we simply set the removal treatment effects to zero.}
+Note that these are not estimated as separate models; we estimate all parameters together, but when we generate predictions from the baseline model, we simply set the removal treatment effects to zero.
 
@@
 comparison is to look for systematic deviations between model predictions and observations, we used a deterministic version of the model (e.g., 
-recruitment is the $\lambda$ of (\ref{eqn:recrProcessModel}), rather than a Poisson random variable with mean $\lambda$). \new{Specifically, we used the recruitment regression to calculate 
+recruitment is the $\lambda$ of (\ref{eqn:recrProcessModel}), rather than a Poisson random variable with mean $\lambda$). Specifically, we used the recruitment regression to calculate 
 the expected number of new recruits at time $t+1$, which we then multiplied by the observed mean size of recruits for the focal species to get expected cover of new recruits. 
-We used the survival regression to predict the survival probability of each genet, and the growth regression to predict the expected size (on arithmetic scale) of each genet at time $t+1$ (note that the
-expected size on arithmetic scale is not exp(expected log size), because of Jensen's inequality). 
-Expected cover at $t+1$ is expected area of new recruits, plus the product of survival probability and expected new size summed over all genets present at time $t$. }
+We used the survival regression to predict the survival probability of each genet, and the growth regression to predict the expected size (on arithmetic scale) of each genet at time $t+1$ (note that the expected size on arithmetic scale is not exp(expected log size), because of Jensen's inequality).
+Expected quadrat-level cover at $t+1$ is expected area of new recruits, plus the product of survival probability and expected new size summed over all genets present at time $t$. 
 Because we do not project forward beyond one time step, we do not need to assign locations to the new recruits. For control plots, we used the baseline version of the model with all 
@@
 \subsection*{Integral projection models}
-
-\comment{PBA: should we move much of this to Appendix A? SPE: DONE}
 
@@
 to compute neighborhood competition as a function of individual size for the IPM.
-
-%$S$ is given by (Eqn. \ref{eqn:survReg}) and $G$ is given by (Eqn. \ref{eqn:growReg}), using an expected neighborhood crowding
-%index calculated from the size distributions of the competing species. In the vital rate regressions and in the IBM, each 
-%individual has a unique neighborhood crowding index based on the spatial locations and sizes of neighboring plants. This 
-%measure cannot be extended to the IPM, which tracks individual size but not location. Instead we used a 
-%mean-field approximation that captures the essential features of spatially-explicit neighborhood interactions \citep{adler_coexistence_2010}. 
-%We found that in both the observed data and IBM simulations, heterospecific individuals were approximately randomly distributed with respect to each other, 
-%but conspecific individuals displayed a non-random, size-dependent patterns: small genets were randomly distributed, while large genets 
-%were segregated from each other. The overdispersion of large conspecific genets was introduced into the IPM 
-%with a `no-overlap' rule \citep{adler_coexistence_2010}. In Appendix \ref{sec:kernelMethods} we explain how the spline competition kernels with the no-overlap rule were used 
-%to compute neighborhood crowding index as a function of individual size for the IPM.  
-%
-%For recruitment in the IPM, the factor $ \Phi_j  =\exp{\left(\gamma_{j,t}^R +  \varphi_{jg}^R + \chi_{jh}^R + 
-%\sum \limits_{m} \omega_{jm}^R \sqrt{C'_{mq,t}} \right)}$
-%in eqn. (\ref{eqn:recrDataModel}) gives the expected 
-%area of new recruits per quadrat, per unit area of parents. To incorporate this recruitment function into the IPM, we assumed that individual fecundity increases 
-%linearly with individual area (equal to $e^u$ for a size-$u$ individual). The recruitment kernel is thus
-%$R_j(v_j,u_j,\vec{n})=c_{0,j}(v_j)e^{u_j}\Phi_j$ \citep{adler_coexistence_2010} where $c_{0,j}$ is the
-%size distribution of new recruits. $\Phi_j$ is calculated from eqn. \ref{eqn:recrDataModel} with total cover values computed from the population densities as $C'_{j,t} = \int e^u n(u_{j,t}) du$.
 
@@
 (Fig. \ref{fig:CoverTrends}), but shrub removal did appear to cause an increase in the cover of \textit{P. spicata}. 
-However, \new{when we tested the effects of removal on cover by analyzing year-to-year changes in (log) cover, a measure of population 
-growth rate, none of the removal treatment effects were significantly different from zero (Table \ref{table:coefficients}). For \textit{P. spicata}, 
+However, when we tested the effects of removal treatments on year-to-year population growth rates ($log(Cover_{t+1}/Cover_t)$) at the quadrat level,
+ none of the removal treatment effects were significantly different from zero (Table \ref{table:coefficients}). For \textit{P. spicata}, 
 the 95\% confidence interval only just contained zero (95\% CI: -0.01, 0.38), providing marginal evidence for a positive effect of removal on population growth. 
-Fitting with weighted least squares to address heteroscedasticity in residuals indicated even less evidence for a positive 
-response of  \textit{P. spicata} to removal.} 
-\comment{SPE: I think you need to explain how did ``analyzing year-to-year changes in (log) cover'' -- what was the response, what were the
-covariates? I think reviewers will also want to know what was done. And without that information, 
-it's not clear what the residuals would be, or what weighted least squares would entail. } 
-However, these treatment comparisons do not directly address our questions about whether 
-responses to removal were adequately predicted by models fit to historical, observational data. \comment{SPE: is it ``historical, observational data'' or was the ``baseline'' model
-fitted to observational data both historical and modern?}
+Fitting with weighted least squares to address the high variance in residuals for plots with very low cover indicated even less evidence for a positive 
+response of  \textit{P. spicata} to removal. However, these treatment comparisons do not directly address our questions about whether 
+responses to removal were adequately predicted by models fit to observational data.
 
@@
 to explain the effect of removals on survival, growth and recruitment without any additional parameters. 
-\new{SPE: As in our previous studies using only historical data, the neighborhood competition coefficients in the 
+As in our previous studies using only the historical data, the neighborhood competition coefficients in the 
 fitted models imply that interspecific competition is much weaker than intraspecific competition, in many
-cases by 1-2 orders of magnitude (Table \ref{table:alphas}).}  
+cases by 1-2 orders of magnitude (Table \ref{table:alphas}).  
 If the removal treatment effects that we added to these models are statistically significant, it would indicate an effect of 
-removals either larger or smaller than the effect predicted by a model based on data from the historical control plots. 
-\comment{SPE: is the baseline model just based on the historical data, or also on the modern data from control plots? 
-For me, ``historical control plots'' is ambiguous on this point.} 
+removals either larger or smaller than the effect predicted by a model based on data from the control plots. 
 For the survival models, the 95\% confidence intervals on the removal effects overlapped zero for all four species (Fig. \ref{fig:VitalRateTest}; 
@@
  meaning that individuals of these species grew more rapidly in removal plots than controls even after accounting for neighborhood interactions 
- (Fig. \ref{fig:VitalRateTest}; Tables \ref{ARTRgrowth} to \ref{PSSPgrowth}). In contrast, \textit{Poa secunda} had lower recruitment in removal 
+ (Fig. \ref{fig:VitalRateTest}; Tables \ref{ARTRgrowth} to \ref{PSSPgrowth}). In contrast, \textit{P. secunda} had lower recruitment in removal 
  than controls plots (upper limit of 95\% CI $<0$; Fig. \ref{fig:VitalRateTest}; Table \ref{table:recruitment}). For the other species, 
@@
 tripartita} removal, after accounting for local neighborhood interactions.
+The other species codes are HECO: \textit{Hesperostipa comata}, POSE: \textit{Poa secunda}, and PSSP: \textit{Pseudoroegneria spicata}.
  Source file: \texttt{treatment\textunderscore test\textunderscore figure.r}.}
@@
 
-To quantify the short-term population-level consequences of removals, \new{we used an IBM to project the cover of each species in each quadrat one year ahead using versions of the model with and without removal treatment effects.
+To quantify the short-term population-level consequences of removals, we used an IBM to project the cover of each species in each quadrat one year ahead using versions of the model with and without removal treatment effects.
 Plotting observed vs.
@@
 spicata} were predictions from the removal model closer to the 1:1 line.
-Using the version of the model with maximum removal treatment effects (removal coefficients set to the 95\% confidence limit furthest from zero) increased differences between baseline and removal model predictions} (Fig.
-\ref{fig:obsVpred_maxCI}).
-\comment{PBA: I think this figure may be useless?
-cut?
-SPE: yes, cut.}
 
@@
 
-\new{Time series of predicted and observed cover, averaged over quadrats, show similar patterns.
+Time series of predicted and observed cover, averaged over quadrats, show similar patterns.
 Predicted population size closely tracked observed population size (Fig.
-\ref{fig:IBM1step}) in most years for all species, but the difference between predictions from the baseline and removal models varied among species.} For \textit{H.
+\ref{fig:IBM1step}) in most years for all species, but the difference between predictions from the baseline and removal models varied among species.
+For \textit{H.
 comata}, differences in predictions from the two models were virtually indistinguishable.
 For \textit{A.
-tripartita}, \textit{Poa secunda} and especially \textit{P.
+tripartita}, \textit{P.
+secunda} and especially \textit{P.
 spicata}, the model with removal effects consistently predicted higher population growth than the baseline model (Fig.
@@
 
-%Our population models based on historical, observational data reveal weak interspecific interactions \citep{adler_coexistence_2010,chu_large_2015} and, as a result, 
-%predict very little competitive release among the four dominant species in a sagebrush steppe. Despite the fact that previous removal studies have reported variable 
-%results, ranging from negative effects of removal on remaining species to complete competitive release, we were suspicious of our model's \new{ability to accurately 
-%describe interspecific interactions, so we conducted a removal experiment to test it.} The results provide substantial support for our model's predictions.
-
 Our population models based on historical, observational data reveal weak interspecific interactions \citep{adler_coexistence_2010,chu_large_2015}. As a result, the 
@@
 data, we were suspicious of our model's ability to accurately describe interspecific interactions, so we conducted a removal experiment to test it. 
-The results provide substantial support for our model's predictions, and for the weakness of interspecific interactions in this community. \comment{SPE has
-revised this paragraph. The previous version is in the tex file, commented out.}
+The results provide substantial support for our model's predictions, and for the weakness of interspecific interactions in this community.
 
@@
 
-\new{How can we reconcile the insensitivity of grasses to sagebrush removal in our study with evidence from rangeland ecology 
+How can we reconcile the insensitivity of grasses to sagebrush removal in our study with evidence from rangeland ecology 
 research showing that sagebrush eradication increases grass production? The most likely explanation involves ecological context: our 
@@
 and accelerating responses as initial sagebrush cover passes some threshold \citep{McDaniel2005}. The cover of 
-sagebrush in our ungrazed quadrats may have fallen below that threshold. }
+sagebrush in our ungrazed quadrats may have fallen below that threshold.
 
@@
 \section*{Author contributions}
-PBA and ARK designed the experiment and supervised data collection with the help of JBT; PBA, SPE, GH and BT analyzed data \new{and 
-developed models}; PBA wrote the first draft of the manuscript and all authors contributed to editing.
+PBA and ARK designed the experiment and supervised data collection with the help of JBT; PBA, SPE, GH and BT analyzed data and 
+developed models; PBA wrote the first draft of the manuscript and all authors contributed to editing.
 
@@
 \centering
-\caption{Summary of fixed effects for the \textit{A. tripartita} survival model. ``logarea" is the effect of plant size, ``Treatment*" is the removal effect, and the ``W.*" coefficients 
-are effects of neighborhood crowding (the $\omega$s in eqn.
+\caption{Summary of posterior distributions for fixed effects for the \textit{A.
+tripartita} survival model.
+``logarea" is the effect of plant size, ``Treatment*" is the removal effect, and the ``W.*" coefficients are effects of neighborhood crowding (the $\omega$s in eqn.
 \ref{eqn:survReg}).
@@
 \centering
-\caption{Summary of fixed effects for the \textit{H. comata} survival model. See Table \ref{ARTRsurvival} for an explanation of coefficient names
+\caption{Summary of posterior distributions for fixed effects for the \textit{H. comata} survival model. See Table \ref{ARTRsurvival} for an explanation of coefficient names
 and column headers.} 
@@
 \centering
-\caption{Summary of fixed effects for the \textit{P. secunda} survival model. See Table \ref{ARTRsurvival} for an explanation of coefficient names
+\caption{Summary of posterior distributions for fixed effects for the \textit{P. secunda} survival model. See Table \ref{ARTRsurvival} for an explanation of coefficient names
 and column headers.} 
@@
 \centering
-\caption{Summary of fixed effects for the \textit{P. spicata} survival model. See Table \ref{ARTRsurvival} for an explanation of coefficient names
+\caption{Summary of posterior distributions for fixed effects for the \textit{P. spicata} survival model. See Table \ref{ARTRsurvival} for an explanation of coefficient names
 and column headers.} 
@@
 \centering
-\caption{Summary of fixed effects for the \textit{A. tripartita} growth model. ``logarea.t0" is the effect of plant size, ``Treatment*" is the removal effect, and the ``W.*" coefficients 
-are effects of neighborhood crowding (the $\omega$s in eqn.
+\caption{Summary of posterior distributions for fixed effects for the \textit{A.
+tripartita} growth model.
+``logarea.t0" is the effect of plant size, ``Treatment*" is the removal effect, and the ``W.*" coefficients are effects of neighborhood crowding (the $\omega$s in eqn.
 \ref{eqn:growReg}.
@@
 \centering
-\caption{Summary of fixed effects for the \textit{H. comata} growth model. See Table \ref{ARTRgrowth} for an explanation of coefficient names
+\caption{Summary of posterior distributions for fixed effects for the \textit{H. comata} growth model. See Table \ref{ARTRgrowth} for an explanation of coefficient names
 and column headers.} 
@@
 \centering
-\caption{Summary of fixed effects for the \textit{P. secunda} growth model. See Table \ref{ARTRgrowth} for an explanation of coefficient names
+\caption{Summary of posterior distributions for fixed effects for the \textit{P. secunda} growth model. See Table \ref{ARTRgrowth} for an explanation of coefficient names
 and column headers.} 
@@
 \centering
-\caption{Summary of fixed effects for the \textit{P. spicata} growth model. See Table \ref{ARTRgrowth} for an explanation of coefficient names
+\caption{Summary of posterior distributions for fixed effects for the \textit{P. spicata} growth model. See Table \ref{ARTRgrowth} for an explanation of coefficient names
 and column headers.} 
@@
 \centering
-\caption{Summary of fixed effects for the recruitment model (symbols correspond to Eqns. \ref{eqn:recrDataModel} and \ref{eqn:recrProcessModel}).} 
+\caption{Summary of posterior distributions for fixed effects for the recruitment model (symbols correspond to Eqns.
+\ref{eqn:recrDataModel} and \ref{eqn:recrProcessModel}).
+The indexing on $\omega$ shown here and used in the computer code for the recruitment model gives $\omega$[3,1] as the effect of species 3 on species 1.
+This is the reverse of the normal convention used in \ref{eqn:recrProcessModel} and shown in \ref{table:alphas}.}
 \label{table:recruitment}
@@
 species affects itself more than others, i.e. if the largest (in magnitude) element in each column is the element on the diagonal of the array.
-The one exception to this pattern is the effect of ARTR on recruitment by POSE, but the nonconforming
-coefficient ($\omega$[3,1] = -0.6589 in Table \ref{table:recruitment}) has low imprecision (sd = 0.3342, Table \ref{table:recruitment}.} 
+The one exception to this pattern is the effect of POSE on recruitment by ARTR, but the nonconforming
+coefficient ($\omega$[3,1] = -0.6589 in Table \ref{table:recruitment}) has low precision (sd = 0.3342, Table \ref{table:recruitment}.} 
 \label{table:alphas}
@@
 \underline{Recruitment} & & & & \\ 
-ARTR  & -0.5881 & 0.0635 & 0.0220 & 0.1164 \\
-HECO   &  -0.4800 & -1.7368& 0.0388 &  -0.3682\\
-POSE  &  -0.6589& -0.1021& -1.8943 &   -0.1953\\
-PSSP  & -0.1867 &-0.3746  & 0.1131& -1.7379\\
+ARTR  & -0.5881 &  -0.4800 & -0.6589 & -0.1867\\
+HECO   &   0.06346 & -1.7368 & -0.1021 & -0.3746\\
+POSE  &   0.0220 &  0.0388 & -1.8943 &  0.1131\\
+PSSP  &  0.1164 & -0.3682 & -0.1953 & -1.7379\\
 \hline
@@
  \centering
-\includegraphics[width=1\textwidth]{obsVpred_quad_year_maxCI}
-\caption{The relationship between observed and predicted cover for each quadrat, year, and species assuming maximum effects of removal.
-The 1:1 line is dashed, the solid black line shows the trend for control plots with predictions from the baseline IBM model (black symbols), the blue line shows the trend for removal plots with predictions from the baseline model (blue symbols), and the red line shows the trend for removal plots with predictions from the removal model (red symbols).
-Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r}.
-}
-\label{fig:obsVpred_maxCI}
-\end{figure}
-
- \begin{figure}[tbp]
- \centering
  \includegraphics[width=1\textwidth]{cover_projections_1step_maxCI}

### 2017-05-16 removals/manuscript/removal_ms.tex
trivia

@@
 of species $j$ is represented by a density function $n_j(u,t)$ which gives the density of genets of size $u$ at time $t$ (``size'' is natural log of genet area, so 
-that $n_j(u,t)h$ is the number of genets whose area is between between $\exp(u)$ and $\exp(u+h)$). The size distribution function at time $t+1$ is given by
+that $n_j(u,t)h$ is the number of genets whose area is between between $\exp(u)$ and $\exp(u+h)$ for $h \ll 1$). The size distribution function at time $t+1$ is given by
 \begin{equation}
@@
 Appendix \ref{AppendixA} for a complete description. As in these previous studies, the IPM uses a mean-field approximation for
-neighborhood comeptition that captures the essential features of spatially-explicit local interactions; 
+neighborhood competition that captures the essential features of spatially-explicit local interactions; 
 in Appendix \ref{sec:kernelMethods} we explain how the spline competition kernels were used 

### 2017-05-30 removals/manuscript/removal_ms.tex
Added text on ARTR response to grass removal, cleaned up files

@@
 To quantify the long-term population-level consequences of permanent competitor removals, we used an IPM to project the long-term average cover of each species.
-We concentrated on grass species' responses to \textit{A.
-tripartita} removal, since the growth regressions indicated our baseline model might underestimate competitive release in this situation (our baseline vital rate models worked well for predicting \textit{A.
-tripartita} responses to grass removal).
-Using the baseline model, with no removal treatment coeffiicients, the IPM predicts little effect of \textit{A.
-tripartita} removal on the cover of any grass species (compare gray and brown boxes in Fig.
+The IPM predicts little effect of \textit{A.
+tripartita} removal on the cover of any grass species when we simulate the baseline model with no removal treatment coeffiicients (compare gray and brown boxes in Fig.
 \ref{fig:IPMresults}).
@@
 
+ \new{The baseline model predicts more dramatic competitive release of \textit{A.
+tripartita} following grass removal: mean cover increases from 5.6\% to 16.2\%.
+Adding removal treatment effects dramatically alters the outcome, with \textit{A.
+tripartita} decreasing towards extinction.
+Both outcomes primarily reflect parameters in the recruitment model.
+The baseline model predicts an increase in cover following grass removal because of the relatively strong negative effects of \textit{Poa secunda} and \textit{H.
+comata} on \textit{A.
+tripartita} recruitment (\ref{table:recruitment}).
+The removal effects model reverses that outcome because the estimated coefficient for the grass removal treatment effect is very large and negative.
+However, all of these recruitment parameters are estimated with considerable uncertainty.
+For example, the effect of \textit{Poa secunda} on \textit{A.
+tripartita} is estimated to be -0.66, but the standard deviation of the estimate is 0.33 and the 95\% credible interval stops just short of zero.
+The grass removal treatment effect has a credible interval that broadly overlaps zero (\ref{table:recruitment}).
+}
+
 \section*{Discussion}
@@
 on \textit{P. secunda} is positive). At the population level, models with and without removal effects projected similar patterns of 
-short (Fig. \ref{fig:IBM1step}) and long-term (Fig. \ref{fig:IPMresults}) dynamics for \textit{H. comata} and \textit{A. tripartita}. 
+short (Fig.
+\ref{fig:IBM1step}) and long-term (Fig.
+\ref{fig:IPMresults}) dynamics for \textit{H.
+comata}.
+\new{ For \textit{A.
+tripartita}, the experimental observations and short-term population projections show little response to removal, while the long-term simulations from the baseline and removal effects models suggest strong but inconsistent responses to removal.
+However, these long-term simulations primarily reflect the influence of recruitment parameters that are highly uncertain.
+We have little confidence in the recruitment model
+for \textit{A.
+tripartita} given the species' low (a mean of 0.63 recruits per quadrat per year, compared to means of 2.6 to 5.3 for the grasses) and episodic recruitment (only 13 \textit{A.
+tripartita} recruits observed since we resumed annual censuses in 2007).
+Given this uncertainty, we put more weight on the direct observations and short-term projections showing little effect of removal on \textit{A.
+tripartita} populations.
+}
 For \textit{P. secunda}, the model with removal effects predicted higher population growth over the short but not the long-term, perhaps 

### 2017-05-30 removals/ipm/IPM-figures.r

@@ -1,20 +1,6 @@
 
 # call from removal_analysis_wrapper.r
 
-# plot means
-simResults <- read.csv(simFile)
-if(nrow(simResults)>3) simResults <- simResults[1:3,]
-
-
-png("ipm/IPMsims.png",height=3,width=4.5,units="in",res=400)
-  par(tcl=-0.2,mgp=c(2,0.5,0),mar=c(3,3,1,1))
-  myCol <- c("black","dodgerblue3","red3")
-  tmp <- barplot(as.matrix(simResults),beside=T,names.arg=rep("",4),col=myCol,ylab="Cover (%)")
-  axis(side=1,at=tmp[2,],sppList,cex.axis=0.9)
-  legend("topright",c("Baseline","Baseline, no ARTR","Removal, no ARTR"),
-         fill=myCol,bty="n",cex=0.9)
-dev.off()
-
 # box plots
 if(!exists("baseline")) baseline <- read.csv("ipm/baselineCover.csv")
 if(!exists("baseline.noARTR")) baseline.noARTR <- read.csv("ipm/baselineCover-noARTR.csv")

### 2017-05-30 removals/removal_analysis_wrapper.r

@@ -194,20 +194,6 @@ sppList <-  c("ARTR","HECO","POSE","PSSP")
 
 source("ipm/get_W_functions.r")  # get neighbor distance decay functions
 
-#start with all 4 species, remove ARTR
-init.species <- c(1:4)
-tlimit <- 400
-burn.in <- 0
-totReps=100
-trtEffects=F
-source("ipm/IPM-setup.r")
-output <- array(NA,c(tlimit,4,totReps))
-for(doRep in 1:totReps){
-  source("ipm/IPM-removeARTR.r")
-  output[,,doRep] <- covSave
-}
-meanCov <- apply(output,MARGIN=c(1,2),FUN=mean)
-
 #no treatment effects, all species
 init.species <- c(1:4)
 tlimit <- 2500
@@ -250,6 +236,22 @@ meanCover4 <- meanCover
 # source("ipm/IPM-getEquilibrium.r")
 # write.csv(covSave[(burn.in+1):tlimit,],"ipm/removalCover-noARTRnoPSSP.csv",row.names=F)
 
+# baseline model (no treatment effects), remove all grasses
+init.species <- c(1)
+trtEffects=F
+max.CI=F
+source("ipm/IPM-setup.r")
+source("ipm/IPM-getEquilibrium.r")
+write.csv(covSave[(burn.in+1):tlimit,],"ipm/baselineCover-noGrass.csv",row.names=F)
+
+# baseline model (with treatment effects), remove all grasses
+init.species <- c(1)
+trtEffects=T
+max.CI=F
+source("ipm/IPM-setup.r")
+source("ipm/IPM-getEquilibrium.r")
+write.csv(covSave[(burn.in+1):tlimit,],"ipm/removalCover-noGrass.csv",row.names=F)
+
 simResults <- rbind(meanCover1,meanCover2,meanCover3) 
 colnames(simResults) <- sppList
 write.csv(simResults,"ipm/simResults-meanCover.csv",row.names=F)
@@ -269,7 +271,13 @@ Rpars$intcpt.yr[,1] <- Rpars$intcpt.yr[,1]+1
 source("ipm/IPM-getEquilibrium.r")
 print(rbind(meanCover1,meanCover)) # compare baseline run with this one
 
-simFile <- "ipm/simResults-meanCover.csv"
+source("ipm/IPM-figures.r") # plot results of ARTR removals
 
-source("ipm/IPM-figures.r")
[…]

### 2017-05-30 removals/manuscript/removal_ms.tex
ms edits by SPE

--- 2017-05-16 removals/manuscript/removal_ms.tex
@@
 tripartita} response to grass removal because the removal treatment effects in the vital rate regressions were never significantly different from zero, indicating that the baseline model effectively predicted competitive release.
-\comment{PBA: Just do it?
-SPE: Just do it, and then explain why there is no difference between ``grass removal'' and ``grass removal with removal treatment effect''.
-}
 
@@
 \ref{fig:CoverTrends}).
-Could the small competitive release following simulated \textit{A.
-tripartita} removal simply reflect the low \textit{A.
+Could the small competitive release \new{of the grasses}
+following simulated \textit{A.
+tripartita} removal simply reflect the low simulated \textit{A.
 tripartita} cover?
 To explore this possibility, we increased \textit{A.
-tripartita} cover in the baseline simulation by increasing average \textit{A.
-tripartita} fecundity.
+tripartita} cover in the baseline simulation by increasing its average fecundity.
 The increased fecundity led to a doubling of \textit{A.
-tripartita} average cover, but had essentially zero effect on the cover of the grasses (changes of less than 0.1\% cover).
+tripartita} average cover, but had essentially zero effect on the cover of the grasses (changes of less than 0.1\% cover), \new{implying that shrub removal from higher cover
+would still have little effect on the grasses.}
+
+\new{The baseline model predicts more dramatic competitive release of \textit{A. tripartita} following grass removal: mean cover increases from 5.6\% to 16.2\%. Adding
+removal treatment effects dramatically alters the outcome, with \textit{A. tripartita} decreasing towards extinction. However, these outcomes both are driven
+by parameters that are estimated with very high uncertainty (Table~\ref{table:recruitment}).
+The baseline model predicts an increase following grass removal because of the relatively strong negative effects of \textit{Poa secunda} and \textit{H.
+comata} on \textit{A.
+tripartita} recruitment (\ref{table:recruitment}).
+The estimated effect of
+\emph{Poa} is -0.66, but the standard deviation of the estimate is 0.33 and the 95\% credible interval stops just short of zero; the 95\% credible interval for the
+\textit{H. comata} effect includes values less than half of the point-estimate. The grass removal treatment effect, which leads to predicted extinction, has a credible 
+interval that broadly overlaps zero (Table~\ref{table:recruitment}). \textit{A. tripartita} recruitment parameters have exceptionally high uncertainty, presumably 
+because \textit{A. tripartita} recruitment is episodic (see \emph{Discussion}), and this produces extreme uncertainty about its long-term 
+response to grass removal.}
 
@@
 on \textit{P. secunda} is positive). At the population level, models with and without removal effects projected similar patterns of 
-short (Fig. \ref{fig:IBM1step}) and long-term (Fig. \ref{fig:IPMresults}) dynamics for \textit{H. comata} and \textit{A. tripartita}. 
+short (Fig. \ref{fig:IBM1step}) and long-term (Fig. \ref{fig:IPMresults}) dynamics for \textit{H. comata}. \new{ For \textit{A. tripartita}, 
+the experimental observations and short-term population projections show little response to removal, while the long-term simulations 
+from the baseline and removal effects models suggest strong (but opposite) effects of removal: a tripling of abundance in the baseline model, and extinction in the
+removal effects model. However, these long-term simulations primarily reflect the influence of recruitment parameters 
+that are highly uncertain (see \emph{Results}). This high uncertainty 
+results from \textit{A. tripartita} recruitment being low (a mean of 0.63 recruits per quadrat per year, compared to means of 2.6 to 5.3 for the grasses) and 
+very episodic (roughly half of the total recruits in the 22 years of historical data were observed in 2 of those years, and only 13 \textit{A. tripartita} 
+recruits were observed since we resumed annual censuses in 2007). Given this uncertainty, we put more weight on the direct observations 
+and short-term projections showing little effect of removal on \textit{A. tripartita} populations.}
 For \textit{P. secunda}, the model with removal effects predicted higher population growth over the short but not the long-term, perhaps 

### 2017-05-31 removals/manuscript/removal_ms.tex
Final run and edits before Ecology submission

@@
 We tested for the effect of removal treatment on this measure of cover change with a linear mixed effects model implemented in the \texttt{lme4} package \citep{Bates2015} of R, including year and quadrat as random effects.
-We considered the removal treatment effect to be significant when the 95\% confidence interval on the estimated coefficient did not overlap zero.
+We considered the removal treatment effect to be significant when the 95\% credible interval on the estimated coefficient did not overlap zero.
 Because variance in residuals of plots with low cover was high for two species, we explored fitting by weighted least squares, to account for the apparent heteroscedasiticity.
@@
 Perhaps large uncertainty in the estimates of these effects makes it difficult to reject the null hypothesis represented by our baseline model.
-To address this possibility, we also simulated a version of the model in which we replaced the point estimate of each removal treatment effect with either the 5\% or 95\% confidence limit for the treatment effect,whichever was farthest from zero.
+To address this possibility, we also simulated a version of the model in which we replaced the point estimate of each removal treatment effect with either the 5\% or 95\% credible limit for the treatment effect, whichever was farthest from zero.
 This version of the model shows the population-level changes we would expect given maximum removal effects.
@@
 To the extent that the baseline model underestimates (or overestimates) the actual competitive release, the results of the second and third scenarios will differ.
-As with the IBM, we repeated these simulations using maximum removal treatment effects represented by parameters values at the 5\% or 95\% confidence limit farthest from zero.
+As with the IBM, we repeated these simulations using maximum removal treatment effects represented by parameters values at the 5\% or 95\% credible limit farthest from zero.
 We do not show simulations of long-term \textit{A.
@@
  none of the removal treatment effects were significantly different from zero (Table \ref{table:coefficients}). For \textit{P. spicata}, 
-the 95\% confidence interval only just contained zero (95\% CI: -0.01, 0.38), providing marginal evidence for a positive effect of removal on population growth. 
+the 95\% credible interval only just contained zero (95\% CI: -0.01, 0.38), providing marginal evidence for a positive effect of removal on population growth. 
 Fitting with weighted least squares to address the high variance in residuals for plots with very low cover indicated even less evidence for a positive 
@@
 removals either larger or smaller than the effect predicted by a model based on data from the control plots. 
-For the survival models, the 95\% confidence intervals on the removal effects overlapped zero for all four species (Fig. \ref{fig:VitalRateTest}; 
+For the survival models, the 95\% credible intervals on the removal effects overlapped zero for all four species (Fig. \ref{fig:VitalRateTest}; 
 Tables \ref{ARTRsurvival} to \ref{PSSPsurvival}), meaning that the survival probabilities of plants in removal plots can be predicted based on data 
-from control plots. For the growth models, the 95\% confidence intervals on the removal effects were positive for \textit{P. secunda}  and \textit{P. spicata},
+from control plots. For the growth models, the 95\% credible intervals on the removal effects were positive for \textit{P. secunda}  and \textit{P. spicata},
  meaning that individuals of these species grew more rapidly in removal plots than controls even after accounting for neighborhood interactions 
@@
  \caption{Effects of removal treatments on the survival, growth, and recruitment of the four dominant species.
-Circles show the estimated coefficient of the removal effect, bars show 95\% confidence intervals.
+Circles show the estimated coefficient of the removal effect, bars show 95\% credible intervals.
 The removal effect shows response of \textit{A.
@@
 \ref{fig:IPMresults}).
-Running a version of the model with maximum removal treatment effects (removal coefficients set to the 95\% confidence limit furthest from zero), resulted in the same qualitative outcome, but a larger increase in \textit{P.
+Running a version of the model with maximum removal treatment effects (removal coefficients set to the 95\% credible limit furthest from zero), resulted in the same qualitative outcome, but a larger increase in \textit{P.
 spicata} (Fig.
@@
 \ref{fig:CoverTrends}).
-Could the small competitive release \new{of the grasses}
+Could the small competitive release of the grasses
 following simulated \textit{A.
@@
 The increased fecundity led to a doubling of \textit{A.
-tripartita} average cover, but had essentially zero effect on the cover of the grasses (changes of less than 0.1\% cover), \new{implying that shrub removal from higher cover
-would still have little effect on the grasses.}
-
-\new{The baseline model predicts more dramatic competitive release of \textit{A. tripartita} following grass removal: mean cover increases from 5.6\% to 16.2\%. Adding
+tripartita} average cover, but had essentially zero effect on the cover of the grasses (changes of less than 0.1\% cover), implying that even if shrub cover were higher its removal
+would still have little effect on the grasses.
+
+The baseline model predicts more dramatic competitive release of \textit{A. tripartita} following grass removal: mean cover increases from 5.6\% to 16.2\%. Adding
 removal treatment effects dramatically alters the outcome, with \textit{A. tripartita} decreasing towards extinction. However, these outcomes both are driven
@@
 because \textit{A. tripartita} recruitment is episodic (see \emph{Discussion}), and this produces extreme uncertainty about its long-term 
-response to grass removal.}
+response to grass removal.
 
@@
 on \textit{P. secunda} is positive). At the population level, models with and without removal effects projected similar patterns of 
-short (Fig. \ref{fig:IBM1step}) and long-term (Fig. \ref{fig:IPMresults}) dynamics for \textit{H. comata}. \new{ For \textit{A. tripartita}, 
+short (Fig. \ref{fig:IBM1step}) and long-term (Fig. \ref{fig:IPMresults}) dynamics for \textit{H. comata}. For \textit{A. tripartita}, 
 the experimental observations and short-term population projections show little response to removal, while the long-term simulations 
@@
 recruits were observed since we resumed annual censuses in 2007). Given this uncertainty, we put more weight on the direct observations 
-and short-term projections showing little effect of removal on \textit{A. tripartita} populations.}
+and short-term projections showing little effect of grass removal on \textit{A. tripartita} populations.
 For \textit{P. secunda}, the model with removal effects predicted higher population growth over the short but not the long-term, perhaps 
@@
 Convergence was observed graphically for all parameters, and confirmed by the Brooks-Gelman statistic \citep{brooks_general_1998}.
-For all three vital rate models, we considered the effect of removal treatments, or other coefficients of interest, to be significant when the 95\% confidence interval (or posterior interval) on the estimate of $\chi$ did not overlap zero.
+For all three vital rate models, we considered the effect of removal treatments, or other coefficients of interest, to be significant when the 95\% credible interval (or posterior interval) on the estimate of $\chi$ did not overlap zero.
 
@@
 \caption{Statistical models of year-to-year changes in log(cover) for the four focal species. Values in square brackets show 95\%
- confidence intervals.}
+ credible intervals.}
 \centering
@@
 \hline
-\multicolumn{5}{l}{\scriptsize{$^*$ 0 outside the confidence interval}}
+\multicolumn{5}{l}{\scriptsize{$^*$ 0 outside the credible interval}}
 \end{tabular}
@@
   model with  \textit{A. tripartita} set to zero and (mean) removal treatment coefficients included, and blue boxes show results with removal treatments 
-  coefficients assigned the outermost value of the 95\% confidence intervals.  Source file: \texttt{ipm\textbackslash IPM-figures.r}.}
+  coefficients assigned the outermost value of the 95\% credible intervals.  Source file: \texttt{ipm\textbackslash IPM-figures.r}.}
   \label{fig:IPMresults-maxCI}

### 2017-05-31 removals/manuscript/removal_ms.tex
Final edits before submission, changed colors in Fig 5

@@
 
-\author[a]{Peter B. Adler\thanks{Corresponding author. Department of Wildland Resources and the Ecology Center, Utah State University, Logan, Utah Email: peter.adler@usu.edu}}
-\author[a]{Andrew Kleinhesselink}
-\author[b]{Giles Hooker}
-\author[c]{J. Bret Taylor}
-\author[a]{Brittany Teller}
-\author[d]{Stephen P. Ellner}
-\affil[a]{Department of Wildland Resources and the Ecology Center, Utah State University, Logan, Utah}
-\affil[b]{Department of Biological Statistics and Computational Biology, Cornell University, Ithaca, New York}
-\affil[c]{USDA, Agricultural Research Service, U. S. Sheep Experiment Station, 19 Office Loop, Dubois, ID, USA}
-\affil[d]{Department of Ecology and Evolutionary Biology, Cornell University, Ithaca, New York}
+\author[1]{Peter B. Adler\thanks{Corresponding author. Department of Wildland Resources and the Ecology Center, Utah State University, Logan, Utah Email: peter.adler@usu.edu}}
+\author[1]{Andrew Kleinhesselink}
+\author[2]{Giles Hooker}
+\author[3]{J. Bret Taylor}
+\author[1]{Brittany Teller}
+\author[4]{Stephen P. Ellner}
+\affil[1]{Department of Wildland Resources and the Ecology Center, Utah State University, Logan, Utah}
+\affil[2]{Department of Biological Statistics and Computational Biology, Cornell University, Ithaca, New York}
+\affil[3]{USDA, Agricultural Research Service, U. S. Sheep Experiment Station, 19 Office Loop, Dubois, ID, USA}
+\affil[4]{Department of Ecology and Evolutionary Biology, Cornell University, Ithaca, New York}
 
@@
 \maketitle
-
-\textbf{\large{Keywords:}} Coexistence, competition, integral projection model, removal experiment, sagebrush steppe.
 
@@
 
+\textbf{\large{Keywords:}} Coexistence, competition, integral projection model, removal experiment, sagebrush steppe.
+
 \section*{Introduction}
@@
 
-\section*{Methods}
+\section*{Materials and Methods}
 
@@
 Because variance in residuals of plots with low cover was high for two species, we explored fitting by weighted least squares, to account for the apparent heteroscedasiticity.
-The results of these tests were qualitatively similar, so opted to only report the simpler analysis.
+The results of these tests were qualitatively similar, so we opted to only report the simpler analysis.
 
@@
 We calculated the equilibrium cover for each species by simulating an additional 2000 time steps and averaging each species' cover over this period.
-For the second scenario, we set the cover of \textit{A.
-tripartita} to zero and then repeated the procedure.
-This simulation projects the competitive release of grasses following \textit{A.
-tripartita} removal according to the baseline model (no removal effects).
+For the second scenario, we simulated our experimental removals by setting either the cover of \textit{A.
+tripartita} to zero or the cover of all three grass species to zero, and then simulating the dynamics of the remaining species as under the first scenario.
+This simulation projects competitive release following removal according to the baseline model (no removal effects).
 For the third scenario, we again set \textit{A.
-triparitia} cover to zero, but this time we projected the model using the vital rate regressions that include removal effects.
+triparitia} or grass cover to zero, but this time we projected the model using the vital rate regressions that include removal effects.
 If our baseline model successfully predicts competitive release observed in the removal experiment, then the results of the second and third scenarios should be similar.
@@
 As with the IBM, we repeated these simulations using maximum removal treatment effects represented by parameters values at the 5\% or 95\% credible limit farthest from zero.
-We do not show simulations of long-term \textit{A.
-tripartita} response to grass removal because the removal treatment effects in the vital rate regressions were never significantly different from zero, indicating that the baseline model effectively predicted competitive release.
 
@@
 During the period of our experiment, annual water-year (October-September) precipitation fell within the 5\% - 95\% quantiles of the 
-long-term mean, although more years fell below the mean than above it (\ref{fig:climate}a). However, annual mean temperatures 
-were consistently warm relative to the long-term mean, and two years were warmer than the 95\% quantile (Figure \ref{fig:climate}b).
+long-term mean, although more years fell below the mean than above it (Fig. \ref{fig:climate}a in Appendix C). However, annual mean temperatures 
+were consistently warm relative to the long-term mean, and two years were warmer than the 95\% quantile (Fig. \ref{fig:climate}b).
 
@@
 The IPM predicts little effect of \textit{A.
-tripartita} removal on the cover of any grass species when we simulate the baseline model with no removal treatment coeffiicients (compare gray and brown boxes in Fig.
+tripartita} removal on the cover of any grass species when we simulate the baseline model with no removal treatment coeffiicients (compare gray and blue boxes in Fig.
 \ref{fig:IPMresults}).
-In fact, projected average cover of \textit{H.
-comata} and \textit{P.
+In fact, projected average cover of \textit{P.
+secunda} and \textit{P.
 spicata} declines slightly following \textit{A.
@@
 Gray boxes show cover simulated by a model with \textit{A.
-tripartita} present and no removal treatment coefficients, brown boxes show results from the same model but with \textit{A.
-tripartita} cover set to zero (a species removal), and green boxes show results from a model with \textit{A.
+tripartita} present and no removal treatment coefficients, blue boxes show results from the same model but with \textit{A.
+tripartita} cover set to zero (a species removal), and red boxes show results from a model with \textit{A.
 tripartita} set to zero and removal treatment coefficients included.
@@
 The equilibrium cover of \textit{A.
-tripartita} simulated by the IPM (6.0\%) is consistent with observed cover during the historical data collection period (6.6\% from 1931-1956), but is low compared to contemporary observed cover (15-25\%, Fig.
+tripartita} simulated by the IPM (5.4\%) is consistent with observed cover during the historical data collection period (6.6\% from 1931-1956), but is low compared to contemporary observed cover (15-25\%, Fig.
 \ref{fig:CoverTrends}).
@@
 The increased fecundity led to a doubling of \textit{A.
-tripartita} average cover, but had essentially zero effect on the cover of the grasses (changes of less than 0.1\% cover), implying that even if shrub cover were higher its removal
-would still have little effect on the grasses.
-
-The baseline model predicts more dramatic competitive release of \textit{A. tripartita} following grass removal: mean cover increases from 5.6\% to 16.2\%. Adding
+tripartita} average cover, but had essentially zero effect on the cover of the grasses (changes of less than 0.1\% cover), implying that even if shrub cover were higher its removal would still have little effect on the grasses.
+
+The baseline model predicts more dramatic competitive release of \textit{A. tripartita} following grass removal: mean cover increases from 5.4\% to 16.2\%. Adding
 removal treatment effects dramatically alters the outcome, with \textit{A. tripartita} decreasing towards extinction. However, these outcomes both are driven
@@
 
-%\newpage
+\newpage
+\renewcommand{\refname}{Literature cited}
 \bibliographystyle{Ecology}
@@
   resulting from random year effects in the demographic models. Gray boxes show cover simulated by a model with \textit{A. tripartita} present and no removal treatment 
-  coefficients, brown boxes show results from the same model but with \textit{A. tripartita} cover set to zero (a species removal), green boxes show results from a 
-  model with  \textit{A. tripartita} set to zero and (mean) removal treatment coefficients included, and blue boxes show results with removal treatments 
+  coefficients, blue boxes show results from the same model but with \textit{A. tripartita} cover set to zero (a species removal), red boxes show results from a 
+  model with  \textit{A. tripartita} set to zero and (mean) removal treatment coefficients included, and orange boxes show results with removal treatments 
   coefficients assigned the outermost value of the 95\% credible intervals.  Source file: \texttt{ipm\textbackslash IPM-figures.r}.}

### 2017-05-31 removals/ipm/IPM-figures.r

@@ -7,7 +7,7 @@ if(!exists("baseline.noARTR")) baseline.noARTR <- read.csv("ipm/baselineCover-no
 if(!exists("removal.noARTR")) removal.noARTR <- read.csv("ipm/removalCover-noARTR.csv")
 if(!exists("removal.noARTR.maxCI")) removal.noARTR.maxCI <- read.csv("ipm/removalCover-noARTR-maxCI.csv")
 
-myCols<-c("darkgrey","darkgoldenrod","darkgreen")
+myCols<-c("darkgrey","blue3","red3")
 sppNames <-c("A. tripartita","H. comata","P. secunda","P. spicata")
 
 png("ipm/boxplots.png",height=3.5, width=8, units="in",res=400)
@@ -28,10 +28,10 @@ png("ipm/boxplots-maxCI.png",height=3.5, width=8, units="in",res=400)
   plot(c(1:17),rep(NA,17),ylim=c(0,10),ylab="Cover (%)",xlab="",xaxt="n")
   for(doSpp in 1:4){
     tmp <- cbind(baseline[,doSpp],baseline.noARTR[,doSpp],removal.noARTR[,doSpp],removal.noARTR.maxCI[,doSpp])
-    boxplot(100*tmp,col=c(myCols,"blue"),add=T, at=(seq(1:4)+(doSpp-1)*4),names=rep("",4),xaxt="n")
+    boxplot(100*tmp,col=c(myCols,"orange"),add=T, at=(seq(1:4)+(doSpp-1)*4),names=rep("",4),xaxt="n")
   }
   axis(side=1,at=c(3,7,11,15),labels=sppNames,font=4)
   abline(v=4.5);abline(v=8.5);abline(v=12.5)
   legend("topright",c("Baseline model","Baseline model, no ARTR","Removal model, no ARTR","Removal model, no ARTR, max CI"),
-         fill=c(myCols,"blue"),bg="white",cex=0.9)
+         fill=c(myCols,"orange"),bg="white",cex=0.9)
 dev.off()

### 2017-07-07 precip/analysis/figure_scripts/plot_significant_effect_comparison.R
adding some old changes

@@ -41,6 +41,8 @@ all_effects <-
   mutate( type = ifelse(type == 'TRUE_FALSE' | type == 'FALSE_TRUE', 'mixed', type )) %>% 
   mutate( sig_label = ifelse(type == 'TRUE_TRUE', 'significant', 'not significant'))
 
+head(all_effects)
+
 overall_cor <- all_effects %>% ungroup %>% summarise( cor = cor(mu.x, mu.y))
 overall_cor
 
@@ -55,7 +57,6 @@ cors$label <- paste0( 'r=', round( cors$cor, 2))
 all_effects <- unique(all_effects)
 
 nrow( all_effects )
-
 
 p1 <- 
   ggplot(all_effects, aes( x = mu.x, y = mu.y, color = Treatment, shape = sig_label,  group = 1 )  ) + 
@@ -78,7 +79,7 @@ p1 <- p1 + geom_text( data = cors, aes( x  = x.pos, y = y.pos , label = label , 
 p1
 
 
-p1 <- p1 + geom_text( data = subset( all_effects, type == 'TRUE_TRUE' & species == 'POSE' & vital_rate == 'recruitment' & Treatment == 'Irrigation'), 
+p1 <- p1 + geom_text( data = subset( all_effects, species == 'POSE' & vital_rate == 'recruitment' & Treatment == 'Irrigation'), 
                 aes( label = paste(species, vital_rate, sep = '\n' )) , adj = -0.1, nudge_y = -0.2, show.legend = F) 
 
 p <- 
@@ -100,6 +101,6 @@ g$p
 
 png( 'figures/parameter_predictions.png' , height = 6, width = 8 , res = 300 , units = 'in')
 
-print( p1 + labs( color = '' )) 
+print( p1 ) 
 
 dev.off()

### 2017-07-07 precip/manuscript/overall_prediction_score.tex

% latex table generated in R 3.3.1 by xtable 1.8-2 package
% Thu Dec 22 04:09:23 2016
\begin{table}[ht]
\centering
\caption{Comparison of model predictions from climate model and year effects model for each species and vital rate.  Two prediction scores are reported, MSE and lppd. Lower MSE indicates improved predictions whereas higher lppd indicates improved predictions.  Instances where the climate model outperformed the random year effects model are marked with "***" in the last column. ARTR = \textit{A. tripartita}, HECO = \textit{H. comata}, POSE = \textit{P. secunda}, PSSP = \textit{P. spicata}.} 
\label{table:overallPreds}
\begin{tabular}{lllrrrl}
  \hline
species & vital\_rate & score & climate model & year effects model & diff & improved \\ 
  \hline
ARTR & growth & lppd & -186.08 & -189.17 & 3.09 & *** \\ 
  ARTR & growth & MSE & 0.55 & 0.55 & -0.00 & *** \\ 
  ARTR & recruitment & lppd & -79.05 & -77.81 & -1.24 &  \\ 
  ARTR & recruitment & MSE & 169.86 & 9.47 & 160.39 &  \\ 
  ARTR & survival & lppd & -37.55 & -36.42 & -1.13 &  \\ 
  ARTR & survival & MSE & 0.06 & 0.06 & 0.00 &  \\ 
  HECO & growth & lppd & -475.54 & -454.36 & -21.18 &  \\ 
  HECO & growth & MSE & 1.26 & 1.18 & 0.09 &  \\ 
  HECO & recruitment & lppd & -149.43 & -151.60 & 2.17 & *** \\ 
  HECO & recruitment & MSE & 288.16 & 235.57 & 52.59 &  \\ 
  HECO & survival & lppd & -158.99 & -147.06 & -11.93 &  \\ 
  HECO & survival & MSE & 0.12 & 0.11 & 0.01 &  \\ 
  POSE & growth & lppd & -1823.71 & -1831.56 & 7.85 & *** \\ 
  POSE & growth & MSE & 1.72 & 1.73 & -0.01 & *** \\ 
  POSE & recruitment & lppd & -260.05 & -257.46 & -2.59 &  \\ 
  POSE & recruitment & MSE & 45.40 & 37.32 & 8.08 &  \\ 
  POSE & survival & lppd & -698.06 & -718.60 & 20.54 & *** \\ 
  POSE & survival & MSE & 0.14 & 0.14 & -0.00 & *** \\ 
  PSSP & growth & lppd & -1232.93 & -1237.92 & 4.99 & *** \\ 
  PSSP & growth & MSE & 1.51 & 1.51 & 0.00 &  \\ 
  PSSP & recruitment & lppd & -271.34 & -273.94 & 2.60 & *** \\ 
  PSSP & recruitment & MSE & 79.09 & 42.68 & 36.41 &  \\ 
  PSSP & survival & lppd & -332.47 & -307.26 & -25.21 &  \\ 
  PSSP & survival & MSE & 0.11 & 0.10 & 0.01 &  \\ 
   \hline
\end{tabular}
\end{table}

### 2017-08-04 precip/manuscript/overall_cover_change_modified.tex
adding a bunch of old edits to scripts and manuscript files

% latex table generated in R 3.3.1 by xtable 1.8-2 package
% Tue Jan 17 10:14:56 2017
\begin{table}[ht]
\centering
\caption{Treatment effects on log cover change for each species from start of experiment in 2011 to last year of experiment in 2016. Intercept gives control effects. ARTR = \textit{A. tripartita}, HECO = \textit{H. comata}, POSE = \textit{P. secunda}, PSSP = \textit{P. spicata}.} 
\label{table:coverChange}
\begin{tabular}{rllll}

  \hline
ARTR & Estimate & Std.Error & t-value & Pr($\textgreater \lvert t \rvert $)   \\ 
  (Intercept) & -0.28 & 0.46 & -0.61 & 0.55 \\ 
  TreatmentDrought & 0.94 & 0.77 & 1.21 & 0.24 \\ 
  TreatmentIrrigation & 0.89 & 0.77 & 1.15 & 0.27 \\
  \hline 
HECO & Estimate & Std.Error & t-value & Pr($\textgreater \lvert t \rvert $) \\ 
  (Intercept) & 0.40 & 0.26 & 1.54 & 0.15 \\ 
  TreatmentDrought & -3.00 & 0.58 & -5.18 & 0.001 \\ 
  TreatmentIrrigation & -0.12 & 0.5 & -0.25 & 0.81 \\ 
    \hline

POSE & Estimate & Std.Error & t-value & Pr($\textgreater \lvert t \rvert $) \\ 
  (Intercept) & -0.72 & 0.46 & -1.57 & 0.13 \\ 
  TreatmentDrought & 0.03 & 0.82 & 0.03 & 0.97 \\ 
  TreatmentIrrigation & 1.15 & 0.78 & 1.47 & 0.16 \\ 
    \hline

PSSP & Estimate & Std.Error & t-value & Pr($\textgreater \lvert t \rvert $) \\ 
  (Intercept) & 0.02 & 0.21 & 0.09 & 0.93 \\ 
  TreatmentDrought & -0.89 & 0.38 & -2.34 & 0.03 \\ 
  TreatmentIrrigation & 0.15 & 0.38 & 0.38 & 0.70 \\ 
   \hline

\end{tabular}
\end{table}

### 2017-08-04 precip/manuscript/overall_prediction_score.tex
adding a bunch of old edits to scripts and manuscript files

@@
 % latex table generated in R 3.3.1 by xtable 1.8-2 package
-% Thu Dec 22 04:09:23 2016
+% Fri Dec 23 14:08:37 2016
 \begin{table}[ht]
 \centering
-\caption{Comparison of model predictions from climate model and year effects model for each species and vital rate.
+\caption{Comparison of model predictions from climate model and no climate model for each species and vital rate.
 Two prediction scores are reported, MSE and lppd.
 Lower MSE indicates improved predictions whereas higher lppd indicates improved predictions.
-Instances where the climate model outperformed the random year effects model are marked with "***" in the last column.
+Instances where the climate model outperformed the no climate model are marked with "***" in the last column.
 ARTR = \textit{A.
@@
   \hline
-species & vital\_rate & score & climate model & year effects model & diff & improved \\ 
+species & vital\_rate & score & climate model & baseline model & diff & improved \\ 
   \hline

### 2017-08-04 precip/manuscript/pgr_predictions.tex
adding a bunch of old edits to scripts and manuscript files

@@
 % latex table generated in R 3.3.1 by xtable 1.8-2 package
-% Thu Dec 22 04:07:36 2016
-\begin{table}[!h]
+% Tue Jan 17 13:20:20 2017
+\begin{table}[H]
 \centering
 \caption{MSE of predicted log cover changes and correlations between log cover changes predicted and observed.
-Predictions for the cover changes in the experimental plots were generated either from the year effects or the climate models.
-Instances where the climate model made better predictions than the year effects model are indicated with the "***".
+Predictions for the cover changes in the experimental plots were generated either from the climate or no climate models.
+Instances where the climate model made better predictions than the no climate model are indicated with the "***".
 ARTR = \textit{A.
@@
   \hline
- & species & stat & year effects model & climate model & diff & improved \\ 
+ & species & stat & baseline model & climate model & diff & improved \\ 
   \hline

### 2017-08-04 precip/R/climate/soilMoistureTreatmentEffects.R

@@ -86,13 +86,21 @@ summary(mTreatment)
 
 library(lsmeans)
 library(texreg)
+library(xtable )
+library(dplyr)
+test <- lsmeans(mTreatment,  ~ Treatment + season + rainfall)
+#test <- lsm(mTreatment, ~ Treatment + season + rainfall )
 
-lsmeans(mTreatment,  ~ Treatment + season + rainfall)
+
+tab <- summary(test)
+
+tab <-  data.frame(tab)
+
+tab <- as.data.frame(tab) %>%  dplyr::select(season, rainfall,  Treatment, lsmean, SE, asymp.LCL, asymp.UCL ) %>% arrange(season, rainfall, Treatment )
 
 statsOutput <- 'manuscript/soil_moisture_model.tex'
 
-
-texreg(mTreatment,caption="soil moisture model",
+texreg(mTreatment,caption="Treatment effects on soil moisture. Intercept refers to drought effects in fall not rainy conditions.  Model fit to the continuously logged soil moisture data as well as the spot measurements collected from all plots in the spring.",
        caption.above=TRUE,file=statsOutput, label = 'table:soil_moisture_model')
 
 

### 2017-08-04 precip/analysis/figure_scripts/plot_cover.R

@@ -3,6 +3,7 @@ library(ggplot2)
 library(stringr)
 library(dplyr)
 library(tidyr)
+library(gridExtra)
 
 load('analysis/figure_scripts/my_plotting_theme.Rdata')
 
@@ -24,6 +25,22 @@ obs_cover <-
   group_by(species, year, Treatment ) %>% 
   summarise( cover_obs = mean(area) )
 
+old_plot <- 
+  ggplot ( obs_cover %>% filter( year < 1960), aes( x = year, y = cover_obs*100/10000)) + 
+  geom_point() + 
+  geom_line() + 
+  ylab( 'Average Cover (%)') + 
+  scale_x_continuous(breaks = seq(1929, 1960, 5)) + 
+  my_theme + theme(strip.background = element_blank(), strip.text = element_blank()) + 
+  facet_wrap( ~ species, nrow = 2) 
+
+old_plot
+
+p <- obs_cover %>% filter(year < 1960) %>% group_by( species) %>% do(  p = old_plot  %+% . + ggtitle(unique(.$species)))
+p$p[[1]] + ylim ( 0, 13) 
+p$p[[2]] + ylim( 0, 5)
+
+
 pred_cover <- 
   pred %>% 
   group_by(species, model, year, Treatment, simulation) %>%
@@ -36,6 +53,7 @@ added <- expand.grid( year = 2011 , species = unique(pred_cover$species), model 
 pred_cover <- merge( added, pred_cover , all.x = T, all.y = T)
 
 all_cover <- merge( obs_cover, pred_cover, all.x = T)
+
 
 all_cover <- 
   all_cover %>% 
@@ -54,6 +72,7 @@ all_cover$type  <- factor( all_cover$type, labels = c('observed', 'predicted'))
 
 all_cover$val <- 100*(all_cover$val/10000)
 
+
 my_plot <- 
   ggplot( subset( all_cover, model == 'climate'), aes( x = year, y = val, linetype = type, color = Treatment)) + 
   geom_point() + 
@@ -68,13 +87,15 @@ g <- all_cover %>% filter( model == 'climate') %>% group_by(species) %>% do( p =
 
 g$p
 
-png( 'figures/predicted_and_observed_cover.png', height = 7, width = 12, units = 'in', res = 300)
+png( 'figures/predicted_and_observed_cover.png', height = 6, width = 8, units = 'in', res = 300)
 print( 
 grid.arrange( g$p[[1]] + guides( color = 'none', linetype = 'none') + ylim(0, 21), 
[…]

### 2017-08-04 precip/analysis/figure_scripts/plot_obs_vs_pred_treatment_parameters_for_presentation.R

@@ -1,0 +1,80 @@
+# plot obs v predicted treatment parameters together 
+library(ggplot2)
+library(dplyr)
+library(tidyr)
+
+fils <- dir('output/', 'predicted_and_observed_par_estimates.*.csv', full.names = T)
+
+dat <- lapply( fils, read.csv)
+
+dat <- do.call(rbind, dat)
+
+load('analysis/figure_scripts/my_plotting_theme.Rdata')
+
+dat$par <- factor(dat$par, labels = c('Intercept', 'Size x Treatment'))
+dat$type <- factor(dat$type, levels = c('predicted_effect', 'observed_effect'), ordered = T)
+
+blank_plot <- 
+  ggplot( data = dat, aes( x = Treatment, y = mu, ymin = lcl, ymax = ucl, color = Treatment, shape = type)) + 
+  geom_point(position = position_dodge(width = 0.3) , size = 4, color = NA) + 
+  geom_errorbar(position = position_dodge(width = 0.3), width = 0.3, color = NA) + 
+  geom_hline(aes(yintercept = 0), linetype = 2, alpha = 0.3 ) +
+  scale_color_manual(values = my_colors[3:4]) + 
+  facet_grid( par ~ species)  + 
+  ylab('Parameter estimate (+/- 95% BCI)') +
+  my_theme + theme(strip.background = element_blank())
+  
+png('figures/blank_treatment_effects.png', height = 5.8, width = 8 , units = 'in', res = 300)
+print(blank_plot)
+dev.off()
+
+par_plot <- 
+  ggplot( data = dat, aes( x = Treatment, y = mu, ymin = lcl, ymax = ucl, color = Treatment, shape = type)) + 
+  geom_point(position = position_dodge(width = 0.3) , size = 4) + 
+  geom_errorbar(position = position_dodge(width = 0.3), width = 0.3) + 
+  geom_hline(aes(yintercept = 0), linetype = 2, alpha = 0.3 ) +
+  scale_color_manual(values = my_colors[3:4]) + 
+  scale_shape_manual(values = c(17,19) ) + 
+  facet_grid( par ~ species)  + 
+  ylab('Treatment effects (+/- 95% Bayesian Credible Interval)') +
+  my_theme + 
+  theme(strip.background = element_blank(), 
+        strip.text = element_text(size = 12), 
+        axis.text = element_text(size = 10), 
+        axis.title = element_text(size = 12), 
+        axis.text.x = element_blank(), 
[…]

### 2017-08-04 precip/analysis/figure_scripts/plot_significant_effect_comparison.R

@@ -52,10 +52,12 @@ cors$x.pos <- c(1.5,1.5)
 cors$y.pos <- c(1.5,1.5)
 cors$label <- paste0( 'r=', round( cors$cor, 2))
 
+cors2 <- all_effects %>% group_by(Treatment) %>% summarise( cor = cor(mu.x, mu.y))
+cors2$x.pos <- c(1.5,1.5) 
+cors2$y.pos <- c(1.9,1.9)
+cors2$label <- paste0( 'r=', round( cors2$cor, 2))
+
 all_effects <- unique(all_effects)
-
-nrow( all_effects )
-
 
 p1 <- 
   ggplot(all_effects, aes( x = mu.x, y = mu.y, color = Treatment, shape = sig_label,  group = 1 )  ) + 
@@ -72,6 +74,50 @@ p1 <-
   scale_shape_manual(values = c(16,8), guide = 'none') + 
   facet_grid( . ~ par  )  + 
   coord_fixed()
+
+p2 <- 
+  ggplot(all_effects, aes( x = mu.x, y = mu.y, color = Treatment,  group = 1 )  ) + 
+  geom_abline(aes( intercept = 0, slope = 1), linetype = 2, color = 'black') + 
+  geom_point(size = 2.5, alpha = 0.7) + 
+  #geom_smooth(method = 'lm', se = F, color = 'darkgray', linetype = 2) +
+  geom_vline(aes(xintercept = 0) , linetype = 2 , alpha = 0.1) + 
+  geom_hline( aes(yintercept = 0 ), linetype = 2, alpha = 0.1)  + 
+  ylab( 'Observed Treatment Effect') + 
+  xlab( 'Predicted Treatment Effect') + 
+  scale_x_continuous(limits = c(-2, 2))  + 
+  scale_y_continuous( limits = c(-2,2)) + 
+  scale_color_manual(values = my_colors[3:4])  + 
+  my_theme + 
+  scale_shape_manual(values = c(16,8), guide = 'none') + 
+  facet_grid( . ~ Treatment  )  + 
+  coord_fixed()
+
+p2_blank <- 
+  ggplot(all_effects, aes( x = mu.x, y = mu.y, color = Treatment, shape = sig_label,  group = 1 )  ) + 
+    geom_abline(aes( intercept = 0, slope = 1), linetype = 2, color = 'black') + 
[…]

### 2017-08-04 precip/analysis/figure_scripts/plot_spring_soil_moisture_spot_measures.R

@@ -121,11 +121,16 @@ aggregate(data = df, VWC ~ Treatment, FUN = 'mean')
 
 # make table for stats output 
 library(xtable)
-
+library(lsmeans)
 statsOutput <- 'manuscript/spot_measurements.tex'
 
-texreg(spot_m,caption="Spring soil moisture",
-       caption.above=TRUE,file=statsOutput, label = 'table:spotVWC')
+test <- lsmeans(spot_m , "Treatment" )
+
+#xtlsm_spot <- xtable(test, caption = '')
+#print(xtlsm_spot, )
+
+texreg(spot_m,caption="Estimated parameters from a mixed effects model fit to the spring soil moisture data. Parameter with s.e. are shown. Intercept gives soil moisture in ambient controls.",
+        caption.above=TRUE,file=statsOutput, label = 'table:spotVWC')
 
 
 aggregate( data = df, VWC ~ Treatment, FUN = 'mean')

### 2017-08-04 precip/analysis/make_prediction_score_tables.R

@@ -1,3 +1,6 @@
+library(dplyr)
+library(tidyr)
+
 # format prediction tables 
 alllppd <- read.csv('output/overall_lppd.csv')
 allMSE <- read.csv('output/all_MSE_scores.csv')
@@ -18,8 +21,10 @@ overall_score_table <-
   arrange( species, vital_rate, score )
 
 
-oaxt <- xtable(overall_score_table, caption = 'Comparison of model predictions from climate model and year effects model for each species and vital rate.  Two prediction scores are reported, MSE and lppd. Lower MSE indicates improved predictions whereas higher lppd indicates improved predictions.  Instances where the climate model outperformed the random year effects model are marked with "***" in the last column. ARTR = \\textit{A. tripartita}, HECO = \\textit{H. comata}, POSE = \\textit{P. secunda}, PSSP = \\textit{P. spicata}.' , 
-label = 'table:overallPreds')
+overall_score_table <- rename( overall_score_table, `no climate model` = `year effects model` )
+
+oaxt <- xtable(overall_score_table, caption = 'Comparison of model predictions from climate model and no climate model for each species and vital rate.  Two prediction scores are reported, MSE and lppd. Lower MSE indicates improved predictions whereas higher lppd indicates improved predictions.  Instances where the climate model outperformed the no climate model are marked with "***" in the last column. ARTR = \\textit{A. tripartita}, HECO = \\textit{H. comata}, POSE = \\textit{P. secunda}, PSSP = \\textit{P. spicata}.' , 
+  label = 'table:overallPreds')
 
 print(oaxt, 'manuscript/overall_prediction_score.tex', type = 'latex', include.rownames = F, caption.placement ="top")
 
@@ -48,17 +53,23 @@ treatment_score_table <-
   arrange( species, vital_rate, Treatment, score )
 
 
-txt <- xtable(treatment_score_table, caption = 'Comparison of model predictions from climate model and year effects model for each species and vital rate and treatment.  Two prediction scores are reported, MSE and lppd. Lower MSE indicates improved predictions whereas higher lppd indicates improved predictions.  Instances where the climate model outperformed the random year effects model are marked with "***" in the last column. ARTR = \\textit{A. tripartita}, HECO = \\textit{H. comata}, POSE = \\textit{P. secunda}, PSSP = \\textit{P. spicata}.' , 
-               label = 'table:treatmentPreds')
+treatment_score_table <- treatment_score_table %>% 
[…]

### 2017-08-04 precip/analysis/select_climate_covariates.R

@@ -1,5 +1,8 @@
 # 
 library(stringr)
+library(dplyr)
+library(tidyr)
+
 rm(list = ls())
 
 cor_files <- dir('output', 'correlations.csv', full.names = T)
@@ -89,7 +92,7 @@ library(xtable)
 
 
 xtcor <- xtable(all_cors, 
-                caption = 'Selected climate variables for each vital rate model for each species. Correlations and p-values between the choosen variables and the intercept of year effects model are shown. For ARTR growth and POSE growth and survival, the correlations between the year effects on size and the soil moisture variables are also given. "f" = fall, "su" = summer, "sp" = spring. ARTR = \\textit{A. tripartita}, HECO = \\textit{H. comata}, POSE = \\textit{P. secunda}, PSSP = \\textit{P. spicata}.', 
+                caption = 'Selected climate variables for each vital rate model for each species. Correlations and p-values between the choosen variables and the intercept of the no climate model are shown. For ARTR growth and POSE growth and survival, the correlations between the year effects on size and the soil moisture variables are also given. "f" = fall, "su" = summer, "sp" = spring. ARTR = \\textit{A. tripartita}, HECO = \\textit{H. comata}, POSE = \\textit{P. secunda}, PSSP = \\textit{P. spicata}.', 
                 label = 'table:strongCor')
 
 print(xtcor, 'manuscript/strong_correlations.tex', type = 'latex', include.rownames = F, caption.placement ="top")

### 2017-08-21 removals/manuscript/ResponseToReviews.tex
Started response to reviews letter.

\documentclass[12pt]{article}

\usepackage{amsmath,amsfonts,amssymb,graphicx,natbib,setspace,authblk} \usepackage{float}
\usepackage[running]{lineno} \usepackage[vmargin=1in,hmargin=1in]{geometry}

\usepackage{enumitem} \setlist{topsep=.125em,itemsep=-0.15em,leftmargin=0.75cm}
\setlength{\parindent}{0.0in} \setlength{\parskip}{0.12in}
\usepackage[compact]{titlesec}

\usepackage{bm,mathrsfs}
\usepackage{mathptmx}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% %%% Just for commenting
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\usepackage[usenames]{color}
\definecolor{ForestGreen}{rgb}{0.1,0.44,0.1} 
\definecolor{blue}{rgb}{0,0,0.7} 
\definecolor{red}{rgb}{0.8,0,0}

\newcommand{\new}{\textcolor{red}} 
\newcommand{\comment}{\textcolor{ForestGreen}}
\newcommand{\response}{\textcolor{blue}}

\renewcommand{\floatpagefraction}{0.98} 
\renewcommand{\topfraction}{0.99} 
\renewcommand{\textfraction}{0.05}
\clubpenalty = 10000 \widowpenalty = 10000
\newcommand{\be}{\begin{equation}} 
\newcommand{\ee}{\end{equation}} 
\newcommand{\ba}{\begin{equation} \begin{aligned}} 
\newcommand{\ea}{\end{aligned} \end{equation}}

\def\X{\mathbf{X}} \def\A{\mathbf{A}} \def\B{\mathbf{B}} \def\C{\mathbf{C}} \def\D{\mathbf{D}}
\def\G{\mathbf{G}} \def\H{\mathbf{H}} \def\N{\mathbf{N}} \def\M{\mathbf{M}} \def\W{\mathbf{\Omega}}
\def\P{\mathbf{P}} \def\V{\mathbf{V}} \def\r{r^{\vphantom{0}}} \def\rbar{\bar{r}^{\vphantom{0}}}
\def\E{\mathbb{E}}

\newcommand{\s}[1]{{#1}^{\#}} 
\newcommand{\f}[1]{{#1}^{\flat}} 
\newcommand{\br}[1]{\langle {#1} \rangle}
\newcommand{\bs}{\backslash}

\newcommand{\numcirc}[1] {\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {#1}}}}

\begin{document} 
\centerline{\large{Response to reviews}}

\large{Original reviews are in black and \response{our responses are in blue.}
\normalsize

Miller's comments:

Your manuscript has been evaluated by two reviewers with expertise in plant competition and population/community ecology. As you will see, they appreciate the value of confronting model predictions with experiments and your very rigorous approach to this problem. I am in agreement with the reviewers about the strengths of the manuscript, though I also agree that there is some room for improvement in pitch and clarity. Both reviewers noted that the fit for Ecology can be strengthened (and the 'confirmatory' impression can be reduced), particularly in the Discussion, by generalizing beyond your study system and addressing bigger-picture issues about the merits of different approaches to estimating competitive interactions. Reviewer 1 further raised some concerns about your analyses, including clarity of methods and accounting for uncertainty in parameter estimates. Reviewer 2 (see attachment) noted that low cover, especially of grasses, likely contributed to weak competitive effects, and suggests some additional or alternative interpretation of the results.

My own main concern overlaps with Reviewer 1 regarding the clarity and interpretation of your statistical analysis. The manuscript sends mixed messages about how you estimated the removal treatment coefficient. There are (at least) two ways to do this. You could (A) fit a full model, including a removal coefficient (Eq 1), to all the data and evaluate significance of the removal term. This is suggested by the sentences at lines 171-173 ('note that these are not estimated as separate models'). Or you could (B) fit the crowding coefficients (your $\omega_jm$) from the control plots, then use them (as fixed values) in a second round of models fit to the removal plots that include a removal term (as a free parameter) and evaluate the significance of this coefficient. This is suggested at lines 263-268 ('would indicate effect larger or smaller than predicted by a model based on data from control plots'). Reviewer 1 and I have the intuition that method B is a better way to evaluate whether observational data can predict the effects of removals. Under method A (if that is what you did), I do not see why the treatment coefficient would be interpreted as the under- or over-estimation of competition. If the full model is fit to all the data, then the removal coefficient should instead reflect the difference between control and removal plots that is not related to competition, especially if both plot types spanned similar ranges of crowding environments, as you suggest (lines 289-293). This could perhaps be some unintended effect of herbicide or spatial location. In your revision, please take care to improve the communication and justification of your methods for testing whether observational data can predict the effects of experimental removals.

\response{ Thoughts: I am not opposed to their suggested test (though it is a little harder to implement), but it might come from a bit of a misunderstanding 
about our data. The removal of species A allows us to test the model's predictions about response of species, B,C, and D to A. But because the cover of A
will be zero in all the removal plots, we can't get an independent estimate of the effects of A on the other species just from those removal plots. Any significant removal effect will be lumped in with the intercept, which is the issue they have noticed. I guess "fixing" the competition terms as they
suggest might help the interpretation a little, but only if the inclusion of the removal plots changes the estimated competition terms. I will check on that.}

Other suggestions:

- The abstract can be strengthened all around. It is very narrowly motivated (first words: 'A previous study').

- line 32: add 'of competitors' following 'Stable coexistence'

- line 33: use of 'we' here is awkward because it is a different 'we' than at line 61

- Figure 5: You incorporate random year effects into these cover estimates, so they are basically the stationary distributions of cover. Your vital rate models included spatial random effects. Can you incorporate these too?

- You find that competitive interactions are weak, and that your models do a good job at predicting them. I could imagine that weak interactions make removal effects easy to predict (nothing happens!). You conclude that the success of this case study supports the use of multi-species models built from observational data. Do you think this conclusion will apply to communities with stronger competition? You might address this issue in your Discussion.

Reviewer: 1

Comments to the Author

Adler et al. describes a removal experiment to test competitive release in sagebrush steppe. It is related to a previous survey, which found very little evidence of interspecific competition, predicting that competitive release would be minimal in this system. Results from the experiment largely confirm predictions based on observational models – there are only a few instances where the removal treatment has an impact on survival, growth, or recruitment above and beyond the effects of interspecific competition. And, for 3 of 4 species, these impacts to not scale up to affect population growth rate or equilibrium cover. I really liked this paper! It has important implications for the use of long-term survey data for assessing community dynamics. It is great to see observational models coupled with experiments to test the bounds of what our models can be used to predict. It is a very thorough examination of competitive release, and I think it is the kind of classic ecological topic that would appeal to a large audience. I have a few major and minor comments below.

Major comments:

Originally I did not realize that you were fitting one model to the survey and removal experiment simultaneously. I thought that you would want to separate the data used to parameterize the model vs. the data used to test the model.  I wonder how much the removal experiment data is influencing parameters (other than the removal effect) in the model. One constraint of using survey data to fit models is that often natural communities do not experience a wide range of conditions (i.e., grass and Artemisia both at low cover or a rainy year with low crowding). Does including the removal experiment data in model fitting alter the estimates of (intra- and inter-specific) competition parameters because it encompasses combinations of the dependent variables that do not exist in the survey? \response{I will check on this}. This may particularly be important for the grass removal plots because they did not go to 0 immediately and thus exhibit variability (such that the model might want to adjust the slope/competition parameter rather than add a removal intercept effect). I think you tried to address this issue with Figure C-3 but I was confused by that graph (see minor comments below), and I think it would be better to either to 1) mathematically argue that the experimental data does not influence parameters other than the removal effect or 2) fit models using only the survey data and show that the parameter estimates are unchanged.

\response{Giles commented via email: fitting on one data set and then testing on another will tend to magnify differences and there aren’t valid tests for them. }

Possibly related to this – where the removal and competition parameters correlated?

\response{Not sure what the reviewer means...perhaps correlated in the MCMC chains?}

Another thing that kept crossing my mind as I read the paper was about the treatment of “non-significant” parameters, i.e., parameters with high uncertainty. As far as I could tell throughout the paper, parameters were never dropped from models, and projections were done using the parameter estimates based on the mean of the MCMC chain values. I was struggling with what it means to use parameters with high uncertainty in projections. You explain that you want to use the deterministic versions of the models (line 186) rather than a draw from the Poisson random variable; however those deterministic values of lambda have uncertainty. In your simulations you could incorporate that uncertainty by taking a random draw from the MCMC chain for each parameter (I don't know if this is standard practice). It seemed that this would somehow “equalize” parameter uncertainty (in competition, yearly intercept, removal parameters, etc.). I’m not sure if this would change the conclusions, the only thing it might do is make the variance larger in Fig. 5, and thus the P. spicata Removal model/no ARTR might not look as different from the other P. spicata models. I’m not saying you need to do this, I’m just curious if you had considered this issue at all.

\response{We can certainly do this if we have to, but it will be a programming pain in the butt...Good arguments to justify ignoring uncertainty in everything except the removal effects? (We address uncertainty in those by projecting models with the 95 \% CI values). }

My last comment, and really the only negative thing I can say about this paper is that it comes across as rather confirmatory in nature. It is absolutely solid science and highly interesting and important but the writing does not punch you with novelty. I think some small changes in wording could highlight that this is quite novel and hasn't been assessed in any other systems.

\response{Seems like the novelty is in the approach--experimental validation of a model based on observational data. We could also point out that while 
removal experiments have a long tradition, they are rarely coupled with a population model and generate only short term results like changes in cover or 
fitness components. Other thoughts?}

Minor comments:

Abstract

Can you mention how many species you were modeling somewhere near the beginning of the abstract?
Lines 19-20: instead of “population models based on vital rate regressions” maybe say “population projections based on vital rate regressions” or “population predictions” so it does not sound like you are fitting new models

Line 20: The first time I read this I was not sure what you meant by “vital rate regressions”. Maybe put this in parenthesis up in lines 7-9?

Line25: while a competition coefficient is often understood to represent resource competition, it can also represent any other density dependent effect like microbial effects, right?

Introduction

Methods

Lines 137-139:  When did you survey the removal plots? What month and for how many years?

Lines 150-173: In this section, I was confused what plots (survey plots or removal experiment) you were using to fit the models. Originally I had thought you would use only the survey plots to fit the models, then hold those parameters constant and add the removal parameter and fit the models to the removal data. I had thought that you would want to separate the data used to parameterize the model vs. the data used to test the model. I then realized that you were fitting the model to all data simultaneously (survey and removal experiment). Maybe be explicit somewhere?

Were the 2011 pretreatment data in the removal plots used as the conditions for t=1 in the model (i.e. for the competitive effects)? I don’t think it should be, because for the majority of the 2011-2012 transition, the target species were growing without competition from the removed species. Thus you might find that when there was a lot of Artemisia pretreatment in 2011, there is a large positive growth response in 2012.

\response{I zeroed out the removal species for the t=1 data}

Line 186: what is A-5?

Lines 181-198: if the removal effect is not “significant” (credible intervals overlap 0) is it still included in projections?

\response{Yes, this relates to the main concern about propagating uncertainty.}

Results

Line 263:  When I read this I wanted to know something about the uncertainty in these parameters – do most credible intervals overlap zero, or are the coefficients small but “significant”?

Lines 277-283: Does the root zone of Artemisia generally fall under its canopy? Or is it larger in diameter than the aboveground canopy? Do you think the primary limiting resource is nutrients/moisture or light?

Line 291: Can you add some explanation to the “break covariances in plant densities” part? It took me a while to figure out what you meant.

Discussion
Line 421: remove one of the “dynamics” words

Tables and Figures:

Fig. 1: I’m confused how the focal species account for 70\% basal and 60\% canopy cover (as stated in the methods), but the mean cover values for the grasses in Fig. 1 range from only 1-3\%. I realize your estimates of cover were highly precise, but it seems that it would be hard to model species that are at such low abundance.

Fig. 4. I imagine you omitted error bars from these graphs b/c it might get messy, but I would like to see them (possibly in the supplement). It would be helpful, especially for P. spicata, to assess how off the predictions were, i.e. do error bars overlap?

Supplementary

Figure C-3: I’m having a hard time interpreting this figure. If you look at the second box from the left on the top row (W.ARTR vs. W.HECO). The y axis is crowding from Artemisia. The red points are plots where Artemisia was removed. Why do the red points range from 0-60 on the y axis? Shouldn't they all be near 0? Or are pretreatment conditions in the removal plots included here?

Reviewer: 2

This study investigates whether models parameterized from historical data sets accurately describes
interspecific interactions by assessing whether additional terms are needed to account for removal
results beyond predictions from non-experimental data sets. Previous work led by Adler using models
based on historical data has produced important results, e.g., on the role of the storage effect and of
niche vs neutral processes in community dynamics, so I applaud their attempt to confirm model validity
and strengthen those results. On the other hand, I do have some concerns about the robustness of the
results, as described below. More broadly, even if the conclusions are robust, the manuscript could use
greater efforts to generalize beyond the validity of these particular models for these species. It would
be much appropriate for a general journal such as Ecology if the results were used to draw conclusions
on when modeling approaches will underestimate or overestimate interaction effects. Focusing the
discussion on such questions would be more appropriate than the current discussion material on
coexistence and inter- vs intraspecific competition, which are not part of the research questions that
end the Introduction.

\response{Not sure what to say about when models will under/over estimate interaction effects. Thoughts?}

One concern in the design is based on the relatively low covers, especially of the grasses. Based on
Figure 1, basal grass cover per species is very low even before removal—1-2\% per species. Finding
significant effects of such low cover on shrubs would indeed be surprising. Given this low expectation,
this does not seem a very strong test of whether model predictions are valid or not for the shrub. Does
the model even predict any effect of grass removals at these levels? \response{Yes, I can look these up}. Should the grass removal
treatment even be included in the study? The parameters estimated for A. tripartita seem to be very
uncertain in any case (pp 16-17), further suggesting that the results for this species do not add the story.
The test of the effects of shrubs on grasses is stronger (mean initial shrub cover was ca. 30\% before
removal) and indeed two out of three tests found significant shrub removal effects on growth beyond
the regressions on neighborhood crowding, as well as on population growth and the removal model
provides better predictions from the IBM for P. spicata. However, the authors argued this result may
not reflect stronger interspecific competition than estimated by the model because the release was no
greater for plants originally under a canopy than outside a sagebrush canopy. However, that argument
only makes sense either if competition is only for light or if the canopy edge reflects the edge of the
rooting zone as well. Given the habitat, I doubt the former is true and the authors also suggest it is
unlikely (line 442). For the latter, is anything known about the relationship of root distribution to
canopy cover in A. tripartita? My guess is that roots extend well beyond the canopy edge in these
habitats, leaving underestimated competitive interactions as a likely explanation for the significant
removal effect.

\response{Low cover should be irrelevant as long as the populations are near equilibrium. I can remind the reviewer the 
grasses are measured with basal cover.}

Given this result for P. spicata, line 401 of the discussion is a bit surprising: “how to reconcile the
insensitivity of grasses to sagebrush removal…..” On the other hand, it does hold for two of the grasses
and some discussion of why the grass species differ in sensitivity to shrub removal would be interesting.
Regardless of which species it holds for, the authors then attempt to reconcile the insensitivity found in
this study with strong effects found in the rangeland ecology literature by noting the relatively low shrub
cover (and small size of A. tripartita) in these ungrazed areas to explain the lack of strong effect of shrub
removal on grasses (lines 401-412). I agree this is a reasonable explanation and would argue that this is
also a possible, even likely, explanation for lack of effect of the grasses on sagebrush as above. In either
case, this reason for weak interspecific interactions (few interspecific neighbors) is quite different from
the ecological explanations in the next paragraph (lines 413-426; e.g., resource partitioning, balance of
facilitation and competition) and has quite different implications from those discussed in the
subsequent paragraph (lines 427-438). If this material stays, it would be helpful to compare and
contrast these different sorts of explanations and their implications more directly.

A few minor points:

Lines 50-53: Are these two sentences partly saying the same thing? If there is full compensation, the
response must depend on how much was removed, i.e., the degree of dominance.

Line 81: should “determine” be replaced by “estimate”?

Line 87: given that you earlier defined competitive release as a population response, should
“competitive release” at the end of this sentence be replaced by “individual response to neighbors”?
The second question more directly addresses the population consequences.

Line 216: “very similar to those used in our previous studies” is a bit confusing. Why not just use the
vital rate functions calculated in the first part of the study? If they are indeed the same, say so rather
than introducing similarity to other published studies again.

Line 234-235: Several confusing points here: a) The distinction between the community reaching
equilibrium (500 time steps) and each species reaching equilibrium cover (“additional 2000 time steps”)
is not clear. b) I assume the random year effects were used for all 2500 time steps? c) were the 3
scenarios applied AFTER the community reached equilibrium but before the 2000 steps, i.e., removal
from an equilibrium community? Or from the beginning of each set of runs?

Fig. 4. Is the baseline removal line missing for H. comata?

I found several of the supplemental figures to be somewhat confusing, most notably Figure C-3. This is
potentially important information about the distribution of crowding coefficients between removals and
controls, but I can’t quite figure out what each combination represents.

\response{I will fix C-3...}

\end{document}

### 2017-08-21 removals/manuscript/ResponseToReviews.tex
More minor edits on initial response letter.

@@
 I guess "fixing" the competition terms as they
-suggest might help the interpretation a little, but only if the inclusion of the removal plots changes the estimated competition terms. I will check on that.}
+suggest might help the interpretation a little, but only if the inclusion of the removal plots changes the estimated competition terms.
+I checked on the growth models for POSE and PSSP and there is almost no difference in any of the parameters between our full model and a model for control plots only.
+The treatment effect effectively soaks up all the difference.
+I think we just have to explain that our design does not allow us to disentangle differences in competitive effects from other effects that go into the intercept.
+We would need partial removal treatments to do that....}
 
@@
 Does including the removal experiment data in model fitting alter the estimates of (intra- and inter-specific) competition parameters because it encompasses combinations of the dependent variables that do not exist in the survey?
-\response{I will check on this}.
+\response{The answer is No}.
 This may particularly be important for the grass removal plots because they did not go to 0 immediately and thus exhibit variability (such that the model might want to adjust the slope/competition parameter rather than add a removal intercept effect).

### 2017-08-29 removals/manuscript/ResponseToReviews.tex
editing the responses doc

@@
 will be zero in all the removal plots, we can't get an independent estimate of the effects of A on the other species just from those removal plots. 
-Any significant removal effect will be lumped in with the intercept, which is the issue they have noticed.
-I guess "fixing" the competition terms as they
-suggest might help the interpretation a little, but only if the inclusion of the removal plots changes the estimated competition terms.
-I checked on the growth models for POSE and PSSP and there is almost no difference in any of the parameters between our full model and a model for control plots only.
-The treatment effect effectively soaks up all the difference.
-I think we just have to explain that our design does not allow us to disentangle differences in competitive effects from other effects that go into the intercept.
-We would need partial removal treatments to do that....}
+Any significant removal effect will be lumped in with the intercept, which is the issue they have noticed. I guess "fixing" the competition terms as they
+suggest might help the interpretation a little, but only if the inclusion of the removal plots changes the estimated competition terms. I checked on the growth models 
+for POSE and PSSP and there is almost no difference in any of the parameters between our full model and a model for control plots only. The treatment effect effectively soaks 
+up all the difference. I think we just have to explain that our design does not allow us to disentangle differences in competitive effects from other 
+effects that go into the intercept. We would need partial removal treatments to do that....}
 
@@
 
+\response{SPE: I think the intuition we're fighting is that our way of testing for a removal effect will tend to systematically \emph{underestimate} differences, i.e. that it will have
+lower power than the suggested alternative approach to detect when there is a removal effect. In order to satisfy Miller and the reviewer, we need to argue that this intuition of theirs
+is incorrect.}
+
 Possibly related to this – where the removal and competition parameters correlated?
@@
 
+\response{SPE: The reviewer's suggestion would be the right thing to do, if we were were using the model to make projections about future populations with and without competitor removals. What we are
+actually doing, however, is using projections of future populations to say something about the model. That ``something'' is that the interspecific competition coefficients are small in an ecologically
+relevant sense: the shift in mean cover due to interspecific competition is (for 3 species) small compared to the natural fluctuations in cover driven by interannual environmental variability. 
+If we made the projections with parameter uncertainty (as suggested) on top of random year effects, we would be conflating parameter uncertainty with environmental variability, ``tilting'' the
+comparison in our favor by overstating the magnitude of natural fluctuations.
+One point of Fig.
+5 is that for \emph{P.
+spicata} the small effect of competition with ARTR each year adds up (in the model)
+to a nontrivial long-term effect.}
+
+\response{I think the real question about parameter uncertainty is this: if you sample a parameter vector at random from the posterior distribution, and re-make Fig. 5 using that parameter vector
+in all years, would the results always look like Fig. 5, or are there other possible outcomes within the set of plausible parameter values? This would be a lot of work, and I don't see any good way to
+summarize it graphically. }
+
 My last comment, and really the only negative thing I can say about this paper is that it comes across as rather confirmatory in nature.
@@
 fitness components. Other thoughts?}
+
+\response{SPE: once again, Peter is being the Voice of the Apocalypse on this project, so it falls to me to be rah-rah. According to 
+\texttt{http://onlinelibrary.wiley.com/doi/10.1111/cobi.12049/abstract}, population models fitted to observational data don't do well at out-of-sample prediction, because
+conditions change and out-of-sample environments are different from in-sample. Here we have a rare, perhaps unique, opportunity to (a) test the out-of-sample predictive power of models fitted
+to observational data when effects of changing environmental conditions are accounted for (in our case, by using the control plots to estimate random year effects), 
+and (b) test the specific conclusion of Adler et al. (2010) that interspecific competition is very weak, relative to intraspecific, in this system. The latter test is important 
+because experimental comparisons of inter- vs intra-specific competition (such as Pantastico-Caldas, M. and D. L. Venable. 1993. Competition in two species of desert annuals along a topographic 
+gradient. Ecology 74: 2192-2203) often (typically?) reach the opposite conclusion. Are experimental tests too artificial, not capturing the processes that actually make 
+inter$\gg$intra, or are models based on observational data too unreliable? The answer has major implications for our understanding of how plant communities operate.}
 
@@
 \response{Yes, this relates to the main concern about propagating uncertainty.}
+
+\response{SPE: as above, I think the issue here is that we are not actually trying to project future population trajectories or the range of possible trajectories; rather, we are
+using projections to obtain an ecologically relevant measure of whether an effect is large or small.}
 
@@
 
-\response{Low cover should be irrelevant as long as the populations are near equilibrium. I can remind the reviewer the 
-grasses are measured with basal cover.}
+\response{Low cover should be irrelevant as long as the populations are near equilibrium. (SPE: why? Is the logic that
+the cover at equilibrium, whatever it might be, creates enough competitive pressure to keep each of the species
+from increasing?) I can remind the reviewer the grasses are measured with basal cover (SPE: do you also have data, from the
+contemporary quadrats, that would let you translate basal cover into canopy cover for the grasses? In the pictures I've seen,
+the total canopy cover for the grasses is higher than the canopy cover of ARTR. In fact, I think it might be useful to add to the
+SI a figure consisting of maybe 4 or 9 typical quadrats photographed from above. People reading the article seem to think (incorrectly) 
+that the community is mostly bare ground with the occasional plant here and there, and they are (partially mis-)interpreting our Results in light
+of that belief.}
 
@@
 contrast these different sorts of explanations and their implications more directly.
+
+\response{SPE: One issue here is that the reviewer is confusing the total effect of competitors with the per-unit-area effects of competitors. 
+Our conclusions are about the per-unit-area effects (how much are you hurt by 1 cm$^2$ of PSSP within your zone of
+competition?), which can be large even when the total effect of a particular competing species is low because that species
+is rare. Rareness of a species is thus not an alternative explanation for finding that per-unit-area effects of (what is the
+antonym of con?)-specific competitors are small.}
 

### 2017-08-29 removals/manuscript/ResponseToReviews.tex
trivia

@@
 
-\response{SPE: once again, Peter is being the Voice of the Apocalypse on this project, so it falls to me to be rah-rah. According to 
-\texttt{http://onlinelibrary.wiley.com/doi/10.1111/cobi.12049/abstract}, population models fitted to observational data don't do well at out-of-sample prediction, because
-conditions change and out-of-sample environments are different from in-sample. Here we have a rare, perhaps unique, opportunity to (a) test the out-of-sample predictive power of models fitted
-to observational data when effects of changing environmental conditions are accounted for (in our case, by using the control plots to estimate random year effects), 
-and (b) test the specific conclusion of Adler et al. (2010) that interspecific competition is very weak, relative to intraspecific, in this system. The latter test is important 
-because experimental comparisons of inter- vs intra-specific competition (such as Pantastico-Caldas, M. and D. L. Venable. 1993. Competition in two species of desert annuals along a topographic 
-gradient. Ecology 74: 2192-2203) often (typically?) reach the opposite conclusion. Are experimental tests too artificial, not capturing the processes that actually make 
-inter$\gg$intra, or are models based on observational data too unreliable? The answer has major implications for our understanding of how plant communities operate.}
+\response{SPE: here are a few more ideas, or maybe expansions on Peter's ideas and those in the paper. According to \texttt{http://onlinelibrary.wiley.com/doi/10.1111/cobi.12049/abstract}, 
+population models fitted to observational data don't do well at out-of-sample prediction, because
+conditions change and consequently out-of-sample environments are different from in-sample environments. Here we have a rare, perhaps unique, opportunity to (a) test the out-of-sample 
+predictive power of models fitted to observational data when effects of changing environmental conditions are accounted for (in our case, by using the control plots to estimate random year effects), 
+and (b) test the specific finding of Adler et al. (2010) that interspecific competition is very weak, relative to intraspecific, in this system. Given the conflicting results 
+about the relative importance of within- and between-species competition, as reviewed in the manuscript, 
+it is very informative to test the Adler et al. (2010) specific conclusions and thereby test the reliability of conclusions from models fitted to observational
+data. Maybe it would also be worth talking about the fact that experimental comparisons of inter- vs intra-specific competition (such as Pantastico-Caldas, M. and D. L. Venable. 1993. 
+Competition in two species of desert annuals along a topographic gradient. Ecology 74: 2192-2203) often (typically?) reach the opposite conclusion from observational studies (i.e, experiments
+suggest that within- and between-species competition are comparable in strength). Are experimental tests too artificial, not capturing the processes that actually make inter$\gg$intra, or are 
+models based on observational data too unreliable? The answer is important for our understanding of how plant communities operate, and for how we should study plant communities in the future.
+The more we can say about our results having general methodological conclusions for future work, the more it will do for our ``general significance'' score.}
 
@@
 that the community is mostly bare ground with the occasional plant here and there, and they are (partially mis-)interpreting our Results in light
-of that belief.}
+of that belief.)}
 
@@
 
+\response{SPE [I just went ahead and made some changes in the manuscript]: a) This was bad phrasing on our part. 
+The community equilibrium is not a constant level of cover for each species, as it would be in a deterministic
+model, but a steady-state pattern of fluctuations over time. The additional 2000 time steps are needed to compute the average cover of each species
+at the community steady-state. We have re-phrased and expanded these sentences to clarify this point. b) Yes, that's right. The paragraph
+has been reorganized to clarify this point. c) The scenarios were applied from the very beginning; the paragraph has been re-organized to 
+clarify this point.}
+
 Fig. 4. Is the baseline removal line missing for H. comata?

### 2017-08-29 removals/manuscript/removal_ms.tex
trivia

@@
 
-\smallskip \textbf{Article type:} Standard Paper
+\smallskip \textbf{Submission type:} Article
 
@@
 We simulated the IPM to estimate equilibrium abundances under three different scenarios.
-For all IPM simulations, we set the spatial random effects to zero, meaning that we are simulating dynamics on a hypothetical, average site.
-The first scenario featured all four species and our baseline model for control plots (no removal effects).
-We initialized all species with arbitrarily low cover and ran the model for 500 time steps (years) to allow the community to reach an equilibrium.
+\new{In all of these simulations, we set the spatial random effects to zero, meaning that we are simulating dynamics on a hypothetical, average site.
 To incorporate temporal environmental variation, at each time step we randomly selected one of the 30 sets of random year effects for the survival, growth and recruitment models (``kernel selection'' in the terminology of \citealt{metcalf_statistical_2015}).
-We calculated the equilibrium cover for each species by simulating an additional 2000 time steps and averaging each species' cover over this period.
+For each scenario, we initialized all species with very low cover and ran the model for 500 time steps (years) to allow the community to reach its equilibrium (i.e., its steady-state pattern of fluctuations in response to environmental variability).
+We calculated the \new{average steady-state cover} for each species by simulating an additional 2000 time steps and averaging each species' cover over this period.} The first scenario featured all four species and our baseline model for control plots (no removal effects).
 For the second scenario, we simulated our experimental removals by setting either the cover of \textit{A.

### 2017-08-30 removals/manuscript/ResponseToReviews.tex
trivia

@@
 This is suggested by the sentences at lines 171-173 ('note that these are not estimated as separate models').
-Or you could (B) fit the crowding coefficients (your $\omega_jm$) from the control plots, then use them (as fixed values) in a second round of models fit to the removal plots that include a removal term (as a free parameter) and evaluate the significance of this coefficient.
+Or you could (B) fit the crowding coefficients (your $\omega_{jm}$) from the control plots, then use them (as fixed values) in a second round of models fit to the removal plots that include a removal term (as a free parameter) and evaluate the significance of this coefficient.
 This is suggested at lines 263-268 ('would indicate effect larger or smaller than predicted by a model based on data from control plots').
@@
 
-My last comment, and really the only negative thing I can say about this paper is that it comes across as rather confirmatory in nature.
-It is absolutely solid science and highly interesting and important but the writing does not punch you with novelty.
-I think some small changes in wording could highlight that this is quite novel and hasn't been assessed in any other systems.
+My last comment, and really the only negative thing I can say about this paper is that it comes across as rather confirmatory in nature. It is absolutely solid science and 
+highly interesting and important but the writing does not punch you with novelty. I think some small changes in wording could highlight that this is quite novel and hasn't 
+been assessed in any other systems.
 

### 2017-09-08 removals/manuscript/ResponseToReviews.tex
Working on response to reviews

@@
 
-\large{Original reviews are in black and \response{our responses are in blue.}
 \normalsize
+
+Dear Tom,
+
+We really appreciated the quality of the reviews we received.
+Both reviewers and yourself clearly "got" the paper -- both its strengths and weaknesses.
+We have revised the ms.
+to highlight those strengths and shore up the weaknesses.
+In particular, we have 1) added text to the Introduction (and Discussion??) to emphasize the importance and novelty of our work, and 2) added new text and supplementary figures to the Methods to justify the decisions we made in designing our data analysis.
+In our point-by-point response below, the original reviewer comments are in black and \response{our responses are in blue.} We hope you fill this version of the manuscript suitable for publication.
+
+Thanks for your work,
+
+Peter Adler (on behalf of all coauthors)
 
@@
 
-\response{ Thoughts: I am not opposed to their suggested test (though it is a little harder to implement), but it might come from a bit of a misunderstanding 
-about our data. The removal of species A allows us to test the model's predictions about response of species, B,C, and D to A. But because the cover of A
-will be zero in all the removal plots, we can't get an independent estimate of the effects of A on the other species just from those removal plots. 
-Any significant removal effect will be lumped in with the intercept, which is the issue they have noticed. I guess "fixing" the competition terms as they
-suggest might help the interpretation a little, but only if the inclusion of the removal plots changes the estimated competition terms. I checked on the growth models 
-for POSE and PSSP and there is almost no difference in any of the parameters between our full model and a model for control plots only. The treatment effect effectively soaks 
-up all the difference. I think we just have to explain that our design does not allow us to disentangle differences in competitive effects from other 
-effects that go into the intercept. We would need partial removal treatments to do that....}
+\response{ This is an insightful comment, and we understand your point.
+There are two reasons why your suggested approaches A and B will not differ in our case.
+First, to answer Reviewer 1's question about this same issue, ``including the removal experiment data in model fitting \emph{does not} alter the estimates of (intra- and inter-specific) competition parameters," or any other parameters for that matter.
+We estimated a version of our P.
+spicata growth model using only data from control plots (and dropping the removal treatment effect).
+The correlation between the parameters estimated by that model and the parameters estimated by a model fit to all data is 0.99983.
+Results for P.
+secunda were similar.
+So there is no evidence of interactions between the removal treatment and other parameters in the model.
+Second, to clarify why all the differences between demographic performance in control and removal plots can be captured by one removal coefficient that only alters the overall intercept, we drew a new figure (Appendix C-???).
+To generate this figure, we fit a version of the growth model for P.
+spicata that did not include crowding from A.
+tripartita, but includes all other terms in our full model.
+We then plotted the residuals from this model against the observed values of A.
+tripartita crowding, using different colors from points for the control plots and the A.
+tripartita removal plots.
+The control plots points stretch out across the x-axis from A.
+tripartita values of zero up to a maximum of 70, and demonstrate a very weak but negative relationship between A.
+tripartita crowding and P.
+spicata growth.
+The slope of this line is the competition coefficient we estimate.
+However, all of the points for the removal plots, where A.
+tripartita cover is zero by design, are stacked on top of each other at x = 0.
+The figure shows that P.
+spicata plants experiencing zero A.
+tripartita crowding are growing faster in removal plots than in control plots, but it also shows why our removals provide no inference about the slope of the fitted line.
+So you are correct that we cannot clearly distinguish between two potential causes of the positive removal effect, one being that our observational data lead to an underestimate of A.
+tripartita competitive effects, and the other being that, in your words, ``the difference between control and removal plots...is not related to competition."  In retrospect, rather than only doing complete A.
+tripartita removals, we should have included partial removals.
+That said, while we now know how to conduct an even stronger test of our model in the future, the test we did conduct does represent a significant advance beyond the existing state of the art.
+
+To clarify these issues, we have added new text to Methods???
+Results???
+Discussion???, including the figure showing the marginal effects of A.
+tripartita crowding.
+Should we also mention how we would do it differently next time, in Discussion???}
 
@@
 
+\response{Steve, want to take a crack at it?}
+
 - line 32: add 'of competitors' following 'Stable coexistence'
 
+\response{Done.}
+
 - line 33: use of 'we' here is awkward because it is a different 'we' than at line 61
+
+\response{Fixed.}
 
@@
 Can you incorporate these too?
+
+\response{We could.
+In fact, for our short-term population projections, which we compare directly to observed cover values, we do include spatial random effects.
+However, for these long-term projections, we just wanted to focus on changes in cover for a hypothetical, average plot.
+If we incorporated the spatial random effects, we would six slightly different versions of this plot, but because none of the other parameters vary spatially, the effects of removals on each species would not change much at all.
+We have added a sentence to the methods justifying our choice to exclude spatial variation DO THIS.
+}
 
@@
 You might address this issue in your Discussion.
+
+\response{ ??? I don't have any good ideas about what to say. We could mention that a next step is testing predictions about strong INTRAspecific competition...????  }
 
@@
 
+\response{Thank you for the positive feedback!}
+
 Major comments:
@@
 Does including the removal experiment data in model fitting alter the estimates of (intra- and inter-specific) competition parameters because it encompasses combinations of the dependent variables that do not exist in the survey?
-\response{The answer is No}.
 This may particularly be important for the grass removal plots because they did not go to 0 immediately and thus exhibit variability (such that the model might want to adjust the slope/competition parameter rather than add a removal intercept effect).
@@
 
-\response{Giles commented via email: fitting on one data set and then testing on another will tend to magnify differences and there aren’t valid tests for them. }
+\response{We addressed this point above, in response to Miller's comments.
+Briefly, adding data from the removal experiments does not alter estimates of competition coefficients, but rather alters estimated of model intercepts because crowding from removed species must equal zero.
+As a result, a separate intercept for removal plots captures the effects of the removals.
+We have clarified this in...DO THIS}
 
-\response{SPE: I think the intuition we're fighting is that our way of testing for a removal effect will tend to systematically \emph{underestimate} differences, i.e. that it will have
-lower power than the suggested alternative approach to detect when there is a removal effect. In order to satisfy Miller and the reviewer, we need to argue that this intuition of theirs
-is incorrect.}
+\response{Giles commented via email: fitting on one data set and then testing on another will tend to magnify differences and there aren’t valid tests for them.
+DO WE WANT TO INCLUDE THIS POINT AS WELL?
+}
 
@@
 
-\response{We can certainly do this if we have to, but it will be a programming pain in the butt...Good arguments to justify ignoring uncertainty in everything except the removal effects?
-(We address uncertainty in those by projecting models with the 95 \% CI values).
-}
-
-\response{SPE: The reviewer's suggestion would be the right thing to do, if we were were using the model to make projections about future populations with and without competitor removals. What we are
-actually doing, however, is using projections of future populations to say something about the model. That ``something'' is that the interspecific competition coefficients are small in an ecologically
-relevant sense: the shift in mean cover due to interspecific competition is (for 3 species) small compared to the natural fluctuations in cover driven by interannual environmental variability. 
+\response{ This is a good point, and we appreciate being pushed to justify our decision to ignore parameter uncertainty. Our goal is to test
+whether competitive release is as small as our previous model predicted. To be conservative, we want to make it as easy as possible to
+model large effects of removals.  Propagating parameter uncertainty through our simulations would add noise, making it \emph{harder} to 
+detect the effects of removals. More formally, our previous result showed that the interspecific competition coefficients are small in an ecologically
+relevant sense: the shift in mean cover due to interspecific competition shown in Fig.
+5 is (for 3 species) small compared to the natural fluctuations in cover driven by interannual environmental variability.
 If we made the projections with parameter uncertainty (as suggested) on top of random year effects, we would be conflating parameter uncertainty with environmental variability, ``tilting'' the
@@
 5 is that for \emph{P.
-spicata} the small effect of competition with ARTR each year adds up (in the model)
-to a nontrivial long-term effect.}
+spicata} the small effect of competition with ARTR each year adds up (in the model) to a nontrivial long-term effect.
 
-\response{I think the real question about parameter uncertainty is this: if you sample a parameter vector at random from the posterior distribution, and re-make Fig. 5 using that parameter vector
-in all years, would the results always look like Fig. 5, or are there other possible outcomes within the set of plausible parameter values? This would be a lot of work, and I don't see any good way to
-summarize it graphically. }
+We have added new text to the Methods to justify our approach...DO THIS...add new paragraph in section: Integral projection models...
+  }
 
@@
 
-\response{Seems like the novelty is in the approach--experimental validation of a model based on observational data. We could also point out that while 
-removal experiments have a long tradition, they are rarely coupled with a population model and generate only short term results like changes in cover or 
-fitness components. Other thoughts?}
+\response{We all agreed that this is the first time we have been accused of underselling a result! We have added a new paragraph to the Introduction
+to highlight the novelty of our approach and to show how it answers a growing chorus of calls for more predictive tests. DO THIS...
+Also add material to Discussion???}
 
@@
 Can you mention how many species you were modeling somewhere near the beginning of the abstract?
+
+\response{Stated in first sentence of abstract.}
+
 Lines 19-20: instead of “population models based on vital rate regressions” maybe say “population projections based on vital rate regressions” or “population predictions” so it does not sound like you are fitting new models
+
+\response{Good suggestion, done.}
 
@@
 
+\response{Done.}
+
 Line25: while a competition coefficient is often understood to represent resource competition, it can also represent any other density dependent effect like microbial effects, right?
+
+\response{Good point, we changed ``resource competition" to ``feedbacks."}
 
@@
 
+\response{We have clarified that this is an Equation.}
+
 Lines 181-198: if the removal effect is not “significant” (credible intervals overlap 0) is it still included in projections?
 
-\response{Yes, this relates to the main concern about propagating uncertainty.}
-
-\response{SPE: as above, I think the issue here is that we are not actually trying to project future population trajectories or the range of possible trajectories; rather, we are
-using projections to obtain an ecologically relevant measure of whether an effect is large or small.}
+\response{Yes, this relates to the main concern about propagating uncertainty. ADDRESS THIS IN TEXT}
 
@@
 
+\response{UGH}
+
 Supplementary
@@
 Or are pretreatment conditions in the removal plots included here?
+
+\response{We revised the legend to clarify that the red points for Artemisia are pre-treatment values.}
 
@@
 
-\response{Not sure what to say about when models will under/over estimate interaction effects. Thoughts?}
+\response{Not sure what to say about when models will under/over estimate interaction effects. Thoughts? Avoid extrapolation? Other platitudes?
+Or simply say that we need more tests of this kind to answer these questions. }
 
@@
 this does not seem a very strong test of whether model predictions are valid or not for the shrub. Does
-the model even predict any effect of grass removals at these levels? \response{Yes, I can look these up}. Should the grass removal
+the model even predict any effect of grass removals at these levels? Should the grass removal
 treatment even be included in the study? The parameters estimated for A. tripartita seem to be very
 uncertain in any case (pp 16-17), further suggesting that the results for this species do not add the story.
+
+\response{We can offer multiple lines of evidence that low grass cover is not the cause of weak grass effects. 
+First, grass cover is high enough to cause strong intraspecific competition. The puzzle is why interspecific effects
+of grasses are so weak when their intraspecific effects are strong. Second, our growth and survival models do not
+depend on average cover but rather neighborhoood scale cover: even if average cover is low, a
+grass species may reach high abundance in a particular neighborhood, as seen in the Appendix figures
+showing distribution of crowding values, and exert strong per-unit-area effects on another plant. The grasses do in fact
+affect Artemisa, as shown in Tables B-2 (marginally significant effect of H. comata on Artemisia survival), B-6 (significant 
+effect of Poa on Artemisia growth, and  B-15 (P. spicata has a positive effect on Artemisia recruitment, while H. comata
+has a negative effect; the 95\% CIs do not overlap zero). Finally,
+grasses are mapped in the currency of basal cover, while for Artemisia canopy cover is mapped. To help readers
+appreciate that small values of basal cover are not trivial, we have added a new supplementary figure showing 
+photos of example quadrats before and after the removal treatment. (ADD FIG NUMBER)  }
+
 The test of the effects of shrubs on grasses is stronger (mean initial shrub cover was ca. 30\% before
@@
 
-\response{Low cover should be irrelevant as long as the populations are near equilibrium. (SPE: why? Is the logic that
-the cover at equilibrium, whatever it might be, creates enough competitive pressure to keep each of the species
-from increasing?) I can remind the reviewer the grasses are measured with basal cover (SPE: do you also have data, from the
-contemporary quadrats, that would let you translate basal cover into canopy cover for the grasses? In the pictures I've seen,
-the total canopy cover for the grasses is higher than the canopy cover of ARTR. In fact, I think it might be useful to add to the
-SI a figure consisting of maybe 4 or 9 typical quadrats photographed from above. People reading the article seem to think (incorrectly) 
-that the community is mostly bare ground with the occasional plant here and there, and they are (partially mis-)interpreting our Results in light
-of that belief.)}
+\response{TO DO: add caveat}
 
@@
 
-\response{SPE: One issue here is that the reviewer is confusing the total effect of competitors with the per-unit-area effects of competitors. 
-Our conclusions are about the per-unit-area effects (how much are you hurt by 1 cm$^2$ of PSSP within your zone of
-competition?), which can be large even when the total effect of a particular competing species is low because that species
-is rare. Rareness of a species is thus not an alternative explanation for finding that per-unit-area effects of (what is the
-antonym of con?)-specific competitors are small.}
+\response{We thank the reviewer for pointing out this apparent contradiction. TO DO: ??? reviewer is right
+that we seem to be using two different arguments in different places...I need to think about this more...}
 
@@
 
-\response{SPE [I just went ahead and made some changes in the manuscript]: a) This was bad phrasing on our part. 
+\response{ This was bad phrasing on our part. 
 The community equilibrium is not a constant level of cover for each species, as it would be in a deterministic
@@
 
-\response{I will fix C-3...}
+\response{We have revised the legend of this figure to better explain what it shows.}
 

### 2017-09-08 removals/manuscript/removal_ms.tex
natbib
Working on response to reviews

@@
 
-\usepackage{amsmath,amsfonts,amssymb,graphicx,natbib,setspace,authblk}
+\usepackage{amsmath,amsfonts,amssymb,graphicx,setspace,authblk}
 \usepackage{float}
@@
 \usepackage[vmargin=1in,hmargin=1in]{geometry}
+
+%%%  Sorting references by year(?)
+\usepackage[authoryear,sectionbib,sort]{natbib}
+%%%
 
@@
 Our removal treatments were 1) removal of all perennial grasses and 2) removal of the dominant shrub, \textit{Artemisia tripartita}.
-We modeled survival, growth and recruitment as functions of the locations, sizes, and species identities of neighboring plants, along with an indicator variable for removal treatment.
-If our ``baseline" model, which accounts for local plant-plant interactions, accurately explains the observed responses to removals, then the fitted removal indicator effect should be non-significant.
+We \new{regressed} survival, growth and recruitment on the locations, sizes, and species identities of neighboring plants, along with an indicator variable for removal treatment.
+If our ``baseline" \new{regression} model, which accounts for local plant-plant interactions, accurately explains the observed responses to removals, then the fitted removal indicator effect should be non-significant.
 For survival and recruitment, the removal treatment effects were never significantly positive, indicating that the baseline model did not underestimate intraspecific competitive effects.
@@
 tripartita} failed to improve the growth regressions, raising questions about the mechanisms driving the positive response to removal.
-For three species, population models based on the vital rate regressions showed little impact of removals on population growth rates or equilibrium cover.
+For three species, population projections based on the vital rate regressions showed little impact of removals on population growth rates or equilibrium cover.
 For \textit{P.
@@
 The ability of models based on observational data to predict population responses to experimental removals increased our confidence that interspecific interactions are in fact weak for at least three of our four study species.
-The implication is that factors besides interspecific resource competition play important roles in determining the structure and dynamics of this community.
+The implication is that factors besides interspecific \new{feedbacks} play important roles in determining the structure and dynamics of this community.
 The success of the models also supports the use of multispecies models based on long-term observational data for  characterizing competitive interactions.
@@
 
-Stable coexistence requires that intraspecific interactions must limit populations more than interspecific interactions \citep{chesson_mechanisms_2000}.
-Whenever we study populations of species that appear to be coexisting stably, such as populations that have co-occurred in close proximity for extended periods, we should expect intraspecific limitations to be stronger than interspecific limitations.
-However, we still might expect interspecific interactions to be strong and negative.
+Stable coexistence \new{of competitors} requires that intraspecific interactions must limit populations more than interspecific interactions \citep{chesson_mechanisms_2000}.
+\new{When studying} populations of species that appear to be coexisting stably, such as populations that have co-occurred in close proximity for extended periods, \new{one} should expect intraspecific limitations to be stronger than interspecific limitations.
+However, interspecific interactions \new{might still} be strong and negative.
 In fact, plant community ecologists often motivate research on coexistence by noting that all plant species compete for water, light, and a few mineral resources (e.g.
@@
 Plant ecologists have conducted dozens of removal experiments and have observed almost every possible result.
-In some cases, competitive release is complete: remaining species fully compensate for removed biomass \citep{leps_nutrient_1999,jutila_effects_2002}.
-In other cases, the degree of competitive release depended on the functional group or dominance of the removed and remaining species \citep{smith_dominant_2003,sala_resource_1989,belsky_effects_1992}.
+In some cases, competitive release was complete: remaining species fully compensated for removed biomass \citep{leps_nutrient_1999,jutila_effects_2002}.
+In other cases, \new{some competitive release occurred but the degree of compensation} depended on the functional group or dominance of the removed and remaining species \citep{smith_dominant_2003,sala_resource_1989,belsky_effects_1992}.
 Finally, even negative effects of removals have been observed \citep{keddy_effects_1989,gilbert_dominant_2009}, consistent with the recent facilitation literature.
@@
 comparison is to look for systematic deviations between model predictions and observations, we used a deterministic version of the model (e.g., 
-recruitment is the $\lambda$ of (\ref{eqn:recrProcessModel}), rather than a Poisson random variable with mean $\lambda$). Specifically, we used the recruitment regression to calculate 
+recruitment is the $\lambda$ of Eqn. \ref{eqn:recrProcessModel}, rather than a Poisson random variable with mean $\lambda$). Specifically, we used the recruitment regression to calculate 
 the expected number of new recruits at time $t+1$, which we then multiplied by the observed mean size of recruits for the focal species to get expected cover of new recruits. 
@@
 spicata}.
-Blue symbols show values from controls plots, red symbols show values from \textit{Artemisia} removal plots.``W.allcov'' refers to the aggregated crowding by all shrubs and perennial grasses beyond the focal dominant species, and ``W.allpts'' refers to the aggregated crowding by forb species.
+\response{Each point compares the crowding experienced by one \textit{P.
+spicata} individual in one year from the species on the x-axis to the species on the y-axis.}   Blue symbols show values from controls plots, red symbols show \new{pre-treatment} values from \textit{Artemisia} removal plots.``W.allcov'' refers to the aggregated crowding by all shrubs and perennial grasses beyond the focal dominant species, and ``W.allpts'' refers to the aggregated crowding by forb species.
 \texttt{Wdistrib\textbackslash exploreSurvivalWs.r}.

### 2017-09-09 removals/manuscript/ResponseToReviews.tex
More revisions...

@@
 
-To clarify these issues, we have added new text to Methods???
+To clarify these issues, we have DO THIS added new text to Methods???
 Results???
@@
 
-\response{Steve, want to take a crack at it?}
+\response{DO THIS Steve, want to take a crack at it?}
 
@@
 
-\response{ ??? I don't have any good ideas about what to say. We could mention that a next step is testing predictions about strong INTRAspecific competition...????  }
+\response{DO THIS ??? I don't have any good ideas about what to say. We could mention that a next step is testing predictions about strong INTRAspecific competition...????  }
 
@@
 As a result, a separate intercept for removal plots captures the effects of the removals.
-We have clarified this in...DO THIS}
+We have clarified this in...refer to new section...DO THIS}
 
@@
 whether competitive release is as small as our previous model predicted. To be conservative, we want to make it as easy as possible to
-model large effects of removals.  Propagating parameter uncertainty through our simulations would add noise, making it \emph{harder} to 
+model large effects of removals. (That is why we included the "maximum removal effect" simulation based on the outer limits of the removal effect
+confidence intervals.) Propagating parameter uncertainty through our simulations would add noise, making it \emph{harder} to 
 detect the effects of removals. More formally, our previous result showed that the interspecific competition coefficients are small in an ecologically
@@
 5 is (for 3 species) small compared to the natural fluctuations in cover driven by interannual environmental variability.
-If we made the projections with parameter uncertainty (as suggested) on top of random year effects, we would be conflating parameter uncertainty with environmental variability, ``tilting'' the
+If, as suggested, we made the projections with parameter uncertainty on top of random year effects, we would be conflating parameter uncertainty with environmental variability, ``tilting'' the
 comparison in our favor by overstating the magnitude of natural fluctuations.
@@
 
-\response{We thank the reviewer for pointing out this apparent contradiction. TO DO: ??? reviewer is right
-that we seem to be using two different arguments in different places...I need to think about this more...}
+\response{We thank the reviewer for pointing out this apparent contradiction. In fact, consistent with 
+the reviewers' critique of our argument, we actually demonstrated that 
+artificially increasing Artemisia cover would not lead to a much bigger competitive release. We have revised this
+section to suggest that overgrazing must do more than simply increase shrub cover, but must also alter
+the per capita competition coefficients, potentially by altering the balance between competition and facilitation
+or changing resource availability. DO THIS}
 
@@
 
+\response{We revised this passage to make it clear that the second sentence involves partial, not full
+compensation.}
+
 Line 81: should “determine” be replaced by “estimate”?
+
+\response{We have added "estimated" later in the sentence to recognize the uncertainty in the model.}
 
@@
 
+\response{We went with ``underestimate the response to decreased crowding" because our recruitment model does not operate 
+at the level of individuals.}
+
 Line 216: “very similar to those used in our previous studies” is a bit confusing. Why not just use the
@@
 than introducing similarity to other published studies again.
+
+\response{We do use the vital rate functions calculated in the first part of \emph{this} study, but they differ in very subtle ways 
+from the models used in our previous studies. We have revised the sentence to clarify this.}
 
@@
 
+\response{It is there, but virtually indistinguishable from the removal effect line; we've added text to the legend to point this out.}
+
 I found several of the supplemental figures to be somewhat confusing, most notably Figure C-3. This is

### 2017-09-09 removals/manuscript/removal_ms.tex
More revisions...

@@
 Compared to traditional field experiments, our approach has the advantage of projecting full life-cycle responses over long time periods, but its reliance on observational data sets limits the strength of our inference.
-We would have more confidence in our model's projections of competitive release if we could experimentally confirm the ability of our models to describe interspecific interactions.
-
-To test whether our sagebrush community model, which is based on observational data, underestimates the strength of interspecific competition and competitive release, we conducted a removal experiment in a sagebrush steppe.
+A successful experimental test could increase our confidence in the model's projections of competitive release.
+
+\new{ The idea that results from observational studies of competition represent hypotheses which should be tested experimentally is not new (Hairston 1980), but such tests remain extremely rare.
+In fact, independent tests of any kind of quantitative prediction are rare in ecology, spurring recent calls for a greater emphasis on prediction (Mouquet, Houlahan, Petchey).
+These authors argue that testing predictions will not only advance applied goals, such as forecasting ecological impacts of global change, but will also accelerate progress in basic research.
+}
+
+\new{ In the spirit of this argument, } we conducted a removal experiment to test whether our sagebrush steppe community model, which is based on observational data, underestimates the strength of interspecific competition and competitive release.
 We studied the responses of three perennial grass species to removal of sagebrush, and the response of sagebrush to removal of all grasses.
-The primary purpose of the experiment was not to directly measure competitive release, which is difficult to do over a short time period in a semiarid ecosystem with low and variable plant cover, but rather to validate the parameters in our model that represent key interspecific interactions and ultimately determine the magnitude of competitive release.
+The primary purpose of the experiment was not to directly measure competitive release, which is difficult to do over a short time period in a semiarid ecosystem with low and variable plant cover, but rather \new{to test whether our model accurately characterizes} the interspecific interactions that ultimately determine the magnitude of competitive release.
 We addressed two research questions: First, are the survival, growth or recruitment of our target species significantly different in control vs.
@@
 Because our demographic models include the effects of neighborhood crowding, they should be able to predict how experimental removal of neighbors alters demographic performance.
-If the full effect of experimental removals is stronger than predicted, it means that our statistical models underestimate competitive release.
+If the full effect of experimental removals is stronger than predicted, it means that our statistical models underestimate \new{the response to decreased crowding}.
 Our second question integrates across the full life cycle: how do short and long-term population-level responses to removals compare to projections from multispecies population models based on observational data?
@@
 
-The kernel is constructed from the fitted survival ($S$), growth ($G$), and recruitment ($R$) models, 
+As in our previous studies of this system \citep{adler_coexistence_2010,adler_forecasting_2012}, the kernel is constructed from the fitted survival ($S$), growth ($G$), and recruitment ($R$) models, 
 \begin{equation}
@@
 \end{equation}
-which are very similar to those used in our previous studies of this system \citep{adler_coexistence_2010,adler_forecasting_2012}; see 
-Appendix \ref{AppendixA} for a complete description. As in these previous studies, the IPM uses a mean-field approximation for
+(see Appendix \ref{AppendixA} for a complete description). The IPM uses a mean-field approximation for
 neighborhood competition that captures the essential features of spatially-explicit local interactions; 
@@
 Red lines and symbols show one-step-ahead predictions from an IBM that includes removal treatment coefficients.
-Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r}.
+\new{For \textit{H.
+comata}, lines showing predictions for the baseline and removal effect models are virtually indistinguishable.} Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r}.
 }
@@
 
+% not sure yet where this should go
+\new{Our study represents a very strong test of a surprising prediction for two reasons. First, the prediction was already published; the experiment 
+represented a very real chance of getting it wrong...Second, we were just testing an observationally based model with new observations, but rather with
+an experimental manipulation, increasing the potential to confront the model with a novel situation outside the historical range of variation...
+Add Steve's point about testing population models--in contrast to the Crone study, we do well when we control for environmental variation with random year effects...}
+
 For three of the four species, we find little or no evidence that models based on observational data underestimate 
@@
 the experimental observations and short-term population projections show little response to removal, while the long-term simulations 
-from the baseline and removal effects models suggest strong (but opposite) effects of removal: a tripling of abundance in the baseline model, and extinction in the
-removal effects model. However, these long-term simulations primarily reflect the influence of recruitment parameters 
+from the baseline and removal effects models suggest strong (but opposite) effects of removal: a tripling of abundance in the baseline model, and extinction in the removal effects model.
+However, these long-term simulations primarily reflect the influence of recruitment parameters
 that are highly uncertain (see \emph{Results}). This high uncertainty 
@@
 and species-specificity of intraspecific density dependence. A useful first step in this direction would be to test 
-whether our model is also successful at predicting the effects of release from intraspecific competition
+whether our model is also successful at predicting the effects of release from intraspecific competition.
 

### 2017-09-10 removals/manuscript/ResponseToReviews.tex
Still working on revisions, just saving to be safe

@@
 
+\response{We added the following sentence, "We repeated mapped censuses of all plots each June from 2012-2016."}
+
 Lines 150-173: In this section, I was confused what plots (survey plots or removal experiment) you were using to fit the models.
@@
 
+\response{We added this sentence to the Methods: "For plants in removal plots in 2012, we set crowding for removed neighbor species to zero, rather than using the pre-treatment (2011) data."}
+
 \response{I zeroed out the removal species for the t=1 data}
@@
 
-\response{Yes, this relates to the main concern about propagating uncertainty. ADDRESS THIS IN TEXT}
+\response{Yes, this relates to the main concern about propagating uncertainty. DO THIS: ADDRESS THIS IN TEXT}
 
@@
 
+\response{Thank you, fixed.}
+
 Tables and Figures:
@@
 I realize your estimates of cover were highly precise, but it seems that it would be hard to model species that are at such low abundance.
+
+\response{We addressed this point in detail in response to Reviewer 2's comment below.}
 
@@
 
-\response{TO DO: add caveat}
+\response{This is a fair point. We added a sentence noting that grasses in interspaces could still experience competition
+if shrub roots extend beyond canopies.}
 
@@
 the reviewers' critique of our argument, we actually demonstrated that 
-artificially increasing Artemisia cover would not lead to a much bigger competitive release. We have revised this
+artificially increasing Artemisia cover would not lead to a much bigger competitive release. We have added text to this 
 section to suggest that overgrazing must do more than simply increase shrub cover, but must also alter
 the per capita competition coefficients, potentially by altering the balance between competition and facilitation
-or changing resource availability. DO THIS}
+or changing resource availability. }
 

### 2017-09-10 removals/manuscript/removal_ms.tex
Still working on revisions, just saving to be safe

@@
 The sagebrush removal was virtually 100\% effective; grass removals required additional herbicide applications in April and May of each year, with fewer grasses remaining each year.
+We repeated mapped censuses of all plots each June from 2012-2016.
 
@@
 quadrat (recruitment) scale, and $\omega_{jm}$ is an interaction coefficient which determines the impact of crowding by species $m$ on the 
-focal species $j$.
+focal species $j$. For plants in removal plots in 2012, we set crowding for removed neighbor species to zero, rather than using the pre-treatment (2011) data.
 
@@
 removal is releasing \textit{P. spicata} from competition, we should expect \textit{P. spicata} individuals that were located beneath \textit{A. tripartita} 
-shrubs that we removed to respond more than individuals located away from any shrubs. Yet we found no evidence for such a difference. 
+shrubs that we removed to respond more than individuals located away from any shrubs. Yet we found no evidence for such a difference. \new{Perhaps 
+\textit{A. tripartita's} roots extend far enough beyond its canopy to compete for resources with grasses located in interspaces.}
 An alternative explanation involves pre-treatment differences in composition between control plots and  the \textit{A. tripartita} 
@@
 research showing that sagebrush eradication increases grass production? The most likely explanation involves ecological context: our 
-study focused on \textit{Artemisia tripartita} in an ungrazed plant communities, whereas most research on sagebrush 
+study focused on \textit{Artemisia tripartita} in an ungrazed plant community, whereas most research on sagebrush 
 and forage production has focused on \textit{Artemisia tridentata} in heavily grazed communities, and often has involved grass seeding 
@@
 a nonlinear fashion, with little increase in production at sites with low initial sagebrush cover, 
-and accelerating responses as initial sagebrush cover passes some threshold \citep{McDaniel2005}. The cover of 
+and accelerating responses as initial sagebrush cover passes some threshold \citep{McDaniel2005}, \new{implying a shift in the nature of interspecific interactions, perhaps due to changes in the balance of competition and facilitation or in resource supply.} The cover of
 sagebrush in our ungrazed quadrats may have fallen below that threshold.
@@
 in this community \citep{adler_weak_2009}. Interactions with above- or belowground enemies can regulate populations and, given 
-enough species-specificity in the effects of these enemies, decouple the dynamics of different species dynamics \citep{hersh_evaluating_2011,
+enough species-specificity in the effects of these enemies, decouple the dynamics of different species \citep{hersh_evaluating_2011,
 janzen_herbivores_1970,connell_role_1971}.  More likely, plants may alter the environment, and even resource availability, 

### 2017-09-10 removals/manuscript/ResponseToReviews.tex
Another intermediate save on revisions

@@
 We estimated a version of our P.
-spicata growth model using only data from control plots (and dropping the removal treatment effect).
+spicata growth model using only data from control plots.
 The correlation between the parameters estimated by that model and the parameters estimated by a model fit to all data is 0.99983.
@@
 If we incorporated the spatial random effects, we would six slightly different versions of this plot, but because none of the other parameters vary spatially, the effects of removals on each species would not change much at all.
-We have added a sentence to the methods justifying our choice to exclude spatial variation DO THIS.
+We have added a sentence to the Methods justifying our choice to exclude spatial variation.
 }
@@
 
-We have added new text to the Methods to justify our approach...DO THIS...add new paragraph in section: Integral projection models...
+We have added a new paragraph to the Methods section "Individual Based Models" that makes these points.
   }
@@
 
+\response{Addressed above....FIX THIS UP}
+
 Were the 2011 pretreatment data in the removal plots used as the conditions for t=1 in the model (i.e.
@@
 
-\response{Yes, this relates to the main concern about propagating uncertainty. DO THIS: ADDRESS THIS IN TEXT}
+\response{Yes, we now clarify this in the new Methods paragraph on parameter uncertainty. Note that most of the parameters close
+to zero are the interspecific interaction effects. If we set those to zero, or dropped them from the simulations, it would further reduce the 
+magnitude of competitive release predicted by the model.}
 
@@
 
+\response{We added this sentence: "The intraspecific interaction effects are large and have credible intervals that do not overlap zero, while the interspecific coefficients are small and the majority have credible intervals that do overlap zero." }
+
 Lines 277-283: Does the root zone of Artemisia generally fall under its canopy?
@@
 
+\response{We altered this section to be clear that competition for light or for soil resources should be stronger under the canopy.
+We also added a caveat in the Discussion (in response to Reviewer 2) about roots extending beyond canopies.
+}
+
 Line 291: Can you add some explanation to the “break covariances in plant densities” part? It took me a while to figure out what you meant.
+
+\response{We replaced "break covariances in plant densities among species" with "result in combinations of neighborhood-scale densities that deviate from naturally occurring patterns of correlation."}
 

### 2017-09-10 removals/manuscript/removal_ms.tex
Another intermediate save on revisions

@@
 
-\new{ The idea that results from observational studies of competition represent hypotheses which should be tested experimentally is not new (Hairston 1980), but such tests remain extremely rare.
-In fact, independent tests of any kind of quantitative prediction are rare in ecology, spurring recent calls for a greater emphasis on prediction (Mouquet, Houlahan, Petchey).
+\new{ The idea that results from observational studies of competition represent hypotheses which should be tested experimentally is not new \citep{Hairston1980}, but such tests remain extremely rare.
+In fact, independent tests of any kind of quantitative prediction are rare in ecology, spurring recent calls for a greater emphasis on prediction \citep{Mouquet2015,Petchey2015,Houlahan2017}.
 These authors argue that testing predictions will not only advance applied goals, such as forecasting ecological impacts of global change, but will also accelerate progress in basic research.
@@
 We then projected the model forward one year, applying the random year effects corresponding to year $t$ and the spatial random effects corresponding to the quadrat of interest. 
-Because the main goal of this
-comparison is to look for systematic deviations between model predictions and observations, we used a deterministic version of the model (e.g., 
-recruitment is the $\lambda$ of Eqn. \ref{eqn:recrProcessModel}, rather than a Poisson random variable with mean $\lambda$). Specifically, we used the recruitment regression to calculate 
-the expected number of new recruits at time $t+1$, which we then multiplied by the observed mean size of recruits for the focal species to get expected cover of new recruits. 
 We used the survival regression to predict the survival probability of each genet, and the growth regression to predict the expected size (on arithmetic scale) of each genet at time $t+1$ (note that the expected size on arithmetic scale is not exp(expected log size), because of Jensen's inequality).
@@
 
-One reason we might observe small effects of the removal treatment is the limited power of our experiment.
-Perhaps large uncertainty in the estimates of these effects makes it difficult to reject the null hypothesis represented by our baseline model.
+\new{Although all the parameters in our model are estimated with uncertainty, we chose not to propage parameter uncertainty through the 
+simulations because we wanted to maximize our ability to detect the population-level consequences of a small removal effect. 
+Our previously published results showed that interspecific competition is weak relative to the effects of interannual environmental variability 
+\citep{Tredennick2017}. If we included parameter uncertainty along with random year effects, we would be conflating parameter uncertainty 
+with environmental variability, overstating the magnitude of natural fluctuations, and making it more difficult to show that competitive release
+is larger than the observationally-based model predicts.
+Therefore, we used estimated means of all parameters, including those with confidence intervals that overlapped zero.
+Based on the same rationale, we used a deterministic version of the model (e.g.,
+recruitment is the $\lambda$ of Eqn. \ref{eqn:recrProcessModel}, rather than a Poisson random variable with mean $\lambda$). 
+Specifically, we used the recruitment regression to calculate the expected number of new recruits at time $t+1$, which we then multiplied 
+by the observed mean size of recruits for the focal species to get expected cover of new recruits. }
+
+Another reason we might observe small effects of the removal treatment is the limited power of our experiment.
+Perhaps large uncertainty in the estimates of the removal effects themselves makes it difficult to reject the null hypothesis represented by our baseline model.
 To address this possibility, we also simulated a version of the model in which we replaced the point estimate of each removal treatment effect with either the 5\% or 95\% credible limit for the treatment effect, whichever was farthest from zero.
@@
 We simulated the IPM to estimate equilibrium abundances under three different scenarios.
-\new{In all of these simulations, we set the spatial random effects to zero, meaning that we are simulating dynamics on a hypothetical, average site.
-To incorporate temporal environmental variation, at each time step we randomly selected one of the 30 sets of random year effects for the survival, growth and recruitment models (``kernel selection'' in the terminology of \citealt{metcalf_statistical_2015}).
+\new{In all of these simulations, we set the spatial random effects to zero, meaning that we are simulating dynamics on a hypothetical, average site.} To incorporate temporal environmental variation, at each time step we randomly selected one of the 30 sets of random year effects for the survival, growth and recruitment models (``kernel selection'' in the terminology of \citealt{metcalf_statistical_2015}).
 For each scenario, we initialized all species with very low cover and ran the model for 500 time steps (years) to allow the community to reach its equilibrium (i.e., its steady-state pattern of fluctuations in response to environmental variability).
@@
 fitted models imply that interspecific competition is much weaker than intraspecific competition, in many
-cases by 1-2 orders of magnitude (Table \ref{table:alphas}).  
+cases by 1-2 orders of magnitude (Table \ref{table:alphas}).  The intraspecific interaction effects are large and have credible intervals that 
+do not overlap zero, while the interspecific coefficients are small and the majority have credible intervals that do overlap zero.
 If the removal treatment effects that we added to these models are statistically significant, it would indicate an effect of 
@@
 If removal is releasing grasses from \textit{A.
-tripartita} competition, we would expect individuals located under the shrub canopies to experience a greater effect of removal.
+tripartita} competition \new{for light or soil resources}, we would expect individuals located under the shrub canopies to experience a greater effect of removal than individuals further away.
 However, the location of individuals with respect to \textit{A.
@@
 
-Third, responses to removals might be poorly predicted by models based on observational data if the removals reduce plant densities (crowding) below the range of historical variation, or if the removals break covariances in plant densities among species.
+Third, responses to removals might be poorly predicted by models based on observational data if the removals reduce plant densities (crowding) below the range of historical variation, or if the removals \new{result in combinations of neighborhood-scale densities that deviate from naturally occurring patterns of correlation}.
 However, the distribution of crowding observed within control plots easily spans the crowding values observed in removal plots (Fig.

### 2017-09-11 removals/manuscript/ResponseToReviews.tex
Done with revisions for now.

@@
 We really appreciated the quality of the reviews we received.
-Both reviewers and yourself clearly "got" the paper -- both its strengths and weaknesses.
+Both reviewers and yourself clearly ``got" the paper -- both its strengths and weaknesses.
 We have revised the ms.
 to highlight those strengths and shore up the weaknesses.
-In particular, we have 1) added text to the Introduction (and Discussion??) to emphasize the importance and novelty of our work, and 2) added new text and supplementary figures to the Methods to justify the decisions we made in designing our data analysis.
+In particular, we have 1) added text to the Introduction and Discussion to emphasize the importance, generality, and novelty of our work, and 2) added new text and supplementary figures to the Methods to justify the decisions we made in designing our data analysis.
 In our point-by-point response below, the original reviewer comments are in black and \response{our responses are in blue.} We hope you fill this version of the manuscript suitable for publication.
@@
 So there is no evidence of interactions between the removal treatment and other parameters in the model.
-Second, to clarify why all the differences between demographic performance in control and removal plots can be captured by one removal coefficient that only alters the overall intercept, we drew a new figure (Appendix C-???).
+Second, to clarify why all the differences between demographic performance in control and removal plots can be captured by one removal coefficient that only alters the overall intercept, we drew a new figure (Appendix C-2).
 To generate this figure, we fit a version of the growth model for P.
@@
 That said, while we now know how to conduct an even stronger test of our model in the future, the test we did conduct does represent a significant advance beyond the existing state of the art.
-
-To clarify these issues, we have DO THIS added new text to Methods???
-Results???
-Discussion???, including the figure showing the marginal effects of A.
-tripartita crowding.
-Should we also mention how we would do it differently next time, in Discussion???}
+To clarify these issues, we have added a new paragraph to the Methods section, along with the new Appendix figure (C-2).
+We also mention in the Discussion how we would do it differently next time.}
 
@@
 
-\response{DO THIS Steve, want to take a crack at it?}
+\response{We revised the abstract to broaden the motivation.}
 
@@
 
-\response{DO THIS ??? I don't have any good ideas about what to say. We could mention that a next step is testing predictions about strong INTRAspecific competition...????  }
+\response{We added a new "lessons learned" paragraph to the Discussion where we address this point (yes, we see no reason why
+multispecies models would not also work in communities with stronger competition), along with other generalizations 
+from our experience on this project.  }
 
@@
 As a result, a separate intercept for removal plots captures the effects of the removals.
-We have clarified this in...refer to new section...DO THIS}
-
-\response{Giles commented via email: fitting on one data set and then testing on another will tend to magnify differences and there aren’t valid tests for them.
-DO WE WANT TO INCLUDE THIS POINT AS WELL?
-}
+We have clarified this in a new Methods paragraph and an additional Appendix figure (C-2).}
 
@@
 
-\response{Not sure what the reviewer means...perhaps correlated in the MCMC chains?}
+\response{We are not sure what correlation the reviewer is interested in (correlations in the MCMC chains?), but perhaps our assurance that
+adding the removal data does not alter the competition coefficients helps allay the implicit concern.}
 
@@
 whether competitive release is as small as our previous model predicted. To be conservative, we want to make it as easy as possible to
-model large effects of removals. (That is why we included the "maximum removal effect" simulation based on the outer limits of the removal effect
+model large effects of removals. (That is why we included the ``maximum removal effect" simulation based on the outer limits of the removal effect
 confidence intervals.) Propagating parameter uncertainty through our simulations would add noise, making it \emph{harder} to 
@@
 spicata} the small effect of competition with ARTR each year adds up (in the model) to a nontrivial long-term effect.
-
 We have added a new paragraph to the Methods section "Individual Based Models" that makes these points.
@@
 \response{We all agreed that this is the first time we have been accused of underselling a result! We have added a new paragraph to the Introduction
-to highlight the novelty of our approach and to show how it answers a growing chorus of calls for more predictive tests. DO THIS...
-Also add material to Discussion???}
-
-\response{SPE: here are a few more ideas, or maybe expansions on Peter's ideas and those in the paper. According to \texttt{http://onlinelibrary.wiley.com/doi/10.1111/cobi.12049/abstract}, 
-population models fitted to observational data don't do well at out-of-sample prediction, because
-conditions change and consequently out-of-sample environments are different from in-sample environments. Here we have a rare, perhaps unique, opportunity to (a) test the out-of-sample 
-predictive power of models fitted to observational data when effects of changing environmental conditions are accounted for (in our case, by using the control plots to estimate random year effects), 
-and (b) test the specific finding of Adler et al. (2010) that interspecific competition is very weak, relative to intraspecific, in this system. Given the conflicting results 
-about the relative importance of within- and between-species competition, as reviewed in the manuscript, 
-it is very informative to test the Adler et al. (2010) specific conclusions and thereby test the reliability of conclusions from models fitted to observational
-data. Maybe it would also be worth talking about the fact that experimental comparisons of inter- vs intra-specific competition (such as Pantastico-Caldas, M. and D. L. Venable. 1993. 
-Competition in two species of desert annuals along a topographic gradient. Ecology 74: 2192-2203) often (typically?) reach the opposite conclusion from observational studies (i.e, experiments
-suggest that within- and between-species competition are comparable in strength). Are experimental tests too artificial, not capturing the processes that actually make inter$\gg$intra, or are 
-models based on observational data too unreliable? The answer is important for our understanding of how plant communities operate, and for how we should study plant communities in the future.
-The more we can say about our results having general methodological conclusions for future work, the more it will do for our ``general significance'' score.}
+to highlight the novelty of our approach and to show how it answers a growing chorus of calls for more predictive tests. Also, in the Discussion, we 
+added some text to further emphasize why our study is a strong test of a prediction. }
 
@@
 
-\response{Addressed above....FIX THIS UP}
+\response{The new Methods paragraph addresses these questions directly.}
 
@@
 \response{We added this sentence to the Methods: "For plants in removal plots in 2012, we set crowding for removed neighbor species to zero, rather than using the pre-treatment (2011) data."}
-
-\response{I zeroed out the removal species for the t=1 data}
 
@@
 
-\response{We added this sentence: "The intraspecific interaction effects are large and have credible intervals that do not overlap zero, while the interspecific coefficients are small and the majority have credible intervals that do overlap zero." }
+\response{We added this sentence: ``The intraspecific interaction effects are large and have credible intervals that do not overlap zero, while the interspecific coefficients are small and the majority have credible intervals that do overlap zero." }
 
@@
 
-\response{We replaced "break covariances in plant densities among species" with "result in combinations of neighborhood-scale densities that deviate from naturally occurring patterns of correlation."}
+\response{We replaced ``break covariances in plant densities among species" with ``result in combinations of neighborhood-scale densities that deviate from naturally occurring patterns of correlation."}
 
@@
 
-\response{UGH}
+\response{The requested figure is included below (Fig. 1). It shows broad overlap, but we think it is too messy to include in the supplement. }
+
+ \begin{figure}[tbp]
+ \centering
+ \includegraphics[width=1\textwidth]{cover_projections_1stepBARS}
+ \caption{ Same as Fig. 4 in the main text, but with error bars showing 1 standard deviation of the mean. }
+ \label{fig:projectionsBars}
+ \end{figure}
 
@@
 
-\response{Not sure what to say about when models will under/over estimate interaction effects. Thoughts? Avoid extrapolation? Other platitudes?
-Or simply say that we need more tests of this kind to answer these questions. }
+\response{We have added new Discussion paragraph to directly address these questions about methodological
+generalizations (lines ???). }
 
@@
 grasses are mapped in the currency of basal cover, while for Artemisia canopy cover is mapped. To help readers
-appreciate that small values of basal cover are not trivial, we have added a new supplementary figure showing 
-photos of example quadrats before and after the removal treatment. (ADD FIG NUMBER)  }
+appreciate that small values of basal cover are not trivial, we have added a new supplementary figure (C-1) showing 
+photos of example quadrats before and after the removal treatment.  }
 

### 2017-09-11 removals/manuscript/removal_ms.tex
Done with revisions for now.

@@
 
-A previous study of four dominant species in a sagebrush steppe community used historical data sets and multispecies population models to demonstrate weak interspecific interactions and predicted that little competitive release would result from removal of heterospecific competitors.
-We conducted a removal experiment to test whether our model, with weak interspecific competition, accurately describes interspecific interactions.
-Our removal treatments were 1) removal of all perennial grasses and 2) removal of the dominant shrub, \textit{Artemisia tripartita}.
+\new{Stable coexistence requires intraspecific limitations to be stronger than interspecific limitations.
+The greater the difference between intra- and interspecific limitations, the more stable the coexistence, and the weaker the competitive release any species should experience following removal of competitors.
+We conducted a removal experiment to test whether a previously estimated model, showing surprisingly weak interspecific competition for four dominant species in a sagebrush steppe, accurately predicts competitive release.} Our removal treatments were 1) removal of all perennial grasses and 2) removal of the dominant shrub, \textit{Artemisia tripartita}.
 We \new{regressed} survival, growth and recruitment on the locations, sizes, and species identities of neighboring plants, along with an indicator variable for removal treatment.
-If our ``baseline" \new{regression} model, which accounts for local plant-plant interactions, accurately explains the observed responses to removals, then the fitted removal indicator effect should be non-significant.
-For survival and recruitment, the removal treatment effects were never significantly positive, indicating that the baseline model did not underestimate intraspecific competitive effects.
-For \textit{Poa secunda} recruitment, the removal treatment was negative, indicating that our baseline model overestimated intraspecific competition.
-For growth, the removal treatment effect was significant and positive for two species, \textit{Poa secunda} and \textit{Pseudoroegneria spicata}, indicating that the baseline model underestimated intraspecific competitive effects.
+If our ``baseline" \new{regression} model, which accounts for local plant-plant interactions, accurately explains the observed responses to removals, then the removal coefficient should be non-significant.
+For survival and recruitment, the removal coefficients were never significantly positive, indicating that the baseline model did not underestimate intraspecific competition.
+For \textit{Poa secunda} recruitment, the removal coefficient was negative, indicating that our baseline model overestimated intraspecific competition.
+For growth, the removal treatment effect was significant and positive for two species, \textit{Poa secunda} and \textit{Pseudoroegneria spicata}, indicating that the baseline model underestimated intraspecific competition.
 However, including information about the location of grass individuals with respect to removed \textit{A.
 tripartita} failed to improve the growth regressions, raising questions about the mechanisms driving the positive response to removal.
-For three species, population projections based on the vital rate regressions showed little impact of removals on population growth rates or equilibrium cover.
+For three species, population projections based on the vital rate regressions showed little impact of removals.
 For \textit{P.
@@
 The sagebrush removal was virtually 100\% effective; grass removals required additional herbicide applications in April and May of each year, with fewer grasses remaining each year.
+Appendix Fig.
+\ref{fig:photos} shows photographs of example quadrats before and after treatment.
 We repeated mapped censuses of all plots each June from 2012-2016.
@@
 We refer to the model that includes the estimated removal treatment effect as the ``removal model".
-Note that these are not estimated as separate models; we estimate all parameters together, but when we generate predictions from the baseline model, we simply set the removal treatment effects to zero.
+
+\new{Note that the ``baseline" and ``removal" models are not estimated separately; we estimate all parameters together, but when we generate predictions from the baseline model, we simply set the removal treatment effects to zero.
+We chose this over an alternative approach of fitting a model only to data from control plots and then making and evaluating predictions for the removal plots.
+The principal advantage of our approach is that the significance test (confidence intervals) for the removal effect is well-defined; we do not know of valid tests for fitting on one data set and testing on another (GILES??).
+In addition, the alternative approach would return essentially the same results because adding data from the removal plots had negligible effects on other model parameters.
+Correlations between the parameters of a model fit only to control plot data and our full model (leaving aside the removal effects) were stronger than 0.999.
+Why don't we find strong interactions between removal treatments and the estimated competition coefficients?
+The answer is that because our treatments involved complete, not just partial, removal of target species, their effects can be captured with an intercept term, as illustrated in Fig.
+\ref{fig:PSSPresids}.
+As a result, if, for example, one of the grasses responds more positively to removal of \textit{A.
+tripartita} than our baseline model predicts, we cannot determine whether this reflects a release from competition (i.e.
+a poorly estimated competition coefficient) or some other effect of our removal treatment  independent of competitor densities.
+}
 
@@
 For each scenario, we initialized all species with very low cover and ran the model for 500 time steps (years) to allow the community to reach its equilibrium (i.e., its steady-state pattern of fluctuations in response to environmental variability).
-We calculated the \new{average steady-state cover} for each species by simulating an additional 2000 time steps and averaging each species' cover over this period.} The first scenario featured all four species and our baseline model for control plots (no removal effects).
+We calculated the \new{average steady-state cover for each species by simulating an additional 2000 time steps and averaging each species' cover over this period.} The first scenario featured all four species and our baseline model for control plots (no removal effects).
 For the second scenario, we simulated our experimental removals by setting either the cover of \textit{A.
@@
 
-Our population models based on historical, observational data reveal weak interspecific interactions \citep{adler_coexistence_2010,chu_large_2015}. As a result, the 
-models predict very little competitive release among the four dominant species in a sagebrush steppe. Because the models were based solely on observational
-data, we were suspicious of our model's ability to accurately describe interspecific interactions, so we conducted a removal experiment to test it. 
+\subsection*{An experimental test of model predictions}
+
+Our population model based on historical, observational data revealed weak interspecific interactions \citep{adler_coexistence_2010,chu_large_2015}.
+As a result, the model predicts very little competitive release among the four dominant species in a sagebrush steppe.
+Because the model was based solely on observational
+data, we were suspicious of its ability to accurately describe the outcome of interspecific interactions, so we conducted a removal experiment to test it. 
+\new{Our approach represents a strong test of a surprising prediction for two reasons. First, the prediction was already published; the experiment 
+gave us a very real opportunity to be publicly wrong. Second, testing an observationally-based model with an experimental manipulation, 
+not just additional observations, further increases the potential to confront the model with a novel situation outside the historical range of variation.}
 The results provide substantial support for our model's predictions, and for the weakness of interspecific interactions in this community.
-
-% not sure yet where this should go
-\new{Our study represents a very strong test of a surprising prediction for two reasons. First, the prediction was already published; the experiment 
-represented a very real chance of getting it wrong...Second, we were just testing an observationally based model with new observations, but rather with
-an experimental manipulation, increasing the potential to confront the model with a novel situation outside the historical range of variation...
-Add Steve's point about testing population models--in contrast to the Crone study, we do well when we control for environmental variation with random year effects...}
 
@@
 
-The weak interspecific interactions among many of the dominant species in our system implies that these species  
+\new{What general lessons can we draw about testing predictions from population models? 
+First, partial removals, rather than complete removal treatments, could 
+have helped us to distinguish between competition-mediated and density-independent effects of removal treatments (e.g. Fig. \ref{fig:PSSPresids}) 
+Future work in other systems might benefit from this insight.
+Second, the success of our models supports the use of long-term observational data for characterizing competitive interactions.
+We see no reason why this approach would not work just as well in a community with stronger
+interspecific interactions. After all, our approach did detect strong intraspecific effects. 
+Finally, our results give cause for optimism about population forecasting.
+\citet{Crone2013} found that matrix projection models for plants failed to forecast population trends due environmental differences 
+between the data collection and forecast periods. Our population models performed quite well after accounting for year-to-year 
+environmental variation with random year effects. }
+
+\subsection*{Implications of weak interspecific interactions}
+
+The weak interspecific interactions among many of the dominant species in our system indicates that these species  
 are limited by quite different factors. This low niche overlap could reflect strong resource partitioning \citep{tilman_resource_1982}, 
@@
 \section*{Acknowledgements}
-Funding was provided by NSF grants DEB-1353078,and  DEB-1054040 to PBA and DEB-1353039 to SPE and GH, and by the Utah Agricultural Experiment Station. 
+Funding was provided by NSF grants DEB-1353078 and DEB-1054040 to PBA and DEB-1353039 to SPE and GH, and by the Utah Agricultural Experiment Station, which approves this contribution as journal paper number 9033.
 The USDA-ARS Sheep Experiment Station generously provided access to historical data and the field experiment site. 
-We thank Robin Snyder for suggestions that improved the manuscript.
+We thank Robin Snyder, Tom Miller, and two anonymous reviewers for suggestions that improved the manuscript.
 
@@
  \centering
+ \includegraphics[width=0.7\textwidth]{plate}
+ \caption{Examples of a grass removal quadrat before treatment in 2011 (A) and four years after treatment in 2015 (B), and an \textit{A.
+tripartita} removal quadrat before treatment in 2011 (C) and again in 2015.
+Basal cover of the perennial grasses in (A) was 4.6\% and canopy cover of  \textit{A.
+tripartita} in (B) was 22.9\%.
+}
+ \label{fig:photos}
+ \end{figure}
+
+ \begin{figure}[h]
+ \centering
+ \includegraphics[width=0.7\textwidth]{PSSP_marginalWARTR}
+ \caption{Marginal effects of \textit{A. tripartita} crowding (W) on \textit{P. spicata} growth. Points are residuals from a model that included all 
+ covariates from our standard model except for \textit{A.
+tripartita} crowding.
+Grey points show (residual) growth of individual plants in control plots, red points show plants in \textit{A.
+tripartita} removal plots.
+The blue symbol shows the mean of the red points.
+Source file: \texttt{growth\textbackslash PSSPgrowth.r}.
+}
+ \label{fig:PSSPresids}
+ \end{figure}
+
+ \begin{figure}[h]
+ \centering
  \includegraphics[width=0.7\textwidth]{CompKernels}
@@
 spicata}.
-\response{Each point compares the crowding experienced by one \textit{P.
+\new{Each point compares the crowding experienced by one \textit{P.
 spicata} individual in one year from the species on the x-axis to the species on the y-axis.}   Blue symbols show values from controls plots, red symbols show \new{pre-treatment} values from \textit{Artemisia} removal plots.``W.allcov'' refers to the aggregated crowding by all shrubs and perennial grasses beyond the focal dominant species, and ``W.allpts'' refers to the aggregated crowding by forb species.
@@
 Blue lines and symbols show one-step-ahead predictions from the baseline IBM (no removal treatment coefficients), averaged across quadrats.
-Red lines and symbols show one-step-ahead predictions from an IBM that includes removal treatment coefficients.
+Red lines and symbols show one-step-ahead predictions from an IBM that includes removal treatment coefficients assigned the outermost value of the 95\% credible intervals.
   Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r}. }

### 2017-09-18 removals/manuscript/ResponseToReviews.tex
Made Giles' small changes

@@
 In particular, we have 1) added text to the Introduction and Discussion to emphasize the importance, generality, and novelty of our work, and 2) added new text and supplementary figures to the Methods to justify the decisions we made in designing our data analysis.
-In our point-by-point response below, the original reviewer comments are in black and \response{our responses are in blue.} We hope you fill this version of the manuscript suitable for publication.
+In our point-by-point response below, the original reviewer comments are in black and \response{our responses are in blue.} We hope you will find this version of the manuscript suitable for publication.
 
@@
 However, for these long-term projections, we just wanted to focus on changes in cover for a hypothetical, average plot.
-If we incorporated the spatial random effects, we would six slightly different versions of this plot, but because none of the other parameters vary spatially, the effects of removals on each species would not change much at all.
+If we incorporated the spatial random effects, we would generate six slightly different versions of this plot, but because none of the other parameters vary spatially, the effects of removals on each species would not change much at all.
 We have added a sentence to the Methods justifying our choice to exclude spatial variation.
@@
 It seemed that this would somehow “equalize” parameter uncertainty (in competition, yearly intercept, removal parameters, etc.).
-I’m not sure if this would change the conclusions, the only thing it might do is make the variance larger in Fig.
+I'm not sure if this would change the conclusions, the only thing it might do is make the variance larger in Fig.
 5, and thus the P.
@@
 spicata models.
-I’m not saying you need to do this, I’m just curious if you had considered this issue at all.
+I'm not saying you need to do this, I'm just curious if you had considered this issue at all.
 
@@
 spicata} the small effect of competition with ARTR each year adds up (in the model) to a nontrivial long-term effect.
-We have added a new paragraph to the Methods section "Individual Based Models" that makes these points.
+We have added a new paragraph to the Methods section ``Individual Based Models" that makes these points.
 }

### 2017-09-18 removals/manuscript/removal_ms.tex
Made Giles' small changes

@@
 If our ``baseline" \new{regression} model, which accounts for local plant-plant interactions, accurately explains the observed responses to removals, then the removal coefficient should be non-significant.
-For survival and recruitment, the removal coefficients were never significantly positive, indicating that the baseline model did not underestimate intraspecific competition.
-For \textit{Poa secunda} recruitment, the removal coefficient was negative, indicating that our baseline model overestimated intraspecific competition.
-For growth, the removal treatment effect was significant and positive for two species, \textit{Poa secunda} and \textit{Pseudoroegneria spicata}, indicating that the baseline model underestimated intraspecific competition.
+For survival and recruitment, the removal coefficients were never significantly positive, indicating that the baseline model did not underestimate interspecific competition.
+For \textit{Poa secunda} recruitment, the removal coefficient was negative, indicating that our baseline model overestimated intspecific competition.
+For growth, the removal treatment effect was significant and positive for two species, \textit{Poa secunda} and \textit{Pseudoroegneria spicata}, indicating that the baseline model underestimated interspecific competition.
 However, including information about the location of grass individuals with respect to removed \textit{A.
@@
 To address this possibility, we also simulated a version of the model in which we replaced the point estimate of each removal treatment effect with either the 5\% or 95\% credible limit for the treatment effect, whichever was farthest from zero.
-This version of the model shows the population-level changes we would expect given maximum removal effects.
+This version of the model shows the population-level changes we would expect given maximum removal effects that are consistent with our data.
 
@@
  the response variable, against the $W$ values of all other species, the independent variables. Because some $W=0$, we conducted two 
- separate regressions. First, using all $W$'s, we fitted a generalized linear model with a logit link function to evaluate whether t
- he probability that the focal species' $W$ equals 0 is influenced by the value of other species' $W$'s. In this model, the dependent 
+ separate regressions. First, using all $W$'s, we fitted a generalized linear model with a logit link function to evaluate whether the 
+ probability that the focal species' $W$ equals 0 is influenced by the value of other species' $W$'s. In this model, the dependent 
  variable is a Bernoulli variate coding for zero versus non-zero value of the focal species' crowding, and the independent 

### 2017-09-22 removals/manuscript/ResponseToReviews.tex
edits on ms

@@
 \response{We addressed this point above, in response to Miller's comments.
-Briefly, adding data from the removal experiments does not alter estimates of competition coefficients, but rather alters estimated of model intercepts because crowding from removed species must equal zero.
+Briefly, adding data from the removal experiments does not alter estimates of competition coefficients, but rather alters only estimates of model intercepts because crowding from removed species must equal zero.
 As a result, a separate intercept for removal plots captures the effects of the removals.
@@
 
-\response{We are not sure what correlation the reviewer is interested in (correlations in the MCMC chains?), but perhaps our assurance that
-adding the removal data does not alter the competition coefficients helps allay the implicit concern.}
+\response{We are not sure what correlation the reviewer is interested in 
+(correlations in the MCMC chains?), but perhaps our assurance that
+adding the removal data does not alter the competition coefficients 
+helps allay the implicit concern.}
 
@@
 
-\response{ This is a good point, and we appreciate being pushed to justify our decision to ignore parameter uncertainty. Our goal is to test
+\response{This is a good point, and we appreciate being pushed to justify our decision to ignore parameter uncertainty. Our goal is to test
 whether competitive release is as small as our previous model predicted. To be conservative, we want to make it as easy as possible to
@@
 5 is (for 3 species) small compared to the natural fluctuations in cover driven by interannual environmental variability.
-If, as suggested, we made the projections with parameter uncertainty on top of random year effects, we would be conflating parameter uncertainty with environmental variability, ``tilting'' the
-comparison in our favor by overstating the magnitude of natural fluctuations.
+If, as suggested, we made the projections with parameter uncertainty on top of random year effects, we would be conflating parameter uncertainty with environmental variability, ``tilting'' the comparison in our favor by overstating the magnitude of natural fluctuations.
 One point of Fig.

### 2017-09-22 removals/manuscript/removal_ms.tex
edits on ms

@@
 Compared to traditional field experiments, our approach has the advantage of projecting full life-cycle responses over long time periods, but its reliance on observational data sets limits the strength of our inference.
-A successful experimental test could increase our confidence in the model's projections of competitive release.
-
-\new{ The idea that results from observational studies of competition represent hypotheses which should be tested experimentally is not new \citep{Hairston1980}, but such tests remain extremely rare.
+A successful experimental test would increase our confidence in the model's projections of competitive release.
+
+\new{The idea that results from observational studies of competition represent hypotheses which should be tested experimentally is not new \citep{Hairston1980}, but such tests remain extremely rare.
 In fact, independent tests of any kind of quantitative prediction are rare in ecology, spurring recent calls for a greater emphasis on prediction \citep{Mouquet2015,Petchey2015,Houlahan2017}.
-These authors argue that testing predictions will not only advance applied goals, such as forecasting ecological impacts of global change, but will also accelerate progress in basic research.
-}
-
-\new{ In the spirit of this argument, } we conducted a removal experiment to test whether our sagebrush steppe community model, which is based on observational data, underestimates the strength of interspecific competition and competitive release.
+These authors argue that testing predictions will not only advance applied goals such as forecasting ecological impacts of global change, but will also accelerate progress in basic research.}
+
+\new{In the spirit of this argument,} we conducted a removal experiment to test whether our sagebrush steppe community model, which is based on observational data, underestimates the strength of interspecific competition and competitive release.
 We studied the responses of three perennial grass species to removal of sagebrush, and the response of sagebrush to removal of all grasses.
-The primary purpose of the experiment was not to directly measure competitive release, which is difficult to do over a short time period in a semiarid ecosystem with low and variable plant cover, but rather \new{to test whether our model accurately characterizes} the interspecific interactions that ultimately determine the magnitude of competitive release.
-We addressed two research questions: First, are the survival, growth or recruitment of our target species significantly different in control vs.
-removal treatments after accounting for differences in local species composition?
-Because our demographic models include the effects of neighborhood crowding, they should be able to predict how experimental removal of neighbors alters demographic performance.
+The primary purpose of the experiment was not to directly measure competitive release, which is difficult to do over a short time period in a semiarid ecosystem with low and variable plant cover, but rather \new{to test whether our model accurately characterizes} the interspecific interactions that determine the magnitude of competitive release.
+We addressed two questions: First, are the survival, growth or recruitment of our target species significantly different in control vs.
+removal treatments, after accounting for differences in local species composition?
+Because our demographic models include the effects of neighborhood crowding, they should be able to predict how experimental removal of neighbors alters demographic rates.
 If the full effect of experimental removals is stronger than predicted, it means that our statistical models underestimate \new{the response to decreased crowding}.
 Our second question integrates across the full life cycle: how do short and long-term population-level responses to removals compare to projections from multispecies population models based on observational data?
-If models based on observational data adequately describe facilitative and competitive interactions, the population dynamics they project should be similar to projections from models that include information from the experimental removal treatments, a result which would validate the use of models and long-term data sets to study interspecific interactions.
+If models based on observational data adequately describe facilitative and competitive interactions, the population dynamics they project should be similar to projections from models that include information from the experimental removal treatments.
+Obtaining that result would validate the use of long-term data sets and modeling to study interspecific interactions.
 
@@
 
-\new{Note that the ``baseline" and ``removal" models are not estimated separately; we estimate all parameters together, but when we generate predictions from the baseline model, we simply set the removal treatment effects to zero.
-We chose this over an alternative approach of fitting a model only to data from control plots and then making and evaluating predictions for the removal plots.
-The principal advantage of our approach is that the significance test (confidence intervals) for the removal effect is well-defined; we do not know of valid tests for fitting on one data set and testing on another (GILES??).
-In addition, the alternative approach would return essentially the same results because adding data from the removal plots had negligible effects on other model parameters.
-Correlations between the parameters of a model fit only to control plot data and our full model (leaving aside the removal effects) were stronger than 0.999.
-Why don't we find strong interactions between removal treatments and the estimated competition coefficients?
-The answer is that because our treatments involved complete, not just partial, removal of target species, their effects can be captured with an intercept term, as illustrated in Fig.
-\ref{fig:PSSPresids}.
-As a result, if, for example, one of the grasses responds more positively to removal of \textit{A.
-tripartita} than our baseline model predicts, we cannot determine whether this reflects a release from competition (i.e.
-a poorly estimated competition coefficient) or some other effect of our removal treatment  independent of competitor densities.
-}
+\new{Note that the ``baseline" and ``removal" models are not estimated separately; we estimate all parameters together, but when we generate predictions from the baseline model, 
+we simply set the removal treatment effects to zero. We chose this over an alternative approach of fitting a model only to data from control plots and then making 
+and evaluating predictions for the removal plots. The principal advantage of our approach is that the significance test (confidence intervals) for the removal effect 
+is well-defined; we do not know of valid tests for fitting on one data set and testing on another (GILES??). In addition, the alternative approach would return 
+essentially the same results because adding data from the removal plots had negligible effects on other model parameters. Correlations between the parameters of 
+a model fit only to control plot data and our full model (leaving aside the removal effects) were stronger than 0.999. 
+Why don't we find strong interactions between removal treatments and the estimated competition coefficients? The answer is that because our treatments 
+involved complete, not just partial, removal of target species, their effects can be captured with an intercept term, as illustrated in Fig. \ref{fig:PSSPresids}. 
+As a result, if, for example, one of the grasses responds more positively to removal of \textit{A. tripartita} than our baseline model predicts, we cannot 
+determine whether this reflects a poorly estimated competition coefficient, or some other effect of the removal treatment besides competitor densities.}
 
@@
 
-To simulate changes in cover in each quadrat from year $t$ to year $t+1$, we initialized the IBM with the actual configuration of the four focal species observed in year $t$ (genet sizes and locations).
+To simulate changes in cover in each quadrat from year $t$ to year $t+1$, we initialized the IBM with the actual 
+configuration of the four focal species observed in year $t$ (genet sizes and locations). 
 We then projected the model forward one year, applying the random year effects corresponding to year $t$ and the spatial random effects corresponding to the quadrat of interest. 
@@
 
-\new{Although all the parameters in our model are estimated with uncertainty, we chose not to propage parameter uncertainty through the 
+\new{Although all the parameters in our model are estimated with uncertainty, we chose not to propage parameter uncertainty through these 
 simulations because we wanted to maximize our ability to detect the population-level consequences of a small removal effect. 
-Our previously published results showed that interspecific competition is weak relative to the effects of interannual environmental variability 
-\citep{Tredennick2017}. If we included parameter uncertainty along with random year effects, we would be conflating parameter uncertainty 
-with environmental variability, overstating the magnitude of natural fluctuations, and making it more difficult to show that competitive release
-is larger than the observationally-based model predicts.
-Therefore, we used estimated means of all parameters, including those with confidence intervals that overlapped zero.
-Based on the same rationale, we used a deterministic version of the model (e.g.,
+We previously found that interspecific competition is weak relative to the effects of interannual environmental variability 
+\citep{Tredennick2017}. Including parameter uncertainty along with random year effects would conflate parameter uncertainty 
+with environmental variability, overstating the magnitude of natural fluctuations and making it more difficult to show that competitive release
+is larger than the observationally-based model predicts. Therefore, we used estimated means of all parameters, including those with 
+confidence intervals that overlapped zero. Based on the same rationale, we used a deterministic version of the model (e.g., 
 recruitment is the $\lambda$ of Eqn. \ref{eqn:recrProcessModel}, rather than a Poisson random variable with mean $\lambda$). 
@@
 For each scenario, we initialized all species with very low cover and ran the model for 500 time steps (years) to allow the community to reach its equilibrium (i.e., its steady-state pattern of fluctuations in response to environmental variability).
-We calculated the \new{average steady-state cover for each species by simulating an additional 2000 time steps and averaging each species' cover over this period.} The first scenario featured all four species and our baseline model for control plots (no removal effects).
+We calculated the \new{average steady-state cover for each species by simulating an additional 2000 time steps and averaging each species' cover over this period.} The first of the three scenarios featured all four species and our baseline model for control plots (no removal effects).
 For the second scenario, we simulated our experimental removals by setting either the cover of \textit{A.
@@
 (Fig. \ref{fig:CoverTrends}), but shrub removal did appear to cause an increase in the cover of \textit{P. spicata}. 
-However, when we tested the effects of removal treatments on year-to-year population growth rates ($log(Cover_{t+1}/Cover_t)$) at the quadrat level,
+However, when we tested the effects of removal treatments on year-to-year population growth rates ($\log(Cover_{t+1}/Cover_t)$) at the quadrat level,
  none of the removal treatment effects were significantly different from zero (Table \ref{table:coefficients}). For \textit{P. spicata}, 
@@
 a nonlinear fashion, with little increase in production at sites with low initial sagebrush cover, 
-and accelerating responses as initial sagebrush cover passes some threshold \citep{McDaniel2005}, \new{implying a shift in the nature of interspecific interactions, perhaps due to changes in the balance of competition and facilitation or in resource supply.} The cover of
+and accelerating responses as initial sagebrush cover passes some threshold \citep{McDaniel2005}, \new{implying a 
+shift in the nature of interspecific interactions, perhaps due to changes in the balance of 
+competition and facilitation or in resource supply.} The cover of 
 sagebrush in our ungrazed quadrats may have fallen below that threshold.
@@
 have helped us to distinguish between competition-mediated and density-independent effects of removal treatments (e.g. Fig. \ref{fig:PSSPresids}) 
-Future work in other systems might benefit from this insight.
-Second, the success of our models supports the use of long-term observational data for characterizing competitive interactions.
-We see no reason why this approach would not work just as well in a community with stronger
+Future work in other systems might benefit from this insight. Second, the success of our models supports 
+the use of long-term observational data for characterizing competitive interactions. We see no reason why 
+this approach would not work just as well in a community with stronger 
 interspecific interactions. After all, our approach did detect strong intraspecific effects. 
@@
 \section*{Acknowledgements}
-Funding was provided by NSF grants DEB-1353078 and DEB-1054040 to PBA and DEB-1353039 to SPE and GH, and by the Utah Agricultural Experiment Station, which approves this contribution as journal paper number 9033.
+Funding was provided by NSF grants DEB-1353078 and DEB-1054040 to PBA and DEB-1353039 to SPE and GH, and 
+by the Utah Agricultural Experiment Station, which approves this contribution as journal paper number 9033. 
 The USDA-ARS Sheep Experiment Station generously provided access to historical data and the field experiment site. 

### 2017-10-06 removals/manuscript/ResponseToReviews.tex
last bits on removals ms

@@
 
-Another thing that kept crossing my mind as I read the paper was about the treatment of “non-significant” parameters, i.e., parameters with high uncertainty.
+Another thing that kept crossing my mind as I read the paper was about the treatment of ``non-significant'' parameters, i.e., parameters with high uncertainty.
 As far as I could tell throughout the paper, parameters were never dropped from models, and projections were done using the parameter estimates based on the mean of the MCMC chain values.
@@
 \response{This is a good point, and we appreciate being pushed to justify our decision to ignore parameter uncertainty. Our goal is to test
-whether competitive release is as small as our previous model predicted. To be conservative, we want to make it as easy as possible to
-model large effects of removals. (That is why we included the ``maximum removal effect" simulation based on the outer limits of the removal effect
-confidence intervals.) Propagating parameter uncertainty through our simulations would add noise, making it \emph{harder} to 
+whether competitive release is as small as our previous model predicted. To be conservative, we want to make it as easy as possible for the model to show 
+large effects from removals. That is why we included the ``maximum removal effect'' simulation based on the outer limits of the removal effect
+confidence intervals. Propagating parameter uncertainty through our simulations would add noise, making it \emph{harder} to 
 detect the effects of removals. More formally, our previous result showed that the interspecific competition coefficients are small in an ecologically
@@
 5 is (for 3 species) small compared to the natural fluctuations in cover driven by interannual environmental variability.
-If, as suggested, we made the projections with parameter uncertainty on top of random year effects, we would be conflating parameter uncertainty with environmental variability, ``tilting'' the comparison in our favor by overstating the magnitude of natural fluctuations.
-One point of Fig.
-5 is that for \emph{P.
-spicata} the small effect of competition with ARTR each year adds up (in the model) to a nontrivial long-term effect.
-We have added a new paragraph to the Methods section ``Individual Based Models" that makes these points.
-}
+If, as suggested, we had made those projections with parameter uncertainty on top of the random year effects,
+we would be conflating parameter uncertainty with environmental variability, ``tilting'' the comparison in our favor by overstating the magnitude 
+of natural fluctuations. One point of Fig. 5 is that for \emph{P. spicata} the small effect of competition with ARTR each year add'' 
+up (in the model) to a nontrivial long-term effect. We have added a new paragraph to the Methods section ``Individual Based Models" that makes these points. }
 
@@
 \response{We all agreed that this is the first time we have been accused of underselling a result! We have added a new paragraph to the Introduction
-to highlight the novelty of our approach and to show how it answers a growing chorus of calls for more predictive tests. Also, in the Discussion, we 
+to highlight the novelty of our approach and to show how it answers a growing chorus of calls for predictive tests. Also, in the Discussion, we 
 added some text to further emphasize why our study is a strong test of a prediction. }
@@
 
-\response{Stated in first sentence of abstract.}
+\response{Stated in first sentence of Abstract.}
 
-Lines 19-20: instead of “population models based on vital rate regressions” maybe say “population projections based on vital rate regressions” or “population predictions” so it does not sound like you are fitting new models
+Lines 19-20: instead of ``population models based on vital rate regressions'' maybe say ``population projections based on vital rate regressions" 
+or ``population predictions'' so it does not sound like you are fitting new models.
 
@@
 
-Line 20: The first time I read this I was not sure what you meant by “vital rate regressions”. Maybe put this in parenthesis up in lines 7-9?
+Line 20: The first time I read this I was not sure what you meant by ``vital rate regressions''. Maybe put this in parenthesis up in lines 7-9?
 
@@
 
-Line25: while a competition coefficient is often understood to represent resource competition, it can also represent any other density dependent effect like microbial effects, right?
+Line 25: while a competition coefficient is often understood to represent resource competition, it can also represent any other 
+density dependent effect like microbial effects, right?
 
-\response{Good point, we changed ``resource competition" to ``feedbacks."}
+\response{Good point, we changed ``resource competition" to ``feedbacks''. }
 
@@
 
-\response{We added the following sentence, "We repeated mapped censuses of all plots each June from 2012-2016."}
+\response{We added the following sentence, ``We repeated mapped censuses of all plots each June from 2012-2016''.}
 
@@
 
-\response{We added this sentence to the Methods: "For plants in removal plots in 2012, we set crowding for removed neighbor species to zero, rather than using the pre-treatment (2011) data."}
+\response{We added this sentence to the Methods: ``For plants in removal plots in 2012, we set crowding for removed neighbor species to zero, rather than using the pre-treatment (2011) data.''}
 
@@
 
-Lines 181-198: if the removal effect is not “significant” (credible intervals overlap 0) is it still included in projections?
+Lines 181-198: if the removal effect is not ``significant'' (credible intervals overlap 0) is it still included in projections?
 
@@
 
-Line 263:  When I read this I wanted to know something about the uncertainty in these parameters – do most credible intervals overlap zero, or are the coefficients small but “significant”?
+Line 263:  When I read this I wanted to know something about the uncertainty in these parameters – do most credible intervals overlap zero, or are the coefficients small but ``significant''?
 
@@
 
-Lines 277-283: Does the root zone of Artemisia generally fall under its canopy?
+Lines 277-283: Does the root zone of \emph{Artemisia} generally fall under its canopy?
 Or is it larger in diameter than the aboveground canopy?
@@
 
-Line 291: Can you add some explanation to the “break covariances in plant densities” part? It took me a while to figure out what you meant.
+Line 291: Can you add some explanation to the ``break covariances in plant densities'' part? It took me a while to figure out what you meant.
 
-\response{We replaced ``break covariances in plant densities among species" with ``result in combinations of neighborhood-scale densities that deviate from naturally occurring patterns of correlation."}
+\response{We replaced ``break covariances in plant densities among species" with ``result in combinations of neighborhood-scale densities that deviate from 
+naturally occurring patterns of correlation.''}
 
 Discussion
-Line 421: remove one of the “dynamics” words
+Line 421: remove one of the ``dynamics” words
 
@@
 
-\response{We addressed this point in detail in response to Reviewer 2's comment below.}
+\response{We address this point in detail in response to Reviewer 2's comment below.}
 
@@
 
-\response{We have added new Discussion paragraph to directly address these questions about methodological
-generalizations (lines ???). }
+\response{We have added a new Discussion paragraph to directly address these questions about methodological 
+generalizations (line 458).}
 
@@
 \response{We can offer multiple lines of evidence that low grass cover is not the cause of weak grass effects. 
-First, grass cover is high enough to cause strong intraspecific competition. The puzzle is why interspecific effects
-of grasses are so weak when their intraspecific effects are strong. Second, our growth and survival models do not
+First, grass cover is high enough to cause the strong intraspecific competition that our data analysis detected. 
+The puzzle is why interspecific effects of grasses are so much weaker than their intraspecific effects. Second, our growth and survival models do not
 depend on average cover but rather neighborhoood scale cover: even if average cover is low, a
@@
 grasses are mapped in the currency of basal cover, while for Artemisia canopy cover is mapped. To help readers
-appreciate that small values of basal cover are not trivial, we have added a new supplementary figure (C-1) showing 
+appreciate that the observed values of grass basal cover are not trivial, we have added a new supplementary figure (C-1) showing 
 photos of example quadrats before and after the removal treatment.  }
@@
 
-\response{This is a fair point. We added a sentence noting that grasses in interspaces could still experience competition
-if shrub roots extend beyond canopies.}
+\response{This is a fair point. We added a sentence noting that grasses in interspaces 
+could still experience competition if shrub roots extend beyond canopies.}
 
-Given this result for P. spicata, line 401 of the discussion is a bit surprising: “how to reconcile the
-insensitivity of grasses to sagebrush removal…..” On the other hand, it does hold for two of the grasses
+Given this result for P. spicata, line 401 of the discussion is a bit surprising: ``how to reconcile the
+insensitivity of grasses to sagebrush removal... .'' On the other hand, it does hold for two of the grasses
 and some discussion of why the grass species differ in sensitivity to shrub removal would be interesting.
@@
 
-Line 81: should “determine” be replaced by “estimate”?
+Line 81: should ``determine'' be replaced by ``estimate''?
 
@@
 Line 87: given that you earlier defined competitive release as a population response, should
-“competitive release” at the end of this sentence be replaced by “individual response to neighbors”?
+``competitive release'' at the end of this sentence be replaced by ``individual response to neighbors''?
 The second question more directly addresses the population consequences.
 
-\response{We went with ``underestimate the response to decreased crowding" because our recruitment model does not operate 
+\response{We went with ``underestimate the response to decreased crowding'' because our recruitment model does not operate 
 at the level of individuals.}
 
-Line 216: “very similar to those used in our previous studies” is a bit confusing. Why not just use the
+Line 216: ``very similar to those used in our previous studies'' is a bit confusing. Why not just use the
 vital rate functions calculated in the first part of the study? If they are indeed the same, say so rather
@@
 Line 234-235: Several confusing points here: a) The distinction between the community reaching
-equilibrium (500 time steps) and each species reaching equilibrium cover (“additional 2000 time steps”)
+equilibrium (500 time steps) and each species reaching equilibrium cover (``additional 2000 time steps'')
 is not clear. b) I assume the random year effects were used for all 2500 time steps? c) were the 3
@@
 
-\response{ This was bad phrasing on our part. 
+\response{This was bad phrasing on our part, which we have corrected. 
 The community equilibrium is not a constant level of cover for each species, as it would be in a deterministic

### 2017-10-06 removals/manuscript/removal_ms.tex
last bits on removals ms

@@
 %%%  Sorting references by year(?)
-\usepackage[authoryear,sectionbib,sort]{natbib}
+\usepackage[authoryear,sort]{natbib}
 %%%
@@
 \usepackage[usenames]{color}
-\newcommand{\new}{\textcolor{red}}
-\newcommand{\comment}{\textcolor{blue}}
+\newcommand{\new}{\textcolor{black}}
+\newcommand{\comment}{\textcolor{black}}
 
@@
 \new{Our approach represents a strong test of a surprising prediction for two reasons. First, the prediction was already published; the experiment 
-gave us a very real opportunity to be publicly wrong. Second, testing an observationally-based model with an experimental manipulation, 
+gave us a very real opportunity to be wrong in public. Second, testing an observationally-based model with an experimental manipulation, 
 not just additional observations, further increases the potential to confront the model with a novel situation outside the historical range of variation.}
@@
 interspecific interactions. After all, our approach did detect strong intraspecific effects. 
-Finally, our results give cause for optimism about population forecasting.
-\citet{Crone2013} found that matrix projection models for plants failed to forecast population trends due environmental differences 
-between the data collection and forecast periods. Our population models performed quite well after accounting for year-to-year 
-environmental variation with random year effects. }
+Finally, our results give some cause for optimism about population forecasting.
+\citet{Crone2013} found that matrix projection models for plants often failed to forecast population trends due 
+to environmental differences between the data collection and forecast periods. Our population models 
+performed quite well, after accounting for year-to-year environmental variation through the inclusion of
+random year effects.}
 

### 2017-10-12 removals/manuscript/removal_ms.tex
trivia

@@
 and evaluating predictions for the removal plots. The principal advantage of our approach is that the significance test (confidence intervals) for the removal effect 
-is well-defined; we do not know of valid tests for fitting on one data set and testing on another (GILES??). In addition, the alternative approach would return 
+is well-defined; we do not know of valid tests for fitting on one data set and testing on another. In addition, the alternative approach would return 
 essentially the same results because adding data from the removal plots had negligible effects on other model parameters. Correlations between the parameters of 

### 2017-10-16 removals/manuscript/ResponseToReviews.tex
Final edits to revision and response letter.

@@
 In particular, we have 1) added text to the Introduction and Discussion to emphasize the importance, generality, and novelty of our work, and 2) added new text and supplementary figures to the Methods to justify the decisions we made in designing our data analysis.
-In our point-by-point response below, the original reviewer comments are in black and \response{our responses are in blue.} We hope you will find this version of the manuscript suitable for publication.
+In our point-by-point response below, the original reviewer comments are in black and \response{our responses are in blue.} In the manuscript file, we have highlighted new and revised sections with red print.
+We hope you will find this version of the manuscript suitable for publication.
 
@@
 So there is no evidence of interactions between the removal treatment and other parameters in the model.
-Second, to clarify why all the differences between demographic performance in control and removal plots can be captured by one removal coefficient that only alters the overall intercept, we drew a new figure (Appendix C-2).
+}
+
+\response{Second, to clarify why all the differences between demographic performance in control and removal plots can be captured by one removal coefficient that only alters the overall intercept, we drew a new figure (Appendix C-2).
 To generate this figure, we fit a version of the growth model for P.
@@
 That said, while we now know how to conduct an even stronger test of our model in the future, the test we did conduct does represent a significant advance beyond the existing state of the art.
-To clarify these issues, we have added a new paragraph to the Methods section, along with the new Appendix figure (C-2).
-We also mention in the Discussion how we would do it differently next time.}
+To clarify these issues, we have added a new paragraph to the Methods section (lines 182-198), along with the new Appendix figure (C-2).
+We also mention in the Discussion how we would do it differently next time (line 460).
+}
 
@@
 If we incorporated the spatial random effects, we would generate six slightly different versions of this plot, but because none of the other parameters vary spatially, the effects of removals on each species would not change much at all.
-We have added a sentence to the Methods justifying our choice to exclude spatial variation.
+We have added a sentence to the Methods clarifying our choice to exclude spatial variation (line 255).
 }
@@
 
-\response{We added a new "lessons learned" paragraph to the Discussion where we address this point (yes, we see no reason why
+\response{We added a new ``lessons learned" paragraph to the Discussion where we address this point (yes, we see no reason why
 multispecies models would not also work in communities with stronger competition), along with other generalizations 
-from our experience on this project.  }
+from our experience on this project (lines 459-470).  }
 
@@
 \response{We addressed this point above, in response to Miller's comments.
-Briefly, adding data from the removal experiments does not alter estimates of competition coefficients, but rather alters only estimates of model intercepts because crowding from removed species must equal zero.
+Briefly, adding data from the removal experiments does not alter estimates of competition coefficients, but rather alters only estimates of model intercepts because crowding from removed species always equals zero.
 As a result, a separate intercept for removal plots captures the effects of the removals.
@@
 we would be conflating parameter uncertainty with environmental variability, ``tilting'' the comparison in our favor by overstating the magnitude 
-of natural fluctuations. One point of Fig. 5 is that for \emph{P. spicata} the small effect of competition with ARTR each year add'' 
-up (in the model) to a nontrivial long-term effect. We have added a new paragraph to the Methods section ``Individual Based Models" that makes these points. }
+of natural fluctuations. One point of Fig. 5 is that for \emph{P. spicata} the small effect of competition with ARTR each year add 
+up (in the model) to a nontrivial long-term effect. We have added a new paragraph to the Methods section ``Individual Based Models" that makes these points (lines 219-231). }
 
@@
 \response{We all agreed that this is the first time we have been accused of underselling a result! We have added a new paragraph to the Introduction
-to highlight the novelty of our approach and to show how it answers a growing chorus of calls for predictive tests. Also, in the Discussion, we 
-added some text to further emphasize why our study is a strong test of a prediction. }
+to highlight the novelty of our approach and to show how it answers a growing chorus of calls for predictive tests (lines 75-81). Also, in the Discussion, we 
+added some text to further emphasize why our study is a strong test of a prediction (lines 399-403). }
 
@@
 
-\response{The new Methods paragraph addresses these questions directly.}
+\response{The new Methods paragraph (lines 182-198) addresses these questions directly.}
 
@@
 
-\response{We replaced ``break covariances in plant densities among species" with ``result in combinations of neighborhood-scale densities that deviate from 
-naturally occurring patterns of correlation.''}
+\response{We replaced ``break covariances in plant densities among species" with ``result in combinations of neighborhood-scale densities that deviate 
+from naturally occurring patterns of correlation.''}
 
@@
 \response{We have added a new Discussion paragraph to directly address these questions about methodological 
-generalizations (line 458).}
+generalizations (line 459-470).}
 
@@
 showing distribution of crowding values, and exert strong per-unit-area effects on another plant. The grasses do in fact
-affect Artemisa, as shown in Tables B-2 (marginally significant effect of H. comata on Artemisia survival), B-6 (significant 
+affect Artemisia, as shown in Tables B-2 (marginally significant effect of H. comata on Artemisia survival), B-6 (significant 
 effect of Poa on Artemisia growth, and  B-15 (P. spicata has a positive effect on Artemisia recruitment, while H. comata
@@
 \response{This is a fair point. We added a sentence noting that grasses in interspaces 
-could still experience competition if shrub roots extend beyond canopies.}
+could still experience competition if shrub roots extend beyond canopies (line 438).}
 
@@
 the per capita competition coefficients, potentially by altering the balance between competition and facilitation
-or changing resource availability. }
+or changing resource availability (lines 455-457). }
 
@@
 
-\response{We have added "estimated" later in the sentence to recognize the uncertainty in the model.}
+\response{We have added ``estimated" later in the sentence to recognize the uncertainty in the model.}
 

### 2017-10-16 removals/manuscript/removal_ms.tex
Final edits to revision and response letter.

@@
 %%%  Sorting references by year(?)
-\usepackage[authoryear,sort]{natbib}
+\usepackage[authoryear]{natbib}
 %%%
@@
 \usepackage[usenames]{color}
-\newcommand{\new}{\textcolor{black}}
+\newcommand{\new}{\textcolor{red}}
 \newcommand{\comment}{\textcolor{black}}
@@
 The greater the difference between intra- and interspecific limitations, the more stable the coexistence, and the weaker the competitive release any species should experience following removal of competitors.
-We conducted a removal experiment to test whether a previously estimated model, showing surprisingly weak interspecific competition for four dominant species in a sagebrush steppe, accurately predicts competitive release.} Our removal treatments were 1) removal of all perennial grasses and 2) removal of the dominant shrub, \textit{Artemisia tripartita}.
+We conducted a removal experiment to test whether a previously estimated model, showing surprisingly weak interspecific competition for four dominant species in a sagebrush steppe, accurately predicts competitive release.} Our treatments were 1) removal of all perennial grasses and 2) removal of the dominant shrub, \textit{Artemisia tripartita}.
 We \new{regressed} survival, growth and recruitment on the locations, sizes, and species identities of neighboring plants, along with an indicator variable for removal treatment.
@@
 For survival and recruitment, the removal coefficients were never significantly positive, indicating that the baseline model did not underestimate interspecific competition.
-For \textit{Poa secunda} recruitment, the removal coefficient was negative, indicating that our baseline model overestimated intspecific competition.
+For \textit{Poa secunda} recruitment, the removal coefficient was negative, indicating that our baseline model overestimated interspecific competition.
 For growth, the removal treatment effect was significant and positive for two species, \textit{Poa secunda} and \textit{Pseudoroegneria spicata}, indicating that the baseline model underestimated interspecific competition.
@@
 Neutral theory \citep{hubbell_unified_2001} goes even further, assuming that interspecific competition is exactly as strong as intraspecific competition.
-Although a growing list of studies has emphasized the role of interspecific facilitation in some plant communities \citep{he_global_2013,brooker_facilitation_2008}, it is unclear whether this research has overturned the default assumption that plants compete intensely with other species.
+Although a growing list of studies has emphasized the role of interspecific facilitation in some plant communities \citep{brooker_facilitation_2008,he_global_2013}, it is unclear whether this research has overturned the default assumption that plants compete intensely with other species.
 
@@
 In some cases, competitive release was complete: remaining species fully compensated for removed biomass \citep{leps_nutrient_1999,jutila_effects_2002}.
-In other cases, \new{some competitive release occurred but the degree of compensation} depended on the functional group or dominance of the removed and remaining species \citep{smith_dominant_2003,sala_resource_1989,belsky_effects_1992}.
+In other cases, \new{some competitive release occurred but the degree of compensation} depended on the functional group or dominance of the removed and remaining species \citep{sala_resource_1989,belsky_effects_1992,smith_dominant_2003}.
 Finally, even negative effects of removals have been observed \citep{keddy_effects_1989,gilbert_dominant_2009}, consistent with the recent facilitation literature.
@@
 In fact, at our sagebrush steppe study site  \citep{adler_coexistence_2010}, our model predicts removal of sagebrush would actually reduce the equilibrium cover of the three dominant grass species very slightly, reflecting the facilitative effect of sagebrush on grass recruitment and very weak competitive effects on growth and survival.
-Since we only studied the most common, co-occurring species, we expected to find evidence of stable coexistence, but the predicted negative effect of sagebrush removal on grass cover is surprising given the long history of sagebrush eradication to increase grass forage production in sagebrush rangelands \citep{Cook1963, Mueggler1958, Robertson1947}.
+Since we only studied the most common, co-occurring species, we expected to find evidence of stable coexistence, but the predicted negative effect of sagebrush removal on grass cover is surprising given the long history of sagebrush eradication to increase grass forage production in sagebrush rangelands \citep{Robertson1947,Mueggler1958,Cook1963}.
 Compared to traditional field experiments, our approach has the advantage of projecting full life-cycle responses over long time periods, but its reliance on observational data sets limits the strength of our inference.
@@
 and evaluating predictions for the removal plots. The principal advantage of our approach is that the significance test (confidence intervals) for the removal effect 
-is well-defined; we do not know of valid tests for fitting on one data set and testing on another. In addition, the alternative approach would return 
+is well-defined; fitting on one data set and testing on another would require developing non-standard null distributions to account for variation in both data sets.
+In addition, the alternative approach would return
 essentially the same results because adding data from the removal plots had negligible effects on other model parameters. Correlations between the parameters of 
@@
 As a result, if, for example, one of the grasses responds more positively to removal of \textit{A. tripartita} than our baseline model predicts, we cannot 
-determine whether this reflects a poorly estimated competition coefficient, or some other effect of the removal treatment besides competitor densities.}
+determine whether this reflects a poorly estimated competition coefficient, or some other effect of the removal treatment unrelated to competitor densities.}
 

### 2017-12-21 removals/survival/PSSPsurvival.R
Fixing buggy data; new version of data/idahomodern/speciesdata uploaded.

@@ -96,3 +96,4 @@ m1 <- inla(survives ~ logarea+ Treatment + W.ARTR + W.HECO + W.POSE + W.PSSP+ W.
 #   control.predictor = list(link = 1),control.compute=list(dic=T,mlik=T),
 #   control.inla = list(h = 1e-10),Ntrials=rep(1,nrow(allD)))
 
+

### 2017-12-22 removals/miscChecks.r
Another data and results update. Still not final (a few quad years missing) but much closer to "correct"

@@ -1,0 +1,118 @@
+
+### plot a size distribution after running an IPM simulation
+### and compare to observed
+
+doSpp=2
+
+plot(1,1,type="n",xlim=c(log(0.15),log(max(maxSize*2))),ylim=c(0,0.3),
+  xlab="Size",ylab="Frequency")
+#simulated
+lines(v[[doSpp]],rowMeans(sizeSave[[doSpp]][,(burn.in+1):tlimit]),col="red")
+
+#observed
+infile <- paste0(root,"\\ExperimentTests\\data\\idaho_modern\\speciesData\\",sppList[doSpp],"\\survD.csv")
+tmp=read.csv(infile)
+obsSize <- table(cut(log(tmp$area),breaks=c(v[[doSpp]]-h[doSpp],6)))
+obsSize <- obsSize/sum(obsSize)
+lines(v[[doSpp]],obsSize,col="black")
+
+### compare baseline and removal predictions for quadrat level cover
+# first run IBM summarize_sims1step.r
+ratio=numeric(4)
+for(i in 1:4){
+  if(i==1) {myTrt="No_grass"}else{myTrt="No_shrub"}
+  removal.base=simD[simD$Treatment==myTrt,(6+i)]
+  removal.trt=simD[simD$Treatment==myTrt,(10+i)]
+  tmp=removal.trt/removal.base
+  ratio[i]=mean(tmp,na.rm=T)
+}
+print(ratio)
+
+### POSE growth models
+# first run POSEgrowth.r
+
+# play with LMER models
+
+#baseline model
+m1.lmer <- lmer(logarea.t1~logarea.t0+Treatment+W.ARTR + W.HECO + W.POSE + W.PSSP+ W.allcov + W.allpts +
+              (logarea.t0|year),data=allD)
+summary(m1.lmer)
+
+# try treatment x size interactions
+m2.lmer <- lmer(logarea.t1~logarea.t0*Treatment+W.ARTR + W.HECO + W.POSE + W.PSSP+ W.allcov + W.allpts +
+              (logarea.t0|year),data=allD)
+summary(m2.lmer) # not significant, but close
+
+m3.lmer<- lmer(logarea.t1~logarea.t0+Treatment+W.ARTR + W.HECO  + W.POSE + W.PSSP+W.POSE*Treatment+ W.allcov + W.allpts+
+              (1|Group)+(logarea.t0|year),data=allD) 
+summary(m3.lmer) # SIGNIFICANT
+
[…]

### 2018-01-04 removals/Manuscript/BetterThanJensen.tex
stuff

% This document is for pdflatex
\documentclass[11pt]{article}

\usepackage{amsmath,amsfonts,amssymb,graphicx,natbib,subfigure,setspace}
\usepackage[vmargin=1in,hmargin=1in]{geometry}

\usepackage{enumitem}
\setlist{topsep=.125em,itemsep=-0.15em,leftmargin=0.75cm}

\usepackage[compact,small]{titlesec}

\renewcommand{\floatpagefraction}{0.96}
\renewcommand{\topfraction}{0.99}
\renewcommand{\textfraction}{0.1}

\setlength{\parskip}{0.1in}
\setlength{\parindent}{0.0in}

\clubpenalty = 10000
\widowpenalty = 10000

\numberwithin{equation}{section}

\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand{\ba}{\begin{equation} \begin{aligned}}
\newcommand{\ea}{\end{aligned} \end{equation}}

\sloppy

\usepackage{palatino}
\renewcommand{\baselinestretch}{1.2}

\begin{document}

Here is the replacement for the Jensen correction based on Gaussian growth residuals.

After the growth model is fitted,

\begin{enumerate}
\item  Extract for each plant in the data set (indexed by $k$) 
its \texttt{logarea.t0} $z_{0,k}$, the growth residual $z_{1,k} - \hat{z}_{1,k}$ (here $z_1$ is
the plant's \texttt{logarea.t1} and $\hat{z}_1$ is the predicted logarea), and the predicted variance
in growth $V_k$.

\item Compute the set of scaled residuals $r_k = (z_{1,k} - \hat{z}_{1,k})/\sqrt{V_k}$ and store them
in a vector \texttt{scaledResiduals}.

\end{enumerate}

Then in computing the one-step-ahead predictions, for each growth model the line 
\begin{verbatim}
plants$area.pred <- plants$surv.prob*exp(plants$logarea.pred + V.pred/2)
\end{verbatim} 
is replaced by the following 
\begin{verbatim}
resid.Pred <- sqrt(V.pred)*scaledResiduals
newJensen <- mean(exp(resid.Pred))
plants$area.pred <- plants$surv.prob*exp(plants$logarea.pred)*newJensen
\end{verbatim}

\end{document}

### 2018-01-17 removals/growth/POSEgrowth.R
Adding growth residuals figure

@@ -114,7 +114,7 @@ rm(m2)
 # does effect diminish with time?
 allD$trtYears <- as.factor(ifelse(allD$Treatment=="No_shrub",
                        as.numeric(as.character(allD$year))-2010,0))
-m3 <- inla(logarea.t1 ~ trtYears + logarea.t0  + W.ARTR + W.HECO + W.POSE + W.PSSP + W.allcov + W.allpts + inARTR +
+m3 <- inla(logarea.t1 ~ trtYears + logarea.t0  + W.ARTR + W.HECO + W.POSE + W.PSSP + W.allcov + W.allpts  +
   f(yearID, model="iid", prior="normal",param=c(0,0.001))+
 #  f(GroupID, model="iid", prior="normal",param=c(0,0.001))+
   f(year, logarea.t0, model="iid", prior="normal",param=c(0,0.001)), data=allD,

### 2018-01-17 removals/growth/PSSPgrowth.R

@@ -112,7 +112,7 @@ rm(m2)
 # does effect diminish with time?
 allD$trtYears <- as.factor(ifelse(allD$Treatment=="No_shrub",
                        as.numeric(as.character(allD$year))-2010,0))
-m3 <- inla(logarea.t1 ~ trtYears + logarea.t0  + W.ARTR + W.HECO + W.POSE + W.PSSP + W.allcov + W.allpts + inARTR +
+m3 <- inla(logarea.t1 ~ trtYears + logarea.t0  + W.ARTR + W.HECO + W.POSE + W.PSSP + W.allcov + W.allpts  +
   f(yearID, model="iid", prior="normal",param=c(0,0.001))+
 #  f(GroupID, model="iid", prior="normal",param=c(0,0.001))+
   f(year, logarea.t0, model="iid", prior="normal",param=c(0,0.001)), data=allD,
@@ -126,9 +126,8 @@ cat(capture.output(print(xtable(m3$summary.fixed,digits=4,caption=paste("Summary
         label=paste0(iSpp,"growth-trtYears")),caption.placement="top")),file=statsOutput,sep="\n",append=T)
 rm(m3)
 
-# show why the removals only give us inference about the intercept
-# fit a model without ARTR crowding, then plot residuals against ARTR crowding
-# look at residuals vs marginal W.ARTR effects
+# Show why the removals only give us inference about the intercept:
+# Fit a model without ARTR crowding, then plot residuals against ARTR crowding
 library(lme4)  # use lmer for convenience
 m0 <- lmer(logarea.t1~logarea.t0+ W.HECO + W.POSE + W.PSSP + W.allcov + W.allpts+
              (1|Group)+(logarea.t0|year),data=allD)

### 2018-01-17 removals/growth/growth_residuals_fig.R

@@ -1,0 +1,42 @@
+
+# call this script from removal_analysis_wrapper.r
+
+if(exists("growth_residuals")==F) stop("You need to fit growth models to get residuals")
+
+spp_names <- c("Artemisia tripartita","Hesperostipa comata","Poa secunda","Pseudoroegneria spicata")
+
+# get pretreatment cover
+covD <- read.csv("C:\\Repos\\ExperimentTests\\removals\\QuadYearCover.csv")
+
+# shrub cover
+shrubD <- subset(covD, species=="Artemisia tripartita" & Treatment=="No_shrub" & year==2011)
+shrubD <- shrubD[,c("quad","cover")]
+
+# grass cover
+grassD <- subset(covD,species!="Artemisia tripartita" & Treatment=="No_grass" & year==2011)
+grassD <- aggregate(grassD$cover,by=list(quad=grassD$quad),FUN="sum")
+names(grassD) <- c("quad","cover")
+
+#merge in cover data
+for(i in 1:4){
+  if(i==1){
+    growth_residuals[[i]] <- merge(growth_residuals[[i]],grassD,all.x=T)
+  }else{
+    growth_residuals[[i]] <- merge(growth_residuals[[i]],shrubD,all.x=T)
+  }
+}
+
+# plot figure
+par(mfrow=c(2,2),tcl=-0.2,mgp=c(2,0.5,0),mar=c(3,3,2,1),oma=c(2,2,0,0))
+for(i in 1:4){
+  if(i==1){
+    keep <- which(growth_residuals[[i]]$Treatment=="No_grass")
+  }else{
+    keep <- which(growth_residuals[[i]]$Treatment=="No_shrub")
+  }
+  plot(growth_residuals[[i]]$cover[keep],growth_residuals[[i]]$resids[keep],xlab="",ylab="")
+  abline(h=0,lty="dashed")
+  mtext(side=3,spp_names[i],line=0.5,adj=0,font=3)
+}
+mtext(side=1, "Pre-treatment cover of removal species",outer=T,cex=1.1)
+mtext(side=2, "Growth residuals",outer=T,cex=1.1)

### 2018-01-18 removals/manuscript/removal_ms.tex
Removed all red highlights from manuscript

@@
 
-\new{Stable coexistence requires intraspecific limitations to be stronger than interspecific limitations.
+Stable coexistence requires intraspecific limitations to be stronger than interspecific limitations.
 The greater the difference between intra- and interspecific limitations, the more stable the coexistence, and the weaker the competitive release any species should experience following removal of competitors.
-We conducted a removal experiment to test whether a previously estimated model, showing surprisingly weak interspecific competition for four dominant species in a sagebrush steppe, accurately predicts competitive release.} Our treatments were 1) removal of all perennial grasses and 2) removal of the dominant shrub, \textit{Artemisia tripartita}.
-We \new{regressed} survival, growth and recruitment on the locations, sizes, and species identities of neighboring plants, along with an indicator variable for removal treatment.
-If our ``baseline" \new{regression} model, which accounts for local plant-plant interactions, accurately explains the observed responses to removals, then the removal coefficient should be non-significant.
+We conducted a removal experiment to test whether a previously estimated model, showing surprisingly weak interspecific competition for four dominant species in a sagebrush steppe, accurately predicts competitive release.
+Our treatments were 1) removal of all perennial grasses and 2) removal of the dominant shrub, \textit{Artemisia tripartita}.
+We regressed survival, growth and recruitment on the locations, sizes, and species identities of neighboring plants, along with an indicator variable for removal treatment.
+If our ``baseline" regression model, which accounts for local plant-plant interactions, accurately explains the observed responses to removals, then the removal coefficient should be non-significant.
 For survival and recruitment, the removal coefficients were never significantly positive, indicating that the baseline model did not underestimate interspecific competition.
@@
 The ability of models based on observational data to predict population responses to experimental removals increased our confidence that interspecific interactions are in fact weak for at least three of our four study species.
-The implication is that factors besides interspecific \new{feedbacks} play important roles in determining the structure and dynamics of this community.
+The implication is that factors besides interspecific feedbacks play important roles in determining the structure and dynamics of this community.
 The success of the models also supports the use of multispecies models based on long-term observational data for characterizing competitive interactions.
@@
 
-Stable coexistence \new{of competitors} requires that intraspecific interactions must limit populations more than interspecific interactions \citep{chesson_mechanisms_2000}.
-\new{When studying} populations of species that appear to be coexisting stably, such as populations that have co-occurred in close proximity for extended periods, \new{one} should expect intraspecific limitations to be stronger than interspecific limitations.
-However, interspecific interactions \new{might still} be strong and negative.
+Stable coexistence of competitors requires that intraspecific interactions must limit populations more than interspecific interactions \citep{chesson_mechanisms_2000}.
+When studying populations of species that appear to be coexisting stably, such as populations that have co-occurred in close proximity for extended periods, one should expect intraspecific limitations to be stronger than interspecific limitations.
+However, interspecific interactions might still be strong and negative.
 In fact, plant community ecologists often motivate research on coexistence by noting that all plant species compete for water, light, and a few mineral resources (e.g.
@@
 In some cases, competitive release was complete: remaining species fully compensated for removed biomass \citep{leps_nutrient_1999,jutila_effects_2002}.
-In other cases, \new{some competitive release occurred but the degree of compensation} depended on the functional group or dominance of the removed and remaining species \citep{sala_resource_1989,belsky_effects_1992,smith_dominant_2003}.
+In other cases, some competitive release occurred but the degree of compensation depended on the functional group or dominance of the removed and remaining species \citep{sala_resource_1989,belsky_effects_1992,smith_dominant_2003}.
 Finally, even negative effects of removals have been observed \citep{keddy_effects_1989,gilbert_dominant_2009}, consistent with the recent facilitation literature.
@@
 
-\new{The idea that results from observational studies of competition represent hypotheses which should be tested experimentally is not new \citep{Hairston1980}, but such tests remain extremely rare.
+The idea that results from observational studies of competition represent hypotheses which should be tested experimentally is not new \citep{Hairston1980}, but such tests remain extremely rare.
 In fact, independent tests of any kind of quantitative prediction are rare in ecology, spurring recent calls for a greater emphasis on prediction \citep{Mouquet2015,Petchey2015,Houlahan2017}.
-These authors argue that testing predictions will not only advance applied goals such as forecasting ecological impacts of global change, but will also accelerate progress in basic research.}
-
-\new{In the spirit of this argument,} we conducted a removal experiment to test whether our sagebrush steppe community model, which is based on observational data, underestimates the strength of interspecific competition and competitive release.
+These authors argue that testing predictions will not only advance applied goals such as forecasting ecological impacts of global change, but will also accelerate progress in basic research.
+
+In the spirit of this argument, we conducted a removal experiment to test whether our sagebrush steppe community model, which is based on observational data, underestimates the strength of interspecific competition and competitive release.
 We studied the responses of three perennial grass species to removal of sagebrush, and the response of sagebrush to removal of all grasses.
-The primary purpose of the experiment was not to directly measure competitive release, which is difficult to do over a short time period in a semiarid ecosystem with low and variable plant cover, but rather \new{to test whether our model accurately characterizes} the interspecific interactions that determine the magnitude of competitive release.
+The primary purpose of the experiment was not to directly measure competitive release, which is difficult to do over a short time period in a semiarid ecosystem with low and variable plant cover, but rather to test whether our model accurately characterizes the interspecific interactions that determine the magnitude of competitive release.
 We addressed two questions: First, are the survival, growth or recruitment of our target species significantly different in control vs.
@@
 Because our demographic models include the effects of neighborhood crowding, they should be able to predict how experimental removal of neighbors alters demographic rates.
-If the full effect of experimental removals is stronger than predicted, it means that our statistical models underestimate \new{the response to decreased crowding}.
+If the full effect of experimental removals is stronger than predicted, it means that our statistical models underestimate the response to decreased crowding.
 Our second question integrates across the full life cycle: how do short and long-term population-level responses to removals compare to projections from multispecies population models based on observational data?
@@
 
-\new{Note that the ``baseline" and ``removal" models are not estimated separately; we estimate all parameters together, but when we generate predictions from the baseline model, 
+Note that the ``baseline" and ``removal" models are not estimated separately; we estimate all parameters together, but when we generate predictions from the baseline model, 
 we simply set the removal treatment effects to zero. We chose this over an alternative approach of fitting a model only to data from control plots and then making 
@@
 As a result, if, for example, one of the grasses responds more positively to removal of \textit{A. tripartita} than our baseline model predicts, we cannot 
-determine whether this reflects a poorly estimated competition coefficient, or some other effect of the removal treatment unrelated to competitor densities.}
+determine whether this reflects a poorly estimated competition coefficient, or some other effect of the removal treatment unrelated to competitor densities.
 
@@
 
-\new{Although all the parameters in our model are estimated with uncertainty, we chose not to propage parameter uncertainty through these 
+Although all the parameters in our model are estimated with uncertainty, we chose not to propage parameter uncertainty through these 
 simulations because we wanted to maximize our ability to detect the population-level consequences of a small removal effect. 
@@
 Specifically, we used the recruitment regression to calculate the expected number of new recruits at time $t+1$, which we then multiplied 
-by the observed mean size of recruits for the focal species to get expected cover of new recruits. }
+by the observed mean size of recruits for the focal species to get expected cover of new recruits.
 
@@
 We simulated the IPM to estimate equilibrium abundances under three different scenarios.
-\new{In all of these simulations, we set the spatial random effects to zero, meaning that we are simulating dynamics on a hypothetical, average site.} To incorporate temporal environmental variation, at each time step we randomly selected one of the 30 sets of random year effects for the survival, growth and recruitment models (``kernel selection'' in the terminology of \citealt{metcalf_statistical_2015}).
+In all of these simulations, we set the spatial random effects to zero, meaning that we are simulating dynamics on a hypothetical, average site.
+To incorporate temporal environmental variation, at each time step we randomly selected one of the 30 sets of random year effects for the survival, growth and recruitment models (``kernel selection'' in the terminology of \citealt{metcalf_statistical_2015}).
 For each scenario, we initialized all species with very low cover and ran the model for 500 time steps (years) to allow the community to reach its equilibrium (i.e., its steady-state pattern of fluctuations in response to environmental variability).
-We calculated the \new{average steady-state cover for each species by simulating an additional 2000 time steps and averaging each species' cover over this period.} The first of the three scenarios featured all four species and our baseline model for control plots (no removal effects).
+We calculated the average steady-state cover for each species by simulating an additional 2000 time steps and averaging each species' cover over this period.
+The first of the three scenarios featured all four species and our baseline model for control plots (no removal effects).
 For the second scenario, we simulated our experimental removals by setting either the cover of \textit{A.
@@
 If removal is releasing grasses from \textit{A.
-tripartita} competition \new{for light or soil resources}, we would expect individuals located under the shrub canopies to experience a greater effect of removal than individuals further away.
+tripartita} competition for light or soil resources, we would expect individuals located under the shrub canopies to experience a greater effect of removal than individuals further away.
 However, the location of individuals with respect to \textit{A.
@@
 
-Third, responses to removals might be poorly predicted by models based on observational data if the removals reduce plant densities (crowding) below the range of historical variation, or if the removals \new{result in combinations of neighborhood-scale densities that deviate from naturally occurring patterns of correlation}.
+Third, responses to removals might be poorly predicted by models based on observational data if the removals reduce plant densities (crowding) below the range of historical variation, or if the removals result in combinations of neighborhood-scale densities that deviate from naturally occurring patterns of correlation.
 However, the distribution of crowding observed within control plots easily spans the crowding values observed in removal plots (Fig.
@@
 Red lines and symbols show one-step-ahead predictions from an IBM that includes removal treatment coefficients.
-\new{For \textit{H.
-comata}, lines showing predictions for the baseline and removal effect models are virtually indistinguishable.} Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r}.
+For \textit{H.
+comata}, lines showing predictions for the baseline and removal effect models are virtually indistinguishable.
+Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r}.
 }
@@
 data, we were suspicious of its ability to accurately describe the outcome of interspecific interactions, so we conducted a removal experiment to test it. 
-\new{Our approach represents a strong test of a surprising prediction for two reasons. First, the prediction was already published; the experiment 
+Our approach represents a strong test of a surprising prediction for two reasons. First, the prediction was already published; the experiment 
 gave us a very real opportunity to be wrong in public. Second, testing an observationally-based model with an experimental manipulation, 
-not just additional observations, further increases the potential to confront the model with a novel situation outside the historical range of variation.}
+not just additional observations, further increases the potential to confront the model with a novel situation outside the historical range of variation.
 The results provide substantial support for our model's predictions, and for the weakness of interspecific interactions in this community.
@@
 removal is releasing \textit{P. spicata} from competition, we should expect \textit{P. spicata} individuals that were located beneath \textit{A. tripartita} 
-shrubs that we removed to respond more than individuals located away from any shrubs. Yet we found no evidence for such a difference. \new{Perhaps 
-\textit{A. tripartita's} roots extend far enough beyond its canopy to compete for resources with grasses located in interspaces.}
+shrubs that we removed to respond more than individuals located away from any shrubs. Yet we found no evidence for such a difference. Perhaps 
+\textit{A. tripartita's} roots extend far enough beyond its canopy to compete for resources with grasses located in interspaces.
 An alternative explanation involves pre-treatment differences in composition between control plots and  the \textit{A. tripartita} 
@@
 a nonlinear fashion, with little increase in production at sites with low initial sagebrush cover, 
-and accelerating responses as initial sagebrush cover passes some threshold \citep{McDaniel2005}, \new{implying a 
+and accelerating responses as initial sagebrush cover passes some threshold \citep{McDaniel2005}, implying a 
 shift in the nature of interspecific interactions, perhaps due to changes in the balance of 
-competition and facilitation or in resource supply.} The cover of 
+competition and facilitation or in resource supply. The cover of 
 sagebrush in our ungrazed quadrats may have fallen below that threshold.
 
-\new{What general lessons can we draw about testing predictions from population models? 
+What general lessons can we draw about testing predictions from population models? 
 First, partial removals, rather than complete removal treatments, could 
@@
 performed quite well, after accounting for year-to-year environmental variation through the inclusion of
-random year effects.}
+random year effects.
 
@@
 spicata}.
-\new{Each point compares the crowding experienced by one \textit{P.
-spicata} individual in one year from the species on the x-axis to the species on the y-axis.}   Blue symbols show values from controls plots, red symbols show \new{pre-treatment} values from \textit{Artemisia} removal plots.``W.allcov'' refers to the aggregated crowding by all shrubs and perennial grasses beyond the focal dominant species, and ``W.allpts'' refers to the aggregated crowding by forb species.
+Each point compares the crowding experienced by one \textit{P.
+spicata} individual in one year from the species on the x-axis to the species on the y-axis.
+Blue symbols show values from controls plots, red symbols show pre-treatment values from \textit{Artemisia} removal plots.``W.allcov'' refers to the aggregated crowding by all shrubs and perennial grasses beyond the focal dominant species, and ``W.allpts'' refers to the aggregated crowding by forb species.
 \texttt{Wdistrib\textbackslash exploreSurvivalWs.r}.

### 2018-01-22 removals/manuscript/removal_ms.tex
Complete set of new results with corrected demography data

@@
   \hline
-(Intercept) & -0.0207 & 0.2261 & -0.4554 & -0.0246 & 0.4367 & -0.0321 & 0.0000 \\ 
-  logarea & 0.7208 & 0.0575 & 0.6109 & 0.7195 & 0.8378 & 0.7171 & 0.0000 \\ 
-  TreatmentNo\_grass & -1.3218 & 0.7343 & -2.7293 & -1.3340 & 0.1563 & -1.3591 & 0.0000 \\ 
-  W.ARTR & -1.5214 & 0.2075 & -1.9391 & -1.5179 & -1.1237 & -1.5107 & 0.0000 \\ 
-  W.HECO & -0.0694 & 0.0423 & -0.1485 & -0.0708 & 0.0178 & -0.0738 & 0.0000 \\ 
-  W.POSE & -0.0049 & 0.0761 & -0.1469 & -0.0077 & 0.1525 & -0.0134 & 0.0000 \\ 
-  W.PSSP & 0.0281 & 0.0275 & -0.0235 & 0.0271 & 0.0849 & 0.0252 & 0.0000 \\ 
-  W.allcov & -0.0055 & 0.0083 & -0.0219 & -0.0054 & 0.0109 & -0.0054 & 0.0000 \\ 
-  W.allpts & 0.1694 & 0.1579 & -0.1386 & 0.1686 & 0.4812 & 0.1672 & 0.0000 \\ 
+(Intercept) & -0.2613 & 0.1344 & -0.5276 & -0.2605 & 0.0001 & -0.2588 & 0.0000 \\ 
+  logarea & 0.7093 & 0.0400 & 0.6315 & 0.7090 & 0.7885 & 0.7085 & 0.0000 \\ 
+  TreatmentNo\_grass & -0.9994 & 0.6453 & -2.2156 & -1.0182 & 0.3228 & -1.0565 & 0.0000 \\ 
+  W.ARTR & -2.8522 & 0.4187 & -3.6968 & -2.8443 & -2.0513 & -2.8284 & 0.0000 \\ 
+  W.HECO & -0.0720 & 0.0615 & -0.1856 & -0.0747 & 0.0565 & -0.0804 & 0.0000 \\ 
+  W.POSE & 0.0617 & 0.0895 & -0.1061 & 0.0588 & 0.2459 & 0.0528 & 0.0000 \\ 
+  W.PSSP & 0.0387 & 0.0475 & -0.0500 & 0.0370 & 0.1369 & 0.0335 & 0.0000 \\ 
+  W.allcov & -0.0065 & 0.0109 & -0.0282 & -0.0065 & 0.0148 & -0.0063 & 0.0000 \\ 
+  W.allpts & 0.1293 & 0.1653 & -0.1956 & 0.1294 & 0.4533 & 0.1296 & 0.0000 \\ 
    \hline
@@
   \hline
-(Intercept) & 1.4189 & 0.2115 & 0.9991 & 1.4190 & 1.8371 & 1.4192 & 0.0000 \\ 
-  logarea & 1.2885 & 0.0813 & 1.1399 & 1.2843 & 1.4601 & 1.2758 & 0.0000 \\ 
-  TreatmentNo\_shrub & 0.3958 & 0.4175 & -0.4193 & 0.3942 & 1.2190 & 0.3909 & 0.0000 \\ 
-  W.ARTR & -0.0064 & 0.0028 & -0.0120 & -0.0064 & -0.0009 & -0.0064 & 0.0000 \\ 
-  W.HECO & -0.6647 & 0.0673 & -0.7996 & -0.6638 & -0.5351 & -0.6619 & 0.0000 \\ 
-  W.POSE & 0.0732 & 0.0534 & -0.0304 & 0.0728 & 0.1790 & 0.0720 & 0.0000 \\ 
-  W.PSSP & 0.0190 & 0.0184 & -0.0167 & 0.0188 & 0.0554 & 0.0185 & 0.0000 \\ 
-  W.allcov & 0.0021 & 0.0047 & -0.0072 & 0.0021 & 0.0113 & 0.0021 & 0.0000 \\ 
-  W.allpts & -0.1185 & 0.0919 & -0.2998 & -0.1183 & 0.0610 & -0.1177 & 0.0000 \\ 
+(Intercept) & 1.4753 & 0.2011 & 1.0764 & 1.4753 & 1.8736 & 1.4753 & 0.0000 \\ 
+  logarea & 1.1991 & 0.0747 & 1.0622 & 1.1954 & 1.3561 & 1.1874 & 0.0000 \\ 
+  TreatmentNo\_shrub & 0.2915 & 0.3900 & -0.4693 & 0.2898 & 1.0610 & 0.2864 & 0.0000 \\ 
+  W.ARTR & -0.0063 & 0.0030 & -0.0121 & -0.0063 & -0.0005 & -0.0063 & 0.0000 \\ 
+  W.HECO & -1.6962 & 0.1458 & -1.9894 & -1.6939 & -1.4163 & -1.6891 & 0.0000 \\ 
+  W.POSE & 0.0649 & 0.0563 & -0.0442 & 0.0644 & 0.1768 & 0.0634 & 0.0000 \\ 
+  W.PSSP & 0.0313 & 0.0360 & -0.0381 & 0.0309 & 0.1031 & 0.0301 & 0.0000 \\ 
+  W.allcov & -0.0012 & 0.0062 & -0.0135 & -0.0012 & 0.0110 & -0.0012 & 0.0000 \\ 
+  W.allpts & -0.1295 & 0.1004 & -0.3273 & -0.1293 & 0.0670 & -0.1290 & 0.0000 \\ 
    \hline
@@
   \hline
-(Intercept) & 1.3162 & 0.1965 & 0.9242 & 1.3171 & 1.7024 & 1.3186 & 0.0000 \\ 
-  logarea & 1.0585 & 0.0640 & 0.9365 & 1.0569 & 1.1902 & 1.0537 & 0.0000 \\ 
-  TreatmentNo\_shrub & -0.2572 & 0.1723 & -0.5949 & -0.2573 & 0.0811 & -0.2576 & 0.0000 \\ 
-  W.ARTR & 0.0001 & 0.0019 & -0.0036 & 0.0001 & 0.0038 & 0.0001 & 0.0000 \\ 
-  W.HECO & -0.0153 & 0.0138 & -0.0424 & -0.0154 & 0.0119 & -0.0154 & 0.0000 \\ 
-  W.POSE & -1.2632 & 0.0816 & -1.4260 & -1.2623 & -1.1053 & -1.2605 & 0.0000 \\ 
-  W.PSSP & 0.0249 & 0.0122 & 0.0011 & 0.0249 & 0.0490 & 0.0247 & 0.0000 \\ 
-  W.allcov & -0.0024 & 0.0026 & -0.0075 & -0.0024 & 0.0029 & -0.0024 & 0.0000 \\ 
-  W.allpts & 0.0018 & 0.0524 & -0.1009 & 0.0018 & 0.1048 & 0.0017 & 0.0000 \\ 
+(Intercept) & 1.4020 & 0.1922 & 1.0176 & 1.4031 & 1.7793 & 1.4050 & 0.0000 \\ 
+  logarea & 1.0362 & 0.0645 & 0.9139 & 1.0343 & 1.1696 & 1.0307 & 0.0000 \\ 
+  TreatmentNo\_shrub & -0.1604 & 0.1622 & -0.4784 & -0.1606 & 0.1583 & -0.1610 & 0.0000 \\ 
+  W.ARTR & 0.0000 & 0.0020 & -0.0038 & 0.0000 & 0.0039 & -0.0000 & 0.0000 \\ 
+  W.HECO & -0.0224 & 0.0208 & -0.0631 & -0.0225 & 0.0185 & -0.0227 & 0.0000 \\ 
+  W.POSE & -2.0073 & 0.1112 & -2.2290 & -2.0062 & -1.7921 & -2.0038 & 0.0000 \\ 
+  W.PSSP & 0.0468 & 0.0238 & 0.0004 & 0.0466 & 0.0939 & 0.0463 & 0.0000 \\ 
+  W.allcov & -0.0061 & 0.0035 & -0.0130 & -0.0061 & 0.0008 & -0.0061 & 0.0000 \\ 
+  W.allpts & -0.0383 & 0.0567 & -0.1494 & -0.0384 & 0.0730 & -0.0385 & 0.0000 \\ 
    \hline
@@
   \hline
-(Intercept) & 1.2121 & 0.1699 & 0.8815 & 1.2103 & 1.5533 & 1.2066 & 0.0000 \\ 
-  logarea & 1.5492 & 0.0976 & 1.3612 & 1.5474 & 1.7473 & 1.5437 & 0.0000 \\ 
-  TreatmentNo\_shrub & -0.1885 & 0.2208 & -0.6237 & -0.1879 & 0.2431 & -0.1867 & 0.0000 \\ 
-  W.ARTR & 0.0101 & 0.0021 & 0.0061 & 0.0101 & 0.0142 & 0.0101 & 0.0000 \\ 
-  W.HECO & 0.0046 & 0.0185 & -0.0319 & 0.0046 & 0.0406 & 0.0047 & 0.0000 \\ 
-  W.POSE & 0.0245 & 0.0384 & -0.0502 & 0.0242 & 0.1007 & 0.0236 & 0.0000 \\ 
-  W.PSSP & -0.4430 & 0.0291 & -0.5010 & -0.4427 & -0.3867 & -0.4420 & 0.0000 \\ 
-  W.allcov & 0.0100 & 0.0028 & 0.0046 & 0.0100 & 0.0155 & 0.0100 & 0.0000 \\ 
-  W.allpts & 0.0943 & 0.0605 & -0.0246 & 0.0944 & 0.2129 & 0.0944 & 0.0000 \\ 
+(Intercept) & 1.1978 & 0.1639 & 0.8783 & 1.1962 & 1.5257 & 1.1933 & 0.0000 \\ 
+  logarea & 1.5138 & 0.0935 & 1.3336 & 1.5122 & 1.7029 & 1.5092 & 0.0000 \\ 
+  TreatmentNo\_shrub & -0.2466 & 0.2023 & -0.6454 & -0.2460 & 0.1489 & -0.2449 & 0.0000 \\ 
+  W.ARTR & 0.0099 & 0.0022 & 0.0057 & 0.0099 & 0.0142 & 0.0099 & 0.0000 \\ 
+  W.HECO & 0.0025 & 0.0289 & -0.0550 & 0.0028 & 0.0586 & 0.0032 & 0.0000 \\ 
+  W.POSE & 0.0517 & 0.0389 & -0.0222 & 0.0508 & 0.1308 & 0.0490 & 0.0000 \\ 
+  W.PSSP & -1.1267 & 0.0739 & -1.2743 & -1.1258 & -0.9838 & -1.1242 & 0.0000 \\ 
+  W.allcov & 0.0131 & 0.0037 & 0.0060 & 0.0131 & 0.0203 & 0.0131 & 0.0000 \\ 
+  W.allpts & 0.0756 & 0.0658 & -0.0540 & 0.0757 & 0.2045 & 0.0760 & 0.0000 \\ 
    \hline
@@
   \hline
-(Intercept) & 0.7378 & 0.2437 & 0.2591 & 0.7378 & 1.2161 & 0.7377 & 0.0000 \\ 
-  logarea.t0 & 0.8663 & 0.0393 & 0.7892 & 0.8663 & 0.9434 & 0.8663 & 0.0000 \\ 
-  TreatmentNo\_grass & 0.1037 & 0.1436 & -0.1783 & 0.1037 & 0.3854 & 0.1037 & 0.0000 \\ 
-  W.ARTR & -0.0364 & 0.0892 & -0.2116 & -0.0364 & 0.1387 & -0.0364 & 0.0000 \\ 
-  W.HECO & 0.0033 & 0.0136 & -0.0233 & 0.0033 & 0.0299 & 0.0033 & 0.0000 \\ 
-  W.POSE & -0.0573 & 0.0268 & -0.1100 & -0.0573 & -0.0047 & -0.0573 & 0.0000 \\ 
-  W.PSSP & 0.0062 & 0.0084 & -0.0102 & 0.0062 & 0.0226 & 0.0062 & 0.0000 \\ 
-  W.allcov & -0.0055 & 0.0030 & -0.0114 & -0.0055 & 0.0004 & -0.0055 & 0.0000 \\ 
-  W.allpts & 0.0029 & 0.0477 & -0.0908 & 0.0029 & 0.0965 & 0.0029 & 0.0000 \\ 
+(Intercept) & 0.7092 & 0.2385 & 0.2396 & 0.7091 & 1.1784 & 0.7090 & 0.0000 \\ 
+  logarea.t0 & 0.8689 & 0.0394 & 0.7912 & 0.8689 & 0.9466 & 0.8689 & 0.0000 \\ 
+  TreatmentNo\_grass & 0.1385 & 0.1446 & -0.1453 & 0.1385 & 0.4221 & 0.1385 & 0.0000 \\ 
+  W.ARTR & -0.1497 & 0.1960 & -0.5345 & -0.1497 & 0.2347 & -0.1497 & 0.0000 \\ 
+  W.HECO & 0.0045 & 0.0198 & -0.0343 & 0.0045 & 0.0433 & 0.0045 & 0.0000 \\ 
+  W.POSE & -0.0614 & 0.0308 & -0.1218 & -0.0614 & -0.0010 & -0.0614 & 0.0000 \\ 
+  W.PSSP & -0.0050 & 0.0158 & -0.0360 & -0.0050 & 0.0260 & -0.0050 & 0.0000 \\ 
+  W.allcov & -0.0052 & 0.0041 & -0.0133 & -0.0052 & 0.0029 & -0.0052 & 0.0000 \\ 
+  W.allpts & 0.0115 & 0.0536 & -0.0938 & 0.0115 & 0.1167 & 0.0115 & 0.0000 \\ 
    \hline
@@
   \hline
-(Intercept) & 0.3782 & 0.0797 & 0.2212 & 0.3783 & 0.5343 & 0.3784 & 0.0000 \\ 
-  logarea.t0 & 0.8272 & 0.0205 & 0.7869 & 0.8272 & 0.8674 & 0.8272 & 0.0000 \\ 
-  TreatmentNo\_shrub & -0.0162 & 0.1336 & -0.2785 & -0.0162 & 0.2460 & -0.0162 & 0.0000 \\ 
-  W.ARTR & -0.0031 & 0.0010 & -0.0051 & -0.0031 & -0.0010 & -0.0031 & 0.0000 \\ 
-  W.HECO & -0.0479 & 0.0195 & -0.0862 & -0.0479 & -0.0098 & -0.0479 & 0.0000 \\ 
-  W.POSE & 0.0242 & 0.0162 & -0.0077 & 0.0242 & 0.0560 & 0.0242 & 0.0000 \\ 
-  W.PSSP & -0.0151 & 0.0065 & -0.0279 & -0.0151 & -0.0023 & -0.0151 & 0.0000 \\ 
-  W.allcov & -0.0038 & 0.0019 & -0.0076 & -0.0038 & 0.0000 & -0.0038 & 0.0000 \\ 
-  W.allpts & -0.0513 & 0.0380 & -0.1258 & -0.0513 & 0.0232 & -0.0513 & 0.0000 \\ 
+(Intercept) & 0.3937 & 0.0815 & 0.2330 & 0.3938 & 0.5537 & 0.3940 & 0.0000 \\ 
+  logarea.t0 & 0.8207 & 0.0213 & 0.7788 & 0.8207 & 0.8626 & 0.8207 & 0.0000 \\ 
+  TreatmentNo\_shrub & 0.0691 & 0.1314 & -0.1891 & 0.0691 & 0.3268 & 0.0691 & 0.0000 \\ 
+  W.ARTR & -0.0042 & 0.0011 & -0.0063 & -0.0042 & -0.0020 & -0.0042 & 0.0000 \\ 
+  W.HECO & -0.2134 & 0.0618 & -0.3348 & -0.2134 & -0.0922 & -0.2135 & 0.0000 \\ 
+  W.POSE & 0.0204 & 0.0178 & -0.0146 & 0.0204 & 0.0553 & 0.0204 & 0.0000 \\ 
+  W.PSSP & -0.0373 & 0.0131 & -0.0630 & -0.0373 & -0.0117 & -0.0373 & 0.0000 \\ 
+  W.allcov & -0.0062 & 0.0027 & -0.0115 & -0.0062 & -0.0010 & -0.0062 & 0.0000 \\ 
+  W.allpts & -0.0212 & 0.0416 & -0.1028 & -0.0212 & 0.0604 & -0.0212 & 0.0000 \\ 
    \hline
@@
 \centering
-\caption{Summary of posterior distributions for fixed effects for the \textit{P. secunda} growth model. See Table \ref{ARTRgrowth} for an explanation of coefficient names
+\caption{Summary of posterior distributions for fixed effects for the \textit{Poa secunda} growth model. See Table \ref{ARTRgrowth} for an explanation of coefficient names
 and column headers.} 
@@
   \hline
-(Intercept) & 0.4923 & 0.0681 & 0.3581 & 0.4923 & 0.6259 & 0.4924 & 0.0000 \\ 
-  logarea.t0 & 0.6750 & 0.0232 & 0.6292 & 0.6750 & 0.7203 & 0.6751 & 0.0000 \\ 
-  TreatmentNo\_shrub & 0.2161 & 0.0644 & 0.0897 & 0.2161 & 0.3425 & 0.2161 & 0.0000 \\ 
-  W.ARTR & -0.0002 & 0.0008 & -0.0019 & -0.0002 & 0.0014 & -0.0002 & 0.0000 \\ 
-  W.HECO & 0.0039 & 0.0063 & -0.0084 & 0.0039 & 0.0161 & 0.0039 & 0.0000 \\ 
-  W.POSE & -0.2641 & 0.0387 & -0.3400 & -0.2641 & -0.1883 & -0.2641 & 0.0000 \\ 
-  W.PSSP & -0.0073 & 0.0056 & -0.0183 & -0.0073 & 0.0038 & -0.0073 & 0.0000 \\ 
-  W.allcov & -0.0003 & 0.0012 & -0.0026 & -0.0003 & 0.0020 & -0.0003 & 0.0000 \\ 
-  W.allpts & -0.0254 & 0.0231 & -0.0709 & -0.0254 & 0.0200 & -0.0254 & 0.0000 \\ 
+(Intercept) & 0.5069 & 0.0625 & 0.3830 & 0.5070 & 0.6299 & 0.5072 & 0.0000 \\ 
+  logarea.t0 & 0.6669 & 0.0228 & 0.6215 & 0.6670 & 0.7116 & 0.6673 & 0.0000 \\ 
+  TreatmentNo\_shrub & 0.1937 & 0.0632 & 0.0696 & 0.1937 & 0.3176 & 0.1937 & 0.0000 \\ 
+  W.ARTR & -0.0004 & 0.0009 & -0.0022 & -0.0004 & 0.0013 & -0.0004 & 0.0000 \\ 
+  W.HECO & 0.0110 & 0.0100 & -0.0087 & 0.0110 & 0.0307 & 0.0110 & 0.0000 \\ 
+  W.POSE & -0.5753 & 0.0623 & -0.6977 & -0.5753 & -0.4530 & -0.5753 & 0.0000 \\ 
+  W.PSSP & -0.0093 & 0.0114 & -0.0316 & -0.0093 & 0.0130 & -0.0093 & 0.0000 \\ 
+  W.allcov & -0.0004 & 0.0016 & -0.0035 & -0.0004 & 0.0028 & -0.0004 & 0.0000 \\ 
+  W.allpts & -0.0340 & 0.0255 & -0.0842 & -0.0340 & 0.0161 & -0.0340 & 0.0000 \\ 
    \hline
@@
   \hline
-(Intercept) & 0.3987 & 0.0660 & 0.2690 & 0.3987 & 0.5281 & 0.3988 & 0.0000 \\ 
-  logarea.t0 & 0.8249 & 0.0149 & 0.7956 & 0.8249 & 0.8543 & 0.8249 & 0.0000 \\ 
-  TreatmentNo\_shrub & 0.1948 & 0.0714 & 0.0546 & 0.1948 & 0.3349 & 0.1948 & 0.0000 \\ 
-  W.ARTR & -0.0025 & 0.0008 & -0.0041 & -0.0025 & -0.0010 & -0.0025 & 0.0000 \\ 
-  W.HECO & -0.0118 & 0.0072 & -0.0260 & -0.0118 & 0.0023 & -0.0118 & 0.0000 \\ 
-  W.POSE & -0.0067 & 0.0130 & -0.0323 & -0.0067 & 0.0189 & -0.0067 & 0.0000 \\ 
-  W.PSSP & -0.1480 & 0.0133 & -0.1740 & -0.1480 & -0.1220 & -0.1480 & 0.0000 \\ 
-  W.allcov & -0.0036 & 0.0011 & -0.0057 & -0.0036 & -0.0015 & -0.0036 & 0.0000 \\ 
-  W.allpts & -0.0522 & 0.0236 & -0.0984 & -0.0522 & -0.0060 & -0.0522 & 0.0000 \\ 
+(Intercept) & 0.3846 & 0.0710 & 0.2425 & 0.3851 & 0.5236 & 0.3858 & 0.0000 \\ 
+  logarea.t0 & 0.8259 & 0.0167 & 0.7929 & 0.8259 & 0.8587 & 0.8260 & 0.0000 \\ 
+  TreatmentNo\_shrub & 0.2262 & 0.0681 & 0.0924 & 0.2262 & 0.3598 & 0.2262 & 0.0000 \\ 
+  W.ARTR & -0.0023 & 0.0008 & -0.0038 & -0.0023 & -0.0007 & -0.0023 & 0.0000 \\ 
+  W.HECO & -0.0201 & 0.0111 & -0.0419 & -0.0201 & 0.0017 & -0.0201 & 0.0000 \\ 
+  W.POSE & -0.0200 & 0.0137 & -0.0469 & -0.0200 & 0.0069 & -0.0200 & 0.0000 \\ 
+  W.PSSP & -0.3845 & 0.0335 & -0.4502 & -0.3845 & -0.3188 & -0.3846 & 0.0000 \\ 
+  W.allcov & -0.0048 & 0.0014 & -0.0076 & -0.0048 & -0.0021 & -0.0048 & 0.0000 \\ 
+  W.allpts & -0.0510 & 0.0256 & -0.1012 & -0.0510 & -0.0008 & -0.0510 & 0.0000 \\ 
    \hline
@@
   \hline
-(Intercept) & 0.4932 & 0.0622 & 0.3700 & 0.4934 & 0.6154 & 0.4937 & 0.0000 \\ 
-  logarea.t0 & 0.6746 & 0.0220 & 0.6306 & 0.6747 & 0.7177 & 0.6750 & 0.0000 \\ 
-  TreatmentNo\_shrub & 0.2370 & 0.0739 & 0.0919 & 0.2370 & 0.3820 & 0.2370 & 0.0000 \\ 
-  W.ARTR & -0.0002 & 0.0008 & -0.0018 & -0.0002 & 0.0014 & -0.0002 & 0.0000 \\ 
-  W.HECO & 0.0038 & 0.0062 & -0.0083 & 0.0038 & 0.0160 & 0.0038 & 0.0000 \\ 
-  W.POSE & -0.2644 & 0.0383 & -0.3395 & -0.2644 & -0.1893 & -0.2644 & 0.0000 \\ 
-  W.PSSP & -0.0075 & 0.0056 & -0.0184 & -0.0075 & 0.0034 & -0.0075 & 0.0000 \\ 
-  W.allcov & -0.0003 & 0.0012 & -0.0026 & -0.0003 & 0.0020 & -0.0003 & 0.0000 \\ 
-  W.allpts & -0.0258 & 0.0229 & -0.0707 & -0.0258 & 0.0191 & -0.0258 & 0.0000 \\ 
-  inARTR & -0.0592 & 0.1133 & -0.2817 & -0.0592 & 0.1632 & -0.0592 & 0.0000 \\ 
+(Intercept) & 0.5080 & 0.0625 & 0.3841 & 0.5081 & 0.6311 & 0.5084 & 0.0000 \\ 
+  logarea.t0 & 0.6665 & 0.0228 & 0.6211 & 0.6666 & 0.7111 & 0.6669 & 0.0000 \\ 
+  TreatmentNo\_shrub & 0.2247 & 0.0744 & 0.0786 & 0.2247 & 0.3707 & 0.2247 & 0.0000 \\ 
+  W.ARTR & -0.0004 & 0.0009 & -0.0022 & -0.0004 & 0.0013 & -0.0004 & 0.0000 \\ 
+  W.HECO & 0.0109 & 0.0100 & -0.0088 & 0.0109 & 0.0306 & 0.0109 & 0.0000 \\ 
+  W.POSE & -0.5770 & 0.0624 & -0.6995 & -0.5770 & -0.4547 & -0.5770 & 0.0000 \\ 
+  W.PSSP & -0.0093 & 0.0114 & -0.0316 & -0.0093 & 0.0130 & -0.0093 & 0.0000 \\ 
+  W.allcov & -0.0004 & 0.0016 & -0.0036 & -0.0004 & 0.0027 & -0.0004 & 0.0000 \\ 
+  W.allpts & -0.0340 & 0.0256 & -0.0842 & -0.0340 & 0.0161 & -0.0340 & 0.0000 \\ 
+  inARTR & -0.0890 & 0.1126 & -0.3100 & -0.0890 & 0.1319 & -0.0890 & 0.0000 \\ 
    \hline
@@
   \hline
-(Intercept) & 0.4129 & 0.0633 & 0.2885 & 0.4128 & 0.5381 & 0.4124 & 0.0000 \\ 
-  logarea.t0 & 0.8261 & 0.0173 & 0.7919 & 0.8262 & 0.8602 & 0.8262 & 0.0000 \\ 
-  TreatmentNo\_shrub & 0.2444 & 0.0798 & 0.0878 & 0.2444 & 0.4009 & 0.2444 & 0.0000 \\ 
-  W.ARTR & -0.0029 & 0.0008 & -0.0044 & -0.0029 & -0.0015 & -0.0029 & 0.0000 \\ 
-  W.HECO & -0.0138 & 0.0068 & -0.0272 & -0.0138 & -0.0005 & -0.0138 & 0.0000 \\ 
-  W.POSE & -0.0071 & 0.0127 & -0.0319 & -0.0071 & 0.0178 & -0.0071 & 0.0000 \\ 
-  W.PSSP & -0.1372 & 0.0126 & -0.1619 & -0.1372 & -0.1125 & -0.1372 & 0.0000 \\ 
-  W.allcov & -0.0039 & 0.0010 & -0.0059 & -0.0039 & -0.0018 & -0.0039 & 0.0000 \\ 
-  W.allpts & -0.0576 & 0.0224 & -0.1017 & -0.0576 & -0.0136 & -0.0576 & 0.0000 \\ 
-  inARTR & -0.1301 & 0.1102 & -0.3465 & -0.1301 & 0.0862 & -0.1301 & 0.0000 \\ 
+(Intercept) & 0.3955 & 0.0606 & 0.2764 & 0.3953 & 0.5153 & 0.3950 & 0.0000 \\ 
+  logarea.t0 & 0.8276 & 0.0168 & 0.7943 & 0.8277 & 0.8608 & 0.8277 & 0.0000 \\ 
+  TreatmentNo\_shrub & 0.2531 & 0.0784 & 0.0991 & 0.2531 & 0.4070 & 0.2531 & 0.0000 \\ 
+  W.ARTR & -0.0027 & 0.0008 & -0.0042 & -0.0027 & -0.0011 & -0.0027 & 0.0000 \\ 
+  W.HECO & -0.0231 & 0.0110 & -0.0447 & -0.0231 & -0.0016 & -0.0231 & 0.0000 \\ 
+  W.POSE & -0.0198 & 0.0138 & -0.0469 & -0.0198 & 0.0072 & -0.0198 & 0.0000 \\ 
+  W.PSSP & -0.3600 & 0.0326 & -0.4239 & -0.3600 & -0.2961 & -0.3600 & 0.0000 \\ 
+  W.allcov & -0.0051 & 0.0014 & -0.0079 & -0.0051 & -0.0024 & -0.0051 & 0.0000 \\ 
+  W.allpts & -0.0557 & 0.0254 & -0.1055 & -0.0557 & -0.0060 & -0.0557 & 0.0000 \\ 
+  inARTR & -0.0649 & 0.1111 & -0.2830 & -0.0649 & 0.1531 & -0.0649 & 0.0000 \\ 
    \hline
@@
   \hline
-(Intercept) & 0.4963 & 0.0616 & 0.3752 & 0.4963 & 0.6172 & 0.4964 & 0.0000 \\ 
-  trtYears1 & 0.0798 & 0.1272 & -0.1699 & 0.0798 & 0.3293 & 0.0798 & 0.0000 \\ 
-  trtYears2 & 0.2425 & 0.1350 & -0.0225 & 0.2425 & 0.5073 & 0.2425 & 0.0000 \\ 
-  trtYears3 & 0.2512 & 0.1554 & -0.0539 & 0.2512 & 0.5561 & 0.2512 & 0.0000 \\ 
-  trtYears4 & 0.2988 & 0.1693 & -0.0337 & 0.2988 & 0.6310 & 0.2988 & 0.0000 \\ 
-  trtYears5 & 0.4787 & 0.1845 & 0.1164 & 0.4787 & 0.8407 & 0.4787 & 0.0000 \\ 
-  logarea.t0 & 0.6731 & 0.0234 & 0.6270 & 0.6732 & 0.7191 & 0.6732 & 0.0000 \\ 
-  W.ARTR & -0.0003 & 0.0009 & -0.0019 & -0.0003 & 0.0014 & -0.0003 & 0.0000 \\ 
-  W.HECO & 0.0035 & 0.0064 & -0.0090 & 0.0035 & 0.0160 & 0.0035 & 0.0000 \\ 
-  W.POSE & -0.2624 & 0.0394 & -0.3398 & -0.2624 & -0.1851 & -0.2624 & 0.0000 \\ 
-  W.PSSP & -0.0080 & 0.0057 & -0.0192 & -0.0080 & 0.0032 & -0.0080 & 0.0000 \\ 
-  W.allcov & -0.0004 & 0.0012 & -0.0027 & -0.0004 & 0.0020 & -0.0004 & 0.0000 \\ 
-  W.allpts & -0.0267 & 0.0236 & -0.0730 & -0.0267 & 0.0196 & -0.0267 & 0.0000 \\ 
-  inARTR & -0.0563 & 0.1169 & -0.2859 & -0.0563 & 0.1730 & -0.0563 & 0.0000 \\ 
+(Intercept) & 0.5079 & 0.0626 & 0.3839 & 0.5080 & 0.6310 & 0.5082 & 0.0000 \\ 
+  trtYears1 & 0.0910 & 0.1169 & -0.1386 & 0.0910 & 0.3203 & 0.0910 & 0.0000 \\ 
+  trtYears2 & 0.1560 & 0.1212 & -0.0819 & 0.1560 & 0.3937 & 0.1560 & 0.0000 \\ 
+  trtYears3 & 0.1891 & 0.1461 & -0.0976 & 0.1891 & 0.4757 & 0.1891 & 0.0000 \\ 
+  trtYears4 & 0.3125 & 0.1608 & -0.0032 & 0.3125 & 0.6280 & 0.3126 & 0.0000 \\ 
+  trtYears5 & 0.3737 & 0.1782 & 0.0239 & 0.3737 & 0.7233 & 0.3737 & 0.0000 \\ 
+  logarea.t0 & 0.6663 & 0.0228 & 0.6207 & 0.6664 & 0.7111 & 0.6666 & 0.0000 \\ 
+  W.ARTR & -0.0005 & 0.0009 & -0.0022 & -0.0005 & 0.0013 & -0.0005 & 0.0000 \\ 
+  W.HECO & 0.0108 & 0.0100 & -0.0089 & 0.0108 & 0.0305 & 0.0108 & 0.0000 \\ 
+  W.POSE & -0.5725 & 0.0624 & -0.6950 & -0.5725 & -0.4501 & -0.5725 & 0.0000 \\ 
+  W.PSSP & -0.0100 & 0.0114 & -0.0323 & -0.0100 & 0.0123 & -0.0100 & 0.0000 \\ 
+  W.allcov & -0.0004 & 0.0016 & -0.0036 & -0.0004 & 0.0028 & -0.0004 & 0.0000 \\ 
+  W.allpts & -0.0347 & 0.0256 & -0.0849 & -0.0347 & 0.0155 & -0.0347 & 0.0000 \\ 
    \hline
@@
   \hline
-(Intercept) & 0.4122 & 0.0626 & 0.2892 & 0.4120 & 0.5364 & 0.4115 & 0.0000 \\ 
-  trtYears1 & 0.3256 & 0.1440 & 0.0428 & 0.3256 & 0.6081 & 0.3257 & 0.0000 \\ 
-  trtYears2 & -0.0353 & 0.1467 & -0.3235 & -0.0353 & 0.2525 & -0.0353 & 0.0000 \\ 
-  trtYears3 & 0.4829 & 0.1586 & 0.1714 & 0.4829 & 0.7940 & 0.4829 & 0.0000 \\ 
-  trtYears4 & 0.5059 & 0.1602 & 0.1915 & 0.5059 & 0.8202 & 0.5059 & 0.0000 \\ 
-  trtYears5 & -0.0588 & 0.1682 & -0.3890 & -0.0588 & 0.2712 & -0.0588 & 0.0000 \\ 
-  logarea.t0 & 0.8266 & 0.0173 & 0.7922 & 0.8267 & 0.8607 & 0.8268 & 0.0000 \\ 
-  W.ARTR & -0.0030 & 0.0008 & -0.0044 & -0.0030 & -0.0015 & -0.0030 & 0.0000 \\ 
-  W.HECO & -0.0139 & 0.0068 & -0.0272 & -0.0139 & -0.0005 & -0.0139 & 0.0000 \\ 
-  W.POSE & -0.0064 & 0.0127 & -0.0313 & -0.0064 & 0.0185 & -0.0064 & 0.0000 \\ 
-  W.PSSP & -0.1362 & 0.0126 & -0.1609 & -0.1362 & -0.1115 & -0.1362 & 0.0000 \\ 
-  W.allcov & -0.0040 & 0.0010 & -0.0061 & -0.0040 & -0.0020 & -0.0040 & 0.0000 \\ 
-  W.allpts & -0.0579 & 0.0224 & -0.1020 & -0.0579 & -0.0139 & -0.0579 & 0.0000 \\ 
-  inARTR & -0.1235 & 0.1104 & -0.3403 & -0.1235 & 0.0931 & -0.1235 & 0.0000 \\ 
+(Intercept) & 0.3948 & 0.0606 & 0.2757 & 0.3946 & 0.5149 & 0.3943 & 0.0000 \\ 
+  trtYears1 & 0.3434 & 0.1341 & 0.0800 & 0.3434 & 0.6065 & 0.3434 & 0.0000 \\ 
+  trtYears2 & -0.0514 & 0.1391 & -0.3245 & -0.0514 & 0.2215 & -0.0513 & 0.0000 \\ 
+  trtYears3 & 0.3547 & 0.1493 & 0.0616 & 0.3547 & 0.6476 & 0.3547 & 0.0000 \\ 
+  trtYears4 & 0.5487 & 0.1519 & 0.2505 & 0.5487 & 0.8467 & 0.5487 & 0.0000 \\ 
+  trtYears5 & -0.0575 & 0.1598 & -0.3711 & -0.0575 & 0.2560 & -0.0575 & 0.0000 \\ 
+  logarea.t0 & 0.8284 & 0.0168 & 0.7951 & 0.8284 & 0.8614 & 0.8285 & 0.0000 \\ 
+  W.ARTR & -0.0026 & 0.0008 & -0.0042 & -0.0026 & -0.0011 & -0.0026 & 0.0000 \\ 
+  W.HECO & -0.0234 & 0.0109 & -0.0447 & -0.0234 & -0.0020 & -0.0234 & 0.0000 \\ 
+  W.POSE & -0.0195 & 0.0137 & -0.0463 & -0.0195 & 0.0073 & -0.0195 & 0.0000 \\ 
+  W.PSSP & -0.3590 & 0.0322 & -0.4223 & -0.3590 & -0.2957 & -0.3590 & 0.0000 \\ 
+  W.allcov & -0.0053 & 0.0014 & -0.0081 & -0.0053 & -0.0026 & -0.0053 & 0.0000 \\ 
+  W.allpts & -0.0567 & 0.0251 & -0.1060 & -0.0567 & -0.0074 & -0.0567 & 0.0000 \\ 
    \hline
@@
    \hline
+
+   \hline
+    $\gamma$[1] & 0.2843 & 0.6342 & -1.0900 & 1.4950 & 1.0279 &   810 \\ 
+    $\gamma$[2] & 3.2828 & 0.4467 & 2.4429 & 4.1940 & 1.0062 &   260 \\ 
+    $\gamma$[3] & 3.2923 & 0.3691 & 2.5680 & 4.0320 & 1.0040 &  1800 \\ 
+    $\gamma$[4] & 2.9362 & 0.3811 & 2.1779 & 3.6462 & 1.0116 &   140 \\ 
+    $\chi$[2,2] & 0.0405 & 0.4077 & -0.7392 & 0.8147 & 1.0011 &  2000 \\ 
+    $\chi$[2,3] & -0.1396 & 0.2658 & -0.6654 & 0.3970 & 1.0015 &  1400 \\ 
+    $\chi$[2,4] & 0.4050 & 0.2439 & -0.0664 & 0.8924 & 1.0010 &  2000 \\ 
+    $\chi$[3,1] & -1.3691 & 0.6755 & -2.7260 & -0.0685 & 1.0093 &   170 \\ 
+      $\omega$[1,1] & -0.5539 & 0.1165 & -0.7517 & -0.2845 & 1.0329 &   240 \\ 
+      $\omega$[1,2] & 0.0654 & 0.0581 & -0.0429 & 0.1898 & 1.0025 &   740 \\ 
+      $\omega$[1,3] & 0.0043 & 0.0389 & -0.0771 & 0.0813 & 1.0019 &  2000 \\ 
+      $\omega$[1,4] & 0.0953 & 0.0410 & 0.0209 & 0.1830 & 1.0120 &   130 \\ 
+      $\omega$[2,1] & -0.4522 & 0.1533 & -0.7461 & -0.1556 & 1.0007 &  2000 \\ 
+      $\omega$[2,2] & -1.5975 & 0.1175 & -1.8260 & -1.3669 & 1.0033 &  2000 \\ 
+      $\omega$[2,3] & -0.0058 & 0.0866 & -0.1806 & 0.1662 & 1.0035 &  2000 \\ 
+      $\omega$[2,4] & -0.3047 & 0.0858 & -0.4794 & -0.1381 & 1.0025 &   740 \\ 
+      $\omega$[3,1] & -0.5030 & 0.2801 & -1.0880 & 0.0341 & 1.0185 &   120 \\ 
+      $\omega$[3,2] & -0.0586 & 0.1571 & -0.3722 & 0.2399 & 1.0053 &   320 \\ 
+      $\omega$[3,3] & -1.7052 & 0.1471 & -1.9770 & -1.3980 & 1.0006 &  2000 \\ 
+      $\omega$[3,4] & -0.0960 & 0.1304 & -0.3682 & 0.1443 & 1.0005 &  2000 \\ 
+      $\omega$[4,1] & -0.1253 & 0.2688 & -0.6245 & 0.3921 & 1.0028 &  2000 \\ 
+      $\omega$[4,2] & -0.3056 & 0.1911 & -0.7052 & 0.0560 & 1.0041 &   420 \\ 
+      $\omega$[4,3] & 0.0335 & 0.1386 & -0.2499 & 0.2999 & 1.0040 &   440 \\ 
+      $\omega$[4,4] & -1.7046 & 0.1597 & -2.0080 & -1.3900 & 1.0101 &   250 \\ 
+     $\theta$[1] & 0.6117 & 0.0708 & 0.4855 & 0.7560 & 1.0006 &  2000 \\ 
+     $\theta$[2] & 1.1862 & 0.1462 & 0.9392 & 1.5000 & 1.0014 &  2000 \\ 
+     $\theta$[3] & 1.1782 & 0.1043 & 0.9872 & 1.4080 & 1.0009 &  2000 \\ 
+     $\theta$[4] & 1.2081 & 0.1060 & 1.0090 & 1.4210 & 1.0006 &  2000 \\ 
+     $p$[1] & 0.8367 & 0.1095 & 0.5280 & 0.9604 & 1.0235 &  2000 \\ 
+     $p$[2] & 0.9995 & 0.0006 & 0.9979 & 1.0000 & 1.0006 &  2000 \\ 
+     $p$[3] & 0.8834 & 0.0634 & 0.7124 & 0.9619 & 1.0054 &   410 \\ 
+     $p$[4] & 0.8100 & 0.1410 & 0.4269 & 0.9650 & 1.0023 &   830 \\ 
+      \hline
 \end{tabular}
@@
 \underline{Survival} & & & & \\        
-ARTR   & -1.5214  & -0.0694  & -0.0049 &  0.0281\\
-HECO   & -0.0064  &  -0.6647 &  0.0732 &  0.0190\\
-POSE  &   0.0001&  -0.0153 &  -1.2632&   0.0249\\
-PSSP  &  0.0101 &  0.0046 & 0.0245 &  -0.4430\\
+ARTR & -2.8522 & -0.0720 & 0.0617 & 0.0387\\
+HECO & -0.0063 & -1.6962 & 0.0649 & 0.0313\\
+POSE & 0.0000 & -0.0224 & -2.0073 & 0.0468\\
+PSSP & 0.0099 & 0.0025 & 0.0517 & -1.1267\\
 \underline{Growth} & & & & \\ 
-ARTR   & -0.0364 &  0.0033& -0.0573&  0.0062\\
-HECO   &  -0.0031 & -0.0479& 0.0242 & -0.0151\\
-POSE  &  -0.0002& 0.0039&  -0.2641&   -0.0073\\
-PSSP  & -0.0025 & -0.0118 & -0.0067 & -0.1480\\
+ARTR & -0.1497 & 0.0045 & -0.0614 & -0.0050\\
+HECO & -0.0042 & -0.2134 & 0.0204 & -0.0373\\
+POSE & -0.0004 & 0.0110 & -0.5753 & -0.0093\\
+PSSP & -0.0023 & -0.0201 & -0.0200 & -0.3845\\
 \underline{Recruitment} & & & & \\ 
-ARTR  & -0.5881 &  -0.4800 & -0.6589 & -0.1867\\
-HECO   &   0.06346 & -1.7368 & -0.1021 & -0.3746\\
-POSE  &   0.0220 &  0.0388 & -1.8943 &  0.1131\\
-PSSP  &  0.1164 & -0.3682 & -0.1953 & -1.7379\\
+ARTR & -0.5539 & -0.4522 & -0.5030 & -0.1253\\
+HECO &  0.0654 & -1.5975 & -0.0586 & -0.3056\\
+POSE & 0.0043 & -0.0058 & -1.7052 & 0.0335\\
+PSSP & 0.0953 & -0.3047 & -0.0960 & -1.7046\\
 \hline

### 2018-01-22 removals/removal_analysis_wrapper.r

@@ -39,9 +39,7 @@ rm(list=tmp)
 trtTests <- data.frame("species"="c","stage"="c","effect"=1,"CI.02.5"=1,"CI.97.5"=1,stringsAsFactors = F)
 
 # read in distance weights
-#dists <- read.csv(paste0(root,"/ExperimentTests/data/idaho_modern/speciesdata/IdahoModDistanceWeights_noExptl.csv"))
-dists <- read.csv(paste0(root,"/ExperimentTests/data/idaho/speciesdata/IdahoDistanceWeights.csv"))
-
+dists <- read.csv(paste0(root,"/ExperimentTests/data/idaho_modern/speciesdata/IdahoModDistanceWeights.csv"))
 
 ###
 ### fit survival models (takes ~ 10 minutes)
@@ -179,8 +177,7 @@ rm(list=tmp)
 sppList <-  c("ARTR","HECO","POSE","PSSP")
 
 # read in distance weights
-#dists <- read.csv(paste0(root,"/ExperimentTests/data/idaho_modern/speciesdata/IdahoModDistanceWeights_noExptl.csv"))
-dists <- read.csv(paste0(root,"/ExperimentTests/data/idaho/speciesdata/IdahoDistanceWeights.csv"))
+dists <- read.csv(paste0(root,"/ExperimentTests/data/idaho_modern/speciesdata/IdahoModDistanceWeights.csv"))
 
 max.CI <- F  # TRUE means use maximum removal effect
 source("ibm/ibm_removal_1step.r")
@@ -219,11 +216,6 @@ meanCover1 <- meanCover
 #no treatment effects, ARTR removal
 init.species <- c(2:4)
 source("ipm/IPM-getEquilibrium.r")
-# SPE 
-par(mfrow=c(2,1)); 
-HECO.base = sizeSave[[2]]; 
-plot(v[[2]],apply(HECO.base,1,mean));
-#
 write.csv(covSave[(burn.in+1):tlimit,],"ipm/baselineCover-noARTR.csv",row.names=F)
 meanCover2 <- meanCover
 
@@ -233,10 +225,6 @@ trtEffects=T
 max.CI=F
 source("ipm/IPM-setup.r")
 source("ipm/IPM-getEquilibrium.r")
-# SPE 
-HECO.trt = sizeSave[[2]]; 
-plot(v[[2]],apply(HECO.trt,1,mean));
-#
[…]

### 2018-01-24 removals/manuscript/removal_ms.tex
Starting to rewrite Results and Discussion

@@
 If our ``baseline" regression model, which accounts for local plant-plant interactions, accurately explains the observed responses to removals, then the removal coefficient should be non-significant.
-For survival and recruitment, the removal coefficients were never significantly positive, indicating that the baseline model did not underestimate interspecific competition.
-For \textit{Poa secunda} recruitment, the removal coefficient was negative, indicating that our baseline model overestimated interspecific competition.
-For growth, the removal treatment effect was significant and positive for two species, \textit{Poa secunda} and \textit{Pseudoroegneria spicata}, indicating that the baseline model underestimated interspecific competition.
-However, including information about the location of grass individuals with respect to removed \textit{A.
-tripartita} failed to improve the growth regressions, raising questions about the mechanisms driving the positive response to removal.
+\new{For survival and recruitment, the removal coefficients were never significantly different from zero.} For growth, the removal treatment effect was significant and positive for two species, \textit{Poa secunda} and \textit{Pseudoroegneria spicata}, indicating that the baseline model underestimated interspecific competition.
+However, including information about pretreatment cover of \textit{A.
+tripartita} at the quadrat or individual scale failed to improve the growth regressions, raising questions about the mechanisms driving the positive response to removal.
 For three species, population projections based on the vital rate regressions showed little impact of removals.
@@
  meaning that individuals of these species grew more rapidly in removal plots than controls even after accounting for neighborhood interactions 
- (Fig. \ref{fig:VitalRateTest}; Tables \ref{ARTRgrowth} to \ref{PSSPgrowth}). In contrast, \textit{P. secunda} had lower recruitment in removal 
- than controls plots (upper limit of 95\% CI $<0$; Fig. \ref{fig:VitalRateTest}; Table \ref{table:recruitment}). For the other species, 
- we could not detect significant removal treatment effects on recruitment.
+ (Fig.
+\ref{fig:VitalRateTest}; Tables \ref{ARTRgrowth} to \ref{PSSPgrowth}).
+\new{ We could not detect significant removal treatment effects on recruitment for any species (Fig.
+\ref{fig:VitalRateTest}; Table \ref{table:recruitment}).}
 
@@
 \ref{fig:W-by-treatment}).
-These pre-existing differences in composition raise the possibility that plots selected for the removal treatments might differ from the control plots in some unobserved covariate that could influence plant performance.
+These pre-existing differences in composition raise the possibility that plots selected for the removal treatments might differ from the control plots in some unobserved covariate that could influence plant composition and performance.
 
@@
 \ref{fig:IPMresults}).
-In fact, projected average cover of \textit{P.
-secunda} and \textit{P.
-spicata} declines slightly following \textit{A.
+In fact, projected average cover of all three grass species declines slightly following \textit{A.
 tripartita} removal.
-Using a model with removal treatment coefficients, the IPM predicts little effect of shrub removal on \textit{Poa secunda} and \textit{H.
-comata}, but a 2-fold increase in the cover of \textit{P.
-spicata} (Fig.
+Using a model with removal treatment coefficients, the IPM projects that \textit{A.
+tripartita} removal will increase equilibrium cover, relative to the baseline scenario, by a factor of 1.4, 1.5 and 3.0 for \textit{H.
+comata}, \textit{Poa secunda} and \textit{P.
+spicata}, respectively (Fig.
 \ref{fig:IPMresults}).
-Running a version of the model with maximum removal treatment effects (removal coefficients set to the 95\% credible limit furthest from zero), resulted in the same qualitative outcome, but a larger increase in \textit{P.
+Running a version of the model with maximum removal treatment effects (removal coefficients set to the 95\% credible limit furthest from zero), resulted in a much larger projected increase in \textbf{H.
+comata}, reflecting its small sample size and corresponding high uncertainty, but little change in the projections for \textit{Poa secunda} and \textit{P.
 spicata} (Fig.
@@
 The increased fecundity led to a doubling of \textit{A.
-tripartita} average cover, but had essentially zero effect on the cover of the grasses (changes of less than 0.1\% cover), implying that even if shrub cover were higher its removal would still have little effect on the grasses.
-
-The baseline model predicts more dramatic competitive release of \textit{A. tripartita} following grass removal: mean cover increases from 5.4\% to 16.2\%. Adding
-removal treatment effects dramatically alters the outcome, with \textit{A. tripartita} decreasing towards extinction. However, these outcomes both are driven
-by parameters that are estimated with very high uncertainty (Table~\ref{table:recruitment}).
+tripartita} average cover, but had essentially zero effect on the cover of the grasses (changes of less than 0.1\% cover),
+% CHECK THIS
+ implying that even if shrub cover were higher its removal would still have little effect on the grasses.
+
+The baseline model predicts more dramatic competitive release of \textit{A. tripartita} following grass removal: mean cover increases from 4\% to over11\%. 
+Adding
+removal treatment effects dramatically alters the outcome, with \textit{A.
+tripartita} decreasing towards extinction.
+However, these outcomes both are driven by parameters that are estimated with very high uncertainty (Table~\ref{table:recruitment}).
 The baseline model predicts an increase following grass removal because of the relatively strong negative effects of \textit{Poa secunda} and \textit{H.
@@
 The estimated effect of
-\emph{Poa} is -0.66, but the standard deviation of the estimate is 0.33 and the 95\% credible interval stops just short of zero; the 95\% credible interval for the
+\emph{Poa} is -0.50, but the standard deviation of the estimate is 0.28 and the 95\% credible interval stops just short of zero; the 95\% credible interval for the
 \textit{H. comata} effect includes values less than half of the point-estimate. The grass removal treatment effect, which leads to predicted extinction, has a credible 
@@
 \section*{Discussion}
-
-\subsection*{An experimental test of model predictions}
 
@@
 not just additional observations, further increases the potential to confront the model with a novel situation outside the historical range of variation.
-The results provide substantial support for our model's predictions, and for the weakness of interspecific interactions in this community.
+\new{We expected that upon completion of this study, we would have clear evidence for the success or failure of our previous model's predictions.
+Unfortunately, our results do not provide that clarity.
+In the sections below, we first summarize the evidence that our experiment demonstrated greater competitive release than our model predicted, we then explain why the experimental results might warrant as much skepticism as our original results, and we end by discussing general lessons for the study of interspecific competition.}
+
+\subsection*{Evidence for greater competitive release than predicted}
 
@@
 The one exception to this pattern is the effect of POSE on recruitment by ARTR, but the nonconforming
-coefficient ($\omega$[3,1] = -0.6589 in Table \ref{table:recruitment}) has low precision (sd = 0.3342, Table \ref{table:recruitment}.} 
+coefficient ($\omega$[3,1] = -0.50 in Table \ref{table:recruitment}) has low precision (sd = 0.28, Table \ref{table:recruitment}.} 
 \label{table:alphas}

### 2018-01-26 removals/manuscript/removal_ms.tex
Intermediate back up on manuscript changes

@@
 
-\title{Weak interspecific interactions in a sagebrush steppe: evidence from observations, models, and experiments}
+\title{Weak interspecific interactions in a sagebrush steppe\new{?: conflicting evidence from observations and experiments} }
 
@@
 If our ``baseline" regression model, which accounts for local plant-plant interactions, accurately explains the observed responses to removals, then the removal coefficient should be non-significant.
-\new{For survival and recruitment, the removal coefficients were never significantly different from zero.} For growth, the removal treatment effect was significant and positive for two species, \textit{Poa secunda} and \textit{Pseudoroegneria spicata}, indicating that the baseline model underestimated interspecific competition.
+\new{For survival, the removal coefficients were never significantly different from zero, and only \textit{A.
+tripartita} showed a (negative) response to removals at the recruitment stage.} For growth, the removal treatment effect was significant and positive for two species, \textit{Poa secunda} and \textit{Pseudoroegneria spicata}, indicating that the baseline model underestimated interspecific competition.
 However, including information about pretreatment cover of \textit{A.
@@
 \ref{fig:VitalRateTest}; Tables \ref{ARTRgrowth} to \ref{PSSPgrowth}).
-\new{ We could not detect significant removal treatment effects on recruitment for any species (Fig.
+\new{ At the recruitment stage, only \textit{A.
+tripartita} showed a significant response to removals, and it was negative (Fig.
 \ref{fig:VitalRateTest}; Table \ref{table:recruitment}).}
@@
 We conducted additional analyses to further explore removal effects on the growth of these two species.
-First, we classified all individuals of these species as located either inside or outside the limits of \textit{A.
+First, \new{we asked whether increases in grass growth following removal were greatest in quadrats or neighborhoods where we removed the most \textit{A.
+tripartita}.
+We did this by taking residuals from the fitted growth models for plants in removal quadrats and plotting those residuals against the pre-treatment, quadrat-level cover of \textit{A.
+tripartita}.
+No species showed the positive relationship we expected to see (Fig.
+\ref{fig:GrowthResidsQuad}).
+In fact, \textit{Poa secunda} residuals were significantly lower in quadrats with high pre-treatment cover of \textit{A.
+tripartita} (individuals in high cover quadrats are growing more slowly than the growth model predicted).
+We repeated this analysis at the individual plant level for grasses present in 2011.
+We calculated pre-treatment crowding from \textit{A.
+tripartita} for each of those individuals, then plotted their growth residuals against crowding for all years in which each plant survived.
+We found no significant relationships between the growth residuals and pre-treatment crowding from \textit{A.
+tripartita} (Fig.
+\ref{fig:GrowthResidsIndivid}).
+}
+
+%we classified all individuals of these species as located either inside or outside the limits of \textit{A.
 tripartita} canopies present in 2011 (before removal).
 If removal is releasing grasses from \textit{A.
-tripartita} competition for light or soil resources, we would expect individuals located under the shrub canopies to experience a greater effect of removal than individuals further away.
+tripartita} competition for light or soil resources, we would expect individuals located under shrub canopies to experience a greater effect of removal than individuals further away.
 However, the location of individuals with respect to \textit{A.
@@
 spicata}, we found no sign of a decrease in the removal effect over time (Tables \ref{table:POSEgrowth-trtYears} and \ref{table:PSSPgrowth-trtYears}).
+In fact, for  \textit{P.
+secunda}, the removal effects increase over time.
 
@@
 These pre-existing differences in composition raise the possibility that plots selected for the removal treatments might differ from the control plots in some unobserved covariate that could influence plant composition and performance.
+\new{On the other hand, some of the control plots that currently contain zero \textit{A.
+tripartita} had high \textit{A.
+tripartita} in the 1900's, and vice versa.
+If abiotic variation, such as soil depth, were the primary control of variation in \textit{A.
+tripartita} cover, such large changes in cover over time would be unlikely.
+}
 
@@
 comata}, differences in predicted cover of removal plots between the baseline and removal models were very small.
-For the other three species, the removal model predicted higher cover in removal plots than the baseline model predicted, but only for \textit{P.
-spicata} were predictions from the removal model closer to the 1:1 line.
+For the other three species, the removal model predicted higher cover in removal plots than the baseline model predicted.
 
@@
 The increased fecundity led to a doubling of \textit{A.
-tripartita} average cover, but had essentially zero effect on the cover of the grasses (changes of less than 0.1\% cover),
-% CHECK THIS
- implying that even if shrub cover were higher its removal would still have little effect on the grasses.
+tripartita} average cover, but had essentially zero effect on the cover of the grasses (changes of less than 0.1\% cover), implying that even if shrub cover were higher its removal would still have little effect on the grasses.
 
 The baseline model predicts more dramatic competitive release of \textit{A. tripartita} following grass removal: mean cover increases from 4\% to over11\%. 
-Adding
-removal treatment effects dramatically alters the outcome, with \textit{A.
+Adding removal treatment effects dramatically alters the outcome, with \textit{A.
 tripartita} decreasing towards extinction.
@@
 Unfortunately, our results do not provide that clarity.
-In the sections below, we first summarize the evidence that our experiment demonstrated greater competitive release than our model predicted, we then explain why the experimental results might warrant as much skepticism as our original results, and we end by discussing general lessons for the study of interspecific competition.}
+In the sections below, we first summarize the evidence that our experiment revealed greater competitive release than our baseline model predicted, we then explain weaknesses in the experimental results that complicate the interpretation, and we end by discussing general lessons for the study of interspecific competition.}
 
@@
 
-For three of the four species, we find little or no evidence that models based on observational data underestimate 
-the intensity of interspecific competition. After accounting 
-for local neighborhood crowding, the removal treatment effects were never significant for \textit{A. tripartita} or \textit{H. comata}; 
-our baseline model, which accounts for local neighborhood composition, adequately explained vital rate responses to removal.  
-\textit{P. secunda} showed mixed responses to \textit{A. tripartita} removal: removal effects were not significant at the survival stage, 
-were positive at the growth stage, and negative at the recruitment stage (where the estimated effect of \textit{A. tripartita} 
-on \textit{P. secunda} is positive). At the population level, models with and without removal effects projected similar patterns of 
-short (Fig. \ref{fig:IBM1step}) and long-term (Fig. \ref{fig:IPMresults}) dynamics for \textit{H. comata}. For \textit{A. tripartita}, 
-the experimental observations and short-term population projections show little response to removal, while the long-term simulations 
-from the baseline and removal effects models suggest strong (but opposite) effects of removal: a tripling of abundance in the baseline model, and extinction in the removal effects model.
-However, these long-term simulations primarily reflect the influence of recruitment parameters
+\new{We evaluated response to removals in three ways: vital rates, short-term population growth rates, and changes in long-term equilibrium cover.
+Two species, \textit{Poa secunda} and \textit{P.
+pseudoroegneria}, showed significant, positive responses to \textit{A.
+tripartita} removal at the level of vital rates.
+Individuals of these species grew 21\% and 24\% faster, respectively, in removal plots than controls, after accounting for variation in local neighborhood composition.
+In other words, our baseline model appeared to underestimate the competitive release of these species following \textit{A.
+tripartita} removal.}
+
+\new{The elevated growth of \textit{Poa secunda} and \textit{P.
+pseudoroegneria} individuals translated into greater short-term population growth rates.
+For these species, the individual-based model with removal effects predicted higher cover than the baseline version of the model.
+The differences in predicted cover of roughly 20\% (\ref{fig:IBM1step}) are comparable to the differences in the individual growth rates, reflecting the fact that at these short time scales changes in individual size influence cover more than survival or recruitment.
+For \textit{A.
+tripartita}, the magnitude of the difference between predictions from the baseline and removal effects models was similar, further emphasizing the short-term influence of growth compared to survival and recruitment, which responded negatively to removals in this species.
+}
+
+\new{Long-term projections using the integral projection model showed even larger responses to removal.
+For the three grasses, projected equilibrium cover in the absence of \textit{A.
+tripartita} is 1.4 to 3-fold higher for the removal effects model compared to the baseline model.
+Although the absolute changes in basal cover are small, these relative changes are substantial, and seem to present compelling evidence that competitive release is stronger than predicted by our baseline model.
+}
+
+\subsection*{Weaknesses in the experimental evidence}
+
+\new{Although the growth regressions for \textit{Poa secunda} and \textit{P.
+pseudoroegneria} and the long-run projections for all three grass species suggest stronger competitive release than our baseline model predicted, a variety of other evidence makes it harder to confidently reject those baseline model predictions.
+Only two of the twelve vital rate tests indicated greater competitive release than predicted.
+Granted, small samples sizes contributed to greater uncertaintly for \textit{A.
+tripartita} and \textit{H.
+comata}, but even for \textit{Poa secunda} and \textit{P.
+pseudoroegneria} the increases in growth were small relative to other sources of variation.
+For example, using quadrat-level cover data, we could not detect treatment differences (Table \ref{table:coefficients}).
+}
+
+\new{The strongest evidence for large competitive release of grasses following \textit{A.
+tripartita} removal comes from the long-term projections of the IPM.
+These projections effectively make removal treatment effects permanent; the removal effects on vital rates observed during the experiment are locked in for the entire 2500 time step simulation.
+Furthermore, these projections ignore parameter error because we wanted to highlight the removal effects.
+Given these strong assumptions, the projections should be viewed with some caution.
+}
+
+\new{For \textit{A. tripartita}, the long-term IPM projections actually show extinction following grass removal.
+However, this result primarily reflects the influence of recruitment parameters 
 that are highly uncertain (see \emph{Results}). This high uncertainty 
-results from \textit{A. tripartita} recruitment being low (a mean of 0.63 recruits per quadrat per year, compared to means of 2.6 to 5.3 for the grasses) and 
-very episodic (roughly half of the total recruits in the 22 years of historical data were observed in 2 of those years, and only 13 \textit{A. tripartita} 
-recruits were observed since we resumed annual censuses in 2007). Given this uncertainty, we put more weight on the direct observations 
-and short-term projections showing little effect of grass removal on \textit{A. tripartita} populations.
-For \textit{P. secunda}, the model with removal effects predicted higher population growth over the short but not the long-term, perhaps 
-because the short-term projections mostly reflect the positive effect of \textit{A. tripartita} removal on growth, while the long-term 
-equilibrium are also sensitive to the negative effects of removal on recruitment. Although the baseline model did 
-not always successfully predict removal effects on vital rates, it does appear to successfully capture the population-level 
-consequences of removals on these three species.
-
-The story is different for the fourth species, \textit{P. spicata}. Both individual growth and projected short- and long-term population 
-growth were higher in the sagebrush removal treatments than in the control plots, suggesting that our analyses of observational data may 
-underestimate the competitive effect of \textit{A. tripartita} on \textit{P. spicata}, and the resulting positive effect of \textit{A. tripartita} 
-removal. However, the mechanism by which \textit{A. tripartita} removal promotes \textit{P. spicata} remains unclear. If  \textit{A. tripartita} 
-removal is releasing \textit{P. spicata} from competition, we should expect \textit{P. spicata} individuals that were located beneath \textit{A. tripartita} 
-shrubs that we removed to respond more than individuals located away from any shrubs. Yet we found no evidence for such a difference. Perhaps 
-\textit{A. tripartita's} roots extend far enough beyond its canopy to compete for resources with grasses located in interspaces.
-An alternative explanation involves pre-treatment differences in composition between control plots and  the \textit{A. tripartita} 
-removal plots, which had higher cover of \textit{A. tripartita} (see 2011 means in Fig. \ref{fig:CoverTrends}) and higher cover of subdominant 
-grasses and shrubs (Fig. \ref{fig:W-by-treatment}). Perhaps these pre-treatment differences reflect subtle differences in 
-edaphic conditions which could promote \textit{P. spicata} growth. In summary, while \textit{P. spicata} appears to 
-show stronger competitive release than our baseline model predicted, the evidence is not conclusive.
-
-How can we reconcile the insensitivity of grasses to sagebrush removal in our study with evidence from rangeland ecology 
-research showing that sagebrush eradication increases grass production? The most likely explanation involves ecological context: our 
-study focused on \textit{Artemisia tripartita} in an ungrazed plant community, whereas most research on sagebrush 
-and forage production has focused on \textit{Artemisia tridentata} in heavily grazed communities, and often has involved grass seeding 
-following sagebrush removal (e.g. \citealt{Cook1963}). \textit{A. tripartita} is smaller in stature than \textit{A. tridentata} 
-and may interact differently with perennial grasses. More importantly, cover of sagebrush can increase dramatically on overgrazed sites, 
-especially under a spring grazing regime. Competitive release of grasses following sagebrush removal often increases in 
-a nonlinear fashion, with little increase in production at sites with low initial sagebrush cover, 
-and accelerating responses as initial sagebrush cover passes some threshold \citep{McDaniel2005}, implying a 
-shift in the nature of interspecific interactions, perhaps due to changes in the balance of 
-competition and facilitation or in resource supply. The cover of 
-sagebrush in our ungrazed quadrats may have fallen below that threshold.
+results from \textit{A. tripartita} recruitment being low
+ and very episodic (roughly half of the total recruits in the 22 years of historical data were observed in 2 of those years, and only 28 \textit{A. tripartita} 
+recruits have been observed since we resumed annual censuses in 2007). Because of this uncertainty, we put more weight on the direct observations 
+and short-term projections showing little effect of grass removal on \textit{A. tripartita} populations. }
+
+\new{ Another reason we view the growth results with some suspicion is that increases in growth were no larger in individual neighborhoods or quadrats with higher pre-treatment \textit{A.
+tripartita} cover.
+If \textit{A.
+tripartita} removal is releasing grasses from competition, shouldn't this effect be larger for grasses located in areas with high \textit{A.
+tripartita} cover?
+The absence of this relationship (Figs.
+\ref{fig:GrowthResidsQuad} and \ref{fig:GrowthResidsIndivid}) raises questions about the mechanism by which \textit{A.
+tripartita} removal promotes grass growth.
+Perhaps \textit{A.
+tripartita's} roots extend far enough beyond its canopy to compete for resources with grasses located in interspaces.
+While this might explain the lack of the relationship at the neighbrhood scale, it seems unlikely to explain the quadrat-level results.
+}
+An alternative explanation involves pre-treatment differences in composition between control plots and  the \textit{A.
+tripartita} removal plots, which had higher cover of \textit{A.
+tripartita} (see 2011 means in Fig.
+\ref{fig:CoverTrends}) and higher cover of subdominant grasses and shrubs (Fig.
+\ref{fig:W-by-treatment}).
+These pre-treatment differences could reflect subtle differences in edaphic conditions which could promote \textit{P.
+spicata} growth.
+\new{However, since the 1900's, many quadrats have experienced dramatic changes in \textit{A.
+tripartita} cover, arguing against viewing \textit{A.
+tripartita} cover as a sensitive indicator of edaphic conditions.
+A final possibility is that our removal treatment influenced grass growth through some other mechanism, unrelated to competition.
+Maybe the loss of shrub cover benefits grasses by reducing small mammal herbivory (FIND REF)?
+Without a better understanding of the mechanism driving increased grass growth in the shrub removal quadrats, we are hesitant to dramatically adjust our prior understanding based on models fit to decades of observational data.
+}
+
+%How can we reconcile the insensitivity of grasses to sagebrush removal in our study with evidence from rangeland ecology 
+%research showing that sagebrush eradication increases grass production? The most likely explanation involves ecological context: our 
+%study focused on \textit{Artemisia tripartita} in an ungrazed plant community, whereas most research on sagebrush 
+%and forage production has focused on \textit{Artemisia tridentata} in heavily grazed communities, and often has involved grass seeding 
+%following sagebrush removal (e.g. \citealt{Cook1963}). \textit{A. tripartita} is smaller in stature than \textit{A. tridentata} 
+%and may interact differently with perennial grasses. More importantly, cover of sagebrush can increase dramatically on overgrazed sites, 
+%especially under a spring grazing regime. Competitive release of grasses following sagebrush removal often increases in 
+%a nonlinear fashion, with little increase in production at sites with low initial sagebrush cover, 
+%and accelerating responses as initial sagebrush cover passes some threshold \citep{McDaniel2005}, implying a 
+%shift in the nature of interspecific interactions, perhaps due to changes in the balance of 
+%competition and facilitation or in resource supply. The cover of 
+%sagebrush in our ungrazed quadrats may have fallen below that threshold.
+
+\subsection*{Lessons for studying competition}
 
@@
 interspecific interactions. After all, our approach did detect strong intraspecific effects.
+
 Finally, our results give some cause for optimism about population forecasting.
@@
 
-\subsection*{Implications of weak interspecific interactions}
-
-The weak interspecific interactions among many of the dominant species in our system indicates that these species  
-are limited by quite different factors. This low niche overlap could reflect strong resource partitioning \citep{tilman_resource_1982}, 
-perhaps combined with variation in the times and locations at which species are taking up those resources. 
-Species-specific responses to spatial and temporal variation in non-resource environmental variables can also weaken interspecific 
-interactions \citep{chesson_mechanisms_2000}, though in a previous study we failed to find evidence of a temporal storage effect 
-in this community \citep{adler_weak_2009}. Interactions with above- or belowground enemies can regulate populations and, given 
-enough species-specificity in the effects of these enemies, decouple the dynamics of different species \citep{hersh_evaluating_2011,
-janzen_herbivores_1970,connell_role_1971}.  More likely, plants may alter the environment, and even resource availability, 
-in ways that facilitate their neighbors. For example, shade cast by \textit{A. tripartita} canopies may reduce 
-evapotranspiration (e.g. \citealt{Barbier2008}). \textit{A. tripartita} removal might offset this facilitative effect by reducing 
-resource uptake, resulting in little net change in water balance following removal. 
-Unfortunately, we did not monitor soil moisture at the fine spatial and temporal resolution necessary to test this hypothesis.
-
-Evidence for weak interspecific interactions and limited competitive release has interesting implications. If most 
-interspecific interactions are weak, then single-species models may be adequate for forecasting population responses 
-to environmental change \citep{adler_forecasting_2012,kleinhesselink_indirect_2015,Chu2016}. Similarly, differences among 
-species in relative abundance should reflect variation in the strength of intraspecific density dependence more than 
-variation in the strength of interspecific interactions \citep{yenni_strong_2012}. A recent global study showing little 
-variation in the abundance of invasive grassland species in native and invaded ranges on different 
-continents  \citep{firn_abundance_2011} is consistent with the idea that intraspecific interactions are the primary determinant 
-of species abundances. In systems where this is true, effective management and conservation requires understanding the sources 
-and species-specificity of intraspecific density dependence. A useful first step in this direction would be to test 
-whether our model is also successful at predicting the effects of release from intraspecific competition.
-
-For many species in other communities, interspecific competition may still be an important driver. In closed-canopy 
-forests where competition for light is intense, we might have found different results than in our system, in which 
-competition for a number of belowground resources is more important than competition for light. However, a 
-number of recent studies from forests have also found that heterospecific competitive effects are dramatically 
-weaker than conspecific effects \citep{comita_asymmetric_2010,Kunstler2016,johnson_conspecific_2012}.  We also 
-expect stronger competition in successional systems, or communities in the early stages of community assembly 
-where the process of competitive exclusion is still underway \citep{kokkoris_patterns_1999}. Finally, we might have
-found stronger evidence of competition if we could have studied rare species in our system, but only if those species 
-are rare because of niche overlap with stronger competitors. If they are rare because they occupy unique, but narrow, niches, 
-interspecific interactions would still be weak. Determining why interspecific interactions and competitive release are 
-strong in some ecological contexts but weak in others remains a fascinating research problem.
+%\subsection*{Implications of weak interspecific interactions}
+%
+%The weak interspecific interactions among many of the dominant species in our system indicates that these species  
+%are limited by quite different factors. This low niche overlap could reflect strong resource partitioning \citep{tilman_resource_1982}, 
+%perhaps combined with variation in the times and locations at which species are taking up those resources. 
+%Species-specific responses to spatial and temporal variation in non-resource environmental variables can also weaken interspecific 
+%interactions \citep{chesson_mechanisms_2000}, though in a previous study we failed to find evidence of a temporal storage effect 
+%in this community \citep{adler_weak_2009}. Interactions with above- or belowground enemies can regulate populations and, given 
+%enough species-specificity in the effects of these enemies, decouple the dynamics of different species \citep{hersh_evaluating_2011,
+%janzen_herbivores_1970,connell_role_1971}.  More likely, plants may alter the environment, and even resource availability, 
+%in ways that facilitate their neighbors. For example, shade cast by \textit{A. tripartita} canopies may reduce 
+%evapotranspiration (e.g. \citealt{Barbier2008}). \textit{A. tripartita} removal might offset this facilitative effect by reducing 
+%resource uptake, resulting in little net change in water balance following removal. 
+%Unfortunately, we did not monitor soil moisture at the fine spatial and temporal resolution necessary to test this hypothesis.
+%
+%Evidence for weak interspecific interactions and limited competitive release has interesting implications. If most 
+%interspecific interactions are weak, then single-species models may be adequate for forecasting population responses 
+%to environmental change \citep{adler_forecasting_2012,kleinhesselink_indirect_2015,Chu2016}. Similarly, differences among 
+%species in relative abundance should reflect variation in the strength of intraspecific density dependence more than 
+%variation in the strength of interspecific interactions \citep{yenni_strong_2012}. A recent global study showing little 
+%variation in the abundance of invasive grassland species in native and invaded ranges on different 
+%continents  \citep{firn_abundance_2011} is consistent with the idea that intraspecific interactions are the primary determinant 
+%of species abundances. In systems where this is true, effective management and conservation requires understanding the sources 
+%and species-specificity of intraspecific density dependence. A useful first step in this direction would be to test 
+%whether our model is also successful at predicting the effects of release from intraspecific competition.
+%
+%For many species in other communities, interspecific competition may still be an important driver. In closed-canopy 
+%forests where competition for light is intense, we might have found different results than in our system, in which 
+%competition for a number of belowground resources is more important than competition for light. However, a 
+%number of recent studies from forests have also found that heterospecific competitive effects are dramatically 
+%weaker than conspecific effects \citep{comita_asymmetric_2010,Kunstler2016,johnson_conspecific_2012}.  We also 
+%expect stronger competition in successional systems, or communities in the early stages of community assembly 
+%where the process of competitive exclusion is still underway \citep{kokkoris_patterns_1999}. Finally, we might have
+%found stronger evidence of competition if we could have studied rare species in our system, but only if those species 
+%are rare because of niche overlap with stronger competitors. If they are rare because they occupy unique, but narrow, niches, 
+%interspecific interactions would still be weak. Determining why interspecific interactions and competitive release are 
+%strong in some ecological contexts but weak in others remains a fascinating research problem.
 
@@
 
-\begin{table}
-\centering
-\caption{Summary of fixed effects for the \textit{P.
+%\begin{table}
+%\centering
+%\caption{Summary of fixed effects for the \textit{P.
 secunda} growth model with individual-level \textit{A.
@@
 }
-\label{table:POSEgrowth-inARTR}
-\begin{tabular}{rrrrrrrr}
-  \hline
- & mean & sd & 0.025quant & 0.5quant & 0.975quant & mode & kld \\ 
-  \hline
-(Intercept) & 0.5080 & 0.0625 & 0.3841 & 0.5081 & 0.6311 & 0.5084 & 0.0000 \\ 
-  logarea.t0 & 0.6665 & 0.0228 & 0.6211 & 0.6666 & 0.7111 & 0.6669 & 0.0000 \\ 
-  TreatmentNo\_shrub & 0.2247 & 0.0744 & 0.0786 & 0.2247 & 0.3707 & 0.2247 & 0.0000 \\ 
-  W.ARTR & -0.0004 & 0.0009 & -0.0022 & -0.0004 & 0.0013 & -0.0004 & 0.0000 \\ 
-  W.HECO & 0.0109 & 0.0100 & -0.0088 & 0.0109 & 0.0306 & 0.0109 & 0.0000 \\ 
-  W.POSE & -0.5770 & 0.0624 & -0.6995 & -0.5770 & -0.4547 & -0.5770 & 0.0000 \\ 
-  W.PSSP & -0.0093 & 0.0114 & -0.0316 & -0.0093 & 0.0130 & -0.0093 & 0.0000 \\ 
-  W.allcov & -0.0004 & 0.0016 & -0.0036 & -0.0004 & 0.0027 & -0.0004 & 0.0000 \\ 
-  W.allpts & -0.0340 & 0.0256 & -0.0842 & -0.0340 & 0.0161 & -0.0340 & 0.0000 \\ 
-  inARTR & -0.0890 & 0.1126 & -0.3100 & -0.0890 & 0.1319 & -0.0890 & 0.0000 \\ 
-   \hline
-\end{tabular}
-\end{table}
-
-\begin{table}
-\centering
-\caption{Summary of fixed effects for the \textit{P.
+%\label{table:POSEgrowth-inARTR}
+%\begin{tabular}{rrrrrrrr}
+%  \hline
+% & mean & sd & 0.025quant & 0.5quant & 0.975quant & mode & kld \\ 
+%  \hline
+%(Intercept) & 0.5080 & 0.0625 & 0.3841 & 0.5081 & 0.6311 & 0.5084 & 0.0000 \\ 
+%  logarea.t0 & 0.6665 & 0.0228 & 0.6211 & 0.6666 & 0.7111 & 0.6669 & 0.0000 \\ 
+%  TreatmentNo\_shrub & 0.2247 & 0.0744 & 0.0786 & 0.2247 & 0.3707 & 0.2247 & 0.0000 \\ 
+%  W.ARTR & -0.0004 & 0.0009 & -0.0022 & -0.0004 & 0.0013 & -0.0004 & 0.0000 \\ 
+%  W.HECO & 0.0109 & 0.0100 & -0.0088 & 0.0109 & 0.0306 & 0.0109 & 0.0000 \\ 
+%  W.POSE & -0.5770 & 0.0624 & -0.6995 & -0.5770 & -0.4547 & -0.5770 & 0.0000 \\ 
+%  W.PSSP & -0.0093 & 0.0114 & -0.0316 & -0.0093 & 0.0130 & -0.0093 & 0.0000 \\ 
+%  W.allcov & -0.0004 & 0.0016 & -0.0036 & -0.0004 & 0.0027 & -0.0004 & 0.0000 \\ 
+%  W.allpts & -0.0340 & 0.0256 & -0.0842 & -0.0340 & 0.0161 & -0.0340 & 0.0000 \\ 
+%  inARTR & -0.0890 & 0.1126 & -0.3100 & -0.0890 & 0.1319 & -0.0890 & 0.0000 \\ 
+%   \hline
+%\end{tabular}
+%\end{table}
+%
+%\begin{table}
+%\centering
+%\caption{Summary of fixed effects for the \textit{P.
 spicata} growth model with individual-level \textit{A.
@@
 See Table \ref{ARTRgrowth} for an explanation of other coefficient names and column headers.}
-\label{table:PSSPgrowth-inARTR}
-\begin{tabular}{rrrrrrrr}
-  \hline
- & mean & sd & 0.025quant & 0.5quant & 0.975quant & mode & kld \\ 
-  \hline
-(Intercept) & 0.3955 & 0.0606 & 0.2764 & 0.3953 & 0.5153 & 0.3950 & 0.0000 \\ 
-  logarea.t0 & 0.8276 & 0.0168 & 0.7943 & 0.8277 & 0.8608 & 0.8277 & 0.0000 \\ 
-  TreatmentNo\_shrub & 0.2531 & 0.0784 & 0.0991 & 0.2531 & 0.4070 & 0.2531 & 0.0000 \\ 
-  W.ARTR & -0.0027 & 0.0008 & -0.0042 & -0.0027 & -0.0011 & -0.0027 & 0.0000 \\ 
-  W.HECO & -0.0231 & 0.0110 & -0.0447 & -0.0231 & -0.0016 & -0.0231 & 0.0000 \\ 
-  W.POSE & -0.0198 & 0.0138 & -0.0469 & -0.0198 & 0.0072 & -0.0198 & 0.0000 \\ 
-  W.PSSP & -0.3600 & 0.0326 & -0.4239 & -0.3600 & -0.2961 & -0.3600 & 0.0000 \\ 
-  W.allcov & -0.0051 & 0.0014 & -0.0079 & -0.0051 & -0.0024 & -0.0051 & 0.0000 \\ 
-  W.allpts & -0.0557 & 0.0254 & -0.1055 & -0.0557 & -0.0060 & -0.0557 & 0.0000 \\ 
-  inARTR & -0.0649 & 0.1111 & -0.2830 & -0.0649 & 0.1531 & -0.0649 & 0.0000 \\ 
-   \hline
-\end{tabular}
-\end{table}
+%\label{table:PSSPgrowth-inARTR}
+%\begin{tabular}{rrrrrrrr}
+%  \hline
+% & mean & sd & 0.025quant & 0.5quant & 0.975quant & mode & kld \\ 
+%  \hline
+%(Intercept) & 0.3955 & 0.0606 & 0.2764 & 0.3953 & 0.5153 & 0.3950 & 0.0000 \\ 
+%  logarea.t0 & 0.8276 & 0.0168 & 0.7943 & 0.8277 & 0.8608 & 0.8277 & 0.0000 \\ 
+%  TreatmentNo\_shrub & 0.2531 & 0.0784 & 0.0991 & 0.2531 & 0.4070 & 0.2531 & 0.0000 \\ 
+%  W.ARTR & -0.0027 & 0.0008 & -0.0042 & -0.0027 & -0.0011 & -0.0027 & 0.0000 \\ 
+%  W.HECO & -0.0231 & 0.0110 & -0.0447 & -0.0231 & -0.0016 & -0.0231 & 0.0000 \\ 
+%  W.POSE & -0.0198 & 0.0138 & -0.0469 & -0.0198 & 0.0072 & -0.0198 & 0.0000 \\ 
+%  W.PSSP & -0.3600 & 0.0326 & -0.4239 & -0.3600 & -0.2961 & -0.3600 & 0.0000 \\ 
+%  W.allcov & -0.0051 & 0.0014 & -0.0079 & -0.0051 & -0.0024 & -0.0051 & 0.0000 \\ 
+%  W.allpts & -0.0557 & 0.0254 & -0.1055 & -0.0557 & -0.0060 & -0.0557 & 0.0000 \\ 
+%  inARTR & -0.0649 & 0.1111 & -0.2830 & -0.0649 & 0.1531 & -0.0649 & 0.0000 \\ 
+%   \hline
+%\end{tabular}
+%\end{table}
 
@@
 
+\begin{figure}[h]
+\centering
+\includegraphics[width=1\textwidth]{growth_residuals_vs_quad_removal_cover}
+\caption{The relationship between residuals from the fitted growth model and quadrat-level pre-treatment cover of the removal species (all grasses removed in quadrats where we studied \textit{A.
+tripartita}, and \textit{A.
+tripartita} removed in the quadrats where we studied the grasses).
+Linear regressions are shown only when statistically significant.
+Source file:  \texttt{growth\textbackslash growth\textunderscore residuals\textunderscore by\textunderscore quad.R}.}
+\label{fig:GrowthResidsQuad}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=1\textwidth]{growth_residuals_vs_Wremoval}
+\caption{The relationship between residuals from the fitted growth model and individual-level pre-treatment crowding exerted by the removal species (all grasses in quadrats where we studied \textit{A.
+tripartita}, and \textit{A.
+tripartita} in the quadrats where we studied the grasses).
+Only individuals present in 2011 in the removal treatment quadrats are included in this analysis.
+If those individuals survived into later years, we did include their performance in those later years.
+None of the relationships were statistically significant.
+Source file:  \texttt{growth\textbackslash growth\textunderscore residuals\textunderscore by\textunderscore individ.R}.}
+\label{fig:GrowthResidsIndivid}
+\end{figure}
+
   \begin{figure}[tbp]

### 2018-01-26 removals/manuscript/removal_ms.tex
Finished first complete draft of re-do

@@
 
-\title{Weak interspecific interactions in a sagebrush steppe\new{?: conflicting evidence from observations and experiments} }
+\title{Weak interspecific interactions in a sagebrush steppe\new{? Conflicting evidence from observations and experiments} }
 
@@
 \new{For survival, the removal coefficients were never significantly different from zero, and only \textit{A.
-tripartita} showed a (negative) response to removals at the recruitment stage.} For growth, the removal treatment effect was significant and positive for two species, \textit{Poa secunda} and \textit{Pseudoroegneria spicata}, indicating that the baseline model underestimated interspecific competition.
-However, including information about pretreatment cover of \textit{A.
+tripartita} showed a (negative) response to removals at the recruitment stage.
+For growth, the removal treatment effect was significant and positive for two species, \textit{Poa secunda} and \textit{Pseudoroegneria spicata}, indicating that the baseline model underestimated interspecific competition.
+For all three grass species, population models based on the vital rate regressions that included removal effects projected 1.4 to 3-fold increases in equilibrium population size relative to the baseline model (no removal effects).} However, including information about pretreatment cover of \textit{A.
 tripartita} at the quadrat or individual scale failed to improve the growth regressions, raising questions about the mechanisms driving the positive response to removal.
-For three species, population projections based on the vital rate regressions showed little impact of removals.
-For \textit{P.
-spicata}, the population models that included effects of \textit{A.
-tripartita} removal projected greater competitive release than the baseline models.
-The ability of models based on observational data to predict population responses to experimental removals increased our confidence that interspecific interactions are in fact weak for at least three of our four study species.
+\new{While the experimental results showed greater competitive release than our baseline models predicted, we will need more tightly controlled experiments focused on underlying mechanisms to conclusively validate or reject our model's predictions.}
+
+%The ability of models based on observational data to predict population responses to experimental removals increased our confidence that interspecific interactions are in fact weak for at least three of our four study species.
 The implication is that factors besides interspecific feedbacks play important roles in determining the structure and dynamics of this community.
@@
 Unfortunately, our results do not provide that clarity.
-In the sections below, we first summarize the evidence that our experiment revealed greater competitive release than our baseline model predicted, we then explain weaknesses in the experimental results that complicate the interpretation, and we end by discussing general lessons for the study of interspecific competition.}
+In the sections below, we first summarize the evidence that our experiment revealed greater competitive release than our baseline model predicted, we then explain weaknesses in the experimental results that complicate the interpretation, and we end by discussing general lessons for the study of interspecific interactions.}
 
@@
 tripartita} is 1.4 to 3-fold higher for the removal effects model compared to the baseline model.
-Although the absolute changes in basal cover are small, these relative changes are substantial, and seem to present compelling evidence that competitive release is stronger than predicted by our baseline model.
+While the absolute changes in basal cover are small, these relative changes are substantial, and seem to present compelling evidence that competitive release is stronger than predicted by our baseline model.
 }
@@
 Only two of the twelve vital rate tests indicated greater competitive release than predicted.
-Granted, small samples sizes contributed to greater uncertaintly for \textit{A.
+Granted, small samples sizes contributed to greater uncertainty for \textit{A.
 tripartita} and \textit{H.
@@
 These projections effectively make removal treatment effects permanent; the removal effects on vital rates observed during the experiment are locked in for the entire 2500 time step simulation.
-Furthermore, these projections ignore parameter error because we wanted to highlight the removal effects.
+Furthermore, these projections ignore parameter error because we wanted to highlight the population consequences of the removal effects.
 Given these strong assumptions, the projections should be viewed with some caution.
@@
 \new{For \textit{A. tripartita}, the long-term IPM projections actually show extinction following grass removal.
-However, this result primarily reflects the influence of recruitment parameters 
-that are highly uncertain (see \emph{Results}). This high uncertainty 
-results from \textit{A. tripartita} recruitment being low
- and very episodic (roughly half of the total recruits in the 22 years of historical data were observed in 2 of those years, and only 28 \textit{A. tripartita} 
-recruits have been observed since we resumed annual censuses in 2007). Because of this uncertainty, we put more weight on the direct observations 
-and short-term projections showing little effect of grass removal on \textit{A. tripartita} populations. }
-
-\new{ Another reason we view the growth results with some suspicion is that increases in growth were no larger in individual neighborhoods or quadrats with higher pre-treatment \textit{A.
+However, this result primarily reflects the influence of recruitment parameters that are highly uncertain (see \emph{Results}).
+This high uncertainty results from \textit{A.
+tripartita} recruitment being low and very episodic (roughly half of the total recruits in the 22 years of historical data were observed in 2 of those years, and only 28 \textit{A.
+tripartita} recruits have been observed since we resumed annual censuses in 2007).
+Because of this uncertainty, we put more weight on the direct observations and short-term projections showing little effect of grass removal on \textit{A.
+tripartita} populations.}
+
+\new{Another reason we view the growth results with some suspicion is that increases in growth were no larger in individual neighborhoods or quadrats with higher pre-treatment \textit{A.
 tripartita} cover.
@@
 A final possibility is that our removal treatment influenced grass growth through some other mechanism, unrelated to competition.
-Maybe the loss of shrub cover benefits grasses by reducing small mammal herbivory (FIND REF)?
-Without a better understanding of the mechanism driving increased grass growth in the shrub removal quadrats, we are hesitant to dramatically adjust our prior understanding based on models fit to decades of observational data.
+Maybe the loss of shrub cover benefits grasses by reducing small mammal herbivory \citep{Orrock2010}?
+Without a better understanding of the mechanism driving increased grass growth in the shrub removal quadrats, we are hesitant to dramatically adjust our prior understanding gained from models fit to decades of observational data.
 }
@@
 
-\subsection*{Lessons for studying competition}
-
-What general lessons can we draw about testing predictions from population models? 
-First, partial removals, rather than complete removal treatments, could 
-have helped us to distinguish between competition-mediated and density-independent effects of removal treatments (e.g. Fig. \ref{fig:PSSPresids}) 
-Future work in other systems might benefit from this insight. Second, the success of our models supports 
-the use of long-term observational data for characterizing competitive interactions. We see no reason why 
-this approach would not work just as well in a community with stronger 
-interspecific interactions. After all, our approach did detect strong intraspecific effects.
-
-Finally, our results give some cause for optimism about population forecasting.
+\subsection*{Lessons for studying interspecific interactions}
+
+\new{A variety of approaches are available for studying interspecific interactions among plants, and each has its advantages and disadvantages.
+Observational datasets have the advantage of being realistic: they reflect all the processes and interactions that drive community dynamics across a range of spatial and temporal scales.
+Their disadvantage is weak inference.
+Distinguishing between correlation and causation is challenging, and simulations of a model fit to observational data may rely heavily on extrapolation.
+Experiments, such as response surface designs (Inouye), offer an alternative approach to studying interactions at the population scale.
+Their chief advantage is strong inference about causality.
+But they sacrifice realism: experiments are typically implemented at fine spatial and temporal scales which exclude potentially important processes.
+Finally, we can study interactions mechanistically by describing patterns of resource use \citep{Dybzinski2007,Silvertown1999}.
+Mechanistic studies have the virtue of focusing on directly on the processes of interest, but rarely address more than one limiting resource at a time.
+}
+
+\new{ Our previous work was based purely on observational data, so we were particularly concerned about the extrapolation issue.
+Specifically, our model projects the performance of individual plants growing in the absence of neighbors.
+Although the observational data set seems to include such situations (Fig.
+\ref{fig:Wscatters}), if these isolated individual are still interacting with neighbors belowground, our model might underestimate the true intrinsic growth rates.
+Therefore, we conducted the removal experiment, hoping to gain the advantages offered by multiple, complementary approaches.
+The positive responses of \textit{Poa secunda} and \textit{P.
+pseudoroegneria} to removals suggests our observational approach might have underestimated intrinsic growth rates after all.
+But removal experiments fall into a gray area between the observational and experimental approach.
+They are an experimental manipulation, but they rely on natural spatial variation in composition.
+We can't remove a plant from a location where it does not occur.
+The reliance on natural variation weakens inference, leaving nagging questions about pre-existing differences among quadrats.
+We could have avoided those concerns by conducting a response surface experiment, planting our study species in randomly selected locations.
+}
+
+\new{We also learned that partial removals, rather than complete removal treatments, could 
+have helped us to distinguish between competition-mediated and density-independent effects of removal treatments (e.g. Fig. \ref{fig:PSSPresids}).
+Partial removals, or experimental gradients in densities, would give us inference about the competition coefficients themselves, whereas our 
+design lumped all treatment effects into the intercept terms.
+Complete removals might be adequate for testing the presence of competition, but do not work as well for quantifying per capita competitive effects.
+Furthermore, while our experiment focused on a few interspecific interactions, ultimately it is the difference in the strength of intra- and interspecific interactions which determines the stability of coexistence and the degree to which the dynamics of co-occurring species are coupled.
+Next time we will also include intraspecific removal treatments.
+}
+
+\new{The difficulty we experienced interpreting our results also made us appreciate the value of mechanistic insight.
+Even if the removal experiment had perfectly validated the predictions of our phenomenological baseline model, we suspect we would still be asking ourselves, how can interspecific interactions really be so weak?
+The only way to rigorously answer that question is to study the factors--resources, natural enemies, environmental constraints--limiting population growth.
+}
+
+Although our results point to limitations of model based on long-term observational data, they also give some cause for optimism about population forecasting.
 \citet{Crone2013} found that matrix projection models for plants often failed to forecast population trends due
@@
 random year effects.
+
+\subsection*{Conclusion}
+
+Experimental removal of \textit{A.
+tripartita} led to greater release of grasses than our previous model predicted.
+However, the mechanisms driving this  release remain unclear, and make us hesitant to conclude that our model based on observational data underestimates the strength of interspecific competition.
+The results do increase our interest in testing model predictions using even more tightly controlled experiments as well as direct study of resource-use.
 

### 2018-01-26 removals/growth/growth_residuals_by_individ.R

@@ -43,7 +43,7 @@ for(i in 1:4){
   
 }
 
-mtext(side=1, "Pre-treatment crowding from removal species (%)",outer=T,line=0.5,cex=1.1)
+mtext(side=1, "Pre-treatment crowding from removal species",outer=T,line=0.5,cex=1.1)
 mtext(side=2, "Growth residuals",outer=T,line=0.5, cex=1.1)
 
 dev.off()

### 2018-01-27 removals/manuscript/removal_ms_SPE.tex
comments on ms

--- 2018-01-26 removals/manuscript/removal_ms.tex
@@
 %%%
+
+\graphicspath{ {figures/} }
 
@@
 \newcommand{\new}{\textcolor{red}}
+\newcommand{\spe}{\textcolor{blue}}
 \newcommand{\comment}{\textcolor{black}}
@@
 For growth, the removal treatment effect was significant and positive for two species, \textit{Poa secunda} and \textit{Pseudoroegneria spicata}, indicating that the baseline model underestimated interspecific competition.
-For all three grass species, population models based on the vital rate regressions that included removal effects projected 1.4 to 3-fold increases in equilibrium population size relative to the baseline model (no removal effects).} However, including information about pretreatment cover of \textit{A.
-tripartita} at the quadrat or individual scale failed to improve the growth regressions, raising questions about the mechanisms driving the positive response to removal.
+For all three grass species, population models based on the vital rate regressions that included removal effects projected 1.4 to 3-fold increases in equilibrium population size relative to the baseline model (no removal effects).}
+%However, including information about pretreatment cover of \textit{A. tripartita} at the quadrat or individual scale failed to 
+improve the growth regressions
+\spe{However, we found no evidence of higher response to removal in quadrats with higher pretreatment cover of \textit{A.
+tripartita}, or by plants with higher pre-treatment crowding by \textit{A.
+tripartita} individuals,}
+raising questions about the mechanisms driving the positive response to removal.
 \new{While the experimental results showed greater competitive release than our baseline models predicted, we will need more tightly controlled experiments focused on underlying mechanisms to conclusively validate or reject our model's predictions.}
@@
 spicata} growth suggest that models based on observational data may underestimate competitive release.
-We conducted additional analyses to further explore removal effects on the growth of these two species.
+We conducted additional analyses to further explore \spe{this hypothesis.}
+%removal effects on the growth of these two species. 
 First, \new{we asked whether increases in grass growth following removal were greatest in quadrats or neighborhoods where we removed the most \textit{A.
@@
 tripartita}.
-No species showed the positive relationship we expected to see (Fig.
+No species showed the positive relationship
+\spe{that would be expected} 
+% we expected to see 
+(Fig.
 \ref{fig:GrowthResidsQuad}).
 In fact, \textit{Poa secunda} residuals were significantly lower in quadrats with high pre-treatment cover of \textit{A.
-tripartita} (individuals in high cover quadrats are growing more slowly than the growth model predicted).
+tripartita} (individuals in high pre-treatment cover quadrats grew more slowly than the growth model
+\spe{including the removal effect} predicted).
 We repeated this analysis at the individual plant level for grasses present in 2011.
@@
 not just additional observations, further increases the potential to confront the model with a novel situation outside the historical range of variation.
-\new{We expected that upon completion of this study, we would have clear evidence for the success or failure of our previous model's predictions.
+\spe{We expected this study to provide} \new{clear evidence for the success or failure of our previous model's predictions.
 Unfortunately, our results do not provide that clarity.
-In the sections below, we first summarize the evidence that our experiment revealed greater competitive release than our baseline model predicted, we then explain weaknesses in the experimental results that complicate the interpretation, and we end by discussing general lessons for the study of interspecific interactions.}
+In the sections below, we first summarize the evidence that our experiment revealed greater competitive release than our baseline model predicted, then explain weaknesses in the experimental results that complicate the interpretation, and end by discussing general lessons for the study of interspecific interactions.}
 
@@
 If \textit{A.
-tripartita} removal is releasing grasses from competition, shouldn't this effect be larger for grasses located in areas with high \textit{A.
+tripartita} removal is releasing grasses from competition, shouldn't this effect be larger for grass genets in areas with high \textit{A.
 tripartita} cover?
@@
 tripartita's} roots extend far enough beyond its canopy to compete for resources with grasses located in interspaces.
-While this might explain the lack of the relationship at the neighbrhood scale, it seems unlikely to explain the quadrat-level results.
-}
+While this might explain the lack of the relationship at the neighborhood scale, it seems unlikely to explain the quadrat-level results.}
 An alternative explanation involves pre-treatment differences in composition between control plots and  the \textit{A.
@@
 tripartita} cover as a sensitive indicator of edaphic conditions.
-A final possibility is that our removal treatment influenced grass growth through some other mechanism, unrelated to competition.
-Maybe the loss of shrub cover benefits grasses by reducing small mammal herbivory \citep{Orrock2010}?
-Without a better understanding of the mechanism driving increased grass growth in the shrub removal quadrats, we are hesitant to dramatically adjust our prior understanding gained from models fit to decades of observational data.
-}
+A final possibility is that our removal treatment influenced grass growth through some other mechanism, unrelated to competition.} 
+\spe{For example, loss of shrub cover might benefits grasses by reducing small mammal herbivory \citep{Orrock2010}. 
+Without a better understanding of the mechanism driving increased grass growth in the shrub removal quadrats, 
+we are hesitant to dramatically adjust our prior understanding gained from models fit to decades of observational data.  }
 
@@
 \new{A variety of approaches are available for studying interspecific interactions among plants, and each has its advantages and disadvantages.
-Observational datasets have the advantage of being realistic: they reflect all the processes and interactions that drive community dynamics across a range of spatial and temporal scales.
+Observational datasets have the advantage of \spe{realism}: they reflect all the processes and interactions that drive community dynamics across a range of spatial and temporal scales.
 Their disadvantage is weak inference.
@@
 
-\new{ Our previous work was based purely on observational data, so we were particularly concerned about the extrapolation issue.
+\new{Our previous work was based purely on observational data, so we were particularly concerned about the extrapolation issue.
 Specifically, our model projects the performance of individual plants growing in the absence of neighbors.
 Although the observational data set seems to include such situations (Fig.
-\ref{fig:Wscatters}), if these isolated individual are still interacting with neighbors belowground, our model might underestimate the true intrinsic growth rates.
+\ref{fig:Wscatters}), if these isolated individual are still interacting with neighbors belowground, our model might underestimate the true intrinsic growth rates \spe{that would occur when near and distant neighbors are removed.}
+\spe{(SPE: I'm not sure if readers will understand the issue here. I think the point you're making is that if our competition kernels
+under-estimate the strength of long-range interactions, they will underestimate the effect of removing all ARTR from a quadrat. 
+And because they are based on observational data, the estimated kernels might do that. Maybe near-neighbor and far-neighbor density 
+are confounded due to quadrat-to-quadrat variation, so it appears that only near neighbors matter. Or maybe because 
+near neighbors matter most, the model that optimally balances
+realism and parsimony (to optimize prediction) only based predictions on close neighbors.)} 
 Therefore, we conducted the removal experiment, hoping to gain the advantages offered by multiple, complementary approaches.
@@
 pseudoroegneria} to removals suggests our observational approach might have underestimated intrinsic growth rates after all.
-But removal experiments fall into a gray area between the observational and experimental approach.
+But removal experiments fall into a gray area between \spe{an observational study and a randomized, controlled experiment}.
 They are an experimental manipulation, but they rely on natural spatial variation in composition.
@@
 The reliance on natural variation weakens inference, leaving nagging questions about pre-existing differences among quadrats.
-We could have avoided those concerns by conducting a response surface experiment, planting our study species in randomly selected locations.
-}
+We could have avoided those concerns by conducting a response surface experiment, planting our study species in randomly selected locations.}
 
 \new{We also learned that partial removals, rather than complete removal treatments, could 
-have helped us to distinguish between competition-mediated and density-independent effects of removal treatments (e.g. Fig. \ref{fig:PSSPresids}).
+have helped to distinguish between competition-mediated and \spe{other} effects of removal treatments (e.g. Fig. \ref{fig:PSSPresids}).
 Partial removals, or experimental gradients in densities, would give us inference about the competition coefficients themselves, whereas our 
@@
 
-\new{The difficulty we experienced interpreting our results also made us appreciate the value of mechanistic insight.
+\new{Our difficulties with interpreting our results also made us appreciate the value of mechanistic insight.
 Even if the removal experiment had perfectly validated the predictions of our phenomenological baseline model, we suspect we would still be asking ourselves, how can interspecific interactions really be so weak?
@@
 
-Although our results point to limitations of model based on long-term observational data, they also give some cause for optimism about population forecasting.
+Although our results point to limitations of models based on long-term observational data, they also give some cause for optimism about population forecasting.
 \citet{Crone2013} found that matrix projection models for plants often failed to forecast population trends due
-to environmental differences between the data collection and forecast periods. Our population models 
-performed quite well, after accounting for year-to-year environmental variation through the inclusion of
-random year effects.
+to environmental differences between the data collection and forecast periods. 
+%Our population models performed quite well after accounting for year-to-year environmental variation through the inclusion of
+%random year effects. 
+\spe{By including random year-effects to account for year-to-year environmental variation, our population models performed quite well
+at forecasting between-year population changes (Figure \ref{fig:obsVpred}). }
 
@@
 \centering
-\includegraphics[width=1\textwidth]{growth_residuals_vs_quad_removal_cover}
+%\includegraphics[width=1\textwidth]{growth_residuals_vs_quad_removal_cover}
 \caption{The relationship between residuals from the fitted growth model and quadrat-level pre-treatment cover of the removal species (all grasses removed in quadrats where we studied \textit{A.
@@
 \centering
-\includegraphics[width=1\textwidth]{growth_residuals_vs_Wremoval}
+%\includegraphics[width=1\textwidth]{growth_residuals_vs_Wremoval}
 \caption{The relationship between residuals from the fitted growth model and individual-level pre-treatment crowding exerted by the removal species (all grasses in quadrats where we studied \textit{A.

### 2018-01-28 removals/manuscript/removal_ms_SPE.tex
removal ms. edits

@@
 tripartita}, or by plants with higher pre-treatment crowding by \textit{A.
-tripartita} individuals,}
+tripartita},}
 raising questions about the mechanisms driving the positive response to removal.
@@
 No species showed the positive relationship
-\spe{that would be expected} 
+\spe{that would be expected (Fig. \ref{fig:GrowthResidsQuad}).} 
 % we expected to see 
-(Fig.
-\ref{fig:GrowthResidsQuad}).
 In fact, \textit{Poa secunda} residuals were significantly lower in quadrats with high pre-treatment cover of \textit{A.
@@
 tripartita} for each of those individuals, then plotted their growth residuals against crowding for all years in which each plant survived.
-We found no significant relationships between the growth residuals and pre-treatment crowding from \textit{A.
+We found no significant relationships between growth residuals and pre-treatment crowding from \textit{A.
 tripartita} (Fig.
@@
 }
+
+\begin{figure}[tbp]
+\centering
+%\includegraphics[width=1\textwidth]{growth_residuals_vs_quad_removal_cover}
+\caption{The relationship between residuals from the fitted growth model and quadrat-level pre-treatment cover of the removal species (all grasses removed in quadrats where we studied \textit{A.
+tripartita}, and \textit{A.
+tripartita} removed in the quadrats where we studied the grasses).
+Linear regressions are shown only when statistically significant.
+Source file:  \texttt{growth\textbackslash growth\textunderscore residuals\textunderscore by\textunderscore quad.R}.}
+\label{fig:GrowthResidsQuad}
+\end{figure}
 
@@
 Second, we evaluated the possibility that removal effects might be transient.
-We removed aboveground plant material, but roots of the removed plants remain in the ground and decompose, potentially creating a flush of mineral nutrients.
+We removed aboveground plant material, but roots of the removed plants remained in the ground potentially creating a flush of mineral nutrients as they decompose.
 We tested for a transient removal effect by adding a year-by-treatment interaction.
@@
 \ref{fig:IPMresults}).
-Running a version of the model with maximum removal treatment effects (removal coefficients set to the 95\% credible limit furthest from zero), resulted in a much larger projected increase in \textbf{H.
+Running a version of the model with maximum removal treatment effects (removal coefficients set to the 95\% credible limit furthest from zero), resulted in a much larger projected increase in \textit{H.
 comata}, reflecting its small sample size and corresponding high uncertainty, but little change in the projections for \textit{Poa secunda} and \textit{P.
@@
 
-\begin{figure}[h]
-\centering
-%\includegraphics[width=1\textwidth]{growth_residuals_vs_quad_removal_cover}
-\caption{The relationship between residuals from the fitted growth model and quadrat-level pre-treatment cover of the removal species (all grasses removed in quadrats where we studied \textit{A.
+%\begin{figure}[h]
+%\centering
+%%\includegraphics[width=1\textwidth]{growth_residuals_vs_quad_removal_cover}
+%\caption{The relationship between residuals from the fitted growth model and quadrat-level pre-treatment cover of the removal species (all grasses removed in quadrats where we studied \textit{A.
 tripartita}, and \textit{A.
@@
 Source file:  \texttt{growth\textbackslash growth\textunderscore residuals\textunderscore by\textunderscore quad.R}.}
-\label{fig:GrowthResidsQuad}
-\end{figure}
+%\label{fig:GrowthResidsQuad}
+%\end{figure}
 

### 2018-02-01 removals/Manuscript/removal_ms_revision1.tex
Incorporated Ellner and Hooker edits into manuscript

--- 2018-01-24 removals/manuscript/removal_ms.tex
@@
 \usepackage[usenames]{color}
-\newcommand{\new}{\textcolor{red}}
+\newcommand{\new}{\textcolor{black}}
 \newcommand{\comment}{\textcolor{black}}
@@
 
-Stable coexistence requires intraspecific limitations to be stronger than interspecific limitations.
+\new{Stable coexistence requires intraspecific limitations to be stronger than interspecific limitations.
 The greater the difference between intra- and interspecific limitations, the more stable the coexistence, and the weaker the competitive release any species should experience following removal of competitors.
-We conducted a removal experiment to test whether a previously estimated model, showing surprisingly weak interspecific competition for four dominant species in a sagebrush steppe, accurately predicts competitive release.
-Our treatments were 1) removal of all perennial grasses and 2) removal of the dominant shrub, \textit{Artemisia tripartita}.
-We regressed survival, growth and recruitment on the locations, sizes, and species identities of neighboring plants, along with an indicator variable for removal treatment.
-If our ``baseline" regression model, which accounts for local plant-plant interactions, accurately explains the observed responses to removals, then the removal coefficient should be non-significant.
-\new{For survival and recruitment, the removal coefficients were never significantly different from zero.} For growth, the removal treatment effect was significant and positive for two species, \textit{Poa secunda} and \textit{Pseudoroegneria spicata}, indicating that the baseline model underestimated interspecific competition.
-However, including information about pretreatment cover of \textit{A.
-tripartita} at the quadrat or individual scale failed to improve the growth regressions, raising questions about the mechanisms driving the positive response to removal.
+We conducted a removal experiment to test whether a previously estimated model, showing surprisingly weak interspecific competition for four dominant species in a sagebrush steppe, accurately predicts competitive release.} Our treatments were 1) removal of all perennial grasses and 2) removal of the dominant shrub, \textit{Artemisia tripartita}.
+We \new{regressed} survival, growth and recruitment on the locations, sizes, and species identities of neighboring plants, along with an indicator variable for removal treatment.
+If our ``baseline" \new{regression} model, which accounts for local plant-plant interactions, accurately explains the observed responses to removals, then the removal coefficient should be non-significant.
+For survival and recruitment, the removal coefficients were never significantly positive, indicating that the baseline model did not underestimate interspecific competition.
+For \textit{Poa secunda} recruitment, the removal coefficient was negative, indicating that our baseline model overestimated interspecific competition.
+For growth, the removal treatment effect was significant and positive for two species, \textit{Poa secunda} and \textit{Pseudoroegneria spicata}, indicating that the baseline model underestimated interspecific competition.
+However, including information about the location of grass individuals with respect to removed \textit{A.
+tripartita} failed to improve the growth regressions, raising questions about the mechanisms driving the positive response to removal.
 For three species, population projections based on the vital rate regressions showed little impact of removals.
@@
 The ability of models based on observational data to predict population responses to experimental removals increased our confidence that interspecific interactions are in fact weak for at least three of our four study species.
-The implication is that factors besides interspecific feedbacks play important roles in determining the structure and dynamics of this community.
+The implication is that factors besides interspecific \new{feedbacks} play important roles in determining the structure and dynamics of this community.
 The success of the models also supports the use of multispecies models based on long-term observational data for characterizing competitive interactions.
@@
 
-Stable coexistence of competitors requires that intraspecific interactions must limit populations more than interspecific interactions \citep{chesson_mechanisms_2000}.
-When studying populations of species that appear to be coexisting stably, such as populations that have co-occurred in close proximity for extended periods, one should expect intraspecific limitations to be stronger than interspecific limitations.
-However, interspecific interactions might still be strong and negative.
+Stable coexistence \new{of competitors} requires that intraspecific interactions must limit populations more than interspecific interactions \citep{chesson_mechanisms_2000}.
+\new{When studying} populations of species that appear to be coexisting stably, such as populations that have co-occurred in close proximity for extended periods, \new{one} should expect intraspecific limitations to be stronger than interspecific limitations.
+However, interspecific interactions \new{might still} be strong and negative.
 In fact, plant community ecologists often motivate research on coexistence by noting that all plant species compete for water, light, and a few mineral resources (e.g.
@@
 In some cases, competitive release was complete: remaining species fully compensated for removed biomass \citep{leps_nutrient_1999,jutila_effects_2002}.
-In other cases, some competitive release occurred but the degree of compensation depended on the functional group or dominance of the removed and remaining species \citep{sala_resource_1989,belsky_effects_1992,smith_dominant_2003}.
+In other cases, \new{some competitive release occurred but the degree of compensation} depended on the functional group or dominance of the removed and remaining species \citep{sala_resource_1989,belsky_effects_1992,smith_dominant_2003}.
 Finally, even negative effects of removals have been observed \citep{keddy_effects_1989,gilbert_dominant_2009}, consistent with the recent facilitation literature.
@@
 
-The idea that results from observational studies of competition represent hypotheses which should be tested experimentally is not new \citep{Hairston1980}, but such tests remain extremely rare.
+\new{The idea that results from observational studies of competition represent hypotheses which should be tested experimentally is not new \citep{Hairston1980}, but such tests remain extremely rare.
 In fact, independent tests of any kind of quantitative prediction are rare in ecology, spurring recent calls for a greater emphasis on prediction \citep{Mouquet2015,Petchey2015,Houlahan2017}.
-These authors argue that testing predictions will not only advance applied goals such as forecasting ecological impacts of global change, but will also accelerate progress in basic research.
-
-In the spirit of this argument, we conducted a removal experiment to test whether our sagebrush steppe community model, which is based on observational data, underestimates the strength of interspecific competition and competitive release.
+These authors argue that testing predictions will not only advance applied goals such as forecasting ecological impacts of global change, but will also accelerate progress in basic research.}
+
+\new{In the spirit of this argument,} we conducted a removal experiment to test whether our sagebrush steppe community model, which is based on observational data, underestimates the strength of interspecific competition and competitive release.
 We studied the responses of three perennial grass species to removal of sagebrush, and the response of sagebrush to removal of all grasses.
-The primary purpose of the experiment was not to directly measure competitive release, which is difficult to do over a short time period in a semiarid ecosystem with low and variable plant cover, but rather to test whether our model accurately characterizes the interspecific interactions that determine the magnitude of competitive release.
+The primary purpose of the experiment was not to directly measure competitive release, which is difficult to do over a short time period in a semiarid ecosystem with low and variable plant cover, but rather \new{to test whether our model accurately characterizes} the interspecific interactions that determine the magnitude of competitive release.
 We addressed two questions: First, are the survival, growth or recruitment of our target species significantly different in control vs.
@@
 Because our demographic models include the effects of neighborhood crowding, they should be able to predict how experimental removal of neighbors alters demographic rates.
-If the full effect of experimental removals is stronger than predicted, it means that our statistical models underestimate the response to decreased crowding.
+If the full effect of experimental removals is stronger than predicted, it means that our statistical models underestimate \new{the response to decreased crowding}.
 Our second question integrates across the full life cycle: how do short and long-term population-level responses to removals compare to projections from multispecies population models based on observational data?
@@
 
-Note that the ``baseline" and ``removal" models are not estimated separately; we estimate all parameters together, but when we generate predictions from the baseline model, 
+\new{Note that the ``baseline" and ``removal" models are not estimated separately; we estimate all parameters together, but when we generate predictions from the baseline model, 
 we simply set the removal treatment effects to zero. We chose this over an alternative approach of fitting a model only to data from control plots and then making 
@@
 As a result, if, for example, one of the grasses responds more positively to removal of \textit{A. tripartita} than our baseline model predicts, we cannot 
-determine whether this reflects a poorly estimated competition coefficient, or some other effect of the removal treatment unrelated to competitor densities.
+determine whether this reflects a poorly estimated competition coefficient, or some other effect of the removal treatment unrelated to competitor densities.}
 
@@
 
-Although all the parameters in our model are estimated with uncertainty, we chose not to propage parameter uncertainty through these 
+\new{Although all the parameters in our model are estimated with uncertainty, we chose not to propage parameter uncertainty through these 
 simulations because we wanted to maximize our ability to detect the population-level consequences of a small removal effect. 
@@
 Specifically, we used the recruitment regression to calculate the expected number of new recruits at time $t+1$, which we then multiplied 
-by the observed mean size of recruits for the focal species to get expected cover of new recruits.
+by the observed mean size of recruits for the focal species to get expected cover of new recruits. }
 
@@
 We simulated the IPM to estimate equilibrium abundances under three different scenarios.
-In all of these simulations, we set the spatial random effects to zero, meaning that we are simulating dynamics on a hypothetical, average site.
-To incorporate temporal environmental variation, at each time step we randomly selected one of the 30 sets of random year effects for the survival, growth and recruitment models (``kernel selection'' in the terminology of \citealt{metcalf_statistical_2015}).
+\new{In all of these simulations, we set the spatial random effects to zero, meaning that we are simulating dynamics on a hypothetical, average site.} To incorporate temporal environmental variation, at each time step we randomly selected one of the 30 sets of random year effects for the survival, growth and recruitment models (``kernel selection'' in the terminology of \citealt{metcalf_statistical_2015}).
 For each scenario, we initialized all species with very low cover and ran the model for 500 time steps (years) to allow the community to reach its equilibrium (i.e., its steady-state pattern of fluctuations in response to environmental variability).
-We calculated the average steady-state cover for each species by simulating an additional 2000 time steps and averaging each species' cover over this period.
-The first of the three scenarios featured all four species and our baseline model for control plots (no removal effects).
+We calculated the \new{average steady-state cover for each species by simulating an additional 2000 time steps and averaging each species' cover over this period.} The first of the three scenarios featured all four species and our baseline model for control plots (no removal effects).
 For the second scenario, we simulated our experimental removals by setting either the cover of \textit{A.
@@
  meaning that individuals of these species grew more rapidly in removal plots than controls even after accounting for neighborhood interactions 
- (Fig.
-\ref{fig:VitalRateTest}; Tables \ref{ARTRgrowth} to \ref{PSSPgrowth}).
-\new{ We could not detect significant removal treatment effects on recruitment for any species (Fig.
-\ref{fig:VitalRateTest}; Table \ref{table:recruitment}).}
+ (Fig. \ref{fig:VitalRateTest}; Tables \ref{ARTRgrowth} to \ref{PSSPgrowth}). In contrast, \textit{P. secunda} had lower recruitment in removal 
+ than controls plots (upper limit of 95\% CI $<0$; Fig. \ref{fig:VitalRateTest}; Table \ref{table:recruitment}). For the other species, 
+ we could not detect significant removal treatment effects on recruitment.
 
@@
 If removal is releasing grasses from \textit{A.
-tripartita} competition for light or soil resources, we would expect individuals located under the shrub canopies to experience a greater effect of removal than individuals further away.
+tripartita} competition \new{for light or soil resources}, we would expect individuals located under the shrub canopies to experience a greater effect of removal than individuals further away.
 However, the location of individuals with respect to \textit{A.
@@
 
-Third, responses to removals might be poorly predicted by models based on observational data if the removals reduce plant densities (crowding) below the range of historical variation, or if the removals result in combinations of neighborhood-scale densities that deviate from naturally occurring patterns of correlation.
+Third, responses to removals might be poorly predicted by models based on observational data if the removals reduce plant densities (crowding) below the range of historical variation, or if the removals \new{result in combinations of neighborhood-scale densities that deviate from naturally occurring patterns of correlation}.
 However, the distribution of crowding observed within control plots easily spans the crowding values observed in removal plots (Fig.
@@
 \ref{fig:W-by-treatment}).
-These pre-existing differences in composition raise the possibility that plots selected for the removal treatments might differ from the control plots in some unobserved covariate that could influence plant composition and performance.
+These pre-existing differences in composition raise the possibility that plots selected for the removal treatments might differ from the control plots in some unobserved covariate that could influence plant performance.
 
@@
 Red lines and symbols show one-step-ahead predictions from an IBM that includes removal treatment coefficients.
-For \textit{H.
-comata}, lines showing predictions for the baseline and removal effect models are virtually indistinguishable.
-Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r}.
+\new{For \textit{H.
+comata}, lines showing predictions for the baseline and removal effect models are virtually indistinguishable.} Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r}.
 }
@@
 \ref{fig:IPMresults}).
-In fact, projected average cover of all three grass species declines slightly following \textit{A.
+In fact, projected average cover of \textit{P.
+secunda} and \textit{P.
+spicata} declines slightly following \textit{A.
 tripartita} removal.
-Using a model with removal treatment coefficients, the IPM projects that \textit{A.
-tripartita} removal will increase equilibrium cover, relative to the baseline scenario, by a factor of 1.4, 1.5 and 3.0 for \textit{H.
-comata}, \textit{Poa secunda} and \textit{P.
-spicata}, respectively (Fig.
+Using a model with removal treatment coefficients, the IPM predicts little effect of shrub removal on \textit{Poa secunda} and \textit{H.
+comata}, but a 2-fold increase in the cover of \textit{P.
+spicata} (Fig.
 \ref{fig:IPMresults}).
-Running a version of the model with maximum removal treatment effects (removal coefficients set to the 95\% credible limit furthest from zero), resulted in a much larger projected increase in \textbf{H.
-comata}, reflecting its small sample size and corresponding high uncertainty, but little change in the projections for \textit{Poa secunda} and \textit{P.
+Running a version of the model with maximum removal treatment effects (removal coefficients set to the 95\% credible limit furthest from zero), resulted in the same qualitative outcome, but a larger increase in \textit{P.
 spicata} (Fig.
@@
 The increased fecundity led to a doubling of \textit{A.
-tripartita} average cover, but had essentially zero effect on the cover of the grasses (changes of less than 0.1\% cover),
-% CHECK THIS
- implying that even if shrub cover were higher its removal would still have little effect on the grasses.
-
-The baseline model predicts more dramatic competitive release of \textit{A. tripartita} following grass removal: mean cover increases from 4\% to over11\%. 
-Adding
-removal treatment effects dramatically alters the outcome, with \textit{A.
-tripartita} decreasing towards extinction.
-However, these outcomes both are driven by parameters that are estimated with very high uncertainty (Table~\ref{table:recruitment}).
+tripartita} average cover, but had essentially zero effect on the cover of the grasses (changes of less than 0.1\% cover), implying that even if shrub cover were higher its removal would still have little effect on the grasses.
+
+The baseline model predicts more dramatic competitive release of \textit{A. tripartita} following grass removal: mean cover increases from 5.4\% to 16.2\%. Adding
+removal treatment effects dramatically alters the outcome, with \textit{A. tripartita} decreasing towards extinction. However, these outcomes both are driven
+by parameters that are estimated with very high uncertainty (Table~\ref{table:recruitment}).
 The baseline model predicts an increase following grass removal because of the relatively strong negative effects of \textit{Poa secunda} and \textit{H.
@@
 The estimated effect of
-\emph{Poa} is -0.50, but the standard deviation of the estimate is 0.28 and the 95\% credible interval stops just short of zero; the 95\% credible interval for the
+\emph{Poa} is -0.66, but the standard deviation of the estimate is 0.33 and the 95\% credible interval stops just short of zero; the 95\% credible interval for the
 \textit{H. comata} effect includes values less than half of the point-estimate. The grass removal treatment effect, which leads to predicted extinction, has a credible 
@@
 
+\subsection*{An experimental test of model predictions}
+
 Our population model based on historical, observational data revealed weak interspecific interactions \citep{adler_coexistence_2010,chu_large_2015}.
@@
 data, we were suspicious of its ability to accurately describe the outcome of interspecific interactions, so we conducted a removal experiment to test it. 
-Our approach represents a strong test of a surprising prediction for two reasons. First, the prediction was already published; the experiment 
+\new{Our approach represents a strong test of a surprising prediction for two reasons. First, the prediction was already published; the experiment 
 gave us a very real opportunity to be wrong in public. Second, testing an observationally-based model with an experimental manipulation, 
-not just additional observations, further increases the potential to confront the model with a novel situation outside the historical range of variation.
-\new{We expected that upon completion of this study, we would have clear evidence for the success or failure of our previous model's predictions.
-Unfortunately, our results do not provide that clarity.
-In the sections below, we first summarize the evidence that our experiment demonstrated greater competitive release than our model predicted, we then explain why the experimental results might warrant as much skepticism as our original results, and we end by discussing general lessons for the study of interspecific competition.}
-
-\subsection*{Evidence for greater competitive release than predicted}
+not just additional observations, further increases the potential to confront the model with a novel situation outside the historical range of variation.}
+The results provide substantial support for our model's predictions, and for the weakness of interspecific interactions in this community.
 
@@
 removal is releasing \textit{P. spicata} from competition, we should expect \textit{P. spicata} individuals that were located beneath \textit{A. tripartita} 
-shrubs that we removed to respond more than individuals located away from any shrubs. Yet we found no evidence for such a difference. Perhaps 
-\textit{A. tripartita's} roots extend far enough beyond its canopy to compete for resources with grasses located in interspaces.
+shrubs that we removed to respond more than individuals located away from any shrubs. Yet we found no evidence for such a difference. \new{Perhaps 
+\textit{A. tripartita's} roots extend far enough beyond its canopy to compete for resources with grasses located in interspaces.}
 An alternative explanation involves pre-treatment differences in composition between control plots and  the \textit{A. tripartita} 
@@
 a nonlinear fashion, with little increase in production at sites with low initial sagebrush cover, 
-and accelerating responses as initial sagebrush cover passes some threshold \citep{McDaniel2005}, implying a 
+and accelerating responses as initial sagebrush cover passes some threshold \citep{McDaniel2005}, \new{implying a 
 shift in the nature of interspecific interactions, perhaps due to changes in the balance of 
-competition and facilitation or in resource supply. The cover of 
+competition and facilitation or in resource supply.} The cover of 
 sagebrush in our ungrazed quadrats may have fallen below that threshold.
 
-What general lessons can we draw about testing predictions from population models? 
+\new{What general lessons can we draw about testing predictions from population models? 
 First, partial removals, rather than complete removal treatments, could 
@@
 performed quite well, after accounting for year-to-year environmental variation through the inclusion of
-random year effects.
+random year effects.}
 
@@
   \hline
-(Intercept) & -0.2613 & 0.1344 & -0.5276 & -0.2605 & 0.0001 & -0.2588 & 0.0000 \\ 
-  logarea & 0.7093 & 0.0400 & 0.6315 & 0.7090 & 0.7885 & 0.7085 & 0.0000 \\ 
-  TreatmentNo\_grass & -0.9994 & 0.6453 & -2.2156 & -1.0182 & 0.3228 & -1.0565 & 0.0000 \\ 
-  W.ARTR & -2.8522 & 0.4187 & -3.6968 & -2.8443 & -2.0513 & -2.8284 & 0.0000 \\ 
-  W.HECO & -0.0720 & 0.0615 & -0.1856 & -0.0747 & 0.0565 & -0.0804 & 0.0000 \\ 
-  W.POSE & 0.0617 & 0.0895 & -0.1061 & 0.0588 & 0.2459 & 0.0528 & 0.0000 \\ 
-  W.PSSP & 0.0387 & 0.0475 & -0.0500 & 0.0370 & 0.1369 & 0.0335 & 0.0000 \\ 
-  W.allcov & -0.0065 & 0.0109 & -0.0282 & -0.0065 & 0.0148 & -0.0063 & 0.0000 \\ 
-  W.allpts & 0.1293 & 0.1653 & -0.1956 & 0.1294 & 0.4533 & 0.1296 & 0.0000 \\ 
+(Intercept) & -0.0207 & 0.2261 & -0.4554 & -0.0246 & 0.4367 & -0.0321 & 0.0000 \\ 
+  logarea & 0.7208 & 0.0575 & 0.6109 & 0.7195 & 0.8378 & 0.7171 & 0.0000 \\ 
+  TreatmentNo\_grass & -1.3218 & 0.7343 & -2.7293 & -1.3340 & 0.1563 & -1.3591 & 0.0000 \\ 
+  W.ARTR & -1.5214 & 0.2075 & -1.9391 & -1.5179 & -1.1237 & -1.5107 & 0.0000 \\ 
+  W.HECO & -0.0694 & 0.0423 & -0.1485 & -0.0708 & 0.0178 & -0.0738 & 0.0000 \\ 
+  W.POSE & -0.0049 & 0.0761 & -0.1469 & -0.0077 & 0.1525 & -0.0134 & 0.0000 \\ 
+  W.PSSP & 0.0281 & 0.0275 & -0.0235 & 0.0271 & 0.0849 & 0.0252 & 0.0000 \\ 
+  W.allcov & -0.0055 & 0.0083 & -0.0219 & -0.0054 & 0.0109 & -0.0054 & 0.0000 \\ 
+  W.allpts & 0.1694 & 0.1579 & -0.1386 & 0.1686 & 0.4812 & 0.1672 & 0.0000 \\ 
    \hline
@@
   \hline
-(Intercept) & 1.4753 & 0.2011 & 1.0764 & 1.4753 & 1.8736 & 1.4753 & 0.0000 \\ 
-  logarea & 1.1991 & 0.0747 & 1.0622 & 1.1954 & 1.3561 & 1.1874 & 0.0000 \\ 
-  TreatmentNo\_shrub & 0.2915 & 0.3900 & -0.4693 & 0.2898 & 1.0610 & 0.2864 & 0.0000 \\ 
-  W.ARTR & -0.0063 & 0.0030 & -0.0121 & -0.0063 & -0.0005 & -0.0063 & 0.0000 \\ 
-  W.HECO & -1.6962 & 0.1458 & -1.9894 & -1.6939 & -1.4163 & -1.6891 & 0.0000 \\ 
-  W.POSE & 0.0649 & 0.0563 & -0.0442 & 0.0644 & 0.1768 & 0.0634 & 0.0000 \\ 
-  W.PSSP & 0.0313 & 0.0360 & -0.0381 & 0.0309 & 0.1031 & 0.0301 & 0.0000 \\ 
-  W.allcov & -0.0012 & 0.0062 & -0.0135 & -0.0012 & 0.0110 & -0.0012 & 0.0000 \\ 
-  W.allpts & -0.1295 & 0.1004 & -0.3273 & -0.1293 & 0.0670 & -0.1290 & 0.0000 \\ 
+(Intercept) & 1.4189 & 0.2115 & 0.9991 & 1.4190 & 1.8371 & 1.4192 & 0.0000 \\ 
+  logarea & 1.2885 & 0.0813 & 1.1399 & 1.2843 & 1.4601 & 1.2758 & 0.0000 \\ 
+  TreatmentNo\_shrub & 0.3958 & 0.4175 & -0.4193 & 0.3942 & 1.2190 & 0.3909 & 0.0000 \\ 
+  W.ARTR & -0.0064 & 0.0028 & -0.0120 & -0.0064 & -0.0009 & -0.0064 & 0.0000 \\ 
+  W.HECO & -0.6647 & 0.0673 & -0.7996 & -0.6638 & -0.5351 & -0.6619 & 0.0000 \\ 
+  W.POSE & 0.0732 & 0.0534 & -0.0304 & 0.0728 & 0.1790 & 0.0720 & 0.0000 \\ 
+  W.PSSP & 0.0190 & 0.0184 & -0.0167 & 0.0188 & 0.0554 & 0.0185 & 0.0000 \\ 
+  W.allcov & 0.0021 & 0.0047 & -0.0072 & 0.0021 & 0.0113 & 0.0021 & 0.0000 \\ 
+  W.allpts & -0.1185 & 0.0919 & -0.2998 & -0.1183 & 0.0610 & -0.1177 & 0.0000 \\ 
    \hline
@@
   \hline
-(Intercept) & 1.4020 & 0.1922 & 1.0176 & 1.4031 & 1.7793 & 1.4050 & 0.0000 \\ 
-  logarea & 1.0362 & 0.0645 & 0.9139 & 1.0343 & 1.1696 & 1.0307 & 0.0000 \\ 
-  TreatmentNo\_shrub & -0.1604 & 0.1622 & -0.4784 & -0.1606 & 0.1583 & -0.1610 & 0.0000 \\ 
-  W.ARTR & 0.0000 & 0.0020 & -0.0038 & 0.0000 & 0.0039 & -0.0000 & 0.0000 \\ 
-  W.HECO & -0.0224 & 0.0208 & -0.0631 & -0.0225 & 0.0185 & -0.0227 & 0.0000 \\ 
-  W.POSE & -2.0073 & 0.1112 & -2.2290 & -2.0062 & -1.7921 & -2.0038 & 0.0000 \\ 
-  W.PSSP & 0.0468 & 0.0238 & 0.0004 & 0.0466 & 0.0939 & 0.0463 & 0.0000 \\ 
-  W.allcov & -0.0061 & 0.0035 & -0.0130 & -0.0061 & 0.0008 & -0.0061 & 0.0000 \\ 
-  W.allpts & -0.0383 & 0.0567 & -0.1494 & -0.0384 & 0.0730 & -0.0385 & 0.0000 \\ 
+(Intercept) & 1.3162 & 0.1965 & 0.9242 & 1.3171 & 1.7024 & 1.3186 & 0.0000 \\ 
+  logarea & 1.0585 & 0.0640 & 0.9365 & 1.0569 & 1.1902 & 1.0537 & 0.0000 \\ 
+  TreatmentNo\_shrub & -0.2572 & 0.1723 & -0.5949 & -0.2573 & 0.0811 & -0.2576 & 0.0000 \\ 
+  W.ARTR & 0.0001 & 0.0019 & -0.0036 & 0.0001 & 0.0038 & 0.0001 & 0.0000 \\ 
+  W.HECO & -0.0153 & 0.0138 & -0.0424 & -0.0154 & 0.0119 & -0.0154 & 0.0000 \\ 
+  W.POSE & -1.2632 & 0.0816 & -1.4260 & -1.2623 & -1.1053 & -1.2605 & 0.0000 \\ 
+  W.PSSP & 0.0249 & 0.0122 & 0.0011 & 0.0249 & 0.0490 & 0.0247 & 0.0000 \\ 
+  W.allcov & -0.0024 & 0.0026 & -0.0075 & -0.0024 & 0.0029 & -0.0024 & 0.0000 \\ 
+  W.allpts & 0.0018 & 0.0524 & -0.1009 & 0.0018 & 0.1048 & 0.0017 & 0.0000 \\ 
    \hline
@@
   \hline
-(Intercept) & 1.1978 & 0.1639 & 0.8783 & 1.1962 & 1.5257 & 1.1933 & 0.0000 \\ 
-  logarea & 1.5138 & 0.0935 & 1.3336 & 1.5122 & 1.7029 & 1.5092 & 0.0000 \\ 
-  TreatmentNo\_shrub & -0.2466 & 0.2023 & -0.6454 & -0.2460 & 0.1489 & -0.2449 & 0.0000 \\ 
-  W.ARTR & 0.0099 & 0.0022 & 0.0057 & 0.0099 & 0.0142 & 0.0099 & 0.0000 \\ 
-  W.HECO & 0.0025 & 0.0289 & -0.0550 & 0.0028 & 0.0586 & 0.0032 & 0.0000 \\ 
-  W.POSE & 0.0517 & 0.0389 & -0.0222 & 0.0508 & 0.1308 & 0.0490 & 0.0000 \\ 
-  W.PSSP & -1.1267 & 0.0739 & -1.2743 & -1.1258 & -0.9838 & -1.1242 & 0.0000 \\ 
-  W.allcov & 0.0131 & 0.0037 & 0.0060 & 0.0131 & 0.0203 & 0.0131 & 0.0000 \\ 
-  W.allpts & 0.0756 & 0.0658 & -0.0540 & 0.0757 & 0.2045 & 0.0760 & 0.0000 \\ 
+(Intercept) & 1.2121 & 0.1699 & 0.8815 & 1.2103 & 1.5533 & 1.2066 & 0.0000 \\ 
+  logarea & 1.5492 & 0.0976 & 1.3612 & 1.5474 & 1.7473 & 1.5437 & 0.0000 \\ 
+  TreatmentNo\_shrub & -0.1885 & 0.2208 & -0.6237 & -0.1879 & 0.2431 & -0.1867 & 0.0000 \\ 
+  W.ARTR & 0.0101 & 0.0021 & 0.0061 & 0.0101 & 0.0142 & 0.0101 & 0.0000 \\ 
+  W.HECO & 0.0046 & 0.0185 & -0.0319 & 0.0046 & 0.0406 & 0.0047 & 0.0000 \\ 
+  W.POSE & 0.0245 & 0.0384 & -0.0502 & 0.0242 & 0.1007 & 0.0236 & 0.0000 \\ 
+  W.PSSP & -0.4430 & 0.0291 & -0.5010 & -0.4427 & -0.3867 & -0.4420 & 0.0000 \\ 
+  W.allcov & 0.0100 & 0.0028 & 0.0046 & 0.0100 & 0.0155 & 0.0100 & 0.0000 \\ 
+  W.allpts & 0.0943 & 0.0605 & -0.0246 & 0.0944 & 0.2129 & 0.0944 & 0.0000 \\ 
    \hline
@@
   \hline
-(Intercept) & 0.7092 & 0.2385 & 0.2396 & 0.7091 & 1.1784 & 0.7090 & 0.0000 \\ 
-  logarea.t0 & 0.8689 & 0.0394 & 0.7912 & 0.8689 & 0.9466 & 0.8689 & 0.0000 \\ 
-  TreatmentNo\_grass & 0.1385 & 0.1446 & -0.1453 & 0.1385 & 0.4221 & 0.1385 & 0.0000 \\ 
-  W.ARTR & -0.1497 & 0.1960 & -0.5345 & -0.1497 & 0.2347 & -0.1497 & 0.0000 \\ 
-  W.HECO & 0.0045 & 0.0198 & -0.0343 & 0.0045 & 0.0433 & 0.0045 & 0.0000 \\ 
-  W.POSE & -0.0614 & 0.0308 & -0.1218 & -0.0614 & -0.0010 & -0.0614 & 0.0000 \\ 
-  W.PSSP & -0.0050 & 0.0158 & -0.0360 & -0.0050 & 0.0260 & -0.0050 & 0.0000 \\ 
-  W.allcov & -0.0052 & 0.0041 & -0.0133 & -0.0052 & 0.0029 & -0.0052 & 0.0000 \\ 
-  W.allpts & 0.0115 & 0.0536 & -0.0938 & 0.0115 & 0.1167 & 0.0115 & 0.0000 \\ 
+(Intercept) & 0.7378 & 0.2437 & 0.2591 & 0.7378 & 1.2161 & 0.7377 & 0.0000 \\ 
+  logarea.t0 & 0.8663 & 0.0393 & 0.7892 & 0.8663 & 0.9434 & 0.8663 & 0.0000 \\ 
+  TreatmentNo\_grass & 0.1037 & 0.1436 & -0.1783 & 0.1037 & 0.3854 & 0.1037 & 0.0000 \\ 
+  W.ARTR & -0.0364 & 0.0892 & -0.2116 & -0.0364 & 0.1387 & -0.0364 & 0.0000 \\ 
+  W.HECO & 0.0033 & 0.0136 & -0.0233 & 0.0033 & 0.0299 & 0.0033 & 0.0000 \\ 
+  W.POSE & -0.0573 & 0.0268 & -0.1100 & -0.0573 & -0.0047 & -0.0573 & 0.0000 \\ 
+  W.PSSP & 0.0062 & 0.0084 & -0.0102 & 0.0062 & 0.0226 & 0.0062 & 0.0000 \\ 
+  W.allcov & -0.0055 & 0.0030 & -0.0114 & -0.0055 & 0.0004 & -0.0055 & 0.0000 \\ 
+  W.allpts & 0.0029 & 0.0477 & -0.0908 & 0.0029 & 0.0965 & 0.0029 & 0.0000 \\ 
    \hline
@@
   \hline
-(Intercept) & 0.3937 & 0.0815 & 0.2330 & 0.3938 & 0.5537 & 0.3940 & 0.0000 \\ 
-  logarea.t0 & 0.8207 & 0.0213 & 0.7788 & 0.8207 & 0.8626 & 0.8207 & 0.0000 \\ 
-  TreatmentNo\_shrub & 0.0691 & 0.1314 & -0.1891 & 0.0691 & 0.3268 & 0.0691 & 0.0000 \\ 
-  W.ARTR & -0.0042 & 0.0011 & -0.0063 & -0.0042 & -0.0020 & -0.0042 & 0.0000 \\ 
-  W.HECO & -0.2134 & 0.0618 & -0.3348 & -0.2134 & -0.0922 & -0.2135 & 0.0000 \\ 
-  W.POSE & 0.0204 & 0.0178 & -0.0146 & 0.0204 & 0.0553 & 0.0204 & 0.0000 \\ 
-  W.PSSP & -0.0373 & 0.0131 & -0.0630 & -0.0373 & -0.0117 & -0.0373 & 0.0000 \\ 
-  W.allcov & -0.0062 & 0.0027 & -0.0115 & -0.0062 & -0.0010 & -0.0062 & 0.0000 \\ 
-  W.allpts & -0.0212 & 0.0416 & -0.1028 & -0.0212 & 0.0604 & -0.0212 & 0.0000 \\ 
+(Intercept) & 0.3782 & 0.0797 & 0.2212 & 0.3783 & 0.5343 & 0.3784 & 0.0000 \\ 
+  logarea.t0 & 0.8272 & 0.0205 & 0.7869 & 0.8272 & 0.8674 & 0.8272 & 0.0000 \\ 
+  TreatmentNo\_shrub & -0.0162 & 0.1336 & -0.2785 & -0.0162 & 0.2460 & -0.0162 & 0.0000 \\ 
+  W.ARTR & -0.0031 & 0.0010 & -0.0051 & -0.0031 & -0.0010 & -0.0031 & 0.0000 \\ 
+  W.HECO & -0.0479 & 0.0195 & -0.0862 & -0.0479 & -0.0098 & -0.0479 & 0.0000 \\ 
+  W.POSE & 0.0242 & 0.0162 & -0.0077 & 0.0242 & 0.0560 & 0.0242 & 0.0000 \\ 
+  W.PSSP & -0.0151 & 0.0065 & -0.0279 & -0.0151 & -0.0023 & -0.0151 & 0.0000 \\ 
+  W.allcov & -0.0038 & 0.0019 & -0.0076 & -0.0038 & 0.0000 & -0.0038 & 0.0000 \\ 
+  W.allpts & -0.0513 & 0.0380 & -0.1258 & -0.0513 & 0.0232 & -0.0513 & 0.0000 \\ 
    \hline
@@
 \centering
-\caption{Summary of posterior distributions for fixed effects for the \textit{Poa secunda} growth model. See Table \ref{ARTRgrowth} for an explanation of coefficient names
+\caption{Summary of posterior distributions for fixed effects for the \textit{P. secunda} growth model. See Table \ref{ARTRgrowth} for an explanation of coefficient names
 and column headers.} 
@@
   \hline
-(Intercept) & 0.5069 & 0.0625 & 0.3830 & 0.5070 & 0.6299 & 0.5072 & 0.0000 \\ 
-  logarea.t0 & 0.6669 & 0.0228 & 0.6215 & 0.6670 & 0.7116 & 0.6673 & 0.0000 \\ 
-  TreatmentNo\_shrub & 0.1937 & 0.0632 & 0.0696 & 0.1937 & 0.3176 & 0.1937 & 0.0000 \\ 
-  W.ARTR & -0.0004 & 0.0009 & -0.0022 & -0.0004 & 0.0013 & -0.0004 & 0.0000 \\ 
-  W.HECO & 0.0110 & 0.0100 & -0.0087 & 0.0110 & 0.0307 & 0.0110 & 0.0000 \\ 
-  W.POSE & -0.5753 & 0.0623 & -0.6977 & -0.5753 & -0.4530 & -0.5753 & 0.0000 \\ 
-  W.PSSP & -0.0093 & 0.0114 & -0.0316 & -0.0093 & 0.0130 & -0.0093 & 0.0000 \\ 
-  W.allcov & -0.0004 & 0.0016 & -0.0035 & -0.0004 & 0.0028 & -0.0004 & 0.0000 \\ 
-  W.allpts & -0.0340 & 0.0255 & -0.0842 & -0.0340 & 0.0161 & -0.0340 & 0.0000 \\ 
+(Intercept) & 0.4923 & 0.0681 & 0.3581 & 0.4923 & 0.6259 & 0.4924 & 0.0000 \\ 
+  logarea.t0 & 0.6750 & 0.0232 & 0.6292 & 0.6750 & 0.7203 & 0.6751 & 0.0000 \\ 
+  TreatmentNo\_shrub & 0.2161 & 0.0644 & 0.0897 & 0.2161 & 0.3425 & 0.2161 & 0.0000 \\ 
+  W.ARTR & -0.0002 & 0.0008 & -0.0019 & -0.0002 & 0.0014 & -0.0002 & 0.0000 \\ 
+  W.HECO & 0.0039 & 0.0063 & -0.0084 & 0.0039 & 0.0161 & 0.0039 & 0.0000 \\ 
+  W.POSE & -0.2641 & 0.0387 & -0.3400 & -0.2641 & -0.1883 & -0.2641 & 0.0000 \\ 
+  W.PSSP & -0.0073 & 0.0056 & -0.0183 & -0.0073 & 0.0038 & -0.0073 & 0.0000 \\ 
+  W.allcov & -0.0003 & 0.0012 & -0.0026 & -0.0003 & 0.0020 & -0.0003 & 0.0000 \\ 
+  W.allpts & -0.0254 & 0.0231 & -0.0709 & -0.0254 & 0.0200 & -0.0254 & 0.0000 \\ 
    \hline
@@
   \hline
-(Intercept) & 0.3846 & 0.0710 & 0.2425 & 0.3851 & 0.5236 & 0.3858 & 0.0000 \\ 
-  logarea.t0 & 0.8259 & 0.0167 & 0.7929 & 0.8259 & 0.8587 & 0.8260 & 0.0000 \\ 
-  TreatmentNo\_shrub & 0.2262 & 0.0681 & 0.0924 & 0.2262 & 0.3598 & 0.2262 & 0.0000 \\ 
-  W.ARTR & -0.0023 & 0.0008 & -0.0038 & -0.0023 & -0.0007 & -0.0023 & 0.0000 \\ 
-  W.HECO & -0.0201 & 0.0111 & -0.0419 & -0.0201 & 0.0017 & -0.0201 & 0.0000 \\ 
-  W.POSE & -0.0200 & 0.0137 & -0.0469 & -0.0200 & 0.0069 & -0.0200 & 0.0000 \\ 
-  W.PSSP & -0.3845 & 0.0335 & -0.4502 & -0.3845 & -0.3188 & -0.3846 & 0.0000 \\ 
-  W.allcov & -0.0048 & 0.0014 & -0.0076 & -0.0048 & -0.0021 & -0.0048 & 0.0000 \\ 
-  W.allpts & -0.0510 & 0.0256 & -0.1012 & -0.0510 & -0.0008 & -0.0510 & 0.0000 \\ 
+(Intercept) & 0.3987 & 0.0660 & 0.2690 & 0.3987 & 0.5281 & 0.3988 & 0.0000 \\ 
+  logarea.t0 & 0.8249 & 0.0149 & 0.7956 & 0.8249 & 0.8543 & 0.8249 & 0.0000 \\ 
+  TreatmentNo\_shrub & 0.1948 & 0.0714 & 0.0546 & 0.1948 & 0.3349 & 0.1948 & 0.0000 \\ 
+  W.ARTR & -0.0025 & 0.0008 & -0.0041 & -0.0025 & -0.0010 & -0.0025 & 0.0000 \\ 
+  W.HECO & -0.0118 & 0.0072 & -0.0260 & -0.0118 & 0.0023 & -0.0118 & 0.0000 \\ 
+  W.POSE & -0.0067 & 0.0130 & -0.0323 & -0.0067 & 0.0189 & -0.0067 & 0.0000 \\ 
+  W.PSSP & -0.1480 & 0.0133 & -0.1740 & -0.1480 & -0.1220 & -0.1480 & 0.0000 \\ 
+  W.allcov & -0.0036 & 0.0011 & -0.0057 & -0.0036 & -0.0015 & -0.0036 & 0.0000 \\ 
+  W.allpts & -0.0522 & 0.0236 & -0.0984 & -0.0522 & -0.0060 & -0.0522 & 0.0000 \\ 
    \hline
@@
   \hline
-(Intercept) & 0.5080 & 0.0625 & 0.3841 & 0.5081 & 0.6311 & 0.5084 & 0.0000 \\ 
-  logarea.t0 & 0.6665 & 0.0228 & 0.6211 & 0.6666 & 0.7111 & 0.6669 & 0.0000 \\ 
-  TreatmentNo\_shrub & 0.2247 & 0.0744 & 0.0786 & 0.2247 & 0.3707 & 0.2247 & 0.0000 \\ 
-  W.ARTR & -0.0004 & 0.0009 & -0.0022 & -0.0004 & 0.0013 & -0.0004 & 0.0000 \\ 
-  W.HECO & 0.0109 & 0.0100 & -0.0088 & 0.0109 & 0.0306 & 0.0109 & 0.0000 \\ 
-  W.POSE & -0.5770 & 0.0624 & -0.6995 & -0.5770 & -0.4547 & -0.5770 & 0.0000 \\ 
-  W.PSSP & -0.0093 & 0.0114 & -0.0316 & -0.0093 & 0.0130 & -0.0093 & 0.0000 \\ 
-  W.allcov & -0.0004 & 0.0016 & -0.0036 & -0.0004 & 0.0027 & -0.0004 & 0.0000 \\ 
-  W.allpts & -0.0340 & 0.0256 & -0.0842 & -0.0340 & 0.0161 & -0.0340 & 0.0000 \\ 
-  inARTR & -0.0890 & 0.1126 & -0.3100 & -0.0890 & 0.1319 & -0.0890 & 0.0000 \\ 
+(Intercept) & 0.4932 & 0.0622 & 0.3700 & 0.4934 & 0.6154 & 0.4937 & 0.0000 \\ 
+  logarea.t0 & 0.6746 & 0.0220 & 0.6306 & 0.6747 & 0.7177 & 0.6750 & 0.0000 \\ 
+  TreatmentNo\_shrub & 0.2370 & 0.0739 & 0.0919 & 0.2370 & 0.3820 & 0.2370 & 0.0000 \\ 
+  W.ARTR & -0.0002 & 0.0008 & -0.0018 & -0.0002 & 0.0014 & -0.0002 & 0.0000 \\ 
+  W.HECO & 0.0038 & 0.0062 & -0.0083 & 0.0038 & 0.0160 & 0.0038 & 0.0000 \\ 
+  W.POSE & -0.2644 & 0.0383 & -0.3395 & -0.2644 & -0.1893 & -0.2644 & 0.0000 \\ 
+  W.PSSP & -0.0075 & 0.0056 & -0.0184 & -0.0075 & 0.0034 & -0.0075 & 0.0000 \\ 
+  W.allcov & -0.0003 & 0.0012 & -0.0026 & -0.0003 & 0.0020 & -0.0003 & 0.0000 \\ 
+  W.allpts & -0.0258 & 0.0229 & -0.0707 & -0.0258 & 0.0191 & -0.0258 & 0.0000 \\ 
+  inARTR & -0.0592 & 0.1133 & -0.2817 & -0.0592 & 0.1632 & -0.0592 & 0.0000 \\ 
    \hline
@@
   \hline
-(Intercept) & 0.3955 & 0.0606 & 0.2764 & 0.3953 & 0.5153 & 0.3950 & 0.0000 \\ 
-  logarea.t0 & 0.8276 & 0.0168 & 0.7943 & 0.8277 & 0.8608 & 0.8277 & 0.0000 \\ 
-  TreatmentNo\_shrub & 0.2531 & 0.0784 & 0.0991 & 0.2531 & 0.4070 & 0.2531 & 0.0000 \\ 
-  W.ARTR & -0.0027 & 0.0008 & -0.0042 & -0.0027 & -0.0011 & -0.0027 & 0.0000 \\ 
-  W.HECO & -0.0231 & 0.0110 & -0.0447 & -0.0231 & -0.0016 & -0.0231 & 0.0000 \\ 
-  W.POSE & -0.0198 & 0.0138 & -0.0469 & -0.0198 & 0.0072 & -0.0198 & 0.0000 \\ 
-  W.PSSP & -0.3600 & 0.0326 & -0.4239 & -0.3600 & -0.2961 & -0.3600 & 0.0000 \\ 
-  W.allcov & -0.0051 & 0.0014 & -0.0079 & -0.0051 & -0.0024 & -0.0051 & 0.0000 \\ 
-  W.allpts & -0.0557 & 0.0254 & -0.1055 & -0.0557 & -0.0060 & -0.0557 & 0.0000 \\ 
-  inARTR & -0.0649 & 0.1111 & -0.2830 & -0.0649 & 0.1531 & -0.0649 & 0.0000 \\ 
+(Intercept) & 0.4129 & 0.0633 & 0.2885 & 0.4128 & 0.5381 & 0.4124 & 0.0000 \\ 
+  logarea.t0 & 0.8261 & 0.0173 & 0.7919 & 0.8262 & 0.8602 & 0.8262 & 0.0000 \\ 
+  TreatmentNo\_shrub & 0.2444 & 0.0798 & 0.0878 & 0.2444 & 0.4009 & 0.2444 & 0.0000 \\ 
+  W.ARTR & -0.0029 & 0.0008 & -0.0044 & -0.0029 & -0.0015 & -0.0029 & 0.0000 \\ 
+  W.HECO & -0.0138 & 0.0068 & -0.0272 & -0.0138 & -0.0005 & -0.0138 & 0.0000 \\ 
+  W.POSE & -0.0071 & 0.0127 & -0.0319 & -0.0071 & 0.0178 & -0.0071 & 0.0000 \\ 
+  W.PSSP & -0.1372 & 0.0126 & -0.1619 & -0.1372 & -0.1125 & -0.1372 & 0.0000 \\ 
+  W.allcov & -0.0039 & 0.0010 & -0.0059 & -0.0039 & -0.0018 & -0.0039 & 0.0000 \\ 
+  W.allpts & -0.0576 & 0.0224 & -0.1017 & -0.0576 & -0.0136 & -0.0576 & 0.0000 \\ 
+  inARTR & -0.1301 & 0.1102 & -0.3465 & -0.1301 & 0.0862 & -0.1301 & 0.0000 \\ 
    \hline
@@
   \hline
-(Intercept) & 0.5079 & 0.0626 & 0.3839 & 0.5080 & 0.6310 & 0.5082 & 0.0000 \\ 
-  trtYears1 & 0.0910 & 0.1169 & -0.1386 & 0.0910 & 0.3203 & 0.0910 & 0.0000 \\ 
-  trtYears2 & 0.1560 & 0.1212 & -0.0819 & 0.1560 & 0.3937 & 0.1560 & 0.0000 \\ 
-  trtYears3 & 0.1891 & 0.1461 & -0.0976 & 0.1891 & 0.4757 & 0.1891 & 0.0000 \\ 
-  trtYears4 & 0.3125 & 0.1608 & -0.0032 & 0.3125 & 0.6280 & 0.3126 & 0.0000 \\ 
-  trtYears5 & 0.3737 & 0.1782 & 0.0239 & 0.3737 & 0.7233 & 0.3737 & 0.0000 \\ 
-  logarea.t0 & 0.6663 & 0.0228 & 0.6207 & 0.6664 & 0.7111 & 0.6666 & 0.0000 \\ 
-  W.ARTR & -0.0005 & 0.0009 & -0.0022 & -0.0005 & 0.0013 & -0.0005 & 0.0000 \\ 
-  W.HECO & 0.0108 & 0.0100 & -0.0089 & 0.0108 & 0.0305 & 0.0108 & 0.0000 \\ 
-  W.POSE & -0.5725 & 0.0624 & -0.6950 & -0.5725 & -0.4501 & -0.5725 & 0.0000 \\ 
-  W.PSSP & -0.0100 & 0.0114 & -0.0323 & -0.0100 & 0.0123 & -0.0100 & 0.0000 \\ 
-  W.allcov & -0.0004 & 0.0016 & -0.0036 & -0.0004 & 0.0028 & -0.0004 & 0.0000 \\ 
-  W.allpts & -0.0347 & 0.0256 & -0.0849 & -0.0347 & 0.0155 & -0.0347 & 0.0000 \\ 
+(Intercept) & 0.4963 & 0.0616 & 0.3752 & 0.4963 & 0.6172 & 0.4964 & 0.0000 \\ 
+  trtYears1 & 0.0798 & 0.1272 & -0.1699 & 0.0798 & 0.3293 & 0.0798 & 0.0000 \\ 
+  trtYears2 & 0.2425 & 0.1350 & -0.0225 & 0.2425 & 0.5073 & 0.2425 & 0.0000 \\ 
+  trtYears3 & 0.2512 & 0.1554 & -0.0539 & 0.2512 & 0.5561 & 0.2512 & 0.0000 \\ 
+  trtYears4 & 0.2988 & 0.1693 & -0.0337 & 0.2988 & 0.6310 & 0.2988 & 0.0000 \\ 
+  trtYears5 & 0.4787 & 0.1845 & 0.1164 & 0.4787 & 0.8407 & 0.4787 & 0.0000 \\ 
+  logarea.t0 & 0.6731 & 0.0234 & 0.6270 & 0.6732 & 0.7191 & 0.6732 & 0.0000 \\ 
+  W.ARTR & -0.0003 & 0.0009 & -0.0019 & -0.0003 & 0.0014 & -0.0003 & 0.0000 \\ 
+  W.HECO & 0.0035 & 0.0064 & -0.0090 & 0.0035 & 0.0160 & 0.0035 & 0.0000 \\ 
+  W.POSE & -0.2624 & 0.0394 & -0.3398 & -0.2624 & -0.1851 & -0.2624 & 0.0000 \\ 
+  W.PSSP & -0.0080 & 0.0057 & -0.0192 & -0.0080 & 0.0032 & -0.0080 & 0.0000 \\ 
+  W.allcov & -0.0004 & 0.0012 & -0.0027 & -0.0004 & 0.0020 & -0.0004 & 0.0000 \\ 
+  W.allpts & -0.0267 & 0.0236 & -0.0730 & -0.0267 & 0.0196 & -0.0267 & 0.0000 \\ 
+  inARTR & -0.0563 & 0.1169 & -0.2859 & -0.0563 & 0.1730 & -0.0563 & 0.0000 \\ 
    \hline
@@
   \hline
-(Intercept) & 0.3948 & 0.0606 & 0.2757 & 0.3946 & 0.5149 & 0.3943 & 0.0000 \\ 
-  trtYears1 & 0.3434 & 0.1341 & 0.0800 & 0.3434 & 0.6065 & 0.3434 & 0.0000 \\ 
-  trtYears2 & -0.0514 & 0.1391 & -0.3245 & -0.0514 & 0.2215 & -0.0513 & 0.0000 \\ 
-  trtYears3 & 0.3547 & 0.1493 & 0.0616 & 0.3547 & 0.6476 & 0.3547 & 0.0000 \\ 
-  trtYears4 & 0.5487 & 0.1519 & 0.2505 & 0.5487 & 0.8467 & 0.5487 & 0.0000 \\ 
-  trtYears5 & -0.0575 & 0.1598 & -0.3711 & -0.0575 & 0.2560 & -0.0575 & 0.0000 \\ 
-  logarea.t0 & 0.8284 & 0.0168 & 0.7951 & 0.8284 & 0.8614 & 0.8285 & 0.0000 \\ 
-  W.ARTR & -0.0026 & 0.0008 & -0.0042 & -0.0026 & -0.0011 & -0.0026 & 0.0000 \\ 
-  W.HECO & -0.0234 & 0.0109 & -0.0447 & -0.0234 & -0.0020 & -0.0234 & 0.0000 \\ 
-  W.POSE & -0.0195 & 0.0137 & -0.0463 & -0.0195 & 0.0073 & -0.0195 & 0.0000 \\ 
-  W.PSSP & -0.3590 & 0.0322 & -0.4223 & -0.3590 & -0.2957 & -0.3590 & 0.0000 \\ 
-  W.allcov & -0.0053 & 0.0014 & -0.0081 & -0.0053 & -0.0026 & -0.0053 & 0.0000 \\ 
-  W.allpts & -0.0567 & 0.0251 & -0.1060 & -0.0567 & -0.0074 & -0.0567 & 0.0000 \\ 
+(Intercept) & 0.4122 & 0.0626 & 0.2892 & 0.4120 & 0.5364 & 0.4115 & 0.0000 \\ 
+  trtYears1 & 0.3256 & 0.1440 & 0.0428 & 0.3256 & 0.6081 & 0.3257 & 0.0000 \\ 
+  trtYears2 & -0.0353 & 0.1467 & -0.3235 & -0.0353 & 0.2525 & -0.0353 & 0.0000 \\ 
+  trtYears3 & 0.4829 & 0.1586 & 0.1714 & 0.4829 & 0.7940 & 0.4829 & 0.0000 \\ 
+  trtYears4 & 0.5059 & 0.1602 & 0.1915 & 0.5059 & 0.8202 & 0.5059 & 0.0000 \\ 
+  trtYears5 & -0.0588 & 0.1682 & -0.3890 & -0.0588 & 0.2712 & -0.0588 & 0.0000 \\ 
+  logarea.t0 & 0.8266 & 0.0173 & 0.7922 & 0.8267 & 0.8607 & 0.8268 & 0.0000 \\ 
+  W.ARTR & -0.0030 & 0.0008 & -0.0044 & -0.0030 & -0.0015 & -0.0030 & 0.0000 \\ 
+  W.HECO & -0.0139 & 0.0068 & -0.0272 & -0.0139 & -0.0005 & -0.0139 & 0.0000 \\ 
+  W.POSE & -0.0064 & 0.0127 & -0.0313 & -0.0064 & 0.0185 & -0.0064 & 0.0000 \\ 
+  W.PSSP & -0.1362 & 0.0126 & -0.1609 & -0.1362 & -0.1115 & -0.1362 & 0.0000 \\ 
+  W.allcov & -0.0040 & 0.0010 & -0.0061 & -0.0040 & -0.0020 & -0.0040 & 0.0000 \\ 
+  W.allpts & -0.0579 & 0.0224 & -0.1020 & -0.0579 & -0.0139 & -0.0579 & 0.0000 \\ 
+  inARTR & -0.1235 & 0.1104 & -0.3403 & -0.1235 & 0.0931 & -0.1235 & 0.0000 \\ 
    \hline
@@
    \hline
-
-   \hline
-    $\gamma$[1] & 0.2843 & 0.6342 & -1.0900 & 1.4950 & 1.0279 &   810 \\ 
-    $\gamma$[2] & 3.2828 & 0.4467 & 2.4429 & 4.1940 & 1.0062 &   260 \\ 
-    $\gamma$[3] & 3.2923 & 0.3691 & 2.5680 & 4.0320 & 1.0040 &  1800 \\ 
-    $\gamma$[4] & 2.9362 & 0.3811 & 2.1779 & 3.6462 & 1.0116 &   140 \\ 
-    $\chi$[2,2] & 0.0405 & 0.4077 & -0.7392 & 0.8147 & 1.0011 &  2000 \\ 
-    $\chi$[2,3] & -0.1396 & 0.2658 & -0.6654 & 0.3970 & 1.0015 &  1400 \\ 
-    $\chi$[2,4] & 0.4050 & 0.2439 & -0.0664 & 0.8924 & 1.0010 &  2000 \\ 
-    $\chi$[3,1] & -1.3691 & 0.6755 & -2.7260 & -0.0685 & 1.0093 &   170 \\ 
-      $\omega$[1,1] & -0.5539 & 0.1165 & -0.7517 & -0.2845 & 1.0329 &   240 \\ 
-      $\omega$[1,2] & 0.0654 & 0.0581 & -0.0429 & 0.1898 & 1.0025 &   740 \\ 
-      $\omega$[1,3] & 0.0043 & 0.0389 & -0.0771 & 0.0813 & 1.0019 &  2000 \\ 
-      $\omega$[1,4] & 0.0953 & 0.0410 & 0.0209 & 0.1830 & 1.0120 &   130 \\ 
-      $\omega$[2,1] & -0.4522 & 0.1533 & -0.7461 & -0.1556 & 1.0007 &  2000 \\ 
-      $\omega$[2,2] & -1.5975 & 0.1175 & -1.8260 & -1.3669 & 1.0033 &  2000 \\ 
-      $\omega$[2,3] & -0.0058 & 0.0866 & -0.1806 & 0.1662 & 1.0035 &  2000 \\ 
-      $\omega$[2,4] & -0.3047 & 0.0858 & -0.4794 & -0.1381 & 1.0025 &   740 \\ 
-      $\omega$[3,1] & -0.5030 & 0.2801 & -1.0880 & 0.0341 & 1.0185 &   120 \\ 
-      $\omega$[3,2] & -0.0586 & 0.1571 & -0.3722 & 0.2399 & 1.0053 &   320 \\ 
-      $\omega$[3,3] & -1.7052 & 0.1471 & -1.9770 & -1.3980 & 1.0006 &  2000 \\ 
-      $\omega$[3,4] & -0.0960 & 0.1304 & -0.3682 & 0.1443 & 1.0005 &  2000 \\ 
-      $\omega$[4,1] & -0.1253 & 0.2688 & -0.6245 & 0.3921 & 1.0028 &  2000 \\ 
-      $\omega$[4,2] & -0.3056 & 0.1911 & -0.7052 & 0.0560 & 1.0041 &   420 \\ 
-      $\omega$[4,3] & 0.0335 & 0.1386 & -0.2499 & 0.2999 & 1.0040 &   440 \\ 
-      $\omega$[4,4] & -1.7046 & 0.1597 & -2.0080 & -1.3900 & 1.0101 &   250 \\ 
-     $\theta$[1] & 0.6117 & 0.0708 & 0.4855 & 0.7560 & 1.0006 &  2000 \\ 
-     $\theta$[2] & 1.1862 & 0.1462 & 0.9392 & 1.5000 & 1.0014 &  2000 \\ 
-     $\theta$[3] & 1.1782 & 0.1043 & 0.9872 & 1.4080 & 1.0009 &  2000 \\ 
-     $\theta$[4] & 1.2081 & 0.1060 & 1.0090 & 1.4210 & 1.0006 &  2000 \\ 
-     $p$[1] & 0.8367 & 0.1095 & 0.5280 & 0.9604 & 1.0235 &  2000 \\ 
-     $p$[2] & 0.9995 & 0.0006 & 0.9979 & 1.0000 & 1.0006 &  2000 \\ 
-     $p$[3] & 0.8834 & 0.0634 & 0.7124 & 0.9619 & 1.0054 &   410 \\ 
-     $p$[4] & 0.8100 & 0.1410 & 0.4269 & 0.9650 & 1.0023 &   830 \\ 
-      \hline
 \end{tabular}
@@
 The one exception to this pattern is the effect of POSE on recruitment by ARTR, but the nonconforming
-coefficient ($\omega$[3,1] = -0.50 in Table \ref{table:recruitment}) has low precision (sd = 0.28, Table \ref{table:recruitment}.} 
+coefficient ($\omega$[3,1] = -0.6589 in Table \ref{table:recruitment}) has low precision (sd = 0.3342, Table \ref{table:recruitment}.} 
 \label{table:alphas}
@@
 \underline{Survival} & & & & \\        
-ARTR & -2.8522 & -0.0720 & 0.0617 & 0.0387\\
-HECO & -0.0063 & -1.6962 & 0.0649 & 0.0313\\
-POSE & 0.0000 & -0.0224 & -2.0073 & 0.0468\\
-PSSP & 0.0099 & 0.0025 & 0.0517 & -1.1267\\
+ARTR   & -1.5214  & -0.0694  & -0.0049 &  0.0281\\
+HECO   & -0.0064  &  -0.6647 &  0.0732 &  0.0190\\
+POSE  &   0.0001&  -0.0153 &  -1.2632&   0.0249\\
+PSSP  &  0.0101 &  0.0046 & 0.0245 &  -0.4430\\
 \underline{Growth} & & & & \\ 
-ARTR & -0.1497 & 0.0045 & -0.0614 & -0.0050\\
-HECO & -0.0042 & -0.2134 & 0.0204 & -0.0373\\
-POSE & -0.0004 & 0.0110 & -0.5753 & -0.0093\\
-PSSP & -0.0023 & -0.0201 & -0.0200 & -0.3845\\
+ARTR   & -0.0364 &  0.0033& -0.0573&  0.0062\\
+HECO   &  -0.0031 & -0.0479& 0.0242 & -0.0151\\
+POSE  &  -0.0002& 0.0039&  -0.2641&   -0.0073\\
+PSSP  & -0.0025 & -0.0118 & -0.0067 & -0.1480\\
 \underline{Recruitment} & & & & \\ 
-ARTR & -0.5539 & -0.4522 & -0.5030 & -0.1253\\
-HECO &  0.0654 & -1.5975 & -0.0586 & -0.3056\\
-POSE & 0.0043 & -0.0058 & -1.7052 & 0.0335\\
-PSSP & 0.0953 & -0.3047 & -0.0960 & -1.7046\\
+ARTR  & -0.5881 &  -0.4800 & -0.6589 & -0.1867\\
+HECO   &   0.06346 & -1.7368 & -0.1021 & -0.3746\\
+POSE  &   0.0220 &  0.0388 & -1.8943 &  0.1131\\
+PSSP  &  0.1164 & -0.3682 & -0.1953 & -1.7379\\
 \hline
@@
 spicata}.
-Each point compares the crowding experienced by one \textit{P.
-spicata} individual in one year from the species on the x-axis to the species on the y-axis.
-Blue symbols show values from controls plots, red symbols show pre-treatment values from \textit{Artemisia} removal plots.``W.allcov'' refers to the aggregated crowding by all shrubs and perennial grasses beyond the focal dominant species, and ``W.allpts'' refers to the aggregated crowding by forb species.
+\new{Each point compares the crowding experienced by one \textit{P.
+spicata} individual in one year from the species on the x-axis to the species on the y-axis.}   Blue symbols show values from controls plots, red symbols show \new{pre-treatment} values from \textit{Artemisia} removal plots.``W.allcov'' refers to the aggregated crowding by all shrubs and perennial grasses beyond the focal dominant species, and ``W.allpts'' refers to the aggregated crowding by forb species.
 \texttt{Wdistrib\textbackslash exploreSurvivalWs.r}.

### 2018-02-01 removals/manuscript/removal_ms.tex
Incorporated Ellner and Hooker edits into manuscript

@@
 For growth, the removal treatment effect was significant and positive for two species, \textit{Poa secunda} and \textit{Pseudoroegneria spicata}, indicating that the baseline model underestimated interspecific competition.
-For all three grass species, population models based on the vital rate regressions that included removal effects projected 1.4 to 3-fold increases in equilibrium population size relative to the baseline model (no removal effects).} However, including information about pretreatment cover of \textit{A.
-tripartita} at the quadrat or individual scale failed to improve the growth regressions, raising questions about the mechanisms driving the positive response to removal.
+For all three grass species, population models based on the vital rate regressions that included removal effects projected 1.4 to 3-fold increases in equilibrium population size relative to the baseline model (no removal effects).}
+However, we found no evidence of higher response to removal in quadrats with higher pretreatment cover of \textit{A.
+tripartita}, or by plants experiencing higher pre-treatment crowding by \textit{A.
+tripartita},
+raising questions about the mechanisms driving the positive response to removal.
 \new{While the experimental results showed greater competitive release than our baseline models predicted, we will need more tightly controlled experiments focused on underlying mechanisms to conclusively validate or reject our model's predictions.}
@@
 spicata} growth suggest that models based on observational data may underestimate competitive release.
-We conducted additional analyses to further explore removal effects on the growth of these two species.
+We conducted additional analyses to further explore this hypothesis.
 First, \new{we asked whether increases in grass growth following removal were greatest in quadrats or neighborhoods where we removed the most \textit{A.
 tripartita}.
-We did this by taking residuals from the fitted growth models for plants in removal quadrats and plotting those residuals against the pre-treatment, quadrat-level cover of \textit{A.
+We did this by taking residuals from the fitted growth model that included removal effects, and plotting the residuals for plants in removal quadrats against the pre-treatment, quadrat-level cover of \textit{A.
 tripartita}.
-No species showed the positive relationship we expected to see (Fig.
+No species showed the positive relationship that would be expected (Fig.
 \ref{fig:GrowthResidsQuad}).
 In fact, \textit{Poa secunda} residuals were significantly lower in quadrats with high pre-treatment cover of \textit{A.
-tripartita} (individuals in high cover quadrats are growing more slowly than the growth model predicted).
+tripartita}: individuals in high pre-treatment cover quadrats grew more slowly than predicted by the growth model with removal effects.
 We repeated this analysis at the individual plant level for grasses present in 2011.
@@
 
-%we classified all individuals of these species as located either inside or outside the limits of \textit{A.
-tripartita} canopies present in 2011 (before removal).
-If removal is releasing grasses from \textit{A.
-tripartita} competition for light or soil resources, we would expect individuals located under shrub canopies to experience a greater effect of removal than individuals further away.
-However, the location of individuals with respect to \textit{A.
-tripartita} canopies had no effect on growth of either species (Tables \ref{table:POSEgrowth-inARTR} and \ref{table:PSSPgrowth-inARTR}), meaning that individuals located outside \textit{A.
-tripartita} canopies responded to removals as much as individuals underneath (removed) canopies.
+\begin{figure}[tbp]
+\centering
+\includegraphics[width=1\textwidth]{growth_residuals_vs_quad_removal_cover}
+\caption{The relationship between residuals from the fitted growth model and quadrat-level pre-treatment cover of the removal species (all grasses removed in quadrats where we studied \textit{A.
+tripartita}, and \textit{A.
+tripartita} removed in the quadrats where we studied the grasses).
+Linear regressions are shown only when statistically significant.
+Source file:  \texttt{growth\textbackslash growth\textunderscore residuals\textunderscore by\textunderscore quad.R}.}
+\label{fig:GrowthResidsQuad}
+\end{figure}
 
 Second, we evaluated the possibility that removal effects might be transient.
-We removed aboveground plant material, but roots of the removed plants remain in the ground and decompose, potentially creating a flush of mineral nutrients.
+We removed aboveground plant material, but roots of the removed plants remained in the ground potentially creating a flush of mineral nutrients as they decompose.
 We tested for a transient removal effect by adding a year-by-treatment interaction.
@@
 In fact, for  \textit{P.
-secunda}, the removal effects increase over time.
+secunda}, the removal effects increased over time.
 
@@
 not just additional observations, further increases the potential to confront the model with a novel situation outside the historical range of variation.
-\new{We expected that upon completion of this study, we would have clear evidence for the success or failure of our previous model's predictions.
+\new{We expected this study to provide clear evidence for the success or failure of our previous model's predictions.
 Unfortunately, our results do not provide that clarity.
-In the sections below, we first summarize the evidence that our experiment revealed greater competitive release than our baseline model predicted, we then explain weaknesses in the experimental results that complicate the interpretation, and we end by discussing general lessons for the study of interspecific interactions.}
+In the sections below, we first summarize the evidence that our experiment revealed greater competitive release than our baseline model predicted, then explain weaknesses in the experimental results that complicate the interpretation, and end by discussing general lessons for the study of interspecific interactions.}
 
@@
 For these species, the individual-based model with removal effects predicted higher cover than the baseline version of the model.
-The differences in predicted cover of roughly 20\% (\ref{fig:IBM1step}) are comparable to the differences in the individual growth rates, reflecting the fact that at these short time scales changes in individual size influence cover more than survival or recruitment.
+The differences in predicted cover of roughly 20\% (Fig.
+\ref{fig:IBM1step}) are comparable to the differences in the individual growth rates, reflecting the fact that at these short time scales changes in individual size influence cover more than survival or recruitment.
 For \textit{A.
@@
 
-\new{The strongest evidence for large competitive release of grasses following \textit{A.
-tripartita} removal comes from the long-term projections of the IPM.
+\new{The strongest evidence that competitive release of grasses following \textit{A.
+tripartita} removal was greater than predicted comes from the long-term projections of the IPM.
 These projections effectively make removal treatment effects permanent; the removal effects on vital rates observed during the experiment are locked in for the entire 2500 time step simulation.
@@
 
-\new{Another reason we view the growth results with some suspicion is that increases in growth were no larger in individual neighborhoods or quadrats with higher pre-treatment \textit{A.
+\new{Another reason we view the growth results with suspicion is that increases in growth were no larger in individual neighborhoods or quadrats with higher pre-treatment \textit{A.
 tripartita} cover.
 If \textit{A.
-tripartita} removal is releasing grasses from competition, shouldn't this effect be larger for grasses located in areas with high \textit{A.
+tripartita} removal is releasing grasses from competition, shouldn't this effect be larger for grass genets in areas with high \textit{A.
 tripartita} cover?
@@
 spicata} growth.
-\new{However, since the 1900's, many quadrats have experienced dramatic changes in \textit{A.
+\new{However, since the 1900's, many quadrats have undergone dramatic changes in \textit{A.
 tripartita} cover, arguing against viewing \textit{A.
@@
 A final possibility is that our removal treatment influenced grass growth through some other mechanism, unrelated to competition. 
-Maybe the loss of shrub cover benefits grasses by reducing small mammal herbivory \citep{Orrock2010}?
-Without a better understanding of the mechanism driving increased grass growth in the shrub removal quadrats, we are hesitant to dramatically adjust our prior understanding gained from models fit to decades of observational data.
-}
+For example, loss of shrub cover might benefits grasses by reducing small mammal herbivory \citep{Orrock2010}. 
+Without a better understanding of the mechanism driving increased grass growth in the shrub removal quadrats, 
+we are hesitant to dramatically adjust our prior understanding gained from models fit to decades of observational data.  }
 
@@
 \new{A variety of approaches are available for studying interspecific interactions among plants, and each has its advantages and disadvantages.
-Observational datasets have the advantage of being realistic: they reflect all the processes and interactions that drive community dynamics across a range of spatial and temporal scales.
+Observational datasets have the advantage of realism: they reflect all the processes and interactions that drive community dynamics across a range of spatial and temporal scales.
 Their disadvantage is weak inference.
 Distinguishing between correlation and causation is challenging, and simulations of a model fit to observational data may rely heavily on extrapolation.
-Experiments, such as response surface designs (Inouye), offer an alternative approach to studying interactions at the population scale.
+Experiments, such as response surface designs \citep{Inouye2001}, offer an alternative approach to studying interactions at the population scale.
 Their chief advantage is strong inference about causality.
@@
 Finally, we can study interactions mechanistically by describing patterns of resource use \citep{Dybzinski2007,Silvertown1999}.
-Mechanistic studies have the virtue of focusing on directly on the processes of interest, but rarely address more than one limiting resource at a time.
-}
-
-\new{ Our previous work was based purely on observational data, so we were particularly concerned about the extrapolation issue.
+Mechanistic studies have the virtue of focusing  directly on the processes of interest, but rarely address more than one limiting resource at a time.
+}
+
+\new{Our previous work was based purely on observational data, so we were particularly concerned about the extrapolation issue.
 Specifically, our model projects the performance of individual plants growing in the absence of neighbors.
 Although the observational data set seems to include such situations (Fig.
-\ref{fig:Wscatters}), if these isolated individual are still interacting with neighbors belowground, our model might underestimate the true intrinsic growth rates.
-Therefore, we conducted the removal experiment, hoping to gain the advantages offered by multiple, complementary approaches.
+\ref{fig:Wscatters}), the data-driven competition kernels we use to calculate neighborhood crowding are quite narrow: they are heavily weighted toward nearby neighbors and essentially ignore the influence of more distant neighbors (Fig.
+\ref{fig:CompKernels}).
+However, if our competition kernels under-estimate the strength of long-range interactions, then our model might underestimate the true intrinsic growth rates that would occur when both near and distant neighbors are removed.
+We conducted the removal experiment to address this kind of concern, with the hope of gaining the advantages offered by multiple, complementary approaches.
 The positive responses of \textit{Poa secunda} and \textit{P.
 pseudoroegneria} to removals suggests our observational approach might have underestimated intrinsic growth rates after all.
-But removal experiments fall into a gray area between the observational and experimental approach.
-They are an experimental manipulation, but they rely on natural spatial variation in composition.
-We can't remove a plant from a location where it does not occur.
+But removal experiments fall into a gray area between an observational study and a randomized, controlled experiment.
+They are an experimental manipulation, but they rely on natural spatial variation in composition; we can only remove plants from locations where they naturally occur.
 The reliance on natural variation weakens inference, leaving nagging questions about pre-existing differences among quadrats.
-We could have avoided those concerns by conducting a response surface experiment, planting our study species in randomly selected locations.
-}
+We could have avoided those concerns by conducting a response surface experiment, planting our study species in randomly selected locations.}
 
 \new{We also learned that partial removals, rather than complete removal treatments, could 
-have helped us to distinguish between competition-mediated and density-independent effects of removal treatments (e.g. Fig. \ref{fig:PSSPresids}).
+have helped to distinguish between competition-mediated and other effects of removal treatments (e.g. Fig. \ref{fig:PSSPresids}).
 Partial removals, or experimental gradients in densities, would give us inference about the competition coefficients themselves, whereas our 
@@
 
-\new{The difficulty we experienced interpreting our results also made us appreciate the value of mechanistic insight.
+\new{Our difficulty interpreting our results also made us appreciate the value of mechanistic insight.
 Even if the removal experiment had perfectly validated the predictions of our phenomenological baseline model, we suspect we would still be asking ourselves, how can interspecific interactions really be so weak?
@@
 
-Although our results point to limitations of model based on long-term observational data, they also give some cause for optimism about population forecasting.
+Although our results point to limitations of models based on long-term observational data, they also give some cause for optimism about population forecasting.
 \citet{Crone2013} found that matrix projection models for plants often failed to forecast population trends due
-to environmental differences between the data collection and forecast periods. Our population models 
-performed quite well, after accounting for year-to-year environmental variation through the inclusion of
-random year effects.
+to environmental differences between the data collection and forecast periods. 
+By including random year-effects to account for year-to-year environmental variation, our population models performed quite well
+at forecasting between-year population changes (Figure \ref{fig:obsVpred}).
 
@@
 The results do increase our interest in testing model predictions using even more tightly controlled experiments as well as direct study of resource-use.
-
-%\subsection*{Implications of weak interspecific interactions}
-%
-%The weak interspecific interactions among many of the dominant species in our system indicates that these species  
-%are limited by quite different factors. This low niche overlap could reflect strong resource partitioning \citep{tilman_resource_1982}, 
-%perhaps combined with variation in the times and locations at which species are taking up those resources. 
-%Species-specific responses to spatial and temporal variation in non-resource environmental variables can also weaken interspecific 
-%interactions \citep{chesson_mechanisms_2000}, though in a previous study we failed to find evidence of a temporal storage effect 
-%in this community \citep{adler_weak_2009}. Interactions with above- or belowground enemies can regulate populations and, given 
-%enough species-specificity in the effects of these enemies, decouple the dynamics of different species \citep{hersh_evaluating_2011,
-%janzen_herbivores_1970,connell_role_1971}.  More likely, plants may alter the environment, and even resource availability, 
-%in ways that facilitate their neighbors. For example, shade cast by \textit{A. tripartita} canopies may reduce 
-%evapotranspiration (e.g. \citealt{Barbier2008}). \textit{A. tripartita} removal might offset this facilitative effect by reducing 
-%resource uptake, resulting in little net change in water balance following removal. 
-%Unfortunately, we did not monitor soil moisture at the fine spatial and temporal resolution necessary to test this hypothesis.
-%
-%Evidence for weak interspecific interactions and limited competitive release has interesting implications. If most 
-%interspecific interactions are weak, then single-species models may be adequate for forecasting population responses 
-%to environmental change \citep{adler_forecasting_2012,kleinhesselink_indirect_2015,Chu2016}. Similarly, differences among 
-%species in relative abundance should reflect variation in the strength of intraspecific density dependence more than 
-%variation in the strength of interspecific interactions \citep{yenni_strong_2012}. A recent global study showing little 
-%variation in the abundance of invasive grassland species in native and invaded ranges on different 
-%continents  \citep{firn_abundance_2011} is consistent with the idea that intraspecific interactions are the primary determinant 
-%of species abundances. In systems where this is true, effective management and conservation requires understanding the sources 
-%and species-specificity of intraspecific density dependence. A useful first step in this direction would be to test 
-%whether our model is also successful at predicting the effects of release from intraspecific competition.
-%
-%For many species in other communities, interspecific competition may still be an important driver. In closed-canopy 
-%forests where competition for light is intense, we might have found different results than in our system, in which 
-%competition for a number of belowground resources is more important than competition for light. However, a 
-%number of recent studies from forests have also found that heterospecific competitive effects are dramatically 
-%weaker than conspecific effects \citep{comita_asymmetric_2010,Kunstler2016,johnson_conspecific_2012}.  We also 
-%expect stronger competition in successional systems, or communities in the early stages of community assembly 
-%where the process of competitive exclusion is still underway \citep{kokkoris_patterns_1999}. Finally, we might have
-%found stronger evidence of competition if we could have studied rare species in our system, but only if those species 
-%are rare because of niche overlap with stronger competitors. If they are rare because they occupy unique, but narrow, niches, 
-%interspecific interactions would still be weak. Determining why interspecific interactions and competitive release are 
-%strong in some ecological contexts but weak in others remains a fascinating research problem.
 
@@
 
-\begin{figure}[h]
-\centering
-\includegraphics[width=1\textwidth]{growth_residuals_vs_quad_removal_cover}
-\caption{The relationship between residuals from the fitted growth model and quadrat-level pre-treatment cover of the removal species (all grasses removed in quadrats where we studied \textit{A.
+%\begin{figure}[h]
+%\centering
+%%\includegraphics[width=1\textwidth]{growth_residuals_vs_quad_removal_cover}
+%\caption{The relationship between residuals from the fitted growth model and quadrat-level pre-treatment cover of the removal species (all grasses removed in quadrats where we studied \textit{A.
 tripartita}, and \textit{A.

### 2018-03-02 removals/manuscript/removal_ms.tex
Reformatted figure legends

@@
 \usepackage[vmargin=1in,hmargin=1in]{geometry}
+\usepackage{caption}
 
@@
 responses to removal were adequately predicted by models fit to observational data.
-
-\begin{figure}[tbp]
-\centering
-\includegraphics[width=1\textwidth]{treatment_trends_cover}
-\caption{Cover trends by treatment for the four dominant species.  Note the difference in y-axis scale between the canopy cover of \textit{A. tripartita} and 
-the basal cover of the dominant grasses. ``Shrub removal'' refers to removal of \textit{A. tripartita}, and ``Grass removal'' refers to removal of all perennial grasses. 
-The cover of species in treatments from which they are removed (e.g. \textit{A. tripartita} cover in shrub removal plots) is included to show values at the start 
-of the experiment in 2011 and the efficacy of the removal treatments. Source file: \texttt{treatment\textunderscore trends\textunderscore removals.r}.}
-\label{fig:CoverTrends}
-\end{figure}
 
@@
 \ref{fig:VitalRateTest}; Table \ref{table:recruitment}).}
-
- \begin{figure}[tbp]
- \centering
- \includegraphics[width=1\textwidth]{treatment_tests}
- \caption{Effects of removal treatments on the survival, growth, and recruitment of the four dominant species.
-Circles show the estimated coefficient of the removal effect, bars show 95\% credible intervals.
-The removal effect shows response of \textit{A.
-tripartita} (``ARTR") to perennial grass removal, and response of the perennial grasses to \textit{A.
-tripartita} removal, after accounting for local neighborhood interactions.
-The other species codes are HECO: \textit{Hesperostipa comata}, POSE: \textit{Poa secunda}, and PSSP: \textit{Pseudoroegneria spicata}.
- Source file: \texttt{treatment\textunderscore test\textunderscore figure.r}.}
- \label{fig:VitalRateTest}
- \end{figure}
 
@@
 
-\begin{figure}[tbp]
-\centering
-\includegraphics[width=1\textwidth]{growth_residuals_vs_quad_removal_cover}
-\caption{The relationship between residuals from the fitted growth model and quadrat-level pre-treatment cover of the removal species (all grasses removed in quadrats where we studied \textit{A.
-tripartita}, and \textit{A.
-tripartita} removed in the quadrats where we studied the grasses).
-Linear regressions are shown only when statistically significant.
-Source file:  \texttt{growth\textbackslash growth\textunderscore residuals\textunderscore by\textunderscore quad.R}.}
-\label{fig:GrowthResidsQuad}
-\end{figure}
-
 Second, we evaluated the possibility that removal effects might be transient.
@@
 
- \begin{figure}[tbp]
- \centering
- \includegraphics[width=1\textwidth]{obsVpred_quad_year}
- \caption{The relationship between observed and predicted cover for each quadrat, year, and species.
-The 1:1 line is dashed, the solid black line shows the trend for control plots with predictions from the baseline IBM model (black symbols), the blue line shows the trend for removal plots with predictions from the baseline model (blue symbols), and the red line shows the trend for removal plots with predictions from the removal model (red symbols).
-Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r}.
-}
- \label{fig:obsVpred}
- \end{figure}
-
 Time series of predicted and observed cover, averaged over quadrats, show similar patterns.
@@
 \ref{fig:IBM1step-maxCI}).
-
- \begin{figure}[tbp]
- \centering
- \includegraphics[width=1\textwidth]{cover_projections_1step}
- \caption{Observed and predicted cover for each of the four modeled species.
-Solid black lines show mean observed cover averaged across control (closed symbols) and removal treatment (open symbols) quadrats.
-Blue lines and symbols show one-step-ahead predictions from the baseline IBM (no removal treatment coefficients), averaged across quadrats.
-Red lines and symbols show one-step-ahead predictions from an IBM that includes removal treatment coefficients.
-For \textit{H.
-comata}, lines showing predictions for the baseline and removal effect models are virtually indistinguishable.
-Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r}.
-}
- \label{fig:IBM1step}
- \end{figure}
 
@@
 \ref{fig:IPMresults-maxCI}).
-
- \begin{figure}[tbp]
- \centering
- \includegraphics[width=1\textwidth]{boxplots}
- \caption{Equilibrium cover of the four dominant species simulated by the IPM.
-Boxplots show interannual variation reflecting random year effects.
-Gray boxes show cover simulated by a model with \textit{A.
-tripartita} present and no removal treatment coefficients, blue boxes show results from the same model but with \textit{A.
-tripartita} cover set to zero (a species removal), and red boxes show results from a model with \textit{A.
-tripartita} set to zero and removal treatment coefficients included.
-Source file: \texttt{ipm\textbackslash IPM-figures.r}.}
- \label{fig:IPMresults}
- \end{figure}
 
@@
 \bibliography{RemovalRefs}
+
+\newpage
+\section*{Figure legends}
+
+\textbf{Figure 1.} Cover trends by treatment for the four dominant species.
+Note the difference in y-axis scale between the canopy cover of \textit{A.
+tripartita} and the basal cover of the dominant grasses.
+``Shrub removal'' refers to removal of \textit{A.
+tripartita}, and ``Grass removal'' refers to removal of all perennial grasses.
+The cover of species in treatments from which they are removed (e.g.
+\textit{A.
+tripartita} cover in shrub removal plots) is included to show values at the start of the experiment in 2011 and the efficacy of the removal treatments.
+Source file: \texttt{treatment\textunderscore trends\textunderscore removals.r}
+
+\vspace{3mm}
+
+\noindent \textbf{Figure 2.} Effects of removal treatments on the survival, growth, and recruitment of the four dominant species.
+Circles show the estimated coefficient of the removal effect, bars show 95\% credible intervals.
+The removal effect shows response of \textit{A.
+tripartita} (``ARTR") to perennial grass removal, and response of the perennial grasses to \textit{A.
+tripartita} removal, after accounting for local neighborhood interactions.
+The other species codes are HECO: \textit{Hesperostipa comata}, POSE: \textit{Poa secunda}, and PSSP: \textit{Pseudoroegneria spicata}.
+ Source file: \texttt{treatment\textunderscore test\textunderscore figure.r}.
+
+ \vspace{3mm}
+
+\noindent \textbf{Figure 3.} The relationship between residuals from the fitted growth model and quadrat-level pre-treatment cover of the removal species (all grasses removed in quadrats where we studied \textit{A.
+tripartita}, and \textit{A.
+tripartita} removed in the quadrats where we studied the grasses).
+Linear regressions are shown only when statistically significant.
+Source file:  \texttt{growth\textbackslash growth\textunderscore residuals\textunderscore by\textunderscore quad.R}.
+
+\vspace{3mm}
+
+\noindent \textbf{Figure 4.} The relationship between observed and predicted cover for each quadrat, year, and species.
+The 1:1 line is dashed, the solid black line shows the trend for control plots with predictions from the baseline IBM model (black symbols), the blue line shows the trend for removal plots with predictions from the baseline model (blue symbols), and the red line shows the trend for removal plots with predictions from the removal model (red symbols).
+Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r}.
+
+\vspace{3mm}
+
+\noindent \textbf{Figure 5.} Observed and predicted cover for each of the four modeled species.
+Solid black lines show mean observed cover averaged across control (closed symbols) and removal treatment (open symbols) quadrats.
+Blue lines and symbols show one-step-ahead predictions from the baseline IBM (no removal treatment coefficients), averaged across quadrats.
+Red lines and symbols show one-step-ahead predictions from an IBM that includes removal treatment coefficients.
+For \textit{H.
+comata}, lines showing predictions for the baseline and removal effect models are virtually indistinguishable.
+Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r}.
+
+\vspace{3mm}
+
+ \noindent\textbf{Figure 6.} Equilibrium cover of the four dominant species simulated by the IPM.
+Boxplots show interannual variation reflecting random year effects.
+Gray boxes show cover simulated by a model with \textit{A.
+tripartita} present and no removal treatment coefficients, blue boxes show results from the same model but with \textit{A.
+tripartita} cover set to zero (a species removal), and red boxes show results from a model with \textit{A.
+tripartita} set to zero and removal treatment coefficients included.
+Source file: \texttt{ipm\textbackslash IPM-figures.r}
+
+\captionsetup[figure]{labelfont=bf, labelsep=none, justification=raggedright,singlelinecheck=false}
+
+\clearpage
+\begin{figure}[tbp]
+\centering
+\includegraphics[width=1\textwidth]{treatment_trends_cover}
+\caption{}
+\label{fig:CoverTrends}
+\end{figure}
+
+\clearpage
+ \begin{figure}[tbp]
+ \centering
+ \includegraphics[width=1\textwidth]{treatment_tests}
+ \caption{}
+ \label{fig:VitalRateTest}
+ \end{figure}
+
+ \clearpage
+ \begin{figure}[tbp]
+ \centering
+ \includegraphics[width=1\textwidth]{growth_residuals_vs_quad_removal_cover}
+ \caption{}
+ \label{fig:GrowthResidsQuad}
+ \end{figure}
+
+ \clearpage
+  \begin{figure}[tbp]
+  \centering
+  \includegraphics[width=1\textwidth]{obsVpred_quad_year}
+  \caption{}
+  \label{fig:obsVpred}
+  \end{figure}
+
+  \clearpage
+   \begin{figure}[tbp]
+   \centering
+   \includegraphics[width=1\textwidth]{cover_projections_1step}
+   \caption{}
+   \label{fig:IBM1step}
+   \end{figure}
+
+   \clearpage
+    \begin{figure}[tbp]
+    \centering
+    \includegraphics[width=1\textwidth]{boxplots}
+    \caption{}
+    \label{fig:IPMresults}
+    \end{figure}
 

### 2018-03-22 removals/manuscript/removal_ms.tex
First pass at final version

@@
 
-\title{Weak interspecific interactions in a sagebrush steppe\new{? Conflicting evidence from observations and experiments} }
+\title{Weak interspecific interactions in a sagebrush steppe? Conflicting evidence from observations and experiments}
 
@@
 If our ``baseline" regression model, which accounts for local plant-plant interactions, accurately explains the observed responses to removals, then the removal coefficient should be non-significant.
-\new{For survival, the removal coefficients were never significantly different from zero, and only \textit{A.
+For survival, the removal coefficients were never significantly different from zero, and only \textit{A.
 tripartita} showed a (negative) response to removals at the recruitment stage.
 For growth, the removal treatment effect was significant and positive for two species, \textit{Poa secunda} and \textit{Pseudoroegneria spicata}, indicating that the baseline model underestimated interspecific competition.
-For all three grass species, population models based on the vital rate regressions that included removal effects projected 1.4 to 3-fold increases in equilibrium population size relative to the baseline model (no removal effects).}
+For all three grass species, population models based on the vital rate regressions that included removal effects projected 1.4 to 3-fold increases in equilibrium population size relative to the baseline model (no removal effects).
 However, we found no evidence of higher response to removal in quadrats with higher pretreatment cover of \textit{A.
@@
 raising questions about the mechanisms driving the positive response to removal.
-\new{While the experimental results showed greater competitive release than our baseline models predicted, we will need more tightly controlled experiments focused on underlying mechanisms to conclusively validate or reject our model's predictions.}
-
-%The ability of models based on observational data to predict population responses to experimental removals increased our confidence that interspecific interactions are in fact weak for at least three of our four study species.
-The implication is that factors besides interspecific feedbacks play important roles in determining the structure and dynamics of this community.
-The success of the models also supports the use of multispecies models based on long-term observational data for characterizing competitive interactions.
-
-\textbf{\large{Keywords:}} Coexistence, competition, integral projection model, removal experiment, sagebrush steppe.
+While our results show the value of combining observations with a simple removal experiment, more tightly controlled experiments focused on underlying mechanisms may be required to conclusively validate or reject predictions from phenomenological models.
+
+\textbf{\large{Keywords:}} Coexistence, competition, facilitation, integral projection model, population dynamics, removal experiment, sagebrush steppe.
 
@@
 The sagebrush removal was virtually 100\% effective; grass removals required additional herbicide applications in April and May of each year, with fewer grasses remaining each year.
-Appendix Fig.
+Appendix S1 Fig.
 \ref{fig:photos} shows photographs of example quadrats before and after treatment.
@@
 Why don't we find strong interactions between removal treatments and the estimated competition coefficients? The answer is that because our treatments 
-involved complete, not just partial, removal of target species, their effects can be captured with an intercept term, as illustrated in Fig. \ref{fig:PSSPresids}. 
+involved complete, not just partial, removal of target species, their effects can be captured with an intercept term, as illustrated in Appendix S1: Fig. \ref{fig:PSSPresids}. 
 As a result, if, for example, one of the grasses responds more positively to removal of \textit{A. tripartita} than our baseline model predicts, we cannot 
@@
 \end{equation}
-(see Appendix \ref{AppendixA} for a complete description). The IPM uses a mean-field approximation for
+(see Appendix S1 for a complete description). The IPM uses a mean-field approximation for
 neighborhood competition that captures the essential features of spatially-explicit local interactions; 
-in Appendix \ref{sec:kernelMethods} we explain how the spline competition kernels were used 
+in Appendix S1: \ref{sec:kernelMethods} we explain how the spline competition kernels were used 
 to compute neighborhood competition as a function of individual size for the IPM.
@@
 During the period of our experiment, annual water-year (October-September) precipitation fell within the 5\% - 95\% quantiles of the 
-long-term mean, although more years fell below the mean than above it (Fig. \ref{fig:climate}a in Appendix C). However, annual mean temperatures 
-were consistently warm relative to the long-term mean, and two years were warmer than the 95\% quantile (Fig. \ref{fig:climate}b).
+long-term mean, although more years fell below the mean than above it (Appendix S1: Fig. \ref{fig:climate}a). However, annual mean temperatures 
+were consistently warm relative to the long-term mean, and two years were warmer than the 95\% quantile (Appendix S1: Fig. \ref{fig:climate}b).
 
@@
 However, when we tested the effects of removal treatments on year-to-year population growth rates ($\log(Cover_{t+1}/Cover_t)$) at the quadrat level,
- none of the removal treatment effects were significantly different from zero (Table \ref{table:coefficients}). For \textit{P. spicata}, 
+ none of the removal treatment effects were significantly different from zero (Appendix S1: Table \ref{table:coefficients}). For \textit{P. spicata}, 
 the 95\% credible interval only just contained zero (95\% CI: -0.01, 0.38), providing marginal evidence for a positive effect of removal on population growth. 
@@
 fitted models imply that interspecific competition is much weaker than intraspecific competition, in many
-cases by 1-2 orders of magnitude (Table \ref{table:alphas}). The intraspecific interaction effects are large and have credible intervals that 
+cases by 1-2 orders of magnitude (Appendix S1: Table \ref{table:alphas}). The intraspecific interaction effects are large and have credible intervals that 
 do not overlap zero, while the interspecific coefficients are small and the majority have credible intervals that do overlap zero.
@@
 For the survival models, the 95\% credible intervals on the removal effects overlapped zero for all four species (Fig. \ref{fig:VitalRateTest}; 
-Tables \ref{ARTRsurvival} to \ref{PSSPsurvival}), meaning that the survival probabilities of plants in removal plots can be predicted based on data 
+Appendix S1: Tables \ref{ARTRsurvival} to \ref{PSSPsurvival}), meaning that the survival probabilities of plants in removal plots can be predicted based on data 
 from control plots. For the growth models, the 95\% credible intervals on the removal effects were positive for \textit{P. secunda}  and \textit{P. spicata},
@@
  (Fig.
-\ref{fig:VitalRateTest}; Tables \ref{ARTRgrowth} to \ref{PSSPgrowth}).
-\new{ At the recruitment stage, only \textit{A.
+\ref{fig:VitalRateTest}; Appendix S1: Tables \ref{ARTRgrowth} to \ref{PSSPgrowth}).
+At the recruitment stage, only \textit{A.
 tripartita} showed a significant response to removals, and it was negative (Fig.
-\ref{fig:VitalRateTest}; Table \ref{table:recruitment}).}
+\ref{fig:VitalRateTest}; Appendix S1: Table \ref{table:recruitment}).
 
@@
 We conducted additional analyses to further explore this hypothesis.
-First, \new{we asked whether increases in grass growth following removal were greatest in quadrats or neighborhoods where we removed the most \textit{A.
+First, we asked whether increases in grass growth following removal were greatest in quadrats or neighborhoods where we removed the most \textit{A.
 tripartita}.
@@
 We found no significant relationships between growth residuals and pre-treatment crowding from \textit{A.
-tripartita} (Fig.
+tripartita} (Appendix S1: Fig.
 \ref{fig:GrowthResidsIndivid}).
-}
 
@@
 secunda}  and \textit{P.
-spicata}, we found no sign of a decrease in the removal effect over time (Tables \ref{table:POSEgrowth-trtYears} and \ref{table:PSSPgrowth-trtYears}).
+spicata}, we found no sign of a decrease in the removal effect over time (Appendix S1: Tables \ref{table:POSEgrowth-trtYears} and \ref{table:PSSPgrowth-trtYears}).
 In fact, for  \textit{P.
@@
 Third, responses to removals might be poorly predicted by models based on observational data if the removals reduce plant densities (crowding) below the range of historical variation, or if the removals result in combinations of neighborhood-scale densities that deviate from naturally occurring patterns of correlation.
-However, the distribution of crowding observed within control plots easily spans the crowding values observed in removal plots (Fig.
-\ref{fig:Wscatters}), and we failed to find evidence for strong interspecific covariances in crowding (Appendix \ref{AppendixA}).
+However, the distribution of crowding observed within control plots easily spans the crowding values observed in removal plots (Appendix S1: Fig.
+\ref{fig:Wscatters}), and we failed to find evidence for strong interspecific covariances in crowding (Appendix S1).
 
@@
 comata} was lower in removals than controls, though it was absent from most \textit{P.
-spicata} neighborhoods in both treatments (Fig.
+spicata} neighborhoods in both treatments (Appendix S1: Fig.
 \ref{fig:W-by-treatment}).
 These pre-existing differences in composition raise the possibility that plots selected for the removal treatments might differ from the control plots in some unobserved covariate that could influence plant composition and performance.
-\new{On the other hand, some of the control plots that currently contain zero \textit{A.
+On the other hand, some of the control plots that currently contain zero \textit{A.
 tripartita} had high \textit{A.
@@
 tripartita} cover, such large changes in cover over time would be unlikely.
-}
 
@@
 secunda} and \textit{P.
-spicata} (Fig.
+spicata} (Appendix S1: Fig.
 \ref{fig:IBM1step-maxCI}).
@@
 comata}, reflecting its small sample size and corresponding high uncertainty, but little change in the projections for \textit{Poa secunda} and \textit{P.
-spicata} (Fig.
+spicata} (Appendix S1: Fig.
 \ref{fig:IPMresults-maxCI}).
@@
 tripartita} decreasing towards extinction.
-However, these outcomes both are driven by parameters that are estimated with very high uncertainty (Table~\ref{table:recruitment}).
+However, these outcomes both are driven by parameters that are estimated with very high uncertainty (Appendix S1: Table \ref{table:recruitment}).
 The baseline model predicts an increase following grass removal because of the relatively strong negative effects of \textit{Poa secunda} and \textit{H.
@@
 \textit{H. comata} effect includes values less than half of the point-estimate. The grass removal treatment effect, which leads to predicted extinction, has a credible 
-interval that broadly overlaps zero (Table~\ref{table:recruitment}). \textit{A. tripartita} recruitment parameters have exceptionally high uncertainty, presumably 
+interval that broadly overlaps zero (Appendix S1: Table\ref{table:recruitment}). \textit{A. tripartita} recruitment parameters have exceptionally high uncertainty, presumably 
 because \textit{A. tripartita} recruitment is episodic (see \emph{Discussion}), and this produces extreme uncertainty about its long-term 
@@
 not just additional observations, further increases the potential to confront the model with a novel situation outside the historical range of variation.
-\new{We expected this study to provide clear evidence for the success or failure of our previous model's predictions.
+We expected this study to provide clear evidence for the success or failure of our previous model's predictions.
 Unfortunately, our results do not provide that clarity.
-In the sections below, we first summarize the evidence that our experiment revealed greater competitive release than our baseline model predicted, then explain weaknesses in the experimental results that complicate the interpretation, and end by discussing general lessons for the study of interspecific interactions.}
+In the sections below, we first summarize the evidence that our experiment revealed greater competitive release than our baseline model predicted, then explain weaknesses in the experimental results that complicate the interpretation, and end by discussing general lessons for the study of interspecific interactions.
 
@@
 
-\new{We evaluated response to removals in three ways: vital rates, short-term population growth rates, and changes in long-term equilibrium cover.
+We evaluated response to removals in three ways: vital rates, short-term population growth rates, and changes in long-term equilibrium cover.
 Two species, \textit{Poa secunda} and \textit{P.
@@
 In other words, our baseline model appeared to underestimate the competitive release of these species following \textit{A.
-tripartita} removal.}
-
-\new{The elevated growth of \textit{Poa secunda} and \textit{P.
+tripartita} removal.
+
+The elevated growth of \textit{Poa secunda} and \textit{P.
 pseudoroegneria} individuals translated into greater short-term population growth rates.
@@
 tripartita}, the magnitude of the difference between predictions from the baseline and removal effects models was similar, further emphasizing the short-term influence of growth compared to survival and recruitment, which responded negatively to removals in this species.
-}
-
-\new{Long-term projections using the integral projection model showed even larger responses to removal.
+
+Long-term projections using the integral projection model showed even larger responses to removal.
 For the three grasses, projected equilibrium cover in the absence of \textit{A.
@@
 While the absolute changes in basal cover are small, these relative changes are substantial, and seem to present compelling evidence that competitive release is stronger than predicted by our baseline model.
-}
 
@@
 
-\new{Although the growth regressions for \textit{Poa secunda} and \textit{P.
+Although the growth regressions for \textit{Poa secunda} and \textit{P.
 pseudoroegneria} and the long-run projections for all three grass species suggest stronger competitive release than our baseline model predicted, a variety of other evidence makes it harder to confidently reject those baseline model predictions.
@@
 pseudoroegneria} the increases in growth were small relative to other sources of variation.
-For example, using quadrat-level cover data, we could not detect treatment differences (Table \ref{table:coefficients}).
-}
-
-\new{The strongest evidence that competitive release of grasses following \textit{A.
+For example, using quadrat-level cover data, we could not detect treatment differences (Appendix S1: Table \ref{table:coefficients}).
+
+The strongest evidence that competitive release of grasses following \textit{A.
 tripartita} removal was greater than predicted comes from the long-term projections of the IPM.
@@
 Given these strong assumptions, the projections should be viewed with some caution.
-}
-
-\new{For \textit{A. tripartita}, the long-term IPM projections actually show extinction following grass removal.
+
+For \textit{A. tripartita}, the long-term IPM projections actually show extinction following grass removal.
 However, this result primarily reflects the influence of recruitment parameters that are highly uncertain (see \emph{Results}).
@@
 Because of this uncertainty, we put more weight on the direct observations and short-term projections showing little effect of grass removal on \textit{A.
-tripartita} populations.}
-
-\new{Another reason we view the growth results with suspicion is that increases in growth were no larger in individual neighborhoods or quadrats with higher pre-treatment \textit{A.
+tripartita} populations.
+
+Another reason we view the growth results with suspicion is that increases in growth were no larger in individual neighborhoods or quadrats with higher pre-treatment \textit{A.
 tripartita} cover.
@@
 tripartita} cover?
-The absence of this relationship (Figs.
-\ref{fig:GrowthResidsQuad} and \ref{fig:GrowthResidsIndivid}) raises questions about the mechanism by which \textit{A.
+The absence of this relationship (Fig.
+\ref{fig:GrowthResidsQuad} and Appendix S1: Fig.
+\ref{fig:GrowthResidsIndivid}) raises questions about the mechanism by which \textit{A.
 tripartita} removal promotes grass growth.
@@
 tripartita's} roots extend far enough beyond its canopy to compete for resources with grasses located in interspaces.
-While this might explain the lack of the relationship at the neighborhood scale, it seems unlikely to explain the quadrat-level results.}
+While this might explain the lack of the relationship at the neighborhood scale, it seems unlikely to explain the quadrat-level results.
+% add bit about distance weights?
 An alternative explanation involves pre-treatment differences in composition between control plots and  the \textit{A.
@@
 tripartita} (see 2011 means in Fig.
-\ref{fig:CoverTrends}) and higher cover of subdominant grasses and shrubs (Fig.
+\ref{fig:CoverTrends}) and higher cover of subdominant grasses and shrubs (Appendix S1: Fig.
 \ref{fig:W-by-treatment}).
@@
 spicata} growth.
-\new{However, since the 1900's, many quadrats have undergone dramatic changes in \textit{A.
+However, since the 1900's, many quadrats have undergone dramatic changes in \textit{A.
 tripartita} cover, arguing against viewing \textit{A.
@@
 Without a better understanding of the mechanism driving increased grass growth in the shrub removal quadrats, 
-we are hesitant to dramatically adjust our prior understanding gained from models fit to decades of observational data.  }
+we are hesitant to dramatically adjust our prior understanding gained from models fit to decades of observational data.
 
@@
 
-\new{A variety of approaches are available for studying interspecific interactions among plants, and each has its advantages and disadvantages.
+A variety of approaches are available for studying interspecific interactions among plants, and each has its advantages and disadvantages.
 Observational datasets have the advantage of realism: they reflect all the processes and interactions that drive community dynamics across a range of spatial and temporal scales.
@@
 Mechanistic studies have the virtue of focusing  directly on the processes of interest, but rarely address more than one limiting resource at a time.
-}
-
-\new{Our previous work was based purely on observational data, so we were particularly concerned about the extrapolation issue.
+
+Our previous work was based purely on observational data, so we were particularly concerned about the extrapolation issue.
 Specifically, our model projects the performance of individual plants growing in the absence of neighbors.
-Although the observational data set seems to include such situations (Fig.
+Although the observational data set seems to include such situations (Appendix S1: Fig.
 \ref{fig:Wscatters}), the data-driven competition kernels we use to calculate neighborhood crowding are quite narrow: they are heavily weighted toward nearby neighbors and essentially ignore the influence of more distant neighbors (Fig.
-\ref{fig:CompKernels}).
+Appendix S1: {fig:CompKernels}).
 However, if our competition kernels under-estimate the strength of long-range interactions, then our model might underestimate the true intrinsic growth rates that would occur when both near and distant neighbors are removed.
+% mention exercise with flat distance weights for ARTR
 We conducted the removal experiment to address this kind of concern, with the hope of gaining the advantages offered by multiple, complementary approaches.
@@
 The reliance on natural variation weakens inference, leaving nagging questions about pre-existing differences among quadrats.
-We could have avoided those concerns by conducting a response surface experiment, planting our study species in randomly selected locations.}
-
-\new{We also learned that partial removals, rather than complete removal treatments, could 
-have helped to distinguish between competition-mediated and other effects of removal treatments (e.g. Fig. \ref{fig:PSSPresids}).
+We could have avoided those concerns by conducting a response surface experiment, planting our study species in randomly selected locations.
+
+We also learned that partial removals, rather than complete removal treatments, could 
+have helped to distinguish between competition-mediated and other effects of removal treatments (e.g. Appendix S1: Fig. \ref{fig:PSSPresids}).
 Partial removals, or experimental gradients in densities, would give us inference about the competition coefficients themselves, whereas our 
@@
 Next time we will also include intraspecific removal treatments.
-}
-
-\new{Our difficulty interpreting our results also made us appreciate the value of mechanistic insight.
+
+Our difficulty interpreting our results also made us appreciate the value of mechanistic insight.
 Even if the removal experiment had perfectly validated the predictions of our phenomenological baseline model, we suspect we would still be asking ourselves, how can interspecific interactions really be so weak?
 The only way to rigorously answer that question is to study the factors--resources, natural enemies, environmental constraints--limiting population growth.
-}
 
@@
 
-\section*{Author contributions}
-PBA and ARK designed the experiment and supervised data collection with the help of JBT; PBA, SPE, GH and BT analyzed data and 
-developed models; PBA wrote the first draft of the manuscript and all authors contributed to editing.
-
 \section*{Acknowledgements}
+PBA and ARK designed the experiment and supervised data collection with the help of JBT; PBA, SPE, GH and BT analyzed data and developed models; PBA wrote the first draft of the manuscript and all authors contributed to editing.
 Funding was provided by NSF grants DEB-1353078 and DEB-1054040 to PBA and DEB-1353039 to SPE and GH, and 
@@
 
-\centerline{\Large \textbf{Appendices}}
-\centerline{Adler et al., ``Weak interspecific interactions''}
+\centerline{\Large \textbf{Appendix S1}}
+\centerline{Adler et al., ``Weak interspecific interactions in a sagebrush steppe? Conflicting evidence from observations and experiments,'' \textit{Ecology}}
 
@@
 
-\renewcommand{\theequation}{A-\arabic{equation}}
-\renewcommand{\thetable}{A-\arabic{table}}
-\renewcommand{\thefigure}{A-\arabic{figure}}
-\renewcommand{\thesection}{\Alph{section}}
+\renewcommand{\theequation}{S\arabic{equation}}
+\renewcommand{\thetable}{S\arabic{table}}
+\renewcommand{\thefigure}{S\arabic{figure}}
+\renewcommand{\thesection}{\arabic{section}}
 
 \section{Detailed Methods} 
-\label{AppendixA}
+\label{AppendixS1}
 
@@
 
-\renewcommand{\theequation}{B-\arabic{equation}}
-\renewcommand{\thetable}{B-\arabic{table}}
-\renewcommand{\thefigure}{B-\arabic{figure}}
-
 \section{Supplementary Tables}
@@
 \end{table}
-
-%\begin{table}
-%\centering
-%\caption{Summary of fixed effects for the \textit{P.
-secunda} growth model with individual-level \textit{A.
-tripartita removal} data (the ``inARTR" coefficient).
-See Table \ref{ARTRgrowth} for an explanation of other coefficient names and column headers.
-}
-%\label{table:POSEgrowth-inARTR}
-%\begin{tabular}{rrrrrrrr}
-%  \hline
-% & mean & sd & 0.025quant & 0.5quant & 0.975quant & mode & kld \\ 
-%  \hline
-%(Intercept) & 0.5080 & 0.0625 & 0.3841 & 0.5081 & 0.6311 & 0.5084 & 0.0000 \\ 
-%  logarea.t0 & 0.6665 & 0.0228 & 0.6211 & 0.6666 & 0.7111 & 0.6669 & 0.0000 \\ 
-%  TreatmentNo\_shrub & 0.2247 & 0.0744 & 0.0786 & 0.2247 & 0.3707 & 0.2247 & 0.0000 \\ 
-%  W.ARTR & -0.0004 & 0.0009 & -0.0022 & -0.0004 & 0.0013 & -0.0004 & 0.0000 \\ 
-%  W.HECO & 0.0109 & 0.0100 & -0.0088 & 0.0109 & 0.0306 & 0.0109 & 0.0000 \\ 
-%  W.POSE & -0.5770 & 0.0624 & -0.6995 & -0.5770 & -0.4547 & -0.5770 & 0.0000 \\ 
-%  W.PSSP & -0.0093 & 0.0114 & -0.0316 & -0.0093 & 0.0130 & -0.0093 & 0.0000 \\ 
-%  W.allcov & -0.0004 & 0.0016 & -0.0036 & -0.0004 & 0.0027 & -0.0004 & 0.0000 \\ 
-%  W.allpts & -0.0340 & 0.0256 & -0.0842 & -0.0340 & 0.0161 & -0.0340 & 0.0000 \\ 
-%  inARTR & -0.0890 & 0.1126 & -0.3100 & -0.0890 & 0.1319 & -0.0890 & 0.0000 \\ 
-%   \hline
-%\end{tabular}
-%\end{table}
-%
-%\begin{table}
-%\centering
-%\caption{Summary of fixed effects for the \textit{P.
-spicata} growth model with individual-level \textit{A.
-tripartita} removal data (the ``inARTR" coefficient).
-See Table \ref{ARTRgrowth} for an explanation of other coefficient names and column headers.}
-%\label{table:PSSPgrowth-inARTR}
-%\begin{tabular}{rrrrrrrr}
-%  \hline
-% & mean & sd & 0.025quant & 0.5quant & 0.975quant & mode & kld \\ 
-%  \hline
-%(Intercept) & 0.3955 & 0.0606 & 0.2764 & 0.3953 & 0.5153 & 0.3950 & 0.0000 \\ 
-%  logarea.t0 & 0.8276 & 0.0168 & 0.7943 & 0.8277 & 0.8608 & 0.8277 & 0.0000 \\ 
-%  TreatmentNo\_shrub & 0.2531 & 0.0784 & 0.0991 & 0.2531 & 0.4070 & 0.2531 & 0.0000 \\ 
-%  W.ARTR & -0.0027 & 0.0008 & -0.0042 & -0.0027 & -0.0011 & -0.0027 & 0.0000 \\ 
-%  W.HECO & -0.0231 & 0.0110 & -0.0447 & -0.0231 & -0.0016 & -0.0231 & 0.0000 \\ 
-%  W.POSE & -0.0198 & 0.0138 & -0.0469 & -0.0198 & 0.0072 & -0.0198 & 0.0000 \\ 
-%  W.PSSP & -0.3600 & 0.0326 & -0.4239 & -0.3600 & -0.2961 & -0.3600 & 0.0000 \\ 
-%  W.allcov & -0.0051 & 0.0014 & -0.0079 & -0.0051 & -0.0024 & -0.0051 & 0.0000 \\ 
-%  W.allpts & -0.0557 & 0.0254 & -0.1055 & -0.0557 & -0.0060 & -0.0557 & 0.0000 \\ 
-%  inARTR & -0.0649 & 0.1111 & -0.2830 & -0.0649 & 0.1531 & -0.0649 & 0.0000 \\ 
-%   \hline
-%\end{tabular}
-%\end{table}
 
@@
 
-\renewcommand{\theequation}{C-\arabic{equation}}
-\renewcommand{\thetable}{C-\arabic{table}}
-\renewcommand{\thefigure}{C-\arabic{figure}}
-
 \section{Supplementary Figures}
@@
 
-%\begin{figure}[h]
-%\centering
-%%\includegraphics[width=1\textwidth]{growth_residuals_vs_quad_removal_cover}
-%\caption{The relationship between residuals from the fitted growth model and quadrat-level pre-treatment cover of the removal species (all grasses removed in quadrats where we studied \textit{A.
-tripartita}, and \textit{A.
-tripartita} removed in the quadrats where we studied the grasses).
-Linear regressions are shown only when statistically significant.
-Source file:  \texttt{growth\textbackslash growth\textunderscore residuals\textunderscore by\textunderscore quad.R}.}
-%\label{fig:GrowthResidsQuad}
-%\end{figure}
-
 \begin{figure}[h]

### 2018-03-23 removals/manuscript/removal_ms.tex
Minor changes to removal ms. final version

@@
 \fi
+
+% cross reference to appendix
+\usepackage{xr}
+\externaldocument{appendix}
 
@@
 
-All data and R code necessary to reproduce our analysis will be deposited in the Dryad Digital Repository once the manuscript is accepted.
-The current version of the computer code and all necessary data are available at \texttt{https://github.com/pbadler/ExperimentTests/tree/master/removals}.
+All data and R code necessary to reproduce our analysis are available on the Dryad Digital Repository (insert url here).
 
@@
 We mapped the new quadrats in June, 2011 and then implemented the removal treatments in September, 2011.
-We randomly assigned 8 of the sixteen new quadrats to the sagebrush removal treatment, and the remaining 8 to the grass removal treatment.
+We randomly assigned eight of the 16 new quadrats to the sagebrush removal treatment, and the remaining eight to the grass removal treatment.
 We cut sagebrush stems at ground level and applied a 5\% solution of Roundup Pro (Glyphosate, Monsanto) to the cut stems.
@@
 The sagebrush removal was virtually 100\% effective; grass removals required additional herbicide applications in April and May of each year, with fewer grasses remaining each year.
-Appendix S1 Fig.
+Appendix S1: Fig.
 \ref{fig:photos} shows photographs of example quadrats before and after treatment.
@@
 The current analysis departs from previous work with the inclusion of a categorical removal treatment effect.
-Although the three vital rate models differ in functional form (see Appendix A for complete model descriptions), all of them have a linear predictor, $\mu$, containing the following terms:
+Although the three vital rate models differ in functional form (see Appendix S1 for complete model descriptions), all of them have a linear predictor, $\mu$, containing the following terms:
 \begin{equation}
@@
 confidence intervals that overlapped zero. Based on the same rationale, we used a deterministic version of the model (e.g., 
-recruitment is the $\lambda$ of Eqn. \ref{eqn:recrProcessModel}, rather than a Poisson random variable with mean $\lambda$). 
+recruitment is the $\lambda$ of Appendix S1: Eqn. \ref{eqn:recrProcessModel}, rather than a Poisson random variable with mean $\lambda$). 
 Specifically, we used the recruitment regression to calculate the expected number of new recruits at time $t+1$, which we then multiplied 
@@
 data, we were suspicious of its ability to accurately describe the outcome of interspecific interactions, so we conducted a removal experiment to test it. 
-Our approach represents a strong test of a surprising prediction for two reasons. First, the prediction was already published; the experiment 
+Our approach represents a strong test of a surprising prediction for two reasons. First, the prediction was already published, so the experiment 
 gave us a very real opportunity to be wrong in public. Second, testing an observationally-based model with an experimental manipulation, 
@@
 
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-% APPENDICES !
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-
-\clearpage 
-\newpage
-
-\setcounter{page}{1}
-\setcounter{equation}{0}
-\setcounter{figure}{0}
-\setcounter{section}{0}
-\setcounter{table}{0}
-
-\centerline{\Large \textbf{Appendix S1}}
-\centerline{Adler et al., ``Weak interspecific interactions in a sagebrush steppe? Conflicting evidence from observations and experiments,'' \textit{Ecology}}
-
-\vspace{0.4in}
-
-\renewcommand{\theequation}{S\arabic{equation}}
-\renewcommand{\thetable}{S\arabic{table}}
-\renewcommand{\thefigure}{S\arabic{figure}}
-\renewcommand{\thesection}{\arabic{section}}
-
-\section{Detailed Methods} 
-\label{AppendixS1}
-
-\subsection{Statistical models of vital rates}
-
-We modeled the survival probability, $S$, of genet $i$ in species $j$ (which is in quadrat group $g$ and removal treatment $h$), 
-from time $t$ to $t+1$  as
-\begin{equation}
-\mbox{logit}(S_{ij,t+1}) = \gamma_{j,t}^S + \varphi_{jg}^S+  \chi_{jh}^S  + \beta_{j,t}^S u_{ij,t} +  
-\sum \limits_{m} \omega_{jm}^S {W}_{ijm,t}
-\label{eqn:survReg}
-\end{equation}
-where $\gamma$ is a time-dependent intercept (year effect), $\varphi$ is the coefficient for the 
-effect of spatial location group, $\chi$ is the removal treatment effect,and  $\beta$ is the coefficient that 
-represents the effect of log genet size, $u$, on survival. 
-$W_{ijm,t}$ is the crowding of individual $i$ in species $j$ by neighbors of species $m$ (explained below), and 
-$\omega_{jm}$ is an interaction coefficient which determines the impact of crowding by species $m$ on the focal species $j$.
-
-Our growth model has a similar structure. The change in genet size from time $t$ to $t+1$ , conditional on survival, is given by:
-\begin{equation}
-u_{ij,t+1} = \gamma_{j,t}^G + \varphi_{jg}^G+  \chi_{jh}^G  + \beta_{j,t}^G u_{ij,t} + 
-\sum \limits_{m} \omega_{jm}^G {W}_{ijm,t} + \varepsilon_{ij,t}^G .
-\label{eqn:growReg}
-\end{equation}
-To capture the non-constant error variance in growth, we modeled the variance $\varepsilon$ about the growth 
-curve (\ref{eqn:growReg}) as a nonlinear function of predicted genet size:
-\begin{equation}
-Var(\varepsilon_{ij,t}^G) = a \exp(b\hat{u}_{ij,t+1}).
-\label{eqn:growVar}
-\end{equation}
-where $\hat{u}_{ij,t+1}$ is the right-hand side of (\ref{eqn:growReg}) without $\varepsilon$.
-
-Our model includes crowding from the four dominant species, \textit{A.
-tripartita}, \textit{H.
-comata}, \textit{P.
-spicata}, and \textit{P.
-secunda}, as well as two covariates representing total crowding from (1) all other perennial grasses and shrubs, which were mapped as polygons, and (2) all forb species, which were mapped as points.
-The next section (\textit{Modeling local crowding}) provides a detailed explanation of how we modeled crowding based on the size of and distance to neighboring plants around each focal genet.
-
-We model recruitment at the quadrat level rather than at the individual genet level because the mapped data do not allow 
-us to determine which recruits were produced by which potential parent genets. We assume that the number of individuals, $y$, of species $j$ recruiting at 
-time $t+1$ in the location $q$ follows a negative binomial distribution:
-\begin{equation}
-y_{jq,t+1}= NegBin(\lambda_{jq,t+1},\theta) 	   
-\label{eqn:recrDataModel}
-\end{equation}
-where $\lambda$ is the mean and $\theta$ is the size parameter (\texttt{dnegbin(p,$\theta$)} in the BUGS language, with $p=\theta/(\theta+\lambda)$). 
-In turn, $\lambda$ depends on the composition of the quadrat in the previous year:
-\begin{equation}
-\lambda_{jq,t+1} = C'_{jq,t} \exp{\left(\gamma_{j,t}^R +  \varphi_{jg}^R + \chi_{jh}^R + 
-\sum \limits_{m} \omega_{jm}^R \sqrt{C'_{mq,t}} \right)}
-\label{eqn:recrProcessModel}
-\end{equation}
-where the superscript $R$ refers to Recruitment, $C'_{jq,t}$ is the `effective cover' (cm$^2$) of species $j$ in quadrat $q$ at time $t$, $\gamma$ 
-is a time-dependent intercept, $\varphi$ is a coefficient for the effect of spatial location group,
-$\chi$ is the removal treatment effect, and $\omega_{jm}$ is the coefficient that determines the effect of species $m$ cover on recruitment by species $j$. 
-Eqn (\ref{eqn:recrProcessModel}) is essentially the Ricker model for discrete time population growth \citep{ricker_stock_1954}, except that we use the square root of cover in the density-dependent term.
-We previously used a Ricker equation \citep{adler_coexistence_2010}, but using the square-root transformation instead gave a lower Deviance Information Criterion (DIC) value \citep{spiegelhalter_bayesian_2002} for the recruitment model.
-Following previous work \citep{adler_coexistence_2010}, we treated year and spatial group variables as random factors, allowing intercepts to vary among years and spatial locations.
-
-Due to the possibility that plants outside the mapped quadrat could contribute recruits to the focal quadrat or interact with plants in the focal quadrat, we estimated effective cover as a mixture of the observed cover, $C$, in the focal quadrat, $q$, and the mean cover, $\bar{C}$, across the spatial location group, $g$, in which the
-quadrat is located: $C'_{jq,t}=p_j C_{jq,t}+(1-p_j) \bar{C}_{jg,t}$, where $p$ is a mixing fraction between 0 and 1 that was estimated as part of fitting the model.
-
-We ran Markov Chain Monte Carlo (MCMC) simulations in WinBUGS 1.4 \citep{lunn_winbugs_2000} to estimate the recruitment model parameters.
-Each model was run for 20,000 iterations and two independent chains with different initial values for parameters.
-We discarded the initial 10,000 samples.
-Convergence was observed graphically for all parameters, and confirmed by the Brooks-Gelman statistic \citep{brooks_general_1998}.
-For all three vital rate models, we considered the effect of removal treatments, or other coefficients of interest, to be significant when the 95\% credible interval (or posterior interval) on the estimate of $\chi$ did not overlap zero.
-
-To incorporate the fitted recruitment function into the IPM, we assumed that individual fecundity increases 
-linearly with individual area (equal to $e^u$ for a size-$u$ individual). The recruitment kernel is thus
-$R_j(v,u,\vec{n})=c_{0,j}(v)e^{u}\Phi_j$ \citep{adler_coexistence_2010} where $c_{0,j}$ is the
-size distribution of new recruits. $\Phi_j$ is calculated from eqn. (\ref{eqn:recrDataModel}) with total cover values computed from the population densities as $C'_{j,t} = \int e^u n_j(u,t) du$.
-
-\subsection{Modeling local crowding}
-We modeled the crowding experienced by a focal genet in each year $t$ as a function of the distance to and size of neighbor genets. In previous work, 
-we assumed that the decay of crowding with neighbor distance followed a Gaussian function \citep{chu_large_2015}, but here we use a data-driven 
-approach \citep{teller_linking_2016}. We model the crowding experienced by genet $i$ of species $j$ from neighbors of species $m$ as the sum of neighbor 
-areas across a set of concentric annuli, $k$, centered at the plant,
-\begin{equation}
-w_{ijmk,t} = F_{jm}(d_{k})A_{ijmk,t}     
-\label{eqn:wik}
-\end{equation}
-where $F_{jm}$ is the competition kernel (described below) for effects of species $m$ on species $j$, 
-$d_{k}$ is the average of the inner and outer radii of annulus $k$, 
-and $A_{ijmk,t}$ is the total area of genets of species $m$ in annulus $k$ around the focal genet in year $t$. The total crowding on 
-the focal genet exerted by species $m$ is
-\begin{equation}
-W_{ijm,t}  =\sum_k {w_{ijmk,t}} .
-\label{eqn:wijm}
-\end{equation} 
-Note that $W_{ijj,t}$ gives intraspecific crowding. The $W$'s are then 
-included as covariates in the survival (\ref{eqn:survReg}) and growth (\ref{eqn:growReg}) regressions.
-
-We assume that competition kernels $F_{jm}(d)$ are non-negative and decreasing, so that distant plants have less effect 
-than close plants. Otherwise, we let the data dictate the shape of the kernel by fitting a spline model 
-using the methods of \citet{teller_linking_2016}. The shape of $F_{jm}$ is determined by a set of spline basis coefficients $\vec{b}_{jm}$
-and a smoothing parameter $\eta$ that controls the complexity of the fitted kernel. 
-Demographic models such as \eqref{eqn:survReg} then have $\gamma$, $\varphi$, $\chi$ , 
-$\beta$, $\boldsymbol{\omega}$, $\vec{b}$ and $\eta$ as parameters to be fitted. We implemented this in the statistical computing environment, \texttt{R}, 
-by making the spline coefficients and $\eta$ the arguments of an objective function that calculates  each $w_{ijmk,t}$ value using the input spline coefficients, 
-calls the model-fitting functions \texttt{lmer} and/or \texttt{glmer} to fit the other parameters in the survival and growth regressions, 
-and returns an approximate AIC value and model degrees of freedom ($df$) for survival and growth combined. We used the $\vec{b}$ values at the smoothest 
-(smallest $df$) local minimum of AIC as a function of $df$, as in \cite{teller_linking_2016}. This approach assumes that one measure of crowding affects 
-survival and growth. In addition, for fitting the kernels we assumed that survival and growth depended only on intraspecific crowding, and thus only fitted the
-within-species competition kernels $F_{jj}$.
-Based on previous work with parametric kernels, \citep{adler_coexistence_2010}, we know that interspecific effects are rarely significantly different from zero, so there is not enough signal to fit interspecific kernels nonparametrically.
-Therefore, we set all $F_{mj}$ equal to $F_{jj}$, meaning that
-the within-species competition kernel for species $j$ is also used to determine the distance-dependence of species $j$ neighbors on all other species. 
-We used data from all historical plots and contemporary control-treatment plots to estimate the competition kernels, which are shown in Fig. \ref{fig:CompKernels}.
-
-Once we had estimated the competitions kernels, we used them to calculate the $w_{ijmk,t}$ values for each individual in each year, 
-and fitted the full survival and growth regressions, which include the interspecific interaction coefficients, $\omega_{jm}$. 
-All genets in a quadrat were included in calculating the $w$'s, but plants located within 5 cm of quadrat edges were not used in fitting the regressions. 
-We fit the models using the \texttt{R-INLA} package for R \citep{rue_approximate_2009}. We used the same approach when exploring variations 
-on the growth regressions, such as the addition of a year-by-treatment interaction or information about the locations of 
-individual grasses relative to removed \textit{A. tripartita}. We also compared the species composition of the control and removal plots at both 
-the quadrat and neighborhood ($W$) scales.
-
-\subsection{Interspecific covariance in local crowding} 
-We explored interspecific covariance in local crowding experienced by individual plants, by regressing the $W$ values exerted by one neighbor species,
- the response variable, against the $W$ values of all other species, the independent variables. Because some $W=0$, we conducted two 
- separate regressions. First, using all $W$'s, we fitted a generalized linear model with a logit link function to evaluate whether the 
- probability that the focal species' $W$ equals 0 is influenced by the value of other species' $W$'s. In this model, the dependent 
- variable is a Bernoulli variate coding for zero versus non-zero value of the focal species' crowding, and the independent 
- variables are the $W$'s for all other species. Second, for the set of records in which the focal species has $W>0$, we 
- performed a linear regression, where the focal species' $W$ is the dependent variable, and the other species' $W$'s are 
- the independent variables. We repeated these regressions for each focal species. Due to large samples size, interspecific $W$ 
- values were often statistically significant predictors of intraspecific. However, they explained very little variance. The 
- maximum reduction in deviance for the generalized linear regressions and $R^2$ for the linear regressions were both less 
- than 8\%. The \texttt{R} code for this analysis is included as ..\texttt{\textbackslash Wdistrib\textbackslash exploreSurvivalWs.r}.
-
-\subsection{Mean field approximation of local crowding for the IPM} 
-\label{sec:kernelMethods} 
-\citet{adler_coexistence_2010} developed a mean field approximation for local crowding when the
-competition kernels are all Gaussian functions, $F_{jm}(d) = e^{-\alpha_{jm} d^2}$. The approximation is explained in 
-the online SI to \citet{adler_coexistence_2010} and in section 5.3 of \citet{Ellner2016}. 
-Here we explain how that approximation was modified for the IPMs in this paper, which
-used fitted nonparametric competition kernels.
-
-The mean field approximation is based on the observed spatial distribution patterns of the species \citep{adler_coexistence_2010}. 
-In both the observed data and IBM simulations, heterospecific individuals were approximately randomly distributed with respect to each other, 
-but conspecific individuals displayed a non-random, size-dependent patterns: small genets were randomly distributed, while large genets 
-were segregated from each other. The overdispersion of large conspecific genets is incorporated into the IPM 
-with a `no-overlap' rule explained below, as in \citep{adler_coexistence_2010}.
-
-For $j \ne m$ (between-species competition), overlap between individuals is allowed. The mean field approximation is 
-that from the perspective of any focal plant in species $j$, individuals of species $m$ are distributed at random in space, 
-independent of each other and of their size.
-
-Consider the region between the circles of radius $x$ and $x+dx$ centered on a focal genet of species $j$. The area of this annulus
-is $2 \pi x \; dx$  to leading order for $dx \approx 0$. A species $m$ genet 
-in the annulus puts competitive pressure $F_{jm}(x)$ times its area on
-the focal genet. The expected total competitive pressure from all such genets 
-is therefore is $F_{jm}(x) 2 \pi x \; dx$ times the expected fractional cover of species $m$ in the annulus 
-(fractional cover is the total area of species $m$ genets, as a fraction of the total area). The excepted fractional cover $C_m$ of species $m$
-in the annulus equals its fractional cover in the habitat as a whole, because of the assumption of random distribution
-spatial distributions. We therefore have $C_m  = \int e^u n_m(u,t) du/A$ where $A$ is the total area of the habitat. 
-The total expected competitive pressure on a species-$j$ genet due to species $m$ is then 
-\begin{equation}
-W_{jm} = \int_0^\infty{C_m F_{jm}(x) 2 \pi x \; dx}  = C_m \left [2 \pi \int_0^{\infty} x F(x) \, dx \right ].
-\label{eqn:wbarm}
-\end{equation} 
-The quantity in square brackets is a constant (that is, it only depends on what the kernel function
-is) so it can be computed once and for all for each kernel used in the IPM. The integral is finite because
-all fitted kernels fall to zero at a finite distance from the focal plant.
-
-Our kernel fitting method only uses competition kernel values at the ``mid-ring'' distances
-halfway between the inner and outer radii of a series of annuli around each focal
-plant, scaled so that the value at the innermost mid-ring distance equals 1. 
-In the IPM we defined the kernel at other distances by linear interpolation between values at 
-mid-ring distances, except that for the innermost ring a kernel value of 1 was specified at the
-outer radius of the ring and at distance $x=0$.
-
-Now consider within-species competition. We assume that conspecifics cannot overlap. Genet shapes are irregular, but we 
-nonetheless implement the no-overlap rule by assuming that a genet of log area $u_i$ is a 
-circle of radius $r_i$ where $\pi r_i^2 = e^{u_i}$. The no-overlap rule is then that the centroids of two conspecific individuals 
-must be separated by at least the sum of their radii.
-
-For any one focal genet, the no-overlap restriction on its neighbors' locations affects 
-only a negligibly small part of the habitat. The expected cover of individuals in the places
-where they can occur (relative to one focal plant) is thus assumed to equal their expected locations
-in the habitat as a whole.
-
-Let $C_m(u)$ be the total cover of species $m$ genets of radius $r$ or smaller, 
-\begin{equation}
-C_m(r) = \int_L^{\log(\pi r^2)}{\! \! \! e^z n_m(z,t) \, dz} .
-\label{eqn:cm}
-\end{equation}
-A focal genet of radius $r$ cannot have any conspecific neighbors centered 
-at distances less than $r$. It can have a neighbor centered at distance $x>r$ if that neighbor's
-radius is no more than $x-r$. Adding up the expected cover of all such possible neighbors
-for a focal genet of radius $r$,    
-\begin{equation}
-W_{mm}(r) = 2 \pi \int_r^{\infty}F_{mm}(x) x C_m(x-r) \, dx
-\label{eqn:wbarmr} 
-\end{equation}
-This integral is again finite and computable because the kernels $F$ fall to 0 at finite $x$.
-
-\clearpage 
-\newpage
-
-\section{Supplementary Tables}
-
-% supplementary tables
-\begin{table}[h]
-\caption{Statistical models of year-to-year changes in log(cover) for the four focal species. Values in square brackets show 95\%
- credible intervals.}
-\centering
-%\begin{center}
-\begin{tabular}{l c c c c }
-\hline
-Species & \textit{A. tripartita} & \textit{H. comata} & \textit{P. secunda} & \textit{P. spicata} \\
-\hline
-(Intercept)           & $-0.03$          & $0.08$           & $-0.14$          & $0.02$           \\
-                      & $[-0.48;\ 0.41]$ & $[-0.14;\ 0.30]$ & $[-0.56;\ 0.27]$ & $[-0.24;\ 0.29]$ \\
-TreatmentNo\_grass    & $-0.05$          &                  &                  &                  \\
-                      & $[-0.74;\ 0.64]$ &                  &                  &                  \\
-TreatmentNo\_shrub    &                  & $0.24$           & $-0.05$          & $0.18$           \\
-                      &                  & $[-0.12;\ 0.60]$ & $[-0.35;\ 0.25]$ & $[-0.01;\ 0.38]$ \\
-\hline
-%AIC                   & 227.74           & 114.65           & 205.41           & 168.17           \\
-%BIC                   & 240.01           & 124.69           & 218.18           & 181.40           \\
-Log Likelihood        & -108.87          & -52.33           & -97.70           & -79.09           \\
-Num. obs.             & 86               & 55               & 95               & 104              \\
-Num. groups: quad     & 19               & 11               & 21               & 22               \\
-Num. groups: year     & 5                & 5                & 5                & 5                \\
-Var: quad (Intercept) & 0.44             & 0.00             & 0.02             & 0.00             \\
-Var: year (Intercept) & 0.00             & 0.02             & 0.19             & 0.07             \\
-Var: Residual         & 0.52             & 0.36             & 0.38             & 0.24             \\
-\hline
-\multicolumn{5}{l}{\scriptsize{$^*$ 0 outside the credible interval}}
-\end{tabular}
-\label{table:coefficients}
-%\end{center}
-\end{table}
-
-% latex table generated in R 3.2.2 by xtable 1.8-2 package
-% Mon Sep 19 08:32:17 2016
-\begin{table}[ht]
-\centering
-\caption{Summary of posterior distributions for fixed effects for the \textit{A.
-tripartita} survival model.
-``logarea" is the effect of plant size, ``Treatment*" is the removal effect, and the ``W.*" coefficients are effects of neighborhood crowding (the $\omega$s in eqn.
-\ref{eqn:survReg}).
-``quant" refers to quantile and ``kld" reports the Kullback-Leibler divergence between the Gaussian and the (simplified) Laplace approximation to the marginal posterior densities.
-}
-\label{ARTRsurvival}
-\begin{tabular}{rrrrrrrr}
-  \hline
- & mean & sd & 0.025quant & 0.5quant & 0.975quant & mode & kld \\ 
-  \hline
-(Intercept) & -0.2613 & 0.1344 & -0.5276 & -0.2605 & 0.0001 & -0.2588 & 0.0000 \\ 
-  logarea & 0.7093 & 0.0400 & 0.6315 & 0.7090 & 0.7885 & 0.7085 & 0.0000 \\ 
-  TreatmentNo\_grass & -0.9994 & 0.6453 & -2.2156 & -1.0182 & 0.3228 & -1.0565 & 0.0000 \\ 
-  W.ARTR & -2.8522 & 0.4187 & -3.6968 & -2.8443 & -2.0513 & -2.8284 & 0.0000 \\ 
-  W.HECO & -0.0720 & 0.0615 & -0.1856 & -0.0747 & 0.0565 & -0.0804 & 0.0000 \\ 
-  W.POSE & 0.0617 & 0.0895 & -0.1061 & 0.0588 & 0.2459 & 0.0528 & 0.0000 \\ 
-  W.PSSP & 0.0387 & 0.0475 & -0.0500 & 0.0370 & 0.1369 & 0.0335 & 0.0000 \\ 
-  W.allcov & -0.0065 & 0.0109 & -0.0282 & -0.0065 & 0.0148 & -0.0063 & 0.0000 \\ 
-  W.allpts & 0.1293 & 0.1653 & -0.1956 & 0.1294 & 0.4533 & 0.1296 & 0.0000 \\ 
-   \hline
-\end{tabular}
-\end{table}
-
-% latex table generated in R 3.2.2 by xtable 1.8-2 package
-% Mon Sep 19 08:33:00 2016
-\begin{table}[ht]
-\centering
-\caption{Summary of posterior distributions for fixed effects for the \textit{H. comata} survival model. See Table \ref{ARTRsurvival} for an explanation of coefficient names
-and column headers.} 
-\label{HECOsurvival}
-\begin{tabular}{rrrrrrrr}
-  \hline
- & mean & sd & 0.025quant & 0.5quant & 0.975quant & mode & kld \\ 
-  \hline
-(Intercept) & 1.4753 & 0.2011 & 1.0764 & 1.4753 & 1.8736 & 1.4753 & 0.0000 \\ 
-  logarea & 1.1991 & 0.0747 & 1.0622 & 1.1954 & 1.3561 & 1.1874 & 0.0000 \\ 
-  TreatmentNo\_shrub & 0.2915 & 0.3900 & -0.4693 & 0.2898 & 1.0610 & 0.2864 & 0.0000 \\ 
-  W.ARTR & -0.0063 & 0.0030 & -0.0121 & -0.0063 & -0.0005 & -0.0063 & 0.0000 \\ 
-  W.HECO & -1.6962 & 0.1458 & -1.9894 & -1.6939 & -1.4163 & -1.6891 & 0.0000 \\ 
-  W.POSE & 0.0649 & 0.0563 & -0.0442 & 0.0644 & 0.1768 & 0.0634 & 0.0000 \\ 
-  W.PSSP & 0.0313 & 0.0360 & -0.0381 & 0.0309 & 0.1031 & 0.0301 & 0.0000 \\ 
-  W.allcov & -0.0012 & 0.0062 & -0.0135 & -0.0012 & 0.0110 & -0.0012 & 0.0000 \\ 
-  W.allpts & -0.1295 & 0.1004 & -0.3273 & -0.1293 & 0.0670 & -0.1290 & 0.0000 \\ 
-   \hline
-\end{tabular}
-\end{table}
-
-% latex table generated in R 3.2.2 by xtable 1.8-2 package
-% Mon Sep 19 08:34:30 2016
-\begin{table}[ht]
-\centering
-\caption{Summary of posterior distributions for fixed effects for the \textit{P. secunda} survival model. See Table \ref{ARTRsurvival} for an explanation of coefficient names
-and column headers.} 
-\label{POSEsurvival}
-\begin{tabular}{rrrrrrrr}
-  \hline
- & mean & sd & 0.025quant & 0.5quant & 0.975quant & mode & kld \\ 
-  \hline
-(Intercept) & 1.4020 & 0.1922 & 1.0176 & 1.4031 & 1.7793 & 1.4050 & 0.0000 \\ 
-  logarea & 1.0362 & 0.0645 & 0.9139 & 1.0343 & 1.1696 & 1.0307 & 0.0000 \\ 
-  TreatmentNo\_shrub & -0.1604 & 0.1622 & -0.4784 & -0.1606 & 0.1583 & -0.1610 & 0.0000 \\ 
-  W.ARTR & 0.0000 & 0.0020 & -0.0038 & 0.0000 & 0.0039 & -0.0000 & 0.0000 \\ 
-  W.HECO & -0.0224 & 0.0208 & -0.0631 & -0.0225 & 0.0185 & -0.0227 & 0.0000 \\ 
-  W.POSE & -2.0073 & 0.1112 & -2.2290 & -2.0062 & -1.7921 & -2.0038 & 0.0000 \\ 
-  W.PSSP & 0.0468 & 0.0238 & 0.0004 & 0.0466 & 0.0939 & 0.0463 & 0.0000 \\ 
-  W.allcov & -0.0061 & 0.0035 & -0.0130 & -0.0061 & 0.0008 & -0.0061 & 0.0000 \\ 
-  W.allpts & -0.0383 & 0.0567 & -0.1494 & -0.0384 & 0.0730 & -0.0385 & 0.0000 \\ 
-   \hline
-\end{tabular}
-\end{table}
-
-% latex table generated in R 3.2.2 by xtable 1.8-2 package
-% Mon Sep 19 08:35:48 2016
-\begin{table}[ht]
-\centering
-\caption{Summary of posterior distributions for fixed effects for the \textit{P. spicata} survival model. See Table \ref{ARTRsurvival} for an explanation of coefficient names
-and column headers.} 
-\label{PSSPsurvival}
-\begin{tabular}{rrrrrrrr}
-  \hline
- & mean & sd & 0.025quant & 0.5quant & 0.975quant & mode & kld \\ 
-  \hline
-(Intercept) & 1.1978 & 0.1639 & 0.8783 & 1.1962 & 1.5257 & 1.1933 & 0.0000 \\ 
-  logarea & 1.5138 & 0.0935 & 1.3336 & 1.5122 & 1.7029 & 1.5092 & 0.0000 \\ 
-  TreatmentNo\_shrub & -0.2466 & 0.2023 & -0.6454 & -0.2460 & 0.1489 & -0.2449 & 0.0000 \\ 
-  W.ARTR & 0.0099 & 0.0022 & 0.0057 & 0.0099 & 0.0142 & 0.0099 & 0.0000 \\ 
-  W.HECO & 0.0025 & 0.0289 & -0.0550 & 0.0028 & 0.0586 & 0.0032 & 0.0000 \\ 
-  W.POSE & 0.0517 & 0.0389 & -0.0222 & 0.0508 & 0.1308 & 0.0490 & 0.0000 \\ 
-  W.PSSP & -1.1267 & 0.0739 & -1.2743 & -1.1258 & -0.9838 & -1.1242 & 0.0000 \\ 
-  W.allcov & 0.0131 & 0.0037 & 0.0060 & 0.0131 & 0.0203 & 0.0131 & 0.0000 \\ 
-  W.allpts & 0.0756 & 0.0658 & -0.0540 & 0.0757 & 0.2045 & 0.0760 & 0.0000 \\ 
-   \hline
-\end{tabular}
-\end{table}
-
-% latex table generated in R 3.2.2 by xtable 1.8-2 package
-% Mon Sep 19 08:36:00 2016
-\begin{table}[ht]
-\centering
-\caption{Summary of posterior distributions for fixed effects for the \textit{A.
-tripartita} growth model.
-``logarea.t0" is the effect of plant size, ``Treatment*" is the removal effect, and the ``W.*" coefficients are effects of neighborhood crowding (the $\omega$s in eqn.
-\ref{eqn:growReg}.
-``quant" refers to quantile and ``kld" reports the Kullback-Leibler divergence between the Gaussian and the (simplified) Laplace approximation to the marginal posterior densities.}
-\label{ARTRgrowth}
-\begin{tabular}{rrrrrrrr}
-  \hline
- & mean & sd & 0.025quant & 0.5quant & 0.975quant & mode & kld \\ 
-  \hline
-(Intercept) & 0.7092 & 0.2385 & 0.2396 & 0.7091 & 1.1784 & 0.7090 & 0.0000 \\ 
-  logarea.t0 & 0.8689 & 0.0394 & 0.7912 & 0.8689 & 0.9466 & 0.8689 & 0.0000 \\ 
-  TreatmentNo\_grass & 0.1385 & 0.1446 & -0.1453 & 0.1385 & 0.4221 & 0.1385 & 0.0000 \\ 
-  W.ARTR & -0.1497 & 0.1960 & -0.5345 & -0.1497 & 0.2347 & -0.1497 & 0.0000 \\ 
-  W.HECO & 0.0045 & 0.0198 & -0.0343 & 0.0045 & 0.0433 & 0.0045 & 0.0000 \\ 
-  W.POSE & -0.0614 & 0.0308 & -0.1218 & -0.0614 & -0.0010 & -0.0614 & 0.0000 \\ 
-  W.PSSP & -0.0050 & 0.0158 & -0.0360 & -0.0050 & 0.0260 & -0.0050 & 0.0000 \\ 
-  W.allcov & -0.0052 & 0.0041 & -0.0133 & -0.0052 & 0.0029 & -0.0052 & 0.0000 \\ 
-  W.allpts & 0.0115 & 0.0536 & -0.0938 & 0.0115 & 0.1167 & 0.0115 & 0.0000 \\ 
-   \hline
-\end{tabular}
-\end{table}
-
-% latex table generated in R 3.2.2 by xtable 1.8-2 package
-% Mon Sep 19 08:36:24 2016
-\begin{table}[ht]
-\centering
-\caption{Summary of posterior distributions for fixed effects for the \textit{H. comata} growth model. See Table \ref{ARTRgrowth} for an explanation of coefficient names
-and column headers.} 
-\label{HECOgrowth}
-\begin{tabular}{rrrrrrrr}
-  \hline
- & mean & sd & 0.025quant & 0.5quant & 0.975quant & mode & kld \\ 
-  \hline
-(Intercept) & 0.3937 & 0.0815 & 0.2330 & 0.3938 & 0.5537 & 0.3940 & 0.0000 \\ 
-  logarea.t0 & 0.8207 & 0.0213 & 0.7788 & 0.8207 & 0.8626 & 0.8207 & 0.0000 \\ 
-  TreatmentNo\_shrub & 0.0691 & 0.1314 & -0.1891 & 0.0691 & 0.3268 & 0.0691 & 0.0000 \\ 
-  W.ARTR & -0.0042 & 0.0011 & -0.0063 & -0.0042 & -0.0020 & -0.0042 & 0.0000 \\ 
-  W.HECO & -0.2134 & 0.0618 & -0.3348 & -0.2134 & -0.0922 & -0.2135 & 0.0000 \\ 
-  W.POSE & 0.0204 & 0.0178 & -0.0146 & 0.0204 & 0.0553 & 0.0204 & 0.0000 \\ 
-  W.PSSP & -0.0373 & 0.0131 & -0.0630 & -0.0373 & -0.0117 & -0.0373 & 0.0000 \\ 
-  W.allcov & -0.0062 & 0.0027 & -0.0115 & -0.0062 & -0.0010 & -0.0062 & 0.0000 \\ 
-  W.allpts & -0.0212 & 0.0416 & -0.1028 & -0.0212 & 0.0604 & -0.0212 & 0.0000 \\ 
-   \hline
-\end{tabular}
-\end{table}
-
-% latex table generated in R 3.2.2 by xtable 1.8-2 package
-% Mon Sep 19 08:36:58 2016
-\begin{table}[ht]
-\centering
-\caption{Summary of posterior distributions for fixed effects for the \textit{Poa secunda} growth model. See Table \ref{ARTRgrowth} for an explanation of coefficient names
-and column headers.} 
-\label{POSEgrowth}
-\begin{tabular}{rrrrrrrr}
-  \hline
- & mean & sd & 0.025quant & 0.5quant & 0.975quant & mode & kld \\ 
-  \hline
-(Intercept) & 0.5069 & 0.0625 & 0.3830 & 0.5070 & 0.6299 & 0.5072 & 0.0000 \\ 
-  logarea.t0 & 0.6669 & 0.0228 & 0.6215 & 0.6670 & 0.7116 & 0.6673 & 0.0000 \\ 
-  TreatmentNo\_shrub & 0.1937 & 0.0632 & 0.0696 & 0.1937 & 0.3176 & 0.1937 & 0.0000 \\ 
-  W.ARTR & -0.0004 & 0.0009 & -0.0022 & -0.0004 & 0.0013 & -0.0004 & 0.0000 \\ 
-  W.HECO & 0.0110 & 0.0100 & -0.0087 & 0.0110 & 0.0307 & 0.0110 & 0.0000 \\ 
-  W.POSE & -0.5753 & 0.0623 & -0.6977 & -0.5753 & -0.4530 & -0.5753 & 0.0000 \\ 
-  W.PSSP & -0.0093 & 0.0114 & -0.0316 & -0.0093 & 0.0130 & -0.0093 & 0.0000 \\ 
-  W.allcov & -0.0004 & 0.0016 & -0.0035 & -0.0004 & 0.0028 & -0.0004 & 0.0000 \\ 
-  W.allpts & -0.0340 & 0.0255 & -0.0842 & -0.0340 & 0.0161 & -0.0340 & 0.0000 \\ 
-   \hline
-\end{tabular}
-\end{table}
-
-% latex table generated in R 3.2.2 by xtable 1.8-2 package
-% Mon Sep 19 08:37:48 2016
-\begin{table}[ht]
-\centering
-\caption{Summary of posterior distributions for fixed effects for the \textit{P. spicata} growth model. See Table \ref{ARTRgrowth} for an explanation of coefficient names
-and column headers.} 
-\label{PSSPgrowth}
-\begin{tabular}{rrrrrrrr}
-  \hline
- & mean & sd & 0.025quant & 0.5quant & 0.975quant & mode & kld \\ 
-  \hline
-(Intercept) & 0.3846 & 0.0710 & 0.2425 & 0.3851 & 0.5236 & 0.3858 & 0.0000 \\ 
-  logarea.t0 & 0.8259 & 0.0167 & 0.7929 & 0.8259 & 0.8587 & 0.8260 & 0.0000 \\ 
-  TreatmentNo\_shrub & 0.2262 & 0.0681 & 0.0924 & 0.2262 & 0.3598 & 0.2262 & 0.0000 \\ 
-  W.ARTR & -0.0023 & 0.0008 & -0.0038 & -0.0023 & -0.0007 & -0.0023 & 0.0000 \\ 
-  W.HECO & -0.0201 & 0.0111 & -0.0419 & -0.0201 & 0.0017 & -0.0201 & 0.0000 \\ 
-  W.POSE & -0.0200 & 0.0137 & -0.0469 & -0.0200 & 0.0069 & -0.0200 & 0.0000 \\ 
-  W.PSSP & -0.3845 & 0.0335 & -0.4502 & -0.3845 & -0.3188 & -0.3846 & 0.0000 \\ 
-  W.allcov & -0.0048 & 0.0014 & -0.0076 & -0.0048 & -0.0021 & -0.0048 & 0.0000 \\ 
-  W.allpts & -0.0510 & 0.0256 & -0.1012 & -0.0510 & -0.0008 & -0.0510 & 0.0000 \\ 
-   \hline
-\end{tabular}
-\end{table}
-
-\begin{table}
-\centering
-\caption{Summary of fixed effects for the \textit{P.
-secunda} growth model with treatment*year effects (the ``trtYears*" coefficients).
-See Table \ref{ARTRgrowth} for an explanation of other coefficient names and column header.}
-\label{table:POSEgrowth-trtYears}
-\begin{tabular}{rrrrrrrr}
-  \hline
- & mean & sd & 0.025quant & 0.5quant & 0.975quant & mode & kld \\ 
-  \hline
-(Intercept) & 0.5079 & 0.0626 & 0.3839 & 0.5080 & 0.6310 & 0.5082 & 0.0000 \\ 
-  trtYears1 & 0.0910 & 0.1169 & -0.1386 & 0.0910 & 0.3203 & 0.0910 & 0.0000 \\ 
-  trtYears2 & 0.1560 & 0.1212 & -0.0819 & 0.1560 & 0.3937 & 0.1560 & 0.0000 \\ 
-  trtYears3 & 0.1891 & 0.1461 & -0.0976 & 0.1891 & 0.4757 & 0.1891 & 0.0000 \\ 
-  trtYears4 & 0.3125 & 0.1608 & -0.0032 & 0.3125 & 0.6280 & 0.3126 & 0.0000 \\ 
-  trtYears5 & 0.3737 & 0.1782 & 0.0239 & 0.3737 & 0.7233 & 0.3737 & 0.0000 \\ 
-  logarea.t0 & 0.6663 & 0.0228 & 0.6207 & 0.6664 & 0.7111 & 0.6666 & 0.0000 \\ 
-  W.ARTR & -0.0005 & 0.0009 & -0.0022 & -0.0005 & 0.0013 & -0.0005 & 0.0000 \\ 
-  W.HECO & 0.0108 & 0.0100 & -0.0089 & 0.0108 & 0.0305 & 0.0108 & 0.0000 \\ 
-  W.POSE & -0.5725 & 0.0624 & -0.6950 & -0.5725 & -0.4501 & -0.5725 & 0.0000 \\ 
-  W.PSSP & -0.0100 & 0.0114 & -0.0323 & -0.0100 & 0.0123 & -0.0100 & 0.0000 \\ 
-  W.allcov & -0.0004 & 0.0016 & -0.0036 & -0.0004 & 0.0028 & -0.0004 & 0.0000 \\ 
-  W.allpts & -0.0347 & 0.0256 & -0.0849 & -0.0347 & 0.0155 & -0.0347 & 0.0000 \\ 
-   \hline
-\end{tabular}
-\end{table}
-
-\begin{table}
-\centering
-\caption{Summary of fixed effects for the \textit{P.
-spicata} growth model with treatment*year effects (the ``trtYears*" coefficients).
-See Table \ref{ARTRgrowth} for an explanation of other coefficient names and column headers.}
-\label{table:PSSPgrowth-trtYears}
-\begin{tabular}{rrrrrrrr}
-  \hline
- & mean & sd & 0.025quant & 0.5quant & 0.975quant & mode & kld \\ 
-  \hline
-(Intercept) & 0.3948 & 0.0606 & 0.2757 & 0.3946 & 0.5149 & 0.3943 & 0.0000 \\ 
-  trtYears1 & 0.3434 & 0.1341 & 0.0800 & 0.3434 & 0.6065 & 0.3434 & 0.0000 \\ 
-  trtYears2 & -0.0514 & 0.1391 & -0.3245 & -0.0514 & 0.2215 & -0.0513 & 0.0000 \\ 
-  trtYears3 & 0.3547 & 0.1493 & 0.0616 & 0.3547 & 0.6476 & 0.3547 & 0.0000 \\ 
-  trtYears4 & 0.5487 & 0.1519 & 0.2505 & 0.5487 & 0.8467 & 0.5487 & 0.0000 \\ 
-  trtYears5 & -0.0575 & 0.1598 & -0.3711 & -0.0575 & 0.2560 & -0.0575 & 0.0000 \\ 
-  logarea.t0 & 0.8284 & 0.0168 & 0.7951 & 0.8284 & 0.8614 & 0.8285 & 0.0000 \\ 
-  W.ARTR & -0.0026 & 0.0008 & -0.0042 & -0.0026 & -0.0011 & -0.0026 & 0.0000 \\ 
-  W.HECO & -0.0234 & 0.0109 & -0.0447 & -0.0234 & -0.0020 & -0.0234 & 0.0000 \\ 
-  W.POSE & -0.0195 & 0.0137 & -0.0463 & -0.0195 & 0.0073 & -0.0195 & 0.0000 \\ 
-  W.PSSP & -0.3590 & 0.0322 & -0.4223 & -0.3590 & -0.2957 & -0.3590 & 0.0000 \\ 
-  W.allcov & -0.0053 & 0.0014 & -0.0081 & -0.0053 & -0.0026 & -0.0053 & 0.0000 \\ 
-  W.allpts & -0.0567 & 0.0251 & -0.1060 & -0.0567 & -0.0074 & -0.0567 & 0.0000 \\ 
-   \hline
-\end{tabular}
-\end{table}
-
-% latex table generated in R 3.2.2 by xtable 1.8-2 package
-% Mon Sep 19 08:37:48 2016
-\begin{table}[ht]
-\centering
-\caption{Summary of posterior distributions for fixed effects for the recruitment model (symbols correspond to Eqns.
-\ref{eqn:recrDataModel} and \ref{eqn:recrProcessModel}).
-The indexing on $\omega$ shown here and used in the computer code for the recruitment model gives $\omega$[3,1] as the effect of species 3 on species 1.
-This is the reverse of the normal convention used in \ref{eqn:recrProcessModel} and shown in \ref{table:alphas}.}
-\label{table:recruitment}
-\begin{tabular}{rrrrrrr}
-  \hline
- & mean & sd & X2.5. & X97.5. & Rhat & n.eff \\ 
-  \hline
- $\gamma$[1] & 0.3341 & 0.7071 & -1.1262 & 1.6391 & 1.0018 &  1100 \\ 
-  $\gamma$[2] & 3.4488 & 0.4468 & 2.5010 & 4.3401 & 1.0904 &    25 \\ 
-  $\gamma$[3] & 3.2440 & 0.3650 & 2.4840 & 3.9190 & 1.0536 &    34 \\ 
- $\gamma$[4] & 2.9054 & 0.3689 & 2.1820 & 3.6160 & 1.0149 &   110 \\ 
-  $\chi$[2,2] & -0.0953 & 0.4093 & -0.8854 & 0.7570 & 1.0010 &  2000 \\ 
- $\chi$[2,3] & -1.2787 & 0.3325 & -1.9370 & -0.6392 & 1.0041 &   420 \\ 
-   $\chi$[2,4] & 0.0951 & 0.2617 & -0.4121 & 0.6063 & 1.0023 &   810 \\ 
-  $\chi$[3,1] & -1.4366 & 0.8544 & -3.1491 & 0.1132 & 1.0006 &  2000 \\ 
-  $\omega$[1,1] & -0.5881 & 0.1018 & -0.7639 & -0.3720 & 1.0197 &    82 \\ 
-  $\omega$[1,2] & 0.0635 & 0.0651 & -0.0528 & 0.1997 & 1.0329 &    59 \\ 
-  $\omega$[1,3] & 0.0220 & 0.0422 & -0.0610 & 0.1040 & 1.0036 &   920 \\ 
-  $\omega$[1,4] & 0.1164 & 0.0456 & 0.0347 & 0.2123 & 1.0197 &    96 \\ 
-  $\omega$[2,1] & -0.4800 & 0.1522 & -0.7746 & -0.1906 & 1.0029 &  2000 \\ 
-  $\omega$[2,2] & -1.7368 & 0.1297 & -1.9841 & -1.4819 & 1.0006 &  2000 \\ 
-  $\omega$[2,3] & 0.0388 & 0.0865 & -0.1287 & 0.2138 & 1.0152 &   100 \\ 
-  $\omega$[2,4] & -0.3682 & 0.0926 & -0.5434 & -0.1881 & 1.0005 &  2000 \\ 
-  $\omega$[3,1] & -0.6589 & 0.3342 & -1.3131 & -0.0155 & 1.0096 &   280 \\ 
-  $\omega$[3,2] & -0.1021 & 0.1981 & -0.4828 & 0.2936 & 1.0089 &   210 \\ 
-  $\omega$[3,3] & -1.8943 & 0.1611 & -2.2060 & -1.5850 & 1.0084 &  2000 \\ 
-  $\omega$[3,4] & -0.1953 & 0.1508 & -0.4968 & 0.0995 & 1.0074 &  2000 \\ 
-  $\omega$[4,1] & -0.1867 & 0.3000 & -0.7608 & 0.4486 & 1.0099 &   190 \\ 
-  $\omega$[4,2] & -0.3746 & 0.1896 & -0.7224 & -0.0115 & 1.0453 &    55 \\ 
-  $\omega$[4,3] & 0.1131 & 0.1437 & -0.1583 & 0.4170 & 1.0321 &    55 \\ 
-  $\omega$[4,4] & -1.7379 & 0.1549 & -2.0400 & -1.4409 & 1.0116 &  1400 \\ 
-   $\theta$[1] & 0.6198 & 0.0740 & 0.4835 & 0.7766 & 1.0008 &  2000 \\ 
-   $\theta$[2] & 1.1197 & 0.1428 & 0.8666 & 1.4450 & 1.0008 &  2000 \\ 
-   $\theta$[3] & 1.1718 & 0.1068 & 0.9685 & 1.3890 & 1.0008 &  2000 \\ 
-  $\theta$[4] & 1.0961 & 0.1060 & 0.9009 & 1.3280 & 1.0022 &   890 \\ 
-  $p$[1] & 0.8137 & 0.1228 & 0.4813 & 0.9525 & 1.0399 &    60 \\ 
-  $p$[2] & 0.9996 & 0.0005 & 0.9984 & 1.0000 & 1.0007 &  2000 \\ 
-  $p$[3] & 0.7906 & 0.1288 & 0.4273 & 0.9382 & 1.0176 &  2000 \\ 
-  $p$[4] & 0.7502 & 0.1480 & 0.3920 & 0.9453 & 1.0150 &  2000 \\ 
-   \hline
-
-   \hline
-    $\gamma$[1] & 0.2843 & 0.6342 & -1.0900 & 1.4950 & 1.0279 &   810 \\ 
-    $\gamma$[2] & 3.2828 & 0.4467 & 2.4429 & 4.1940 & 1.0062 &   260 \\ 
-    $\gamma$[3] & 3.2923 & 0.3691 & 2.5680 & 4.0320 & 1.0040 &  1800 \\ 
-    $\gamma$[4] & 2.9362 & 0.3811 & 2.1779 & 3.6462 & 1.0116 &   140 \\ 
-    $\chi$[2,2] & 0.0405 & 0.4077 & -0.7392 & 0.8147 & 1.0011 &  2000 \\ 
-    $\chi$[2,3] & -0.1396 & 0.2658 & -0.6654 & 0.3970 & 1.0015 &  1400 \\ 
-    $\chi$[2,4] & 0.4050 & 0.2439 & -0.0664 & 0.8924 & 1.0010 &  2000 \\ 
-    $\chi$[3,1] & -1.3691 & 0.6755 & -2.7260 & -0.0685 & 1.0093 &   170 \\ 
-      $\omega$[1,1] & -0.5539 & 0.1165 & -0.7517 & -0.2845 & 1.0329 &   240 \\ 
-      $\omega$[1,2] & 0.0654 & 0.0581 & -0.0429 & 0.1898 & 1.0025 &   740 \\ 
-      $\omega$[1,3] & 0.0043 & 0.0389 & -0.0771 & 0.0813 & 1.0019 &  2000 \\ 
-      $\omega$[1,4] & 0.0953 & 0.0410 & 0.0209 & 0.1830 & 1.0120 &   130 \\ 
-      $\omega$[2,1] & -0.4522 & 0.1533 & -0.7461 & -0.1556 & 1.0007 &  2000 \\ 
-      $\omega$[2,2] & -1.5975 & 0.1175 & -1.8260 & -1.3669 & 1.0033 &  2000 \\ 
-      $\omega$[2,3] & -0.0058 & 0.0866 & -0.1806 & 0.1662 & 1.0035 &  2000 \\ 
-      $\omega$[2,4] & -0.3047 & 0.0858 & -0.4794 & -0.1381 & 1.0025 &   740 \\ 
-      $\omega$[3,1] & -0.5030 & 0.2801 & -1.0880 & 0.0341 & 1.0185 &   120 \\ 
-      $\omega$[3,2] & -0.0586 & 0.1571 & -0.3722 & 0.2399 & 1.0053 &   320 \\ 
-      $\omega$[3,3] & -1.7052 & 0.1471 & -1.9770 & -1.3980 & 1.0006 &  2000 \\ 
-      $\omega$[3,4] & -0.0960 & 0.1304 & -0.3682 & 0.1443 & 1.0005 &  2000 \\ 
-      $\omega$[4,1] & -0.1253 & 0.2688 & -0.6245 & 0.3921 & 1.0028 &  2000 \\ 
-      $\omega$[4,2] & -0.3056 & 0.1911 & -0.7052 & 0.0560 & 1.0041 &   420 \\ 
-      $\omega$[4,3] & 0.0335 & 0.1386 & -0.2499 & 0.2999 & 1.0040 &   440 \\ 
-      $\omega$[4,4] & -1.7046 & 0.1597 & -2.0080 & -1.3900 & 1.0101 &   250 \\ 
-     $\theta$[1] & 0.6117 & 0.0708 & 0.4855 & 0.7560 & 1.0006 &  2000 \\ 
-     $\theta$[2] & 1.1862 & 0.1462 & 0.9392 & 1.5000 & 1.0014 &  2000 \\ 
-     $\theta$[3] & 1.1782 & 0.1043 & 0.9872 & 1.4080 & 1.0009 &  2000 \\ 
-     $\theta$[4] & 1.2081 & 0.1060 & 1.0090 & 1.4210 & 1.0006 &  2000 \\ 
-     $p$[1] & 0.8367 & 0.1095 & 0.5280 & 0.9604 & 1.0235 &  2000 \\ 
-     $p$[2] & 0.9995 & 0.0006 & 0.9979 & 1.0000 & 1.0006 &  2000 \\ 
-     $p$[3] & 0.8834 & 0.0634 & 0.7124 & 0.9619 & 1.0054 &   410 \\ 
-     $p$[4] & 0.8100 & 0.1410 & 0.4269 & 0.9650 & 1.0023 &   830 \\ 
-      \hline
-\end{tabular}
-\end{table}
-
-\newpage 
-\begin{table}[ht]
-\centering
-\caption{Summary of competition coefficients in the demographic models. These are the ``W.abcd'' fixed effects coefficients from
-the Tables in this section summarizing the survival, growth and recruitment models for each species.  
-In each $4 \times 4$ array of coefficients, the entry in row $i$, column $j$ is the 
-coefficient for the effect of species $j$ cover on species $i$. Intraspecific competition is weak relative to intraspecific competition if each
-species affects itself more than others, i.e. if the largest (in magnitude) element in each column is the element on the diagonal of the array.
-The one exception to this pattern is the effect of POSE on recruitment by ARTR, but the nonconforming
-coefficient ($\omega$[3,1] = -0.50 in Table \ref{table:recruitment}) has low precision (sd = 0.28, Table \ref{table:recruitment}.} 
-\label{table:alphas}
-\begin{tabular}{lrrrr}
-       & ARTR &  HECO &  POSE &  PSSP \\
-\hline 
-\underline{Survival} & & & & \\        
-ARTR & -2.8522 & -0.0720 & 0.0617 & 0.0387\\
-HECO & -0.0063 & -1.6962 & 0.0649 & 0.0313\\
-POSE & 0.0000 & -0.0224 & -2.0073 & 0.0468\\
-PSSP & 0.0099 & 0.0025 & 0.0517 & -1.1267\\
-\underline{Growth} & & & & \\ 
-ARTR & -0.1497 & 0.0045 & -0.0614 & -0.0050\\
-HECO & -0.0042 & -0.2134 & 0.0204 & -0.0373\\
-POSE & -0.0004 & 0.0110 & -0.5753 & -0.0093\\
-PSSP & -0.0023 & -0.0201 & -0.0200 & -0.3845\\
-\underline{Recruitment} & & & & \\ 
-ARTR & -0.5539 & -0.4522 & -0.5030 & -0.1253\\
-HECO &  0.0654 & -1.5975 & -0.0586 & -0.3056\\
-POSE & 0.0043 & -0.0058 & -1.7052 & 0.0335\\
-PSSP & 0.0953 & -0.3047 & -0.0960 & -1.7046\\
-\hline
-\end{tabular}
-\end{table}
-
-\clearpage
-
-\section{Supplementary Figures}
-
- \begin{figure}[h]
- \centering
- \includegraphics[width=0.7\textwidth]{plate}
- \caption{Examples of a grass removal quadrat before treatment in 2011 (A) and four years after treatment in 2015 (B), and an \textit{A.
-tripartita} removal quadrat before treatment in 2011 (C) and again in 2015.
-Basal cover of the perennial grasses in (A) was 4.6\% and canopy cover of  \textit{A.
-tripartita} in (B) was 22.9\%.
-}
- \label{fig:photos}
- \end{figure}
-
- \begin{figure}[h]
- \centering
- \includegraphics[width=0.7\textwidth]{PSSP_marginalWARTR}
- \caption{Marginal effects of \textit{A. tripartita} crowding (W) on \textit{P. spicata} growth. Points are residuals from a model that included all 
- covariates from our standard model except for \textit{A.
-tripartita} crowding.
-Grey points show (residual) growth of individual plants in control plots, red points show plants in \textit{A.
-tripartita} removal plots.
-The blue symbol shows the mean of the red points.
-Source file: \texttt{growth\textbackslash PSSPgrowth.r}.
-}
- \label{fig:PSSPresids}
- \end{figure}
-
- \begin{figure}[h]
- \centering
- \includegraphics[width=0.7\textwidth]{CompKernels}
- \caption{Competition kernels for the four dominant species. Source file: \texttt{Wdistrib\textbackslash exploreSurvivalWs.r}. }
- \label{fig:CompKernels}
- \end{figure}
-
-  \begin{figure}[h]
-  \centering
-  \includegraphics[width=1\textwidth]{climate}
-  \caption{Annual precipitation (a) and mean temperature (b) during the period of the experiment, shown against the long-term (1927-2016) means (solid blue and red lines) and 5\% and 95\% quantiles (dashed lines).
-Source file:  \texttt{climate\textunderscore fig.r}.}
-  \label{fig:climate}
-  \end{figure}
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=1\textwidth]{growth_residuals_vs_Wremoval}
-\caption{The relationship between residuals from the fitted growth model and individual-level pre-treatment crowding exerted by the removal species (all grasses in quadrats where we studied \textit{A.
-tripartita}, and \textit{A.
-tripartita} in the quadrats where we studied the grasses).
-Only individuals present in 2011 in the removal treatment quadrats are included in this analysis.
-If those individuals survived into later years, we did include their performance in those later years.
-None of the relationships were statistically significant.
-Source file:  \texttt{growth\textbackslash growth\textunderscore residuals\textunderscore by\textunderscore individ.R}.}
-\label{fig:GrowthResidsIndivid}
-\end{figure}
-
-  \begin{figure}[tbp]
-  \centering
-  \includegraphics[width=1\textwidth]{PSSP_W_scatters}
-  \caption{Bivariate scatter plots comparing crowding exerted by each neighbor species on \textit{P.
-spicata}.
-Each point compares the crowding experienced by one \textit{P.
-spicata} individual in one year from the species on the x-axis to the species on the y-axis.
-Blue symbols show values from controls plots, red symbols show pre-treatment values from \textit{Artemisia} removal plots.``W.allcov'' refers to the aggregated crowding by all shrubs and perennial grasses beyond the focal dominant species, and ``W.allpts'' refers to the aggregated crowding by forb species.
-\texttt{Wdistrib\textbackslash exploreSurvivalWs.r}.
-}
-  \label{fig:Wscatters}
-  \end{figure}
-
-  \begin{figure}[tbp]
- \centering
- \includegraphics[width=1\textwidth]{PSSP_W_byTrt}
- \caption{Distribution of crowding exerted by each neighbor species on \textit{P.
-spicata} individuals in controls plots (blue) and removal quadrats (red) in 2011, before \textit{A.
-tripartita}  removals were conducted.
-Inset bar graphs show the probability that a particular neighbor species was present ($W>0$) in a focal plant's local neighborhood in control (blue) and removal (red) quadrats.
-``W.allcov'' refers to the aggregated crowding by all shrubs and perennial grasses beyond the focal dominant species, and ``W.allpts'' refers to the aggregated crowding by forb species.
-\texttt{Wdistrib\textbackslash exploreSurvivalWs.r}.
-}
- \label{fig:W-by-treatment}
- \end{figure}
-
- \begin{figure}[tbp]
- \centering
- \includegraphics[width=1\textwidth]{cover_projections_1step_maxCI}
- \caption{Observed and predicted cover for each of the four modeled species assuming maximum effects of removal. 
-  Solid black lines show mean observed cover averaged across control (closed symbols) and removal treatment (open symbols) quadrats.
-Blue lines and symbols show one-step-ahead predictions from the baseline IBM (no removal treatment coefficients), averaged across quadrats.
-Red lines and symbols show one-step-ahead predictions from an IBM that includes removal treatment coefficients assigned the outermost value of the 95\% credible intervals.
-  Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r}. }
- \label{fig:IBM1step-maxCI}
- \end{figure}
-
-  \begin{figure}[tbp]
-  \centering
-  \includegraphics[width=1\textwidth]{boxplots-maxCI}
-  \caption{Equilibrium cover of the four dominant species simulated by the IPM with maximum removal treatment effects. Boxplots show interannual variation in cover, 
-  resulting from random year effects in the demographic models. Gray boxes show cover simulated by a model with \textit{A. tripartita} present and no removal treatment 
-  coefficients, blue boxes show results from the same model but with \textit{A. tripartita} cover set to zero (a species removal), red boxes show results from a 
-  model with  \textit{A. tripartita} set to zero and (mean) removal treatment coefficients included, and orange boxes show results with removal treatments 
-  coefficients assigned the outermost value of the 95\% credible intervals.  Source file: \texttt{ipm\textbackslash IPM-figures.r}.}
-  \label{fig:IPMresults-maxCI}
-  \end{figure}
-
 \end{document}

### 2018-04-02 removals/Manuscript/appendix.tex
Adding stand alone appendix

--- 2018-01-26 removals/manuscript/removal_ms.tex
@@
 
-\title{Weak interspecific interactions in a sagebrush steppe\new{? Conflicting evidence from observations and experiments} }
-
-\author[1]{Peter B. Adler\thanks{Corresponding author. Department of Wildland Resources and the Ecology Center, Utah State University, Logan, Utah Email: peter.adler@usu.edu}}
-\author[1]{Andrew Kleinhesselink}
-\author[2]{Giles Hooker}
-\author[3]{J. Bret Taylor}
-\author[1]{Brittany Teller}
-\author[4]{Stephen P. Ellner}
-\affil[1]{Department of Wildland Resources and the Ecology Center, Utah State University, Logan, Utah}
-\affil[2]{Department of Biological Statistics and Computational Biology, Cornell University, Ithaca, New York}
-\affil[3]{USDA, Agricultural Research Service, U. S. Sheep Experiment Station, 19 Office Loop, Dubois, ID, USA}
-\affil[4]{Department of Ecology and Evolutionary Biology, Cornell University, Ithaca, New York}
-
-\renewcommand\Authands{ and }
-
-% \date{Last compile: \today}
-
-\sloppy
-
-\renewcommand{\baselinestretch}{1.25}
-
 \begin{document}
-
-\maketitle
-
-\bigskip \textbf{Running title:} Weak interspecific interactions
-
-\smallskip \textbf{Submission type:} Article
-
-\newpage
-
-\begin{doublespacing}
-
-\linenumbers
-
-\section*{Abstract}
-
-Stable coexistence requires intraspecific limitations to be stronger than interspecific limitations.
-The greater the difference between intra- and interspecific limitations, the more stable the coexistence, and the weaker the competitive release any species should experience following removal of competitors.
-We conducted a removal experiment to test whether a previously estimated model, showing surprisingly weak interspecific competition for four dominant species in a sagebrush steppe, accurately predicts competitive release.
-Our treatments were 1) removal of all perennial grasses and 2) removal of the dominant shrub, \textit{Artemisia tripartita}.
-We regressed survival, growth and recruitment on the locations, sizes, and species identities of neighboring plants, along with an indicator variable for removal treatment.
-If our ``baseline" regression model, which accounts for local plant-plant interactions, accurately explains the observed responses to removals, then the removal coefficient should be non-significant.
-\new{For survival, the removal coefficients were never significantly different from zero, and only \textit{A.
-tripartita} showed a (negative) response to removals at the recruitment stage.
-For growth, the removal treatment effect was significant and positive for two species, \textit{Poa secunda} and \textit{Pseudoroegneria spicata}, indicating that the baseline model underestimated interspecific competition.
-For all three grass species, population models based on the vital rate regressions that included removal effects projected 1.4 to 3-fold increases in equilibrium population size relative to the baseline model (no removal effects).} However, including information about pretreatment cover of \textit{A.
-tripartita} at the quadrat or individual scale failed to improve the growth regressions, raising questions about the mechanisms driving the positive response to removal.
-\new{While the experimental results showed greater competitive release than our baseline models predicted, we will need more tightly controlled experiments focused on underlying mechanisms to conclusively validate or reject our model's predictions.}
-
-%The ability of models based on observational data to predict population responses to experimental removals increased our confidence that interspecific interactions are in fact weak for at least three of our four study species.
-The implication is that factors besides interspecific feedbacks play important roles in determining the structure and dynamics of this community.
-The success of the models also supports the use of multispecies models based on long-term observational data for characterizing competitive interactions.
-
-\textbf{\large{Keywords:}} Coexistence, competition, integral projection model, removal experiment, sagebrush steppe.
-
-\section*{Introduction}
-
-Stable coexistence of competitors requires that intraspecific interactions must limit populations more than interspecific interactions \citep{chesson_mechanisms_2000}.
-When studying populations of species that appear to be coexisting stably, such as populations that have co-occurred in close proximity for extended periods, one should expect intraspecific limitations to be stronger than interspecific limitations.
-However, interspecific interactions might still be strong and negative.
-In fact, plant community ecologists often motivate research on coexistence by noting that all plant species compete for water, light, and a few mineral resources (e.g.
-\citealt{silvertown_plant_2004}).
-Resource competition plays a central role in both Tilman's mechanistic resource competition models \citep{tilman_resource_1982} and Grime's Competitor-Stress tolerator-Ruderal strategies \citep{grime_plant_1979}.
-Neutral theory \citep{hubbell_unified_2001} goes even further, assuming that interspecific competition is exactly as strong as intraspecific competition.
-Although a growing list of studies has emphasized the role of interspecific facilitation in some plant communities \citep{brooker_facilitation_2008,he_global_2013}, it is unclear whether this research has overturned the default assumption that plants compete intensely with other species.
-
-Competitive release provides an intuitive, population-level measure of the strength of interspecific competition.
-Following removal of one or more species from a community, how much will the abundance of the remaining species increase?
-Large increases in the abundance of the remaining species represent a large competitive release, and imply strong interspecific interaction.
-Plant ecologists have conducted dozens of removal experiments and have observed almost every possible result.
-In some cases, competitive release was complete: remaining species fully compensated for removed biomass \citep{leps_nutrient_1999,jutila_effects_2002}.
-In other cases, some competitive release occurred but the degree of compensation depended on the functional group or dominance of the removed and remaining species \citep{sala_resource_1989,belsky_effects_1992,smith_dominant_2003}.
-Finally, even negative effects of removals have been observed \citep{keddy_effects_1989,gilbert_dominant_2009}, consistent with the recent facilitation literature.
-Results from removal experiments do come with caveats.
-Experiments typically do not run long enough to document long-term outcomes, responses may only be measured at one life stage whereas population responses depend on the whole life cycle, and removal treatments require disturbance which may cause a variety of impacts unrelated to resource competition \citep{aarssen_neighbour_1990}.
-
-An alternative approach to studying competitive release is to fit multispecies population models that estimate species interactions using long-term, observational data, and then simulate population responses to removals.
-Applying this approach to arid and semiarid plant communities, we found evidence of strong intraspecific limitation and weak interspecific interactions, leading to very stable coexistence and implying little competitive release \citep{adler_coexistence_2010,chu_large_2015}.
-In fact, at our sagebrush steppe study site  \citep{adler_coexistence_2010}, our model predicts removal of sagebrush would actually reduce the equilibrium cover of the three dominant grass species very slightly, reflecting the facilitative effect of sagebrush on grass recruitment and very weak competitive effects on growth and survival.
-Since we only studied the most common, co-occurring species, we expected to find evidence of stable coexistence, but the predicted negative effect of sagebrush removal on grass cover is surprising given the long history of sagebrush eradication to increase grass forage production in sagebrush rangelands \citep{Robertson1947,Mueggler1958,Cook1963}.
-Compared to traditional field experiments, our approach has the advantage of projecting full life-cycle responses over long time periods, but its reliance on observational data sets limits the strength of our inference.
-A successful experimental test would increase our confidence in the model's projections of competitive release.
-
-The idea that results from observational studies of competition represent hypotheses which should be tested experimentally is not new \citep{Hairston1980}, but such tests remain extremely rare.
-In fact, independent tests of any kind of quantitative prediction are rare in ecology, spurring recent calls for a greater emphasis on prediction \citep{Mouquet2015,Petchey2015,Houlahan2017}.
-These authors argue that testing predictions will not only advance applied goals such as forecasting ecological impacts of global change, but will also accelerate progress in basic research.
-
-In the spirit of this argument, we conducted a removal experiment to test whether our sagebrush steppe community model, which is based on observational data, underestimates the strength of interspecific competition and competitive release.
-We studied the responses of three perennial grass species to removal of sagebrush, and the response of sagebrush to removal of all grasses.
-The primary purpose of the experiment was not to directly measure competitive release, which is difficult to do over a short time period in a semiarid ecosystem with low and variable plant cover, but rather to test whether our model accurately characterizes the interspecific interactions that determine the magnitude of competitive release.
-We addressed two questions: First, are the survival, growth or recruitment of our target species significantly different in control vs.
-removal treatments, after accounting for differences in local species composition?
-Because our demographic models include the effects of neighborhood crowding, they should be able to predict how experimental removal of neighbors alters demographic rates.
-If the full effect of experimental removals is stronger than predicted, it means that our statistical models underestimate the response to decreased crowding.
-Our second question integrates across the full life cycle: how do short and long-term population-level responses to removals compare to projections from multispecies population models based on observational data?
-If models based on observational data adequately describe facilitative and competitive interactions, the population dynamics they project should be similar to projections from models that include information from the experimental removal treatments.
-Obtaining that result would validate the use of long-term data sets and modeling to study interspecific interactions.
-
-\section*{Materials and Methods}
-
-\subsection*{Study site and data set description}
-
-The U.S.
-Sheep Experiment Station (USSES) is located 9.6 km north of Dubois, Idaho (44.2\degree N, 112.1\degree W), 1500 m above sea level.
-During the period of data collection (1926 \textendash 1957), mean annual precipitation was 270 mm and mean temperatures ranged from -8\degree C (January) to 21\degree C (July).
-The vegetation is dominated by the shrub, \textit{Artemisia tripartita}, and the C3  perennial bunchgrasses \textit{Hesperostipa comata}, \textit{Pseudoroegneria spicata},  and \textit{Poa secunda}.
-These four species, the focus of our models, accounted for over 70\% of basal cover (grasses) and 60\% of canopy cover (shrubs and forbs).
-
-Scientists at the USSES established 26 1-m$^2$ quadrats between 1926 and 1932.
-Eighteen quadrats were distributed among four ungrazed exclosures, and eight were distributed in two paddocks grazed at medium intensity spring through fall.
-All quadrats were located on similar topography and soils.
-In most years until 1957, all individual plants in each quadrat were mapped using a pantograph (Blaisdell 1958).
-The historical data set is public and available online \citep{zachmann_mapped_2010}.
-In 2007, we located 14 of the original quadrats, all of which are inside permanent livestock exclosures, and resumed annual mapped censusing using the traditional pantograph method.
-
-We extracted data on survival, growth, and recruitment from the mapped quadrats based on plants' spatial locations.
-Our approach tracks genets, which may be composed of multiple polygons, as they fragment and/or coalesce.
-Each mapped polygon is classified as a surviving genet or a new recruit based on its spatial location relative to genets present in previous years \citep{lauenroth_demography_2008}.
-We modeled vital rates using data from 22 year-to-year transitions between 1926 and 1957, and nine year-to-year transitions from 2007 to 2016.
-Only four quadrats were observed for the first two transitions in the 1920's, while at least 14 quadrats were observed for all subsequent transitions.
-
-All data and R code necessary to reproduce our analysis will be deposited in the Dryad Digital Repository once the manuscript is accepted.
-The current version of the computer code and all necessary data are available at \texttt{https://github.com/pbadler/ExperimentTests/tree/master/removals}.
-
-\subsection*{Removal experiment}
-
-In spring 2011, in a large exclosure containing six of the historical permanent quadrats, we established an additional 16 quadrats for the removal treatments.
-New quadrats were selected to be similar to historical plots, but also needed to have enough of the focal species present so that removal experiments could be conducted.
-We evaluated potential locations for the new quadrats near the original quadrats and then used only those locations with at least 5\% canopy cover of \textit{Artemisia tripartita} and at least 3\% combined basal cover of the three focal perennial grasses.
-We rejected locations falling on hill slopes, areas with greater than 20\% bare rock, or with over 10\% cover of the woody shrubs \textit{Purshia tridentata} or \textit{Amelanchier utahensis}.
-We mapped the new quadrats in June, 2011 and then implemented the removal treatments in September, 2011.
-We randomly assigned 8 of the sixteen new quadrats to the sagebrush removal treatment, and the remaining 8 to the grass removal treatment.
-We cut sagebrush stems at ground level and applied a 5\% solution of Roundup Pro (Glyphosate, Monsanto) to the cut stems.
-In the grass removal plots, we painted the same herbicide solution on all perennial grasses.
-We removed plants 50 cm beyond the quadrat boundaries to minimize edge effects.
-The sagebrush removal was virtually 100\% effective; grass removals required additional herbicide applications in April and May of each year, with fewer grasses remaining each year.
-Appendix Fig.
-\ref{fig:photos} shows photographs of example quadrats before and after treatment.
-We repeated mapped censuses of all plots each June from 2012-2016.
-
-Although the primary focus of our analysis is comparison of observed removal responses with responses predicted by our population models, we also conducted a more traditional statistical analysis of differences in cover trends on control vs.
-removal plots.
-The response variable for this analysis was year-to-year change in cover at the quadrat scale, calculated as $log(Cover_{t+1}/Cover_t)$.
-We tested for the effect of removal treatment on this measure of cover change with a linear mixed effects model implemented in the \texttt{lme4} package \citep{Bates2015} of R, including year and quadrat as random effects.
-We considered the removal treatment effect to be significant when the 95\% credible interval on the estimated coefficient did not overlap zero.
-Because variance in residuals of plots with low cover was high for two species, we explored fitting by weighted least squares, to account for the apparent heteroscedasiticity.
-The results of these tests were qualitatively similar, so we opted to only report the simpler analysis.
-
-\subsection*{Overview of statistical models for vital rates}
-
-Our regression models for survival, growth and recruitment follow previous work \citep{adler_coexistence_2010,chu_large_2015}.
-The current analysis departs from previous work with the inclusion of a categorical removal treatment effect.
-Although the three vital rate models differ in functional form (see Appendix A for complete model descriptions), all of them have a linear predictor, $\mu$, containing the following terms:
-\begin{equation}
-\mu_{ij} = \gamma_j+ \varphi_{jg}+  \chi_j  + \sum \limits_{m} \omega_{jm} {W}_{ijm}
-\label{eqn:survReg}
-\end{equation}
-where $i$ refers to an individual genet (for survival and growth) or quadrat (for recruitment) of species $j$, $\gamma$ is a time-dependent intercept (year effect), $\varphi$ is the coefficient for the effect of spatial location (`$g$' represents a group of nearby quadrats located within one pasture or grazing exclosure), $\chi$ is the removal treatment effect,
-$W_{ijm,t}$ is the crowding on species $j$ exerted by neighbors of species $m$ at the individual (survival and growth) or
-quadrat (recruitment) scale, and $\omega_{jm}$ is an interaction coefficient which determines the impact of crowding by species $m$ on the 
-focal species $j$. For plants in removal plots in 2012, we set crowding for removed neighbor species to zero, rather than using the pre-treatment (2011) data.
-
-If the conspecific and heterospecific crowding effects, the $\omega_{jm}$'s, explain demographic rates in removal treatment plots as well as they do in control plots, then there will be no need for the removal treatment effect, and the estimated coefficient will be small and not significantly different from zero.
-However, if demographic rates in removal plots differ from control plots even after accounting for crowding effects, then the removal treatment effect will be significant.
-For example, we might find that, holding crowding constant, grasses grow more quickly in the sagebrush removal plots than in the control plots.
-This result would indicate that predictions from a ``baseline model'', a model including all estimated parameters except for the removal treatment effect, underestimates competitive release.
-We refer to the model that includes the estimated removal treatment effect as the ``removal model".
-
-Note that the ``baseline" and ``removal" models are not estimated separately; we estimate all parameters together, but when we generate predictions from the baseline model, 
-we simply set the removal treatment effects to zero. We chose this over an alternative approach of fitting a model only to data from control plots and then making 
-and evaluating predictions for the removal plots. The principal advantage of our approach is that the significance test (confidence intervals) for the removal effect 
-is well-defined; fitting on one data set and testing on another would require developing non-standard null distributions to account for variation in both data sets.
-In addition, the alternative approach would return
-essentially the same results because adding data from the removal plots had negligible effects on other model parameters. Correlations between the parameters of 
-a model fit only to control plot data and our full model (leaving aside the removal effects) were stronger than 0.999. 
-Why don't we find strong interactions between removal treatments and the estimated competition coefficients? The answer is that because our treatments 
-involved complete, not just partial, removal of target species, their effects can be captured with an intercept term, as illustrated in Fig. \ref{fig:PSSPresids}. 
-As a result, if, for example, one of the grasses responds more positively to removal of \textit{A. tripartita} than our baseline model predicts, we cannot 
-determine whether this reflects a poorly estimated competition coefficient, or some other effect of the removal treatment unrelated to competitor densities.
-
-\subsection*{Individual-based models}
-The vital rate regressions provide a qualitative test of removal treatment effects: Does removal have an effect on a vital rate that goes above and beyond the effects of species interactions that are explicit in the model?
-However, the population-level effects of removal integrate the effects of the removal treatment on all three vital rates.
-We used an individual-based model (IBM) to predict changes in population size (cover) from one year to the next in removal and control quadrats, and compared these predictions with observed changes.
-
-To simulate changes in cover in each quadrat from year $t$ to year $t+1$, we initialized the IBM with the actual 
-configuration of the four focal species observed in year $t$ (genet sizes and locations). 
-We then projected the model forward one year, applying the random year effects corresponding to year $t$ and the spatial random effects corresponding to the quadrat of interest. 
-We used the survival regression to predict the survival probability of each genet, and the growth regression to predict the expected size (on arithmetic scale) of each genet at time $t+1$ (note that the expected size on arithmetic scale is not exp(expected log size), because of Jensen's inequality).
-Expected quadrat-level cover at $t+1$ is expected area of new recruits, plus the product of survival probability and expected new size summed over all genets present at time $t$. 
-Because we do not project forward beyond one time step, we do not need to assign locations to the new recruits. For control plots, we used the baseline version of the model with all 
-removal treatment effects set to zero. For removal treatment plots, we generated predictions from both the baseline and removal versions of the model, 
-with and without treatment effects set to zero, respectively.
-
-Although all the parameters in our model are estimated with uncertainty, we chose not to propage parameter uncertainty through these 
-simulations because we wanted to maximize our ability to detect the population-level consequences of a small removal effect. 
-We previously found that interspecific competition is weak relative to the effects of interannual environmental variability 
-\citep{Tredennick2017}. Including parameter uncertainty along with random year effects would conflate parameter uncertainty 
-with environmental variability, overstating the magnitude of natural fluctuations and making it more difficult to show that competitive release
-is larger than the observationally-based model predicts. Therefore, we used estimated means of all parameters, including those with 
-confidence intervals that overlapped zero. Based on the same rationale, we used a deterministic version of the model (e.g., 
-recruitment is the $\lambda$ of Eqn. \ref{eqn:recrProcessModel}, rather than a Poisson random variable with mean $\lambda$). 
-Specifically, we used the recruitment regression to calculate the expected number of new recruits at time $t+1$, which we then multiplied 
-by the observed mean size of recruits for the focal species to get expected cover of new recruits.
-
-Another reason we might observe small effects of the removal treatment is the limited power of our experiment.
-Perhaps large uncertainty in the estimates of the removal effects themselves makes it difficult to reject the null hypothesis represented by our baseline model.
-To address this possibility, we also simulated a version of the model in which we replaced the point estimate of each removal treatment effect with either the 5\% or 95\% credible limit for the treatment effect, whichever was farthest from zero.
-This version of the model shows the population-level changes we would expect given maximum removal effects that are consistent with our data.
-
-\subsection*{Integral projection models}
-
-To quantify the long-term population-level consequences of the removal effects, we built and simulated an integral projection model (IPM). The population 
-of species $j$ is represented by a density function $n_j(u,t)$ which gives the density of genets of size $u$ at time $t$ (``size'' is natural log of genet area, so 
-that $n_j(u,t)h$ is the number of genets whose area is between between $\exp(u)$ and $\exp(u+h)$ for $h \ll 1$). The size distribution function at time $t+1$ is given by
-\begin{equation}
-n_j(v,t+1)=\int\limits_{L_j}^{U_j} k_j (v,u,{\vec{n}})n_j(u,t)du   
-\label{eqn:IPM}
-\end{equation}
-where the kernel $k_j (v_j,u_j,\vec{n}_j)$ describes all possible transitions from size $u$ to $v$, and $\vec{n}$ 
-is the set of size-distribution functions for all species in the community. The integral is evaluated over a range of sizes [$L_j,U_j$] 
-that extends past the range of observed sizes.
-
-As in our previous studies of this system \citep{adler_coexistence_2010,adler_forecasting_2012}, the kernel is constructed from the fitted survival ($S$), growth ($G$), and recruitment ($R$) models, 
-\begin{equation}
-k_j(v,u,\vec{n})=S_j(u,\vec{n})G_j(v,u,\vec{n})+R_j(v,u,\vec{n}) 
-\label{eqn:IPM} 
-\end{equation}
-(see Appendix \ref{AppendixA} for a complete description). The IPM uses a mean-field approximation for
-neighborhood competition that captures the essential features of spatially-explicit local interactions; 
-in Appendix \ref{sec:kernelMethods} we explain how the spline competition kernels were used 
-to compute neighborhood competition as a function of individual size for the IPM.
-
-We simulated the IPM to estimate equilibrium abundances under three different scenarios.
-In all of these simulations, we set the spatial random effects to zero, meaning that we are simulating dynamics on a hypothetical, average site.
-To incorporate temporal environmental variation, at each time step we randomly selected one of the 30 sets of random year effects for the survival, growth and recruitment models (``kernel selection'' in the terminology of \citealt{metcalf_statistical_2015}).
-For each scenario, we initialized all species with very low cover and ran the model for 500 time steps (years) to allow the community to reach its equilibrium (i.e., its steady-state pattern of fluctuations in response to environmental variability).
-We calculated the average steady-state cover for each species by simulating an additional 2000 time steps and averaging each species' cover over this period.
-The first of the three scenarios featured all four species and our baseline model for control plots (no removal effects).
-For the second scenario, we simulated our experimental removals by setting either the cover of \textit{A.
-tripartita} to zero or the cover of all three grass species to zero, and then simulating the dynamics of the remaining species as under the first scenario.
-This simulation projects competitive release following removal according to the baseline model (no removal effects).
-For the third scenario, we again set \textit{A.
-triparitia} or grass cover to zero, but this time we projected the model using the vital rate regressions that include removal effects.
-If our baseline model successfully predicts competitive release observed in the removal experiment, then the results of the second and third scenarios should be similar.
-To the extent that the baseline model underestimates (or overestimates) the actual competitive release, the results of the second and third scenarios will differ.
-As with the IBM, we repeated these simulations using maximum removal treatment effects represented by parameters values at the 5\% or 95\% credible limit farthest from zero.
-
-\section*{Results}
-
-During the period of our experiment, annual water-year (October-September) precipitation fell within the 5\% - 95\% quantiles of the 
-long-term mean, although more years fell below the mean than above it (Fig. \ref{fig:climate}a in Appendix C). However, annual mean temperatures 
-were consistently warm relative to the long-term mean, and two years were warmer than the 95\% quantile (Fig. \ref{fig:climate}b).
-
-\subsection*{Cover trends}
-
-Trends in cover by species and removal treatment were variable. For example, grass removal had little effect on the cover of \textit{A. tripartita} 
-(Fig. \ref{fig:CoverTrends}), but shrub removal did appear to cause an increase in the cover of \textit{P. spicata}. 
-However, when we tested the effects of removal treatments on year-to-year population growth rates ($\log(Cover_{t+1}/Cover_t)$) at the quadrat level,
- none of the removal treatment effects were significantly different from zero (Table \ref{table:coefficients}). For \textit{P. spicata}, 
-the 95\% credible interval only just contained zero (95\% CI: -0.01, 0.38), providing marginal evidence for a positive effect of removal on population growth. 
-Fitting with weighted least squares to address the high variance in residuals for plots with very low cover indicated even less evidence for a positive 
-response of  \textit{P. spicata} to removal. However, these treatment comparisons do not directly address our questions about whether 
-responses to removal were adequately predicted by models fit to observational data.
-
-\begin{figure}[tbp]
-\centering
-\includegraphics[width=1\textwidth]{treatment_trends_cover}
-\caption{Cover trends by treatment for the four dominant species.  Note the difference in y-axis scale between the canopy cover of \textit{A. tripartita} and 
-the basal cover of the dominant grasses. ``Shrub removal'' refers to removal of \textit{A. tripartita}, and ``Grass removal'' refers to removal of all perennial grasses. 
-The cover of species in treatments from which they are removed (e.g. \textit{A. tripartita} cover in shrub removal plots) is included to show values at the start 
-of the experiment in 2011 and the efficacy of the removal treatments. Source file: \texttt{treatment\textunderscore trends\textunderscore removals.r}.}
-\label{fig:CoverTrends}
-\end{figure}
-
-\subsection*{Vital rates}
-
-Our vital rate models take into account the effects of local species composition and should therefore be able 
-to explain the effect of removals on survival, growth and recruitment without any additional parameters. 
-As in our previous studies using only the historical data, the neighborhood competition coefficients in the 
-fitted models imply that interspecific competition is much weaker than intraspecific competition, in many
-cases by 1-2 orders of magnitude (Table \ref{table:alphas}). The intraspecific interaction effects are large and have credible intervals that 
-do not overlap zero, while the interspecific coefficients are small and the majority have credible intervals that do overlap zero.
-If the removal treatment effects that we added to these models are statistically significant, it would indicate an effect of 
-removals either larger or smaller than the effect predicted by a model based on data from the control plots. 
-For the survival models, the 95\% credible intervals on the removal effects overlapped zero for all four species (Fig. \ref{fig:VitalRateTest}; 
-Tables \ref{ARTRsurvival} to \ref{PSSPsurvival}), meaning that the survival probabilities of plants in removal plots can be predicted based on data 
-from control plots. For the growth models, the 95\% credible intervals on the removal effects were positive for \textit{P. secunda}  and \textit{P. spicata},
- meaning that individuals of these species grew more rapidly in removal plots than controls even after accounting for neighborhood interactions 
- (Fig.
-\ref{fig:VitalRateTest}; Tables \ref{ARTRgrowth} to \ref{PSSPgrowth}).
-\new{ At the recruitment stage, only \textit{A.
-tripartita} showed a significant response to removals, and it was negative (Fig.
-\ref{fig:VitalRateTest}; Table \ref{table:recruitment}).}
-
- \begin{figure}[tbp]
- \centering
- \includegraphics[width=1\textwidth]{treatment_tests}
- \caption{Effects of removal treatments on the survival, growth, and recruitment of the four dominant species.
-Circles show the estimated coefficient of the removal effect, bars show 95\% credible intervals.
-The removal effect shows response of \textit{A.
-tripartita} (``ARTR") to perennial grass removal, and response of the perennial grasses to \textit{A.
-tripartita} removal, after accounting for local neighborhood interactions.
-The other species codes are HECO: \textit{Hesperostipa comata}, POSE: \textit{Poa secunda}, and PSSP: \textit{Pseudoroegneria spicata}.
- Source file: \texttt{treatment\textunderscore test\textunderscore figure.r}.}
- \label{fig:VitalRateTest}
- \end{figure}
-
-The positive removal effects on \textit{P.
-secunda}  and \textit{P.
-spicata} growth suggest that models based on observational data may underestimate competitive release.
-We conducted additional analyses to further explore removal effects on the growth of these two species.
-First, \new{we asked whether increases in grass growth following removal were greatest in quadrats or neighborhoods where we removed the most \textit{A.
-tripartita}.
-We did this by taking residuals from the fitted growth models for plants in removal quadrats and plotting those residuals against the pre-treatment, quadrat-level cover of \textit{A.
-tripartita}.
-No species showed the positive relationship we expected to see (Fig.
-\ref{fig:GrowthResidsQuad}).
-In fact, \textit{Poa secunda} residuals were significantly lower in quadrats with high pre-treatment cover of \textit{A.
-tripartita} (individuals in high cover quadrats are growing more slowly than the growth model predicted).
-We repeated this analysis at the individual plant level for grasses present in 2011.
-We calculated pre-treatment crowding from \textit{A.
-tripartita} for each of those individuals, then plotted their growth residuals against crowding for all years in which each plant survived.
-We found no significant relationships between the growth residuals and pre-treatment crowding from \textit{A.
-tripartita} (Fig.
-\ref{fig:GrowthResidsIndivid}).
-}
-
-%we classified all individuals of these species as located either inside or outside the limits of \textit{A.
-tripartita} canopies present in 2011 (before removal).
-If removal is releasing grasses from \textit{A.
-tripartita} competition for light or soil resources, we would expect individuals located under shrub canopies to experience a greater effect of removal than individuals further away.
-However, the location of individuals with respect to \textit{A.
-tripartita} canopies had no effect on growth of either species (Tables \ref{table:POSEgrowth-inARTR} and \ref{table:PSSPgrowth-inARTR}), meaning that individuals located outside \textit{A.
-tripartita} canopies responded to removals as much as individuals underneath (removed) canopies.
-
-Second, we evaluated the possibility that removal effects might be transient.
-We removed aboveground plant material, but roots of the removed plants remain in the ground and decompose, potentially creating a flush of mineral nutrients.
-We tested for a transient removal effect by adding a year-by-treatment interaction.
-For both  \textit{P.
-secunda}  and \textit{P.
-spicata}, we found no sign of a decrease in the removal effect over time (Tables \ref{table:POSEgrowth-trtYears} and \ref{table:PSSPgrowth-trtYears}).
-In fact, for  \textit{P.
-secunda}, the removal effects increase over time.
-
-Third, responses to removals might be poorly predicted by models based on observational data if the removals reduce plant densities (crowding) below the range of historical variation, or if the removals result in combinations of neighborhood-scale densities that deviate from naturally occurring patterns of correlation.
-However, the distribution of crowding observed within control plots easily spans the crowding values observed in removal plots (Fig.
-\ref{fig:Wscatters}), and we failed to find evidence for strong interspecific covariances in crowding (Appendix \ref{AppendixA}).
-
-Finally, we evaluated pre-existing differences in community composition in removal and control plots.
-In selecting removal plots, our criteria required candidate plots to include at least a certain amount of our target removal species.
-Because we adopted the historical plots as our controls, the same criteria were not applied (a control plot might not contain any \textit{A.
-tripartita}).
-To quantify resulting differences in pre-treatment (2011) composition, we compared neighborhood crowding at the individual plant level (the $W_{jm}$'s of the growth regressions) across treatments.
-\textit{A.
-tripartita} crowding and crowding by all subdominant shrub and grass species appeared greater in removal plots than in the controls, whereas crowding by \textit{H.
-comata} was lower in removals than controls, though it was absent from most \textit{P.
-spicata} neighborhoods in both treatments (Fig.
-\ref{fig:W-by-treatment}).
-These pre-existing differences in composition raise the possibility that plots selected for the removal treatments might differ from the control plots in some unobserved covariate that could influence plant composition and performance.
-\new{On the other hand, some of the control plots that currently contain zero \textit{A.
-tripartita} had high \textit{A.
-tripartita} in the 1900's, and vice versa.
-If abiotic variation, such as soil depth, were the primary control of variation in \textit{A.
-tripartita} cover, such large changes in cover over time would be unlikely.
-}
-
-\subsection*{Population-level responses}
-
-To quantify the short-term population-level consequences of removals, we used an IBM to project the cover of each species in each quadrat one year ahead using versions of the model with and without removal treatment effects.
-Plotting observed vs.
-predicted cover for each quadrat in each year (Fig.
-\ref{fig:obsVpred}) shows points clustered around the 1:1 line, indicating reasonable accuracy and little bias for both control and removal quadrats.
-For \textit{H.
-comata}, differences in predicted cover of removal plots between the baseline and removal models were very small.
-For the other three species, the removal model predicted higher cover in removal plots than the baseline model predicted.
-
- \begin{figure}[tbp]
- \centering
- \includegraphics[width=1\textwidth]{obsVpred_quad_year}
- \caption{The relationship between observed and predicted cover for each quadrat, year, and species.
-The 1:1 line is dashed, the solid black line shows the trend for control plots with predictions from the baseline IBM model (black symbols), the blue line shows the trend for removal plots with predictions from the baseline model (blue symbols), and the red line shows the trend for removal plots with predictions from the removal model (red symbols).
-Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r}.
-}
- \label{fig:obsVpred}
- \end{figure}
-
-Time series of predicted and observed cover, averaged over quadrats, show similar patterns.
-Predicted population size closely tracked observed population size (Fig.
-\ref{fig:IBM1step}) in most years for all species, but the difference between predictions from the baseline and removal models varied among species.
-For \textit{H.
-comata}, differences in predictions from the two models were virtually indistinguishable.
-For \textit{A.
-tripartita}, \textit{P.
-secunda} and especially \textit{P.
-spicata}, the model with removal effects consistently predicted higher population growth than the baseline model (Fig.
-\ref{fig:IBM1step}).
-Using the version of the model with maximum removal treatment effects, the patterns were qualitatively similar, with a somewhat greater contrast between predictions of the baseline and removal models for \textit{P.
-secunda} and \textit{P.
-spicata} (Fig.
-\ref{fig:IBM1step-maxCI}).
-
- \begin{figure}[tbp]
- \centering
- \includegraphics[width=1\textwidth]{cover_projections_1step}
- \caption{Observed and predicted cover for each of the four modeled species.
-Solid black lines show mean observed cover averaged across control (closed symbols) and removal treatment (open symbols) quadrats.
-Blue lines and symbols show one-step-ahead predictions from the baseline IBM (no removal treatment coefficients), averaged across quadrats.
-Red lines and symbols show one-step-ahead predictions from an IBM that includes removal treatment coefficients.
-For \textit{H.
-comata}, lines showing predictions for the baseline and removal effect models are virtually indistinguishable.
-Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r}.
-}
- \label{fig:IBM1step}
- \end{figure}
-
-To quantify the long-term population-level consequences of permanent competitor removals, we used an IPM to project the long-term average cover of each species.
-The IPM predicts little effect of \textit{A.
-tripartita} removal on the cover of any grass species when we simulate the baseline model with no removal treatment coeffiicients (compare gray and blue boxes in Fig.
-\ref{fig:IPMresults}).
-In fact, projected average cover of all three grass species declines slightly following \textit{A.
-tripartita} removal.
-Using a model with removal treatment coefficients, the IPM projects that \textit{A.
-tripartita} removal will increase equilibrium cover, relative to the baseline scenario, by a factor of 1.4, 1.5 and 3.0 for \textit{H.
-comata}, \textit{Poa secunda} and \textit{P.
-spicata}, respectively (Fig.
-\ref{fig:IPMresults}).
-Running a version of the model with maximum removal treatment effects (removal coefficients set to the 95\% credible limit furthest from zero), resulted in a much larger projected increase in \textbf{H.
-comata}, reflecting its small sample size and corresponding high uncertainty, but little change in the projections for \textit{Poa secunda} and \textit{P.
-spicata} (Fig.
-\ref{fig:IPMresults-maxCI}).
-
- \begin{figure}[tbp]
- \centering
- \includegraphics[width=1\textwidth]{boxplots}
- \caption{Equilibrium cover of the four dominant species simulated by the IPM.
-Boxplots show interannual variation reflecting random year effects.
-Gray boxes show cover simulated by a model with \textit{A.
-tripartita} present and no removal treatment coefficients, blue boxes show results from the same model but with \textit{A.
-tripartita} cover set to zero (a species removal), and red boxes show results from a model with \textit{A.
-tripartita} set to zero and removal treatment coefficients included.
-Source file: \texttt{ipm\textbackslash IPM-figures.r}.}
- \label{fig:IPMresults}
- \end{figure}
-
-The equilibrium cover of \textit{A.
-tripartita} simulated by the IPM (5.4\%) is consistent with observed cover during the historical data collection period (6.6\% from 1931-1956), but is low compared to contemporary observed cover (15-25\%, Fig.
-\ref{fig:CoverTrends}).
-Could the small competitive release of the grasses
-following simulated \textit{A.
-tripartita} removal simply reflect the low simulated \textit{A.
-tripartita} cover?
-To explore this possibility, we increased \textit{A.
-tripartita} cover in the baseline simulation by increasing its average fecundity.
-The increased fecundity led to a doubling of \textit{A.
-tripartita} average cover, but had essentially zero effect on the cover of the grasses (changes of less than 0.1\% cover), implying that even if shrub cover were higher its removal would still have little effect on the grasses.
-
-The baseline model predicts more dramatic competitive release of \textit{A. tripartita} following grass removal: mean cover increases from 4\% to over11\%. 
-Adding removal treatment effects dramatically alters the outcome, with \textit{A.
-tripartita} decreasing towards extinction.
-However, these outcomes both are driven by parameters that are estimated with very high uncertainty (Table~\ref{table:recruitment}).
-The baseline model predicts an increase following grass removal because of the relatively strong negative effects of \textit{Poa secunda} and \textit{H.
-comata} on \textit{A.
-tripartita} recruitment (\ref{table:recruitment}).
-The estimated effect of
-\emph{Poa} is -0.50, but the standard deviation of the estimate is 0.28 and the 95\% credible interval stops just short of zero; the 95\% credible interval for the
-\textit{H. comata} effect includes values less than half of the point-estimate. The grass removal treatment effect, which leads to predicted extinction, has a credible 
-interval that broadly overlaps zero (Table~\ref{table:recruitment}). \textit{A. tripartita} recruitment parameters have exceptionally high uncertainty, presumably 
-because \textit{A. tripartita} recruitment is episodic (see \emph{Discussion}), and this produces extreme uncertainty about its long-term 
-response to grass removal.
-
-\section*{Discussion}
-
-Our population model based on historical, observational data revealed weak interspecific interactions \citep{adler_coexistence_2010,chu_large_2015}.
-As a result, the model predicts very little competitive release among the four dominant species in a sagebrush steppe.
-Because the model was based solely on observational
-data, we were suspicious of its ability to accurately describe the outcome of interspecific interactions, so we conducted a removal experiment to test it. 
-Our approach represents a strong test of a surprising prediction for two reasons. First, the prediction was already published; the experiment 
-gave us a very real opportunity to be wrong in public. Second, testing an observationally-based model with an experimental manipulation, 
-not just additional observations, further increases the potential to confront the model with a novel situation outside the historical range of variation.
-\new{We expected that upon completion of this study, we would have clear evidence for the success or failure of our previous model's predictions.
-Unfortunately, our results do not provide that clarity.
-In the sections below, we first summarize the evidence that our experiment revealed greater competitive release than our baseline model predicted, we then explain weaknesses in the experimental results that complicate the interpretation, and we end by discussing general lessons for the study of interspecific interactions.}
-
-\subsection*{Evidence for greater competitive release than predicted}
-
-\new{We evaluated response to removals in three ways: vital rates, short-term population growth rates, and changes in long-term equilibrium cover.
-Two species, \textit{Poa secunda} and \textit{P.
-pseudoroegneria}, showed significant, positive responses to \textit{A.
-tripartita} removal at the level of vital rates.
-Individuals of these species grew 21\% and 24\% faster, respectively, in removal plots than controls, after accounting for variation in local neighborhood composition.
-In other words, our baseline model appeared to underestimate the competitive release of these species following \textit{A.
-tripartita} removal.}
-
-\new{The elevated growth of \textit{Poa secunda} and \textit{P.
-pseudoroegneria} individuals translated into greater short-term population growth rates.
-For these species, the individual-based model with removal effects predicted higher cover than the baseline version of the model.
-The differences in predicted cover of roughly 20\% (\ref{fig:IBM1step}) are comparable to the differences in the individual growth rates, reflecting the fact that at these short time scales changes in individual size influence cover more than survival or recruitment.
-For \textit{A.
-tripartita}, the magnitude of the difference between predictions from the baseline and removal effects models was similar, further emphasizing the short-term influence of growth compared to survival and recruitment, which responded negatively to removals in this species.
-}
-
-\new{Long-term projections using the integral projection model showed even larger responses to removal.
-For the three grasses, projected equilibrium cover in the absence of \textit{A.
-tripartita} is 1.4 to 3-fold higher for the removal effects model compared to the baseline model.
-While the absolute changes in basal cover are small, these relative changes are substantial, and seem to present compelling evidence that competitive release is stronger than predicted by our baseline model.
-}
-
-\subsection*{Weaknesses in the experimental evidence}
-
-\new{Although the growth regressions for \textit{Poa secunda} and \textit{P.
-pseudoroegneria} and the long-run projections for all three grass species suggest stronger competitive release than our baseline model predicted, a variety of other evidence makes it harder to confidently reject those baseline model predictions.
-Only two of the twelve vital rate tests indicated greater competitive release than predicted.
-Granted, small samples sizes contributed to greater uncertainty for \textit{A.
-tripartita} and \textit{H.
-comata}, but even for \textit{Poa secunda} and \textit{P.
-pseudoroegneria} the increases in growth were small relative to other sources of variation.
-For example, using quadrat-level cover data, we could not detect treatment differences (Table \ref{table:coefficients}).
-}
-
-\new{The strongest evidence for large competitive release of grasses following \textit{A.
-tripartita} removal comes from the long-term projections of the IPM.
-These projections effectively make removal treatment effects permanent; the removal effects on vital rates observed during the experiment are locked in for the entire 2500 time step simulation.
-Furthermore, these projections ignore parameter error because we wanted to highlight the population consequences of the removal effects.
-Given these strong assumptions, the projections should be viewed with some caution.
-}
-
-\new{For \textit{A. tripartita}, the long-term IPM projections actually show extinction following grass removal.
-However, this result primarily reflects the influence of recruitment parameters that are highly uncertain (see \emph{Results}).
-This high uncertainty results from \textit{A.
-tripartita} recruitment being low and very episodic (roughly half of the total recruits in the 22 years of historical data were observed in 2 of those years, and only 28 \textit{A.
-tripartita} recruits have been observed since we resumed annual censuses in 2007).
-Because of this uncertainty, we put more weight on the direct observations and short-term projections showing little effect of grass removal on \textit{A.
-tripartita} populations.}
-
-\new{Another reason we view the growth results with some suspicion is that increases in growth were no larger in individual neighborhoods or quadrats with higher pre-treatment \textit{A.
-tripartita} cover.
-If \textit{A.
-tripartita} removal is releasing grasses from competition, shouldn't this effect be larger for grasses located in areas with high \textit{A.
-tripartita} cover?
-The absence of this relationship (Figs.
-\ref{fig:GrowthResidsQuad} and \ref{fig:GrowthResidsIndivid}) raises questions about the mechanism by which \textit{A.
-tripartita} removal promotes grass growth.
-Perhaps \textit{A.
-tripartita's} roots extend far enough beyond its canopy to compete for resources with grasses located in interspaces.
-While this might explain the lack of the relationship at the neighbrhood scale, it seems unlikely to explain the quadrat-level results.
-}
-An alternative explanation involves pre-treatment differences in composition between control plots and  the \textit{A.
-tripartita} removal plots, which had higher cover of \textit{A.
-tripartita} (see 2011 means in Fig.
-\ref{fig:CoverTrends}) and higher cover of subdominant grasses and shrubs (Fig.
-\ref{fig:W-by-treatment}).
-These pre-treatment differences could reflect subtle differences in edaphic conditions which could promote \textit{P.
-spicata} growth.
-\new{However, since the 1900's, many quadrats have experienced dramatic changes in \textit{A.
-tripartita} cover, arguing against viewing \textit{A.
-tripartita} cover as a sensitive indicator of edaphic conditions.
-A final possibility is that our removal treatment influenced grass growth through some other mechanism, unrelated to competition.
-Maybe the loss of shrub cover benefits grasses by reducing small mammal herbivory \citep{Orrock2010}?
-Without a better understanding of the mechanism driving increased grass growth in the shrub removal quadrats, we are hesitant to dramatically adjust our prior understanding gained from models fit to decades of observational data.
-}
-
-%How can we reconcile the insensitivity of grasses to sagebrush removal in our study with evidence from rangeland ecology 
-%research showing that sagebrush eradication increases grass production? The most likely explanation involves ecological context: our 
-%study focused on \textit{Artemisia tripartita} in an ungrazed plant community, whereas most research on sagebrush 
-%and forage production has focused on \textit{Artemisia tridentata} in heavily grazed communities, and often has involved grass seeding 
-%following sagebrush removal (e.g. \citealt{Cook1963}). \textit{A. tripartita} is smaller in stature than \textit{A. tridentata} 
-%and may interact differently with perennial grasses. More importantly, cover of sagebrush can increase dramatically on overgrazed sites, 
-%especially under a spring grazing regime. Competitive release of grasses following sagebrush removal often increases in 
-%a nonlinear fashion, with little increase in production at sites with low initial sagebrush cover, 
-%and accelerating responses as initial sagebrush cover passes some threshold \citep{McDaniel2005}, implying a 
-%shift in the nature of interspecific interactions, perhaps due to changes in the balance of 
-%competition and facilitation or in resource supply. The cover of 
-%sagebrush in our ungrazed quadrats may have fallen below that threshold.
-
-\subsection*{Lessons for studying interspecific interactions}
-
-\new{A variety of approaches are available for studying interspecific interactions among plants, and each has its advantages and disadvantages.
-Observational datasets have the advantage of being realistic: they reflect all the processes and interactions that drive community dynamics across a range of spatial and temporal scales.
-Their disadvantage is weak inference.
-Distinguishing between correlation and causation is challenging, and simulations of a model fit to observational data may rely heavily on extrapolation.
-Experiments, such as response surface designs (Inouye), offer an alternative approach to studying interactions at the population scale.
-Their chief advantage is strong inference about causality.
-But they sacrifice realism: experiments are typically implemented at fine spatial and temporal scales which exclude potentially important processes.
-Finally, we can study interactions mechanistically by describing patterns of resource use \citep{Dybzinski2007,Silvertown1999}.
-Mechanistic studies have the virtue of focusing on directly on the processes of interest, but rarely address more than one limiting resource at a time.
-}
-
-\new{ Our previous work was based purely on observational data, so we were particularly concerned about the extrapolation issue.
-Specifically, our model projects the performance of individual plants growing in the absence of neighbors.
-Although the observational data set seems to include such situations (Fig.
-\ref{fig:Wscatters}), if these isolated individual are still interacting with neighbors belowground, our model might underestimate the true intrinsic growth rates.
-Therefore, we conducted the removal experiment, hoping to gain the advantages offered by multiple, complementary approaches.
-The positive responses of \textit{Poa secunda} and \textit{P.
-pseudoroegneria} to removals suggests our observational approach might have underestimated intrinsic growth rates after all.
-But removal experiments fall into a gray area between the observational and experimental approach.
-They are an experimental manipulation, but they rely on natural spatial variation in composition.
-We can't remove a plant from a location where it does not occur.
-The reliance on natural variation weakens inference, leaving nagging questions about pre-existing differences among quadrats.
-We could have avoided those concerns by conducting a response surface experiment, planting our study species in randomly selected locations.
-}
-
-\new{We also learned that partial removals, rather than complete removal treatments, could 
-have helped us to distinguish between competition-mediated and density-independent effects of removal treatments (e.g. Fig. \ref{fig:PSSPresids}).
-Partial removals, or experimental gradients in densities, would give us inference about the competition coefficients themselves, whereas our 
-design lumped all treatment effects into the intercept terms.
-Complete removals might be adequate for testing the presence of competition, but do not work as well for quantifying per capita competitive effects.
-Furthermore, while our experiment focused on a few interspecific interactions, ultimately it is the difference in the strength of intra- and interspecific interactions which determines the stability of coexistence and the degree to which the dynamics of co-occurring species are coupled.
-Next time we will also include intraspecific removal treatments.
-}
-
-\new{The difficulty we experienced interpreting our results also made us appreciate the value of mechanistic insight.
-Even if the removal experiment had perfectly validated the predictions of our phenomenological baseline model, we suspect we would still be asking ourselves, how can interspecific interactions really be so weak?
-The only way to rigorously answer that question is to study the factors--resources, natural enemies, environmental constraints--limiting population growth.
-}
-
-Although our results point to limitations of model based on long-term observational data, they also give some cause for optimism about population forecasting.
-\citet{Crone2013} found that matrix projection models for plants often failed to forecast population trends due
-to environmental differences between the data collection and forecast periods. Our population models 
-performed quite well, after accounting for year-to-year environmental variation through the inclusion of
-random year effects.
-
-\subsection*{Conclusion}
-
-Experimental removal of \textit{A.
-tripartita} led to greater release of grasses than our previous model predicted.
-However, the mechanisms driving this  release remain unclear, and make us hesitant to conclude that our model based on observational data underestimates the strength of interspecific competition.
-The results do increase our interest in testing model predictions using even more tightly controlled experiments as well as direct study of resource-use.
-
-%\subsection*{Implications of weak interspecific interactions}
-%
-%The weak interspecific interactions among many of the dominant species in our system indicates that these species  
-%are limited by quite different factors. This low niche overlap could reflect strong resource partitioning \citep{tilman_resource_1982}, 
-%perhaps combined with variation in the times and locations at which species are taking up those resources. 
-%Species-specific responses to spatial and temporal variation in non-resource environmental variables can also weaken interspecific 
-%interactions \citep{chesson_mechanisms_2000}, though in a previous study we failed to find evidence of a temporal storage effect 
-%in this community \citep{adler_weak_2009}. Interactions with above- or belowground enemies can regulate populations and, given 
-%enough species-specificity in the effects of these enemies, decouple the dynamics of different species \citep{hersh_evaluating_2011,
-%janzen_herbivores_1970,connell_role_1971}.  More likely, plants may alter the environment, and even resource availability, 
-%in ways that facilitate their neighbors. For example, shade cast by \textit{A. tripartita} canopies may reduce 
-%evapotranspiration (e.g. \citealt{Barbier2008}). \textit{A. tripartita} removal might offset this facilitative effect by reducing 
-%resource uptake, resulting in little net change in water balance following removal. 
-%Unfortunately, we did not monitor soil moisture at the fine spatial and temporal resolution necessary to test this hypothesis.
-%
-%Evidence for weak interspecific interactions and limited competitive release has interesting implications. If most 
-%interspecific interactions are weak, then single-species models may be adequate for forecasting population responses 
-%to environmental change \citep{adler_forecasting_2012,kleinhesselink_indirect_2015,Chu2016}. Similarly, differences among 
-%species in relative abundance should reflect variation in the strength of intraspecific density dependence more than 
-%variation in the strength of interspecific interactions \citep{yenni_strong_2012}. A recent global study showing little 
-%variation in the abundance of invasive grassland species in native and invaded ranges on different 
-%continents  \citep{firn_abundance_2011} is consistent with the idea that intraspecific interactions are the primary determinant 
-%of species abundances. In systems where this is true, effective management and conservation requires understanding the sources 
-%and species-specificity of intraspecific density dependence. A useful first step in this direction would be to test 
-%whether our model is also successful at predicting the effects of release from intraspecific competition.
-%
-%For many species in other communities, interspecific competition may still be an important driver. In closed-canopy 
-%forests where competition for light is intense, we might have found different results than in our system, in which 
-%competition for a number of belowground resources is more important than competition for light. However, a 
-%number of recent studies from forests have also found that heterospecific competitive effects are dramatically 
-%weaker than conspecific effects \citep{comita_asymmetric_2010,Kunstler2016,johnson_conspecific_2012}.  We also 
-%expect stronger competition in successional systems, or communities in the early stages of community assembly 
-%where the process of competitive exclusion is still underway \citep{kokkoris_patterns_1999}. Finally, we might have
-%found stronger evidence of competition if we could have studied rare species in our system, but only if those species 
-%are rare because of niche overlap with stronger competitors. If they are rare because they occupy unique, but narrow, niches, 
-%interspecific interactions would still be weak. Determining why interspecific interactions and competitive release are 
-%strong in some ecological contexts but weak in others remains a fascinating research problem.
-
-\section*{Author contributions}
-PBA and ARK designed the experiment and supervised data collection with the help of JBT; PBA, SPE, GH and BT analyzed data and 
-developed models; PBA wrote the first draft of the manuscript and all authors contributed to editing.
-
-\section*{Acknowledgements}
-Funding was provided by NSF grants DEB-1353078 and DEB-1054040 to PBA and DEB-1353039 to SPE and GH, and 
-by the Utah Agricultural Experiment Station, which approves this contribution as journal paper number 9033. 
-The USDA-ARS Sheep Experiment Station generously provided access to historical data and the field experiment site. 
-We thank Robin Snyder, Tom Miller, and two anonymous reviewers for suggestions that improved the manuscript.
-
-\newpage
-\renewcommand{\refname}{Literature cited}
-\bibliographystyle{Ecology}
-\bibliography{RemovalRefs}
-
-\end{doublespacing}
-
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-% APPENDICES !
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-
-\clearpage 
-\newpage
 
@@
 
-\centerline{\Large \textbf{Appendices}}
-\centerline{Adler et al., ``Weak interspecific interactions''}
-
-\vspace{0.4in}
-
-\renewcommand{\theequation}{A-\arabic{equation}}
-\renewcommand{\thetable}{A-\arabic{table}}
-\renewcommand{\thefigure}{A-\arabic{figure}}
-\renewcommand{\thesection}{\Alph{section}}
+\centerline{\Large \textbf{Appendix S1}}
+
+\vspace{0.2in}
+
+\noindent \textbf{Adler et al., ``Weak interspecific interactions in a sagebrush steppe? Conflicting evidence from observations and experiments,'' \textit{Ecology}. }
+
+\vspace{0.2in}
+
+\renewcommand{\theequation}{S\arabic{equation}}
+\renewcommand{\thetable}{S\arabic{table}}
+\renewcommand{\thefigure}{S\arabic{figure}}
+\renewcommand{\thesection}{\arabic{section}}
 
 \section{Detailed Methods} 
-\label{AppendixA}
+\label{AppendixS1}
 
@@
 
+\renewcommand{\refname}{Literature cited}
+\bibliographystyle{Ecology}
+\bibliography{RemovalRefs}
+
 \clearpage 
 \newpage
-
-\renewcommand{\theequation}{B-\arabic{equation}}
-\renewcommand{\thetable}{B-\arabic{table}}
-\renewcommand{\thefigure}{B-\arabic{figure}}
 
@@
 \centering
-\includegraphics[width=1\textwidth]{growth_residuals_vs_quad_removal_cover}
-\caption{The relationship between residuals from the fitted growth model and quadrat-level pre-treatment cover of the removal species (all grasses removed in quadrats where we studied \textit{A.
-tripartita}, and \textit{A.
-tripartita} removed in the quadrats where we studied the grasses).
-Linear regressions are shown only when statistically significant.
-Source file:  \texttt{growth\textbackslash growth\textunderscore residuals\textunderscore by\textunderscore quad.R}.}
-\label{fig:GrowthResidsQuad}
-\end{figure}
-
-\begin{figure}[h]
-\centering
 \includegraphics[width=1\textwidth]{growth_residuals_vs_Wremoval}

### 2018-04-02 removals/Manuscript/appendix.tex
Added better than Jensen section

@@
  than 8\%. The \texttt{R} code for this analysis is included as ..\texttt{\textbackslash Wdistrib\textbackslash exploreSurvivalWs.r}.
+
+\subsection{Correcting for Jensen's inequality in one-step-ahead projections}
+
+We used an individual-based model, based on the vital rate regressions, to project total cover one-year ahead in each quadrat.
+We modeled expected growth and variance in growth on the log scale, meaning that when we exponentiate to transform cover into arithmetic space the expected variance in cover is not normally distributed, and Jensen's inequality creates a downward bias in predicted cover.
+Therefore, we developed a correction that takes into account the relationship between plant size and expected variance in growth.
+
+ After the growth model is fitted,
+
+ \begin{enumerate}
+ \item  Extract for each plant in the data set (indexed by $k$) 
+ its \texttt{logarea.t0} $z_{0,k}$, the growth residual $z_{1,k} - \hat{z}_{1,k}$ (here $z_1$ is
+ the plant's \texttt{logarea.t1} and $\hat{z}_1$ is the predicted logarea), and the predicted variance
+ in growth $V_k$.
+
+ \item Compute the set of scaled residuals $r_k = (z_{1,k} - \hat{z}_{1,k})/\sqrt{V_k}$ and store them
+ in a vector \texttt{scaledResiduals}.
+
+ \end{enumerate}
+
+ Then in computing the one-step-ahead predictions, for each growth model the line 
+ \begin{verbatim}
+ plants$area.pred <- plants$surv.prob*exp(plants$logarea.pred + V.pred/2)
+ \end{verbatim} 
+ is replaced by the following 
+ \begin{verbatim}
+ resid.Pred <- sqrt(V.pred)*scaledResiduals
+ newJensen <- mean(exp(resid.Pred))
+ plants$area.pred <- plants$surv.prob*exp(plants$logarea.pred)*newJensen
+ \end{verbatim}
 

### 2018-04-05 removals/Manuscript/appendix.tex
writing up btJ

--- 2018-04-02 removals/Manuscript/appendix.tex
@@
 
+\linespread{1.15}
+
 \begin{document}
@@
 \renewcommand{\thesection}{\arabic{section}}
+
+\linespread{1.75}
 
@@
  than 8\%. The \texttt{R} code for this analysis is included as ..\texttt{\textbackslash Wdistrib\textbackslash exploreSurvivalWs.r}.
+
+\subsection{Correcting for Jensen's inequality in one-step-ahead projections} 
+One-year-ahead projections of cover in each quadrat, based on its state in the current year, were made using the regression models for survival and growth. 
+For each genet present in year $t$, we computed its probability of survival and its expected total cover in year $t+1$, incorporating the random 
+year-effects for year $t$ and quadrat group effects. The projected cover in year $t+1$ is the sum over all plants in year $t$ of survival probability times  
+projected mean size in year $t+1$ conditional on survival.
+
+Projections of mean cover next year have to take account of the fact that our demographic models use log-transformed cover as the size measure, 
+and project expected growth and variance in growth on log scale. Random variations in realized growth (on log scale) are modeled as zero-mean Gaussian deviations from the 
+regression line, with size-dependent variance. But when we transform those projections back to cover on arithmetic scale, the variability in growth is no longer Gaussian, and 
+its mean is no longer zero becauses of Jensen's inequality. The expected size of an individual next year (conditional on survival) is always larger than the value 
+obtained by exponentiating its expected log-transformed size next year, because the exponential function is concave up.
+
+We therefore developed a correction that takes into account the relationship between plant size and expected variance in growth on log scale, and the 
+effect of back-transformation to arithmetic scale, as follows.
+
+The first step is a data-based estimate of the distribution of random variability in growth for each species. After the growth model for a species was 
+fitted, we extracted for each observed size transition in the data set (indexed by $k$) the initial log-transformed size $z_{0,k}$ and the final log-transformed size $z_{1,k}$. 
+We used the growth model to compute the expected log-transformed new size $\hat{z}_{1,k}$ and the predicted variance in growth $V_k$, and from those, we 
+computed the set of \emph{scaled growth residuals} 
+\be
+r_k = (z_{1,k} - \hat{z}_{1,k})/\sqrt{V_k} 
+\ee 
+with $k$ running across all observed size transitions in the data set for that species. We treat the $r_k$ values as a sample from a normalized 
+distribution specifying the shape of random growth variability on log scale, and used them to correct one-step-ahead cover 
+projections, as follows.
+
+For individual $j$ present in year $t$ with size $z_j(t)$ let $\hat{z}_j(t+1)$ be its projected mean log-transformed size, and $V_j(t)$ the predicted variance in growth
+for that individual, based on its size in year $t$. The projected mean cover on arithmetic scale for individual $j$ in year $t+1$, conditional on survival, 
+is then computed as the average of the set of values 
+\be
+\exp\left(\hat{z}_j(t+1) + \sqrt{V_j(t)}r_k \right).
+\ee
+The values being exponentiated are possible outcomes of growth with random variability on log scale, their exp's are possible outcomes on arithmetic scale, and
+the average of those values is the projected mean cover for the individual, conditional on survival.
 

### 2018-04-07 removals/Manuscript/appendix.tex
Finalized appendix

@@
 \subsection{Correcting for Jensen's inequality in one-step-ahead projections} 
-One-year-ahead projections of cover in each quadrat, based on its state in the current year, were made using the regression models for survival and growth. 
+One-year-ahead projections of cover in each quadrat, based on its state in the current year, were made using the regression models for survival, growth and recruitment. 
 For each genet present in year $t$, we computed its probability of survival and its expected total cover in year $t+1$, incorporating the random 
 year-effects for year $t$ and quadrat group effects. The projected cover in year $t+1$ is the sum over all plants in year $t$ of survival probability times  
-projected mean size in year $t+1$ conditional on survival.
+projected mean size in year $t+1$ conditional on survival, plus the number of expected recruits multiplied by mean recruit size.
 

### 2018-04-07 removals/Manuscript/appendix.tex
Really finalzing appendix

@@
 regression line, with size-dependent variance. But when we transform those projections back to cover on arithmetic scale, the variability in growth is no longer Gaussian, and 
-its mean is no longer zero becauses of Jensen's inequality. The expected size of an individual next year (conditional on survival) is always larger than the value 
+its mean is no longer zero because of Jensen's inequality. The expected size of an individual next year (conditional on survival) is always larger than the value 
 obtained by exponentiating its expected log-transformed size next year, because the exponential function is concave up.

### 2018-05-15 precip/R/recruitment/get_recruitment.R
Comment out some data preparation steps with climate
Some of the links to the original datadrivers climate data are broken
just commenting those out so we can get to work on this

@@ -48,7 +48,6 @@ for(i in 1:Nspp){
   }
 }
 D2[is.na(D2)]=0  # replace missing values 
-
 D2$Period <- 'Modern'
 
 # merge in treatment data
@@ -59,6 +58,7 @@ D2 <- merge(D2,tmp, all.x=T)
 # combine old and new data
 D=rbind(D,D2)
 rm(D2)
+
 
 # get rid of removal treatments
 D <- subset(D,Treatment!="No_grass" & Treatment!="No_shrub")
@@ -82,7 +82,6 @@ tmpD=D[,c("quad","year","Group",paste("cov.",sppList,sep=""))]
 tmpD=aggregate(tmpD[,4:NCOL(tmpD)],by=list("year"=tmpD$year,
   "Group"=tmpD$Group),FUN=mean)
 names(tmpD)[3:NCOL(tmpD)]=paste("Gcov.",sppList,sep="")
-
 D=merge(D,tmpD,all.x=T)
 
 # assign indicator variables -------------------------------------------------------------------------------- # 

### 2018-05-17 precip/R/climate/aggregate_VWC_data.R
More small changes to replicate ARK analysis

@@ -4,21 +4,21 @@
 #
 #######################################################################################
 
-rm(list = ls()) 
-
-detachAllPackages <- function() {
-  
-  basic.packages <- c("package:stats","package:graphics","package:grDevices","package:utils","package:datasets","package:methods","package:base")
-  
-  package.list <- search()[ifelse(unlist(gregexpr("package:",search()))==1,TRUE,FALSE)]
-  
-  package.list <- setdiff(package.list,basic.packages)
-  
-  if (length(package.list)>0)  for (package in package.list) detach(package, character.only=TRUE)
-  
-}
-
-detachAllPackages()
+# rm(list = ls()) 
+# 
+# detachAllPackages <- function() {
+#   
+#   basic.packages <- c("package:stats","package:graphics","package:grDevices","package:utils","package:datasets","package:methods","package:base")
+#   
+#   package.list <- search()[ifelse(unlist(gregexpr("package:",search()))==1,TRUE,FALSE)]
+#   
+#   package.list <- setdiff(package.list,basic.packages)
+#   
+#   if (length(package.list)>0)  for (package in package.list) detach(package, character.only=TRUE)
+#   
+# }
+# 
+# detachAllPackages()
 
 library( ggplot2 ) 
 library(tidyr)

### 2018-05-17 precip/R/climate/prepare_climate_covariates.R

@@ -85,7 +85,7 @@ q_VWC <-
          #VWC.a.0 = lag( VWC.a.1,4), 
          #VWC.a.l = lag( VWC.a.0,4)) %>%
   filter( season == 'spring') %>% # plants are measured at the end of spring each year 
-  select( Treatment, Period, year, season, starts_with("VWC")) %>%
+  dplyr::select( Treatment, Period, year, season, starts_with("VWC")) %>%
   ungroup() %>% 
   gather( var, val, starts_with('VWC')) %>% 
   filter( !is.na(val)) %>%
@@ -106,7 +106,7 @@ q_precip <-
          P.su.0 = lag(P.su.1, 4), 
          P.su.l = lag(P.su.0, 4)) %>% 
   filter( season == 'spring') %>% # plants are measured at the end of spring each year 
-  select( Treatment, Period, year, season, starts_with("P"))
+  dplyr::select( Treatment, Period, year, season, starts_with("P"))
 
 q_temp <- 
   seasonal_clim %>% 
@@ -126,7 +126,7 @@ q_temp <-
           T.w.0 = lag(T.w.1, 4), 
           T.w.l = lag(T.w.0, 4)) %>% 
   filter( season == 'spring') %>% 
-  select( Treatment, Period, year, season, starts_with("T."))
+  dplyr::select( Treatment, Period, year, season, starts_with("T."))
 
 allClim <- 
   q_precip %>% 

### 2018-05-17 precip/R/climate/soilMoistureTreatmentEffects.R

@@ -1,6 +1,6 @@
 # calculate treatment effects compared to control soil moisture 
 
-rm(list = ls()) 
+#rm(list = ls()) 
 
 library( ggplot2 ) 
 library(tidyr)

### 2018-05-17 precip/analysis/analysis_pipeline.R

@@ -19,6 +19,7 @@ library(zoo)
 library(texreg)
 library(xtable)
 library(gridExtra)
+library(MASS)
 
 # run data preparation files first --------------------------- # 
 

### 2018-05-17 precip/analysis/figure_scripts/plot_spring_soil_moisture_spot_measures.R

@@ -1,12 +1,12 @@
 ##### Spring 2015 soil moisture
 ##### 
 
-rm(list = ls())
-library(ggplot2 )
-library(dplyr)
-library(tidyr)
-library(lme4)
-library(texreg)
+# rm(list = ls())
+# library(ggplot2 )
+# library(dplyr)
+# library(tidyr)
+# library(lme4)
+# library(texreg)
 
 load('analysis/figure_scripts/my_plotting_theme.Rdata')
 

### 2018-05-17 precip/analysis/treatment_trends_precip.R

@@ -1,11 +1,11 @@
-library(lme4)
-library(dplyr)
-library(tidyr)
+# library(lme4)
+# library(dplyr)
+# library(tidyr)
 
 # call from removal_analysis_wrapper.r
-root <- "~"
+root <- my_path
 sppList=c("Artemisia tripartita","Hesperostipa comata","Poa secunda","Pseudoroegneria spicata")
-dataDir <- file.path(root,"/driversdata/data/idaho_modern/")
+dataDir <- file.path(root,"driversdata/data/idaho_modern//")
 
 load('analysis/figure_scripts/my_plotting_theme.Rdata')
 

### 2018-05-22 precip/analysis/fit_stan_treatment_models.R
Got Andy's code running up to stan fits

@@ -1,4 +1,5 @@
-rm(list = ls())
+#rm(list = ls())
+
 library(rstan)
 
 df <- expand.grid(species = c('ARTR', 'HECO', 'POSE', 'PSSP'), vital_rate = c('growth', 'recruitment', 'survival'))

### 2018-05-26 precip/manuscript/overall_prediction_score.tex
still trying to get everything to run

@@
-% latex table generated in R 3.3.1 by xtable 1.8-2 package
-% Fri Dec 23 14:08:37 2016
+% latex table generated in R 3.4.1 by xtable 1.8-2 package
+% Sat May 26 11:15:59 2018
 \begin{table}[ht]
@@
   \hline
-species & vital\_rate & score & climate model & baseline model & diff & improved \\ 
+species & vital\_rate & score & climate model & no climate model & diff & improved \\ 
   \hline
-ARTR & growth & lppd & -186.08 & -189.17 & 3.09 & *** \\ 
-  ARTR & growth & MSE & 0.55 & 0.55 & -0.00 & *** \\ 
-  ARTR & recruitment & lppd & -79.05 & -77.81 & -1.24 &  \\ 
-  ARTR & recruitment & MSE & 169.86 & 9.47 & 160.39 &  \\ 
-  ARTR & survival & lppd & -37.55 & -36.42 & -1.13 &  \\ 
-  ARTR & survival & MSE & 0.06 & 0.06 & 0.00 &  \\ 
-  HECO & growth & lppd & -475.54 & -454.36 & -21.18 &  \\ 
-  HECO & growth & MSE & 1.26 & 1.18 & 0.09 &  \\ 
-  HECO & recruitment & lppd & -149.43 & -151.60 & 2.17 & *** \\ 
-  HECO & recruitment & MSE & 288.16 & 235.57 & 52.59 &  \\ 
-  HECO & survival & lppd & -158.99 & -147.06 & -11.93 &  \\ 
-  HECO & survival & MSE & 0.12 & 0.11 & 0.01 &  \\ 
-  POSE & growth & lppd & -1823.71 & -1831.56 & 7.85 & *** \\ 
-  POSE & growth & MSE & 1.72 & 1.73 & -0.01 & *** \\ 
-  POSE & recruitment & lppd & -260.05 & -257.46 & -2.59 &  \\ 
-  POSE & recruitment & MSE & 45.40 & 37.32 & 8.08 &  \\ 
-  POSE & survival & lppd & -698.06 & -718.60 & 20.54 & *** \\ 
+ARTR & growth & lppd & -231.22 & -233.95 & 2.73 & *** \\ 
+  ARTR & growth & MSE & 0.82 & 0.79 & 0.04 &  \\ 
+  ARTR & recruitment & lppd & -92.08 & -90.12 & -1.97 &  \\ 
+  ARTR & recruitment & MSE & 56.48 & 7.15 & 49.33 &  \\ 
+  ARTR & survival & lppd & -44.65 & -42.49 & -2.16 &  \\ 
+  ARTR & survival & MSE & 0.07 & 0.07 & 0.00 &  \\ 
+  HECO & growth & lppd & -480.47 & -464.58 & -15.89 &  \\ 
+  HECO & growth & MSE & 1.12 & 1.06 & 0.06 &  \\ 
+  HECO & recruitment & lppd & -159.48 & -159.11 & -0.37 &  \\ 
+  HECO & recruitment & MSE & 238.43 & 152.52 & 85.91 &  \\ 
+  HECO & survival & lppd & -194.94 & -179.77 & -15.17 &  \\ 
+  HECO & survival & MSE & 0.13 & 0.12 & 0.01 &  \\ 
+  POSE & growth & lppd & -1951.59 & -1967.73 & 16.14 & *** \\ 
+  POSE & growth & MSE & 1.76 & 1.78 & -0.02 & *** \\ 
+  POSE & recruitment & lppd & -266.88 & -266.56 & -0.32 &  \\ 
+  POSE & recruitment & MSE & 45.94 & 37.76 & 8.17 &  \\ 
+  POSE & survival & lppd & -784.24 & -805.86 & 21.63 & *** \\ 
   POSE & survival & MSE & 0.14 & 0.14 & -0.00 & *** \\ 
-  PSSP & growth & lppd & -1232.93 & -1237.92 & 4.99 & *** \\ 
-  PSSP & growth & MSE & 1.51 & 1.51 & 0.00 &  \\ 
-  PSSP & recruitment & lppd & -271.34 & -273.94 & 2.60 & *** \\ 
-  PSSP & recruitment & MSE & 79.09 & 42.68 & 36.41 &  \\ 
-  PSSP & survival & lppd & -332.47 & -307.26 & -25.21 &  \\ 
-  PSSP & survival & MSE & 0.11 & 0.10 & 0.01 &  \\ 
+  PSSP & growth & lppd & -1308.13 & -1318.27 & 10.15 & *** \\ 
+  PSSP & growth & MSE & 1.51 & 1.52 & -0.01 & *** \\ 
+  PSSP & recruitment & lppd & -293.20 & -296.26 & 3.06 & *** \\ 
+  PSSP & recruitment & MSE & 76.96 & 40.47 & 36.49 &  \\ 
+  PSSP & survival & lppd & -407.91 & -372.77 & -35.14 &  \\ 
+  PSSP & survival & MSE & 0.12 & 0.11 & 0.01 &  \\ 
    \hline

### 2018-05-26 precip/analysis/figure_scripts/plot_obs_vs_pred_treatment_parameters.R

@@ -152,7 +152,7 @@ for( i in 1:length(gs_fits)){
   
   muhat <- matrix( ncol = nrow(pred_matrix), nrow = nrow( bt ) )
   
-  for( j in 1:nrow(bt)) { 
+  for( j in 1:nrow(bt)) { i
     muhat[j , ] <- pred_matrix%*%bt[j, ]
   }
   

### 2018-05-28 precip/manuscript/overall_prediction_score.tex
saving latest run

@@
 % latex table generated in R 3.4.1 by xtable 1.8-2 package
-% Sat May 26 11:15:59 2018
+% Mon May 28 03:54:23 2018
 \begin{table}[ht]
@@
   \hline
-ARTR & growth & lppd & -231.22 & -233.95 & 2.73 & *** \\ 
-  ARTR & growth & MSE & 0.82 & 0.79 & 0.04 &  \\ 
+ARTR & growth & lppd & -233.79 & -233.95 & 0.16 & *** \\ 
+  ARTR & growth & MSE & 0.89 & 0.79 & 0.10 &  \\ 
   ARTR & recruitment & lppd & -92.08 & -90.12 & -1.97 &  \\ 
@@
   HECO & recruitment & MSE & 238.43 & 152.52 & 85.91 &  \\ 
-  HECO & survival & lppd & -194.94 & -179.77 & -15.17 &  \\ 
+  HECO & survival & lppd & -189.88 & -179.77 & -10.11 &  \\ 
   HECO & survival & MSE & 0.13 & 0.12 & 0.01 &  \\ 
-  POSE & growth & lppd & -1951.59 & -1967.73 & 16.14 & *** \\ 
-  POSE & growth & MSE & 1.76 & 1.78 & -0.02 & *** \\ 
+  POSE & growth & lppd & -1954.94 & -1967.73 & 12.79 & *** \\ 
+  POSE & growth & MSE & 1.77 & 1.78 & -0.02 & *** \\ 
   POSE & recruitment & lppd & -266.88 & -266.56 & -0.32 &  \\ 
@@
   POSE & survival & MSE & 0.14 & 0.14 & -0.00 & *** \\ 
-  PSSP & growth & lppd & -1308.13 & -1318.27 & 10.15 & *** \\ 
-  PSSP & growth & MSE & 1.51 & 1.52 & -0.01 & *** \\ 
+  PSSP & growth & lppd & -1325.05 & -1318.27 & -6.78 &  \\ 
+  PSSP & growth & MSE & 1.55 & 1.52 & 0.04 &  \\ 
   PSSP & recruitment & lppd & -293.20 & -296.26 & 3.06 & *** \\ 

### 2018-05-29 removals/Manuscript/appendix.tex
Corrected Table 12 in appendix

@@
 \usepackage{caption}
+\usepackage{longtable}
 
@@
 
-\noindent \textbf{Adler et al., ``Weak interspecific interactions in a sagebrush steppe? Conflicting evidence from observations and experiments,'' \textit{Ecology}. }
+\noindent \textbf{Adler et al., ``Weak interspecific interactions in a sagebrush steppe? Conflicting evidence from observations and experiments,'' \textit{Ecology}. 2018. }
 
@@
 % Mon Sep 19 08:37:48 2016
-\begin{table}[ht]
+\clearpage
 \centering
+\begin{longtable}{ r r r r r r r }
 \caption{Summary of posterior distributions for fixed effects for the recruitment model (symbols correspond to Eqns.
@@
 The indexing on $\omega$ shown here and used in the computer code for the recruitment model gives $\omega$[3,1] as the effect of species 3 on species 1.
-This is the reverse of the normal convention used in \ref{eqn:recrProcessModel} and shown in \ref{table:alphas}.}
-\label{table:recruitment}
-\begin{tabular}{rrrrrrr}
+This is the reverse of the normal convention used in \ref{eqn:recrProcessModel} and shown in \ref{table:alphas}.} \label{table:recruitment}  \\
+%   \multicolumn{7}{c}%
   \hline
@@
   \hline
+  \endfirsthead
+  \multicolumn{7}{c}%
+  {\tablename\ \thetable\ -- \textit{Continued from previous page}} \\
+  \hline
+   & mean & sd & X2.5. & X97.5. & Rhat & n.eff \\
+  \hline
+  \endhead
+  \hline \multicolumn{7}{r}{\textit{Continued on next page}} \\
+  \endfoot
+  \hline
+  \endlastfoot
  $\gamma$[1] & 0.3341 & 0.7071 & -1.1262 & 1.6391 & 1.0018 &  1100 \\ 
@@
   $p$[4] & 0.7502 & 0.1480 & 0.3920 & 0.9453 & 1.0150 &  2000 \\ 
-   \hline
-
-   \hline
     $\gamma$[1] & 0.2843 & 0.6342 & -1.0900 & 1.4950 & 1.0279 &   810 \\ 
@@
       \hline
-\end{tabular}
-\end{table}
+\end{longtable}
 
@@
 The one exception to this pattern is the effect of POSE on recruitment by ARTR, but the nonconforming
-coefficient ($\omega$[3,1] = -0.50 in Table \ref{table:recruitment}) has low precision (sd = 0.28, Table \ref{table:recruitment}.} 
+coefficient ($\omega$[3,1] = -0.50 in Table \ref{table:recruitment}) has low precision (sd = 0.28, Table \ref{table:recruitment}).} 
 \label{table:alphas}

### 2018-06-01 precip/manuscript/overall_prediction_score.tex
Finally got all scripts to run

@@
-% latex table generated in R 3.4.1 by xtable 1.8-2 package
-% Mon May 28 03:54:23 2018
+% latex table generated in R 3.5.0 by xtable 1.8-2 package
+% Fri Jun 01 10:20:08 2018
 \begin{table}[ht]
@@
   POSE & recruitment & MSE & 45.94 & 37.76 & 8.17 &  \\ 
-  POSE & survival & lppd & -784.24 & -805.86 & 21.63 & *** \\ 
+  POSE & survival & lppd & -778.89 & -805.86 & 26.98 & *** \\ 
   POSE & survival & MSE & 0.14 & 0.14 & -0.00 & *** \\ 
-  PSSP & growth & lppd & -1325.05 & -1318.27 & -6.78 &  \\ 
-  PSSP & growth & MSE & 1.55 & 1.52 & 0.04 &  \\ 
+  PSSP & growth & lppd & -1308.13 & -1318.27 & 10.15 & *** \\ 
+  PSSP & growth & MSE & 1.51 & 1.52 & -0.01 & *** \\ 
   PSSP & recruitment & lppd & -293.20 & -296.26 & 3.06 & *** \\ 

### 2018-06-01 precip/manuscript/pgr_predictions.tex
Finally got all scripts to run

@@
-% latex table generated in R 3.3.1 by xtable 1.8-2 package
-% Tue Jan 17 13:20:20 2017
+% latex table generated in R 3.5.0 by xtable 1.8-2 package
+% Fri Jun 01 10:57:20 2018
 \begin{table}[H]
@@
   \hline
- & species & stat & baseline model & climate model & diff & improved \\ 
+ & species & stat & no climate model & climate model & diff & improved \\ 
   \hline
-1 & ARTR & cor & 0.48 & 0.19 & -0.29 &  \\ 
-  2 & ARTR & MSE & 0.30 & 0.30 & 0.00 &  \\ 
-  3 & HECO & cor & 0.29 & 0.22 & -0.07 &  \\ 
-  4 & HECO & MSE & 0.49 & 0.57 & 0.07 &  \\ 
-  5 & POSE & cor & 0.45 & 0.53 & 0.07 & *** \\ 
-  6 & POSE & MSE & 0.42 & 0.41 & -0.01 & *** \\ 
-  7 & PSSP & cor & 0.36 & 0.38 & 0.03 & *** \\ 
-  8 & PSSP & MSE & 0.39 & 0.39 & -0.01 & *** \\ 
+1 & ARTR & cor & 0.43 & 0.12 & -0.31 &  \\ 
+  2 & ARTR & MSE & 0.27 & 0.39 & 0.12 &  \\ 
+  3 & HECO & cor & 0.29 & 0.30 & 0.01 & *** \\ 
+  4 & HECO & MSE & 0.50 & 0.51 & 0.01 &  \\ 
+  5 & POSE & cor & 0.47 & 0.58 & 0.10 & *** \\ 
+  6 & POSE & MSE & 0.41 & 0.40 & -0.02 & *** \\ 
+  7 & PSSP & cor & 0.38 & 0.35 & -0.03 &  \\ 
+  8 & PSSP & MSE & 0.41 & 0.40 & -0.01 & *** \\ 
    \hline

### 2018-06-01 precip/manuscript/pgr_predictions_by_treatment.tex
adding a bunch of old edits to scripts and manuscript files
Finally got all scripts to run

@@
-% latex table generated in R 3.3.1 by xtable 1.8-2 package
-% Thu Dec 22 04:07:33 2016
+% latex table generated in R 3.5.0 by xtable 1.8-2 package
+% Fri Jun 01 10:57:23 2018
 \begin{table}[H]
@@
   \hline
-1 & ARTR & Control & cor & 0.36 & -0.02 & -0.39 &  \\ 
-  2 & ARTR & Control & MSE & 0.25 & 0.26 & 0.00 &  \\ 
-  3 & ARTR & Drought & cor & 0.60 & 0.43 & -0.17 &  \\ 
-  4 & ARTR & Drought & MSE & 0.35 & 0.38 & 0.03 &  \\ 
-  5 & ARTR & Irrigation & cor & 0.61 & 0.09 & -0.52 &  \\ 
-  6 & ARTR & Irrigation & MSE & 0.32 & 0.29 & -0.04 & *** \\ 
-  7 & HECO & Control & cor & 0.46 & 0.27 & -0.19 &  \\ 
-  8 & HECO & Control & MSE & 0.29 & 0.32 & 0.04 &  \\ 
-  9 & HECO & Drought & cor & -0.24 & 0.72 & 0.95 & *** \\ 
-  10 & HECO & Drought & MSE & 2.07 & 1.50 & -0.58 & *** \\ 
-  11 & HECO & Irrigation & cor & 0.53 & -0.18 & -0.71 &  \\ 
-  12 & HECO & Irrigation & MSE & 0.52 & 0.86 & 0.34 &  \\ 
-  13 & POSE & Control & cor & 0.43 & 0.48 & 0.05 & *** \\ 
-  14 & POSE & Control & MSE & 0.38 & 0.40 & 0.01 &  \\ 
-  15 & POSE & Drought & cor & 0.56 & 0.72 & 0.16 & *** \\ 
-  16 & POSE & Drought & MSE & 0.45 & 0.46 & 0.00 &  \\ 
-  17 & POSE & Irrigation & cor & 0.55 & 0.41 & -0.14 &  \\ 
-  18 & POSE & Irrigation & MSE & 0.47 & 0.42 & -0.05 & *** \\ 
-  19 & PSSP & Control & cor & 0.28 & 0.38 & 0.10 & *** \\ 
-  20 & PSSP & Control & MSE & 0.34 & 0.33 & -0.01 & *** \\ 
-  21 & PSSP & Drought & cor & 0.48 & 0.45 & -0.03 &  \\ 
-  22 & PSSP & Drought & MSE & 0.51 & 0.51 & 0.00 &  \\ 
-  23 & PSSP & Irrigation & cor & 0.43 & 0.30 & -0.12 &  \\ 
-  24 & PSSP & Irrigation & MSE & 0.38 & 0.37 & -0.01 & *** \\ 
+1 & ARTR & Control & cor & 0.31 & -0.00 & -0.31 &  \\ 
+  2 & ARTR & Control & MSE & 0.23 & 0.31 & 0.08 &  \\ 
+  3 & ARTR & Drought & cor & 0.52 & 0.19 & -0.34 &  \\ 
+  4 & ARTR & Drought & MSE & 0.33 & 0.34 & 0.01 &  \\ 
+  5 & ARTR & Irrigation & cor & 0.55 & -0.06 & -0.61 &  \\ 
+  6 & ARTR & Irrigation & MSE & 0.29 & 0.61 & 0.32 &  \\ 
+  7 & HECO & Control & cor & 0.52 & 0.30 & -0.22 &  \\ 
+  8 & HECO & Control & MSE & 0.28 & 0.31 & 0.03 &  \\ 
+  9 & HECO & Drought & cor & 0.08 & 0.75 & 0.67 & *** \\ 
+  10 & HECO & Drought & MSE & 2.13 & 1.40 & -0.73 & *** \\ 
+  11 & HECO & Irrigation & cor & 0.53 & 0.04 & -0.49 &  \\ 
+  12 & HECO & Irrigation & MSE & 0.52 & 0.69 & 0.18 &  \\ 
+  13 & POSE & Control & cor & 0.50 & 0.55 & 0.05 & *** \\ 
+  14 & POSE & Control & MSE & 0.38 & 0.39 & 0.01 &  \\ 
+  15 & POSE & Drought & cor & 0.54 & 0.77 & 0.23 & *** \\ 
+  16 & POSE & Drought & MSE & 0.44 & 0.41 & -0.03 & *** \\ 
+  17 & POSE & Irrigation & cor & 0.51 & 0.35 & -0.16 &  \\ 
+  18 & POSE & Irrigation & MSE & 0.48 & 0.40 & -0.08 & *** \\ 
+  19 & PSSP & Control & cor & 0.32 & 0.36 & 0.03 & *** \\ 
+  20 & PSSP & Control & MSE & 0.35 & 0.34 & -0.01 & *** \\ 
+  21 & PSSP & Drought & cor & 0.52 & 0.42 & -0.11 &  \\ 
+  22 & PSSP & Drought & MSE & 0.53 & 0.53 & -0.01 & *** \\ 
+  23 & PSSP & Irrigation & cor & 0.45 & 0.27 & -0.18 &  \\ 
+  24 & PSSP & Irrigation & MSE & 0.39 & 0.38 & -0.01 & *** \\ 
    \hline

### 2018-06-01 precip/manuscript/pgr_predictions.tex
Final run (?) of year effects branch

@@
 % latex table generated in R 3.5.0 by xtable 1.8-2 package
-% Fri Jun 01 10:57:20 2018
+% Fri Jun 01 15:59:31 2018
 \begin{table}[H]
@@
   2 & ARTR & MSE & 0.27 & 0.39 & 0.12 &  \\ 
-  3 & HECO & cor & 0.29 & 0.30 & 0.01 & *** \\ 
+  3 & HECO & cor & 0.29 & 0.30 & 0.02 & *** \\ 
   4 & HECO & MSE & 0.50 & 0.51 & 0.01 &  \\ 
   5 & POSE & cor & 0.47 & 0.58 & 0.10 & *** \\ 
-  6 & POSE & MSE & 0.41 & 0.40 & -0.02 & *** \\ 
+  6 & POSE & MSE & 0.41 & 0.40 & -0.01 & *** \\ 
   7 & PSSP & cor & 0.38 & 0.35 & -0.03 &  \\ 

### 2018-06-01 precip/manuscript/pgr_predictions_by_treatment.tex
Final run (?) of year effects branch

@@
 % latex table generated in R 3.5.0 by xtable 1.8-2 package
-% Fri Jun 01 10:57:23 2018
+% Fri Jun 01 15:59:34 2018
 \begin{table}[H]
@@
   \hline
-1 & ARTR & Control & cor & 0.31 & -0.00 & -0.31 &  \\ 
+1 & ARTR & Control & cor & 0.31 & 0.00 & -0.31 &  \\ 
   2 & ARTR & Control & MSE & 0.23 & 0.31 & 0.08 &  \\ 
-  3 & ARTR & Drought & cor & 0.52 & 0.19 & -0.34 &  \\ 
+  3 & ARTR & Drought & cor & 0.52 & 0.19 & -0.33 &  \\ 
   4 & ARTR & Drought & MSE & 0.33 & 0.34 & 0.01 &  \\ 
-  5 & ARTR & Irrigation & cor & 0.55 & -0.06 & -0.61 &  \\ 
+  5 & ARTR & Irrigation & cor & 0.54 & -0.06 & -0.60 &  \\ 
   6 & ARTR & Irrigation & MSE & 0.29 & 0.61 & 0.32 &  \\ 
@@
   8 & HECO & Control & MSE & 0.28 & 0.31 & 0.03 &  \\ 
-  9 & HECO & Drought & cor & 0.08 & 0.75 & 0.67 & *** \\ 
-  10 & HECO & Drought & MSE & 2.13 & 1.40 & -0.73 & *** \\ 
-  11 & HECO & Irrigation & cor & 0.53 & 0.04 & -0.49 &  \\ 
+  9 & HECO & Drought & cor & 0.07 & 0.75 & 0.68 & *** \\ 
+  10 & HECO & Drought & MSE & 2.14 & 1.39 & -0.74 & *** \\ 
+  11 & HECO & Irrigation & cor & 0.53 & 0.05 & -0.48 &  \\ 
   12 & HECO & Irrigation & MSE & 0.52 & 0.69 & 0.18 &  \\ 
@@
   14 & POSE & Control & MSE & 0.38 & 0.39 & 0.01 &  \\ 
-  15 & POSE & Drought & cor & 0.54 & 0.77 & 0.23 & *** \\ 
-  16 & POSE & Drought & MSE & 0.44 & 0.41 & -0.03 & *** \\ 
-  17 & POSE & Irrigation & cor & 0.51 & 0.35 & -0.16 &  \\ 
+  15 & POSE & Drought & cor & 0.55 & 0.77 & 0.22 & *** \\ 
+  16 & POSE & Drought & MSE & 0.44 & 0.44 & -0.00 & *** \\ 
+  17 & POSE & Irrigation & cor & 0.52 & 0.35 & -0.17 &  \\ 
   18 & POSE & Irrigation & MSE & 0.48 & 0.40 & -0.08 & *** \\ 
-  19 & PSSP & Control & cor & 0.32 & 0.36 & 0.03 & *** \\ 
+  19 & PSSP & Control & cor & 0.33 & 0.36 & 0.03 & *** \\ 
   20 & PSSP & Control & MSE & 0.35 & 0.34 & -0.01 & *** \\ 
-  21 & PSSP & Drought & cor & 0.52 & 0.42 & -0.11 &  \\ 
+  21 & PSSP & Drought & cor & 0.53 & 0.42 & -0.11 &  \\ 
   22 & PSSP & Drought & MSE & 0.53 & 0.53 & -0.01 & *** \\ 

### 2019-04-09 precip/R/calculate_cover_per_plot.R
Commit everything before reorganization

@@ -11,11 +11,8 @@ species_codes <-   function(x){
   toupper( lapply( temp_split, paste0, collapse = ''))
 }
 
-my_path <- '~/Dropbox' # this should be set within the overall analysis script 
-root <- my_path
-
 sppList=c("ARTR","HECO","POSE","PSSP")
-dataDir <- file.path(root,"driversdata/data")
+dataDir <- file.path(my_path,"driversdata/data")
 
 old <- read_csv(file.path(dataDir, 'idaho/allrecords_cover.csv'))
 new <- read_csv(file.path(dataDir, 'idaho_modern/allrecords_cover.csv'))

### 2019-04-09 precip/analysis/IBM_simulation.R

@@ -3,9 +3,9 @@ library(tidyverse)
 
 # functions ---------------------------------------------------------------------------- 
 
-get_recruit_area <- function(spp) { 
+get_recruit_area <- function(spp, my_path) { 
   
-  dat <- read.csv(paste0('~/Dropbox/driversdata/data/idaho/speciesData/', as.character(spp), '/recSize.csv'))
+  dat <- read.csv(paste0(my_path, '/data/idaho/speciesData/', as.character(spp), '/recSize.csv'))
   
   dat$area
 
@@ -94,7 +94,7 @@ for(i in 1:nrow( model_list)){
   G <- exp( G*Y_scale + Y_center ) # tranform to cm scale
   R <- rstan::extract( rfit, 'IBM_Y_hat')$IBM_Y_hat
   
-  a <- get_recruit_area(spp)
+  a <- get_recruit_area(spp, my_path)
   
   K <- S*G  # survival by size 
   R <- R*median(a)

### 2019-04-09 precip/analysis/analysis_pipeline.R

@@ -6,7 +6,7 @@
 
 rm(list = ls() ) 
 
-setwd("~/Dropbox/projects/ExperimentTests/precip")
+setwd("~/Dropbox/projects/ExperimentTests/precip") # this is set automatically if you open this as an R project
 my_path <- "~/Dropbox" # this is needed to get to the driversdata directory
 
 library(rstan)
@@ -35,7 +35,7 @@ rm(list=setdiff(ls(), "my_path")) # clean up, but leave my_path
 
 # 2. Cover trends
 
-source('analysis/analysis/plot_cover.R')
+source('analysis/plot_cover.R')
 rm(list=setdiff(ls(), "my_path")) # clean up, but leave my_path
 
 # ----- Model Fitting/Selection ---------------------------------- # 
